### 2023-12-18 index.qmd
set up project and moved lit review from website over

---
title: "Uncertainty Literature Review"
author: Harriet Mason
bibliography: references.bib
date: last-modified
format: pdf
editor_options: 
  chunk_output_type: console
---
## Format Notes
##### Usual Format
1. Introduction
2. Notation (if relevant)
3-5. Details of literature review broken into several sub sections. First usually provides context to the problem and the following sections answer questions that context highlights.
6. Future Work/ Conclusion

##### Usual Format for this paper
1. Introduction
2. Definitions of Uncertainty
  - The more taxonomies I read the more convinced I am the "defining uncertainty" is the best one. Not only is it the most complete, but it also identifies WHY some uncertainties are difficult to visualise (specifically which part of the taxonomy) and suggests HOW you can best communicate these uncertainties.
    - A lot of other papers just make a comment like "oh some of these are hard to communicate" but that paper actually goes into detail instead of hand waving the problem away.
    - It also explains why the uncertainty visualisation literature focuses on a specific subset of uncertainty cases.
  - Examples
    - Null Values
    - Variance
    - Hypothesis Testing
    - Distribution
  - Combining uncertainty is also a reccurring theme in these papers
3. Visualisation in the context of uncertainty communication
  - Basically I want a brief section highlighting the importance of visualisation and discussing it in relation to other forms of uncertainty communication.
  - especially considering some uncertainty visualisation papers discuss using just a text based method to communicate uncerainty
4. Specific Section on Uncertainty Visualisation
  - There is also two kinds of papers, those that compare elements of a plot (like mapping uncertainty to an aesthetic) and those that compare some expression of a distribution (like a scatter plot compared to a box plot).
  - try to map back to david speigleholder paper i.e. uncertainty means lots of different things
5. Future work/Conclusion

# Random thoughts
- I think it is also important to highlight if uncertainty is meta information or if it is a variable in of itself. Meta information (such as contiunity, exhangeability, etc) are usually set to specific visual features (e.g. VSUP clearly sees uncertainty as meta information)
- A lot of work that evaluates uncertainty visualisations checks to see if the mean can be gathered from a visualisation, but if we view uncertainty as meta-information, then with a high uncertainty we wouldn't want the mean to be visible
  - Also an example of why the motivation matters for these things. If you don't know "WHY" someone is using a visualisation for uncertainty and just pick the "best" one, you are going to get a visualisation that does a poor job of its task.
  - put uncertainty in the context of other meta-data of variables

```{r}
#| echo: false
#| warning: false
#| message: false

# Load libraries
library(tidyverse)
library(tsibbledata)
library(Vizumap)
library(fable)
```
# Introduction
Think back to the last time you made some sort of data visualisation. What was the purpose of that visualisation? Was it to better understand your data? Was it to help you make a decision? Was it to to communicate that decision to someone else? Now think about the last time you expressed some form of uncertainty. Was it a set of numerical confidence intervals? Maybe they were expressed as a set of values in a table. Did you consider visualising your uncertainty instead? There are many stages in our analysis that benefit from the power of data visualisation, however this does not mean it is always done with success. Visualization is an important step in exploratory data analysis and it is often utilised to **learn** what is important about a data set. The importance of data driven discovery is highlighted by data sets such as Anscombe's quartet [@anscombe] or the Datasaurus Dozen [@datasaurpkg]. Each of the pairwise plots in these data sets have the same summary statistics but strikingly different information when visualised. Anscombe quartet is shown in @fig-anscombe, because describing the data is never the same as seeing it. Instead of having to repeatedly check endless hypothesis to find interesting numerical features, visualisations **tell** us what is important about our data. This powerful aspect of data visualisation is poorly or seldom used in later stages when we are communicating our findings, specifically with respect to uncertainty.

```{r}
#| echo: false
#| warning: false
#| message: false
#| label: fig-anscombe
#| fig-cap: "The four scatter plots that make up Anscombe's quartet. The four scatter plots are visually distinct but have the same mean, standard deviation, and correlation. The visualisation highlights the importance of plotting your data to identify interesting features that are hidden by other summary statistics."
tibble(x = c(anscombe$x1, anscombe$x2,
             anscombe$x3, anscombe$x4),
       y = c(anscombe$y1, anscombe$y2,
             anscombe$y3, anscombe$y4),
       Plot = c(rep("Plot 1",11), rep("Plot 2",11), 
                rep("Plot 3",11), rep("Plot 4",11))) %>%
  ggplot(aes(x,y)) +
  geom_point(aes(fill=Plot), colour="black", 
             size=3, pch=21, alpha=0.75) +
  facet_wrap(~Plot) +
  theme_classic() +
  theme(aspect.ratio = 1,
        legend.position = "none") +
  scale_fill_brewer(type = "qual", palette = 4)
```

##### Secondary Nature of Uncertainty
An overwhelming consensus among visualisation authors seems to be that uncertainty is secondary to estimations. There is a belief held by those that work with data that the uncertainty associated with an estimate (the noise) only exists to hide the estimate itself (the signal). From this point of view uncertainty is optional information that only exists to hinder our message. Therefore despite its alleged importance, when simplifying a plot uncertainty the first thing to go. This belief is also reflected in the development of new uncertainty visualisations. Often when trying to visualise multi-dimensional data, uncertainty is relegated to unimportant aesthetics in the plot, often of lower importance than the estimate where it is easily ignored [@Correll2018; @Lucchesi2017]. Sometimes uncertainty is not relegated to an undesirable aesthetic but instead incorporated using interactivity to allow users to explore the complicated space themselves [@Potter2009]. Even the literature about uncertainty communication expresses an implicit belief that it is of secondary importance to the estimates or context of the data.

####  no visual distinction between sample/pop variance
@Bella2005 found that most participants were ignorant to the fact that error bars are used for both confidence intervals and standard error bars, two wildly different indicators of precision. 
**also the paper sherry sent me**

# Definitions of Uncertainty
##### No clear definition of uncertainty
This issues surrounding uncertainty visualisation is not helped by the fact that the term "uncertainty" lacks a commonly accepted definition in the literature. @Lipshitz1997 even commented that “there are almost as many definitions of uncertainty as there are treatments of the subject” . This mishmash of terminology leads to a large body of work, all claiming to finding the best visualisation or expression of of "uncertainty" but most don't even seem to agree on what uncertainty is. The most encompassing definition of uncertainty I have seen comes from @utypo who define uncertainty as **"any deviation from the unachievable ideal of completely deterministic knowledge of the relevant system"**. This definition encapsulates many concepts adjacent to randomness such as predictions, probability mass distributions (PMF), estimate error, and any data that is not a set of deterministic outcomes. This is the definition of uncertainty I will use for the rest of this report.

##### Uncertainty taxonomy
More commonly uncertainty is defined using a taxonomy rather than a strict definition. There are a few taxonomies for uncertainty, but, just like the definition, most of them are a subset of the one laid out by @utypo. To complete the definition I will include that taxonomy here. @fig-taxonomy is an illustration of the taxonomy presented by @utypo. In this taxonomy, there are three things we need to consider for each "uncertainty" we encounter through the modelling process. First, we need to consider the source of the uncertainty. Is this uncertainty coming from inaccurate measurements or a poorly defined model? This is the *location* of the uncertainty. Second, consider how well you can quantify this uncertainty. Do you know exactly how much measurement error there is in each observation or are you not even aware if there is a measurement error? This is the *level* of your uncertainty, and it ranges from discrete to total ignorance. Finally, consider how this uncertainty came into existence. Is it a result of a naturally random process (epistemic) or is it due to imperfect information and could be improved (aleatory). This is the *nature* of your uncertainty. @utypo then goes on to describe mapping our uncertainty in a 3D space that is defined by its location, level, and nature, but I think the taxonomy is more easily understood as a series of questions we need to consider when we are trying to quantify uncertainty.

![Depicts an illustration of the taxonomy described in @utypo. From right to left the drawing shows the location, level and nature of uncertainty with examples of that category underneath. A specific source of uncertainty from the location can be mapped to a level of ignorance that can increase or decrease (i.e. moving up or down the green line) depending on the nature of the uncertainty. Identifying the location, level and nature of your uncertainty allows you to better understand it. ](taxonomyvis.jpeg){#fig-taxonomy}

##### You should not ignore "uncertainty" that seems unrelated to your problem
While information about the sources of our uncertainty and the type of uncertainty may seem like an unimportant secondary step in uncertainty visualisation, communicating these features of uncertainty helps decision makers make more informed choices. @Padilla2021 found that low forecaster confidence or high model uncertainty both contribute to more conservative judgements by decision makers. Failing to communicate the nature of your uncertainty can result in underestimation or overestimation of failure probabilities [@Kiureghian2009]. Additionally @Gustafson2019 found that the framing of our uncertainty, (i.e. if the source of uncertainty is from a lack of knowledge, approximations, unknown unknowns, or disagreement among parties) was found to not have a detrimental effect on the belief in the estimates, perceived credibility, or behavioural intentions of the decision makers. This means communicating secondary information about your uncertainty can provide additional benefits to decision makers without additional negative repercussions.
# Unorganised

##### The benefits of Unvertainty visualisaiton
Visualisations can provide a more complete picture of a risk than numerical summaries alone. Even something as simple as sketching a distribution before recalling statistics or making predictions can greatly increase the accuracy of those measures [@Hullman2018; @Goldstein2014]. While there is some evidence that confidence intervals provided in text form only are less likely to be misinterpreted than graphics [@Savelli2013], text is insufficient to express more complicated aspects of a distribution, such as mass. Expressing uncertainty verbally decreases the perceived reliability and trustworthiness of the source [@VanderBles2020]. Any confusion caused by expressing uncertainty as a visualisation could also be due to a lack of exposure, since @Kay2016 found people repeatedly exposed to the same uncertainty visualisations quickly get better at making judgements. Additionally, visualisation allow for interactive graphics that provide a more in depth understanding of probability [@Potter2009; @Ancker2009] and infographics that make uncertainty more accessible for people with poor numeracy skills [@Ancker2009]. 
**application in energy**
The use of uncertainty in high dimensional environments is especially important in energy data. Large models that incorporate spatial-temporal data from many sources and systems are used to predict energy uses in the short and long term. Understanding how to improve and make better decisions in these models is imperative in both the daily operation of the energy sector as well as in the transition from fossil fuels to clean energy. The energy sector needs better heuristics to make energy supply analysis less costly to conduct [@Stenclik2021], therefore it is an incredibly relevant application of uncertainty visualisation techniques.

##### People dont visualise uncertainty but it is considered important
Despite these benefits, there is evidence that we don't visualise uncertainty as often as we should. A survey conducted by @Hullman2020a found that majority of visualisation authors agreed that expressing uncertainty is important and should be done more often than it currently is, some even agreed that failing to do so is tantamount to fraud. Despite this, only a quarter of respondents included uncertainty in 50% or more of their visualisations [@Hullman2020a]. Meaning participants were convinced that visualising uncertainty is morally important but were able to provide self sufficient reasoning that allows them to avoid doing it. Some economists suggest that visualisation authors are responding to incentives that make it tempting to avoid visualising uncertainty, even if those incentives are based more in perception than reality [@Manski2020]. The study by @Hullman2020a found that the most common reasons authors don't visualise uncertainty despite knowing it's moral importance are: not wanting to overwhelm the audience; an inability to calculate the uncertainty; a lack of access to the uncertainty information; and not wanting to make their data seem questionable [@Hullman2020a].

##### Effect of uncertainty visualisation on decision making
If decision markers are not presented with the uncertainty about an estimate the data analysts have, for all intents and purposes, made the decision for the decision maker. Upon further interviews @Hullman2020a found that authors believed uncertainty would overwhelm the audience and make their data seem questionable because decision makers are unable to understand uncertainty. This belief, while pervasive, is not true. There is some research that suggests laypeople cannot understand complicated concepts in statistical thinking (such as trick questions on hypothesis tests or the difference between Frequentist and Bayesian thinking) [@Hoekstra2014; @Bella2005] but there is a large amount of research suggesting that presenting uncertainty information improves decision making, both experimentally [@Joslyn2012; @Savelli2013; @Kay2016; @Fernandes2018] and in practice [@Al-Kassab2014]. As a matter of fact, doing what many authors currently do (providing only a deterministic outcome with no uncertainty) causes decision makers to be *less* decisive and have completely unbounded expectations on an outcome [@Savelli2013]. This reality cannot be avoided by providing secondary or non-specific information such as explaining calculations [@Joslyn2012], explaining the advantages of a recommendation [@Joslyn2012], or expressing uncertainty in vague terms [@Erev_1990; @Olson_1997], all of which are undesirable for decision makers and lead to measurably worse decisions [@Joslyn2012; @Erev_1990; @Olson_1997]. One of the most popular depictions of uncertainty for decision making is a quantile dotplot, shown in @fig-quantdot. It is important to avoid whittling down the problem **too** much. Providing a categorical decision alone is somewhat useless [@Joslyn2012], and visualising a single estimate is akin to providing a decision or expressing no uncertainty at all.

```{r}
#| echo: false
#| warning: false
#| message: false
#| label: fig-quantdot
#| fig-cap: "This plot depicts an example of a a quantile dotplot that expresses the uncertainty associated with a daily maximum temperature. The probability associated with each temperature is expressed with discrete countable bins and the predicted temperature is expressed with a line. Discretised depictions such as this, make decision making in the face of uncertainty easier for the viewer."
set.seed(1)
dotplot_data <- tibble(temp = 10 + round(rnorm(40, mean=10, sd=2))) 
mid <- round(mean(dotplot_data$temp), 2)
dotplot_data %>%
  ggplot(aes(x=temp)) +
  geom_dotplot(binwidth = 0.75, fill="grey") +
  theme_classic() +
  geom_segment(x=round(mid), xend=round(mid), 
               y=-1, yend=0.7, size=2) +
  geom_label(x=round(mid), y=0.7, label=paste0(mid, "°C"),
             fill = "black", fontface = "bold",
             colour="white") +
  scale_x_continuous(breaks = seq(10, 30, 5), 
                     limits = c(10, 30)) +
  theme(axis.line.y = element_blank(),
        axis.ticks.y = element_blank(),
        axis.title.y = element_blank(),
        axis.text.y = element_blank()) +
  labs(x = "Tomorrow's Daily Max Temperature")
```

##### Avoiding uncertainty creates a lack of trust
Not only does communicating uncertainty improve decisions but the mistrust created by communicating certainty in uncertain situations can be exploited. A 6-month survey of anti-mask groups on Facebook during the COVID-19 pandemic showed that the anti-maskers thought carefully about their grammar of graphics and made persuasive visualisations using the same data as pro-mask groups. They did this by exploiting information ignored by the pro-maskers [@Lee2021]. It is understood that deceptive plots can lead viewers to come to incorrect conclusions or significantly overstate effects or risks [@Pandey2015; @Padilla2022] but these incorrect takeaways cannot be mitigated with instructions in how to correctly understand the plot [@Boone2018]. This evidence indicates we are more likely than not to hurt our message when we ignore uncertainty information and trying to raise the general public's plot literacy is an insufficient strategy to curb conspiracy theories and misguided scientific communication.  In direct contrast to this, displaying numerical estimates of uncertainty information has shown to lead to greater trust in predictions [@Joslyn2012; @VanderBles2020]. While @Han2009 found people have more worry when presented with uncertainty regarding health outcomes, this worry is not a bad thing if the concern is warranted given the ambiguous situation.

##### Poeple might simply be chosing to not visualise uncertainty
The disconnect between the research in support of visualising uncertainty and the consensus against it may not be entirely driven by a lack of understanding of the literature. For example, at least one interviewee from the study by @Hullman2020a claimed that expertise implies that the signal being conveyed is significant, but also said they would omit uncertainty if it obfuscated the message they were trying to convey.  Other authors who were capable of calculating and and representing uncertainty well did not do it, and were unable to provide a self-satisfying reason why [@Hullman2020a]. These conflicting motivations are acknowledged in the paper itself where @Hullman2020a says:

> "It is worth noting that many authors seemed confident in stating rationales, as though they perceived them to be truths that do not require examples to demonstrate. It is possible that rationales for omission represent ingrained beliefs more than conclusions authors have drawn from concrete experiences attempting to convey uncertainty".

#### Continuous vs discrete displays
There are other distributional features, such as discreteness, that are important to consider in uncertainty visualisation. There is a reasonable amount of evidence that cumulative displays or discrete displays (such as a quantile dot-plots or histograms) are the best ways to express mass for decision making and probability estimates [@Fernandes2018; @Hofmann2012; Kay2016; @Hullman2018; @kale2019decision].

#### Distributions shoud be in a single plot
Elements from a single distribution should be displayed using a single plot, since displaying the features of one distribution across multiple plots makes the information hard to combine and results in some details (such as the estimate error) being completely ignored [@moritz2017trust; Correll2018].

### Perceptual tasks wrt uncertainty
Not only should we considered the hierarchy in the information we display, but we also need to consider the heuristics that connect some pieces of information to specific elementary tasks. For example @Hofmann2012 showed that polar co-ordinates are more effective than cartesian co-ordinates when considering data that depicts a 360 degree direction (a case where polar co-ordinates has a natural interpretation). Uncertainty also has a natural mapping that should be considered when we express it in a plot. Error is best mapped to fuzziness, location, and colour value; arrangement, size and transparency are an OK second choice; but saturation, hue, orientation and shape are unacceptable and have no intuitive connection to variance [@Maceachren2012]. No only do the graphical elements we map our features to matter, but the direction matters too. Graphical elements that are more fuzzy (fuzziness), further from centre (location), lighter (colour value), poorly arranged (arrangement), smaller (size), more transparent (transparency) are perceived to be more uncertain [@Maceachren2012]. This idea also extends to interval estimation, where questions about probability are best answered with gradients and questions about start and end times are easiest to answer with ambiguation [@Gschwandtnei2016]. Heuristics can work against us just as much as they can work for us. The sine illusion can cause the confidence interval of a smoothed sine curve to seem wider at the peaks than the troughs, causing us to underestimate uncertainty associated with changing values [@Vanderplas2015]. Therefore we should not only keep the hierarchy of information in mind when we map features of our distribution, but also take advantage of these intuitive mappings when we can.

# Bibliography

```{r, include=FALSE, eval=FALSE}
library(spelling)
qmd <- "ideas/confirmation/confirmationreport.qmd"
ignore <- readLines("WORDLIST")
check_spelling <- spell_check_files(
  qmd,
  ignore = ignore,
  lang = "en_GB"
)
if (nrow(check_spelling) > 0) {
  print(check_spelling)
  stop("Check spelling in Qmd files!")
}
```

### 2023-12-18 writingdumpster.qmd
updated dumpster file

--- 2023-12-18 index.qmd
@@
 ---
-title: "Uncertainty Literature Review"
-author: Harriet Mason
-bibliography: references.bib
-date: last-modified
-format: pdf
-editor_options: 
-  chunk_output_type: console
+title: "Writing Trash Bin"
+author: "Harriet Mason"
+format: html
 ---
-## Format Notes
-##### Usual Format
-1. Introduction
-2. Notation (if relevant)
-3-5. Details of literature review broken into several sub sections. First usually provides context to the problem and the following sections answer questions that context highlights.
-6. Future Work/ Conclusion
+# Why this exists
+I have an information hoarding problem, this helps me throw out sections of the literature review that are not necessarily relevant without actually throwing it out.
+Knowing that I can come back to this at any point and pull information back, makes me more comfortable with removing it from the report.
 
-##### Usual Format for this paper
-1. Introduction
-2. Definitions of Uncertainty
+# General Thoughts For Actual Review
+##### Introduction
+##### Defining Uncertainty
   - The more taxonomies I read the more convinced I am the "defining uncertainty" is the best one.
@@
   - Combining uncertainty is also a reccurring theme in these papers
-3. Visualisation in the context of uncertainty communication
-  - Basically I want a brief section highlighting the importance of visualisation and discussing it in relation to other forms of uncertainty communication.
-  - especially considering some uncertainty visualisation papers discuss using just a text based method to communicate uncerainty
-4. Specific Section on Uncertainty Visualisation
-  - There is also two kinds of papers, those that compare elements of a plot (like mapping uncertainty to an aesthetic) and those that compare some expression of a distribution (like a scatter plot compared to a box plot).
-  - try to map back to david speigleholder paper i.e. uncertainty means lots of different things
-5. Future work/Conclusion
-
-# Random thoughts
 - I think it is also important to highlight if uncertainty is meta information or if it is a variable in of itself.
@@
 
-```{r}
-#| echo: false
-#| warning: false
-#| message: false
+##### Visualisation as a form of Uncertainty communication
+  - Basically I want a brief section highlighting the importance of visualisation and discussing it in relation to other forms of uncertainty communication.
+  - especially considering some uncertainty visualisation papers discuss using just a text based method to communicate uncerainty
 
-# Load libraries
-library(tidyverse)
-library(tsibbledata)
-library(Vizumap)
-library(fable)
-```
-# Introduction
-Think back to the last time you made some sort of data visualisation.
-What was the purpose of that visualisation?
-Was it to better understand your data?
-Was it to help you make a decision?
-Was it to to communicate that decision to someone else?
-Now think about the last time you expressed some form of uncertainty.
-Was it a set of numerical confidence intervals?
-Maybe they were expressed as a set of values in a table.
-Did you consider visualising your uncertainty instead?
-There are many stages in our analysis that benefit from the power of data visualisation, however this does not mean it is always done with success.
-Visualization is an important step in exploratory data analysis and it is often utilised to **learn** what is important about a data set.
-The importance of data driven discovery is highlighted by data sets such as Anscombe's quartet [@anscombe] or the Datasaurus Dozen [@datasaurpkg].
-Each of the pairwise plots in these data sets have the same summary statistics but strikingly different information when visualised.
-Anscombe quartet is shown in @fig-anscombe, because describing the data is never the same as seeing it.
-Instead of having to repeatedly check endless hypothesis to find interesting numerical features, visualisations **tell** us what is important about our data.
-This powerful aspect of data visualisation is poorly or seldom used in later stages when we are communicating our findings, specifically with respect to uncertainty.
+##### Uncertainty visualisation best practices
+  - There is also two kinds of papers, those that compare elements of a plot (like mapping uncertainty to an aesthetic) and those that compare some expression of a distribution (like a scatter plot compared to a box plot).
+  - try to map back to david speigleholder paper i.e. uncertainty means lots of different things
 
-```{r}
-#| echo: false
-#| warning: false
-#| message: false
-#| label: fig-anscombe
-#| fig-cap: "The four scatter plots that make up Anscombe's quartet.
-The four scatter plots are visually distinct but have the same mean, standard deviation, and correlation.
-The visualisation highlights the importance of plotting your data to identify interesting features that are hidden by other summary statistics."
-tibble(x = c(anscombe$x1, anscombe$x2,
-             anscombe$x3, anscombe$x4),
-       y = c(anscombe$y1, anscombe$y2,
-             anscombe$y3, anscombe$y4),
-       Plot = c(rep("Plot 1",11), rep("Plot 2",11), 
-                rep("Plot 3",11), rep("Plot 4",11))) %>%
-  ggplot(aes(x,y)) +
-  geom_point(aes(fill=Plot), colour="black", 
-             size=3, pch=21, alpha=0.75) +
-  facet_wrap(~Plot) +
-  theme_classic() +
-  theme(aspect.ratio = 1,
-        legend.position = "none") +
-  scale_fill_brewer(type = "qual", palette = 4)
-```
-
-##### Secondary Nature of Uncertainty
-An overwhelming consensus among visualisation authors seems to be that uncertainty is secondary to estimations.
-There is a belief held by those that work with data that the uncertainty associated with an estimate (the noise) only exists to hide the estimate itself (the signal).
-From this point of view uncertainty is optional information that only exists to hinder our message.
-Therefore despite its alleged importance, when simplifying a plot uncertainty the first thing to go.
-This belief is also reflected in the development of new uncertainty visualisations.
-Often when trying to visualise multi-dimensional data, uncertainty is relegated to unimportant aesthetics in the plot, often of lower importance than the estimate where it is easily ignored [@Correll2018; @Lucchesi2017].
-Sometimes uncertainty is not relegated to an undesirable aesthetic but instead incorporated using interactivity to allow users to explore the complicated space themselves [@Potter2009].
-Even the literature about uncertainty communication expresses an implicit belief that it is of secondary importance to the estimates or context of the data.
-
-####  no visual distinction between sample/pop variance
-@Bella2005 found that most participants were ignorant to the fact that error bars are used for both confidence intervals and standard error bars, two wildly different indicators of precision. 
-**also the paper sherry sent me**
-
-# Definitions of Uncertainty
-##### No clear definition of uncertainty
-This issues surrounding uncertainty visualisation is not helped by the fact that the term "uncertainty" lacks a commonly accepted definition in the literature.
-@Lipshitz1997 even commented that “there are almost as many definitions of uncertainty as there are treatments of the subject” .
-This mishmash of terminology leads to a large body of work, all claiming to finding the best visualisation or expression of of "uncertainty" but most don't even seem to agree on what uncertainty is.
-The most encompassing definition of uncertainty I have seen comes from @utypo who define uncertainty as **"any deviation from the unachievable ideal of completely deterministic knowledge of the relevant system"**.
-This definition encapsulates many concepts adjacent to randomness such as predictions, probability mass distributions (PMF), estimate error, and any data that is not a set of deterministic outcomes.
-This is the definition of uncertainty I will use for the rest of this report.
-
-##### Uncertainty taxonomy
-More commonly uncertainty is defined using a taxonomy rather than a strict definition.
-There are a few taxonomies for uncertainty, but, just like the definition, most of them are a subset of the one laid out by @utypo.
-To complete the definition I will include that taxonomy here.
-@fig-taxonomy is an illustration of the taxonomy presented by @utypo.
-In this taxonomy, there are three things we need to consider for each "uncertainty" we encounter through the modelling process.
-First, we need to consider the source of the uncertainty.
-Is this uncertainty coming from inaccurate measurements or a poorly defined model?
-This is the *location* of the uncertainty.
-Second, consider how well you can quantify this uncertainty.
-Do you know exactly how much measurement error there is in each observation or are you not even aware if there is a measurement error?
-This is the *level* of your uncertainty, and it ranges from discrete to total ignorance.
-Finally, consider how this uncertainty came into existence.
-Is it a result of a naturally random process (epistemic) or is it due to imperfect information and could be improved (aleatory).
-This is the *nature* of your uncertainty.
-@utypo then goes on to describe mapping our uncertainty in a 3D space that is defined by its location, level, and nature, but I think the taxonomy is more easily understood as a series of questions we need to consider when we are trying to quantify uncertainty.
-
-![Depicts an illustration of the taxonomy described in @utypo.
-From right to left the drawing shows the location, level and nature of uncertainty with examples of that category underneath.
-A specific source of uncertainty from the location can be mapped to a level of ignorance that can increase or decrease (i.e.
-moving up or down the green line) depending on the nature of the uncertainty.
-Identifying the location, level and nature of your uncertainty allows you to better understand it.
-](taxonomyvis.jpeg){#fig-taxonomy}
-
-##### You should not ignore "uncertainty" that seems unrelated to your problem
-While information about the sources of our uncertainty and the type of uncertainty may seem like an unimportant secondary step in uncertainty visualisation, communicating these features of uncertainty helps decision makers make more informed choices.
-@Padilla2021 found that low forecaster confidence or high model uncertainty both contribute to more conservative judgements by decision makers.
-Failing to communicate the nature of your uncertainty can result in underestimation or overestimation of failure probabilities [@Kiureghian2009].
-Additionally @Gustafson2019 found that the framing of our uncertainty, (i.e.
+# Previous literature review
+#####Taxonomy Benefits
+- Distinctions at this taxonomy level seem to be important also untangling helps to understand it
+  - @Padilla2021 found that low forecaster confidence or high model uncertainty both contribute to more conservative judgements by decision makers.
+  - Failing to communicate the nature of your uncertainty can result in underestimation or overestimation of failure probabilities [@Kiureghian2009]
+  - @Gustafson2019 found that the framing of our uncertainty, (i.e.
 if the source of uncertainty is from a lack of knowledge, approximations, unknown unknowns, or disagreement among parties) was found to not have a detrimental effect on the belief in the estimates, perceived credibility, or behavioural intentions of the decision makers.
-This means communicating secondary information about your uncertainty can provide additional benefits to decision makers without additional negative repercussions.
-# Unorganised
 
@@
 Therefore we should not only keep the hierarchy of information in mind when we map features of our distribution, but also take advantage of these intuitive mappings when we can.
-
-# Bibliography
-
-```{r, include=FALSE, eval=FALSE}
-library(spelling)
-qmd <- "ideas/confirmation/confirmationreport.qmd"
-ignore <- readLines("WORDLIST")
-check_spelling <- spell_check_files(
-  qmd,
-  ignore = ignore,
-  lang = "en_GB"
-)
-if (nrow(check_spelling) > 0) {
-  print(check_spelling)
-  stop("Check spelling in Qmd files!")
-}
-```

### 2023-12-18 index.qmd
culled literature reivew

@@
 ---
-## Format Notes
-##### Usual Format
-1. Introduction
-2. Notation (if relevant)
-3-5. Details of literature review broken into several sub sections. First usually provides context to the problem and the following sections answer questions that context highlights.
-6. Future Work/ Conclusion
 
-##### Usual Format for this paper
-1. Introduction
-2. Definitions of Uncertainty
-  - The more taxonomies I read the more convinced I am the "defining uncertainty" is the best one.
-Not only is it the most complete, but it also identifies WHY some uncertainties are difficult to visualise (specifically which part of the taxonomy) and suggests HOW you can best communicate these uncertainties.
-    - A lot of other papers just make a comment like "oh some of these are hard to communicate" but that paper actually goes into detail instead of hand waving the problem away.
-    - It also explains why the uncertainty visualisation literature focuses on a specific subset of uncertainty cases.
-  - Examples
-    - Null Values
-    - Variance
-    - Hypothesis Testing
-    - Distribution
-  - Combining uncertainty is also a reccurring theme in these papers
-3. Visualisation in the context of uncertainty communication
-  - Basically I want a brief section highlighting the importance of visualisation and discussing it in relation to other forms of uncertainty communication.
-  - especially considering some uncertainty visualisation papers discuss using just a text based method to communicate uncerainty
-4. Specific Section on Uncertainty Visualisation
-  - There is also two kinds of papers, those that compare elements of a plot (like mapping uncertainty to an aesthetic) and those that compare some expression of a distribution (like a scatter plot compared to a box plot).
-  - try to map back to david speigleholder paper i.e. uncertainty means lots of different things
-5. Future work/Conclusion
-
-# Random thoughts
-- I think it is also important to highlight if uncertainty is meta information or if it is a variable in of itself.
-Meta information (such as contiunity, exhangeability, etc) are usually set to specific visual features (e.g.
-VSUP clearly sees uncertainty as meta information)
-- A lot of work that evaluates uncertainty visualisations checks to see if the mean can be gathered from a visualisation, but if we view uncertainty as meta-information, then with a high uncertainty we wouldn't want the mean to be visible
-  - Also an example of why the motivation matters for these things.
-If you don't know "WHY" someone is using a visualisation for uncertainty and just pick the "best" one, you are going to get a visualisation that does a poor job of its task.
-  - put uncertainty in the context of other meta-data of variables
-
-```{r}
-#| echo: false
-#| warning: false
-#| message: false
-
-# Load libraries
-library(tidyverse)
-library(tsibbledata)
-library(Vizumap)
-library(fable)
-```
-# Introduction
+# 1. Introduction
 Think back to the last time you made some sort of data visualisation.
@@
 The visualisation highlights the importance of plotting your data to identify interesting features that are hidden by other summary statistics."
+library(tidyverse)
 tibble(x = c(anscombe$x1, anscombe$x2,
@@
 
-##### Secondary Nature of Uncertainty
-An overwhelming consensus among visualisation authors seems to be that uncertainty is secondary to estimations.
-There is a belief held by those that work with data that the uncertainty associated with an estimate (the noise) only exists to hide the estimate itself (the signal).
-From this point of view uncertainty is optional information that only exists to hinder our message.
-Therefore despite its alleged importance, when simplifying a plot uncertainty the first thing to go.
-This belief is also reflected in the development of new uncertainty visualisations.
-Often when trying to visualise multi-dimensional data, uncertainty is relegated to unimportant aesthetics in the plot, often of lower importance than the estimate where it is easily ignored [@Correll2018; @Lucchesi2017].
-Sometimes uncertainty is not relegated to an undesirable aesthetic but instead incorporated using interactivity to allow users to explore the complicated space themselves [@Potter2009].
-Even the literature about uncertainty communication expresses an implicit belief that it is of secondary importance to the estimates or context of the data.
-
-####  no visual distinction between sample/pop variance
-@Bella2005 found that most participants were ignorant to the fact that error bars are used for both confidence intervals and standard error bars, two wildly different indicators of precision. 
-**also the paper sherry sent me**
-
-# Definitions of Uncertainty
-##### No clear definition of uncertainty
+# 2. Defining Uncertainty
 This issues surrounding uncertainty visualisation is not helped by the fact that the term "uncertainty" lacks a commonly accepted definition in the literature.
@@
 
-##### Uncertainty taxonomy
+## 2.1 The role of context in defining uncertainty
+- (no visual distinction between sample/pop variance) @Bella2005 found that most participants were ignorant to the fact that error bars are used for both confidence intervals and standard error bars, two wildly different indicators of precision.
+(**also the paper sherry sent me**)
+- **notes from the Xiao-li Meng papers**
+- **papers that talk about task specific stuff**
+
+## 2.2 Taxonomies of Uncertainty
 More commonly uncertainty is defined using a taxonomy rather than a strict definition.
-There are a few taxonomies for uncertainty, but, just like the definition, most of them are a subset of the one laid out by @utypo.
-To complete the definition I will include that taxonomy here.
+
 @fig-taxonomy is an illustration of the taxonomy presented by @utypo.
@@
 
-##### You should not ignore "uncertainty" that seems unrelated to your problem
-While information about the sources of our uncertainty and the type of uncertainty may seem like an unimportant secondary step in uncertainty visualisation, communicating these features of uncertainty helps decision makers make more informed choices.
-@Padilla2021 found that low forecaster confidence or high model uncertainty both contribute to more conservative judgements by decision makers.
-Failing to communicate the nature of your uncertainty can result in underestimation or overestimation of failure probabilities [@Kiureghian2009].
-Additionally @Gustafson2019 found that the framing of our uncertainty, (i.e.
-if the source of uncertainty is from a lack of knowledge, approximations, unknown unknowns, or disagreement among parties) was found to not have a detrimental effect on the belief in the estimates, perceived credibility, or behavioural intentions of the decision makers.
-This means communicating secondary information about your uncertainty can provide additional benefits to decision makers without additional negative repercussions.
-# Unorganised
+# 3. Visualisation as a form of uncertainty comunication
+## 3.1 General attitudes towards uncertainty
+## 3.2 Attitudes towards uncertainty visualisation
 
-##### The benefits of Unvertainty visualisaiton
-Visualisations can provide a more complete picture of a risk than numerical summaries alone.
-Even something as simple as sketching a distribution before recalling statistics or making predictions can greatly increase the accuracy of those measures [@Hullman2018; @Goldstein2014].
-While there is some evidence that confidence intervals provided in text form only are less likely to be misinterpreted than graphics [@Savelli2013], text is insufficient to express more complicated aspects of a distribution, such as mass.
-Expressing uncertainty verbally decreases the perceived reliability and trustworthiness of the source [@VanderBles2020].
-Any confusion caused by expressing uncertainty as a visualisation could also be due to a lack of exposure, since @Kay2016 found people repeatedly exposed to the same uncertainty visualisations quickly get better at making judgements.
-Additionally, visualisation allow for interactive graphics that provide a more in depth understanding of probability [@Potter2009; @Ancker2009] and infographics that make uncertainty more accessible for people with poor numeracy skills [@Ancker2009].
-**application in energy**
-The use of uncertainty in high dimensional environments is especially important in energy data.
-Large models that incorporate spatial-temporal data from many sources and systems are used to predict energy uses in the short and long term.
-Understanding how to improve and make better decisions in these models is imperative in both the daily operation of the energy sector as well as in the transition from fossil fuels to clean energy.
-The energy sector needs better heuristics to make energy supply analysis less costly to conduct [@Stenclik2021], therefore it is an incredibly relevant application of uncertainty visualisation techniques.
+# 4. Uncertainty Visualisation Best Practices
 
-##### People dont visualise uncertainty but it is considered important
-Despite these benefits, there is evidence that we don't visualise uncertainty as often as we should.
-A survey conducted by @Hullman2020a found that majority of visualisation authors agreed that expressing uncertainty is important and should be done more often than it currently is, some even agreed that failing to do so is tantamount to fraud.
-Despite this, only a quarter of respondents included uncertainty in 50% or more of their visualisations [@Hullman2020a].
-Meaning participants were convinced that visualising uncertainty is morally important but were able to provide self sufficient reasoning that allows them to avoid doing it.
-Some economists suggest that visualisation authors are responding to incentives that make it tempting to avoid visualising uncertainty, even if those incentives are based more in perception than reality [@Manski2020].
-The study by @Hullman2020a found that the most common reasons authors don't visualise uncertainty despite knowing it's moral importance are: not wanting to overwhelm the audience; an inability to calculate the uncertainty; a lack of access to the uncertainty information; and not wanting to make their data seem questionable [@Hullman2020a].
-
-##### Effect of uncertainty visualisation on decision making
-If decision markers are not presented with the uncertainty about an estimate the data analysts have, for all intents and purposes, made the decision for the decision maker.
-Upon further interviews @Hullman2020a found that authors believed uncertainty would overwhelm the audience and make their data seem questionable because decision makers are unable to understand uncertainty.
-This belief, while pervasive, is not true.
-There is some research that suggests laypeople cannot understand complicated concepts in statistical thinking (such as trick questions on hypothesis tests or the difference between Frequentist and Bayesian thinking) [@Hoekstra2014; @Bella2005] but there is a large amount of research suggesting that presenting uncertainty information improves decision making, both experimentally [@Joslyn2012; @Savelli2013; @Kay2016; @Fernandes2018] and in practice [@Al-Kassab2014].
-As a matter of fact, doing what many authors currently do (providing only a deterministic outcome with no uncertainty) causes decision makers to be *less* decisive and have completely unbounded expectations on an outcome [@Savelli2013].
-This reality cannot be avoided by providing secondary or non-specific information such as explaining calculations [@Joslyn2012], explaining the advantages of a recommendation [@Joslyn2012], or expressing uncertainty in vague terms [@Erev_1990; @Olson_1997], all of which are undesirable for decision makers and lead to measurably worse decisions [@Joslyn2012; @Erev_1990; @Olson_1997].
-One of the most popular depictions of uncertainty for decision making is a quantile dotplot, shown in @fig-quantdot.
-It is important to avoid whittling down the problem **too** much.
-Providing a categorical decision alone is somewhat useless [@Joslyn2012], and visualising a single estimate is akin to providing a decision or expressing no uncertainty at all.
-
-```{r}
-#| echo: false
-#| warning: false
-#| message: false
-#| label: fig-quantdot
-#| fig-cap: "This plot depicts an example of a a quantile dotplot that expresses the uncertainty associated with a daily maximum temperature.
-The probability associated with each temperature is expressed with discrete countable bins and the predicted temperature is expressed with a line.
-Discretised depictions such as this, make decision making in the face of uncertainty easier for the viewer."
-set.seed(1)
-dotplot_data <- tibble(temp = 10 + round(rnorm(40, mean=10, sd=2))) 
-mid <- round(mean(dotplot_data$temp), 2)
-dotplot_data %>%
-  ggplot(aes(x=temp)) +
-  geom_dotplot(binwidth = 0.75, fill="grey") +
-  theme_classic() +
-  geom_segment(x=round(mid), xend=round(mid), 
-               y=-1, yend=0.7, size=2) +
-  geom_label(x=round(mid), y=0.7, label=paste0(mid, "°C"),
-             fill = "black", fontface = "bold",
-             colour="white") +
-  scale_x_continuous(breaks = seq(10, 30, 5), 
-                     limits = c(10, 30)) +
-  theme(axis.line.y = element_blank(),
-        axis.ticks.y = element_blank(),
-        axis.title.y = element_blank(),
-        axis.text.y = element_blank()) +
-  labs(x = "Tomorrow's Daily Max Temperature")
-```
-
-##### Avoiding uncertainty creates a lack of trust
-Not only does communicating uncertainty improve decisions but the mistrust created by communicating certainty in uncertain situations can be exploited.
-A 6-month survey of anti-mask groups on Facebook during the COVID-19 pandemic showed that the anti-maskers thought carefully about their grammar of graphics and made persuasive visualisations using the same data as pro-mask groups.
-They did this by exploiting information ignored by the pro-maskers [@Lee2021].
-It is understood that deceptive plots can lead viewers to come to incorrect conclusions or significantly overstate effects or risks [@Pandey2015; @Padilla2022] but these incorrect takeaways cannot be mitigated with instructions in how to correctly understand the plot [@Boone2018].
-This evidence indicates we are more likely than not to hurt our message when we ignore uncertainty information and trying to raise the general public's plot literacy is an insufficient strategy to curb conspiracy theories and misguided scientific communication.
-In direct contrast to this, displaying numerical estimates of uncertainty information has shown to lead to greater trust in predictions [@Joslyn2012; @VanderBles2020].
-While @Han2009 found people have more worry when presented with uncertainty regarding health outcomes, this worry is not a bad thing if the concern is warranted given the ambiguous situation.
-
-##### Poeple might simply be chosing to not visualise uncertainty
-The disconnect between the research in support of visualising uncertainty and the consensus against it may not be entirely driven by a lack of understanding of the literature.
-For example, at least one interviewee from the study by @Hullman2020a claimed that expertise implies that the signal being conveyed is significant, but also said they would omit uncertainty if it obfuscated the message they were trying to convey.
-Other authors who were capable of calculating and and representing uncertainty well did not do it, and were unable to provide a self-satisfying reason why [@Hullman2020a].
-These conflicting motivations are acknowledged in the paper itself where @Hullman2020a says:
-
-> "It is worth noting that many authors seemed confident in stating rationales, as though they perceived them to be truths that do not require examples to demonstrate.
-It is possible that rationales for omission represent ingrained beliefs more than conclusions authors have drawn from concrete experiences attempting to convey uncertainty".
-
-#### Continuous vs discrete displays
-There are other distributional features, such as discreteness, that are important to consider in uncertainty visualisation.
-There is a reasonable amount of evidence that cumulative displays or discrete displays (such as a quantile dot-plots or histograms) are the best ways to express mass for decision making and probability estimates [@Fernandes2018; @Hofmann2012; Kay2016; @Hullman2018; @kale2019decision].
-
-#### Distributions shoud be in a single plot
-Elements from a single distribution should be displayed using a single plot, since displaying the features of one distribution across multiple plots makes the information hard to combine and results in some details (such as the estimate error) being completely ignored [@moritz2017trust; Correll2018].
-
-### Perceptual tasks wrt uncertainty
-Not only should we considered the hierarchy in the information we display, but we also need to consider the heuristics that connect some pieces of information to specific elementary tasks.
-For example @Hofmann2012 showed that polar co-ordinates are more effective than cartesian co-ordinates when considering data that depicts a 360 degree direction (a case where polar co-ordinates has a natural interpretation).
-Uncertainty also has a natural mapping that should be considered when we express it in a plot.
-Error is best mapped to fuzziness, location, and colour value; arrangement, size and transparency are an OK second choice; but saturation, hue, orientation and shape are unacceptable and have no intuitive connection to variance [@Maceachren2012].
-No only do the graphical elements we map our features to matter, but the direction matters too.
-Graphical elements that are more fuzzy (fuzziness), further from centre (location), lighter (colour value), poorly arranged (arrangement), smaller (size), more transparent (transparency) are perceived to be more uncertain [@Maceachren2012].
-This idea also extends to interval estimation, where questions about probability are best answered with gradients and questions about start and end times are easiest to answer with ambiguation [@Gschwandtnei2016].
-Heuristics can work against us just as much as they can work for us.
-The sine illusion can cause the confidence interval of a smoothed sine curve to seem wider at the peaks than the troughs, causing us to underestimate uncertainty associated with changing values [@Vanderplas2015].
-Therefore we should not only keep the hierarchy of information in mind when we map features of our distribution, but also take advantage of these intuitive mappings when we can.
+# 5. Future work/Conclusion
 

### 2023-12-18 index.qmd
updated notes and sections after mendeley paper sort

@@
 
-## 2.2 Taxonomies of Uncertainty
+## 2.2 Taxonomies
+### 2.2.1 Taxonomies of Uncertainty
 More commonly uncertainty is defined using a taxonomy rather than a strict definition.
@@
 
-# 3. Visualisation as a form of uncertainty comunication
-## 3.1 General attitudes towards uncertainty
-## 3.2 Attitudes towards uncertainty visualisation
+### 2.2.1 Taxonomies of Visual Uncertainty
 
-# 4. Uncertainty Visualisation Best Practices
+# 3.General Attitudes...
+## 3.1 Towards uncertainty in general
+## 3.2 Towards uncertainty visualisation
+
+# 4 Uncertainty Visualisation Best Practices
+# 4.1 The current landscape of uncertainty visualisation research
+# 4.2 Take aways from the literature
 

### 2023-12-18 writingdumpster.qmd
updated notes and sections after mendeley paper sort

@@
 # General Thoughts For Actual Review
+##### General
+- I might need to go back through and add in notes about general visualisation stuff and bad papers for motivations
+- could also help to include a bunch of papers that talk about density visualisation without mentioning uncertainty visualisation in the "context" section. (for now have left those unsorted)
+
 ##### Introduction
+
 ##### Defining Uncertainty
@@
 ##### Visualisation as a form of Uncertainty communication
+- A LOT OF the work in uncertainty communication has a sub-field that is visualisation specific
   - Basically I want a brief section highlighting the importance of visualisation and discussing it in relation to other forms of uncertainty communication.

### 2023-12-19 index.qmd
added in some bits for context section

@@
 ## 2.1 The role of context in defining uncertainty
+When statistics is, at its core, the study of uncertainty, discussing uncertainty visualisation as though it is separate from other statistical visualisations is inherrently confusing.
+This distinction is largely ignored rather than clarified in the literature, and it leads to adjacent unanswered questions.
+Is our goal in uncertainty communication to acknowledge and convey the inherrent randomness that exists in all statistical calculations and models?
+Is it to supress signal that could be the result of random variance?
+Is it to find the best way to convey the mass function of a random variable?
+A survey of the literature will leave the answers to these questions unclear, however this inherrent issue has not gone unnoticed.
+
+**need for task based understanding**
+- several review papers acknowledge this issue is causing 
+- **papers that talk about task specific stuff**
+- This problem of a task based philosiphy extends all the way back to philisophical uncerstandings of statistics and variance.
+- **notes from the Xiao-li Meng papers**
+- **notes from john paper**
+- While the statistics side brings up the purpose of a model or estimate for a philisophical consideration, it is a real, practical, and urgent problem for uncertainty visualisation. 
+- **go through papers with endless different goals**
+- **review paper that has an entire paragraph discussing differnet findings for box plots based on the task**
+
+**need for statistics links**
 - (no visual distinction between sample/pop variance) @Bella2005 found that most participants were ignorant to the fact that error bars are used for both confidence intervals and standard error bars, two wildly different indicators of precision.
 (**also the paper sherry sent me**)
-- **notes from the Xiao-li Meng papers**
-- **papers that talk about task specific stuff**
 

### 2024-01-04 index.qmd
worked on first draft of context section

@@
 # 2. Defining Uncertainty
-This issues surrounding uncertainty visualisation is not helped by the fact that the term "uncertainty" lacks a commonly accepted definition in the literature.
-@Lipshitz1997 even commented that “there are almost as many definitions of uncertainty as there are treatments of the subject” .
-This mishmash of terminology leads to a large body of work, all claiming to finding the best visualisation or expression of of "uncertainty" but most don't even seem to agree on what uncertainty is.
-The most encompassing definition of uncertainty I have seen comes from @utypo who define uncertainty as **"any deviation from the unachievable ideal of completely deterministic knowledge of the relevant system"**.
-This definition encapsulates many concepts adjacent to randomness such as predictions, probability mass distributions (PMF), estimate error, and any data that is not a set of deterministic outcomes.
-This is the definition of uncertainty I will use for the rest of this report.
 
-## 2.1 The role of context in defining uncertainty
-When statistics is, at its core, the study of uncertainty, discussing uncertainty visualisation as though it is separate from other statistical visualisations is inherrently confusing.
-This distinction is largely ignored rather than clarified in the literature, and it leads to adjacent unanswered questions.
+## 2.1 Uncertainty is goal dependent
+Statistics is, at its core, the study of uncertainty.
+Therefore discussing uncertainty visualisation a a separate sub domain to "normal" data visualisation is inherently confusing.
+What is typically meant by "uncertainty" visualisation is "noise", that is, we want to present some signal cushioned by its natural variance.
+Unfortunately, this distinction between "signal" and "noise" is entirely goal dependent.
+
+The entire process of data analysis, from deciding what should be observed as data through to communicating that data in a plot is governed by human decision and the goal of analysis.
+This conceptual way of thinking about uncertainty has been rattling around in the philisophical discussions of uncertainty for quite a while.
+
+@Otsuka2023 points out that the philosophy of statistics is build upon taking real world entities and boiling them down into probabilistic objects and this ontological process is largely dependent on our goals.
+For exmaple, when we record coin flips we typically want to model the behavior with a binomial distribution, so we ignore the outcome of a coin landing on its side despite the fact it is a real world possibility.
+
+@Meng2014 shows that the entire concept of signal and noise are relative and the noise for one goal can be the signal for another.
+This break down is done by considering the data source, the "resolution" of the problem and data, and the phases involved in the analysis.
+
+@Carlin2023
+
+Since uncertainty visualisation is largely goal dependent, hard and fast uncertainty visualisation rules are difficult to design.
+Additionally communication of statistics cannot ignore the ontological process discussed by @Otsuka2023 because the goal of communication is to map those statistical objects back to real world entities.
+
+After all, the distinction between uncertainty and signal is largely determined by the goal of an analysis rather than any inherent properties of our data.
+
+This leads to several unanswered questions.
 Is our goal in uncertainty communication to acknowledge and convey the inherrent randomness that exists in all statistical calculations and models?
@@
 A survey of the literature will leave the answers to these questions unclear, however this inherrent issue has not gone unnoticed.
+
+- Should the visualisations change based on these goals?
 
@@
 (**also the paper sherry sent me**)
+
+Uncertainty can only be defines within a specific problem.
+There is not a universal motivation agnostic "uncertainty" similarly as there is no motivation agnostic "signal".
+Therefore, in our following definitions of uncertainty, we will assume that the uncertainties we discuss are connected to some specific question.
 
@@
 # 4 Uncertainty Visualisation Best Practices
+In this section we will first go over the current state of uncertainty visualisation research, which is currently a collection of evaluation studies and review papers, and discuss what it is doing well and what could be done better.
+Following that, we will provide general reccomendations from the literature.
+
 # 4.1 The current landscape of uncertainty visualisation research
+In section 2.1 it was mentioned that uncertainty must be defined within a specific motivating question, otherwise it inherently does not make sense.
+A large difficulty with the uncertainty visualisation evalusation studies is that this rule is not followed.
+There are a shockingly large number of evaluation studies that seem to pay no attention to the information that is relevant to the question they are asking, and show participants a selection of seemingly random visualisations from a statistical point of view.
+
+- inferential uncertainty and outcome uncertainty ARE NOT THE SAME THING??? they visualise DIFFERENT DISTRIBUTIONS
+
+We define two primary motivations for uncertainty visualisation.
+  1) To prevent deterministic conclusions from a random signal (uncertainty as noise)
+  2) To convey information about a variance, probability, or other random event (uncertainty as signal)
+
 # 4.2 Take aways from the literature

### 2024-01-04 index.qmd
halfway through first draft of context section

@@
 @Otsuka2023 points out that the philosophy of statistics is build upon taking real world entities and boiling them down into probabilistic objects and this ontological process is largely dependent on our goals.
-For exmaple, when we record coin flips we typically want to model the behavior with a binomial distribution, so we ignore the outcome of a coin landing on its side despite the fact it is a real world possibility.
+For example, when we record coin flips we typically want to model the behavior with a binomial distribution, so we ignore the outcome of a coin landing on its side despite the fact it is a real world possibility.
 
@@
 This break down is done by considering the data source, the "resolution" of the problem and data, and the phases involved in the analysis.
+@Meng2021 highlights that data needs to be treated as an output from another process rahter than an value neutral input.
 
-@Carlin2023
+@Carlin2023 highlights a range of pitfalls that come up frequently when statisticians use regression models and they connect these pitfalls to the teaching of statistical methods as goal agnostic tools.
+@Carlin2023 suggest that every research question belongs to one of three distinct types: descriptive, predictive, and causal, each of which has its own appropriate statistical methods.
+
+@Amar2005
 

### 2024-01-04 paper.qmd
renamed paper and added notes from meeting

--- 2024-01-04 index.qmd
@@
 The entire process of data analysis, from deciding what should be observed as data through to communicating that data in a plot is governed by human decision and the goal of analysis.
-This conceptual way of thinking about uncertainty has been rattling around in the philisophical discussions of uncertainty for quite a while.
+This conceptual way of thinking about uncertainty has been rattling around in the philosophical discussions of uncertainty for quite a while.
 
@@
 
-# 4 Uncertainty Visualisation Best Practices
 In this section we will first go over the current state of uncertainty visualisation research, which is currently a collection of evaluation studies and review papers, and discuss what it is doing well and what could be done better.
@@
 
-# 4.1 The current landscape of uncertainty visualisation research
+# 4 The current landscape of uncertainty visualisation research
+# 4.1 The landscape
 In section 2.1 it was mentioned that uncertainty must be defined within a specific motivating question, otherwise it inherently does not make sense.
@@
 
-# 4.2 Take aways from the literature
+# 4.2 Key points from the literature
+
+# 5 Great Examples
+- best things you have seen and why they are the best
+- can't have a best practices because you need experimental results and my point is that the experiments aren't testing the right things
 

### 2024-01-04 paper.qmd
finished context section and moved bits that didnt fit to other sections

--- 2024-01-04 index.qmd
@@
 
+- @Suh2023 found that the main issue with deployment of AI was communication between data science and subject matter experts, since subject matter experts didn't want technical presentations that didn't explain exactly when and how a model would fail.
+
 ```{r}
@@
 # 2. Defining Uncertainty
+Uncertainty visualisation is not made any easier by the fact that the term "uncertainty" lacks a commonly accepted definition in the literature.
+@Lipshitz1997 even commented that “there are almost as many definitions of uncertainty as there are treatments of the subject”.
+This mishmash of terminology leads to a large body of work, all claiming to finding the best visualisation or expression of of "uncertainty" but most don't even seem to agree on what uncertainty is.
 
-## 2.1 Uncertainty is goal dependent
-Statistics is, at its core, the study of uncertainty.
-Therefore discussing uncertainty visualisation a a separate sub domain to "normal" data visualisation is inherently confusing.
-What is typically meant by "uncertainty" visualisation is "noise", that is, we want to present some signal cushioned by its natural variance.
-Unfortunately, this distinction between "signal" and "noise" is entirely goal dependent.
-
-The entire process of data analysis, from deciding what should be observed as data through to communicating that data in a plot is governed by human decision and the goal of analysis.
-This conceptual way of thinking about uncertainty has been rattling around in the philisophical discussions of uncertainty for quite a while.
-
-@Otsuka2023 points out that the philosophy of statistics is build upon taking real world entities and boiling them down into probabilistic objects and this ontological process is largely dependent on our goals.
-For exmaple, when we record coin flips we typically want to model the behavior with a binomial distribution, so we ignore the outcome of a coin landing on its side despite the fact it is a real world possibility.
-
-@Meng2014 shows that the entire concept of signal and noise are relative and the noise for one goal can be the signal for another.
-This break down is done by considering the data source, the "resolution" of the problem and data, and the phases involved in the analysis.
-
-@Carlin2023
-
-Since uncertainty visualisation is largely goal dependent, hard and fast uncertainty visualisation rules are difficult to design.
-Additionally communication of statistics cannot ignore the ontological process discussed by @Otsuka2023 because the goal of communication is to map those statistical objects back to real world entities.
-
-After all, the distinction between uncertainty and signal is largely determined by the goal of an analysis rather than any inherent properties of our data.
-
-This leads to several unanswered questions.
-Is our goal in uncertainty communication to acknowledge and convey the inherrent randomness that exists in all statistical calculations and models?
-Is it to supress signal that could be the result of random variance?
-Is it to find the best way to convey the mass function of a random variable?
-A survey of the literature will leave the answers to these questions unclear, however this inherrent issue has not gone unnoticed.
-
-- Should the visualisations change based on these goals?
-
-**need for task based understanding**
-- several review papers acknowledge this issue is causing 
-- **papers that talk about task specific stuff**
-- This problem of a task based philosiphy extends all the way back to philisophical uncerstandings of statistics and variance.
-- **notes from the Xiao-li Meng papers**
-- **notes from john paper**
-- While the statistics side brings up the purpose of a model or estimate for a philisophical consideration, it is a real, practical, and urgent problem for uncertainty visualisation. 
-- **go through papers with endless different goals**
-- **review paper that has an entire paragraph discussing differnet findings for box plots based on the task**
-
-**need for statistics links**
-- (no visual distinction between sample/pop variance) @Bella2005 found that most participants were ignorant to the fact that error bars are used for both confidence intervals and standard error bars, two wildly different indicators of precision.
-(**also the paper sherry sent me**)
-
-Uncertainty can only be defines within a specific problem.
-There is not a universal motivation agnostic "uncertainty" similarly as there is no motivation agnostic "signal".
-Therefore, in our following definitions of uncertainty, we will assume that the uncertainties we discuss are connected to some specific question.
+## 2.1 Motivating Questions
+Signal and noise can only be untangled in the presence of a motivating question.
+While it is seldom mentioned in the uncertainty visualisation literature, but is still important to understand that "uncertainty" does not exist without a motivating question.
+The entire process of data analysis, from deciding what should be observed as data through to communicating that data in a plot is governed by human decision and the goal of an analysis.
+At the philosophical level, applied statistics is simply taking real world entities and boiling them down into probabilistic objects, an ontological process that is largely dependent on our goals [@Otsuka2023].
+When we move onto data provenance the issue pursists, as what is kept as data and what is tossed away is determined by the motivation of an analysis and what was previously noise can be shown to become signal depending on the resolution of the question [@Meng2014].
+After moving onto modelling this issue continues as each research question can be can be categorised as descriptive, predictive, or causal, each of which has its own appropriate statistical methods and motivation agnostic model selection leads to signal devoid of meaning [@Carlin2023].
+Even at the final stages of visualisation a lack of understanding of the motivating question make it difficult to untangle what is signal and what is noise, leaving many uncertainty visualisation studies with conflicting results [@Kinkeldey2014].
+These cases highlight that the signal that is being conveyed and the noise that hampens that signal cannot be untangled and therefore communicated without a predetermined motivating question.
+Therefore any discussion of uncertainty must be done within the context of a motivating question as it is exactly what makes the distinction between signal and noise.
 
 ## 2.2 Taxonomies
+Within the context of a motivating uncertainty can be classified and 
 ### 2.2.1 Taxonomies of Uncertainty
+
 More commonly uncertainty is defined using a taxonomy rather than a strict definition.
@@
 
+*Notes from other section*
+- The usually uncertainty is only considered at the final stage but it needs to be considered through all stages.
+The big data paradox shrinks confidence intervals but magnifies bias [@Bradley2021] because the confidence intervals shrink due to the sample size but the imprecise target group creates bias.
+- @Meng2021 highlights that data needs to be treated as an output from another process rahter than an value neutral input.
+- If the distinction between noise and uncertainty is largely task dependent, considerations of quantifying and expressing uncertainty should be considered for the entire duraction of a project; another concept that has been independently identified by several authors [@Kinkeldey2014; @Refsgaard2007]
+- The most encompassing definition of uncertainty I have seen comes from @utypo who define uncertainty as **"any deviation from the unachievable ideal of completely deterministic knowledge of the relevant system"**.
+This definition encapsulates many concepts adjacent to randomness such as predictions, probability mass distributions (PMF), estimate error, and any data that is not a set of deterministic outcomes.
+This is the definition of uncertainty I will use for the rest of this report.
+- While quantification of the uncertainty that arrises from statistical philosophy and data provenance are beyond the scope of this paper, how this uncertainty is communicated is not.
+The ontological process discussed by @Otsuka2023 where information is tossed as we boil our information down into statistical objects requires us to map our statistical objects back to real world entities.
+This means the noise introduced and signal lost in earlier stages of the data analysis need to be identified, quantified and expressed in terms of real world outcomes.
+Visualisation is often the final stage of larger data science project and the down stream effects of poor abstraction and problem at the lower levels become impossible to ignore [@Munzner2009].
+
 ### 2.2.1 Taxonomies of Visual Uncertainty
@@
 
-# 4 Uncertainty Visualisation Best Practices
+# 4 The current landscape of uncertainty visualisation research
 In this section we will first go over the current state of uncertainty visualisation research, which is currently a collection of evaluation studies and review papers, and discuss what it is doing well and what could be done better.
 Following that, we will provide general reccomendations from the literature.
-
-# 4.1 The current landscape of uncertainty visualisation research
+# 4.1 The landscape
 In section 2.1 it was mentioned that uncertainty must be defined within a specific motivating question, otherwise it inherently does not make sense.
@@
 
-# 4.2 Take aways from the literature
+Notes from previous sections that were moved here
+- The concept of uncertainty being task dependent is *particularly* salient for uncertainty visualisation, and it is repeatedly identified as a problem in previous reviews of the uncertainty visualisation literature [@Kinkeldey2014; @Hullman2016] as well as across many sub domains and applications [@Wallsten1997; @Munzner2009; @Fischhoff2014; @Meng2021; @Amar2005].
+The fact that this conclusion is repeatedly reached shows both the importance and the lack of acknowledgement this concept receives.
+- @Fischhoff2014 discusses how decisions with different levels of flexibility should have different uncertainty visualisations.
+If the user has no choice they should simply be made aware of the risk, if they have several options, the user needs to see the uncertainty in each, if they need to come up with the options themselves, they need to see how things work and how the system fails.
+- @Amar2005 suggested a taxonomy for information visualisation based on the types of tasks we use visualisations for and suggest 10 "analytical primitives" that we can then map to visualisations.
+- Given that uncertainty estimation and communication seem to be largely task depenent, what questions should we be testing?
+This leads to several unanswered questions.
+Is our goal in uncertainty communication to acknowledge and convey the inherrent randomness that exists in all statistical calculations and models?
+Is it to supress signal that could be the result of random variance?
+Is it to find the best way to convey the mass function of a random variable?
+A survey of the literature will leave the answers to these questions unclear, however this inherrent issue has not gone unnoticed.
+- Statistics is, at its core, the study of uncertainty.
+Therefore discussing uncertainty visualisation a a separate sub domain to "normal" data visualisation is inherently confusing.
+What is typically meant by "uncertainty" visualisation is "noise", that is, we want to present some signal cushioned by its natural variance.
+Unfortunately, this distinction between "signal" and "noise" is entirely goal dependent.
+- @Bella2005 found that most participants were ignorant to the fact that error bars are used for both confidence intervals and standard error bars, two wildly different indicators of precision.
+(**also the paper sherry sent me**)
+# 4.2 Key points from the literature
+
+# 5 Great Examples
+- best things you have seen and why they are the best
+- can't have a best practices because you need experimental results and my point is that the experiments aren't testing the right things
 

### 2024-01-04 writingdumpster.qmd
finished context section and moved bits that didnt fit to other sections

@@
 Knowing that I can come back to this at any point and pull information back, makes me more comfortable with removing it from the report.
+
+# context section outtakes
+This conceptual way of thinking about uncertainty has been rattling around in the different areas of data science for a while.
+For example, when we record coin flips we typically want to model the behavior with a binomial distribution, so we ignore the outcome of a coin landing on its side despite the fact it is a real world possibility.
+@Wallsten1997 argue that the best method for evaluating or combining subjective probabilities depends on the uncertainty the decision maker wants to represent and why it matters.
+@Munzner2009 created a nested model for visualisation that highlighted how pitfalls in upstream levels, such as problem characterisation and data abstraction, can cause downstream effects and damage the effectivness of a visualisation.
+@Fischhoff2014 looks at uncertainty visualisation for decision making decides that we should have different ways of communicating uncertainty based off what the user is supposed to do with it.
 

### 2024-01-07 paper.qmd
added in notes from some overview papers

@@
 ---
+
+- Can discuss each uncertainty in isolation using distributions, ranges, and lists of alternatives, but the total cumulative uncertainty from every stage is more difficult (near impossible) @Spiegelhalter2017
+- "cannot assess the quality of risk communication unless the objectives are clear" circles back to motivation. He assumes we are fulfilling a duty to inform. @Spiegelhalter2017
+- the Affect heuristic comes to play in how people understand risk @Spiegelhalter2017
+- avoids the topic of risk at times of accute crisis because numerical risk would be inappropriate @Spiegelhalter2017
+- positive vs negative framing matters, can try to avoid it @Spiegelhalter2017
+- no preference between percent or frequency although (small interaction with numeracy skills though) @Spiegelhalter2017
+- Need careful specification of the reference class (e.g. % chance of rain should be fractions of situations like this in which is rains - over both time and location) @Spiegelhalter2017
+- Need to make time interval clear (50% each year) @Spiegelhalter2017
+- understanding better in a table than text @Spiegelhalter2017
+- if expressing as a frequency, note that larger numerators suggest a larger risk, exteme version of this is ignoring the denominator all together @Spiegelhalter2017
+- Graphical features that improve accuracy, features that facilitate behavioural change, and features that viewers like may be different (are not necessarily the same). @Spiegelhalter2017
+- also should consider graphical literacy @Spiegelhalter2017
+- graphs are good for gists and different graphics are good at different things (i.e. trends vs comparison etc) but hard to avoid framing (show information as part of a whole) @Spiegelhalter2017
+- scatting in pictograph increases the impression of randomness but makes it difficult to count @Spiegelhalter2017
+- people with high numeracy count pictographs but low numeracy assess area @Spiegelhalter2017
+- simplifying a multui-outcome array to two outcomes improved understanding @Spiegelhalter2017
+- using words to communicate numerical risk is generally discouraged and using words without numerical interpretation can lead to lower accuracy and higher risk aversion @Spiegelhalter2017
+- heath had a wide range of different metrics they adopted to communicate chronic risk of adverse events in the future @Spiegelhalter2017
+- hard to communicate small risks because people struggle to discrimintate between 1 in 10 000 and 1 in 100 000.
+Common solution is to use a logarithmic scale (if drawn vertically these are known as risk ladders).
+design of scale important because risk is associated with position rather than magnitude @Spiegelhalter2017
+- compares a bunch of stuff to the chance of getting struck by lightning since they are on a common scale (which is kind of funny) @Spiegelhalter2017
+- hierarchy for level of precision for epistemic uncertainty: 1) numbers given to appropriate levels of precision 2) a distribution range 3) a measure of statistical significance 4) verbal quantifiers to numbers 5) refusal to give a number unless the evidence is good enough @Spiegelhalter2017
+- Approaches to communicating limited understanding include 1) explicit model uncertainty 2) Qualitative scales expressed as strength of evidence (e.g.
+high to low certainty of evidence) 3) Acknowledge limitations and their possible impact 4) Acknowledged ignorance @Spiegelhalter2017
+- He then went through and explained how uncertainty is communicated across a wide range of areas @Spiegelhalter2017
+- Hurricane risk is not only the path, but also the storm surge and wind speed. @Spiegelhalter2017
+- coloured maps for earthquake risk are easily interpretable @Spiegelhalter2017
+- low-probability high-impact events are particularly difficult to communicate risks about @Spiegelhalter2017
+- industrial accidents have a "tollerable" risk chart based on risk and number of fatalities @Spiegelhalter2017
+
+- General issues when communicating risks: clear objectives, segment audience into target groups, use plain language and limit info to only what is necessary, allow for different levels of interest knowledge and numeracy (a gist level, numerical level, then evidence and uncertainty) @Spiegelhalter2017
+- Numerical risks: Use absolute risks (but relative risk when catastrophic event), for single unique event use percent chance, express change as a proportion, frequency, or percentage (and be clear about reference class), avoid framing bias by showing risk with and without the outcome, keep the denominator fixed when making comparisons, explicit about time interval, For more knowledgeable audiences consider providing epistemic uncertainty and confidence in analysis @Spiegelhalter2017
 
@@
 This mishmash of terminology leads to a large body of work, all claiming to finding the best visualisation or expression of of "uncertainty" but most don't even seem to agree on what uncertainty is.
-
-## 2.1 Motivating Questions
-Signal and noise can only be untangled in the presence of a motivating question.
-While it is seldom mentioned in the uncertainty visualisation literature, but is still important to understand that "uncertainty" does not exist without a motivating question.
-The entire process of data analysis, from deciding what should be observed as data through to communicating that data in a plot is governed by human decision and the goal of an analysis.
-At the philosophical level, applied statistics is simply taking real world entities and boiling them down into probabilistic objects, an ontological process that is largely dependent on our goals [@Otsuka2023].
-When we move onto data provenance the issue pursists, as what is kept as data and what is tossed away is determined by the motivation of an analysis and what was previously noise can be shown to become signal depending on the resolution of the question [@Meng2014].
-After moving onto modelling this issue continues as each research question can be can be categorised as descriptive, predictive, or causal, each of which has its own appropriate statistical methods and motivation agnostic model selection leads to signal devoid of meaning [@Carlin2023].
-Even at the final stages of visualisation a lack of understanding of the motivating question make it difficult to untangle what is signal and what is noise, leaving many uncertainty visualisation studies with conflicting results [@Kinkeldey2014].
-These cases highlight that the signal that is being conveyed and the noise that hampens that signal cannot be untangled and therefore communicated without a predetermined motivating question.
-Therefore any discussion of uncertainty must be done within the context of a motivating question as it is exactly what makes the distinction between signal and noise.
+- mentions uncertainty can be variance, precision, accuracy, reliability, or related concepts @Hullman2016
 
 ## 2.2 Taxonomies
-Within the context of a motivating uncertainty can be classified and 
+Papers that mention uncertainty often use a single sentance definition that suits the purpose of the paper, while papers that seek to define uncertainty use a taxonomy. While
+
+Most definitions of uncertainty are done with a taxonomy. Most papers that provide a definition of uncert Within the context of a motivating uncertainty can be classified and
+
 ### 2.2.1 Taxonomies of Uncertainty
@@
 ### 2.2.1 Taxonomies of Visual Uncertainty
+  - Most research had focused on developing taxonomies of uncertainty and developing methods to depict uncertainty visually @Kinkeldey2014
+- 5 dichotomous categories of uncertainty can be identified (1) explicit/implicit (directly mapping or showing multiple outcomes) (2) intrinsic/extrinsic (using existing symbols e.g.
+colour value, or new objects e.g.
+grids) (3) visually integral/separable (can or cannot be separated from the data) (4) coincidence/adjacent (if data and uncertainty are in integrated or separate views (5) static/dynamic (animation/interaction) @Kinkeldey2014
+- Based on those they only use (4), (2), and (5), left out (1) because most visualisations are explicit, and (3) corresponds to (1) in most cases. @Kinkeldey2014
+- They sugest that future typologies should take different categoeis of tasks into account (1) communication tasks (2) analytical tasks (3) exploratory takss.
+List task-centred typology as a main topic for future research in the field to facilitate practical guidelines in this field @Kinkeldey2014
 
+- @Spiegelhalter2017 Establishes that risk is known probabilities and uncertainty is estimated probabilities (i.e.
+more real world).
+Suggest uncertainty is what we have when you cannot accurately quantify risk
+- Highlights the lack of a clear definition of uncertainty "it depends on who you ask" @Spiegelhalter2017
+- ambiguity is also poorly defined, as it can be basically the same as uncertainty (uncertain probabilities) or contested values @Spiegelhalter2017
+- A lot of taxonomies for uncertainty, simple is "known knowns, known unknowns, and unknown unknowns", ka (known, unknown, nad unknowable), (risk, uncertainty, ignorance).
+This also breaks down into the framework of aleatory uncertainty (unavoidable randomness), epistemic uncertatiny (uncertainty about structure, still quantified), ontological uncertainty (uncertain about entire modelling process).
+@Spiegelhalter2017
 # 3.General Attitudes...
+I dont think this is worth a section but it is mentioned a lot in .
+- the role of unertainty visualisation is to communicate accurate information
+- problem of communicating to a group thing (avoid communicating inaccurate information)
+- how people feel about uncertainty is irrelevant
+  - natural disasters make people uncomfortable but to include it in a conversation about whether or not we should express it is rediculous.
+
 ## 3.1 Towards uncertainty in general
@@
 
-# 4 The current landscape of uncertainty visualisation research
+# 4 LITERATURE SUMMARY
+## intro
 In this section we will first go over the current state of uncertainty visualisation research, which is currently a collection of evaluation studies and review papers, and discuss what it is doing well and what could be done better.
 Following that, we will provide general reccomendations from the literature.
-# 4.1 The landscape
+
+- methods used in uncertainty visualisation evaluation remain ad hoc.
+studies approach issue from a usability perspective instead of asking WHY representations DO or DO NOT work.
+Studies do not follow any methodology commonly agreed upon.
+@Kinkeldey2014
+- Lack of formalisation and eigour in empirical methdos is an issue that is much broader than uncertainty visualisation (extends to info-vis and related domains).
+Therefore this paper is a starting point, not a definitive summary @Kinkeldey2014
+- The question in these papers is often "does method a work better than method B" which is an engineering approach, where the toal is to improve a particular product rather than create general principals @Kinkeldey2014
+- Authors then try to generalise beyond the specific constraints of the test even though the framing does not allow for it @Kinkeldey2014
+- Alternatively, we should ask how/why does method A work better than method B, which grounds work in perceptual and cognative theory and provides a framwork @Kinkeldey2014
+
+# 4.?? Noise vs Signal (question +)
+The first issue that arises with uncertainty research is that uncertainty is frequently treated as a signal, not noise.
+This discrepancy is also noted by @Kinkeldey2014 who, in their own literature review on uncertainty methods, comment that the literature seems unsure if uncertainty is simply another variable or if it should be treated differently.
+This is a fair question to ask.
+If uncertainty is simply a signal we are trying to convey, why should it be treated any different to any other signal in a visualisation?
+
+- The literature also does not discuss if uncertainty is another variable, or if it needs to be treated differently (i,.e.
+is it metadata of another variable).
+Most papers studies on uncertainty dont consider this issue, and the question remains, should someone have two separate values, or an integrated view uncertain data view @Kinkeldey2014
+
+This is best seen with a simple look at some papers
+(Make into table)
+Questions where uncertainty is a signal
+
+Questions where uncertainty is a noise
+
+## 4.?? Greater Infoviz Issues
+- Since there are so many papers that treat uncertainty as a signal, not a noise, issues in the uncertainty visualisation literature also help to understand some difficulties in the information visualisation literature as large.
+
+- Suggests that it is straightforward to show a value but much more complex to show uncertainty @Hullman2016 (I think this is interesting in relation to how people SEE uncertainty visualisation.
+Why is uncertainty hard to visualise?)
+- Authors provide little justification for their chosen response models (e.g. absolute accuracy vs relative measures) @Hullman2016
+- When asking questions, researchers should focus on events that can be repeated (because of a misunderstanding of confidence intervals) @Hullman2016
+- Uncertainty defined as a by product of the process of making a visualisation, from data collection to commucating @Hullman2016
+- Some questions (e.g. what are the chances that the no.6 bus will arrive first) you can elicit the viewers subjective probability distribution @Hullman2016
+# 4.?? Tasks and Motivating Questions
+## 4.1?? In other statistics
+What is meant by "uncertainty" may seem obvious to some, but when you attempt to quantify or visualise it you will quickly find yourself asking, "uncertainty about...
+what?".
+Do you mean uncertainty on an estimate?
+On a forecast?
+How many steps ahead is this forecast?
+Are we only considering the uncertainty in the estimate or in the parameters or are we considering the possibility of measurement error or biased inputs?
+Signal and noise can only be untangled in the presence of a motivating question.
+
+The idea that uncertainty can only be defined in the presence of a motivating question is well grounded in most areas of statistics.
+The entire process of data analysis, from deciding what should be observed as data through to communicating that data in a plot is governed by human decision and the goal of an analysis.
+At the philosophical level, applied statistics is simply taking real world entities and boiling them down into probabilistic objects, an ontological process that is largely dependent on our goals [@Otsuka2023].
+When we move onto data provenance the issue persists, as what is kept as data and what is tossed away is determined by the motivation of an analysis and what was previously noise can be shown to become signal depending on the resolution of the question [@Meng2014].
+After moving onto modelling this issue continues as each research question can be can be categorised as descriptive, predictive, or causal, each of which has its own appropriate statistical methods and motivation agnostic model selection leads to signal devoid of meaning [@Carlin2023].
+Even at the final stages of visualisation a lack of understanding of the motivating question make it difficult to untangle what is signal and what is noise, leaving many uncertainty visualisation studies with conflicting results [@Kinkeldey2014].
+These cases highlight that uncertainty is defined at *every* stage in relation to our motivating question, from data collection to visualisation.
+Discussions of uncertainty cannot be had if we are not clear *what* we are uncertain about.
+Once it is established what we are uncertainty about, we can consider the other elements of uncertainty that need to be defined.
+
+## 4.1?? General task notes
+- Mentioned alleatory vs epistemic uncertainty, in the visualisation literature alleatory uncertainty is more commonly the focus.
+also mentions dependence and joint pdfs as a consideration @Hullman2016
+- Most studies involve intrinsic approaches such as colour, transparency, and those that used extrinsic techniques typically used glyphs, error bars, grid based techniques, or contouring @Kinkeldey2014
+- most studies used coincident approaches, static visualisations, are applied to a specific domain and dont have generalised results @Kinkeldey2014
+- The tasks for participants were most commonly value retrival (some where uncertainty and data value are retrived separately, others where they are some separately).
+also also be asked to extract a relative value @Kinkeldey2014
+- How does the complexity of uncertainty relate to the cateogies of user and task @Kinkeldey2014
+- Several studies probide evidence that the usability of uncertainty representations can be highly user and task dependent @Kinkeldey2014
+- The nature of the task plays an important role for the usability of uncertainty techniques, this may explain many of the insonsitent outcomes from the studies under review since two studies assessing sifferent techniques and user groups are not comparable @Kinkeldey2014
+- Doing a follow up publication to discuss issues with reasoning and decision making based on uncertainty visualisations which is not considered or discussed in this paper @Kinkeldey2014
+- Main take away is that "we need to systematize future empirical studies on uncertainty visualisation to better enable comparison and generalisation of the findings"
+- One way to advance this goal is a taxonomy, however existing taxonomies are focused on data types, uncertainty categories, and representation types @Kinkeldey2014
+- he lists a bunch of tasks that were tested with uncertainty visualisationsat the end @Kinkeldey2014
+
+- risk is uncertainty as a signal, and when uncertainty communication is spoken about this is typically what is discussed, uncertainty as signal supression is what is usually meant when we talk about visualisations but that area is functionally ignored.
+This idea, that uncertainty can only be defined in the face of a motivating question, is well grounded in the literature but seldom explicitly mentioned
+
 In section 2.1 it was mentioned that uncertainty must be defined within a specific motivating question, otherwise it inherently does not make sense.
@@
 
+Uncertainty as signal papers have incredibly predictable results
+Uncertainty as noise papers should follow a similar protocol to the line up papers
+
 Notes from previous sections that were moved here
@@
 - @Amar2005 suggested a taxonomy for information visualisation based on the types of tasks we use visualisations for and suggest 10 "analytical primitives" that we can then map to visualisations.
-- Given that uncertainty estimation and communication seem to be largely task depenent, what questions should we be testing?
-This leads to several unanswered questions.
-Is our goal in uncertainty communication to acknowledge and convey the inherrent randomness that exists in all statistical calculations and models?
-Is it to supress signal that could be the result of random variance?
-Is it to find the best way to convey the mass function of a random variable?
-A survey of the literature will leave the answers to these questions unclear, however this inherrent issue has not gone unnoticed.
 - Statistics is, at its core, the study of uncertainty.
@@
 (**also the paper sherry sent me**)
-# 4.2 Key points from the literature
 
+- Need to compare to an approripate ground truth, but this is a philisophical exercse @Hullman2016
+## 4.?? Incomparable plots
+- But when different techniques are to be compared, it is important that the scenarios and datasets are informationally equivalent, i.e.
+according to Larkin and Simon (1987, p.
+67) ‘[t]wo representations are informationally equivalent if all of the information in the one is also inferable from the other, and vice versa’.
+A goal in testing, then, is often to determine whether they are also computationally equivalent, or whether one depiction has an advantage over another, i.e.
+‘[t]wo representations are computationally equivalent if they are informationally equivalent and, in addition, any inference that can be drawn easily and quickly from the information given explicitly in the one can also be drawn easily and quickly from the information given explicitly in the other, and vice versa’ (Larkin and Simon 1987, p.
+67).
+@Kinkeldey2014
+
+- The way we ask questions and the types of questions we ask are selected with little justification. This paper makes suggestions to reduce the noise in the data from these papers. @Hullman2016
+
+- Most commonly check abolute measures of accuracy, where accuracy is abs(subjective error - actual probabiltiy).
+Some papers also check relative measures of accuracy and ask subjects to find regions of least certainty or to rank targets by uncertainty.
+studies also consider response time and confidence rating.
+some studies also ask users for their own expression of uncertainty (e.g.
+by asking when sketches are uncertainty and ranking various depictions of uncertainty).
+Some studies ask subjects to make some decision using the data.
+ometimes things like complexity of task, degree of visual overload, ease of use, visual appeal, and preferences @Hullman2016
+# 4.?? Source of Uncertainty
+- rarely used extension is aggregation of uncertainty over an area (retrieve overall estimation from a spatial distribution of uncertainty).
+Other tasks include comparisons, rankings.
+also have a task called "search" where participants have to identify entities that fulfill certainty characteristics (high or low values) @Kinkeldey2014
+- Whether to use classed or unclassed uncertainty is rarely discussed @Kinkeldey2014.
+e.g.
+land map with remotely sensed images has ambiguity in the boundaries, which can result from multiple sources (land cover class, measurement error, images from different dates).
+If you combine them all it might not make sense for someone who only needs one.
+
+## 4.?? Importance (and unimportance) of Meta information
+- Specifies uncertainty can be represented by three components attribute (what) position (where) and temporal (when) @Kinkeldey2014
+- Studies dealing with uncertainty information typically deal with attribute but rarely position and time @Kinkeldey2014
+- 10/44 studies involve multiple types of uncertainty @Kinkeldey2014
+
+- When we boil down data into a statistical object, we boil away many aspects on context. Birth and coin flips become an object that is one in the same, and often the visualisation 
+- Why are spatial and temporal data considered special?
+What is the statistical element that remains after they have been away?
+Inexchangeability.
+Temporal data is one dimensional inexchangeability, spatial data has two dimensional inexchangeability.
+  - There are other things that have inexchangeability, and not every question related to spatial or temporal data requires the inexchangeability.
+  - 
+- Yes the data type is relevant but its not the most important thing
+  - one with temporal, one with that temporal paper but you didn't know it was temporal, one with the bus time cdf
+  - screenshot from these three papers, the connecting factor is the question, not the data type
+  - The data type: spatial, temporal, etc, usually is mentioned as the 
+  - Not a feature of the data, but it is a feature of the question "how do I show uncertainty on a map?".
+This is a fine question to ask, but the lack of specificity in what is being done, and the zoning in on functionally irrelevant elements leads the literature as a huge mess.
+
+# 4.?? Required Changes and Considerations to the Experimental Space 
+- If you just keep using rankings, you need to compare everything one by one forever to get an idea of the space @Hullman2016
+- people sometimes ignore uncertainty and instead look at the pattern (e.g.
+distance between means) and use a heuristic to answer the question.
+be proactive about possible heuristics, look for signes of heuristics in responses, ask subjects to describe their strategy, consider including degree of heuristic as a dependent variable @Hullman2016
+- People believe in the law of small numbers but dont believe in the lar of large numbers @Hullman2016
+- experiencing an event makes your percieved probability higher than if you see a description of it @Hullman2016
+- Blur and fog are suggested because they are metaphors for uncertainty, however the usefulness of these metaphors is rarely investigated @Kinkeldey2014
+- Frequency thing means that more effective communication to participants in a study means more effective science and we have a roll on effect @Hullman2016
+- Probability judgements are less relative than other psychophysical judgements, 0%, 50% and 100% all have well understood meanings.
+Subjective probability distributions are spontaneous, and subjects are constructing a distribution upon being asked rather than articulating a fully formed mental representation of their beliefs.
+Therefore it is subject to heuristics.
+consider anchoring a scale with farmiliar probabilities .
+not sure if being asked for the entire distribution helps, e.g.
+the graphical one reduces noise but asking people for an interval estimation leads to overprecision and overconfidence @Hullman2016
+
+- If the goal is to compare visualisation, ground truth is easy, if the goal is to evaluate how accurate the subjective probability distributions are, the ground truth becomes complex.
+ground truth issue e.g.
+if you provide a sample from a distribution, is the true value the mean of that sample or the population mean?
+@Hullman2016
+- believed that offering enough decisions with varying rewards you can observe the true subjective probability, however whether or not probability is the right representation of subjective uncertainty is a topic of debate @Hullman2016
+- she also discusses graphical interface (users have to draw the probability distribution) and framing probabilities as natural frequencies are better choices because they reduce noise and stuff.@Hullman2016
+- Consider incorperating utility functions. Include "probability-coherence" checks @Hullman2016
+- evidence that people do not follow the axioms of probability when conveying subjective probabilities. How do you deal with this as an analyst (e.g. if the probabilities do not sum to 1 @Hullman2016
+- If someone answers a question correctly it is hard to tell if it was because of the information provided or if the person was using a heuristic @Hullman2016
+- You cant tell if subjects in an experiment do poorly because the visualisation was bad or because the question was misinterpreted @Hullman2016
+
+- Visualisations risk: consider a good summary table as a visualisation, use multiple formats because no single representation suits all members of an audience, illuminate graphics with words and numbers, design graphics to allow a part-to-whole comparison on an appropriate scale, narrative labels are important (show magnitude through tick marks), use narrative, images and metaphors to gain/retain attention but dont arouse undue emotion, assume low numeracy of a general public audience and adopt a less-is-more approach by reducing the need for inderence and making clear and explicit comparison, be cautious about interactivity and animations (may introduce uncessary complexity), avoid chart junk (like 3d bar charts), assess the needs of the audience, and iterate towards a final design.
 # 5 Great Examples
@@
 - can't have a best practices because you need experimental results and my point is that the experiments aren't testing the right things
+- VSUP, the map interactive map that susuan showed me, the pixel map
 

### 2024-01-07 writingdumpster.qmd
added in notes from some overview papers

@@
 # context section outtakes
+Given that uncertainty estimation and communication seem to be largely task depenent, what questions should we be testing?
+This leads to several unanswered questions.
+Is our goal in uncertainty communication to acknowledge and convey the inherrent randomness that exists in all statistical calculations and models?
+Is it to supress signal that could be the result of random variance?
+Is it to find the best way to convey the mass function of a random variable?
+A survey of the literature will leave the answers to these questions unclear, however this inherrent issue has not gone unnoticed.
 This conceptual way of thinking about uncertainty has been rattling around in the different areas of data science for a while.

### 2024-01-08 paper.qmd
sorted notes from review paper

@@
 ---
-
-- Can discuss each uncertainty in isolation using distributions, ranges, and lists of alternatives, but the total cumulative uncertainty from every stage is more difficult (near impossible) @Spiegelhalter2017
-- "cannot assess the quality of risk communication unless the objectives are clear" circles back to motivation. He assumes we are fulfilling a duty to inform. @Spiegelhalter2017
-- the Affect heuristic comes to play in how people understand risk @Spiegelhalter2017
-- avoids the topic of risk at times of accute crisis because numerical risk would be inappropriate @Spiegelhalter2017
-- positive vs negative framing matters, can try to avoid it @Spiegelhalter2017
-- no preference between percent or frequency although (small interaction with numeracy skills though) @Spiegelhalter2017
-- Need careful specification of the reference class (e.g. % chance of rain should be fractions of situations like this in which is rains - over both time and location) @Spiegelhalter2017
-- Need to make time interval clear (50% each year) @Spiegelhalter2017
-- understanding better in a table than text @Spiegelhalter2017
-- if expressing as a frequency, note that larger numerators suggest a larger risk, exteme version of this is ignoring the denominator all together @Spiegelhalter2017
-- Graphical features that improve accuracy, features that facilitate behavioural change, and features that viewers like may be different (are not necessarily the same). @Spiegelhalter2017
-- also should consider graphical literacy @Spiegelhalter2017
-- graphs are good for gists and different graphics are good at different things (i.e. trends vs comparison etc) but hard to avoid framing (show information as part of a whole) @Spiegelhalter2017
-- scatting in pictograph increases the impression of randomness but makes it difficult to count @Spiegelhalter2017
-- people with high numeracy count pictographs but low numeracy assess area @Spiegelhalter2017
-- simplifying a multui-outcome array to two outcomes improved understanding @Spiegelhalter2017
-- using words to communicate numerical risk is generally discouraged and using words without numerical interpretation can lead to lower accuracy and higher risk aversion @Spiegelhalter2017
-- heath had a wide range of different metrics they adopted to communicate chronic risk of adverse events in the future @Spiegelhalter2017
-- hard to communicate small risks because people struggle to discrimintate between 1 in 10 000 and 1 in 100 000.
-Common solution is to use a logarithmic scale (if drawn vertically these are known as risk ladders).
-design of scale important because risk is associated with position rather than magnitude @Spiegelhalter2017
-- compares a bunch of stuff to the chance of getting struck by lightning since they are on a common scale (which is kind of funny) @Spiegelhalter2017
-- hierarchy for level of precision for epistemic uncertainty: 1) numbers given to appropriate levels of precision 2) a distribution range 3) a measure of statistical significance 4) verbal quantifiers to numbers 5) refusal to give a number unless the evidence is good enough @Spiegelhalter2017
-- Approaches to communicating limited understanding include 1) explicit model uncertainty 2) Qualitative scales expressed as strength of evidence (e.g.
-high to low certainty of evidence) 3) Acknowledge limitations and their possible impact 4) Acknowledged ignorance @Spiegelhalter2017
-- He then went through and explained how uncertainty is communicated across a wide range of areas @Spiegelhalter2017
-- Hurricane risk is not only the path, but also the storm surge and wind speed. @Spiegelhalter2017
-- coloured maps for earthquake risk are easily interpretable @Spiegelhalter2017
-- low-probability high-impact events are particularly difficult to communicate risks about @Spiegelhalter2017
-- industrial accidents have a "tollerable" risk chart based on risk and number of fatalities @Spiegelhalter2017
-
-- General issues when communicating risks: clear objectives, segment audience into target groups, use plain language and limit info to only what is necessary, allow for different levels of interest knowledge and numeracy (a gist level, numerical level, then evidence and uncertainty) @Spiegelhalter2017
-- Numerical risks: Use absolute risks (but relative risk when catastrophic event), for single unique event use percent chance, express change as a proportion, frequency, or percentage (and be clear about reference class), avoid framing bias by showing risk with and without the outcome, keep the denominator fixed when making comparisons, explicit about time interval, For more knowledgeable audiences consider providing epistemic uncertainty and confidence in analysis @Spiegelhalter2017
 
@@
 
+- industrial accidents have a "tollerable" risk chart based on risk and number of fatalities @Spiegelhalter2017
 # 2. Defining Uncertainty
+## 2.1 Coloquially
+## 2.2 Through Taxonomies
 Uncertainty visualisation is not made any easier by the fact that the term "uncertainty" lacks a commonly accepted definition in the literature.
@@
 
-## 2.2 Taxonomies
 Papers that mention uncertainty often use a single sentance definition that suits the purpose of the paper, while papers that seek to define uncertainty use a taxonomy. While
-
-Most definitions of uncertainty are done with a taxonomy. Most papers that provide a definition of uncert Within the context of a motivating uncertainty can be classified and
-
-### 2.2.1 Taxonomies of Uncertainty
 
@@
 Visualisation is often the final stage of larger data science project and the down stream effects of poor abstraction and problem at the lower levels become impossible to ignore [@Munzner2009].
-
-### 2.2.1 Taxonomies of Visual Uncertainty
-  - Most research had focused on developing taxonomies of uncertainty and developing methods to depict uncertainty visually @Kinkeldey2014
-- 5 dichotomous categories of uncertainty can be identified (1) explicit/implicit (directly mapping or showing multiple outcomes) (2) intrinsic/extrinsic (using existing symbols e.g.
-colour value, or new objects e.g.
-grids) (3) visually integral/separable (can or cannot be separated from the data) (4) coincidence/adjacent (if data and uncertainty are in integrated or separate views (5) static/dynamic (animation/interaction) @Kinkeldey2014
-- Based on those they only use (4), (2), and (5), left out (1) because most visualisations are explicit, and (3) corresponds to (1) in most cases. @Kinkeldey2014
-- They sugest that future typologies should take different categoeis of tasks into account (1) communication tasks (2) analytical tasks (3) exploratory takss.
-List task-centred typology as a main topic for future research in the field to facilitate practical guidelines in this field @Kinkeldey2014
-
+- Approaches to communicating limited understanding include 1) explicit model uncertainty 2) Qualitative scales expressed as strength of evidence (e.g.
+high to low certainty of evidence) 3) Acknowledge limitations and their possible impact 4) Acknowledged ignorance @Spiegelhalter2017
+- hierarchy for level of precision for epistemic uncertainty: 1) numbers given to appropriate levels of precision 2) a distribution range 3) a measure of statistical significance 4) verbal quantifiers to numbers 5) refusal to give a number unless the evidence is good enough @Spiegelhalter2017
 - @Spiegelhalter2017 Establishes that risk is known probabilities and uncertainty is estimated probabilities (i.e.
@@
 @Spiegelhalter2017
-# 3.General Attitudes...
+- Can discuss each uncertainty in isolation using distributions, ranges, and lists of alternatives, but the total cumulative uncertainty from every stage is more difficult (near impossible) @Spiegelhalter2017
+# 3.General Attitude Towards Uncertainty + Its communication
 I dont think this is worth a section but it is mentioned a lot in .
@@
   - natural disasters make people uncomfortable but to include it in a conversation about whether or not we should express it is rediculous.
+- Numerical risks: Use absolute risks (but relative risk when catastrophic event), for single unique event use percent chance, express change as a proportion, frequency, or percentage (and be clear about reference class), avoid framing bias by showing risk with and without the outcome, keep the denominator fixed when making comparisons, explicit about time interval, For more knowledgeable audiences consider providing epistemic uncertainty and confidence in analysis @Spiegelhalter2017
 
-## 3.1 Towards uncertainty in general
-## 3.2 Towards uncertainty visualisation
+- General issues when communicating risks: clear objectives, segment audience into target groups, use plain language and limit info to only what is necessary, allow for different levels of interest knowledge and numeracy (a gist level, numerical level, then evidence and uncertainty) @Spiegelhalter2017
+
+- compares a bunch of stuff to the chance of getting struck by lightning since they are on a common scale (which is kind of funny) @Spiegelhalter2017
+- hard to communicate small risks because people struggle to discrimintate between 1 in 10 000 and 1 in 100 000.
+Common solution is to use a logarithmic scale (if drawn vertically these are known as risk ladders).
+design of scale important because risk is associated with position rather than magnitude @Spiegelhalter2017
+- heath had a wide range of different metrics they adopted to communicate chronic risk of adverse events in the future @Spiegelhalter2017
+- using words to communicate numerical risk is generally discouraged and using words without numerical interpretation can lead to lower accuracy and higher risk aversion @Spiegelhalter2017
+- the Affect heuristic comes to play in how people understand risk @Spiegelhalter2017
+- avoids the topic of risk at times of accute crisis because numerical risk would be inappropriate @Spiegelhalter2017
+- positive vs negative framing matters, can try to avoid it @Spiegelhalter2017
+- no preference between percent or frequency although (small interaction with numeracy skills though) @Spiegelhalter2017
+- Need careful specification of the reference class (e.g. % chance of rain should be fractions of situations like this in which is rains - over both time and location) @Spiegelhalter2017
+- Need to make time interval clear (50% each year) @Spiegelhalter2017
+- understanding better in a table than text @Spiegelhalter2017
+- if expressing as a frequency, note that larger numerators suggest a larger risk, exteme version of this is ignoring the denominator all together @Spiegelhalter2017
 
 # 4 LITERATURE SUMMARY
+# 4. Similar work (other literature reviews)
+- If you just keep using rankings, you need to compare everything one by one forever to get an idea of the space @Hullman2016
+- people sometimes ignore uncertainty and instead look at the pattern (e.g.
+distance between means) and use a heuristic to answer the question.
+be proactive about possible heuristics, look for signes of heuristics in responses, ask subjects to describe their strategy, consider including degree of heuristic as a dependent variable @Hullman2016
+- People believe in the law of small numbers but dont believe in the lar of large numbers @Hullman2016
+- experiencing an event makes your percieved probability higher than if you see a description of it @Hullman2016
+- Blur and fog are suggested because they are metaphors for uncertainty, however the usefulness of these metaphors is rarely investigated @Kinkeldey2014
+- Frequency thing means that more effective communication to participants in a study means more effective science and we have a roll on effect @Hullman2016
+- Probability judgements are less relative than other psychophysical judgements, 0%, 50% and 100% all have well understood meanings.
+Subjective probability distributions are spontaneous, and subjects are constructing a distribution upon being asked rather than articulating a fully formed mental representation of their beliefs.
+Therefore it is subject to heuristics.
+consider anchoring a scale with farmiliar probabilities .
+not sure if being asked for the entire distribution helps, e.g.
+the graphical one reduces noise but asking people for an interval estimation leads to overprecision and overconfidence @Hullman2016
+- If the goal is to compare visualisation, ground truth is easy, if the goal is to evaluate how accurate the subjective probability distributions are, the ground truth becomes complex.
+ground truth issue e.g.
+if you provide a sample from a distribution, is the true value the mean of that sample or the population mean?
+@Hullman2016
+- believed that offering enough decisions with varying rewards you can observe the true subjective probability, however whether or not probability is the right representation of subjective uncertainty is a topic of debate @Hullman2016
+- she also discusses graphical interface (users have to draw the probability distribution) and framing probabilities as natural frequencies are better choices because they reduce noise and stuff.@Hullman2016
+- Consider incorperating utility functions. Include "probability-coherence" checks @Hullman2016
+- evidence that people do not follow the axioms of probability when conveying subjective probabilities. How do you deal with this as an analyst (e.g. if the probabilities do not sum to 1 @Hullman2016
+- If someone answers a question correctly it is hard to tell if it was because of the information provided or if the person was using a heuristic @Hullman2016
+- You cant tell if subjects in an experiment do poorly because the visualisation was bad or because the question was misinterpreted @Hullman2016
+- Visualisations risk: consider a good summary table as a visualisation, use multiple formats because no single representation suits all members of an audience, illuminate graphics with words and numbers, design graphics to allow a part-to-whole comparison on an appropriate scale, narrative labels are important (show magnitude through tick marks), use narrative, images and metaphors to gain/retain attention but dont arouse undue emotion, assume low numeracy of a general public audience and adopt a less-is-more approach by reducing the need for inderence and making clear and explicit comparison, be cautious about interactivity and animations (may introduce uncessary complexity), avoid chart junk (like 3d bar charts), assess the needs of the audience, and iterate towards a final design.
+@Spiegelhalter2017
+- Graphical features that improve accuracy, features that facilitate behavioural change, and features that viewers like may be different (are not necessarily the same). @Spiegelhalter2017
+- also should consider graphical literacy @Spiegelhalter2017
+- graphs are good for gists and different graphics are good at different things (i.e. trends vs comparison etc) but hard to avoid framing (show information as part of a whole) @Spiegelhalter2017
+- scatting in pictograph increases the impression of randomness but makes it difficult to count @Spiegelhalter2017
+- people with high numeracy count pictographs but low numeracy assess area @Spiegelhalter2017
+# 4.?? Taxonomies of Visual Uncertainty
+  - Most research had focused on developing taxonomies of uncertainty and developing methods to depict uncertainty visually @Kinkeldey2014
+- 5 dichotomous categories of uncertainty can be identified (1) explicit/implicit (directly mapping or showing multiple outcomes) (2) intrinsic/extrinsic (using existing symbols e.g.
+colour value, or new objects e.g.
+grids) (3) visually integral/separable (can or cannot be separated from the data) (4) coincidence/adjacent (if data and uncertainty are in integrated or separate views (5) static/dynamic (animation/interaction) @Kinkeldey2014
+- Based on those they only use (4), (2), and (5), left out (1) because most visualisations are explicit, and (3) corresponds to (1) in most cases. @Kinkeldey2014
+- They sugest that future typologies should take different categoeis of tasks into account (1) communication tasks (2) analytical tasks (3) exploratory takss.
+List task-centred typology as a main topic for future research in the field to facilitate practical guidelines in this field @Kinkeldey2014
 ## intro
@@
 - Alternatively, we should ask how/why does method A work better than method B, which grounds work in perceptual and cognative theory and provides a framwork @Kinkeldey2014
+
+- low-probability high-impact events are particularly difficult to communicate risks about @Spiegelhalter2017
 
@@
 - Some questions (e.g. what are the chances that the no.6 bus will arrive first) you can elicit the viewers subjective probability distribution @Hullman2016
+- communication tasks: This category comprises map reading tasks involving data and uncertainty value retrieval (which location is most uncertain?).
+For tasks from this category, visualisation techniques can be chosen following the traditional rules from cartography.
+@Kinkeldey2014 (basically saying why dont you just use what is already here if you are treating uncertainty as a signal, which I agree with)
 # 4.?? Tasks and Motivating Questions
+- "cannot assess the quality of risk communication unless the objectives are clear" circles back to motivation. He assumes we are fulfilling a duty to inform. @Spiegelhalter2017
 ## 4.1?? In other statistics
@@
 
-## 4.?? Importance (and unimportance) of Meta information
+# 4.?? Importance (and unimportance) of Secondary and Meta information
 - Specifies uncertainty can be represented by three components attribute (what) position (where) and temporal (when) @Kinkeldey2014
@@
 
-# 4.?? Required Changes and Considerations to the Experimental Space 
-- If you just keep using rankings, you need to compare everything one by one forever to get an idea of the space @Hullman2016
-- people sometimes ignore uncertainty and instead look at the pattern (e.g.
-distance between means) and use a heuristic to answer the question.
-be proactive about possible heuristics, look for signes of heuristics in responses, ask subjects to describe their strategy, consider including degree of heuristic as a dependent variable @Hullman2016
-- People believe in the law of small numbers but dont believe in the lar of large numbers @Hullman2016
-- experiencing an event makes your percieved probability higher than if you see a description of it @Hullman2016
-- Blur and fog are suggested because they are metaphors for uncertainty, however the usefulness of these metaphors is rarely investigated @Kinkeldey2014
-- Frequency thing means that more effective communication to participants in a study means more effective science and we have a roll on effect @Hullman2016
-- Probability judgements are less relative than other psychophysical judgements, 0%, 50% and 100% all have well understood meanings.
-Subjective probability distributions are spontaneous, and subjects are constructing a distribution upon being asked rather than articulating a fully formed mental representation of their beliefs.
-Therefore it is subject to heuristics.
-consider anchoring a scale with farmiliar probabilities .
-not sure if being asked for the entire distribution helps, e.g.
-the graphical one reduces noise but asking people for an interval estimation leads to overprecision and overconfidence @Hullman2016
-
-- If the goal is to compare visualisation, ground truth is easy, if the goal is to evaluate how accurate the subjective probability distributions are, the ground truth becomes complex.
-ground truth issue e.g.
-if you provide a sample from a distribution, is the true value the mean of that sample or the population mean?
-@Hullman2016
-- believed that offering enough decisions with varying rewards you can observe the true subjective probability, however whether or not probability is the right representation of subjective uncertainty is a topic of debate @Hullman2016
-- she also discusses graphical interface (users have to draw the probability distribution) and framing probabilities as natural frequencies are better choices because they reduce noise and stuff.@Hullman2016
-- Consider incorperating utility functions. Include "probability-coherence" checks @Hullman2016
-- evidence that people do not follow the axioms of probability when conveying subjective probabilities. How do you deal with this as an analyst (e.g. if the probabilities do not sum to 1 @Hullman2016
-- If someone answers a question correctly it is hard to tell if it was because of the information provided or if the person was using a heuristic @Hullman2016
-- You cant tell if subjects in an experiment do poorly because the visualisation was bad or because the question was misinterpreted @Hullman2016
-
-- Visualisations risk: consider a good summary table as a visualisation, use multiple formats because no single representation suits all members of an audience, illuminate graphics with words and numbers, design graphics to allow a part-to-whole comparison on an appropriate scale, narrative labels are important (show magnitude through tick marks), use narrative, images and metaphors to gain/retain attention but dont arouse undue emotion, assume low numeracy of a general public audience and adopt a less-is-more approach by reducing the need for inderence and making clear and explicit comparison, be cautious about interactivity and animations (may introduce uncessary complexity), avoid chart junk (like 3d bar charts), assess the needs of the audience, and iterate towards a final design.
+- Hurricane risk is not only the path, but also the storm surge and wind speed. @Spiegelhalter2017
+- coloured maps for earthquake risk are easily interpretable @Spiegelhalter2017
 # 5 Great Examples

### 2024-01-09 paper.qmd
added experiment paper checklist

@@
 # 4 LITERATURE SUMMARY
+# Experiment paper checklist
+General
+- Specific Field of Research
+- Question asked
+- Risk or Uncertainty: 
+- Noise vs signal (risk vs uncertainty)
+- Major task goal
+    - e.g. identify a bias or misunderstanding, compare two visualisation methods
+- Minor task goal
+    - Communication(value retrieval)
+    - Analytical(complex consideration)
+    - Exploration(free for all)
+
+Uncertainty Considerations
+	- Source
+	- Level (statistical-ignorance)
+	- Nature (Epistemic/Aleatory)
+
+Visual Uncertainty taxonomy
+- Explicit/implicit
+    - (directly mapping or showing multiple outcomes)
+- Intrinsic/Extrinsic 
+    - (using existing symbols e.g. colour value, or new objects e.g. grids)
+- Visually integral/separable 
+    - (can or cannot be separated from the data) 
+- Coincidence/adjacent 
+    - (if data and uncertainty are in integrated or separate views
+- Static/dynamic 
+    - (animation/interaction)
+
+Other plot considerations
+	- Dimension of data
+	- Dimension of uncertainty
+	- Feature mappings
+
+Extra (not necessarily recorded)
+	- Metrics used (and recorded)
+	- Possible Heuristics
+	- What is the ground truth?
+	- Participant literacy
+
 # 4. Similar work (other literature reviews)

### 2024-01-09 paper.qmd
added some comments

@@
 @Kinkeldey2014 (basically saying why dont you just use what is already here if you are treating uncertainty as a signal, which I agree with)
+- papers usually have same variance if there is a signal question, and use the mean as a proxy to answer the question.
+Really should ask a changing signal question where the signal IS impacted by the uncertainty information.
+- Weirdly When setting their ground truth it seems like a lot of papers EXPECT participants to completely ignore uncertainty information as noise.
+Like the participants are actually considered to be incorrect if they dampen their signal understanding because of noise considerations which I find FUNNY.
+- I think the questions where uncertainty is a signal, not a noise, allow the authors of visualisation experiments to ignore the context and motivation that usually governs a visualisation.
 # 4.?? Tasks and Motivating Questions

### 2024-01-16 paper.qmd
added notes

@@
 
+- if the experiment DOES use noise as a signal, it is often very unclear how the participants should include that in their respose and what would be a valid response from participants (e.g.
+the likelihood to how much question and the cone of uncertainty questions).
+separate to the data, uncertainty has its own goal.
+Is it to impact confidence in the estimate (like a bayesian thing)?
+to supress signal (like a barely noticiable difference line up plot thing)?
+to facilitate more efficient decisions (so accounting for risk thing)?
+
+- will it impact their decision at all?
+there are a couple of papers where you can answer all the questions by just ignoring the uncertainty information because of the way they have set it up.
+e.g.
+difference in means when the variance is the same, experiments that check if someone can still get the signal and make sure the uncertainty information isnt in the way.
+
 - The literature also does not discuss if uncertainty is another variable, or if it needs to be treated differently (i,.e.
@@
 
+- I think the dimension of the uncertainty should be considered every time there is aggregation from multiple sources of uncertainty and you want to maintain those sources.
+You can project the uncertainty down into one dimension if the source information is in of itself unimportant.
 # 4.?? Importance (and unimportance) of Secondary and Meta information

### 2024-01-23 paper.qmd
started doing summary of general ideas (no fine details)

@@
 # 1. Introduction
+From entertainment choices to news articles to insurance plans, the modern citizen is so over run with information in every aspect of their life it can be overwhelming.
+So overwhelming in fact, that it is not uncommon for people to become so overwhelmed by the information they refuse to engage with it at all.
+In this overflow of information, tools that can effectively summarise information down into simple and clear ideas become more valuable.
+While websites that summarise insurance plans or hotel options down into a clear table make a living off it, few tools are as powerful or effective as a visualisation.
+Information visualisations remain one of the most powerful tools for fast and reliable science communication.
+Effectively leveraging visualisations to make them more effective is therefore key in facilitating simple and clear communication of science and other ideas.
+
 Think back to the last time you made some sort of data visualisation.
@@
 - industrial accidents have a "tollerable" risk chart based on risk and number of fatalities @Spiegelhalter2017
+
 # 2. Defining Uncertainty
-## 2.1 Coloquially
-## 2.2 Through Taxonomies
 Uncertainty visualisation is not made any easier by the fact that the term "uncertainty" lacks a commonly accepted definition in the literature.
-@Lipshitz1997 even commented that “there are almost as many definitions of uncertainty as there are treatments of the subject”.
+@Spiegelhalter2017 even commented that the definition of uncertainty essentially "depends on who you ask".
 This mishmash of terminology leads to a large body of work, all claiming to finding the best visualisation or expression of of "uncertainty" but most don't even seem to agree on what uncertainty is.
-- mentions uncertainty can be variance, precision, accuracy, reliability, or related concepts @Hullman2016
+In this section we will establish exactly what we mean when we talk about uncertainty.
 
-Papers that mention uncertainty often use a single sentance definition that suits the purpose of the paper, while papers that seek to define uncertainty use a taxonomy. While
+## 2.1 Coloquial Definitions
+The most common definition of uncertainty used in papers that discuss visualising it are colloquial.
+Most definitions make some nod towards variance, precision, trust, error, missing values or another related concepts.
+Since *uncertainty*, unlike *probability*, does not have a strict definition that is established in mathematics, there is large variety in its usage.
+Some consider it to be synonymous with probabiltiy, some consider it to be an encompassing umbrella term of which probability is only an element, and some consider it to by something else entirely.
+This lack of a definition leads to uncertainty being swapped out freely with other vague words such as ambiguity and risk.
+Definitions presented in papers that discuss uncertainty visualisation want it to be encompassing and include everything a layperson might consider when trying to consider "uncertainty".
+The issue with these broad definitions is is that in order to visualize something, it needs to be quantified, and in order to quantify something, we need to know exactly what it is we are counting.
+This discrepancy appears regularly in the uncertainty visualisation literature.
+Often a vague definition such as "uncertainty is anything that isn't deterministic", or even a refusal to define uncertainty at all in an experiments introduction will be accompanied by a highly specific quantification of uncertainty as a probability density distribution of a forecast when providing the visualisation.
+Ironically it seems that many of the papers that refer to themselves as "uncertainty" papers are just suffering from a form of uncertainty, a lack of precision.[^1]
 
-More commonly uncertainty is defined using a taxonomy rather than a strict definition.
+This lack of precision has caused several pernicious issues in the literature, including different representation of uncertainty being compared despite not being realistic substitutes (with no acknowledgement of this from the authors); a lack of understanding as to *why* uncertainty should be included in a visualisation; a large swath of literature focused on a minute aspect of the entire uncertainty space with little attention to creative solutions to real world visualisation problems; and a swath of newly designed visualisations that are disconnected from the problems that they hope to solve.
+These problems means the literature either needs to transition away from the broad umbrella term "uncertainty", or consider the full range of sources of uncertainty and how they can be communicated when designing an experiment.[^2]
+
+[^1]: Flesh out note - I think several of the sentences in this paragraph need examples (or quatified evidence) from the spreadsheet definition part
+[^2]: I am not sure if I should mention this here because I go into depth in the literature review and these comments might come across as unjustified if they are left in this section.
+Maybe this is more of a conclusion paragraph.
+
+## 2.2 Taxonomy Definitions
+While colloquial definitions of uncertainty are useful when trying to communicate with laypeople who don't understand strict statistical definitions, these definitions are not helpful when we are trying to work out exactly what should be shown on a plot.
+When authors try to make a considerable effort to define what we mean when we talk about uncertainty, they often express it in the form of a taxonomy.
+
+When reading any literature about uncertainty or an adjacent topic you will be overrun with more ways to organise uncertainty than you will know what to do with.
+The problem is so pervasive that it is almost impossible for two randomly selected papers that both claim to be about uncertainty will use the same definition.
+
+Some taxonomies contain overlapping terminology but also have alternative definitions and categories.
+For example *___* organises uncertainty into the two categories, Aleatroy (inherrent randomness) and epistemic (lack of knowledge), but *___* organises uncertainty into the three categories aleatory (inherrent randomness in a forecast) vs epistemic (uncertainty in structure and parameters of statistical models) vs ontological (uncertainty about the entire modelling process as a description of reality).
+Other taxonomies disucuss the degree of measurability
+
+Other taxonomies conflate features of the data with features of the uncertainty:
+- measurement (uncertainty on a single varible), spatial (uncertainty in spatial data), temporal (uncertainty in a time data)
+
+there are even taxonomies
+
+What is worse is that the reason for using one taxonomy over another, or why a new taxonomy was established in the first place is rarely discussed.
+This leaves new authors in the field with a wide array of choices and no information that would allow them to determine which is most effective for them.
+The complete mess of this literature makes it more enticing for authors to just define their own definition of uncertainty, which only adds to this problem.
+This is not done in ignorance of the issue either, many papers that comment on the unending list of uncertainty definitions will still provide a new taxonomy of their own design.
+I would almost find this situation comical if I were not one of the people forced to sort through this mess.
+
+There are a handful of problems that is created by this lack of consensus. 
+The first issues is that 
+Which leaves the definition of aleatory and epistemic uncertainty confused.
+
+ Below is a list of ways some papers have organised uncertainty:
+
+- scientific methods (variance, distribution) vs human judgments (disagreement among parties)
+- Random vs systematic uncertainty
+- statistical (can provide a continuous pdf) vs bounded (know values will fall in a certain range) uncertainty
+
+- accuracy vs precision
+- Uncertainty from discovery, assessment and execution
+- risk vs uncertainty
+- Source of uncertainty (model structure, parameters, forcing functions, intial state, model operation) and diagnostic (past/current) vs prognostic (forecasting)
+- uncertainty is organised according to Satience (legal moral, societal, institutional, proprietary, situational) and level of severity (high, medium, low)
+
+There are several issues that become apparent when looking at this list. First of all, 
+Looking at these taxonomies there are three key features of interest that become apparent. People seem to be interested in the:
+
+1) What stage of the analyis or what group is causing the uncertainty (the location)
+2) How well the uncertainty can be quantified (the level), and
+3) Whether or not the uncertainty is believed to be caused by natural variance or a lack of information (Nature)
+
+ .
+Additionally, since many papers will go on to define uncertainty as a probability or a density in their method, the actual organisation of "uncertainty" is irrelevant to their results since a probability distribution is
+
+or where it exists in the broader scope of uncertainty communication.
+If a taxonomy is presented it is rarely considered when discussing differences in calculating, conceptualising,  The definition is also rarely dicussed beyond the introudction of a paper.
+to its place in the broader scope of of these definitions are presented with little justification as to why
+
+Type A evaluation of standard uncertainty, and the other which is based on scientific judgement (such as experience and specifications), called the Type B evaluation of standard uncertainty.
+
+It can be seen that while these taxonomies are both attempting to organise similar aspects of uncertainty, they have little overlap.
+The reason for this is relatively simple.
+Uncertainty is often only defined in so far as it is needed.
+
+Ultimately these taxonomies end up alienating those that actually work with uncertainty by being too narrow.
+Uncertainty is a broad and complicated topic and while oversimplified taxonomies might help us hold the concept of uncertainty in our head, they do little to help with the practical question of how to quantify, visualise, or communicate uncertainty.
+
+- mention that utypo taxonomy contains all these taxonomies and can be easily built upon without making a new taxonomy.
 
@@
 
+This way of conceptualising and organising uncertainty is not a fringe idea and is backed up by several related ideas in statistics. 
+- also makes it easier for people to communicate exactly what they need visualisation tools for. Look at geography papers and how their specific uncertainty considerations have been ignored for years.
+- can organise other "uncertainties" according to this framework
+- it also gives us rules for combining uncertainty since statistical uncertainty can be combined with staitstical uncertainty, and outcome uncertainty can be combined with outcome uncertainty.
+Additionally, this definition of uncertainty is helpful in understanding *how* uncertainty should be communicated and the consideration we should make.
+
 *Notes from other section*
@@
 high to low certainty of evidence) 3) Acknowledge limitations and their possible impact 4) Acknowledged ignorance @Spiegelhalter2017
+
 - hierarchy for level of precision for epistemic uncertainty: 1) numbers given to appropriate levels of precision 2) a distribution range 3) a measure of statistical significance 4) verbal quantifiers to numbers 5) refusal to give a number unless the evidence is good enough @Spiegelhalter2017
-- @Spiegelhalter2017 Establishes that risk is known probabilities and uncertainty is estimated probabilities (i.e.
-more real world).
-Suggest uncertainty is what we have when you cannot accurately quantify risk
-- Highlights the lack of a clear definition of uncertainty "it depends on who you ask" @Spiegelhalter2017
-- ambiguity is also poorly defined, as it can be basically the same as uncertainty (uncertain probabilities) or contested values @Spiegelhalter2017
+
 - A lot of taxonomies for uncertainty, simple is "known knowns, known unknowns, and unknown unknowns", ka (known, unknown, nad unknowable), (risk, uncertainty, ignorance).
@@
 - can't have a best practices because you need experimental results and my point is that the experiments aren't testing the right things
-- VSUP, the map interactive map that susuan showed me, the pixel map
+- VSUP
+- the map interactive map that susuan showed me
+- the pixel map
+- The new york times wealthy people that end up in different economics classes, white vs black people
+- the animated surface classifications from that satelite images experimental paper
+- the climate change scenario uncertainty
 

### 2024-01-23 writingdumpster.qmd
started doing summary of general ideas (no fine details)

@@
 Knowing that I can come back to this at any point and pull information back, makes me more comfortable with removing it from the report.
+# Definition outtakes
+## Random
+- mentions uncertainty can be variance, precision, accuracy, reliability, or related concepts @Hullman2016
+-  @Spiegelhalter2017 defines risk as risk known probabilities and uncertainty as estimated probabilities, therefore uncertainty is what we have when you cannot accurately quantify risk.
 
+## Taxonomies
+To avoid alienating lay people with confusing complexity, many of these taxonomies are overly simplistic.
+Most taxonomies seek to organise uncertainty based off where the uncertainty is coming from.
+For example, @Spiegelhalter2017 organises uncertainty into three categories:
+  - Aleatory uncertainty: inevitable unpredictability due to unforeseeable factors 
+  - Epistemic uncertainty: uncertainty about the structure and parameters of statistical models
+  - Ontological uncertainty: uncertainty about the entire modelling process as a description of reality.
+These three categories can be considered to be uncertainty when deciding which model we should use (ontological), uncertainty surrounding parameter estimates (epistemic), and uncertainty surrounding a forecast (aleatory), however it is not hard to find another, very different, example of a taxonomy that organises uncertainty based on what is seen to be causing it.
+@Gustafson2019 organises uncertainty into the four groups:
+  - Deficient: Lack of knowledge
+  – Technical: modelling approximations and measurement error
+  – Scientific Unknown unknowns
+  – Consensus: Disagreement among parties
+However the additional comments about the inherent variability of aleatory uncertainty results in a bit of confusion as to if our cateogirisation should change if we decide to treat the parameters of a statistical model to be inherently random (as is done in Bayesian statistics) or if we beleive a forecast variability to be caused by a lack of knowledge.
+
+He kept the taxonomy to three levels to maintain simplicity, however this organisation methods conflates amount of uncertainty present in an estimate with what is causing that uncertainty, and in doing so makes it hard for us to use this definition to better understand uncertainty within our own analysis.
 # context section outtakes

### 2024-01-24 paper.qmd
taxonomy summary draft edits

@@
 The problem is so pervasive that it is almost impossible for two randomly selected papers that both claim to be about uncertainty will use the same definition.
-
-Some taxonomies contain overlapping terminology but also have alternative definitions and categories.
-For example *___* organises uncertainty into the two categories, Aleatroy (inherrent randomness) and epistemic (lack of knowledge), but *___* organises uncertainty into the three categories aleatory (inherrent randomness in a forecast) vs epistemic (uncertainty in structure and parameters of statistical models) vs ontological (uncertainty about the entire modelling process as a description of reality).
-Other taxonomies disucuss the degree of measurability
-
-Other taxonomies conflate features of the data with features of the uncertainty:
-- measurement (uncertainty on a single varible), spatial (uncertainty in spatial data), temporal (uncertainty in a time data)
-
-there are even taxonomies
-
 What is worse is that the reason for using one taxonomy over another, or why a new taxonomy was established in the first place is rarely discussed.
@@
 
-There are a handful of problems that is created by this lack of consensus. 
-The first issues is that 
-Which leaves the definition of aleatory and epistemic uncertainty confused.
-
- Below is a list of ways some papers have organised uncertainty:
-
+Examples of Taxonomies used to untanlge the features of uncertainty:
+- Aleatroy (inherrent randomness) and epistemic (lack of knowledge)
+- measurement (uncertainty on a single varible), spatial (uncertainty in spatial data), temporal (uncertainty in a time data)
+- aleatory (inherrent randomness in a forecast) vs epistemic (uncertainty in structure and parameters of statistical models) vs ontological (uncertainty about the entire modelling process as a description of reality)
 - scientific methods (variance, distribution) vs human judgments (disagreement among parties)
@@
 - statistical (can provide a continuous pdf) vs bounded (know values will fall in a certain range) uncertainty
-
 - accuracy vs precision
@@
 
-There are several issues that become apparent when looking at this list. First of all, 
+This taxonomy issue creates several problems in the literature.
+Ideally, if we wanted to combine the literature on uncertainty visualisation, each paper would state the same definition of uncertainty and explain which aspect of uncertainty they are visualising, so the experiments would shift around the space while the definition of uncertainty and what it contains would remains stationary.
+This would make it easy for the typical reader to use the literature to find an appropriate visualisation for the particular category or case of uncertainty.
+Instead, one has to comb through the method of a particular design to understand if the uncertainty you are working with is amenable to this method.
+Another issue this creates is that the "space" of uncertainty visualisation is unclear.
+This means that several areas of uncertainty visualisation that have a real need for creative visualisation tools are left untouched, while the visualisations that are considered "uncertainty visualisations" simply because they are an expression of a probability mass function dominate the literature.
+Finally the taxonomies do little to help data analysts or statisticians untangle, estimate and understand the uncertainty that may be in their own projects.
+The endless set of constantly changing definitions implies their *is no* set meaning to the term "uncertainty" and something that cannot be defined certainly cannot be quantified for communication or estimation.
+This leads people to be less likely to visualise uncertainty due to its imprecision and difficulty to commincate.
+
+This does not mean there is nothing to be learned from these taxonomies, each in their own way show what apsect of uncertainty is important to certain cases.
+
 Looking at these taxonomies there are three key features of interest that become apparent. People seem to be interested in the:

### 2024-01-24 writingdumpster.qmd
taxonomy summary draft edits

@@
 ## Taxonomies
+Some taxonomies contain overlapping terminology but also have alternative definitions and categories.
+For example *___* organises uncertainty into the two categories, Aleatroy (inherrent randomness) and epistemic (lack of knowledge), but *___* organises uncertainty into the three categories aleatory (inherrent randomness in a forecast) vs epistemic (uncertainty in structure and parameters of statistical models) vs ontological (uncertainty about the entire modelling process as a description of reality).
+
 To avoid alienating lay people with confusing complexity, many of these taxonomies are overly simplistic.

### 2024-01-25 paper.qmd
updated taxonomy visualisation and uncertainty sections and added the mathematical framework bit

@@
 @Spiegelhalter2017 even commented that the definition of uncertainty essentially "depends on who you ask".
+The disarray of organisation and definitions within the uncertainty literature is a common sentiment expressed by reviewers[@Spiegelhalter2017; @Kinkeldey2014].
 This mishmash of terminology leads to a large body of work, all claiming to finding the best visualisation or expression of of "uncertainty" but most don't even seem to agree on what uncertainty is.
@@
 
+The most encompassing definition of uncertainty I have seen comes from @utypo who define uncertainty as **"any deviation from the unachievable ideal of completely deterministic knowledge of the relevant system"**.
+This definition encapsulates many concepts adjacent to randomness such as predictions, probability mass distributions (PMF), estimate error, and any data that is not a set of deterministic outcomes.
+This is the definition of uncertainty I will use for the rest of this report.
+
 [^1]: Flesh out note - I think several of the sentences in this paragraph need examples (or quatified evidence) from the spreadsheet definition part
@@
 ## 2.2 Taxonomy Definitions
+# 2.2.1 Taxonomies of Uncertainty
 While colloquial definitions of uncertainty are useful when trying to communicate with laypeople who don't understand strict statistical definitions, these definitions are not helpful when we are trying to work out exactly what should be shown on a plot.
@@
 Examples of Taxonomies used to untanlge the features of uncertainty:
-- Aleatroy (inherrent randomness) and epistemic (lack of knowledge)
+- aleatory uncertainty (unavoidable randomness), epistemic uncertatiny (uncertainty about structure, still quantified), ontological uncertainty (uncertain about entire modelling process).
+@Spiegelhalter2017
 - measurement (uncertainty on a single varible), spatial (uncertainty in spatial data), temporal (uncertainty in a time data)
@@
 - risk vs uncertainty
+- "known knowns, known unknowns, and unknown unknowns", (risk, uncertainty, ignorance)
 - Source of uncertainty (model structure, parameters, forcing functions, intial state, model operation) and diagnostic (past/current) vs prognostic (forecasting)
 - uncertainty is organised according to Satience (legal moral, societal, institutional, proprietary, situational) and level of severity (high, medium, low)
+- Level of precision in epistemic uncertainty: 1) numbers given to appropriate levels of precision 2) a distribution range 3) a measure of statistical significance 4) verbal quantifiers to numbers 5) refusal to give a number unless the evidence is good enough @Spiegelhalter2017
 
@@
 This does not mean there is nothing to be learned from these taxonomies, each in their own way show what apsect of uncertainty is important to certain cases.
-
-Looking at these taxonomies there are three key features of interest that become apparent. People seem to be interested in the:
-
-1) What stage of the analyis or what group is causing the uncertainty (the location)
-2) How well the uncertainty can be quantified (the level), and
-3) Whether or not the uncertainty is believed to be caused by natural variance or a lack of information (Nature)
-
- .
-Additionally, since many papers will go on to define uncertainty as a probability or a density in their method, the actual organisation of "uncertainty" is irrelevant to their results since a probability distribution is
-
-or where it exists in the broader scope of uncertainty communication.
-If a taxonomy is presented it is rarely considered when discussing differences in calculating, conceptualising,  The definition is also rarely dicussed beyond the introudction of a paper.
-to its place in the broader scope of of these definitions are presented with little justification as to why
-
-Type A evaluation of standard uncertainty, and the other which is based on scientific judgement (such as experience and specifications), called the Type B evaluation of standard uncertainty.
-
-It can be seen that while these taxonomies are both attempting to organise similar aspects of uncertainty, they have little overlap.
-The reason for this is relatively simple.
-Uncertainty is often only defined in so far as it is needed.
-
-Ultimately these taxonomies end up alienating those that actually work with uncertainty by being too narrow.
-Uncertainty is a broad and complicated topic and while oversimplified taxonomies might help us hold the concept of uncertainty in our head, they do little to help with the practical question of how to quantify, visualise, or communicate uncertainty.
-
-- mention that utypo taxonomy contains all these taxonomies and can be easily built upon without making a new taxonomy.
+There are a handful of overlapping features that indicate which elements of uncertainty are of interest to statistics and related applied fields.
+@utypo identified these similarities and used them to design a taxonomy that encapsulates a large proportion of the ad hoc definitions that currently overrun the literature, and is the definition of uncertainty that we will work with for the remainder of this paper.
 
@@
 First, we need to consider the source of the uncertainty.
-Is this uncertainty coming from inaccurate measurements or a poorly defined model?
+Is this uncertainty coming from the data (for example, from an inaccurate measurement), from the array of model choices we have, or from the assumptions on a parameter?
 This is the *location* of the uncertainty.
@@
 Finally, consider how this uncertainty came into existence.
-Is it a result of a naturally random process (epistemic) or is it due to imperfect information and could be improved (aleatory).
+Is it a result of a naturally random process (aleatory) or is it due to imperfect information and could be improved (epistemic).
 This is the *nature* of your uncertainty.
-@utypo then goes on to describe mapping our uncertainty in a 3D space that is defined by its location, level, and nature, but I think the taxonomy is more easily understood as a series of questions we need to consider when we are trying to quantify uncertainty.
+@utypo then goes on to describe mapping our uncertainty in a 3D space that is defined by its location, level, and nature, but I think the taxonomy is more easily understood as a series of questions we should to consider when we are trying to quantify uncertainty.
+Location asks "where and at what stage in the modelling process is this uncertainty coming from", level asks "how well can we quantify this uncertainty", and nature asks "can we reduce this uncertainty with improves knowledge?".
+While this taxonomy specifically applies to the uncertainty in the model, it is easy to fit most of the uncertainty we would consider quantifying into this format.
 
@@
 
-This way of conceptualising and organising uncertainty is not a fringe idea and is backed up by several related ideas in statistics. 
+*The taxonomy* we will use in this paper may seem complicated and difficult to grasp at first glance, but it essentially boils down to three dimensions to consider.Uncertainty is a broad and complicated topic and while oversimplified taxonomies might help us hold the concept of uncertainty in our head, they do little to help with the practical question of how to quantify, visualise, or communicate uncertainty.
 - also makes it easier for people to communicate exactly what they need visualisation tools for. Look at geography papers and how their specific uncertainty considerations have been ignored for years.
+- Additionally, this definition of uncertainty is helpful in understanding *how* uncertainty should be communicated and the consideration we should make.
 - can organise other "uncertainties" according to this framework
-- it also gives us rules for combining uncertainty since statistical uncertainty can be combined with staitstical uncertainty, and outcome uncertainty can be combined with outcome uncertainty.
-Additionally, this definition of uncertainty is helpful in understanding *how* uncertainty should be communicated and the consideration we should make.
 
-*Notes from other section*
+*The location* axis of the taxonomy lines up neatly with the typical data analysis pipeline and its importance is backed up by a several philisophical perspectives on statistics as well as discussions on the practical needs of uncertainty visualisation.
 - The usually uncertainty is only considered at the final stage but it needs to be considered through all stages.
@@
 - If the distinction between noise and uncertainty is largely task dependent, considerations of quantifying and expressing uncertainty should be considered for the entire duraction of a project; another concept that has been independently identified by several authors [@Kinkeldey2014; @Refsgaard2007]
-- The most encompassing definition of uncertainty I have seen comes from @utypo who define uncertainty as **"any deviation from the unachievable ideal of completely deterministic knowledge of the relevant system"**.
-This definition encapsulates many concepts adjacent to randomness such as predictions, probability mass distributions (PMF), estimate error, and any data that is not a set of deterministic outcomes.
-This is the definition of uncertainty I will use for the rest of this report.
 - While quantification of the uncertainty that arrises from statistical philosophy and data provenance are beyond the scope of this paper, how this uncertainty is communicated is not.
@@
 Visualisation is often the final stage of larger data science project and the down stream effects of poor abstraction and problem at the lower levels become impossible to ignore [@Munzner2009].
+
+*The level* axis forces us to identify how quantifiable the uncertainty is. 
+- it also gives us rules for combining uncertainty since statistical uncertainty can be combined with staitstical uncertainty, and outcome uncertainty can be combined with outcome uncertainty. 
+- Can discuss each uncertainty in isolation using distributions, ranges, and lists of alternatives, but the total cumulative uncertainty from every stage is more difficult (near impossible) @Spiegelhalter2017
 - Approaches to communicating limited understanding include 1) explicit model uncertainty 2) Qualitative scales expressed as strength of evidence (e.g.
@@
 
-- hierarchy for level of precision for epistemic uncertainty: 1) numbers given to appropriate levels of precision 2) a distribution range 3) a measure of statistical significance 4) verbal quantifiers to numbers 5) refusal to give a number unless the evidence is good enough @Spiegelhalter2017
+*The nature axis* allows us to easily communicate if the uncertainty can be reduced or if it needs to be accepted.
+While this level if not of particular importance to the visualisation of uncertainty, that is not the only purpose of the taxonomy.
+The nature of the uncertainty is an important element in expressing uncertainty, especually to stakeholders, and connecting the nature to the location makes it easier to understand *how* uncertainty can be reduced.
 
-- A lot of taxonomies for uncertainty, simple is "known knowns, known unknowns, and unknown unknowns", ka (known, unknown, nad unknowable), (risk, uncertainty, ignorance).
-This also breaks down into the framework of aleatory uncertainty (unavoidable randomness), epistemic uncertatiny (uncertainty about structure, still quantified), ontological uncertainty (uncertain about entire modelling process).
-@Spiegelhalter2017
-- Can discuss each uncertainty in isolation using distributions, ranges, and lists of alternatives, but the total cumulative uncertainty from every stage is more difficult (near impossible) @Spiegelhalter2017
+### 2.2.2 Taxonomies of Visual Uncertainty
+There are some taxonomies of visual uncertainty.
+These taxonmies seem to seek to define "uncertainty visualisation" and a combined term rather than as two separate words.
+We largely disagree with this conceptualization of uncertainty visualisations as it blurs the line between the mathematical estimations and assumptions (the uncertainty part) and the visual depiction of those mathematical objects (the visualisation part).
+While we will not focus on these taxonomies, it is still important to mention them and highlight the reasons they are not good practice.
+
+There are a small handful of taxonomies (or typologies) that are specifically for uncertainty visualisations.
+@Kinkeldey2014 identified give categories of uncertainty based on the work of other literature reivews of uncertainty visualisation.
+The five categories are: (1) explicit/implicit (directly mapping or showing multiple outcomes) (2) intrinsic/extrinsic (using existing symbols e.g.
+colour value, or new objects e.g.
+grids) (3) visually integral/separable (can or cannot be separated from the data) (4) coincidence/adjacent (if data and uncertainty are in integrated or separate views (5) static/dynamic (animation/interaction).
+Of the groups they identified, they actually only used (4), (2), and (5), left out (1) because most visualisations are explicit, and (3) corresponds to (1) in most cases [@Kinkeldey2014].
+
+@Potter2012 organised several existing uncertainty visualisations into groups based on the dimensionality of the data (1D, 2D, 3D, and No Dimension) and the dimensionality of the uncertainty (Scalar, Vector, Tensor).
+However, because the term "PDF", a statistical object that describes a random variable that is typically a one dimensional function, is used to describe both the data and the uncertainty for all dimensions, it is hard to understand how this dimensionality should work.
+
+@Grewal2021 created a taxonomy that mapped uncertainty visualisations to some point in a 2D space that was defined by the axis "domain expertise" and "continum of discreteness" that scaled from "point estimate" to "continuous distribtion".
+
+@Thomson2005 titled their paper "A typology for visualizing uncertainty" however it is a taxonomy for the specific types of uncertainty that we may want to visualise and a way to calculate that uncertainty.
+While this paper is more about mathematics than visualisation, it is clearly written as a precursor to visualisation.
+
+These taxonomies are useful in the sense that they often identify areas where there is an over or under supply of uncertainty visualisations.
+Additionally taxonomies do highlight a concept of uncertainty dimensionality that was not directly covered in the uncertainty taxonomy, however the distinction between uncertainty as a single value versus as a PDF could be considered a sub-consideration of uncertainty that's level is statistical so it can be quantified.
+While uncertainty visualisation taxonomies are not the standard practice (generating a taxonomy of uncertainty and then developing methods to depict that uncertainty is) [@Kinkeldey2014], it is still important to acknowledge the existence of "uncertainty visualisation typologies" and why they may miss the mark in what is needed from the uncertainty visualisaiton literature.
+
+The first issue with these taxonomies is that most of these taxonomies are created by observing existing uncertainty visualisations.
+This means under researched gaps in the taxonomy could be due to a lack of need just as well as a lack of attention, and the benefit of trying to create visualisations that land in uncharted areas of the taxonomy are unclear.
+
+The second issue is that there does not seem to be anything these taxonomies offer that would not be better established by separate taxonomies for uncertainty and visualisation.
+These taxonomies seem to depict uncertainty visualisation as a field within itself, rather than a specific case of information visualisation that requires more attention to mathematics than usual.
+The difference between "uncertainty visualisations" and "data visualisations" is not technically in the visual element, it is mathematical.
+@Kinkeldey2014 almost acknowledges this in their own paper that discusses an uncertainty visualsiation taxonomy when they claim "future typologies should take different categories of tasks into account (1) communication tasks (2) analytical tasks (3) exploratory task", a common typology for information visualisation in general.
+The process of understanding and estimating uncertainty requires knowledge of the data, the statistical methods used to make an estimate, and the assumptions of a model.
+Visualising the statistics that represent uncertainty should be no different than depicting the statistics that represent any other element of a graphic, and therefore there seems to be no reason to have a separate taxonomy for uncertainty visualisation.
+
 # 3.General Attitude Towards Uncertainty + Its communication
@@
 
+- low-probability high-impact events are particularly difficult to communicate risks about @Spiegelhalter2017
 # 4 LITERATURE SUMMARY
-# Experiment paper checklist
-General
-- Specific Field of Research
-- Question asked
-- Risk or Uncertainty: 
-- Noise vs signal (risk vs uncertainty)
-- Major task goal
-    - e.g. identify a bias or misunderstanding, compare two visualisation methods
-- Minor task goal
-    - Communication(value retrieval)
-    - Analytical(complex consideration)
-    - Exploration(free for all)
-
-Uncertainty Considerations
-	- Source
-	- Level (statistical-ignorance)
-	- Nature (Epistemic/Aleatory)
-
-Visual Uncertainty taxonomy
-- Explicit/implicit
-    - (directly mapping or showing multiple outcomes)
-- Intrinsic/Extrinsic 
-    - (using existing symbols e.g. colour value, or new objects e.g. grids)
-- Visually integral/separable 
-    - (can or cannot be separated from the data) 
-- Coincidence/adjacent 
-    - (if data and uncertainty are in integrated or separate views
-- Static/dynamic 
-    - (animation/interaction)
-
-Other plot considerations
-	- Dimension of data
-	- Dimension of uncertainty
-	- Feature mappings
-
-Extra (not necessarily recorded)
-	- Metrics used (and recorded)
-	- Possible Heuristics
-	- What is the ground truth?
-	- Participant literacy
-
-# 4. Similar work (other literature reviews)
+- methods used in uncertainty visualisation evaluation remain ad hoc.
+studies approach issue from a usability perspective instead of asking WHY representations DO or DO NOT work.
+Studies do not follow any methodology commonly agreed upon.
+@Kinkeldey2014
+- Lack of formalisation and eigour in empirical methdos is an issue that is much broader than uncertainty visualisation (extends to info-vis and related domains).
+Therefore this paper is a starting point, not a definitive summary @Kinkeldey2014
+- The question in these papers is often "does method a work better than method B" which is an engineering approach, where the toal is to improve a particular product rather than create general principals @Kinkeldey2014
+- Authors then try to generalise beyond the specific constraints of the test even though the framing does not allow for it @Kinkeldey2014
+- Alternatively, we should ask how/why does method A work better than method B, which grounds work in perceptual and cognative theory and provides a framwork @Kinkeldey2014
+# 4.?? Similar work (other literature reviews)
 - If you just keep using rankings, you need to compare everything one by one forever to get an idea of the space @Hullman2016
@@
 - people with high numeracy count pictographs but low numeracy assess area @Spiegelhalter2017
-# 4.?? Taxonomies of Visual Uncertainty
-  - Most research had focused on developing taxonomies of uncertainty and developing methods to depict uncertainty visually @Kinkeldey2014
-- 5 dichotomous categories of uncertainty can be identified (1) explicit/implicit (directly mapping or showing multiple outcomes) (2) intrinsic/extrinsic (using existing symbols e.g.
-colour value, or new objects e.g.
-grids) (3) visually integral/separable (can or cannot be separated from the data) (4) coincidence/adjacent (if data and uncertainty are in integrated or separate views (5) static/dynamic (animation/interaction) @Kinkeldey2014
-- Based on those they only use (4), (2), and (5), left out (1) because most visualisations are explicit, and (3) corresponds to (1) in most cases. @Kinkeldey2014
-- They sugest that future typologies should take different categoeis of tasks into account (1) communication tasks (2) analytical tasks (3) exploratory takss.
-List task-centred typology as a main topic for future research in the field to facilitate practical guidelines in this field @Kinkeldey2014
-## intro
-In this section we will first go over the current state of uncertainty visualisation research, which is currently a collection of evaluation studies and review papers, and discuss what it is doing well and what could be done better.
-Following that, we will provide general reccomendations from the literature.
+# 4.1 Plot Comparison Papers
+# 4.1.1 The problem of incomparable plots
+Visual inference can be seen as a process that combines multiple fields, from mathematics to psychology, to convey meaning.
+Unfortunately which tasks belong to mathematics, computer science, or psychology is poorly defined.
+The uncertainty visualisation taxonomies make the blur between these fields.
+A visualisation is, in a lot of ways, a statistic, or at the very least statistics are calculated prior to the visualisation being calculated, however this may be invisible to the user.
+A density plot does not just "appear" from the data, usually a smoothing function (that should be calibrated according to the sample size however how this should be done is a different matter of dispute) generates an estimated density function (statistics) which is then depicted on a plot using a line (design or computer science) which is then converted into information in your head using perceptual tasks and heuristics (psychology).
 
-- methods used in uncertainty visualisation evaluation remain ad hoc.
-studies approach issue from a usability perspective instead of asking WHY representations DO or DO NOT work.
-Studies do not follow any methodology commonly agreed upon.
-@Kinkeldey2014
-- Lack of formalisation and eigour in empirical methdos is an issue that is much broader than uncertainty visualisation (extends to info-vis and related domains).
-Therefore this paper is a starting point, not a definitive summary @Kinkeldey2014
-- The question in these papers is often "does method a work better than method B" which is an engineering approach, where the toal is to improve a particular product rather than create general principals @Kinkeldey2014
-- Authors then try to generalise beyond the specific constraints of the test even though the framing does not allow for it @Kinkeldey2014
-- Alternatively, we should ask how/why does method A work better than method B, which grounds work in perceptual and cognative theory and provides a framwork @Kinkeldey2014
+These overlapping but distinct fields create a high burden of spanning expertise to understand if two visualisation are equivalent in each field.
+This is largely caused by the concept of "information" differing dramatically between fields.
+According to mathematics, a sufficient statistic for a parameter contains the same amount of information as an entire sample.
+Therefore, if you are trying to estimate a population mean, the sample mean and the entire data set are both equivalent, however a similar statistic, such as the median, is not and will become a worse and worse approximation for the mean the more skewed your underlying distribution in.
+Computer science considered two pieces of information to be equivalent if they come from the same data [^This is the justification that I have come across in several computer science field papers, I am not sure if this is true for the whole field so this is definitely a question for Sarah].
+This definition of information is much looser and is contained within the mathematical definition of information.
 
-- low-probability high-impact events are particularly difficult to communicate risks about @Spiegelhalter2017
+Most uncertainty visualisation papers show that the definition of information *should* be the one defined by mathematics, not the one defined by computer science.
+A close look at the visual inference literature shows this to be true, but a large amount of noise is added to the research through visual heuristics.
+This is simply because, if one visualisation depicts a "significant statistic" and another visualisation shows a similar statistic, then the graphic that shows the significant statistic holds more relevant information to the question and will lead to more accurate interpretations.
+If the information that is depicted on a plot cannot be used to *mathematically* generate a best estimate for a specific statistic, I do wonder how the visual system would be expected to cover that gap if *not* for heuristics.
 
+While mathematical information *should* be considered the starting point in considering if two visualisations are equivalent, it is not the only consideration.
+A large amount of work in perceptual tasks, attention, and psychology surrounding charts shows that all the information depicted is not paid attention to in equal measure.
+Elements such as colour can have sizable impacts on the way a graphic is percieved.
+Mathematics identifies the information that is *shown* while psychology identifies the difference in information *recieved*.
+A failure to understand this distinction leads to graphics that are different not only in the way information is depicted but in the information itself.
+This problem is pervasive in uncertainty visualisation literature.
 # 4.?? Noise vs Signal (question +)
@@
 - I think the questions where uncertainty is a signal, not a noise, allow the authors of visualisation experiments to ignore the context and motivation that usually governs a visualisation.
-# 4.?? Tasks and Motivating Questions
+# 4.?? Tasks and Motivating Questions (uncertainty about.. .what?)
 - "cannot assess the quality of risk communication unless the objectives are clear" circles back to motivation. He assumes we are fulfilling a duty to inform. @Spiegelhalter2017
+## 4.?? What are you uncertain about? A mathematical considerations
+- Should add plot comparison mathematical framework example here
+- take a plot and work through the entire example as a flow chart thing
 ## 4.1?? In other statistics
@@
 - can't have a best practices because you need experimental results and my point is that the experiments aren't testing the right things
+- pixel map
+  - Similar to what could be good but its computation isnt quite what I would like
+  - heaps of sample depictions are animated and that is not always possible, but this is a clear indicator of how you can do a sample visualisaiton
+  - hasnt been tested so it is unclear if this actually works at signal supression
 - VSUP
-- the map interactive map that susuan showed me
-- the pixel map
-- The new york times wealthy people that end up in different economics classes, white vs black people
+  - The purpose of uncertainty is clear in the design and it is signal supression
+  - Supression is not the only uncertainty feature we should be interested in (shape is also important which is why I have beef with over smoothing and something visualisation is very good to express) but by being clear in its goal it knows what it is
+  - It addresses several key issues with the bivariate map that arrise in general and especially when it is used for visualising uncertainty 
+- the census dot map (has been taken down but CNN replicated it [here](https://edition.cnn.com/interactive/2021/us/census-race-ethnicity-map/))
+  - I love a good depiction of a sample
+  - You can see the sparse regions and the pop 
+  - it is similar to the pixel map (could also be singal supression but it is not tested)
+  - the sample depiction works well with the interactivity and you can see the signal that is appropriate for a particular level of zoomed in.
+If you zoom out the points turn into solid colours and show clear signal.
+- The new york times class mobility and race animated plot [here](https://www.nytimes.com/interactive/2018/03/19/upshot/race-class-white-and-black-men.html) 
 - the animated surface classifications from that satelite images experimental paper
-- the climate change scenario uncertainty
+- the climate change scenario uncertainty. There are quite a few of these:
+  - [e.g.1](https://earthobservatory.nasa.gov/features/GlobalWarming/page5.php)
+  - [e.g.2](https://climatechange.chicago.gov/sites/production/files/2016-07/scenariotempgraph_0.jpg)
+  - there are a few examples on the [climate change scenario wiki](https://en.wikipedia.org/wiki/Climate_change_scenario)
+  - Good example of uncertainty that has uncertainty from two different sources (human decision in the form of inputs and statistical uncertainty), one of which cant be quantified.
+  - a good example in how you can combine two different  types of uncertainty
+
+Notes
+- Since a lot of the good visualisations are made by random people creating a new visualisation (likely through back and forth communication which is HOW you get good visualisations) 
+- A lot of these show the data instead of an inference which is probably better for this scenario because sample size is important for uncertainty.
+It is not always appropriate and sometimes dots are not exactly equal to one draw.
+  - comments about smoothing in density plots and bin width in histograms.
+I do think the sample size should be communicated to people.
+While it is not the only thing that matters it seems kind of pointless to go to such lengths to hide it.
+- Sometimes points are used for things that are NOT samples which can get confusing and honestly I dont like it.
 
 # 5. Future work/Conclusion
+
+# Appendix (or information I am holding)
+## Experiment paper checklist
+General
+- Specific Field of Research
+- Question asked
+- Risk or Uncertainty: 
+- Noise vs signal (risk vs uncertainty)
+- Major task goal
+    - e.g. identify a bias or misunderstanding, compare two visualisation methods
+- Minor task goal
+    - Communication(value retrieval)
+    - Analytical(complex consideration)
+    - Exploration(free for all)
+
+Uncertainty Considerations
+	- Source
+	- Level (statistical-ignorance)
+	- Nature (Epistemic/Aleatory)
+
+Visual Uncertainty taxonomy
+- Explicit/implicit
+    - (directly mapping or showing multiple outcomes)
+- Intrinsic/Extrinsic 
+    - (using existing symbols e.g. colour value, or new objects e.g. grids)
+- Visually integral/separable 
+    - (can or cannot be separated from the data) 
+- Coincidence/adjacent 
+    - (if data and uncertainty are in integrated or separate views
+- Static/dynamic 
+    - (animation/interaction)
+
+Other plot considerations
+	- Dimension of data
+	- Dimension of uncertainty
+	- Feature mappings
+
+Extra (not necessarily recorded)
+	- Metrics used (and recorded)
+	- Possible Heuristics
+	- What is the ground truth?
+	- Participant literacy
 

### 2024-01-26 paper.qmd
finished the taxonomy summary bit and added the drawing

@@
 ---
-title: "Uncertainty Literature Review"
+title: "The Noisy Work of Uncertainty Visualisation"
 author: Harriet Mason
@@
 
+*I need to fix the taxonomy illustration because inputs and model uncertainty need to be swapped*
+
 ![Depicts an illustration of the taxonomy described in @utypo.
@@
 
-*The taxonomy* we will use in this paper may seem complicated and difficult to grasp at first glance, but it essentially boils down to three dimensions to consider.Uncertainty is a broad and complicated topic and while oversimplified taxonomies might help us hold the concept of uncertainty in our head, they do little to help with the practical question of how to quantify, visualise, or communicate uncertainty.
-- also makes it easier for people to communicate exactly what they need visualisation tools for. Look at geography papers and how their specific uncertainty considerations have been ignored for years.
-- Additionally, this definition of uncertainty is helpful in understanding *how* uncertainty should be communicated and the consideration we should make.
-- can organise other "uncertainties" according to this framework
+*The taxonomy* we will use in this paper may seem complicated and difficult to grasp at first glance, but it essentially boils down to three dimensions to consider.
+Uncertainty is a broad and complicated topic and while oversimplified taxonomies might help us hold the concept of uncertainty in our head, they do little to help with the practical question of how to quantify, visualise, or communicate uncertainty.
+Organising uncertainty according to its location, level and nature makes allows us to consider uncertainty in the miriad of ways it can arise in a data analysis project and better understand *how* this uncertainty should be communicated.
+With a detailed understanding of the sources of uncertainty and the ways it can be quantified, we can better understand the gaps in the current set of visualisation tools and assess if these gaps need to be filled.
 
-*The location* axis of the taxonomy lines up neatly with the typical data analysis pipeline and its importance is backed up by a several philisophical perspectives on statistics as well as discussions on the practical needs of uncertainty visualisation.
-- The usually uncertainty is only considered at the final stage but it needs to be considered through all stages.
-The big data paradox shrinks confidence intervals but magnifies bias [@Bradley2021] because the confidence intervals shrink due to the sample size but the imprecise target group creates bias.
-- @Meng2021 highlights that data needs to be treated as an output from another process rahter than an value neutral input.
-- If the distinction between noise and uncertainty is largely task dependent, considerations of quantifying and expressing uncertainty should be considered for the entire duraction of a project; another concept that has been independently identified by several authors [@Kinkeldey2014; @Refsgaard2007]
-- While quantification of the uncertainty that arrises from statistical philosophy and data provenance are beyond the scope of this paper, how this uncertainty is communicated is not.
-The ontological process discussed by @Otsuka2023 where information is tossed as we boil our information down into statistical objects requires us to map our statistical objects back to real world entities.
-This means the noise introduced and signal lost in earlier stages of the data analysis need to be identified, quantified and expressed in terms of real world outcomes.
-Visualisation is often the final stage of larger data science project and the down stream effects of poor abstraction and problem at the lower levels become impossible to ignore [@Munzner2009].
+*The location* axis of the taxonomy lines up neatly with the typical data analysis pipeline.
+Multiple authors in the uncertainty literature have commented on the need to consider quantifying and expressing uncertainty at every stage of a project [@Kinkeldey2014; @Refsgaard2007].
+@fig-loc shows the connection between the typical data analysis pipeline and the location element of the uncertainty taxonomy, along with examples of how uncertainty could arise from each of these steps.
+The relevance of every stage of the uncertainty taxonomy is backed up by both theoretical and practical work in statistics and adjacent fields.
+@Munzner2009, in her nested model for visualisation, comments on how even something as low level as poor problem abstraction can cause down stream effects and become impossible to ignore.
+@Otsuka2023 highlights that the choices made in the ontological process of boiling real world events down into statistical objects can influence later stages of analysis due to what was considred signal and what was considered noise.
+@Meng2021 stresses the importance of treating data as an input from another process with noise and variance, rather than as a neutral input.
+High level stages such as model selection and estimation have obvious introductions of uncertainty through the choice of models and nature of estimation [^I'm not sure if I need a citation here because this feels obvious].
+Location is also acknowledged as an important element of consideration in the taxonomies of several visualsiation papers [^ I will probably need a citation for which ones].
+
+[Illustration of the connection between the data generating process and the location of the uncertainty analysis. Several examples are collected from @utypo and @Munzner2009](location.png){#fig-loc}
 
 *The level* axis forces us to identify how quantifiable the uncertainty is.
-- it also gives us rules for combining uncertainty since statistical uncertainty can be combined with staitstical uncertainty, and outcome uncertainty can be combined with outcome uncertainty. 
-- Can discuss each uncertainty in isolation using distributions, ranges, and lists of alternatives, but the total cumulative uncertainty from every stage is more difficult (near impossible) @Spiegelhalter2017
-- Approaches to communicating limited understanding include 1) explicit model uncertainty 2) Qualitative scales expressed as strength of evidence (e.g.
-high to low certainty of evidence) 3) Acknowledge limitations and their possible impact 4) Acknowledged ignorance @Spiegelhalter2017
+This element is important because establishing how quantifiable uncertainty is informs us how it can be communicated, something that @Spiegelhalter2017 also identified in the form of "precision".
+Statistical uncertainty can be quantified, so it can be expressed using a PDF, a variance, an error estimate, or something similar while outcome uncertainty cant be quantified but each scenario may be simulated.
+Acknowledged ignorance can only be stated in the form of an assumption.
+Visualisations of climate change scenario uncertainty typically combine the outcome uncertainty of carbon emissions with the and statistical uncertainty of the noisy measurement and modelling system to create a cohesive illustration of the uncertainty surrounding climate change.
+It is commonly noted that each uncertainty must be discussed in isolation, but combining the uncertainty from every stage is near impossible [@Spiegelhalter2017].
+The level axis and climate change visualisations provide tools for how to consider combining these uncertainties.
+Uncertainties that are the same level but from multiple locations can likely be combined to establish an overall model uncertainty.
+While uncertainties at different levels cannot be combined, they can still be visualised simultaneously as shown in the climate scenario uncertainty visualisations.
 
@@
 While this level if not of particular importance to the visualisation of uncertainty, that is not the only purpose of the taxonomy.
-The nature of the uncertainty is an important element in expressing uncertainty, especually to stakeholders, and connecting the nature to the location makes it easier to understand *how* uncertainty can be reduced.
+The nature of the uncertainty is an important element in expressing uncertainty, especially to stakeholders, and connecting the nature to the location makes it easier to understand *how* uncertainty can be reduced.
+
+Mapping uncertainty into the 3D space defined by location, level, and nature can expand beyond the narrow case of statistical modelling it was defined for.
+*@___* included "consensus" as a taxonomy element, however it was not included in the final definition, but it can easily be added as a location that would align with the communication stage of the data pipeline.
+Several authors [*@_+_@Spiegelhalter2017*] considered the method by which uncertainty was quantified (e.g.
+by a single value, a function, simulated outcomes, etc), which can be be thought of as a more precise categorisation of the level axis.
+This flexibility means that the use of this framework is not limited by a specific case and should be used going forward to define what aspect of the uncertainty space a contribution exists in.
+
+*Should I add in a bit where uncertainty papers are organised according to which uncertainties they visualise or should I leave that until the lit review section?*
 
@@
 There are some taxonomies of visual uncertainty.
-These taxonmies seem to seek to define "uncertainty visualisation" and a combined term rather than as two separate words.
+These taxonomies seem to seek to define "uncertainty visualisation" and a combined term rather than as two separate words.
 We largely disagree with this conceptualization of uncertainty visualisations as it blurs the line between the mathematical estimations and assumptions (the uncertainty part) and the visual depiction of those mathematical objects (the visualisation part).
@@
 There are a small handful of taxonomies (or typologies) that are specifically for uncertainty visualisations.
-@Kinkeldey2014 identified give categories of uncertainty based on the work of other literature reivews of uncertainty visualisation.
+@Kinkeldey2014 identified give categories of uncertainty based on the work of other literature reviews of uncertainty visualisation.
 The five categories are: (1) explicit/implicit (directly mapping or showing multiple outcomes) (2) intrinsic/extrinsic (using existing symbols e.g.
@@
 This problem is pervasive in uncertainty visualisation literature.
+
+- The big data paradox shrinks confidence intervals but magnifies bias [@Bradley2021] because the confidence intervals shrink due to the sample size but the imprecise target group creates bias.
+
 # 4.?? Noise vs Signal (question +)

### 2024-01-26 paper.qmd
formatting edits so knit works

@@
 ## 2.2 Taxonomy Definitions
-# 2.2.1 Taxonomies of Uncertainty
+### 2.2.1 Taxonomies of Uncertainty
 While colloquial definitions of uncertainty are useful when trying to communicate with laypeople who don't understand strict statistical definitions, these definitions are not helpful when we are trying to work out exactly what should be shown on a plot.
@@
 Multiple authors in the uncertainty literature have commented on the need to consider quantifying and expressing uncertainty at every stage of a project [@Kinkeldey2014; @Refsgaard2007].
-@fig-loc shows the connection between the typical data analysis pipeline and the location element of the uncertainty taxonomy, along with examples of how uncertainty could arise from each of these steps.
+@fig-location shows the connection between the typical data analysis pipeline and the location element of the uncertainty taxonomy, along with examples of how uncertainty could arise from each of these steps.
 The relevance of every stage of the uncertainty taxonomy is backed up by both theoretical and practical work in statistics and adjacent fields.
@@
 
-[Illustration of the connection between the data generating process and the location of the uncertainty analysis. Several examples are collected from @utypo and @Munzner2009](location.png){#fig-loc}
+[Illustration of the connection between the data generating process and the location of the uncertainty analysis.
+Several examples are collected from utypo and Munzner2009.
+](location.png){#fig-location}
 

### 2024-01-30 paper.qmd
added a philosophy part to the uncertainty definition part, and did a lot of the general attitudes towards uncertainty

@@
 ---
-title: "The Noisy Work of Uncertainty Visualisation"
+title: "Noisy Work: A Review of The Uncertainty Visualisation Literature"
 author: Harriet Mason
@@
 Examples of Taxonomies used to untanlge the features of uncertainty:
+
 - aleatory uncertainty (unavoidable randomness), epistemic uncertatiny (uncertainty about structure, still quantified), ontological uncertainty (uncertain about entire modelling process).
@@
 
-[Illustration of the connection between the data generating process and the location of the uncertainty analysis.
+![Illustration of the connection between the data generating process and the location of the uncertainty analysis.
 Several examples are collected from utypo and Munzner2009.
@@
 by a single value, a function, simulated outcomes, etc), which can be be thought of as a more precise categorisation of the level axis.
+This taxonomy also ignores contextual information about the data, such as spatial or temporal considerations.
 This flexibility means that the use of this framework is not limited by a specific case and should be used going forward to define what aspect of the uncertainty space a contribution exists in.
 
-*Should I add in a bit where uncertainty papers are organised according to which uncertainties they visualise or should I leave that until the lit review section?*
+The taxonomies that cannot be contained within the @utypo taxonomy are not a new uncertainty dimensions, but rather they are conflating adjacent information with uncertainty.
+A good example of this is the distinction between spatial, temporal, and estimated uncertainty.
+When we create statistical models and quantify uncertainty, the spatial or temporal aspects are often boilled away [@Otsuka2023], and the only element of it that remains is the inexchangeability of our data.
+For this reason, there realistically is no such thing as "spatial" uncertainty in the same sense that there is no such thing as "photograph" uncertainty or "coin toss" uncertainty or "baby birth" uncertainty.
+Maps and photographs boil down to the same statistical objects in the same way that coin tosses and baby births boil down to the same statistical objects.
+Once we have removed the physical nature these objects are near identical when considered statistically.
+Of course the nature of them can change based on what we choose to observe (i.e.
+if we consider a coin landing on its side to be an outcome or if we consider intersex babies as an outcome in births) but this means the contextual information of our data, e.g.
+whether or not it is spatial or temporal, is only of importance at two stages of the data analysis pipeline.
+It is first important when we *boil it down* to our data through *observation* and again when we *map our statistical information back to its real world information* to *communicate* it.
+The only way the context uniquely contributes to uncertainty is in the "boil down" stage, however spatial or temporal uncertainty never refers to that specific cause of uncertainty.
+It is typically in reference to measurement error, data aggregation, statistical modelling, or other steps that occur *after* the boil down stage, and are therefore not unique to spatial or temporal data.
+While the spatial or temporal features *may* become relevant at the final communication stage, however this has little do with quantifying uncertainty, and the difficulties in communicating a complicated statistical model through a complicated map is already well discussed in the visualisation literature, and in later sections we will discuss how this concept is related to uncertainty visualisation.
+
+In this sense we are not proposing that the current conceptualisation of data according to its physical properties is necessarily wrong, or that it leads to incorrect conclusions, we are merely suggesting for a different way of slicing the metaphysical cake of uncertainty that better aligns with what is *actually happening* when we perform a statistical analysis.
+This metaphysical cake is illustrated below as @fig-cake.
+In changing  what the actual boundaries are when we work with these problems, we improve our understanding of what we are actually doing when we conceptualise uncertainty, and in the process allow for more creative solutions creativity (e.g.
+by borrowing concepts from spatial visualisation to visualise uncertainty in photographs generated).
+However, since uncertainty is almost synonymous with the concept of "statistics", this adjustment in how we conceptualise and organise statistical methods would likely benefit other domains within the field.
+
+![The metaphysical cake we are slicing differently](metaphysicalcake.jpg){#fig-cake}
+
+It is important to note that this taxonomy is the closest we can get to mathematical definition of uncertainty.
+While individual *expressions of uncertainty*, such as a confidence interval or a PDF are mathematically defined, the term *uncertainty* cannot be defined mathematically because it is a layperson word that has been adopted by statistics to describe any deviation from complete determinism, whether mathematically defined or not.
+While the broader field of uncertainty visualisation does need to be more closely inspected, specifically through the scope of mathematical definitions, this issue cannot be solved with a more specific definition of uncertainty through mathematics.
+Instead or establishing a new definition of uncertainty that is strict enough to prevent the current array of misunderstandings, we should focus on inspecting how the current literature *deviates* from established mathematical definitions, why these deviations occur, and understand how these deviations can lead to larger problems in uncertainty communication.
 
@@
 
-# 3.General Attitude Towards Uncertainty + Its communication
+# 3.General Attitude Towards Uncertainty (Uncertainty Communication Boradly)
+It would be impossible to discuss uncertainty visualisation without mentioning the concept of uncertainty communication and discussing how visualisation fits into that broader scope.
+Uncertainty visualisation is related to uncertainty communication in the same way that visualisation more broadly is related communication more broadly.
+Visualisation is fast, it can communicate more complicated ideas that what can be done with text alone, and
+
+*psychology of uncertainty communication*
+
+It is difficult to read a paper that discusses uncertainty that does not make a nod towards uncertainty bias.
+In the case of uncertainty visualisation, these nods are not always relevant.
+Similarly to how contextual information is conflated with uncertainty when we try to define it, human attitudes to risk are often commented on in the discussions of uncertainty visualisation.
+Ironically, despite being mentioned in a large proportion of papers that discuss uncertainty risk aversion, has very to do with the psychology of uncertainty.
+As a matter of fact, it has so little to do with uncertainty communication that we would not even mention it in this paper were it not for the fact that ignoring it would suggest we had missed an important section of the literature.
+
+Risk-aversion is the tendancy of people to prefer outcomes of low uncertainty to those outcomes with high uncertainty despute the outcomes with high uncertainty having a higher or equal expected outcome.
+
+Uncertainty information is typically included in some form of statistical communication to offer some transparency to the viewer.
+Data analysis is a long and complicated process, so much so that even experts in the field can find themselves lost with a single incorrect assumption, so uncertainty can help pull back the curtain on this process and prevent people from understanding it as a human agnostic, entirely deterministic process.
+When science authors discuss "risk avoidance" in the uncertainty communication conversations, it holds with it some implication that heuristics such as risk avoidance are relevant in every conversation about uncertainty communication, however that is not true.
+
+Risk and uncertainty are not perfect synonyms of one another.
+@Spiegelhalter2017 makes the distinction clear by specifying *risk* as *a precise random event* (such as a coin flip) while uncertainty is imprecision (such as the estimated probability associated with a coin that might be biased).
+Additionally, risk avoidance is observed when people have a choice between a *deterministic* outcome and a *risky* event that has higher expected value.
+How this psychological heuristic is therefore related to all uncertainty visualisation despite uncertainty communication being different in both the scenario (i.e.
+it is highly unlikely that we need to communicate a choice between a perfectly deterministic and an uncertain event) and in the type of uncertainty (since risk and uncertainty are not synonyms).
+
+The regular conflation of risk and uncertainty despite clear and specific differences, without any explanation of how these terms are related, implies that a large portion of uncertainty literature is simply reporting information from any research that is uncertainty adjacent without considerations for the specifics of how these concepts would be related to uncertainty.
+
+It is not even clear that this behaviour is entirely irrational either, considering risk aversion is, at its core, a trade off between bias and variance.
+If risk-aversion is truly "irrational" behaviour, every statistical textbook that discusses the use of a biased but consistent estimator in a fite sample case (a common example would be specific cases of the maximum likelihood estimator) should immediately toss those chapters out and the authors should feel embarassed for that *clearly* irrational line of thinking.
+Of course I am making a joke but this does highlight a clear problem in the way risk-aversion is discussed in the literature.
+Discussing risk-aversion as though it is some diversion from a perfectly rational, mathematical choice, implies there is *no value to a decrease in variance* which is not even true *within mathematics*.
+The concept of risk aversion treats risk as "noise", as though it is something that only exists to be ignored, but if the common place use of biased MLE's is an indication of anything, it is that *ignoring uncertainty* is the irrational choice.
+
+Selecting a biased is technically statisticians doing "risk aversion", but when used to ironically when it is used to select models it is not viewed as an irrational human fault.
+
+This issue, where psychological considerations seem to be entirely devorced from what is *actually* being discussed extends to the conversation around trust as well.
+The idea of measuring or being concerned about audiences perception of trust, without first establishing if what we are communicating is *trustworthy*, .
+The goal of uncertainty visualisation should not be to be trustworthy, but rather to be transparent (which can facilitate trust).
+
 I dont think this is worth a section but it is mentioned a lot in .
-- the role of unertainty visualisation is to communicate accurate information
 - problem of communicating to a group thing (avoid communicating inaccurate information)
@@
 design of scale important because risk is associated with position rather than magnitude @Spiegelhalter2017
-- heath had a wide range of different metrics they adopted to communicate chronic risk of adverse events in the future @Spiegelhalter2017
 - using words to communicate numerical risk is generally discouraged and using words without numerical interpretation can lead to lower accuracy and higher risk aversion @Spiegelhalter2017
@@
 - low-probability high-impact events are particularly difficult to communicate risks about @Spiegelhalter2017
+
 # 4 LITERATURE SUMMARY
@@
 - Alternatively, we should ask how/why does method A work better than method B, which grounds work in perceptual and cognative theory and provides a framwork @Kinkeldey2014
+
 # 4.?? Similar work (other literature reviews)
@@
 - people with high numeracy count pictographs but low numeracy assess area @Spiegelhalter2017
+
 # 4.1 Plot Comparison Papers
@@
 - I think the questions where uncertainty is a signal, not a noise, allow the authors of visualisation experiments to ignore the context and motivation that usually governs a visualisation.
+
 # 4.?? Tasks and Motivating Questions (uncertainty about.. .what?)
@@
 ## 4.1?? General task notes
+- heath had a wide range of different metrics they adopted to communicate chronic risk of adverse events in the future @Spiegelhalter2017 (use this for motivation based statistics since they define new statistics based off what they need)
 - Mentioned alleatory vs epistemic uncertainty, in the visualisation literature alleatory uncertainty is more commonly the focus.

### 2024-02-01 paper.qmd
added in taxonomy screenshots from papers
added link to twitter thing tom found

@@
 - Level of precision in epistemic uncertainty: 1) numbers given to appropriate levels of precision 2) a distribution range 3) a measure of statistical significance 4) verbal quantifiers to numbers 5) refusal to give a number unless the evidence is good enough @Spiegelhalter2017
+
+![Example of different uncertainty taxonomies](paperscreenshots/collage1.jpg){#fig-location}
 
@@
 While this paper is more about mathematics than visualisation, it is clearly written as a precursor to visualisation.
+
+![Example of different uncertainty visualisation taxonomies](paperscreenshots/collage2.jpg){#fig-location}
 
@@
   - a good example in how you can combine two different  types of uncertainty
+  - [This](https://www.esri.com/arcgis-blog/products/arcgis-pro/mapping/graphical-cartograms-in-arcgis-pro/) has some good maps and some bonkers ones
 

### 2024-02-02 paper.qmd
made some edits to the general attitudes

@@
 # 3.General Attitude Towards Uncertainty (Uncertainty Communication Boradly)
-It would be impossible to discuss uncertainty visualisation without mentioning the concept of uncertainty communication and discussing how visualisation fits into that broader scope.
-Uncertainty visualisation is related to uncertainty communication in the same way that visualisation more broadly is related communication more broadly.
-Visualisation is fast, it can communicate more complicated ideas that what can be done with text alone, and
+It would be impossible to discuss uncertainty visualisation without mentioning the broader conversation around uncertainty communication.
+The conversation around uncertainty communication is often centered around the psychology is communicating risks.
 
-*psychology of uncertainty communication*
+Most research into general uncertainty communication seems to focus specifically on how we communicate risks.
+Risk and uncertainty are not perfect synonyms of one another.
+@Spiegelhalter2017 makes the distinction clear by specifying *risk* as *a precise random event* (such as a coin flip) while uncertainty is imprecision (such as the estimated probability associated with a coin that might be biased).
+For this reason, risk is often seen as a specific case of uncertainty and that well defined sub area has a lot of detailled research.
+Some markers of probability that have common uses, probabilities such as 0 (impossible), 0.5 (a fair coin flip), 1 (certain) are easy for people to have an internal sense of [@Spiegelhalter2017].
+It is incredibly hard to communicate small risks, because people cannot differentiate between 1/100 an 1/1000, and communicating low probability but high impact risks are particularly challenging [@Spiegelhalter2017].
+This issue can be somewhat alleviated by communicating the likelihood of an event in relation to another event on a similar scale (such as being struck by lightning) [@Spiegelhalter2017].
+The framing of how we communicate risks matters due to the affect heuristic, so there will be a difference in peoples behaviours if, for example, you tell people they have a 90% chance of survival vs a 10% chance of death, and when expressing a frequency, a larger numerator communicates a larger risk [@Spiegelhalter2017].
+The reference class of a particular risk should be explicitly stated to avoid confusion, for example if we are communicating the chance of rain, the time period over which we are expecting this chance of rain (i.e.
+50% chance today or this week) needs to be specified [@Spiegelhalter2017].
+The problem with a risk focus is that it is only a small piece of the uncertainty pie, and very few of these findings can be reasonably assumed to go beyond risk communication.
+Even less can be assumed to translate to uncertainty visualisation.
 
-It is difficult to read a paper that discusses uncertainty that does not make a nod towards uncertainty bias.
-In the case of uncertainty visualisation, these nods are not always relevant.
-Similarly to how contextual information is conflated with uncertainty when we try to define it, human attitudes to risk are often commented on in the discussions of uncertainty visualisation.
-Ironically, despite being mentioned in a large proportion of papers that discuss uncertainty risk aversion, has very to do with the psychology of uncertainty.
-As a matter of fact, it has so little to do with uncertainty communication that we would not even mention it in this paper were it not for the fact that ignoring it would suggest we had missed an important section of the literature.
+*this is all currently from the spieglehatter paper because that is the only paper I have moved notes over from mendeley before I started doing the summary write up.
+this will not all be from one paper, dont worry.*
 
-Risk-aversion is the tendancy of people to prefer outcomes of low uncertainty to those outcomes with high uncertainty despute the outcomes with high uncertainty having a higher or equal expected outcome.
-
+Risk and uncertainty are quite different in practice.
 Uncertainty information is typically included in some form of statistical communication to offer some transparency to the viewer.
@@
 
-Risk and uncertainty are not perfect synonyms of one another.
-@Spiegelhalter2017 makes the distinction clear by specifying *risk* as *a precise random event* (such as a coin flip) while uncertainty is imprecision (such as the estimated probability associated with a coin that might be biased).
+There are some elements of risk communication that seem to translate to uncertaitny communication more broadly.
+Using words to communicate risk (as well as uncertainty more broadly) is discouraged because people can misinterpret the actual risk involved, so when possible numerical estimates of risk should be communicated, however this can make information harder to understand for those with poor numeracy skills [@Spiegelhalter2017].
+
+likely due uncertainty and risk being near synonymous, it is difficult to read a paper that discusses uncertainty that does not make a nod towards human attitudes towards risk.
+Even when the uncertainty being discussed in that particular paper has very little to do with risk.
+
+Risk-aversion is the tendancy of people to prefer outcomes of low uncertainty to those outcomes with high uncertainty despute the outcomes with high uncertainty having a higher or equal expected outcome.
+The story of uncertainty communication is littered with conflated ideas, and that issue is highly visible here.
+Despite being mentioned in a large proportion of papers that discuss uncertainty risk aversion, seems to have very little to do communicating uncertainty.
+As a matter of fact, it has so little to do with uncertainty communication that we would not even mention it in this paper were it not for the fact that ignoring it would suggest we had missed an important section of the literature.
+
  Additionally, risk avoidance is observed when people have a choice between a *deterministic* outcome and a *risky* event that has higher expected value.
@@
 
-- compares a bunch of stuff to the chance of getting struck by lightning since they are on a common scale (which is kind of funny) @Spiegelhalter2017
-- hard to communicate small risks because people struggle to discrimintate between 1 in 10 000 and 1 in 100 000.
-Common solution is to use a logarithmic scale (if drawn vertically these are known as risk ladders).
-design of scale important because risk is associated with position rather than magnitude @Spiegelhalter2017
-- using words to communicate numerical risk is generally discouraged and using words without numerical interpretation can lead to lower accuracy and higher risk aversion @Spiegelhalter2017
-- the Affect heuristic comes to play in how people understand risk @Spiegelhalter2017
-- avoids the topic of risk at times of accute crisis because numerical risk would be inappropriate @Spiegelhalter2017
-- positive vs negative framing matters, can try to avoid it @Spiegelhalter2017
-- no preference between percent or frequency although (small interaction with numeracy skills though) @Spiegelhalter2017
-- Need careful specification of the reference class (e.g. % chance of rain should be fractions of situations like this in which is rains - over both time and location) @Spiegelhalter2017
-- Need to make time interval clear (50% each year) @Spiegelhalter2017
-- understanding better in a table than text @Spiegelhalter2017
-- if expressing as a frequency, note that larger numerators suggest a larger risk, exteme version of this is ignoring the denominator all together @Spiegelhalter2017
-
-- low-probability high-impact events are particularly difficult to communicate risks about @Spiegelhalter2017
-
+Uncertainty visualisation is related to uncertainty communication in the same way that visualisation more broadly is related communication more broadly.
+Visualisation is fast, it can communicate more complicated ideas that what can be done with text alone, and
 # 4 LITERATURE SUMMARY
@@
 - take a plot and work through the entire example as a flow chart thing
+- tasks e.g.
+"What is the average of X" (how well lined up you are moving down the data pipeline) and resolution (uncertainty is very often moving back a resolution level in statisitcs - also can be an assumption - bootstrap is a replacement for an assumption, not a resolution change, but bootstrapping is centered on the data)  e.g.
+"What is the noise around the estimate of the value of X"
 ## 4.1?? In other statistics

### 2024-02-05 paper.qmd
finished general uncertainty communication bit

@@
 Risk and uncertainty are not perfect synonyms of one another.
-@Spiegelhalter2017 makes the distinction clear by specifying *risk* as *a precise random event* (such as a coin flip) while uncertainty is imprecision (such as the estimated probability associated with a coin that might be biased).
+@Spiegelhalter2017 makes the distinction clear by specifying *risk* as *a precise random event* (such as a coin flip) while *uncertainty* (also sometimes referred to as abiguity) is imprecision (such as the estimated probability associated with a coin that might be biased).
 For this reason, risk is often seen as a specific case of uncertainty and that well defined sub area has a lot of detailled research.
@@
 
-Risk and uncertainty are quite different in practice.
-Uncertainty information is typically included in some form of statistical communication to offer some transparency to the viewer.
-Data analysis is a long and complicated process, so much so that even experts in the field can find themselves lost with a single incorrect assumption, so uncertainty can help pull back the curtain on this process and prevent people from understanding it as a human agnostic, entirely deterministic process.
-When science authors discuss "risk avoidance" in the uncertainty communication conversations, it holds with it some implication that heuristics such as risk avoidance are relevant in every conversation about uncertainty communication, however that is not true.
+Risk and uncertainty are not only different in how they are communicated, but also why they are communicated.
+Risk communication focuses on communicating probability around events.
+These events are typically a binary event that is known to be random, so a version of communication that does not mention risks (e.g.
+telling someone they will or will not get cancer instead of communicating their risks about it) inherently carries with it a lack of transparency.
+Additionally risk communication is often communicated to avoid or informa about an unfavorable outcome, so framing can be important if we want to influence decisions in a certainty way.
+These motivations do not extend to the communication of uncertainty information more broadly.
+While uncertainty information *is* often used to improve decisions, it seems the main motivation in its communication is to offer transparency to the viewer about a statistical analysis.
+That is, it acts as a form of statistical "hedging" for signals found in an analysis.
+Interviews with experts in statistic back up this primary motivation, as ignoring uncertainty information of often expressed as being similar to fraud or lying and the goal of "improving decisions" is only seen as a secondary outcome of an appropriately hedged signal [@Hullman2020a; @Manski2020].
+This concept of uncertainty means that the work on communicating risk does not cover the field of uncertainty communication as a whole.
+It covers the specific case where the signal itself *is* uncertain, but this work does not naturally extend to the goal of hedging a signal (i.e.
+when we want to communicate uncertainty as noise).
+Unfortunately understanding the relevant noise seems to be its own problem, as statisticians and visualisation authors seem to be unaware or confused in how to calculate the uncertainty related to an analysis [@Hullman2020a].
+Additionally, communicating uncertainty such that it conveys the appropriate singal supression involves combining uncertainty from multiple different sources, something that is considered to be difficult by even experts in uncertainty communication [@Spiegelhalter2017].
 
 There are some elements of risk communication that seem to translate to uncertaitny communication more broadly.
-Using words to communicate risk (as well as uncertainty more broadly) is discouraged because people can misinterpret the actual risk involved, so when possible numerical estimates of risk should be communicated, however this can make information harder to understand for those with poor numeracy skills [@Spiegelhalter2017].
+Using words to communicate risk is discouraged because people can misinterpret the actual risk involved, so when possible numerical estimates of risk should be communicated, however this can make information harder to understand for those with poor numeracy skills [@Spiegelhalter2017].
+People performing better and preferring with numerical estimates of risk translates to uncertainty communication more broadly *inc (citations for this)*.
+Risk communication also needs to have clear objectives, use plain language, limit information to only what is necessary, and segment the audience to allow for differences in interest and knowledge [@Spiegelhalter2017].
+These general concepts of communication also extend to communication of uncertainty.
+On the other hand, it is not sensible to assume that probability specific concepts such as framing, intuitiveness of probabilities, or reference classes are still relevant when we consider uncertainty communication as signal supression, not as a synonym of risk or probability communication.
+In this same vein, most uncertainty communication papers very often discuss risk-aversion as something that needs to be considered despite it being unclear exactly how risk-aversion is related to uncertainty communication *(citations for the papers that do this)*.
 
-likely due uncertainty and risk being near synonymous, it is difficult to read a paper that discusses uncertainty that does not make a nod towards human attitudes towards risk.
-Even when the uncertainty being discussed in that particular paper has very little to do with risk.
+In the same sense that risk and uncertainty are different, risk-aversion and amiguity-aversion (which can be considered the uncertainty version of risk aversion) are also not the same.
+Risk-aversion is the tendency of people to prefer outcomes of low uncertainty to those outcomes with high uncertainty despite the outcomes with high uncertainty having a higher or equal expected outcome.
+In the case of communicating risk, the only reason this heuristic would be relevant was if risk-aversion was a mistake that needed to somehow be corrected for, rather than a heuristic that *reveals the utility of decreased uncertainty*.
+Risk-aversion is often treated as a mistake or something that should be avoided by the uncertainty visualisation literature even though that is not necessarily true, depending on what is causing the risk-aversion (e.g.
+is it due to poorer estimates or increased awareness of negative outcomes).
+If a particular communication method leads people to place a greater value on certainty or make worse estimates, that is the aspect of importance, not risk-aversion which conflates these factors.
+The work that advocates for transparency in risk-communication gives you whiplash when it's subtext argues that transparency leads to "incorrect" conclusions.
+Risk-aversion has a large number of confounding factors that make it difficult to understand if it is something to be considered in risk communication, ambiguity-aversion has a similar problem.
+Ambiguity-aversion is the tendency of people to prefer to take on certain risks rather than unknown risks.
+It is essentially risk-aversion applied to risk.
+For this reason discussions of ambiguity-aversion have the same issues as the conversations about risk-aversion, where they do not consider ambiguity-aversion to be an estimate of the utility of decreased uncertainty, it contains a subtext that argues against transparency, and it is considered a valuable insight in of itself despite conflating many factors.
 
-Risk-aversion is the tendancy of people to prefer outcomes of low uncertainty to those outcomes with high uncertainty despute the outcomes with high uncertainty having a higher or equal expected outcome.
-The story of uncertainty communication is littered with conflated ideas, and that issue is highly visible here.
-Despite being mentioned in a large proportion of papers that discuss uncertainty risk aversion, seems to have very little to do communicating uncertainty.
-As a matter of fact, it has so little to do with uncertainty communication that we would not even mention it in this paper were it not for the fact that ignoring it would suggest we had missed an important section of the literature.
-
- Additionally, risk avoidance is observed when people have a choice between a *deterministic* outcome and a *risky* event that has higher expected value.
-How this psychological heuristic is therefore related to all uncertainty visualisation despite uncertainty communication being different in both the scenario (i.e.
-it is highly unlikely that we need to communicate a choice between a perfectly deterministic and an uncertain event) and in the type of uncertainty (since risk and uncertainty are not synonyms).
-
-The regular conflation of risk and uncertainty despite clear and specific differences, without any explanation of how these terms are related, implies that a large portion of uncertainty literature is simply reporting information from any research that is uncertainty adjacent without considerations for the specifics of how these concepts would be related to uncertainty.
-
-It is not even clear that this behaviour is entirely irrational either, considering risk aversion is, at its core, a trade off between bias and variance.
-If risk-aversion is truly "irrational" behaviour, every statistical textbook that discusses the use of a biased but consistent estimator in a fite sample case (a common example would be specific cases of the maximum likelihood estimator) should immediately toss those chapters out and the authors should feel embarassed for that *clearly* irrational line of thinking.
-Of course I am making a joke but this does highlight a clear problem in the way risk-aversion is discussed in the literature.
-Discussing risk-aversion as though it is some diversion from a perfectly rational, mathematical choice, implies there is *no value to a decrease in variance* which is not even true *within mathematics*.
+The idea that risk and ambiguity aversion reveal the utility of decreased variance seems to seldom be considered in the literature, therefore it should be mentioned here, otherwise this issue might continue.
+However, considering risk aversion is at its core, a trade off between bias and variance, it is bizzare to assume it to be a mistake when statisticians perform these trade offs all the time.
+If risk-aversion is truly "irrational" behaviour, every statistical textbook that discusses the use of a biased but consistent estimator in a finite sample case (a common example would be specific cases of the maximum likelihood estimator) should be tossed out.
+Discussing risk and ambiguity aversion as though it is some diversion from a perfectly rational, mathematical choice, implies there is *no value to a decrease in variance which is not even true within mathematics*.
 The concept of risk aversion treats risk as "noise", as though it is something that only exists to be ignored, but if the common place use of biased MLE's is an indication of anything, it is that *ignoring uncertainty* is the irrational choice.
 
-Selecting a biased is technically statisticians doing "risk aversion", but when used to ironically when it is used to select models it is not viewed as an irrational human fault.
+Trust is another concept that is related to uncertainty communication, but fact that it is often front and center in the discussion has a somewhat ominous subtext.
+If the purpose of displaying uncertainty information is to appropriately hedge a signal with noise, then it should be assumed that trust is only related to uncertainty communication through increased transparency and honesty.
+It does seem to be the case that uncertainty communication increases trust because it is a proxy for *(add citations)*.
+Unfortunately a large proportion of the literature discusses trust as something that is directly related to uncertainty visualisation rather than as an observable product.
+In viewing uncertainty communication as directly related to trust, not related through the proxy of transparency, several unobserved variables are conflated.
+The amount of uncertainty that is present in the information, whether or not the information is trustworthy or makes sense, prior beliefs of the participants, and the trustworthiness of the source are a few examples of variables that are conflated when authors directly consider trust to be of direct interest.
+Similarly to the issues in risk-aversion directly considering trust, and not transparency, as the metric of importance in uncertainty communication leads to a questionable subtext that argues against transparency.
+Being concerned about audiences perception of trust, without first establishing if what we are communicating is *trustworthy*, leads to the implication that details that result in information not being considered trustworthy should be avoided.
 
-This issue, where psychological considerations seem to be entirely devorced from what is *actually* being discussed extends to the conversation around trust as well.
-The idea of measuring or being concerned about audiences perception of trust, without first establishing if what we are communicating is *trustworthy*, .
-The goal of uncertainty visualisation should not be to be trustworthy, but rather to be transparent (which can facilitate trust).
+Authors that imply that a decrease in trust or an increase in ambiguity or risk aversion are metrics of importance in of themselves do not understand the purpose of uncertainty communication as methods to increase transparency.
+Science communication should be primarily concerned with accuracy, setting trust and risk-aversion as the variables of interest implicitly encourages statisticians to set trust and risk-aversion as the primary goals of communication.
+The issue of trust being divorced from trustworthiness has been commented on by other authors [@ONeill2018], however the issue still persists in the uncertainty visualisation literature [@Zhao2023].
+Additionally, it seems that risk-aversion being divorced from the value of risk has gone unnoticed.
 
-I dont think this is worth a section but it is mentioned a lot in .
-- problem of communicating to a group thing (avoid communicating inaccurate information)
-- how people feel about uncertainty is irrelevant
-  - natural disasters make people uncomfortable but to include it in a conversation about whether or not we should express it is rediculous.
-- Numerical risks: Use absolute risks (but relative risk when catastrophic event), for single unique event use percent chance, express change as a proportion, frequency, or percentage (and be clear about reference class), avoid framing bias by showing risk with and without the outcome, keep the denominator fixed when making comparisons, explicit about time interval, For more knowledgeable audiences consider providing epistemic uncertainty and confidence in analysis @Spiegelhalter2017
+While this is a focus on uncertainty visualisation, these issues in uncertainty communication need to be acknowledged because they bleed into the uncertainty visualisation literature.
+Uncertainty visualisation are compared on their ability to communicate explicit risk, elicit trust, and prevent risk or ambiguity avoidance, continuing these issues in uncertainty communication at large.
+While uncertainty visualisation can make it easier to communicate the complicated details of uncertainty information, it carries some challenges that are unique to the visualisaition, specifically with respect to the "signal supression" concept.
+In simple estimates or verbal communication, the signal is often easy to identify because it is what we are explicitly saying.
+Visualisations are used in both data exploration and communication.
+This means what exactly is a *signal* in any particular visualisation is hard to identify, since we often let the visualisation *tell us* what the signal is.
+Additionally, you cannot add noise to *every single possible* signal one might take from a visualisation.
+Two people looking at the same visualisation might, just by chance, develop two entirely different insights.
+These unique and facinating challenges that are faced by viewing uncertainty visualisaiton through the lense of "signal supression" have been almost completely untouched by the literature.
 
-- General issues when communicating risks: clear objectives, segment audience into target groups, use plain language and limit info to only what is necessary, allow for different levels of interest knowledge and numeracy (a gist level, numerical level, then evidence and uncertainty) @Spiegelhalter2017
-
-Uncertainty visualisation is related to uncertainty communication in the same way that visualisation more broadly is related communication more broadly.
-Visualisation is fast, it can communicate more complicated ideas that what can be done with text alone, and
 # 4 LITERATURE SUMMARY

### 2024-02-06 paper.qmd
added in bias variance trade off comments and analogy and started the similar work section

@@
 
-# 4 LITERATURE SUMMARY
+# 4 Uncertainty Visualisation Literature Summary
+
+## 4.1 Similar work
+There are a wealth of uncertainty communication and visualisation literature reviews that already exist.
+
+@Kinkeldey2014 summarised the literature specifically for geopsatial uncertainty visualisations and attempted to describe similarities in their characteristics.
+They found large difficulties in the ability
+
 - methods used in uncertainty visualisation evaluation remain ad hoc.
@@
 - Alternatively, we should ask how/why does method A work better than method B, which grounds work in perceptual and cognative theory and provides a framwork @Kinkeldey2014
+- Blur and fog are suggested because they are metaphors for uncertainty, however the usefulness of these metaphors is rarely investigated @Kinkeldey2014
 
-# 4.?? Similar work (other literature reviews)
 - If you just keep using rankings, you need to compare everything one by one forever to get an idea of the space @Hullman2016
@@
 - experiencing an event makes your percieved probability higher than if you see a description of it @Hullman2016
-- Blur and fog are suggested because they are metaphors for uncertainty, however the usefulness of these metaphors is rarely investigated @Kinkeldey2014
 - Frequency thing means that more effective communication to participants in a study means more effective science and we have a roll on effect @Hullman2016
@@
 - You cant tell if subjects in an experiment do poorly because the visualisation was bad or because the question was misinterpreted @Hullman2016
-- Visualisations risk: consider a good summary table as a visualisation, use multiple formats because no single representation suits all members of an audience, illuminate graphics with words and numbers, design graphics to allow a part-to-whole comparison on an appropriate scale, narrative labels are important (show magnitude through tick marks), use narrative, images and metaphors to gain/retain attention but dont arouse undue emotion, assume low numeracy of a general public audience and adopt a less-is-more approach by reducing the need for inderence and making clear and explicit comparison, be cautious about interactivity and animations (may introduce uncessary complexity), avoid chart junk (like 3d bar charts), assess the needs of the audience, and iterate towards a final design.
-@Spiegelhalter2017
-- Graphical features that improve accuracy, features that facilitate behavioural change, and features that viewers like may be different (are not necessarily the same). @Spiegelhalter2017
-- also should consider graphical literacy @Spiegelhalter2017
-- graphs are good for gists and different graphics are good at different things (i.e. trends vs comparison etc) but hard to avoid framing (show information as part of a whole) @Spiegelhalter2017
-- scatting in pictograph increases the impression of randomness but makes it difficult to count @Spiegelhalter2017
-- people with high numeracy count pictographs but low numeracy assess area @Spiegelhalter2017
+
+@Spiegelhalter2017 presents an literature reivew of uncertainty communication as a whole and considers uncertainty visualisation as a subfield of communication as a whole.
+In this review he has several key findings that are of particular interest.
+First, the best visualisation is highly dependent on your audience, and depending on how broad your audience is, there may not be a best visualisation.
+No single representation will be accessible to all members of an audience, and even if most audience members can read a plot, they may not read it in the same way.
+For example, people with high numeracy skills count pictographs, but people with low numeracy assess area, and seemingly unrelated elements may not equally influence all these methods of reading, for example scattering in a pictograph makes counting particularly more difficult.
+For these reasons, it is better to understand your audience and iterate towards a final design, rather than only consider one option.
+Second, the best visualisation for one task is unlikely to be the best visualisation for another task.
+Graphics are good for "gists" and different graphics are good for different "gists", e.g.
+one plot may we good at identifying trend, while another might be good for comparison @Spiegelhalter2017.
+Therefore, if we change the goal of our visualisation, and therefore the metric,
+
+Visualisations that improve accuracy, visualisations that improve behavioural change, and visualisation that viewers like may not necessarily all be the same visualisation @Spiegelhalter2017.
+
+He highlights a theme that is a common theme in the uncertainty visualisation literature, which is no visualisation will cover enough information, graphics are good for "gists" and different graphics are good at different things (e.g.
+identifying trends vs comparison).
+we consider a summary table as a visualisation
+
+illuminate graphics with words and numbers, 
+design graphics to allow a part-to-whole comparison on an appropriate scale, 
+include appropriate narrative labels (such as magnitudes through tick marks), 
+use images and metaphors to gain/retain attention but dont arouse undue emotion, 
+assume low numeracy of a general public audience and adopt a less-is-more approach by reducing the need for inference and making clear and explicit comparison, 
+be cautious about interactivity and animations (may introduce unnecessary complexity), 
+avoid chart junk (like 3d bar charts),
+
+These literature reviews highlight a common theme,
+In contrast to these literature reviews, we do not have a specific domain or tool analysis we intend to perform.
+Rather, the point of this literature review is to understand the *purpose* of visualising uncertainty (i.e.
+signal supression), and understand if the current visualisation tools are sufficient to achieve that purpose and if not, where there is need for improvement.
+Additionally, rather than discussing "uncertainty visualisation" as its own field, we view it as a subfield of both uncertainty and visualisation, and understand if this change in conceptualisation can alleviate or help us understand some of the issues faced by previous authors attempting to summarise the uncertainty visualisation literature.
 
@@
 # 4.?? Tasks and Motivating Questions (uncertainty about.. .what?)
-- "cannot assess the quality of risk communication unless the objectives are clear" circles back to motivation. He assumes we are fulfilling a duty to inform. @Spiegelhalter2017
+
+### A navigating a forest analogy
+When communicating uncertainty information, it is important to ask if we should consider the bias to be variance.
+We often talk about the "bias variance trade off" but we rarely discuss bias as "cashed in" variance, which in some sense, it is.
+If you are struggling to understand how bias is just traded off variance, let us walk through an analogy.
+
+Let us consider the data analysis pipeline a dense forest that you need to cross to get to your house.
+To prevent yourself from getting *too* lost you decide to bring some poles and rope so you can mark where you are going.
+You know your rope is long enough to get you to your house with enough wiggle room that you can reach the 5 houses either side of your house.
+This problem is drawn in @fig-forest1.
+@fig-forest2 depicts the typical data analysis pipeline and resulting uncertainty using this forest analogy.
+In this method, we navigate the forest by putting one of our poles down every time we notice the trees have changed (i.e.
+we have move to a new section of the forest).
+As we move down the line and we put our poles down, we cannot be certain we are always walking in the direction of our house, but we can do small checks or validation to make sure we are vaguely heading in the right direction.
+When we get to the final stage, we have enough rope remaining to get to the three houses directly in front of us, and we state that band as our "uncertainty".
+@fig-forest3 depicts the bias variance trade off in the case where we make no decisions, record as much information as possible, and check every single model with every single variable and keep every alternative as a possible option.
+This method will result in no bias (although it is often impossible to check the entirety of these spaces so it cannot) but a huge amount of variance.
+In an effort to avoid putting down any poles, we have managed to navigate the forest, but have essentially no idea which of the 10 houses are ours.
+Of course, with sufficient feedback we can whittle down this vast array of choices (which is what a neural network can do) but feedback from an increased sample size can only influence variance, it cannot influence bias that was previously variance.
+@fig-forest4 shows an alternative way to perform the data analysis process to consider variance at stages beyond model uncertainty (such as the scenario uncertainty that comes with different human choices in climate change models).
+In this scenario you navigate the forest to a certain point, and then clone yourself three times and each put down a new pole at different distances to each other.
+All four versions of yourself will then walk in an identical fashion from the pole to the end of the forest, and each have your own path with your own confidence interval.
+This is how the [climate scenario uncertainty](https://climatechange.chicago.gov/sites/production/files/2016-07/scenariotempgraph_0.jpg) is constructed.
+
+::: {#fig-forest layout="[[59,-2, 39], [59,-2, 39]]" layout-valign="bottom"}
+
+![Data analysis Goal](forestwalk/fig1.jpg){#fig-forest1}
+
+![The typical modelling process](forestwalk/fig2.jpg){#fig-forest2}
+
+![Making zero assumption at every stage](forestwalk/fig3.jpg){#fig-forest3}
+
+![An example of "scenario uncertainty" with a confidence interval](forestwalk/fig4.jpg){#fig-forest4}
+
+An analogy of walking through the forest to help understand how our conceptualization of uncertainty and bias can influence what we consider to be uncertainty and what we consider to be bias. 
+:::
+
+This analogy may seem silly but it is explaining a very real problem that exists in uncertainty communication.
+The bias and variance trade off is commonly discussed in modelling, but rarely mentioned when we discuss communicating uncertainty.
+The reality is that every point in our data analysis, is full of decisions.
+Every time we make a decision at some point in this pipeline, we are impacting this bias and variance trade off.
+We have swapped out the variance from a large range of choices for the bias that may arise from an incorrect decision.
+Validation methods (such as cross validation, assumption checks, etc) help to guide the choices we make in the data analysis pipeline, but they can never offer any kind of certainty or truth considering validation methods themselves are also only estimates.
+When we consider the "uncertainty" of a model we usually only consider the variance of the final stage, the prediction error for a specific model and a specific set of features, and ignore the variance (that has been converted into bias by our selections) that was introduced at earlier stages (e.g.
+such as in data collection, when cleaning the data, or when considering alternative models).
+This is *well known* in modelling, it is why big tech companies try to collect as much data as they can and then put it in a neural network and hope they have enough data and enough feedback to avoid the this bias at *all* stages of the analysis.
+Recording *everything* means we don't even have to put any poles down at the data collection stage.
+Sometimes these statements are couched in a single assumption, such as when a time series model's confidence interval is constructed "assuming previous trends will continue into the future".
+This trade off issue means the uncertainty we typically estimate, in the form of a confidence inteval or traces or whatever you choose, has confidence intervals that are too small for what you actually consider the goal of the analysis to be.
+Unfortunately factors that typically decrease the variance, such as sample size, can only have an effect at the levels where that information is used to inform the choices.
+So confidence intervals that are incredibly small but built on a biased sample will be misleading and will be too small to *actually* predict future data.
+It is not uncommon for true results or models produced by a different team to sit entirely outside of the confidence band of another.
+This might be obvious when it is read, but the conversation about whether or not we should consider what was previously variance, to be variance when we communicate our analysis, is something that is rarely discussed although it is important.
+
+Therefore, we must ask ourselves, *why* are we visualsing uncertainty?
+Uncertainty is only thought of as the final stages because to give a point prediction for a random process feels silly.
+By hiding the selections we made at earlier stages, when we traded that variance in for bias, we functionally made *point predictions* but it was just invisible to us.
+
 ## 4.?? What are you uncertain about? A mathematical considerations
@@
 "What is the noise around the estimate of the value of X"
+- "cannot assess the quality of risk communication unless the objectives are clear" circles back to motivation. He assumes we are fulfilling a duty to inform. @Spiegelhalter2017
+
 ## 4.1?? In other statistics

### 2024-02-07 paper.qmd
added aside comments in paper, also explanation for mathematical definition of uncertainty, also updated gitignore

@@
 # 1. Introduction
+::: aside
+I have not touched this section since the confirmation so it would be best to ignore it
+:::
+
 From entertainment choices to news articles to insurance plans, the modern citizen is so over run with information in every aspect of their life it can be overwhelming.
@@
 # 2. Defining Uncertainty
+::: aside
+I have thought more about the idea of defining uncertainty mathematically and I actually *DO* think it is possible.
+I want the record to show the reason I *did not* do this earlier is that *nobody* in the uncertainty literature has defined uncertainty mathematically, and I thought I was supposed to highlight problems in the literature review, not solve them.
+I have had an idea of how to formalise uncertainty after seeing Xiao-li Meng's work, but an uncertainty formalisation would solve *several* open questions he has postulated in previous papers, and would be so new it would need to be a paper in of itself, not as a sub-section in a literature review.
+A mathematical definition would be *completely* new and *very complicated*, and I think its difficulty is pretty clear by the fact that *nobody* and *I really mean nobody*, not spiegelhatter or wilkinson or mceachren or any statistician working in this space, has provided an overarching mathematical definition.
+This is work that would take a long time (and it is the reason I started reading philosophy books because I was actually trying to do this) and even if I did do it, it *should not* be in this paper.
+I think it might be possible based on the reading I have been doing over the past year to try and formalise this work, *but it is not a quick thing that can just be added to this paper*.
+The reason I wanted to do a summary of the papers using the larger taxonomy of uncertainty was because it is encompasing of all the taxonomies, and it allows me to show  that there are *no* formal rules to the uncertainty that is used in uncertainty communication and visualisation.
+Here I am trying to make the point that the lack of a formal definition or of mathematical considerations *has created problems in the literature*.
+To clearly and mathematically define uncertainty will *be solving this problem*.
+Something I was told *not to do* because this is a literature review.
+If a mathematical definition that we are all happy with *needs* to be in this paper, then there is a reasonable chance this paper will never be finished.
+:::
+
 Uncertainty visualisation is not made any easier by the fact that the term "uncertainty" lacks a commonly accepted definition in the literature.
@@
 ## 2.1 Coloquial Definitions
+::: aside
+I am not sure how to collect these together and show that they are highly varied.
+This might be a situation for the tidytext clustering or something.
+There is an additional problem is that the definition for uncertainty is *completely* unrelated to the uncertainty that is quantified in the graphics.
+I think the tidytex can help with the definitions being all over the place, but I am not sure how to identify the lack of correlation between the colloquial definitions of uncertainty and the mathematical quantificaitons.
+:::
+
 The most common definition of uncertainty used in papers that discuss visualising it are colloquial.
@@
 ### 2.2.1 Taxonomies of Uncertainty
+::: aside
+I actually think I agree with the sentiment in the grammar of graphics that a taxonomy is more of an opinion and we can reclassify it and is therefore not thorough enough to remain unchanged and be build upon.
+That being said, if this is a literature review and all existing literature use written definitions or taxnomies, that is what I have to work with.
+The reason I selected the larger taxonomy is because the other taxonomies all *fit within it* and it can be used to define a space that shows the definitions are inconsistent.
+:::
+
 While colloquial definitions of uncertainty are useful when trying to communicate with laypeople who don't understand strict statistical definitions, these definitions are not helpful when we are trying to work out exactly what should be shown on a plot.
@@
 ![Example of different uncertainty taxonomies](paperscreenshots/collage1.jpg){#fig-location}
+
+::: aside
+This figure is not going to be in the final version of the paper.
+It is here as a placeholder.
+Ideally, I would replace with with a graphic that shows where all these taxonomies land in the location, level, nature taxonomy (since it is encompassing) and show they are very inconsistent and a new definition is pulled out for almost every paper (even in statistics papers)
+:::
 
@@
 ](taxonomyvis.jpeg){#fig-taxonomy}
+
+::: aside
+I had this here to explain why these are the considerations people make in statistics and how they relate to elements of the taxonomy to try and provide some justification, but If I am just going to show the definitions are inconsitent, I might get rid of this section.
+I also largely included it so you guys can see what the taxonomies are considering
+:::
 
@@
 
+#### Context comments
+::: aside
+I have this here to explain why context based taxonomies can be ignored when we are considering uncertainty.
+I did not go into too much detail because it would involve discussing visualisation which I am trying to avoid at this point.
+I actually think this quote from the grammar of graphics is exactly what I was trying to say.
+*"The system in this book is not a model for a GIS, however, because geography and statistics differ in a crucial respect.
+Geography is anchored in real space-time and statistics in abstract dimensions.
+This is a distinction along a continuum rather than a sharp break; after all, there is a whole field called spatial statistics (Cressie, 1991).
+But this difference in focus clearly means that a system optimized to handle geography will not be graceful when dealing with statistical graphics"*.
+Basically, my point is this section is still discussing the definitions of unertainty, we have not gotten to disussing visualisation so we are still in the abstract *statistics* domain.
+The only parts of spatial statistics that we care about in this domain are the statistical properties that come from spatial information, and *none* of those statistical properties generate uncertainty.
+This is not unique to the properties of spatial information, other statistics information that is important for modelling and assumptions, such as non-linearity, class levels, etc do not contribute to uncertainty.
+Operations on data, inherrent randomness, and missing information generate uncertainty, which is what we are talking about here.
+A confidence interval is a confidence interval whether it came from spatial data, temporal, or cross sectional data.
+*This does not mean I am saying spatial and temporal information is not valuable ever*, I am just saying it is not valuable *yet*.
+This information will be *highly relevant* when we discuss context and translating from the abstract domain back into real world for communication.
+Since we are still in the abstract domain, things that are anchored in real space-time create confusion that is evident in the literature.
+The confusion that is created in the literature was going to be the topic I had later.
+Spatial and temporal uncertainty papers limit how they communicate their uncertainty because they are considering what can be visualised, not what they should *try* to visualise based on the uncertainty abstraction.
+Since the context is only relevant *with respect to uncertainty* when we visualise it, I was going to discuss this issue in the visualisation literature review section.
+:::
+
 The taxonomies that cannot be contained within the @utypo taxonomy are not a new uncertainty dimensions, but rather they are conflating adjacent information with uncertainty.
@@
 
+#### Maths comments I might relate back to an actual mathematical definition
+::: aside
+ In the "semantics" section of the book, Leeland Wilkinson provides a series of definitions of words related to uncertainty and mathematical definitions of examples of uncertainty, but those are not mathematical definitions of uncertainty.
+Every *mathematical* definition of uncertainty is a definition of *probability* OR a definition of an *example* of uncertainty, such as a confidence interval or PDF.
+I included this here, however, I actually now believe there actually is a mathematical definition, but it would not fit in this paper.
+If this paper remains a *here are the discussions of uncertainty and there is no mathematical definition* I might leave this here as an explanation as to why, but not suggest that there *isnt* a mathematical definition.
+:::
+
 It is important to note that this taxonomy is the closest we can get to mathematical definition of uncertainty.
@@
 Instead or establishing a new definition of uncertainty that is strict enough to prevent the current array of misunderstandings, we should focus on inspecting how the current literature *deviates* from established mathematical definitions, why these deviations occur, and understand how these deviations can lead to larger problems in uncertainty communication.
+
+#### An explanation of the building blocks for my mathematical definition
+::: aside
+I wrote this section between me deciding it would be possible to mathematically define uncertainty and me deciding it would be too large for this paper.
+I am aware that this is not a formal definition and I cannot say this is the unifying feature of uncertainty without some proof.
+But, this is something that *can be* defined in mathematical terms and *can* lead to an encompassing definition of uncertainty (i.e.
+it would be something that the groups of confidence intervals, error bars, and missing data using an information filtration) but that is beyond the scope of this paper.
+This way of thinking about uncertainty has resolved a lot of my issues with the field and a lot of other issues that have been raised by authors in their summary of the literature.
+This is very similar to Xiao-li Meng's work that asserts that bias and variance are actually the same thing, with a couple of changes.
+:::
+
+The first problem in the conception and attempts to offer a definition of uncertainty, is that bias *and* variance are both technically uncertainty.
+
+Basically the idea is that every analysis has some goal, and every choice we make that cuts out several alternatives is a "restriction" that trades the variance of multiple options into a potential bias.
+This occurs at *every stage* of the data analysis pipeline, from collecting data to communication.
+Bias comes from "incorrect information" being passed along the pipeline, while variance comes from "holding on to information".
+The randomness that is "irriduciable error" (if it exists or not is not relevant to our definition) cannot be differentiated from the bias and variance (I say both bias and variance because at the end of the pipeline it is impossible to know if error is being caused by bias or variance) unless we literally know the function that generated the data.
+We typically talk about bias and variance in terms of the model, but it exists at every stage of the pipeline.
+Since every step of the analysis pipeline has an almost infinite set of choices (there is an infinite set of things we can observe, an infinite number of choices that can be made in data cleaning, etc) each of these levels is functionally its *own* random variable.
+When you make your choice you completely shrink the variance of that level to 0 (it basically goes from a distribution of choices to a single outcome of that distribution) and, the amount this *guess* is wrong by, is bias.
+
+This idea is very similar to the bias-variance trade off, but it is slightly different because we are assuming there is not a trade-off perse, but that bias is just "cashed in" variance.
+If selections at each stage are made using feedback, we can reduce the bias without increasing the variance, but it "costs" some data.
+Assumptions let you reduce the variance with a low data "cost" but there is a higher chance of being incorrect (since there is less validation) and introducing bias.
+Flexible models like neural networks can reduce the variance but it comes at a high data "cost".
+If you have a *lot* of data it is better to avoid as many assumptions as possible because increased certainty that comes from more data can only effect error that is still in the form of variance (i.e.
+we are considering alternatives).
+This is well established in modelling, it is why big tech companies try to collect as much data as they can and then put it in a neural network and hope they have enough data and enough feedback to avoid the bias at *all* stages of the analysis.
+Recording *everything* and putting it in the most flexible modle possible means we don't even have to put any poles down, even at the data collection stage
+
+Validation methods (such as cross validation, assumption checks, etc) help to guide the choices we make in the data analysis pipeline, but they can never offer any kind of certainty or truth considering validation methods themselves are also only estimates.
+Once a variance has been cashed in for bias, its error cannot be reduced by more data.
+
+When we consider the "uncertainty" of a model we usually only consider the variance of the final stage, the prediction error for *a specific model and a specific set of features*, and ignore the variance that has been converted into bias by our selections that was introduced at earlier stages of the pipeline.
+Since uncertainty is basically "how true is this estimate or statement", if we had not cashed in our variance for a single outcome of bias, it would appear as variance.
+The climate scenario uncertainty plots are an example of bias that was converted back to variance for the sake of the visualisation (time series forecasts typically assume away the bias by stating "we are assuming previous trends will continue into the future").
+Prediction error is a useful metric because it allows us to see if the bias introduced in earlier stages expressed as variance through an error, however it still cannot account for bias introduced at the sampling stage as that bias also exists in the test set.
+
+Prediction error is not always feasible or useful in a statistical analysis (such as when we produce a visualisation) so it is unclear exactly *how* we should reintroduce that bias as variance to communicate the pair as uncertainty.
+It is not uncommon for uncertainty to only be considered at the final stage of an analysis as a confidence interval or estimate range, but this is typically misleading as earlier bias has not been converted back to variance.
+Factors that typically decrease the variance, such as sample size, can only have an effect on something that is still in a state of variance.
+This means that an established model that is build on more and more data will have confidence intervals that are incredibly small but if it is build on some bias (such as a biased sample) the confidence intervals will become misleading and will be too small to *actually* predict future data.
+It is not uncommon for true results or models produced by a different teams to sit entirely outside of the confidence band of another.
+By hiding the selections we made at earlier stages, when we traded that variance in for bias, we functionally made *point predictions* (they were based on educated guess but point predictions none the less) but *these* point predictions are often invisible to us.
+
+#### A Forest analogy
+Let us consider the data analysis pipeline a dense forest that you need to cross to get to your house.
+To prevent yourself from getting *too* lost you decide to bring some poles and rope so you can mark where you are going.
+You know your rope is long enough to get you to your house with enough wiggle room that you can reach the 5 houses either side of your house.
+This problem is drawn in @fig-forest1.
+@fig-forest2 depicts the typical data analysis pipeline and resulting uncertainty using this forest analogy.
+In this method, we navigate the forest by putting one of our poles down every time we notice the trees have changed (i.e.
+we have move to a new section of the forest).
+As we move down the line and we put our poles down, we cannot be certain we are always walking in the direction of our house, but we can do small checks or validation to make sure we are vaguely heading in the right direction.
+When we get to the final stage, we have enough rope remaining to get to the three houses directly in front of us, and we state that band as our "uncertainty".
+@fig-forest3 depicts the bias variance trade off in the case where we make no decisions, record as much information as possible, and check every single model with every single variable and keep every alternative as a possible option.
+This method will result in no bias (although it is often impossible to check the entirety of these spaces so it cannot) but a huge amount of variance.
+In an effort to avoid putting down any poles, we have managed to navigate the forest, but have essentially no idea which of the 10 houses are ours.
+Of course, with sufficient feedback we can whittle down this vast array of choices (which is what a neural network can do) but feedback from an increased sample size can only influence variance, it cannot influence bias that was previously variance.
+@fig-forest4 shows an alternative way to perform the data analysis process to consider variance at stages beyond model uncertainty (such as the scenario uncertainty that comes with different human choices in climate change models).
+In this scenario you navigate the forest to a certain point, and then clone yourself three times and each put down a new pole at different distances to each other.
+All four versions of yourself will then walk in an identical fashion from the pole to the end of the forest, and each have your own path with your own confidence interval.
+This is how the [climate scenario uncertainty](https://climatechange.chicago.gov/sites/production/files/2016-07/scenariotempgraph_0.jpg) is constructed.
+
+::: {#fig-forest layout="[[59,-2, 39], [59,-2, 39]]" layout-valign="bottom"}
+
+![Data analysis Goal](forestwalk/fig1.jpg){#fig-forest1}
+
+![The typical modelling process](forestwalk/fig2.jpg){#fig-forest2}
+
+![Making zero assumption at every stage](forestwalk/fig3.jpg){#fig-forest3}
+
+![An example of "scenario uncertainty" with a confidence interval](forestwalk/fig4.jpg){#fig-forest4}
+
+An analogy of walking through the forest to help understand how our conceptualization of uncertainty and bias can influence what we consider to be uncertainty and what we consider to be bias. 
+:::
+
+#### The influence of the "goal" of an analysis
+Prediction error is also useful because the *goal* of our analysis is clearly defined in a metric.
+This is not always the case.
+When you attempt to quantify or visualise it you will quickly find yourself asking, "uncertainty about...
+what?".
+Do you mean uncertainty on an estimate?
+On a forecast?
+How many steps ahead is this forecast?
+Are we only considering the uncertainty in the estimate or in the parameters or are we considering the possibility of measurement error or biased inputs?
+These questions start to highlight another invisible problem in uncertainty definitions, signal and noise can only be untangled in the presence of a motivating question.
+
+The idea that uncertainty can only be defined in the presence of a motivating question is well grounded in most areas of statistics.
+The entire process of data analysis, from deciding what should be observed as data through to communicating that data in a plot is governed by human decision and the goal of an analysis.
+At the philosophical level, applied statistics is simply taking real world entities and boiling them down into probabilistic objects, an ontological process that is largely dependent on our goals [@Otsuka2023].
+When we move onto data provenance the issue persists, as what is kept as data and what is tossed away is determined by the motivation of an analysis and what was previously noise can be shown to become signal depending on the resolution of the question [@Meng2014].
+After moving onto modelling this issue continues as each research question can be can be categorised as descriptive, predictive, or causal, each of which has its own appropriate statistical methods and motivation agnostic model selection leads to signal devoid of meaning [@Carlin2023].
+Even at the final stages of visualisation a lack of understanding of the motivating question make it difficult to untangle what is signal and what is noise, leaving many uncertainty visualisation studies with conflicting results [@Kinkeldey2014].
+These cases highlight that uncertainty is defined at *every* stage in relation to our motivating question.
+Discussions of uncertainty cannot be had if we are not clear *what* we are uncertain about.
+Once it is established what we are uncertainty about, we can consider the other elements of uncertainty that need to be defined.
+
+This concept is a little harder to define mathematically, but we *can* consider that every analysis has some metric that is the target "goal" of the analysis.
+Refusing to define this target goal from the begining does not mean you avoid this problem, it just increases the bias and variance issues described earlier, because there is no way to check assumptions or correct the variance with data.
+This is actually why I was thinking about the concept of *suffient visual statistics*.
+If a graphic has a piece of information mapped to it (which we should know because we made the graphic ourselves) then that information can be extracted with a percecptual task and the graph has a *sufficient visual statistic*.
+I also think the idea of sufficient statistics is useful because it establishes that we can only know if we have extracted all the relevant information from our *if a metric we are collecting data for has been defined*.
+If a piece of information cannot be extracted with a perceptual task, it can only be found with assumptions and we introduce a bias at the communication stage.
+If the information does exist but it reuires a heuristic or some collection of perceptual tasks that might be harder for some people than others, we have maintained variance at the communication stage.
+A complicated visualisation with a *huge* amount of information can technically be used to answer a lot of questions, but the mental overload would create variance.
+This is why it is important to have your *goal* in mind when designing a visualisation and not assume it can or should be able to do everything.
+I suspect we can work out retroactively which visualisation contain which visual statistics, by systematically working our way through the grammar of graphics and then information that is contained in simple perceptual tasks.
+
+Even if this cannot be used to give a mathematical definition of uncertainty, it is still soemthing that *should* be considered.
+This might be obvious when it is read, but the conversation about whether or not we should consider what was previously variance to be variance when we communicate our analysis, or if avoiding defining a *goal* metric can have detrimental effects on our communication are things that are rarely discussed although it is important.
+
+::: aside
+Di said this has already been shown and I believe it, but then I don't understand why the uncertainty visualisation literature seems to ignore it.
+This is what I was trying to show with the plot and question section of the excel spreadsheet.
+I was trying to highlight that when you ask participants to answer a question, to mark them as correct or incorrect you need to define a metric, and if the plot was created using a pipeline that *ignores* that metric, it will be passing on incorrect information and lead to mistakes when it is being read.
+Expecting any plot to be able to answer any question that has the basis of some random metric simply because the metric is related to uncertainty and so is the plot seems ridiculous to me, but its harder to find a paper that *doesn't* do this than a paper that *does*.
+:::
 
@@
 # 3.General Attitude Towards Uncertainty (Uncertainty Communication Boradly)
+::: aside
+This section is supposed to be an explanation of the general literature, but the definition problem makes it hard to sort the general literature.
+While there is that overview from spiegelhatter, it is mostly a discussion on risk communication, communicating ambiguous information is not discussed much.
+The focus on risk (uncertainty as signal) and not uncertainty (uncertainty as noise) is an issue, but the bigger problem is caused by the vagueness of uncertainty issue.
+Since a paper will be titled "uncertainty communication" it could be any number of things mathematically, but 90% of the time is it just communicating risks or uncertainty as signal (and very rarely uncertainty as noise).
+I can only tell which one a paper is *once I am in the method section* and even then it is sometimes not explained.
+Also, since I haven't been using this metric since the beginning I need to go back and check this for all the papers.
+I also think this section would benefit from the table, I can quantify how many papers communicate uncertaitny as signal, consider risk-aversion and truth but dont mention transparency, etc.
+I think the only way to back up a lot of my general comments is with these tables that show how widespread these issues are.
+This process is tedious but because definitions for key words are redefined within every paper, I dont think a clustering method will help
+:::
+
 It would be impossible to discuss uncertainty visualisation without mentioning the broader conversation around uncertainty communication.
@@
 For this reason, risk is often seen as a specific case of uncertainty and that well defined sub area has a lot of detailled research.
-Some markers of probability that have common uses, probabilities such as 0 (impossible), 0.5 (a fair coin flip), 1 (certain) are easy for people to have an internal sense of [@Spiegelhalter2017].
+Some markers of probability that have common uses, probabilities such as 0 (impossible), 0.5 (a fair coin flip), 1 (certain) are easy for people to have an internal sense of *hullman paper*.
 It is incredibly hard to communicate small risks, because people cannot differentiate between 1/100 an 1/1000, and communicating low probability but high impact risks are particularly challenging [@Spiegelhalter2017].
@@
 this will not all be from one paper, dont worry.*
+
+::: aside
+This section still needs some citations from papers that communicate uncertainty, but the are *hard* to organise because of the ambiguous uncertainty definition problem.
+So they will all be communicating something different but you cant find out what it is until you are knee deep in the method section.
+:::
 
@@
 
+What is meant by "uncertainty" may seem obvious to some, but when you attempt to quantify or visualise it you will quickly find yourself asking, "uncertainty about...
+what?".
+Do you mean uncertainty on an estimate?
+On a forecast?
+How many steps ahead is this forecast?
+Are we only considering the uncertainty in the estimate or in the parameters or are we considering the possibility of measurement error or biased inputs?
+Signal and noise can only be untangled in the presence of a motivating question.
+
+The idea that uncertainty can only be defined in the presence of a motivating question is well grounded in most areas of statistics.
+The entire process of data analysis, from deciding what should be observed as data through to communicating that data in a plot is governed by human decision and the goal of an analysis.
+At the philosophical level, applied statistics is simply taking real world entities and boiling them down into probabilistic objects, an ontological process that is largely dependent on our goals [@Otsuka2023].
+When we move onto data provenance the issue persists, as what is kept as data and what is tossed away is determined by the motivation of an analysis and what was previously noise can be shown to become signal depending on the resolution of the question [@Meng2014].
+After moving onto modelling this issue continues as each research question can be can be categorised as descriptive, predictive, or causal, each of which has its own appropriate statistical methods and motivation agnostic model selection leads to signal devoid of meaning [@Carlin2023].
+Even at the final stages of visualisation a lack of understanding of the motivating question make it difficult to untangle what is signal and what is noise, leaving many uncertainty visualisation studies with conflicting results [@Kinkeldey2014].
+These cases highlight that uncertainty is defined at *every* stage in relation to our motivating question, from data collection to visualisation.
+Discussions of uncertainty cannot be had if we are not clear *what* we are uncertain about.
+Once it is established what we are uncertainty about, we can consider the other elements of uncertainty that need to be defined.
+
 # 4 Uncertainty Visualisation Literature Summary
@@
 ## 4.1 Similar work
+
+::: aside
+Again, this is why I had that table, because these papers are impossible to organise without some metrics.
+Things that are highly important for uncertainty as I understand it, are not mentioned at all or only mentioned knee deep in the method, so reading and organising these papers such that the information can be combined is a bit of a nightmare.
+If I *dont* find this information uncertainty, I will end up with the same problem as *all* the other uncertainty visualisation literature reviews, which is that the information is *all conflicting and it is impossible to identify any results*.
+The other literature reviews *clearly state this*.
+:::
+
 There are a wealth of uncertainty communication and visualisation literature reviews that already exist.
@@
 
-### A navigating a forest analogy
-When communicating uncertainty information, it is important to ask if we should consider the bias to be variance.
-We often talk about the "bias variance trade off" but we rarely discuss bias as "cashed in" variance, which in some sense, it is.
-If you are struggling to understand how bias is just traded off variance, let us walk through an analogy.
-
-Let us consider the data analysis pipeline a dense forest that you need to cross to get to your house.
-To prevent yourself from getting *too* lost you decide to bring some poles and rope so you can mark where you are going.
-You know your rope is long enough to get you to your house with enough wiggle room that you can reach the 5 houses either side of your house.
-This problem is drawn in @fig-forest1.
-@fig-forest2 depicts the typical data analysis pipeline and resulting uncertainty using this forest analogy.
-In this method, we navigate the forest by putting one of our poles down every time we notice the trees have changed (i.e.
-we have move to a new section of the forest).
-As we move down the line and we put our poles down, we cannot be certain we are always walking in the direction of our house, but we can do small checks or validation to make sure we are vaguely heading in the right direction.
-When we get to the final stage, we have enough rope remaining to get to the three houses directly in front of us, and we state that band as our "uncertainty".
-@fig-forest3 depicts the bias variance trade off in the case where we make no decisions, record as much information as possible, and check every single model with every single variable and keep every alternative as a possible option.
-This method will result in no bias (although it is often impossible to check the entirety of these spaces so it cannot) but a huge amount of variance.
-In an effort to avoid putting down any poles, we have managed to navigate the forest, but have essentially no idea which of the 10 houses are ours.
-Of course, with sufficient feedback we can whittle down this vast array of choices (which is what a neural network can do) but feedback from an increased sample size can only influence variance, it cannot influence bias that was previously variance.
-@fig-forest4 shows an alternative way to perform the data analysis process to consider variance at stages beyond model uncertainty (such as the scenario uncertainty that comes with different human choices in climate change models).
-In this scenario you navigate the forest to a certain point, and then clone yourself three times and each put down a new pole at different distances to each other.
-All four versions of yourself will then walk in an identical fashion from the pole to the end of the forest, and each have your own path with your own confidence interval.
-This is how the [climate scenario uncertainty](https://climatechange.chicago.gov/sites/production/files/2016-07/scenariotempgraph_0.jpg) is constructed.
-
-::: {#fig-forest layout="[[59,-2, 39], [59,-2, 39]]" layout-valign="bottom"}
-
-![Data analysis Goal](forestwalk/fig1.jpg){#fig-forest1}
-
-![The typical modelling process](forestwalk/fig2.jpg){#fig-forest2}
-
-![Making zero assumption at every stage](forestwalk/fig3.jpg){#fig-forest3}
-
-![An example of "scenario uncertainty" with a confidence interval](forestwalk/fig4.jpg){#fig-forest4}
-
-An analogy of walking through the forest to help understand how our conceptualization of uncertainty and bias can influence what we consider to be uncertainty and what we consider to be bias. 
-:::
-
-This analogy may seem silly but it is explaining a very real problem that exists in uncertainty communication.
-The bias and variance trade off is commonly discussed in modelling, but rarely mentioned when we discuss communicating uncertainty.
-The reality is that every point in our data analysis, is full of decisions.
-Every time we make a decision at some point in this pipeline, we are impacting this bias and variance trade off.
-We have swapped out the variance from a large range of choices for the bias that may arise from an incorrect decision.
-Validation methods (such as cross validation, assumption checks, etc) help to guide the choices we make in the data analysis pipeline, but they can never offer any kind of certainty or truth considering validation methods themselves are also only estimates.
-When we consider the "uncertainty" of a model we usually only consider the variance of the final stage, the prediction error for a specific model and a specific set of features, and ignore the variance (that has been converted into bias by our selections) that was introduced at earlier stages (e.g.
-such as in data collection, when cleaning the data, or when considering alternative models).
-This is *well known* in modelling, it is why big tech companies try to collect as much data as they can and then put it in a neural network and hope they have enough data and enough feedback to avoid the this bias at *all* stages of the analysis.
-Recording *everything* means we don't even have to put any poles down at the data collection stage.
-Sometimes these statements are couched in a single assumption, such as when a time series model's confidence interval is constructed "assuming previous trends will continue into the future".
-This trade off issue means the uncertainty we typically estimate, in the form of a confidence inteval or traces or whatever you choose, has confidence intervals that are too small for what you actually consider the goal of the analysis to be.
-Unfortunately factors that typically decrease the variance, such as sample size, can only have an effect at the levels where that information is used to inform the choices.
-So confidence intervals that are incredibly small but built on a biased sample will be misleading and will be too small to *actually* predict future data.
-It is not uncommon for true results or models produced by a different team to sit entirely outside of the confidence band of another.
-This might be obvious when it is read, but the conversation about whether or not we should consider what was previously variance, to be variance when we communicate our analysis, is something that is rarely discussed although it is important.
-
-Therefore, we must ask ourselves, *why* are we visualsing uncertainty?
-Uncertainty is only thought of as the final stages because to give a point prediction for a random process feels silly.
-By hiding the selections we made at earlier stages, when we traded that variance in for bias, we functionally made *point predictions* but it was just invisible to us.
-
 ## 4.?? What are you uncertain about? A mathematical considerations
@@
 - "cannot assess the quality of risk communication unless the objectives are clear" circles back to motivation. He assumes we are fulfilling a duty to inform. @Spiegelhalter2017
-
-## 4.1?? In other statistics
-What is meant by "uncertainty" may seem obvious to some, but when you attempt to quantify or visualise it you will quickly find yourself asking, "uncertainty about...
-what?".
-Do you mean uncertainty on an estimate?
-On a forecast?
-How many steps ahead is this forecast?
-Are we only considering the uncertainty in the estimate or in the parameters or are we considering the possibility of measurement error or biased inputs?
-Signal and noise can only be untangled in the presence of a motivating question.
-
-The idea that uncertainty can only be defined in the presence of a motivating question is well grounded in most areas of statistics.
-The entire process of data analysis, from deciding what should be observed as data through to communicating that data in a plot is governed by human decision and the goal of an analysis.
-At the philosophical level, applied statistics is simply taking real world entities and boiling them down into probabilistic objects, an ontological process that is largely dependent on our goals [@Otsuka2023].
-When we move onto data provenance the issue persists, as what is kept as data and what is tossed away is determined by the motivation of an analysis and what was previously noise can be shown to become signal depending on the resolution of the question [@Meng2014].
-After moving onto modelling this issue continues as each research question can be can be categorised as descriptive, predictive, or causal, each of which has its own appropriate statistical methods and motivation agnostic model selection leads to signal devoid of meaning [@Carlin2023].
-Even at the final stages of visualisation a lack of understanding of the motivating question make it difficult to untangle what is signal and what is noise, leaving many uncertainty visualisation studies with conflicting results [@Kinkeldey2014].
-These cases highlight that uncertainty is defined at *every* stage in relation to our motivating question, from data collection to visualisation.
-Discussions of uncertainty cannot be had if we are not clear *what* we are uncertain about.
-Once it is established what we are uncertainty about, we can consider the other elements of uncertainty that need to be defined.
 

### 2024-02-10 paper.qmd
moved maths definition stuff and context stuff down

--- 2024-02-06 paper.qmd
@@
 ### 2.2.1 Taxonomies of Uncertainty
+::: aside
+I actually think I agree with the sentiment in the grammar of graphics that a taxonomy is more of an opinion and we can reclassify it and is therefore not thorough enough to remain unchanged and be build upon.
+That being said, if this is a literature review and all existing literature use written definitions or taxnomies, that is what I have to work with.
+The larger taxonomy can be used to show that the other taxonomies fit within it, but it is in of itself not tied to anything except work *that paper itself* identifies as transient.
+:::
+
 While colloquial definitions of uncertainty are useful when trying to communicate with laypeople who don't understand strict statistical definitions, these definitions are not helpful when we are trying to work out exactly what should be shown on a plot.
@@
 
-The taxonomies that cannot be contained within the @utypo taxonomy are not a new uncertainty dimensions, but rather they are conflating adjacent information with uncertainty.
-A good example of this is the distinction between spatial, temporal, and estimated uncertainty.
-When we create statistical models and quantify uncertainty, the spatial or temporal aspects are often boilled away [@Otsuka2023], and the only element of it that remains is the inexchangeability of our data.
-For this reason, there realistically is no such thing as "spatial" uncertainty in the same sense that there is no such thing as "photograph" uncertainty or "coin toss" uncertainty or "baby birth" uncertainty.
-Maps and photographs boil down to the same statistical objects in the same way that coin tosses and baby births boil down to the same statistical objects.
-Once we have removed the physical nature these objects are near identical when considered statistically.
-Of course the nature of them can change based on what we choose to observe (i.e.
-if we consider a coin landing on its side to be an outcome or if we consider intersex babies as an outcome in births) but this means the contextual information of our data, e.g.
-whether or not it is spatial or temporal, is only of importance at two stages of the data analysis pipeline.
-It is first important when we *boil it down* to our data through *observation* and again when we *map our statistical information back to its real world information* to *communicate* it.
-The only way the context uniquely contributes to uncertainty is in the "boil down" stage, however spatial or temporal uncertainty never refers to that specific cause of uncertainty.
-It is typically in reference to measurement error, data aggregation, statistical modelling, or other steps that occur *after* the boil down stage, and are therefore not unique to spatial or temporal data.
-While the spatial or temporal features *may* become relevant at the final communication stage, however this has little do with quantifying uncertainty, and the difficulties in communicating a complicated statistical model through a complicated map is already well discussed in the visualisation literature, and in later sections we will discuss how this concept is related to uncertainty visualisation.
-
-In this sense we are not proposing that the current conceptualisation of data according to its physical properties is necessarily wrong, or that it leads to incorrect conclusions, we are merely suggesting for a different way of slicing the metaphysical cake of uncertainty that better aligns with what is *actually happening* when we perform a statistical analysis.
-This metaphysical cake is illustrated below as @fig-cake.
-In changing  what the actual boundaries are when we work with these problems, we improve our understanding of what we are actually doing when we conceptualise uncertainty, and in the process allow for more creative solutions creativity (e.g.
-by borrowing concepts from spatial visualisation to visualise uncertainty in photographs generated).
-However, since uncertainty is almost synonymous with the concept of "statistics", this adjustment in how we conceptualise and organise statistical methods would likely benefit other domains within the field.
-
-![The metaphysical cake we are slicing differently](metaphysicalcake.jpg){#fig-cake}
-
-It is important to note that this taxonomy is the closest we can get to mathematical definition of uncertainty.
-While individual *expressions of uncertainty*, such as a confidence interval or a PDF are mathematically defined, the term *uncertainty* cannot be defined mathematically because it is a layperson word that has been adopted by statistics to describe any deviation from complete determinism, whether mathematically defined or not.
-While the broader field of uncertainty visualisation does need to be more closely inspected, specifically through the scope of mathematical definitions, this issue cannot be solved with a more specific definition of uncertainty through mathematics.
-Instead or establishing a new definition of uncertainty that is strict enough to prevent the current array of misunderstandings, we should focus on inspecting how the current literature *deviates* from established mathematical definitions, why these deviations occur, and understand how these deviations can lead to larger problems in uncertainty communication.
+It is important to note that this taxonomy is thorough but it is not a mathematical definition of uncertainty.
+While individual *expressions of uncertainty*, such as a confidence interval or a PDF are mathematically defined, the term *uncertainty* has not been defined mathematically.
+This is largely because it is a layperson word that has been adopted by statistics to describe any deviation from complete determinism, however this does not mean that a mathematical definition that encompases missing values, alternative models, and prediction intervals *cannot* be found.
+However, that task is far beyond the scope of this work.
+In this work, we will investigate the array of misunderstandings and focus on inspecting how the current literature *deviates* from already established mathematical definitions, why these deviations occur, and understand how these deviations can lead to larger problems in uncertainty communication.
 
@@
 
+::: aside
+I should add something in about the actual quantified uncertainty vs the discussed uncertanty thing (i.e. what is defined with words and what is defined with maths)
+:::
+
 # 3.General Attitude Towards Uncertainty (Uncertainty Communication Boradly)
@@
 ## 4.1 Similar work
+
 There are a wealth of uncertainty communication and visualisation literature reviews that already exist.
@@
 You can project the uncertainty down into one dimension if the source information is in of itself unimportant.
+
+# 4.?? Contextual Inormation
+I have this here to explain why context based taxonomies can be ignored when we are considering uncertainty.
+I did not go into too much detail because it would involve discussing visualisation which I am trying to avoid at this point.
+I actually think this quote from the grammar of graphics is exactly what I was trying to say.
+*"The system in this book is not a model for a GIS, however, because geography and statistics differ in a crucial respect.
+Geography is anchored in real space-time and statistics in abstract dimensions.
+This is a distinction along a continuum rather than a sharp break; after all, there is a whole field called spatial statistics (Cressie, 1991).
+But this difference in focus clearly means that a system optimized to handle geography will not be graceful when dealing with statistical graphics"*.
+Basically, my point is this section is still discussing the definitions of unertainty, we have not gotten to disussing visualisation so we are still in the abstract *statistics* domain.
+The only parts of spatial statistics that we care about in this domain are the statistical properties that come from spatial information, and *none* of those statistical properties generate uncertainty.
+This is not unique to the properties of spatial information, other statistics information that is important for modelling and assumptions, such as non-linearity, class levels, etc do not contribute to uncertainty.
+Operations on data, inherrent randomness, and missing information generate uncertainty, which is what we are talking about here.
+A confidence interval is a confidence interval whether it came from spatial data, temporal, or cross sectional data.
+*This does not mean I am saying spatial and temporal information is not valuable ever*, I am just saying it is not valuable *yet*.
+This information will be *highly relevant* when we discuss context and translating from the abstract domain back into real world for communication.
+Since we are still in the abstract domain, things that are anchored in real space-time create confusion that is evident in the literature.
+The confusion that is created in the literature was going to be the topic I had later.
+Spatial and temporal uncertainty papers limit how they communicate their uncertainty because they are considering what can be visualised, not what they should *try* to visualise based on the uncertainty abstraction.
+Since the context is only relevant *with respect to uncertainty* when we visualise it, I was going to discuss this issue in the visualisation literature review section.
+
+The taxonomies that cannot be contained within the @utypo taxonomy are not a new uncertainty dimensions, but rather they are conflating adjacent information with uncertainty.
+A good example of this is the distinction between spatial, temporal, and estimated uncertainty.
+When we create statistical models and quantify uncertainty, the spatial or temporal aspects are often boilled away [@Otsuka2023], and the only element of it that remains is the inexchangeability of our data.
+For this reason, there realistically is no such thing as "spatial" uncertainty in the same sense that there is no such thing as "photograph" uncertainty or "coin toss" uncertainty or "baby birth" uncertainty.
+Maps and photographs boil down to the same statistical objects in the same way that coin tosses and baby births boil down to the same statistical objects.
+Once we have removed the physical nature these objects are near identical when considered statistically.
+Of course the nature of them can change based on what we choose to observe (i.e.
+if we consider a coin landing on its side to be an outcome or if we consider intersex babies as an outcome in births) but this means the contextual information of our data, e.g.
+whether or not it is spatial or temporal, is only of importance at two stages of the data analysis pipeline.
+It is first important when we *boil it down* to our data through *observation* and again when we *map our statistical information back to its real world information* to *communicate* it.
+The only way the context uniquely contributes to uncertainty is in the "boil down" stage, however spatial or temporal uncertainty never refers to that specific cause of uncertainty.
+It is typically in reference to measurement error, data aggregation, statistical modelling, or other steps that occur *after* the boil down stage, and are therefore not unique to spatial or temporal data.
+While the spatial or temporal features *may* become relevant at the final communication stage, however this has little do with quantifying uncertainty, and the difficulties in communicating a complicated statistical model through a complicated map is already well discussed in the visualisation literature, and in later sections we will discuss how this concept is related to uncertainty visualisation.
+
+In this sense we are not proposing that the current conceptualisation of data according to its physical properties is necessarily wrong, or that it leads to incorrect conclusions, we are merely suggesting for a different way of slicing the metaphysical cake of uncertainty that better aligns with what is *actually happening* when we perform a statistical analysis.
+This metaphysical cake is illustrated below as @fig-cake.
+In changing  what the actual boundaries are when we work with these problems, we improve our understanding of what we are actually doing when we conceptualise uncertainty, and in the process allow for more creative solutions creativity (e.g.
+by borrowing concepts from spatial visualisation to visualise uncertainty in photographs generated).
+However, since uncertainty is almost synonymous with the concept of "statistics", this adjustment in how we conceptualise and organise statistical methods would likely benefit other domains within the field.
+
+![The metaphysical cake we are slicing differently](metaphysicalcake.jpg){#fig-cake}
+
 # 4.?? Importance (and unimportance) of Secondary and Meta information

### 2024-02-12 paper.qmd
finished similar work section and moved it to be beneath the introduction

@@
 
-# 2. Defining Uncertainty
+# 2. Similar work
+There are a wealth of uncertainty communication and visualisation literature reviews that already exist. Interestingly they all seem to have identified similar pitfalls or fallen into one themselves.
+
+@Kinkeldey2014 summarised the literature specifically for geopsatial uncertainty visualisations and attempted to describe similarities in their characteristics.
+They found most experiments the methods for uncertainty visualisation evaluation to be adhoc, with no commonly agreed upon methodology or formalisation.
+They found that experiments took on an engineering approach, where plots were evaluated with the goal to improve a particular representation, rather than to investigate why particular representations do or do not work and create general principals.
+Authors would then ignore the engineering perspective they had taken on, and try to generalise beyond the specific constraints of their experiment.
+While they did discuss the results of some perceptual task experiments they found that even those results were somewhat conflicting and difficult to combine in any meaningful way.
+They also stress the importance of selecting the correct audience, as numerical and visual literacy can differ vaslty between subpopulations so the results from one populations may not reflect the results from another.
+The paper contains an entire paragraph where experiments that compared box-plot representations are rattled off, each experiment finding results that conflicted with its predecessor.
+Ultimately they suggested the visualisation typologies move away from data types, uncertainty categories, and representation types and towards "task-centered typologies" to better enable generalisations of the findings.
+
+@Hullman2016 reviewed the uncertainty visualisation literature more broadly (not specifically for geospatial uncertainty) and offered advice on methodology and analysis of results for other uncertainty visualisation experimenters.
+She found similar issues in the literature and also noticed the somewhat ad-hoc methodology though the lack of justification in the method for the particular questions participants were required to answer, and the lack of generalisability of comparing plots one by one to create a ranking.
+She also discussed the role of heuristics in the analysis of results and suggested authors look for signs of them in responses and ask subjects to describe their strategy in their response.
+What was particularly interesting was people's tendancy to ignore uncertainty in a plot all togehter, and use a pattern in the signal (such as distance between means) to answer questions that require uncertainty information.
+She also discussed the lack of clarity in exactly what "uncertainty" authors are trying to ask about.
+Whether the participants are supposed to answer according to some ground truth or their subjective probability is unclear, as is whether the *uncertainty* depicted represents the uncertainty of a sampling or population distribution.
+Ultimately all these issues become confliated and it becomes impossible to tell if subjects did poorly in an experiment because they misunderstood a visualisation, because the question was misinterpreted, or because they used a heuristic.
+
+@Spiegelhalter2017 presents an literature reivew of uncertainty communication as a whole and considers uncertainty visualisation as a subfield of communication as a whole.
+In this review he has several key findings that are of particular interest.
+First, the best visualisation is highly dependent on your audience, and depending on how broad your audience is, there may not be a best visualisation.
+No single representation will be accessible to all members of an audience, and even if most audience members can read a plot, they may not read it in the same way.
+For example, people with high numeracy skills count pictographs, but people with low numeracy assess area, and seemingly unrelated elements may not equally influence all these methods of reading, for example scattering in a pictograph makes counting particularly more difficult.
+For these reasons, it is better to understand your audience and iterate towards a final design, rather than only consider one option.
+Second, the best visualisation for one task is unlikely to be the best visualisation for another task.
+Graphics are good for "gists" and different graphics are good for different "gists", e.g.
+one plot may we good at identifying trend, while another might be good for comparison.
+Therefore, if we change the goal of our visualisation, and therefore the metric we are trying to evaluate, we implicitly change the best visualisation as no visualisation will both cover and summarise enough information to be the best at answering every question.
+Finally, the visualisation advice he gives is more specifically about general visualisation practice rather than advice specifically for uncertainty visualisation.
+Common sentiments such as avoiding chart junk, considering a table in place of a visualsiaiton, use of narrative labels, managing emotion, less is more approach, etc, are the main concepts expressed.
+The only uncertainty specific advice given is that the graphic should allow "part-to-whole" comparison on an appropriate scale, which seems to be an extension of the framing concept discussed when speaking about uncertainty communication more broadly.
+
+These literature reviews highlight a common theme, the field of uncertainty visualisation is near impossible to navigate.
+There is so much noise in the experimental results that general patterns are difficult to identify and extract for a common purpose.
+This noise is attributed to several things, the results dependence on tasks, the difficulty in understanding the specifics of uncertainty and the conflation of signal and noise from many steps in the experimental design.
+Additionally these reviews often try to look at uncertainty visualisation from a specific perspective (e.g.
+geospatial or bayesian modelling) and find that even within these sub-fields the data is too noise to come to any meaningfull conclusions.
+
+In contrast to these literature reviews, we do not have a specific domain or tool analysis we intend to perform.
+Rather, the point of this literature review is to understand the *purpose* of visualising uncertainty, and understand if the current visualisation tools are sufficient to achieve that purpose and if not, where there is need for improvement.
+Additionally, rather than discussing "uncertainty visualisation" as its own field, we view it as a subfield of both uncertainty and visualisation, and understand if this change in conceptualisation can alleviate or help us to find some signal in the noise of the field.
+Ultimately, this paper should serve as a guide for future uncertaitny visualisation authors to prevent their work from becoming noise in what is already an incredibly noisy field.
+
+# 3. Defining Uncertainty
 Uncertainty visualisation is not made any easier by the fact that the term "uncertainty" lacks a commonly accepted definition in the literature.
@@
 
-## 2.1 Coloquial Definitions
+## 3.1 Coloquial Definitions
 ::: aside
@@
 
-## 2.2 Taxonomy Definitions
-### 2.2.1 Taxonomies of Uncertainty
+## 3.2 Taxonomy Definitions
+### 3.2.1 Taxonomies of Uncertainty
 ::: aside
@@
 
-### 2.2.2 Taxonomies of Visual Uncertainty
+### 3.2.2 Taxonomies of Visual Uncertainty
 There are some taxonomies of visual uncertainty.
@@
 
-# 3.General Attitude Towards Uncertainty (Uncertainty Communication Boradly)
+# 5.General Attitude Towards Uncertainty (Uncertainty Communication Boradly)
 It would be impossible to discuss uncertainty visualisation without mentioning the broader conversation around uncertainty communication.
@@
 
-# 4 Uncertainty Visualisation Literature Summary
-
-## 4.1 Similar work
-
-There are a wealth of uncertainty communication and visualisation literature reviews that already exist.
-
-@Kinkeldey2014 summarised the literature specifically for geopsatial uncertainty visualisations and attempted to describe similarities in their characteristics.
-They found large difficulties in the ability
-
-- methods used in uncertainty visualisation evaluation remain ad hoc.
-studies approach issue from a usability perspective instead of asking WHY representations DO or DO NOT work.
-Studies do not follow any methodology commonly agreed upon.
-@Kinkeldey2014
-- Lack of formalisation and eigour in empirical methdos is an issue that is much broader than uncertainty visualisation (extends to info-vis and related domains).
-Therefore this paper is a starting point, not a definitive summary @Kinkeldey2014
-- The question in these papers is often "does method a work better than method B" which is an engineering approach, where the toal is to improve a particular product rather than create general principals @Kinkeldey2014
-- Authors then try to generalise beyond the specific constraints of the test even though the framing does not allow for it @Kinkeldey2014
-- Alternatively, we should ask how/why does method A work better than method B, which grounds work in perceptual and cognative theory and provides a framwork @Kinkeldey2014
-- Blur and fog are suggested because they are metaphors for uncertainty, however the usefulness of these metaphors is rarely investigated @Kinkeldey2014
-
-- If you just keep using rankings, you need to compare everything one by one forever to get an idea of the space @Hullman2016
-- people sometimes ignore uncertainty and instead look at the pattern (e.g.
-distance between means) and use a heuristic to answer the question.
-be proactive about possible heuristics, look for signes of heuristics in responses, ask subjects to describe their strategy, consider including degree of heuristic as a dependent variable @Hullman2016
+### Extra notes for this section
 - People believe in the law of small numbers but dont believe in the lar of large numbers @Hullman2016
@@
 the graphical one reduces noise but asking people for an interval estimation leads to overprecision and overconfidence @Hullman2016
-- If the goal is to compare visualisation, ground truth is easy, if the goal is to evaluate how accurate the subjective probability distributions are, the ground truth becomes complex.
-ground truth issue e.g.
-if you provide a sample from a distribution, is the true value the mean of that sample or the population mean?
-@Hullman2016
 - believed that offering enough decisions with varying rewards you can observe the true subjective probability, however whether or not probability is the right representation of subjective uncertainty is a topic of debate @Hullman2016
@@
 - evidence that people do not follow the axioms of probability when conveying subjective probabilities. How do you deal with this as an analyst (e.g. if the probabilities do not sum to 1 @Hullman2016
-- If someone answers a question correctly it is hard to tell if it was because of the information provided or if the person was using a heuristic @Hullman2016
-- You cant tell if subjects in an experiment do poorly because the visualisation was bad or because the question was misinterpreted @Hullman2016
 
-@Spiegelhalter2017 presents an literature reivew of uncertainty communication as a whole and considers uncertainty visualisation as a subfield of communication as a whole.
-In this review he has several key findings that are of particular interest.
-First, the best visualisation is highly dependent on your audience, and depending on how broad your audience is, there may not be a best visualisation.
-No single representation will be accessible to all members of an audience, and even if most audience members can read a plot, they may not read it in the same way.
-For example, people with high numeracy skills count pictographs, but people with low numeracy assess area, and seemingly unrelated elements may not equally influence all these methods of reading, for example scattering in a pictograph makes counting particularly more difficult.
-For these reasons, it is better to understand your audience and iterate towards a final design, rather than only consider one option.
-Second, the best visualisation for one task is unlikely to be the best visualisation for another task.
-Graphics are good for "gists" and different graphics are good for different "gists", e.g.
-one plot may we good at identifying trend, while another might be good for comparison @Spiegelhalter2017.
-Therefore, if we change the goal of our visualisation, and therefore the metric,
-
-Visualisations that improve accuracy, visualisations that improve behavioural change, and visualisation that viewers like may not necessarily all be the same visualisation @Spiegelhalter2017.
-
-He highlights a theme that is a common theme in the uncertainty visualisation literature, which is no visualisation will cover enough information, graphics are good for "gists" and different graphics are good at different things (e.g.
-identifying trends vs comparison).
+@Spiegelhalter2017
 we consider a summary table as a visualisation
-
 illuminate graphics with words and numbers, 
 design graphics to allow a part-to-whole comparison on an appropriate scale, 
-include appropriate narrative labels (such as magnitudes through tick marks), 
 use images and metaphors to gain/retain attention but dont arouse undue emotion, 
@@
 
-These literature reviews highlight a common theme,
-In contrast to these literature reviews, we do not have a specific domain or tool analysis we intend to perform.
-Rather, the point of this literature review is to understand the *purpose* of visualising uncertainty (i.e.
-signal supression), and understand if the current visualisation tools are sufficient to achieve that purpose and if not, where there is need for improvement.
-Additionally, rather than discussing "uncertainty visualisation" as its own field, we view it as a subfield of both uncertainty and visualisation, and understand if this change in conceptualisation can alleviate or help us understand some of the issues faced by previous authors attempting to summarise the uncertainty visualisation literature.
+# 4 Uncertainty Visualisation Literature Summary
+The previous sections of this paper established several issues in uncertainty visualisaiton.
+## Types of Uncertainty Visualisation Papers
 
-# 4.1 Plot Comparison Papers
-# 4.1.1 The problem of incomparable plots
+## Uncertainty Visualisation and the Grammar of Graphics
+- Since there are so many papers that treat uncertainty as a signal, not a noise, issues in the uncertainty visualisation literature also help to understand some difficulties in the information visualisation literature as large.
+
+- Suggests that it is straightforward to show a value but much more complex to show uncertainty @Hullman2016 (I think this is interesting in relation to how people SEE uncertainty visualisation.
+Why is uncertainty hard to visualise?)
+- Authors provide little justification for their chosen response models (e.g. absolute accuracy vs relative measures) @Hullman2016
+- When asking questions, researchers should focus on events that can be repeated (because of a misunderstanding of confidence intervals) @Hullman2016
+- Uncertainty defined as a by product of the process of making a visualisation, from data collection to commucating @Hullman2016
+- Some questions (e.g. what are the chances that the no.6 bus will arrive first) you can elicit the viewers subjective probability distribution @Hullman2016
+- communication tasks: This category comprises map reading tasks involving data and uncertainty value retrieval (which location is most uncertain?).
+For tasks from this category, visualisation techniques can be chosen following the traditional rules from cartography.
+@Kinkeldey2014 (basically saying why dont you just use what is already here if you are treating uncertainty as a signal, which I agree with)
+- papers usually have same variance if there is a signal question, and use the mean as a proxy to answer the question.
+Really should ask a changing signal question where the signal IS impacted by the uncertainty information.
+- Weirdly When setting their ground truth it seems like a lot of papers EXPECT participants to completely ignore uncertainty information as noise.
+Like the participants are actually considered to be incorrect if they dampen their signal understanding because of noise considerations which I find FUNNY.
+- I think the questions where uncertainty is a signal, not a noise, allow the authors of visualisation experiments to ignore the context and motivation that usually governs a visualisation.
+
 Visual inference can be seen as a process that combines multiple fields, from mathematics to psychology, to convey meaning.
@@
 
-# 4.?? Noise vs Signal (question +)
+- But when different techniques are to be compared, it is important that the scenarios and datasets are informationally equivalent, i.e.
+according to Larkin and Simon (1987, p.
+67) ‘[t]wo representations are informationally equivalent if all of the information in the one is also inferable from the other, and vice versa’.
+A goal in testing, then, is often to determine whether they are also computationally equivalent, or whether one depiction has an advantage over another, i.e.
+‘[t]wo representations are computationally equivalent if they are informationally equivalent and, in addition, any inference that can be drawn easily and quickly from the information given explicitly in the one can also be drawn easily and quickly from the information given explicitly in the other, and vice versa’ (Larkin and Simon 1987, p.
+67).
+@Kinkeldey2014
+
+- The way we ask questions and the types of questions we ask are selected with little justification. This paper makes suggestions to reduce the noise in the data from these papers. @Hullman2016
+
+- Most commonly check abolute measures of accuracy, where accuracy is abs(subjective error - actual probabiltiy).
+Some papers also check relative measures of accuracy and ask subjects to find regions of least certainty or to rank targets by uncertainty.
+studies also consider response time and confidence rating.
+some studies also ask users for their own expression of uncertainty (e.g.
+by asking when sketches are uncertainty and ranking various depictions of uncertainty).
+Some studies ask subjects to make some decision using the data.
+ometimes things like complexity of task, degree of visual overload, ease of use, visual appeal, and preferences @Hullman2016
+
+## Noise and Signal in Uncertainty Visualisation
+
 The first issue that arises with uncertainty research is that uncertainty is frequently treated as a signal, not noise.
@@
 
-## 4.?? Greater Infoviz Issues
-- Since there are so many papers that treat uncertainty as a signal, not a noise, issues in the uncertainty visualisation literature also help to understand some difficulties in the information visualisation literature as large.
-
-- Suggests that it is straightforward to show a value but much more complex to show uncertainty @Hullman2016 (I think this is interesting in relation to how people SEE uncertainty visualisation.
-Why is uncertainty hard to visualise?)
-- Authors provide little justification for their chosen response models (e.g. absolute accuracy vs relative measures) @Hullman2016
-- When asking questions, researchers should focus on events that can be repeated (because of a misunderstanding of confidence intervals) @Hullman2016
-- Uncertainty defined as a by product of the process of making a visualisation, from data collection to commucating @Hullman2016
-- Some questions (e.g. what are the chances that the no.6 bus will arrive first) you can elicit the viewers subjective probability distribution @Hullman2016
-- communication tasks: This category comprises map reading tasks involving data and uncertainty value retrieval (which location is most uncertain?).
-For tasks from this category, visualisation techniques can be chosen following the traditional rules from cartography.
-@Kinkeldey2014 (basically saying why dont you just use what is already here if you are treating uncertainty as a signal, which I agree with)
-- papers usually have same variance if there is a signal question, and use the mean as a proxy to answer the question.
-Really should ask a changing signal question where the signal IS impacted by the uncertainty information.
-- Weirdly When setting their ground truth it seems like a lot of papers EXPECT participants to completely ignore uncertainty information as noise.
-Like the participants are actually considered to be incorrect if they dampen their signal understanding because of noise considerations which I find FUNNY.
-- I think the questions where uncertainty is a signal, not a noise, allow the authors of visualisation experiments to ignore the context and motivation that usually governs a visualisation.
-
-# 4.?? Tasks and Motivating Questions (uncertainty about.. .what?)
-
-## 4.?? What are you uncertain about? A mathematical considerations
 - Should add plot comparison mathematical framework example here
@@
 
-## 4.1?? General task notes
 - heath had a wide range of different metrics they adopted to communicate chronic risk of adverse events in the future @Spiegelhalter2017 (use this for motivation based statistics since they define new statistics based off what they need)
@@
 - Need to compare to an approripate ground truth, but this is a philisophical exercse @Hullman2016
-## 4.?? Incomparable plots
-- But when different techniques are to be compared, it is important that the scenarios and datasets are informationally equivalent, i.e.
-according to Larkin and Simon (1987, p.
-67) ‘[t]wo representations are informationally equivalent if all of the information in the one is also inferable from the other, and vice versa’.
-A goal in testing, then, is often to determine whether they are also computationally equivalent, or whether one depiction has an advantage over another, i.e.
-‘[t]wo representations are computationally equivalent if they are informationally equivalent and, in addition, any inference that can be drawn easily and quickly from the information given explicitly in the one can also be drawn easily and quickly from the information given explicitly in the other, and vice versa’ (Larkin and Simon 1987, p.
-67).
-@Kinkeldey2014
 
-- The way we ask questions and the types of questions we ask are selected with little justification. This paper makes suggestions to reduce the noise in the data from these papers. @Hullman2016
-
-- Most commonly check abolute measures of accuracy, where accuracy is abs(subjective error - actual probabiltiy).
-Some papers also check relative measures of accuracy and ask subjects to find regions of least certainty or to rank targets by uncertainty.
-studies also consider response time and confidence rating.
-some studies also ask users for their own expression of uncertainty (e.g.
-by asking when sketches are uncertainty and ranking various depictions of uncertainty).
-Some studies ask subjects to make some decision using the data.
-ometimes things like complexity of task, degree of visual overload, ease of use, visual appeal, and preferences @Hullman2016
-# 4.?? Source of Uncertainty
+## An Unexplored Space (Different visualisations for location and stuff)
 - rarely used extension is aggregation of uncertainty over an area (retrieve overall estimation from a spatial distribution of uncertainty).
@@
 
-# 4.?? Contextual Inormation
+## Contextual Inormation
 I have this here to explain why context based taxonomies can be ignored when we are considering uncertainty.
@@
 This is a fine question to ask, but the lack of specificity in what is being done, and the zoning in on functionally irrelevant elements leads the literature as a huge mess.
-
 - Hurricane risk is not only the path, but also the storm surge and wind speed. @Spiegelhalter2017
 - coloured maps for earthquake risk are easily interpretable @Spiegelhalter2017
+
 # 5 Great Examples
@@
   - [This](https://www.esri.com/arcgis-blog/products/arcgis-pro/mapping/graphical-cartograms-in-arcgis-pro/) has some good maps and some bonkers ones 
-
+- Exceedance probability map (example where uncertainty that *should* be conveyed as signal is expressed as a signal. It is a typical maths problem that involves)
 Notes

### 2024-02-12 writingdumpster.qmd
finished similar work section and moved it to be beneath the introduction

@@
 Knowing that I can come back to this at any point and pull information back, makes me more comfortable with removing it from the report.
-# Definition outtakes
-## Random
+
+# Moved over notes
 - mentions uncertainty can be variance, precision, accuracy, reliability, or related concepts @Hullman2016

### 2024-02-13 paper.qmd
updated paper

@@
 
-# 5.General Attitude Towards Uncertainty (Uncertainty Communication Boradly)
+# 4.General Attitude Towards Uncertainty (Uncertainty Communication Boradly)
 It would be impossible to discuss uncertainty visualisation without mentioning the broader conversation around uncertainty communication.
@@
 
-# 4 Uncertainty Visualisation Literature Summary
-The previous sections of this paper established several issues in uncertainty visualisaiton.
+# 5. Uncertainty Visualisation Literature Summary
+The previous sections of this paper established several issues in uncertainty communication at large.
+In this section, we will discuss the current literature on uncertainty visualisation, how the uncertainty communication issues more broadly manifest in the specific case of visualisation, and what general lessons we can take from the existing work.
+
 ## Types of Uncertainty Visualisation Papers
+Evaluation experiments are the standard rule for visualisations because the human brain is not as reliable as mathematical calculation.
+Therefore, user studies often aim to assess the limitations, biases, and heuristics of our mental calculator so that we can better understand the problems we may encounter when we plot our data.
+This is not to say any paper that suggests a visualisation without an evaluation experiment is completely lacking in justification, and there are many papers that suggest a novel visualisation without an evaluation study.
+Sometimes these papers are a preliminary step in finding a solution for common problems and intend to evaluate the visualsiation in later work.
+Until that later work is done, it is often difficult to accept one particular representation.
+While justification could come in the form of discussing established concepts in visualisation, such as the hierarchy of perceptual tasks, even these may need an evaluation study to back up their claims.
+
+The reasoning for this is obvious.
+There are a large number heuristics and biases that are not obvious to us when designing visualisation.
+Additionally, these heuristics and baises can change depending on the larger scope of the graphic and the population we are communicating with.
+Additionally, since there is often a miriad of ways to visualise any particular mathematical object, to adopt specific visualisation that is not already common practice needs reasonable justification, lest we run into a range of unforseen heuristic pitfalls.
+
+There are some papers that suggest a novel plot that does not serve to fix a specific visualisation issue.
+These papers suggest ways to include even *more* information in a visualisation, and shift the information overload from the visualisation author to the viewer.
+Clever methods of combining several types of information, such as VSUPs or tours, can better help us understand how our brains process information in a graphic, but visualisations that simply seek to display as much information as possible fail to understand that visualisations are primarily as a tool to summarise information.
+Poor understanding of our goals in generating a visualisation will not be remedied by overloading the viewer.
+
+The remainder of section 4 will focus on the results and methods of uncertainty visualisation evaluation experiments.
+There are six main types of evaluation experiments in the uncertainty visualisation literature, they are:
+-  Plot Comparisons: These papers describe an uncertainty visualisation evaluation experiment, where they compare two or more uncertainty visualisaitons on a specific set of questions.
+These questions can be about specific statistics (e.g.
+What is the probability that A>B?) or about a secondary goal such as trust or decision making (e.g.
+How much do you trust this estimate?).
+- Perceptual Tasks: These experiments identify perceptual tasks that are/could be associated with uncertainty, and test its association with uncertainty, check it's number of distinguishable levels, and/or check how quickly/accurately people can draw the uncertainty information from the graphic.
+- Heuristic Checks: These experiments are checking if a perceptual bias that has been identified through public use of a particular plot, other evaluation experiments, or gut feeling identified through public use of a plot or in a different
+
+Heuristic checks: do people have this bias when looking at this plot?
+Meta papers: Uncertainty visualisation experiments should actually be done like this.
+Qualatative analysis: Look at this visualisaiton for this purpose, tell us what you are looking at and why.
+Can people get X: Draw/look at this uncertainty viusalisation, what is the mean of this distribution?
+
+Each of these types of papers may have distinct goals in their visualisation, but their problems and conflicting results share a common source.
+Some of these issues come from issues deeper than the visualisation field (e.g.
+the lacking definitions of uncertainty) while some are specific to visualisation (e.g.
+the difficulty in identifying the signal of a graphic).
+Below we will go through the five key issues in uncertainty visualisation, describe how they mainfest in the uncertainty visualisation literature, and offer some tentative solutions to these problems.
+
+We will also comment on some of the uncertainty visualisation papers that suggest a visualisation without an evaluation experiment.
+Some of these plots may offer a solution to the unique problems we mention in this paper and these visualisations will be discussed in "Part 6: Great Examples".
+We will also consider these papers to illustrate the issues that visualisations are aware of and working to fix, and discuss whether or not these directions of research are likely to bear fruit, and why.
+
+- perceptual tasks need to know what you are uncertain about
+- perceptual tasks don't consider
 
 ## Uncertainty Visualisation and the Grammar of Graphics
+The first issue in uncertainty visualisation is there does not seem to be a clear rule for what makes a particular visualisation a depiction of uncertainty.
+
 - Since there are so many papers that treat uncertainty as a signal, not a noise, issues in the uncertainty visualisation literature also help to understand some difficulties in the information visualisation literature as large.
@@
 ## Noise and Signal in Uncertainty Visualisation
-
 The first issue that arises with uncertainty research is that uncertainty is frequently treated as a signal, not noise.
@@
 
+- new papers dont seem to be aware of this problem.
 - if the experiment DOES use noise as a signal, it is often very unclear how the participants should include that in their respose and what would be a valid response from participants (e.g.
@@
 
-# 4.?? Importance (and unimportance) of Secondary and Meta information
 - Specifies uncertainty can be represented by three components attribute (what) position (where) and temporal (when) @Kinkeldey2014

### 2024-02-16 paper.qmd
added summary paper

@@
 The only uncertainty specific advice given is that the graphic should allow "part-to-whole" comparison on an appropriate scale, which seems to be an extension of the framing concept discussed when speaking about uncertainty communication more broadly.
+
+@Griethe2006 provided a summary of the uncertainty visualisation literature and did not take any particular angle in the analysis.
+She makes the point that there is no universally accepted definition of uncertainty and no unicersally accepted way to display it.
+She defined uncertainty as "the degree to which the lack of knowledge about the amount of error is responsible for hesitancy in accepting results and observations without caution".
+She then explains that there are two ways to visualise uncertianty, implicitly and explicitly.
+When a visualisaion is implicit, the uncertainty is considered an uninteresting feature of the data, and instead of mapping some quantification of uncertainty to a visual feature, data is filtered out if it does not meet a specified "certainty" threshold.
+These statements clearly points towards a signal/noise dichotomy with normal visualisation and uncertainty visualisation.
+Additionally she notes that the motivation of the analysis is important, by stating "one has to answer the questions which kinds of uncertainty are relevant from the data’s perspective, which kinds are helpful from the perspective of the visualization steps and which kinds are practicably extractable from the perspective of the available methods?", and "The decision of which of these techniques to choose strongly depends on the intended goal" when commenting on which uncertainty visualisation techniques should be used.
+She clearly sees uncertainty as additional information or data, and therefore sees the uncertainty visualisation problem as a specific case of the high dimensional visualisation issues.
+She mentions that there are detailled classifications of uncertainty but the literature and subsequent techniques are not ballanced, and instead concentrate on scalar values.
+In facing this lack of ballance she creates a new classification system for uncertainty visualisation.
+She commented that uncertainty appears at every step in the analyis pipeline, and commented on some the mathematical definitions that seek to quantify uncertianty, but did not explain how these mathematical definitions can be combined across levels.
+Unfortunately this quantification is needed since she mentioned that in order to visualise uncertainty, it needs to be measurable.
 

### 2024-02-19 paper.qmd
finished types of papers section

@@
 Ultimately, this paper should serve as a guide for future uncertaitny visualisation authors to prevent their work from becoming noise in what is already an incredibly noisy field.
+
+There are several frameworks that the uncertainty visualsiation problem is approached from.
+One approaches it from the concept of risk, and tries to answer how to best express uncertainty in a visualisation such that the viewer can acurately extract and interpret the risks involved.
+Another approach is to consider uncertainty visualsiation as a specific sub-field of high-dimensional data visualsiation, and therefore uncertainty is just another variable that needs to be expressed.
 
@@
 ## Types of Uncertainty Visualisation Papers
+There are two broad categories of uncertainty visualisation experiments, depending on what they believe the "problem" in uncertainty visualisation to be.
+
+The first type of paper believes uncertainty to be fundamentally differently to other variables due to psychological elements of uncertainty, such as risk aversion.
+These papers seek to investigate uncertainty visualisaiton by assessing which visualisation methods facilitate these "risk" related biases and check that perceptual tasks are not impacted by the fact that "uncertainty" is what is being visualised.
+At least this is often how the motivation is described.
+Due to the fact that uncertainty is poorly defined and the psychological effects of uncertainty is never justifiably connected to the scientific communication of it, as was discussed in ssections 3 and 4, these papers are all over the place in terms of methodology.
+Since it is unclear exactly *what* they are testing for, since it is unclear *how* uncertainty information is suppoesdly different from any other information, these papers are poorly justified and attempt to test for the secondary effects of a psychological outcome that has actually never been justified or shown to exist.
+
+The second type of paper perceives uncertainty to be the same as any other variable, and therefore the issue facing uncertainty visualisation is actually a dimensionality problem.
+These experiments evaluate different ways to include even *more* information than would be put in a typical visualisation.
+Some of these methods, such as a bivariate pallet or glyph map, create graphics that are overloaded with information and offer incorrect or meaningless takeaways at first glance.
+This is not to say *every* visualisation method that sees it as a dimensionality problem is bad.
+Clever methods of combining several types of information, such as value supressing uncertainty pallets (VSUPs), can better help us understand how our brains process information in a graphic.
+However this is because visualisations that simply seek to display as much information as possible fail to understand visualisation are a tool to summarise information using visual features, and once it fails to summarise information it is no longer a good visualisation.
+Failling to understand this key aspect of visualsiation leads to graphics that *contain* the correct amount of information, but that volume of information cannot be readily accessed by the viewer.
+
+In addition to the way the papers *see* the uncertainty visualisation problem, these papers can also be organised according to the *goal* of the experiment.
 Evaluation experiments are the standard rule for visualisations because the human brain is not as reliable as mathematical calculation.
@@
 
-There are some papers that suggest a novel plot that does not serve to fix a specific visualisation issue.
-These papers suggest ways to include even *more* information in a visualisation, and shift the information overload from the visualisation author to the viewer.
-Clever methods of combining several types of information, such as VSUPs or tours, can better help us understand how our brains process information in a graphic, but visualisations that simply seek to display as much information as possible fail to understand that visualisations are primarily as a tool to summarise information.
-Poor understanding of our goals in generating a visualisation will not be remedied by overloading the viewer.
-
 The remainder of section 4 will focus on the results and methods of uncertainty visualisation evaluation experiments.
 There are six main types of evaluation experiments in the uncertainty visualisation literature, they are:
+- Association perceptual tasks: These experiments identify perceptual tasks that are/could be associated with uncertainty, and test its association with uncertainty, check it's number of distinguishable levels, and/or check how quickly/accurately people can draw the uncertainty information from the graphic.
+- Value perceptual tasks: Participants are shown some depiction of uncertainty and are asked to extract a value, make a comparison, or do something simple to show that the graphic is readable.  
+- Heuristic Checks: These experiments are checking if a perceptual bias that has been identified through public use of a particular plot, other evaluation experiments, or gut feeling identified through public use of a plot or in a different.
 -  Plot Comparisons: These papers describe an uncertainty visualisation evaluation experiment, where they compare two or more uncertainty visualisaitons on a specific set of questions.
@@
 How much do you trust this estimate?).
-- Perceptual Tasks: These experiments identify perceptual tasks that are/could be associated with uncertainty, and test its association with uncertainty, check it's number of distinguishable levels, and/or check how quickly/accurately people can draw the uncertainty information from the graphic.
-- Heuristic Checks: These experiments are checking if a perceptual bias that has been identified through public use of a particular plot, other evaluation experiments, or gut feeling identified through public use of a plot or in a different
+- Qualitative analysis: These experiments show participants an uncertainty visualisation for an open ended or specific task, and asks them to describe how the visualisation was used to come to a conclusion, or they are asked to describe what they see in the visualisation.
+- Method papers: These papers discuss issues with the evaluation methods of uncertainty visualisaiton and suggest changes or improvements, typically with results indicating the benefit of the change.
 
-Heuristic checks: do people have this bias when looking at this plot?
-Meta papers: Uncertainty visualisation experiments should actually be done like this.
-Qualatative analysis: Look at this visualisaiton for this purpose, tell us what you are looking at and why.
-Can people get X: Draw/look at this uncertainty viusalisation, what is the mean of this distribution?
-
-Each of these types of papers may have distinct goals in their visualisation, but their problems and conflicting results share a common source.
-Some of these issues come from issues deeper than the visualisation field (e.g.
-the lacking definitions of uncertainty) while some are specific to visualisation (e.g.
-the difficulty in identifying the signal of a graphic).
+These experiments are not mutually exclusive within a paper, and a single study will often contain several experiments, usually from several of these categories.
+These distinctions are important because each have their own issues in how they relate to uncertainty more broadly.
 Below we will go through the five key issues in uncertainty visualisation, describe how they mainfest in the uncertainty visualisation literature, and offer some tentative solutions to these problems.
@@
 We will also consider these papers to illustrate the issues that visualisations are aware of and working to fix, and discuss whether or not these directions of research are likely to bear fruit, and why.
-
-- perceptual tasks need to know what you are uncertain about
-- perceptual tasks don't consider
 
@@
   - There are other things that have inexchangeability, and not every question related to spatial or temporal data requires the inexchangeability.
-  - 
+- perceptual tasks need to know what you are uncertain about.
 - Yes the data type is relevant but its not the most important thing

### 2024-02-20 paper.qmd
worked on uncertainty vis and grammar of graphics section

@@
 The first issue in uncertainty visualisation is there does not seem to be a clear rule for what makes a particular visualisation a depiction of uncertainty.
+This problem seems to start at the lack of the mathematical definition of uncertainty which then bleeds into a lack of a definition for an "uncertainty visualisation".
+For example @Leland2005 mentions that popular graphics, such as pie charts and bar charts omit uncertainty, however at least one or both of these charts are used in a large number of uncertainty visualisation experiments.
+How can a chart omit uncertainty according to the grammar of graphics, but be used to describe it in several evaluation studies?
+The answer circles back to the definition problem we discussed earlier.
+
+There are two ways to think of uncertainty in graphics:
+1) A parameter is defined and graphed, and uncertainty is depicted as additional information in that graphic.
+Therefore uncertainty visualisation is a layering a "normal" visualisation with some aesthetic changes that incorperate uncertainty.
+2) Uncertainty is defined (using a PDF, confidence interval, etc) and graphed using whatever method is required.
+The difference between these two conceptualisations stems from the two different ways to percieve the uncertainty visualisation problems.
+Method (1) requires thinking of uncertainty as a dimensionality problem, while (2) sees uncertainty as something that is defined and visualised on its own.
+For example, someone who percieves of uncertainty visualisation as (1) would see a pie chart as a normal graphic, and a pie chart with blurred boundaries as an uncertainty visualisation, while someone who uses method (2) would see a pie chart that depicts probabilities of an event as an uncertainty graphic and blurred boundaries would be pointless.
+
+These two differing philosophies diverge in *where* in the visualisation pipeline they believe uncertainty to be defined.
+Those who are using method (1) believe uncertainty is defined at the geometry or aesthetics stages shown in @fig-pipeline, while (2) believes uncertainty is defined at the statistics stage.
+Unfortunately there is no single person or experiment that is consistent in any one belief.
+Papers will simultanously discuss uncertainty as though a visualisation of variance is an uncertainty visualisation but also assume a bar char of variances is *not* an uncertainty visualisation.
+This difficulty and lack of consistency comes from the desire to discuss "uncertainty visualisation" as though it is a single term and not the combination of "uncertainty" and "visualisation".
+If we split the terms and look at the quantification and visualisation of uncertainty separately, the results of many uncertainty visualisation papers become self evident.
+
+![The grammar of graphics data analysis pipeline](grammarofgraphicspipeline.png){#fig-pipeline}
+
+Regardless of how you see uncertainty visualisations, there are two things that must be done in order to depict uncertainty in a graphic.
+1) Mathematically estimate uncertainty
+2) Visualise uncertainty
+There are numerous ways to depict uncertainty mathematically.
+Examples of mathematically defined objects that are used to express uncertainty inlude: confidence intervals, sampling distributions, bias, variance, prediction error, value ranges, quantiles etc.
+These things can all be specificed, calculated, and set as the ground truth in a visualisation experiment.
+There are also numerous ways to depict uncertainty visually.
+The computed values can be mapped to a position, length, colour, area, etc.
+This process, of quantifying what you are visualising and then attributing it to a visual attribute is not unique to uncertainty visualisation.
+In this sense, uncertainty is exactly the same as any other variable, and if in this sense, uncertainty visualisation is exactly the same as normal visualisation.
+
+There are also numerous ways to something we have mathematically depicted.
 

### 2024-02-21 paper.qmd
lot of additions to the grammar of graphics section

--- 2024-02-19 paper.qmd
@@
 ## Types of Uncertainty Visualisation Papers
-There are two broad categories of uncertainty visualisation experiments, depending on what they believe the "problem" in uncertainty visualisation to be.
-
-The first type of paper believes uncertainty to be fundamentally differently to other variables due to psychological elements of uncertainty, such as risk aversion.
-These papers seek to investigate uncertainty visualisaiton by assessing which visualisation methods facilitate these "risk" related biases and check that perceptual tasks are not impacted by the fact that "uncertainty" is what is being visualised.
-At least this is often how the motivation is described.
-Due to the fact that uncertainty is poorly defined and the psychological effects of uncertainty is never justifiably connected to the scientific communication of it, as was discussed in ssections 3 and 4, these papers are all over the place in terms of methodology.
-Since it is unclear exactly *what* they are testing for, since it is unclear *how* uncertainty information is suppoesdly different from any other information, these papers are poorly justified and attempt to test for the secondary effects of a psychological outcome that has actually never been justified or shown to exist.
-
-The second type of paper perceives uncertainty to be the same as any other variable, and therefore the issue facing uncertainty visualisation is actually a dimensionality problem.
-These experiments evaluate different ways to include even *more* information than would be put in a typical visualisation.
-Some of these methods, such as a bivariate pallet or glyph map, create graphics that are overloaded with information and offer incorrect or meaningless takeaways at first glance.
-This is not to say *every* visualisation method that sees it as a dimensionality problem is bad.
-Clever methods of combining several types of information, such as value supressing uncertainty pallets (VSUPs), can better help us understand how our brains process information in a graphic.
-However this is because visualisations that simply seek to display as much information as possible fail to understand visualisation are a tool to summarise information using visual features, and once it fails to summarise information it is no longer a good visualisation.
-Failling to understand this key aspect of visualsiation leads to graphics that *contain* the correct amount of information, but that volume of information cannot be readily accessed by the viewer.
-
-In addition to the way the papers *see* the uncertainty visualisation problem, these papers can also be organised according to the *goal* of the experiment.
+Uncertainty visualisation papers can be organised according to the *goal* of the experiment.
 Evaluation experiments are the standard rule for visualisations because the human brain is not as reliable as mathematical calculation.
@@
 The first issue in uncertainty visualisation is there does not seem to be a clear rule for what makes a particular visualisation a depiction of uncertainty.
+This problem seems to start at the lack of the mathematical definition of uncertainty which then bleeds into a lack of a definition for an "uncertainty visualisation".
+For example @Leland2005 mentions that popular graphics, such as pie charts and bar charts omit uncertainty, however at least one or both of these charts are used in a large number of uncertainty visualisation experiments.
+How can a chart omit uncertainty according to the grammar of graphics, but be used to describe it in several evaluation studies?
+The answer to this question is related to the definition problem discussed earlier, but to understand how, we will need to go through the grammar of graphics, and its absence in the uncertainty visualisation literature.
 
-- Since there are so many papers that treat uncertainty as a signal, not a noise, issues in the uncertainty visualisation literature also help to understand some difficulties in the information visualisation literature as large.
+When creating a graphic there are several tasks that must be completed in a specific order, regardless of whether or not we are working with a uncertainty visualisation or not.
+@Leland2005 depicts these steps as a pipeline, shown in @fig-pipeline.
+This pipeline shows every step from data observation to rendering of a graphic, but we are going to focus on the tail end of the pipeline specifically on statistics, geometry, coordinates, and aesthetics.
+
+![The grammar of graphics data analysis pipeline](grammarofgraphicspipeline.png){#fig-pipeline}
+
+In order to visually communicate uncertainty we must:  
+1) Mathematically quantify uncertainty in some way. 
+2) Visually express this quantification using some geometry, axis, aesthetic, and rendering.
+3) Interpret the visual expression to extract the information using perceptual tasks.
+
+Each of these steps have a large amount of research independent of the "uncertainty visualisation" subfield.
+There are numerous ways to depict uncertainty mathematically.
+Examples of mathematically defined objects that are used to express uncertainty inlude: confidence intervals, sampling distributions, bias, variance, prediction error, value ranges, quantiles etc.
+These definitions all exist in mathematics independent of the field of visualisation.
+Once we have some quantification of uncertainty, we need to express it visually.
+There are numerous options in combinations of geometry, axis, and aesthetics and modern graphical software will allow most of these combinations.
+There is also a large amount of research in information visualisation and psychology into how the choice of visual expression impacts a persons ability or ease with which they extract relevant information.
+In this sense, once we have a mathematical expression of uncertainty, the visualisation of that uncertainty is identical to the visualisation process of any other variable, or at least this is how it is treated in the uncertainty visualisation experiments.
+
+This pipeline leads to two obvious facts that are *well* established in mathematics and visualisation.
+First, a graphic that depicts one statistic does not necessary contain information about a different, possibly irrelevant, statistic.
+For example, if you compare two graphics and one depicts the mean and the other depicts a median and you ask viewers to extract the median value, the graphic that *actually* depicts that statistic will perform better.
+People cannot percieve information that is not depicted.
+This is not necessarily a feature of visualisation, this is simply a fact of mathematics.
+Additionally, if information that can be used to calculate a statistic, but not the statistic itself, is shown, people will need to use a heuristic to extract that information which will decrease the precision with which that information is communicated.
+Second, there is a hierarchy to perceptual tasks where extracting visual information in some forms is easier than others.
+The hierarchy established 40 years ago by @Cleveland1984 is:
+1) position along a common scale
+2) position along a non-aligned scale
+3) length
+4) angle/slope
+5) area
+6) volume
+7) colour
+This hierarchy has had additions and adjustments over the years.
+For example, this hierarchy is a general rule, but the hierarchy can change from individual to individual.
+Additionally, there are other graphical rules to consider such as guestalt heuristics, broader methods of extraction, and attention principles.
+These establishe visualisation concepts allow us to anticipate the ease with which certain pieces of information will be extracted from a plot.
+A bizarre feature of uncertainty visualisation literature is that many of the evaluation studies seem to be written as though these mathematical facts and established visualisaiton rules do not exist.
+
+The papers that seem to be the most unaware of these established rules in visualisation are the plot comparison papers.
+The most common type of uncertainty visualisation evaluation study is a "comparison of plots".
+The papers that show several uncertainty visualisations and ask participants to extract statistics.
+In these papers will typically ask separate questions about the signal (e.g.
+what is the mean of this distribution) and the uncertainty (e.g.
+what is the variance of this distribution).
+Then the accuracy, speed, or confidence of the participants responses are compared across the plots.
+The overwhelming majority of these papers depict different mathematical expressions of both the signal and the uncertainty.
+In 100% of these papers, the plot that has mapped the statistic in question to some visual feature will outperform those that don't.
+When the statistics used to depict uncertainty *are* the same (which is rare) 100% of those papers will follow the hierarchy of perceptual tasks that has already been established in the visualisation literature.
+If the experiment participant needs to extract a probability and one graphic uses position along a common scale, while the other uses colour or literally any other perceptual task, the position depiction will always outperform the other.
+Strangely, none of these papers comment that their study seems to simply follow established perceptual task rules.
+Instead they opt to comment on how a "bar chat outperforms a mean with confidence intervals" not seeming to connect that it was only on questions related to the median.
+This makes it clear we have a serious problem with the classification of "uncertainty visualisation".
+The visualisations follow identical rules to those established already in the visualisation literature, and experiments that view uncertianty visualisation as somehow special or different from normal visualisation are simply repeating existing established research in information visualisation.
+This stems from a class of "uncertainty visualisation" that seems to have no rules for what belongs in the group, and no consideration for the mathematics that is actually depicted in these plots.
+Uncertainty visualisation experiments do not work to build upon existing visualisation literature or differentiate uncertainty visualisaiton as a new field, but seem to exist with shocking indifference to the field it belongs to.
+
+This brings us to a second problem in these papers.
+Since there is no rules for what is or is not an uncertainty visualisation, graphics that are compared in this sub-dicipline are often *wildly* different.
+If we were to compare two different visualisations, we would like to change as few features as possible so we can better understand the cause of the difference in perception.
+If several graphics are compared that are different statistically, geometrically, and aesthetically, there is *no way to combine the results to establish generic principles for the field*.
+Uncertainty visualisaiton papers only seem to require that the plots depict "uncertainty" or typically belong to the class of "uncertainty visualisation", some of which are simply invented by the author.
+The complete lack of formality is easy to see if we actually plot the visualisations compared across evaluation experiments according to the grammar of graphics *in the figure below*:
+
+**INCLUDE A GRAPHIC THAT SHOWS WHERE EACH TYPE OF UNCERTAINTY VISUALISATION SITS IN THE GRAMMAR OF GRAPHICS AND HOW DIFFERENT THEY ARE ACROSS MULTIPLE STUDIES**
+
+This lack of formality has severly cost the field.
+This lack of formalisation is not unique to uncertainty visualisation and the wider field of information visualisation suffers form similar problems [@Kinkeldey2014].
+Several uncertainty visualisation literature reviews have commented on the immense difficulty of combining results across experiments, but none have been able to pinpoint the reason.
+@Kinkeldey2014 suggested a task based taxonomy that organised the plots by the question or purpose of the visualisation, @Spiegelhalter2017 simply ignored the existence of uncertainty visualisation as a subfield and just suggested we follow established rules within uncertainty visualisation (which was actually good advice), @Hullman2016 commented on the large amount of noise generated from uncertainty visualisation evaluation methods that led to a difficulty in combining results, @Griethe2006 did not meaningfully combine the work and instead listed findings in the field, but also did not comment on the lack of generalisable results.
+While we largely agree with the difficulties of these authors, but instead of a task based taxonomy or better questions, we believe uncertainty visualisation, as it is now, should not exist.
+Any generalisable contribution made by these papers is already an established result in information visualisaiton, and therefore is not specific to "uncertainty visualisation".
+Any "uncertainty visualisation" specific result is not generalisable beyond the experimental conditions because of the poor formalisation in the definitions of uncertainty and uncertainty visualisations.
+
+Exactly *why* uncertainty visualisation is different from typical visualisation is rarely discussed, however there does seem to be some subtextual arguments put forth in the motivation of these expermients.
+These motivations often discuss at least one of two reasons uncertainty visualisation authors seem to think uncertainty visualisation should be treated differently to normal visualisation.
+1) Uncertainty is fundamentally differently to other variables due to psychological heuristics involved in the interpretation of uncertainty, such as risk aversion.
+2) Uncertainty is not of interest in of itself and is typically layered on at the end of a the visualisation pipeline, so uncertainty visualisation is the field of adding error information into already established graphics.
+
+Papers that are justified using the first type of motivation rarely explain exactly how the psychological apsects of uncertainty impct the participants in a way that inhibits their ability to read information from a graphic.
+Since uncertainty is poorly defined and the psychological effects of uncertainty is never justifiably connected to the scientific communication of it, as was discussed in ssections 3 and 4, it is unclear exactly *what* these papers are testing for.
+It is almost as though they *feel* that uncertainty visualisations *should* be different, but do not consider *how* uncertainty information is supposedly different beyond decision making criteria such as *risk avoidance* which is not something science communicators have any business trying to control, as was discussed in previous sections.
+For these reasons, experiments that test for secondary factors (such as influence on decision making) are confounding far too many variables to establish any generalisiable take away, and experiments that ask participants to extract values fail to notice there is no longer a reason for "uncertainty" to be any different to any other type of visualisation.
+
+Papers that are justified using the second type of motivation clearly fit within the sub-field of high dimensional visualisation, but also rarely comment on the connection.
+These experiments evaluate different ways to include even *more* information than would be put in a typical visualisation.
+Some of these methods, such as a bivariate pallet or glyph map, create graphics that are overloaded with information and offer incorrect or meaningless takeaways at first glance.
+This is not to say *every* visualisation method that sees it as a dimensionality problem is bad.
+Clever methods of combining several types of information, such as value supressing uncertainty pallets (VSUPs), can better help us understand how our brains process information in a graphic.
+For example, the VSUP allows us to test is colour value and saturation can be combined such that we mentally combine signal and noise together as one variable.
+Since the take aways from the papers are not usually related to uncertainty specifically, and often related to interesting perceptual tasks, there is also no justification for why this work is "uncertainty visualisation" work and not simply perceptual task studies that can be used by the rest of the information visualisation community.
+
+These two perceptualisations of *why* uncertainty visualisation is different from normal visualisation is also what leads to differing opinions on what is an uncertainty visualisation.
+For example, someone who percieves of uncertainty visualisation as a dimensionality problem would see a pie chart as a normal graphic, and a pie chart with blurred boundaries as an uncertainty visualisation, while someone who sees uncertainty visualisation as psychologically differentwould see a pie chart as an uncertainty graphic so long as it depicts probabilities.
+This may lead you to believe that those who believe uncertainty is unique due to its psychological properties believe uncertainty should be defined at the mathematical stage, while those that believe it is an addition to an established visualisation think uncertainty is defined in a graphic at the aesthetic stage.
+This is not the case.
+As shown by the *grammar of graphics fig* there is no single person or experiment that is consistent in any one belief.
+This difficulty and lack of consistency comes from the desire to discuss "uncertainty visualisation" as though it is a single term and not the combination of "uncertainty" and "visualisation".
+If we split the terms and look at the quantification and visualisation of uncertainty separately, the results of many uncertainty visualisation papers become self evident.
+
+This is not to say any paper that discusses uncertainty visualisation is completely pointless because the results already exist. Perceptual task experiments that identify
+
+and fruitless endeavor. Rather, we want to established that the current literature seems to be disconnected from the actual motivations of uncertainty visualisation.
 

### 2024-02-21 paper.qmd
Finished the grammar of graphics section

--- 2024-02-19 paper.qmd
@@
 *The location* axis of the taxonomy lines up neatly with the typical data analysis pipeline.
-Multiple authors in the uncertainty literature have commented on the need to consider quantifying and expressing uncertainty at every stage of a project [@Kinkeldey2014; @Refsgaard2007].
+Multiple authors in the uncertainty literature have commented on the need to consider quantifying and expressing uncertainty at every stage of a project [@Kinkeldey2014; @Hullman2016; @Refsgaard2007].
 @fig-location shows the connection between the typical data analysis pipeline and the location element of the uncertainty taxonomy, along with examples of how uncertainty could arise from each of these steps.
@@
 Additionally, since there is often a miriad of ways to visualise any particular mathematical object, to adopt specific visualisation that is not already common practice needs reasonable justification, lest we run into a range of unforseen heuristic pitfalls.
+
+::: aside
+I should maybe change this to relatability to uncertainty, value extraction, decision making and trust, and meta papers or something. Maybe perception and evaluation papers?
+:::
 
@@
 
-## Uncertainty Visualisation and the Grammar of Graphics
+## Perception Experiments and a Poor Definition
+Even the most simple experiments are not free from the issues in the uncertainty definition.
+
+Papers that consider unique aspects of uncertainty in their experiment such as experiemnts that identify the usefulness of specific perceptual tasks, experiments that identify problematic heuristics in commonly used uncertainty visualisations, and experiments that consider unique problems in uncertainty experiments all have useful results.
+These papers will be discussed in the upcomming sections.
+
+Perceptual task experiments that identify visual aesthetics that are intuitively connected to uncertainty are useful as they establish an inutive psychological connection.
+These papers are also not without their flaws, however their flaws have little to do with issues in the grammar of graphics as they are well defined in this space and more to do with the lack.
+Since uncertainty and signal are typically mapped to different aesthetics, it is not always clear what exactly the visualisation is suggesting we should be uncertain about, and when a perceptual task that should be considered something like a train line has "uncertainty" added to it, it is not
+
+- rarely used extension is aggregation of uncertainty over an area (retrieve overall estimation from a spatial distribution of uncertainty).
+Other tasks include comparisons, rankings.
+also have a task called "search" where participants have to identify entities that fulfill certainty characteristics (high or low values) @Kinkeldey2014
+- Whether to use classed or unclassed uncertainty is rarely discussed @Kinkeldey2014.
+e.g.
+land map with remotely sensed images has ambiguity in the boundaries, which can result from multiple sources (land cover class, measurement error, images from different dates).
+If you combine them all it might not make sense for someone who only needs one.
+
+- I think the dimension of the uncertainty should be considered every time there is aggregation from multiple sources of uncertainty and you want to maintain those sources.
+You can project the uncertainty down into one dimension if the source information is in of itself unimportant.
+
+### Contextual Inormation
+I have this here to explain why context based taxonomies can be ignored when we are considering uncertainty.
+I did not go into too much detail because it would involve discussing visualisation which I am trying to avoid at this point.
+I actually think this quote from the grammar of graphics is exactly what I was trying to say.
+*"The system in this book is not a model for a GIS, however, because geography and statistics differ in a crucial respect.
+Geography is anchored in real space-time and statistics in abstract dimensions.
+This is a distinction along a continuum rather than a sharp break; after all, there is a whole field called spatial statistics (Cressie, 1991).
+But this difference in focus clearly means that a system optimized to handle geography will not be graceful when dealing with statistical graphics"*.
+Basically, my point is this section is still discussing the definitions of unertainty, we have not gotten to disussing visualisation so we are still in the abstract *statistics* domain.
+The only parts of spatial statistics that we care about in this domain are the statistical properties that come from spatial information, and *none* of those statistical properties generate uncertainty.
+This is not unique to the properties of spatial information, other statistics information that is important for modelling and assumptions, such as non-linearity, class levels, etc do not contribute to uncertainty.
+Operations on data, inherrent randomness, and missing information generate uncertainty, which is what we are talking about here.
+A confidence interval is a confidence interval whether it came from spatial data, temporal, or cross sectional data.
+*This does not mean I am saying spatial and temporal information is not valuable ever*, I am just saying it is not valuable *yet*.
+This information will be *highly relevant* when we discuss context and translating from the abstract domain back into real world for communication.
+Since we are still in the abstract domain, things that are anchored in real space-time create confusion that is evident in the literature.
+The confusion that is created in the literature was going to be the topic I had later.
+Spatial and temporal uncertainty papers limit how they communicate their uncertainty because they are considering what can be visualised, not what they should *try* to visualise based on the uncertainty abstraction.
+Since the context is only relevant *with respect to uncertainty* when we visualise it, I was going to discuss this issue in the visualisation literature review section.
+
+The taxonomies that cannot be contained within the @utypo taxonomy are not a new uncertainty dimensions, but rather they are conflating adjacent information with uncertainty.
+A good example of this is the distinction between spatial, temporal, and estimated uncertainty.
+When we create statistical models and quantify uncertainty, the spatial or temporal aspects are often boilled away [@Otsuka2023], and the only element of it that remains is the inexchangeability of our data.
+For this reason, there realistically is no such thing as "spatial" uncertainty in the same sense that there is no such thing as "photograph" uncertainty or "coin toss" uncertainty or "baby birth" uncertainty.
+Maps and photographs boil down to the same statistical objects in the same way that coin tosses and baby births boil down to the same statistical objects.
+Once we have removed the physical nature these objects are near identical when considered statistically.
+Of course the nature of them can change based on what we choose to observe (i.e.
+if we consider a coin landing on its side to be an outcome or if we consider intersex babies as an outcome in births) but this means the contextual information of our data, e.g.
+whether or not it is spatial or temporal, is only of importance at two stages of the data analysis pipeline.
+It is first important when we *boil it down* to our data through *observation* and again when we *map our statistical information back to its real world information* to *communicate* it.
+The only way the context uniquely contributes to uncertainty is in the "boil down" stage, however spatial or temporal uncertainty never refers to that specific cause of uncertainty.
+It is typically in reference to measurement error, data aggregation, statistical modelling, or other steps that occur *after* the boil down stage, and are therefore not unique to spatial or temporal data.
+While the spatial or temporal features *may* become relevant at the final communication stage, however this has little do with quantifying uncertainty, and the difficulties in communicating a complicated statistical model through a complicated map is already well discussed in the visualisation literature, and in later sections we will discuss how this concept is related to uncertainty visualisation.
+
+In this sense we are not proposing that the current conceptualisation of data according to its physical properties is necessarily wrong, or that it leads to incorrect conclusions, we are merely suggesting for a different way of slicing the metaphysical cake of uncertainty that better aligns with what is *actually happening* when we perform a statistical analysis.
+This metaphysical cake is illustrated below as @fig-cake.
+In changing  what the actual boundaries are when we work with these problems, we improve our understanding of what we are actually doing when we conceptualise uncertainty, and in the process allow for more creative solutions creativity (e.g.
+by borrowing concepts from spatial visualisation to visualise uncertainty in photographs generated).
+However, since uncertainty is almost synonymous with the concept of "statistics", this adjustment in how we conceptualise and organise statistical methods would likely benefit other domains within the field.
+
+![The metaphysical cake we are slicing differently](metaphysicalcake.jpg){#fig-cake}
+
+- Specifies uncertainty can be represented by three components attribute (what) position (where) and temporal (when) @Kinkeldey2014
+- Studies dealing with uncertainty information typically deal with attribute but rarely position and time @Kinkeldey2014
+- 10/44 studies involve multiple types of uncertainty @Kinkeldey2014
+
+- When we boil down data into a statistical object, we boil away many aspects on context. Birth and coin flips become an object that is one in the same, and often the visualisation 
+- Why are spatial and temporal data considered special?
+What is the statistical element that remains after they have been away?
+Inexchangeability.
+Temporal data is one dimensional inexchangeability, spatial data has two dimensional inexchangeability.
+  - There are other things that have inexchangeability, and not every question related to spatial or temporal data requires the inexchangeability.
+- perceptual tasks need to know what you are uncertain about.
+- Yes the data type is relevant but its not the most important thing
+  - one with temporal, one with that temporal paper but you didn't know it was temporal, one with the bus time cdf
+  - screenshot from these three papers, the connecting factor is the question, not the data type
+  - The data type: spatial, temporal, etc, usually is mentioned as the 
+  - Not a feature of the data, but it is a feature of the question "how do I show uncertainty on a map?".
+This is a fine question to ask, but the lack of specificity in what is being done, and the zoning in on functionally irrelevant elements leads the literature as a huge mess.
+- Hurricane risk is not only the path, but also the storm surge and wind speed. @Spiegelhalter2017
+- coloured maps for earthquake risk are easily interpretable @Spiegelhalter2017
+
+## Evaluation Experiments and The Grammar of Graphics
 The first issue in uncertainty visualisation is there does not seem to be a clear rule for what makes a particular visualisation a depiction of uncertainty.
+This problem seems to start at the lack of the mathematical definition of uncertainty which then bleeds into a lack of a definition for an "uncertainty visualisation".
+For example @Leland2005 mentions that popular graphics, such as pie charts and bar charts omit uncertainty, however at least one or both of these charts are used in a large number of uncertainty visualisation experiments.
+How can a chart omit uncertainty according to the grammar of graphics, but be used to describe it in several evaluation studies?
+The answer to this question is related to the definition problem discussed earlier, but to understand how, we will need to go through the grammar of graphics, and its absence in the uncertainty visualisation literature.
 
-- Since there are so many papers that treat uncertainty as a signal, not a noise, issues in the uncertainty visualisation literature also help to understand some difficulties in the information visualisation literature as large.
-
-- Suggests that it is straightforward to show a value but much more complex to show uncertainty @Hullman2016 (I think this is interesting in relation to how people SEE uncertainty visualisation.
-Why is uncertainty hard to visualise?)
-- Authors provide little justification for their chosen response models (e.g. absolute accuracy vs relative measures) @Hullman2016
-- When asking questions, researchers should focus on events that can be repeated (because of a misunderstanding of confidence intervals) @Hullman2016
-- Uncertainty defined as a by product of the process of making a visualisation, from data collection to commucating @Hullman2016
-- Some questions (e.g. what are the chances that the no.6 bus will arrive first) you can elicit the viewers subjective probability distribution @Hullman2016
-- communication tasks: This category comprises map reading tasks involving data and uncertainty value retrieval (which location is most uncertain?).
-For tasks from this category, visualisation techniques can be chosen following the traditional rules from cartography.
-@Kinkeldey2014 (basically saying why dont you just use what is already here if you are treating uncertainty as a signal, which I agree with)
-- papers usually have same variance if there is a signal question, and use the mean as a proxy to answer the question.
-Really should ask a changing signal question where the signal IS impacted by the uncertainty information.
-- Weirdly When setting their ground truth it seems like a lot of papers EXPECT participants to completely ignore uncertainty information as noise.
-Like the participants are actually considered to be incorrect if they dampen their signal understanding because of noise considerations which I find FUNNY.
-- I think the questions where uncertainty is a signal, not a noise, allow the authors of visualisation experiments to ignore the context and motivation that usually governs a visualisation.
+::: aside
+I think I want the "in order to visualise uncertainty you need to quantify it and the quantification does not align with the definition" point here since this section is now about value extraction.
+:::
 
@@
 A visualisation is, in a lot of ways, a statistic, or at the very least statistics are calculated prior to the visualisation being calculated, however this may be invisible to the user.
-A density plot does not just "appear" from the data, usually a smoothing function (that should be calibrated according to the sample size however how this should be done is a different matter of dispute) generates an estimated density function (statistics) which is then depicted on a plot using a line (design or computer science) which is then converted into information in your head using perceptual tasks and heuristics (psychology).
-
+A density plot does not just "appear" from the data, usually a smoothing function generates an estimated density function, which is then depicted on a plot using a line, which is then converted into information in your head using perceptual tasks and heuristics.
 These overlapping but distinct fields create a high burden of spanning expertise to understand if two visualisation are equivalent in each field.
@@
 Therefore, if you are trying to estimate a population mean, the sample mean and the entire data set are both equivalent, however a similar statistic, such as the median, is not and will become a worse and worse approximation for the mean the more skewed your underlying distribution in.
-Computer science considered two pieces of information to be equivalent if they come from the same data [^This is the justification that I have come across in several computer science field papers, I am not sure if this is true for the whole field so this is definitely a question for Sarah].
+Computer science considered two pieces of information to be equivalent if they come from the same data.
 This definition of information is much looser and is contained within the mathematical definition of information.
+Thanksfully, we can turn to the grammar of graphics to try and organise and untangle the information in our graphics.
 
-Most uncertainty visualisation papers show that the definition of information *should* be the one defined by mathematics, not the one defined by computer science.
-A close look at the visual inference literature shows this to be true, but a large amount of noise is added to the research through visual heuristics.
-This is simply because, if one visualisation depicts a "significant statistic" and another visualisation shows a similar statistic, then the graphic that shows the significant statistic holds more relevant information to the question and will lead to more accurate interpretations.
-If the information that is depicted on a plot cannot be used to *mathematically* generate a best estimate for a specific statistic, I do wonder how the visual system would be expected to cover that gap if *not* for heuristics.
+When creating a graphic there are several tasks that must be completed in a specific order, regardless of whether or not we are working with a uncertainty visualisation or not.
+@Leland2005 depicts these steps as a pipeline, shown in @fig-pipeline.
+This pipeline shows every step from data observation to rendering of a graphic, but we are going to focus on the tail end of the pipeline specifically on statistics, geometry, coordinates, and aesthetics.
 
-While mathematical information *should* be considered the starting point in considering if two visualisations are equivalent, it is not the only consideration.
-A large amount of work in perceptual tasks, attention, and psychology surrounding charts shows that all the information depicted is not paid attention to in equal measure.
-Elements such as colour can have sizable impacts on the way a graphic is percieved.
-Mathematics identifies the information that is *shown* while psychology identifies the difference in information *recieved*.
-A failure to understand this distinction leads to graphics that are different not only in the way information is depicted but in the information itself.
-This problem is pervasive in uncertainty visualisation literature.
+![The grammar of graphics data analysis pipeline](grammarofgraphicspipeline.png){#fig-pipeline}
 
-- The big data paradox shrinks confidence intervals but magnifies bias [@Bradley2021] because the confidence intervals shrink due to the sample size but the imprecise target group creates bias.
+In order to visually communicate uncertainty we must:  
+1) Mathematically quantify uncertainty in some way. 
+2) Visually express this quantification using some geometry, axis, aesthetic, and rendering.
+3) Interpret the visual expression to extract the information using perceptual tasks.
 
-- But when different techniques are to be compared, it is important that the scenarios and datasets are informationally equivalent, i.e.
-according to Larkin and Simon (1987, p.
-67) ‘[t]wo representations are informationally equivalent if all of the information in the one is also inferable from the other, and vice versa’.
-A goal in testing, then, is often to determine whether they are also computationally equivalent, or whether one depiction has an advantage over another, i.e.
-‘[t]wo representations are computationally equivalent if they are informationally equivalent and, in addition, any inference that can be drawn easily and quickly from the information given explicitly in the one can also be drawn easily and quickly from the information given explicitly in the other, and vice versa’ (Larkin and Simon 1987, p.
-67).
-@Kinkeldey2014
+Each of these steps have a large amount of research independent of the "uncertainty visualisation" subfield.
+There are numerous ways to depict uncertainty mathematically.
+Examples of mathematically defined objects that are used to express uncertainty inlude: confidence intervals, sampling distributions, bias, variance, prediction error, value ranges, quantiles etc.
+These definitions all exist in mathematics independent of the field of visualisation.
+Once we have some quantification of uncertainty, we need to express it visually.
+There are numerous options in combinations of geometry, axis, and aesthetics and modern graphical software will allow most of these combinations.
+There is also a large amount of research in information visualisation and psychology into how the choice of visual expression impacts a persons ability or ease with which they extract relevant information.
+In this sense, once we have a mathematical expression of uncertainty, the visualisation of that uncertainty is identical to the visualisation process of any other variable, or at least this is how it is treated in the uncertainty visualisation experiments.
 
-- The way we ask questions and the types of questions we ask are selected with little justification. This paper makes suggestions to reduce the noise in the data from these papers. @Hullman2016
+This distinction is important because lower levels of the pipeline influence higher levels.
+Visualisation authors are almost unanimous in commenting that the information in two plots must be the same in order for the visual techniques to be compared [@Cleveland1984; @Kinkeldey2014] *this citation has way more, check confirmation*.
+@Kinkeldey2014 adopts a definition from the 80s that suggests two graphics are informationally equivalent if all the information in one plot is inferable from the other and vice versa, and computationally equivalent if that information can be extracted from both plots with similar easy and speed.
+This definition illustrates a similar idea to the grammar of graphics pipeline, shown in @fig-pipeline.
+First the mathematical information in a plot is established, and then that mathematical information is visually represented.
+Two plots that depict mathematical information that cannot be inferred from each other, cannot be compared on their visual elements.
+To prevent this requirement from being *too* limiting, when using a ground truth statistic we can consider two plots to be informationally equivalent if both visualisations depict a sufficient statistic.
+For example, a depiction of a sample and its average are equivalent if we are trying to infer the population mean.
+This does not mean they are computationally equivalent.
+Additionally, two plots cannot be compared on their visual elements if they contain different information as well.
+The reason for this is obvious.
+If you compare two graphics and one depicts the mean and the other depicts a median and you ask viewers to extract the median value, the graphic that *actually* depicts that statistic will perform better.
+People cannot percieve information that is not depicted.
 
-- Most commonly check abolute measures of accuracy, where accuracy is abs(subjective error - actual probabiltiy).
-Some papers also check relative measures of accuracy and ask subjects to find regions of least certainty or to rank targets by uncertainty.
-studies also consider response time and confidence rating.
-some studies also ask users for their own expression of uncertainty (e.g.
-by asking when sketches are uncertainty and ranking various depictions of uncertainty).
-Some studies ask subjects to make some decision using the data.
-ometimes things like complexity of task, degree of visual overload, ease of use, visual appeal, and preferences @Hullman2016
+Computational equivalency in visualisation is also something that has been well researched.
+For simple tasks such as value extraction, there is a hierarchy to perceptual tasks where extracting visual information in some forms is easier than others.
+The hierarchy established 40 years ago by @Cleveland1984 is:
+1) position along a common scale
+2) position along a non-aligned scale
+3) length
+4) angle/slope
+5) area
+6) volume
+7) colour
+This hierarchy has had additions and adjustments over the years.
+For example, this hierarchy is a general rule, but the hierarchy can change from individual to individual.
+Additionally, there are other graphical rules to consider such as guestalt heuristics, broader methods of extraction, and attention principles.
+These established visualisation concepts allow us to anticipate the ease with which certain pieces of information will be extracted from a plot.
+
+A bizarre feature of uncertainty visualisation literature is that it does not work to build upon these existing principle or identify the ways in which uncertainty visualisations may diverge from these rules, but rather it seems to exist with complete indifference to the research in information visualisation at large.
+The most common uncertainty visualisation evaluation study asks participants to perform some value abstraction either explicitly by asking participants to report values or vaguely by asking participants to find regions of least certainty, rank plots, report their confidence, and/or make decisions according to uncertainty information.
+Experimenters then compare these results based on the participants accuracy, preference, confidence, and/or speed @Hullman2016.
+Most commonly, these value abstraction papers compare two different visualisations of uncertainty on a set of these questions, however, since there is no rules for what is or is not an uncertainty visualisation the graphics that are compared often differ in both computation *and* information.
+Uncertainty visualisaiton papers only seem to require that the plots depict "uncertainty" which has already been established to be poorly defined.
+The complete lack of formality is easy to see if we actually plot the visualisations compared across evaluation experiments according to the grammar of graphics *in the figure below*:
+
+**INCLUDE A GRAPHIC THAT SHOWS WHERE EACH TYPE OF UNCERTAINTY VISUALISATION SITS IN THE GRAMMAR OF GRAPHICS AND HOW DIFFERENT THEY ARE ACROSS MULTIPLE STUDIES**
+
+Reading the uncertainty visalisation literature will make the importance of information and computation incredibly salient.
+If plots that have different information are compared, whichever directly depicts the ground truth statistic (or has information that can easily be convered to that ground truth statistic) will perform the best *cite*.
+If plots that have the same information are compared, the plot that requires a perceptual task that is higher on the hierarchy will perform the best *cite*.
+Therefore, when it comes to any task that involves value abstraction, many uncertainty visualisation evaluation studies are simple repeating already established results in visualisation.
+
+Several uncertainty visualisation literature reviews have commented on the immense difficulty of combining results across experiments, but none have been able to pinpoint the reason.
+@Kinkeldey2014 suggested a task based taxonomy that organised the plots by the question or purpose of the visualisation, and even suggested that communication tasks (such as value retrival of uncertainty values) should just follow traditional cartography rules, @Spiegelhalter2017 simply ignored the existence of uncertainty visualisation as a subfield and just suggested we follow established rules within uncertainty visualisation, @Hullman2016 said that commented on the large amount of noise generated from uncertainty visualisation evaluation methods that led to a difficulty in combining results, @Griethe2006 did not meaningfully combine the work and instead listed findings in the field, but also did not comment on the lack of generalisable results.
+While we largely agree with the difficulties of these authors, we do not agree that the solution is a task based taxonomy or better questions.
+While we agree the lack of attention to tasks and appropriate questions has generated problems in the field, there is a much more pressing issue, which is that uncertainty visualisation authors don't seem to understand *why* uncertainty visualisation different to typical visualisations.
+
+The class of "uncertainty visualisation" does not seem to have any rules for what belongs in the group and what makes it different to a "typical" visualisation is rarely discussed.
+Instead one of two general motivations for the field are provided in the subtext.
+1) Uncertainty is fundamentally differently to other variables due to psychological heuristics involved in the interpretation of uncertainty (e.g. risk aversion).
+2) Uncertainty is not of interest in of itself and is typically layered on at the end of a the visualisation pipeline, so uncertainty visualisation is the field of adding error information into already established graphics.
+Neither of these motivations are remotely relevant when we are evaluating the simple task of value extraction.
+The psychological effects of uncertainty are only relevant for decision making and have nothing to do with value abstraction, and the information overload issue is not specific to uncertainty and is simply an example of a larger dimensionality issue.
+Therefore, these explanations for the difference fall short of explaining *why* uncertainty should be treated any differently than any other variable.
+
+Despite these results, it is clear that authors intuitively *feel* that uncertainty is different even though this difference has not been apparent in existing uncertainty visualisation experiments.
+@Kinkeldey2014 also noticed that it is not clear whether or not uncertainty is just another variable, as many value extraction papers treats it as such in their experimental design and uncertainty may need to be in a variable class of its own.
+@Hullman2016 commented that it is straightforward to show a value but it is much more complex to show uncertainty, but did not explain *why* uncertainty information faces more difficulty.
+So, if we cannot see any difference to uncertainty information when it coems to value abstraction, *when* is uncertainty information different?
 
 ## Noise and Signal in Uncertainty Visualisation
-The first issue that arises with uncertainty research is that uncertainty is frequently treated as a signal, not noise.
+Uncertainty is noise, not signal.
+While this comment may seem obvious, what may not be obvious is that the value abstraction experiments are treating uncertainty as *signal*.
+That is why the experimental results are identical to the existing information visualisation literature.
+
 This discrepancy is also noted by @Kinkeldey2014 who, in their own literature review on uncertainty methods, comment that the literature seems unsure if uncertainty is simply another variable or if it should be treated differently.
@@
 If uncertainty is simply a signal we are trying to convey, why should it be treated any different to any other signal in a visualisation?
+
+- I think the questions where uncertainty is a signal, not a noise, allow the authors of visualisation experiments to ignore the context and motivation that usually governs a visualisation.
+- Weirdly When setting their ground truth it seems like a lot of papers EXPECT participants to completely ignore uncertainty information as noise.
+Like the participants are actually considered to be incorrect if they dampen their signal understanding because of noise considerations which I find FUNNY.
+(this feels like it is more as a communicating uncertainty issues)
+- papers usually have same variance if there is a signal question, and use the mean as a proxy to answer the question.
+Really should ask a changing signal question where the signal IS impacted by the uncertainty information.
+- When asking questions, researchers should focus on events that can be repeated (because of a misunderstanding of confidence intervals) @Hullman2016 (moved here because it is clear she is touching on the whole "what are we uncertain about" thing)
 
@@
 - Need to compare to an approripate ground truth, but this is a philisophical exercse @Hullman2016
-
-## An Unexplored Space (Different visualisations for location and stuff)
-- rarely used extension is aggregation of uncertainty over an area (retrieve overall estimation from a spatial distribution of uncertainty).
-Other tasks include comparisons, rankings.
-also have a task called "search" where participants have to identify entities that fulfill certainty characteristics (high or low values) @Kinkeldey2014
-- Whether to use classed or unclassed uncertainty is rarely discussed @Kinkeldey2014.
-e.g.
-land map with remotely sensed images has ambiguity in the boundaries, which can result from multiple sources (land cover class, measurement error, images from different dates).
-If you combine them all it might not make sense for someone who only needs one.
-
-- I think the dimension of the uncertainty should be considered every time there is aggregation from multiple sources of uncertainty and you want to maintain those sources.
-You can project the uncertainty down into one dimension if the source information is in of itself unimportant.
-
-## Contextual Inormation
-I have this here to explain why context based taxonomies can be ignored when we are considering uncertainty.
-I did not go into too much detail because it would involve discussing visualisation which I am trying to avoid at this point.
-I actually think this quote from the grammar of graphics is exactly what I was trying to say.
-*"The system in this book is not a model for a GIS, however, because geography and statistics differ in a crucial respect.
-Geography is anchored in real space-time and statistics in abstract dimensions.
-This is a distinction along a continuum rather than a sharp break; after all, there is a whole field called spatial statistics (Cressie, 1991).
-But this difference in focus clearly means that a system optimized to handle geography will not be graceful when dealing with statistical graphics"*.
-Basically, my point is this section is still discussing the definitions of unertainty, we have not gotten to disussing visualisation so we are still in the abstract *statistics* domain.
-The only parts of spatial statistics that we care about in this domain are the statistical properties that come from spatial information, and *none* of those statistical properties generate uncertainty.
-This is not unique to the properties of spatial information, other statistics information that is important for modelling and assumptions, such as non-linearity, class levels, etc do not contribute to uncertainty.
-Operations on data, inherrent randomness, and missing information generate uncertainty, which is what we are talking about here.
-A confidence interval is a confidence interval whether it came from spatial data, temporal, or cross sectional data.
-*This does not mean I am saying spatial and temporal information is not valuable ever*, I am just saying it is not valuable *yet*.
-This information will be *highly relevant* when we discuss context and translating from the abstract domain back into real world for communication.
-Since we are still in the abstract domain, things that are anchored in real space-time create confusion that is evident in the literature.
-The confusion that is created in the literature was going to be the topic I had later.
-Spatial and temporal uncertainty papers limit how they communicate their uncertainty because they are considering what can be visualised, not what they should *try* to visualise based on the uncertainty abstraction.
-Since the context is only relevant *with respect to uncertainty* when we visualise it, I was going to discuss this issue in the visualisation literature review section.
-
-The taxonomies that cannot be contained within the @utypo taxonomy are not a new uncertainty dimensions, but rather they are conflating adjacent information with uncertainty.
-A good example of this is the distinction between spatial, temporal, and estimated uncertainty.
-When we create statistical models and quantify uncertainty, the spatial or temporal aspects are often boilled away [@Otsuka2023], and the only element of it that remains is the inexchangeability of our data.
-For this reason, there realistically is no such thing as "spatial" uncertainty in the same sense that there is no such thing as "photograph" uncertainty or "coin toss" uncertainty or "baby birth" uncertainty.
-Maps and photographs boil down to the same statistical objects in the same way that coin tosses and baby births boil down to the same statistical objects.
-Once we have removed the physical nature these objects are near identical when considered statistically.
-Of course the nature of them can change based on what we choose to observe (i.e.
-if we consider a coin landing on its side to be an outcome or if we consider intersex babies as an outcome in births) but this means the contextual information of our data, e.g.
-whether or not it is spatial or temporal, is only of importance at two stages of the data analysis pipeline.
-It is first important when we *boil it down* to our data through *observation* and again when we *map our statistical information back to its real world information* to *communicate* it.
-The only way the context uniquely contributes to uncertainty is in the "boil down" stage, however spatial or temporal uncertainty never refers to that specific cause of uncertainty.
-It is typically in reference to measurement error, data aggregation, statistical modelling, or other steps that occur *after* the boil down stage, and are therefore not unique to spatial or temporal data.
-While the spatial or temporal features *may* become relevant at the final communication stage, however this has little do with quantifying uncertainty, and the difficulties in communicating a complicated statistical model through a complicated map is already well discussed in the visualisation literature, and in later sections we will discuss how this concept is related to uncertainty visualisation.
-
-In this sense we are not proposing that the current conceptualisation of data according to its physical properties is necessarily wrong, or that it leads to incorrect conclusions, we are merely suggesting for a different way of slicing the metaphysical cake of uncertainty that better aligns with what is *actually happening* when we perform a statistical analysis.
-This metaphysical cake is illustrated below as @fig-cake.
-In changing  what the actual boundaries are when we work with these problems, we improve our understanding of what we are actually doing when we conceptualise uncertainty, and in the process allow for more creative solutions creativity (e.g.
-by borrowing concepts from spatial visualisation to visualise uncertainty in photographs generated).
-However, since uncertainty is almost synonymous with the concept of "statistics", this adjustment in how we conceptualise and organise statistical methods would likely benefit other domains within the field.
-
-![The metaphysical cake we are slicing differently](metaphysicalcake.jpg){#fig-cake}
-
-- Specifies uncertainty can be represented by three components attribute (what) position (where) and temporal (when) @Kinkeldey2014
-- Studies dealing with uncertainty information typically deal with attribute but rarely position and time @Kinkeldey2014
-- 10/44 studies involve multiple types of uncertainty @Kinkeldey2014
-
-- When we boil down data into a statistical object, we boil away many aspects on context. Birth and coin flips become an object that is one in the same, and often the visualisation 
-- Why are spatial and temporal data considered special?
-What is the statistical element that remains after they have been away?
-Inexchangeability.
-Temporal data is one dimensional inexchangeability, spatial data has two dimensional inexchangeability.
-  - There are other things that have inexchangeability, and not every question related to spatial or temporal data requires the inexchangeability.
-- perceptual tasks need to know what you are uncertain about.
-- Yes the data type is relevant but its not the most important thing
-  - one with temporal, one with that temporal paper but you didn't know it was temporal, one with the bus time cdf
-  - screenshot from these three papers, the connecting factor is the question, not the data type
-  - The data type: spatial, temporal, etc, usually is mentioned as the 
-  - Not a feature of the data, but it is a feature of the question "how do I show uncertainty on a map?".
-This is a fine question to ask, but the lack of specificity in what is being done, and the zoning in on functionally irrelevant elements leads the literature as a huge mess.
-- Hurricane risk is not only the path, but also the storm surge and wind speed. @Spiegelhalter2017
-- coloured maps for earthquake risk are easily interpretable @Spiegelhalter2017
 

### 2024-02-21 writingdumpster.qmd
lot of additions to the grammar of graphics section
Finished the grammar of graphics section

@@
 -  @Spiegelhalter2017 defines risk as risk known probabilities and uncertainty as estimated probabilities, therefore uncertainty is what we have when you cannot accurately quantify risk.
+- Authors provide little justification for their chosen response models (e.g. absolute accuracy vs relative measures) @Hullman2016 
+- Some questions (e.g. what are the chances that the no.6 bus will arrive first) you can elicit the viewers subjective probability distribution @Hullman2016
+- The big data paradox shrinks confidence intervals but magnifies bias [@Bradley2021] because the confidence intervals shrink due to the sample size but the imprecise target group creates bias.
+- The way we ask questions and the types of questions we ask are selected with little justification. This paper makes suggestions to reduce the noise in the data from these papers. @Hullman2016
+
+# Main section outtakes
+While mathematical information *should* be considered the starting point in considering if two visualisations are equivalent, it is not the only consideration.
+A large amount of work in perceptual tasks, attention, and psychology surrounding charts shows that all the information depicted is not paid attention to in equal measure.
+Elements such as colour can have sizable impacts on the way a graphic is percieved.
+Mathematics identifies the information that is *shown* while psychology identifies the difference in information *recieved*.
+A failure to understand this distinction leads to graphics that are different not only in the way information is depicted but in the information itself.
+This problem is pervasive in uncertainty visualisation literature.
+
+This is not to say any paper that discusses uncertainty visualisation is completely pointless because the results already exist.
+We are simply pointing out that treating uncertainty graphics as special simply because they contain uncertainty, even if that uncertainty is not related to the task, leads to results that are already established in information visualisation.
+
+These two perceptualisations of *why* uncertainty visualisation is different from normal visualisation is also what leads to differing opinions on what is an uncertainty visualisation.
+For example, someone who percieves of uncertainty visualisation as a dimensionality problem would see a pie chart as a normal graphic, and a pie chart with blurred boundaries as an uncertainty visualisation (even though blurring the boundary between segments of a pie chart makes no sense when you think of a pie chart as a polar co-ordinate bar chart), while someone who sees uncertainty visualisation as psychologically different would see a pie chart as an uncertainty graphic so long as it depicts probabilities.
+This may lead you to believe that those who believe uncertainty is unique due to its psychological properties believe uncertainty should be defined at the mathematical stage, while those that believe it is an addition to an established visualisation think uncertainty is defined in a graphic at the aesthetic stage.
+This is not the case.
+As shown by the *grammar of graphics fig* there is no single person or experiment that is consistent in any one belief.
+This difficulty and lack of consistency comes from the desire to discuss "uncertainty visualisation" as though it is a single term and not the combination of "uncertainty" and "visualisation".
+If we split the terms and look at the quantification and visualisation of uncertainty separately, the results of many uncertainty visualisation papers become self evident.
+
+Papers that are justified using the first type of motivation rarely explain exactly how the psychological apsects of uncertainty impct the participants in a way that inhibits their ability to read information from a graphic.
+Since uncertainty is poorly defined and the psychological effects of uncertainty is never justifiably connected to the scientific communication of it, as was discussed in ssections 3 and 4, it is unclear exactly *what* these papers are testing for.
+It is almost as though they *feel* that uncertainty visualisations *should* be different, but do not consider *how* uncertainty information is supposedly different beyond decision making criteria such as *risk avoidance* which is not something science communicators have any business trying to control, as was discussed in previous sections.
+For these reasons, experiments that test for secondary factors (such as influence on decision making) are confounding far too many variables to establish any generalisiable take away, and experiments that ask participants to extract values fail to notice there is no longer a reason for "uncertainty" to be any different to any other type of visualisation.
+
+Papers that are justified using the second type of motivation clearly fit within the sub-field of high dimensional visualisation, but also rarely comment on the connection.
+These experiments evaluate different ways to include even *more* information than would be put in a typical visualisation.
+Some of these methods, such as a bivariate pallet or glyph map, create graphics that are overloaded with information and offer incorrect or meaningless takeaways at first glance.
+This is not to say *every* visualisation method that sees it as a dimensionality problem is bad.
+Clever methods of combining several types of information, such as value supressing uncertainty pallets (VSUPs), can better help us understand how our brains process information in a graphic.
+For example, the VSUP allows us to test is colour value and saturation can be combined such that we mentally combine signal and noise together as one variable.
+Since the take aways from the papers are not usually related to uncertainty specifically, and often related to interesting perceptual tasks, there is also no justification for why this work is "uncertainty visualisation" work and not simply perceptual task studies that can be used by the rest of the information visualisation community.
+
+This stems from a class of "uncertainty visualisation" that seems to have no rules for what belongs in the group, and no consideration for the mathematics that is actually depicted in these plots.
+Uncertainty visualisation experiments do not work to build upon existing visualisation literature or differentiate uncertainty visualisaiton as a new field, but seem to exist with shocking indifference to the field it belongs to.
+
+uncertainty visualisation, as it is now, should not exist.
+Any generalisable contribution made by these papers is already an established result in information visualisaiton, and therefore is not specific to "uncertainty visualisation".
+Any "uncertainty visualisation" specific result is not generalisable beyond the experimental conditions because of the poor formalisation in the definitions of uncertainty and uncertainty visualisations.
+
+This lack of formalisation is not unique to uncertainty visualisation and the wider field of information visualisation suffers form similar problems [@Kinkeldey2014].
+
+This makes it clear we have a serious problem with the classification of "uncertainty visualisation".
+The visualisations follow identical rules to those established already in the visualisation literature, and experiments that view uncertianty visualisation as somehow special or different from normal visualisation are simply repeating existing established research in information visualisation.
+
+If several graphics are compared that are different statistically, geometrically, and aesthetically, there is *no way to combine the results to establish generic principles for the field*.
+
+The overwhelming majority of these papers depict different mathematical expressions of both the signal and the uncertainty.
+In 100% of these papers, the plot that has mapped the statistic in question to some visual feature will outperform those that don't.
+When the statistics used to depict uncertainty *are* the same (which is rare) 100% of those papers will follow the hierarchy of perceptual tasks that has already been established in the visualisation literature.
+If the experiment participant needs to extract a probability and one graphic uses position along a common scale, while the other uses colour or literally any other perceptual task, the position depiction will always outperform the other.
+
+If we were to compare two different visualisations, we would like to change as few features as possible so we can better understand the cause of the difference in perception.
+
+Additionally, if information that can be used to calculate a statistic, but not the statistic itself, is shown, people will need to use a heuristic to extract that information which will decrease the precision with which that information is communicated.
+If the information that is depicted on a plot cannot be used to *mathematically* generate a best estimate for a specific statistic, it is questionable how uncertainty visualisation authors expect participants to cover that gap if *not* for heuristics.
+
+In these papers will typically ask separate questions about the signal (e.g. what is the mean of this distribution) and the uncertainty (e.g. what is the variance of this distribution).
+
+Failling to understand this key aspect of visualsiation leads to graphics that *contain* the correct amount of information, but that volume of information cannot be readily accessed by the viewer.
 

### 2024-02-22 paper.qmd
moved and reworked a bunch of the context section

@@
 
+## 3.3 Contextual Information
+Some uncertainty work, specifically uncertainty visualisation, focus on the context surrounding the uncertainty.
+These are things like spatial data, temporal data, category data, etc.
+While other discussions of uncertainty may consider this distinction more relevant to their work, here we will consider this contextual information.
+Contextual information is important for interpretation and understanding of graphics, but it does not in of itself generate uncertainty.
+This is not unique to the properties of spatial information, other statistics information that is important for modelling and assumptions, such as non-linearity, class levels, etc do not contribute to uncertainty.
+Operations on data, inherrent randomness, and assumptions that create bias generate uncertainty, which is what we are talking about here.
+A confidence interval is a confidence interval whether it came from spatial data, temporal, or cross sectional data.
+
+Papers that stress the importance of spatial or temporal contexts may be conflating contextual information with uncertainty.
+@fig-spatial depicts the role of contextual information in graphics, as something that helps us understand the information in the statistical graphic.
+When we create statistical models and quantify uncertainty, the spatial or temporal aspects are often boilled away [@Otsuka2023], and the only element of it that remains is the inexchangeability of our data.
+For this reason, there realistically is no such thing as "spatial" uncertainty in the same sense that there is no such thing as "photograph" uncertainty or "coin toss" uncertainty or "baby birth" uncertainty.
+Maps and photographs boil down to the same statistical objects in the same way that coin tosses and baby births boil down to the same statistical objects.
+Once we have removed the physical nature these objects are near identical when considered statistically.
+Of course the nature of them can change based on what we choose to observe (i.e.
+if we consider a coin landing on its side to be an outcome or if we consider intersex babies as an outcome in births) but this means the contextual information of our data, e.g.
+whether or not it is spatial or temporal, is only of importance at two stages of the data analysis pipeline.
+It is first important when we *boil it down* to our data through *observation* and again when we *map our statistical information back to its real world information* to *communicate* it.
+The only way the context uniquely contributes to uncertainty is in the "boil down" stage, however spatial or temporal uncertainty never refers to that specific cause of uncertainty.
+@Kinkeldey2014 discussed how uncertainty  can be represented by three components attribute (what) position (where) and temporal (when) and that studied typically deal with uncertainty around attribute but rarely position and time, however it is never specified what considerations should cause attribute uncertainty to be different to position or temporal uncertainty.
+@Kay2016 did an experiment that showed uncertainty around bus arrival times, however the visualisation used in the experiment, shown in @fig-bustime, is indistinguishable from most work that would be considered "attribute" uncertainty.
+
+![The uncertainty visualisation used by @Kay2016 to show the uncertainty around a bus arrival time prediction.](bustime.png){#fig-bustime}
+
+It is typically in reference to measurement error, data aggregation, statistical modelling, or other steps that occur *after* the boil down stage, and are therefore not unique to spatial or temporal data.
+While the spatial or temporal features *may* become relevant at the final communication stage, however this has little do with quantifying uncertainty, and the difficulties in communicating a complicated statistical model through a complicated map is already well discussed in the visualisation literature, and in later sections we will discuss how this concept is related to uncertainty visualisation.
+
+![Data pipeline with visualisations to show why we are ignoring contextual information for the time being.
+The visualisations that depict the data are similar to typical statistical graphics.
+The purpose of the spatial information is to provide context to the information we extract from the statistical aspect of the graphic.](spatialinfo.jpeg){#fig-spatial}
+
+This does not mean we are saying spatial and temporal information is never of importance.
+Spatial information is *highly relevant* when we translate from the abstract domain back into real world for communication.
+This position is best explained by @Leland2005 in *The Grammar of Graphics*:
+
+> "Geography is anchored in real space-time and statistics in abstract dimensions.
+This is a distinction along a continuum rather than a sharp break...
+but this difference in focus clearly means that a system optimized to handle geography will not be graceful when dealing with statistical graphics."
+
+This paper focuses on uncertainty, specifically uncertainty that is still in the the abstract *statistics* domain.
+Since we are still in the abstract domain, things that are anchored in real space-time create confusion that is evident in the literature.
+Spatial and temporal uncertainty papers limit how they communicate their uncertainty because they are considering what can be visualised inside the context provided by maps and, not what they should *try* to visualise based on the uncertainty abstraction.
+Additionally, since uncertainty is defined in the abstract domain, of the issues faced by spatial uncertainty visualisations are likely shared by other graphics that only differ in contextual information.
+
+In this sense we are not proposing that the current conceptualisation of data according to its physical properties is necessarily wrong, or that it leads to conclusions that are not necessarily specific to spatial or temporal data, and it may lead us to miss relevant information that is outside our domain.
+This is not only the case for uncertainty data.
+@Slingsby2023 discussed using a gridded glyphmap for spatial modelling, however the statistical elements are identical to a fluxuation diagram depicting $f(infection|time, age, longitude, latitude)$ presented in @Wickham2011.
+The plot by @Slingsby2023 is simply a special case of the graphic by @Wickham2011, where a separate visualisation is made for each time point, the continuous function $infection | age$ has replaced the categorical variable $happy$, a fade is added for the population size, and a map is shown in the background.
+@Slingsby2023 presents other maps in the paper, however they are also special cases of the product plots framework.
+This is not to say this extension is uninteresting, on the contrary we believe identifying cases where particular graphics are useful is a worthwhile endevour, but these plots seem to be have been developed unaware of the existing statistical framework.
+When we separate the field of visualisation along lines that may not be of particular relevance, we increase the likelihood of "reinventing the wheel" in each sub-dicipline.
+
+```{r}
+#| output: asis
+#| echo: false
+img_files <- fs::dir_ls("prodplots", glob="*.png")
+cat("::: {layout-ncol=2}\n",
+    glue::glue("![]({img_files})\n\n\n"),
+    ":::",
+    sep = ""
+)
+```
+
 # 4.General Attitude Towards Uncertainty (Uncertainty Communication Boradly)
@@
 
-## Perception Experiments and a Poor Definition
-Even the most simple experiments are not free from the issues in the uncertainty definition.
-
-Papers that consider unique aspects of uncertainty in their experiment such as experiemnts that identify the usefulness of specific perceptual tasks, experiments that identify problematic heuristics in commonly used uncertainty visualisations, and experiments that consider unique problems in uncertainty experiments all have useful results.
-These papers will be discussed in the upcomming sections.
-
-Perceptual task experiments that identify visual aesthetics that are intuitively connected to uncertainty are useful as they establish an inutive psychological connection.
-These papers are also not without their flaws, however their flaws have little to do with issues in the grammar of graphics as they are well defined in this space and more to do with the lack.
-Since uncertainty and signal are typically mapped to different aesthetics, it is not always clear what exactly the visualisation is suggesting we should be uncertain about, and when a perceptual task that should be considered something like a train line has "uncertainty" added to it, it is not
-
-- rarely used extension is aggregation of uncertainty over an area (retrieve overall estimation from a spatial distribution of uncertainty).
-Other tasks include comparisons, rankings.
-also have a task called "search" where participants have to identify entities that fulfill certainty characteristics (high or low values) @Kinkeldey2014
-- Whether to use classed or unclassed uncertainty is rarely discussed @Kinkeldey2014.
-e.g.
-land map with remotely sensed images has ambiguity in the boundaries, which can result from multiple sources (land cover class, measurement error, images from different dates).
-If you combine them all it might not make sense for someone who only needs one.
-
-- I think the dimension of the uncertainty should be considered every time there is aggregation from multiple sources of uncertainty and you want to maintain those sources.
-You can project the uncertainty down into one dimension if the source information is in of itself unimportant.
-
-### Contextual Inormation
-I have this here to explain why context based taxonomies can be ignored when we are considering uncertainty.
-I did not go into too much detail because it would involve discussing visualisation which I am trying to avoid at this point.
-I actually think this quote from the grammar of graphics is exactly what I was trying to say.
-*"The system in this book is not a model for a GIS, however, because geography and statistics differ in a crucial respect.
-Geography is anchored in real space-time and statistics in abstract dimensions.
-This is a distinction along a continuum rather than a sharp break; after all, there is a whole field called spatial statistics (Cressie, 1991).
-But this difference in focus clearly means that a system optimized to handle geography will not be graceful when dealing with statistical graphics"*.
-Basically, my point is this section is still discussing the definitions of unertainty, we have not gotten to disussing visualisation so we are still in the abstract *statistics* domain.
-The only parts of spatial statistics that we care about in this domain are the statistical properties that come from spatial information, and *none* of those statistical properties generate uncertainty.
-This is not unique to the properties of spatial information, other statistics information that is important for modelling and assumptions, such as non-linearity, class levels, etc do not contribute to uncertainty.
-Operations on data, inherrent randomness, and missing information generate uncertainty, which is what we are talking about here.
-A confidence interval is a confidence interval whether it came from spatial data, temporal, or cross sectional data.
-*This does not mean I am saying spatial and temporal information is not valuable ever*, I am just saying it is not valuable *yet*.
-This information will be *highly relevant* when we discuss context and translating from the abstract domain back into real world for communication.
-Since we are still in the abstract domain, things that are anchored in real space-time create confusion that is evident in the literature.
-The confusion that is created in the literature was going to be the topic I had later.
-Spatial and temporal uncertainty papers limit how they communicate their uncertainty because they are considering what can be visualised, not what they should *try* to visualise based on the uncertainty abstraction.
-Since the context is only relevant *with respect to uncertainty* when we visualise it, I was going to discuss this issue in the visualisation literature review section.
-
-The taxonomies that cannot be contained within the @utypo taxonomy are not a new uncertainty dimensions, but rather they are conflating adjacent information with uncertainty.
-A good example of this is the distinction between spatial, temporal, and estimated uncertainty.
-When we create statistical models and quantify uncertainty, the spatial or temporal aspects are often boilled away [@Otsuka2023], and the only element of it that remains is the inexchangeability of our data.
-For this reason, there realistically is no such thing as "spatial" uncertainty in the same sense that there is no such thing as "photograph" uncertainty or "coin toss" uncertainty or "baby birth" uncertainty.
-Maps and photographs boil down to the same statistical objects in the same way that coin tosses and baby births boil down to the same statistical objects.
-Once we have removed the physical nature these objects are near identical when considered statistically.
-Of course the nature of them can change based on what we choose to observe (i.e.
-if we consider a coin landing on its side to be an outcome or if we consider intersex babies as an outcome in births) but this means the contextual information of our data, e.g.
-whether or not it is spatial or temporal, is only of importance at two stages of the data analysis pipeline.
-It is first important when we *boil it down* to our data through *observation* and again when we *map our statistical information back to its real world information* to *communicate* it.
-The only way the context uniquely contributes to uncertainty is in the "boil down" stage, however spatial or temporal uncertainty never refers to that specific cause of uncertainty.
-It is typically in reference to measurement error, data aggregation, statistical modelling, or other steps that occur *after* the boil down stage, and are therefore not unique to spatial or temporal data.
-While the spatial or temporal features *may* become relevant at the final communication stage, however this has little do with quantifying uncertainty, and the difficulties in communicating a complicated statistical model through a complicated map is already well discussed in the visualisation literature, and in later sections we will discuss how this concept is related to uncertainty visualisation.
-
-In this sense we are not proposing that the current conceptualisation of data according to its physical properties is necessarily wrong, or that it leads to incorrect conclusions, we are merely suggesting for a different way of slicing the metaphysical cake of uncertainty that better aligns with what is *actually happening* when we perform a statistical analysis.
-This metaphysical cake is illustrated below as @fig-cake.
-In changing  what the actual boundaries are when we work with these problems, we improve our understanding of what we are actually doing when we conceptualise uncertainty, and in the process allow for more creative solutions creativity (e.g.
-by borrowing concepts from spatial visualisation to visualise uncertainty in photographs generated).
-However, since uncertainty is almost synonymous with the concept of "statistics", this adjustment in how we conceptualise and organise statistical methods would likely benefit other domains within the field.
-
-![The metaphysical cake we are slicing differently](metaphysicalcake.jpg){#fig-cake}
-
-- Specifies uncertainty can be represented by three components attribute (what) position (where) and temporal (when) @Kinkeldey2014
-- Studies dealing with uncertainty information typically deal with attribute but rarely position and time @Kinkeldey2014
-- 10/44 studies involve multiple types of uncertainty @Kinkeldey2014
-
-- When we boil down data into a statistical object, we boil away many aspects on context. Birth and coin flips become an object that is one in the same, and often the visualisation 
-- Why are spatial and temporal data considered special?
-What is the statistical element that remains after they have been away?
-Inexchangeability.
-Temporal data is one dimensional inexchangeability, spatial data has two dimensional inexchangeability.
-  - There are other things that have inexchangeability, and not every question related to spatial or temporal data requires the inexchangeability.
-- perceptual tasks need to know what you are uncertain about.
-- Yes the data type is relevant but its not the most important thing
-  - one with temporal, one with that temporal paper but you didn't know it was temporal, one with the bus time cdf
-  - screenshot from these three papers, the connecting factor is the question, not the data type
-  - The data type: spatial, temporal, etc, usually is mentioned as the 
-  - Not a feature of the data, but it is a feature of the question "how do I show uncertainty on a map?".
-This is a fine question to ask, but the lack of specificity in what is being done, and the zoning in on functionally irrelevant elements leads the literature as a huge mess.
-- Hurricane risk is not only the path, but also the storm surge and wind speed. @Spiegelhalter2017
-- coloured maps for earthquake risk are easily interpretable @Spiegelhalter2017
-
 ## Evaluation Experiments and The Grammar of Graphics
@@
 
+## Perception Experiments, Question Problems, and Task Specificity
+Uncertainty, like any other variable we are going to plot, has a natural mappings that should be considered when we express it in a plot.
+In the same way that mapping the colours red to hot and blue to cold, many variables, including uncertainty, have an intuitive mapping.
+
+The choice of what we map uncertainty to will impact how easily
+
+Even the most simple experiments are not free from the issues in the uncertainty definition.
+Some experiments directly relate the internal 
+Papers that consider unique aspects of uncertainty in their experiment such as experiemnts that identify the usefulness of specific perceptual tasks, experiments that identify problematic heuristics in commonly used uncertainty visualisations, and experiments that consider unique problems in uncertainty experiments all have useful results.
+These papers will be discussed in the upcomming sections.
+
+Perceptual task experiments that identify visual aesthetics that are intuitively connected to uncertainty are useful as they establish an inutive psychological connection.
+These papers are also not without their flaws, however their flaws have little to do with issues in the grammar of graphics as they are well defined in this space and more to do with the lack.
+Since uncertainty and signal are typically mapped to different aesthetics, it is not always clear what exactly the visualisation is suggesting we should be uncertain about, and when a perceptual task that should be considered something like a train line has "uncertainty" added to it, it is not
+
+-  Error is best mapped to fuzziness, location, and colour value; arrangement, size and transparency are an OK second choice; but saturation, hue, orientation and shape are unacceptable and have no intuitive connection to variance [@Maceachren2012].
+No only do the graphical elements we map our features to matter, but the direction matters too.
+Graphical elements that are more fuzzy (fuzziness), further from centre (location), lighter (colour value), poorly arranged (arrangement), smaller (size), more transparent (transparency) are perceived to be more uncertain [@Maceachren2012].
+This idea also extends to interval estimation, where questions about probability are best answered with gradients and questions about start and end times are easiest to answer with ambiguation [@Gschwandtnei2016].
+Heuristics can work against us just as much as they can work for us.
+The sine illusion can cause the confidence interval of a smoothed sine curve to seem wider at the peaks than the troughs, causing us to underestimate uncertainty associated with changing values [@Vanderplas2015]
+
+- @Nathonours found that a scatter plot is better than a line plot if you want to convey the correlation between two time series but we cannot be sure if this was influenced by swapping the distribution depicted or by dropping the irrelevant feature of inexchangeability.
+
+- rarely used extension is aggregation of uncertainty over an area (retrieve overall estimation from a spatial distribution of uncertainty).
+Other tasks include comparisons, rankings.
+also have a task called "search" where participants have to identify entities that fulfill certainty characteristics (high or low values) @Kinkeldey2014
+- Whether to use classed or unclassed uncertainty is rarely discussed @Kinkeldey2014.
+e.g.
+land map with remotely sensed images has ambiguity in the boundaries, which can result from multiple sources (land cover class, measurement error, images from different dates).
+If you combine them all it might not make sense for someone who only needs one.
+
+- I think the dimension of the uncertainty should be considered every time there is aggregation from multiple sources of uncertainty and you want to maintain those sources.
+You can project the uncertainty down into one dimension if the source information is in of itself unimportant.
+
 ## Noise and Signal in Uncertainty Visualisation
@@
 That is why the experimental results are identical to the existing information visualisation literature.
+
+The view of uncertainty as a dimensionality problem is not quite right.
+It is important to keep in mind that just because information is in a graphic, that does not mean it will be "seen".
+The phenomena of inattentional blindness shows that there is no perception without attention and it is powerful enough that participants can fail to be aware of random objects appearing on a screen or a gorilla walking through a basketball game [@simons1999gorillas; @mack2003inattentional].
+Therefore if information is mapped to graphical elements that are so low on hierarchy they can be ignored, they might as well not be there at all.
+The people who don't plot uncertainty because they think its unimportant and the people who relegate uncertainty to the lowest ranking aesthetic on the list are both saying the same thing, "I think the uncertainty is unimportant".
+Including uncertainty is worth very little if no attention is left to see it.
+This does not mean we cannot put a *large* amount of information in a graphic.
+Glyph maps can be used to depict the multi-dimensional information in spatial-temporal data by mapping line plots to map locations, but we still need to decide what information is important.
+The map can either present trends in the global or local variance depending on whether or not the line plot is scaled globally or locally [@Wickham2012], however smaller details are almost impossible to convey.
+No matter how much information we try and put in a plot, there will always be only a handful of key takeaways.
+
+- depicts a bivariate map which uses a bivariate colour palette that is created by blending two single hue colour palettes.
+One colour represents the variable of interest while the other represents the sampling error of that variable.
+There are two immediate problems with this method.
+First of all, uncertainty is being expressed with hue and saturation which @Maceachren2012 found to be the worst aesthetics to map to uncertainty to as they don't have an intuitive interpretation.
+Value has a natural connection to uncertainty (lighter values equate to higher uncertainty and darker values equate to more certainty) so it is a much more appropriate choice.
+While the `Vizumap` data does depict areas of light and darkness, they are largely irrelevant to the uncertainty measure causing our heuristics to lead us to the incorrect conclusions.
+The Value-Suppressing Uncertainty Palettes (VSUP) shown in @fig-vsup maps estimates to the hue and error to the value thereby creating a more intuitive plot [@Correll2018].
+Additionally, at high levels of uncertainty VSUP only has one output colour, which prevents viewers from decrypting any particular value and also avoids enforcing a binary encoding of significance [@Correll2018].
+Unfortunately VSUP are not easy to combine with packages like `Vizumap` which leaves it still somewhat difficult to express this encoding in practice, however the combination of a bivariate map with VSUP has shown to improve decisions in the face of uncertainty [@Correll2018].
+
+- The line-up protocol is an application on this idea and considers each visualisation to be a single outcome of some larger distribution [@Buja2009; @Wickham2010; @Chowdhury; @Hofmann2012].
+By generating a sample of visualisations from a hypothetical distribution, visualisation authors can check if perceived patterns are real or merely the result of chance.
+
+The final consideration when deciding what information to depict is whether or not additional information will clutter the graph.
+Our visualisation should aim to show enough information to solve a task while avoiding irrelevant distracting information [@kosslyn2006graph].
+It also seems that people will discount information they perceive as discountable if they are able.
+Including mean estimates on depictions of mass can cause people to discount the uncertainty information and use the difference between means as a proxy for the probability distribution [@Kale2021].
 
@@
 - Need to compare to an approripate ground truth, but this is a philisophical exercse @Hullman2016
+
+## Bad Question stuff?
+In the discussion of the HOPs, error bar, and violin plot study @Hullman2015 note:
+
+> "In light of the common usage of error bars for presenting multiple independent distributions, it is noteworthy how poorly subjects using these representations performed on tasks asking them to assess how reliably a draw from one variable was larger than the other...Many subjects reported values less than 50%, which are not plausible given that the mean of B was larger than the mean of A."
+
+- The way we ask questions and the types of questions we ask are selected with little justification. This paper makes suggestions to reduce the noise in the data from these papers. @Hullman2016
 

### 2024-02-22 writingdumpster.qmd
moved and reworked a bunch of the context section

@@
 - The big data paradox shrinks confidence intervals but magnifies bias [@Bradley2021] because the confidence intervals shrink due to the sample size but the imprecise target group creates bias.
-- The way we ask questions and the types of questions we ask are selected with little justification. This paper makes suggestions to reduce the noise in the data from these papers. @Hullman2016
+- coloured maps for earthquake risk are easily interpretable @Spiegelhalter2017
+- Hurricane risk is not only the path, but also the storm surge and wind speed. @Spiegelhalter2017
 

### 2024-02-22 paper.qmd
updated task specific section

@@
 Discussing risk and ambiguity aversion as though it is some diversion from a perfectly rational, mathematical choice, implies there is *no value to a decrease in variance which is not even true within mathematics*.
+There have been other discussions on the appropriateness of discussing risk-aversion as a bias [@Vranas2000], but few have made the point that if uncertainty holds some *value* there is not technically a "right" decision at all.
 The concept of risk aversion treats risk as "noise", as though it is something that only exists to be ignored, but if the common place use of biased MLE's is an indication of anything, it is that *ignoring uncertainty* is the irrational choice.
@@
 
+## Uncertainty and Task Specificity
+When we consider mapping uncertainty, a reasonable questions that should come to mind is "uncertainty about what?".
+Every single thing concept or activity in statistics does not have uncertainty.
+For example, descriptive statistics cannot have uncertainty as they are just describing statistics as they are, the sample mean is the sample mean, we are not uncertain about its value.
+A sample variance describes dispersion, which is also how we describe uncertainty, but the sample variance in of itself does not have any uncertainty.
+Most of the mathematical uncertainty concepts have this idea embedded in them.
+For example, a confidence interval is an interval drawn around the sample mean that will theoretically contain the *populaton mean* 95% of the time.
+This issue has been noticed several times, however it is often brought up as a *task* dependence rather than an understanding that uncertainty is inherrently linked to inference on a particular statistic.
+@Wallsten1997 argue that the best method for evaluating or combining subjective probabilities depends on the uncertainty the decision maker wants to represent and why it matters.
+@Munzner2009 created a nested model for visualisation that highlighted how the first mistake that can be made in a visualisation is in the problem characterisation, and failling to do it well can cause downstream effects and damage the effectivness of a visualisation.
+@Fischhoff2014 looks at uncertainty visualisation for decision making decides that we should have different ways of communicating uncertainty based off what the user is supposed to do with it.
+This should make it clear that uncertainty only exists when we are performing inference.
+Therefore it only makes sense to describe uncertainty in terms of something to be uncertain *about*, it is not some a latent feature of statistics and data.
+
+We feel the need to make this point clear because a *large* amount of uncertainty research seems to be confused by the purpose of uncertainty visualisation, what it is related to, and why we visualise it.
+Often, the authors are even more confused about what they are doing than the participants.
+There are enough papers that fail to understand these basic principals of uncertainty that this entire paper could simply be a list with explanations.
+Instead, we have a handful of highlights presented below.
+- @Zhao2023 displayed a model prediction with uncertainty and took participants using the model prediction as a sign of trust, failing to notice that there is no reason for uncertainty information to influence someone to *use their own prediction*.
+If the Beuro of Meterology was uncertain about their rain prediction, that would not be sufficient conditions for me to decide to make my own rain predictions.
+Despite this, the authors seemed to assume that the uncertainty information *should* do that.
+- @Hofman2020 asked questions about discusses the difference in showing inferential or outcome (prediction) intervals and then found that participants shown the inferential distribution performed poorly when asked questions about predictions.
+They found that showing the relevant distribution to the question reduced the error.
+This result should be astoundingly obvious if it was well understood that uncertainty is not some latent variable and is only relevant for specific ground truth statistics.
+This papers *percieved* importants highlights an underlying belief that uncertainty is linked to a specific variable or data set, not the variable we are drawing inference on.
+- @Padilla2021 found that high uncertainty in the model estimates (calculated uncertainty) and low forecaster confidence (which is typically an expression of suspected bias in the model) both caused participants to have decreased confidence in their results and suggested modelers express both if they are relevant.
+@Kale2019 discussed the importance of communicating decisions made in the data analysis pipeline and being aware of the alternatives.
+While there is nothing wrong with the core ideas of these papers, the fact that they were published indicates that the general population was not already aware that choices introduced early in the data analysis pipeline create bias and therefore uncertainty in our final values.
+This indicates a general failure to see bias and variance both as sources of uncertainty.
+- @Wickham2011 suggests their product plot framework, which includes histograms, should have a way to measure uncertainty, but does not consider that a histogram is *already* a depiction of mass and would already be considered an uncertainty visualisation were our statistic of interest the population mean.
+To suggest that a histogram needs uncertainty bounds is similar to suggesting a density or box plot need uncertainty bounds.
+- @Griethe2006 commented that "if visualization is used as a means to explore a data volume or to communicate its contents the uncertainty has to be included".
+@Potter2010 aimed to create a summary plot that "concisely presented data with uncertaninty information".
+These papers fail to realise the exploration step of an analysis, which includes descriptive statistics, exploratory data visualisaiton, and unsupervised machine learning techniques, is performed without a prior hypothesis.
+Therefore, it is performed to *suggest* a direction for future inference work or communication, which *do* contain uncertainty and does not contain uncertainty within itself.
+This confusion would not occur if it was understood that uncertainty does not exist without inference.
+- @Ibrekk1987 asked participants for the "best" estimate while displaying a population density function.
+The participants proided the mode instead of the mean, although there was no indication the mean was what was being asked about.
+- @Boukhelifa2012 tried to quantify the strength of the intuitive connection between "sketchiness" and uncertainty.
+Since they had seen literature that explained uncertainty was task specific, they tested the connection for multiple "tasks", what they failled to understand, is that the "task" dependency which is likely just a proxy of the evaluation experiment problem discussed in the previous section.
+This misunderstanding led to "sketchiness" being used to depict uncertainty in the categories of bar charts, in graphic networks, and in train lines, with little consideration or explanation as to *how* these things would be uncertainty.
+I imagine it is quite hard to be "uncertain" about the existence of a train line.
+Participants rightfully assumed the sketchiness therefore represented something else, such alternative options or simply ignored it.
+
+![The graphics displayed by @Boukhelifa2012 to identify if there is an intuitive connection between sketchiness and uncertainty.](sketchiness.png){#fig-sketchy}
+
+The reality is that a lot of uncertainty visualisation authors do not seem to have an intuitive understanding of uncertainties connection to inference, when inference is being performed, and how to design experiments that capture this relationship.
+
+This misunderstanding is not due to laziness or the fault of the authors, but rather is likely caused by the absence of a strict definition of uncertainty.
+The imprecise and confusing set of existing definitions are creating a field in which the authors themselves do not know what they are testing.
+They do not even seem to understand what uncertainty is.
+Uncertainty examples include imputed data, model selection, inherrent randomness, biased sampling, etc, not because these things *are* uncertainty, but because they *create* uncertainty when we perform inference.
+These things are not "uncertainty" in of themselves, but rather contribute to the distance between the final estimate and the statistic we want to perform inference on.
+The field of uncertainty communication is in desperate need of a unifying mathematical definition that will allow the bias and variance introduced through various stages of the data analysis pipeline to be compared or quantified.
+
+Most uncertainty definitions speak about this vague concept of uncertainty but do not
+
+The fact that uncertainty only exists when we are performing inference also creates issues with the questions of an experiment.
+If uncertainty is depicted and uncertainty is always related to .
+The authors of these experiments often comment on the results from the perspective that the participants have made a conceptual mistake, and at no point do they question why we would depict uncertainty for an unrelated ground truth.
+The results of these experiments essentially show that participants can be "tricked" by failling to understand uncertainty (showing that people can be tricked if you purposely lead them astray should not constitute interesting research).
+Ironically the authors fail to realise they have made this same mistake when they did not consider displaying uncertainty for a fundamentally unrelated
+
+assume the participants have made some sort of mistake, rather than questioning their own understanding of uncertainty.
+
+must be related to the statistic we wish to draw inference on
+
+Reguardless of whether or not the visualisation and associated uncertainty is related to the statistic we are trying to draw inference from, participants will *assume* they are.
+
+@Hofman2020 suggests authors depictions of the
+
+This concept is the first issue that uncertainty experiments run into, specifically in perceptual task experiments.
+
+It is rarely mentioned in the uncertainty literature, but
+
+The choice of what we map uncertainty to will impact how easily
+
+Even the most simple experiments are not free from the issues in the uncertainty definition.
+Some experiments directly relate the internal
+
+-  Error is best mapped to fuzziness, location, and colour value; arrangement, size and transparency are an OK second choice; but saturation, hue, orientation and shape are unacceptable and have no intuitive connection to variance [@Maceachren2012].
+No only do the graphical elements we map our features to matter, but the direction matters too.
+Graphical elements that are more fuzzy (fuzziness), further from centre (location), lighter (colour value), poorly arranged (arrangement), smaller (size), more transparent (transparency) are perceived to be more uncertain [@Maceachren2012].
+This idea also extends to interval estimation, where questions about probability are best answered with gradients and questions about start and end times are easiest to answer with ambiguation [@Gschwandtnei2016].
+Heuristics can work against us just as much as they can work for us.
+The sine illusion can cause the confidence interval of a smoothed sine curve to seem wider at the peaks than the troughs, causing us to underestimate uncertainty associated with changing values [@Vanderplas2015]
+
+- @Nathonours found that a scatter plot is better than a line plot if you want to convey the correlation between two time series but we cannot be sure if this was influenced by swapping the distribution depicted or by dropping the irrelevant feature of inexchangeability.
+
+- rarely used extension is aggregation of uncertainty over an area (retrieve overall estimation from a spatial distribution of uncertainty).
+Other tasks include comparisons, rankings.
+also have a task called "search" where participants have to identify entities that fulfill certainty characteristics (high or low values) @Kinkeldey2014
+- Whether to use classed or unclassed uncertainty is rarely discussed @Kinkeldey2014.
+e.g.
+land map with remotely sensed images has ambiguity in the boundaries, which can result from multiple sources (land cover class, measurement error, images from different dates).
+If you combine them all it might not make sense for someone who only needs one.
+
+- I think the dimension of the uncertainty should be considered every time there is aggregation from multiple sources of uncertainty and you want to maintain those sources.
+You can project the uncertainty down into one dimension if the source information is in of itself unimportant.
+
 ## Evaluation Experiments and The Grammar of Graphics
@@
 So, if we cannot see any difference to uncertainty information when it coems to value abstraction, *when* is uncertainty information different?
-
-## Perception Experiments, Question Problems, and Task Specificity
-Uncertainty, like any other variable we are going to plot, has a natural mappings that should be considered when we express it in a plot.
-In the same way that mapping the colours red to hot and blue to cold, many variables, including uncertainty, have an intuitive mapping.
-
-The choice of what we map uncertainty to will impact how easily
-
-Even the most simple experiments are not free from the issues in the uncertainty definition.
-Some experiments directly relate the internal 
-Papers that consider unique aspects of uncertainty in their experiment such as experiemnts that identify the usefulness of specific perceptual tasks, experiments that identify problematic heuristics in commonly used uncertainty visualisations, and experiments that consider unique problems in uncertainty experiments all have useful results.
-These papers will be discussed in the upcomming sections.
-
-Perceptual task experiments that identify visual aesthetics that are intuitively connected to uncertainty are useful as they establish an inutive psychological connection.
-These papers are also not without their flaws, however their flaws have little to do with issues in the grammar of graphics as they are well defined in this space and more to do with the lack.
-Since uncertainty and signal are typically mapped to different aesthetics, it is not always clear what exactly the visualisation is suggesting we should be uncertain about, and when a perceptual task that should be considered something like a train line has "uncertainty" added to it, it is not
-
--  Error is best mapped to fuzziness, location, and colour value; arrangement, size and transparency are an OK second choice; but saturation, hue, orientation and shape are unacceptable and have no intuitive connection to variance [@Maceachren2012].
-No only do the graphical elements we map our features to matter, but the direction matters too.
-Graphical elements that are more fuzzy (fuzziness), further from centre (location), lighter (colour value), poorly arranged (arrangement), smaller (size), more transparent (transparency) are perceived to be more uncertain [@Maceachren2012].
-This idea also extends to interval estimation, where questions about probability are best answered with gradients and questions about start and end times are easiest to answer with ambiguation [@Gschwandtnei2016].
-Heuristics can work against us just as much as they can work for us.
-The sine illusion can cause the confidence interval of a smoothed sine curve to seem wider at the peaks than the troughs, causing us to underestimate uncertainty associated with changing values [@Vanderplas2015]
-
-- @Nathonours found that a scatter plot is better than a line plot if you want to convey the correlation between two time series but we cannot be sure if this was influenced by swapping the distribution depicted or by dropping the irrelevant feature of inexchangeability.
-
-- rarely used extension is aggregation of uncertainty over an area (retrieve overall estimation from a spatial distribution of uncertainty).
-Other tasks include comparisons, rankings.
-also have a task called "search" where participants have to identify entities that fulfill certainty characteristics (high or low values) @Kinkeldey2014
-- Whether to use classed or unclassed uncertainty is rarely discussed @Kinkeldey2014.
-e.g.
-land map with remotely sensed images has ambiguity in the boundaries, which can result from multiple sources (land cover class, measurement error, images from different dates).
-If you combine them all it might not make sense for someone who only needs one.
-
-- I think the dimension of the uncertainty should be considered every time there is aggregation from multiple sources of uncertainty and you want to maintain those sources.
-You can project the uncertainty down into one dimension if the source information is in of itself unimportant.
 

### 2024-02-22 writingdumpster.qmd
updated task specific section

@@
 For example, when we record coin flips we typically want to model the behavior with a binomial distribution, so we ignore the outcome of a coin landing on its side despite the fact it is a real world possibility.
-@Wallsten1997 argue that the best method for evaluating or combining subjective probabilities depends on the uncertainty the decision maker wants to represent and why it matters.
-@Munzner2009 created a nested model for visualisation that highlighted how pitfalls in upstream levels, such as problem characterisation and data abstraction, can cause downstream effects and damage the effectivness of a visualisation.
-@Fischhoff2014 looks at uncertainty visualisation for decision making decides that we should have different ways of communicating uncertainty based off what the user is supposed to do with it.
 
@@
 - Distinctions at this taxonomy level seem to be important also untangling helps to understand it
-  - @Padilla2021 found that low forecaster confidence or high model uncertainty both contribute to more conservative judgements by decision makers.
   - Failing to communicate the nature of your uncertainty can result in underestimation or overestimation of failure probabilities [@Kiureghian2009]

### 2024-02-23 paper.qmd
finished the task wordvomit and future work  sections

@@
 date: last-modified
+toc: true
+number-sections: true
 format: pdf
@@
 
-# 1. Introduction
-::: aside
-I have not touched this section since the confirmation so it would be best to ignore it
-:::
+# Introduction
+
+**Aside: I have not touched this section since the confirmation so it would be best to ignore it**
 
@@
 ## 3.1 Coloquial Definitions
-::: aside
-I am not sure how to collect these together and show that they are highly varied.
+**Aside: I am not sure how to collect these together and show that they are highly varied.
 This might be a situation for the tidytext clustering or something.
-There is an additional problem is that the definition for uncertainty is *completely* unrelated to the uncertainty that is quantified in the graphics.
-I think the tidytex can help with the definitions being all over the place, but I am not sure how to identify the lack of correlation between the colloquial definitions of uncertainty and the mathematical quantificaitons.
-:::
+There is an additional problem is that the definition for uncertainty is completely unrelated to the uncertainty that is quantified in the graphics.
+I think the tidytex can help with the definitions being all over the place, but I am not sure how to identify the lack of correlation between the colloquial definitions of uncertainty and the mathematical quantificaitons.**
 
@@
 ### 3.2.1 Taxonomies of Uncertainty
-::: aside
-I actually think I agree with the sentiment in the grammar of graphics that a taxonomy is more of an opinion and we can reclassify it and is therefore not thorough enough to remain unchanged and be build upon.
+**Aside: I actually think I agree with the sentiment in the grammar of graphics that a taxonomy is more of an opinion and we can reclassify it and is therefore not thorough enough to remain unchanged and be build upon.
 That being said, if this is a literature review and all existing literature use written definitions or taxnomies, that is what I have to work with.
-The larger taxonomy can be used to show that the other taxonomies fit within it, but it is in of itself not tied to anything except work *that paper itself* identifies as transient.
-:::
+The larger taxonomy can be used to show that the other taxonomies fit within it, but it is in of itself not tied to anything except work that paper itself identifies as transient.**
 
@@
 
-::: aside
-This figure is not going to be in the final version of the paper.
+**Aside: This figure is not going to be in the final version of the paper.
 It is here as a placeholder.
-Ideally, I would replace with with a graphic that shows where all these taxonomies land in the location, level, nature taxonomy (since it is encompassing) and show they are very inconsistent and a new definition is pulled out for almost every paper (even in statistics papers)
-:::
+Ideally, I would replace with with a graphic that shows where all these taxonomies land in the location, level, nature taxonomy (since it is encompassing) and show they are very inconsistent and a new definition is pulled out for almost every paper (even in statistics papers)**
 
@@
 
-::: aside
-I had this here to explain why these are the considerations people make in statistics and how they relate to elements of the taxonomy to try and provide some justification, but If I am just going to show the definitions are inconsitent, I might get rid of this section.
-I also largely included it so you guys can see what the taxonomies are considering
-:::
+**Aside: I had this here to explain why these are the considerations people make in statistics and how they relate to elements of the taxonomy to try and provide some justification, but If I am just going to show the definitions are inconsitent, I might get rid of this section.
+I also largely included it so you guys can see what the taxonomies are considering**
 
@@
 
-::: aside
-I should add something in about the actual quantified uncertainty vs the discussed uncertanty thing (i.e. what is defined with words and what is defined with maths)
-:::
+**Aside: I should add something in about the actual quantified uncertainty vs the discussed uncertanty thing (i.e. what is defined with words and what is defined with maths)**
 
@@
 
-*this is all currently from the spieglehatter paper because that is the only paper I have moved notes over from mendeley before I started doing the summary write up.
-this will not all be from one paper, dont worry.*
+**Aside:this is all currently from the spieglehatter paper because that is the only paper I have moved notes over from mendeley before I started doing the summary write up.
+this will not all be from one paper, dont worry.**
 
-::: aside
-This section still needs some citations from papers that communicate uncertainty, but the are *hard* to organise because of the ambiguous uncertainty definition problem.
-So they will all be communicating something different but you cant find out what it is until you are knee deep in the method section.
-:::
+**Aside:This section still needs some citations from papers that communicate uncertainty, but the are *hard* to organise because of the ambiguous uncertainty definition problem.
+So they will all be communicating something different but you cant find out what it is until you are knee deep in the method section.**
 
@@
 
-::: aside
-I should maybe change this to relatability to uncertainty, value extraction, decision making and trust, and meta papers or something. Maybe perception and evaluation papers?
-:::
+**Aside: I should maybe change this to relatability to uncertainty, value extraction, decision making and trust, and meta papers or something. Maybe perception and evaluation papers?**
 
@@
 
-## Uncertainty and Task Specificity
+## Uncertainty and Inference
 When we consider mapping uncertainty, a reasonable questions that should come to mind is "uncertainty about what?".
@@
 To suggest that a histogram needs uncertainty bounds is similar to suggesting a density or box plot need uncertainty bounds.
-- @Griethe2006 commented that "if visualization is used as a means to explore a data volume or to communicate its contents the uncertainty has to be included".
+- @`Griethe2006 commented that "if visualization is used as a means to explore a data volume or to communicate its contents the uncertainty has to be included".
 @Potter2010 aimed to create a summary plot that "concisely presented data with uncertaninty information".
@@
 The imprecise and confusing set of existing definitions are creating a field in which the authors themselves do not know what they are testing.
-They do not even seem to understand what uncertainty is.
-Uncertainty examples include imputed data, model selection, inherrent randomness, biased sampling, etc, not because these things *are* uncertainty, but because they *create* uncertainty when we perform inference.
+Uncertainty examples include imputed data, model selection, inherent randomness, biased sampling, etc, not because these things *are* uncertainty, but because they *create* uncertainty *when we perform inference*.
 These things are not "uncertainty" in of themselves, but rather contribute to the distance between the final estimate and the statistic we want to perform inference on.
-The field of uncertainty communication is in desperate need of a unifying mathematical definition that will allow the bias and variance introduced through various stages of the data analysis pipeline to be compared or quantified.
+The field of uncertainty communication is in desperate need of a unifying mathematical definition that firmly identifies exactly how each of these things contribute to uncertainty.
+This may be a considerable task, but as of right now, the list of things that are "uncertainty" continues to grow, and the best methods of uncerstanding and quantifying them are ad-hoc at best.
+@Thomson2005 suggests a mathematical formula for *examples* of uncertainty, @Meng2014 mathematically defined the variance introduced to a model by the array of model choices, information theory tries to quantify uncertainty using the idea of entropy.
+None of these existing methods are thorough enough for an analyst to understand what causes uncertainty, and quantify it for communication.
+This limitation becomes apparent when you look at the limited scope of what the uncertainty visualisation literatrue has *actually* focused on visualising.
+Almost every uncertainty visualisation paper contains a broad and encapsulating definition of uncertainty in the introduction, followed by a visualisation of a normally distributed probablity density function.
 
-Most uncertainty definitions speak about this vague concept of uncertainty but do not
+**Aside: Include table that shows broad uncertainty definition and then PDF visualisation**
 
-The fact that uncertainty only exists when we are performing inference also creates issues with the questions of an experiment.
-If uncertainty is depicted and uncertainty is always related to .
-The authors of these experiments often comment on the results from the perspective that the participants have made a conceptual mistake, and at no point do they question why we would depict uncertainty for an unrelated ground truth.
-The results of these experiments essentially show that participants can be "tricked" by failling to understand uncertainty (showing that people can be tricked if you purposely lead them astray should not constitute interesting research).
-Ironically the authors fail to realise they have made this same mistake when they did not consider displaying uncertainty for a fundamentally unrelated
+In order to visualise something, it needs to be quantifiable.
+In order to quanitfy something, it needs to be mathematically defined.
+The broad classes of what is consdiered "uncertainty" is not currently quantifiable in any way that is not ad-hoc.
+The only apsects of uncertaitny that are currently quantifiable are confidence intervals, prediction intervals, and related terms.
+Even then, these quantifications often only capture the uncertainty in our model or its assumptions, it ignores bias often introduced in earlier stages of the analysis.
+Broader and more complicated concepts such as the effects of assumptions, imputed missing variables, and model choices remain difficult to meaningfully quantify beyond ad-hoc methods.
+This means that the current literature of uncertainty visualisation papers are doing one of two things:
+1) Evaluating multiple visual representations of a single mathematical expression of uncertainty (e.g.
+a PDF, confidence interval or error).
+These papers are purposefully vague, and would be clearer in their intention if the title was aligned with the actual mathematical object they are visualising.
+For example, a paper that is titled "evaluation of uncertainty visualisations" that only considers PDF visualisation should realistically be titled "evaluation of PDF visualisations".
+2) Evaluating multiple visual representations of *multiple* different mathematical expression of uncertainty.
+Since these experiments visualise different mathematical objects, some of which are closer to the ground truth statistic, the results of these papers can be easily predicted with an understanding of how to get the ground truth statistic from a plot and the perceptual tasks required to do so.
 
-assume the participants have made some sort of mistake, rather than questioning their own understanding of uncertainty.
-
-must be related to the statistic we wish to draw inference on
-
-Reguardless of whether or not the visualisation and associated uncertainty is related to the statistic we are trying to draw inference from, participants will *assume* they are.
-
-@Hofman2020 suggests authors depictions of the
-
-This concept is the first issue that uncertainty experiments run into, specifically in perceptual task experiments.
-
-It is rarely mentioned in the uncertainty literature, but
-
-The choice of what we map uncertainty to will impact how easily
-
-Even the most simple experiments are not free from the issues in the uncertainty definition.
-Some experiments directly relate the internal
-
--  Error is best mapped to fuzziness, location, and colour value; arrangement, size and transparency are an OK second choice; but saturation, hue, orientation and shape are unacceptable and have no intuitive connection to variance [@Maceachren2012].
-No only do the graphical elements we map our features to matter, but the direction matters too.
-Graphical elements that are more fuzzy (fuzziness), further from centre (location), lighter (colour value), poorly arranged (arrangement), smaller (size), more transparent (transparency) are perceived to be more uncertain [@Maceachren2012].
-This idea also extends to interval estimation, where questions about probability are best answered with gradients and questions about start and end times are easiest to answer with ambiguation [@Gschwandtnei2016].
-Heuristics can work against us just as much as they can work for us.
-The sine illusion can cause the confidence interval of a smoothed sine curve to seem wider at the peaks than the troughs, causing us to underestimate uncertainty associated with changing values [@Vanderplas2015]
-
-- @Nathonours found that a scatter plot is better than a line plot if you want to convey the correlation between two time series but we cannot be sure if this was influenced by swapping the distribution depicted or by dropping the irrelevant feature of inexchangeability.
-
-- rarely used extension is aggregation of uncertainty over an area (retrieve overall estimation from a spatial distribution of uncertainty).
-Other tasks include comparisons, rankings.
-also have a task called "search" where participants have to identify entities that fulfill certainty characteristics (high or low values) @Kinkeldey2014
-- Whether to use classed or unclassed uncertainty is rarely discussed @Kinkeldey2014.
-e.g.
-land map with remotely sensed images has ambiguity in the boundaries, which can result from multiple sources (land cover class, measurement error, images from different dates).
-If you combine them all it might not make sense for someone who only needs one.
-
-- I think the dimension of the uncertainty should be considered every time there is aggregation from multiple sources of uncertainty and you want to maintain those sources.
-You can project the uncertainty down into one dimension if the source information is in of itself unimportant.
+It is clear that the vague definition is causing across the board misunderstandings of uncertainty, confusing experiment methodologies and motivations, and a literature that is difficult to navigate.
+This is also causing a large body of experiments that have easily predictable results that can be easily explained by the existing visualisation.
+This problem will be the focus of the following section.
 
@@
 
-::: aside
-I think I want the "in order to visualise uncertainty you need to quantify it and the quantification does not align with the definition" point here since this section is now about value extraction.
-:::
+**Aside: I think I want the "in order to visualise uncertainty you need to quantify it and the quantification does not align with the definition" point here since this section is now about value extraction.**
 
@@
 Most commonly, these value abstraction papers compare two different visualisations of uncertainty on a set of these questions, however, since there is no rules for what is or is not an uncertainty visualisation the graphics that are compared often differ in both computation *and* information.
+@Hullman2018 asks participants questions that would require a prediction interval but provided a sampling distribution and argued that the two could be considered equivalent so long as participants understood the difference between a confidence interval and a prediction interval, knew the sample size, and knew the formula to convert the confidence interval to a prediction interval.
+The authors themselves note that this is rediculous, which begs the question.
 Uncertainty visualisaiton papers only seem to require that the plots depict "uncertainty" which has already been established to be poorly defined.
@@
 
-**INCLUDE A GRAPHIC THAT SHOWS WHERE EACH TYPE OF UNCERTAINTY VISUALISATION SITS IN THE GRAMMAR OF GRAPHICS AND HOW DIFFERENT THEY ARE ACROSS MULTIPLE STUDIES**
+**Aside: INCLUDE A GRAPHIC THAT SHOWS WHERE EACH TYPE OF UNCERTAINTY VISUALISATION SITS IN THE GRAMMAR OF GRAPHICS AND HOW DIFFERENT THEY ARE ACROSS MULTIPLE STUDIES**
 
@@
 
+- rarely used extension is aggregation of uncertainty over an area (retrieve overall estimation from a spatial distribution of uncertainty).
+Other tasks include comparisons, rankings.
+also have a task called "search" where participants have to identify entities that fulfill certainty characteristics (high or low values) @Kinkeldey2014
+- Whether to use classed or unclassed uncertainty is rarely discussed @Kinkeldey2014.
+e.g.
+land map with remotely sensed images has ambiguity in the boundaries, which can result from multiple sources (land cover class, measurement error, images from different dates).
+If you combine them all it might not make sense for someone who only needs one.
+
 - depicts a bivariate map which uses a bivariate colour palette that is created by blending two single hue colour palettes.
@@
 Unfortunately VSUP are not easy to combine with packages like `Vizumap` which leaves it still somewhat difficult to express this encoding in practice, however the combination of a bivariate map with VSUP has shown to improve decisions in the face of uncertainty [@Correll2018].
+
+- @Patrick2023 investigated peoples ability to identify a signal and compared it to typical hypothesis tests
 
@@
 
-# 5 Great Examples
+# Great Examples
 - best things you have seen and why they are the best
@@
 
-# 5. Future work/Conclusion
+# Future work/Conclusion
+This paper has identified several issues in the uncertainty visualisation literature and with those issues, we have several suggestions for future work that would likely be fruitful for the field.
 
-# Appendix (or information I am holding)
-## Experiment paper checklist
-General
-- Specific Field of Research
-- Question asked
-- Risk or Uncertainty: 
-- Noise vs signal (risk vs uncertainty)
-- Major task goal
-    - e.g. identify a bias or misunderstanding, compare two visualisation methods
-- Minor task goal
-    - Communication(value retrieval)
-    - Analytical(complex consideration)
-    - Exploration(free for all)
+Our first suggestion for future work is a mathematical defintion of uncertainty.
+Specifically, a definition that has a place for all the concepts that currently sit under the "uncertainty" umbrella, explains how these concepts relate to uncertianty, and also allows us to quantify them in a way where they can be combined or visualised.
+This work would not only benefit uncertainty visualisaiton, but also statistics more broadly.
+There is currently a lot of work that quantifies the uncertainty through bias and variance at various stages of the data analysis pipeline separately.
+Elements such as imputed data, assumptions, sampling methods, and analyst decisions all have their own independent work, quantification and discussions, but methods to brand between these concepts are few and far between.
+A conceptual framework that allows us to combine these results would be incredibly beneficial to the field.
 
-Uncertainty Considerations
-	- Source
-	- Level (statistical-ignorance)
-	- Nature (Epistemic/Aleatory)
+The concept of uncertainty should also be formalised within the grammar of graphics.
+@Leland2005 referred to uncertainty as "semantics", however the depth of integration required to understand problems in uncertainty visualisation suggest they *need* to be formally placed within the framework.
+Displaying a distribution is not as simple as displaying the data.
+For example, the distinction between a sample and a smothed density function is not cut and dry.
+One may view them both as the same data but one is represented as a point, the other is binned according to some function and then represented as a line.
+However, this distinction blurs as n gets larger and larger and, depending on the collision modifier used, depictions of the sample will appear identical to depictions of the mass.
+This is obviously true due to large sample theory, the backbone of statistics.
+The issue of blurred elements of the grammar of graphics in the case of distribution visualisations does not end here.
+Visualisations that depict mass such as a confidence interval, a boxplot, a PDF, a letterbox plot, a violin plot, a histogram, a quantile dot plot, etc, all show some version of the mass of a variable at different resolutions, but whether these veriables differ in something as low down in the grammar pipeline as the variable stage, or later at the geometry stage is unclear.
+The lack of formalisation in this way allows authors to compare graphics that *seem* to be similar mathematically but are actually conveying very different information.
+Not only would a formalisation of uncertainty visualisation within the grammar of graphics allow us to iron out some of these confusions, but they would make it easier to understand when existing visualisation methods apply and can be used to explain our results.
 
-Visual Uncertainty taxonomy
-- Explicit/implicit
-    - (directly mapping or showing multiple outcomes)
-- Intrinsic/Extrinsic 
-    - (using existing symbols e.g. colour value, or new objects e.g. grids)
-- Visually integral/separable 
-    - (can or cannot be separated from the data) 
-- Coincidence/adjacent 
-    - (if data and uncertainty are in integrated or separate views
-- Static/dynamic 
-    - (animation/interaction)
+```{r}
+# visualisation of a dot plot with increasing sample size
+```
 
-Other plot considerations
-	- Dimension of data
-	- Dimension of uncertainty
-	- Feature mappings
+If we are going to consider uncertainty as noise, not signal, there needs to be a way to identify this signal supression in an experimental design.
+It is clear that value retrival and questions with a specific ground truth are not appropriate, as uncertainty visualisation will simply follow the existing visualisation rules.
+Testing a secondary feature of a plot, that is, how strong a signal is becomes a little more complicated.
+Line up protocols have been used in adjacent work that look at the strength of the signal in a plot, and this idea of identifying if plots have some "barely noticiable differences" could be utilised.
+There is also the possibility that uncertainty visualisation evaluations will need to swap to a qualitative methodology where participants are allowed to freely comment on what they notice in graphics until we establish how the existence of noise can be observed.
 
-Extra (not necessarily recorded)
-	- Metrics used (and recorded)
-	- Possible Heuristics
-	- What is the ground truth?
-	- Participant literacy
+Additionally, information visualisation needs a more precise concept of the difference between two plots, and that framework should be utilised in experiments.
+Other fields of science employ marginal changes when designing experiments to ensure it is well understood *what* aspect of their experiment is contributing to their results.
+Visualisation has the grammar of graphics, however we have already discussed how this can break down in the case of distribution visualisation.
+
+If a visualisation researcher would prefer to perform experiments rather than formalise methods, the task dependency many authors in uncertainty visualisation mention would be a useful direction for research.
+Unfortunately the current visualisation evaluation literature is *far* too noisy to allow us to identify what might be the cause of this task dependence.
+It is unclear if it is a conflation between task and the ground truth/displayed statistics concepts discussed earlier, or an actual difference in the way we percieve graphics.
+A good starting place for this idea would be to identify if there is task dependency in perceptual tasks, since all perceptual task research to date has been done on value retrival.
+That is, an experiment that compares things like area, position, colour and other perceptual tasks on value exaction, comparisons, and other simple tasks.
+It is also clear that the the number of potential tasks that can be performed on a visualisation increases with with the number of observations.
+A single observation is limited to value extraction, two observations can be compared, multiple observations allow for shapes or global statistics to be extracted.
+The interaction between sample size and task is of particular interest to the uncertainty visualisation community, as uncertainty can be expressed through multiple observations using a sample, or through a single value using an error.
+Of course, this is limited by the fact that there also isnt a definition for what is a "task" and given the mess created by the lack of formalisation in uncertainty visualisation, it may be wise to formalise that concept before performing these experiemnts.
+Reguardless, these are directions of research would be fruitful to the uncertainty visualisation community even if it appears on the surface to be research that is only beneficial to the "normal" visualisation community.
 

### 2024-02-23 writingdumpster.qmd
finished the task wordvomit and future work  sections

@@
 # Moved over notes
+- Error is best mapped to fuzziness, location, and colour value; arrangement, size and transparency are an OK second choice; but saturation, hue, orientation and shape are unacceptable and have no intuitive connection to variance [@Maceachren2012].
+- @Nathonours found that a scatter plot is better than a line plot if you want to convey the correlation between two time series but we cannot be sure if this was influenced by swapping the distribution depicted or by dropping the irrelevant feature of inexchangeability.
+- No only do the graphical elements we map our features to matter, but the direction matters too.
+Graphical elements that are more fuzzy (fuzziness), further from centre (location), lighter (colour value), poorly arranged (arrangement), smaller (size), more transparent (transparency) are perceived to be more uncertain [@Maceachren2012].
+- This idea also extends to interval estimation, where questions about probability are best answered with gradients and questions about start and end times are easiest to answer with ambiguation [@Gschwandtnei2016].
+- Heuristics can work against us just as much as they can work for us.
+The sine illusion can cause the confidence interval of a smoothed sine curve to seem wider at the peaks than the troughs, causing us to underestimate uncertainty associated with changing values [@Vanderplas2015]
 - mentions uncertainty can be variance, precision, accuracy, reliability, or related concepts @Hullman2016
@@
 - Hurricane risk is not only the path, but also the storm surge and wind speed. @Spiegelhalter2017
+
+# Experiment paper checklist
+General
+- Specific Field of Research
+- Question asked
+- Risk or Uncertainty: 
+- Noise vs signal (risk vs uncertainty)
+- Major task goal
+    - e.g. identify a bias or misunderstanding, compare two visualisation methods
+- Minor task goal
+    - Communication(value retrieval)
+    - Analytical(complex consideration)
+    - Exploration(free for all)
+
+Uncertainty Considerations
+	- Source
+	- Level (statistical-ignorance)
+	- Nature (Epistemic/Aleatory)
+
+Visual Uncertainty taxonomy
+- Explicit/implicit
+    - (directly mapping or showing multiple outcomes)
+- Intrinsic/Extrinsic 
+    - (using existing symbols e.g. colour value, or new objects e.g. grids)
+- Visually integral/separable 
+    - (can or cannot be separated from the data) 
+- Coincidence/adjacent 
+    - (if data and uncertainty are in integrated or separate views
+- Static/dynamic 
+    - (animation/interaction)
+
+Other plot considerations
+	- Dimension of data
+	- Dimension of uncertainty
+	- Feature mappings
+
+Extra (not necessarily recorded)
+	- Metrics used (and recorded)
+	- Possible Heuristics
+	- What is the ground truth?
+	- Participant literacy
 

### 2024-02-25 paper.qmd
small changes

@@
 ---
-
-# Introduction
+# Background
+## Introduction
 
@@
 
-# 2. Similar work
+## Similar work
 There are a wealth of uncertainty communication and visualisation literature reviews that already exist. Interestingly they all seem to have identified similar pitfalls or fallen into one themselves.
@@
 None of these existing methods are thorough enough for an analyst to understand what causes uncertainty, and quantify it for communication.
+Realistically, this problem is well known, and put simply by Freeman Dyson in his fameous Birds and Frogs speech:
+
+> Rigorous theorems are the best way to give a subject intellectual depth and precision.
+Until you can prove rigorous theorems, you do not fully understand the meaning of your concepts.
+- Freeman Dyson (Birds and Frogs speech)
+
 This limitation becomes apparent when you look at the limited scope of what the uncertainty visualisation literatrue has *actually* focused on visualising.

### 2024-02-26 paper.qmd
some word vomit of noise and signal, also some added citations

@@
 ## Noise and Signal in Uncertainty Visualisation
-Uncertainty is noise, not signal.
-While this comment may seem obvious, what may not be obvious is that the value abstraction experiments are treating uncertainty as *signal*.
-That is why the experimental results are identical to the existing information visualisation literature.
+Uncertainty is noise.
+This may be an obvious statement, but what is not obvious is that the existing literature sees noise as a signal.
+When we directly ask "What is the varinance of this variable" the variance *becomes* the signal.
+Noise and signal are not an inherrent property of variance, or meta-data or the aspects of an analysis that are usually secondary considerations, it is a property of the data that *promotes* a message we are trying to infer and the data that *supresses* it.
+The replication problem in uncertainty visualisation, where experiments are simply repeating established visualisation work, comes from the misunderstandings established in previous sections.
+Uncertainty is poorly defined, therefore authors do not understand it is not an inherrent property of the data and is actually related to a particular inference question, this incorrect view that uncertainty is always variance or probabilities leads to the class of "uncertainty visualisations" that depict different mathematical objects or visual features with little in the sense of consistent rules, this complicated class has caused visualisation authors to miss that they are simply repeating the established evaluation results from normal information visualisations.
+This situation is a bit of a mess.
+It also highlights a unique and facinating problem faced by uncertainty visualisation.
+If asking direct questions about uncertainty causes us to treat it as a signal, how do we evaluate uncertainty as *noise*?
+
+Uncertainty evaluation studies are not all value extraction.
+There are cases where authors *know* the uncertainty information is relevant, but are not quite sure how to capture it.
+These studies most commonly evaluate a visualiation on its ability to convey information for a decision or how well it induces trust.
+
+*decision and trust paragraph*
+
+There are other studies that try to include uncertainty information without directly asking for it.
+This is an interesting approach, but these papers are often still looking for a "signal" which is specified in the form of a "ground truth" that is used to evaluate the plot.
+Unfortunately this usually comes in the form of slightly cryptic questions that create a large amount of noise on the interpretation side [@Hullman2016].
+- @Hofmann2012 showed two distributions in 20 different visualisations (a lineup protocol) using a jittered sample, a density plot, a histogram, and a box plot.
+Participants were asked to report in which of the plots was "the blue group furthest to the right" and to provide reasoning from a multiple choice list and grade their certainty.
+The experiment set up is shown in @fig-right.
+The participants answers were then compared to a ground truth where the correct plot had a right shifted mean.
+By comparing the results to a ground truth statistic and marking participants as "wrong" or "right", the error from the participants that had an alternative interpretation to the concept of "furthest right" was conflated with the error from a poor design.
+
+![This shows the user interface for the experiment performed by @Hofmann2012. The question of "furthest to the right" is open to interpretation. ](furthestright.png){#fig-right}
+
+Additionally, the use of a ground truth highlights how participants were supposed to *use* the uncertainty information. In this informa
+
+- The way we ask questions and the types of questions we ask are selected with little justification. This paper makes suggestions to reduce the noise in the data from these papers. @Hullman2016
 
@@
 - Need to compare to an approripate ground truth, but this is a philisophical exercse @Hullman2016
-
-## Bad Question stuff?
-In the discussion of the HOPs, error bar, and violin plot study @Hullman2015 note:
-
-> "In light of the common usage of error bars for presenting multiple independent distributions, it is noteworthy how poorly subjects using these representations performed on tasks asking them to assess how reliably a draw from one variable was larger than the other...Many subjects reported values less than 50%, which are not plausible given that the mean of B was larger than the mean of A."
-
-- The way we ask questions and the types of questions we ask are selected with little justification. This paper makes suggestions to reduce the noise in the data from these papers. @Hullman2016
 

### 2024-02-26 writingdumpster.qmd
some word vomit of noise and signal, also some added citations

@@
 # Main section outtakes
+Visualisation experiments are highly sensitive to the way questions are asked.
+For example, a study by @Hullman2015 that compared HOPs, error bar, and violin plots found that many participants reported that group A was more likely to be higher than group B, which was not not plausible given that the mean of B was larger than the mean of A.
+The authors attributed this obvious mistake to the visual representation, but we believe it was more likely caused by the distributions being shown in the order A then B while the question swapped the order by asking "What is P(B>A)?".
+This flip is minor, but must be done *after* the participants have worked out the probability.
+Since the primary task is done they forget to complete the secondary task, similar to forgetting your bank card in an ATM.
+I suspect this was the cause of the mistake as I, when attempting this question myself, made the exact same mistake.
+I also constalty lead my bank card in ATMs.
+
 While mathematical information *should* be considered the starting point in considering if two visualisations are equivalent, it is not the only consideration.

### 2024-02-26 paper.qmd
finished word vomit of noise vs signal section

@@
 telling someone they will or will not get cancer instead of communicating their risks about it) inherently carries with it a lack of transparency.
-Additionally risk communication is often communicated to avoid or informa about an unfavorable outcome, so framing can be important if we want to influence decisions in a certainty way.
+Additionally risk communication is often communicated to avoid or information about an unfavorable outcome, so framing can be important if we want to influence decisions in a certainty way.
 These motivations do not extend to the communication of uncertainty information more broadly.
@@
 when we want to communicate uncertainty as noise).
+
 Unfortunately understanding the relevant noise seems to be its own problem, as statisticians and visualisation authors seem to be unaware or confused in how to calculate the uncertainty related to an analysis [@Hullman2020a].
@@
 
-Trust is another concept that is related to uncertainty communication, but fact that it is often front and center in the discussion has a somewhat ominous subtext.
-If the purpose of displaying uncertainty information is to appropriately hedge a signal with noise, then it should be assumed that trust is only related to uncertainty communication through increased transparency and honesty.
-It does seem to be the case that uncertainty communication increases trust because it is a proxy for *(add citations)*.
-Unfortunately a large proportion of the literature discusses trust as something that is directly related to uncertainty visualisation rather than as an observable product.
-In viewing uncertainty communication as directly related to trust, not related through the proxy of transparency, several unobserved variables are conflated.
-The amount of uncertainty that is present in the information, whether or not the information is trustworthy or makes sense, prior beliefs of the participants, and the trustworthiness of the source are a few examples of variables that are conflated when authors directly consider trust to be of direct interest.
-Similarly to the issues in risk-aversion directly considering trust, and not transparency, as the metric of importance in uncertainty communication leads to a questionable subtext that argues against transparency.
-Being concerned about audiences perception of trust, without first establishing if what we are communicating is *trustworthy*, leads to the implication that details that result in information not being considered trustworthy should be avoided.
-
-Authors that imply that a decrease in trust or an increase in ambiguity or risk aversion are metrics of importance in of themselves do not understand the purpose of uncertainty communication as methods to increase transparency.
-Science communication should be primarily concerned with accuracy, setting trust and risk-aversion as the variables of interest implicitly encourages statisticians to set trust and risk-aversion as the primary goals of communication.
-The issue of trust being divorced from trustworthiness has been commented on by other authors [@ONeill2018], however the issue still persists in the uncertainty visualisation literature [@Zhao2023].
-Additionally, it seems that risk-aversion being divorced from the value of risk has gone unnoticed.
-
 While this is a focus on uncertainty visualisation, these issues in uncertainty communication need to be acknowledged because they bleed into the uncertainty visualisation literature.
@@
 Signal and noise can only be untangled in the presence of a motivating question.
-
-The idea that uncertainty can only be defined in the presence of a motivating question is well grounded in most areas of statistics.
-The entire process of data analysis, from deciding what should be observed as data through to communicating that data in a plot is governed by human decision and the goal of an analysis.
-At the philosophical level, applied statistics is simply taking real world entities and boiling them down into probabilistic objects, an ontological process that is largely dependent on our goals [@Otsuka2023].
-When we move onto data provenance the issue persists, as what is kept as data and what is tossed away is determined by the motivation of an analysis and what was previously noise can be shown to become signal depending on the resolution of the question [@Meng2014].
-After moving onto modelling this issue continues as each research question can be can be categorised as descriptive, predictive, or causal, each of which has its own appropriate statistical methods and motivation agnostic model selection leads to signal devoid of meaning [@Carlin2023].
-Even at the final stages of visualisation a lack of understanding of the motivating question make it difficult to untangle what is signal and what is noise, leaving many uncertainty visualisation studies with conflicting results [@Kinkeldey2014].
-These cases highlight that uncertainty is defined at *every* stage in relation to our motivating question, from data collection to visualisation.
-Discussions of uncertainty cannot be had if we are not clear *what* we are uncertain about.
-Once it is established what we are uncertainty about, we can consider the other elements of uncertainty that need to be defined.
 
@@
 @Fischhoff2014 looks at uncertainty visualisation for decision making decides that we should have different ways of communicating uncertainty based off what the user is supposed to do with it.
+Health have a wide range of metrics used to communicate different risks [@@Spiegelhalter2017] indicating they utilise new statistics based off the specific aspect of risk that is relevant to the issue.
 This should make it clear that uncertainty only exists when we are performing inference.
 Therefore it only makes sense to describe uncertainty in terms of something to be uncertain *about*, it is not some a latent feature of statistics and data.
+
+The idea that uncertainty can only be defined in the presence of a motivating question is well grounded in most areas of statistics.
+The entire process of data analysis, from deciding what should be observed as data through to communicating that data in a plot is governed by human decision and the goal of an analysis.
+At the philosophical level, applied statistics is simply taking real world entities and boiling them down into probabilistic objects, an ontological process that is largely dependent on our goals [@Otsuka2023].
+When we move onto data provenance the issue persists, as what is kept as data and what is tossed away is determined by the motivation of an analysis and what was previously noise can be shown to become signal depending on the resolution of the question [@Meng2014].
+After moving onto modelling this issue continues as each research question can be can be categorised as descriptive, predictive, or causal, each of which has its own appropriate statistical methods and motivation agnostic model selection leads to signal devoid of meaning [@Carlin2023].
+Even at the final stages of visualisation a lack of understanding of the motivating question make it difficult to untangle what is signal and what is noise, leaving many uncertainty visualisation studies with conflicting results [@Kinkeldey2014].
+These cases highlight that uncertainty is defined at *every* stage in relation to our motivating question, from data collection to visualisation.
+Discussions of uncertainty cannot be had if we are not clear *what* we are uncertain about.
+Once it is established what we are uncertainty about, we can consider the other elements of uncertainty that need to be defined.
 
@@
 Instead, we have a handful of highlights presented below.
-- @Zhao2023 displayed a model prediction with uncertainty and took participants using the model prediction as a sign of trust, failing to notice that there is no reason for uncertainty information to influence someone to *use their own prediction*.
-If the Beuro of Meterology was uncertain about their rain prediction, that would not be sufficient conditions for me to decide to make my own rain predictions.
-Despite this, the authors seemed to assume that the uncertainty information *should* do that.
 - @Hofman2020 asked questions about discusses the difference in showing inferential or outcome (prediction) intervals and then found that participants shown the inferential distribution performed poorly when asked questions about predictions.
@@
 To suggest that a histogram needs uncertainty bounds is similar to suggesting a density or box plot need uncertainty bounds.
-- @`Griethe2006 commented that "if visualization is used as a means to explore a data volume or to communicate its contents the uncertainty has to be included".
+- @Griethe2006 commented that "if visualization is used as a means to explore a data volume or to communicate its contents the uncertainty has to be included".
 @Potter2010 aimed to create a summary plot that "concisely presented data with uncertaninty information".
@@
 
+There is also the common view that uncertainty as a dimensionality problem.
+This perspective is not quite right.
+It is important to keep in mind that just because information is in a graphic, that does not mean it will be "seen".
+The phenomena of inattentional blindness shows that there is no perception without attention and it is powerful enough that participants can fail to be aware of random objects appearing on a screen or a gorilla walking through a basketball game [@simons1999gorillas; @mack2003inattentional].
+Therefore if information is mapped to graphical elements that are so low on hierarchy they can be ignored, they might as well not be there at all.
+The people who don't plot uncertainty because they think its unimportant and the people who relegate uncertainty to the lowest ranking aesthetic on the list are both saying the same thing, "I think the uncertainty is unimportant".
+Including uncertainty is worth very little if no attention is left to see it.
+This does not mean we cannot put a *large* amount of information in a graphic.
+Glyph maps can be used to depict the multi-dimensional information in spatial-temporal data by mapping line plots to map locations, but we still need to decide what information is important.
+The map can either present trends in the global or local variance depending on whether or not the line plot is scaled globally or locally [@Wickham2012], however smaller details are almost impossible to convey.
+No matter how much information we try and put in a plot, there will always be only a handful of key takeaways.
+
+This issue has come up in many different forms in the uncertainty visualisation literature.
+Typically as a "task dependence" on the performance of uncertainty visualisations.
+It is repeatedly identified as a problem in previous reviews of the uncertainty visualisation literature [@Kinkeldey2014; @Hullman2016] as well as across many sub domains and applications [@Wallsten1997; @Fischhoff2014; @Meng2021; @Amar2005].
+The fact that this conclusion is repeatedly reached shows both the importance and the lack of acknowledgement this concept receives.
+@Fischhoff2014  takes this approach by considerting how decisions with different levels of flexibility should have different uncertainty visualisations.
+If the user has no choice they should simply be made aware of the risk, if they have several options, the user needs to see the uncertainty in each, if they need to come up with the options themselves, they need to see how things work and how the system fails.
+@Kinkeldey2014 suggested that the nature of the task plays an important role and believed the task dependence for uncertainty visualisation could explain some of the inconsistency in outcomes.
+
 Despite these results, it is clear that authors intuitively *feel* that uncertainty is different even though this difference has not been apparent in existing uncertainty visualisation experiments.
@@
 These studies most commonly evaluate a visualiation on its ability to convey information for a decision or how well it induces trust.
+The motivation behind these experiments are obvious, if we cannot measure uncertainty *directy* maybe we can measure its secondary affects.
+The issue is that in order to say if some plots are *better* or *worse* than others, is to ascribe value to uncertainty information.
 
-*decision and trust paragraph*
+Decision making questions that typically involve uncertainty discuss the concept of "risk-aversion" as though it is a mistake.
+The irony of an uncertainty visualisation pper expressing the importance of uncertainty visualisation, but setting up a ground truth or optimum strategy in which participants *should* ignore the uncertainty information is not lose on us.
+This perspective does not make sense because it implies participants should make decisions entirely based on the average.
+A possible soltion for this is to provide a utility framework for a particular experiment [@Hullman2016], however it is unclear how easily participants could adopt a utility framework that is different to their own.
+@Cheong2016 introduced a very simple optimum strategy into their experiment on bushfire uncertainty visualisations, where participants should choose to evacuate a house as soon as its likelihood of catching fire is 50% or higher.
+Interestingly, despite this very obvious and simple tactic to maximize payout from the experiment, it seems like many participants did not adopt it, instead acting much more warily as though they were considering whether or not they would *actually* evacuate in the event of a fire.
+These papers also very rarely consider what the optimum decision would be in a given scenario, and a difficulty with anticipating the best strategy is conflated with the failings of the uncertainty visualisation.
+Ultimately, we agree with @Hullman2016 who suggested that what determines an approripate ground truth is largely a philosophical exercise.
 
-There are other studies that try to include uncertainty information without directly asking for it.
+Authors also occasionally use trust as a proxy for uncertainty.
+Similar to measuring uncertainty through decision quality, this method also has its own series of issues.
+If the purpose of displaying uncertainty information is to appropriately hedge a signal with noise, then it should be assumed that trust is only related to uncertainty communication through increased transparency and honesty.
+It does seem to be the case that uncertainty communication increases trust because it is a proxy for transparency *(add citations)*.
+Unfortunately a large proportion of the literature discusses trust as something that is directly related to uncertainty visualisation rather than as an observable by product.
+In viewing uncertainty communication as directly related to trust, not related through the proxy of transparency, several unobserved variables are conflated.
+The amount of uncertainty that is present in the information, whether or not the information is trustworthy or makes sense, prior beliefs of the participants, and the trustworthiness of the source are a few examples of variables that are conflated when authors directly consider trust to be of direct interest.
+Considering trust, and not transparency, as the metric of importance in uncertainty communication leads to a questionable subtext that argues against transparency.
+@Hullman2020a found that author simultaneously argued that failing to visualise uncertainty was akin to fraud, but also many avoided uncertainty visualisation because they didn't want their work to come across as "untrustworthy".Being concerned about audiences perception of trust, without first establishing if what we are communicating is *trustworthy*, leads to the implication that details that result in information not being considered trustworthy should be avoided.
+Authors that imply that a decrease in trust or an increase in ambiguity or risk aversion are metrics of importance in of themselves do not understand the purpose of uncertainty communication as methods to increase transparency.
+Science communication should be primarily concerned with accuracy, setting trust and risk-aversion as the variables of interest implicitly encourages statisticians to set trust and risk-aversion as the primary goals of communication.
+The issue of trust being divorced from trustworthiness has been commented on by other authors [@ONeill2018], however the issue still persists in the uncertainty visualisation literature [@Zhao2023].
+
+Additionally, some studies use a ground truth in these trust experiments, which makes for a very confusing read on exactly how participants are supposed to use the information. 
+- @Zhao2023 displayed a model prediction with uncertainty and took participants using the model prediction as a sign of trust.
+They reported that visualising uncertainty information caused participants to trust the model in the low variance case, but the results in the high variance case were inconclusive.
+The discussion made it clear the authors thought the uncertainty information should make the visualisation more trustworthy, but conflating trust and the use of a prediction implied uncertaitny information should somehow influence participants to use their own prediction.
+If the Beuro of Meterology was uncertain about their rain prediction, that would not be sufficient conditions for me to decide to make my own rain predictions.
+Despite this, the authors seemed to assume that the uncertainty information *should* have an influence on that, showing they had not deeply considered how uncertainty inforormation should infludence the choices of the participants.
+
+A similar measure to trust is using "confidence" in an extracted value or a decision.
+Interestingly, "confidence" is also used to try and capture the clarity of a message in a normal visualisation.
+Confidence cannot simultaneously be a measure of clarity of visualisation *and* a way to capture the uncertainty expressed in a visualisation.
+
+There are other studies that try to include uncertainty information without directly asking for it but do not go down the route of asking for a decision or levels of trust.
+These papers often try to ask a question that should utilise both the uncertainty and signal in the response.
 This is an interesting approach, but these papers are often still looking for a "signal" which is specified in the form of a "ground truth" that is used to evaluate the plot.
@@
 The participants answers were then compared to a ground truth where the correct plot had a right shifted mean.
-By comparing the results to a ground truth statistic and marking participants as "wrong" or "right", the error from the participants that had an alternative interpretation to the concept of "furthest right" was conflated with the error from a poor design.
+By comparing the results to a ground truth statistic and marking participants as "wrong" or "right", the error from the participants that had an alternative interpretation to the concept of "furthest right" was conflated with the error from a the visualisation choice.
+- @Padilla2017 made several confusing assumption when determining the ground truth of their experiments that were assessing storm visualisations.
+Experiment 2 assumed that the participants should provide responses that indicated they perceived the intensity of the storm to be independent of the uncertainty distribution of its location.
+This assumption seems counterintuitive as best, however experiment 3 asked participants to "decide which oil rig will receive more damage based on the depicted forecast of the hurricane path" for which they *were* supposed to incorperate uncertainty information and not assume they were independent.
+The authors seemed to be unaware of these confounding mistakes in their assumptions of how the participants were supposed to utilise the uncertainty information.
+- @Sanyal2009 mapped uncertainty to dots and signal to a 3D surface and asked participants to identify areas of high and low signal and high and low uncertainty.
+Participants were not asked to combine that information in any way, and the signal and the noise were treated as separate variables.
+- @Correll2014 asked participants to extract the mean and variance from four uncertainty visualisations, bar charts with error bars, box plots, gradient plots, and violin plots (that were adjusted to make sure the mathematical information they were showing was the same) from two side by side distributions.
+Participants were also asked to answer a question that was supposed to incorperate both the signal and the noise, such as "How likely is candidate B to win the election?" when the two distributions indicated voter preference.
+Participants were not able to answer the question about likelihood in term of probability, but were instead given seven options from 1=Outcome will be most in favor of A to 7=Outcome will be most in favor of B.
+The ground truth statistic for this question was a scalar multiple of Cohen’s d, indicating participants were supposed to incorperate uncertainty information using a very specific formula that was likely unknown to them but assumed to be used implicitly.
+- @Cheong2016 tested multiple different visual representations of uncertainty for representing the likelihood of a house being burned down based on its location.
 
@@
 
-Additionally, the use of a ground truth highlights how participants were supposed to *use* the uncertainty information. In this informa
+It is unclear how participants are supposed to incorperate uncertainty information into their responses when uncertainty visualisation authors themselves are unsure.
+Other authors interpret this difficulty as a combination of the uncertainty and task complexity [@Kinkeldey2014], and not as a general confusion.
+All these examples sever to illustrate that *we dont know how to measure uncertainty* and if we don't know how to measure it, *we cannot evaluate the quality of a visualisation*.
+It could be argued that a well done uncertainty visualisations should have an imperceptible signal unless the signal would be identified with a hypothesis, but this idea also has a miriad of issues that come with it.
+There is already a series of issues with the reject/do not reject concepts in hypothesis testing, so it is already questionable to set the results of a hypothesis test as the ground truth in an uncertainty visualisation experiment, however these may be additional problems that go beyond the binary nature of the issue.
+@Patrick2023 compared people ability to recognise patterns in a residual plot to typical statistical tests and found human viewers looking at a plot were less sensitive than the typical residual tests.
+These experiments utilised the lineup protocol which has been suggested as a method to check if perceived patterns are real or merely the result of chance [@Buja2009; @Wickham2010; @Chowdhury].
+This concept bears similarity to the goal of uncertainty visualisation, except uncertainty visualisation wants to achieve this within a single plot.
 
-- The way we ask questions and the types of questions we ask are selected with little justification. This paper makes suggestions to reduce the noise in the data from these papers. @Hullman2016
-
-The view of uncertainty as a dimensionality problem is not quite right.
-It is important to keep in mind that just because information is in a graphic, that does not mean it will be "seen".
-The phenomena of inattentional blindness shows that there is no perception without attention and it is powerful enough that participants can fail to be aware of random objects appearing on a screen or a gorilla walking through a basketball game [@simons1999gorillas; @mack2003inattentional].
-Therefore if information is mapped to graphical elements that are so low on hierarchy they can be ignored, they might as well not be there at all.
-The people who don't plot uncertainty because they think its unimportant and the people who relegate uncertainty to the lowest ranking aesthetic on the list are both saying the same thing, "I think the uncertainty is unimportant".
-Including uncertainty is worth very little if no attention is left to see it.
-This does not mean we cannot put a *large* amount of information in a graphic.
-Glyph maps can be used to depict the multi-dimensional information in spatial-temporal data by mapping line plots to map locations, but we still need to decide what information is important.
-The map can either present trends in the global or local variance depending on whether or not the line plot is scaled globally or locally [@Wickham2012], however smaller details are almost impossible to convey.
-No matter how much information we try and put in a plot, there will always be only a handful of key takeaways.
-
-- rarely used extension is aggregation of uncertainty over an area (retrieve overall estimation from a spatial distribution of uncertainty).
-Other tasks include comparisons, rankings.
-also have a task called "search" where participants have to identify entities that fulfill certainty characteristics (high or low values) @Kinkeldey2014
-- Whether to use classed or unclassed uncertainty is rarely discussed @Kinkeldey2014.
-e.g.
-land map with remotely sensed images has ambiguity in the boundaries, which can result from multiple sources (land cover class, measurement error, images from different dates).
-If you combine them all it might not make sense for someone who only needs one.
-
-- depicts a bivariate map which uses a bivariate colour palette that is created by blending two single hue colour palettes.
-One colour represents the variable of interest while the other represents the sampling error of that variable.
-There are two immediate problems with this method.
-First of all, uncertainty is being expressed with hue and saturation which @Maceachren2012 found to be the worst aesthetics to map to uncertainty to as they don't have an intuitive interpretation.
-Value has a natural connection to uncertainty (lighter values equate to higher uncertainty and darker values equate to more certainty) so it is a much more appropriate choice.
-While the `Vizumap` data does depict areas of light and darkness, they are largely irrelevant to the uncertainty measure causing our heuristics to lead us to the incorrect conclusions.
-The Value-Suppressing Uncertainty Palettes (VSUP) shown in @fig-vsup maps estimates to the hue and error to the value thereby creating a more intuitive plot [@Correll2018].
-Additionally, at high levels of uncertainty VSUP only has one output colour, which prevents viewers from decrypting any particular value and also avoids enforcing a binary encoding of significance [@Correll2018].
-Unfortunately VSUP are not easy to combine with packages like `Vizumap` which leaves it still somewhat difficult to express this encoding in practice, however the combination of a bivariate map with VSUP has shown to improve decisions in the face of uncertainty [@Correll2018].
-
-- @Patrick2023 investigated peoples ability to identify a signal and compared it to typical hypothesis tests
-
-- The line-up protocol is an application on this idea and considers each visualisation to be a single outcome of some larger distribution [@Buja2009; @Wickham2010; @Chowdhury; @Hofmann2012].
-By generating a sample of visualisations from a hypothetical distribution, visualisation authors can check if perceived patterns are real or merely the result of chance.
-
-The final consideration when deciding what information to depict is whether or not additional information will clutter the graph.
-Our visualisation should aim to show enough information to solve a task while avoiding irrelevant distracting information [@kosslyn2006graph].
-It also seems that people will discount information they perceive as discountable if they are able.
-Including mean estimates on depictions of mass can cause people to discount the uncertainty information and use the difference between means as a proxy for the probability distribution [@Kale2021].
-
-This discrepancy is also noted by @Kinkeldey2014 who, in their own literature review on uncertainty methods, comment that the literature seems unsure if uncertainty is simply another variable or if it should be treated differently.
-This is a fair question to ask.
-If uncertainty is simply a signal we are trying to convey, why should it be treated any different to any other signal in a visualisation?
-
-- I think the questions where uncertainty is a signal, not a noise, allow the authors of visualisation experiments to ignore the context and motivation that usually governs a visualisation.
-- Weirdly When setting their ground truth it seems like a lot of papers EXPECT participants to completely ignore uncertainty information as noise.
-Like the participants are actually considered to be incorrect if they dampen their signal understanding because of noise considerations which I find FUNNY.
-(this feels like it is more as a communicating uncertainty issues)
-- papers usually have same variance if there is a signal question, and use the mean as a proxy to answer the question.
-Really should ask a changing signal question where the signal IS impacted by the uncertainty information.
-- When asking questions, researchers should focus on events that can be repeated (because of a misunderstanding of confidence intervals) @Hullman2016 (moved here because it is clear she is touching on the whole "what are we uncertain about" thing)
-
-- new papers dont seem to be aware of this problem.
-- if the experiment DOES use noise as a signal, it is often very unclear how the participants should include that in their respose and what would be a valid response from participants (e.g.
-the likelihood to how much question and the cone of uncertainty questions).
-separate to the data, uncertainty has its own goal.
-Is it to impact confidence in the estimate (like a bayesian thing)?
-to supress signal (like a barely noticiable difference line up plot thing)?
-to facilitate more efficient decisions (so accounting for risk thing)?
-
-- will it impact their decision at all?
-there are a couple of papers where you can answer all the questions by just ignoring the uncertainty information because of the way they have set it up.
-e.g.
-difference in means when the variance is the same, experiments that check if someone can still get the signal and make sure the uncertainty information isnt in the way.
-
-- The literature also does not discuss if uncertainty is another variable, or if it needs to be treated differently (i,.e.
-is it metadata of another variable).
-Most papers studies on uncertainty dont consider this issue, and the question remains, should someone have two separate values, or an integrated view uncertain data view @Kinkeldey2014
-
-This is best seen with a simple look at some papers
-(Make into table)
-Questions where uncertainty is a signal
-
-Questions where uncertainty is a noise
-
-- Should add plot comparison mathematical framework example here
-- take a plot and work through the entire example as a flow chart thing
-- tasks e.g.
-"What is the average of X" (how well lined up you are moving down the data pipeline) and resolution (uncertainty is very often moving back a resolution level in statisitcs - also can be an assumption - bootstrap is a replacement for an assumption, not a resolution change, but bootstrapping is centered on the data)  e.g.
-"What is the noise around the estimate of the value of X"
-- "cannot assess the quality of risk communication unless the objectives are clear" circles back to motivation. He assumes we are fulfilling a duty to inform. @Spiegelhalter2017
-
-- heath had a wide range of different metrics they adopted to communicate chronic risk of adverse events in the future @Spiegelhalter2017 (use this for motivation based statistics since they define new statistics based off what they need)
-- Mentioned alleatory vs epistemic uncertainty, in the visualisation literature alleatory uncertainty is more commonly the focus.
-also mentions dependence and joint pdfs as a consideration @Hullman2016
-- Most studies involve intrinsic approaches such as colour, transparency, and those that used extrinsic techniques typically used glyphs, error bars, grid based techniques, or contouring @Kinkeldey2014
-- most studies used coincident approaches, static visualisations, are applied to a specific domain and dont have generalised results @Kinkeldey2014
-- The tasks for participants were most commonly value retrival (some where uncertainty and data value are retrived separately, others where they are some separately).
-also also be asked to extract a relative value @Kinkeldey2014
-- How does the complexity of uncertainty relate to the cateogies of user and task @Kinkeldey2014
-- Several studies probide evidence that the usability of uncertainty representations can be highly user and task dependent @Kinkeldey2014
-- The nature of the task plays an important role for the usability of uncertainty techniques, this may explain many of the insonsitent outcomes from the studies under review since two studies assessing sifferent techniques and user groups are not comparable @Kinkeldey2014
-- Doing a follow up publication to discuss issues with reasoning and decision making based on uncertainty visualisations which is not considered or discussed in this paper @Kinkeldey2014
-- Main take away is that "we need to systematize future empirical studies on uncertainty visualisation to better enable comparison and generalisation of the findings"
-- One way to advance this goal is a taxonomy, however existing taxonomies are focused on data types, uncertainty categories, and representation types @Kinkeldey2014
-- he lists a bunch of tasks that were tested with uncertainty visualisationsat the end @Kinkeldey2014
-
-- risk is uncertainty as a signal, and when uncertainty communication is spoken about this is typically what is discussed, uncertainty as signal supression is what is usually meant when we talk about visualisations but that area is functionally ignored.
-This idea, that uncertainty can only be defined in the face of a motivating question, is well grounded in the literature but seldom explicitly mentioned
-
-In section 2.1 it was mentioned that uncertainty must be defined within a specific motivating question, otherwise it inherently does not make sense.
-A large difficulty with the uncertainty visualisation evalusation studies is that this rule is not followed.
-There are a shockingly large number of evaluation studies that seem to pay no attention to the information that is relevant to the question they are asking, and show participants a selection of seemingly random visualisations from a statistical point of view.
-
-- inferential uncertainty and outcome uncertainty ARE NOT THE SAME THING??? they visualise DIFFERENT DISTRIBUTIONS
-
-We define two primary motivations for uncertainty visualisation.
-  1) To prevent deterministic conclusions from a random signal (uncertainty as noise)
-  2) To convey information about a variance, probability, or other random event (uncertainty as signal)
-
-Uncertainty as signal papers have incredibly predictable results
-Uncertainty as noise papers should follow a similar protocol to the line up papers
-
-Notes from previous sections that were moved here
-- The concept of uncertainty being task dependent is *particularly* salient for uncertainty visualisation, and it is repeatedly identified as a problem in previous reviews of the uncertainty visualisation literature [@Kinkeldey2014; @Hullman2016] as well as across many sub domains and applications [@Wallsten1997; @Munzner2009; @Fischhoff2014; @Meng2021; @Amar2005].
-The fact that this conclusion is repeatedly reached shows both the importance and the lack of acknowledgement this concept receives.
-- @Fischhoff2014 discusses how decisions with different levels of flexibility should have different uncertainty visualisations.
-If the user has no choice they should simply be made aware of the risk, if they have several options, the user needs to see the uncertainty in each, if they need to come up with the options themselves, they need to see how things work and how the system fails.
-- @Amar2005 suggested a taxonomy for information visualisation based on the types of tasks we use visualisations for and suggest 10 "analytical primitives" that we can then map to visualisations.
-- Statistics is, at its core, the study of uncertainty.
-Therefore discussing uncertainty visualisation a a separate sub domain to "normal" data visualisation is inherently confusing.
-What is typically meant by "uncertainty" visualisation is "noise", that is, we want to present some signal cushioned by its natural variance.
-Unfortunately, this distinction between "signal" and "noise" is entirely goal dependent.
-- @Bella2005 found that most participants were ignorant to the fact that error bars are used for both confidence intervals and standard error bars, two wildly different indicators of precision.
-(**also the paper sherry sent me**)
-
-- Need to compare to an approripate ground truth, but this is a philisophical exercse @Hullman2016
+The problem of trying to assess uncertainty without knowing *why* we are depiting uncertainty has been noticed by other authors.
+@Spiegelhalter2017 noted we "cannot assess the quality of risk communication unless the objectives are clear".
 
 # Great Examples
+This is not to say all experiments related to uncertainty are unusable.
+Some work that identifies heuristics are both interesting and particularly related to visualisation of uncertain information.
+@Newman2012 found that participants were more likely to view points within the bar as more likely than points outside of the bar in bar charts with error bars.
+Similar effects have been identified in other types of uncertainty displayed.
+@Padilla2017 found that points that were on an outcome of an ensemble display were perceived as more likely than points not on an outcome, even when the point that was not on a specific outcome of the ensemble was closer to the mean of the uncertainty distribution.
+
 - best things you have seen and why they are the best
@@
 
-# Future work/Conclusion
+# Future work
 This paper has identified several issues in the uncertainty visualisation literature and with those issues, we have several suggestions for future work that would likely be fruitful for the field.
@@
 Of course, this is limited by the fact that there also isnt a definition for what is a "task" and given the mess created by the lack of formalisation in uncertainty visualisation, it may be wise to formalise that concept before performing these experiemnts.
+@Amar2005 suggested a taxonomy for information visualisation based on the types of tasks we use visualisations for and suggest 10 "analytical primitives" that we can then map to visualisations, which could be a good starting point.
 Reguardless, these are directions of research would be fruitful to the uncertainty visualisation community even if it appears on the surface to be research that is only beneficial to the "normal" visualisation community.

### 2024-02-26 writingdumpster.qmd
finished word vomit of noise vs signal section

@@
 # Moved over notes
+- @Bella2005 found that most participants were ignorant to the fact that error bars are used for both confidence intervals and standard error bars, two wildly different indicators of precision.
+(**also the paper sherry sent me**)
+- Several studies probide evidence that the usability of uncertainty representations can be highly user and task dependent @Kinkeldey2014
+- The tasks for participants were most commonly value retrival (some where uncertainty and data value are retrived separately, others where they are some separately).
+also also be asked to extract a relative value @Kinkeldey2014
+- most studies used coincident approaches, static visualisations, are applied to a specific domain and dont have generalised results @Kinkeldey2014
+- Most studies involve intrinsic approaches such as colour, transparency, and those that used extrinsic techniques typically used glyphs, error bars, grid based techniques, or contouring @Kinkeldey2014
+- Mentioned alleatory vs epistemic uncertainty, in the visualisation literature alleatory uncertainty is more commonly the focus.
+also mentions dependence and joint pdfs as a consideration @Hullman2016
+- When asking questions, researchers should focus on events that can be repeated (because of a misunderstanding of confidence intervals) @Hullman2016 (moved here because it is clear she is touching on the whole "what are we uncertain about" thing)
+- we should aim to show enough information to solve a task while avoiding irrelevant distracting information [@kosslyn2006graph]. 
+- people will discount information they perceive as discountable if they are able.
+Including mean estimates on depictions of mass can cause people to discount the uncertainty information and use the difference between means as a proxy for the probability distribution [@Kale2021].
+- rarely used extension is aggregation of uncertainty over an area (retrieve overall estimation from a spatial distribution of uncertainty).
+Other tasks include comparisons, rankings.
+also have a task called "search" where participants have to identify entities that fulfill certainty characteristics (high or low values) @Kinkeldey2014
+- Whether to use classed or unclassed uncertainty is rarely discussed @Kinkeldey2014.
+e.g.
+land map with remotely sensed images has ambiguity in the boundaries, which can result from multiple sources (land cover class, measurement error, images from different dates).
+If you combine them all it might not make sense for someone who only needs one.
+- The way we ask questions and the types of questions we ask are selected with little justification. This paper makes suggestions to reduce the noise in the data from these papers. @Hullman2016
 - Error is best mapped to fuzziness, location, and colour value; arrangement, size and transparency are an OK second choice; but saturation, hue, orientation and shape are unacceptable and have no intuitive connection to variance [@Maceachren2012].
@@
 # Main section outtakes
+- tasks e.g.
+"What is the average of X" (how well lined up you are moving down the data pipeline) and resolution (uncertainty is very often moving back a resolution level in statisitcs - also can be an assumption - bootstrap is a replacement for an assumption, not a resolution change, but bootstrapping is centered on the data)  e.g.
+"What is the noise around the estimate of the value of X"
 Visualisation experiments are highly sensitive to the way questions are asked.
@@
 # Previous literature review
+
+##### Bivariate map
+Depicts a bivariate map which uses a bivariate colour palette that is created by blending two single hue colour palettes.
+One colour represents the variable of interest while the other represents the sampling error of that variable.
+There are two immediate problems with this method.
+First of all, uncertainty is being expressed with hue and saturation which @Maceachren2012 found to be the worst aesthetics to map to uncertainty to as they don't have an intuitive interpretation.
+Value has a natural connection to uncertainty (lighter values equate to higher uncertainty and darker values equate to more certainty) so it is a much more appropriate choice.
+While the `Vizumap` data does depict areas of light and darkness, they are largely irrelevant to the uncertainty measure causing our heuristics to lead us to the incorrect conclusions.
+The Value-Suppressing Uncertainty Palettes (VSUP) shown in @fig-vsup maps estimates to the hue and error to the value thereby creating a more intuitive plot [@Correll2018].
+Additionally, at high levels of uncertainty VSUP only has one output colour, which prevents viewers from decrypting any particular value and also avoids enforcing a binary encoding of significance [@Correll2018].
+Unfortunately VSUP are not easy to combine with packages like `Vizumap` which leaves it still somewhat difficult to express this encoding in practice, however the combination of a bivariate map with VSUP has shown to improve decisions in the face of uncertainty [@Correll2018].
+
 #####Taxonomy Benefits

### 2024-02-27 paper.qmd
some write up of great examples section

@@
 # Great Examples
-This is not to say all experiments related to uncertainty are unusable.
-Some work that identifies heuristics are both interesting and particularly related to visualisation of uncertain information.
+Of course we do not believe every uncertainty visualisation is unusable and every experiment is build upon misguided assumptions.
+Despite the common problems detialled in the previous sections, there is some interesting work in the uncertainty visualisation space.
+This can come in the form of an uncertainty visualisation that identifies or fixes pitfalls in the previous sections (without identifying them) or does an experiment that identifies useful information that avoids these pitfalls.
+
+A lot of work that identifies heuristics or biases in plots do not hinge on the previous assumptions.
+This work also provides useful insights for experiments by highlights pitfalls participants might fall into when they review the results of evaluation experiments [@Hullman2016].
 @Newman2012 found that participants were more likely to view points within the bar as more likely than points outside of the bar in bar charts with error bars.
@@
 @Padilla2017 found that points that were on an outcome of an ensemble display were perceived as more likely than points not on an outcome, even when the point that was not on a specific outcome of the ensemble was closer to the mean of the uncertainty distribution.
+The sine illusion can cause the confidence interval of a smoothed sine curve to seem wider at the peaks than the troughs, causing us to underestimate uncertainty associated with changing values [@Vanderplas2015]
 
-- best things you have seen and why they are the best
-- can't have a best practices because you need experimental results and my point is that the experiments aren't testing the right things
+One example of a great uncertainty visualisation method that was designed with signal supression in mind is the value supressing uncertainty pallet (VSUP).
+Depicts a bivariate map which uses a bivariate colour palette that is created by blending two single hue colour palettes.
+One colour represents the variable of interest while the other represents the sampling error of that variable.
+There are two immediate problems with this method.
+First of all, uncertainty is being expressed with hue and saturation which @Maceachren2012 found to be the worst aesthetics to map to uncertainty to as they don't have an intuitive interpretation.
+Value has a natural connection to uncertainty (lighter values equate to higher uncertainty and darker values equate to more certainty) so it is a much more appropriate choice.
+While the `Vizumap` data does depict areas of light and darkness, they are largely irrelevant to the uncertainty measure causing our heuristics to lead us to the incorrect conclusions.
+The Value-Suppressing Uncertainty Palettes (VSUP) shown in @fig-vsup maps estimates to the hue and error to the value thereby creating a more intuitive plot [@Correll2018].
+Additionally, at high levels of uncertainty VSUP only has one output colour, which prevents viewers from decrypting any particular value and also avoids enforcing a binary encoding of significance [@Correll2018].
+Unfortunately VSUP are not easy to combine with packages like `Vizumap` which leaves it still somewhat difficult to express this encoding in practice, however the combination of a bivariate map with VSUP has shown to improve decisions in the face of uncertainty [@Correll2018].
+
 - pixel map

### 2024-02-27 writingdumpster.qmd
some write up of great examples section

--- 2024-02-26 writingdumpster.qmd
@@
 - This idea also extends to interval estimation, where questions about probability are best answered with gradients and questions about start and end times are easiest to answer with ambiguation [@Gschwandtnei2016].
-- Heuristics can work against us just as much as they can work for us.
-The sine illusion can cause the confidence interval of a smoothed sine curve to seem wider at the peaks than the troughs, causing us to underestimate uncertainty associated with changing values [@Vanderplas2015]
+
 - mentions uncertainty can be variance, precision, accuracy, reliability, or related concepts @Hullman2016
@@
 # Previous literature review
+
 #####Taxonomy Benefits

### 2024-02-27 paper.qmd
formatting edits during meeting

@@
 
-# 3. Defining Uncertainty
+# Defining Uncertainty
 Uncertainty visualisation is not made any easier by the fact that the term "uncertainty" lacks a commonly accepted definition in the literature.
@@
 
-## 3.1 Coloquial Definitions
+## Coloquial Definitions
 **Aside: I am not sure how to collect these together and show that they are highly varied.
@@
 
-## 3.2 Taxonomy Definitions
-### 3.2.1 Taxonomies of Uncertainty
+## Taxonomy Definitions
+### Taxonomies of Uncertainty
 **Aside: I actually think I agree with the sentiment in the grammar of graphics that a taxonomy is more of an opinion and we can reclassify it and is therefore not thorough enough to remain unchanged and be build upon.
@@
 
-### 3.2.2 Taxonomies of Visual Uncertainty
+### Taxonomies of Visual Uncertainty
 There are some taxonomies of visual uncertainty.
@@
 
-## 3.3 Contextual Information
+## Contextual Information
 Some uncertainty work, specifically uncertainty visualisation, focus on the context surrounding the uncertainty.
@@
 
-# 4.General Attitude Towards Uncertainty (Uncertainty Communication Boradly)
+# General Attitude Towards Uncertainty (Uncertainty Communication Boradly)
 It would be impossible to discuss uncertainty visualisation without mentioning the broader conversation around uncertainty communication.
@@
 
-# 5. Uncertainty Visualisation Literature Summary
+# Uncertainty Visualisation Literature Summary
 The previous sections of this paper established several issues in uncertainty communication at large.

### 2024-02-27 paper.qmd
word vomit of great ideas and updated edit notes

@@
   chunk_output_type: console
+abstract: In reviewing the literature of uncertainty visualisation
 ---
-# Background
-## Introduction
 
-**Aside: I have not touched this section since the confirmation so it would be best to ignore it**
+# Introduction
 
-From entertainment choices to news articles to insurance plans, the modern citizen is so over run with information in every aspect of their life it can be overwhelming.
-So overwhelming in fact, that it is not uncommon for people to become so overwhelmed by the information they refuse to engage with it at all.
-In this overflow of information, tools that can effectively summarise information down into simple and clear ideas become more valuable.
-While websites that summarise insurance plans or hotel options down into a clear table make a living off it, few tools are as powerful or effective as a visualisation.
-Information visualisations remain one of the most powerful tools for fast and reliable science communication.
-Effectively leveraging visualisations to make them more effective is therefore key in facilitating simple and clear communication of science and other ideas.
+> Edit note: this section is untouched since the confirmation. Might need a rewrite
 
@@
 This powerful aspect of data visualisation is poorly or seldom used in later stages when we are communicating our findings, specifically with respect to uncertainty.
-
-- @Suh2023 found that the main issue with deployment of AI was communication between data science and subject matter experts, since subject matter experts didn't want technical presentations that didn't explain exactly when and how a model would fail.
 
@@
 
-- industrial accidents have a "tollerable" risk chart based on risk and number of fatalities @Spiegelhalter2017
+# Similar work
+> Edit note: I think with the addition of more summary papers I should just summarise common themes of issues the papers ran into instead of detailling the work paper by paper.
 
-## Similar work
 There are a wealth of uncertainty communication and visualisation literature reviews that already exist. Interestingly they all seem to have identified similar pitfalls or fallen into one themselves.
@@
 ## Coloquial Definitions
-**Aside: I am not sure how to collect these together and show that they are highly varied.
-This might be a situation for the tidytext clustering or something.
-There is an additional problem is that the definition for uncertainty is completely unrelated to the uncertainty that is quantified in the graphics.
-I think the tidytex can help with the definitions being all over the place, but I am not sure how to identify the lack of correlation between the colloquial definitions of uncertainty and the mathematical quantificaitons.**
-
 The most common definition of uncertainty used in papers that discuss visualising it are colloquial.
@@
 ### Taxonomies of Uncertainty
-**Aside: I actually think I agree with the sentiment in the grammar of graphics that a taxonomy is more of an opinion and we can reclassify it and is therefore not thorough enough to remain unchanged and be build upon.
-That being said, if this is a literature review and all existing literature use written definitions or taxnomies, that is what I have to work with.
-The larger taxonomy can be used to show that the other taxonomies fit within it, but it is in of itself not tied to anything except work that paper itself identifies as transient.**
+> Edit Note: A lot of this is going to be moved to the uncertainty definition specific section of the vis lit review and the point here should just be "this is a mess" instead of trying to find any reason in it.
+My opinion and goal for this section has changed a lot since writing this (specifically deciding that the encapsulating taxonomy doesnt fix the problem and a maths definition is needed) so it is one of the sections that needs big rewrites.
 
@@
 
-**Aside: This figure is not going to be in the final version of the paper.
-It is here as a placeholder.
-Ideally, I would replace with with a graphic that shows where all these taxonomies land in the location, level, nature taxonomy (since it is encompassing) and show they are very inconsistent and a new definition is pulled out for almost every paper (even in statistics papers)**
+> Edit note: I dont think I will keep this figure in the paper. The goal was to show they are inconsistent but I think a sentance or two in the text will achieve that.
 
@@
 
-*I need to fix the taxonomy illustration because inputs and model uncertainty need to be swapped*
-
-![Depicts an illustration of the taxonomy described in @utypo.
+![*I need to fix the taxonomy illustration because inputs and model uncertainty need to be swapped*.
+Depicts an illustration of the taxonomy described in @utypo.
 From right to left the drawing shows the location, level and nature of uncertainty with examples of that category underneath.
@@
 
-**Aside: I had this here to explain why these are the considerations people make in statistics and how they relate to elements of the taxonomy to try and provide some justification, but If I am just going to show the definitions are inconsitent, I might get rid of this section.
-I also largely included it so you guys can see what the taxonomies are considering**
+> Edit note: Everything from this point down was used to try and provide some justification for why this dimensionality concept could work, but I have since changed my mind and am probably going to get rid of this section unless it fits somewhere else.
 
@@
 Mapping uncertainty into the 3D space defined by location, level, and nature can expand beyond the narrow case of statistical modelling it was defined for.
-*@___* included "consensus" as a taxonomy element, however it was not included in the final definition, but it can easily be added as a location that would align with the communication stage of the data pipeline.
-Several authors [*@_+_@Spiegelhalter2017*] considered the method by which uncertainty was quantified (e.g.
+*(find the citation for this)* included "consensus" as a taxonomy element, however it was not included in the final definition, but it can easily be added as a location that would align with the communication stage of the data pipeline.
+Several authors [@Spiegelhalter2017] *(I have other papers)* considered the method by which uncertainty was quantified (e.g.
 by a single value, a function, simulated outcomes, etc), which can be be thought of as a more precise categorisation of the level axis.
@@
 ### Taxonomies of Visual Uncertainty
+> Edit Notes: Similar to the "similar works" section, I think this section would be better off with a short sentance about each paper and comments on their similarities because a paragraph for each paper is too much
+
 There are some taxonomies of visual uncertainty.
@@
 
-**Aside: I should add something in about the actual quantified uncertainty vs the discussed uncertanty thing (i.e. what is defined with words and what is defined with maths)**
+## Contextual Information
+> Edit Note: This section should probably be moved to the visualisation specific section, although that depends on the tone that I end up with after the rewrites of the earlier sections.
 
-## Contextual Information
 Some uncertainty work, specifically uncertainty visualisation, focus on the context surrounding the uncertainty.
@@
 # General Attitude Towards Uncertainty (Uncertainty Communication Boradly)
+
+> Edit note: this section needs a huge rewrite.
+The idea of this section blends way too much with the measurement discussions in the "noise and signal" section.
+I have already taken paragraphs from here and just pasted it in the noise section so it might not be coherrent.
+Additionaly this section goes on SEVERAL long rambles that realistically should have been a sentance or not included at all.
+
 It would be impossible to discuss uncertainty visualisation without mentioning the broader conversation around uncertainty communication.
@@
 
-### Extra notes for this section
-- People believe in the law of small numbers but dont believe in the lar of large numbers @Hullman2016
-- experiencing an event makes your percieved probability higher than if you see a description of it @Hullman2016
-- Frequency thing means that more effective communication to participants in a study means more effective science and we have a roll on effect @Hullman2016
-- Probability judgements are less relative than other psychophysical judgements, 0%, 50% and 100% all have well understood meanings.
-Subjective probability distributions are spontaneous, and subjects are constructing a distribution upon being asked rather than articulating a fully formed mental representation of their beliefs.
-Therefore it is subject to heuristics.
-consider anchoring a scale with farmiliar probabilities .
-not sure if being asked for the entire distribution helps, e.g.
-the graphical one reduces noise but asking people for an interval estimation leads to overprecision and overconfidence @Hullman2016
-- believed that offering enough decisions with varying rewards you can observe the true subjective probability, however whether or not probability is the right representation of subjective uncertainty is a topic of debate @Hullman2016
-- she also discusses graphical interface (users have to draw the probability distribution) and framing probabilities as natural frequencies are better choices because they reduce noise and stuff.@Hullman2016
-- Consider incorperating utility functions. Include "probability-coherence" checks @Hullman2016
-- evidence that people do not follow the axioms of probability when conveying subjective probabilities. How do you deal with this as an analyst (e.g. if the probabilities do not sum to 1 @Hullman2016
-
-@Spiegelhalter2017
-we consider a summary table as a visualisation
-illuminate graphics with words and numbers, 
-design graphics to allow a part-to-whole comparison on an appropriate scale, 
-use images and metaphors to gain/retain attention but dont arouse undue emotion, 
-assume low numeracy of a general public audience and adopt a less-is-more approach by reducing the need for inference and making clear and explicit comparison, 
-be cautious about interactivity and animations (may introduce unnecessary complexity), 
-avoid chart junk (like 3d bar charts),
-
 # Uncertainty Visualisation Literature Summary
@@
 ## Types of Uncertainty Visualisation Papers
+
+> Edit notes: Originally I was planning to have the discussion following this section relate to these "types" of papers, but that didn't end up happening.
+Depending on what I do with later sections, this might need a rewrite or to just be removed, since I'm not sure how relevant this is to the main ideas of the paper.
+
 Uncertainty visualisation papers can be organised according to the *goal* of the experiment.
@@
 Additionally, since there is often a miriad of ways to visualise any particular mathematical object, to adopt specific visualisation that is not already common practice needs reasonable justification, lest we run into a range of unforseen heuristic pitfalls.
-
-**Aside: I should maybe change this to relatability to uncertainty, value extraction, decision making and trust, and meta papers or something. Maybe perception and evaluation papers?**
 
@@
 
-## Uncertainty and Inference
+## Defining uncertainty with resepct to inference
 When we consider mapping uncertainty, a reasonable questions that should come to mind is "uncertainty about what?".
@@
 
-**Aside: Include table that shows broad uncertainty definition and then PDF visualisation**
+> Edit notes: Need some way to show uncertainty visualisation papers provide sweeping definitions and then visualise a simple PDF everytime.
 
@@
 
-## Evaluation Experiments and The Grammar of Graphics
+## Uncertainty evaluation experiments repeat visualisation studies
 The first issue in uncertainty visualisation is there does not seem to be a clear rule for what makes a particular visualisation a depiction of uncertainty.
@@
 
-**Aside: I think I want the "in order to visualise uncertainty you need to quantify it and the quantification does not align with the definition" point here since this section is now about value extraction.**
+> Edit notes: The "in order to visualise uncertainty you need to quantify it and the quantification does not align with the definition" point could also work here since this section is now about value extraction.
 
@@
 This pipeline shows every step from data observation to rendering of a graphic, but we are going to focus on the tail end of the pipeline specifically on statistics, geometry, coordinates, and aesthetics.
+
+> Edit notes: Grammar of graphics is just a placeholder, Di suggested to replace this with my own pipeline visualisation based on some data wrangling papers.
 
@@
 The hierarchy established 40 years ago by @Cleveland1984 is:
-1) position along a common scale
-2) position along a non-aligned scale
-3) length
-4) angle/slope
-5) area
-6) volume
+1) position along a common scale. 
+2) position along a non-aligned scale. 
+3) length. 
+4) angle/slope. 
+5) area. 
+6) volume. 
 7) colour
+
+> Edit notes: need to add the notes from the susan lit review that has the more recent perceptual task list
+
 This hierarchy has had additions and adjustments over the years.
@@
 
-**Aside: INCLUDE A GRAPHIC THAT SHOWS WHERE EACH TYPE OF UNCERTAINTY VISUALISATION SITS IN THE GRAMMAR OF GRAPHICS AND HOW DIFFERENT THEY ARE ACROSS MULTIPLE STUDIES**
+> Edit note: need to include a graphic that illustrates the wide spread of mathematical information and visual techniques that are visualised in these experiments.
 
@@
 There is also the common view that uncertainty as a dimensionality problem.
-This perspective is not quite right.
+This perspective is not quite right and the literature seems to be about 50/50 on whether uncertainty is a psychological or dimensionality problem.
+@Kinkeldey2014 also identified this issue when they highlighted that the literature makes it unclear if uncertainty is a variable in of itself, something that should be interpreted with the main variable, or if it is metadata.
 It is important to keep in mind that just because information is in a graphic, that does not mean it will be "seen".
@@
 
-## Noise and Signal in Uncertainty Visualisation
+## Measuring uncertainty as noise, not signal
 Uncertainty is noise.
@@
 # Great Examples
+
+> Edit notes: this section seems to go off a bit on dimensionality of visual tasks and stuff that might be better put in the noise and signal section about measuring uncertainty.
+
 Of course we do not believe every uncertainty visualisation is unusable and every experiment is build upon misguided assumptions.
@@
 This can come in the form of an uncertainty visualisation that identifies or fixes pitfalls in the previous sections (without identifying them) or does an experiment that identifies useful information that avoids these pitfalls.
+
+Uncertainty that is created by steps that are further down the pipeline are often ignored by the uncertainty visualisation literature, however there are visualisations that make these considerations.
+If uncertainty had a more encapsulating mathematical definition, the uncertainty created at earlier stages might be able to be visualised with the model uncertainty that is typically expressed through confidence intervals.
+That being said, even with a definition that allows us to combine these uncertainties into a single value, it still might be desirable to see which stage of the analysis is creating uncertainty in the visualisaiton.
+One example of a visualisation that includes that information is visualisations of climate scenario uncertainty, as shown in @fig-climatescenario.
+Uncertainty is introduced into this model by the choices of individuals which is difficult to model or predict and is also unlikely to be constant over time.
+Therefore specific scenarios are modelled and shown, each with their own model uncertainty.
+This highlights a method that can be used to visualise uncertainty from earlier in the pipeline that is often forgotten or ignored via assumptions.
+
+![I will just make my own version of a climate scenario plot, this one is just here from a random site as a place holder](climatescenario.png){fig-climatescenario}
 
@@
 
-One example of a great uncertainty visualisation method that was designed with signal supression in mind is the value supressing uncertainty pallet (VSUP).
-Depicts a bivariate map which uses a bivariate colour palette that is created by blending two single hue colour palettes.
-One colour represents the variable of interest while the other represents the sampling error of that variable.
-There are two immediate problems with this method.
-First of all, uncertainty is being expressed with hue and saturation which @Maceachren2012 found to be the worst aesthetics to map to uncertainty to as they don't have an intuitive interpretation.
-Value has a natural connection to uncertainty (lighter values equate to higher uncertainty and darker values equate to more certainty) so it is a much more appropriate choice.
-While the `Vizumap` data does depict areas of light and darkness, they are largely irrelevant to the uncertainty measure causing our heuristics to lead us to the incorrect conclusions.
-The Value-Suppressing Uncertainty Palettes (VSUP) shown in @fig-vsup maps estimates to the hue and error to the value thereby creating a more intuitive plot [@Correll2018].
-Additionally, at high levels of uncertainty VSUP only has one output colour, which prevents viewers from decrypting any particular value and also avoids enforcing a binary encoding of significance [@Correll2018].
-Unfortunately VSUP are not easy to combine with packages like `Vizumap` which leaves it still somewhat difficult to express this encoding in practice, however the combination of a bivariate map with VSUP has shown to improve decisions in the face of uncertainty [@Correll2018].
+There are a number of visualisation methods that do take the approach of uncertainty as signal suppression.
+These visualisation examples try to establish uncertainty as something that should be interpreted *with* the signal and take several approaches to do so.
+@fig-maps depicts the array of approaches that have been used to visualise uncertainty in a map.
 
-- pixel map
-  - Similar to what could be good but its computation isnt quite what I would like
-  - heaps of sample depictions are animated and that is not always possible, but this is a clear indicator of how you can do a sample visualisaiton
-  - hasnt been tested so it is unclear if this actually works at signal supression
-- VSUP
-  - The purpose of uncertainty is clear in the design and it is signal supression
-  - Supression is not the only uncertainty feature we should be interested in (shape is also important which is why I have beef with over smoothing and something visualisation is very good to express) but by being clear in its goal it knows what it is
-  - It addresses several key issues with the bivariate map that arrise in general and especially when it is used for visualising uncertainty 
-- the census dot map (has been taken down but CNN replicated it [here](https://edition.cnn.com/interactive/2021/us/census-race-ethnicity-map/))
-  - I love a good depiction of a sample
-  - You can see the sparse regions and the pop 
-  - it is similar to the pixel map (could also be singal supression but it is not tested)
-  - the sample depiction works well with the interactivity and you can see the signal that is appropriate for a particular level of zoomed in.
-If you zoom out the points turn into solid colours and show clear signal.
-- The new york times class mobility and race animated plot [here](https://www.nytimes.com/interactive/2018/03/19/upshot/race-class-white-and-black-men.html) 
-- the animated surface classifications from that satelite images experimental paper
-- the climate change scenario uncertainty. There are quite a few of these:
-  - [e.g.1](https://earthobservatory.nasa.gov/features/GlobalWarming/page5.php)
-  - [e.g.2](https://climatechange.chicago.gov/sites/production/files/2016-07/scenariotempgraph_0.jpg)
-  - there are a few examples on the [climate change scenario wiki](https://en.wikipedia.org/wiki/Climate_change_scenario)
-  - Good example of uncertainty that has uncertainty from two different sources (human decision in the form of inputs and statistical uncertainty), one of which cant be quantified.
-  - a good example in how you can combine two different  types of uncertainty
-  - [This](https://www.esri.com/arcgis-blog/products/arcgis-pro/mapping/graphical-cartograms-in-arcgis-pro/) has some good maps and some bonkers ones 
-- Exceedance probability map (example where uncertainty that *should* be conveyed as signal is expressed as a signal. It is a typical maths problem that involves)
-Notes
-- Since a lot of the good visualisations are made by random people creating a new visualisation (likely through back and forth communication which is HOW you get good visualisations) 
-- A lot of these show the data instead of an inference which is probably better for this scenario because sample size is important for uncertainty.
-It is not always appropriate and sometimes dots are not exactly equal to one draw.
-  - comments about smoothing in density plots and bin width in histograms.
-I do think the sample size should be communicated to people.
-While it is not the only thing that matters it seems kind of pointless to go to such lengths to hide it.
-- Sometimes points are used for things that are NOT samples which can get confusing and honestly I dont like it.
+The first approach to embedding noise in the signal is to adjust the signal visualisation so that it includes the noise.
+This method has been particularly popular with map visualisations.
+To visualise uncertainty such that it does its job of signal supression the uncertainty and the signal need to be interpreted in conjunction with each other.
+There are several suggested ways to go about this, most of which assume the "no uncertianty" version of the visualisation is a simple choropleth map.
+There are two methods to include uncertainty in a map.
+The first is to include uncertainty as error and map our signal and noise to a two dimensional *colour* space, which is the method adopted by the bivariate map and the value supressing uncertainty pallet [@Correll2018].
+These methods rely on the assumption that colour hue and value exist mentally as a two dimensional colour space and that distance in this colour space is linear (or at least appears linear using our scale).
+However, for uncertainty to be interpreted in the visualisation *with* the signal, our brain would need to be able to flatted this two dimensional space into a single space of "signal validity", and there is simply not evidence that we can do this.
+This visualisation method brings up an interesting concept though.
+If there are two aesthetics we can use that are flattened into a single channel in our mind, then plotting measurement and error to each of those aesthetics is a viable method to visualise uncertainty.
+The idea of two perceptual tasks flattening into one variable in the mind of the viewer may be wishful thinking, but it is not impossible given we are not certain on how the perceptual tasks are mapped within the human brain.
+@Sterzik2023 found that when a value was mapped to the textures of stippling, hatching, and triangles, and found that the difference between two points on this one dimensional texture was actually a 2D space (likely "business" and light/darkness).
+This idea of two aesthetics flattening into one cohesive variable in the perceptual space is particularly interesting for uncertainty visualisation.
+
+An alternative method to visualising uncertainty in a map is to combine uncertainty and signal at the earlier stage so the "supressed signal" is represented by a single variable.
+This statistic can then be expressed in a one dimensional colour space, which is a method addopted by the Baysian surprise metric map [@Ndlovu2023] and the excedance probability map [Lucchesi2017].
+These methods do not require us to compress colour hue and value into a single variable to interpret signal and noise as a single variable, it is done explicitly for us before the visualisaiton.
+These methods require more specificity in *how* the uncertainty contains important information.
+
+::: {#fig-maps}
+
+![](mappals/img1.png)
+
+![](mappals/img2.png)
+
+![](mappals/img4.png)
+![](mappals/img3.png)
+
+![](mappals/img5.png)
+
+*I am going to replace this with my own R coded maps so they are all on the same data and signal/noise.* Four different ways to include error information in a map.
+A typical chloropleth map is considered the map that ignores uncertainty and simply visualises the signal.
+The bivariate map visualises the uncertainty, but does so using two different visual channels, which gives the perception of two different signals (or one signal in the assumptions the colors contrast with each other).
+An adjustment to the bivariate map is the VSUP which is similar to the bivariate plot except it merges high uncertainty values in the palette to reduce our ability to discern values that have a high enough error to not be a valid signal.
+The Bayesian surprise metric is similar to the lineup protocol, where the colour value is the difference between the observers posterior and prior beliefs.
+The exceedance probability map is a direct visualisation of the probability of a specific hypothesis.
+:::
+
+These maps make the importance of combining uncertainty and signal in a single visual channel clear.
+A chloropleth map will show signal that is not valid inference because of high uncertainty.
+At the other end of the spectrum, the bivariate map will show signal that is not always interesting because it forces us to interpret uncertainty and signal differently.
+The value supressing uncertainty pallet was clearly designed with hypothesis testing in mind, because it combines highly uncertain values.
+The main difference between the VSUP and the bivariate colour map is actually the subtle combination of signal and noise through discernability.
+In this sense, the uncertainty visualisation maps that combine uncertainty and signal before it is conveyed visually simplify the visualisation and make it easier to interpret value supressed signal.
+This is what is done by the exceedence probability map and the Bayesian surprise metric map.
+They are more explicit in how the uncertainty should supress the signal since they make the specific hypothesis explicit.
+
+Another method that has been used is to visualise samples from the distribution we are trying to draw inference on instead of estimates along with a variance or error.
+This has been used in maps with the pixel map [Lucchesi2017], but is more commonly used in animation with the HOPs plots [@Hullman2015] or similar concepts [@Blenkinsop2000] .
+The difficulty with resampling methods is that they are typically conflated with animation even though this does not necessarily need to be true, as can be seen in the @fig-pixelmap.
+Even though @Kale2018 found that users were more sensitive to the underlying trend when shown the HOPs plot over the static outcoms plot, this effect went away when the speed of the animation increased, indicating that performance difference might be due to overplotting in the static case and a smaller sample size in the static condition would produce the same result.
+Therefore, so long as the sample size of the number of outcomes is kept to a level where individual samples are still visible, there is no evidence that an outcome plot *needs* to be animated.
+Visualisations that opt to express a signal as a sample rather than an estimate have the potential to supress signal since it is not explicitly visualised, however this has yet to be shown in evaluation experiments.
+This is not to say that visualisations of mass would not be able to perform signal supression, but a sample can easily be expressed using aesthetics such as colour on a map and mass visualisation often struggling with issues such as over or under smoothing.
+
+![](pixelmap.png){fig-pixelmap}
+
+A final method to perform signal supression is simply to visualise the data if it is available.
+An example of this is shown in @fig-census for racial distributions spatially in america, an interactive version of the plot can be found [here](https://edition.cnn.com/interactive/2021/us/census-race-ethnicity-map/).
+This map shows the typical causes of uncertainty in a spatial model, i.e.
+regions where data is sparse, ethnically diverse areas, uneven distribution of points within boundaries, etc, but it avoids the need to create a visualisation with a specific signal in mind.
+This is the technique typically employed by exploratory data analysis, which means it's lack of a specified signal means there is both *no* uncertainty (since we are technically not performing inference) but in the event we *do* implicitly perform inference, there is some hedging.
+In this sense, the best uncertainty visualisation you can get without specifying a signal you want to convey is visualising the data itself.
+The census dot map's addition of interactivity also allows users to zoom in and see the details that caused "uncertainty" in the form of inconsistent colours at lower resolutions when they were zoomed out.
+The raw data can also be used in a similar method to the HOPs plot, where a statistic of importance is implied by the visualisation, but the signal is supressed using the data itself instead of resampling methods.
+This method is used by the New York Times with their class mobility plot shown in fig, an animated version of which can be found [here](https://www.nytimes.com/interactive/2018/03/19/upshot/race-class-white-and-black-men.html).
+This does not mean that visualising raw data instead of implementing sampling techniques is always a valid uncertainty visualisation that will prevent insignificant signal from getting through.
+@Buja2009 illustrated how groups that appear linearly separable in a linear discriminant analysis (LDA) visualisation of the data can actually be the result of a LDA performed on too many variables, something that was not clear from the visualisation until the lineup protocol was implemented.
+
+![](censusdotmap.png){fig-census}
+![](classmobility.png){fig-census}
+
+These great examples show there are visualisations that are being made that have a good conceptualisation of uncertainty and express it as signal supression, however this work is not being done formally within visualisation.
+These industry visualisation are likely made with a large amount of back and forth between different levels of an organisation, something that is important for good visual communication [@Spiegelhalter2017].
 
@@
 ```{r}
-# visualisation of a dot plot with increasing sample size
+#| include: false
+
+# put a visualisation of a dot plot with increasing sample size here to show my point
 ```

### 2024-02-27 writingdumpster.qmd
word vomit of great ideas and updated edit notes

--- 2024-02-26 writingdumpster.qmd
@@
 # Moved over notes
+- make sure quantified vs discussed uncertainty is mentioned
+- industrial accidents have a "tollerable" risk chart based on risk and number of fatalities @Spiegelhalter2017
+- @Suh2023 found that the main issue with deployment of AI was communication between data science and subject matter experts, since subject matter experts didn't want technical presentations that didn't explain exactly when and how a model would fail.
+- People believe in the law of small numbers but dont believe in the lar of large numbers @Hullman2016
+- experiencing an event makes your percieved probability higher than if you see a description of it @Hullman2016
+- Frequency thing means that more effective communication to participants in a study means more effective science and we have a roll on effect @Hullman2016
+- Probability judgements are less relative than other psychophysical judgements, 0%, 50% and 100% all have well understood meanings.
+Subjective probability distributions are spontaneous, and subjects are constructing a distribution upon being asked rather than articulating a fully formed mental representation of their beliefs.
+Therefore it is subject to heuristics.
+consider anchoring a scale with farmiliar probabilities .
+not sure if being asked for the entire distribution helps, e.g.
+the graphical one reduces noise but asking people for an interval estimation leads to overprecision and overconfidence @Hullman2016
+- believed that offering enough decisions with varying rewards you can observe the true subjective probability, however whether or not probability is the right representation of subjective uncertainty is a topic of debate @Hullman2016
+- she also discusses graphical interface (users have to draw the probability distribution) and framing probabilities as natural frequencies are better choices because they reduce noise and stuff.@Hullman2016
+- Consider incorperating utility functions. Include "probability-coherence" checks @Hullman2016
+- evidence that people do not follow the axioms of probability when conveying subjective probabilities. How do you deal with this as an analyst (e.g. if the probabilities do not sum to 1 @Hullman2016
+- we consider a summary table as a visualisation, illuminate graphics with words and numbers, design graphics to allow a part-to-whole comparison on an appropriate scale, use images and metaphors to gain/retain attention but dont arouse undue emotion, assume low numeracy of a general public audience and adopt a less-is-more approach by reducing the need for inference and making clear and explicit comparison, be cautious about interactivity and animations (may introduce unnecessary complexity), avoid chart junk (like 3d bar charts) @Spiegelhalter2017
+- @Maceachren2012 found to be the worst aesthetics to map to uncertainty to as they don't have an intuitive interpretation.
 - @Bella2005 found that most participants were ignorant to the fact that error bars are used for both confidence intervals and standard error bars, two wildly different indicators of precision.
@@
 # Previous literature review
-
-##### Bivariate map
-Depicts a bivariate map which uses a bivariate colour palette that is created by blending two single hue colour palettes.
-One colour represents the variable of interest while the other represents the sampling error of that variable.
-There are two immediate problems with this method.
-First of all, uncertainty is being expressed with hue and saturation which @Maceachren2012 found to be the worst aesthetics to map to uncertainty to as they don't have an intuitive interpretation.
-Value has a natural connection to uncertainty (lighter values equate to higher uncertainty and darker values equate to more certainty) so it is a much more appropriate choice.
-While the `Vizumap` data does depict areas of light and darkness, they are largely irrelevant to the uncertainty measure causing our heuristics to lead us to the incorrect conclusions.
-The Value-Suppressing Uncertainty Palettes (VSUP) shown in @fig-vsup maps estimates to the hue and error to the value thereby creating a more intuitive plot [@Correll2018].
-Additionally, at high levels of uncertainty VSUP only has one output colour, which prevents viewers from decrypting any particular value and also avoids enforcing a binary encoding of significance [@Correll2018].
-Unfortunately VSUP are not easy to combine with packages like `Vizumap` which leaves it still somewhat difficult to express this encoding in practice, however the combination of a bivariate map with VSUP has shown to improve decisions in the face of uncertainty [@Correll2018].
+##### Introduction addition
+From entertainment choices to news articles to insurance plans, the modern citizen is so over run with information in every aspect of their life it can be overwhelming.
+So overwhelming in fact, that it is not uncommon for people to become so overwhelmed by the information they refuse to engage with it at all.
+In this overflow of information, tools that can effectively summarise information down into simple and clear ideas become more valuable.
+While websites that summarise insurance plans or hotel options down into a clear table make a living off it, few tools are as powerful or effective as a visualisation.
+Information visualisations remain one of the most powerful tools for fast and reliable science communication.
+Effectively leveraging visualisations to make them more effective is therefore key in facilitating simple and clear communication of science and other ideas.
 

### 2024-02-27 paper.qmd
formatting changes to paper

@@
 
-**Aside:this is all currently from the spieglehatter paper because that is the only paper I have moved notes over from mendeley before I started doing the summary write up.
-this will not all be from one paper, dont worry.**
-
-**Aside:This section still needs some citations from papers that communicate uncertainty, but the are *hard* to organise because of the ambiguous uncertainty definition problem.
-So they will all be communicating something different but you cant find out what it is until you are knee deep in the method section.**
+> Edit note: Currently these notes are just from spiegelhater because this is the only paper I have moved notes over from.
+Other sources will be used here lol.
+Working out what exactly they are communicating is a bit of a pain though.
 
@@
 The hierarchy established 40 years ago by @Cleveland1984 is:
+
 1) position along a common scale. 

### 2024-02-29 paper.qmd
paper updates from yesterday

@@
 Instead, we have a handful of highlights presented below.
+- @wu2023rational recognised it is difficult to understand when differences in performance stem from information asymmetry.
 - @Hofman2020 asked questions about discusses the difference in showing inferential or outcome (prediction) intervals and then found that participants shown the inferential distribution performed poorly when asked questions about predictions.
@@
 
+- The process of working out whether or not a visualisation shows a piece of information, and if it does, what perceptual task it is mapped to and what assumptions or heuristics need to be used, is *incredibly* time consuming.
+While this method does predict the results for these experiments with remarkable accuracy, to do it for a large number of plots, especially when the ground truth statistics and visualisation are incredibly complicated is just not feasible for a reviewer to do.
+
 ## Uncertainty evaluation experiments repeat visualisation studies
@@
 @Hullman2016 commented that it is straightforward to show a value but it is much more complex to show uncertainty, but did not explain *why* uncertainty information faces more difficulty.
-So, if we cannot see any difference to uncertainty information when it coems to value abstraction, *when* is uncertainty information different?
+After all, confidence intervals, minimums, maxmimums, variance, and any other statistic that can be considered related to "uncertainty" is just that, a statisitc.
+There is nothing special about these statistics that should differentiate them from a mean or median and imply their visualisation should be "special" in a way that would warrant its own field.
+Especially because it seems as though (at least in the case of visual inference) they *are* no different from any other value.
+So, if we cannot see any difference to uncertainty information when it comes to value abstraction, *when* is uncertainty information different?
 
@@
 - @Cheong2016 tested multiple different visual representations of uncertainty for representing the likelihood of a house being burned down based on its location.
+Their payment scheme, which paid out $0.10 for a correct choice (i.e.
+staying when the house was not burned down or leaving when the house was burned down) and 0 for an incorrect choice (i.e.leaving when the house didnt burn down or staying when the house burned down), meant participants were incentivised to base their entire leave/stay decision on whether or not the likelihood of a fire at their house is above or below 50%.
+The authors failled to recognise that this "complicated decision making task" boilled down to a simple value extraction problem, which the text made easiest to extract.
+The results from the experiment indicated that many of the participants also failled to recognise the obvious and simply strategy.
+
+- @Blenkinsop2000 tailored their questions according to the visualisation shown, which implies an understanding that different visualisations contain different information.
+Despite this the authors still compared the visualisations according to which were "useful for uncertainty visualisation".
+
+- @Blenkinsop2000 did try to include uncertainty as noise, asking participants to search for a specific outcome (the land classified as grass) in a random .
+However the participants failing to identify the signal (something that would be expected of an uncertainty plot) was seen as a sign the task was too complicated.
+While the task in this experiment *was* certainty far too complicated (and poorly communicated) this discussion indicates the authors did not consider the signal suppression to be the goal of the visualisation, and rather an annoying distraction.
+
+- @Boone2018
+
+- @wu2023rational attempted to remedy this issue caused by designing a "rational agent benchmark" such that participants in an experiment are compared to a rational agent.
+This suggestion is an insufficient band-aid solution that is built upon the same invalid assumptions as the rest of the literature, and simply digs the field deeper into these problems, rather than solve them.
+
+- @wu2023rational almost seems to recognise the problem of decision making experiments with payment schemes or correct/incorrect answers, but instead of understanding this boils these tasks down to value extraction, they suggest participants should be compared to a "rational agent".
+This rational agent acts as a benchmark, in the same way a ground truth statistic does for value extraction experiments, except the authors don't seem to realise that is what they have done.
+This paper highlights many decision making tasks are simply a value extraction experiment without realising it, this paper included.
 
@@
 @Patrick2023 compared people ability to recognise patterns in a residual plot to typical statistical tests and found human viewers looking at a plot were less sensitive than the typical residual tests.
+This increased sensisitivity could be the result of another aspect of uncertainty that is ignored.
+Statstical tests are typically built upon a series of assumptions, and it is difficult to identify *which* assumption failure caused the data to fail the statistical tests.
+Lineup plots allow us to see *how* our data is different to the assumed distribution, and better understand *why* our data may have failled a test ^[Did Patricks paper compare statistical tests to lineup protocols in the event an earlier assumption of the tests, such as, independent observation?
+I think he did check some.
+I wonder if you can quantify the information difference you get from a rejected statistical test vs a rejected line plot].
 These experiments utilised the lineup protocol which has been suggested as a method to check if perceived patterns are real or merely the result of chance [@Buja2009; @Wickham2010; @Chowdhury].
-This concept bears similarity to the goal of uncertainty visualisation, except uncertainty visualisation wants to achieve this within a single plot.
+This concept bears similarity to the goal of uncertainty visualisation, but it is not quite the same.
+@fig-hypyvs shows the conceptual difference between the lineup protocol and uncertainty visualisation.
+A lineup protocol displays the uncertainty about the null and identifies if the true data plot is identifiable (and therefore significiantly difference) while an uncertainty plot displays the variance of the estimated value and assesses if the null of "no signal" is within this plot (e.g.
+if the error bars overlap with zero).
+^[Does this flip have mathematical implications in how the plots should be assessed?] This implies a visible signal in an uncertainty plot should indicate some divergence from the null.
+
+![Difference between the null plot vs the uncertainty plot.](hypvs.jpeg){#fig-hypvs}
 
@@
 This has been used in maps with the pixel map [Lucchesi2017], but is more commonly used in animation with the HOPs plots [@Hullman2015] or similar concepts [@Blenkinsop2000] .
-The difficulty with resampling methods is that they are typically conflated with animation even though this does not necessarily need to be true, as can be seen in the @fig-pixelmap.
+The difficulty with resampling methods is that they are typically conflated with animation even though this does not necessarily need to be true, as can be seen in the @fig-pixelmap and the New York Times class mobility figure, shown in @fig-nyt, an animated version of which can be found [here](https://www.nytimes.com/interactive/2018/03/19/upshot/race-class-white-and-black-men.html).
 Even though @Kale2018 found that users were more sensitive to the underlying trend when shown the HOPs plot over the static outcoms plot, this effect went away when the speed of the animation increased, indicating that performance difference might be due to overplotting in the static case and a smaller sample size in the static condition would produce the same result.
@@
 This is not to say that visualisations of mass would not be able to perform signal supression, but a sample can easily be expressed using aesthetics such as colour on a map and mass visualisation often struggling with issues such as over or under smoothing.
+These sampling methods can show the messiness of the data that sits behind a model.
 
 ![](pixelmap.png){fig-pixelmap}
+![](classmobility.png){fig-nyt}
 
@@
 The raw data can also be used in a similar method to the HOPs plot, where a statistic of importance is implied by the visualisation, but the signal is supressed using the data itself instead of resampling methods.
-This method is used by the New York Times with their class mobility plot shown in fig, an animated version of which can be found [here](https://www.nytimes.com/interactive/2018/03/19/upshot/race-class-white-and-black-men.html).
 This does not mean that visualising raw data instead of implementing sampling techniques is always a valid uncertainty visualisation that will prevent insignificant signal from getting through.
@@
 ![](censusdotmap.png){fig-census}
-![](classmobility.png){fig-census}
 

### 2024-03-01 paper.qmd
Big restructure of paper

@@
 
-# Introduction
+# Background
+
+## Introduction
 
@@
 
-# Similar work
+## Similar work
 > Edit note: I think with the addition of more summary papers I should just summarise common themes of issues the papers ran into instead of detailling the work paper by paper.
@@
 
-# Defining Uncertainty
+## Types of Uncertainty Visualisation Papers
+
+> Edit notes: Originally I was planning to have the discussion following this section relate to these "types" of papers, but that didn't end up happening.
+Depending on what I do with later sections, this might need a rewrite or to just be removed, since I'm not sure how relevant this is to the main ideas of the paper.
+
+Uncertainty visualisation papers can be organised according to the *goal* of the experiment.
+Evaluation experiments are the standard rule for visualisations because the human brain is not as reliable as mathematical calculation.
+Therefore, user studies often aim to assess the limitations, biases, and heuristics of our mental calculator so that we can better understand the problems we may encounter when we plot our data.
+This is not to say any paper that suggests a visualisation without an evaluation experiment is completely lacking in justification, and there are many papers that suggest a novel visualisation without an evaluation study.
+Sometimes these papers are a preliminary step in finding a solution for common problems and intend to evaluate the visualsiation in later work.
+Until that later work is done, it is often difficult to accept one particular representation.
+While justification could come in the form of discussing established concepts in visualisation, such as the hierarchy of perceptual tasks, even these may need an evaluation study to back up their claims.
+
+The reasoning for this is obvious.
+There are a large number heuristics and biases that are not obvious to us when designing visualisation.
+Additionally, these heuristics and baises can change depending on the larger scope of the graphic and the population we are communicating with.
+Additionally, since there is often a miriad of ways to visualise any particular mathematical object, to adopt specific visualisation that is not already common practice needs reasonable justification, lest we run into a range of unforseen heuristic pitfalls.
+
+The remainder of section 4 will focus on the results and methods of uncertainty visualisation evaluation experiments.
+There are six main types of evaluation experiments in the uncertainty visualisation literature, they are:
+- Association perceptual tasks: These experiments identify perceptual tasks that are/could be associated with uncertainty, and test its association with uncertainty, check it's number of distinguishable levels, and/or check how quickly/accurately people can draw the uncertainty information from the graphic.
+- Value perceptual tasks: Participants are shown some depiction of uncertainty and are asked to extract a value, make a comparison, or do something simple to show that the graphic is readable.  
+- Heuristic Checks: These experiments are checking if a perceptual bias that has been identified through public use of a particular plot, other evaluation experiments, or gut feeling identified through public use of a plot or in a different.
+-  Plot Comparisons: These papers describe an uncertainty visualisation evaluation experiment, where they compare two or more uncertainty visualisaitons on a specific set of questions.
+These questions can be about specific statistics (e.g.
+What is the probability that A>B?) or about a secondary goal such as trust or decision making (e.g.
+How much do you trust this estimate?).
+- Qualitative analysis: These experiments show participants an uncertainty visualisation for an open ended or specific task, and asks them to describe how the visualisation was used to come to a conclusion, or they are asked to describe what they see in the visualisation.
+- Method papers: These papers discuss issues with the evaluation methods of uncertainty visualisaiton and suggest changes or improvements, typically with results indicating the benefit of the change.
+
+These experiments are not mutually exclusive within a paper, and a single study will often contain several experiments, usually from several of these categories.
+These distinctions are important because each have their own issues in how they relate to uncertainty more broadly.
+Below we will go through the five key issues in uncertainty visualisation, describe how they mainfest in the uncertainty visualisation literature, and offer some tentative solutions to these problems.
+
+We will also comment on some of the uncertainty visualisation papers that suggest a visualisation without an evaluation experiment.
+Some of these plots may offer a solution to the unique problems we mention in this paper and these visualisations will be discussed in "Part 6: Great Examples".
+We will also consider these papers to illustrate the issues that visualisations are aware of and working to fix, and discuss whether or not these directions of research are likely to bear fruit, and why.
+
+# The Uncertainty Definition Problem
+## Definitions in Existing Work
 Uncertainty visualisation is not made any easier by the fact that the term "uncertainty" lacks a commonly accepted definition in the literature.
@@
 
-## Coloquial Definitions
 The most common definition of uncertainty used in papers that discuss visualising it are colloquial.
@@
 
-## Taxonomy Definitions
-### Taxonomies of Uncertainty
+### Taxonomy Definitions
 > Edit Note: A lot of this is going to be moved to the uncertainty definition specific section of the vis lit review and the point here should just be "this is a mess" instead of trying to find any reason in it.
@@
 
-### Taxonomies of Visual Uncertainty
-> Edit Notes: Similar to the "similar works" section, I think this section would be better off with a short sentance about each paper and comments on their similarities because a paragraph for each paper is too much
-
-There are some taxonomies of visual uncertainty.
-These taxonomies seem to seek to define "uncertainty visualisation" and a combined term rather than as two separate words.
-We largely disagree with this conceptualization of uncertainty visualisations as it blurs the line between the mathematical estimations and assumptions (the uncertainty part) and the visual depiction of those mathematical objects (the visualisation part).
-While we will not focus on these taxonomies, it is still important to mention them and highlight the reasons they are not good practice.
-
-There are a small handful of taxonomies (or typologies) that are specifically for uncertainty visualisations.
-@Kinkeldey2014 identified give categories of uncertainty based on the work of other literature reviews of uncertainty visualisation.
-The five categories are: (1) explicit/implicit (directly mapping or showing multiple outcomes) (2) intrinsic/extrinsic (using existing symbols e.g.
-colour value, or new objects e.g.
-grids) (3) visually integral/separable (can or cannot be separated from the data) (4) coincidence/adjacent (if data and uncertainty are in integrated or separate views (5) static/dynamic (animation/interaction).
-Of the groups they identified, they actually only used (4), (2), and (5), left out (1) because most visualisations are explicit, and (3) corresponds to (1) in most cases [@Kinkeldey2014].
-
-@Potter2012 organised several existing uncertainty visualisations into groups based on the dimensionality of the data (1D, 2D, 3D, and No Dimension) and the dimensionality of the uncertainty (Scalar, Vector, Tensor).
-However, because the term "PDF", a statistical object that describes a random variable that is typically a one dimensional function, is used to describe both the data and the uncertainty for all dimensions, it is hard to understand how this dimensionality should work.
-
-@Grewal2021 created a taxonomy that mapped uncertainty visualisations to some point in a 2D space that was defined by the axis "domain expertise" and "continum of discreteness" that scaled from "point estimate" to "continuous distribtion".
-
-@Thomson2005 titled their paper "A typology for visualizing uncertainty" however it is a taxonomy for the specific types of uncertainty that we may want to visualise and a way to calculate that uncertainty.
-While this paper is more about mathematics than visualisation, it is clearly written as a precursor to visualisation.
-
-![Example of different uncertainty visualisation taxonomies](paperscreenshots/collage2.jpg){#fig-location}
-
-These taxonomies are useful in the sense that they often identify areas where there is an over or under supply of uncertainty visualisations.
-Additionally taxonomies do highlight a concept of uncertainty dimensionality that was not directly covered in the uncertainty taxonomy, however the distinction between uncertainty as a single value versus as a PDF could be considered a sub-consideration of uncertainty that's level is statistical so it can be quantified.
-While uncertainty visualisation taxonomies are not the standard practice (generating a taxonomy of uncertainty and then developing methods to depict that uncertainty is) [@Kinkeldey2014], it is still important to acknowledge the existence of "uncertainty visualisation typologies" and why they may miss the mark in what is needed from the uncertainty visualisaiton literature.
-
-The first issue with these taxonomies is that most of these taxonomies are created by observing existing uncertainty visualisations.
-This means under researched gaps in the taxonomy could be due to a lack of need just as well as a lack of attention, and the benefit of trying to create visualisations that land in uncharted areas of the taxonomy are unclear.
-
-The second issue is that there does not seem to be anything these taxonomies offer that would not be better established by separate taxonomies for uncertainty and visualisation.
-These taxonomies seem to depict uncertainty visualisation as a field within itself, rather than a specific case of information visualisation that requires more attention to mathematics than usual.
-The difference between "uncertainty visualisations" and "data visualisations" is not technically in the visual element, it is mathematical.
-@Kinkeldey2014 almost acknowledges this in their own paper that discusses an uncertainty visualsiation taxonomy when they claim "future typologies should take different categories of tasks into account (1) communication tasks (2) analytical tasks (3) exploratory task", a common typology for information visualisation in general.
-The process of understanding and estimating uncertainty requires knowledge of the data, the statistical methods used to make an estimate, and the assumptions of a model.
-Visualising the statistics that represent uncertainty should be no different than depicting the statistics that represent any other element of a graphic, and therefore there seems to be no reason to have a separate taxonomy for uncertainty visualisation.
-
-## Contextual Information
-> Edit Note: This section should probably be moved to the visualisation specific section, although that depends on the tone that I end up with after the rewrites of the earlier sections.
-
-Some uncertainty work, specifically uncertainty visualisation, focus on the context surrounding the uncertainty.
-These are things like spatial data, temporal data, category data, etc.
-While other discussions of uncertainty may consider this distinction more relevant to their work, here we will consider this contextual information.
-Contextual information is important for interpretation and understanding of graphics, but it does not in of itself generate uncertainty.
-This is not unique to the properties of spatial information, other statistics information that is important for modelling and assumptions, such as non-linearity, class levels, etc do not contribute to uncertainty.
-Operations on data, inherrent randomness, and assumptions that create bias generate uncertainty, which is what we are talking about here.
-A confidence interval is a confidence interval whether it came from spatial data, temporal, or cross sectional data.
-
-Papers that stress the importance of spatial or temporal contexts may be conflating contextual information with uncertainty.
-@fig-spatial depicts the role of contextual information in graphics, as something that helps us understand the information in the statistical graphic.
-When we create statistical models and quantify uncertainty, the spatial or temporal aspects are often boilled away [@Otsuka2023], and the only element of it that remains is the inexchangeability of our data.
-For this reason, there realistically is no such thing as "spatial" uncertainty in the same sense that there is no such thing as "photograph" uncertainty or "coin toss" uncertainty or "baby birth" uncertainty.
-Maps and photographs boil down to the same statistical objects in the same way that coin tosses and baby births boil down to the same statistical objects.
-Once we have removed the physical nature these objects are near identical when considered statistically.
-Of course the nature of them can change based on what we choose to observe (i.e.
-if we consider a coin landing on its side to be an outcome or if we consider intersex babies as an outcome in births) but this means the contextual information of our data, e.g.
-whether or not it is spatial or temporal, is only of importance at two stages of the data analysis pipeline.
-It is first important when we *boil it down* to our data through *observation* and again when we *map our statistical information back to its real world information* to *communicate* it.
-The only way the context uniquely contributes to uncertainty is in the "boil down" stage, however spatial or temporal uncertainty never refers to that specific cause of uncertainty.
-@Kinkeldey2014 discussed how uncertainty  can be represented by three components attribute (what) position (where) and temporal (when) and that studied typically deal with uncertainty around attribute but rarely position and time, however it is never specified what considerations should cause attribute uncertainty to be different to position or temporal uncertainty.
-@Kay2016 did an experiment that showed uncertainty around bus arrival times, however the visualisation used in the experiment, shown in @fig-bustime, is indistinguishable from most work that would be considered "attribute" uncertainty.
-
-![The uncertainty visualisation used by @Kay2016 to show the uncertainty around a bus arrival time prediction.](bustime.png){#fig-bustime}
-
-It is typically in reference to measurement error, data aggregation, statistical modelling, or other steps that occur *after* the boil down stage, and are therefore not unique to spatial or temporal data.
-While the spatial or temporal features *may* become relevant at the final communication stage, however this has little do with quantifying uncertainty, and the difficulties in communicating a complicated statistical model through a complicated map is already well discussed in the visualisation literature, and in later sections we will discuss how this concept is related to uncertainty visualisation.
-
-![Data pipeline with visualisations to show why we are ignoring contextual information for the time being.
-The visualisations that depict the data are similar to typical statistical graphics.
-The purpose of the spatial information is to provide context to the information we extract from the statistical aspect of the graphic.](spatialinfo.jpeg){#fig-spatial}
-
-This does not mean we are saying spatial and temporal information is never of importance.
-Spatial information is *highly relevant* when we translate from the abstract domain back into real world for communication.
-This position is best explained by @Leland2005 in *The Grammar of Graphics*:
-
-> "Geography is anchored in real space-time and statistics in abstract dimensions.
-This is a distinction along a continuum rather than a sharp break...
-but this difference in focus clearly means that a system optimized to handle geography will not be graceful when dealing with statistical graphics."
-
-This paper focuses on uncertainty, specifically uncertainty that is still in the the abstract *statistics* domain.
-Since we are still in the abstract domain, things that are anchored in real space-time create confusion that is evident in the literature.
-Spatial and temporal uncertainty papers limit how they communicate their uncertainty because they are considering what can be visualised inside the context provided by maps and, not what they should *try* to visualise based on the uncertainty abstraction.
-Additionally, since uncertainty is defined in the abstract domain, of the issues faced by spatial uncertainty visualisations are likely shared by other graphics that only differ in contextual information.
-
-In this sense we are not proposing that the current conceptualisation of data according to its physical properties is necessarily wrong, or that it leads to conclusions that are not necessarily specific to spatial or temporal data, and it may lead us to miss relevant information that is outside our domain.
-This is not only the case for uncertainty data.
-@Slingsby2023 discussed using a gridded glyphmap for spatial modelling, however the statistical elements are identical to a fluxuation diagram depicting $f(infection|time, age, longitude, latitude)$ presented in @Wickham2011.
-The plot by @Slingsby2023 is simply a special case of the graphic by @Wickham2011, where a separate visualisation is made for each time point, the continuous function $infection | age$ has replaced the categorical variable $happy$, a fade is added for the population size, and a map is shown in the background.
-@Slingsby2023 presents other maps in the paper, however they are also special cases of the product plots framework.
-This is not to say this extension is uninteresting, on the contrary we believe identifying cases where particular graphics are useful is a worthwhile endevour, but these plots seem to be have been developed unaware of the existing statistical framework.
-When we separate the field of visualisation along lines that may not be of particular relevance, we increase the likelihood of "reinventing the wheel" in each sub-dicipline.
-
-```{r}
-#| output: asis
-#| echo: false
-img_files <- fs::dir_ls("prodplots", glob="*.png")
-cat("::: {layout-ncol=2}\n",
-    glue::glue("![]({img_files})\n\n\n"),
-    ":::",
-    sep = ""
-)
-```
-
-# General Attitude Towards Uncertainty (Uncertainty Communication Boradly)
-
-> Edit note: this section needs a huge rewrite.
-The idea of this section blends way too much with the measurement discussions in the "noise and signal" section.
-I have already taken paragraphs from here and just pasted it in the noise section so it might not be coherrent.
-Additionaly this section goes on SEVERAL long rambles that realistically should have been a sentance or not included at all.
-
-It would be impossible to discuss uncertainty visualisation without mentioning the broader conversation around uncertainty communication.
-The conversation around uncertainty communication is often centered around the psychology is communicating risks.
-
-Most research into general uncertainty communication seems to focus specifically on how we communicate risks.
-Risk and uncertainty are not perfect synonyms of one another.
-@Spiegelhalter2017 makes the distinction clear by specifying *risk* as *a precise random event* (such as a coin flip) while *uncertainty* (also sometimes referred to as abiguity) is imprecision (such as the estimated probability associated with a coin that might be biased).
-For this reason, risk is often seen as a specific case of uncertainty and that well defined sub area has a lot of detailled research.
-Some markers of probability that have common uses, probabilities such as 0 (impossible), 0.5 (a fair coin flip), 1 (certain) are easy for people to have an internal sense of *hullman paper*.
-It is incredibly hard to communicate small risks, because people cannot differentiate between 1/100 an 1/1000, and communicating low probability but high impact risks are particularly challenging [@Spiegelhalter2017].
-This issue can be somewhat alleviated by communicating the likelihood of an event in relation to another event on a similar scale (such as being struck by lightning) [@Spiegelhalter2017].
-The framing of how we communicate risks matters due to the affect heuristic, so there will be a difference in peoples behaviours if, for example, you tell people they have a 90% chance of survival vs a 10% chance of death, and when expressing a frequency, a larger numerator communicates a larger risk [@Spiegelhalter2017].
-The reference class of a particular risk should be explicitly stated to avoid confusion, for example if we are communicating the chance of rain, the time period over which we are expecting this chance of rain (i.e.
-50% chance today or this week) needs to be specified [@Spiegelhalter2017].
-The problem with a risk focus is that it is only a small piece of the uncertainty pie, and very few of these findings can be reasonably assumed to go beyond risk communication.
-Even less can be assumed to translate to uncertainty visualisation.
-
-> Edit note: Currently these notes are just from spiegelhater because this is the only paper I have moved notes over from.
-Other sources will be used here lol.
-Working out what exactly they are communicating is a bit of a pain though.
-
-Risk and uncertainty are not only different in how they are communicated, but also why they are communicated.
-Risk communication focuses on communicating probability around events.
-These events are typically a binary event that is known to be random, so a version of communication that does not mention risks (e.g.
-telling someone they will or will not get cancer instead of communicating their risks about it) inherently carries with it a lack of transparency.
-Additionally risk communication is often communicated to avoid or information about an unfavorable outcome, so framing can be important if we want to influence decisions in a certainty way.
-These motivations do not extend to the communication of uncertainty information more broadly.
-While uncertainty information *is* often used to improve decisions, it seems the main motivation in its communication is to offer transparency to the viewer about a statistical analysis.
-That is, it acts as a form of statistical "hedging" for signals found in an analysis.
-Interviews with experts in statistic back up this primary motivation, as ignoring uncertainty information of often expressed as being similar to fraud or lying and the goal of "improving decisions" is only seen as a secondary outcome of an appropriately hedged signal [@Hullman2020a; @Manski2020].
-This concept of uncertainty means that the work on communicating risk does not cover the field of uncertainty communication as a whole.
-It covers the specific case where the signal itself *is* uncertain, but this work does not naturally extend to the goal of hedging a signal (i.e.
-when we want to communicate uncertainty as noise).
-
-Unfortunately understanding the relevant noise seems to be its own problem, as statisticians and visualisation authors seem to be unaware or confused in how to calculate the uncertainty related to an analysis [@Hullman2020a].
-Additionally, communicating uncertainty such that it conveys the appropriate singal supression involves combining uncertainty from multiple different sources, something that is considered to be difficult by even experts in uncertainty communication [@Spiegelhalter2017].
-
-There are some elements of risk communication that seem to translate to uncertaitny communication more broadly.
-Using words to communicate risk is discouraged because people can misinterpret the actual risk involved, so when possible numerical estimates of risk should be communicated, however this can make information harder to understand for those with poor numeracy skills [@Spiegelhalter2017].
-People performing better and preferring with numerical estimates of risk translates to uncertainty communication more broadly *inc (citations for this)*.
-Risk communication also needs to have clear objectives, use plain language, limit information to only what is necessary, and segment the audience to allow for differences in interest and knowledge [@Spiegelhalter2017].
-These general concepts of communication also extend to communication of uncertainty.
-On the other hand, it is not sensible to assume that probability specific concepts such as framing, intuitiveness of probabilities, or reference classes are still relevant when we consider uncertainty communication as signal supression, not as a synonym of risk or probability communication.
-In this same vein, most uncertainty communication papers very often discuss risk-aversion as something that needs to be considered despite it being unclear exactly how risk-aversion is related to uncertainty communication *(citations for the papers that do this)*.
-
-In the same sense that risk and uncertainty are different, risk-aversion and amiguity-aversion (which can be considered the uncertainty version of risk aversion) are also not the same.
-Risk-aversion is the tendency of people to prefer outcomes of low uncertainty to those outcomes with high uncertainty despite the outcomes with high uncertainty having a higher or equal expected outcome.
-In the case of communicating risk, the only reason this heuristic would be relevant was if risk-aversion was a mistake that needed to somehow be corrected for, rather than a heuristic that *reveals the utility of decreased uncertainty*.
-Risk-aversion is often treated as a mistake or something that should be avoided by the uncertainty visualisation literature even though that is not necessarily true, depending on what is causing the risk-aversion (e.g.
-is it due to poorer estimates or increased awareness of negative outcomes).
-If a particular communication method leads people to place a greater value on certainty or make worse estimates, that is the aspect of importance, not risk-aversion which conflates these factors.
-The work that advocates for transparency in risk-communication gives you whiplash when it's subtext argues that transparency leads to "incorrect" conclusions.
-Risk-aversion has a large number of confounding factors that make it difficult to understand if it is something to be considered in risk communication, ambiguity-aversion has a similar problem.
-Ambiguity-aversion is the tendency of people to prefer to take on certain risks rather than unknown risks.
-It is essentially risk-aversion applied to risk.
-For this reason discussions of ambiguity-aversion have the same issues as the conversations about risk-aversion, where they do not consider ambiguity-aversion to be an estimate of the utility of decreased uncertainty, it contains a subtext that argues against transparency, and it is considered a valuable insight in of itself despite conflating many factors.
-
-The idea that risk and ambiguity aversion reveal the utility of decreased variance seems to seldom be considered in the literature, therefore it should be mentioned here, otherwise this issue might continue.
-However, considering risk aversion is at its core, a trade off between bias and variance, it is bizzare to assume it to be a mistake when statisticians perform these trade offs all the time.
-If risk-aversion is truly "irrational" behaviour, every statistical textbook that discusses the use of a biased but consistent estimator in a finite sample case (a common example would be specific cases of the maximum likelihood estimator) should be tossed out.
-Discussing risk and ambiguity aversion as though it is some diversion from a perfectly rational, mathematical choice, implies there is *no value to a decrease in variance which is not even true within mathematics*.
-There have been other discussions on the appropriateness of discussing risk-aversion as a bias [@Vranas2000], but few have made the point that if uncertainty holds some *value* there is not technically a "right" decision at all.
-The concept of risk aversion treats risk as "noise", as though it is something that only exists to be ignored, but if the common place use of biased MLE's is an indication of anything, it is that *ignoring uncertainty* is the irrational choice.
-
-While this is a focus on uncertainty visualisation, these issues in uncertainty communication need to be acknowledged because they bleed into the uncertainty visualisation literature.
-Uncertainty visualisation are compared on their ability to communicate explicit risk, elicit trust, and prevent risk or ambiguity avoidance, continuing these issues in uncertainty communication at large.
-While uncertainty visualisation can make it easier to communicate the complicated details of uncertainty information, it carries some challenges that are unique to the visualisaition, specifically with respect to the "signal supression" concept.
-In simple estimates or verbal communication, the signal is often easy to identify because it is what we are explicitly saying.
-Visualisations are used in both data exploration and communication.
-This means what exactly is a *signal* in any particular visualisation is hard to identify, since we often let the visualisation *tell us* what the signal is.
-Additionally, you cannot add noise to *every single possible* signal one might take from a visualisation.
-Two people looking at the same visualisation might, just by chance, develop two entirely different insights.
-These unique and facinating challenges that are faced by viewing uncertainty visualisaiton through the lense of "signal supression" have been almost completely untouched by the literature.
-
-What is meant by "uncertainty" may seem obvious to some, but when you attempt to quantify or visualise it you will quickly find yourself asking, "uncertainty about...
-what?".
-Do you mean uncertainty on an estimate?
-On a forecast?
-How many steps ahead is this forecast?
-Are we only considering the uncertainty in the estimate or in the parameters or are we considering the possibility of measurement error or biased inputs?
-Signal and noise can only be untangled in the presence of a motivating question.
-
-# Uncertainty Visualisation Literature Summary
-The previous sections of this paper established several issues in uncertainty communication at large.
-In this section, we will discuss the current literature on uncertainty visualisation, how the uncertainty communication issues more broadly manifest in the specific case of visualisation, and what general lessons we can take from the existing work.
-
-## Types of Uncertainty Visualisation Papers
-
-> Edit notes: Originally I was planning to have the discussion following this section relate to these "types" of papers, but that didn't end up happening.
-Depending on what I do with later sections, this might need a rewrite or to just be removed, since I'm not sure how relevant this is to the main ideas of the paper.
-
-Uncertainty visualisation papers can be organised according to the *goal* of the experiment.
-Evaluation experiments are the standard rule for visualisations because the human brain is not as reliable as mathematical calculation.
-Therefore, user studies often aim to assess the limitations, biases, and heuristics of our mental calculator so that we can better understand the problems we may encounter when we plot our data.
-This is not to say any paper that suggests a visualisation without an evaluation experiment is completely lacking in justification, and there are many papers that suggest a novel visualisation without an evaluation study.
-Sometimes these papers are a preliminary step in finding a solution for common problems and intend to evaluate the visualsiation in later work.
-Until that later work is done, it is often difficult to accept one particular representation.
-While justification could come in the form of discussing established concepts in visualisation, such as the hierarchy of perceptual tasks, even these may need an evaluation study to back up their claims.
-
-The reasoning for this is obvious.
-There are a large number heuristics and biases that are not obvious to us when designing visualisation.
-Additionally, these heuristics and baises can change depending on the larger scope of the graphic and the population we are communicating with.
-Additionally, since there is often a miriad of ways to visualise any particular mathematical object, to adopt specific visualisation that is not already common practice needs reasonable justification, lest we run into a range of unforseen heuristic pitfalls.
-
-The remainder of section 4 will focus on the results and methods of uncertainty visualisation evaluation experiments.
-There are six main types of evaluation experiments in the uncertainty visualisation literature, they are:
-- Association perceptual tasks: These experiments identify perceptual tasks that are/could be associated with uncertainty, and test its association with uncertainty, check it's number of distinguishable levels, and/or check how quickly/accurately people can draw the uncertainty information from the graphic.
-- Value perceptual tasks: Participants are shown some depiction of uncertainty and are asked to extract a value, make a comparison, or do something simple to show that the graphic is readable.  
-- Heuristic Checks: These experiments are checking if a perceptual bias that has been identified through public use of a particular plot, other evaluation experiments, or gut feeling identified through public use of a plot or in a different.
--  Plot Comparisons: These papers describe an uncertainty visualisation evaluation experiment, where they compare two or more uncertainty visualisaitons on a specific set of questions.
-These questions can be about specific statistics (e.g.
-What is the probability that A>B?) or about a secondary goal such as trust or decision making (e.g.
-How much do you trust this estimate?).
-- Qualitative analysis: These experiments show participants an uncertainty visualisation for an open ended or specific task, and asks them to describe how the visualisation was used to come to a conclusion, or they are asked to describe what they see in the visualisation.
-- Method papers: These papers discuss issues with the evaluation methods of uncertainty visualisaiton and suggest changes or improvements, typically with results indicating the benefit of the change.
-
-These experiments are not mutually exclusive within a paper, and a single study will often contain several experiments, usually from several of these categories.
-These distinctions are important because each have their own issues in how they relate to uncertainty more broadly.
-Below we will go through the five key issues in uncertainty visualisation, describe how they mainfest in the uncertainty visualisation literature, and offer some tentative solutions to these problems.
-
-We will also comment on some of the uncertainty visualisation papers that suggest a visualisation without an evaluation experiment.
-Some of these plots may offer a solution to the unique problems we mention in this paper and these visualisations will be discussed in "Part 6: Great Examples".
-We will also consider these papers to illustrate the issues that visualisations are aware of and working to fix, and discuss whether or not these directions of research are likely to bear fruit, and why.
-
-## Defining uncertainty with resepct to inference
+## The role of inference
 When we consider mapping uncertainty, a reasonable questions that should come to mind is "uncertainty about what?".
@@
 @Fischhoff2014 looks at uncertainty visualisation for decision making decides that we should have different ways of communicating uncertainty based off what the user is supposed to do with it.
-Health have a wide range of metrics used to communicate different risks [@@Spiegelhalter2017] indicating they utilise new statistics based off the specific aspect of risk that is relevant to the issue.
+Health have a wide range of metrics used to communicate different risks [@Spiegelhalter2017] indicating they utilise new statistics based off the specific aspect of risk that is relevant to the issue.
 This should make it clear that uncertainty only exists when we are performing inference.
@@
 
-## Uncertainty evaluation experiments repeat visualisation studies
+# The Uncertainty Visualisation Definition Problem
+
+> Edit Notes: Similar to the "similar works" section, I think this section would be better off with a short sentance about each paper and comments on their similarities because a paragraph for each paper is too much
+
+## Uncertainty Visualisations in Existing work
+
+### Taxonomies of Uncertainty Visualisations
+There are some taxonomies of visual uncertainty.
+These taxonomies seem to seek to define "uncertainty visualisation" and a combined term rather than as two separate words.
+We largely disagree with this conceptualization of uncertainty visualisations as it blurs the line between the mathematical estimations and assumptions (the uncertainty part) and the visual depiction of those mathematical objects (the visualisation part).
+While we will not focus on these taxonomies, it is still important to mention them and highlight the reasons they are not good practice.
+
+There are a small handful of taxonomies (or typologies) that are specifically for uncertainty visualisations.
+@Kinkeldey2014 identified give categories of uncertainty based on the work of other literature reviews of uncertainty visualisation.
+The five categories are: (1) explicit/implicit (directly mapping or showing multiple outcomes) (2) intrinsic/extrinsic (using existing symbols e.g.
+colour value, or new objects e.g.
+grids) (3) visually integral/separable (can or cannot be separated from the data) (4) coincidence/adjacent (if data and uncertainty are in integrated or separate views (5) static/dynamic (animation/interaction).
+Of the groups they identified, they actually only used (4), (2), and (5), left out (1) because most visualisations are explicit, and (3) corresponds to (1) in most cases [@Kinkeldey2014].
+
+@Potter2012 organised several existing uncertainty visualisations into groups based on the dimensionality of the data (1D, 2D, 3D, and No Dimension) and the dimensionality of the uncertainty (Scalar, Vector, Tensor).
+However, because the term "PDF", a statistical object that describes a random variable that is typically a one dimensional function, is used to describe both the data and the uncertainty for all dimensions, it is hard to understand how this dimensionality should work.
+
+@Grewal2021 created a taxonomy that mapped uncertainty visualisations to some point in a 2D space that was defined by the axis "domain expertise" and "continum of discreteness" that scaled from "point estimate" to "continuous distribtion".
+
+@Thomson2005 titled their paper "A typology for visualizing uncertainty" however it is a taxonomy for the specific types of uncertainty that we may want to visualise and a way to calculate that uncertainty.
+While this paper is more about mathematics than visualisation, it is clearly written as a precursor to visualisation.
+
+![Example of different uncertainty visualisation taxonomies](paperscreenshots/collage2.jpg){#fig-location}
+
+These taxonomies are useful in the sense that they often identify areas where there is an over or under supply of uncertainty visualisations.
+Additionally taxonomies do highlight a concept of uncertainty dimensionality that was not directly covered in the uncertainty taxonomy, however the distinction between uncertainty as a single value versus as a PDF could be considered a sub-consideration of uncertainty that's level is statistical so it can be quantified.
+While uncertainty visualisation taxonomies are not the standard practice (generating a taxonomy of uncertainty and then developing methods to depict that uncertainty is) [@Kinkeldey2014], it is still important to acknowledge the existence of "uncertainty visualisation typologies" and why they may miss the mark in what is needed from the uncertainty visualisaiton literature.
+
+The first issue with these taxonomies is that most of these taxonomies are created by observing existing uncertainty visualisations.
+This means under researched gaps in the taxonomy could be due to a lack of need just as well as a lack of attention, and the benefit of trying to create visualisations that land in uncharted areas of the taxonomy are unclear.
+
+The second issue is that there does not seem to be anything these taxonomies offer that would not be better established by separate taxonomies for uncertainty and visualisation.
+These taxonomies seem to depict uncertainty visualisation as a field within itself, rather than a specific case of information visualisation that requires more attention to mathematics than usual.
+The difference between "uncertainty visualisations" and "data visualisations" is not technically in the visual element, it is mathematical.
+@Kinkeldey2014 almost acknowledges this in their own paper that discusses an uncertainty visualsiation taxonomy when they claim "future typologies should take different categories of tasks into account (1) communication tasks (2) analytical tasks (3) exploratory task", a common typology for information visualisation in general.
+The process of understanding and estimating uncertainty requires knowledge of the data, the statistical methods used to make an estimate, and the assumptions of a model.
+Visualising the statistics that represent uncertainty should be no different than depicting the statistics that represent any other element of a graphic, and therefore there seems to be no reason to have a separate taxonomy for uncertainty visualisation.
+
 The first issue in uncertainty visualisation is there does not seem to be a clear rule for what makes a particular visualisation a depiction of uncertainty.
@@
 The answer to this question is related to the definition problem discussed earlier, but to understand how, we will need to go through the grammar of graphics, and its absence in the uncertainty visualisation literature.
+
+### Contextual Information
+> Edit Note: This section should probably be moved to the visualisation specific section, although that depends on the tone that I end up with after the rewrites of the earlier sections.
+
+Some uncertainty work, specifically uncertainty visualisation, focus on the context surrounding the uncertainty.
+These are things like spatial data, temporal data, category data, etc.
+While other discussions of uncertainty may consider this distinction more relevant to their work, here we will consider this contextual information.
+Contextual information is important for interpretation and understanding of graphics, but it does not in of itself generate uncertainty.
+This is not unique to the properties of spatial information, other statistics information that is important for modelling and assumptions, such as non-linearity, class levels, etc do not contribute to uncertainty.
+Operations on data, inherrent randomness, and assumptions that create bias generate uncertainty, which is what we are talking about here.
+A confidence interval is a confidence interval whether it came from spatial data, temporal, or cross sectional data.
+
+Papers that stress the importance of spatial or temporal contexts may be conflating contextual information with uncertainty.
+@fig-spatial depicts the role of contextual information in graphics, as something that helps us understand the information in the statistical graphic.
+When we create statistical models and quantify uncertainty, the spatial or temporal aspects are often boilled away [@Otsuka2023], and the only element of it that remains is the inexchangeability of our data.
+For this reason, there realistically is no such thing as "spatial" uncertainty in the same sense that there is no such thing as "photograph" uncertainty or "coin toss" uncertainty or "baby birth" uncertainty.
+Maps and photographs boil down to the same statistical objects in the same way that coin tosses and baby births boil down to the same statistical objects.
+Once we have removed the physical nature these objects are near identical when considered statistically.
+Of course the nature of them can change based on what we choose to observe (i.e.
+if we consider a coin landing on its side to be an outcome or if we consider intersex babies as an outcome in births) but this means the contextual information of our data, e.g.
+whether or not it is spatial or temporal, is only of importance at two stages of the data analysis pipeline.
+It is first important when we *boil it down* to our data through *observation* and again when we *map our statistical information back to its real world information* to *communicate* it.
+The only way the context uniquely contributes to uncertainty is in the "boil down" stage, however spatial or temporal uncertainty never refers to that specific cause of uncertainty.
+@Kinkeldey2014 discussed how uncertainty  can be represented by three components attribute (what) position (where) and temporal (when) and that studied typically deal with uncertainty around attribute but rarely position and time, however it is never specified what considerations should cause attribute uncertainty to be different to position or temporal uncertainty.
+@Kay2016 did an experiment that showed uncertainty around bus arrival times, however the visualisation used in the experiment, shown in @fig-bustime, is indistinguishable from most work that would be considered "attribute" uncertainty.
+
+![The uncertainty visualisation used by @Kay2016 to show the uncertainty around a bus arrival time prediction.](bustime.png){#fig-bustime}
+
+It is typically in reference to measurement error, data aggregation, statistical modelling, or other steps that occur *after* the boil down stage, and are therefore not unique to spatial or temporal data.
+While the spatial or temporal features *may* become relevant at the final communication stage, however this has little do with quantifying uncertainty, and the difficulties in communicating a complicated statistical model through a complicated map is already well discussed in the visualisation literature, and in later sections we will discuss how this concept is related to uncertainty visualisation.
+
+![Data pipeline with visualisations to show why we are ignoring contextual information for the time being.
+The visualisations that depict the data are similar to typical statistical graphics.
+The purpose of the spatial information is to provide context to the information we extract from the statistical aspect of the graphic.](spatialinfo.jpeg){#fig-spatial}
+
+This does not mean we are saying spatial and temporal information is never of importance.
+Spatial information is *highly relevant* when we translate from the abstract domain back into real world for communication.
+This position is best explained by @Leland2005 in *The Grammar of Graphics*:
+
+> "Geography is anchored in real space-time and statistics in abstract dimensions.
+This is a distinction along a continuum rather than a sharp break...
+but this difference in focus clearly means that a system optimized to handle geography will not be graceful when dealing with statistical graphics."
+
+This paper focuses on uncertainty, specifically uncertainty that is still in the the abstract *statistics* domain.
+Since we are still in the abstract domain, things that are anchored in real space-time create confusion that is evident in the literature.
+Spatial and temporal uncertainty papers limit how they communicate their uncertainty because they are considering what can be visualised inside the context provided by maps and, not what they should *try* to visualise based on the uncertainty abstraction.
+Additionally, since uncertainty is defined in the abstract domain, of the issues faced by spatial uncertainty visualisations are likely shared by other graphics that only differ in contextual information.
+
+In this sense we are not proposing that the current conceptualisation of data according to its physical properties is necessarily wrong, or that it leads to conclusions that are not necessarily specific to spatial or temporal data, and it may lead us to miss relevant information that is outside our domain.
+This is not only the case for uncertainty data.
+@Slingsby2023 discussed using a gridded glyphmap for spatial modelling, however the statistical elements are identical to a fluxuation diagram depicting $f(infection|time, age, longitude, latitude)$ presented in @Wickham2011.
+The plot by @Slingsby2023 is simply a special case of the graphic by @Wickham2011, where a separate visualisation is made for each time point, the continuous function $infection | age$ has replaced the categorical variable $happy$, a fade is added for the population size, and a map is shown in the background.
+@Slingsby2023 presents other maps in the paper, however they are also special cases of the product plots framework.
+This is not to say this extension is uninteresting, on the contrary we believe identifying cases where particular graphics are useful is a worthwhile endevour, but these plots seem to be have been developed unaware of the existing statistical framework.
+When we separate the field of visualisation along lines that may not be of particular relevance, we increase the likelihood of "reinventing the wheel" in each sub-dicipline.
+
+```{r}
+#| output: asis
+#| echo: false
+img_files <- fs::dir_ls("prodplots", glob="*.png")
+cat("::: {layout-ncol=2}\n",
+    glue::glue("![]({img_files})\n\n\n"),
+    ":::",
+    sep = ""
+)
+```
+
+## "Uncertainty Visualisation" is not defined
 
@@
 In this sense, once we have a mathematical expression of uncertainty, the visualisation of that uncertainty is identical to the visualisation process of any other variable, or at least this is how it is treated in the uncertainty visualisation experiments.
+
+### Uncertainty as signal
 
@@
 
-## Measuring uncertainty as noise, not signal
+# Measureing noise, not signal
+
+## Uncertainty vs Risk
 Uncertainty is noise.
@@
 
+> Edit note: this section needs a huge rewrite.
+The idea of this section blends way too much with the measurement discussions in the "noise and signal" section.
+I have already taken paragraphs from here and just pasted it in the noise section so it might not be coherrent.
+Additionaly this section goes on SEVERAL long rambles that realistically should have been a sentance or not included at all.
+
+It would be impossible to discuss uncertainty visualisation without mentioning the broader conversation around uncertainty communication.
+The conversation around uncertainty communication is often centered around the psychology is communicating risks.
+
+Most research into general uncertainty communication seems to focus specifically on how we communicate risks.
+Risk and uncertainty are not perfect synonyms of one another.
+@Spiegelhalter2017 makes the distinction clear by specifying *risk* as *a precise random event* (such as a coin flip) while *uncertainty* (also sometimes referred to as abiguity) is imprecision (such as the estimated probability associated with a coin that might be biased).
+For this reason, risk is often seen as a specific case of uncertainty and that well defined sub area has a lot of detailled research.
+Some markers of probability that have common uses, probabilities such as 0 (impossible), 0.5 (a fair coin flip), 1 (certain) are easy for people to have an internal sense of *hullman paper*.
+It is incredibly hard to communicate small risks, because people cannot differentiate between 1/100 an 1/1000, and communicating low probability but high impact risks are particularly challenging [@Spiegelhalter2017].
+This issue can be somewhat alleviated by communicating the likelihood of an event in relation to another event on a similar scale (such as being struck by lightning) [@Spiegelhalter2017].
+The framing of how we communicate risks matters due to the affect heuristic, so there will be a difference in peoples behaviours if, for example, you tell people they have a 90% chance of survival vs a 10% chance of death, and when expressing a frequency, a larger numerator communicates a larger risk [@Spiegelhalter2017].
+The reference class of a particular risk should be explicitly stated to avoid confusion, for example if we are communicating the chance of rain, the time period over which we are expecting this chance of rain (i.e.
+50% chance today or this week) needs to be specified [@Spiegelhalter2017].
+The problem with a risk focus is that it is only a small piece of the uncertainty pie, and very few of these findings can be reasonably assumed to go beyond risk communication.
+Even less can be assumed to translate to uncertainty visualisation.
+
+> Edit note: Currently these notes are just from spiegelhater because this is the only paper I have moved notes over from.
+Other sources will be used here lol.
+Working out what exactly they are communicating is a bit of a pain though.
+
+Risk and uncertainty are not only different in how they are communicated, but also why they are communicated.
+Risk communication focuses on communicating probability around events.
+These events are typically a binary event that is known to be random, so a version of communication that does not mention risks (e.g.
+telling someone they will or will not get cancer instead of communicating their risks about it) inherently carries with it a lack of transparency.
+Additionally risk communication is often communicated to avoid or information about an unfavorable outcome, so framing can be important if we want to influence decisions in a certainty way.
+These motivations do not extend to the communication of uncertainty information more broadly.
+While uncertainty information *is* often used to improve decisions, it seems the main motivation in its communication is to offer transparency to the viewer about a statistical analysis.
+That is, it acts as a form of statistical "hedging" for signals found in an analysis.
+Interviews with experts in statistic back up this primary motivation, as ignoring uncertainty information of often expressed as being similar to fraud or lying and the goal of "improving decisions" is only seen as a secondary outcome of an appropriately hedged signal [@Hullman2020a; @Manski2020].
+This concept of uncertainty means that the work on communicating risk does not cover the field of uncertainty communication as a whole.
+It covers the specific case where the signal itself *is* uncertain, but this work does not naturally extend to the goal of hedging a signal (i.e.
+when we want to communicate uncertainty as noise).
+
+Unfortunately understanding the relevant noise seems to be its own problem, as statisticians and visualisation authors seem to be unaware or confused in how to calculate the uncertainty related to an analysis [@Hullman2020a].
+Additionally, communicating uncertainty such that it conveys the appropriate singal supression involves combining uncertainty from multiple different sources, something that is considered to be difficult by even experts in uncertainty communication [@Spiegelhalter2017].
+
+There are some elements of risk communication that seem to translate to uncertaitny communication more broadly.
+Using words to communicate risk is discouraged because people can misinterpret the actual risk involved, so when possible numerical estimates of risk should be communicated, however this can make information harder to understand for those with poor numeracy skills [@Spiegelhalter2017].
+People performing better and preferring with numerical estimates of risk translates to uncertainty communication more broadly *inc (citations for this)*.
+Risk communication also needs to have clear objectives, use plain language, limit information to only what is necessary, and segment the audience to allow for differences in interest and knowledge [@Spiegelhalter2017].
+These general concepts of communication also extend to communication of uncertainty.
+On the other hand, it is not sensible to assume that probability specific concepts such as framing, intuitiveness of probabilities, or reference classes are still relevant when we consider uncertainty communication as signal supression, not as a synonym of risk or probability communication.
+In this same vein, most uncertainty communication papers very often discuss risk-aversion as something that needs to be considered despite it being unclear exactly how risk-aversion is related to uncertainty communication *(citations for the papers that do this)*.
+
+In the same sense that risk and uncertainty are different, risk-aversion and amiguity-aversion (which can be considered the uncertainty version of risk aversion) are also not the same.
+Risk-aversion is the tendency of people to prefer outcomes of low uncertainty to those outcomes with high uncertainty despite the outcomes with high uncertainty having a higher or equal expected outcome.
+In the case of communicating risk, the only reason this heuristic would be relevant was if risk-aversion was a mistake that needed to somehow be corrected for, rather than a heuristic that *reveals the utility of decreased uncertainty*.
+Risk-aversion is often treated as a mistake or something that should be avoided by the uncertainty visualisation literature even though that is not necessarily true, depending on what is causing the risk-aversion (e.g.
+is it due to poorer estimates or increased awareness of negative outcomes).
+If a particular communication method leads people to place a greater value on certainty or make worse estimates, that is the aspect of importance, not risk-aversion which conflates these factors.
+The work that advocates for transparency in risk-communication gives you whiplash when it's subtext argues that transparency leads to "incorrect" conclusions.
+Risk-aversion has a large number of confounding factors that make it difficult to understand if it is something to be considered in risk communication, ambiguity-aversion has a similar problem.
+Ambiguity-aversion is the tendency of people to prefer to take on certain risks rather than unknown risks.
+It is essentially risk-aversion applied to risk.
+For this reason discussions of ambiguity-aversion have the same issues as the conversations about risk-aversion, where they do not consider ambiguity-aversion to be an estimate of the utility of decreased uncertainty, it contains a subtext that argues against transparency, and it is considered a valuable insight in of itself despite conflating many factors.
+
+The idea that risk and ambiguity aversion reveal the utility of decreased variance seems to seldom be considered in the literature, therefore it should be mentioned here, otherwise this issue might continue.
+However, considering risk aversion is at its core, a trade off between bias and variance, it is bizzare to assume it to be a mistake when statisticians perform these trade offs all the time.
+If risk-aversion is truly "irrational" behaviour, every statistical textbook that discusses the use of a biased but consistent estimator in a finite sample case (a common example would be specific cases of the maximum likelihood estimator) should be tossed out.
+Discussing risk and ambiguity aversion as though it is some diversion from a perfectly rational, mathematical choice, implies there is *no value to a decrease in variance which is not even true within mathematics*.
+There have been other discussions on the appropriateness of discussing risk-aversion as a bias [@Vranas2000], but few have made the point that if uncertainty holds some *value* there is not technically a "right" decision at all.
+The concept of risk aversion treats risk as "noise", as though it is something that only exists to be ignored, but if the common place use of biased MLE's is an indication of anything, it is that *ignoring uncertainty* is the irrational choice.
+
+While this is a focus on uncertainty visualisation, these issues in uncertainty communication need to be acknowledged because they bleed into the uncertainty visualisation literature.
+Uncertainty visualisation are compared on their ability to communicate explicit risk, elicit trust, and prevent risk or ambiguity avoidance, continuing these issues in uncertainty communication at large.
+While uncertainty visualisation can make it easier to communicate the complicated details of uncertainty information, it carries some challenges that are unique to the visualisaition, specifically with respect to the "signal supression" concept.
+In simple estimates or verbal communication, the signal is often easy to identify because it is what we are explicitly saying.
+Visualisations are used in both data exploration and communication.
+This means what exactly is a *signal* in any particular visualisation is hard to identify, since we often let the visualisation *tell us* what the signal is.
+Additionally, you cannot add noise to *every single possible* signal one might take from a visualisation.
+Two people looking at the same visualisation might, just by chance, develop two entirely different insights.
+These unique and facinating challenges that are faced by viewing uncertainty visualisaiton through the lense of "signal supression" have been almost completely untouched by the literature.
+
+What is meant by "uncertainty" may seem obvious to some, but when you attempt to quantify or visualise it you will quickly find yourself asking, "uncertainty about...
+what?".
+Do you mean uncertainty on an estimate?
+On a forecast?
+How many steps ahead is this forecast?
+Are we only considering the uncertainty in the estimate or in the parameters or are we considering the possibility of measurement error or biased inputs?
+Signal and noise can only be untangled in the presence of a motivating question.
+
+The previous sections of this paper established several issues in uncertainty communication at large.
+In this section, we will discuss the current literature on uncertainty visualisation, how the uncertainty communication issues more broadly manifest in the specific case of visualisation, and what general lessons we can take from the existing work.
+
+## Decision Making
+
 Decision making questions that typically involve uncertainty discuss the concept of "risk-aversion" as though it is a mistake.
@@
 
+## Trust
 Authors also occasionally use trust as a proxy for uncertainty.
@@
 
+## Other attemps to capture uncertainty
 There are other studies that try to include uncertainty information without directly asking for it but do not go down the route of asking for a decision or levels of trust.
@@
 
+## Uncertainty is not hypothesis testing
 It is unclear how participants are supposed to incorperate uncertainty information into their responses when uncertainty visualisation authors themselves are unsure.

### 2024-03-01 paper.qmd
added notes and adjusted some bits

@@
 
-> Edit note: this section is untouched since the confirmation. Might need a rewrite
+> Edit note: this section is untouched since the confirmation. Might need a rewrite. should comment that visualisation helps us understand *how* our data is different.
 
@@
 
-- @Boone2018
-
-- @wu2023rational attempted to remedy this issue caused by designing a "rational agent benchmark" such that participants in an experiment are compared to a rational agent.
-This suggestion is an insufficient band-aid solution that is built upon the same invalid assumptions as the rest of the literature, and simply digs the field deeper into these problems, rather than solve them.
-
-- @wu2023rational almost seems to recognise the problem of decision making experiments with payment schemes or correct/incorrect answers, but instead of understanding this boils these tasks down to value extraction, they suggest participants should be compared to a "rational agent".
-This rational agent acts as a benchmark, in the same way a ground truth statistic does for value extraction experiments, except the authors don't seem to realise that is what they have done.
-This paper highlights many decision making tasks are simply a value extraction experiment without realising it, this paper included.
+- @Ibrekk1987 asked participants for the "best estimate" which was provided in terms of most probable value (mode), the mean and the median, the "ground truth" statistic used was the mean, despite the "best estimate" largely depending on the loss function of the model and the methods used.
+
+- @Ibrekk1987 used several displays of a probability density functions, since “formally equivalent representations are often not psychologically equivalent”.
+
+- @Ibrekk1987 highlights that in the face of a vague question, participants will use the plot to decide what the authors mean. The authors take this as a
+
+- @Boone2018 organised responses from participants by the approach they used and tried to account for participants prior beliefs (although the assessment of those beliefs used the word "likely" which is open to interpretation).
+
+There is so little understanding of information in uncertainty graphics that participants are frequently provided with information that is insufficient to answer the question at all, an issue that is regularly missed by the experimenters.
+In order to test whether or not people incorrectly interpreted the cone of uncertainty that is used to communicate cone tasks, experimenters regularly do not give participants enough information to answer the questions [@Padilla2017; @Boone2018].
+The cone of uncertainty provides a 95% confidence interval for the *eye* of a hurricane, which allows us to know *where the eye of the storm will likely go*, this does not, give us enough information to answer if the storm will *hit* a particular point, because this requires information about the size and intensity of the storm, *and* an assumption about whether or not the size and intensity will change over time.
+Since these experiments are often trying to *test* for common misreadings of the plot, such as only thinking the area inside the cone will be damaged or believing the cone represents size and the storm is getting bigger.
+The authors often fail to recognize that the exact information that has been witheld for testing is necessary for answering their questions, and considering that information is necessary to mark participants as correct or incorrect, it is hard to understand exactly how they evaluated the experiment participants.
+@Ibrekk1987 expected participants to calculate the mean using a pie chart that had numerical bins combined (e.g.
+segment one was 0-2 inches of snow, and segment two was 2-4 inches of snow) so it could not be done by calculation.
+The authors also expected participants to use a CDF to calculate the mean, which needs to be converted *back* into a PDF to do so mathematically, so it was unclear how participants were supposed to do this calculation visually unless they were to just make a blind guess.
+The authors interpreted this as the participants misinterpreting how the CDF works, not seeming to realise they gave them an impossible task.
+In light of this, the main take away from these papers seems to be "if you do not give people the necessary information to answer a question, they will be unable to answer it", which does not seem interesting enough to justify several papers.
 
@@
 
+There are also some interesting qualitative studies that take an investigative stance to the use of uncertainty in decision making rather than an evaluation stance.
+@Daradkeh2015 presented participants with ten investment alternatives and asked participants "from among available alternatives, which alternative do you prefer the most", and were asked to think aloud and consider the uncertainty in their decision making.
+This study was qualitative and rather than setting a "correct" answer that would have forced a value of risk on the participants, the experimenters goal was to observe and organise the methods people use when making decisions in the face of uncertainty.
+This study was an excellent example in a useful experimental design.
+They highlighted the specific aspects of uncertainty that participants typically considered, such as the range of outcomes that are above/below a certain threshold, minimum and maximum values, the risk of a loss, etc, and mapped where in the decision making process participants made these considerations.
+
 ![I will just make my own version of a climate scenario plot, this one is just here from a random site as a place holder](climatescenario.png){fig-climatescenario}
@@
 Reguardless, these are directions of research would be fruitful to the uncertainty visualisation community even if it appears on the surface to be research that is only beneficial to the "normal" visualisation community.
+
+# Conclusion
+In this paper we have highlighted a series of misunderstandings, confusions, and methodical errors in the uncertainty literature.
+
+Many of these issues are not problems in isolation, so long as the authors themselves are aware of it and state these assumptions in their papers.
+Our problem with this body is work is that the authors themselves seem to be oblivious to this vast array of problems, considering not a single one of the problems discussed here are mentioned explicitly in a single visualisation paper.
 

### 2024-03-01 writingdumpster.qmd
added notes and adjusted some bits

@@
 # Moved over notes
+- @Daradkeh2015 found that people tended to to make decisions on a small number of salient pieces of information rather than on systematic evaluations of the costs and benefits of choices.
+- @Daradkeh2015 found that people tended to to make decisions on a small number of salient pieces of information rather than on systematic evaluations of the costs and benefits of choices.
+@Daradkeh2015 found that people tended to ignore or give little weight probability information and instead focused on the severity of undersierable outcomes.
+@Daradkeh2015 found that people tended to ignore or give little weight probability information and instead focused on the severity of undersierable outcomes.
+- @wu2023rational attempted to remedy this issue caused by designing a "rational agent benchmark" such that participants in an experiment are compared to a rational agent.
+This suggestion is an insufficient band-aid solution that is built upon the same invalid assumptions as the rest of the literature, and simply digs the field deeper into these problems, rather than solve them.
+- @wu2023rational almost seems to recognise the problem of decision making experiments with payment schemes or correct/incorrect answers, but instead of understanding this boils these tasks down to value extraction, they suggest participants should be compared to a "rational agent".
+This rational agent acts as a benchmark, in the same way a ground truth statistic does for value extraction experiments, except the authors don't seem to realise that is what they have done.
+This paper highlights many decision making tasks are simply a value extraction experiment without realising it, this paper included.
 - make sure quantified vs discussed uncertainty is mentioned

### 2024-03-03 paper.qmd
update introduction

@@
 ## Introduction
-
-> Edit note: this section is untouched since the confirmation. Might need a rewrite. should comment that visualisation helps us understand *how* our data is different.
-
-Think back to the last time you made some sort of data visualisation.
-What was the purpose of that visualisation?
-Was it to better understand your data?
-Was it to help you make a decision?
-Was it to to communicate that decision to someone else?
-Now think about the last time you expressed some form of uncertainty.
-Was it a set of numerical confidence intervals?
-Maybe they were expressed as a set of values in a table.
-Did you consider visualising your uncertainty instead?
+From entertainment choices to news articles to insurance plans, the modern citizen is so over run with information in every aspect of their life it can be overwhelming.
+In this overflow of information, tools that can effectively summarise information down into simple and clear ideas become more valuable.
+Information visualisations remain one of the most powerful tools for fast and reliable science communication.
+
 There are many stages in our analysis that benefit from the power of data visualisation, however this does not mean it is always done with success.
@@
 Anscombe quartet is shown in @fig-anscombe, because describing the data is never the same as seeing it.
-Instead of having to repeatedly check endless hypothesis to find interesting numerical features, visualisations **tell** us what is important about our data.
-This powerful aspect of data visualisation is poorly or seldom used in later stages when we are communicating our findings, specifically with respect to uncertainty.
+Instead of having to repeatedly check endless hypothesis to find interesting numerical features, visualisations **tell** us what is important about our data and **how** it might diverge from what we expect.
 
@@
 
+Uncertainty visualisation is a somewhat new subfield of visualisation, with early papers that specifically reference an "uncertainty visualisation" popping up in the late 90s and early 2000s (*Cite Gap 1*).
+Despite the youth of the field, commonly used uncertainty visualisations such as a box plot or histogram have been around for almost as long as statistical graphics themselves (*Cite Gap 2*).
+Early mentions of "uncertainty visualisation" appear in computer science papers discussing uncertainty visualisation as a subfield of geospatial information visualisation, specifically for the many uncertainties that come with satelite images (*Cite Gap 1*).
+In the modern day, uncertainty visualisation has exploded into its own field that applies to a wide range of contexts and is no longer bound by geospatial information (*Cite Gap 1*).
+
 ## Similar work
-> Edit note: I think with the addition of more summary papers I should just summarise common themes of issues the papers ran into instead of detailling the work paper by paper.
-
-There are a wealth of uncertainty communication and visualisation literature reviews that already exist. Interestingly they all seem to have identified similar pitfalls or fallen into one themselves.
+::: aside
+I have a large backlog of uncertainty literature review summaries in my Mendely that need to be added here. That is very high on my todo list once the readable draft is done.
+:::
+
+Any field that experiences an explosion of new research will also invite a series of literature reviews that seek to combine and summarise that information.
+Given that uncertainty visualisation has been around for a few decades now, there is already a wealth of uncertainty visualisation literature reivews.
+Interestingly these papers seem to have identified a series of common pitfalls in uncertainty visualisation and its evaluation.
 
@@
 Another approach is to consider uncertainty visualsiation as a specific sub-field of high-dimensional data visualsiation, and therefore uncertainty is just another variable that needs to be expressed.
+
+> Edit note: just summarise common themes of issues the papers ran into instead of detailling the work paper by paper.
 

### 2024-03-03 writingdumpster.qmd
update introduction

@@
 	- Participant literacy
+
+# Citation Gaps
+Cite Gap 1: When did uncertainty visualisaiton become its own subfield of visualisation?
+
+Cite Gap 2: When was the first "uncertainty visualisation" used (likely a box plot of histogram)
 
@@
 # Previous literature review
-##### Introduction addition
-From entertainment choices to news articles to insurance plans, the modern citizen is so over run with information in every aspect of their life it can be overwhelming.
-So overwhelming in fact, that it is not uncommon for people to become so overwhelmed by the information they refuse to engage with it at all.
-In this overflow of information, tools that can effectively summarise information down into simple and clear ideas become more valuable.
-While websites that summarise insurance plans or hotel options down into a clear table make a living off it, few tools are as powerful or effective as a visualisation.
-Information visualisations remain one of the most powerful tools for fast and reliable science communication.
-Effectively leveraging visualisations to make them more effective is therefore key in facilitating simple and clear communication of science and other ideas.
 

### 2024-03-09 paper.qmd
updated the similar work section

@@
 ::: aside
-I have a large backlog of uncertainty literature review summaries in my Mendely that need to be added here. That is very high on my todo list once the readable draft is done.
+I have a large backlog of uncertainty literature review summaries in my Mendely that need to be added here. There are a lot of them.
 :::
 
-Any field that experiences an explosion of new research will also invite a series of literature reviews that seek to combine and summarise that information.
-Given that uncertainty visualisation has been around for a few decades now, there is already a wealth of uncertainty visualisation literature reivews.
-Interestingly these papers seem to have identified a series of common pitfalls in uncertainty visualisation and its evaluation.
-
-@Kinkeldey2014 summarised the literature specifically for geopsatial uncertainty visualisations and attempted to describe similarities in their characteristics.
-They found most experiments the methods for uncertainty visualisation evaluation to be adhoc, with no commonly agreed upon methodology or formalisation.
-They found that experiments took on an engineering approach, where plots were evaluated with the goal to improve a particular representation, rather than to investigate why particular representations do or do not work and create general principals.
-Authors would then ignore the engineering perspective they had taken on, and try to generalise beyond the specific constraints of their experiment.
-While they did discuss the results of some perceptual task experiments they found that even those results were somewhat conflicting and difficult to combine in any meaningful way.
-They also stress the importance of selecting the correct audience, as numerical and visual literacy can differ vaslty between subpopulations so the results from one populations may not reflect the results from another.
-The paper contains an entire paragraph where experiments that compared box-plot representations are rattled off, each experiment finding results that conflicted with its predecessor.
-Ultimately they suggested the visualisation typologies move away from data types, uncertainty categories, and representation types and towards "task-centered typologies" to better enable generalisations of the findings.
-
-@Hullman2016 reviewed the uncertainty visualisation literature more broadly (not specifically for geospatial uncertainty) and offered advice on methodology and analysis of results for other uncertainty visualisation experimenters.
-She found similar issues in the literature and also noticed the somewhat ad-hoc methodology though the lack of justification in the method for the particular questions participants were required to answer, and the lack of generalisability of comparing plots one by one to create a ranking.
-She also discussed the role of heuristics in the analysis of results and suggested authors look for signs of them in responses and ask subjects to describe their strategy in their response.
-What was particularly interesting was people's tendancy to ignore uncertainty in a plot all togehter, and use a pattern in the signal (such as distance between means) to answer questions that require uncertainty information.
-She also discussed the lack of clarity in exactly what "uncertainty" authors are trying to ask about.
-Whether the participants are supposed to answer according to some ground truth or their subjective probability is unclear, as is whether the *uncertainty* depicted represents the uncertainty of a sampling or population distribution.
-Ultimately all these issues become confliated and it becomes impossible to tell if subjects did poorly in an experiment because they misunderstood a visualisation, because the question was misinterpreted, or because they used a heuristic.
-
-@Spiegelhalter2017 presents an literature reivew of uncertainty communication as a whole and considers uncertainty visualisation as a subfield of communication as a whole.
-In this review he has several key findings that are of particular interest.
-First, the best visualisation is highly dependent on your audience, and depending on how broad your audience is, there may not be a best visualisation.
-No single representation will be accessible to all members of an audience, and even if most audience members can read a plot, they may not read it in the same way.
-For example, people with high numeracy skills count pictographs, but people with low numeracy assess area, and seemingly unrelated elements may not equally influence all these methods of reading, for example scattering in a pictograph makes counting particularly more difficult.
-For these reasons, it is better to understand your audience and iterate towards a final design, rather than only consider one option.
-Second, the best visualisation for one task is unlikely to be the best visualisation for another task.
-Graphics are good for "gists" and different graphics are good for different "gists", e.g.
-one plot may we good at identifying trend, while another might be good for comparison.
-Therefore, if we change the goal of our visualisation, and therefore the metric we are trying to evaluate, we implicitly change the best visualisation as no visualisation will both cover and summarise enough information to be the best at answering every question.
-Finally, the visualisation advice he gives is more specifically about general visualisation practice rather than advice specifically for uncertainty visualisation.
-Common sentiments such as avoiding chart junk, considering a table in place of a visualsiaiton, use of narrative labels, managing emotion, less is more approach, etc, are the main concepts expressed.
-The only uncertainty specific advice given is that the graphic should allow "part-to-whole" comparison on an appropriate scale, which seems to be an extension of the framing concept discussed when speaking about uncertainty communication more broadly.
-
-@Griethe2006 provided a summary of the uncertainty visualisation literature and did not take any particular angle in the analysis.
-She makes the point that there is no universally accepted definition of uncertainty and no unicersally accepted way to display it.
-She defined uncertainty as "the degree to which the lack of knowledge about the amount of error is responsible for hesitancy in accepting results and observations without caution".
-She then explains that there are two ways to visualise uncertianty, implicitly and explicitly.
-When a visualisaion is implicit, the uncertainty is considered an uninteresting feature of the data, and instead of mapping some quantification of uncertainty to a visual feature, data is filtered out if it does not meet a specified "certainty" threshold.
-These statements clearly points towards a signal/noise dichotomy with normal visualisation and uncertainty visualisation.
-Additionally she notes that the motivation of the analysis is important, by stating "one has to answer the questions which kinds of uncertainty are relevant from the data’s perspective, which kinds are helpful from the perspective of the visualization steps and which kinds are practicably extractable from the perspective of the available methods?", and "The decision of which of these techniques to choose strongly depends on the intended goal" when commenting on which uncertainty visualisation techniques should be used.
-She clearly sees uncertainty as additional information or data, and therefore sees the uncertainty visualisation problem as a specific case of the high dimensional visualisation issues.
-She mentions that there are detailled classifications of uncertainty but the literature and subsequent techniques are not ballanced, and instead concentrate on scalar values.
-In facing this lack of ballance she creates a new classification system for uncertainty visualisation.
-She commented that uncertainty appears at every step in the analyis pipeline, and commented on some the mathematical definitions that seek to quantify uncertianty, but did not explain how these mathematical definitions can be combined across levels.
-Unfortunately this quantification is needed since she mentioned that in order to visualise uncertainty, it needs to be measurable.
-
-These literature reviews highlight a common theme, the field of uncertainty visualisation is near impossible to navigate.
-There is so much noise in the experimental results that general patterns are difficult to identify and extract for a common purpose.
-This noise is attributed to several things, the results dependence on tasks, the difficulty in understanding the specifics of uncertainty and the conflation of signal and noise from many steps in the experimental design.
-Additionally these reviews often try to look at uncertainty visualisation from a specific perspective (e.g.
-geospatial or bayesian modelling) and find that even within these sub-fields the data is too noise to come to any meaningfull conclusions.
-
-In contrast to these literature reviews, we do not have a specific domain or tool analysis we intend to perform.
-Rather, the point of this literature review is to understand the *purpose* of visualising uncertainty, and understand if the current visualisation tools are sufficient to achieve that purpose and if not, where there is need for improvement.
-Additionally, rather than discussing "uncertainty visualisation" as its own field, we view it as a subfield of both uncertainty and visualisation, and understand if this change in conceptualisation can alleviate or help us to find some signal in the noise of the field.
+Any field that experiences an explosion of new research will also invite a series of literature reviews that seek to combine and summaries that information.
+Given that uncertainty visualisation has been around for a few decades now, there is already a wealth of uncertainty visualisation literature reviews.
+
+Interestingly, these reviews often do not offer overarching rules for tried and tested uncertainty visualisation, but rather comment on the *difficulties* faced when trying combine the papers from this field.
+@Kinkeldey2014 found most experiments the methods for uncertainty visualisation evaluation to be adhoc, with no commonly agreed upon methodology or formalisation and no greater goal of describing general principals.
+@Hullman2016 commented on the difficulty in taking overarching themes from uncertainty visualisation, as several conflated issues make it unclear if subjects did poorly in an experiment because they misunderstood a visualisation, because the question was misinterpreted, or because they used a heuristic.
+@Spiegelhalter2017 commented that different plots are good for different things, and disagreed with the goal of identifying a universal "best" plot for all people and circumstances.
+@Griethe2006 was unable to find common themes, but instead listed the findings and opinions of each uncertainty visualisation paper.
+
+There are a handful of reasons that are used to explain why it is so difficult to generalise the findings in the uncertainty visualisation literature.
+Some papers suggested or agree with the idea that visualisation typologies should move away from data types, uncertainty categories, and representation types and towards "task-centered typologies" as this will help generalise results [@Kinkeldey2014; @Hullman2016].
+Some authors may not attribute the task to the *reason* the literature cannot be easily summarised, but do mention that the choice of best visualisation is highly dependent on the specific goal of that visualisation [@Griethe2006; @Spiegelhalter2017].
+Others comment that visualisations are highly dependent on the audience and there is no such thing as a "best" visualisation that will be accessible to all members [@Kinkeldey2014; @Spiegelhalter2017].
+These difficulties in noisy uncertainty visualisation findings arises reguardless of the direction the analysis was approached, be it looking at geospatial data [@Kinkeldey2014], or uncertainty visualisation as a whole [@Hullman2016; @Spiegelhalter2017; @Griethe2006].
+
+This review believe all these sources of noise have a single root cause, a lack of a cohesive and encompassing definition of uncertainty.
+The lack of an econompasing definition is touched on by every uncertainty visualisation directly [@Spiegelhalter2017; @Griethe2006] or indirectly by describing a miriad of ways it can be considered in the literature [@Kinkeldey2014; @Hullman2016], although it is never commented on as a source of the noise in the field.
+
+In understanding the definition problem, we will understand the *purpose* of visualising uncertainty, and understand if the current visualisation tools are sufficient to achieve that purpose and if not, where there is need for improvement.
+In understanding this purpose, we will also consider whether or not "uncertainty visualisation" is a valid sub-field of visualisation itself, or if that distinction has only arisen as a by product of the confused definition of uncertainty.
 Ultimately, this paper should serve as a guide for future uncertaitny visualisation authors to prevent their work from becoming noise in what is already an incredibly noisy field.
-
-There are several frameworks that the uncertainty visualsiation problem is approached from.
-One approaches it from the concept of risk, and tries to answer how to best express uncertainty in a visualisation such that the viewer can acurately extract and interpret the risks involved.
-Another approach is to consider uncertainty visualsiation as a specific sub-field of high-dimensional data visualsiation, and therefore uncertainty is just another variable that needs to be expressed.
 
@@
 
+There are several frameworks that the uncertainty visualsiation problem is approached from.
+One approaches it from the concept of risk, and tries to answer how to best express uncertainty in a visualisation such that the viewer can acurately extract and interpret the risks involved.
+Another approach is to consider uncertainty visualsiation as a specific sub-field of high-dimensional data visualsiation, and therefore uncertainty is just another variable that needs to be expressed.
+
 ### Uncertainty as signal

### 2024-03-09 writingdumpster.qmd
updated the similar work section

@@
 
+# Citation Gaps
+Cite Gap 1: When did uncertainty visualisaiton become its own subfield of visualisation?
+
+Cite Gap 2: When was the first "uncertainty visualisation" used (likely a box plot of histogram)
+
 # Moved over notes
+- @Griethe2006 clearly sees uncertainty as additional information or data, and therefore sees the uncertainty visualisation problem as a specific case of the high dimensional visualisation issues. 
+- @Griethe2006 mentions that there are detailled classifications of uncertainty but the literature and subsequent techniques are not ballanced, and instead concentrate on scalar values.
+In facing this lack of ballance she creates a new classification system for uncertainty visualisation.
+- @Griethe2006 commented that uncertainty appears at every step in the analyis pipeline, and commented on some the mathematical definitions that seek to quantify uncertianty, but did not explain how these mathematical definitions can be combined across levels.
+Unfortunately this quantification is needed since she mentioned that in order to visualise uncertainty, it needs to be measurable.
+- @Griethe2006 defined uncertainty as "the degree to which the lack of knowledge about the amount of error is responsible for hesitancy in accepting results and observations without caution". 
+- @Griethe2006 found that there there are two ways to visualise uncertianty, implicitly and explicitly.
+When a visualisaion is implicit, the uncertainty is considered an uninteresting feature of the data, and instead of mapping some quantification of uncertainty to a visual feature, data is filtered out if it does not meet a specified "certainty" threshold.
+These statements clearly points towards a signal/noise dichotomy with normal visualisation and uncertainty visualisation.
+- @Spiegelhalter2017 even if most audience members can read a plot, they may not read it in the same way.
+People with high numeracy skills count pictographs, but people with low numeracy assess area, and seemingly unrelated elements may not equally influence all these methods of reading, for example scattering in a pictograph makes counting particularly more difficult.
+*also add that perceptual task audience thing*
+- @Spiegelhalter2017 it is better to understand your audience and iterate towards a final design, rather than only consider one option.
+- @Spiegelhalter2017 common sentiments such as avoiding chart junk, considering a table in place of a visualsiaiton, use of narrative labels, managing emotion, less is more approach, etc, are the main concepts expressed.
+- @Spiegelhalter2017 suggests that the graphic should allow "part-to-whole" comparison on an appropriate scale, which seems to be an extension of the framing concept discussed when speaking about uncertainty communication more broadly.
+- @Hullman2016 commented on the lack of justification of particular questions used in experiments and the lack of generalisability in using plot ranking as a response technique.
+- @Hullman2016 discussed the lack of clarity in exactly what "uncertainty" authors are trying to ask about.
+Whether the participants are supposed to answer according to some ground truth or their subjective probability is unclear, as is whether the *uncertainty* depicted represents the uncertainty of a sampling or population distribution.
+- @Hullman2016 discussed the role of heuristics in the analysis of results and suggested authors look for signs of them in responses and ask subjects to describe their strategy in their response. 
+- @Hullman2016 What was particularly interesting was people's tendancy to ignore uncertainty in a plot all together, and use a pattern in the signal (such as distance between means) to answer questions that require uncertainty information.
+- @Kinkeldey2014 stress the importance of selecting the correct audience, as numerical and visual literacy can differ vaslty between subpopulations so the results from one populations may not reflect the results from another.
 - @Daradkeh2015 found that people tended to to make decisions on a small number of salient pieces of information rather than on systematic evaluations of the costs and benefits of choices.
@@
 	- Participant literacy
-
-# Citation Gaps
-Cite Gap 1: When did uncertainty visualisaiton become its own subfield of visualisation?
-
-Cite Gap 2: When was the first "uncertainty visualisation" used (likely a box plot of histogram)
 

### 2024-03-09 paper.qmd
finished definitions in existing work section

@@
 
-> Edit note: just summarise common themes of issues the papers ran into instead of detailling the work paper by paper.
+# The Definition Problem
+## Definitions in Existing Work
+Uncertainty visualisation is made particularly difficult by the term "uncertainty" lacking a commonly accepted definition in the literature.
+This mishmash of terminology leads to a large body of work, all claiming to finding the best visualisation or expression of of "uncertainty" but most don't even seem to agree on what uncertainty is.
 
-## Types of Uncertainty Visualisation Papers
+Definitions presented in papers that discuss uncertainty visualisation want it to be encompassing and include everything a layperson might consider when trying to consider "uncertainty".
+Some consider it to be synonymous with specific terms defined in mathematics, such as probability, variance, error, or precision.
+Others consider it to be an encompassing umbrella term of which for these mathematically defined objects are only examples, and include concepts that are somewhat related, such as missing values (*Cite Gap 3*).
+This wide array of uncertainty definitions becomes a snake eating its own tail, as many authors opt to redefine uncertainty themselves[], ignore defining uncertainty entirely [], or pick a definition from a seemingly randomly previously published papers [], all of which make variance in uncertainty definitions even worse (*Cite Gap 3.5*).
 
-> Edit notes: Originally I was planning to have the discussion following this section relate to these "types" of papers, but that didn't end up happening.
-Depending on what I do with later sections, this might need a rewrite or to just be removed, since I'm not sure how relevant this is to the main ideas of the paper.
+While these vague definitions might be OK for polite conversation, they are *not* foundation on which to build an entire sub-field of visualisation.
+In order to visualise uncertainty, it needs to be quantifiable [@Griethe2006; Leland2005], in order to quantify uncertainty, it needs to be mathematically defined.
+The uncertainty visualisation literature is completely awash with papers that define uncertainty using a vague encompassing phrase, (e.g.
+@utypo defined it as any deviation from complete determinism) but go on to *quantify* uncertainty as a PDF, error, or some other easily quantified mathematical object when it comes time to do the visualisation (*Cite Gap 4*).
+This definition swap happens so subtly that it seems to go unnoticed by authors writing the papers, and the fact that "uncertainty visualisation" methods dont even notice they are exclusively focus on depictions of easily quantifiable uncertainty, such as error [], mass [], or variance [] (*Cite Gap 5*).
+This has left the expressions of uncertainty that are hard to quantify or visually differentiate untouched, despite many papers calling for their invention (*Cite Gap 6*).
+There is likely only confusion and obfuscation if the vague definitions of "uncertainty" continue to be used and authors should try to be more specific with the mathematical object they are visualising, *or* a strict definition of uncertainty needs to be defined.
+The current method of titling papers and defining uncertainty forces readers to comb through the method of a particular visualisation experiment to understand if the uncertainty visualised is amenable to their needs.
 
-Uncertainty visualisation papers can be organised according to the *goal* of the experiment.
-Evaluation experiments are the standard rule for visualisations because the human brain is not as reliable as mathematical calculation.
-Therefore, user studies often aim to assess the limitations, biases, and heuristics of our mental calculator so that we can better understand the problems we may encounter when we plot our data.
-This is not to say any paper that suggests a visualisation without an evaluation experiment is completely lacking in justification, and there are many papers that suggest a novel visualisation without an evaluation study.
-Sometimes these papers are a preliminary step in finding a solution for common problems and intend to evaluate the visualsiation in later work.
-Until that later work is done, it is often difficult to accept one particular representation.
-While justification could come in the form of discussing established concepts in visualisation, such as the hierarchy of perceptual tasks, even these may need an evaluation study to back up their claims.
-
-The reasoning for this is obvious.
-There are a large number heuristics and biases that are not obvious to us when designing visualisation.
-Additionally, these heuristics and baises can change depending on the larger scope of the graphic and the population we are communicating with.
-Additionally, since there is often a miriad of ways to visualise any particular mathematical object, to adopt specific visualisation that is not already common practice needs reasonable justification, lest we run into a range of unforseen heuristic pitfalls.
-
-The remainder of section 4 will focus on the results and methods of uncertainty visualisation evaluation experiments.
-There are six main types of evaluation experiments in the uncertainty visualisation literature, they are:
-- Association perceptual tasks: These experiments identify perceptual tasks that are/could be associated with uncertainty, and test its association with uncertainty, check it's number of distinguishable levels, and/or check how quickly/accurately people can draw the uncertainty information from the graphic.
-- Value perceptual tasks: Participants are shown some depiction of uncertainty and are asked to extract a value, make a comparison, or do something simple to show that the graphic is readable.  
-- Heuristic Checks: These experiments are checking if a perceptual bias that has been identified through public use of a particular plot, other evaluation experiments, or gut feeling identified through public use of a plot or in a different.
--  Plot Comparisons: These papers describe an uncertainty visualisation evaluation experiment, where they compare two or more uncertainty visualisaitons on a specific set of questions.
-These questions can be about specific statistics (e.g.
-What is the probability that A>B?) or about a secondary goal such as trust or decision making (e.g.
-How much do you trust this estimate?).
-- Qualitative analysis: These experiments show participants an uncertainty visualisation for an open ended or specific task, and asks them to describe how the visualisation was used to come to a conclusion, or they are asked to describe what they see in the visualisation.
-- Method papers: These papers discuss issues with the evaluation methods of uncertainty visualisaiton and suggest changes or improvements, typically with results indicating the benefit of the change.
-
-These experiments are not mutually exclusive within a paper, and a single study will often contain several experiments, usually from several of these categories.
-These distinctions are important because each have their own issues in how they relate to uncertainty more broadly.
-Below we will go through the five key issues in uncertainty visualisation, describe how they mainfest in the uncertainty visualisation literature, and offer some tentative solutions to these problems.
-
-We will also comment on some of the uncertainty visualisation papers that suggest a visualisation without an evaluation experiment.
-Some of these plots may offer a solution to the unique problems we mention in this paper and these visualisations will be discussed in "Part 6: Great Examples".
-We will also consider these papers to illustrate the issues that visualisations are aware of and working to fix, and discuss whether or not these directions of research are likely to bear fruit, and why.
-
-# The Uncertainty Definition Problem
-## Definitions in Existing Work
-Uncertainty visualisation is not made any easier by the fact that the term "uncertainty" lacks a commonly accepted definition in the literature.
-@Spiegelhalter2017 even commented that the definition of uncertainty essentially "depends on who you ask".
-The disarray of organisation and definitions within the uncertainty literature is a common sentiment expressed by reviewers[@Spiegelhalter2017; @Kinkeldey2014].
-This mishmash of terminology leads to a large body of work, all claiming to finding the best visualisation or expression of of "uncertainty" but most don't even seem to agree on what uncertainty is.
-In this section we will establish exactly what we mean when we talk about uncertainty.
-
-The most common definition of uncertainty used in papers that discuss visualising it are colloquial.
-Most definitions make some nod towards variance, precision, trust, error, missing values or another related concepts.
-Since *uncertainty*, unlike *probability*, does not have a strict definition that is established in mathematics, there is large variety in its usage.
-Some consider it to be synonymous with probabiltiy, some consider it to be an encompassing umbrella term of which probability is only an element, and some consider it to by something else entirely.
-This lack of a definition leads to uncertainty being swapped out freely with other vague words such as ambiguity and risk.
-Definitions presented in papers that discuss uncertainty visualisation want it to be encompassing and include everything a layperson might consider when trying to consider "uncertainty".
-The issue with these broad definitions is is that in order to visualize something, it needs to be quantified, and in order to quantify something, we need to know exactly what it is we are counting.
-This discrepancy appears regularly in the uncertainty visualisation literature.
-Often a vague definition such as "uncertainty is anything that isn't deterministic", or even a refusal to define uncertainty at all in an experiments introduction will be accompanied by a highly specific quantification of uncertainty as a probability density distribution of a forecast when providing the visualisation.
-Ironically it seems that many of the papers that refer to themselves as "uncertainty" papers are just suffering from a form of uncertainty, a lack of precision.[^1]
-
-This lack of precision has caused several pernicious issues in the literature, including different representation of uncertainty being compared despite not being realistic substitutes (with no acknowledgement of this from the authors); a lack of understanding as to *why* uncertainty should be included in a visualisation; a large swath of literature focused on a minute aspect of the entire uncertainty space with little attention to creative solutions to real world visualisation problems; and a swath of newly designed visualisations that are disconnected from the problems that they hope to solve.
-These problems means the literature either needs to transition away from the broad umbrella term "uncertainty", or consider the full range of sources of uncertainty and how they can be communicated when designing an experiment.[^2]
-
-The most encompassing definition of uncertainty I have seen comes from @utypo who define uncertainty as **"any deviation from the unachievable ideal of completely deterministic knowledge of the relevant system"**.
-This definition encapsulates many concepts adjacent to randomness such as predictions, probability mass distributions (PMF), estimate error, and any data that is not a set of deterministic outcomes.
-This is the definition of uncertainty I will use for the rest of this report.
-
-[^1]: Flesh out note - I think several of the sentences in this paragraph need examples (or quatified evidence) from the spreadsheet definition part
-[^2]: I am not sure if I should mention this here because I go into depth in the literature review and these comments might come across as unjustified if they are left in this section.
-Maybe this is more of a conclusion paragraph.
+Some papers have recognised these colloquial definitions are lacking and have made attempts to formalise the space with taxonomies, however these taxonomies come with their own issues.
 
 ### Taxonomy Definitions
-> Edit Note: A lot of this is going to be moved to the uncertainty definition specific section of the vis lit review and the point here should just be "this is a mess" instead of trying to find any reason in it.
-My opinion and goal for this section has changed a lot since writing this (specifically deciding that the encapsulating taxonomy doesnt fix the problem and a maths definition is needed) so it is one of the sections that needs big rewrites.
-
-While colloquial definitions of uncertainty are useful when trying to communicate with laypeople who don't understand strict statistical definitions, these definitions are not helpful when we are trying to work out exactly what should be shown on a plot.
-When authors try to make a considerable effort to define what we mean when we talk about uncertainty, they often express it in the form of a taxonomy.
-
-When reading any literature about uncertainty or an adjacent topic you will be overrun with more ways to organise uncertainty than you will know what to do with.
-The problem is so pervasive that it is almost impossible for two randomly selected papers that both claim to be about uncertainty will use the same definition.
-What is worse is that the reason for using one taxonomy over another, or why a new taxonomy was established in the first place is rarely discussed.
-This leaves new authors in the field with a wide array of choices and no information that would allow them to determine which is most effective for them.
-The complete mess of this literature makes it more enticing for authors to just define their own definition of uncertainty, which only adds to this problem.
-This is not done in ignorance of the issue either, many papers that comment on the unending list of uncertainty definitions will still provide a new taxonomy of their own design.
-I would almost find this situation comical if I were not one of the people forced to sort through this mess.
-
-Examples of Taxonomies used to untanlge the features of uncertainty:
-
-- aleatory uncertainty (unavoidable randomness), epistemic uncertatiny (uncertainty about structure, still quantified), ontological uncertainty (uncertain about entire modelling process).
-@Spiegelhalter2017
-- measurement (uncertainty on a single varible), spatial (uncertainty in spatial data), temporal (uncertainty in a time data)
-- aleatory (inherrent randomness in a forecast) vs epistemic (uncertainty in structure and parameters of statistical models) vs ontological (uncertainty about the entire modelling process as a description of reality)
-- scientific methods (variance, distribution) vs human judgments (disagreement among parties)
-- Random vs systematic uncertainty
-- statistical (can provide a continuous pdf) vs bounded (know values will fall in a certain range) uncertainty
-- accuracy vs precision
-- Uncertainty from discovery, assessment and execution
-- risk vs uncertainty
-- "known knowns, known unknowns, and unknown unknowns", (risk, uncertainty, ignorance)
-- Source of uncertainty (model structure, parameters, forcing functions, intial state, model operation) and diagnostic (past/current) vs prognostic (forecasting)
-- uncertainty is organised according to Satience (legal moral, societal, institutional, proprietary, situational) and level of severity (high, medium, low)
-- Level of precision in epistemic uncertainty: 1) numbers given to appropriate levels of precision 2) a distribution range 3) a measure of statistical significance 4) verbal quantifiers to numbers 5) refusal to give a number unless the evidence is good enough @Spiegelhalter2017
-
-![Example of different uncertainty taxonomies](paperscreenshots/collage1.jpg){#fig-location}
-
-> Edit note: I dont think I will keep this figure in the paper. The goal was to show they are inconsistent but I think a sentance or two in the text will achieve that.
-
-This taxonomy issue creates several problems in the literature.
-Ideally, if we wanted to combine the literature on uncertainty visualisation, each paper would state the same definition of uncertainty and explain which aspect of uncertainty they are visualising, so the experiments would shift around the space while the definition of uncertainty and what it contains would remains stationary.
-This would make it easy for the typical reader to use the literature to find an appropriate visualisation for the particular category or case of uncertainty.
-Instead, one has to comb through the method of a particular design to understand if the uncertainty you are working with is amenable to this method.
-Another issue this creates is that the "space" of uncertainty visualisation is unclear.
-This means that several areas of uncertainty visualisation that have a real need for creative visualisation tools are left untouched, while the visualisations that are considered "uncertainty visualisations" simply because they are an expression of a probability mass function dominate the literature.
-Finally the taxonomies do little to help data analysts or statisticians untangle, estimate and understand the uncertainty that may be in their own projects.
-The endless set of constantly changing definitions implies their *is no* set meaning to the term "uncertainty" and something that cannot be defined certainly cannot be quantified for communication or estimation.
-This leads people to be less likely to visualise uncertainty due to its imprecision and difficulty to commincate.
-
-This does not mean there is nothing to be learned from these taxonomies, each in their own way show what apsect of uncertainty is important to certain cases.
-There are a handful of overlapping features that indicate which elements of uncertainty are of interest to statistics and related applied fields.
-@utypo identified these similarities and used them to design a taxonomy that encapsulates a large proportion of the ad hoc definitions that currently overrun the literature, and is the definition of uncertainty that we will work with for the remainder of this paper.
-
-@fig-taxonomy is an illustration of the taxonomy presented by @utypo.
-In this taxonomy, there are three things we need to consider for each "uncertainty" we encounter through the modelling process.
-First, we need to consider the source of the uncertainty.
-Is this uncertainty coming from the data (for example, from an inaccurate measurement), from the array of model choices we have, or from the assumptions on a parameter?
-This is the *location* of the uncertainty.
-Second, consider how well you can quantify this uncertainty.
-Do you know exactly how much measurement error there is in each observation or are you not even aware if there is a measurement error?
-This is the *level* of your uncertainty, and it ranges from discrete to total ignorance.
-Finally, consider how this uncertainty came into existence.
-Is it a result of a naturally random process (aleatory) or is it due to imperfect information and could be improved (epistemic).
-This is the *nature* of your uncertainty.
-@utypo then goes on to describe mapping our uncertainty in a 3D space that is defined by its location, level, and nature, but I think the taxonomy is more easily understood as a series of questions we should to consider when we are trying to quantify uncertainty.
-Location asks "where and at what stage in the modelling process is this uncertainty coming from", level asks "how well can we quantify this uncertainty", and nature asks "can we reduce this uncertainty with improves knowledge?".
-While this taxonomy specifically applies to the uncertainty in the model, it is easy to fit most of the uncertainty we would consider quantifying into this format.
+Much in the same way that there are almost as many definitions of uncertainty as there are papers on the topic, the field is also over run with taxonomies.
+Taxonomies split uncertainty based on an endless stream of ever changing boundaries, such as whether the uncertainty is due to true randomness or a lack of knowledge [@Spiegelhalter2017], if the uncertainty is in the the attribute, spatial elements, or temporal element of the data [], whether the uncertainty is scientific (e.g.
+error) or human (e.g.
+disagreement among parties) [], if the uncertainty is random or systematic [], statistical or bounded [], accuracy or precision [], if the uncertainty is about the past or the future [], the stage of the data analysis pipeline it comes from [], the severity of the uncertainty [], how quantifiable the uncertainty is [@Spiegelhalter2017], etc., etc., etc.
+(*Cite Gap 7*).
+In their own way, each of these taxonomies show an apsect of uncertainty that an author felt was important to differentiatet.
+@utypo identified many common themes among this wide array of taxonomies and used them to design an encapsulating taxonomy, depicted in @fig-taxonomy.
+This definition organises uncertainty based on three axis, location (where the uncertianty is coming from), level (how quantifiable it is), and nature (if it comes from true randomness or incomplete knowledge).
 
@@
 
-> Edit note: Everything from this point down was used to try and provide some justification for why this dimensionality concept could work, but I have since changed my mind and am probably going to get rid of this section unless it fits somewhere else.
+When considering the use of taxonomies to define something inherrently vague, such as the uncertainty in an uncertainty visualisation, it is helpful to consider this quote from *The Grammar of Graphics*:
 
-*The taxonomy* we will use in this paper may seem complicated and difficult to grasp at first glance, but it essentially boils down to three dimensions to consider.
-Uncertainty is a broad and complicated topic and while oversimplified taxonomies might help us hold the concept of uncertainty in our head, they do little to help with the practical question of how to quantify, visualise, or communicate uncertainty.
-Organising uncertainty according to its location, level and nature makes allows us to consider uncertainty in the miriad of ways it can arise in a data analysis project and better understand *how* this uncertainty should be communicated.
-With a detailed understanding of the sources of uncertainty and the ways it can be quantified, we can better understand the gaps in the current set of visualisation tools and assess if these gaps need to be filled.
+> "Taxonomies are useful to scientists when they lead to new theory or stimulate insights into a problem that previous theorizing might conceal.
+Classification for its own sake, however, is as unproductive in design as it is in science.
+[@Leland2005]
 
-*The location* axis of the taxonomy lines up neatly with the typical data analysis pipeline.
-Multiple authors in the uncertainty literature have commented on the need to consider quantifying and expressing uncertainty at every stage of a project [@Kinkeldey2014; @Hullman2016; @Refsgaard2007].
-@fig-location shows the connection between the typical data analysis pipeline and the location element of the uncertainty taxonomy, along with examples of how uncertainty could arise from each of these steps.
-The relevance of every stage of the uncertainty taxonomy is backed up by both theoretical and practical work in statistics and adjacent fields.
-@Munzner2009, in her nested model for visualisation, comments on how even something as low level as poor problem abstraction can cause down stream effects and become impossible to ignore.
-@Otsuka2023 highlights that the choices made in the ontological process of boiling real world events down into statistical objects can influence later stages of analysis due to what was considred signal and what was considered noise.
-@Meng2021 stresses the importance of treating data as an input from another process with noise and variance, rather than as a neutral input.
-High level stages such as model selection and estimation have obvious introductions of uncertainty through the choice of models and nature of estimation [^I'm not sure if I need a citation here because this feels obvious].
-Location is also acknowledged as an important element of consideration in the taxonomies of several visualsiation papers [^ I will probably need a citation for which ones].
+Therefore, while taxonomies can be helpful to map out the space of "things that we consider to be related to uncertainty", a taxonomy is *not* a definition and do little to help statisticians untangle and estimate the uncertainty in their projects.
+That being said, these taxonomies can hint towards what is important when we think about uncertainty.
+The location axis of the @utypo taxonomy lines up neatly with the typical data analysis pipeline and multiple authors in the uncertainty literature have commented on the need to consider quantifying and expressing uncertainty at every stage of a project [@Kinkeldey2014; @Hullman2016; @Refsgaard2007] including stages as early as conceptualising the problem [@Otsuka2023] or collecting the data [@Meng2021].
+The level axis from the @utypo taxonomy also hints towards an important consideration in uncertainty, because establishing how quantifiable uncertainty is informs us how it can be communicated, something that @Spiegelhalter2017 identified in the form of "precision".
+It is commonly noted that each source of uncertainty in an analysis must be discussed in isolation, but combining the uncertainty from every stage is near impossible [@Spiegelhalter2017] (*Cite Gap 8*).
+It is clear that elements of these taxonomies are identified in many other comments on uncertainty and represent *real* considerations that need to be made when defining uncertainty, even if that definition is yet to be published.
 
-![Illustration of the connection between the data generating process and the location of the uncertainty analysis.
-Several examples are collected from utypo and Munzner2009.
-](location.png){#fig-location}
-
-*The level* axis forces us to identify how quantifiable the uncertainty is.
-This element is important because establishing how quantifiable uncertainty is informs us how it can be communicated, something that @Spiegelhalter2017 also identified in the form of "precision".
-Statistical uncertainty can be quantified, so it can be expressed using a PDF, a variance, an error estimate, or something similar while outcome uncertainty cant be quantified but each scenario may be simulated.
-Acknowledged ignorance can only be stated in the form of an assumption.
-Visualisations of climate change scenario uncertainty typically combine the outcome uncertainty of carbon emissions with the and statistical uncertainty of the noisy measurement and modelling system to create a cohesive illustration of the uncertainty surrounding climate change.
-It is commonly noted that each uncertainty must be discussed in isolation, but combining the uncertainty from every stage is near impossible [@Spiegelhalter2017].
-The level axis and climate change visualisations provide tools for how to consider combining these uncertainties.
-Uncertainties that are the same level but from multiple locations can likely be combined to establish an overall model uncertainty.
-While uncertainties at different levels cannot be combined, they can still be visualised simultaneously as shown in the climate scenario uncertainty visualisations.
-
-*The nature axis* allows us to easily communicate if the uncertainty can be reduced or if it needs to be accepted.
-While this level if not of particular importance to the visualisation of uncertainty, that is not the only purpose of the taxonomy.
-The nature of the uncertainty is an important element in expressing uncertainty, especially to stakeholders, and connecting the nature to the location makes it easier to understand *how* uncertainty can be reduced.
-
-Mapping uncertainty into the 3D space defined by location, level, and nature can expand beyond the narrow case of statistical modelling it was defined for.
-*(find the citation for this)* included "consensus" as a taxonomy element, however it was not included in the final definition, but it can easily be added as a location that would align with the communication stage of the data pipeline.
-Several authors [@Spiegelhalter2017] *(I have other papers)* considered the method by which uncertainty was quantified (e.g.
-by a single value, a function, simulated outcomes, etc), which can be be thought of as a more precise categorisation of the level axis.
-This taxonomy also ignores contextual information about the data, such as spatial or temporal considerations.
-This flexibility means that the use of this framework is not limited by a specific case and should be used going forward to define what aspect of the uncertainty space a contribution exists in.
-
-It is important to note that this taxonomy is thorough but it is not a mathematical definition of uncertainty.
-While individual *expressions of uncertainty*, such as a confidence interval or a PDF are mathematically defined, the term *uncertainty* has not been defined mathematically.
-This is largely because it is a layperson word that has been adopted by statistics to describe any deviation from complete determinism, however this does not mean that a mathematical definition that encompases missing values, alternative models, and prediction intervals *cannot* be found.
-However, that task is far beyond the scope of this work.
-In this work, we will investigate the array of misunderstandings and focus on inspecting how the current literature *deviates* from already established mathematical definitions, why these deviations occur, and understand how these deviations can lead to larger problems in uncertainty communication.
+The task of providing an encapsulating mathematical definition of uncertainty is far beyond the scope of this work, however we will discuss an important but commonly misunderstood feature of uncertainty; its relationship to inference.
 
@@
 
-# The Uncertainty Visualisation Definition Problem
+# The Visualisation Problem
 
@@
 
-## "Uncertainty Visualisation" is not defined
+### Types of Uncertainty Visualisation Papers
+
+> Edit notes: Originally I was planning to have the discussion following this section relate to these "types" of papers, but that didn't end up happening.
+Depending on what I do with later sections, this might need a rewrite or to just be removed, since I'm not sure how relevant this is to the main ideas of the paper.
+
+Uncertainty visualisation papers can be organised according to the *goal* of the experiment.
+Evaluation experiments are the standard rule for visualisations because the human brain is not as reliable as mathematical calculation.
+Therefore, user studies often aim to assess the limitations, biases, and heuristics of our mental calculator so that we can better understand the problems we may encounter when we plot our data.
+This is not to say any paper that suggests a visualisation without an evaluation experiment is completely lacking in justification, and there are many papers that suggest a novel visualisation without an evaluation study.
+Sometimes these papers are a preliminary step in finding a solution for common problems and intend to evaluate the visualsiation in later work.
+Until that later work is done, it is often difficult to accept one particular representation.
+While justification could come in the form of discussing established concepts in visualisation, such as the hierarchy of perceptual tasks, even these may need an evaluation study to back up their claims.
+
+The reasoning for this is obvious.
+There are a large number heuristics and biases that are not obvious to us when designing visualisation.
+Additionally, these heuristics and baises can change depending on the larger scope of the graphic and the population we are communicating with.
+Additionally, since there is often a miriad of ways to visualise any particular mathematical object, to adopt specific visualisation that is not already common practice needs reasonable justification, lest we run into a range of unforseen heuristic pitfalls.
+
+The remainder of section 4 will focus on the results and methods of uncertainty visualisation evaluation experiments.
+There are six main types of evaluation experiments in the uncertainty visualisation literature, they are:
+- Association perceptual tasks: These experiments identify perceptual tasks that are/could be associated with uncertainty, and test its association with uncertainty, check it's number of distinguishable levels, and/or check how quickly/accurately people can draw the uncertainty information from the graphic.
+- Value perceptual tasks: Participants are shown some depiction of uncertainty and are asked to extract a value, make a comparison, or do something simple to show that the graphic is readable.  
+- Heuristic Checks: These experiments are checking if a perceptual bias that has been identified through public use of a particular plot, other evaluation experiments, or gut feeling identified through public use of a plot or in a different.
+-  Plot Comparisons: These papers describe an uncertainty visualisation evaluation experiment, where they compare two or more uncertainty visualisaitons on a specific set of questions.
+These questions can be about specific statistics (e.g.
+What is the probability that A>B?) or about a secondary goal such as trust or decision making (e.g.
+How much do you trust this estimate?).
+- Qualitative analysis: These experiments show participants an uncertainty visualisation for an open ended or specific task, and asks them to describe how the visualisation was used to come to a conclusion, or they are asked to describe what they see in the visualisation.
+- Method papers: These papers discuss issues with the evaluation methods of uncertainty visualisaiton and suggest changes or improvements, typically with results indicating the benefit of the change.
+
+These experiments are not mutually exclusive within a paper, and a single study will often contain several experiments, usually from several of these categories.
+These distinctions are important because each have their own issues in how they relate to uncertainty more broadly.
+Below we will go through the five key issues in uncertainty visualisation, describe how they mainfest in the uncertainty visualisation literature, and offer some tentative solutions to these problems.
+
+We will also comment on some of the uncertainty visualisation papers that suggest a visualisation without an evaluation experiment.
+Some of these plots may offer a solution to the unique problems we mention in this paper and these visualisations will be discussed in "Part 6: Great Examples".
+We will also consider these papers to illustrate the issues that visualisations are aware of and working to fix, and discuss whether or not these directions of research are likely to bear fruit, and why.
+
+## Uncertainty Visualisation is not defined
 
@@
 
-# Measureing noise, not signal
+# The Measureing Noise Problem
 

### 2024-03-09 writingdumpster.qmd
finished definitions in existing work section

@@
 
+Cite Gap 3: Some consider it to be synonymous with specific terms defined in mathematics, such as probability, variance, error, or precision.
+Others consider it to be an encompassing umbrella term of which for these mathematically defined objects are only examples, and include concepts that are somewhat related, such as missing values
+
+Cite Gap 3.5: This wide array of uncertainty definitions becomes a snake eating its own tail, as many authors opt to redefine uncertainty themselves[], ignore defining uncertainty entirely [], or pick a definition from a seemingly randomly previously published papers [], all of which make variance in uncertainty definitions even worse
+
+Cite Gap 4:The uncertainty visualisation literature is completely awash with papers that define uncertainty using a vague encompassing phrase but go on to *quantify* uncertainty as a PDF, error, or some other easily quantified mathematical object when it comes time to do the visualisation.
+
+Cite Gap 5: This definition swap happens so subtly that it seems to go unnoticed by authors writing the papers, and the fact that "uncertainty visualisation" methods dont even notice they are exclusively focus on depictions of easily quantifiable uncertainty, such as error [], mass [], or variance []
+
+Cite Gap 6: This has left the expressions of uncertainty that are hard to quantify or visually differentiate untouched, despite many papers calling for their invention
+
+Cite Gap 7: Taxonomies split uncertainty based on an endless stream of arbitrary lines in the sand, such as whether the uncertainty is due to true randomness or a lack of knowledge [@Spiegelhalter2017], if the uncertainty is in the the attribute, spatial elements, or temporal element of the data [], whether the uncertainty is scientific (e.g.
+error) or human (e.g.
+disagreement among parties) [], if the uncertainty is random or systematic [], statistical or bounded [], accuracy or precision [], if the uncertainty is about the past or the future [], the stage of the data analysis pipeline it comes from [], the severity of the uncertainty [], how quantifiable the uncertainty is [@Spiegelhalter2017], and many, many, many more examples.
+
+Cite Gap 8: It is commonly noted that each uncertainty must be discussed in isolation, but combining the uncertainty from every stage is near impossible [@Spiegelhalter2017]
+
 # Moved over notes
+- @Munzner2009, in her nested model for visualisation, comments on how even something as low level as poor problem abstraction can cause down stream effects and become impossible to ignore. 
+- @Otsuka2023 highlights that the choices made in the ontological process of boiling real world events down into statistical objects can influence later stages of analysis due to what was considred signal and what was considered noise.
+- @Meng2021 stresses the importance of treating data as an input from another process with noise and variance, rather than as a neutral input.
 - @Griethe2006 clearly sees uncertainty as additional information or data, and therefore sees the uncertainty visualisation problem as a specific case of the high dimensional visualisation issues. 

### 2024-03-09 paper.qmd
several minor edits

@@
 # The Definition Problem
-## Definitions in Existing Work
+## Colloquial Definitions
 Uncertainty visualisation is made particularly difficult by the term "uncertainty" lacking a commonly accepted definition in the literature.
@@
 
-### Taxonomy Definitions
+## Taxonomy Definitions
 Much in the same way that there are almost as many definitions of uncertainty as there are papers on the topic, the field is also over run with taxonomies.
@@
 
-## The role of inference
+## Inference and Uncertainty
 When we consider mapping uncertainty, a reasonable questions that should come to mind is "uncertainty about what?".
@@
 
-# The Visualisation Problem
+# The Uncertainty Visualisation Problem
 
@@
 
-## Uncertainty Visualisations in Existing work
-
+## Definitions of "Uncertainty Visualisation" 
 ### Taxonomies of Uncertainty Visualisations

### 2024-03-11 paper.qmd
added intro to vis bit and some changes to inference bit

@@
 This review believe all these sources of noise have a single root cause, a lack of a cohesive and encompassing definition of uncertainty.
-The lack of an econompasing definition is touched on by every uncertainty visualisation directly [@Spiegelhalter2017; @Griethe2006] or indirectly by describing a miriad of ways it can be considered in the literature [@Kinkeldey2014; @Hullman2016], although it is never commented on as a source of the noise in the field.
+The absence of an econompasing definition of uncertainty is mentioned by every uncertainty visualisation directly [@Spiegelhalter2017; @Griethe2006] or indirectly by describing a miriad of ways it can be considered in the literature [@Kinkeldey2014; @Hullman2016], although it is never commented on as a source of the noise in the field.
 
@@
 In understanding this purpose, we will also consider whether or not "uncertainty visualisation" is a valid sub-field of visualisation itself, or if that distinction has only arisen as a by product of the confused definition of uncertainty.
-Ultimately, this paper should serve as a guide for future uncertaitny visualisation authors to prevent their work from becoming noise in what is already an incredibly noisy field.
+Ultimately, this paper should serve as a guide for future uncertainty visualisation authors to prevent their work from becoming noise in what is already an incredibly noisy field.
 
-# The Definition Problem
+# What is Uncertainty?
 ## Colloquial Definitions
@@
 (*Cite Gap 7*).
-In their own way, each of these taxonomies show an apsect of uncertainty that an author felt was important to differentiatet.
+In their own way, each of these taxonomies show an apsect of uncertainty that an author felt was important to differentiate.
 @utypo identified many common themes among this wide array of taxonomies and used them to design an encapsulating taxonomy, depicted in @fig-taxonomy.
@@
 
-## Inference and Uncertainty
-When we consider mapping uncertainty, a reasonable questions that should come to mind is "uncertainty about what?".
-Every single thing concept or activity in statistics does not have uncertainty.
-For example, descriptive statistics cannot have uncertainty as they are just describing statistics as they are, the sample mean is the sample mean, we are not uncertain about its value.
-A sample variance describes dispersion, which is also how we describe uncertainty, but the sample variance in of itself does not have any uncertainty.
+## Defining uncertainty with respect to inference
+What exactly is uncertainty, then?
+If we were to consider making this overarching definition, what would it need in order to be "encapsulating"?
+Well, let us consider some comments about uncertainty or common sentiments that come up time and time again in the literature.
+
+We know uncertainty cannot be entirely explained by individual concepts that *have* been defined, such as probability, variance, error, missing values, precision, or related concepts that pop up throughout the data analysis pipeline.
+We also know it likely has some relationship with the *task* we are considering, given that many authors have attributed the noise in uncertainty visualisation research as a type of "task" dependence.
+Additionally, it should be clear that we wouldn't consider uncertainty to be an appropriate term for *everything* in statistics would we?
+
+@Otsuka2023 spends the first chapter of his book discussing the place of descriptive statistics in the philosophy of the field.
+For example, do descriptive statistics have uncertainty?
+Descriptive statistics simply describe features of your sample as they are.
+The sample mean is the sample mean, the sample size is the sample size, and the sample variance is its sample variance.
+We would not say we are "uncertain" about these values unless we are questioning our own computation skills.
+
 Most of the mathematical uncertainty concepts have this idea embedded in them.
@@
 Once it is established what we are uncertainty about, we can consider the other elements of uncertainty that need to be defined.
+
+Variance and probability are typically considered the closest of these concepts to uncertainty, however they clearly do not fully encapsulate the concept.
+@Spiegelhalter2017 makes the reason for this distinction clear, as in the absence of something to draw inference on, variance and probability describe precise random events (such as a coin flip), while uncertainty is imprecision (such as the estimated probability associated with a coin that might be biased).
+A sample variance describes dispersion, which is also how we describe uncertainty, but the sample variance in of itself does not have any uncertainty.
 
@@
 
-# The Uncertainty Visualisation Problem
+# What is an Uncertainty Visualisation?
+The lack of a consistent definition of uncertainty means that every statement that builds upon "uncertainty" will be ill-formed.
+This includes statements about uncertainty visualisation.
+This appears in the literature as a mountain of seemingly conflicting statements.
+For example @Leland2005 mentions that popular graphics, such as pie charts and bar charts omit uncertainty, however at least one or both of these charts are used in most "uncertainty visualisation" experiments [] (*Cite Gap __*).
+
+Attempts to rectify these conflicts always flutter about the surface and do not seem to grasp the core of this problem.
+For example, @wu2023rational noticed a series of papers with poorly defined tasks and suggested comparing study participants to a "rational agent" to help authors understand and remedy this issue.
+@Kim2019 believed an understanding of participants prior beliefs could assist in interpreting uncertainty visualisation results that tend to be incredibly noisy.
+What these suggestions fail to understand is that the entire practice of identifying "true" facts about "unertainty visualisation" is futile.
+It is impossible to make a statement about the properties of an "uncertainty visualisation" in the same way it is impossible to comment on $\frac 0 0 = 1$.
+Division by zero is undefined, so the statement is ill formed, similarly, "uncertainty" is not defined, so statements *about* "uncertainty" including statements about "uncertainty visualisation" are ill formed.
+Ill-formed statements can neither be true nor false and cannot be part of a logical argument, so experiments and research that build upon this ill-formed concept are neither true nor false, they simply exist.
+Someone who is less familiar with the rules of logical argument may beleive this understanding of the problem to an insufferable nitpick that creates impossible to live up to standards, however the consequences of this definitional issue are *very* real and costly.
+The studies that investigate "uncertainty visualisation" may be identifying *something*, however there is good reason to believe that *something* is neither interesting nor does can it contribute to a larger theory of visualisation.
+This is something we illustrate in the following this section.
+
+## Common misconceptions about uncertainty visualisation
+
+### Defining "uncertainty visualisation" as one word
 
@@
 
-## Definitions of "Uncertainty Visualisation" 
-### Taxonomies of Uncertainty Visualisations
 There are some taxonomies of visual uncertainty.
@@
 
-The first issue in uncertainty visualisation is there does not seem to be a clear rule for what makes a particular visualisation a depiction of uncertainty.
-This problem seems to start at the lack of the mathematical definition of uncertainty which then bleeds into a lack of a definition for an "uncertainty visualisation".
-For example @Leland2005 mentions that popular graphics, such as pie charts and bar charts omit uncertainty, however at least one or both of these charts are used in a large number of uncertainty visualisation experiments.
-How can a chart omit uncertainty according to the grammar of graphics, but be used to describe it in several evaluation studies?
-The answer to this question is related to the definition problem discussed earlier, but to understand how, we will need to go through the grammar of graphics, and its absence in the uncertainty visualisation literature.
-
-### Contextual Information
+### Misunderstanding the role of contextual information
 > Edit Note: This section should probably be moved to the visualisation specific section, although that depends on the tone that I end up with after the rewrites of the earlier sections.
@@
 ```
-
-### Types of Uncertainty Visualisation Papers
-
-> Edit notes: Originally I was planning to have the discussion following this section relate to these "types" of papers, but that didn't end up happening.
-Depending on what I do with later sections, this might need a rewrite or to just be removed, since I'm not sure how relevant this is to the main ideas of the paper.
-
-Uncertainty visualisation papers can be organised according to the *goal* of the experiment.
-Evaluation experiments are the standard rule for visualisations because the human brain is not as reliable as mathematical calculation.
-Therefore, user studies often aim to assess the limitations, biases, and heuristics of our mental calculator so that we can better understand the problems we may encounter when we plot our data.
-This is not to say any paper that suggests a visualisation without an evaluation experiment is completely lacking in justification, and there are many papers that suggest a novel visualisation without an evaluation study.
-Sometimes these papers are a preliminary step in finding a solution for common problems and intend to evaluate the visualsiation in later work.
-Until that later work is done, it is often difficult to accept one particular representation.
-While justification could come in the form of discussing established concepts in visualisation, such as the hierarchy of perceptual tasks, even these may need an evaluation study to back up their claims.
-
-The reasoning for this is obvious.
-There are a large number heuristics and biases that are not obvious to us when designing visualisation.
-Additionally, these heuristics and baises can change depending on the larger scope of the graphic and the population we are communicating with.
-Additionally, since there is often a miriad of ways to visualise any particular mathematical object, to adopt specific visualisation that is not already common practice needs reasonable justification, lest we run into a range of unforseen heuristic pitfalls.
-
-The remainder of section 4 will focus on the results and methods of uncertainty visualisation evaluation experiments.
-There are six main types of evaluation experiments in the uncertainty visualisation literature, they are:
-- Association perceptual tasks: These experiments identify perceptual tasks that are/could be associated with uncertainty, and test its association with uncertainty, check it's number of distinguishable levels, and/or check how quickly/accurately people can draw the uncertainty information from the graphic.
-- Value perceptual tasks: Participants are shown some depiction of uncertainty and are asked to extract a value, make a comparison, or do something simple to show that the graphic is readable.  
-- Heuristic Checks: These experiments are checking if a perceptual bias that has been identified through public use of a particular plot, other evaluation experiments, or gut feeling identified through public use of a plot or in a different.
--  Plot Comparisons: These papers describe an uncertainty visualisation evaluation experiment, where they compare two or more uncertainty visualisaitons on a specific set of questions.
-These questions can be about specific statistics (e.g.
-What is the probability that A>B?) or about a secondary goal such as trust or decision making (e.g.
-How much do you trust this estimate?).
-- Qualitative analysis: These experiments show participants an uncertainty visualisation for an open ended or specific task, and asks them to describe how the visualisation was used to come to a conclusion, or they are asked to describe what they see in the visualisation.
-- Method papers: These papers discuss issues with the evaluation methods of uncertainty visualisaiton and suggest changes or improvements, typically with results indicating the benefit of the change.
-
-These experiments are not mutually exclusive within a paper, and a single study will often contain several experiments, usually from several of these categories.
-These distinctions are important because each have their own issues in how they relate to uncertainty more broadly.
-Below we will go through the five key issues in uncertainty visualisation, describe how they mainfest in the uncertainty visualisation literature, and offer some tentative solutions to these problems.
-
-We will also comment on some of the uncertainty visualisation papers that suggest a visualisation without an evaluation experiment.
-Some of these plots may offer a solution to the unique problems we mention in this paper and these visualisations will be discussed in "Part 6: Great Examples".
-We will also consider these papers to illustrate the issues that visualisations are aware of and working to fix, and discuss whether or not these directions of research are likely to bear fruit, and why.
 
@@
 
-# The Measureing Noise Problem
-
-## Uncertainty vs Risk
+# Noise and Signal
+## Uncertainty versus Risk
 Uncertainty is noise.
@@
 The conversation around uncertainty communication is often centered around the psychology is communicating risks.
-
-Most research into general uncertainty communication seems to focus specifically on how we communicate risks.
-Risk and uncertainty are not perfect synonyms of one another.
-@Spiegelhalter2017 makes the distinction clear by specifying *risk* as *a precise random event* (such as a coin flip) while *uncertainty* (also sometimes referred to as abiguity) is imprecision (such as the estimated probability associated with a coin that might be biased).
-For this reason, risk is often seen as a specific case of uncertainty and that well defined sub area has a lot of detailled research.
-Some markers of probability that have common uses, probabilities such as 0 (impossible), 0.5 (a fair coin flip), 1 (certain) are easy for people to have an internal sense of *hullman paper*.
-It is incredibly hard to communicate small risks, because people cannot differentiate between 1/100 an 1/1000, and communicating low probability but high impact risks are particularly challenging [@Spiegelhalter2017].
-This issue can be somewhat alleviated by communicating the likelihood of an event in relation to another event on a similar scale (such as being struck by lightning) [@Spiegelhalter2017].
-The framing of how we communicate risks matters due to the affect heuristic, so there will be a difference in peoples behaviours if, for example, you tell people they have a 90% chance of survival vs a 10% chance of death, and when expressing a frequency, a larger numerator communicates a larger risk [@Spiegelhalter2017].
-The reference class of a particular risk should be explicitly stated to avoid confusion, for example if we are communicating the chance of rain, the time period over which we are expecting this chance of rain (i.e.
-50% chance today or this week) needs to be specified [@Spiegelhalter2017].
-The problem with a risk focus is that it is only a small piece of the uncertainty pie, and very few of these findings can be reasonably assumed to go beyond risk communication.
-Even less can be assumed to translate to uncertainty visualisation.
 
@@
 
+## Suggestions for measuring uncertainty
+### Types of Uncertainty Visualisation Papers
+
+> Edit notes: Originally I was planning to have the discussion following this section relate to these "types" of papers, but that didn't end up happening.
+Depending on what I do with later sections, this might need a rewrite or to just be removed, since I'm not sure how relevant this is to the main ideas of the paper.
+
+Uncertainty visualisation papers can be organised according to the *goal* of the experiment.
+Evaluation experiments are the standard rule for visualisations because the human brain is not as reliable as mathematical calculation.
+Therefore, user studies often aim to assess the limitations, biases, and heuristics of our mental calculator so that we can better understand the problems we may encounter when we plot our data.
+This is not to say any paper that suggests a visualisation without an evaluation experiment is completely lacking in justification, and there are many papers that suggest a novel visualisation without an evaluation study.
+Sometimes these papers are a preliminary step in finding a solution for common problems and intend to evaluate the visualsiation in later work.
+Until that later work is done, it is often difficult to accept one particular representation.
+While justification could come in the form of discussing established concepts in visualisation, such as the hierarchy of perceptual tasks, even these may need an evaluation study to back up their claims.
+
+The reasoning for this is obvious.
+There are a large number heuristics and biases that are not obvious to us when designing visualisation.
+Additionally, these heuristics and baises can change depending on the larger scope of the graphic and the population we are communicating with.
+Additionally, since there is often a miriad of ways to visualise any particular mathematical object, to adopt specific visualisation that is not already common practice needs reasonable justification, lest we run into a range of unforseen heuristic pitfalls.
+
+The remainder of section 4 will focus on the results and methods of uncertainty visualisation evaluation experiments.
+There are six main types of evaluation experiments in the uncertainty visualisation literature, they are:
+- Association perceptual tasks: These experiments identify perceptual tasks that are/could be associated with uncertainty, and test its association with uncertainty, check it's number of distinguishable levels, and/or check how quickly/accurately people can draw the uncertainty information from the graphic.
+- Value perceptual tasks: Participants are shown some depiction of uncertainty and are asked to extract a value, make a comparison, or do something simple to show that the graphic is readable.  
+- Heuristic Checks: These experiments are checking if a perceptual bias that has been identified through public use of a particular plot, other evaluation experiments, or gut feeling identified through public use of a plot or in a different.
+-  Plot Comparisons: These papers describe an uncertainty visualisation evaluation experiment, where they compare two or more uncertainty visualisaitons on a specific set of questions.
+These questions can be about specific statistics (e.g.
+What is the probability that A>B?) or about a secondary goal such as trust or decision making (e.g.
+How much do you trust this estimate?).
+- Qualitative analysis: These experiments show participants an uncertainty visualisation for an open ended or specific task, and asks them to describe how the visualisation was used to come to a conclusion, or they are asked to describe what they see in the visualisation.
+- Method papers: These papers discuss issues with the evaluation methods of uncertainty visualisaiton and suggest changes or improvements, typically with results indicating the benefit of the change.
+
+These experiments are not mutually exclusive within a paper, and a single study will often contain several experiments, usually from several of these categories.
+These distinctions are important because each have their own issues in how they relate to uncertainty more broadly.
+Below we will go through the five key issues in uncertainty visualisation, describe how they mainfest in the uncertainty visualisation literature, and offer some tentative solutions to these problems.
+
+We will also comment on some of the uncertainty visualisation papers that suggest a visualisation without an evaluation experiment.
+Some of these plots may offer a solution to the unique problems we mention in this paper and these visualisations will be discussed in "Part 6: Great Examples".
+We will also consider these papers to illustrate the issues that visualisations are aware of and working to fix, and discuss whether or not these directions of research are likely to bear fruit, and why.
+
 # Great Examples

### 2024-03-11 writingdumpster.qmd
added intro to vis bit and some changes to inference bit

@@
 # Moved over notes
+- Some markers of probability that have common uses, probabilities such as 0 (impossible), 0.5 (a fair coin flip), 1 (certain) are easy for people to have an internal sense of *hullman paper*. 
+- It is incredibly hard to communicate small risks, because people cannot differentiate between 1/100 an 1/1000, and communicating low probability but high impact risks are particularly challenging [@Spiegelhalter2017].
+This issue can be somewhat alleviated by communicating the likelihood of an event in relation to another event on a similar scale (such as being struck by lightning) [@Spiegelhalter2017].
+- The framing of how we communicate risks matters due to the affect heuristic, so there will be a difference in peoples behaviours if, for example, you tell people they have a 90% chance of survival vs a 10% chance of death, and when expressing a frequency, a larger numerator communicates a larger risk [@Spiegelhalter2017].
+- The reference class of a particular risk should be explicitly stated to avoid confusion, for example if we are communicating the chance of rain, the time period over which we are expecting this chance of rain (i.e.
+50% chance today or this week) needs to be specified [@Spiegelhalter2017].
 - @Munzner2009, in her nested model for visualisation, comments on how even something as low level as poor problem abstraction can cause down stream effects and become impossible to ignore. 
@@
 # Main section outtakes
+- What exactly is uncertainty, then?
+Let us consider the concepts related to uncertainty that have been discussed so far.
+We know uncertainty cannot be entirely explained by individual concepts that *have* been defined, such as probability, variance, error, missing values, precision, or related concepts that pop up throughout the data analysis pipeline.
+We also know it likely has some relationship with the *task* we are considering, given that many authors have attributed the noise in uncertainty visualisation research as a type of "task" dependence.
+Additionally, it should be clear that we wouldn't consider uncertainty to be an appropriate term for *everything* in statistics.
+For example, I wouldn't describe the population parameters of a distribution I have simulated to be "uncertain"
+
 - tasks e.g.

### 2024-03-11 paper.qmd
finished uncertainty is inference section

@@
 # What is Uncertainty?
-## Colloquial Definitions
+## Colloquial definitions
 Uncertainty visualisation is made particularly difficult by the term "uncertainty" lacking a commonly accepted definition in the literature.
@@
 
-## Taxonomy Definitions
+## Taxonomy definitions
 Much in the same way that there are almost as many definitions of uncertainty as there are papers on the topic, the field is also over run with taxonomies.
@@
 If we were to consider making this overarching definition, what would it need in order to be "encapsulating"?
-Well, let us consider some comments about uncertainty or common sentiments that come up time and time again in the literature.
+Well, let us consider what might *not* be considered uncertain in order to understand this concept a little better.
 
-We know uncertainty cannot be entirely explained by individual concepts that *have* been defined, such as probability, variance, error, missing values, precision, or related concepts that pop up throughout the data analysis pipeline.
-We also know it likely has some relationship with the *task* we are considering, given that many authors have attributed the noise in uncertainty visualisation research as a type of "task" dependence.
-Additionally, it should be clear that we wouldn't consider uncertainty to be an appropriate term for *everything* in statistics would we?
+@Otsuka2023 spends the first chapter of his book discussing the place of descriptive statistics in the philosophy of the field and in doing so, highlights an interesting connection between inference and uncertainty.
+Descriptive statistics simply describe our sample as it is and summarises large data down into an easy to swallow format.
+Descriptive statistics are not seen as the primary goal of modern statistics, however this was not always the case.
+Around the 19th century in England, *posivitism* was the popular philosophical approach to science (positivists included famous statisticians such as Francis Galton and Karl Pearson) and practitioners of the approach believed statistics ended with descriptive statistics as science must be based on actual experience and observations, therefore anything that refers to the unobservable (such as new observations or population statistics) is not true science [@Otsuka2023].
+By its very nature, descriptive statistics *cannot* be used to make inferences about the data because it simply exists to summaries the data, to use it to make statements about *new* data is incorrect useage.
+In order to make statements about population statistics, future values, or new observations we need to perform inference, which requires the assumption of the "uniformity of nature" (i.e.
+that unobserved phenomena should be similar to observed phenomena) [@Otsuka2023].
+This subtle shift, from descriptive statistics to inferential statistics was historically shunned *due to the fact it introduced the unknowable*, or in other words, uncertainty.
 
-@Otsuka2023 spends the first chapter of his book discussing the place of descriptive statistics in the philosophy of the field.
-For example, do descriptive statistics have uncertainty?
-Descriptive statistics simply describe features of your sample as they are.
-The sample mean is the sample mean, the sample size is the sample size, and the sample variance is its sample variance.
-We would not say we are "uncertain" about these values unless we are questioning our own computation skills.
+This philosophical understanding of statistics highlights that descriptive statistics do not have uncertainty, however some readers may disagreee with that statement.
+Variance and probability are typically considered stand ins for "uncertainty", it is often how we choose to measure it, and since probability and variance exist in descriptive statistics, descriptive statistics *must* have uncertainty.
+This is not necessarily true, and related distinction was made by @Spiegelhalter2017  commented on a differentiation between precise random events (such as the probability of a coin flip), and uncertainty (such as the estimated probability associated with a coin that might be biased).
+A sample variance is not unknown, and therefore it is not uncertain, rather it is a precise description of dispersion.
+If we were to discuss drawing a new observation, or estimating the true mean of a population *then* the the variance would become relevant in our discussions of uncertainty, but in isolation it is not uncertain.
 
-Most of the mathematical uncertainty concepts have this idea embedded in them.
-For example, a confidence interval is an interval drawn around the sample mean that will theoretically contain the *populaton mean* 95% of the time.
-This issue has been noticed several times, however it is often brought up as a *task* dependence rather than an understanding that uncertainty is inherrently linked to inference on a particular statistic.
+The idea that inference is related to uncertainty pops up in most discussions of uncertainty, however, due to the now *implicit* understanding we are performing inferential statistics, it is often brought up as a *task* or *goal* dependence.
+This is mentioned both by authors at a specific stage of an analysis, and by authors looking at an entire field or at all stages of the analysis pipeline.
+@Otsuka2023 suggested that the process of observing data to perform statistics is largely dependent on our goals, because the process of boiling real world entities down into probabilistic objects depends on the relationship we seek to identify with our data.
+@Meng2014 commented what is kept as data and what is tossed away is determined by the motivation of an analysis and what was previously noise can be shown to become signal depending on the the question we seek to answer.
+@Carlin2023 mentions that each research question can be can be categorised as descriptive, predictive, or causal, each of which has its own appropriate statistical methods and motivation agnostic model selection leads to statistical analysis that is devoid of meaning.
 @Wallsten1997 argue that the best method for evaluating or combining subjective probabilities depends on the uncertainty the decision maker wants to represent and why it matters.
@@
 Health have a wide range of metrics used to communicate different risks [@Spiegelhalter2017] indicating they utilise new statistics based off the specific aspect of risk that is relevant to the issue.
-This should make it clear that uncertainty only exists when we are performing inference.
-Therefore it only makes sense to describe uncertainty in terms of something to be uncertain *about*, it is not some a latent feature of statistics and data.
+These examples show that authors that discuss uncertainty believe it is there is an important relationship between uncertainty and what we decide to keep in our analysis and throw away, i.e.
+the task or goal.
+The task or goal is, in essence, the *statistic you are drawing inference on*.
+Were this not true, we would have no reason to differentiate between a prediction interval and a sampling distribution because they could both be considered "uncertainty about the mean", however they are different because their inferential goals are different.
 
-The idea that uncertainty can only be defined in the presence of a motivating question is well grounded in most areas of statistics.
-The entire process of data analysis, from deciding what should be observed as data through to communicating that data in a plot is governed by human decision and the goal of an analysis.
-At the philosophical level, applied statistics is simply taking real world entities and boiling them down into probabilistic objects, an ontological process that is largely dependent on our goals [@Otsuka2023].
-When we move onto data provenance the issue persists, as what is kept as data and what is tossed away is determined by the motivation of an analysis and what was previously noise can be shown to become signal depending on the resolution of the question [@Meng2014].
-After moving onto modelling this issue continues as each research question can be can be categorised as descriptive, predictive, or causal, each of which has its own appropriate statistical methods and motivation agnostic model selection leads to signal devoid of meaning [@Carlin2023].
-Even at the final stages of visualisation a lack of understanding of the motivating question make it difficult to untangle what is signal and what is noise, leaving many uncertainty visualisation studies with conflicting results [@Kinkeldey2014].
-These cases highlight that uncertainty is defined at *every* stage in relation to our motivating question, from data collection to visualisation.
-Discussions of uncertainty cannot be had if we are not clear *what* we are uncertain about.
-Once it is established what we are uncertainty about, we can consider the other elements of uncertainty that need to be defined.
+These examples highlight two key concepts that seem to be true about the relationship between uncertainty and inference:    
+1) Uncertainty is the by-product of inference, as we seek to draw conclusions about something unknowable.    
+2) Uncertainty must be defined with respect to a *specific* statistic we wish to draw inference on, there is no such thing as a *general* uncertainty that is not linked to a motivating question.
 
-Variance and probability are typically considered the closest of these concepts to uncertainty, however they clearly do not fully encapsulate the concept.
-@Spiegelhalter2017 makes the reason for this distinction clear, as in the absence of something to draw inference on, variance and probability describe precise random events (such as a coin flip), while uncertainty is imprecision (such as the estimated probability associated with a coin that might be biased).
-A sample variance describes dispersion, which is also how we describe uncertainty, but the sample variance in of itself does not have any uncertainty.
-
+## Mistakes when we misunderstand uncertainty
 We feel the need to make this point clear because a *large* amount of uncertainty research seems to be confused by the purpose of uncertainty visualisation, what it is related to, and why we visualise it.

### 2024-03-11 writingdumpster.qmd
finished uncertainty is inference section

@@
 # Main section outtakes
+- This is likely because The idea that uncertainty can only be defined in the presence of a motivating question is well grounded in most areas of statistics.
+The entire process of data analysis, from deciding what should be observed as data through to communicating that data in a plot is governed by human decision and the goal of an analysis.
+- We know uncertainty cannot be entirely explained by individual concepts that *have* been defined, such as probability, variance, error, missing values, precision, or related concepts that pop up throughout the data analysis pipeline.
+We also know it likely has some relationship with the *task* we are considering, given that many authors have attributed the noise in uncertainty visualisation research as a type of "task" dependence.
+Additionally, it should be clear that we wouldn't consider uncertainty to be an appropriate term for *everything* in statistics would we?
 - What exactly is uncertainty, then?

### 2024-03-13 paper.qmd
rendered for meeting
finshed rewrite of mistakes made section

@@
 ## Similar work
-::: aside
-I have a large backlog of uncertainty literature review summaries in my Mendely that need to be added here. There are a lot of them.
-:::
 
@@
 @Fischhoff2014 looks at uncertainty visualisation for decision making decides that we should have different ways of communicating uncertainty based off what the user is supposed to do with it.
-Health have a wide range of metrics used to communicate different risks [@Spiegelhalter2017] indicating they utilise new statistics based off the specific aspect of risk that is relevant to the issue.
 These examples show that authors that discuss uncertainty believe it is there is an important relationship between uncertainty and what we decide to keep in our analysis and throw away, i.e.
@@
 
-## Mistakes when we misunderstand uncertainty
-We feel the need to make this point clear because a *large* amount of uncertainty research seems to be confused by the purpose of uncertainty visualisation, what it is related to, and why we visualise it.
-Often, the authors are even more confused about what they are doing than the participants.
-There are enough papers that fail to understand these basic principals of uncertainty that this entire paper could simply be a list with explanations.
-Instead, we have a handful of highlights presented below.
-- @wu2023rational recognised it is difficult to understand when differences in performance stem from information asymmetry.
-- @Hofman2020 asked questions about discusses the difference in showing inferential or outcome (prediction) intervals and then found that participants shown the inferential distribution performed poorly when asked questions about predictions.
-They found that showing the relevant distribution to the question reduced the error.
-This result should be astoundingly obvious if it was well understood that uncertainty is not some latent variable and is only relevant for specific ground truth statistics.
-This papers *percieved* importants highlights an underlying belief that uncertainty is linked to a specific variable or data set, not the variable we are drawing inference on.
-- @Padilla2021 found that high uncertainty in the model estimates (calculated uncertainty) and low forecaster confidence (which is typically an expression of suspected bias in the model) both caused participants to have decreased confidence in their results and suggested modelers express both if they are relevant.
-@Kale2019 discussed the importance of communicating decisions made in the data analysis pipeline and being aware of the alternatives.
-While there is nothing wrong with the core ideas of these papers, the fact that they were published indicates that the general population was not already aware that choices introduced early in the data analysis pipeline create bias and therefore uncertainty in our final values.
-This indicates a general failure to see bias and variance both as sources of uncertainty.
-- @Wickham2011 suggests their product plot framework, which includes histograms, should have a way to measure uncertainty, but does not consider that a histogram is *already* a depiction of mass and would already be considered an uncertainty visualisation were our statistic of interest the population mean.
-To suggest that a histogram needs uncertainty bounds is similar to suggesting a density or box plot need uncertainty bounds.
-- @Griethe2006 commented that "if visualization is used as a means to explore a data volume or to communicate its contents the uncertainty has to be included".
-@Potter2010 aimed to create a summary plot that "concisely presented data with uncertaninty information".
-These papers fail to realise the exploration step of an analysis, which includes descriptive statistics, exploratory data visualisaiton, and unsupervised machine learning techniques, is performed without a prior hypothesis.
-Therefore, it is performed to *suggest* a direction for future inference work or communication, which *do* contain uncertainty and does not contain uncertainty within itself.
-This confusion would not occur if it was understood that uncertainty does not exist without inference.
-- @Ibrekk1987 asked participants for the "best" estimate while displaying a population density function.
-The participants proided the mode instead of the mean, although there was no indication the mean was what was being asked about.
-- @Boukhelifa2012 tried to quantify the strength of the intuitive connection between "sketchiness" and uncertainty.
-Since they had seen literature that explained uncertainty was task specific, they tested the connection for multiple "tasks", what they failled to understand, is that the "task" dependency which is likely just a proxy of the evaluation experiment problem discussed in the previous section.
-This misunderstanding led to "sketchiness" being used to depict uncertainty in the categories of bar charts, in graphic networks, and in train lines, with little consideration or explanation as to *how* these things would be uncertainty.
-I imagine it is quite hard to be "uncertain" about the existence of a train line.
-Participants rightfully assumed the sketchiness therefore represented something else, such alternative options or simply ignored it.
-
-![The graphics displayed by @Boukhelifa2012 to identify if there is an intuitive connection between sketchiness and uncertainty.](sketchiness.png){#fig-sketchy}
-
-The reality is that a lot of uncertainty visualisation authors do not seem to have an intuitive understanding of uncertainties connection to inference, when inference is being performed, and how to design experiments that capture this relationship.
-
-This misunderstanding is not due to laziness or the fault of the authors, but rather is likely caused by the absence of a strict definition of uncertainty.
-The imprecise and confusing set of existing definitions are creating a field in which the authors themselves do not know what they are testing.
+With this understanding it becomes clear to see why uncertainty is tied to an endless string of examples in the data analysis pipeline.
 Uncertainty examples include imputed data, model selection, inherent randomness, biased sampling, etc, not because these things *are* uncertainty, but because they *create* uncertainty *when we perform inference*.
@@
 The field of uncertainty communication is in desperate need of a unifying mathematical definition that firmly identifies exactly how each of these things contribute to uncertainty.
-This may be a considerable task, but as of right now, the list of things that are "uncertainty" continues to grow, and the best methods of uncerstanding and quantifying them are ad-hoc at best.
+This may be a considerable task, but as of right now, the list of things that are "uncertainty" continues to grow, and the best methods of understanding and quantifying them are ad-hoc at best.
 @Thomson2005 suggests a mathematical formula for *examples* of uncertainty, @Meng2014 mathematically defined the variance introduced to a model by the array of model choices, information theory tries to quantify uncertainty using the idea of entropy.
 None of these existing methods are thorough enough for an analyst to understand what causes uncertainty, and quantify it for communication.
-Realistically, this problem is well known, and put simply by Freeman Dyson in his fameous Birds and Frogs speech:
+The need for a mathematical definition of this field is put simply by Freeman Dyson in his fameous Birds and Frogs speech:
 
-> Rigorous theorems are the best way to give a subject intellectual depth and precision.
-Until you can prove rigorous theorems, you do not fully understand the meaning of your concepts.
-- Freeman Dyson (Birds and Frogs speech)
-
-This limitation becomes apparent when you look at the limited scope of what the uncertainty visualisation literatrue has *actually* focused on visualising.
-Almost every uncertainty visualisation paper contains a broad and encapsulating definition of uncertainty in the introduction, followed by a visualisation of a normally distributed probablity density function.
-
-> Edit notes: Need some way to show uncertainty visualisation papers provide sweeping definitions and then visualise a simple PDF everytime.
+> "Rigorous theorems are the best way to give a subject intellectual depth and precision. Until you can prove rigorous theorems, you do not fully understand the meaning of your concepts." @Dyson2010
 
@@
 Broader and more complicated concepts such as the effects of assumptions, imputed missing variables, and model choices remain difficult to meaningfully quantify beyond ad-hoc methods.
-This means that the current literature of uncertainty visualisation papers are doing one of two things:
-1) Evaluating multiple visual representations of a single mathematical expression of uncertainty (e.g.
-a PDF, confidence interval or error).
-These papers are purposefully vague, and would be clearer in their intention if the title was aligned with the actual mathematical object they are visualising.
-For example, a paper that is titled "evaluation of uncertainty visualisations" that only considers PDF visualisation should realistically be titled "evaluation of PDF visualisations".
-2) Evaluating multiple visual representations of *multiple* different mathematical expression of uncertainty.
-Since these experiments visualise different mathematical objects, some of which are closer to the ground truth statistic, the results of these papers can be easily predicted with an understanding of how to get the ground truth statistic from a plot and the perceptual tasks required to do so.
 
-It is clear that the vague definition is causing across the board misunderstandings of uncertainty, confusing experiment methodologies and motivations, and a literature that is difficult to navigate.
-This is also causing a large body of experiments that have easily predictable results that can be easily explained by the existing visualisation.
-This problem will be the focus of the following section.
+## Mistakes when we misunderstand uncertainty
+The previous section established that uncertainty is a by product of the assumptions we make when we perform inference, additionally, our quantification of uncertainty is largely task dependent, therefore the uncertainty associated with *one* distribution (which we can also view as a task or goal dependence).
+When uncertainty visualisaiton authors directly stare at this concept, they almost unanimously agree with these concepts, however when these details are pushed the the periphery of a problem, we start to get research that confuses itself.
+In these cases, the authors are more confused about what they are doing than the participants.
 
-- The process of working out whether or not a visualisation shows a piece of information, and if it does, what perceptual task it is mapped to and what assumptions or heuristics need to be used, is *incredibly* time consuming.
-While this method does predict the results for these experiments with remarkable accuracy, to do it for a large number of plots, especially when the ground truth statistics and visualisation are incredibly complicated is just not feasible for a reviewer to do.
+There are enough papers that fail to understand these basic principals of uncertainty that this entire paper could simply be a list with explanations. Instead, we have a handful of highlights.
+
+@Hofman2020 performed an experiment where some participants were shown a sampling distribution of the mean, and others were shown a prediction interval around the mean, for the effectiveness of a particular treatment.
+The authors reported that the group that was shown the sampling distribution were willing to pay more and ascribed a higher likelihood to the belief that the treatment was more effective than a control, than the group that was shown a prediction interval.
+A prediction interval indicates the uncertainty associated with treatment *for a particular person* and the sampling interval shown the uncertainty associated with the *average effectivness of the treatment*.
+It is already established practice in medicine to use a wide range of metrics to communicate different risks to best communicate the relevant information to a patient for a particular decision [@Spiegelhalter2017].
+The perceived importance of this work by @Hofman2020 hinges on a misunderstanding that the prediction and confidence interval *are* interchangeable despite performing inference on *very different* statistics.
+This misunderstanding is so widespread that this finding, that shows the confidence and prediction intervals are *not* interchangeable, is considered groundbreaking enough to be published in CHI.
+
+@Boukhelifa2012 tried to quantify the strength of the intuitive connection between a line attribute called "sketchiness" and uncertainty.
+The authors of this study were aware that there was some kind of task dependence with respect to displaying uncertainty, so they tested the connection for multiple "tasks".
+The authors did not seem to understand the "task" dependence is likely a dependence on the inference for a particular statistic, so the "task" dependence was interpreted as *context* dependence.
+@fig-sketchy depicts the six scenarios that participants were shown and asked to interpret what they believe the squiggly line indicates.
+The authors misunderstanding led to "sketchiness" being used to depict uncertainty in the categories of bar charts, in graphic networks, and in train lines, with little consideration or explanation as to *how* these things would be uncertainty.
+It seems a tad ridiculous for the creators of a rail network map to be "uncertain" about the existence of a train line, however we are unable to guess at what the authors of the paper believed the uncertainty would be.
+The most likely scenario is that the authors did not consider the specific estimate that the uncertainty was representing, which is what led to visualisations that do not make conceptual sense.
+Participants rightfully assumed the sketchiness therefore represented something else, such alternative options or simply ignored it.
+
+![The graphics displayed by @Boukhelifa2012 to identify if there is an intuitive connection between sketchiness and uncertainty.](sketchiness.png){#fig-sketchy}
+
+Many authors do not understand bias *is* uncertainty (something that has been proven mathematically by @Meng2014) and whether we view uncertainty as bias or variance depends on the resolution of our problem.
+@Padilla2021 found that high uncertainty in the model estimates (calculated uncertainty) and low forecaster confidence (which is typically an expression of suspected bias in the model) both caused participants to have decreased confidence in their results and suggested modelers express both if they are relevant.
+@Kale2019 discussed the importance of communicating decisions made in the data analysis pipeline and being aware of the alternatives.
+While there is nothing wrong with the results of these papers, the fact that they were published shows some level of surprise by these results.
+This work indicates that there was a significant number of authors in the field were not already aware that choices introduced early in the data analysis pipeline create bias and therefore uncertainty in our final values.
+The surprise that both bias and variance contribute to final uncertainty shows that authors do not understand both will contribute to the *distance* of your estimate to the value you are drawing inference on (AKA error).
+
+The misunderstanding that descriptive statistics also extends to visualisation.
+The exploration step of an analysis, which includes descriptive statistics, exploratory data visualisaiton, and unsupervised machine learning techniques, is performed without a prior hypothesis, however misunderstandings of this fact appears frequently in the literature.
+@Griethe2006 commented that "if visualization is used as a means to explore a data volume or to communicate its contents the uncertainty has to be included".
+@Potter2010 aimed to create a summary plot that "concisely presented data with uncertainty information" to create an exploratory visualisation tool that visualised uncertainty.
+
+This confusion (or indifference) also creates many imprecisely named papers.
+Many papers will boast a title that claims to be about uncertainty visualisaiton, but simply depicts different visual representations of a PDF, however it takes reading the methodology to find this out (*Cite Gap 9*).
+While this is a minor by product of the definition issue discussed earlier, it does result in a literature that is needlessly difficult to navigate.
+
+The reality is that a lot of uncertainty visualisation authors do not seem to have an intuitive understanding of uncertainties connection to inference, when inference is being performed, and how to design experiments that capture this relationship.
+This misunderstanding is not due to laziness or the fault of the authors, but rather is likely caused by the absence of a strict definition of uncertainty.The imprecise and confusing set of existing definitions are creating a field in which the authors themselves do not know what they are testing.
 
@@
 This appears in the literature as a mountain of seemingly conflicting statements.
-For example @Leland2005 mentions that popular graphics, such as pie charts and bar charts omit uncertainty, however at least one or both of these charts are used in most "uncertainty visualisation" experiments [] (*Cite Gap __*).
+For example @Leland2005 mentions that popular graphics, such as pie charts and bar charts omit uncertainty, however at least one or both of these charts are used in most "uncertainty visualisation" experiments [@Ibrekk1987] (*Cite Gap __*).
+Cite Gap: Confusion about what "uncertainty visualisation" is
+- @Wickham2011 suggests their product plot framework, which includes histograms, should have a way to measure uncertainty, but does not consider that a histogram is *already* a depiction of mass and would already be considered an uncertainty visualisation were our statistic of interest the population mean.
+To suggest that a histogram needs uncertainty bounds is similar to suggesting a density or box plot need uncertainty bounds.
 
@@
 Ill-formed statements can neither be true nor false and cannot be part of a logical argument, so experiments and research that build upon this ill-formed concept are neither true nor false, they simply exist.
-Someone who is less familiar with the rules of logical argument may beleive this understanding of the problem to an insufferable nitpick that creates impossible to live up to standards, however the consequences of this definitional issue are *very* real and costly.
+Someone who is less familiar with the rules of logical argument may believe this understanding of the problem to an insufferable nitpick that creates impossible to live up to standards, however the consequences of this definitional issue are *very* real and costly.
 The studies that investigate "uncertainty visualisation" may be identifying *something*, however there is good reason to believe that *something* is neither interesting nor does can it contribute to a larger theory of visualisation.
@@
 
-## Uncertainty Visualisation is not defined
+## Mathematically incomparable objects
 
@@
 Another approach is to consider uncertainty visualsiation as a specific sub-field of high-dimensional data visualsiation, and therefore uncertainty is just another variable that needs to be expressed.
+
+Cite gap __: represent different maths objects in visualisations
+- @wu2023rational recognised it is difficult to understand when differences in performance stem from information asymmetry.
 
@@
 # Noise and Signal
+
+## Types of Uncertainty Visualisation Papers
+
+> Edit notes: Originally I was planning to have the discussion following this section relate to these "types" of papers, but that didn't end up happening.
+Depending on what I do with later sections, this might need a rewrite or to just be removed, since I'm not sure how relevant this is to the main ideas of the paper.
+
+Uncertainty visualisation papers can be organised according to the *goal* of the experiment.
+Evaluation experiments are the standard rule for visualisations because the human brain is not as reliable as mathematical calculation.
+Therefore, user studies often aim to assess the limitations, biases, and heuristics of our mental calculator so that we can better understand the problems we may encounter when we plot our data.
+This is not to say any paper that suggests a visualisation without an evaluation experiment is completely lacking in justification, and there are many papers that suggest a novel visualisation without an evaluation study.
+Sometimes these papers are a preliminary step in finding a solution for common problems and intend to evaluate the visualsiation in later work.
+Until that later work is done, it is often difficult to accept one particular representation.
+While justification could come in the form of discussing established concepts in visualisation, such as the hierarchy of perceptual tasks, even these may need an evaluation study to back up their claims.
+
+The reasoning for this is obvious.
+There are a large number heuristics and biases that are not obvious to us when designing visualisation.
+Additionally, these heuristics and baises can change depending on the larger scope of the graphic and the population we are communicating with.
+Additionally, since there is often a miriad of ways to visualise any particular mathematical object, to adopt specific visualisation that is not already common practice needs reasonable justification, lest we run into a range of unforseen heuristic pitfalls.
+
+The remainder of section 4 will focus on the results and methods of uncertainty visualisation evaluation experiments.
+There are six main types of evaluation experiments in the uncertainty visualisation literature, they are:
+- Association perceptual tasks: These experiments identify perceptual tasks that are/could be associated with uncertainty, and test its association with uncertainty, check it's number of distinguishable levels, and/or check how quickly/accurately people can draw the uncertainty information from the graphic.
+- Value perceptual tasks: Participants are shown some depiction of uncertainty and are asked to extract a value, make a comparison, or do something simple to show that the graphic is readable.  
+- Heuristic Checks: These experiments are checking if a perceptual bias that has been identified through public use of a particular plot, other evaluation experiments, or gut feeling identified through public use of a plot or in a different.
+-  Plot Comparisons: These papers describe an uncertainty visualisation evaluation experiment, where they compare two or more uncertainty visualisaitons on a specific set of questions.
+These questions can be about specific statistics (e.g.
+What is the probability that A>B?) or about a secondary goal such as trust or decision making (e.g.
+How much do you trust this estimate?).
+- Qualitative analysis: These experiments show participants an uncertainty visualisation for an open ended or specific task, and asks them to describe how the visualisation was used to come to a conclusion, or they are asked to describe what they see in the visualisation.
+- Method papers: These papers discuss issues with the evaluation methods of uncertainty visualisaiton and suggest changes or improvements, typically with results indicating the benefit of the change.
+
+These experiments are not mutually exclusive within a paper, and a single study will often contain several experiments, usually from several of these categories.
+These distinctions are important because each have their own issues in how they relate to uncertainty more broadly.
+Below we will go through the five key issues in uncertainty visualisation, describe how they mainfest in the uncertainty visualisation literature, and offer some tentative solutions to these problems.
+
+We will also comment on some of the uncertainty visualisation papers that suggest a visualisation without an evaluation experiment.
+Some of these plots may offer a solution to the unique problems we mention in this paper and these visualisations will be discussed in "Part 6: Great Examples".
+We will also consider these papers to illustrate the issues that visualisations are aware of and working to fix, and discuss whether or not these directions of research are likely to bear fruit, and why.
+
 ## Uncertainty versus Risk
@@
 The conversation around uncertainty communication is often centered around the psychology is communicating risks.
+
+## Types of Uncertainty Visualisation Papers
+
+> Edit notes: Originally I was planning to have the discussion following this section relate to these "types" of papers, but that didn't end up happening.
+Depending on what I do with later sections, this might need a rewrite or to just be removed, since I'm not sure how relevant this is to the main ideas of the paper.
+
+Uncertainty visualisation papers can be organised according to the *goal* of the experiment.
+Evaluation experiments are the standard rule for visualisations because the human brain is not as reliable as mathematical calculation.
+Therefore, user studies often aim to assess the limitations, biases, and heuristics of our mental calculator so that we can better understand the problems we may encounter when we plot our data.
+This is not to say any paper that suggests a visualisation without an evaluation experiment is completely lacking in justification, and there are many papers that suggest a novel visualisation without an evaluation study.
+Sometimes these papers are a preliminary step in finding a solution for common problems and intend to evaluate the visualsiation in later work.
+Until that later work is done, it is often difficult to accept one particular representation.
+While justification could come in the form of discussing established concepts in visualisation, such as the hierarchy of perceptual tasks, even these may need an evaluation study to back up their claims.
+
+The reasoning for this is obvious.
+There are a large number heuristics and biases that are not obvious to us when designing visualisation.
+Additionally, these heuristics and baises can change depending on the larger scope of the graphic and the population we are communicating with.
+Additionally, since there is often a miriad of ways to visualise any particular mathematical object, to adopt specific visualisation that is not already common practice needs reasonable justification, lest we run into a range of unforseen heuristic pitfalls.
+
+The remainder of section 4 will focus on the results and methods of uncertainty visualisation evaluation experiments.
+There are six main types of evaluation experiments in the uncertainty visualisation literature, they are:
+- Association perceptual tasks: These experiments identify perceptual tasks that are/could be associated with uncertainty, and test its association with uncertainty, check it's number of distinguishable levels, and/or check how quickly/accurately people can draw the uncertainty information from the graphic.
+- Value perceptual tasks: Participants are shown some depiction of uncertainty and are asked to extract a value, make a comparison, or do something simple to show that the graphic is readable.  
+- Heuristic Checks: These experiments are checking if a perceptual bias that has been identified through public use of a particular plot, other evaluation experiments, or gut feeling identified through public use of a plot or in a different.
+-  Plot Comparisons: These papers describe an uncertainty visualisation evaluation experiment, where they compare two or more uncertainty visualisaitons on a specific set of questions.
+These questions can be about specific statistics (e.g.
+What is the probability that A>B?) or about a secondary goal such as trust or decision making (e.g.
+How much do you trust this estimate?).
+- Qualitative analysis: These experiments show participants an uncertainty visualisation for an open ended or specific task, and asks them to describe how the visualisation was used to come to a conclusion, or they are asked to describe what they see in the visualisation.
+- Method papers: These papers discuss issues with the evaluation methods of uncertainty visualisaiton and suggest changes or improvements, typically with results indicating the benefit of the change.
+
+These experiments are not mutually exclusive within a paper, and a single study will often contain several experiments, usually from several of these categories.
+These distinctions are important because each have their own issues in how they relate to uncertainty more broadly.
+Below we will go through the five key issues in uncertainty visualisation, describe how they mainfest in the uncertainty visualisation literature, and offer some tentative solutions to these problems.
+
+We will also comment on some of the uncertainty visualisation papers that suggest a visualisation without an evaluation experiment.
+Some of these plots may offer a solution to the unique problems we mention in this paper and these visualisations will be discussed in "Part 6: Great Examples".
+We will also consider these papers to illustrate the issues that visualisations are aware of and working to fix, and discuss whether or not these directions of research are likely to bear fruit, and why.
 
@@
 ## Decision Making
-
 Decision making questions that typically involve uncertainty discuss the concept of "risk-aversion" as though it is a mistake.
@@
 Unfortunately this usually comes in the form of slightly cryptic questions that create a large amount of noise on the interpretation side [@Hullman2016].
+
+Cite gap __: Weird questions to capture uncertainty
+- @Ibrekk1987 asked participants for the "best" estimate while displaying a population density function.
+The participants proided the mode instead of the mean, although there was no indication the mean was what was being asked about.
 - @Hofmann2012 showed two distributions in 20 different visualisations (a lineup protocol) using a jittered sample, a density plot, a histogram, and a box plot.
@@
 
-## Uncertainty is not hypothesis testing
+## Uncertainty and hypothesis testing
 It is unclear how participants are supposed to incorperate uncertainty information into their responses when uncertainty visualisation authors themselves are unsure.
@@
 ## Suggestions for measuring uncertainty
-### Types of Uncertainty Visualisation Papers
-
-> Edit notes: Originally I was planning to have the discussion following this section relate to these "types" of papers, but that didn't end up happening.
-Depending on what I do with later sections, this might need a rewrite or to just be removed, since I'm not sure how relevant this is to the main ideas of the paper.
-
-Uncertainty visualisation papers can be organised according to the *goal* of the experiment.
-Evaluation experiments are the standard rule for visualisations because the human brain is not as reliable as mathematical calculation.
-Therefore, user studies often aim to assess the limitations, biases, and heuristics of our mental calculator so that we can better understand the problems we may encounter when we plot our data.
-This is not to say any paper that suggests a visualisation without an evaluation experiment is completely lacking in justification, and there are many papers that suggest a novel visualisation without an evaluation study.
-Sometimes these papers are a preliminary step in finding a solution for common problems and intend to evaluate the visualsiation in later work.
-Until that later work is done, it is often difficult to accept one particular representation.
-While justification could come in the form of discussing established concepts in visualisation, such as the hierarchy of perceptual tasks, even these may need an evaluation study to back up their claims.
-
-The reasoning for this is obvious.
-There are a large number heuristics and biases that are not obvious to us when designing visualisation.
-Additionally, these heuristics and baises can change depending on the larger scope of the graphic and the population we are communicating with.
-Additionally, since there is often a miriad of ways to visualise any particular mathematical object, to adopt specific visualisation that is not already common practice needs reasonable justification, lest we run into a range of unforseen heuristic pitfalls.
-
-The remainder of section 4 will focus on the results and methods of uncertainty visualisation evaluation experiments.
-There are six main types of evaluation experiments in the uncertainty visualisation literature, they are:
-- Association perceptual tasks: These experiments identify perceptual tasks that are/could be associated with uncertainty, and test its association with uncertainty, check it's number of distinguishable levels, and/or check how quickly/accurately people can draw the uncertainty information from the graphic.
-- Value perceptual tasks: Participants are shown some depiction of uncertainty and are asked to extract a value, make a comparison, or do something simple to show that the graphic is readable.  
-- Heuristic Checks: These experiments are checking if a perceptual bias that has been identified through public use of a particular plot, other evaluation experiments, or gut feeling identified through public use of a plot or in a different.
--  Plot Comparisons: These papers describe an uncertainty visualisation evaluation experiment, where they compare two or more uncertainty visualisaitons on a specific set of questions.
-These questions can be about specific statistics (e.g.
-What is the probability that A>B?) or about a secondary goal such as trust or decision making (e.g.
-How much do you trust this estimate?).
-- Qualitative analysis: These experiments show participants an uncertainty visualisation for an open ended or specific task, and asks them to describe how the visualisation was used to come to a conclusion, or they are asked to describe what they see in the visualisation.
-- Method papers: These papers discuss issues with the evaluation methods of uncertainty visualisaiton and suggest changes or improvements, typically with results indicating the benefit of the change.
-
-These experiments are not mutually exclusive within a paper, and a single study will often contain several experiments, usually from several of these categories.
-These distinctions are important because each have their own issues in how they relate to uncertainty more broadly.
-Below we will go through the five key issues in uncertainty visualisation, describe how they mainfest in the uncertainty visualisation literature, and offer some tentative solutions to these problems.
-
-We will also comment on some of the uncertainty visualisation papers that suggest a visualisation without an evaluation experiment.
-Some of these plots may offer a solution to the unique problems we mention in this paper and these visualisations will be discussed in "Part 6: Great Examples".
-We will also consider these papers to illustrate the issues that visualisations are aware of and working to fix, and discuss whether or not these directions of research are likely to bear fruit, and why.
 

### 2024-03-13 writingdumpster.qmd
finshed rewrite of mistakes made section

@@
 Cite Gap 8: It is commonly noted that each uncertainty must be discussed in isolation, but combining the uncertainty from every stage is near impossible [@Spiegelhalter2017]
+
+Cite Gap 9: Many papers will boast a title that claims to be about uncertainty visualisaiton, but simply depicts different visual representations of a PDF, however it takes reading the methodology to find this out
 
@@
 # Main section outtakes
+- The process of working out whether or not a visualisation shows a piece of information, and if it does, what perceptual task it is mapped to and what assumptions or heuristics need to be used, is *incredibly* time consuming.
+While this method does predict the results for these experiments with remarkable accuracy, to do it for a large number of plots, especially when the ground truth statistics and visualisation are incredibly complicated is just not feasible for a reviewer to do.
 - This is likely because The idea that uncertainty can only be defined in the presence of a motivating question is well grounded in most areas of statistics.

### 2024-03-14 paper.qmd
rewrote intro to uncert vis

@@
 
-# What is an Uncertainty Visualisation?
-The lack of a consistent definition of uncertainty means that every statement that builds upon "uncertainty" will be ill-formed.
-This includes statements about uncertainty visualisation.
-This appears in the literature as a mountain of seemingly conflicting statements.
+# What is Uncertainty Visualisation?
+If we simply built the idea of uncertainty visualisation up from "uncertainty" which we already know has a precarious definition we would understand the field of "uncertainty" visualisation must also be precarious.
+You cannot quantify something that is not defined, and you can't visualise something you cannot quantify.
+If we cannot quantify the full concept uncertainty in any meaningful way (other than with examples such as missing values or a PDF) then there is no such thing as an "uncertainty" visualisation.
+Unfortunately this problem is not quite that simple.
+
+A large drive of the field of uncertainty visualisation, has been the idea that "uncertainty visualisation" (defined as one word) is somehow different to "uncertainty" "visualisation" (separately defined as two words).
+When "uncertainty visualisation" is defined as a single word we lose some nuance in our understanding of the driving force behind some issues, and problems with visualisation and problems with uncertainty become conflated.
+This appears in the literature as a mountain of seemingly conflicting statements about what is, or is not, an "uncertainty visualisation".
 For example @Leland2005 mentions that popular graphics, such as pie charts and bar charts omit uncertainty, however at least one or both of these charts are used in most "uncertainty visualisation" experiments [@Ibrekk1987] (*Cite Gap __*).
-Cite Gap: Confusion about what "uncertainty visualisation" is
-- @Wickham2011 suggests their product plot framework, which includes histograms, should have a way to measure uncertainty, but does not consider that a histogram is *already* a depiction of mass and would already be considered an uncertainty visualisation were our statistic of interest the population mean.
-To suggest that a histogram needs uncertainty bounds is similar to suggesting a density or box plot need uncertainty bounds.
+@Wickham2011 suggests their product plot framework, which includes histograms, should have a way to measure uncertainty, but does not consider that a histogram is *already* a depiction of mass and would already be considered an uncertainty visualisation were our statistic of interest the population mean.
 
-Attempts to rectify these conflicts always flutter about the surface and do not seem to grasp the core of this problem.
-For example, @wu2023rational noticed a series of papers with poorly defined tasks and suggested comparing study participants to a "rational agent" to help authors understand and remedy this issue.
-@Kim2019 believed an understanding of participants prior beliefs could assist in interpreting uncertainty visualisation results that tend to be incredibly noisy.
-What these suggestions fail to understand is that the entire practice of identifying "true" facts about "unertainty visualisation" is futile.
-It is impossible to make a statement about the properties of an "uncertainty visualisation" in the same way it is impossible to comment on $\frac 0 0 = 1$.
-Division by zero is undefined, so the statement is ill formed, similarly, "uncertainty" is not defined, so statements *about* "uncertainty" including statements about "uncertainty visualisation" are ill formed.
-Ill-formed statements can neither be true nor false and cannot be part of a logical argument, so experiments and research that build upon this ill-formed concept are neither true nor false, they simply exist.
-Someone who is less familiar with the rules of logical argument may believe this understanding of the problem to an insufferable nitpick that creates impossible to live up to standards, however the consequences of this definitional issue are *very* real and costly.
-The studies that investigate "uncertainty visualisation" may be identifying *something*, however there is good reason to believe that *something* is neither interesting nor does can it contribute to a larger theory of visualisation.
-This is something we illustrate in the following this section.
+These conflicting ideas and lacking definition of what is means for a visualisation to be an "uncertainty visualisation" is a conerstone of some clearly questionable results within this field.
 
-## Common misconceptions about uncertainty visualisation
-
-### Defining "uncertainty visualisation" as one word
+## Taxonomies of uncertainty visualisation
 
@@
 
-### Misunderstanding the role of contextual information
+## Contextual Information
 > Edit Note: This section should probably be moved to the visualisation specific section, although that depends on the tone that I end up with after the rewrites of the earlier sections.
@@
 So, if we cannot see any difference to uncertainty information when it comes to value abstraction, *when* is uncertainty information different?
+
+## Fixes within "uncertainty visualisation"
+Attempts to rectify these conflicts always flutter about the surface and do not seem to grasp the core of this problem.
+For example, @wu2023rational noticed a series of papers with poorly defined tasks and suggested comparing study participants to a "rational agent" to help authors understand and remedy this issue.
+@Kim2019 believed an understanding of participants prior beliefs could assist in interpreting uncertainty visualisation results that tend to be incredibly noisy.
+What these suggestions fail to understand is that the entire practice of identifying "true" facts about "unertainty visualisation" is futile.
+It is impossible to make a statement about the properties of an "uncertainty visualisation" in the same way it is impossible to comment on $\frac 0 0 = 1$.
+Division by zero is undefined, so the statement is ill formed, similarly, "uncertainty" is not defined, so statements *about* "uncertainty" including statements about "uncertainty visualisation" are ill formed.
+Ill-formed statements can neither be true nor false and cannot be part of a logical argument, so experiments and research that build upon this ill-formed concept are neither true nor false, they simply exist.
+Someone who is less familiar with the rules of logical argument may believe this understanding of the problem to an insufferable nitpick that creates impossible to live up to standards, however the consequences of this definitional issue are *very* real and costly.
+The studies that investigate "uncertainty visualisation" may be identifying *something*, however there is good reason to believe that *something* is neither interesting nor does can it contribute to a larger theory of visualisation.
+This is something we illustrate in the following this section.
 

### 2024-03-14 paper.qmd
small edits

@@
 This appears in the literature as a mountain of seemingly conflicting statements about what is, or is not, an "uncertainty visualisation".
-For example @Leland2005 mentions that popular graphics, such as pie charts and bar charts omit uncertainty, however at least one or both of these charts are used in most "uncertainty visualisation" experiments [@Ibrekk1987] (*Cite Gap __*).
+For example @Leland2005 mentions that popular graphics, such as pie charts and bar charts omit uncertainty, however at least one or both of these charts are used in most "uncertainty visualisation" experiments [@Ibrekk1987] (*Cite Gap 10*).
 @Wickham2011 suggests their product plot framework, which includes histograms, should have a way to measure uncertainty, but does not consider that a histogram is *already* a depiction of mass and would already be considered an uncertainty visualisation were our statistic of interest the population mean.
 
-These conflicting ideas and lacking definition of what is means for a visualisation to be an "uncertainty visualisation" is a conerstone of some clearly questionable results within this field.
+These conflicting ideas and lacking definition of what is means for a visualisation to be an "uncertainty visualisation" is a cornerstone of some clearly questionable results within this field.
 
@@
 
-> Edit Notes: Similar to the "similar works" section, I think this section would be better off with a short sentance about each paper and comments on their similarities because a paragraph for each paper is too much
+Similar to "uncertainty" exactly what defines an "uncertainty visualisation" is not clear.
+Often, the visualisaitons are simply categorised according to features, however exactly what *is* or *is not* an uncertainty visualisation is rarely explained.
 
@@
 
-### Uncertainty as signal
+## Uncertainty as signal
 
@@
 
+## The current summaries of findings
 ## Uncertainty versus Risk

### 2024-03-14 writingdumpster.qmd
small edits

@@
 
-Cite Gap 3: Some consider it to be synonymous with specific terms defined in mathematics, such as probability, variance, error, or precision.
-Others consider it to be an encompassing umbrella term of which for these mathematically defined objects are only examples, and include concepts that are somewhat related, such as missing values
+Cite Gap 3: Some consider it to be synonymous with specific terms defined in mathematics, such as probability [*HERE*], variance[*HERE*], error[*HERE*], or precision[*HERE*].
+Others consider it to be an encompassing umbrella term of which for these mathematically defined objects are only examples[*HERE*], and include concepts that are somewhat related, such as missing values[*HERE*]
 
-Cite Gap 3.5: This wide array of uncertainty definitions becomes a snake eating its own tail, as many authors opt to redefine uncertainty themselves[], ignore defining uncertainty entirely [], or pick a definition from a seemingly randomly previously published papers [], all of which make variance in uncertainty definitions even worse
+Cite Gap 3.5: This wide array of uncertainty definitions becomes a snake eating its own tail, as many authors opt to redefine uncertainty themselves[*HERE*], ignore defining uncertainty entirely [*HERE*], or pick a definition from a seemingly randomly previously published papers [*HERE*], all of which make variance in uncertainty definitions even worse
 
@@
 
-Cite Gap 5: This definition swap happens so subtly that it seems to go unnoticed by authors writing the papers, and the fact that "uncertainty visualisation" methods dont even notice they are exclusively focus on depictions of easily quantifiable uncertainty, such as error [], mass [], or variance []
+Cite Gap 5: This definition swap happens so subtly that it seems to go unnoticed by authors writing the papers, and the fact that "uncertainty visualisation" methods dont even notice they are exclusively focus on depictions of easily quantifiable uncertainty, such as error [*HERE*], mass [*HERE*], or variance [*HERE*]
 
-Cite Gap 6: This has left the expressions of uncertainty that are hard to quantify or visually differentiate untouched, despite many papers calling for their invention
+Cite Gap 6: This has left the expressions of uncertainty that are hard to quantify or visually differentiate untouched, despite many papers calling for their invention [*HERE*]
 
-Cite Gap 7: Taxonomies split uncertainty based on an endless stream of arbitrary lines in the sand, such as whether the uncertainty is due to true randomness or a lack of knowledge [@Spiegelhalter2017], if the uncertainty is in the the attribute, spatial elements, or temporal element of the data [], whether the uncertainty is scientific (e.g.
+Cite Gap 7: Taxonomies split uncertainty based on an endless stream of arbitrary lines in the sand, such as whether the uncertainty is due to true randomness or a lack of knowledge [@Spiegelhalter2017 *HERE*], if the uncertainty is in the the attribute, spatial elements, or temporal element of the data [*HERE*], whether the uncertainty is scientific (e.g.
 error) or human (e.g.
-disagreement among parties) [], if the uncertainty is random or systematic [], statistical or bounded [], accuracy or precision [], if the uncertainty is about the past or the future [], the stage of the data analysis pipeline it comes from [], the severity of the uncertainty [], how quantifiable the uncertainty is [@Spiegelhalter2017], and many, many, many more examples.
+disagreement among parties) [*HERE*], if the uncertainty is random or systematic [*HERE*], statistical or bounded [], accuracy or precision [*HERE*], if the uncertainty is about the past or the future [*HERE*], the stage of the data analysis pipeline it comes from [*HERE*], the severity of the uncertainty [*HERE*], how quantifiable the uncertainty is [@Spiegelhalter2017], and many, many, many more examples.
 
-Cite Gap 8: It is commonly noted that each uncertainty must be discussed in isolation, but combining the uncertainty from every stage is near impossible [@Spiegelhalter2017]
+Cite Gap 8: It is commonly noted that each uncertainty must be discussed in isolation, but combining the uncertainty from every stage is near impossible [@Spiegelhalter2017 *MORE HERE*]
 
 Cite Gap 9: Many papers will boast a title that claims to be about uncertainty visualisaiton, but simply depicts different visual representations of a PDF, however it takes reading the methodology to find this out
+
+Cite Gap 9: For example @Leland2005 mentions that popular graphics, such as pie charts and bar charts omit uncertainty, however at least one or both of these charts are used in most "uncertainty visualisation" experiments [@Ibrekk1987 *MORE HERE*]
+
+Cite Extra: INCORRECT INTERPRETATION OF STUDIES.
+People will use the best available information to them.
+If you ask for a maximum, and give them a confidence interval...
+they will give you the maximum number that they are aware is still plausible.
+Instead of being shocked the participants were wrong, it is important to ask "what was required for them to be RIGHT".
+  - HAVE SOME PERSPECTIVE. If you are paying people below minimum wage to do your experiment, they don't care. If a problem requires extensive mental labour to solve, they will liekly not do it.
+  - This result being reported as "people interpret a confidence interval as a maximum and minimum despite being told it was a confidence interval" is simply not true. It is a false equivalence. 
+  - all these studies show is if you give people the wrong information... they will be wrong. Shocking.
 

### 2024-03-15 paper.qmd
finished taxonomy section

@@
 
+## Establishing misunderstood concepts
+Before discussing the issues facing uncertainty visualisation due to a poor concept *of* uncertainty visualisation, we want to establish two ideas that should be considered in conjunction with the lacking definition of uncertainty.
+The first is the sidestepping of the "uncertainty" definition problem by suggesting "uncertainty visualisation" is inherently different.
+Some authors may consider "uncertainty" to be poorly defined, but "uncertainty visualisation" to be well established.
+We will spend the next section explaining why that is not the case by discussing several taxonomies of uncertainty visualisation.
+The second is a conflation of the contextual information with the visual information.
+This problem is not, in isolation an issue, however it does seem to be a *key* contributor to the misunderstanding that uncertainty is an attribute of data, and not a by product of inference.
+Both of these issues need to be understood in order for us to fully understand the issues surrounding uncertainty visualisation.
+
+### Definitions of "uncertainty visualisation"
 A large drive of the field of uncertainty visualisation, has been the idea that "uncertainty visualisation" (defined as one word) is somehow different to "uncertainty" "visualisation" (separately defined as two words).
@@
 @Wickham2011 suggests their product plot framework, which includes histograms, should have a way to measure uncertainty, but does not consider that a histogram is *already* a depiction of mass and would already be considered an uncertainty visualisation were our statistic of interest the population mean.
-
 These conflicting ideas and lacking definition of what is means for a visualisation to be an "uncertainty visualisation" is a cornerstone of some clearly questionable results within this field.
 
-## Taxonomies of uncertainty visualisation
+Since the concept of "uncertainty visualisation" is harder to define than "uncertainty", "uncertainty visualisation" taxonomies are typically less thorough than taxonomies for "uncertainty".
+They also tend to blur the line between the mathematical elements of uncertainty and the visual depiction of those mathematical objects.
+While we will not focus on these taxonomies, it is still important to mention them and highlight what we can learn from their success' and failures..
 
-Similar to "uncertainty" exactly what defines an "uncertainty visualisation" is not clear.
-Often, the visualisaitons are simply categorised according to features, however exactly what *is* or *is not* an uncertainty visualisation is rarely explained.
+Some taxonomies of uncertainty highlight how confusing most authors find the concept, and how difficult it is to articulate the rules of an uncertainty visualisation when uncertainty itself if not defined.
+@Potter2012 organised several existing uncertainty visualisations into groups based on the dimensionality of the data (1D, 2D, 3D, and No Dimension) and the dimensionality of the uncertainty (Scalar, Vector, Tensor).
+However, because the term "PDF", a statistical object that describes a random variable that is typically a one dimensional function, is used to describe both the data and the uncertainty for all dimensions, it is hard to understand how this should work.
+@Grewal2021 created a taxonomy that mapped uncertainty visualisations to some point in a 2D space defined by the "domain expertise" and "continuum of discreteness" (that scaled from "point estimate" to "continuous distribution").
+This conceptualisation could be considered a visual extension of the law of large numbers, because as a point estimate is a single sample, as the sample size gets larger the distribution will appear to be a continuous.
+However, assumes every point estimate is a sample from the same distribution (a mean is not technically a sample from a prediction interval, but it is a sample from the sampling distribution) and it conflates discreteness from a too small a number of observations, discreteness from a precision problem, discreteness because that is the true nature of the variable, and discreteness as a visualisation choice.
+@Griethe2006 organised uncertainty visualisations into two cases (1) a hypothesis test was preformed to confirm the validity of the visualisation and (2) the visualisation has uncertainty depicted.
+This conceptualization conflates hypothesis testing and confidence intervals, and while they are related concepts, a hypothesis test is not necessarily uncertainty.
+Of course the authors of these papers did not intend for the taxonomy to be a complete description of uncertainty visualisations, however these distinctions that are often ignored are subtly important.
+(*Cite Gap 10*)
 
-There are some taxonomies of visual uncertainty.
-These taxonomies seem to seek to define "uncertainty visualisation" and a combined term rather than as two separate words.
-We largely disagree with this conceptualization of uncertainty visualisations as it blurs the line between the mathematical estimations and assumptions (the uncertainty part) and the visual depiction of those mathematical objects (the visualisation part).
-While we will not focus on these taxonomies, it is still important to mention them and highlight the reasons they are not good practice.
+There are a small handful of taxonomies that highlight some interesting features in the way uncertainty is visualised. @Kinkeldey2014 categorised uncertainty according to whether or not it was:
+1. Implicit (a sample) or explicit (a depiction of mass)
+2. Intrinsic (alter existing symbols symbols to represent uncertainty) or extrinsic (add new objects to represent uncertainty)
+3. visually integral or separable (the uncertainty can be separated from the data and read independently) 
+4. coincidence (uncertainty and data are represented in the same plot) or adjacent
+5. static or dynamic (animated, interactive ect)
+Of the groups they identified, they actually only used (4), (2), and (5), left out (1) because most visualisations are explicit, and (3) corresponds to (1) in most cases [@Kinkeldey2014].
+This taxonomy suggests interesting considerations for a visualisation of a distribution (even though it was not intended to be a visualisation of a distribution).
+A similar version of this taxonomy was presented by @uncertchap2022 who commented that visualisation can be organised into two categories, "graphical annotations of distributional properties" and "visual encodings of uncertainty" which functionally aline with the intrinsic/extrinsic distinction.
+The first four categories boil down to the question "should uncertainty and signal be conveyed together as one variable or separately as two".
+This consideration will become more relevant later in this review.
 
-There are a small handful of taxonomies (or typologies) that are specifically for uncertainty visualisations.
-@Kinkeldey2014 identified give categories of uncertainty based on the work of other literature reviews of uncertainty visualisation.
-The five categories are: (1) explicit/implicit (directly mapping or showing multiple outcomes) (2) intrinsic/extrinsic (using existing symbols e.g.
-colour value, or new objects e.g.
-grids) (3) visually integral/separable (can or cannot be separated from the data) (4) coincidence/adjacent (if data and uncertainty are in integrated or separate views (5) static/dynamic (animation/interaction).
-Of the groups they identified, they actually only used (4), (2), and (5), left out (1) because most visualisations are explicit, and (3) corresponds to (1) in most cases [@Kinkeldey2014].
+![This visualisation is here as a placeholder.
+I am going to replace it with an R coded visualisation that shows the 5 considerations for showing uncertainty in a plot according to @Kinkeldey2014 ](paperscreenshots/collage2.jpg){#fig-location}
 
-@Potter2012 organised several existing uncertainty visualisations into groups based on the dimensionality of the data (1D, 2D, 3D, and No Dimension) and the dimensionality of the uncertainty (Scalar, Vector, Tensor).
-However, because the term "PDF", a statistical object that describes a random variable that is typically a one dimensional function, is used to describe both the data and the uncertainty for all dimensions, it is hard to understand how this dimensionality should work.
-
-@Grewal2021 created a taxonomy that mapped uncertainty visualisations to some point in a 2D space that was defined by the axis "domain expertise" and "continum of discreteness" that scaled from "point estimate" to "continuous distribtion".
-
-@Thomson2005 titled their paper "A typology for visualizing uncertainty" however it is a taxonomy for the specific types of uncertainty that we may want to visualise and a way to calculate that uncertainty.
-While this paper is more about mathematics than visualisation, it is clearly written as a precursor to visualisation.
-
-![Example of different uncertainty visualisation taxonomies](paperscreenshots/collage2.jpg){#fig-location}
-
-These taxonomies are useful in the sense that they often identify areas where there is an over or under supply of uncertainty visualisations.
-Additionally taxonomies do highlight a concept of uncertainty dimensionality that was not directly covered in the uncertainty taxonomy, however the distinction between uncertainty as a single value versus as a PDF could be considered a sub-consideration of uncertainty that's level is statistical so it can be quantified.
-While uncertainty visualisation taxonomies are not the standard practice (generating a taxonomy of uncertainty and then developing methods to depict that uncertainty is) [@Kinkeldey2014], it is still important to acknowledge the existence of "uncertainty visualisation typologies" and why they may miss the mark in what is needed from the uncertainty visualisaiton literature.
-
-The first issue with these taxonomies is that most of these taxonomies are created by observing existing uncertainty visualisations.
-This means under researched gaps in the taxonomy could be due to a lack of need just as well as a lack of attention, and the benefit of trying to create visualisations that land in uncharted areas of the taxonomy are unclear.
-
-The second issue is that there does not seem to be anything these taxonomies offer that would not be better established by separate taxonomies for uncertainty and visualisation.
-These taxonomies seem to depict uncertainty visualisation as a field within itself, rather than a specific case of information visualisation that requires more attention to mathematics than usual.
+Most of these taxonomies are created by observing existing uncertainty visualisations, which means they suffer from many of the same conceptual issues that the field has with uncertainty in general.
+Additionally, there does not seem to be anything these taxonomies offer that would not be better established by separate taxonomies for uncertainty and visualisation.
 The difference between "uncertainty visualisations" and "data visualisations" is not technically in the visual element, it is mathematical.
@@
 The process of understanding and estimating uncertainty requires knowledge of the data, the statistical methods used to make an estimate, and the assumptions of a model.
-Visualising the statistics that represent uncertainty should be no different than depicting the statistics that represent any other element of a graphic, and therefore there seems to be no reason to have a separate taxonomy for uncertainty visualisation.
+Visualising the statistics that represent uncertainty should be no different than depicting the statistics that represent any other element of a graphic, this is something we will show in the following sections of this paper.
 
-## Contextual Information
-> Edit Note: This section should probably be moved to the visualisation specific section, although that depends on the tone that I end up with after the rewrites of the earlier sections.
-
-Some uncertainty work, specifically uncertainty visualisation, focus on the context surrounding the uncertainty.
-These are things like spatial data, temporal data, category data, etc.
+### Contextual Information
+Some uncertainty work, specifically uncertainty visualisation, focus on the context surrounding the uncertainty, such as the spatial or temporal nature of the estimate.
 While other discussions of uncertainty may consider this distinction more relevant to their work, here we will consider this contextual information.
 Contextual information is important for interpretation and understanding of graphics, but it does not in of itself generate uncertainty.
+Its relevance to the field of "uncertainty visualisation" is through the percieved lack of available channels.
+For example, a choropleth map has already used colour to represent an estimate, and the concept of space to provide contextual information, there is little remaining room tany spatial
+
 This is not unique to the properties of spatial information, other statistics information that is important for modelling and assumptions, such as non-linearity, class levels, etc do not contribute to uncertainty.

### 2024-03-15 writingdumpster.qmd
finished taxonomy section

@@
 
-Cite Extra: INCORRECT INTERPRETATION OF STUDIES.
-People will use the best available information to them.
-If you ask for a maximum, and give them a confidence interval...
-they will give you the maximum number that they are aware is still plausible.
-Instead of being shocked the participants were wrong, it is important to ask "what was required for them to be RIGHT".
-  - HAVE SOME PERSPECTIVE. If you are paying people below minimum wage to do your experiment, they don't care. If a problem requires extensive mental labour to solve, they will liekly not do it.
-  - This result being reported as "people interpret a confidence interval as a maximum and minimum despite being told it was a confidence interval" is simply not true. It is a false equivalence. 
-  - all these studies show is if you give people the wrong information... they will be wrong. Shocking.
+Cite Gap 10: Uncertainty Visualisation Taxonomies
+
+# Extra Section Cite Gaps
+Cite Gap - New Section: Logical fallacy interpretation of mathematically unanswerable questions.
+If I tell you my cousins age, and then ask you to guess my age, it would not be unreasonable to guess the age I provided for my cousin.
+Citing this work to state "people incorrectly believe my cousins age is my age" would
+- Deterministic questions about a random event cannot be answered. 
+- @uncertchap2022 took the results of other paper to mean "when participants viewed mean temperature forecasts that included 95% confidence intervals depicted as bars with end caps, they incorrectly believed that the error bars represented high and low temperatures" (this one will depend on that study)
 
 # Moved over notes
+@Thomson2005 titled their paper "A typology for visualizing uncertainty" however it is a taxonomy for the specific types of uncertainty that we may want to visualise and a way to calculate that uncertainty.
+While this paper is more about mathematics than visualisation, it is clearly written as a precursor to visualisation.
 - Some markers of probability that have common uses, probabilities such as 0 (impossible), 0.5 (a fair coin flip), 1 (certain) are easy for people to have an internal sense of *hullman paper*. 

### 2024-03-17 paper.qmd
small edit

@@
 ### Contextual Information
+One of the key components of uncertainty and inference is the noise and signal dichotomy.
+We can only identify which pieces of information contain signal in relation to a signal of interest, the uncertainty around that signal (or noise) is not a
+
 Some uncertainty work, specifically uncertainty visualisation, focus on the context surrounding the uncertainty, such as the spatial or temporal nature of the estimate.

### 2024-03-18 paper.qmd
rewrote contextual information section

--- 2024-03-15 paper.qmd
@@
 This is mentioned both by authors at a specific stage of an analysis, and by authors looking at an entire field or at all stages of the analysis pipeline.
-@Otsuka2023 suggested that the process of observing data to perform statistics is largely dependent on our goals, because the process of boiling real world entities down into probabilistic objects depends on the relationship we seek to identify with our data.
+@Otsuka2023 suggested that the process of observing data to perform statistics is largely dependent on our goals, because the process of boiling real world entities down into probabilistic objects (or "probabilistic kind" as he puts it) depends on the relationship we seek to identify with our data.
 @Meng2014 commented what is kept as data and what is tossed away is determined by the motivation of an analysis and what was previously noise can be shown to become signal depending on the the question we seek to answer.
@@
 ### Contextual Information
-Some uncertainty work, specifically uncertainty visualisation, focus on the context surrounding the uncertainty, such as the spatial or temporal nature of the estimate.
+One of the key components of uncertainty and inference is the noise and signal dichotomy.
+When @Otsuka2023 discusses the information we boil away to create a "probabilistic kind" he uses the example of a coin flip and the sex of baby births as two things that are *very* different in reality, but it is only through the lense of a Bernoulli distribution that we can view these two things as the same.
+The fact is that this boiling down, of a real world event into data, is not something that happens independent of the statisticians, but is just as much as product of our interests (the particular relationship we hope to draw inference on) as it is a product of the event's natural properties.
+This is expressed simply in the text with an extension to the coin flipping example:
+
+> "Recall that we modeled a coin toss by a Bernoulli distribution.
+That was because we supposed the coin will land either heads or tails and ignored the possibility of its landing on its edge.
+If we are to take the latter possibility into account, we should use a multinational distribution with three categories.
+As this example shows, the question of which statistical model we should use is not just determined by the physical nature of the object under study, but rather depends on the modeler’s interests and intentions." @Otsuka2023
+
+With this in mind, let us turn to the concept of "contextual" information.
+For the purposes of this review, we will consider contexual information to be anything that makes it easy for the audience to connect the data from a probablistic kind to its real world parallel, but is largely irrelevant in the abstract probability space.
+A key example of this is spatial information, such as borders, road locations, terrains, etc.
+Spatial information is *highly relevant* when we translate from the abstract domain back into real world for communication.
+It can help provide reasons for *why* there may be specific characteristics in our data, however it is not required information for the model.
+@fig-spatial shows the proces of boilling information down into data, visualising it, and including contextual information to improve our scientific reasoning.
+The concept that contextual information is not the same as statistical information is also highlighted in *The Grammar of Graphics* which states:
+
+> "Geography is anchored in real space-time and statistics in abstract dimensions.
+This is a distinction along a continuum rather than a sharp break...
+but this difference in focus clearly means that a system optimized to handle geography will not be graceful when dealing with statistical graphics." @Leland2005
+
+![Data pipeline with visualisations to show why we are ignoring contextual information for the time being.
+The visualisations that depict the data are similar to typical statistical graphics.
+The purpose of the spatial information is to provide context to the information we extract from the statistical aspect of the graphic.](spatialinfo.jpeg){#fig-spatial}
+
+For the remaining of this paper, we are not going to differentiate uncertainty visualisation using contextual information, despite the fact that it is common in the field.
+Our reasoning for this is that organising the field according to contextual information, promotes the idea that uncertainty is a concept specific to data qualities and ignores the role of inferential statistics.
+
 While other discussions of uncertainty may consider this distinction more relevant to their work, here we will consider this contextual information.
@@
 Its relevance to the field of "uncertainty visualisation" is through the percieved lack of available channels.
-For example, a choropleth map has already used colour to represent an estimate, and the concept of space to provide contextual information, there is little remaining room tany spatial
-
-This is not unique to the properties of spatial information, other statistics information that is important for modelling and assumptions, such as non-linearity, class levels, etc do not contribute to uncertainty.
-Operations on data, inherrent randomness, and assumptions that create bias generate uncertainty, which is what we are talking about here.
+For example, a choropleth map has already used colour to represent an estimate, and position to represent spatial information, which is sometimes relevant to the probabilistic kind, and other times it is simply contextual information (such as when the modelling is done by using a state as a category rather than its geographic information).
+These two cases, when the spatial information is model relevant and when it is contextual, are rarely separated.
+It can lead authors to believe the spatial information is always of the highest importance, and thus it is always given the position axis, even when it is not, especially with respect to the uncertainty in a model.
 A confidence interval is a confidence interval whether it came from spatial data, temporal, or cross sectional data.
 
-Papers that stress the importance of spatial or temporal contexts may be conflating contextual information with uncertainty.
-@fig-spatial depicts the role of contextual information in graphics, as something that helps us understand the information in the statistical graphic.
-When we create statistical models and quantify uncertainty, the spatial or temporal aspects are often boilled away [@Otsuka2023], and the only element of it that remains is the inexchangeability of our data.
-For this reason, there realistically is no such thing as "spatial" uncertainty in the same sense that there is no such thing as "photograph" uncertainty or "coin toss" uncertainty or "baby birth" uncertainty.
-Maps and photographs boil down to the same statistical objects in the same way that coin tosses and baby births boil down to the same statistical objects.
-Once we have removed the physical nature these objects are near identical when considered statistically.
-Of course the nature of them can change based on what we choose to observe (i.e.
-if we consider a coin landing on its side to be an outcome or if we consider intersex babies as an outcome in births) but this means the contextual information of our data, e.g.
-whether or not it is spatial or temporal, is only of importance at two stages of the data analysis pipeline.
-It is first important when we *boil it down* to our data through *observation* and again when we *map our statistical information back to its real world information* to *communicate* it.
-The only way the context uniquely contributes to uncertainty is in the "boil down" stage, however spatial or temporal uncertainty never refers to that specific cause of uncertainty.
+Similar to the confusing definition of uncertainty, the constant conflation of statistical and contextual information also creates a wide range of problems.
 @Kinkeldey2014 discussed how uncertainty  can be represented by three components attribute (what) position (where) and temporal (when) and that studied typically deal with uncertainty around attribute but rarely position and time, however it is never specified what considerations should cause attribute uncertainty to be different to position or temporal uncertainty.
@@
 
-It is typically in reference to measurement error, data aggregation, statistical modelling, or other steps that occur *after* the boil down stage, and are therefore not unique to spatial or temporal data.
-While the spatial or temporal features *may* become relevant at the final communication stage, however this has little do with quantifying uncertainty, and the difficulties in communicating a complicated statistical model through a complicated map is already well discussed in the visualisation literature, and in later sections we will discuss how this concept is related to uncertainty visualisation.
-
-![Data pipeline with visualisations to show why we are ignoring contextual information for the time being.
-The visualisations that depict the data are similar to typical statistical graphics.
-The purpose of the spatial information is to provide context to the information we extract from the statistical aspect of the graphic.](spatialinfo.jpeg){#fig-spatial}
-
-This does not mean we are saying spatial and temporal information is never of importance.
-Spatial information is *highly relevant* when we translate from the abstract domain back into real world for communication.
-This position is best explained by @Leland2005 in *The Grammar of Graphics*:
-
-> "Geography is anchored in real space-time and statistics in abstract dimensions.
-This is a distinction along a continuum rather than a sharp break...
-but this difference in focus clearly means that a system optimized to handle geography will not be graceful when dealing with statistical graphics."
-
 This paper focuses on uncertainty, specifically uncertainty that is still in the the abstract *statistics* domain.
-Since we are still in the abstract domain, things that are anchored in real space-time create confusion that is evident in the literature.
 Spatial and temporal uncertainty papers limit how they communicate their uncertainty because they are considering what can be visualised inside the context provided by maps and, not what they should *try* to visualise based on the uncertainty abstraction.
-Additionally, since uncertainty is defined in the abstract domain, of the issues faced by spatial uncertainty visualisations are likely shared by other graphics that only differ in contextual information.
+When we view a spatial-temporal visualisation through the lense of trying to visualise the distribution of $P(X|longitude, latitude, time)$ there is no reason this is any different from any other high dimensional visualisation problem.
 
@@
 ```
+
+The focus on contextual information has also created a rift in exactly *why* uncertainty visualisation is a field of interest. The two most common justifications are:
+1) Uncertainty is fundamentally differently to other variables due to psychological heuristics involved in the interpretation of uncertainty (e.g. risk aversion).
+2) Uncertainty is not of interest in of itself and is typically layered on at the end of a the visualisation pipeline, so uncertainty visualisation is the field of adding error information into already established graphics and *thus* uncertainty information is a high dimensional visualisation problem.
+Papers will often discuss uncertainty in relation to one of these motivating reasons, the evaluation experiments motivated by (1) often perform different visualisations of PDFs, while the papers motivated by (2) will focus on trying to impute uncertainty as error within one of the existing channels, such as colour (*Cite Gap 11*).
+These dual motivations that co-exist with uncertainty visualisation authors rarely mentioning the others, create confusion in the field as to its purpose.
+@Kinkeldey2014 also identified this issue when they highlighted that the literature makes it unclear if uncertainty is a variable in of itself, something that should be interpreted with the main variable, or if it is metadata.
+
+For these reasons, the implication that uncertainty is not inference related, the barriers that cause authors to ignore existing visualisation methods, and the confusion created in the motivation of uncertainty visualization as a whole, we do not believe it is fruitful to continue separating the field of uncertainty visualisation according to its contextual information.
+Therefore, the rest of this review will not separate uncertainty visualisations based on this information.
 
@@
 The class of "uncertainty visualisation" does not seem to have any rules for what belongs in the group and what makes it different to a "typical" visualisation is rarely discussed.
-Instead one of two general motivations for the field are provided in the subtext.
-1) Uncertainty is fundamentally differently to other variables due to psychological heuristics involved in the interpretation of uncertainty (e.g. risk aversion).
-2) Uncertainty is not of interest in of itself and is typically layered on at the end of a the visualisation pipeline, so uncertainty visualisation is the field of adding error information into already established graphics.
-Neither of these motivations are remotely relevant when we are evaluating the simple task of value extraction.
+*removed bit about two motivations* Neither of these motivations are remotely relevant when we are evaluating the simple task of value extraction.
 The psychological effects of uncertainty are only relevant for decision making and have nothing to do with value abstraction, and the information overload issue is not specific to uncertainty and is simply an example of a larger dimensionality issue.
@@
 This perspective is not quite right and the literature seems to be about 50/50 on whether uncertainty is a psychological or dimensionality problem.
-@Kinkeldey2014 also identified this issue when they highlighted that the literature makes it unclear if uncertainty is a variable in of itself, something that should be interpreted with the main variable, or if it is metadata.
 It is important to keep in mind that just because information is in a graphic, that does not mean it will be "seen".

### 2024-03-18 writingdumpster.qmd
rewrote contextual information section

@@
 Cite Gap 10: Uncertainty Visualisation Taxonomies
+
+Cite Gap 11: Papers will often discuss uncertainty in relation to one of these motivating reasons, the evaluation experiments motivated by (1) often perform different visualisations of PDFs, while the papers motivated by (2) will focus on trying to impute uncertainty as error within one of the existing channels, such as colour
 

### 2024-03-18 paper.qmd
finished information asymetry examples

@@
 ---
+# Current Todo List
+This is a list of tasks that need to be completed before I consider the paper complete. Please keep this in mind when reading.
 
@@
 
-## Mistakes when we misunderstand uncertainty
+## Mistakes made when we misunderstand uncertainty
 The previous section established that uncertainty is a by product of the assumptions we make when we perform inference, additionally, our quantification of uncertainty is largely task dependent, therefore the uncertainty associated with *one* distribution (which we can also view as a task or goal dependence).
@@
 @Potter2010 aimed to create a summary plot that "concisely presented data with uncertainty information" to create an exploratory visualisation tool that visualised uncertainty.
+Exploratory visualisations often differ from descriptive statistics because the explicit statistic we are drawing inference on is less explicit.
+A versatile visualisation such as a scatter plot allows for a viewer to consider several hypothesis at once, each of which may have its own associated uncertainty.
+@Hullman2021 argue that there is no such thing as a "model-free" visualisation, therefore visualisation require robust visualisations of uncertainty as we are always performing inference.
+While we agree with this sentiment, it is clear that uncertainty *cannot* be defined without a specific motivating question, and therefore trying to include uncertainty in an exploratory visualisation stage (which by definition does not have a hypothesis) is not possible.
 
@@
 
-## Mathematically incomparable objects
+## Mistakes made when we misunderstand uncertainty visualisation
+In the previous section, we highlighted several problematic experimental designs or pieces of research that were the result of misunderstandings around the definition of uncertainty.
+Here, we will discuss the issues that arise due to a misunderstanding of "uncertainty visualisation".
+The two primary problems are the regular comparisons of mathematically incomparable plots, and reproducing results that are already widely understood in the larger information visualisation literature.
 
-> Edit notes: The "in order to visualise uncertainty you need to quantify it and the quantification does not align with the definition" point could also work here since this section is now about value extraction.
+### Information asymetry in evaluated plots
+The erroneous notion that "uncertainty visualisations" are a class of visualisations that depict "uncertainty" (something that is undefined) has allowed authors to compare plots that depict wildly different information with little consideration of the value of this practice.
 
-Visual inference can be seen as a process that combines multiple fields, from mathematics to psychology, to convey meaning.
-Unfortunately which tasks belong to mathematics, computer science, or psychology is poorly defined.
-The uncertainty visualisation taxonomies make the blur between these fields.
-A visualisation is, in a lot of ways, a statistic, or at the very least statistics are calculated prior to the visualisation being calculated, however this may be invisible to the user.
-A density plot does not just "appear" from the data, usually a smoothing function generates an estimated density function, which is then depicted on a plot using a line, which is then converted into information in your head using perceptual tasks and heuristics.
-These overlapping but distinct fields create a high burden of spanning expertise to understand if two visualisation are equivalent in each field.
-This is largely caused by the concept of "information" differing dramatically between fields.
-According to mathematics, a sufficient statistic for a parameter contains the same amount of information as an entire sample.
-Therefore, if you are trying to estimate a population mean, the sample mean and the entire data set are both equivalent, however a similar statistic, such as the median, is not and will become a worse and worse approximation for the mean the more skewed your underlying distribution in.
-Computer science considered two pieces of information to be equivalent if they come from the same data.
-This definition of information is much looser and is contained within the mathematical definition of information.
-Thanksfully, we can turn to the grammar of graphics to try and organise and untangle the information in our graphics.
+Visualisation authors are almost unanimous in commenting that the "information" in two plots must be the same in order for the visual techniques to be compared [@Cleveland1984; @Kinkeldey2014] (*Cite gap 12*).
+However, what makes two visualisations equal in "information" is not consistent and is regularly mentioned without it being clear exactly what this information is.
+This is easily illustrated by the wide array of comments made by visualisation authors when they explain why two representations are being compared in their evaluation study.
+@Ibrekk1987 displayed 6 PDFs in their experiment because "formally equivalent representations are often not psychologically equivalent", but do not explain why what makes a representation "formally equivalent" or why only one representation was presented for the 95% CI, box plot, and CDF.
+@Hofman2020 comments that "theoretically" the sampling distribution of the mean and the prediction interval of a new observation are equal "so long as one knows the sample size", but does not recognise the several assumptions and statistical knowledge that is required to compare the two.
+*Other examples* (*Cite Gap 13*).
+@Kinkeldey2014 adopts an existing definition also that suggests two graphics are informationally equivalent if all the information in one plot is inferable from the other and vice-versa, but adds that two plots are computationally equivalent if that information can be extracted from both plots with similar easy and speed.
+It is clear this lacking definition of "information" in a visual is reminicent of our problems with the definition of "uncertainty".
 
-When creating a graphic there are several tasks that must be completed in a specific order, regardless of whether or not we are working with a uncertainty visualisation or not.
-@Leland2005 depicts these steps as a pipeline, shown in @fig-pipeline.
-This pipeline shows every step from data observation to rendering of a graphic, but we are going to focus on the tail end of the pipeline specifically on statistics, geometry, coordinates, and aesthetics.
+Unlike uncertainty, the concept of the "information" in a plot is not entirely avoided and works such as *The Grammar of Graphics* attempt to outline exactly what information is contained within a plot.
+When creating a graphic there are several tasks that must be completed in a specific order, regardless of whether or not we are working with a uncertainty visualisation or not, this pipeline is illustrated in @fig-pipeline, and contains all the steps of the grammar of graphics [@Leland2005].
+First the mathematical information in a plot is established in the "varset" setps, and then that mathematical information is visually represented, which is defined in the "graph" steps.
+Since two plots cannot be compared on their visual elements if they contain different information (because you will never be able to identify if the difference is because of informational asymetry or visual processing) we could understand two plots as being the same if their "varset" steps are identical.
+However because uncertainty is not defined mathematically, we cannot reliably add it in at the "statistic" stage.
+Additionally, the grammar of graphics was not designed to extend to "uncertainty visualisations" as @Leland2005 referred to uncertainty as "semantics", so concepts that are common to uncertainty mathematically, such as resampling do not quite fit within the framework.
+For example, if we have one set of data and two graphics, where one graphic represents the data as a series of points, and the the other depicts a smoothed density using a line.
+These two plots differ in *both* the statistic and geometry stages of the grammar of graphics, however as n (the number of observations in our sample) gets larger, these two plots will become visually indistinguishable (depending on the collision modifier used).
+This result aligns nicely with large sample theory, because as n increases the data and mass should become mathematically indistinguishable, and many uncertainty visualisation experiments notice an interaction between sample size and graphic effectiveness so this result is likely also practically true [@Kale2018; @Newburger2022; @Hofmann2012] but despite the increasing similarity in what is actually depicted in the graphic, the sample and mass visualisations would remain different visualisations within the grammar of graphics framework.
+Additionally, authors rarely attribute this visual version of large sample theory to results where it is clearly applicable.
+@Kale2018 found that users were more sensitive to the underlying trend when shown the HOPs plot over the static outcoms plot, this effect went away when the speed of the animation increased, indicating that performance difference might be due to overplotting in the static case and a smaller sample size in the static condition would produce the same result.
+The issue of blurred elements of the grammar of graphics in the case of distribution visualisations does not end here.
+Visualisations that depict mass such as a confidence interval, a boxplot, a PDF, a letterbox plot, a violin plot, a histogram, a quantile dot plot, etc, all show some version of the mass of a variable at different resolutions, but whether these graphics differ in something as low down in the grammar pipeline as the variable stage, or later at the geometry stage is unclear.
 
-> Edit notes: Grammar of graphics is just a placeholder, Di suggested to replace this with my own pipeline visualisation based on some data wrangling papers.
+![The grammar of graphics data analysis pipeline from @Leland2005.
+the "varset" stages are the mathematical components of the graphic, the mathematical components define the "information" in the graph, while the "graph" stages are the visual elements that determine the "computational" cost of the graph.](grammarofgraphicspipeline.png){#fig-pipeline}
 
-![The grammar of graphics data analysis pipeline](grammarofgraphicspipeline.png){#fig-pipeline}
+Alternative information consideration within well defined frameworks may also be worthwhile to consider.
+@wu2023rational tried to eliminate information asymetry by esentially looking at a visualisation and identifying what information can be extracted and then using a "rational agent benchmark" to determine how a rational person would use that information for a decision making task.
+While this solution is interesting, it may have different results depending on who is applying it as they may use different methods or notice different elements when extracting information from a graphic.
+The mathematically defined concept of sufficient statistics is also scarcely mentioned likely because it is specific to a particular ground truth statistic, however uncertainty *is* specific to a particular statistic, so it could be a useful avenue for understanding when two plots differ in their ability to answer a particular question.
 
-In order to visually communicate uncertainty we must:  
-1) Mathematically quantify uncertainty in some way.   
-2) Visually express this quantification using some geometry, axis, aesthetic, and rendering.  
-3) Interpret the visual expression to extract the information using perceptual tasks.
+Ultimately, this all points to confusion as to when two visualisation contain information asymmetry, a problem that becomes more pronounced in uncertainty visualisation.
+@Ibrekk1987 compared a 6 visualisations of a PDF,  a box plot, a CDF and a mean with a confidence interval on the basis that all of them were "uncertainty visualisations".
+They found that people are better at extracting the mean from a plot when they are shown a plot that contains a mean with a confidence interval than when they are shown a PDF, a box plot, or a CDF (where the mean could not be read off the plot).
+@Hofman2020 and @Zhang2022 compared prediction and sampling distributions because they are both "uncertainty" that is typically depicted around the mean.
+They found that people are better at answering questions about a prediction interval when shown a prediction interval instead of a sampling distribution.
+@Hullman2015 compared the static error bars and violin plots of the marginal distributions of two variables ($A$ and $B$) to an animated plot that depicted a single outcome of the joint distribution of $A$ and $B$ in each frame of an animation, a comparison that was justified because all plots were "uncertainty visualisations".
+They found that the plot that depicted outcomes from $P(A, B)$ where the distance between the two outcomes was equivalent to $a-b$ was better at answering the question "What is $P(B<A)$" than violin or error bar plots depicting $P(A)$ and $P(B)$ separately.
+*Other examples* (*Cite Gap 14*).
 
-Each of these steps have a large amount of research independent of the "uncertainty visualisation" subfield.
-There are numerous ways to depict uncertainty mathematically.
-Examples of mathematically defined objects that are used to express uncertainty inlude: confidence intervals, sampling distributions, bias, variance, prediction error, value ranges, quantiles etc.
-These definitions all exist in mathematics independent of the field of visualisation.
-Once we have some quantification of uncertainty, we need to express it visually.
-There are numerous options in combinations of geometry, axis, and aesthetics and modern graphical software will allow most of these combinations.
-There is also a large amount of research in information visualisation and psychology into how the choice of visual expression impacts a persons ability or ease with which they extract relevant information.
-In this sense, once we have a mathematical expression of uncertainty, the visualisation of that uncertainty is identical to the visualisation process of any other variable, or at least this is how it is treated in the uncertainty visualisation experiments.
+This collection of examples starts to paint a pretty clear picture.
+Visualisations with rather shocking information asymetry are regularly compared because they are both "uncertainty visualisations".
+This results in a series of experiments where the visual aspects of the graphic are not even required to anticipate the experimental results.
+This issue likely results in the lacking mathematical definition preventing us from defining and quantifying "uncertainty" mathematically and then including this quantification in a well defined framework such as the grammar of graphics.
 
-There are several frameworks that the uncertainty visualsiation problem is approached from.
-One approaches it from the concept of risk, and tries to answer how to best express uncertainty in a visualisation such that the viewer can acurately extract and interpret the risks involved.
-Another approach is to consider uncertainty visualsiation as a specific sub-field of high-dimensional data visualsiation, and therefore uncertainty is just another variable that needs to be expressed.
+### Repeating perceptual task experiments
+Given the previous section, one might consider problems in uncertainty visualisation to be fixed if we could restrict evaluation experiments to cases where the information in each graphic is almost identical.
+Unfortunately this still often results in uninteresting or
 
-Cite gap __: represent different maths objects in visualisations
-- @wu2023rational recognised it is difficult to understand when differences in performance stem from information asymmetry.
-
-## Uncertainty as signal
-
-This distinction is important because lower levels of the pipeline influence higher levels.
-Visualisation authors are almost unanimous in commenting that the information in two plots must be the same in order for the visual techniques to be compared [@Cleveland1984; @Kinkeldey2014] *this citation has way more, check confirmation*.
-@Kinkeldey2014 adopts a definition from the 80s that suggests two graphics are informationally equivalent if all the information in one plot is inferable from the other and vice versa, and computationally equivalent if that information can be extracted from both plots with similar easy and speed.
-This definition illustrates a similar idea to the grammar of graphics pipeline, shown in @fig-pipeline.
-First the mathematical information in a plot is established, and then that mathematical information is visually represented.
-Two plots that depict mathematical information that cannot be inferred from each other, cannot be compared on their visual elements.
-To prevent this requirement from being *too* limiting, when using a ground truth statistic we can consider two plots to be informationally equivalent if both visualisations depict a sufficient statistic.
-For example, a depiction of a sample and its average are equivalent if we are trying to infer the population mean.
-This does not mean they are computationally equivalent.
-Additionally, two plots cannot be compared on their visual elements if they contain different information as well.
-The reason for this is obvious.
-If you compare two graphics and one depicts the mean and the other depicts a median and you ask viewers to extract the median value, the graphic that *actually* depicts that statistic will perform better.
-People cannot percieve information that is not depicted.
+Once we have a mathematical expression of uncertainty, the visualisation of that uncertainty is identical to the visualisation process of any other variable, or at least this is how it is treated in the uncertainty visualisation experiments.
 
@@
 
-# Noise and Signal
+# Uncertainty as noise and Uncertainty as signal
 
@@
 The difficulty with resampling methods is that they are typically conflated with animation even though this does not necessarily need to be true, as can be seen in the @fig-pixelmap and the New York Times class mobility figure, shown in @fig-nyt, an animated version of which can be found [here](https://www.nytimes.com/interactive/2018/03/19/upshot/race-class-white-and-black-men.html).
-Even though @Kale2018 found that users were more sensitive to the underlying trend when shown the HOPs plot over the static outcoms plot, this effect went away when the speed of the animation increased, indicating that performance difference might be due to overplotting in the static case and a smaller sample size in the static condition would produce the same result.
 Therefore, so long as the sample size of the number of outcomes is kept to a level where individual samples are still visible, there is no evidence that an outcome plot *needs* to be animated.
@@
 The concept of uncertainty should also be formalised within the grammar of graphics.
-@Leland2005 referred to uncertainty as "semantics", however the depth of integration required to understand problems in uncertainty visualisation suggest they *need* to be formally placed within the framework.
-Displaying a distribution is not as simple as displaying the data.
-For example, the distinction between a sample and a smothed density function is not cut and dry.
-One may view them both as the same data but one is represented as a point, the other is binned according to some function and then represented as a line.
-However, this distinction blurs as n gets larger and larger and, depending on the collision modifier used, depictions of the sample will appear identical to depictions of the mass.
-This is obviously true due to large sample theory, the backbone of statistics.
-The issue of blurred elements of the grammar of graphics in the case of distribution visualisations does not end here.
-Visualisations that depict mass such as a confidence interval, a boxplot, a PDF, a letterbox plot, a violin plot, a histogram, a quantile dot plot, etc, all show some version of the mass of a variable at different resolutions, but whether these veriables differ in something as low down in the grammar pipeline as the variable stage, or later at the geometry stage is unclear.
-The lack of formalisation in this way allows authors to compare graphics that *seem* to be similar mathematically but are actually conveying very different information.
 Not only would a formalisation of uncertainty visualisation within the grammar of graphics allow us to iron out some of these confusions, but they would make it easier to understand when existing visualisation methods apply and can be used to explain our results.

### 2024-03-18 paper.qmd
updated some of the perceptual tasks section

@@
 ### Information asymetry in evaluated plots
-The erroneous notion that "uncertainty visualisations" are a class of visualisations that depict "uncertainty" (something that is undefined) has allowed authors to compare plots that depict wildly different information with little consideration of the value of this practice.
-
 Visualisation authors are almost unanimous in commenting that the "information" in two plots must be the same in order for the visual techniques to be compared [@Cleveland1984; @Kinkeldey2014] (*Cite gap 12*).
@@
 Given the previous section, one might consider problems in uncertainty visualisation to be fixed if we could restrict evaluation experiments to cases where the information in each graphic is almost identical.
-Unfortunately this still often results in uninteresting or
+Unfortunately this still often results in uninteresting or obvious results.
 
-Once we have a mathematical expression of uncertainty, the visualisation of that uncertainty is identical to the visualisation process of any other variable, or at least this is how it is treated in the uncertainty visualisation experiments.
+If two graphics are visually different but identical in the information they contain, they must differ in how that information is depicted. 
+Once we have a mathematical expression of uncertainty, the visualisation of that uncertainty is theoretically identical to the visualisation process of any other variable.
+For simple tasks such as value extraction, there is a hierarchy to perceptual tasks where extracting visual information in some forms is easier than others.
+The hierarchy was originally established 40 years ago by @Cleveland1984, below is an updated version summarised by @Vanderplas2020:
 
-Computational equivalency in visualisation is also something that has been well researched.
-For simple tasks such as value extraction, there is a hierarchy to perceptual tasks where extracting visual information in some forms is easier than others.
-The hierarchy established 40 years ago by @Cleveland1984 is:
+1) Position along a common scale. 
+2) Position along a non-aligned scale. 
+3) Length, direction, angle, slope
+4) Area
+5) Volume, density, curvature
+6) Shading, color saturation, color hue
+7) Discriminable shape
+8) Indiscriminable shape
 
-1) position along a common scale. 
-2) position along a non-aligned scale. 
-3) length. 
-4) angle/slope. 
-5) area. 
-6) volume. 
-7) colour
+This hierarchy is a good general rule, however it can change from person to person [@Davis2022] Additionally, there are other graphical rules to consider such as guestalt heuristics, broader methods of extraction, and attention principles [@Vanderplas2020].
+These established visualisation concepts allow us to anticipate the ease with which certain pieces of information will be extracted from a plot.
+We can use these concepts to understand the computational complexity of a graphic.
+A bizarre feature of the uncertainty visualisation literature is that it does not work to build upon these existing principles or identify the ways in which uncertainty visualisations may diverge from these rules.
+These building block concepts of visualisation are seldom mentioned.
 
-> Edit notes: need to add the notes from the susan lit review that has the more recent perceptual task list
+It is difficult to find examples of uncertainty visualisation experiments where the plots do contain the same information, however when they do, the results align with existing information visualisation research.
 
-This hierarchy has had additions and adjustments over the years.
-For example, this hierarchy is a general rule, but the hierarchy can change from individual to individual.
-Additionally, there are other graphical rules to consider such as guestalt heuristics, broader methods of extraction, and attention principles.
-These established visualisation concepts allow us to anticipate the ease with which certain pieces of information will be extracted from a plot.
+*Other examples* (*Cite Gap 15*).
 
-A bizarre feature of uncertainty visualisation literature is that it does not work to build upon these existing principle or identify the ways in which uncertainty visualisations may diverge from these rules, but rather it seems to exist with complete indifference to the research in information visualisation at large.
-The most common uncertainty visualisation evaluation study asks participants to perform some value abstraction either explicitly by asking participants to report values or vaguely by asking participants to find regions of least certainty, rank plots, report their confidence, and/or make decisions according to uncertainty information.
-Experimenters then compare these results based on the participants accuracy, preference, confidence, and/or speed @Hullman2016.
-Most commonly, these value abstraction papers compare two different visualisations of uncertainty on a set of these questions, however, since there is no rules for what is or is not an uncertainty visualisation the graphics that are compared often differ in both computation *and* information.
-@Hullman2018 asks participants questions that would require a prediction interval but provided a sampling distribution and argued that the two could be considered equivalent so long as participants understood the difference between a confidence interval and a prediction interval, knew the sample size, and knew the formula to convert the confidence interval to a prediction interval.
-The authors themselves note that this is rediculous, which begs the question.
-Uncertainty visualisaiton papers only seem to require that the plots depict "uncertainty" which has already been established to be poorly defined.
-The complete lack of formality is easy to see if we actually plot the visualisations compared across evaluation experiments according to the grammar of graphics *in the figure below*:
+If plots that have different information are compared, whichever directly depicts the ground truth statistic (or has information that can easily be convered to that ground truth statistic) will perform the best.
+If plots that have the same information are compared, the plot that requires a perceptual task that is higher on the hierarchy will perform the best.
+Therefore, when it comes to any task that involves value abstraction, many uncertainty visualisation evaluation studies are simple repeating already established results in statistics or visualisation.
 
-> Edit note: need to include a graphic that illustrates the wide spread of mathematical information and visual techniques that are visualised in these experiments.
-
-Reading the uncertainty visalisation literature will make the importance of information and computation incredibly salient.
-If plots that have different information are compared, whichever directly depicts the ground truth statistic (or has information that can easily be convered to that ground truth statistic) will perform the best *cite*.
-If plots that have the same information are compared, the plot that requires a perceptual task that is higher on the hierarchy will perform the best *cite*.
-Therefore, when it comes to any task that involves value abstraction, many uncertainty visualisation evaluation studies are simple repeating already established results in visualisation.
-
-Several uncertainty visualisation literature reviews have commented on the immense difficulty of combining results across experiments, but none have been able to pinpoint the reason.
-@Kinkeldey2014 suggested a task based taxonomy that organised the plots by the question or purpose of the visualisation, and even suggested that communication tasks (such as value retrival of uncertainty values) should just follow traditional cartography rules, @Spiegelhalter2017 simply ignored the existence of uncertainty visualisation as a subfield and just suggested we follow established rules within uncertainty visualisation, @Hullman2016 said that commented on the large amount of noise generated from uncertainty visualisation evaluation methods that led to a difficulty in combining results, @Griethe2006 did not meaningfully combine the work and instead listed findings in the field, but also did not comment on the lack of generalisable results.
-While we largely agree with the difficulties of these authors, we do not agree that the solution is a task based taxonomy or better questions.
-While we agree the lack of attention to tasks and appropriate questions has generated problems in the field, there is a much more pressing issue, which is that uncertainty visualisation authors don't seem to understand *why* uncertainty visualisation different to typical visualisations.
-
-The class of "uncertainty visualisation" does not seem to have any rules for what belongs in the group and what makes it different to a "typical" visualisation is rarely discussed.
-*removed bit about two motivations* Neither of these motivations are remotely relevant when we are evaluating the simple task of value extraction.
+Therefore, these explanations for the difference fall short of explaining *why* uncertainty should be treated any differently than any other variable.
 The psychological effects of uncertainty are only relevant for decision making and have nothing to do with value abstraction, and the information overload issue is not specific to uncertainty and is simply an example of a larger dimensionality issue.
-Therefore, these explanations for the difference fall short of explaining *why* uncertainty should be treated any differently than any other variable.
-
-There is also the common view that uncertainty as a dimensionality problem.
-This perspective is not quite right and the literature seems to be about 50/50 on whether uncertainty is a psychological or dimensionality problem.
-It is important to keep in mind that just because information is in a graphic, that does not mean it will be "seen".
-The phenomena of inattentional blindness shows that there is no perception without attention and it is powerful enough that participants can fail to be aware of random objects appearing on a screen or a gorilla walking through a basketball game [@simons1999gorillas; @mack2003inattentional].
-Therefore if information is mapped to graphical elements that are so low on hierarchy they can be ignored, they might as well not be there at all.
-The people who don't plot uncertainty because they think its unimportant and the people who relegate uncertainty to the lowest ranking aesthetic on the list are both saying the same thing, "I think the uncertainty is unimportant".
-Including uncertainty is worth very little if no attention is left to see it.
-This does not mean we cannot put a *large* amount of information in a graphic.
-Glyph maps can be used to depict the multi-dimensional information in spatial-temporal data by mapping line plots to map locations, but we still need to decide what information is important.
-The map can either present trends in the global or local variance depending on whether or not the line plot is scaled globally or locally [@Wickham2012], however smaller details are almost impossible to convey.
-No matter how much information we try and put in a plot, there will always be only a handful of key takeaways.
-
-This issue has come up in many different forms in the uncertainty visualisation literature.
-Typically as a "task dependence" on the performance of uncertainty visualisations.
-It is repeatedly identified as a problem in previous reviews of the uncertainty visualisation literature [@Kinkeldey2014; @Hullman2016] as well as across many sub domains and applications [@Wallsten1997; @Fischhoff2014; @Meng2021; @Amar2005].
-The fact that this conclusion is repeatedly reached shows both the importance and the lack of acknowledgement this concept receives.
-@Fischhoff2014  takes this approach by considerting how decisions with different levels of flexibility should have different uncertainty visualisations.
-If the user has no choice they should simply be made aware of the risk, if they have several options, the user needs to see the uncertainty in each, if they need to come up with the options themselves, they need to see how things work and how the system fails.
-@Kinkeldey2014 suggested that the nature of the task plays an important role and believed the task dependence for uncertainty visualisation could explain some of the inconsistency in outcomes.
 
@@
 
-## Fixes within "uncertainty visualisation"
-Attempts to rectify these conflicts always flutter about the surface and do not seem to grasp the core of this problem.
-For example, @wu2023rational noticed a series of papers with poorly defined tasks and suggested comparing study participants to a "rational agent" to help authors understand and remedy this issue.
-@Kim2019 believed an understanding of participants prior beliefs could assist in interpreting uncertainty visualisation results that tend to be incredibly noisy.
-What these suggestions fail to understand is that the entire practice of identifying "true" facts about "unertainty visualisation" is futile.
-It is impossible to make a statement about the properties of an "uncertainty visualisation" in the same way it is impossible to comment on $\frac 0 0 = 1$.
-Division by zero is undefined, so the statement is ill formed, similarly, "uncertainty" is not defined, so statements *about* "uncertainty" including statements about "uncertainty visualisation" are ill formed.
-Ill-formed statements can neither be true nor false and cannot be part of a logical argument, so experiments and research that build upon this ill-formed concept are neither true nor false, they simply exist.
-Someone who is less familiar with the rules of logical argument may believe this understanding of the problem to an insufferable nitpick that creates impossible to live up to standards, however the consequences of this definitional issue are *very* real and costly.
-The studies that investigate "uncertainty visualisation" may be identifying *something*, however there is good reason to believe that *something* is neither interesting nor does can it contribute to a larger theory of visualisation.
-This is something we illustrate in the following this section.
+# Uncertainty as noise and Uncertainty as signal
+In this section we will discuss how *most* attempts to measure the effects of uncertainty are either value extraction experiments with an extra step, or are measuring uncertainty using a proxy that does not actually represent uncertainty.
+In this section we will talk about the issues of trying to measure uncertainty, the real purpose of uncertainty visualisation, and suggestions in how we can better measure uncertainty in our visualisation experiments.
 
-# Uncertainty as noise and Uncertainty as signal
+## Most experiments are value extraction experiments with extra steps
+Those with a keen eye might have read the previous sections and noticed our statements were only relevant to evaluation experiments where participants must extract a particular value from a plot.
+Of course, these experiments are not reflective of "real world" uses of uncertainty visualisations, we do not want to visualise uncertainty so that we can answer questions such as *what is the probability of even A*.
+In reality uncertainty is useful for decision making and eliciting trust from an audience.
+This section will explain why, despite appearing to be different, these experiments are actually still value extraction experiments and thus have the same pitfalls.
 
-## Types of Uncertainty Visualisation Papers
+### Types of Uncertainty Visualisation Papers
 
@@
 Depending on what I do with later sections, this might need a rewrite or to just be removed, since I'm not sure how relevant this is to the main ideas of the paper.
+
+The uncertainty visualisation evaluation study asks participants to perform some value abstraction either explicitly by asking participants to report values or vaguely by asking participants to find regions of least certainty, rank plots, report their confidence, and/or make decisions according to uncertainty information [@Hullman201].
+Experimenters then compare these results based on the participants accuracy, preference, confidence, and/or speed [@Hullman2016].
 

### 2024-03-18 writingdumpster.qmd
finished information asymetry examples
updated some of the perceptual tasks section

@@
 
-# Extra Section Cite Gaps
+Cite Gap 12: Visualisation authors are almost unanimous in commenting that the "information" in two plots must be the same in order for the visual techniques to be compared
+
+Cite Gap 13: Examples of what is "information" in a visualisation
+
+Cite Gap 14: Examples of mathematically obvious results
+
+Cite Gap 15: Examples of replicated perceptual task experiments
+
+## Extra Section Cite Gaps
 Cite Gap - New Section: Logical fallacy interpretation of mathematically unanswerable questions.
@@
 # Moved over notes
+@Kim2019 believed an understanding of participants prior beliefs could assist in interpreting uncertainty visualisation results that tend to be incredibly noisy.
 @Thomson2005 titled their paper "A typology for visualizing uncertainty" however it is a taxonomy for the specific types of uncertainty that we may want to visualise and a way to calculate that uncertainty.
@@
 # Main section outtakes
+There is also the common view that uncertainty as a dimensionality problem.
+This perspective is not quite right and the literature seems to be about 50/50 on whether uncertainty is a psychological or dimensionality problem.
+It is important to keep in mind that just because information is in a graphic, that does not mean it will be "seen".
+The phenomena of inattentional blindness shows that there is no perception without attention and it is powerful enough that participants can fail to be aware of random objects appearing on a screen or a gorilla walking through a basketball game [@simons1999gorillas; @mack2003inattentional].
+Therefore if information is mapped to graphical elements that are so low on hierarchy they can be ignored, they might as well not be there at all.
+The people who don't plot uncertainty because they think its unimportant and the people who relegate uncertainty to the lowest ranking aesthetic on the list are both saying the same thing, "I think the uncertainty is unimportant".
+Including uncertainty is worth very little if no attention is left to see it.
+This does not mean we cannot put a *large* amount of information in a graphic.
+Glyph maps can be used to depict the multi-dimensional information in spatial-temporal data by mapping line plots to map locations, but we still need to decide what information is important.
+The map can either present trends in the global or local variance depending on whether or not the line plot is scaled globally or locally [@Wickham2012], however smaller details are almost impossible to convey.
+No matter how much information we try and put in a plot, there will always be only a handful of key takeaways.
+
+This issue has come up in many different forms in the uncertainty visualisation literature.
+Typically as a "task dependence" on the performance of uncertainty visualisations.
+It is repeatedly identified as a problem in previous reviews of the uncertainty visualisation literature [@Kinkeldey2014; @Hullman2016] as well as across many sub domains and applications [@Wallsten1997; @Fischhoff2014; @Meng2021; @Amar2005].
+The fact that this conclusion is repeatedly reached shows both the importance and the lack of acknowledgement this concept receives.
+@Fischhoff2014  takes this approach by considerting how decisions with different levels of flexibility should have different uncertainty visualisations.
+If the user has no choice they should simply be made aware of the risk, if they have several options, the user needs to see the uncertainty in each, if they need to come up with the options themselves, they need to see how things work and how the system fails.
+@Kinkeldey2014 suggested that the nature of the task plays an important role and believed the task dependence for uncertainty visualisation could explain some of the inconsistency in outcomes.
+
+Several uncertainty visualisation literature reviews have commented on the immense difficulty of combining results across experiments, but none have been able to pinpoint the reason.
+@Kinkeldey2014 suggested a task based taxonomy that organised the plots by the question or purpose of the visualisation, and even suggested that communication tasks (such as value retrival of uncertainty values) should just follow traditional cartography rules, @Spiegelhalter2017 simply ignored the existence of uncertainty visualisation as a subfield and just suggested we follow established rules within uncertainty visualisation, @Hullman2016 said that commented on the large amount of noise generated from uncertainty visualisation evaluation methods that led to a difficulty in combining results, @Griethe2006 did not meaningfully combine the work and instead listed findings in the field, but also did not comment on the lack of generalisable results.
+While we largely agree with the difficulties of these authors, we do not agree that the solution is a task based taxonomy or better questions.
+While we agree the lack of attention to tasks and appropriate questions has generated problems in the field, there is a much more pressing issue, which is that uncertainty visualisation authors don't seem to understand *why* uncertainty visualisation different to typical visualisations.
+
+Attempts to rectify these conflicts always flutter about the surface and do not seem to grasp the core of this problem.
+What these suggestions fail to understand is that the entire practice of identifying "true" facts about "unertainty visualisation" is futile.
+It is impossible to make a statement about the properties of an "uncertainty visualisation" in the same way it is impossible to comment on $\frac 0 0 = 1$.
+Division by zero is undefined, so the statement is ill formed, similarly, "uncertainty" is not defined, so statements *about* "uncertainty" including statements about "uncertainty visualisation" are ill formed.
+Ill-formed statements can neither be true nor false and cannot be part of a logical argument, so experiments and research that build upon this ill-formed concept are neither true nor false, they simply exist.
+Someone who is less familiar with the rules of logical argument may believe this understanding of the problem to an insufferable nitpick that creates impossible to live up to standards, however the consequences of this definitional issue are *very* real and costly.
+The studies that investigate "uncertainty visualisation" may be identifying *something*, however there is good reason to believe that *something* is neither interesting nor does can it contribute to a larger theory of visualisation.
+This is something we illustrate in the following this section.
+
 - The process of working out whether or not a visualisation shows a piece of information, and if it does, what perceptual task it is mapped to and what assumptions or heuristics need to be used, is *incredibly* time consuming.

### 2024-03-19 paper.qmd
finished perceptual task section

@@
 It is difficult to find examples of uncertainty visualisation experiments where the plots do contain the same information, however when they do, the results align with existing information visualisation research.
-
+Technically, a PDF and a mean with confidence intervals both have enough information to extract the mean of the distribution, however they both have a very different computational cost.
+To extract the mean using a PDF, a participant would need to identify the point along the x axis that splits the area under the curve in half.
+If a participant is provided with a mean with a confidence interval, extracting the average is a simple task of reading the position on an aligned scale.
+@Ibrekk1987 found that when asking participants for the "best estimate" (which they thought should be interpreted as the mean of the distribution) of a skewed distribution, participants provided the mean when given a mean with confidence intervals and the mode when given a PDF [@Ibrekk1987].
+Similar results to this occur over and over again in the uncertainty visualisation literature.
+@Gschwandtnei2016 found that visualisations where the start time of an interval could literally be read off the plot (error bars, centered error bars, and ambiguation) performed better than the plots (accumulated probability, gradient, and violin) where the start time involved some guesswork because the drop off was gradual.
+@Cheong2016 found that participants were better at answering questions when they were explicitly given the relevant probability in text rather than when they needed to read it off a map.
 *Other examples* (*Cite Gap 15*).
 
-If plots that have different information are compared, whichever directly depicts the ground truth statistic (or has information that can easily be convered to that ground truth statistic) will perform the best.
-If plots that have the same information are compared, the plot that requires a perceptual task that is higher on the hierarchy will perform the best.
-Therefore, when it comes to any task that involves value abstraction, many uncertainty visualisation evaluation studies are simple repeating already established results in statistics or visualisation.
+These results show that uncertainty is not technically different to any other variable.
+When trying to anticipate the results of these studies, we can use the same principles of information equivalence and difficulty of relevant mental tasks to understand which plots will outperform others.
+This leads us to wonder...
+why is uncertainty visualisation a field at all?
 
-Therefore, these explanations for the difference fall short of explaining *why* uncertainty should be treated any differently than any other variable.
-The psychological effects of uncertainty are only relevant for decision making and have nothing to do with value abstraction, and the information overload issue is not specific to uncertainty and is simply an example of a larger dimensionality issue.
-
-Despite these results, it is clear that authors intuitively *feel* that uncertainty is different even though this difference has not been apparent in existing uncertainty visualisation experiments.
+It is clear that authors intuitively *feel* that uncertainty is different even though this difference has not been captured in existing uncertainty visualisation experiments.
+Some authors cited uncetainty was special because of its psychological qualities, but we have yet to see any evidence of that here.
 @Kinkeldey2014 also noticed that it is not clear whether or not uncertainty is just another variable, as many value extraction papers treats it as such in their experimental design and uncertainty may need to be in a variable class of its own.
@@
 There is nothing special about these statistics that should differentiate them from a mean or median and imply their visualisation should be "special" in a way that would warrant its own field.
-Especially because it seems as though (at least in the case of visual inference) they *are* no different from any other value.
-So, if we cannot see any difference to uncertainty information when it comes to value abstraction, *when* is uncertainty information different?
+Especially when the core take aways from uncertainty evaluation experiments do not seem to differ at all from the established results in information visualisation.
+So, what is it that makes authors believe uncertainty is different?
 
-# Uncertainty as noise and Uncertainty as signal
+# Are you measuing uncertainty as noise or signal?
+Evaluation experiments While uncertainty information *is* often used to improve decisions, it seems the main motivation in its communication is to offer transparency to the viewer about a statistical analysis.
+That is, it acts as a form of statistical "hedging" for signals found in an analysis.
+Interviews with experts in statistic back up this primary motivation, as ignoring uncertainty information of often expressed as being similar to fraud or lying and the goal of "improving decisions" is only seen as a secondary outcome of an appropriately hedged signal [@Hullman2020a; @Manski2020].
+This concept of uncertainty means that the work on communicating risk does not cover the field of uncertainty communication as a whole.
+It covers the specific case where the signal itself *is* uncertain, but this work does not naturally extend to the goal of hedging a signal (i.e.
+when we want to communicate uncertainty as noise).
+
+Unfortunately understanding the relevant noise seems to be its own problem, as statisticians and visualisation authors seem to be unaware or confused in how to calculate the uncertainty related to an analysis [@Hullman2020a].
+Additionally, communicating uncertainty such that it conveys the appropriate singal supression involves combining uncertainty from multiple different sources, something that is considered to be difficult by even experts in uncertainty communication [@Spiegelhalter2017].
+
 In this section we will discuss how *most* attempts to measure the effects of uncertainty are either value extraction experiments with an extra step, or are measuring uncertainty using a proxy that does not actually represent uncertainty.
@@
 
-## The current summaries of findings
 ## Uncertainty versus Risk
@@
 These motivations do not extend to the communication of uncertainty information more broadly.
-While uncertainty information *is* often used to improve decisions, it seems the main motivation in its communication is to offer transparency to the viewer about a statistical analysis.
-That is, it acts as a form of statistical "hedging" for signals found in an analysis.
-Interviews with experts in statistic back up this primary motivation, as ignoring uncertainty information of often expressed as being similar to fraud or lying and the goal of "improving decisions" is only seen as a secondary outcome of an appropriately hedged signal [@Hullman2020a; @Manski2020].
-This concept of uncertainty means that the work on communicating risk does not cover the field of uncertainty communication as a whole.
-It covers the specific case where the signal itself *is* uncertain, but this work does not naturally extend to the goal of hedging a signal (i.e.
-when we want to communicate uncertainty as noise).
-
-Unfortunately understanding the relevant noise seems to be its own problem, as statisticians and visualisation authors seem to be unaware or confused in how to calculate the uncertainty related to an analysis [@Hullman2020a].
-Additionally, communicating uncertainty such that it conveys the appropriate singal supression involves combining uncertainty from multiple different sources, something that is considered to be difficult by even experts in uncertainty communication [@Spiegelhalter2017].
 

### 2024-03-19 writingdumpster.qmd
finished perceptual task section

@@
 # Moved over notes
+- @Blenkinsop2000 asked different questions for each uncertainty visualisation, indicating they had some awareness that the question has a significant impact on whether or not the visualisation should be used.
+- @Gschwandtnei2016 When answering “What is the statistical probability (in %) that the interval has already ended at the marked point in time (red line)?” many answers gave the estimated probability to which the marked point in time falls into the interval.
+This is correct when asking for the probability to which the interval has already started at the marked position.
+However, in case of the probability to which the interval has already ended at the marked position, this answer is the inverted probability of what we were looking for.
+They believe this is their own mistake so they so they don’t present these results.
+
+@Sanyal2009 users performed significantly well using Glyph-size when the task was to search for locations of least uncertainty.
+However, both Glyph-color and Surface-color performed better than Glyph- size when the task was to search for locations of high uncertainty.
+We did not expect to find a significant difference between the two search tasks since both the tasks were designed to find extremes in the data
 @Kim2019 believed an understanding of participants prior beliefs could assist in interpreting uncertainty visualisation results that tend to be incredibly noisy.
@@
 # Main section outtakes
+These results provide some insights that applies to visualisation more broadly but nothing specific to uncertainty informaiton.
+It is clear that audiences primarily interact with visualisations through their "System 1" brain [@daniel2017thinking].
+This aligns with other authors comments that visualisation is primarily about "gists" [Spiegelhalter2017].
+In this sense, asking participants a question that requires them to shift to their "System 2" is entirely contrary to *why* we use visualisation.
+The exact reasons we describe visualisation to be so powerful, that they are able to communicate complicated information fast and efficiently *is a product of the system 1 brain* [@daniel2017thinking], however this means displaying information using visualisations have the same weakness' as system 1 thinking, and visualisation authors need to be aware of this.
+Asking participants to answer questions that require shifting to their system 2 brain is not in the spirit of visualisation, and does not reflect how visualisations will be used in practice, and will very likely result in incorrectly answered questions.
+There is a large number of examples where participants got "confused" but the problem is more likely that the question required participants to shift to their system 2 brain.
+
 There is also the common view that uncertainty as a dimensionality problem.

### 2024-03-19 paper.qmd
removed abstract

@@
   chunk_output_type: console
-abstract: In reviewing the literature of uncertainty visualisation
 ---
@@
 # Are you measuing uncertainty as noise or signal?
-Evaluation experiments While uncertainty information *is* often used to improve decisions, it seems the main motivation in its communication is to offer transparency to the viewer about a statistical analysis.
-That is, it acts as a form of statistical "hedging" for signals found in an analysis.
+Most evaluation experiments fail to capture uncertainty as we want to measure it, as noise. Some authors may read this and think that obviously, we dont use
+
+Uncertainty acts as a form of statistical "hedging" for signals found in an analysis.
 Interviews with experts in statistic back up this primary motivation, as ignoring uncertainty information of often expressed as being similar to fraud or lying and the goal of "improving decisions" is only seen as a secondary outcome of an appropriately hedged signal [@Hullman2020a; @Manski2020].
+
+While uncertainty information *is* often used to improve decisions, it seems the main motivation in its communication is to offer transparency to the viewer about a statistical analysis.
+
 This concept of uncertainty means that the work on communicating risk does not cover the field of uncertainty communication as a whole.
@@
 
-In this section we will discuss how *most* attempts to measure the effects of uncertainty are either value extraction experiments with an extra step, or are measuring uncertainty using a proxy that does not actually represent uncertainty.
-In this section we will talk about the issues of trying to measure uncertainty, the real purpose of uncertainty visualisation, and suggestions in how we can better measure uncertainty in our visualisation experiments.
-
-## Most experiments are value extraction experiments with extra steps
-Those with a keen eye might have read the previous sections and noticed our statements were only relevant to evaluation experiments where participants must extract a particular value from a plot.
-Of course, these experiments are not reflective of "real world" uses of uncertainty visualisations, we do not want to visualise uncertainty so that we can answer questions such as *what is the probability of even A*.
-In reality uncertainty is useful for decision making and eliciting trust from an audience.
-This section will explain why, despite appearing to be different, these experiments are actually still value extraction experiments and thus have the same pitfalls.
-
-### Types of Uncertainty Visualisation Papers
-
-> Edit notes: Originally I was planning to have the discussion following this section relate to these "types" of papers, but that didn't end up happening.
-Depending on what I do with later sections, this might need a rewrite or to just be removed, since I'm not sure how relevant this is to the main ideas of the paper.
-
-The uncertainty visualisation evaluation study asks participants to perform some value abstraction either explicitly by asking participants to report values or vaguely by asking participants to find regions of least certainty, rank plots, report their confidence, and/or make decisions according to uncertainty information [@Hullman201].
-Experimenters then compare these results based on the participants accuracy, preference, confidence, and/or speed [@Hullman2016].
-
-Uncertainty visualisation papers can be organised according to the *goal* of the experiment.
-Evaluation experiments are the standard rule for visualisations because the human brain is not as reliable as mathematical calculation.
-Therefore, user studies often aim to assess the limitations, biases, and heuristics of our mental calculator so that we can better understand the problems we may encounter when we plot our data.
-This is not to say any paper that suggests a visualisation without an evaluation experiment is completely lacking in justification, and there are many papers that suggest a novel visualisation without an evaluation study.
-Sometimes these papers are a preliminary step in finding a solution for common problems and intend to evaluate the visualsiation in later work.
-Until that later work is done, it is often difficult to accept one particular representation.
-While justification could come in the form of discussing established concepts in visualisation, such as the hierarchy of perceptual tasks, even these may need an evaluation study to back up their claims.
-
-The reasoning for this is obvious.
-There are a large number heuristics and biases that are not obvious to us when designing visualisation.
-Additionally, these heuristics and baises can change depending on the larger scope of the graphic and the population we are communicating with.
-Additionally, since there is often a miriad of ways to visualise any particular mathematical object, to adopt specific visualisation that is not already common practice needs reasonable justification, lest we run into a range of unforseen heuristic pitfalls.
-
-The remainder of section 4 will focus on the results and methods of uncertainty visualisation evaluation experiments.
-There are six main types of evaluation experiments in the uncertainty visualisation literature, they are:
-- Association perceptual tasks: These experiments identify perceptual tasks that are/could be associated with uncertainty, and test its association with uncertainty, check it's number of distinguishable levels, and/or check how quickly/accurately people can draw the uncertainty information from the graphic.
-- Value perceptual tasks: Participants are shown some depiction of uncertainty and are asked to extract a value, make a comparison, or do something simple to show that the graphic is readable.  
-- Heuristic Checks: These experiments are checking if a perceptual bias that has been identified through public use of a particular plot, other evaluation experiments, or gut feeling identified through public use of a plot or in a different.
--  Plot Comparisons: These papers describe an uncertainty visualisation evaluation experiment, where they compare two or more uncertainty visualisaitons on a specific set of questions.
-These questions can be about specific statistics (e.g.
-What is the probability that A>B?) or about a secondary goal such as trust or decision making (e.g.
-How much do you trust this estimate?).
-- Qualitative analysis: These experiments show participants an uncertainty visualisation for an open ended or specific task, and asks them to describe how the visualisation was used to come to a conclusion, or they are asked to describe what they see in the visualisation.
-- Method papers: These papers discuss issues with the evaluation methods of uncertainty visualisaiton and suggest changes or improvements, typically with results indicating the benefit of the change.
-
-These experiments are not mutually exclusive within a paper, and a single study will often contain several experiments, usually from several of these categories.
-These distinctions are important because each have their own issues in how they relate to uncertainty more broadly.
-Below we will go through the five key issues in uncertainty visualisation, describe how they mainfest in the uncertainty visualisation literature, and offer some tentative solutions to these problems.
-
-We will also comment on some of the uncertainty visualisation papers that suggest a visualisation without an evaluation experiment.
-Some of these plots may offer a solution to the unique problems we mention in this paper and these visualisations will be discussed in "Part 6: Great Examples".
-We will also consider these papers to illustrate the issues that visualisations are aware of and working to fix, and discuss whether or not these directions of research are likely to bear fruit, and why.
-
-## Uncertainty versus Risk
 Uncertainty is noise.
@@
 
-> Edit note: this section needs a huge rewrite.
-The idea of this section blends way too much with the measurement discussions in the "noise and signal" section.
-I have already taken paragraphs from here and just pasted it in the noise section so it might not be coherrent.
-Additionaly this section goes on SEVERAL long rambles that realistically should have been a sentance or not included at all.
-
 It would be impossible to discuss uncertainty visualisation without mentioning the broader conversation around uncertainty communication.
 The conversation around uncertainty communication is often centered around the psychology is communicating risks.
+
+In this section we will discuss how *most* attempts to measure the effects of uncertainty are either value extraction experiments with an extra step, or are measuring uncertainty using a proxy that does not actually represent uncertainty.
+In this section we will talk about the issues of trying to measure uncertainty, the real purpose of uncertainty visualisation, and suggestions in how we can better measure uncertainty in our visualisation experiments.
+
+## Most experiments are value extraction experiments with extra steps
+Those with a keen eye might have read the previous sections and noticed our statements were only relevant to evaluation experiments where participants must extract a particular value from a plot.
+Of course, these experiments are not reflective of "real world" uses of uncertainty visualisations, we do not want to visualise uncertainty so that we can answer questions such as *what is the probability of even A*.
+In reality uncertainty is useful for decision making and eliciting trust from an audience.
+This section will explain why, despite appearing to be different, these experiments are actually still value extraction experiments and thus have the same pitfalls.
 
@@
 
-## Decision Making
+### Decision making
 Decision making questions that typically involve uncertainty discuss the concept of "risk-aversion" as though it is a mistake.
@@
 
-## Trust
+### Trust and confidence
 Authors also occasionally use trust as a proxy for uncertainty.
@@
 
-## Other attemps to capture uncertainty
+### Other attemps to capture uncertainty
 There are other studies that try to include uncertainty information without directly asking for it but do not go down the route of asking for a decision or levels of trust.
@@
 
-## Uncertainty and hypothesis testing
+## Hypothesis testing versus uncertainty
 It is unclear how participants are supposed to incorperate uncertainty information into their responses when uncertainty visualisation authors themselves are unsure.
@@
 @Spiegelhalter2017 noted we "cannot assess the quality of risk communication unless the objectives are clear".
-
-## Suggestions for measuring uncertainty
 

### 2024-03-20 paper.qmd
rewrote noise vs signal introduction

@@
 Therefore, while taxonomies can be helpful to map out the space of "things that we consider to be related to uncertainty", a taxonomy is *not* a definition and do little to help statisticians untangle and estimate the uncertainty in their projects.
+This is obvious in practice, as not understanding how to calculate uncertainty is one of the leading reasons cited by visualisation authors when explaining why they don't include it in their visualisation [@Hullman2020a].
 That being said, these taxonomies can hint towards what is important when we think about uncertainty.
@@
 After all, confidence intervals, minimums, maxmimums, variance, and any other statistic that can be considered related to "uncertainty" is just that, a statisitc.
+If we want to visualise the relationship between a temperature estimate and our forcasted variance on that esitmate, can we not just use a scatter plot, as we would any other pair of variables?
 There is nothing special about these statistics that should differentiate them from a mean or median and imply their visualisation should be "special" in a way that would warrant its own field.
@@
 
-# Are you measuing uncertainty as noise or signal?
-Most evaluation experiments fail to capture uncertainty as we want to measure it, as noise. Some authors may read this and think that obviously, we dont use
+# Uncertainty is noise, not signal
+Uncertainty is noise, but most evaluation experiments measure it as signal.
+Noise and signal are not an inherrent property of variance, or meta-data or the aspects of an analysis that are usually secondary considerations, it is a property of the data that *promotes* a message we are trying to infer and the data that *supresses* it, if that message is invalid.
+When we ask the viewer of a plot to look at data and extract a value, we are asking them to perform inference on that value.
+There will be noise associated with that answer and that is uncertainty.
+If we ask direct question about some uncertainty metric, we have turned the uncertainty into signal because that is what the participants are drawing inference on.
 
-Uncertainty acts as a form of statistical "hedging" for signals found in an analysis.
-Interviews with experts in statistic back up this primary motivation, as ignoring uncertainty information of often expressed as being similar to fraud or lying and the goal of "improving decisions" is only seen as a secondary outcome of an appropriately hedged signal [@Hullman2020a; @Manski2020].
-
-While uncertainty information *is* often used to improve decisions, it seems the main motivation in its communication is to offer transparency to the viewer about a statistical analysis.
-
-This concept of uncertainty means that the work on communicating risk does not cover the field of uncertainty communication as a whole.
-It covers the specific case where the signal itself *is* uncertain, but this work does not naturally extend to the goal of hedging a signal (i.e.
-when we want to communicate uncertainty as noise).
-
-Unfortunately understanding the relevant noise seems to be its own problem, as statisticians and visualisation authors seem to be unaware or confused in how to calculate the uncertainty related to an analysis [@Hullman2020a].
-Additionally, communicating uncertainty such that it conveys the appropriate singal supression involves combining uncertainty from multiple different sources, something that is considered to be difficult by even experts in uncertainty communication [@Spiegelhalter2017].
-
-Uncertainty is noise.
-This may be an obvious statement, but what is not obvious is that the existing literature sees noise as a signal.
-When we directly ask "What is the varinance of this variable" the variance *becomes* the signal.
-Noise and signal are not an inherrent property of variance, or meta-data or the aspects of an analysis that are usually secondary considerations, it is a property of the data that *promotes* a message we are trying to infer and the data that *supresses* it.
-The replication problem in uncertainty visualisation, where experiments are simply repeating established visualisation work, comes from the misunderstandings established in previous sections.
+This is likely the cause of the contradiction between uncertainty visualisation papers following identical results to normal information visualisation papers, and authors believing uncertainty is inherently *different* to normal visualisation.
+Each of the sections in this paper likely contribute in some way to this contradiction in conceptualisation of uncertainty visualisaiton and the results of the studies in the field.
 Uncertainty is poorly defined, therefore authors do not understand it is not an inherrent property of the data and is actually related to a particular inference question, this incorrect view that uncertainty is always variance or probabilities leads to the class of "uncertainty visualisations" that depict different mathematical objects or visual features with little in the sense of consistent rules, this complicated class has caused visualisation authors to miss that they are simply repeating the established evaluation results from normal information visualisations.
@@
 
-Uncertainty evaluation studies are not all value extraction.
-There are cases where authors *know* the uncertainty information is relevant, but are not quite sure how to capture it.
-These studies most commonly evaluate a visualiation on its ability to convey information for a decision or how well it induces trust.
-The motivation behind these experiments are obvious, if we cannot measure uncertainty *directy* maybe we can measure its secondary affects.
-The issue is that in order to say if some plots are *better* or *worse* than others, is to ascribe value to uncertainty information.
+Uncertainty, acting as it is intended, is fundamentally transparency to the viewer about a statistical analysis.
+There are many secondary benefits that come with this improved transparency, such as better decisions, more trust in the results and more confidence in the authors.
+These secondary benefits are, however, *not* uncertainty.
+The following sections will discuss the issues and limitations in measuring uncertainty through these secondary metrics and provide suggestions as to how future studies should consider measuring uncertainty.
 
-It would be impossible to discuss uncertainty visualisation without mentioning the broader conversation around uncertainty communication.
-The conversation around uncertainty communication is often centered around the psychology is communicating risks.
+## Issues with the current methods of measuring uncertainty
+### Types of Uncertainty Visualisation Papers
+Thankfully, we do not need to survey the literature to identify the motivations and evaluation metrics used in every uncertainty visualisation experiment, @Hullman2019 has done it for us.
 
-In this section we will discuss how *most* attempts to measure the effects of uncertainty are either value extraction experiments with an extra step, or are measuring uncertainty using a proxy that does not actually represent uncertainty.
-In this section we will talk about the issues of trying to measure uncertainty, the real purpose of uncertainty visualisation, and suggestions in how we can better measure uncertainty in our visualisation experiments.
+L1: Behavioral Targets
+- Performance (241; 64.8%): how effectively a user can extract information, make inferences, or make decisions with a visualization.
+- Interpretation & Semantics (64; 17.2%): the ease with which a user associates uncertainty with an encoding.
+- Quality of User Experience (67; 1.8%): the user’s valuation of the visualization, such as their preference or satisfaction.
 
-## Most experiments are value extraction experiments with extra steps
-Those with a keen eye might have read the previous sections and noticed our statements were only relevant to evaluation experiments where participants must extract a particular value from a plot.
-Of course, these experiments are not reflective of "real world" uses of uncertainty visualisations, we do not want to visualise uncertainty so that we can answer questions such as *what is the probability of even A*.
-In reality uncertainty is useful for decision making and eliciting trust from an audience.
-This section will explain why, despite appearing to be different, these experiments are actually still value extraction experiments and thus have the same pitfalls.
+L2: Expected Effects
+Accuracy (134; 36%): Difference from a ground truth response (e.g., [37, 40]).
+- Confidence/Accuracy Alignment (5; 1.3%): Correlation between confidence and accuracy (e.g., [11]).
+- Impact on Decision-Making (3; 0.8%): Effects on decisions where no ground truth is implied (e.g., [15]).
+- Decision Quality (14; 3.8%): Difference from a rational deci- sion standard (e.g., [23, 68]).
+- Memorability (3; 0.8%): How well a fact is remembered (e.g., [39]).
+- Risk Avoidance (23; 6.2%): The degree to which a user’s re- sponse attempts to avoid risk (e.g., [35]).
+- Intuitiveness/Reading ease (31; 8.3%): How “naturally” a vi- sualization supports correct interpretations (e.g., [7]).
+- Awareness of Judgment Process (8; 2.1%): The user’s ability to recognize aspects of their judgment (e.g., [15]).
+- Learnability (0): The user’s ability to improve their perfor- mance over time
+- Confidence (24; 6.5%): The degree of belief in the validity or truth of a judgment, data set, visualization, etc. (e.g., [16]).
+- Satisfaction (38; 10.2%): The user’s aesthetic valuation of a vi- sualization (e.g., [22, 61]).
+- Interaction (4; 1.1%): How much interaction a visualization receives in time, number of clicks, etc.
+Efficiency (37; 9.9%): How well a visualization supports quick judgments (e.g., [59]).
+- Understand internal model (48; 12.9%): Seeking to understand how a judgment is made (e.g. [13, 87]).
 
-## Types of Uncertainty Visualisation Papers
+L3: Evaluation goal
+- Compare impacts of multiple uncertainty visualizations (255; 68.5%): Results of one or more tasks are used to rank two or more uncertainty visualizations (e.g., [11, 40, 53]).
+- Determine the impact of presenting uncertainty (98; 26.3%): At least one visualization that does not contain uncertainty infor- mation is evaluated (e.g., [68]).
+- Validate effectiveness of an uncertainty vis (5; 1.3%): A user study is designed to confirm that a visualization improves some response(s). (e.g. [80]).
+- Understand why/how a visualization works (54; 14.5%): Re- sponses to a visualization(s) are analyzed to identify or confirm some judgment mechanism (e.g., [87]).
+- Understand interactions with user characteristics (53; 14.2%): The effect of properties of the user (expertise, location, etc.) on responses is analyzed (e.g., [1]).
 
-> Edit notes: Originally I was planning to have the discussion following this section relate to these "types" of papers, but that didn't end up happening.
-Depending on what I do with later sections, this might need a rewrite or to just be removed, since I'm not sure how relevant this is to the main ideas of the paper.
+L4: Measures
+(A) Directly elicated measures
+- Decision measures (31; 8.3%) 
+  – Decision (31; 8.3%): A hypothetical choice action (e.g., saying when they would leave for the bus [23]).
+  - Allocation (0%): An amount of currency wagered to en- dorse a choice (e.g., [74]).
+- Attribute-based measures (141; 37.9%) 
+  – Value of [nominal | ordinal | continuous] variable (67; 18%): Specifying a value along a provided scale (e.g., ask- ing the user to estimate the forecasted temperature from a display [46]).
+  – Count (15; 4%): Specifying a count (e.g., of features of a certain type).
+  – Probability (29; 7.8%): A probability specified as a probability or frequency (e.g., [40]) or as a Likert-style question asking for the relative probability of an event (e.g., [1]).
+  – Variance (4; 1.1%): A value or set of values describing how much an outcome can vary (i.e., a range, standard de- viation, standard error, etc. as in [31]).
+  – Rank by [data attribute | uncertainty attribute] (22; 5.9%): The top k features sorted by a numerical attribute.
+- Self-report measures (161; 43.3%) 
+  – Spontaneous Impression (16; 4.3%): An unconstrained description of one’s reaction to a visualization (e.g., [8]).
+  – Subjective Confidence (28; 7.5%): A valuation of one’s confidence in information or a judgment.
+  – Perceived Risk/Reliability (15; 4%): How risky, or con- versely how reliable, one believes an outcome is.
+  – Perceived Ability/Effort (29; 7.8%): How hard a visual- ization is to read or use.
+  – Self-reported Satisfaction (41; 10.8%): An aesthetic val- uation of a visualization.
+  – Self-reported Intention (10; 2.7%): A reported intention to use or recommend information or an artifact (e.g., [17]).
+  – Willingness to Pay (6; 1.6%): A reported intention in- volving allocating money to an information or artifact(s).
+  – Self-reported Action (17; 4.6%): Whether one used, re- membered, etc. information or an artifact.
+  - Galvanized Skin Response (GSR) (1; 0.3%): Change in the electrical resistance of the skin.
+  – Interaction logs (35; 8.9%): A log of a user’s behavior as they use a visualization.
+  – Eye movement (2; 0.5%): An eye tracker log of voluntary or involuntary eye movement.
+
+(B) Derived measures
 
@@
 
-### Decision making
+### Decision making and risk aversion
 Decision making questions that typically involve uncertainty discuss the concept of "risk-aversion" as though it is a mistake.
@@
 
-## Hypothesis testing versus uncertainty
+## Suggestions to measure uncertainty
+### Heuristic checks
+- Have a better hypothesis? Like its more specific. They are looking for a very specific thing *wrong* with a visualisation.
+- We wary of trying to control something and observe it at the same time
+
+A lot of work that identifies heuristics or biases in plots do not hinge on the previous assumptions.
+This work also provides useful insights for experiments by highlights pitfalls participants might fall into when they review the results of evaluation experiments [@Hullman2016].
+@Newman2012 found that participants were more likely to view points within the bar as more likely than points outside of the bar in bar charts with error bars.
+Similar effects have been identified in other types of uncertainty displayed.
+@Padilla2017 found that points that were on an outcome of an ensemble display were perceived as more likely than points not on an outcome, even when the point that was not on a specific outcome of the ensemble was closer to the mean of the uncertainty distribution.
+The sine illusion can cause the confidence interval of a smoothed sine curve to seem wider at the peaks than the troughs, causing us to underestimate uncertainty associated with changing values [@Vanderplas2015]
+
+### Just noticiable signal
+Connecting our evaluation of uncertainty back to the reasons uncertainty is believed to be different is likely the key to improving evaluation studies.
+Uncertainty acts as a form of statistical "hedging" for signals found in an analysis.
+Interviews with experts in statistic back up this primary motivation, as ignoring uncertainty information of often expressed as being similar to fraud or lying and the goal of "improving decisions" is only seen as a secondary outcome of an appropriately hedged signal [@Hullman2020a; @Manski2020].
+
+### Thinking "smaller picture" with
+Many visualisation experiments try to compare two plots with *several* differences, but do not seem to be interested in the *mechanisms* by which we extract information from visualisations.
+Let us consider a decision making experiment, there are several steps that Simply asking "Do people make better decisions with
+
+**Experiment rant from confirmation**
+
+### Comparisons to lineup plots and hypothesis testing
 It is unclear how participants are supposed to incorperate uncertainty information into their responses when uncertainty visualisation authors themselves are unsure.
@@
 ![I will just make my own version of a climate scenario plot, this one is just here from a random site as a place holder](climatescenario.png){fig-climatescenario}
-
-A lot of work that identifies heuristics or biases in plots do not hinge on the previous assumptions.
-This work also provides useful insights for experiments by highlights pitfalls participants might fall into when they review the results of evaluation experiments [@Hullman2016].
-@Newman2012 found that participants were more likely to view points within the bar as more likely than points outside of the bar in bar charts with error bars.
-Similar effects have been identified in other types of uncertainty displayed.
-@Padilla2017 found that points that were on an outcome of an ensemble display were perceived as more likely than points not on an outcome, even when the point that was not on a specific outcome of the ensemble was closer to the mean of the uncertainty distribution.
-The sine illusion can cause the confidence interval of a smoothed sine curve to seem wider at the peaks than the troughs, causing us to underestimate uncertainty associated with changing values [@Vanderplas2015]
 

### 2024-03-20 paper.qmd
rewrote inrepretation and semantics section

@@
 ## Issues with the current methods of measuring uncertainty
-### Types of Uncertainty Visualisation Papers
-Thankfully, we do not need to survey the literature to identify the motivations and evaluation metrics used in every uncertainty visualisation experiment, @Hullman2019 has done it for us.
+The overarching goal of most uncertainty visualisation evaluation papers belong to one of: performance (how effectively a participant can extract information from a plot), interpretation and semantics (the ease with which a particular visualisation is associated with uncertainty), and quality of user experience (if the participants liked the plot or not)[@Hullman2019].
+@Hullman2019 found that the majority of papers evaluate visualisations on performance (approximately 65% of the papers they surveyed) or interpretation and semantics (approximately 17%) and both of these evaluation goals will run into problems because of their conceptialisation of uncertainty.
+
+### Performance
 
@@
 
-L2: Expected Effects
-Accuracy (134; 36%): Difference from a ground truth response (e.g., [37, 40]).
+L2: 
+Value extraction
+- Accuracy (134; 36%): Difference from a ground truth response (e.g., [37, 40]).
 - Confidence/Accuracy Alignment (5; 1.3%): Correlation between confidence and accuracy (e.g., [11]).
+
+Decision and risk avoidance
+- Decision Quality (14; 3.8%): Difference from a rational deci- sion standard (e.g., [23, 68]).
+- Risk Avoidance (23; 6.2%): The degree to which a user’s response attempts to avoid risk (e.g., [35]).
+
+Trust
+- Confidence (24; 6.5%): The degree of belief in the validity or truth of a judgment, data set, visualization, etc. (e.g., [16]).
+
+Qualitative
 - Impact on Decision-Making (3; 0.8%): Effects on decisions where no ground truth is implied (e.g., [15]).
-- Decision Quality (14; 3.8%): Difference from a rational deci- sion standard (e.g., [23, 68]).
+- Awareness of Judgment Process (8; 2.1%): The user’s ability to recognize aspects of their judgment (e.g., [15]).
+- Understand internal model (48; 12.9%): Seeking to understand how a judgment is made (e.g. [13, 87]).
+- Efficiency (37; 9.9%): How well a visualization supports quick judgments (e.g., [59]).
+
+Semantics
+- Intuitiveness/Reading ease (31; 8.3%): How “naturally” a visualization supports correct interpretations (e.g., [7]).
+
+?? Dont care
 - Memorability (3; 0.8%): How well a fact is remembered (e.g., [39]).
-- Risk Avoidance (23; 6.2%): The degree to which a user’s re- sponse attempts to avoid risk (e.g., [35]).
-- Intuitiveness/Reading ease (31; 8.3%): How “naturally” a vi- sualization supports correct interpretations (e.g., [7]).
-- Awareness of Judgment Process (8; 2.1%): The user’s ability to recognize aspects of their judgment (e.g., [15]).
-- Learnability (0): The user’s ability to improve their perfor- mance over time
-- Confidence (24; 6.5%): The degree of belief in the validity or truth of a judgment, data set, visualization, etc. (e.g., [16]).
-- Satisfaction (38; 10.2%): The user’s aesthetic valuation of a vi- sualization (e.g., [22, 61]).
+- Learnability (0): The user’s ability to improve their performance over time
+- Satisfaction (38; 10.2%): The user’s aesthetic valuation of a visualization (e.g., [22, 61]).
 - Interaction (4; 1.1%): How much interaction a visualization receives in time, number of clicks, etc.
-Efficiency (37; 9.9%): How well a visualization supports quick judgments (e.g., [59]).
-- Understand internal model (48; 12.9%): Seeking to understand how a judgment is made (e.g. [13, 87]).
 
@@
 - Compare impacts of multiple uncertainty visualizations (255; 68.5%): Results of one or more tasks are used to rank two or more uncertainty visualizations (e.g., [11, 40, 53]).
-- Determine the impact of presenting uncertainty (98; 26.3%): At least one visualization that does not contain uncertainty infor- mation is evaluated (e.g., [68]).
+- Determine the impact of presenting uncertainty (98; 26.3%): At least one visualization that does not contain uncertainty information is evaluated (e.g., [68]).
 - Validate effectiveness of an uncertainty vis (5; 1.3%): A user study is designed to confirm that a visualization improves some response(s). (e.g. [80]).
@@
 - Understand interactions with user characteristics (53; 14.2%): The effect of properties of the user (expertise, location, etc.) on responses is analyzed (e.g., [1]).
-
-L4: Measures
-(A) Directly elicated measures
-- Decision measures (31; 8.3%) 
-  – Decision (31; 8.3%): A hypothetical choice action (e.g., saying when they would leave for the bus [23]).
-  - Allocation (0%): An amount of currency wagered to en- dorse a choice (e.g., [74]).
-- Attribute-based measures (141; 37.9%) 
-  – Value of [nominal | ordinal | continuous] variable (67; 18%): Specifying a value along a provided scale (e.g., ask- ing the user to estimate the forecasted temperature from a display [46]).
-  – Count (15; 4%): Specifying a count (e.g., of features of a certain type).
-  – Probability (29; 7.8%): A probability specified as a probability or frequency (e.g., [40]) or as a Likert-style question asking for the relative probability of an event (e.g., [1]).
-  – Variance (4; 1.1%): A value or set of values describing how much an outcome can vary (i.e., a range, standard de- viation, standard error, etc. as in [31]).
-  – Rank by [data attribute | uncertainty attribute] (22; 5.9%): The top k features sorted by a numerical attribute.
-- Self-report measures (161; 43.3%) 
-  – Spontaneous Impression (16; 4.3%): An unconstrained description of one’s reaction to a visualization (e.g., [8]).
-  – Subjective Confidence (28; 7.5%): A valuation of one’s confidence in information or a judgment.
-  – Perceived Risk/Reliability (15; 4%): How risky, or con- versely how reliable, one believes an outcome is.
-  – Perceived Ability/Effort (29; 7.8%): How hard a visual- ization is to read or use.
-  – Self-reported Satisfaction (41; 10.8%): An aesthetic val- uation of a visualization.
-  – Self-reported Intention (10; 2.7%): A reported intention to use or recommend information or an artifact (e.g., [17]).
-  – Willingness to Pay (6; 1.6%): A reported intention in- volving allocating money to an information or artifact(s).
-  – Self-reported Action (17; 4.6%): Whether one used, re- membered, etc. information or an artifact.
-  - Galvanized Skin Response (GSR) (1; 0.3%): Change in the electrical resistance of the skin.
-  – Interaction logs (35; 8.9%): A log of a user’s behavior as they use a visualization.
-  – Eye movement (2; 0.5%): An eye tracker log of voluntary or involuntary eye movement.
-
-(B) Derived measures
 
@@
 
-### Decision making and risk aversion
+#### Decision making and risk aversion
 Decision making questions that typically involve uncertainty discuss the concept of "risk-aversion" as though it is a mistake.
@@
 
-### Trust and confidence
+#### Trust and confidence
 Authors also occasionally use trust as a proxy for uncertainty.
@@
 
-### Other attemps to capture uncertainty
+#### Deterministic questions
+- You cannot make a deterministic statements about a random variable (citation: the entire field of probability).
+Many of these issues just boil down to a really pervasive misunderstanding of statistics.
+I do not expect the general population to understand this fact about random variables, however *I would expect* researchers investigating *methods of statisticl communication* would at least understand enough to ask appropriate questions that *are answerable*.
+- while we are here, you cant answer a probability question with a deterministic. I cannot say yes or no to a probability question without providing a threshold. 
+- Basically, you need a probability threshold to answer these questions, something it does not seem like the authors are collecting, and I am unsure if the authors are aware of this themselves because they never mention it.
+
+#### Other (questionable) attemps to capture uncertainty
 There are other studies that try to include uncertainty information without directly asking for it but do not go down the route of asking for a decision or levels of trust.
@@
 Cite gap __: Weird questions to capture uncertainty
+- @Correll2018 “place your 5 ships on the safest locations on the board.” (minimise danger)
 - @Ibrekk1987 asked participants for the "best" estimate while displaying a population density function.
@@
 ![This shows the user interface for the experiment performed by @Hofmann2012. The question of "furthest to the right" is open to interpretation. ](furthestright.png){#fig-right}
+
+### Interpretation and semantics
+Interpretation and semantics experiments are seeking to identify a dimension (or visual task) that uncertainty naturally maps to.
+These experiments inherently view uncertainty as a variable that is separate to the variable on which we have mapped our signal.
+For example, lets say we have a map were maximum daily temperature is presented using points where the colour (red for hot and blue for cold) of the point is associated with the temperature, and the bluriness of that point is associated with the variance *of* that temperature.
+It is highly likely that our brains will not flatten that into a single variable depicting noise and signal, but rather *separately* extract the temperature (colour) and uncertainty (blur) information as two independent variables.
+If the variables are extracted separately, there is no guarentee that the uncertainty will act as an approriate signal supressor.
+This problem has been noticed by others in the field, that typically use this method (specifically in the spatial uncertainty context) and a desire for representations that integrate uncertainty and signal is one of the reasons for the invention of the value-supressing uncertainty pallet [@Correll2018].
+
+Value-supressing uncertainty pallets (VSUP) were developed as a method that would allow the signal and the noise to be interpreted together such that insights gained by the viewers of a plot are appropriately *supressed* by the uncertainty.
+Hence the name of the pallet.
+@fig-maps depicted this map colouring approach and several other extensions on the typical choropleth map that differ in where in the visualisation process they combine the noise and signal information into a single concept of "valid signal".
+The first and most basic map is a simple choropleth map where each value (the colouring of each local government area) has no associated "uncertainty".
+The next map (which embodies the approach taken by the semantics experiments) is the bivariate map which maps the signal to colour value and the uncertainty to colour hue.
+If visualisation was performing signal supression then that would mean the two dimensional space defined by colour value and colour hue can be mentally "flattened" into a single dimension of "valid value" that can captures the signal the our brain would need to be able to flatted this two dimensional space into a single space of "signal validity".
+The idea of two perceptual tasks flattening into one variable in the mind of the viewer may be wishful thinking, but it is not impossible given we are not certain on how the perceptual tasks are mapped within the human brain.
+@Sterzik2023 found that when a value was mapped to the textures of stippling, hatching, and triangles, and found that the difference between two points on this one dimensional texture was actually a 2D space (likely "texture business" and "light/darkness").
+That being said, if we look at the visual signals presented by the bivariate map, where the contrasting light and dark areas actually has no important meaning, it is unlikely this occurs for colour value and hue (or at least it doesnt occur in a way that is useful for uncertainty visualisation).
+Instead of hoping that uncertainty might collapse signal values into a single dimension, we can do some of that work ourselves, by using a VSUP which collapses the colour space such that high uncertainty values cannot be extracted.
+It is unclear how useful VSUPs are at actually combining signal and noise and therefore supressing plot level insights, as they have only been tested on simple value extraction tasks that require evaluating a single point [@Correll2018; @Ndlovu2023] rather than looking for spatial relationships (which is arguably what maps are for).
+Following along with this trend, the next way we might consider visualising uncertainty is to combine uncertainty and signal at the earlier stage so the "supressed signal" is represented by a single variable.
+This statistic can then be expressed in a one dimensional colour space, which is a method addopted by the Baysian surprise metric map [@Ndlovu2023] and the excedance probability map [Lucchesi2017].
+.
+
+::: {#fig-maps}
+
+![](mappals/img1.png)
+
+![](mappals/img2.png)
+
+![](mappals/img4.png)
+![](mappals/img3.png)
+
+![](mappals/img5.png)
+
+*This plot is the highest priortiy to for replacing with my own R coded maps, becuase the effects of these different mapings is really not visible using these methods* Five different ways to include error information in a map.
+A typical chloropleth map is considered the map that ignores uncertainty and simply visualises the signal.
+The bivariate map visualises the uncertainty, but does so using two different visual channels, which gives the perception of two different signals (or one signal in the assumptions the colors contrast with each other).
+An adjustment to the bivariate map is the VSUP which is similar to the bivariate plot except it merges high uncertainty values in the palette to reduce our ability to discern values that have a high enough error to not be a valid signal.
+The Bayesian surprise metric is similar to the lineup protocol, where the colour value is the difference between the observers posterior and prior beliefs.
+The exceedance probability map is a direct visualisation of the probability of a specific hypothesis.
+:::
+
+These maps make the importance of combining uncertainty and signal in a single visual channel clear.
+A chloropleth map will show signal that is not valid inference because of high uncertainty.
+At the other end of the spectrum, the bivariate map will show signal that is not always interesting because it forces us to interpret uncertainty and signal separately.
+As we move through these methods, it seems that the validity of any overarching insight becomes more visible at the cost of our ability to extract particular values of signal or noise.
+Therefore, given that the primary goal of visualisation *is* insights [@North2006], visualisation authors should err on the side of representing supressed signal as a single variable, rather than visualising uncertainty separately using two different channels.
 
@@
 ![I will just make my own version of a climate scenario plot, this one is just here from a random site as a place holder](climatescenario.png){fig-climatescenario}
-
-There are a number of visualisation methods that do take the approach of uncertainty as signal suppression.
-These visualisation examples try to establish uncertainty as something that should be interpreted *with* the signal and take several approaches to do so.
-@fig-maps depicts the array of approaches that have been used to visualise uncertainty in a map.
-
-The first approach to embedding noise in the signal is to adjust the signal visualisation so that it includes the noise.
-This method has been particularly popular with map visualisations.
-To visualise uncertainty such that it does its job of signal supression the uncertainty and the signal need to be interpreted in conjunction with each other.
-There are several suggested ways to go about this, most of which assume the "no uncertianty" version of the visualisation is a simple choropleth map.
-There are two methods to include uncertainty in a map.
-The first is to include uncertainty as error and map our signal and noise to a two dimensional *colour* space, which is the method adopted by the bivariate map and the value supressing uncertainty pallet [@Correll2018].
-These methods rely on the assumption that colour hue and value exist mentally as a two dimensional colour space and that distance in this colour space is linear (or at least appears linear using our scale).
-However, for uncertainty to be interpreted in the visualisation *with* the signal, our brain would need to be able to flatted this two dimensional space into a single space of "signal validity", and there is simply not evidence that we can do this.
-This visualisation method brings up an interesting concept though.
-If there are two aesthetics we can use that are flattened into a single channel in our mind, then plotting measurement and error to each of those aesthetics is a viable method to visualise uncertainty.
-The idea of two perceptual tasks flattening into one variable in the mind of the viewer may be wishful thinking, but it is not impossible given we are not certain on how the perceptual tasks are mapped within the human brain.
-@Sterzik2023 found that when a value was mapped to the textures of stippling, hatching, and triangles, and found that the difference between two points on this one dimensional texture was actually a 2D space (likely "business" and light/darkness).
-This idea of two aesthetics flattening into one cohesive variable in the perceptual space is particularly interesting for uncertainty visualisation.
-
-An alternative method to visualising uncertainty in a map is to combine uncertainty and signal at the earlier stage so the "supressed signal" is represented by a single variable.
-This statistic can then be expressed in a one dimensional colour space, which is a method addopted by the Baysian surprise metric map [@Ndlovu2023] and the excedance probability map [Lucchesi2017].
-These methods do not require us to compress colour hue and value into a single variable to interpret signal and noise as a single variable, it is done explicitly for us before the visualisaiton.
-These methods require more specificity in *how* the uncertainty contains important information.
-
-::: {#fig-maps}
-
-![](mappals/img1.png)
-
-![](mappals/img2.png)
-
-![](mappals/img4.png)
-![](mappals/img3.png)
-
-![](mappals/img5.png)
-
-*I am going to replace this with my own R coded maps so they are all on the same data and signal/noise.* Four different ways to include error information in a map.
-A typical chloropleth map is considered the map that ignores uncertainty and simply visualises the signal.
-The bivariate map visualises the uncertainty, but does so using two different visual channels, which gives the perception of two different signals (or one signal in the assumptions the colors contrast with each other).
-An adjustment to the bivariate map is the VSUP which is similar to the bivariate plot except it merges high uncertainty values in the palette to reduce our ability to discern values that have a high enough error to not be a valid signal.
-The Bayesian surprise metric is similar to the lineup protocol, where the colour value is the difference between the observers posterior and prior beliefs.
-The exceedance probability map is a direct visualisation of the probability of a specific hypothesis.
-:::
-
-These maps make the importance of combining uncertainty and signal in a single visual channel clear.
-A chloropleth map will show signal that is not valid inference because of high uncertainty.
-At the other end of the spectrum, the bivariate map will show signal that is not always interesting because it forces us to interpret uncertainty and signal differently.
-The value supressing uncertainty pallet was clearly designed with hypothesis testing in mind, because it combines highly uncertain values.
-The main difference between the VSUP and the bivariate colour map is actually the subtle combination of signal and noise through discernability.
-In this sense, the uncertainty visualisation maps that combine uncertainty and signal before it is conveyed visually simplify the visualisation and make it easier to interpret value supressed signal.
-This is what is done by the exceedence probability map and the Bayesian surprise metric map.
-They are more explicit in how the uncertainty should supress the signal since they make the specific hypothesis explicit.
 

### 2024-03-21 paper.qmd
updated some of the decision bit

@@
 ## Issues with the current methods of measuring uncertainty
-The overarching goal of most uncertainty visualisation evaluation papers belong to one of: performance (how effectively a participant can extract information from a plot), interpretation and semantics (the ease with which a particular visualisation is associated with uncertainty), and quality of user experience (if the participants liked the plot or not)[@Hullman2019].
-@Hullman2019 found that the majority of papers evaluate visualisations on performance (approximately 65% of the papers they surveyed) or interpretation and semantics (approximately 17%) and both of these evaluation goals will run into problems because of their conceptialisation of uncertainty.
-
-### Performance
-
-L1: Behavioral Targets
-- Performance (241; 64.8%): how effectively a user can extract information, make inferences, or make decisions with a visualization.
-- Interpretation & Semantics (64; 17.2%): the ease with which a user associates uncertainty with an encoding.
-- Quality of User Experience (67; 1.8%): the user’s valuation of the visualization, such as their preference or satisfaction.
-
-L2: 
-Value extraction
-- Accuracy (134; 36%): Difference from a ground truth response (e.g., [37, 40]).
-- Confidence/Accuracy Alignment (5; 1.3%): Correlation between confidence and accuracy (e.g., [11]).
-
-Decision and risk avoidance
-- Decision Quality (14; 3.8%): Difference from a rational deci- sion standard (e.g., [23, 68]).
-- Risk Avoidance (23; 6.2%): The degree to which a user’s response attempts to avoid risk (e.g., [35]).
-
-Trust
-- Confidence (24; 6.5%): The degree of belief in the validity or truth of a judgment, data set, visualization, etc. (e.g., [16]).
-
-Qualitative
-- Impact on Decision-Making (3; 0.8%): Effects on decisions where no ground truth is implied (e.g., [15]).
-- Awareness of Judgment Process (8; 2.1%): The user’s ability to recognize aspects of their judgment (e.g., [15]).
-- Understand internal model (48; 12.9%): Seeking to understand how a judgment is made (e.g. [13, 87]).
-- Efficiency (37; 9.9%): How well a visualization supports quick judgments (e.g., [59]).
-
-Semantics
-- Intuitiveness/Reading ease (31; 8.3%): How “naturally” a visualization supports correct interpretations (e.g., [7]).
-
-?? Dont care
-- Memorability (3; 0.8%): How well a fact is remembered (e.g., [39]).
-- Learnability (0): The user’s ability to improve their performance over time
-- Satisfaction (38; 10.2%): The user’s aesthetic valuation of a visualization (e.g., [22, 61]).
-- Interaction (4; 1.1%): How much interaction a visualization receives in time, number of clicks, etc.
-
-L3: Evaluation goal
-- Compare impacts of multiple uncertainty visualizations (255; 68.5%): Results of one or more tasks are used to rank two or more uncertainty visualizations (e.g., [11, 40, 53]).
-- Determine the impact of presenting uncertainty (98; 26.3%): At least one visualization that does not contain uncertainty information is evaluated (e.g., [68]).
-- Validate effectiveness of an uncertainty vis (5; 1.3%): A user study is designed to confirm that a visualization improves some response(s). (e.g. [80]).
-- Understand why/how a visualization works (54; 14.5%): Re- sponses to a visualization(s) are analyzed to identify or confirm some judgment mechanism (e.g., [87]).
-- Understand interactions with user characteristics (53; 14.2%): The effect of properties of the user (expertise, location, etc.) on responses is analyzed (e.g., [1]).
-
 Uncertainty visualisation papers can be organised according to the *goal* of the experiment.
@@
 Sometimes these papers are a preliminary step in finding a solution for common problems and intend to evaluate the visualsiation in later work.
-Until that later work is done, it is often difficult to accept one particular representation.
-While justification could come in the form of discussing established concepts in visualisation, such as the hierarchy of perceptual tasks, even these may need an evaluation study to back up their claims.
+The reasoning for this is obvious.
+There are often heuristics and biases that are not obvious to us when designing visualisation.
+Additionally, these heuristics and baises can change depending on the larger scope of the graphic and the population we are communicating with [@Spiegelhalter2017; @Kinkeldey2014].
 
-The reasoning for this is obvious.
-There are a large number heuristics and biases that are not obvious to us when designing visualisation.
-Additionally, these heuristics and baises can change depending on the larger scope of the graphic and the population we are communicating with.
-Additionally, since there is often a miriad of ways to visualise any particular mathematical object, to adopt specific visualisation that is not already common practice needs reasonable justification, lest we run into a range of unforseen heuristic pitfalls.
+The overarching goal of most uncertainty visualisation evaluation papers belong to one of: performance (how effectively a participant can extract information from a plot), interpretation and semantics (the ease with which a particular visualisation is associated with uncertainty), and quality of user experience (if the participants liked the plot or not)[@Hullman2019].
+@Hullman2019 found that the majority of papers evaluate visualisations on performance (approximately 65% of the papers they surveyed) or interpretation and semantics (approximately 17%) and both of these evaluation goals will run into problems because of their conceptualisation of uncertainty.
 
-The remainder of section 4 will focus on the results and methods of uncertainty visualisation evaluation experiments.
-There are six main types of evaluation experiments in the uncertainty visualisation literature, they are:
-- Association perceptual tasks: These experiments identify perceptual tasks that are/could be associated with uncertainty, and test its association with uncertainty, check it's number of distinguishable levels, and/or check how quickly/accurately people can draw the uncertainty information from the graphic.
-- Value perceptual tasks: Participants are shown some depiction of uncertainty and are asked to extract a value, make a comparison, or do something simple to show that the graphic is readable.  
-- Heuristic Checks: These experiments are checking if a perceptual bias that has been identified through public use of a particular plot, other evaluation experiments, or gut feeling identified through public use of a plot or in a different.
--  Plot Comparisons: These papers describe an uncertainty visualisation evaluation experiment, where they compare two or more uncertainty visualisaitons on a specific set of questions.
-These questions can be about specific statistics (e.g.
-What is the probability that A>B?) or about a secondary goal such as trust or decision making (e.g.
-How much do you trust this estimate?).
-- Qualitative analysis: These experiments show participants an uncertainty visualisation for an open ended or specific task, and asks them to describe how the visualisation was used to come to a conclusion, or they are asked to describe what they see in the visualisation.
-- Method papers: These papers discuss issues with the evaluation methods of uncertainty visualisaiton and suggest changes or improvements, typically with results indicating the benefit of the change.
+### Performance
+There are many metrics experimenters use to evaluate the performance of a visualisation.
+Each metric, such as can use accuracy, decision quality, confidence, trust and others each have their own issues which prevent them from accurately capturing the effects of uncertainty in a visualisation.
 
-These experiments are not mutually exclusive within a paper, and a single study will often contain several experiments, usually from several of these categories.
-These distinctions are important because each have their own issues in how they relate to uncertainty more broadly.
-Below we will go through the five key issues in uncertainty visualisation, describe how they mainfest in the uncertainty visualisation literature, and offer some tentative solutions to these problems.
+By far the most common metric used is accuracy, which is used by approximately 36% of evaluation studies [@Hullman2019].
+In previous sections, we pointed out issues that arise when accuracy (of extracting probability or variance) is used as the primary metric in evaluating uncertainty visualisations, the work simply replicates known concepts in information visualisation.
+Subjective metrics, such as how much participants liked the aesthetic representation of a visualisation, or metrics with an unclear motivation or use, such as how memorable the information in a plot is or how much a user interacted with a plot, are ignored in this section.
+Below we will explain in detail why experiments that try to measure uncertainty using methods such as decision making, trust, or other types of questions, also fail to capture the effects of visualising uncertainty in a plot.
 
-We will also comment on some of the uncertainty visualisation papers that suggest a visualisation without an evaluation experiment.
-Some of these plots may offer a solution to the unique problems we mention in this paper and these visualisations will be discussed in "Part 6: Great Examples".
-We will also consider these papers to illustrate the issues that visualisations are aware of and working to fix, and discuss whether or not these directions of research are likely to bear fruit, and why.
+#### Decision making and risk aversion
+Decision making tasks are often described by authors as a more realistic version how plots are used in practice.
+While we do not disagree with this statement, exactly *why* these tasks are different to value extraction tasks is never explained.
+While the task may be more similar to how plots are actually used, that does not necessarily mean it is a better *experimental* environement for evaluating plots.
+Let us consider this in more detail.
+A normal value extraction task involves:
+1) correctly interpret the question
+2) extract the specified value
+3) report the value
+On the other hand, a decision making task involves:
+1) Correctly interpret the question
+  2) Make a choice that involves
+  - If no ground truth is in the experiment
+      a) using *personal* risk utility function to identify a threshold
+      b) extracting that threshold value from the plot
+      c) making a judgement based on that value
+  - If a group truth is specified (through right/wrong answers or a payment sceme)
+      a) Use the payment sceme to work out the choice threshold that will maximise payout
+      b) Over-ride the personal risk utility function adopt the experimenters risk utility function to identify a threshold
+      c) extracting that threshold value from the plot
+      d) making a judgement based on that value
+The key aspect that separates a decision making task from a value extraction task, for uncertainty visualisations, is the
 
-> Edit note: Currently these notes are just from spiegelhater because this is the only paper I have moved notes over from.
-Other sources will be used here lol.
-Working out what exactly they are communicating is a bit of a pain though.
+Decision making questions that typically involve uncertainty discuss the concept of "risk-aversion" as though it is a mistake.
+The irony of an uncertainty visualisation pper expressing the importance of uncertainty visualisation, but setting up a ground truth or optimum strategy in which participants *should* ignore the uncertainty information is not lose on us.
+This perspective does not make sense because it implies participants should make decisions entirely based on the average.
+A possible soltion for this is to provide a utility framework for a particular experiment [@Hullman2016], however it is unclear how easily participants could adopt a utility framework that is different to their own.
+@Cheong2016 introduced a very simple optimum strategy into their experiment on bushfire uncertainty visualisations, where participants should choose to evacuate a house as soon as its likelihood of catching fire is 50% or higher.
+Interestingly, despite this very obvious and simple tactic to maximize payout from the experiment, it seems like many participants did not adopt it, instead acting much more warily as though they were considering whether or not they would *actually* evacuate in the event of a fire.
+These papers also very rarely consider what the optimum decision would be in a given scenario, and a difficulty with anticipating the best strategy is conflated with the failings of the uncertainty visualisation.
+Ultimately, we agree with @Hullman2016 who suggested that what determines an approripate ground truth is largely a philosophical exercise.
 
@@
 In this section, we will discuss the current literature on uncertainty visualisation, how the uncertainty communication issues more broadly manifest in the specific case of visualisation, and what general lessons we can take from the existing work.
-
-#### Decision making and risk aversion
-Decision making questions that typically involve uncertainty discuss the concept of "risk-aversion" as though it is a mistake.
-The irony of an uncertainty visualisation pper expressing the importance of uncertainty visualisation, but setting up a ground truth or optimum strategy in which participants *should* ignore the uncertainty information is not lose on us.
-This perspective does not make sense because it implies participants should make decisions entirely based on the average.
-A possible soltion for this is to provide a utility framework for a particular experiment [@Hullman2016], however it is unclear how easily participants could adopt a utility framework that is different to their own.
-@Cheong2016 introduced a very simple optimum strategy into their experiment on bushfire uncertainty visualisations, where participants should choose to evacuate a house as soon as its likelihood of catching fire is 50% or higher.
-Interestingly, despite this very obvious and simple tactic to maximize payout from the experiment, it seems like many participants did not adopt it, instead acting much more warily as though they were considering whether or not they would *actually* evacuate in the event of a fire.
-These papers also very rarely consider what the optimum decision would be in a given scenario, and a difficulty with anticipating the best strategy is conflated with the failings of the uncertainty visualisation.
-Ultimately, we agree with @Hullman2016 who suggested that what determines an approripate ground truth is largely a philosophical exercise.
 

### 2024-03-21 paper.qmd
finished decision making section

@@
 A normal value extraction task involves:
+
 1) correctly interpret the question
@@
 3) report the value
-On the other hand, a decision making task involves:
+
+On the other hand, a decision making task using an uncertainty visualisation involves:
+
 1) Correctly interpret the question
-  2) Make a choice that involves
-  - If no ground truth is in the experiment
-      a) using *personal* risk utility function to identify a threshold
-      b) extracting that threshold value from the plot
-      c) making a judgement based on that value
-  - If a group truth is specified (through right/wrong answers or a payment sceme)
-      a) Use the payment sceme to work out the choice threshold that will maximise payout
-      b) Over-ride the personal risk utility function adopt the experimenters risk utility function to identify a threshold
-      c) extracting that threshold value from the plot
-      d) making a judgement based on that value
-The key aspect that separates a decision making task from a value extraction task, for uncertainty visualisations, is the
+2) Use some risk utility function to set a threshold for an "acceptable level of risk"
+3) Extract the value for from the plot that will be compared to the threshold
+4) Make a judgement based off that value
 
-Decision making questions that typically involve uncertainty discuss the concept of "risk-aversion" as though it is a mistake.
-The irony of an uncertainty visualisation pper expressing the importance of uncertainty visualisation, but setting up a ground truth or optimum strategy in which participants *should* ignore the uncertainty information is not lose on us.
-This perspective does not make sense because it implies participants should make decisions entirely based on the average.
-A possible soltion for this is to provide a utility framework for a particular experiment [@Hullman2016], however it is unclear how easily participants could adopt a utility framework that is different to their own.
+The key aspect that separates a decision making task from a value extraction task, for uncertainty visualisations, is the inclusion of a utility function.
+While this might seem like a small change, it actually completely warps the experiment in unexpected.
+
+The first issue is that individuals will all have their own individual risk utility function, that is, how much they want to avoid or engage with risks or uncertainty.
+Therefore decision making experiments have an additional layer of noise that value extraction experiments do not.
+We cannot be sure if particpants answered differently from the ground truth because a visualisation was difficult to read *or* because the participants risk utility function did not align with the one set by the authors.
+A possible soltion for this is to provide a utility framework for a particular experiment [@Hullman2016], however this method seems to just introduce more noise.
 @Cheong2016 introduced a very simple optimum strategy into their experiment on bushfire uncertainty visualisations, where participants should choose to evacuate a house as soon as its likelihood of catching fire is 50% or higher.
+If the participants *did* correctly identify the optimum strategy and answer accordingly, the "decision making experiment" was actually *just a value extraction experiment*.
+The decision making aspect ceased to matter, all that mattered was identifying if the probability of a fire was greater than 50%.
 Interestingly, despite this very obvious and simple tactic to maximize payout from the experiment, it seems like many participants did not adopt it, instead acting much more warily as though they were considering whether or not they would *actually* evacuate in the event of a fire.
-These papers also very rarely consider what the optimum decision would be in a given scenario, and a difficulty with anticipating the best strategy is conflated with the failings of the uncertainty visualisation.
-Ultimately, we agree with @Hullman2016 who suggested that what determines an approripate ground truth is largely a philosophical exercise.
+This means it is likely that introducing a payment scheme or a utility function creates too much mental labour for the participants to answer "correctly", and they would rather stick with their own risk utility function that feels *natural* than go through the mental labour of adopting another.
+Setting up an experiment that requires participants to do laborious calculations or strenuous mental effort for a small marginal benefit is unlikely to result in sucessful participant responses because it is not in the "spirit" of visualisation.
 
-Risk and uncertainty are not only different in how they are communicated, but also why they are communicated.
-Risk communication focuses on communicating probability around events.
-These events are typically a binary event that is known to be random, so a version of communication that does not mention risks (e.g.
-telling someone they will or will not get cancer instead of communicating their risks about it) inherently carries with it a lack of transparency.
-Additionally risk communication is often communicated to avoid or information about an unfavorable outcome, so framing can be important if we want to influence decisions in a certainty way.
-These motivations do not extend to the communication of uncertainty information more broadly.
+It is clear that audiences primarily interact with visualisations through their "System 1" brain [@daniel2017thinking].
+This aligns with other authors comments that visualisation is primarily about "gists" [Spiegelhalter2017].
+In this sense, asking participants a question that requires them to shift to their "System 2" is entirely contrary to *why* we use visualisation.
+The exact reasons we describe visualisation to be so powerful, that they are able to communicate complicated information fast and efficiently *is a product of the system 1 brain* [@daniel2017thinking], however this means displaying information using visualisations have the same weakness' as system 1 thinking, and visualisation authors need to be aware of this.
+Asking participants to answer questions that require complicated calculations does not reflect how visualisations are used, and will liekly be ignored by the participants (as seen in many of these decision making tasks).
+While we agree with @Hullman2016 who suggested that what determines an approripate ground truth is largely a philosophical exercise, we differ in the sense that we do not think it should be done, lest you create a needlessly complicated value extraction task.
 
-There are some elements of risk communication that seem to translate to uncertaitny communication more broadly.
-Using words to communicate risk is discouraged because people can misinterpret the actual risk involved, so when possible numerical estimates of risk should be communicated, however this can make information harder to understand for those with poor numeracy skills [@Spiegelhalter2017].
-People performing better and preferring with numerical estimates of risk translates to uncertainty communication more broadly *inc (citations for this)*.
-Risk communication also needs to have clear objectives, use plain language, limit information to only what is necessary, and segment the audience to allow for differences in interest and knowledge [@Spiegelhalter2017].
-These general concepts of communication also extend to communication of uncertainty.
-On the other hand, it is not sensible to assume that probability specific concepts such as framing, intuitiveness of probabilities, or reference classes are still relevant when we consider uncertainty communication as signal supression, not as a synonym of risk or probability communication.
-In this same vein, most uncertainty communication papers very often discuss risk-aversion as something that needs to be considered despite it being unclear exactly how risk-aversion is related to uncertainty communication *(citations for the papers that do this)*.
+Additionally, several papers mention "risk-aversion" in participant decisions as a negative by-product of a particular visualisation design, however that displays a failure to understand that risk aversion is *not* a mistake, but rather it *reveals the utility of decreased uncertainty*.
+It may be interesting to investigate why a particular visualisation elicits increased risk-aversion, for example is it due to poorer estimates or increased awareness of negative outcomes, however simply reporting this as "risk aversion" conflates these factors.
+The work that advocates for transparency in risk-communication gives you whiplash when it's subtext argues that transparency leads to "incorrect" conclusions.
+There have been other discussions on the appropriateness of discussing risk-aversion as a bias [@Vranas2000], but few have made the point that if uncertainty holds some *value* there is not technically a "right" decision at all.
 
-In the same sense that risk and uncertainty are different, risk-aversion and amiguity-aversion (which can be considered the uncertainty version of risk aversion) are also not the same.
-Risk-aversion is the tendency of people to prefer outcomes of low uncertainty to those outcomes with high uncertainty despite the outcomes with high uncertainty having a higher or equal expected outcome.
-In the case of communicating risk, the only reason this heuristic would be relevant was if risk-aversion was a mistake that needed to somehow be corrected for, rather than a heuristic that *reveals the utility of decreased uncertainty*.
-Risk-aversion is often treated as a mistake or something that should be avoided by the uncertainty visualisation literature even though that is not necessarily true, depending on what is causing the risk-aversion (e.g.
-is it due to poorer estimates or increased awareness of negative outcomes).
-If a particular communication method leads people to place a greater value on certainty or make worse estimates, that is the aspect of importance, not risk-aversion which conflates these factors.
-The work that advocates for transparency in risk-communication gives you whiplash when it's subtext argues that transparency leads to "incorrect" conclusions.
-Risk-aversion has a large number of confounding factors that make it difficult to understand if it is something to be considered in risk communication, ambiguity-aversion has a similar problem.
-Ambiguity-aversion is the tendency of people to prefer to take on certain risks rather than unknown risks.
-It is essentially risk-aversion applied to risk.
-For this reason discussions of ambiguity-aversion have the same issues as the conversations about risk-aversion, where they do not consider ambiguity-aversion to be an estimate of the utility of decreased uncertainty, it contains a subtext that argues against transparency, and it is considered a valuable insight in of itself despite conflating many factors.
+The final issue with decision making tasks is that calculating "optimum" choices is not a task we use uncertainty for, it is a task that utilises *risk*.
+Risk and uncertainty are slightly different, as risk is known probabilities and uncertainty is unknown probabilities, so uncertainty is what you get when you cannot accurately define risk [@Spiegelhalter2017].
+Communicating risk allows people to weigh up options and make an optimum decision, uncertainty hedges information to let people know there may not even be an optimum strategy.
+In this sense, many "uncertainty" decision making experiments are actually just risk decision making experiments.
 
-The idea that risk and ambiguity aversion reveal the utility of decreased variance seems to seldom be considered in the literature, therefore it should be mentioned here, otherwise this issue might continue.
-However, considering risk aversion is at its core, a trade off between bias and variance, it is bizzare to assume it to be a mistake when statisticians perform these trade offs all the time.
-If risk-aversion is truly "irrational" behaviour, every statistical textbook that discusses the use of a biased but consistent estimator in a finite sample case (a common example would be specific cases of the maximum likelihood estimator) should be tossed out.
-Discussing risk and ambiguity aversion as though it is some diversion from a perfectly rational, mathematical choice, implies there is *no value to a decrease in variance which is not even true within mathematics*.
-There have been other discussions on the appropriateness of discussing risk-aversion as a bias [@Vranas2000], but few have made the point that if uncertainty holds some *value* there is not technically a "right" decision at all.
-The concept of risk aversion treats risk as "noise", as though it is something that only exists to be ignored, but if the common place use of biased MLE's is an indication of anything, it is that *ignoring uncertainty* is the irrational choice.
-
-While this is a focus on uncertainty visualisation, these issues in uncertainty communication need to be acknowledged because they bleed into the uncertainty visualisation literature.
-Uncertainty visualisation are compared on their ability to communicate explicit risk, elicit trust, and prevent risk or ambiguity avoidance, continuing these issues in uncertainty communication at large.
-While uncertainty visualisation can make it easier to communicate the complicated details of uncertainty information, it carries some challenges that are unique to the visualisaition, specifically with respect to the "signal supression" concept.
-In simple estimates or verbal communication, the signal is often easy to identify because it is what we are explicitly saying.
-Visualisations are used in both data exploration and communication.
-This means what exactly is a *signal* in any particular visualisation is hard to identify, since we often let the visualisation *tell us* what the signal is.
-Additionally, you cannot add noise to *every single possible* signal one might take from a visualisation.
-Two people looking at the same visualisation might, just by chance, develop two entirely different insights.
-These unique and facinating challenges that are faced by viewing uncertainty visualisaiton through the lense of "signal supression" have been almost completely untouched by the literature.
-
-What is meant by "uncertainty" may seem obvious to some, but when you attempt to quantify or visualise it you will quickly find yourself asking, "uncertainty about...
-what?".
-Do you mean uncertainty on an estimate?
-On a forecast?
-How many steps ahead is this forecast?
-Are we only considering the uncertainty in the estimate or in the parameters or are we considering the possibility of measurement error or biased inputs?
-Signal and noise can only be untangled in the presence of a motivating question.
-
-The previous sections of this paper established several issues in uncertainty communication at large.
-In this section, we will discuss the current literature on uncertainty visualisation, how the uncertainty communication issues more broadly manifest in the specific case of visualisation, and what general lessons we can take from the existing work.
+This leaves us with several issues in decision making experiments.
+They are noisy because of participants individual utility functions, and when that is not the case they are thinly veiled and over complicated value extraction experiments, and on top of that, they dont even evaluate uncertainty as uncertainty, they evaluate it as risk.
+For these reasons, uncertainty visualisation experiments should either be designed with these caveates in mind and be designed to untangle why a particular visualisation will elicit a specific decision, or the field should steer away using decision quality and related metrics as a basis for evaluating visualisation performance.
 
@@
 ## Suggestions to measure uncertainty
+While uncertainty visualisation can make it easier to communicate the complicated details of uncertainty information, it carries some challenges that are unique to the visualisaition, specifically with respect to the "signal supression" concept.
+In simple estimates or verbal communication, the signal is often easy to identify because it is what we are explicitly saying.
+Visualisations are used in both data exploration and communication.
+This means what exactly is a *signal* in any particular visualisation is hard to identify, since we often let the visualisation *tell us* what the signal is.
+Additionally, you cannot add noise to *every single possible* signal one might take from a visualisation.
+Two people looking at the same visualisation might, just by chance, develop two entirely different insights.
+These unique and facinating challenges that are faced by viewing uncertainty visualisaiton through the lense of "signal supression" have been almost completely untouched by the literature.
+
 ### Heuristic checks

### 2024-03-21 writingdumpster.qmd
finished decision making section

@@
 # Main section outtakes
-These results provide some insights that applies to visualisation more broadly but nothing specific to uncertainty informaiton.
-It is clear that audiences primarily interact with visualisations through their "System 1" brain [@daniel2017thinking].
-This aligns with other authors comments that visualisation is primarily about "gists" [Spiegelhalter2017].
-In this sense, asking participants a question that requires them to shift to their "System 2" is entirely contrary to *why* we use visualisation.
-The exact reasons we describe visualisation to be so powerful, that they are able to communicate complicated information fast and efficiently *is a product of the system 1 brain* [@daniel2017thinking], however this means displaying information using visualisations have the same weakness' as system 1 thinking, and visualisation authors need to be aware of this.
-Asking participants to answer questions that require shifting to their system 2 brain is not in the spirit of visualisation, and does not reflect how visualisations will be used in practice, and will very likely result in incorrectly answered questions.
-There is a large number of examples where participants got "confused" but the problem is more likely that the question required participants to shift to their system 2 brain.
+While this is a focus on uncertainty visualisation, these issues in uncertainty communication need to be acknowledged because they bleed into the uncertainty visualisation literature.
+Uncertainty visualisation are compared on their ability to communicate explicit risk, elicit trust, and prevent risk or ambiguity avoidance, continuing these issues in uncertainty communication at large.
 
+Ambiguity-aversion, uncertainties risk-avoidance counterpart, has a similar problem.
+Ambiguity-aversion is the tendency of people to prefer to take on certain risks rather than unknown risks.
+It is essentially risk-aversion applied to risk.
+For this reason discussions of ambiguity-aversion have the same issues as the conversations about risk-aversion, where they do not consider ambiguity-aversion to be an estimate of the utility of decreased certainty.
+Assuming that participants should act as rational agents implies they should do as rational agents do and *ignore* the uncertainty information.
+
+The idea that risk and ambiguity aversion reveal the utility of decreased variance seems to seldom be considered in the literature, therefore it should be mentioned here, otherwise this issue might continue.
+However, considering risk aversion is at its core, a trade off between bias and variance, it is bizzare to assume it to be a mistake when statisticians perform these trade offs all the time.
+If risk-aversion is truly "irrational" behaviour, every statistical textbook that discusses the use of a biased but consistent estimator in a finite sample case (a common example would be specific cases of the maximum likelihood estimator) should be tossed out.
+Discussing risk and ambiguity aversion as though it is some diversion from a perfectly rational, mathematical choice, implies there is *no value to a decrease in variance which is not even true within mathematics*.
+There are some elements of risk communication that seem to translate to uncertaitny communication more broadly.
+Using words to communicate risk is discouraged because people can misinterpret the actual risk involved, so when possible numerical estimates of risk should be communicated [@Spiegelhalter2017] the same rules apply to uncertainty (*Cite gap __*).
+People performing better and preferring with numerical estimates of risk translates to uncertainty communication more broadly .
+Risk communication also needs to have clear objectives, use plain language, limit information to only what is necessary, and segment the audience to allow for differences in interest and knowledge [@Spiegelhalter2017].
+These general concepts of communication also extend to communication of uncertainty.
+On the other hand, it is not sensible to assume that probability specific concepts such as framing, intuitiveness of probabilities, or reference classes are still relevant when we consider uncertainty communication as signal supression, not as a synonym of risk or probability communication.
+In this same vein, most uncertainty communication papers very often discuss risk-aversion as something that needs to be considered despite it being unclear exactly how risk-aversion is related to uncertainty communication *(citations for the papers that do this)*.
+Very often the decisions made in decision making tasks are compared to some "correct" choice, typically the choice of a rational agent.
+That is, visualisations that elicit choices that are most closely aligned with those of rational agents are seen as the best.
+There is a critical error with this way of thinking.
+If a group truth is specified (through right/wrong answers or a payment scheme) to correctly answer the question, the user needs to work out the threshold that will maximise payout, over-ride their personal risk utility function to adopt this threshold, and report values that align with this threshold.
+This is just a value extraction task with additional unnecessary extra steps.
+Additionally, some authors speak about plots exhibiting "risk-avoidance" which is comical, considering a rational agent who displays no risk-avoiding will simply make the choice that maximise expected value, or in other words **completely ignores the uncertainty information**.
+The ground truths in these experiments *encourage* participants to think of uncertainty information as useless noise that should be ignored, the way a rational agent will.
 There is also the common view that uncertainty as a dimensionality problem.

### 2024-03-22 paper.qmd
Finished the performance section rewrite

@@
 The first issue is that individuals will all have their own individual risk utility function, that is, how much they want to avoid or engage with risks or uncertainty.
-Therefore decision making experiments have an additional layer of noise that value extraction experiments do not.
+Therefore decision making experiments that have an additional layer of noise that value extraction experiments do not.
 We cannot be sure if particpants answered differently from the ground truth because a visualisation was difficult to read *or* because the participants risk utility function did not align with the one set by the authors.
-A possible soltion for this is to provide a utility framework for a particular experiment [@Hullman2016], however this method seems to just introduce more noise.
-@Cheong2016 introduced a very simple optimum strategy into their experiment on bushfire uncertainty visualisations, where participants should choose to evacuate a house as soon as its likelihood of catching fire is 50% or higher.
+Several authors have offered solutions to this issue, however the problem is deeper than any of them seem to realise.
+
+@Hullman2016 suggested providing a utility framework for each experiment, to instruct the participants in how to account for the uncertainty information.
+This is a method that seems to have been adopted by @Fernandes2018 who describe the following scenario to participants who are trying to maximise the coins in their experiment:
+> Subjects gain coins for every minute they are able to continue an activity that is valuable to them (e.g., watching TV at home) before going to the bus stop, and gain a bonus for arriving at their intended destination early.
+Subjects incur a coin penalty for time spent waiting at a stop for a bus to arrive.
+
+Any payment scheme that is incorporated into an experiment will also implicitly set up a utility framework, as it is assumed participants will try to maximise their payments.
+In tasks such as this, the ground truth of each question is typically selected to be the value a ration agent would select and visualisations are evaluated on the basis of how far the participants repsponses are from that ground truth.
+The problem with this, is that all the complexity of the question is in *working out* the best response, and it has little to do with the visualisation itself.
+The visualisation aspect of these studies becomes a value extraction experiment.
+
+@Cheong2016 tested multiple different visual representations of uncertainty for representing the likelihood of a house being burned down based on its location.
+Their payment scheme, which paid out $0.10 for a correct choice (i.e.
+staying when the house was not burned down or leaving when the house was burned down) and 0 for an incorrect choice (i.e.leaving when the house didnt burn down or staying when the house burned down), meant participants were incentivised to base their entire leave/stay decision on whether or not the likelihood of a fire at their house is above or below 50%.
 If the participants *did* correctly identify the optimum strategy and answer accordingly, the "decision making experiment" was actually *just a value extraction experiment*.
@@
 This means it is likely that introducing a payment scheme or a utility function creates too much mental labour for the participants to answer "correctly", and they would rather stick with their own risk utility function that feels *natural* than go through the mental labour of adopting another.
-Setting up an experiment that requires participants to do laborious calculations or strenuous mental effort for a small marginal benefit is unlikely to result in sucessful participant responses because it is not in the "spirit" of visualisation.
+Setting up an experiment that requires participants to do laborious calculations or strenuous mental effort for a small marginal benefit is unlikely to result in successful participant responses because it is not in the "spirit" of visualisation.
+(*Cite Gap 16*)
 
@@
 The exact reasons we describe visualisation to be so powerful, that they are able to communicate complicated information fast and efficiently *is a product of the system 1 brain* [@daniel2017thinking], however this means displaying information using visualisations have the same weakness' as system 1 thinking, and visualisation authors need to be aware of this.
-Asking participants to answer questions that require complicated calculations does not reflect how visualisations are used, and will liekly be ignored by the participants (as seen in many of these decision making tasks).
+Asking participants to answer questions that require complicated calculations does not reflect how visualisations are used, and will liekly be ignored by the participants.
 While we agree with @Hullman2016 who suggested that what determines an approripate ground truth is largely a philosophical exercise, we differ in the sense that we do not think it should be done, lest you create a needlessly complicated value extraction task.
@@
 In this sense, many "uncertainty" decision making experiments are actually just risk decision making experiments.
+Not only this, but the inclusion of uncertainty information should *not* impact the choices of a rational agent.
+A well behaved rational angent *should* ignore uncertainty information, so long as the estimate provided is still the best case.
+This point seems to be lost on many uncertainty visualisation authors.
+@Zhao2023 displayed a model's prediction and its estimated uncertainty and asked participants if they wanted to submit the model estimate as their answer, or make their own prediction.
+At no point did the authors seem to realise a rational agent would accept the model estimate every time, no matter how uncertain the information was, and participants were actually disqualified from the study for taking this approach to answering the questions.
+(*Cite Gap 17*)
 
 This leaves us with several issues in decision making experiments.
-They are noisy because of participants individual utility functions, and when that is not the case they are thinly veiled and over complicated value extraction experiments, and on top of that, they dont even evaluate uncertainty as uncertainty, they evaluate it as risk.
+They are noisy because of participants individual utility functions and when that is not the case they are thinly veiled and over complicated value extraction experiments.
+Additionally, authors dont even seem to agree how participants should incoreperate uncertainty information, with some authors designing complicated utility functions, and others not realising the best response to their experiment is to ignore uncertainty information entirely.
 For these reasons, uncertainty visualisation experiments should either be designed with these caveates in mind and be designed to untangle why a particular visualisation will elicit a specific decision, or the field should steer away using decision quality and related metrics as a basis for evaluating visualisation performance.
@@
 #### Trust and confidence
-Authors also occasionally use trust as a proxy for uncertainty.
-Similar to measuring uncertainty through decision quality, this method also has its own series of issues.
+Authors also often measure the impact of visualising uncertainty on participants trust in an estimate.
+Trust is a by product of displaying uncertainty rather than the goal of it, and viewing the relationship in the converse direction can lead to misguided research.
+
 If the purpose of displaying uncertainty information is to appropriately hedge a signal with noise, then it should be assumed that trust is only related to uncertainty communication through increased transparency and honesty.
-It does seem to be the case that uncertainty communication increases trust because it is a proxy for transparency *(add citations)*.
-Unfortunately a large proportion of the literature discusses trust as something that is directly related to uncertainty visualisation rather than as an observable by product.
-In viewing uncertainty communication as directly related to trust, not related through the proxy of transparency, several unobserved variables are conflated.
-The amount of uncertainty that is present in the information, whether or not the information is trustworthy or makes sense, prior beliefs of the participants, and the trustworthiness of the source are a few examples of variables that are conflated when authors directly consider trust to be of direct interest.
-Considering trust, and not transparency, as the metric of importance in uncertainty communication leads to a questionable subtext that argues against transparency.
-@Hullman2020a found that author simultaneously argued that failing to visualise uncertainty was akin to fraud, but also many avoided uncertainty visualisation because they didn't want their work to come across as "untrustworthy".Being concerned about audiences perception of trust, without first establishing if what we are communicating is *trustworthy*, leads to the implication that details that result in information not being considered trustworthy should be avoided.
-Authors that imply that a decrease in trust or an increase in ambiguity or risk aversion are metrics of importance in of themselves do not understand the purpose of uncertainty communication as methods to increase transparency.
-Science communication should be primarily concerned with accuracy, setting trust and risk-aversion as the variables of interest implicitly encourages statisticians to set trust and risk-aversion as the primary goals of communication.
-The issue of trust being divorced from trustworthiness has been commented on by other authors [@ONeill2018], however the issue still persists in the uncertainty visualisation literature [@Zhao2023].
+Considering trust, and not transparency, as the metric of importance in uncertainty communication can lead to a questionable subtext that argues against transparency, something that has been noticed by several other authors [@Spiegelhalter2017; ONeill2018].
+@Hullman2020a found that author simultaneously argued that failing to visualise uncertainty was akin to fraud, but also many avoided uncertainty visualisation because they didn't want their work to come across as "untrustworthy".
+These authors are optimising *trust* rather than *transparency*, which means they opt to leave out uncertainty information when it does exactly what it is supposed to, decrease certainty in conclusions.
 
-Additionally, some studies use a ground truth in these trust experiments, which makes for a very confusing read on exactly how participants are supposed to use the information. 
-- @Zhao2023 displayed a model prediction with uncertainty and took participants using the model prediction as a sign of trust.
+Science communication should be primarily concerned with accuracy.
+Setting trust and risk-aversion as the variables of interest implicitly encourages statisticians to set trust and risk-aversion as the primary goals of communication.
+The issue of trust being divorced from trustworthiness has been commented on by other authors [@ONeill2018], however the issue still persists in the uncertainty visualisation literature.
+@Zhao2023 displayed a several variations of a visualisation of a model prediction and its uncertainty and took participants using the model prediction as a sign of trust.
 They reported that visualising uncertainty information caused participants to trust the model in the low variance case, but the results in the high variance case were inconclusive.
-The discussion made it clear the authors thought the uncertainty information should make the visualisation more trustworthy, but conflating trust and the use of a prediction implied uncertaitny information should somehow influence participants to use their own prediction.
-If the Beuro of Meterology was uncertain about their rain prediction, that would not be sufficient conditions for me to decide to make my own rain predictions.
-Despite this, the authors seemed to assume that the uncertainty information *should* have an influence on that, showing they had not deeply considered how uncertainty inforormation should infludence the choices of the participants.
+The discussion made it clear the authors thought the uncertainty information should make the visualisation more trustworthy, but conflating trust and the use of a prediction implied uncertaitny information should somehow influence participants to use their own prediction, even though a prediction being uncertainty does not necessarily mean it is incorrect.
+Despite this, the authors seemed to assume that the uncertainty information *should* have an influence on that, showing they had not deeply considered *how* uncertainty information should influence the choices of the participants.
+(*Cite Gap 18*)
 
@@
 
-#### Deterministic questions
-- You cannot make a deterministic statements about a random variable (citation: the entire field of probability).
-Many of these issues just boil down to a really pervasive misunderstanding of statistics.
-I do not expect the general population to understand this fact about random variables, however *I would expect* researchers investigating *methods of statisticl communication* would at least understand enough to ask appropriate questions that *are answerable*.
-- while we are here, you cant answer a probability question with a deterministic. I cannot say yes or no to a probability question without providing a threshold. 
-- Basically, you need a probability threshold to answer these questions, something it does not seem like the authors are collecting, and I am unsure if the authors are aware of this themselves because they never mention it.
+#### Other (questionable) attemps to capture uncertainty
+There is also a swath of studies that are aware a question that boils down to a value extraction experiment, or a question that should be answered by literally ignoring uncertainty information is not what we want when we consider uncertainty visualisations.
+These papers often try to ask a question that should utilise both the uncertainty and signal in the response, however this is rarely what actually occurs.
+This method typically results in in cryptic or confusing questions that create a large amount of noise on the interpretation side of the analysis [@Hullman2016].
 
-#### Other (questionable) attemps to capture uncertainty
-There are other studies that try to include uncertainty information without directly asking for it but do not go down the route of asking for a decision or levels of trust.
-These papers often try to ask a question that should utilise both the uncertainty and signal in the response.
-This is an interesting approach, but these papers are often still looking for a "signal" which is specified in the form of a "ground truth" that is used to evaluate the plot.
-Unfortunately this usually comes in the form of slightly cryptic questions that create a large amount of noise on the interpretation side [@Hullman2016].
+Some authors opt for asking slightly vague questions that imply a use of uncertainty, but compare it to a ground truth that is very specific.
+@Ibrekk1987 asked participants for the "best estimate" which was evaluated in accuracy by comparing it to the mean, however the "best estimate" depends on the loss function we are using, and a loss function of minimised error was not implied by the question.
+@Hofmann2012 showed two distributions in 20 different visualisations (a lineup protocol) using a jittered sample, a density plot, a histogram, and a box plot and asked participants.
+Participants were asked to report in which of the plots was "the blue group furthest to the right" The experiment set up is shown in @fig-right.
+The participants answers were then compared to a ground truth where the correct plot had a blue distribtuon with a right shifted mean.
+By comparing the results to a ground truth statistic and marking participants as "wrong" or "right", the error from the participants that had an alternative interpretation to the concept of "furthest right" was conflated with the error from a the visualisation choice.
+These papers make it unclear if the participants got the answers wrong because they misunderstood the question or because of something related to the plot.
+Therefore, this method leads to inconclusive results about the plot design, and is not advised.
 
-Cite gap __: Weird questions to capture uncertainty
-- @Correll2018 “place your 5 ships on the safest locations on the board.” (minimise danger)
-- @Ibrekk1987 asked participants for the "best" estimate while displaying a population density function.
-The participants proided the mode instead of the mean, although there was no indication the mean was what was being asked about.
-- @Hofmann2012 showed two distributions in 20 different visualisations (a lineup protocol) using a jittered sample, a density plot, a histogram, and a box plot.
-Participants were asked to report in which of the plots was "the blue group furthest to the right" and to provide reasoning from a multiple choice list and grade their certainty.
-The experiment set up is shown in @fig-right.
-The participants answers were then compared to a ground truth where the correct plot had a right shifted mean.
-By comparing the results to a ground truth statistic and marking participants as "wrong" or "right", the error from the participants that had an alternative interpretation to the concept of "furthest right" was conflated with the error from a the visualisation choice.
-- @Padilla2017 made several confusing assumption when determining the ground truth of their experiments that were assessing storm visualisations.
-Experiment 2 assumed that the participants should provide responses that indicated they perceived the intensity of the storm to be independent of the uncertainty distribution of its location.
-This assumption seems counterintuitive as best, however experiment 3 asked participants to "decide which oil rig will receive more damage based on the depicted forecast of the hurricane path" for which they *were* supposed to incorperate uncertainty information and not assume they were independent.
-The authors seemed to be unaware of these confounding mistakes in their assumptions of how the participants were supposed to utilise the uncertainty information.
-- @Sanyal2009 mapped uncertainty to dots and signal to a 3D surface and asked participants to identify areas of high and low signal and high and low uncertainty.
-Participants were not asked to combine that information in any way, and the signal and the noise were treated as separate variables.
-- @Correll2014 asked participants to extract the mean and variance from four uncertainty visualisations, bar charts with error bars, box plots, gradient plots, and violin plots (that were adjusted to make sure the mathematical information they were showing was the same) from two side by side distributions.
-Participants were also asked to answer a question that was supposed to incorperate both the signal and the noise, such as "How likely is candidate B to win the election?" when the two distributions indicated voter preference.
+![This shows the user interface for the experiment performed by @Hofmann2012. The question of "furthest to the right" is open to interpretation. ](furthestright.png){#fig-right}
+
+Another method used by authors is to ask a deterministic question about a random event.
+@Padilla2017 provided participants with a visualisation of the cone of uncertainty and asked then to "decide which oil rig will receive more damage based on the depicted forecast of the hurricane path".
+The cone of uncertainty provides a 60% confidence interval for the location of the eye of a hurricane, which allows us to know the area where the eye of the storm will go, it does not given any information about the intensity of a storm, the size of a storm, or even if a location will be hit.
+This inclusion of determinism seems to cause the authors to stumble themselves, as they are not consistent with their assumptions.
+In their first experiment @Padilla2017 indicated the correct answer was to assume that the storm was equally intense no matter how far from the center of the distribution an oil rig was, however answering their third experiment correctly hinged on assuming the intensity of the storm at a particular point (which in this experiment they phrased as damage) *does* change in intensity as you move away from the center of the distribution.
+Given these conflicting assumptions, it is unclear how the participants were supposed to adjust the probabilistic path information to answer a deterministic question about which oil rig would receive the most damage.
+The flipside of this is asking participants for a deterministic answer to a probabilistic question.
+@Correll2014 asked participants "how likely is candidate B to win the election?" when the two distributions indicated voter preference.
 Participants were not able to answer the question about likelihood in term of probability, but were instead given seven options from 1=Outcome will be most in favor of A to 7=Outcome will be most in favor of B.
 The ground truth statistic for this question was a scalar multiple of Cohen’s d, indicating participants were supposed to incorperate uncertainty information using a very specific formula that was likely unknown to them but assumed to be used implicitly.
-- @Cheong2016 tested multiple different visual representations of uncertainty for representing the likelihood of a house being burned down based on its location.
-Their payment scheme, which paid out $0.10 for a correct choice (i.e.
-staying when the house was not burned down or leaving when the house was burned down) and 0 for an incorrect choice (i.e.leaving when the house didnt burn down or staying when the house burned down), meant participants were incentivised to base their entire leave/stay decision on whether or not the likelihood of a fire at their house is above or below 50%.
-The authors failled to recognise that this "complicated decision making task" boilled down to a simple value extraction problem, which the text made easiest to extract.
-The results from the experiment indicated that many of the participants also failled to recognise the obvious and simply strategy.
 
-- @Blenkinsop2000 tailored their questions according to the visualisation shown, which implies an understanding that different visualisations contain different information.
-Despite this the authors still compared the visualisations according to which were "useful for uncertainty visualisation".
+The final method used by authors is to just explicitly ask about uncertainty and signal information separately.
+@Sanyal2009 mapped uncertainty to dots and signal to a 3D surface and asked participants to identify areas of high and low signal and high and low uncertainty.
+Participants were not asked to combine that information in any way, and the signal and the noise were treated as separate variables.
+@Correll2014 asked participants to separately extract the mean and variance from four uncertainty visualisations.
+These methods explicitly view the uncertainty and signal as two separate variables that should be extracted from a plot, and not two variables that should be interpreted together.
+Even viewing these questions as a routine check to make sure the signal information isn't impacted by the uncertainty is counter intuitive, because the whole point *of* the uncertainty is to impact the signal information.
 
-- @Blenkinsop2000 did try to include uncertainty as noise, asking participants to search for a specific outcome (the land classified as grass) in a random .
-However the participants failing to identify the signal (something that would be expected of an uncertainty plot) was seen as a sign the task was too complicated.
-While the task in this experiment *was* certainty far too complicated (and poorly communicated) this discussion indicates the authors did not consider the signal suppression to be the goal of the visualisation, and rather an annoying distraction.
-
-- @Ibrekk1987 asked participants for the "best estimate" which was provided in terms of most probable value (mode), the mean and the median, the "ground truth" statistic used was the mean, despite the "best estimate" largely depending on the loss function of the model and the methods used.
-
-- @Ibrekk1987 used several displays of a probability density functions, since “formally equivalent representations are often not psychologically equivalent”.
-
-- @Ibrekk1987 highlights that in the face of a vague question, participants will use the plot to decide what the authors mean. The authors take this as a
-
-- @Boone2018 organised responses from participants by the approach they used and tried to account for participants prior beliefs (although the assessment of those beliefs used the word "likely" which is open to interpretation).
-
-There is so little understanding of information in uncertainty graphics that participants are frequently provided with information that is insufficient to answer the question at all, an issue that is regularly missed by the experimenters.
-In order to test whether or not people incorrectly interpreted the cone of uncertainty that is used to communicate cone tasks, experimenters regularly do not give participants enough information to answer the questions [@Padilla2017; @Boone2018].
-The cone of uncertainty provides a 95% confidence interval for the *eye* of a hurricane, which allows us to know *where the eye of the storm will likely go*, this does not, give us enough information to answer if the storm will *hit* a particular point, because this requires information about the size and intensity of the storm, *and* an assumption about whether or not the size and intensity will change over time.
-Since these experiments are often trying to *test* for common misreadings of the plot, such as only thinking the area inside the cone will be damaged or believing the cone represents size and the storm is getting bigger.
-The authors often fail to recognize that the exact information that has been witheld for testing is necessary for answering their questions, and considering that information is necessary to mark participants as correct or incorrect, it is hard to understand exactly how they evaluated the experiment participants.
-@Ibrekk1987 expected participants to calculate the mean using a pie chart that had numerical bins combined (e.g.
-segment one was 0-2 inches of snow, and segment two was 2-4 inches of snow) so it could not be done by calculation.
-The authors also expected participants to use a CDF to calculate the mean, which needs to be converted *back* into a PDF to do so mathematically, so it was unclear how participants were supposed to do this calculation visually unless they were to just make a blind guess.
-The authors interpreted this as the participants misinterpreting how the CDF works, not seeming to realise they gave them an impossible task.
-In light of this, the main take away from these papers seems to be "if you do not give people the necessary information to answer a question, they will be unable to answer it", which does not seem interesting enough to justify several papers.
-
-![This shows the user interface for the experiment performed by @Hofmann2012. The question of "furthest to the right" is open to interpretation. ](furthestright.png){#fig-right}
+These examples are a bit complete mishmash of methods, however they point to a larger issue that goes beyond decision making, trust and confidence experiments.
+Authors *have no idea* how to evaluate the effects of uncertainty in an uncertainty visualisation.
 
@@
 ## Suggestions to measure uncertainty
-While uncertainty visualisation can make it easier to communicate the complicated details of uncertainty information, it carries some challenges that are unique to the visualisaition, specifically with respect to the "signal supression" concept.
+So, the current methods of measuring or understanding the role of uncertainty in a visualisation is quesitonable at best, however this is not because visualiation authors are missing the mark, but rather uncertainty is *particularly* difficult to express in a visualisation.
 In simple estimates or verbal communication, the signal is often easy to identify because it is what we are explicitly saying.
-Visualisations are used in both data exploration and communication.
+Unlike statistical models, visualisations are used in both data exploration and communication.
 This means what exactly is a *signal* in any particular visualisation is hard to identify, since we often let the visualisation *tell us* what the signal is.
 Additionally, you cannot add noise to *every single possible* signal one might take from a visualisation.
-Two people looking at the same visualisation might, just by chance, develop two entirely different insights.
-These unique and facinating challenges that are faced by viewing uncertainty visualisaiton through the lense of "signal supression" have been almost completely untouched by the literature.
+Two people looking at the same visualisation might, just by chance, develop two entirely different insights and draw inference on two completely different statistics.
+These unique and fascinating challenges that are faced by uncertainty visualisaiton have been completely untouched by the literature.
+This section will cover some interesting research in uncertainty visualisation and suggestions for better ways to measure uncertainty.
 
@@
 ### Just noticiable signal
-Connecting our evaluation of uncertainty back to the reasons uncertainty is believed to be different is likely the key to improving evaluation studies.
-Uncertainty acts as a form of statistical "hedging" for signals found in an analysis.
-Interviews with experts in statistic back up this primary motivation, as ignoring uncertainty information of often expressed as being similar to fraud or lying and the goal of "improving decisions" is only seen as a secondary outcome of an appropriately hedged signal [@Hullman2020a; @Manski2020].
+Alternatively visualisation research could shift away from the accuracy concept all together and either allow for open ended responses, or ask questions that are more abstract to try and better replicate the large structural understanding that comes with the insights gained from visualisation.
 
@@
 **Experiment rant from confirmation**
+The field could take a more "exploratory" approach to experimentation and ask multiple accuracy questions for multiple visualisations with different visual encodings and try to identify the *trade-off* in the information a particular plot conveys.
+Data visualisation is commonly utilised as a tool in data exploration, so it is not uncommon for a data analyst to make a plot with only a vague goal and pull out a large number of adjacent observations.
+This experimental framework could replicate this process.
 
@@
 # Great Examples 
+(good decision making example) @Boone2018 organised responses from participants by the approach they used and tried to account for participants prior beliefs (the method of eliciting prior beleifs was, however, questionable).
 

### 2024-03-22 writingdumpster.qmd
Finished the performance section rewrite

@@
 
+Cite Gap 16: Examples of value extraction decision making tasks
+
+Cite Gap 17: Examples of studies where the correct choice was to *ignore* the uncertainty information
+
+Cite Gap 18: Examples of studies where authors measure trust
+
 ## Extra Section Cite Gaps
@@
 # Moved over notes
+
+- @Ibrekk1987 expected participants to calculate the mean using a pie chart that had numerical bins combined (e.g.
+segment one was 0-2 inches of snow, and segment two was 2-4 inches of snow) so it could not be done by calculation.
+The authors also expected participants to use a CDF to calculate the mean, which needs to be converted *back* into a PDF to do so mathematically, so it was unclear how participants were supposed to do this calculation visually unless they were to just make a blind guess.
+The authors interpreted this as the participants misinterpreting how the CDF works, not seeming to realise they gave them an impossible task.
+- @Blenkinsop2000 tailored their questions according to the visualisation shown, which implies an understanding that different visualisations contain different information.
+Despite this the authors still compared the visualisations according to which were "useful for uncertainty visualisation".
 - @Blenkinsop2000 asked different questions for each uncertainty visualisation, indicating they had some awareness that the question has a significant impact on whether or not the visualisation should be used.
@@
 They believe this is their own mistake so they so they don’t present these results.
-
+- @Ibrekk1987 used several displays of a probability density functions, since “formally equivalent representations are often not psychologically equivalent”.
+- @Ibrekk1987 highlights that in the face of a vague question, participants will use the plot to decide what the authors mean. 
 @Sanyal2009 users performed significantly well using Glyph-size when the task was to search for locations of least uncertainty.
@@
 # Main section outtakes
+@Hofmann2012 showed two distributions in 20 different visualisations (a lineup protocol) using a jittered sample, a density plot, a histogram, and a box plot and asked participants.
+Participants were asked to report in which of the plots was "the blue group furthest to the right" and to provide reasoning from a multiple choice list and grade their certainty.
+The experiment set up is shown in @fig-right.
+The participants answers were then compared to a ground truth where the correct plot had a right shifted mean.
+By comparing the results to a ground truth statistic and marking participants as "wrong" or "right", the error from the participants that had an alternative interpretation to the concept of "furthest right" was conflated with the error from a the visualisation choice.
+
 While this is a focus on uncertainty visualisation, these issues in uncertainty communication need to be acknowledged because they bleed into the uncertainty visualisation literature.

### 2024-03-22 paper.qmd
finished rewrite

@@
 Additionally, authors rarely attribute this visual version of large sample theory to results where it is clearly applicable.
-@Kale2018 found that users were more sensitive to the underlying trend when shown the HOPs plot over the static outcoms plot, this effect went away when the speed of the animation increased, indicating that performance difference might be due to overplotting in the static case and a smaller sample size in the static condition would produce the same result.
+@Kale2018 found that users were more sensitive to the underlying trend when shown the HOPs plot over the static outcoms plot, this effect went away when the speed of the animation increased, indicating that performance difference might be due to overplotting in the static case and a smaller sample size in the static condition might produce the same result.
 The issue of blurred elements of the grammar of graphics in the case of distribution visualisations does not end here.
@@
 Uncertainty is noise, but most evaluation experiments measure it as signal.
+Uncertainty acts as a form of statistical "hedging" for signals found in an analysis.
+Interviews with experts in statistic back up this primary motivation, as ignoring uncertainty information of often expressed as being similar to fraud or lying [@Hullman2020a; @Manski2020].
 Noise and signal are not an inherrent property of variance, or meta-data or the aspects of an analysis that are usually secondary considerations, it is a property of the data that *promotes* a message we are trying to infer and the data that *supresses* it, if that message is invalid.
@@
 
-### Heuristic checks
-- Have a better hypothesis? Like its more specific. They are looking for a very specific thing *wrong* with a visualisation.
-- We wary of trying to control something and observe it at the same time
-
-A lot of work that identifies heuristics or biases in plots do not hinge on the previous assumptions.
+### More specific hypothesis
+Heuristic checks are useful because they look at unknown pitfalls that might exist in interpretation of current plots.
+Since the hypothesis for these experiments are usually quite specific, e.g.
+"do people perceive a an outcome that is within the bar as more likely than one outside it, even if both outcomes are the same distance from the mean?".
+This means they are less likely to fall into the trap of trying to answer questions that are *far* too broad to be answered with a single experiment (e.g.
+"is a scatter plot better at showing uncertainty than a box plot").
 This work also provides useful insights for experiments by highlights pitfalls participants might fall into when they review the results of evaluation experiments [@Hullman2016].
@@
 @Padilla2017 found that points that were on an outcome of an ensemble display were perceived as more likely than points not on an outcome, even when the point that was not on a specific outcome of the ensemble was closer to the mean of the uncertainty distribution.
-The sine illusion can cause the confidence interval of a smoothed sine curve to seem wider at the peaks than the troughs, causing us to underestimate uncertainty associated with changing values [@Vanderplas2015]
+The sine illusion can cause the confidence interval of a smoothed sine curve to seem wider at the peaks than the troughs, causing us to underestimate uncertainty associated with changing values [@Vanderplas2015].
 
-### Just noticiable signal
-Alternatively visualisation research could shift away from the accuracy concept all together and either allow for open ended responses, or ask questions that are more abstract to try and better replicate the large structural understanding that comes with the insights gained from visualisation.
+In a similar vein, experiments that verify smaller aspects of plot design might be more useful to the field in the long run because it helps contribute to a larger working theory of "how do we see visualisations".
+Many visualisation experiments try to compare two plots with several differences, but do not seem to be interested in the mechanisms by which we extract information from visualisations.
+Small perceptual tasks that seek to answer small but highly relevant questions (for example, if colour hue and colour value can be perceived as a single signal suppressed variable) would be useful to the field.
 
-### Thinking "smaller picture" with
-Many visualisation experiments try to compare two plots with *several* differences, but do not seem to be interested in the *mechanisms* by which we extract information from visualisations.
-Let us consider a decision making experiment, there are several steps that Simply asking "Do people make better decisions with
-
-**Experiment rant from confirmation**
-The field could take a more "exploratory" approach to experimentation and ask multiple accuracy questions for multiple visualisations with different visual encodings and try to identify the *trade-off* in the information a particular plot conveys.
+### Qualitative Studies
+Alternatively visualisation research could shift away from the accuracy concept all together ask questions that allow for open ended responses.
+This method can enlighten authors as to *how* the uncertainty information was used by the participants.
+@Hofmann2012 tried to capture this by asking participants why they considered a particular plot to be more "right shifted", however this qualitative assessment does not seem to have made it into the final paper.
+@Daradkeh2015 presented participants with ten investment alternatives and asked participants "from among available alternatives, which alternative do you prefer the most", and were asked to think aloud and consider the uncertainty in their decision making.
+The experimenters goal was to observe and organise the methods people use when making decisions in the face of uncertainty.
+This study was an excellent example in a useful experimental design.
+They highlighted the specific aspects of uncertainty that participants typically considered, such as the range of outcomes that are above/below a certain threshold, minimum and maximum values, the risk of a loss, etc, and mapped where in the decision making process participants made these considerations.
 Data visualisation is commonly utilised as a tool in data exploration, so it is not uncommon for a data analyst to make a plot with only a vague goal and pull out a large number of adjacent observations.
@@
 
-### Comparisons to lineup plots and hypothesis testing
-It is unclear how participants are supposed to incorperate uncertainty information into their responses when uncertainty visualisation authors themselves are unsure.
-Other authors interpret this difficulty as a combination of the uncertainty and task complexity [@Kinkeldey2014], and not as a general confusion.
-All these examples sever to illustrate that *we dont know how to measure uncertainty* and if we don't know how to measure it, *we cannot evaluate the quality of a visualisation*.
-It could be argued that a well done uncertainty visualisations should have an imperceptible signal unless the signal would be identified with a hypothesis, but this idea also has a miriad of issues that come with it.
-There is already a series of issues with the reject/do not reject concepts in hypothesis testing, so it is already questionable to set the results of a hypothesis test as the ground truth in an uncertainty visualisation experiment, however these may be additional problems that go beyond the binary nature of the issue.
+### Just noticiable signal
+It could be argued that a well done uncertainty visualisations should have an imperceptible signal unless the signal would be identified with a hypothesis test, almost like a reverse line up protocol, but this idea also has some issues that should be considered.
+The reject or do not reject concepts in hypothesis testing do not offer a complete image of uncertainty, and exploration of uncertainty visualisation largely stems from a desire to move away from this binary framework.
+Additionally, setting up an uncertainty visualisation where the participants are expected to notice the signal once the data behind the visualisation passes a hypothesis test implies the signal *is* noticiable to a human at that level.
 @Patrick2023 compared people ability to recognise patterns in a residual plot to typical statistical tests and found human viewers looking at a plot were less sensitive than the typical residual tests.
-This increased sensisitivity could be the result of another aspect of uncertainty that is ignored.
-Statstical tests are typically built upon a series of assumptions, and it is difficult to identify *which* assumption failure caused the data to fail the statistical tests.
-Lineup plots allow us to see *how* our data is different to the assumed distribution, and better understand *why* our data may have failled a test ^[Did Patricks paper compare statistical tests to lineup protocols in the event an earlier assumption of the tests, such as, independent observation?
-I think he did check some.
-I wonder if you can quantify the information difference you get from a rejected statistical test vs a rejected line plot].
 These experiments utilised the lineup protocol which has been suggested as a method to check if perceived patterns are real or merely the result of chance [@Buja2009; @Wickham2010; @Chowdhury].
@@
 @fig-hypyvs shows the conceptual difference between the lineup protocol and uncertainty visualisation.
+
+![Difference between the null plot vs the uncertainty plot.
 A lineup protocol displays the uncertainty about the null and identifies if the true data plot is identifiable (and therefore significiantly difference) while an uncertainty plot displays the variance of the estimated value and assesses if the null of "no signal" is within this plot (e.g.
 if the error bars overlap with zero).
-^[Does this flip have mathematical implications in how the plots should be assessed?] This implies a visible signal in an uncertainty plot should indicate some divergence from the null.
-
-![Difference between the null plot vs the uncertainty plot.](hypvs.jpeg){#fig-hypvs}
-
-The problem of trying to assess uncertainty without knowing *why* we are depiting uncertainty has been noticed by other authors.
-@Spiegelhalter2017 noted we "cannot assess the quality of risk communication unless the objectives are clear".
+](hypvs.jpeg){#fig-hypvs}
 
 # Great Examples 
-(good decision making example) @Boone2018 organised responses from participants by the approach they used and tried to account for participants prior beliefs (the method of eliciting prior beleifs was, however, questionable).
+Despite the common problems detailed in the previous sections, there is some interesting work in the uncertainty visualisation space.
+This can come in the form of an uncertainty visualisation that attempts to visualise a typically ignored aspect of uncertainty, or an experiment that avoids the pitfalls detailled in the previous sections.
 
-> Edit notes: this section seems to go off a bit on dimensionality of visual tasks and stuff that might be better put in the noise and signal section about measuring uncertainty.
+The literature focus on "quantifiable" uncertainty leaves the variance that occurs at early stages in our analysis that is difficult to quantify ignored and forgotten.
+Some authors have chosen to focus on these unquantifiable cases and look at expressing more complicated cases of uncertainty.
+@Tierney2023 builds upon the tidy data pricipals to allow users to handle missing values.
+This includes data plots with a missing value "shadow" that allows visualisation authors to identify if the variables used in a plot have any structure in their missing values, which would contribute to uncertainty.
+Another example of uncertainty that is often ignored is the uncertainty resulting from human choices.
+Climate scenario uncertainty, shown in @fig-climatescenario, attempts to display the range of climate change outcomes that can result from a range of best and worse case human choices.
 
-Of course we do not believe every uncertainty visualisation is unusable and every experiment is build upon misguided assumptions.
-Despite the common problems detialled in the previous sections, there is some interesting work in the uncertainty visualisation space.
-This can come in the form of an uncertainty visualisation that identifies or fixes pitfalls in the previous sections (without identifying them) or does an experiment that identifies useful information that avoids these pitfalls.
+![Example of a I will just make my own version of a climate scenario plot, this one is just here from a random site as a place holder](climatescenario.png){fig-climatescenario}
 
-Uncertainty that is created by steps that are further down the pipeline are often ignored by the uncertainty visualisation literature, however there are visualisations that make these considerations.
-If uncertainty had a more encapsulating mathematical definition, the uncertainty created at earlier stages might be able to be visualised with the model uncertainty that is typically expressed through confidence intervals.
-That being said, even with a definition that allows us to combine these uncertainties into a single value, it still might be desirable to see which stage of the analysis is creating uncertainty in the visualisaiton.
-One example of a visualisation that includes that information is visualisations of climate scenario uncertainty, as shown in @fig-climatescenario.
-Uncertainty is introduced into this model by the choices of individuals which is difficult to model or predict and is also unlikely to be constant over time.
-Therefore specific scenarios are modelled and shown, each with their own model uncertainty.
-This highlights a method that can be used to visualise uncertainty from earlier in the pipeline that is often forgotten or ignored via assumptions.
-
-There are also some interesting qualitative studies that take an investigative stance to the use of uncertainty in decision making rather than an evaluation stance.
-@Daradkeh2015 presented participants with ten investment alternatives and asked participants "from among available alternatives, which alternative do you prefer the most", and were asked to think aloud and consider the uncertainty in their decision making.
-This study was qualitative and rather than setting a "correct" answer that would have forced a value of risk on the participants, the experimenters goal was to observe and organise the methods people use when making decisions in the face of uncertainty.
-This study was an excellent example in a useful experimental design.
-They highlighted the specific aspects of uncertainty that participants typically considered, such as the range of outcomes that are above/below a certain threshold, minimum and maximum values, the risk of a loss, etc, and mapped where in the decision making process participants made these considerations.
-
-![I will just make my own version of a climate scenario plot, this one is just here from a random site as a place holder](climatescenario.png){fig-climatescenario}
-
-Another method that has been used is to visualise samples from the distribution we are trying to draw inference on instead of estimates along with a variance or error.
-This has been used in maps with the pixel map [Lucchesi2017], but is more commonly used in animation with the HOPs plots [@Hullman2015] or similar concepts [@Blenkinsop2000] .
-The difficulty with resampling methods is that they are typically conflated with animation even though this does not necessarily need to be true, as can be seen in the @fig-pixelmap and the New York Times class mobility figure, shown in @fig-nyt, an animated version of which can be found [here](https://www.nytimes.com/interactive/2018/03/19/upshot/race-class-white-and-black-men.html).
-Therefore, so long as the sample size of the number of outcomes is kept to a level where individual samples are still visible, there is no evidence that an outcome plot *needs* to be animated.
+Another visualisation method that has a lot of potential is visualisation of samples.
 Visualisations that opt to express a signal as a sample rather than an estimate have the potential to supress signal since it is not explicitly visualised, however this has yet to be shown in evaluation experiments.
 This is not to say that visualisations of mass would not be able to perform signal supression, but a sample can easily be expressed using aesthetics such as colour on a map and mass visualisation often struggling with issues such as over or under smoothing.
-These sampling methods can show the messiness of the data that sits behind a model.
+These sampling methods can show more of the messiness of the data that sits behind a model.
+This may not have a detrimental effect on the viewers ability to extract global statistics, as it seems they can be extracted from a visualisation of a sample with ease [Franconeri2021].
+Sample visualisations have been used in maps with the pixel map, shown in @fig-pixelmap [Lucchesi2017], but is more commonly used in animation with the HOPs plots [@Hullman2015] or similar concepts [@Blenkinsop2000].
+This method has also been adopted by visualisation authors outside of academia as can be seen in the  and the New York Times class mobility figure, shown in @fig-nyt, an animated version of which can be found [here](https://www.nytimes.com/interactive/2018/03/19/upshot/race-class-white-and-black-men.html).
 
@@
 
-A final method to perform signal supression is simply to visualise the data if it is available.
+A final method to perform signal supression is simply to visualise the data if it is available and relevant to the uncertainty distribution.
 An example of this is shown in @fig-census for racial distributions spatially in america, an interactive version of the plot can be found [here](https://edition.cnn.com/interactive/2021/us/census-race-ethnicity-map/).
-This map shows the typical causes of uncertainty in a spatial model, i.e.
-regions where data is sparse, ethnically diverse areas, uneven distribution of points within boundaries, etc, but it avoids the need to create a visualisation with a specific signal in mind.
+This map shows the typical causes of uncertainty in a spatial model, (e.g.
+regions where data is sparse, ethnically diverse areas, uneven distribution of points within boundaries, etc) but it avoids the need to create a visualisation with a specific signal in mind.
 This is the technique typically employed by exploratory data analysis, which means it's lack of a specified signal means there is both *no* uncertainty (since we are technically not performing inference) but in the event we *do* implicitly perform inference, there is some hedging.
@@
 The census dot map's addition of interactivity also allows users to zoom in and see the details that caused "uncertainty" in the form of inconsistent colours at lower resolutions when they were zoomed out.
-The raw data can also be used in a similar method to the HOPs plot, where a statistic of importance is implied by the visualisation, but the signal is supressed using the data itself instead of resampling methods.
 This does not mean that visualising raw data instead of implementing sampling techniques is always a valid uncertainty visualisation that will prevent insignificant signal from getting through.
 @Buja2009 illustrated how groups that appear linearly separable in a linear discriminant analysis (LDA) visualisation of the data can actually be the result of a LDA performed on too many variables, something that was not clear from the visualisation until the lineup protocol was implemented.
+However it is simple but effective option that seems to be laregly overlooked by the uncertainty visualisaiton community.
 
 ![](censusdotmap.png){fig-census}
-
-These great examples show there are visualisations that are being made that have a good conceptualisation of uncertainty and express it as signal supression, however this work is not being done formally within visualisation.
-These industry visualisation are likely made with a large amount of back and forth between different levels of an organisation, something that is important for good visual communication [@Spiegelhalter2017].
 
@@
 Our first suggestion for future work is a mathematical defintion of uncertainty.
-Specifically, a definition that has a place for all the concepts that currently sit under the "uncertainty" umbrella, explains how these concepts relate to uncertianty, and also allows us to quantify them in a way where they can be combined or visualised.
-This work would not only benefit uncertainty visualisaiton, but also statistics more broadly.
 There is currently a lot of work that quantifies the uncertainty through bias and variance at various stages of the data analysis pipeline separately.
@@
 The concept of uncertainty should also be formalised within the grammar of graphics.
-Not only would a formalisation of uncertainty visualisation within the grammar of graphics allow us to iron out some of these confusions, but they would make it easier to understand when existing visualisation methods apply and can be used to explain our results.
+This formalisation would allow uncertainty visualisation authors to have a clear understanding of what is or is not an uncertaitny visualisation.
+Additionally placing uncertainty visualisation in the framework that is used to understand existing information visualisation research would help authors understand when existing methods can be used to explain their results.
+Incorperating uncertainty into the grammar of graphics will also give a more precise concept of the information contained within a plot.
+Other fields of science employ marginal changes when designing experiments to ensure it is well understood *what* aspect of their experiment is contributing to their results, and a better sense of what "marginal" is in the case of uncertainty visualisaion would greatly help the field.
 
-```{r}
-#| include: false
-
-# put a visualisation of a dot plot with increasing sample size here to show my point
-```
-
+Additionally the experimental practices within uncertainty visualisation need to be less ad hoc and more standardised.
 If we are going to consider uncertainty as noise, not signal, there needs to be a way to identify this signal supression in an experimental design.
-It is clear that value retrival and questions with a specific ground truth are not appropriate, as uncertainty visualisation will simply follow the existing visualisation rules.
-Testing a secondary feature of a plot, that is, how strong a signal is becomes a little more complicated.
-Line up protocols have been used in adjacent work that look at the strength of the signal in a plot, and this idea of identifying if plots have some "barely noticiable differences" could be utilised.
+As the literature currently exists, there is no way to combine papers to get a meaningful sense of how uncertainty information is understood by a viewer.
 There is also the possibility that uncertainty visualisation evaluations will need to swap to a qualitative methodology where participants are allowed to freely comment on what they notice in graphics until we establish how the existence of noise can be observed.
 
-Additionally, information visualisation needs a more precise concept of the difference between two plots, and that framework should be utilised in experiments.
-Other fields of science employ marginal changes when designing experiments to ensure it is well understood *what* aspect of their experiment is contributing to their results.
-Visualisation has the grammar of graphics, however we have already discussed how this can break down in the case of distribution visualisation.
-
-If a visualisation researcher would prefer to perform experiments rather than formalise methods, the task dependency many authors in uncertainty visualisation mention would be a useful direction for research.
-Unfortunately the current visualisation evaluation literature is *far* too noisy to allow us to identify what might be the cause of this task dependence.
-It is unclear if it is a conflation between task and the ground truth/displayed statistics concepts discussed earlier, or an actual difference in the way we percieve graphics.
-A good starting place for this idea would be to identify if there is task dependency in perceptual tasks, since all perceptual task research to date has been done on value retrival.
-That is, an experiment that compares things like area, position, colour and other perceptual tasks on value exaction, comparisons, and other simple tasks.
-It is also clear that the the number of potential tasks that can be performed on a visualisation increases with with the number of observations.
+If an uncertainty visualisation researcher would prefer to perform experiments rather than formalise methods, there are options there too.
+It would be interesting to know if any perceptual tasks that can be mapped to two different visual tasks condense into a single dimension when looking for overarching signal in a plot.
+Alternatively, the task dependency many authors in uncertainty visualisation mention would be a useful direction to consider.
+It is clear that the the number of potential tasks that can be performed on a visualisation increases with with the number of observations.
 A single observation is limited to value extraction, two observations can be compared, multiple observations allow for shapes or global statistics to be extracted.
@@
 @Amar2005 suggested a taxonomy for information visualisation based on the types of tasks we use visualisations for and suggest 10 "analytical primitives" that we can then map to visualisations, which could be a good starting point.
-Reguardless, these are directions of research would be fruitful to the uncertainty visualisation community even if it appears on the surface to be research that is only beneficial to the "normal" visualisation community.
-
-# Conclusion
-In this paper we have highlighted a series of misunderstandings, confusions, and methodical errors in the uncertainty literature.
-
-Many of these issues are not problems in isolation, so long as the authors themselves are aware of it and state these assumptions in their papers.
-Our problem with this body is work is that the authors themselves seem to be oblivious to this vast array of problems, considering not a single one of the problems discussed here are mentioned explicitly in a single visualisation paper.
+Regardless, these are directions of research would be fruitful to the uncertainty visualisation community even if it appears on the surface to be research that is only beneficial to the "normal" visualisation community.
 

### 2024-03-22 writingdumpster.qmd
finished rewrite

@@
 # Moved over notes
-
+- These industry visualisation are likely made with a large amount of back and forth between different levels of an organisation, something that is important for good visual communication [@Spiegelhalter2017].
+- @Boone2018 organised responses from participants by the approach they used and tried to account for participants prior beliefs (the method of eliciting prior beleifs was, however, questionable). 
+- @Correll2018 “place your 5 ships on the safest locations on the board.” (minimise danger)
+- @Blenkinsop2000 did try to include uncertainty as noise, asking participants to search for a specific outcome (the land classified as grass) in a random .
+However the participants failing to identify the signal (something that would be expected of an uncertainty plot) was seen as a sign the task was too complicated.
+While the task in this experiment *was* certainty far too complicated (and poorly communicated) this discussion indicates the authors did not consider the signal suppression to be the goal of the visualisation, and rather an annoying distraction.
 - @Ibrekk1987 expected participants to calculate the mean using a pie chart that had numerical bins combined (e.g.
@@
 # Main section outtakes
+Conclusion
+In this paper we have highlighted a series of misunderstandings, confusions, and methodical errors in the uncertainty literature.
+Many of these issues are not problems in isolation, so long as the authors themselves are aware of it and state these assumptions in their papers.
+Our problem with this body is work is that the authors themselves seem to be oblivious to this vast array of problems, considering not a single one of the problems discussed here are mentioned explicitly in a single visualisation paper.
+
+It is clear that value retrival and questions with a specific ground truth are not appropriate, as uncertainty visualisation will simply follow the existing visualisation rules.
+Testing a secondary feature of a plot, that is, how strong a signal is becomes a little more complicated.
+Line up protocols have been used in adjacent work that look at the strength of the signal in a plot, and this idea of identifying if plots have some "barely noticiable differences" could be utilised.
+
+Specifically, a definition that has a place for all the concepts that currently sit under the "uncertainty" umbrella, explains how these concepts relate to uncertianty, and also allows us to quantify them in a way where they can be combined or visualised.
+This work would not only benefit uncertainty visualisaiton, but also statistics more broadly.
+
+The problem of trying to assess uncertainty without knowing *why* we are depiting uncertainty has been noticed by other authors.
+@Spiegelhalter2017 noted we "cannot assess the quality of risk communication unless the objectives are clear".
+
+Does the flip (hypothesis test vs uncertainty visualisation) have mathematical implications in how the plots should be assessed?
+This implies a visible signal in an uncertainty plot should indicate some divergence from the null.
+
+This increased sensisitivity could be the result of another aspect of uncertainty that is ignored.
+Statstical tests are typically built upon a series of assumptions, and it is difficult to identify *which* assumption failure caused the data to fail the statistical tests.
+Lineup plots allow us to see *how* our data is different to the assumed distribution, and better understand *why* our data may have failled a test ^[Did Patricks paper compare statistical tests to lineup protocols in the event an earlier assumption of the tests, such as, independent observation?
+I think he did check some.
+I wonder if you can quantify the information difference you get from a rejected statistical test vs a rejected line plot].
+
+It is unclear how participants are supposed to incorperate uncertainty information into their responses when uncertainty visualisation authors themselves are unsure.
+Other authors interpret this difficulty as a combination of the uncertainty and task complexity [@Kinkeldey2014], and not as a general confusion.
+All these examples sever to illustrate that *we dont know how to measure uncertainty* and if we don't know how to measure it, *we cannot evaluate the quality of a visualisation*.
+
+Using the most commonly presented methods In the absense of an encapsulating mathematical definition this will not change.
+That being said, even with a definition that allows us to combine these uncertainties into a single value, it would still be beneficial for authors to develop specialty visualisations for different sources of uncertainty, so analysts can utilise visual tools to understand the uncertainty in their data.
+still might be desirable to see which stage of the analysis is creating uncertainty in the visualisaiton.
+
 @Hofmann2012 showed two distributions in 20 different visualisations (a lineup protocol) using a jittered sample, a density plot, a histogram, and a box plot and asked participants.

### 2024-03-22 paper.qmd
spell check

@@
 ---
-# Current Todo List
-This is a list of tasks that need to be completed before I consider the paper complete. Please keep this in mind when reading.
 
@@
 Despite the youth of the field, commonly used uncertainty visualisations such as a box plot or histogram have been around for almost as long as statistical graphics themselves (*Cite Gap 2*).
-Early mentions of "uncertainty visualisation" appear in computer science papers discussing uncertainty visualisation as a subfield of geospatial information visualisation, specifically for the many uncertainties that come with satelite images (*Cite Gap 1*).
+Early mentions of "uncertainty visualisation" appear in computer science papers discussing uncertainty visualisation as a subfield of geospatial information visualisation, specifically for the many uncertainties that come with satellite images (*Cite Gap 1*).
 In the modern day, uncertainty visualisation has exploded into its own field that applies to a wide range of contexts and is no longer bound by geospatial information (*Cite Gap 1*).
@@
 Interestingly, these reviews often do not offer overarching rules for tried and tested uncertainty visualisation, but rather comment on the *difficulties* faced when trying combine the papers from this field.
-@Kinkeldey2014 found most experiments the methods for uncertainty visualisation evaluation to be adhoc, with no commonly agreed upon methodology or formalisation and no greater goal of describing general principals.
+@Kinkeldey2014 found most experiments the methods for uncertainty visualisation evaluation to be ad hoc, with no commonly agreed upon methodology or formalisation and no greater goal of describing general principals.
 @Hullman2016 commented on the difficulty in taking overarching themes from uncertainty visualisation, as several conflated issues make it unclear if subjects did poorly in an experiment because they misunderstood a visualisation, because the question was misinterpreted, or because they used a heuristic.
@@
 There are a handful of reasons that are used to explain why it is so difficult to generalise the findings in the uncertainty visualisation literature.
-Some papers suggested or agree with the idea that visualisation typologies should move away from data types, uncertainty categories, and representation types and towards "task-centered typologies" as this will help generalise results [@Kinkeldey2014; @Hullman2016].
+Some papers suggested or agree with the idea that visualisation typologies should move away from data types, uncertainty categories, and representation types and towards "task-centred typologies" as this will help generalise results [@Kinkeldey2014; @Hullman2016].
 Some authors may not attribute the task to the *reason* the literature cannot be easily summarised, but do mention that the choice of best visualisation is highly dependent on the specific goal of that visualisation [@Griethe2006; @Spiegelhalter2017].
 Others comment that visualisations are highly dependent on the audience and there is no such thing as a "best" visualisation that will be accessible to all members [@Kinkeldey2014; @Spiegelhalter2017].
-These difficulties in noisy uncertainty visualisation findings arises reguardless of the direction the analysis was approached, be it looking at geospatial data [@Kinkeldey2014], or uncertainty visualisation as a whole [@Hullman2016; @Spiegelhalter2017; @Griethe2006].
+These difficulties in noisy uncertainty visualisation findings arises regardless of the direction the analysis was approached, be it looking at geospatial data [@Kinkeldey2014], or uncertainty visualisation as a whole [@Hullman2016; @Spiegelhalter2017; @Griethe2006].
 
 This review believe all these sources of noise have a single root cause, a lack of a cohesive and encompassing definition of uncertainty.
-The absence of an econompasing definition of uncertainty is mentioned by every uncertainty visualisation directly [@Spiegelhalter2017; @Griethe2006] or indirectly by describing a miriad of ways it can be considered in the literature [@Kinkeldey2014; @Hullman2016], although it is never commented on as a source of the noise in the field.
+The absence of an encompassing definition of uncertainty is mentioned by every uncertainty visualisation directly [@Spiegelhalter2017; @Griethe2006] or indirectly by describing a myriad of ways it can be considered in the literature [@Kinkeldey2014; @Hullman2016], although it is never commented on as a source of the noise in the field.
 
@@
 @utypo defined it as any deviation from complete determinism) but go on to *quantify* uncertainty as a PDF, error, or some other easily quantified mathematical object when it comes time to do the visualisation (*Cite Gap 4*).
-This definition swap happens so subtly that it seems to go unnoticed by authors writing the papers, and the fact that "uncertainty visualisation" methods dont even notice they are exclusively focus on depictions of easily quantifiable uncertainty, such as error [], mass [], or variance [] (*Cite Gap 5*).
+This definition swap happens so subtly that it seems to go unnoticed by authors writing the papers, and the fact that "uncertainty visualisation" methods don't even notice they are exclusively focus on depictions of easily quantifiable uncertainty, such as error [], mass [], or variance [] (*Cite Gap 5*).
 This has left the expressions of uncertainty that are hard to quantify or visually differentiate untouched, despite many papers calling for their invention (*Cite Gap 6*).
@@
 (*Cite Gap 7*).
-In their own way, each of these taxonomies show an apsect of uncertainty that an author felt was important to differentiate.
+In their own way, each of these taxonomies show an aspect of uncertainty that an author felt was important to differentiate.
 @utypo identified many common themes among this wide array of taxonomies and used them to design an encapsulating taxonomy, depicted in @fig-taxonomy.
-This definition organises uncertainty based on three axis, location (where the uncertianty is coming from), level (how quantifiable it is), and nature (if it comes from true randomness or incomplete knowledge).
+This definition organises uncertainty based on three axis, location (where the uncertainty is coming from), level (how quantifiable it is), and nature (if it comes from true randomness or incomplete knowledge).
 
@@
 
-When considering the use of taxonomies to define something inherrently vague, such as the uncertainty in an uncertainty visualisation, it is helpful to consider this quote from *The Grammar of Graphics*:
+When considering the use of taxonomies to define something inherently vague, such as the uncertainty in an uncertainty visualisation, it is helpful to consider this quote from *The Grammar of Graphics*:
 
@@
 Descriptive statistics are not seen as the primary goal of modern statistics, however this was not always the case.
-Around the 19th century in England, *posivitism* was the popular philosophical approach to science (positivists included famous statisticians such as Francis Galton and Karl Pearson) and practitioners of the approach believed statistics ended with descriptive statistics as science must be based on actual experience and observations, therefore anything that refers to the unobservable (such as new observations or population statistics) is not true science [@Otsuka2023].
-By its very nature, descriptive statistics *cannot* be used to make inferences about the data because it simply exists to summaries the data, to use it to make statements about *new* data is incorrect useage.
+Around the 19th century in England, *positivism* was the popular philosophical approach to science (positivists included famous statisticians such as Francis Galton and Karl Pearson) and practitioners of the approach believed statistics ended with descriptive statistics as science must be based on actual experience and observations, therefore anything that refers to the unobservable (such as new observations or population statistics) is not true science [@Otsuka2023].
+By its very nature, descriptive statistics *cannot* be used to make inferences about the data because it simply exists to summaries the data, to use it to make statements about *new* data is incorrect usage.
 In order to make statements about population statistics, future values, or new observations we need to perform inference, which requires the assumption of the "uniformity of nature" (i.e.
@@
 
-This philosophical understanding of statistics highlights that descriptive statistics do not have uncertainty, however some readers may disagreee with that statement.
+This philosophical understanding of statistics highlights that descriptive statistics do not have uncertainty, however some readers may disagree with that statement.
 Variance and probability are typically considered stand ins for "uncertainty", it is often how we choose to measure it, and since probability and variance exist in descriptive statistics, descriptive statistics *must* have uncertainty.
@@
 @Wallsten1997 argue that the best method for evaluating or combining subjective probabilities depends on the uncertainty the decision maker wants to represent and why it matters.
-@Munzner2009 created a nested model for visualisation that highlighted how the first mistake that can be made in a visualisation is in the problem characterisation, and failling to do it well can cause downstream effects and damage the effectivness of a visualisation.
+@Munzner2009 created a nested model for visualisation that highlighted how the first mistake that can be made in a visualisation is in the problem characterisation, and failing to do it well can cause downstream effects and damage the effectiveness of a visualisation.
 @Fischhoff2014 looks at uncertainty visualisation for decision making decides that we should have different ways of communicating uncertainty based off what the user is supposed to do with it.
@@
 None of these existing methods are thorough enough for an analyst to understand what causes uncertainty, and quantify it for communication.
-The need for a mathematical definition of this field is put simply by Freeman Dyson in his fameous Birds and Frogs speech:
+The need for a mathematical definition of this field is put simply by Freeman Dyson in his famous Birds and Frogs speech:
 
@@
 In order to visualise something, it needs to be quantifiable.
-In order to quanitfy something, it needs to be mathematically defined.
-The broad classes of what is consdiered "uncertainty" is not currently quantifiable in any way that is not ad-hoc.
-The only apsects of uncertaitny that are currently quantifiable are confidence intervals, prediction intervals, and related terms.
+In order to quantify something, it needs to be mathematically defined.
+The broad classes of what is considered "uncertainty" is not currently quantifiable in any way that is not ad-hoc.
+The only aspects of uncertainty that are currently quantifiable are confidence intervals, prediction intervals, and related terms.
 Even then, these quantifications often only capture the uncertainty in our model or its assumptions, it ignores bias often introduced in earlier stages of the analysis.
@@
 The previous section established that uncertainty is a by product of the assumptions we make when we perform inference, additionally, our quantification of uncertainty is largely task dependent, therefore the uncertainty associated with *one* distribution (which we can also view as a task or goal dependence).
-When uncertainty visualisaiton authors directly stare at this concept, they almost unanimously agree with these concepts, however when these details are pushed the the periphery of a problem, we start to get research that confuses itself.
+When uncertainty visualisation authors directly stare at this concept, they almost unanimously agree with these concepts, however when these details are pushed the the periphery of a problem, we start to get research that confuses itself.
 In these cases, the authors are more confused about what they are doing than the participants.
@@
 The authors reported that the group that was shown the sampling distribution were willing to pay more and ascribed a higher likelihood to the belief that the treatment was more effective than a control, than the group that was shown a prediction interval.
-A prediction interval indicates the uncertainty associated with treatment *for a particular person* and the sampling interval shown the uncertainty associated with the *average effectivness of the treatment*.
+A prediction interval indicates the uncertainty associated with treatment *for a particular person* and the sampling interval shown the uncertainty associated with the *average effectiveness of the treatment*.
 It is already established practice in medicine to use a wide range of metrics to communicate different risks to best communicate the relevant information to a patient for a particular decision [@Spiegelhalter2017].
@@
 Many authors do not understand bias *is* uncertainty (something that has been proven mathematically by @Meng2014) and whether we view uncertainty as bias or variance depends on the resolution of our problem.
-@Padilla2021 found that high uncertainty in the model estimates (calculated uncertainty) and low forecaster confidence (which is typically an expression of suspected bias in the model) both caused participants to have decreased confidence in their results and suggested modelers express both if they are relevant.
+@Padilla2021 found that high uncertainty in the model estimates (calculated uncertainty) and low forecaster confidence (which is typically an expression of suspected bias in the model) both caused participants to have decreased confidence in their results and suggested modellers express both if they are relevant.
 @Kale2019 discussed the importance of communicating decisions made in the data analysis pipeline and being aware of the alternatives.
@@
 The misunderstanding that descriptive statistics also extends to visualisation.
-The exploration step of an analysis, which includes descriptive statistics, exploratory data visualisaiton, and unsupervised machine learning techniques, is performed without a prior hypothesis, however misunderstandings of this fact appears frequently in the literature.
+The exploration step of an analysis, which includes descriptive statistics, exploratory data visualisation, and unsupervised machine learning techniques, is performed without a prior hypothesis, however misunderstandings of this fact appears frequently in the literature.
 @Griethe2006 commented that "if visualization is used as a means to explore a data volume or to communicate its contents the uncertainty has to be included".
@@
 This confusion (or indifference) also creates many imprecisely named papers.
-Many papers will boast a title that claims to be about uncertainty visualisaiton, but simply depicts different visual representations of a PDF, however it takes reading the methodology to find this out (*Cite Gap 9*).
+Many papers will boast a title that claims to be about uncertainty visualisation, but simply depicts different visual representations of a PDF, however it takes reading the methodology to find this out (*Cite Gap 9*).
 While this is a minor by product of the definition issue discussed earlier, it does result in a literature that is needlessly difficult to navigate.
@@
 4. coincidence (uncertainty and data are represented in the same plot) or adjacent
-5. static or dynamic (animated, interactive ect)
+5. static or dynamic (animated, interactive etc)
 Of the groups they identified, they actually only used (4), (2), and (5), left out (1) because most visualisations are explicit, and (3) corresponds to (1) in most cases [@Kinkeldey2014].
 This taxonomy suggests interesting considerations for a visualisation of a distribution (even though it was not intended to be a visualisation of a distribution).
-A similar version of this taxonomy was presented by @uncertchap2022 who commented that visualisation can be organised into two categories, "graphical annotations of distributional properties" and "visual encodings of uncertainty" which functionally aline with the intrinsic/extrinsic distinction.
+A similar version of this taxonomy was presented by @uncertchap2022 who commented that visualisation can be organised into two categories, "graphical annotations of distributional properties" and "visual encodings of uncertainty" which functionally align with the intrinsic/extrinsic distinction.
 The first four categories boil down to the question "should uncertainty and signal be conveyed together as one variable or separately as two".
@@
 The difference between "uncertainty visualisations" and "data visualisations" is not technically in the visual element, it is mathematical.
-@Kinkeldey2014 almost acknowledges this in their own paper that discusses an uncertainty visualsiation taxonomy when they claim "future typologies should take different categories of tasks into account (1) communication tasks (2) analytical tasks (3) exploratory task", a common typology for information visualisation in general.
+@Kinkeldey2014 almost acknowledges this in their own paper that discusses an uncertainty visualisation taxonomy when they claim "future typologies should take different categories of tasks into account (1) communication tasks (2) analytical tasks (3) exploratory task", a common typology for information visualisation in general.
 The process of understanding and estimating uncertainty requires knowledge of the data, the statistical methods used to make an estimate, and the assumptions of a model.
@@
 One of the key components of uncertainty and inference is the noise and signal dichotomy.
-When @Otsuka2023 discusses the information we boil away to create a "probabilistic kind" he uses the example of a coin flip and the sex of baby births as two things that are *very* different in reality, but it is only through the lense of a Bernoulli distribution that we can view these two things as the same.
+When @Otsuka2023 discusses the information we boil away to create a "probabilistic kind" he uses the example of a coin flip and the sex of baby births as two things that are *very* different in reality, but it is only through the lens of a Bernoulli distribution that we can view these two things as the same.
 The fact is that this boiling down, of a real world event into data, is not something that happens independent of the statisticians, but is just as much as product of our interests (the particular relationship we hope to draw inference on) as it is a product of the event's natural properties.
@@
 With this in mind, let us turn to the concept of "contextual" information.
-For the purposes of this review, we will consider contexual information to be anything that makes it easy for the audience to connect the data from a probablistic kind to its real world parallel, but is largely irrelevant in the abstract probability space.
+For the purposes of this review, we will consider contextual information to be anything that makes it easy for the audience to connect the data from a probabilistic kind to its real world parallel, but is largely irrelevant in the abstract probability space.
 A key example of this is spatial information, such as borders, road locations, terrains, etc.
@@
 It can help provide reasons for *why* there may be specific characteristics in our data, however it is not required information for the model.
-@fig-spatial shows the proces of boilling information down into data, visualising it, and including contextual information to improve our scientific reasoning.
+@fig-spatial shows the process of boiling information down into data, visualising it, and including contextual information to improve our scientific reasoning.
 The concept that contextual information is not the same as statistical information is also highlighted in *The Grammar of Graphics* which states:
@@
 Contextual information is important for interpretation and understanding of graphics, but it does not in of itself generate uncertainty.
-Its relevance to the field of "uncertainty visualisation" is through the percieved lack of available channels.
+Its relevance to the field of "uncertainty visualisation" is through the perceived lack of available channels.
 For example, a choropleth map has already used colour to represent an estimate, and position to represent spatial information, which is sometimes relevant to the probabilistic kind, and other times it is simply contextual information (such as when the modelling is done by using a state as a category rather than its geographic information).
@@
 @Kinkeldey2014 discussed how uncertainty  can be represented by three components attribute (what) position (where) and temporal (when) and that studied typically deal with uncertainty around attribute but rarely position and time, however it is never specified what considerations should cause attribute uncertainty to be different to position or temporal uncertainty.
-@Kay2016 did an experiment that showed uncertainty around bus arrival times, however the visualisation used in the experiment, shown in @fig-bustime, is indistinguishable from most work that would be considered "attribute" uncertainty.
+@Kay2016 did an experiment that showed uncertainty around bus arrival times, however the visualisation used in the experiment, shown in @fig-bus, is indistinguishable from most work that would be considered "attribute" uncertainty.
 
-![The uncertainty visualisation used by @Kay2016 to show the uncertainty around a bus arrival time prediction.](bustime.png){#fig-bustime}
+![The uncertainty visualisation used by @Kay2016 to show the uncertainty around a bus arrival time prediction.](bustime.png){#fig-bus}
 
@@
 Spatial and temporal uncertainty papers limit how they communicate their uncertainty because they are considering what can be visualised inside the context provided by maps and, not what they should *try* to visualise based on the uncertainty abstraction.
-When we view a spatial-temporal visualisation through the lense of trying to visualise the distribution of $P(X|longitude, latitude, time)$ there is no reason this is any different from any other high dimensional visualisation problem.
+When we view a spatial-temporal visualisation through the lens of trying to visualise the distribution of $P(X|longitude, latitude, time)$ there is no reason this is any different from any other high dimensional visualisation problem.
 
@@
 This is not only the case for uncertainty data.
-@Slingsby2023 discussed using a gridded glyphmap for spatial modelling, however the statistical elements are identical to a fluxuation diagram depicting $f(infection|time, age, longitude, latitude)$ presented in @Wickham2011.
+@Slingsby2023 discussed using a gridded glyph map for spatial modelling, however the statistical elements are identical to a fluctuation diagram depicting $f(infection|time, age, longitude, latitude)$ presented in @Wickham2011.
 The plot by @Slingsby2023 is simply a special case of the graphic by @Wickham2011, where a separate visualisation is made for each time point, the continuous function $infection | age$ has replaced the categorical variable $happy$, a fade is added for the population size, and a map is shown in the background.
 @Slingsby2023 presents other maps in the paper, however they are also special cases of the product plots framework.
-This is not to say this extension is uninteresting, on the contrary we believe identifying cases where particular graphics are useful is a worthwhile endevour, but these plots seem to be have been developed unaware of the existing statistical framework.
-When we separate the field of visualisation along lines that may not be of particular relevance, we increase the likelihood of "reinventing the wheel" in each sub-dicipline.
+This is not to say this extension is uninteresting, on the contrary we believe identifying cases where particular graphics are useful is a worthwhile endeavour, but these plots seem to be have been developed unaware of the existing statistical framework.
+When we separate the field of visualisation along lines that may not be of particular relevance, we increase the likelihood of "reinventing the wheel" in each sub-discipline.
 
@@
 
-### Information asymetry in evaluated plots
+### Information asymmetry in evaluated plots
 Visualisation authors are almost unanimous in commenting that the "information" in two plots must be the same in order for the visual techniques to be compared [@Cleveland1984; @Kinkeldey2014] (*Cite gap 12*).
@@
 @Kinkeldey2014 adopts an existing definition also that suggests two graphics are informationally equivalent if all the information in one plot is inferable from the other and vice-versa, but adds that two plots are computationally equivalent if that information can be extracted from both plots with similar easy and speed.
-It is clear this lacking definition of "information" in a visual is reminicent of our problems with the definition of "uncertainty".
+It is clear this lacking definition of "information" in a visual is reminiscent of our problems with the definition of "uncertainty".
 
@@
 When creating a graphic there are several tasks that must be completed in a specific order, regardless of whether or not we are working with a uncertainty visualisation or not, this pipeline is illustrated in @fig-pipeline, and contains all the steps of the grammar of graphics [@Leland2005].
-First the mathematical information in a plot is established in the "varset" setps, and then that mathematical information is visually represented, which is defined in the "graph" steps.
-Since two plots cannot be compared on their visual elements if they contain different information (because you will never be able to identify if the difference is because of informational asymetry or visual processing) we could understand two plots as being the same if their "varset" steps are identical.
+First the mathematical information in a plot is established in the "varset" steps, and then that mathematical information is visually represented, which is defined in the "graph" steps.
+Since two plots cannot be compared on their visual elements if they contain different information (because you will never be able to identify if the difference is because of informational asymmetry or visual processing) we could understand two plots as being the same if their "varset" steps are identical.
 However because uncertainty is not defined mathematically, we cannot reliably add it in at the "statistic" stage.
@@
 Additionally, authors rarely attribute this visual version of large sample theory to results where it is clearly applicable.
-@Kale2018 found that users were more sensitive to the underlying trend when shown the HOPs plot over the static outcoms plot, this effect went away when the speed of the animation increased, indicating that performance difference might be due to overplotting in the static case and a smaller sample size in the static condition might produce the same result.
+@Kale2018 found that users were more sensitive to the underlying trend when shown the HOPs plot over the static outcomes plot, this effect went away when the speed of the animation increased, indicating that performance difference might be due to overplotting in the static case and a smaller sample size in the static condition might produce the same result.
 The issue of blurred elements of the grammar of graphics in the case of distribution visualisations does not end here.
@@
 Alternative information consideration within well defined frameworks may also be worthwhile to consider.
-@wu2023rational tried to eliminate information asymetry by esentially looking at a visualisation and identifying what information can be extracted and then using a "rational agent benchmark" to determine how a rational person would use that information for a decision making task.
+@wu2023rational tried to eliminate information asymmetry by essentially looking at a visualisation and identifying what information can be extracted and then using a "rational agent benchmark" to determine how a rational person would use that information for a decision making task.
 While this solution is interesting, it may have different results depending on who is applying it as they may use different methods or notice different elements when extracting information from a graphic.
@@
 This collection of examples starts to paint a pretty clear picture.
-Visualisations with rather shocking information asymetry are regularly compared because they are both "uncertainty visualisations".
+Visualisations with rather shocking information asymmetry are regularly compared because they are both "uncertainty visualisations".
 This results in a series of experiments where the visual aspects of the graphic are not even required to anticipate the experimental results.
@@
 5) Volume, density, curvature
-6) Shading, color saturation, color hue
+6) Shading, colour saturation, colour hue
 7) Discriminable shape
@@
 
-This hierarchy is a good general rule, however it can change from person to person [@Davis2022] Additionally, there are other graphical rules to consider such as guestalt heuristics, broader methods of extraction, and attention principles [@Vanderplas2020].
+This hierarchy is a good general rule, however it can change from person to person [@Davis2022] Additionally, there are other graphical rules to consider such as gestalt principles, broader methods of extraction, and attention principles [@Vanderplas2020].
 These established visualisation concepts allow us to anticipate the ease with which certain pieces of information will be extracted from a plot.
@@
 Similar results to this occur over and over again in the uncertainty visualisation literature.
-@Gschwandtnei2016 found that visualisations where the start time of an interval could literally be read off the plot (error bars, centered error bars, and ambiguation) performed better than the plots (accumulated probability, gradient, and violin) where the start time involved some guesswork because the drop off was gradual.
+@Gschwandtnei2016 found that visualisations where the start time of an interval could literally be read off the plot (error bars, centred error bars, and ambiguation) performed better than the plots (accumulated probability, gradient, and violin) where the start time involved some guesswork because the drop off was gradual.
 @Cheong2016 found that participants were better at answering questions when they were explicitly given the relevant probability in text rather than when they needed to read it off a map.
@@
 It is clear that authors intuitively *feel* that uncertainty is different even though this difference has not been captured in existing uncertainty visualisation experiments.
-Some authors cited uncetainty was special because of its psychological qualities, but we have yet to see any evidence of that here.
+Some authors cited uncertainty was special because of its psychological qualities, but we have yet to see any evidence of that here.
 @Kinkeldey2014 also noticed that it is not clear whether or not uncertainty is just another variable, as many value extraction papers treats it as such in their experimental design and uncertainty may need to be in a variable class of its own.
 @Hullman2016 commented that it is straightforward to show a value but it is much more complex to show uncertainty, but did not explain *why* uncertainty information faces more difficulty.
-After all, confidence intervals, minimums, maxmimums, variance, and any other statistic that can be considered related to "uncertainty" is just that, a statisitc.
-If we want to visualise the relationship between a temperature estimate and our forcasted variance on that esitmate, can we not just use a scatter plot, as we would any other pair of variables?
+After all, confidence intervals, minimums, maximums, variance, and any other statistic that can be considered related to "uncertainty" is just that, a statistic.
+If we want to visualise the relationship between a temperature estimate and our forecasted variance on that estimate, can we not just use a scatter plot, as we would any other pair of variables?
 There is nothing special about these statistics that should differentiate them from a mean or median and imply their visualisation should be "special" in a way that would warrant its own field.
-Especially when the core take aways from uncertainty evaluation experiments do not seem to differ at all from the established results in information visualisation.
+Especially when the core takeaways from uncertainty evaluation experiments do not seem to differ at all from the established results in information visualisation.
 So, what is it that makes authors believe uncertainty is different?
@@
 Interviews with experts in statistic back up this primary motivation, as ignoring uncertainty information of often expressed as being similar to fraud or lying [@Hullman2020a; @Manski2020].
-Noise and signal are not an inherrent property of variance, or meta-data or the aspects of an analysis that are usually secondary considerations, it is a property of the data that *promotes* a message we are trying to infer and the data that *supresses* it, if that message is invalid.
+Noise and signal are not an inherent property of variance, or meta-data or the aspects of an analysis that are usually secondary considerations, it is a property of the data that *promotes* a message we are trying to infer and the data that *suppresses* it, if that message is invalid.
 When we ask the viewer of a plot to look at data and extract a value, we are asking them to perform inference on that value.
@@
 This is likely the cause of the contradiction between uncertainty visualisation papers following identical results to normal information visualisation papers, and authors believing uncertainty is inherently *different* to normal visualisation.
-Each of the sections in this paper likely contribute in some way to this contradiction in conceptualisation of uncertainty visualisaiton and the results of the studies in the field.
-Uncertainty is poorly defined, therefore authors do not understand it is not an inherrent property of the data and is actually related to a particular inference question, this incorrect view that uncertainty is always variance or probabilities leads to the class of "uncertainty visualisations" that depict different mathematical objects or visual features with little in the sense of consistent rules, this complicated class has caused visualisation authors to miss that they are simply repeating the established evaluation results from normal information visualisations.
+Each of the sections in this paper likely contribute in some way to this contradiction in conceptualisation of uncertainty visualisation and the results of the studies in the field.
+Uncertainty is poorly defined, therefore authors do not understand it is not an inherent property of the data and is actually related to a particular inference question, this incorrect view that uncertainty is always variance or probabilities leads to the class of "uncertainty visualisations" that depict different mathematical objects or visual features with little in the sense of consistent rules, this complicated class has caused visualisation authors to miss that they are simply repeating the established evaluation results from normal information visualisations.
 This situation is a bit of a mess.
-It also highlights a unique and facinating problem faced by uncertainty visualisation.
+It also highlights a unique and fascinating problem faced by uncertainty visualisation.
 If asking direct questions about uncertainty causes us to treat it as a signal, how do we evaluate uncertainty as *noise*?
@@
 This is not to say any paper that suggests a visualisation without an evaluation experiment is completely lacking in justification, and there are many papers that suggest a novel visualisation without an evaluation study.
-Sometimes these papers are a preliminary step in finding a solution for common problems and intend to evaluate the visualsiation in later work.
+Sometimes these papers are a preliminary step in finding a solution for common problems and intend to evaluate the visualisation in later work.
 The reasoning for this is obvious.
 There are often heuristics and biases that are not obvious to us when designing visualisation.
-Additionally, these heuristics and baises can change depending on the larger scope of the graphic and the population we are communicating with [@Spiegelhalter2017; @Kinkeldey2014].
+Additionally, these heuristics and biases can change depending on the larger scope of the graphic and the population we are communicating with [@Spiegelhalter2017; @Kinkeldey2014].
 
@@
 While we do not disagree with this statement, exactly *why* these tasks are different to value extraction tasks is never explained.
-While the task may be more similar to how plots are actually used, that does not necessarily mean it is a better *experimental* environement for evaluating plots.
+While the task may be more similar to how plots are actually used, that does not necessarily mean it is a better *experimental* environment for evaluating plots.
 Let us consider this in more detail.
@@
 Therefore decision making experiments that have an additional layer of noise that value extraction experiments do not.
-We cannot be sure if particpants answered differently from the ground truth because a visualisation was difficult to read *or* because the participants risk utility function did not align with the one set by the authors.
+We cannot be sure if participants answered differently from the ground truth because a visualisation was difficult to read *or* because the participants risk utility function did not align with the one set by the authors.
 Several authors have offered solutions to this issue, however the problem is deeper than any of them seem to realise.
@@
 Any payment scheme that is incorporated into an experiment will also implicitly set up a utility framework, as it is assumed participants will try to maximise their payments.
-In tasks such as this, the ground truth of each question is typically selected to be the value a ration agent would select and visualisations are evaluated on the basis of how far the participants repsponses are from that ground truth.
+In tasks such as this, the ground truth of each question is typically selected to be the value a ration agent would select and visualisations are evaluated on the basis of how far the participants responses are from that ground truth.
 The problem with this, is that all the complexity of the question is in *working out* the best response, and it has little to do with the visualisation itself.
@@
 Their payment scheme, which paid out $0.10 for a correct choice (i.e.
-staying when the house was not burned down or leaving when the house was burned down) and 0 for an incorrect choice (i.e.leaving when the house didnt burn down or staying when the house burned down), meant participants were incentivised to base their entire leave/stay decision on whether or not the likelihood of a fire at their house is above or below 50%.
+staying when the house was not burned down or leaving when the house was burned down) and 0 for an incorrect choice (i.e.leaving when the house didn't burn down or staying when the house burned down), meant participants were incentivised to base their entire leave/stay decision on whether or not the likelihood of a fire at their house is above or below 50%.
 If the participants *did* correctly identify the optimum strategy and answer accordingly, the "decision making experiment" was actually *just a value extraction experiment*.
@@
 The exact reasons we describe visualisation to be so powerful, that they are able to communicate complicated information fast and efficiently *is a product of the system 1 brain* [@daniel2017thinking], however this means displaying information using visualisations have the same weakness' as system 1 thinking, and visualisation authors need to be aware of this.
-Asking participants to answer questions that require complicated calculations does not reflect how visualisations are used, and will liekly be ignored by the participants.
-While we agree with @Hullman2016 who suggested that what determines an approripate ground truth is largely a philosophical exercise, we differ in the sense that we do not think it should be done, lest you create a needlessly complicated value extraction task.
+Asking participants to answer questions that require complicated calculations does not reflect how visualisations are used, and will likely be ignored by the participants.
+While we agree with @Hullman2016 who suggested that what determines an appropriate ground truth is largely a philosophical exercise, we differ in the sense that we do not think it should be done, lest you create a needlessly complicated value extraction task.
 
@@
 Not only this, but the inclusion of uncertainty information should *not* impact the choices of a rational agent.
-A well behaved rational angent *should* ignore uncertainty information, so long as the estimate provided is still the best case.
+A well behaved rational agent *should* ignore uncertainty information, so long as the estimate provided is still the best case.
 This point seems to be lost on many uncertainty visualisation authors.
@@
 They are noisy because of participants individual utility functions and when that is not the case they are thinly veiled and over complicated value extraction experiments.
-Additionally, authors dont even seem to agree how participants should incoreperate uncertainty information, with some authors designing complicated utility functions, and others not realising the best response to their experiment is to ignore uncertainty information entirely.
-For these reasons, uncertainty visualisation experiments should either be designed with these caveates in mind and be designed to untangle why a particular visualisation will elicit a specific decision, or the field should steer away using decision quality and related metrics as a basis for evaluating visualisation performance.
+Additionally, authors don't even seem to agree how participants should incorporate uncertainty information, with some authors designing complicated utility functions, and others not realising the best response to their experiment is to ignore uncertainty information entirely.
+For these reasons, uncertainty visualisation experiments should either be designed with these caveats in mind and be designed to untangle why a particular visualisation will elicit a specific decision, or the field should steer away using decision quality and related metrics as a basis for evaluating visualisation performance.
 
@@
 They reported that visualising uncertainty information caused participants to trust the model in the low variance case, but the results in the high variance case were inconclusive.
-The discussion made it clear the authors thought the uncertainty information should make the visualisation more trustworthy, but conflating trust and the use of a prediction implied uncertaitny information should somehow influence participants to use their own prediction, even though a prediction being uncertainty does not necessarily mean it is incorrect.
+The discussion made it clear the authors thought the uncertainty information should make the visualisation more trustworthy, but conflating trust and the use of a prediction implied uncertainty information should somehow influence participants to use their own prediction, even though a prediction being uncertainty does not necessarily mean it is incorrect.
 Despite this, the authors seemed to assume that the uncertainty information *should* have an influence on that, showing they had not deeply considered *how* uncertainty information should influence the choices of the participants.
@@
 
-#### Other (questionable) attemps to capture uncertainty
+#### Other (questionable) attempts to capture uncertainty
 There is also a swath of studies that are aware a question that boils down to a value extraction experiment, or a question that should be answered by literally ignoring uncertainty information is not what we want when we consider uncertainty visualisations.
@@
 @Ibrekk1987 asked participants for the "best estimate" which was evaluated in accuracy by comparing it to the mean, however the "best estimate" depends on the loss function we are using, and a loss function of minimised error was not implied by the question.
-@Hofmann2012 showed two distributions in 20 different visualisations (a lineup protocol) using a jittered sample, a density plot, a histogram, and a box plot and asked participants.
+@Hofmann2012 showed two distributions in 20 different visualisations (a line-up protocol) using a jittered sample, a density plot, a histogram, and a box plot and asked participants.
 Participants were asked to report in which of the plots was "the blue group furthest to the right" The experiment set up is shown in @fig-right.
-The participants answers were then compared to a ground truth where the correct plot had a blue distribtuon with a right shifted mean.
+The participants answers were then compared to a ground truth where the correct plot had a blue distribution with a right shifted mean.
 By comparing the results to a ground truth statistic and marking participants as "wrong" or "right", the error from the participants that had an alternative interpretation to the concept of "furthest right" was conflated with the error from a the visualisation choice.
@@
 This inclusion of determinism seems to cause the authors to stumble themselves, as they are not consistent with their assumptions.
-In their first experiment @Padilla2017 indicated the correct answer was to assume that the storm was equally intense no matter how far from the center of the distribution an oil rig was, however answering their third experiment correctly hinged on assuming the intensity of the storm at a particular point (which in this experiment they phrased as damage) *does* change in intensity as you move away from the center of the distribution.
+In their first experiment @Padilla2017 indicated the correct answer was to assume that the storm was equally intense no matter how far from the centre of the distribution an oil rig was, however answering their third experiment correctly hinged on assuming the intensity of the storm at a particular point (which in this experiment they phrased as damage) *does* change in intensity as you move away from the centre of the distribution.
 Given these conflicting assumptions, it is unclear how the participants were supposed to adjust the probabilistic path information to answer a deterministic question about which oil rig would receive the most damage.
-The flipside of this is asking participants for a deterministic answer to a probabilistic question.
+The flip side of this is asking participants for a deterministic answer to a probabilistic question.
 @Correll2014 asked participants "how likely is candidate B to win the election?" when the two distributions indicated voter preference.
-Participants were not able to answer the question about likelihood in term of probability, but were instead given seven options from 1=Outcome will be most in favor of A to 7=Outcome will be most in favor of B.
-The ground truth statistic for this question was a scalar multiple of Cohen’s d, indicating participants were supposed to incorperate uncertainty information using a very specific formula that was likely unknown to them but assumed to be used implicitly.
+Participants were not able to answer the question about likelihood in term of probability, but were instead given seven options from 1=Outcome will be most in favour of A to 7=Outcome will be most in favour of B.
+The ground truth statistic for this question was a scalar multiple of Cohen’s d, indicating participants were supposed to incorporate uncertainty information using a very specific formula that was likely unknown to them but assumed to be used implicitly.
 
@@
 These experiments inherently view uncertainty as a variable that is separate to the variable on which we have mapped our signal.
-For example, lets say we have a map were maximum daily temperature is presented using points where the colour (red for hot and blue for cold) of the point is associated with the temperature, and the bluriness of that point is associated with the variance *of* that temperature.
+For example, lets say we have a map were maximum daily temperature is presented using points where the colour (red for hot and blue for cold) of the point is associated with the temperature, and the blurriness of that point is associated with the variance *of* that temperature.
 It is highly likely that our brains will not flatten that into a single variable depicting noise and signal, but rather *separately* extract the temperature (colour) and uncertainty (blur) information as two independent variables.
-If the variables are extracted separately, there is no guarentee that the uncertainty will act as an approriate signal supressor.
-This problem has been noticed by others in the field, that typically use this method (specifically in the spatial uncertainty context) and a desire for representations that integrate uncertainty and signal is one of the reasons for the invention of the value-supressing uncertainty pallet [@Correll2018].
+If the variables are extracted separately, there is no guarantee that the uncertainty will act as an appropriate signal suppressor.
+This problem has been noticed by others in the field, that typically use this method (specifically in the spatial uncertainty context) and a desire for representations that integrate uncertainty and signal is one of the reasons for the invention of the value-suppressing uncertainty pallet [@Correll2018].
 
-Value-supressing uncertainty pallets (VSUP) were developed as a method that would allow the signal and the noise to be interpreted together such that insights gained by the viewers of a plot are appropriately *supressed* by the uncertainty.
+Value-suppressing uncertainty pallets (VSUP) were developed as a method that would allow the signal and the noise to be interpreted together such that insights gained by the viewers of a plot are appropriately *suppressed* by the uncertainty.
 Hence the name of the pallet.
@@
 The next map (which embodies the approach taken by the semantics experiments) is the bivariate map which maps the signal to colour value and the uncertainty to colour hue.
-If visualisation was performing signal supression then that would mean the two dimensional space defined by colour value and colour hue can be mentally "flattened" into a single dimension of "valid value" that can captures the signal the our brain would need to be able to flatted this two dimensional space into a single space of "signal validity".
+If visualisation was performing signal suppression then that would mean the two dimensional space defined by colour value and colour hue can be mentally "flattened" into a single dimension of "valid value" that can captures the signal the our brain would need to be able to flatted this two dimensional space into a single space of "signal validity".
 The idea of two perceptual tasks flattening into one variable in the mind of the viewer may be wishful thinking, but it is not impossible given we are not certain on how the perceptual tasks are mapped within the human brain.
 @Sterzik2023 found that when a value was mapped to the textures of stippling, hatching, and triangles, and found that the difference between two points on this one dimensional texture was actually a 2D space (likely "texture business" and "light/darkness").
-That being said, if we look at the visual signals presented by the bivariate map, where the contrasting light and dark areas actually has no important meaning, it is unlikely this occurs for colour value and hue (or at least it doesnt occur in a way that is useful for uncertainty visualisation).
+That being said, if we look at the visual signals presented by the bivariate map, where the contrasting light and dark areas actually has no important meaning, it is unlikely this occurs for colour value and hue (or at least it doesn't occur in a way that is useful for uncertainty visualisation).
 Instead of hoping that uncertainty might collapse signal values into a single dimension, we can do some of that work ourselves, by using a VSUP which collapses the colour space such that high uncertainty values cannot be extracted.
-It is unclear how useful VSUPs are at actually combining signal and noise and therefore supressing plot level insights, as they have only been tested on simple value extraction tasks that require evaluating a single point [@Correll2018; @Ndlovu2023] rather than looking for spatial relationships (which is arguably what maps are for).
-Following along with this trend, the next way we might consider visualising uncertainty is to combine uncertainty and signal at the earlier stage so the "supressed signal" is represented by a single variable.
-This statistic can then be expressed in a one dimensional colour space, which is a method addopted by the Baysian surprise metric map [@Ndlovu2023] and the excedance probability map [Lucchesi2017].
+It is unclear how useful VSUPs are at actually combining signal and noise and therefore suppressing plot level insights, as they have only been tested on simple value extraction tasks that require evaluating a single point [@Correll2018; @Ndlovu2023] rather than looking for spatial relationships (which is arguably what maps are for).
+Following along with this trend, the next way we might consider visualising uncertainty is to combine uncertainty and signal at the earlier stage so the "suppressed signal" is represented by a single variable.
+This statistic can then be expressed in a one dimensional colour space, which is a method adopted by the Bayesian surprise metric map [@Ndlovu2023] and the exceedance probability map [Lucchesi2017].
 .
@@
 
-*This plot is the highest priortiy to for replacing with my own R coded maps, becuase the effects of these different mapings is really not visible using these methods* Five different ways to include error information in a map.
-A typical chloropleth map is considered the map that ignores uncertainty and simply visualises the signal.
-The bivariate map visualises the uncertainty, but does so using two different visual channels, which gives the perception of two different signals (or one signal in the assumptions the colors contrast with each other).
+*This plot is the highest priority to for replacing with my own R coded maps, because the effects of these different mappings is really not visible using these methods* Five different ways to include error information in a map.
+A typical choropleth map is considered the map that ignores uncertainty and simply visualises the signal.
+The bivariate map visualises the uncertainty, but does so using two different visual channels, which gives the perception of two different signals (or one signal in the assumptions the colours contrast with each other).
 An adjustment to the bivariate map is the VSUP which is similar to the bivariate plot except it merges high uncertainty values in the palette to reduce our ability to discern values that have a high enough error to not be a valid signal.
-The Bayesian surprise metric is similar to the lineup protocol, where the colour value is the difference between the observers posterior and prior beliefs.
+The Bayesian surprise metric is similar to the line-up protocol, where the colour value is the difference between the observers posterior and prior beliefs.
 The exceedance probability map is a direct visualisation of the probability of a specific hypothesis.
@@
 These maps make the importance of combining uncertainty and signal in a single visual channel clear.
-A chloropleth map will show signal that is not valid inference because of high uncertainty.
+A choropleth map will show signal that is not valid inference because of high uncertainty.
 At the other end of the spectrum, the bivariate map will show signal that is not always interesting because it forces us to interpret uncertainty and signal separately.
 As we move through these methods, it seems that the validity of any overarching insight becomes more visible at the cost of our ability to extract particular values of signal or noise.
-Therefore, given that the primary goal of visualisation *is* insights [@North2006], visualisation authors should err on the side of representing supressed signal as a single variable, rather than visualising uncertainty separately using two different channels.
+Therefore, given that the primary goal of visualisation *is* insights [@North2006], visualisation authors should err on the side of representing suppressed signal as a single variable, rather than visualising uncertainty separately using two different channels.
 
 ## Suggestions to measure uncertainty
-So, the current methods of measuring or understanding the role of uncertainty in a visualisation is quesitonable at best, however this is not because visualiation authors are missing the mark, but rather uncertainty is *particularly* difficult to express in a visualisation.
+So, the current methods of measuring or understanding the role of uncertainty in a visualisation is questionable at best, however this is not because visualisation authors are missing the mark, but rather uncertainty is *particularly* difficult to express in a visualisation.
 In simple estimates or verbal communication, the signal is often easy to identify because it is what we are explicitly saying.
@@
 Two people looking at the same visualisation might, just by chance, develop two entirely different insights and draw inference on two completely different statistics.
-These unique and fascinating challenges that are faced by uncertainty visualisaiton have been completely untouched by the literature.
+These unique and fascinating challenges that are faced by uncertainty visualisation have been completely untouched by the literature.
 This section will cover some interesting research in uncertainty visualisation and suggestions for better ways to measure uncertainty.
@@
 
-### Just noticiable signal
+### Just noticeable signal
 It could be argued that a well done uncertainty visualisations should have an imperceptible signal unless the signal would be identified with a hypothesis test, almost like a reverse line up protocol, but this idea also has some issues that should be considered.
 The reject or do not reject concepts in hypothesis testing do not offer a complete image of uncertainty, and exploration of uncertainty visualisation largely stems from a desire to move away from this binary framework.
-Additionally, setting up an uncertainty visualisation where the participants are expected to notice the signal once the data behind the visualisation passes a hypothesis test implies the signal *is* noticiable to a human at that level.
+Additionally, setting up an uncertainty visualisation where the participants are expected to notice the signal once the data behind the visualisation passes a hypothesis test implies the signal *is* noticeable to a human at that level.
 @Patrick2023 compared people ability to recognise patterns in a residual plot to typical statistical tests and found human viewers looking at a plot were less sensitive than the typical residual tests.
-These experiments utilised the lineup protocol which has been suggested as a method to check if perceived patterns are real or merely the result of chance [@Buja2009; @Wickham2010; @Chowdhury].
+These experiments utilised the line-up protocol which has been suggested as a method to check if perceived patterns are real or merely the result of chance [@Buja2009; @Wickham2010; @Chowdhury].
 This concept bears similarity to the goal of uncertainty visualisation, but it is not quite the same.
-@fig-hypyvs shows the conceptual difference between the lineup protocol and uncertainty visualisation.
+@fig-hypyvs shows the conceptual difference between the line-up protocol and uncertainty visualisation.
 
 ![Difference between the null plot vs the uncertainty plot.
-A lineup protocol displays the uncertainty about the null and identifies if the true data plot is identifiable (and therefore significiantly difference) while an uncertainty plot displays the variance of the estimated value and assesses if the null of "no signal" is within this plot (e.g.
+A line-up protocol displays the uncertainty about the null and identifies if the true data plot is identifiable (and therefore significantly difference) while an uncertainty plot displays the variance of the estimated value and assesses if the null of "no signal" is within this plot (e.g.
 if the error bars overlap with zero).
@@
 Despite the common problems detailed in the previous sections, there is some interesting work in the uncertainty visualisation space.
-This can come in the form of an uncertainty visualisation that attempts to visualise a typically ignored aspect of uncertainty, or an experiment that avoids the pitfalls detailled in the previous sections.
+This can come in the form of an uncertainty visualisation that attempts to visualise a typically ignored aspect of uncertainty, or an experiment that avoids the pitfalls detailed in the previous sections.
 
@@
 Some authors have chosen to focus on these unquantifiable cases and look at expressing more complicated cases of uncertainty.
-@Tierney2023 builds upon the tidy data pricipals to allow users to handle missing values.
+@Tierney2023 builds upon the tidy data principles to allow users to handle missing values.
 This includes data plots with a missing value "shadow" that allows visualisation authors to identify if the variables used in a plot have any structure in their missing values, which would contribute to uncertainty.
@@
 Another visualisation method that has a lot of potential is visualisation of samples.
-Visualisations that opt to express a signal as a sample rather than an estimate have the potential to supress signal since it is not explicitly visualised, however this has yet to be shown in evaluation experiments.
-This is not to say that visualisations of mass would not be able to perform signal supression, but a sample can easily be expressed using aesthetics such as colour on a map and mass visualisation often struggling with issues such as over or under smoothing.
+Visualisations that opt to express a signal as a sample rather than an estimate have the potential to suppress signal since it is not explicitly visualised, however this has yet to be shown in evaluation experiments.
+This is not to say that visualisations of mass would not be able to perform signal suppression, but a sample can easily be expressed using aesthetics such as colour on a map and mass visualisation often struggling with issues such as over or under smoothing.
 These sampling methods can show more of the messiness of the data that sits behind a model.
@@
 
-A final method to perform signal supression is simply to visualise the data if it is available and relevant to the uncertainty distribution.
-An example of this is shown in @fig-census for racial distributions spatially in america, an interactive version of the plot can be found [here](https://edition.cnn.com/interactive/2021/us/census-race-ethnicity-map/).
+A final method to perform signal suppression is simply to visualise the data if it is available and relevant to the uncertainty distribution.
+An example of this is shown in @fig-census for racial distributions spatially in America, an interactive version of the plot can be found [here](https://edition.cnn.com/interactive/2021/us/census-race-ethnicity-map/).
 This map shows the typical causes of uncertainty in a spatial model, (e.g.
@@
 This does not mean that visualising raw data instead of implementing sampling techniques is always a valid uncertainty visualisation that will prevent insignificant signal from getting through.
-@Buja2009 illustrated how groups that appear linearly separable in a linear discriminant analysis (LDA) visualisation of the data can actually be the result of a LDA performed on too many variables, something that was not clear from the visualisation until the lineup protocol was implemented.
-However it is simple but effective option that seems to be laregly overlooked by the uncertainty visualisaiton community.
+@Buja2009 illustrated how groups that appear linearly separable in a linear discriminant analysis (LDA) visualisation of the data can actually be the result of a LDA performed on too many variables, something that was not clear from the visualisation until the line-up protocol was implemented.
+However it is simple but effective option that seems to be largely overlooked by the uncertainty visualisation community.
 
@@
 
-Our first suggestion for future work is a mathematical defintion of uncertainty.
+Our first suggestion for future work is a mathematical definition of uncertainty.
 There is currently a lot of work that quantifies the uncertainty through bias and variance at various stages of the data analysis pipeline separately.
@@
 The concept of uncertainty should also be formalised within the grammar of graphics.
-This formalisation would allow uncertainty visualisation authors to have a clear understanding of what is or is not an uncertaitny visualisation.
+This formalisation would allow uncertainty visualisation authors to have a clear understanding of what is or is not an uncertainty visualisation.
 Additionally placing uncertainty visualisation in the framework that is used to understand existing information visualisation research would help authors understand when existing methods can be used to explain their results.
-Incorperating uncertainty into the grammar of graphics will also give a more precise concept of the information contained within a plot.
-Other fields of science employ marginal changes when designing experiments to ensure it is well understood *what* aspect of their experiment is contributing to their results, and a better sense of what "marginal" is in the case of uncertainty visualisaion would greatly help the field.
+incorporating uncertainty into the grammar of graphics will also give a more precise concept of the information contained within a plot.
+Other fields of science employ marginal changes when designing experiments to ensure it is well understood *what* aspect of their experiment is contributing to their results, and a better sense of what "marginal" is in the case of uncertainty visualisation would greatly help the field.
 
 Additionally the experimental practices within uncertainty visualisation need to be less ad hoc and more standardised.
-If we are going to consider uncertainty as noise, not signal, there needs to be a way to identify this signal supression in an experimental design.
+If we are going to consider uncertainty as noise, not signal, there needs to be a way to identify this signal suppression in an experimental design.
 As the literature currently exists, there is no way to combine papers to get a meaningful sense of how uncertainty information is understood by a viewer.
@@
 The interaction between sample size and task is of particular interest to the uncertainty visualisation community, as uncertainty can be expressed through multiple observations using a sample, or through a single value using an error.
-Of course, this is limited by the fact that there also isnt a definition for what is a "task" and given the mess created by the lack of formalisation in uncertainty visualisation, it may be wise to formalise that concept before performing these experiemnts.
+Of course, this is limited by the fact that there also isn't a definition for what is a "task" and given the mess created by the lack of formalisation in uncertainty visualisation, it may be wise to formalise that concept before performing these experiments.
 @Amar2005 suggested a taxonomy for information visualisation based on the types of tasks we use visualisations for and suggest 10 "analytical primitives" that we can then map to visualisations, which could be a good starting point.
@@
 library(spelling)
-qmd <- "ideas/confirmation/confirmationreport.qmd"
+qmd <- "paper.qmd"
 ignore <- readLines("WORDLIST")

### 2024-03-25 paper.qmd
image + caption updates

@@
 
-Uncertainty visualisation is a somewhat new subfield of visualisation, with early papers that specifically reference an "uncertainty visualisation" popping up in the late 90s and early 2000s (*Cite Gap 1*).
-Despite the youth of the field, commonly used uncertainty visualisations such as a box plot or histogram have been around for almost as long as statistical graphics themselves (*Cite Gap 2*).
-Early mentions of "uncertainty visualisation" appear in computer science papers discussing uncertainty visualisation as a subfield of geospatial information visualisation, specifically for the many uncertainties that come with satellite images (*Cite Gap 1*).
-In the modern day, uncertainty visualisation has exploded into its own field that applies to a wide range of contexts and is no longer bound by geospatial information (*Cite Gap 1*).
+Uncertainty visualisation is a somewhat new subfield of visualisation.
+Early papers that specifically reference "uncertainty visualisation" start popping up in the late 80s [@Ibrekk1987], with papers in the early 90s (specifically in the area of geospatial information visualisation) recgonising it as a necessary subfield of visualisation [@MacEachren1992; Carr1992].
+Despite the youth of the specific sub-field, commonly used uncertainty visualisations such as a box plot or histogram have been around for almost as long as statistical graphics themselves.
+In the modern day, uncertainty visualisation has exploded into its own field that applies to a wide range of contexts.
 
@@
 Some consider it to be synonymous with specific terms defined in mathematics, such as probability, variance, error, or precision.
-Others consider it to be an encompassing umbrella term of which for these mathematically defined objects are only examples, and include concepts that are somewhat related, such as missing values (*Cite Gap 3*).
-This wide array of uncertainty definitions becomes a snake eating its own tail, as many authors opt to redefine uncertainty themselves[], ignore defining uncertainty entirely [], or pick a definition from a seemingly randomly previously published papers [], all of which make variance in uncertainty definitions even worse (*Cite Gap 3.5*).
+Others consider it to be an encompassing umbrella term of which for these mathematically defined objects are only examples, and include concepts that are somewhat related, such as missing values (*Cite Gap 3: Organise paper definitions*).
+This wide array of uncertainty definitions becomes a snake eating its own tail, as many authors opt to redefine uncertainty themselves[], ignore defining uncertainty entirely [], or pick a definition from a seemingly randomly previously published papers [], all of which make variance in uncertainty definitions even worse (*Cite Gap 3: Organise paper definitions*).
 
@@
 The uncertainty visualisation literature is completely awash with papers that define uncertainty using a vague encompassing phrase, (e.g.
-@utypo defined it as any deviation from complete determinism) but go on to *quantify* uncertainty as a PDF, error, or some other easily quantified mathematical object when it comes time to do the visualisation (*Cite Gap 4*).
-This definition swap happens so subtly that it seems to go unnoticed by authors writing the papers, and the fact that "uncertainty visualisation" methods don't even notice they are exclusively focus on depictions of easily quantifiable uncertainty, such as error [], mass [], or variance [] (*Cite Gap 5*).
+@utypo defined it as any deviation from complete determinism) but go on to *quantify* uncertainty as a PDF, error, or some other easily quantified mathematical object when it comes time to do the visualisation (*Cite Gap 4: Organise uncertainty quantification*).
+This definition swap happens so subtly that it seems to go unnoticed by authors writing the papers, and the fact that "uncertainty visualisation" methods don't even notice they are exclusively focus on depictions of easily quantifiable uncertainty, such as error [], mass [], or variance [] (*Cite Gap 4: Organise uncertainty quantification*).
 This has left the expressions of uncertainty that are hard to quantify or visually differentiate untouched, despite many papers calling for their invention (*Cite Gap 6*).
@@
 disagreement among parties) [], if the uncertainty is random or systematic [], statistical or bounded [], accuracy or precision [], if the uncertainty is about the past or the future [], the stage of the data analysis pipeline it comes from [], the severity of the uncertainty [], how quantifiable the uncertainty is [@Spiegelhalter2017], etc., etc., etc.
-(*Cite Gap 7*).
+(*Cite Gap 7: Uncertainty Taxonomies*).
 In their own way, each of these taxonomies show an aspect of uncertainty that an author felt was important to differentiate.
@@
 The level axis from the @utypo taxonomy also hints towards an important consideration in uncertainty, because establishing how quantifiable uncertainty is informs us how it can be communicated, something that @Spiegelhalter2017 identified in the form of "precision".
-It is commonly noted that each source of uncertainty in an analysis must be discussed in isolation, but combining the uncertainty from every stage is near impossible [@Spiegelhalter2017] (*Cite Gap 8*).
+It is commonly noted that each source of uncertainty in an analysis must be discussed in isolation, but combining the uncertainty from every stage is near impossible [@Spiegelhalter2017].
 It is clear that elements of these taxonomies are identified in many other comments on uncertainty and represent *real* considerations that need to be made when defining uncertainty, even if that definition is yet to be published.
@@
 
-![The graphics displayed by @Boukhelifa2012 to identify if there is an intuitive connection between sketchiness and uncertainty.](sketchiness.png){#fig-sketchy}
+![The graphics displayed by @Boukhelifa2012 to identify if there is an intuitive connection between sketchiness and uncertainty.
+These graphic were made without concern as to what (i.e.
+which statistic) in the image is supposed to be uncertain.
+This leads to the images being difficult to connect to uncertainty even if we assume tha that is what it is supposed to represent.
+For example, it is unclear how you can be uncertain about a train line (S5).](sketchiness.png){#fig-sketchy width=50%}
 
@@
 This confusion (or indifference) also creates many imprecisely named papers.
-Many papers will boast a title that claims to be about uncertainty visualisation, but simply depicts different visual representations of a PDF, however it takes reading the methodology to find this out (*Cite Gap 9*).
+Many papers will boast a title that claims to be about uncertainty visualisation, but simply depicts different visual representations of a PDF, however it takes reading the methodology to find this out (*Cite Gap 4: Organise uncertainty quantification*).
 While this is a minor by product of the definition issue discussed earlier, it does result in a literature that is needlessly difficult to navigate.
@@
 This appears in the literature as a mountain of seemingly conflicting statements about what is, or is not, an "uncertainty visualisation".
-For example @Leland2005 mentions that popular graphics, such as pie charts and bar charts omit uncertainty, however at least one or both of these charts are used in most "uncertainty visualisation" experiments [@Ibrekk1987] (*Cite Gap 10*).
+For example @Leland2005 mentions that popular graphics, such as pie charts and bar charts omit uncertainty, however at least one or both of these charts are used in most "uncertainty visualisation" experiments [@Ibrekk1987] (*Cite Gap 8: Bar/pie uncertainty graphics*).
 @Wickham2011 suggests their product plot framework, which includes histograms, should have a way to measure uncertainty, but does not consider that a histogram is *already* a depiction of mass and would already be considered an uncertainty visualisation were our statistic of interest the population mean.
@@
 Of course the authors of these papers did not intend for the taxonomy to be a complete description of uncertainty visualisations, however these distinctions that are often ignored are subtly important.
-(*Cite Gap 10*)
+(*Cite Gap 9: Other uncertainty vis taxonomies*)
 
@@
 
-![This visualisation is here as a placeholder.
-I am going to replace it with an R coded visualisation that shows the 5 considerations for showing uncertainty in a plot according to @Kinkeldey2014 ](paperscreenshots/collage2.jpg){#fig-location}
+![Planned visualisation: Add an R coded visualisation that shows the 5 considerations for showing uncertainty in a plot according to @Kinkeldey2014.
+One that has the consideration, and one that doesn't (try and keep as much the same between visuals as possible)](notice.jpeg){#fig-location width=50%}
 
@@
 It can help provide reasons for *why* there may be specific characteristics in our data, however it is not required information for the model.
-@fig-spatial shows the process of boiling information down into data, visualising it, and including contextual information to improve our scientific reasoning.
+@fig-datapipe shows the process of boiling information down into data, visualising it, and choosing whether or not to including contextual information to improve our scientific reasoning.
 The concept that contextual information is not the same as statistical information is also highlighted in *The Grammar of Graphics* which states:
@@
 
-![Data pipeline with visualisations to show why we are ignoring contextual information for the time being.
+::: {#fig-datapipe layout-ncol=2}
+
+![Australia example](spatialinfo.jpeg)
+
+![USA example](spatialinfo2.jpeg)
+
+Data pipeline with visualisations to show why we are ignoring contextual information in the case of uncertainty visualisation.
 The visualisations that depict the data are similar to typical statistical graphics.
-The purpose of the spatial information is to provide context to the information we extract from the statistical aspect of the graphic.](spatialinfo.jpeg){#fig-spatial}
+The Australia example depicts a heat map for temperature in Australia.
+The statistical level of this graphic is no different than a typical density plot that could be made with  exist with any data.
+The geographic information is merely used to provide real world context to the statistical message we extract from graphic, however that addition is of great importance.
+The USA example depicts a map showing the voting proportions in a US election.
+This map illustrates another example of a spatial graphic, however the location of the data is only loosely related to the real world geographic location.
+The graphic maintains readability because the proportion of votes matters as the US state level, and the number of people in a state and its proportion of votes are far more important for drawing inference on who won an election than specific spatial information.
+Notice: Both Graphic needs to be fixed up.
+:::
 
@@
 
-![The uncertainty visualisation used by @Kay2016 to show the uncertainty around a bus arrival time prediction.](bustime.png){#fig-bus}
+![The uncertainty visualisation used by @Kay2016 to show the uncertainty around a bus arrival time prediction.
+The data being a "time" estimate does not prohibit the authors from using a density plot, a visualisation method that is typically used for attributes.
+The time element is irrelevant because the viewer is concerned about their estimated wait time NOW, not how this estimate will change over time.
+The predictions relationship to any predictors is assumed to be irrelevant in this visualisation.](bustime.png){#fig-bus width=50%}
 
@@
 @Slingsby2023 discussed using a gridded glyph map for spatial modelling, however the statistical elements are identical to a fluctuation diagram depicting $f(infection|time, age, longitude, latitude)$ presented in @Wickham2011.
+A side by side depiction of the two graphics is displayed in @fig-prodplots.
 The plot by @Slingsby2023 is simply a special case of the graphic by @Wickham2011, where a separate visualisation is made for each time point, the continuous function $infection | age$ has replaced the categorical variable $happy$, a fade is added for the population size, and a map is shown in the background.
@@
 
-```{r}
-#| output: asis
-#| echo: false
-img_files <- fs::dir_ls("prodplots", glob="*.png")
-cat("::: {layout-ncol=2}\n",
-    glue::glue("![]({img_files})\n\n\n"),
-    ":::",
-    sep = ""
-)
-```
+::: {#fig-prodplots}
+
+![Gridded glyphmap](prodplots/img1.png)
+![Fluctuation diagram](prodplots/img2.png)
+
+A side by side comparison of the gridded glyphmap and the fluctuation diagram.
+One is considered a spatial visualisation technique and the other is considered a normal statistical diagram.
+The product plot and the gridded glyph map are visualisations depicting the relationship $f(w|x,y,z)$ and the similarity in the two plots shows that this three (or 4) variable relationship is the key feature of statistical importance, not the spatial nature of the data.
+The limitations faced by spatial visualisation are often identical to the limitations faces by information visualisation and creating unnecessary divides within the field makes it more difficult to navigate and find relevant statistical visualisations.
+:::
 
@@
 2) Uncertainty is not of interest in of itself and is typically layered on at the end of a the visualisation pipeline, so uncertainty visualisation is the field of adding error information into already established graphics and *thus* uncertainty information is a high dimensional visualisation problem.
-Papers will often discuss uncertainty in relation to one of these motivating reasons, the evaluation experiments motivated by (1) often perform different visualisations of PDFs, while the papers motivated by (2) will focus on trying to impute uncertainty as error within one of the existing channels, such as colour (*Cite Gap 11*).
+(*Cite Gap 10: Reason uncertainty visualisation is different*)
+Papers will often discuss uncertainty in relation to one of these motivating reasons, the evaluation experiments motivated by (1) often perform different visualisations of PDFs, while the papers motivated by (2) will focus on trying to impute uncertainty as error within one of the existing channels, such as colour (*Cite Gap 10*).
 These dual motivations that co-exist with uncertainty visualisation authors rarely mentioning the others, create confusion in the field as to its purpose.
@@
 ### Information asymmetry in evaluated plots
-Visualisation authors are almost unanimous in commenting that the "information" in two plots must be the same in order for the visual techniques to be compared [@Cleveland1984; @Kinkeldey2014] (*Cite gap 12*).
+Visualisation authors are almost unanimous in commenting that the "information" in two plots must be the same in order for the visual techniques to be compared [@Cleveland1984; @Kinkeldey2014] (*Cite gap 12: Equal information*).
 However, what makes two visualisations equal in "information" is not consistent and is regularly mentioned without it being clear exactly what this information is.
@@
 @Hofman2020 comments that "theoretically" the sampling distribution of the mean and the prediction interval of a new observation are equal "so long as one knows the sample size", but does not recognise the several assumptions and statistical knowledge that is required to compare the two.
-*Other examples* (*Cite Gap 13*).
+(*Cite Gap 13: examples of what is defined as "information"*).
 @Kinkeldey2014 adopts an existing definition also that suggests two graphics are informationally equivalent if all the information in one plot is inferable from the other and vice-versa, but adds that two plots are computationally equivalent if that information can be extracted from both plots with similar easy and speed.
@@
 
-![The grammar of graphics data analysis pipeline from @Leland2005.
+![(Notice: Need to replace with a more relevant data pipeline that includes all the uncertainty stuff and how uncertainty arises at each stage.) The grammar of graphics data analysis pipeline from @Leland2005.
 the "varset" stages are the mathematical components of the graphic, the mathematical components define the "information" in the graph, while the "graph" stages are the visual elements that determine the "computational" cost of the graph.](grammarofgraphicspipeline.png){#fig-pipeline}
@@
 They found that the plot that depicted outcomes from $P(A, B)$ where the distance between the two outcomes was equivalent to $a-b$ was better at answering the question "What is $P(B<A)$" than violin or error bar plots depicting $P(A)$ and $P(B)$ separately.
-*Other examples* (*Cite Gap 14*).
+(*Cite Gap 14: Examples of mathematically obvious results*).
 
@@
 @Cheong2016 found that participants were better at answering questions when they were explicitly given the relevant probability in text rather than when they needed to read it off a map.
-*Other examples* (*Cite Gap 15*).
+(*Cite Gap 15: Examples of replicated perceptual task experiments*).
 
@@
 Setting up an experiment that requires participants to do laborious calculations or strenuous mental effort for a small marginal benefit is unlikely to result in successful participant responses because it is not in the "spirit" of visualisation.
-(*Cite Gap 16*)
+(*Cite Gap 16: Examples of value extraction decision making tasks*)
 
@@
 At no point did the authors seem to realise a rational agent would accept the model estimate every time, no matter how uncertain the information was, and participants were actually disqualified from the study for taking this approach to answering the questions.
-(*Cite Gap 17*)
+(*Cite Gap 17: Examples of studies where the correct choice was to ignore the uncertainty information*)
 
@@
 Despite this, the authors seemed to assume that the uncertainty information *should* have an influence on that, showing they had not deeply considered *how* uncertainty information should influence the choices of the participants.
-(*Cite Gap 18*)
+(*Cite Gap 18: examples of studies where authors measure trust*)
 
@@
 
-![This shows the user interface for the experiment performed by @Hofmann2012. The question of "furthest to the right" is open to interpretation. ](furthestright.png){#fig-right}
+![This shows the user interface for the experiment performed by @Hofmann2012. The question of "furthest to the right" is open to interpretation.  ](furthestright.png){#fig-right width=50%}
 
@@
 
-::: {#fig-maps}
+::: {#fig-maps layout-ncol=2}
 
-![](mappals/img1.png)
+![Choropleth map](mappals/img1.png)
 
-![](mappals/img2.png)
+![Bivariate map](mappals/img2.png)
 
-![](mappals/img4.png)
-![](mappals/img3.png)
+![VSUP map](mappals/img4.png)
 
-![](mappals/img5.png)
+![Bayesian surprise map](mappals/img3.png)
 
-*This plot is the highest priority to for replacing with my own R coded maps, because the effects of these different mappings is really not visible using these methods* Five different ways to include error information in a map.
+![Exceedance probability map](mappals/img5.png)
+
+Five different ways to include error information in a map.
 A typical choropleth map is considered the map that ignores uncertainty and simply visualises the signal.
@@
 The exceedance probability map is a direct visualisation of the probability of a specific hypothesis.
+Notice: This plot is the highest priority to for replacing with my own R coded maps, because the effects of these different mappings is really not visible using these methods.
 :::
@@
 This concept bears similarity to the goal of uncertainty visualisation, but it is not quite the same.
-@fig-hypyvs shows the conceptual difference between the line-up protocol and uncertainty visualisation.
+@fig-hypvs shows the conceptual difference between the line-up protocol and uncertainty visualisation.
 
-![Difference between the null plot vs the uncertainty plot.
+![Notice: I will fix this image up, it is currently a quick scribble.
+Difference between the null plot vs the uncertainty plot.
 A line-up protocol displays the uncertainty about the null and identifies if the true data plot is identifiable (and therefore significantly difference) while an uncertainty plot displays the variance of the estimated value and assesses if the null of "no signal" is within this plot (e.g.
-if the error bars overlap with zero).
-](hypvs.jpeg){#fig-hypvs}
+if the error bars overlap with zero).](hypvs.jpeg){#fig-hypvs}
 
@@
 
-![Example of a I will just make my own version of a climate scenario plot, this one is just here from a random site as a place holder](climatescenario.png){fig-climatescenario}
+![Notice: I will make my own Eversion of a climate scenario plot, this one is just here from a random site as a place holder.
+A climate scenario plot shows the forecasts of different emission scenarios depending on human choices.
+The model visualisation wants to show the impact of human choice on the visualisation, rather than assume it away, so five different cases are shown in the model.
+Each of the case studies also depicts the statistical uncertainty of the model using a confidence interval.
+](climatescenario.png){#fig-climatescenario}
 
@@
 This may not have a detrimental effect on the viewers ability to extract global statistics, as it seems they can be extracted from a visualisation of a sample with ease [Franconeri2021].
-Sample visualisations have been used in maps with the pixel map, shown in @fig-pixelmap [Lucchesi2017], but is more commonly used in animation with the HOPs plots [@Hullman2015] or similar concepts [@Blenkinsop2000].
-This method has also been adopted by visualisation authors outside of academia as can be seen in the  and the New York Times class mobility figure, shown in @fig-nyt, an animated version of which can be found [here](https://www.nytimes.com/interactive/2018/03/19/upshot/race-class-white-and-black-men.html).
+Sample visualisations have been used in maps with the pixel map, shown in @fig-samples [Lucchesi2017], but is more commonly used in animation with the HOPs plots [@Hullman2015] or similar concepts [@Blenkinsop2000].
+This method has also been adopted by visualisation authors outside of academia as can be seen in the  and the New York Times class mobility figure, shown in @fig-samples, an animated version of which can be found [here](https://www.nytimes.com/interactive/2018/03/19/upshot/race-class-white-and-black-men.html).
 
-![](pixelmap.png){fig-pixelmap}
-![](classmobility.png){fig-nyt}
+::: {#fig-samples layout-ncol=2}
+
+![Pixel map](pixelmap.png){#fig-pixelmap}
+
+![Outcome plot](classmobility.png){#fig-nyt}
+
+Notice: Will replace both plots with my own r visualisations.
+The pixel map depicts a map of the different poverty rates in local government areas in the US state of Missouri.
+The NYT visualisation shows an screencap of an animation that depicts the economic status of white and black boys who were raised in wealthy families.
+By displaying a psudo-sample rather than an estimate and a variance, the visualisation does some signal suppression as the true poverty rate is masked and can only be extracted by taking the global statistic of a sample.
+You can also depict the signal using explicit values, as the the NYT visualisation did, but it is important that the main visual features depict the uncertainty in the estimate.
+:::
 
@@
 
-![](censusdotmap.png){fig-census}
+![Census dot map](censusdotmap.png){#fig-census}
 

### 2024-03-25 writingdumpster.qmd
image + caption updates

@@
 # Citation Gaps
-Cite Gap 1: When did uncertainty visualisaiton become its own subfield of visualisation?
 
-Cite Gap 2: When was the first "uncertainty visualisation" used (likely a box plot of histogram)
+Cite Gap 3: Organise Uncertainty definitions
+- Some consider it to be synonymous with specific terms defined in mathematics, such as probability [*HERE*], variance[*HERE*], error[*HERE*], or precision[*HERE*].
+Others consider it to be an encompassing umbrella term of which for these mathematically defined objects are only examples[*HERE*], and include concepts that are somewhat related, such as missing values[*HERE*]
+- This wide array of uncertainty definitions becomes a snake eating its own tail, as many authors opt to redefine uncertainty themselves[*HERE*], ignore defining uncertainty entirely [*HERE*], or pick a definition from a seemingly randomly previously published papers [*HERE*], all of which make variance in uncertainty definitions even worse
 
-Cite Gap 3: Some consider it to be synonymous with specific terms defined in mathematics, such as probability [*HERE*], variance[*HERE*], error[*HERE*], or precision[*HERE*].
-Others consider it to be an encompassing umbrella term of which for these mathematically defined objects are only examples[*HERE*], and include concepts that are somewhat related, such as missing values[*HERE*]
-
-Cite Gap 3.5: This wide array of uncertainty definitions becomes a snake eating its own tail, as many authors opt to redefine uncertainty themselves[*HERE*], ignore defining uncertainty entirely [*HERE*], or pick a definition from a seemingly randomly previously published papers [*HERE*], all of which make variance in uncertainty definitions even worse
-
-Cite Gap 4:The uncertainty visualisation literature is completely awash with papers that define uncertainty using a vague encompassing phrase but go on to *quantify* uncertainty as a PDF, error, or some other easily quantified mathematical object when it comes time to do the visualisation.
-
-Cite Gap 5: This definition swap happens so subtly that it seems to go unnoticed by authors writing the papers, and the fact that "uncertainty visualisation" methods dont even notice they are exclusively focus on depictions of easily quantifiable uncertainty, such as error [*HERE*], mass [*HERE*], or variance [*HERE*]
+Cite Gap 4: Organise uncertainty quantification
+- The uncertainty visualisation literature is completely awash with papers that define uncertainty using a vague encompassing phrase but go on to *quantify* uncertainty as a PDF, error, or some other easily quantified mathematical object when it comes time to do the visualisation.
+- This definition swap happens so subtly that it seems to go unnoticed by authors writing the papers, and the fact that "uncertainty visualisation" methods dont even notice they are exclusively focus on depictions of easily quantifiable uncertainty, such as error [*HERE*], mass [*HERE*], or variance [*HERE*]
+- Many papers will boast a title that claims to be about uncertainty visualisaiton, but simply depicts different visual representations of a PDF, however it takes reading the methodology to find this out
 
@@
 
-Cite Gap 7: Taxonomies split uncertainty based on an endless stream of arbitrary lines in the sand, such as whether the uncertainty is due to true randomness or a lack of knowledge [@Spiegelhalter2017 *HERE*], if the uncertainty is in the the attribute, spatial elements, or temporal element of the data [*HERE*], whether the uncertainty is scientific (e.g.
+Cite Gap 7: Uncertainty Taxonomies
+- Taxonomies split uncertainty based on an endless stream of arbitrary lines in the sand, such as whether the uncertainty is due to true randomness or a lack of knowledge [@Spiegelhalter2017 *HERE*], if the uncertainty is in the the attribute, spatial elements, or temporal element of the data [*HERE*], whether the uncertainty is scientific (e.g.
 error) or human (e.g.
@@
 
-Cite Gap 8: It is commonly noted that each uncertainty must be discussed in isolation, but combining the uncertainty from every stage is near impossible [@Spiegelhalter2017 *MORE HERE*]
+Cite Gap 8: Bar/pie uncertainty graphics
+- For example @Leland2005 mentions that popular graphics, such as pie charts and bar charts omit uncertainty, however at least one or both of these charts are used in most "uncertainty visualisation" experiments [@Ibrekk1987 *MORE HERE*]
 
-Cite Gap 9: Many papers will boast a title that claims to be about uncertainty visualisaiton, but simply depicts different visual representations of a PDF, however it takes reading the methodology to find this out
+Cite Gap 9: Other uncertainty vis taxonomies
 
-Cite Gap 9: For example @Leland2005 mentions that popular graphics, such as pie charts and bar charts omit uncertainty, however at least one or both of these charts are used in most "uncertainty visualisation" experiments [@Ibrekk1987 *MORE HERE*]
+Cite Gap 10: Reason uncertainty visualisation is different
+- Papers will often discuss uncertainty in relation to one of these motivating reasons, the evaluation experiments motivated by (1) often perform different visualisations of PDFs, while the papers motivated by (2) will focus on trying to impute uncertainty as error within one of the existing channels, such as colour
 
-Cite Gap 10: Uncertainty Visualisation Taxonomies
-
-Cite Gap 11: Papers will often discuss uncertainty in relation to one of these motivating reasons, the evaluation experiments motivated by (1) often perform different visualisations of PDFs, while the papers motivated by (2) will focus on trying to impute uncertainty as error within one of the existing channels, such as colour
-
-Cite Gap 12: Visualisation authors are almost unanimous in commenting that the "information" in two plots must be the same in order for the visual techniques to be compared
+Cite Gap 12: Equal information
+- Visualisation authors are almost unanimous in commenting that the "information" in two plots must be the same in order for the visual techniques to be compared
 

### 2024-03-26 writingdumpster.qmd
sorted many notes and moved hurricane comment over

@@
 
-Cite Gap 3: Organise Uncertainty definitions
-- Some consider it to be synonymous with specific terms defined in mathematics, such as probability [*HERE*], variance[*HERE*], error[*HERE*], or precision[*HERE*].
+### Cite Gap 3: Uncertainty definition + signal supression
+> Some consider it to be synonymous with specific terms defined in mathematics, such as probability [*HERE*], variance[*HERE*], error[*HERE*], or precision[*HERE*].
 Others consider it to be an encompassing umbrella term of which for these mathematically defined objects are only examples[*HERE*], and include concepts that are somewhat related, such as missing values[*HERE*]
-- This wide array of uncertainty definitions becomes a snake eating its own tail, as many authors opt to redefine uncertainty themselves[*HERE*], ignore defining uncertainty entirely [*HERE*], or pick a definition from a seemingly randomly previously published papers [*HERE*], all of which make variance in uncertainty definitions even worse
+> This wide array of uncertainty definitions becomes a snake eating its own tail, as many authors opt to redefine uncertainty themselves[*HERE*], ignore defining uncertainty entirely [*HERE*], or pick a definition from a seemingly randomly previously published papers [*HERE*], all of which make variance in uncertainty definitions even worse
 
-Cite Gap 4: Organise uncertainty quantification
-- The uncertainty visualisation literature is completely awash with papers that define uncertainty using a vague encompassing phrase but go on to *quantify* uncertainty as a PDF, error, or some other easily quantified mathematical object when it comes time to do the visualisation.
-- This definition swap happens so subtly that it seems to go unnoticed by authors writing the papers, and the fact that "uncertainty visualisation" methods dont even notice they are exclusively focus on depictions of easily quantifiable uncertainty, such as error [*HERE*], mass [*HERE*], or variance [*HERE*]
-- Many papers will boast a title that claims to be about uncertainty visualisaiton, but simply depicts different visual representations of a PDF, however it takes reading the methodology to find this out
+- @Thomson2005 titled their paper "A typology for visualizing uncertainty" however it is a taxonomy for the specific types of uncertainty that we may want to visualise and a way to calculate that uncertainty.
+While this paper is more about mathematics than visualisation, it is clearly written as a precursor to visualisation.
+- @Griethe2006 commented that uncertainty appears at every step in the analyis pipeline, and commented on some the mathematical definitions that seek to quantify uncertianty, but did not explain how these mathematical definitions can be combined across levels.
+Unfortunately this quantification is needed since she mentioned that in order to visualise uncertainty, it needs to be measurable.
+- @Griethe2006 defined uncertainty as "the degree to which the lack of knowledge about the amount of error is responsible for hesitancy in accepting results and observations without caution".
 
-Cite Gap 6: This has left the expressions of uncertainty that are hard to quantify or visually differentiate untouched, despite many papers calling for their invention [*HERE*]
+### Cite Gap 4: Organise uncertainty quantification
+> The uncertainty visualisation literature is completely awash with papers that define uncertainty using a vague encompassing phrase but go on to *quantify* uncertainty as a PDF, error, or some other easily quantified mathematical object when it comes time to do the visualisation.
+> This definition swap happens so subtly that it seems to go unnoticed by authors writing the papers, and the fact that "uncertainty visualisation" methods dont even notice they are exclusively focus on depictions of easily quantifiable uncertainty, such as error [*HERE*], mass [*HERE*], or variance [*HERE*]
+> Many papers will boast a title that claims to be about uncertainty visualisaiton, but simply depicts different visual representations of a PDF, however it takes reading the methodology to find this out
 
-Cite Gap 7: Uncertainty Taxonomies
-- Taxonomies split uncertainty based on an endless stream of arbitrary lines in the sand, such as whether the uncertainty is due to true randomness or a lack of knowledge [@Spiegelhalter2017 *HERE*], if the uncertainty is in the the attribute, spatial elements, or temporal element of the data [*HERE*], whether the uncertainty is scientific (e.g.
+### Cite Gap 6: All PDF no Missing values
+> This has left the expressions of uncertainty that are hard to quantify or visually differentiate untouched, despite many papers calling for their invention [*HERE*]
+
+- @Griethe2006 mentions that there are detailled classifications of uncertainty but the literature and subsequent techniques are not ballanced, and instead concentrate on scalar values.
+In facing this lack of ballance she creates a new classification system for uncertainty visualisation.
+- Mentioned alleatory vs epistemic uncertainty, in the visualisation literature alleatory uncertainty is more commonly the focus.
+also mentions dependence and joint pdfs as a consideration @Hullman2016
+
+### Cite Gap 7: Uncertainty Taxonomies
+> Taxonomies split uncertainty based on an endless stream of arbitrary lines in the sand, such as whether the uncertainty is due to true randomness or a lack of knowledge [@Spiegelhalter2017 *HERE*], if the uncertainty is in the the attribute, spatial elements, or temporal element of the data [*HERE*], whether the uncertainty is scientific (e.g.
 error) or human (e.g.
@@
 
-Cite Gap 8: Bar/pie uncertainty graphics
-- For example @Leland2005 mentions that popular graphics, such as pie charts and bar charts omit uncertainty, however at least one or both of these charts are used in most "uncertainty visualisation" experiments [@Ibrekk1987 *MORE HERE*]
+### Cite Gap 8: Bar/pie uncertainty graphics
+> For example @Leland2005 mentions that popular graphics, such as pie charts and bar charts omit uncertainty, however at least one or both of these charts are used in most "uncertainty visualisation" experiments [@Ibrekk1987 *MORE HERE*]
 
-Cite Gap 9: Other uncertainty vis taxonomies
+### Cite Gap 9: Vis uncertainty Taxonomies
+- Most studies involve intrinsic approaches such as colour, transparency, and those that used extrinsic techniques typically used glyphs, error bars, grid based techniques, or contouring @Kinkeldey2014
 
-Cite Gap 10: Reason uncertainty visualisation is different
-- Papers will often discuss uncertainty in relation to one of these motivating reasons, the evaluation experiments motivated by (1) often perform different visualisations of PDFs, while the papers motivated by (2) will focus on trying to impute uncertainty as error within one of the existing channels, such as colour
+### Cite Gap 10: High dimensional vs psychological
+> Papers will often discuss uncertainty in relation to one of these motivating reasons, the evaluation experiments motivated by (1) often perform different visualisations of PDFs, while the papers motivated by (2) will focus on trying to impute uncertainty as error within one of the existing channels, such as colour
 
-Cite Gap 12: Equal information
-- Visualisation authors are almost unanimous in commenting that the "information" in two plots must be the same in order for the visual techniques to be compared
+- @Griethe2006 clearly sees uncertainty as additional information or data, and therefore sees the uncertainty visualisation problem as a specific case of the high dimensional visualisation issues.
 
-Cite Gap 13: Examples of what is "information" in a visualisation
+### Cite Gap 12 & 13 & 14: Information & Mathematical stuff
+- we should aim to show enough information to solve a task while avoiding irrelevant distracting information [@kosslyn2006graph]. 
+- @Ibrekk1987 used several displays of a probability density functions, since “formally equivalent representations are often not psychologically equivalent”.
+- @Hullman2016 commented on the lack of justification of particular questions used in experiments and the lack of generalisability in using plot ranking as a response technique.
+- @Hullman2016 discussed the lack of clarity in exactly what "uncertainty" authors are trying to ask about.
+Whether the participants are supposed to answer according to some ground truth or their subjective probability is unclear, as is whether the *uncertainty* depicted represents the uncertainty of a sampling or population distribution.
+- @Bella2005 found that most participants were ignorant to the fact that error bars are used for both confidence intervals and standard error bars, two wildly different indicators of precision.
+(Note: Uncertainty intervals can only be validly compared to a constant value, you compare Data to a theoretical distribution, OR uncertainty to a theoretical threshold (not data).
+The participants should convert the distributions to A - B and then see if the 95% of THAT distribution overlaps with 0)
+- @Blenkinsop2000 tailored their questions according to the visualisation shown, which implies an understanding that different visualisations contain different information.
+Despite this the authors still compared the visualisations according to which were "useful for uncertainty visualisation".
+- When asking questions, researchers should focus on events that can be repeated (because of a misunderstanding of confidence intervals) @Hullman2016 (moved here because it is clear she is touching on the whole "what are we uncertain about" thing)
 
-Cite Gap 14: Examples of mathematically obvious results
-
-Cite Gap 15: Examples of replicated perceptual task experiments
-
-Cite Gap 16: Examples of value extraction decision making tasks
-
-Cite Gap 17: Examples of studies where the correct choice was to *ignore* the uncertainty information
-
-Cite Gap 18: Examples of studies where authors measure trust
-
-## Extra Section Cite Gaps
-Cite Gap - New Section: Logical fallacy interpretation of mathematically unanswerable questions.
-If I tell you my cousins age, and then ask you to guess my age, it would not be unreasonable to guess the age I provided for my cousin.
-Citing this work to state "people incorrectly believe my cousins age is my age" would
-- Deterministic questions about a random event cannot be answered. 
-- @uncertchap2022 took the results of other paper to mean "when participants viewed mean temperature forecasts that included 95% confidence intervals depicted as bars with end caps, they incorrectly believed that the error bars represented high and low temperatures" (this one will depend on that study)
-
-# Moved over notes
-- These industry visualisation are likely made with a large amount of back and forth between different levels of an organisation, something that is important for good visual communication [@Spiegelhalter2017].
-- @Boone2018 organised responses from participants by the approach they used and tried to account for participants prior beliefs (the method of eliciting prior beleifs was, however, questionable). 
-- @Correll2018 “place your 5 ships on the safest locations on the board.” (minimise danger)
-- @Blenkinsop2000 did try to include uncertainty as noise, asking participants to search for a specific outcome (the land classified as grass) in a random .
-However the participants failing to identify the signal (something that would be expected of an uncertainty plot) was seen as a sign the task was too complicated.
-While the task in this experiment *was* certainty far too complicated (and poorly communicated) this discussion indicates the authors did not consider the signal suppression to be the goal of the visualisation, and rather an annoying distraction.
+### Cite Gap 15: Value Extraction (Examples of replicated perceptual task experiments)
+- @Correll2018 “place your 5 ships on the safest locations on the board.” (minimise danger) It is a convoluted value extraction task, where participants basicaly have to look at the palette and see that if it is in the lower uncertainty, and lower error it is better, but higher uncertainty is better
 - @Ibrekk1987 expected participants to calculate the mean using a pie chart that had numerical bins combined (e.g.
@@
 The authors interpreted this as the participants misinterpreting how the CDF works, not seeming to realise they gave them an impossible task.
-- @Blenkinsop2000 tailored their questions according to the visualisation shown, which implies an understanding that different visualisations contain different information.
-Despite this the authors still compared the visualisations according to which were "useful for uncertainty visualisation".
-- @Blenkinsop2000 asked different questions for each uncertainty visualisation, indicating they had some awareness that the question has a significant impact on whether or not the visualisation should be used.
-- @Gschwandtnei2016 When answering “What is the statistical probability (in %) that the interval has already ended at the marked point in time (red line)?” many answers gave the estimated probability to which the marked point in time falls into the interval.
-This is correct when asking for the probability to which the interval has already started at the marked position.
-However, in case of the probability to which the interval has already ended at the marked position, this answer is the inverted probability of what we were looking for.
-They believe this is their own mistake so they so they don’t present these results.
-- @Ibrekk1987 used several displays of a probability density functions, since “formally equivalent representations are often not psychologically equivalent”.
-- @Ibrekk1987 highlights that in the face of a vague question, participants will use the plot to decide what the authors mean. 
-@Sanyal2009 users performed significantly well using Glyph-size when the task was to search for locations of least uncertainty.
+- @Sanyal2009 users performed significantly well using Glyph-size when the task was to search for locations of least uncertainty.
 However, both Glyph-color and Surface-color performed better than Glyph- size when the task was to search for locations of high uncertainty.
 We did not expect to find a significant difference between the two search tasks since both the tasks were designed to find extremes in the data
-@Kim2019 believed an understanding of participants prior beliefs could assist in interpreting uncertainty visualisation results that tend to be incredibly noisy.
-@Thomson2005 titled their paper "A typology for visualizing uncertainty" however it is a taxonomy for the specific types of uncertainty that we may want to visualise and a way to calculate that uncertainty.
-While this paper is more about mathematics than visualisation, it is clearly written as a precursor to visualisation.
-- Some markers of probability that have common uses, probabilities such as 0 (impossible), 0.5 (a fair coin flip), 1 (certain) are easy for people to have an internal sense of *hullman paper*. 
+- The tasks for participants were most commonly value retrival (some where uncertainty and data value are retrived separately, others where they are some separately).
+also also be asked to extract a relative value @Kinkeldey2014
+
+### Cite Gap 16: Decision making stuff 
+- @Boone2018 organised responses from participants by the approach they used
+- @Hullman2016 suggested authors look for signs of heuristics in responses and ask subjects to describe their strategy in their response. 
+- @Hullman2016 What was particularly interesting was people's tendancy to ignore uncertainty in a plot all together, and use a pattern in the signal (such as distance between means) to answer questions that require uncertainty information.
+- @Kim2019 believed an understanding of participants prior beliefs could assist in interpreting uncertainty visualisation results that tend to be incredibly noisy.
+- @Daradkeh2015 found that people tended to to make decisions on a small number of salient pieces of information rather than on systematic evaluations of the costs and benefits of choices.
+- @Daradkeh2015 found that people tended to ignore or give little weight probability information and instead focused on the severity of undersierable outcomes.
+- believed that offering enough decisions with varying rewards you can observe the true subjective probability, however whether or not probability is the right representation of subjective uncertainty is a topic of debate @Hullman2016
+- she also discusses graphical interface (users have to draw the probability distribution) and framing probabilities as natural frequencies are better choices because they reduce noise in responses [@Hullman2016]
+- Consider incorporating utility functions. Include "probability-coherence" checks [@Hullman2016]
+- evidence that people do not follow the axioms of probability when conveying subjective probabilities. How do you deal with this as an analyst (e.g. if the probabilities do not sum to 1 @Hullman2016
+
+### Cite Gap 17: Please Ignore uncertainty
+- @Blenkinsop2000 did try to include uncertainty as noise, asking participants to search for a specific outcome (the land classified as grass) in a random .
+However the participants failing to identify the signal (something that would be expected of an uncertainty plot) was seen as a sign the task was too complicated.
+While the task in this experiment *was* certainty far too complicated (and poorly communicated) this discussion indicates the authors did not consider the signal suppression to be the goal of the visualisation, and rather an annoying distraction.
+
+### Cite Gap 18: Trust stuff
+
+# Additions
+
+### Addition topic 1: Logical fallacy interpretation of mathematically unanswerable questions. 
+- If I tell you my cousins age, and then ask you to guess my age, it would not be unreasonable to guess the age I provided for my cousin.
+Citing this work to state "people incorrectly believe my cousins age is my age" would
+- Deterministic questions about a random event cannot be answered. 
+- @uncertchap2022 took the results of other paper to mean "when participants viewed mean temperature forecasts that included 95% confidence intervals depicted as bars with end caps, they incorrectly believed that the error bars represented high and low temperatures" (this one will depend on that study)
+- @Boone2018 the method of eliciting prior beleifs was, however questionable. Said people did or did not think something was likely but the question was weird.
+
+### Addition topic 2: Uncertainty vis communication considerations
+#### Unvertainty
+- @Spiegelhalter2017 it is better to understand your audience and iterate towards a final design, rather than only consider one option.
+- Some markers of probability that have common uses, probabilities such as 0 (impossible), 0.5 (a fair coin flip), 1 (certain) are easy for people to have an internal sense of [@Hullman2016]
 - It is incredibly hard to communicate small risks, because people cannot differentiate between 1/100 an 1/1000, and communicating low probability but high impact risks are particularly challenging [@Spiegelhalter2017].
@@
 50% chance today or this week) needs to be specified [@Spiegelhalter2017].
-- @Munzner2009, in her nested model for visualisation, comments on how even something as low level as poor problem abstraction can cause down stream effects and become impossible to ignore. 
-- @Otsuka2023 highlights that the choices made in the ontological process of boiling real world events down into statistical objects can influence later stages of analysis due to what was considred signal and what was considered noise.
-- @Meng2021 stresses the importance of treating data as an input from another process with noise and variance, rather than as a neutral input.
-- @Griethe2006 clearly sees uncertainty as additional information or data, and therefore sees the uncertainty visualisation problem as a specific case of the high dimensional visualisation issues. 
-- @Griethe2006 mentions that there are detailled classifications of uncertainty but the literature and subsequent techniques are not ballanced, and instead concentrate on scalar values.
-In facing this lack of ballance she creates a new classification system for uncertainty visualisation.
-- @Griethe2006 commented that uncertainty appears at every step in the analyis pipeline, and commented on some the mathematical definitions that seek to quantify uncertianty, but did not explain how these mathematical definitions can be combined across levels.
-Unfortunately this quantification is needed since she mentioned that in order to visualise uncertainty, it needs to be measurable.
-- @Griethe2006 defined uncertainty as "the degree to which the lack of knowledge about the amount of error is responsible for hesitancy in accepting results and observations without caution". 
-- @Griethe2006 found that there there are two ways to visualise uncertianty, implicitly and explicitly.
-When a visualisaion is implicit, the uncertainty is considered an uninteresting feature of the data, and instead of mapping some quantification of uncertainty to a visual feature, data is filtered out if it does not meet a specified "certainty" threshold.
-These statements clearly points towards a signal/noise dichotomy with normal visualisation and uncertainty visualisation.
-- @Spiegelhalter2017 even if most audience members can read a plot, they may not read it in the same way.
-People with high numeracy skills count pictographs, but people with low numeracy assess area, and seemingly unrelated elements may not equally influence all these methods of reading, for example scattering in a pictograph makes counting particularly more difficult.
-*also add that perceptual task audience thing*
-- @Spiegelhalter2017 it is better to understand your audience and iterate towards a final design, rather than only consider one option.
-- @Spiegelhalter2017 common sentiments such as avoiding chart junk, considering a table in place of a visualsiaiton, use of narrative labels, managing emotion, less is more approach, etc, are the main concepts expressed.
-- @Spiegelhalter2017 suggests that the graphic should allow "part-to-whole" comparison on an appropriate scale, which seems to be an extension of the framing concept discussed when speaking about uncertainty communication more broadly.
-- @Hullman2016 commented on the lack of justification of particular questions used in experiments and the lack of generalisability in using plot ranking as a response technique.
-- @Hullman2016 discussed the lack of clarity in exactly what "uncertainty" authors are trying to ask about.
-Whether the participants are supposed to answer according to some ground truth or their subjective probability is unclear, as is whether the *uncertainty* depicted represents the uncertainty of a sampling or population distribution.
-- @Hullman2016 discussed the role of heuristics in the analysis of results and suggested authors look for signs of them in responses and ask subjects to describe their strategy in their response. 
-- @Hullman2016 What was particularly interesting was people's tendancy to ignore uncertainty in a plot all together, and use a pattern in the signal (such as distance between means) to answer questions that require uncertainty information.
-- @Kinkeldey2014 stress the importance of selecting the correct audience, as numerical and visual literacy can differ vaslty between subpopulations so the results from one populations may not reflect the results from another.
-- @Daradkeh2015 found that people tended to to make decisions on a small number of salient pieces of information rather than on systematic evaluations of the costs and benefits of choices.
-- @Daradkeh2015 found that people tended to to make decisions on a small number of salient pieces of information rather than on systematic evaluations of the costs and benefits of choices.
-@Daradkeh2015 found that people tended to ignore or give little weight probability information and instead focused on the severity of undersierable outcomes.
-@Daradkeh2015 found that people tended to ignore or give little weight probability information and instead focused on the severity of undersierable outcomes.
-- @wu2023rational attempted to remedy this issue caused by designing a "rational agent benchmark" such that participants in an experiment are compared to a rational agent.
-This suggestion is an insufficient band-aid solution that is built upon the same invalid assumptions as the rest of the literature, and simply digs the field deeper into these problems, rather than solve them.
-- @wu2023rational almost seems to recognise the problem of decision making experiments with payment schemes or correct/incorrect answers, but instead of understanding this boils these tasks down to value extraction, they suggest participants should be compared to a "rational agent".
-This rational agent acts as a benchmark, in the same way a ground truth statistic does for value extraction experiments, except the authors don't seem to realise that is what they have done.
-This paper highlights many decision making tasks are simply a value extraction experiment without realising it, this paper included.
-- make sure quantified vs discussed uncertainty is mentioned
 - industrial accidents have a "tollerable" risk chart based on risk and number of fatalities @Spiegelhalter2017
-- @Suh2023 found that the main issue with deployment of AI was communication between data science and subject matter experts, since subject matter experts didn't want technical presentations that didn't explain exactly when and how a model would fail.
 - People believe in the law of small numbers but dont believe in the lar of large numbers @Hullman2016
@@
 the graphical one reduces noise but asking people for an interval estimation leads to overprecision and overconfidence @Hullman2016
-- believed that offering enough decisions with varying rewards you can observe the true subjective probability, however whether or not probability is the right representation of subjective uncertainty is a topic of debate @Hullman2016
-- she also discusses graphical interface (users have to draw the probability distribution) and framing probabilities as natural frequencies are better choices because they reduce noise and stuff.@Hullman2016
-- Consider incorperating utility functions. Include "probability-coherence" checks @Hullman2016
-- evidence that people do not follow the axioms of probability when conveying subjective probabilities. How do you deal with this as an analyst (e.g. if the probabilities do not sum to 1 @Hullman2016
-- we consider a summary table as a visualisation, illuminate graphics with words and numbers, design graphics to allow a part-to-whole comparison on an appropriate scale, use images and metaphors to gain/retain attention but dont arouse undue emotion, assume low numeracy of a general public audience and adopt a less-is-more approach by reducing the need for inference and making clear and explicit comparison, be cautious about interactivity and animations (may introduce unnecessary complexity), avoid chart junk (like 3d bar charts) @Spiegelhalter2017
-- @Maceachren2012 found to be the worst aesthetics to map to uncertainty to as they don't have an intuitive interpretation.
-- @Bella2005 found that most participants were ignorant to the fact that error bars are used for both confidence intervals and standard error bars, two wildly different indicators of precision.
-(**also the paper sherry sent me**)
-- Several studies probide evidence that the usability of uncertainty representations can be highly user and task dependent @Kinkeldey2014
-- The tasks for participants were most commonly value retrival (some where uncertainty and data value are retrived separately, others where they are some separately).
-also also be asked to extract a relative value @Kinkeldey2014
+#### Visualisation
+- @Spiegelhalter2017 even if most audience members can read a plot, they may not read it in the same way.
+People with high numeracy skills count pictographs, but people with low numeracy assess area, and seemingly unrelated elements may not equally influence all these methods of reading, for example scattering in a pictograph makes counting particularly more difficult.
+*also add that perceptual task audience thing*
+- @Spiegelhalter2017 common sentiments such as avoiding chart junk, considering a table in place of a visualsiaiton, use of narrative labels, managing emotion, less is more approach, etc, are the main concepts expressed.
+- @Spiegelhalter2017 suggests that the graphic should allow "part-to-whole" comparison on an appropriate scale, which seems to be an extension of the framing concept discussed when speaking about uncertainty communication more broadly.
+- @Hullman2016 discussed the role of heuristics in the analysis of results
+- @Kinkeldey2014 stress the importance of selecting the correct audience, as numerical and visual literacy can differ vaslty between subpopulations so the results from one populations may not reflect the results from another.
+- we consider a summary table as a visualisation, illuminate graphics with words and numbers, design graphics to allow a part-to-whole comparison on an appropriate scale, use images and metaphors to gain/retain attention but don't arouse undue emotion, assume low numeracy of a general public audience and adopt a less-is-more approach by reducing the need for inference and making clear and explicit comparison, be cautious about interactivity and animations (may introduce unnecessary complexity), avoid chart junk (like 3d bar charts) @Spiegelhalter2017
+- @Maceachren2012 found __????_ to be the worst aesthetics to map to uncertainty to as they don't have an intuitive interpretation.
+
+### Addition topic 3: Power of uncert vis (why it failled)
+- @Suh2023 found that the main issue with deployment of AI was communication between data science and subject matter experts, since subject matter experts didn't want technical presentations that didn't explain exactly when and how a model would fail.
+
+### Addition topic 5: Participants getting confused by overcomplicated tasks
+- @Gschwandtnei2016 When answering “What is the statistical probability (in %) that the interval has already ended at the marked point in time (red line)?” many answers gave the estimated probability to which the marked point in time falls into the interval.
+This is correct when asking for the probability to which the interval has already started at the marked position.
+However, in case of the probability to which the interval has already ended at the marked position, this answer is the inverted probability of what we were looking for.
+They believe this is their own mistake so they so they don’t present these results.
+- @Ibrekk1987 highlights that in the face of a vague question, participants will use the plot to decide what the authors mean.
+
+### Addition topic 5: Context stuff
 - most studies used coincident approaches, static visualisations, are applied to a specific domain and dont have generalised results @Kinkeldey2014
-- Most studies involve intrinsic approaches such as colour, transparency, and those that used extrinsic techniques typically used glyphs, error bars, grid based techniques, or contouring @Kinkeldey2014
-- Mentioned alleatory vs epistemic uncertainty, in the visualisation literature alleatory uncertainty is more commonly the focus.
-also mentions dependence and joint pdfs as a consideration @Hullman2016
-- When asking questions, researchers should focus on events that can be repeated (because of a misunderstanding of confidence intervals) @Hullman2016 (moved here because it is clear she is touching on the whole "what are we uncertain about" thing)
-- we should aim to show enough information to solve a task while avoiding irrelevant distracting information [@kosslyn2006graph]. 
+
+# Moved over notes
 - people will discount information they perceive as discountable if they are able.
@@
 - Hurricane risk is not only the path, but also the storm surge and wind speed. @Spiegelhalter2017
-
-# Experiment paper checklist
-General
-- Specific Field of Research
-- Question asked
-- Risk or Uncertainty: 
-- Noise vs signal (risk vs uncertainty)
-- Major task goal
-    - e.g. identify a bias or misunderstanding, compare two visualisation methods
-- Minor task goal
-    - Communication(value retrieval)
-    - Analytical(complex consideration)
-    - Exploration(free for all)
-
-Uncertainty Considerations
-	- Source
-	- Level (statistical-ignorance)
-	- Nature (Epistemic/Aleatory)
-
-Visual Uncertainty taxonomy
-- Explicit/implicit
-    - (directly mapping or showing multiple outcomes)
-- Intrinsic/Extrinsic 
-    - (using existing symbols e.g. colour value, or new objects e.g. grids)
-- Visually integral/separable 
-    - (can or cannot be separated from the data) 
-- Coincidence/adjacent 
-    - (if data and uncertainty are in integrated or separate views
-- Static/dynamic 
-    - (animation/interaction)
-
-Other plot considerations
-	- Dimension of data
-	- Dimension of uncertainty
-	- Feature mappings
-
-Extra (not necessarily recorded)
-	- Metrics used (and recorded)
-	- Possible Heuristics
-	- What is the ground truth?
-	- Participant literacy
 
@@
 Therefore we should not only keep the hierarchy of information in mind when we map features of our distribution, but also take advantage of these intuitive mappings when we can.
+
+# Experiment paper checklist
+General
+- Specific Field of Research
+- Question asked
+- Risk or Uncertainty: 
+- Noise vs signal (risk vs uncertainty)
+- Major task goal
+    - e.g. identify a bias or misunderstanding, compare two visualisation methods
+- Minor task goal
+    - Communication(value retrieval)
+    - Analytical(complex consideration)
+    - Exploration(free for all)
+
+Uncertainty Considerations
+	- Source
+	- Level (statistical-ignorance)
+	- Nature (Epistemic/Aleatory)
+
+Visual Uncertainty taxonomy
+- Explicit/implicit
+    - (directly mapping or showing multiple outcomes)
+- Intrinsic/Extrinsic 
+    - (using existing symbols e.g. colour value, or new objects e.g. grids)
+- Visually integral/separable 
+    - (can or cannot be separated from the data) 
+- Coincidence/adjacent 
+    - (if data and uncertainty are in integrated or separate views
+- Static/dynamic 
+    - (animation/interaction)
+
+Other plot considerations
+	- Dimension of data
+	- Dimension of uncertainty
+	- Feature mappings
+
+Extra (not necessarily recorded)
+	- Metrics used (and recorded)
+	- Possible Heuristics
+	- What is the ground truth?
+	- Participant literacy

### 2024-03-26 writingdumpster.qmd
moved notes around

@@
 ---
-# Why this exists
-I have an information hoarding problem, this helps me throw out sections of the literature review that are not necessarily relevant without actually throwing it out.
-Knowing that I can come back to this at any point and pull information back, makes me more comfortable with removing it from the report.
+# Introduction
+### Motivation for uncert vis
+- @Suh2023 found that the main issue with deployment of AI was communication between data science and subject matter experts, since subject matter experts didn't want technical presentations that didn't explain exactly when and how a model would fail.
+- (maybe include some of the noise vs signal stuff)
 
-# Citation Gaps
-
-### Cite Gap 3: Uncertainty definition + signal supression
-> Some consider it to be synonymous with specific terms defined in mathematics, such as probability [*HERE*], variance[*HERE*], error[*HERE*], or precision[*HERE*].
-Others consider it to be an encompassing umbrella term of which for these mathematically defined objects are only examples[*HERE*], and include concepts that are somewhat related, such as missing values[*HERE*]
-> This wide array of uncertainty definitions becomes a snake eating its own tail, as many authors opt to redefine uncertainty themselves[*HERE*], ignore defining uncertainty entirely [*HERE*], or pick a definition from a seemingly randomly previously published papers [*HERE*], all of which make variance in uncertainty definitions even worse
-
+# Definition Notes
+## General
+- @uncertchap2022 People do not over react when an ensemble member barely misses their town and People overreact more when 1/9 enselbles appear to be hitting their town than when 1/33 ensembles.
+- @uncertchap2022 Correll and Gleicher [9] found that violin plots (mapping probability to area or width at a given y position) and gradient plots (mapping probability to opacity) lead to more intuitive assessments of value likelihood and “ability to surprise,” which are more closely aligned with statistical defini- tions of uncertainty." wtf is the statistical definition of uncertainty
 - @Thomson2005 titled their paper "A typology for visualizing uncertainty" however it is a taxonomy for the specific types of uncertainty that we may want to visualise and a way to calculate that uncertainty.
@@
 Unfortunately this quantification is needed since she mentioned that in order to visualise uncertainty, it needs to be measurable.
-- @Griethe2006 defined uncertainty as "the degree to which the lack of knowledge about the amount of error is responsible for hesitancy in accepting results and observations without caution".
+- mentions uncertainty can be variance, precision, accuracy, reliability, or related concepts @Hullman2016
+- @uncertchap2022 "can enter the analysis pipeline during the mea- surement, modeling, and forecasting phases"
 
-### Cite Gap 4: Organise uncertainty quantification
+## Probability/variance/error etc (Cite Gap 3)
+> Some consider it to be synonymous with specific terms defined in mathematics, such as probability [*HERE*], variance[*HERE*], error[*HERE*], or precision[*HERE*].
+Others consider it to be an encompassing umbrella term of which for these mathematically defined objects are only examples[*HERE*], and include concepts that are somewhat related, such as missing values[*HERE*]
+
+## Redefined uncertainty (cite gap 3)
+> This wide array of uncertainty definitions becomes a snake eating its own tail, as many authors opt to redefine uncertainty themselves[*HERE*], ignore defining uncertainty entirely [*HERE*], or pick a definition from a seemingly randomly previously published papers [*HERE*], all of which make variance in uncertainty definitions even worse
+
+#### Uncertainty Taxonomies (Cite Gap 7)
+> Taxonomies split uncertainty based on an endless stream of arbitrary lines in the sand, such as whether the uncertainty is due to true randomness or a lack of knowledge [@Spiegelhalter2017 *HERE*], if the uncertainty is in the the attribute, spatial elements, or temporal element of the data [*HERE*], whether the uncertainty is scientific (e.g.
+error) or human (e.g.
+disagreement among parties) [*HERE*], if the uncertainty is random or systematic [*HERE*], statistical or bounded [], accuracy or precision [*HERE*], if the uncertainty is about the past or the future [*HERE*], the stage of the data analysis pipeline it comes from [*HERE*], the severity of the uncertainty [*HERE*], how quantifiable the uncertainty is [@Spiegelhalter2017], and many, many, many more examples.
+
+- @uncertchap2022 mention aleatric, epistemic, ontological divide and also differentiates risk with uncertainty
+- Mentioned alleatory vs epistemic uncertainty, in the visualisation literature alleatory uncertainty is more commonly the focus.
+also mentions dependence and joint pdfs as a consideration @Hullman2016
+
+# Visual Uncertainty Definition
+
+### Context relevance
+- most studies used coincident approaches, static visualisations, are applied to a specific domain and dont have generalised results @Kinkeldey2014
+
+### How was uncertainty quantified (Cite Gap 4)
 > The uncertainty visualisation literature is completely awash with papers that define uncertainty using a vague encompassing phrase but go on to *quantify* uncertainty as a PDF, error, or some other easily quantified mathematical object when it comes time to do the visualisation.
@@
 
-### Cite Gap 6: All PDF no Missing values
+- @uncertchap2022 do specify that they are going to refer to quantified uncertainty that can be visualised ("most commonly as a probability distribution")
+- @Griethe2006 mentions that there are detailled classifications of uncertainty but the literature and subsequent techniques are not ballanced, and instead concentrate on scalar values.
+In facing this lack of ballance she creates a new classification system for uncertainty visualisation.
+
+### We need cool uncertainty visuailsations (Cite Gap 6)
 > This has left the expressions of uncertainty that are hard to quantify or visually differentiate untouched, despite many papers calling for their invention [*HERE*]
 
-- @Griethe2006 mentions that there are detailled classifications of uncertainty but the literature and subsequent techniques are not ballanced, and instead concentrate on scalar values.
-In facing this lack of ballance she creates a new classification system for uncertainty visualisation.
-- Mentioned alleatory vs epistemic uncertainty, in the visualisation literature alleatory uncertainty is more commonly the focus.
-also mentions dependence and joint pdfs as a consideration @Hullman2016
+- Whether to use classed or unclassed uncertainty is rarely discussed @Kinkeldey2014.
+e.g.
+land map with remotely sensed images has ambiguity in the boundaries, which can result from multiple sources (land cover class, measurement error, images from different dates).
+If you combine them all it might not make sense for someone who only needs one.
+- Most studies involve intrinsic approaches such as colour, transparency, and those that used extrinsic techniques typically used glyphs, error bars, grid based techniques, or contouring @Kinkeldey2014
+- rarely used extension is aggregation of uncertainty over an area (retrieve overall estimation from a spatial distribution of uncertainty).
+Other tasks include comparisons, rankings.
+also have a task called "search" where participants have to identify entities that fulfill certainty characteristics (high or low values) @Kinkeldey2014
 
-### Cite Gap 7: Uncertainty Taxonomies
-> Taxonomies split uncertainty based on an endless stream of arbitrary lines in the sand, such as whether the uncertainty is due to true randomness or a lack of knowledge [@Spiegelhalter2017 *HERE*], if the uncertainty is in the the attribute, spatial elements, or temporal element of the data [*HERE*], whether the uncertainty is scientific (e.g.
-error) or human (e.g.
-disagreement among parties) [*HERE*], if the uncertainty is random or systematic [*HERE*], statistical or bounded [], accuracy or precision [*HERE*], if the uncertainty is about the past or the future [*HERE*], the stage of the data analysis pipeline it comes from [*HERE*], the severity of the uncertainty [*HERE*], how quantifiable the uncertainty is [@Spiegelhalter2017], and many, many, many more examples.
-
-### Cite Gap 8: Bar/pie uncertainty graphics
+### Bar/pie uncertainty graphics (Cite Gap 8)
 > For example @Leland2005 mentions that popular graphics, such as pie charts and bar charts omit uncertainty, however at least one or both of these charts are used in most "uncertainty visualisation" experiments [@Ibrekk1987 *MORE HERE*]
 
-### Cite Gap 9: Vis uncertainty Taxonomies
-- Most studies involve intrinsic approaches such as colour, transparency, and those that used extrinsic techniques typically used glyphs, error bars, grid based techniques, or contouring @Kinkeldey2014
+### Vis uncertainty Taxonomies (Cite Gap 9)
+- @uncertchap2022 two broad categories of uncertainty visualisation techniques, 1) graphical annotation (mean, error bar, hops) 2) map probability of cI to a visual encoding channel 3) hybrid approaches (e.g.
+contour boxplot, probability density and interval plot)
 
-### Cite Gap 10: High dimensional vs psychological
+### High dimensional vs psychological motivation (Cite Gap 10)
 > Papers will often discuss uncertainty in relation to one of these motivating reasons, the evaluation experiments motivated by (1) often perform different visualisations of PDFs, while the papers motivated by (2) will focus on trying to impute uncertainty as error within one of the existing channels, such as colour
@@
 
-### Cite Gap 12 & 13 & 14: Information & Mathematical stuff
+# Noise vs Signal
+- @uncertchap2022 "Effectively communicating uncertainty is necessary for establishing scientific transparency"
+- @Griethe2006 defined uncertainty as "the degree to which the lack of knowledge about the amount of error is responsible for hesitancy in accepting results and observations without caution".
+
+# Experiment stuff
+### Information & Mathematical stuff (Cite Gap 12 & 13 & 14)
 - we should aim to show enough information to solve a task while avoiding irrelevant distracting information [@kosslyn2006graph]. 
@@
 - When asking questions, researchers should focus on events that can be repeated (because of a misunderstanding of confidence intervals) @Hullman2016 (moved here because it is clear she is touching on the whole "what are we uncertain about" thing)
+- @Nathonours found that a scatter plot is better than a line plot if you want to convey the correlation between two time series but we cannot be sure if this was influenced by swapping the distribution depicted or by dropping the irrelevant feature of inexchangeability.
 
-### Cite Gap 15: Value Extraction (Examples of replicated perceptual task experiments)
+### Logical fallacy interpretation
+- If I tell you my cousins age, and then ask you to guess my age, it would not be unreasonable to guess the age I provided for my cousin.
+Citing this work to state "people incorrectly believe my cousins age is my age" would
+- Deterministic questions about a random event cannot be answered. 
+- @uncertchap2022 took the results of other paper to mean "when participants viewed mean temperature forecasts that included 95% confidence intervals depicted as bars with end caps, they incorrectly believed that the error bars represented high and low temperatures" (this one will depend on that study)
+- @Boone2018 the method of eliciting prior beleifs was, however questionable. Said people did or did not think something was likely but the question was weird.
+- @uncertchap2022 Sources 3, 54, 55, 56 also used to suggest "people interpret uncertainty information as deterministic information".
+Exists with visualisation and not with text.
+Suggests this is because of a "visual-spatial bias".
+suggest HOPs to aleviate this and I refuse to entertain the notion.
+
+### Participants getting confused by overcomplicated tasks
+- @Gschwandtnei2016 When answering “What is the statistical probability (in %) that the interval has already ended at the marked point in time (red line)?” many answers gave the estimated probability to which the marked point in time falls into the interval.
+This is correct when asking for the probability to which the interval has already started at the marked position.
+However, in case of the probability to which the interval has already ended at the marked position, this answer is the inverted probability of what we were looking for.
+They believe this is their own mistake so they so they don’t present these results.
+- @Ibrekk1987 highlights that in the face of a vague question, participants will use the plot to decide what the authors mean.
+- @uncertchap2022 visual boundaries = cognative categories.
+Use the cone of uncertainty work as a justification of this.
+I don't even think THOSE papers found evidence of this.
+Comment that "Viewers understandably assume the value of a boundary is meaningful, particularly when the information about how the visualization was generated is insufficient" and "Viewers have no choice but to assume that the scientists who made the forecast are indicating an important distinction with the boundary".
+Shows they know providing irrelevant information is stupid, specify that the issue is that there is rarely a justification for the boundary.
+YES.
+
+### Value Extraction (Cite Gap 15)
+- @uncertchap2022 Attribute substitution: people will substitute a hard mental calculation for an easy one (attribute substitution heuristic).
+Name "deterministic constructual error" when people try to substitute uncertainty information for deterministic information (e.g.
+error bars for high and low temperature even when the visualisation was explained - from icing study.
+This makes sense because temperature is GIVEN as high and lows so it is a natural interpretation).
+I think this result is in line with the idea that people give the best estimate based on what the visualisation suggests.
+i.e.
+people might have assumed you meant "what is the lowest possible temperature".
+There is a level of "asking a deterministic question for a probabilistic question".
+- @uncertchap2022 Quantile dot plot created so people dont have to visually calculate the area under the curve (so authors are aware of the mental challenge of some of these things) Decrease variance with respect to density plots (it makes sense because you can directly count and quantify raher than guesstimate) (both come from the same source, when-ish is my bus)
+- Authors provide little justification for their chosen response models (e.g. absolute accuracy vs relative measures) @Hullman2016 
 - @Correll2018 “place your 5 ships on the safest locations on the board.” (minimise danger) It is a convoluted value extraction task, where participants basicaly have to look at the palette and see that if it is in the lower uncertainty, and lower error it is better, but higher uncertainty is better
@@
 
-### Cite Gap 16: Decision making stuff 
+### Decision making stuff (Cite Gap 16)
 - @Boone2018 organised responses from participants by the approach they used
@@
 - evidence that people do not follow the axioms of probability when conveying subjective probabilities. How do you deal with this as an analyst (e.g. if the probabilities do not sum to 1 @Hullman2016
+- people will discount information they perceive as discountable if they are able.
+Including mean estimates on depictions of mass can cause people to discount the uncertainty information and use the difference between means as a proxy for the probability distribution [@Kale2021].
+- Some questions (e.g. what are the chances that the no.6 bus will arrive first) you can elicit the viewers subjective probability distribution @Hullman2016
 
-### Cite Gap 17: Please Ignore uncertainty
+### Please Ignore uncertainty (Cite Gap 17)
 - @Blenkinsop2000 did try to include uncertainty as noise, asking participants to search for a specific outcome (the land classified as grass) in a random .
@@
 
-### Cite Gap 18: Trust stuff
+### Trust stuff (Cite Gap 18)
+- @uncertchap2022  mentions "other studies have found that people trust icon arrays more than other common visualization techniques" without commenting on whether or not that trust is good?
 
-# Additions
-
-### Addition topic 1: Logical fallacy interpretation of mathematically unanswerable questions. 
-- If I tell you my cousins age, and then ask you to guess my age, it would not be unreasonable to guess the age I provided for my cousin.
-Citing this work to state "people incorrectly believe my cousins age is my age" would
-- Deterministic questions about a random event cannot be answered. 
-- @uncertchap2022 took the results of other paper to mean "when participants viewed mean temperature forecasts that included 95% confidence intervals depicted as bars with end caps, they incorrectly believed that the error bars represented high and low temperatures" (this one will depend on that study)
-- @Boone2018 the method of eliciting prior beleifs was, however questionable. Said people did or did not think something was likely but the question was weird.
-
-### Addition topic 2: Uncertainty vis communication considerations
-#### Unvertainty
+# Communication considerations
+## Uncertainty
 - @Spiegelhalter2017 it is better to understand your audience and iterate towards a final design, rather than only consider one option.
@@
 the graphical one reduces noise but asking people for an interval estimation leads to overprecision and overconfidence @Hullman2016
-#### Visualisation
+- The big data paradox shrinks confidence intervals but magnifies bias [@Bradley2021] because the confidence intervals shrink due to the sample size but the imprecise target group creates bias.
+
+## Visualisation
+- @uncertchap2022 Icon array and quantile dot plots both support area and frequency based judgements.
+- @uncertchap2022 frequency framing: (e.g. 1 out of 10 rather than 10%) better (plot: icon array, quantile dot plot). 
 - @Spiegelhalter2017 even if most audience members can read a plot, they may not read it in the same way.
@@
 - @Maceachren2012 found __????_ to be the worst aesthetics to map to uncertainty to as they don't have an intuitive interpretation.
-
-### Addition topic 3: Power of uncert vis (why it failled)
-- @Suh2023 found that the main issue with deployment of AI was communication between data science and subject matter experts, since subject matter experts didn't want technical presentations that didn't explain exactly when and how a model would fail.
-
-### Addition topic 5: Participants getting confused by overcomplicated tasks
-- @Gschwandtnei2016 When answering “What is the statistical probability (in %) that the interval has already ended at the marked point in time (red line)?” many answers gave the estimated probability to which the marked point in time falls into the interval.
-This is correct when asking for the probability to which the interval has already started at the marked position.
-However, in case of the probability to which the interval has already ended at the marked position, this answer is the inverted probability of what we were looking for.
-They believe this is their own mistake so they so they don’t present these results.
-- @Ibrekk1987 highlights that in the face of a vague question, participants will use the plot to decide what the authors mean.
-
-### Addition topic 5: Context stuff
-- most studies used coincident approaches, static visualisations, are applied to a specific domain and dont have generalised results @Kinkeldey2014
-
-# Moved over notes
-- people will discount information they perceive as discountable if they are able.
-Including mean estimates on depictions of mass can cause people to discount the uncertainty information and use the difference between means as a proxy for the probability distribution [@Kale2021].
-- rarely used extension is aggregation of uncertainty over an area (retrieve overall estimation from a spatial distribution of uncertainty).
-Other tasks include comparisons, rankings.
-also have a task called "search" where participants have to identify entities that fulfill certainty characteristics (high or low values) @Kinkeldey2014
-- Whether to use classed or unclassed uncertainty is rarely discussed @Kinkeldey2014.
-e.g.
-land map with remotely sensed images has ambiguity in the boundaries, which can result from multiple sources (land cover class, measurement error, images from different dates).
-If you combine them all it might not make sense for someone who only needs one.
-- The way we ask questions and the types of questions we ask are selected with little justification. This paper makes suggestions to reduce the noise in the data from these papers. @Hullman2016
 - Error is best mapped to fuzziness, location, and colour value; arrangement, size and transparency are an OK second choice; but saturation, hue, orientation and shape are unacceptable and have no intuitive connection to variance [@Maceachren2012].
-- @Nathonours found that a scatter plot is better than a line plot if you want to convey the correlation between two time series but we cannot be sure if this was influenced by swapping the distribution depicted or by dropping the irrelevant feature of inexchangeability.
 - No only do the graphical elements we map our features to matter, but the direction matters too.
@@
 - This idea also extends to interval estimation, where questions about probability are best answered with gradients and questions about start and end times are easiest to answer with ambiguation [@Gschwandtnei2016].
-
-- mentions uncertainty can be variance, precision, accuracy, reliability, or related concepts @Hullman2016
--  @Spiegelhalter2017 defines risk as risk known probabilities and uncertainty as estimated probabilities, therefore uncertainty is what we have when you cannot accurately quantify risk.
-- Authors provide little justification for their chosen response models (e.g. absolute accuracy vs relative measures) @Hullman2016 
-- Some questions (e.g. what are the chances that the no.6 bus will arrive first) you can elicit the viewers subjective probability distribution @Hullman2016
-- The big data paradox shrinks confidence intervals but magnifies bias [@Bradley2021] because the confidence intervals shrink due to the sample size but the imprecise target group creates bias.
 - coloured maps for earthquake risk are easily interpretable @Spiegelhalter2017
-- Hurricane risk is not only the path, but also the storm surge and wind speed. @Spiegelhalter2017
 

### 2024-03-31 paper.qmd
sorted many notes and moved hurricane comment over
revised Intro

@@
 
+<!-- TODO 
+- convert to American spelling
+- should it be data-driven or two words
+- May not need to reference the same paper again and again
+-->
+
 # Background
 
-## Introduction
+<!-- ## Introduction-->
 From entertainment choices to news articles to insurance plans, the modern citizen is so over run with information in every aspect of their life it can be overwhelming.
@@
 
-There are many stages in our analysis that benefit from the power of data visualisation, however this does not mean it is always done with success.
-Visualization is an important step in exploratory data analysis and it is often utilised to **learn** what is important about a data set.
-The importance of data driven discovery is highlighted by data sets such as Anscombe's quartet [@anscombe] or the Datasaurus Dozen [@datasaurpkg].
-Each of the pairwise plots in these data sets have the same summary statistics but strikingly different information when visualised.
-Anscombe quartet is shown in @fig-anscombe, because describing the data is never the same as seeing it.
-Instead of having to repeatedly check endless hypothesis to find interesting numerical features, visualisations **tell** us what is important about our data and **how** it might diverge from what we expect.
+There are many stages in our analysis that might benefit from the power of data visualisation, however this does not mean it is appropriately utilized or done with with effectiveness.
+(@Vanderplas2020 has helpful advice here.) Visualization is an important step in exploratory data analysis and it is often utilised to **learn** what is important about a data set.
+The importance of data driven discovery has been highlighted through the data sets Anscombe's quartet [@anscombe] and the Datasaurus Dozen [@datasaurpkg].
+Each of the pairwise plots in these data sets have the same summary statistics but strikingly different information, as revealed when visualised.
+<!-- Anscombe quartet is shown in @fig-anscombe, because describing the data is never the same as seeing it.
+Instead of having to repeatedly check endless hypothesis to find interesting numerical features: DON'T NEED MUCH EMPHASIS ON THIS DATA - VERY WELL-KNOWN -->
+Visualisations can **tell** us what is important about our data and **how** it might diverge from pre-conceived notions or what we expect.
 
@@
 #| message: false
+#| eval: false
 #| label: fig-anscombe
@@
 
-Uncertainty visualisation is a somewhat new subfield of visualisation.
-Early papers that specifically reference "uncertainty visualisation" start popping up in the late 80s [@Ibrekk1987], with papers in the early 90s (specifically in the area of geospatial information visualisation) recgonising it as a necessary subfield of visualisation [@MacEachren1992; Carr1992].
-Despite the youth of the specific sub-field, commonly used uncertainty visualisations such as a box plot or histogram have been around for almost as long as statistical graphics themselves.
-In the modern day, uncertainty visualisation has exploded into its own field that applies to a wide range of contexts.
+Uncertainty visualisation is a relatively new field in research.
+Early papers that specifically reference "uncertainty visualisation" appear in the late 80s [@Ibrekk1987], with geospatial information visualisation literature in the early 90s declaring this to be essential aspect of information display (@MacEachren1992; @Carr1992).
+Despite the new terminology visualisation of uncertainty has been present since the earliest times.
+For example, box plots or histograms can be considered to be displaying uncertainty in the sense of variability in observations sampled from a population distribution.
+<!--In the modern day, uncertainty visualisation has exploded into its own field that applies to a wide range of contexts.-->
 
-## Similar work
+<!-- ## Similar work-->
 
-Any field that experiences an explosion of new research will also invite a series of literature reviews that seek to combine and summaries that information.
+<!-- Any field that experiences an explosion of new research will also invite a series of literature reviews that seek to combine and summaries that information.
 Given that uncertainty visualisation has been around for a few decades now, there is already a wealth of uncertainty visualisation literature reviews.
+-->
 
-Interestingly, these reviews often do not offer overarching rules for tried and tested uncertainty visualisation, but rather comment on the *difficulties* faced when trying combine the papers from this field.
-@Kinkeldey2014 found most experiments the methods for uncertainty visualisation evaluation to be ad hoc, with no commonly agreed upon methodology or formalisation and no greater goal of describing general principals.
-@Hullman2016 commented on the difficulty in taking overarching themes from uncertainty visualisation, as several conflated issues make it unclear if subjects did poorly in an experiment because they misunderstood a visualisation, because the question was misinterpreted, or because they used a heuristic.
+However, uncertainty visualisation takes many forms today, It can be considered to expand the application of data visualisation towards making better decisions, and research is concerned about how the construction of plots and information contained might affect the resulting decisions.
+With the abundance of publications it is timely to consider a review of the state-of-the-art.
+In fact, there have already been several reviews published.
+
+Interestingly, these reviews often do not offer overarching rules for tried and tested uncertainty visualisation, but rather comment on the *difficulties* faced when trying to summarise the papers from this field.
+@Kinkeldey2014 found most experiments on the methods for uncertainty visualisation evaluation to be ad hoc, with no commonly agreed upon methodology or formalisation and no greater goal of describing general principals.
+@Hullman2016 commented on the difficulty in taking overarching themes from uncertainty visualisation, as several conflated issues make it unclear if subjects did poorly in an experiment because they misunderstood a visualisation, because the question was misinterpreted, or because they used a specific heuristic.
 @Spiegelhalter2017 commented that different plots are good for different things, and disagreed with the goal of identifying a universal "best" plot for all people and circumstances.
-@Griethe2006 was unable to find common themes, but instead listed the findings and opinions of each uncertainty visualisation paper.
+@Griethe2006 was unable to find common themes, but instead listed the findings and opinions of <!--each uncertainty visualisation--> a collection of papers.
 
-There are a handful of reasons that are used to explain why it is so difficult to generalise the findings in the uncertainty visualisation literature.
-Some papers suggested or agree with the idea that visualisation typologies should move away from data types, uncertainty categories, and representation types and towards "task-centred typologies" as this will help generalise results [@Kinkeldey2014; @Hullman2016].
-Some authors may not attribute the task to the *reason* the literature cannot be easily summarised, but do mention that the choice of best visualisation is highly dependent on the specific goal of that visualisation [@Griethe2006; @Spiegelhalter2017].
-Others comment that visualisations are highly dependent on the audience and there is no such thing as a "best" visualisation that will be accessible to all members [@Kinkeldey2014; @Spiegelhalter2017].
-These difficulties in noisy uncertainty visualisation findings arises regardless of the direction the analysis was approached, be it looking at geospatial data [@Kinkeldey2014], or uncertainty visualisation as a whole [@Hullman2016; @Spiegelhalter2017; @Griethe2006].
+There are several reasons provided to explain the difficulty in generalising <!--the findings in--> the uncertainty visualisation literature.
+Some suggested or agree with the idea that visualisation typologies should move away from data types, uncertainty categories, and representation types and towards "task-centred typologies" [@Kinkeldey2014; @Hullman2016].
+Other papers indirectly hint at this <!--Some authors may not attribute the task to the *reason* the literature cannot be easily summarised, but do mention that--> by arguing that the choice of best visualisation is highly dependent on the specific goal [@Griethe2006; @Spiegelhalter2017].
+Others contend that visualisations are highly dependent on the audience and there is no such thing as a "best" visualisation that will be accessible to all members [@Kinkeldey2014; @Spiegelhalter2017].
+These concerns appear <!--difficulties in noisy uncertainty visualisation findings arises regardless of the direction the analysis was approached, be it looking--> across all application areas,  <!--, in geospatial data [@Kinkeldey2014], or uncertainty visualisation as a whole [@Hullman2016; @Spiegelhalter2017; @Griethe2006].--> and an underlying thread is that the problem arises from a lack of a cohesive and encompassing **definition of uncertainty**.
 
-This review believe all these sources of noise have a single root cause, a lack of a cohesive and encompassing definition of uncertainty.
-The absence of an encompassing definition of uncertainty is mentioned by every uncertainty visualisation directly [@Spiegelhalter2017; @Griethe2006] or indirectly by describing a myriad of ways it can be considered in the literature [@Kinkeldey2014; @Hullman2016], although it is never commented on as a source of the noise in the field.
+<!--This review believe all these sources of noise have a single root cause, a lack of a cohesive and encompassing definition of uncertainty.
+The absence of an encompassing definition of uncertainty is mentioned by every uncertainty visualisation directly [@Spiegelhalter2017; @Griethe2006] or indirectly by describing a myriad of ways it can be considered in the literature [@Kinkeldey2014; @Hullman2016], although it is never commented on as a source of the noise in the field.-->
 
-In understanding the definition problem, we will understand the *purpose* of visualising uncertainty, and understand if the current visualisation tools are sufficient to achieve that purpose and if not, where there is need for improvement.
-In understanding this purpose, we will also consider whether or not "uncertainty visualisation" is a valid sub-field of visualisation itself, or if that distinction has only arisen as a by product of the confused definition of uncertainty.
-Ultimately, this paper should serve as a guide for future uncertainty visualisation authors to prevent their work from becoming noise in what is already an incredibly noisy field.
+The review provided here attempts to address these issues.
+We begin with understanding the definition problem, to obtain and understanding of the *purpose* of visualising uncertainty, and understand if the current visualisation tools are sufficient to achieve that purpose and if not, where there is need for new methodology.
+We also posit that "uncertainty visualisation" is too diffuse to be considered a field in itself, but rather should be dissected into smaller components.
+The ideal role of this review is to provide a guide for research in these topics, to help synthesize this noisy field.
+<!-- XXX Its not clear that these last two sentences are helpful here.--> <!-- In understanding this purpose, we will also consider whether or not "uncertainty visualisation" is a valid sub-field of visualisation itself, or if that distinction has only arisen as a by product of the confused definition of uncertainty.
+Ultimately, this paper should serve as a guide for future uncertainty visualisation authors to prevent their work from becoming noise in what is already an incredibly noisy field.-->
 
@@
 Given these conflicting assumptions, it is unclear how the participants were supposed to adjust the probabilistic path information to answer a deterministic question about which oil rig would receive the most damage.
+Other authors have commented on the complexity of communicating hurricane risk because the path, storm surge and wind speed are all important and cannot be ignored [@Spiegelhalter2017].
 The flip side of this is asking participants for a deterministic answer to a probabilistic question.

### 2024-03-31 paper.qmd
small re-wording of last intro paragraph

@@
 We begin with understanding the definition problem, to obtain and understanding of the *purpose* of visualising uncertainty, and understand if the current visualisation tools are sufficient to achieve that purpose and if not, where there is need for new methodology.
-We also posit that "uncertainty visualisation" is too diffuse to be considered a field in itself, but rather should be dissected into smaller components.
-The ideal role of this review is to provide a guide for research in these topics, to help synthesize this noisy field.
-<!-- XXX Its not clear that these last two sentences are helpful here.--> <!-- In understanding this purpose, we will also consider whether or not "uncertainty visualisation" is a valid sub-field of visualisation itself, or if that distinction has only arisen as a by product of the confused definition of uncertainty.
+We also posit that "uncertainty visualisation" is too diffuse to be considered a field in itself, but rather should be dissected into smaller elements.
+The ideal role of this review to help synthesize this noisy field, by collecting together essential components, summarising different viewpoints the current literature, and provide a guide for potential research topics.
+<!-- In understanding this purpose, we will also consider whether or not "uncertainty visualisation" is a valid sub-field of visualisation itself, or if that distinction has only arisen as a by product of the confused definition of uncertainty.
 Ultimately, this paper should serve as a guide for future uncertainty visualisation authors to prevent their work from becoming noise in what is already an incredibly noisy field.-->

### 2024-03-31 paper.qmd
tidied section 2, and Future work - look for XXX on questions/comments

@@
 ---
-title: "Noisy Work: A Review of The Uncertainty Visualisation Literature"
+title: "The Noisy Work of Uncertainty Visualisation Research: A Review"
 author: Harriet Mason
@@
 # What is Uncertainty?
-## Colloquial definitions
-Uncertainty visualisation is made particularly difficult by the term "uncertainty" lacking a commonly accepted definition in the literature.
+<!--## Colloquial definitions - what's in this section is not colloquial definitions, it's more overview-->
+
+Uncertainty visualisation is directed towards communication - to the general public when disasters may hit, to decision makers in control centers, and a whole host of other public-facing tasks - needing an understanding of what it is when the term "uncertainty" is used.
+<!--is made particularly difficult by the term "uncertainty" lacking a commonly accepted definition in the literature.
 This mishmash of terminology leads to a large body of work, all claiming to finding the best visualisation or expression of of "uncertainty" but most don't even seem to agree on what uncertainty is.
-
-Definitions presented in papers that discuss uncertainty visualisation want it to be encompassing and include everything a layperson might consider when trying to consider "uncertainty".
-Some consider it to be synonymous with specific terms defined in mathematics, such as probability, variance, error, or precision.
-Others consider it to be an encompassing umbrella term of which for these mathematically defined objects are only examples, and include concepts that are somewhat related, such as missing values (*Cite Gap 3: Organise paper definitions*).
-This wide array of uncertainty definitions becomes a snake eating its own tail, as many authors opt to redefine uncertainty themselves[], ignore defining uncertainty entirely [], or pick a definition from a seemingly randomly previously published papers [], all of which make variance in uncertainty definitions even worse (*Cite Gap 3: Organise paper definitions*).
+-->
+Definitions in the literature appear seemingly view it to be encompassing and include everything the general public might consider to be uncertain.
+Some works (XXX eg) focus more narrowly on specific terms defined mathematically, such as probability, variance, error, or precision.
+Others (XXX eg)  <!--consider it to be an encompassing umbrella term of which for these mathematically defined objects are only examples, and-->  include broader loosely related elements, such as missing values (*Cite Gap 3: Organise paper definitions*).
+<!--This wide array of uncertainty definitions becomes a snake eating its own tail, as many authors opt to redefine uncertainty themselves[], ignore defining uncertainty entirely [], or pick a definition from a seemingly randomly previously published papers [], all of which make variance in uncertainty definitions even worse (*Cite Gap 3: Organise paper definitions*).
 
@@
 The current method of titling papers and defining uncertainty forces readers to comb through the method of a particular visualisation experiment to understand if the uncertainty visualised is amenable to their needs.
+-->
+Several attempts to address this gap are made in the papers describing taxonomies. 
+<!-- Some papers have recognised these colloquial definitions are lacking and have made attempts to formalise the space with taxonomies, however these taxonomies come with their own issues. -->
 
-Some papers have recognised these colloquial definitions are lacking and have made attempts to formalise the space with taxonomies, however these taxonomies come with their own issues.
-
-## Taxonomy definitions
-Much in the same way that there are almost as many definitions of uncertainty as there are papers on the topic, the field is also over run with taxonomies.
-Taxonomies split uncertainty based on an endless stream of ever changing boundaries, such as whether the uncertainty is due to true randomness or a lack of knowledge [@Spiegelhalter2017], if the uncertainty is in the the attribute, spatial elements, or temporal element of the data [], whether the uncertainty is scientific (e.g.
+## Definitions in taxonomies
+<!--Much in the same way that there are almost as many definitions of uncertainty as there are papers on the topic, the field is also over run with taxonomies.--> Taxonomies split uncertainty based on an endless stream of ever changing boundaries, such as whether the uncertainty is due to true randomness or a lack of knowledge [@Spiegelhalter2017], if the uncertainty is in the  attribute, spatial elements, or temporal element of the data [], whether the uncertainty is scientific (e.g.
 error) or human (e.g.
@@
 # Future work
-This paper has identified several issues in the uncertainty visualisation literature and with those issues, we have several suggestions for future work that would likely be fruitful for the field.
+This paper has identified gaps in the uncertainty visualisation literature that could be filled to progress the field.
+<!--and with those issues, we have several suggestions for future work that would likely be fruitful for the field.-->
 
-Our first suggestion for future work is a mathematical definition of uncertainty.
-There is currently a lot of work that quantifies the uncertainty through bias and variance at various stages of the data analysis pipeline separately.
-Elements such as imputed data, assumptions, sampling methods, and analyst decisions all have their own independent work, quantification and discussions, but methods to brand between these concepts are few and far between.
-A conceptual framework that allows us to combine these results would be incredibly beneficial to the field.
+<!--Our first suggestion for future work is--> *Each new development should be accompanied by a mathematical definition of the uncertainty being addressed.* Utilizing bias and variance to define uncertainty is well-established, but it is absent concerns about imputed data, assumptions, sampling methods, and even the human element of analyst decisions.
+<!--There is currently a lot of work that quantifies the uncertainty through bias and variance at various stages of the data analysis pipeline separately.
+Elements such as  imputed data, assumptions, sampling methods, and analyst decisions all have their own independent work, quantification and discussions, but methods to brand between these concepts are few and far between.--> A conceptual framework that allows us to combine these components is needed.
 
-The concept of uncertainty should also be formalised within the grammar of graphics.
-This formalisation would allow uncertainty visualisation authors to have a clear understanding of what is or is not an uncertainty visualisation.
+*The concept of uncertainty should be formalised within the grammar of graphics.* This formalisation would allow uncertainty visualisation authors to have a clear understanding of what is or is not an uncertainty visualisation.
 Additionally placing uncertainty visualisation in the framework that is used to understand existing information visualisation research would help authors understand when existing methods can be used to explain their results.
@@
 
-Additionally the experimental practices within uncertainty visualisation need to be less ad hoc and more standardised.
-If we are going to consider uncertainty as noise, not signal, there needs to be a way to identify this signal suppression in an experimental design.
+XXX Is data pipeline connected with the grammar of graphics? Should this be a recommendation?
+
+*Experimental practices on uncertainty visualisation need to be standardised.* If we are going to consider uncertainty as noise, not signal, there needs to be a way to identify this signal suppression in an experimental design.
 As the literature currently exists, there is no way to combine papers to get a meaningful sense of how uncertainty information is understood by a viewer.
@@
 Regardless, these are directions of research would be fruitful to the uncertainty visualisation community even if it appears on the surface to be research that is only beneficial to the "normal" visualisation community.
+XXX Not sure what this paragraph is recommending?
 

### 2024-04-01 paper.qmd
fixed a small citation error

@@
 Uncertainty visualisation is a relatively new field in research.
-Early papers that specifically reference "uncertainty visualisation" appear in the late 80s [@Ibrekk1987], with geospatial information visualisation literature in the early 90s declaring this to be essential aspect of information display (@MacEachren1992; @Carr1992).
+Early papers that specifically reference "uncertainty visualisation" appear in the late 80s [@Ibrekk1987], with geospatial information visualisation literature in the early 90s declaring this to be essential aspect of information display [@MacEachren1992; @Carr1992].
 Despite the new terminology visualisation of uncertainty has been present since the earliest times.

### 2024-04-01 paper.qmd
data-driven

@@
 (@Vanderplas2020 has helpful advice here.) Visualization is an important step in exploratory data analysis and it is often utilised to **learn** what is important about a data set.
-The importance of data driven discovery has been highlighted through the data sets Anscombe's quartet [@anscombe] and the Datasaurus Dozen [@datasaurpkg].
+The importance of data-driven discovery has been highlighted through the data sets Anscombe's quartet [@anscombe] and the Datasaurus Dozen [@datasaurpkg].
 Each of the pairwise plots in these data sets have the same summary statistics but strikingly different information, as revealed when visualised.
@@
 
-![*I need to fix the taxonomy illustration because inputs and model uncertainty need to be swapped*.
-Depicts an illustration of the taxonomy described in @utypo.
+![Depicts an illustration of the taxonomy described in @utypo.
 From right to left the drawing shows the location, level and nature of uncertainty with examples of that category underneath.

### 2024-04-01 writingdumpster.qmd
sorted the misc notes in the writing dumpster

@@
 # Introduction
+- @Spiegelhalter2017 noted we "cannot assess the quality of risk communication unless the objectives are clear".
+- @Fischhoff2014  takes this approach by considerting how decisions with different levels of flexibility should have different uncertainty visualisations.
+If the user has no choice they should simply be made aware of the risk, if they have several options, the user needs to see the uncertainty in each, if they need to come up with the options themselves, they need to see how things work and how the system fails.
+
 ### Motivation for uncert vis
@@
 - (maybe include some of the noise vs signal stuff)
+- The use of uncertainty in high dimensional environments is especially important in energy data.
+Large models that incorporate spatial-temporal data from many sources and systems are used to predict energy uses in the short and long term.
+Understanding how to improve and make better decisions in these models is imperative in both the daily operation of the energy sector as well as in the transition from fossil fuels to clean energy.
+The energy sector needs better heuristics to make energy supply analysis less costly to conduct [@Stenclik2021], therefore it is an incredibly relevant application of uncertainty visualisation techniques.
 
 # Definition Notes
+- Failing to communicate the nature of your uncertainty can result in underestimation or overestimation of failure probabilities [@Kiureghian2009]
+
 ## General
@@
 
-## Probability/variance/error etc (Cite Gap 3)
-> Some consider it to be synonymous with specific terms defined in mathematics, such as probability [*HERE*], variance[*HERE*], error[*HERE*], or precision[*HERE*].
-Others consider it to be an encompassing umbrella term of which for these mathematically defined objects are only examples[*HERE*], and include concepts that are somewhat related, such as missing values[*HERE*]
-
 ## Redefined uncertainty (cite gap 3)
-> This wide array of uncertainty definitions becomes a snake eating its own tail, as many authors opt to redefine uncertainty themselves[*HERE*], ignore defining uncertainty entirely [*HERE*], or pick a definition from a seemingly randomly previously published papers [*HERE*], all of which make variance in uncertainty definitions even worse
 
 #### Uncertainty Taxonomies (Cite Gap 7)
-> Taxonomies split uncertainty based on an endless stream of arbitrary lines in the sand, such as whether the uncertainty is due to true randomness or a lack of knowledge [@Spiegelhalter2017 *HERE*], if the uncertainty is in the the attribute, spatial elements, or temporal element of the data [*HERE*], whether the uncertainty is scientific (e.g.
-error) or human (e.g.
-disagreement among parties) [*HERE*], if the uncertainty is random or systematic [*HERE*], statistical or bounded [], accuracy or precision [*HERE*], if the uncertainty is about the past or the future [*HERE*], the stage of the data analysis pipeline it comes from [*HERE*], the severity of the uncertainty [*HERE*], how quantifiable the uncertainty is [@Spiegelhalter2017], and many, many, many more examples.
 
+- @Gustafson2019 organises uncertainty into the four groups: Deficient: Lack of knowledge; Technical: modelling approximations and measurement error; Scientific Unknown unknowns; Consensus: Disagreement among parties
 - @uncertchap2022 mention aleatric, epistemic, ontological divide and also differentiates risk with uncertainty
@@
 - @Griethe2006 clearly sees uncertainty as additional information or data, and therefore sees the uncertainty visualisation problem as a specific case of the high dimensional visualisation issues. 
+- It is important to keep in mind that just because information is in a graphic, that does not mean it will be "seen".
+The phenomena of inattentional blindness shows that there is no perception without attention and it is powerful enough that participants can fail to be aware of random objects appearing on a screen or a gorilla walking through a basketball game [@simons1999gorillas; @mack2003inattentional].
+Including uncertainty is worth very little if no attention is left to see it.
+- Glyph maps can be used to depict the multi-dimensional information in spatial-temporal data by mapping line plots to map locations, but we still need to decide what information is important.
+The map can either present trends in the global or local variance depending on whether or not the line plot is scaled globally or locally [@Wickham2012], however smaller details are almost impossible to convey.
+No matter how much information we try and put in a plot, there will always be only a handful of key takeaways.
+- Elements from a single distribution should be displayed using a single plot, since displaying the features of one distribution across multiple plots makes the information hard to combine and results in some details (such as the estimate error) being completely ignored [@moritz2017trust; Correll2018].
 
 # Noise vs Signal
+- Upon further interviews @Hullman2020a found that authors believed uncertainty would overwhelm the audience and make their data seem questionable because decision makers are unable to understand uncertainty.
 - @uncertchap2022 "Effectively communicating uncertainty is necessary for establishing scientific transparency"
 - @Griethe2006 defined uncertainty as "the degree to which the lack of knowledge about the amount of error is responsible for hesitancy in accepting results and observations without caution". 
+- confidence intervals provided in text form only are less likely to be misinterpreted than graphics [@Savelli2013], text is insufficient to express more complicated aspects of a distribution, such as mass.
+  - This kind of harks back to the "if our main concern is accuracy, we should show the values in a table" issue.
+- A survey conducted by @Hullman2020a found that majority of visualisation authors agreed that expressing uncertainty is important and should be done more often than it currently is, some even agreed that failing to do so is tantamount to fraud.
+Despite this, only a quarter of respondents included uncertainty in 50% or more of their visualisations [@Hullman2020a].
+- Some economists suggest that visualisation authors are responding to incentives that make it tempting to avoid visualising uncertainty, even if those incentives are based more in perception than reality [@Manski2020].
+- The study by @Hullman2020a found that the most common reasons authors don't visualise uncertainty despite knowing it's moral importance are: not wanting to overwhelm the audience; an inability to calculate the uncertainty; a lack of access to the uncertainty information; and not wanting to make their data seem questionable [@Hullman2020a].
+- At least one interviewee from the study by @Hullman2020a claimed that expertise implies that the signal being conveyed is significant, but also said they would omit uncertainty if it obfuscated the message they were trying to convey.
+Other authors who were capable of calculating and and representing uncertainty well did not do it, and were unable to provide a self-satisfying reason why [@Hullman2020a].
+These conflicting motivations are acknowledged in the paper itself where @Hullman2020a says:
+
+> "It is worth noting that many authors seemed confident in stating rationales, as though they perceived them to be truths that do not require examples to demonstrate.
+It is possible that rationales for omission represent ingrained beliefs more than conclusions authors have drawn from concrete experiences attempting to convey uncertainty".
 
@@
 ### Value Extraction (Cite Gap 15)
+- This idea also extends to interval estimation, where questions about probability are best answered with gradients and questions about start and end times are easiest to answer with ambiguation [@Gschwandtnei2016].
 - @uncertchap2022 Attribute substitution: people will substitute a hard mental calculation for an easy one (attribute substitution heuristic).
@@
 ### Decision making stuff (Cite Gap 16)
+- Providing a categorical decision alone is somewhat useless [@Joslyn2012], and visualising a single estimate is akin to providing a decision or expressing no uncertainty at all. 
+- This reality cannot be avoided by providing secondary or non-specific information such as explaining calculations [@Joslyn2012], explaining the advantages of a recommendation [@Joslyn2012], or expressing uncertainty in vague terms [@Erev_1990; @Olson_1997], all of which are undesirable for decision makers and lead to measurably worse decisions [@Joslyn2012; @Erev_1990; @Olson_1997].
+- There is some research that suggests laypeople cannot understand complicated concepts in statistical thinking (such as trick questions on hypothesis tests or the difference between Frequentist and Bayesian thinking) [@Hoekstra2014; @Bella2005] but there is a large amount of research suggesting that presenting uncertainty information improves decision making, both experimentally [@Joslyn2012; @Savelli2013; @Kay2016; @Fernandes2018] and in practice [@Al-Kassab2014].
+As a matter of fact, doing what many authors currently do (providing only a deterministic outcome with no uncertainty) causes decision makers to be *less* decisive and have completely unbounded expectations on an outcome [@Savelli2013].
+
 - @Boone2018 organised responses from participants by the approach they used
@@
 - @uncertchap2022  mentions "other studies have found that people trust icon arrays more than other common visualization techniques" without commenting on whether or not that trust is good?
+- A 6-month survey of anti-mask groups on Facebook during the COVID-19 pandemic showed that the anti-maskers thought carefully about their grammar of graphics and made persuasive visualisations using the same data as pro-mask groups.
+They did this by exploiting information ignored by the pro-maskers [@Lee2021].
+- displaying numerical estimates of uncertainty information has shown to lead to greater trust in predictions [@Joslyn2012; @VanderBles2020].
+While @Han2009 found people have more worry when presented with uncertainty regarding health outcomes, this worry is not a bad thing if the concern is warranted given the ambiguous situation.
 
@@
 ## Uncertainty
+- Expressing uncertainty verbally decreases the perceived reliability and trustworthiness of the source [@VanderBles2020].
+- @Gustafson2019 found that the framing of our uncertainty, (i.e.
+if the source of uncertainty is from a lack of knowledge, approximations, unknown unknowns, or disagreement among parties) was found to not have a detrimental effect on the belief in the estimates, perceived credibility, or behavioural intentions of the decision makers.
+- Risk communication also needs to have clear objectives, use plain language, limit information to only what is necessary, and segment the audience to allow for differences in interest and knowledge [@Spiegelhalter2017].
 - @Spiegelhalter2017 it is better to understand your audience and iterate towards a final design, rather than only consider one option.
@@
 ## Visualisation
+- Error is best mapped to fuzziness, location, and colour value; arrangement, size and transparency are an OK second choice; but saturation, hue, orientation and shape are unacceptable and have no intuitive connection to variance [@Maceachren2012].
+No only do the graphical elements we map our features to matter, but the direction matters too.
+Graphical elements that are more fuzzy (fuzziness), further from centre (location), lighter (colour value), poorly arranged (arrangement), smaller (size), more transparent (transparency) are perceived to be more uncertain [@Maceachren2012].
+- There is a reasonable amount of evidence that cumulative displays or discrete displays (such as a quantile dot-plots or histograms) are the best ways to express mass for decision making and probability estimates [@Fernandes2018; @Hofmann2012; Kay2016; @Hullman2018; @kale2019decision].
+- Visualisation allow for interactive graphics that provide a more in depth understanding of probability [@Potter2009; @Ancker2009] and 
+- Infographics that make uncertainty more accessible for people with poor numeracy skills [@Ancker2009]. 
+- Any confusion caused by expressing uncertainty as a visualisation could also be due to a lack of exposure, since @Kay2016 found people repeatedly exposed to the same uncertainty visualisations quickly get better at making judgements.
+- Even something as simple as sketching a distribution before recalling statistics or making predictions can greatly increase the accuracy of those measures [@Hullman2018; @Goldstein2014].
 - @uncertchap2022 Icon array and quantile dot plots both support area and frequency based judgements.
@@
 - coloured maps for earthquake risk are easily interpretable @Spiegelhalter2017
-
-# Main section outtakes
-Conclusion
-In this paper we have highlighted a series of misunderstandings, confusions, and methodical errors in the uncertainty literature.
-Many of these issues are not problems in isolation, so long as the authors themselves are aware of it and state these assumptions in their papers.
-Our problem with this body is work is that the authors themselves seem to be oblivious to this vast array of problems, considering not a single one of the problems discussed here are mentioned explicitly in a single visualisation paper.
-
-It is clear that value retrival and questions with a specific ground truth are not appropriate, as uncertainty visualisation will simply follow the existing visualisation rules.
-Testing a secondary feature of a plot, that is, how strong a signal is becomes a little more complicated.
-Line up protocols have been used in adjacent work that look at the strength of the signal in a plot, and this idea of identifying if plots have some "barely noticiable differences" could be utilised.
-
-Specifically, a definition that has a place for all the concepts that currently sit under the "uncertainty" umbrella, explains how these concepts relate to uncertianty, and also allows us to quantify them in a way where they can be combined or visualised.
-This work would not only benefit uncertainty visualisaiton, but also statistics more broadly.
-
-The problem of trying to assess uncertainty without knowing *why* we are depiting uncertainty has been noticed by other authors.
-@Spiegelhalter2017 noted we "cannot assess the quality of risk communication unless the objectives are clear".
-
-Does the flip (hypothesis test vs uncertainty visualisation) have mathematical implications in how the plots should be assessed?
-This implies a visible signal in an uncertainty plot should indicate some divergence from the null.
-
-This increased sensisitivity could be the result of another aspect of uncertainty that is ignored.
-Statstical tests are typically built upon a series of assumptions, and it is difficult to identify *which* assumption failure caused the data to fail the statistical tests.
-Lineup plots allow us to see *how* our data is different to the assumed distribution, and better understand *why* our data may have failled a test ^[Did Patricks paper compare statistical tests to lineup protocols in the event an earlier assumption of the tests, such as, independent observation?
-I think he did check some.
-I wonder if you can quantify the information difference you get from a rejected statistical test vs a rejected line plot].
-
-It is unclear how participants are supposed to incorperate uncertainty information into their responses when uncertainty visualisation authors themselves are unsure.
-Other authors interpret this difficulty as a combination of the uncertainty and task complexity [@Kinkeldey2014], and not as a general confusion.
-All these examples sever to illustrate that *we dont know how to measure uncertainty* and if we don't know how to measure it, *we cannot evaluate the quality of a visualisation*.
-
-Using the most commonly presented methods In the absense of an encapsulating mathematical definition this will not change.
-That being said, even with a definition that allows us to combine these uncertainties into a single value, it would still be beneficial for authors to develop specialty visualisations for different sources of uncertainty, so analysts can utilise visual tools to understand the uncertainty in their data.
-still might be desirable to see which stage of the analysis is creating uncertainty in the visualisaiton.
-
-@Hofmann2012 showed two distributions in 20 different visualisations (a lineup protocol) using a jittered sample, a density plot, a histogram, and a box plot and asked participants.
-Participants were asked to report in which of the plots was "the blue group furthest to the right" and to provide reasoning from a multiple choice list and grade their certainty.
-The experiment set up is shown in @fig-right.
-The participants answers were then compared to a ground truth where the correct plot had a right shifted mean.
-By comparing the results to a ground truth statistic and marking participants as "wrong" or "right", the error from the participants that had an alternative interpretation to the concept of "furthest right" was conflated with the error from a the visualisation choice.
-
-While this is a focus on uncertainty visualisation, these issues in uncertainty communication need to be acknowledged because they bleed into the uncertainty visualisation literature.
-Uncertainty visualisation are compared on their ability to communicate explicit risk, elicit trust, and prevent risk or ambiguity avoidance, continuing these issues in uncertainty communication at large.
-
-Ambiguity-aversion, uncertainties risk-avoidance counterpart, has a similar problem.
-Ambiguity-aversion is the tendency of people to prefer to take on certain risks rather than unknown risks.
-It is essentially risk-aversion applied to risk.
-For this reason discussions of ambiguity-aversion have the same issues as the conversations about risk-aversion, where they do not consider ambiguity-aversion to be an estimate of the utility of decreased certainty.
-Assuming that participants should act as rational agents implies they should do as rational agents do and *ignore* the uncertainty information.
-
-The idea that risk and ambiguity aversion reveal the utility of decreased variance seems to seldom be considered in the literature, therefore it should be mentioned here, otherwise this issue might continue.
-However, considering risk aversion is at its core, a trade off between bias and variance, it is bizzare to assume it to be a mistake when statisticians perform these trade offs all the time.
-If risk-aversion is truly "irrational" behaviour, every statistical textbook that discusses the use of a biased but consistent estimator in a finite sample case (a common example would be specific cases of the maximum likelihood estimator) should be tossed out.
-Discussing risk and ambiguity aversion as though it is some diversion from a perfectly rational, mathematical choice, implies there is *no value to a decrease in variance which is not even true within mathematics*.
-There are some elements of risk communication that seem to translate to uncertaitny communication more broadly.
-Using words to communicate risk is discouraged because people can misinterpret the actual risk involved, so when possible numerical estimates of risk should be communicated [@Spiegelhalter2017] the same rules apply to uncertainty (*Cite gap __*).
-People performing better and preferring with numerical estimates of risk translates to uncertainty communication more broadly .
-Risk communication also needs to have clear objectives, use plain language, limit information to only what is necessary, and segment the audience to allow for differences in interest and knowledge [@Spiegelhalter2017].
-These general concepts of communication also extend to communication of uncertainty.
-On the other hand, it is not sensible to assume that probability specific concepts such as framing, intuitiveness of probabilities, or reference classes are still relevant when we consider uncertainty communication as signal supression, not as a synonym of risk or probability communication.
-In this same vein, most uncertainty communication papers very often discuss risk-aversion as something that needs to be considered despite it being unclear exactly how risk-aversion is related to uncertainty communication *(citations for the papers that do this)*.
-Very often the decisions made in decision making tasks are compared to some "correct" choice, typically the choice of a rational agent.
-That is, visualisations that elicit choices that are most closely aligned with those of rational agents are seen as the best.
-There is a critical error with this way of thinking.
-If a group truth is specified (through right/wrong answers or a payment scheme) to correctly answer the question, the user needs to work out the threshold that will maximise payout, over-ride their personal risk utility function to adopt this threshold, and report values that align with this threshold.
-This is just a value extraction task with additional unnecessary extra steps.
-Additionally, some authors speak about plots exhibiting "risk-avoidance" which is comical, considering a rational agent who displays no risk-avoiding will simply make the choice that maximise expected value, or in other words **completely ignores the uncertainty information**.
-The ground truths in these experiments *encourage* participants to think of uncertainty information as useless noise that should be ignored, the way a rational agent will.
-There is also the common view that uncertainty as a dimensionality problem.
-This perspective is not quite right and the literature seems to be about 50/50 on whether uncertainty is a psychological or dimensionality problem.
-It is important to keep in mind that just because information is in a graphic, that does not mean it will be "seen".
-The phenomena of inattentional blindness shows that there is no perception without attention and it is powerful enough that participants can fail to be aware of random objects appearing on a screen or a gorilla walking through a basketball game [@simons1999gorillas; @mack2003inattentional].
-Therefore if information is mapped to graphical elements that are so low on hierarchy they can be ignored, they might as well not be there at all.
-The people who don't plot uncertainty because they think its unimportant and the people who relegate uncertainty to the lowest ranking aesthetic on the list are both saying the same thing, "I think the uncertainty is unimportant".
-Including uncertainty is worth very little if no attention is left to see it.
-This does not mean we cannot put a *large* amount of information in a graphic.
-Glyph maps can be used to depict the multi-dimensional information in spatial-temporal data by mapping line plots to map locations, but we still need to decide what information is important.
-The map can either present trends in the global or local variance depending on whether or not the line plot is scaled globally or locally [@Wickham2012], however smaller details are almost impossible to convey.
-No matter how much information we try and put in a plot, there will always be only a handful of key takeaways.
-
-This issue has come up in many different forms in the uncertainty visualisation literature.
-Typically as a "task dependence" on the performance of uncertainty visualisations.
-It is repeatedly identified as a problem in previous reviews of the uncertainty visualisation literature [@Kinkeldey2014; @Hullman2016] as well as across many sub domains and applications [@Wallsten1997; @Fischhoff2014; @Meng2021; @Amar2005].
-The fact that this conclusion is repeatedly reached shows both the importance and the lack of acknowledgement this concept receives.
-@Fischhoff2014  takes this approach by considerting how decisions with different levels of flexibility should have different uncertainty visualisations.
-If the user has no choice they should simply be made aware of the risk, if they have several options, the user needs to see the uncertainty in each, if they need to come up with the options themselves, they need to see how things work and how the system fails.
-@Kinkeldey2014 suggested that the nature of the task plays an important role and believed the task dependence for uncertainty visualisation could explain some of the inconsistency in outcomes.
-
-Several uncertainty visualisation literature reviews have commented on the immense difficulty of combining results across experiments, but none have been able to pinpoint the reason.
-@Kinkeldey2014 suggested a task based taxonomy that organised the plots by the question or purpose of the visualisation, and even suggested that communication tasks (such as value retrival of uncertainty values) should just follow traditional cartography rules, @Spiegelhalter2017 simply ignored the existence of uncertainty visualisation as a subfield and just suggested we follow established rules within uncertainty visualisation, @Hullman2016 said that commented on the large amount of noise generated from uncertainty visualisation evaluation methods that led to a difficulty in combining results, @Griethe2006 did not meaningfully combine the work and instead listed findings in the field, but also did not comment on the lack of generalisable results.
-While we largely agree with the difficulties of these authors, we do not agree that the solution is a task based taxonomy or better questions.
-While we agree the lack of attention to tasks and appropriate questions has generated problems in the field, there is a much more pressing issue, which is that uncertainty visualisation authors don't seem to understand *why* uncertainty visualisation different to typical visualisations.
-
-Attempts to rectify these conflicts always flutter about the surface and do not seem to grasp the core of this problem.
-What these suggestions fail to understand is that the entire practice of identifying "true" facts about "unertainty visualisation" is futile.
-It is impossible to make a statement about the properties of an "uncertainty visualisation" in the same way it is impossible to comment on $\frac 0 0 = 1$.
-Division by zero is undefined, so the statement is ill formed, similarly, "uncertainty" is not defined, so statements *about* "uncertainty" including statements about "uncertainty visualisation" are ill formed.
-Ill-formed statements can neither be true nor false and cannot be part of a logical argument, so experiments and research that build upon this ill-formed concept are neither true nor false, they simply exist.
-Someone who is less familiar with the rules of logical argument may believe this understanding of the problem to an insufferable nitpick that creates impossible to live up to standards, however the consequences of this definitional issue are *very* real and costly.
-The studies that investigate "uncertainty visualisation" may be identifying *something*, however there is good reason to believe that *something* is neither interesting nor does can it contribute to a larger theory of visualisation.
-This is something we illustrate in the following this section.
-
-- The process of working out whether or not a visualisation shows a piece of information, and if it does, what perceptual task it is mapped to and what assumptions or heuristics need to be used, is *incredibly* time consuming.
-While this method does predict the results for these experiments with remarkable accuracy, to do it for a large number of plots, especially when the ground truth statistics and visualisation are incredibly complicated is just not feasible for a reviewer to do.
-- This is likely because The idea that uncertainty can only be defined in the presence of a motivating question is well grounded in most areas of statistics.
-The entire process of data analysis, from deciding what should be observed as data through to communicating that data in a plot is governed by human decision and the goal of an analysis.
-- We know uncertainty cannot be entirely explained by individual concepts that *have* been defined, such as probability, variance, error, missing values, precision, or related concepts that pop up throughout the data analysis pipeline.
-We also know it likely has some relationship with the *task* we are considering, given that many authors have attributed the noise in uncertainty visualisation research as a type of "task" dependence.
-Additionally, it should be clear that we wouldn't consider uncertainty to be an appropriate term for *everything* in statistics would we?
-- What exactly is uncertainty, then?
-Let us consider the concepts related to uncertainty that have been discussed so far.
-We know uncertainty cannot be entirely explained by individual concepts that *have* been defined, such as probability, variance, error, missing values, precision, or related concepts that pop up throughout the data analysis pipeline.
-We also know it likely has some relationship with the *task* we are considering, given that many authors have attributed the noise in uncertainty visualisation research as a type of "task" dependence.
-Additionally, it should be clear that we wouldn't consider uncertainty to be an appropriate term for *everything* in statistics.
-For example, I wouldn't describe the population parameters of a distribution I have simulated to be "uncertain"
-
-- tasks e.g.
-"What is the average of X" (how well lined up you are moving down the data pipeline) and resolution (uncertainty is very often moving back a resolution level in statisitcs - also can be an assumption - bootstrap is a replacement for an assumption, not a resolution change, but bootstrapping is centered on the data)  e.g.
-"What is the noise around the estimate of the value of X"
-Visualisation experiments are highly sensitive to the way questions are asked.
-For example, a study by @Hullman2015 that compared HOPs, error bar, and violin plots found that many participants reported that group A was more likely to be higher than group B, which was not not plausible given that the mean of B was larger than the mean of A.
-The authors attributed this obvious mistake to the visual representation, but we believe it was more likely caused by the distributions being shown in the order A then B while the question swapped the order by asking "What is P(B>A)?".
-This flip is minor, but must be done *after* the participants have worked out the probability.
-Since the primary task is done they forget to complete the secondary task, similar to forgetting your bank card in an ATM.
-I suspect this was the cause of the mistake as I, when attempting this question myself, made the exact same mistake.
-I also constalty lead my bank card in ATMs.
-
-While mathematical information *should* be considered the starting point in considering if two visualisations are equivalent, it is not the only consideration.
-A large amount of work in perceptual tasks, attention, and psychology surrounding charts shows that all the information depicted is not paid attention to in equal measure.
-Elements such as colour can have sizable impacts on the way a graphic is percieved.
-Mathematics identifies the information that is *shown* while psychology identifies the difference in information *recieved*.
-A failure to understand this distinction leads to graphics that are different not only in the way information is depicted but in the information itself.
-This problem is pervasive in uncertainty visualisation literature.
-
-This is not to say any paper that discusses uncertainty visualisation is completely pointless because the results already exist.
-We are simply pointing out that treating uncertainty graphics as special simply because they contain uncertainty, even if that uncertainty is not related to the task, leads to results that are already established in information visualisation.
-
-These two perceptualisations of *why* uncertainty visualisation is different from normal visualisation is also what leads to differing opinions on what is an uncertainty visualisation.
-For example, someone who percieves of uncertainty visualisation as a dimensionality problem would see a pie chart as a normal graphic, and a pie chart with blurred boundaries as an uncertainty visualisation (even though blurring the boundary between segments of a pie chart makes no sense when you think of a pie chart as a polar co-ordinate bar chart), while someone who sees uncertainty visualisation as psychologically different would see a pie chart as an uncertainty graphic so long as it depicts probabilities.
-This may lead you to believe that those who believe uncertainty is unique due to its psychological properties believe uncertainty should be defined at the mathematical stage, while those that believe it is an addition to an established visualisation think uncertainty is defined in a graphic at the aesthetic stage.
-This is not the case.
-As shown by the *grammar of graphics fig* there is no single person or experiment that is consistent in any one belief.
-This difficulty and lack of consistency comes from the desire to discuss "uncertainty visualisation" as though it is a single term and not the combination of "uncertainty" and "visualisation".
-If we split the terms and look at the quantification and visualisation of uncertainty separately, the results of many uncertainty visualisation papers become self evident.
-
-Papers that are justified using the first type of motivation rarely explain exactly how the psychological apsects of uncertainty impct the participants in a way that inhibits their ability to read information from a graphic.
-Since uncertainty is poorly defined and the psychological effects of uncertainty is never justifiably connected to the scientific communication of it, as was discussed in ssections 3 and 4, it is unclear exactly *what* these papers are testing for.
-It is almost as though they *feel* that uncertainty visualisations *should* be different, but do not consider *how* uncertainty information is supposedly different beyond decision making criteria such as *risk avoidance* which is not something science communicators have any business trying to control, as was discussed in previous sections.
-For these reasons, experiments that test for secondary factors (such as influence on decision making) are confounding far too many variables to establish any generalisiable take away, and experiments that ask participants to extract values fail to notice there is no longer a reason for "uncertainty" to be any different to any other type of visualisation.
-
-Papers that are justified using the second type of motivation clearly fit within the sub-field of high dimensional visualisation, but also rarely comment on the connection.
-These experiments evaluate different ways to include even *more* information than would be put in a typical visualisation.
-Some of these methods, such as a bivariate pallet or glyph map, create graphics that are overloaded with information and offer incorrect or meaningless takeaways at first glance.
-This is not to say *every* visualisation method that sees it as a dimensionality problem is bad.
-Clever methods of combining several types of information, such as value supressing uncertainty pallets (VSUPs), can better help us understand how our brains process information in a graphic.
-For example, the VSUP allows us to test is colour value and saturation can be combined such that we mentally combine signal and noise together as one variable.
-Since the take aways from the papers are not usually related to uncertainty specifically, and often related to interesting perceptual tasks, there is also no justification for why this work is "uncertainty visualisation" work and not simply perceptual task studies that can be used by the rest of the information visualisation community.
-
-This stems from a class of "uncertainty visualisation" that seems to have no rules for what belongs in the group, and no consideration for the mathematics that is actually depicted in these plots.
-Uncertainty visualisation experiments do not work to build upon existing visualisation literature or differentiate uncertainty visualisaiton as a new field, but seem to exist with shocking indifference to the field it belongs to.
-
-uncertainty visualisation, as it is now, should not exist.
-Any generalisable contribution made by these papers is already an established result in information visualisaiton, and therefore is not specific to "uncertainty visualisation".
-Any "uncertainty visualisation" specific result is not generalisable beyond the experimental conditions because of the poor formalisation in the definitions of uncertainty and uncertainty visualisations.
-
-This lack of formalisation is not unique to uncertainty visualisation and the wider field of information visualisation suffers form similar problems [@Kinkeldey2014].
-
-This makes it clear we have a serious problem with the classification of "uncertainty visualisation".
-The visualisations follow identical rules to those established already in the visualisation literature, and experiments that view uncertianty visualisation as somehow special or different from normal visualisation are simply repeating existing established research in information visualisation.
-
-If several graphics are compared that are different statistically, geometrically, and aesthetically, there is *no way to combine the results to establish generic principles for the field*.
-
-The overwhelming majority of these papers depict different mathematical expressions of both the signal and the uncertainty.
-In 100% of these papers, the plot that has mapped the statistic in question to some visual feature will outperform those that don't.
-When the statistics used to depict uncertainty *are* the same (which is rare) 100% of those papers will follow the hierarchy of perceptual tasks that has already been established in the visualisation literature.
-If the experiment participant needs to extract a probability and one graphic uses position along a common scale, while the other uses colour or literally any other perceptual task, the position depiction will always outperform the other.
-
-If we were to compare two different visualisations, we would like to change as few features as possible so we can better understand the cause of the difference in perception.
-
-Additionally, if information that can be used to calculate a statistic, but not the statistic itself, is shown, people will need to use a heuristic to extract that information which will decrease the precision with which that information is communicated.
-If the information that is depicted on a plot cannot be used to *mathematically* generate a best estimate for a specific statistic, it is questionable how uncertainty visualisation authors expect participants to cover that gap if *not* for heuristics.
-
-In these papers will typically ask separate questions about the signal (e.g. what is the mean of this distribution) and the uncertainty (e.g. what is the variance of this distribution).
-
-Failling to understand this key aspect of visualsiation leads to graphics that *contain* the correct amount of information, but that volume of information cannot be readily accessed by the viewer.
-
-## Taxonomies
-Some taxonomies contain overlapping terminology but also have alternative definitions and categories.
-For example *___* organises uncertainty into the two categories, Aleatroy (inherrent randomness) and epistemic (lack of knowledge), but *___* organises uncertainty into the three categories aleatory (inherrent randomness in a forecast) vs epistemic (uncertainty in structure and parameters of statistical models) vs ontological (uncertainty about the entire modelling process as a description of reality).
-
-To avoid alienating lay people with confusing complexity, many of these taxonomies are overly simplistic.
-Most taxonomies seek to organise uncertainty based off where the uncertainty is coming from.
-For example, @Spiegelhalter2017 organises uncertainty into three categories:
-  - Aleatory uncertainty: inevitable unpredictability due to unforeseeable factors 
-  - Epistemic uncertainty: uncertainty about the structure and parameters of statistical models
-  - Ontological uncertainty: uncertainty about the entire modelling process as a description of reality.
-These three categories can be considered to be uncertainty when deciding which model we should use (ontological), uncertainty surrounding parameter estimates (epistemic), and uncertainty surrounding a forecast (aleatory), however it is not hard to find another, very different, example of a taxonomy that organises uncertainty based on what is seen to be causing it.
-@Gustafson2019 organises uncertainty into the four groups:
-  - Deficient: Lack of knowledge
-  – Technical: modelling approximations and measurement error
-  – Scientific Unknown unknowns
-  – Consensus: Disagreement among parties
-However the additional comments about the inherent variability of aleatory uncertainty results in a bit of confusion as to if our cateogirisation should change if we decide to treat the parameters of a statistical model to be inherently random (as is done in Bayesian statistics) or if we beleive a forecast variability to be caused by a lack of knowledge.
-
-He kept the taxonomy to three levels to maintain simplicity, however this organisation methods conflates amount of uncertainty present in an estimate with what is causing that uncertainty, and in doing so makes it hard for us to use this definition to better understand uncertainty within our own analysis.
-# context section outtakes
-Given that uncertainty estimation and communication seem to be largely task depenent, what questions should we be testing?
-This leads to several unanswered questions.
-Is our goal in uncertainty communication to acknowledge and convey the inherrent randomness that exists in all statistical calculations and models?
-Is it to supress signal that could be the result of random variance?
-Is it to find the best way to convey the mass function of a random variable?
-A survey of the literature will leave the answers to these questions unclear, however this inherrent issue has not gone unnoticed.
-This conceptual way of thinking about uncertainty has been rattling around in the different areas of data science for a while.
-For example, when we record coin flips we typically want to model the behavior with a binomial distribution, so we ignore the outcome of a coin landing on its side despite the fact it is a real world possibility.
-
-# General Thoughts For Actual Review
-##### General
-- I might need to go back through and add in notes about general visualisation stuff and bad papers for motivations
-- could also help to include a bunch of papers that talk about density visualisation without mentioning uncertainty visualisation in the "context" section. (for now have left those unsorted)
-
-##### Introduction
-
-##### Defining Uncertainty
-  - The more taxonomies I read the more convinced I am the "defining uncertainty" is the best one.
-Not only is it the most complete, but it also identifies WHY some uncertainties are difficult to visualise (specifically which part of the taxonomy) and suggests HOW you can best communicate these uncertainties.
-    - A lot of other papers just make a comment like "oh some of these are hard to communicate" but that paper actually goes into detail instead of hand waving the problem away.
-    - It also explains why the uncertainty visualisation literature focuses on a specific subset of uncertainty cases.
-  - Examples
-    - Null Values
-    - Variance
-    - Hypothesis Testing
-    - Distribution
-  - Combining uncertainty is also a reccurring theme in these papers
-- I think it is also important to highlight if uncertainty is meta information or if it is a variable in of itself.
-Meta information (such as contiunity, exhangeability, etc) are usually set to specific visual features (e.g.
-VSUP clearly sees uncertainty as meta information)
-- A lot of work that evaluates uncertainty visualisations checks to see if the mean can be gathered from a visualisation, but if we view uncertainty as meta-information, then with a high uncertainty we wouldn't want the mean to be visible
-  - Also an example of why the motivation matters for these things.
-If you don't know "WHY" someone is using a visualisation for uncertainty and just pick the "best" one, you are going to get a visualisation that does a poor job of its task.
-  - put uncertainty in the context of other meta-data of variables
-
-##### Visualisation as a form of Uncertainty communication
-- A LOT OF the work in uncertainty communication has a sub-field that is visualisation specific
-  - Basically I want a brief section highlighting the importance of visualisation and discussing it in relation to other forms of uncertainty communication.
-  - especially considering some uncertainty visualisation papers discuss using just a text based method to communicate uncerainty
-
-##### Uncertainty visualisation best practices
-  - There is also two kinds of papers, those that compare elements of a plot (like mapping uncertainty to an aesthetic) and those that compare some expression of a distribution (like a scatter plot compared to a box plot).
-  - try to map back to david speigleholder paper i.e. uncertainty means lots of different things
-
-# Previous literature review
-
-#####Taxonomy Benefits
-- Distinctions at this taxonomy level seem to be important also untangling helps to understand it
-  - Failing to communicate the nature of your uncertainty can result in underestimation or overestimation of failure probabilities [@Kiureghian2009]
-  - @Gustafson2019 found that the framing of our uncertainty, (i.e.
-if the source of uncertainty is from a lack of knowledge, approximations, unknown unknowns, or disagreement among parties) was found to not have a detrimental effect on the belief in the estimates, perceived credibility, or behavioural intentions of the decision makers.
-
-##### The benefits of Unvertainty visualisaiton
-Visualisations can provide a more complete picture of a risk than numerical summaries alone.
-Even something as simple as sketching a distribution before recalling statistics or making predictions can greatly increase the accuracy of those measures [@Hullman2018; @Goldstein2014].
-While there is some evidence that confidence intervals provided in text form only are less likely to be misinterpreted than graphics [@Savelli2013], text is insufficient to express more complicated aspects of a distribution, such as mass.
-Expressing uncertainty verbally decreases the perceived reliability and trustworthiness of the source [@VanderBles2020].
-Any confusion caused by expressing uncertainty as a visualisation could also be due to a lack of exposure, since @Kay2016 found people repeatedly exposed to the same uncertainty visualisations quickly get better at making judgements.
-Additionally, visualisation allow for interactive graphics that provide a more in depth understanding of probability [@Potter2009; @Ancker2009] and infographics that make uncertainty more accessible for people with poor numeracy skills [@Ancker2009].
-**application in energy**
-The use of uncertainty in high dimensional environments is especially important in energy data.
-Large models that incorporate spatial-temporal data from many sources and systems are used to predict energy uses in the short and long term.
-Understanding how to improve and make better decisions in these models is imperative in both the daily operation of the energy sector as well as in the transition from fossil fuels to clean energy.
-The energy sector needs better heuristics to make energy supply analysis less costly to conduct [@Stenclik2021], therefore it is an incredibly relevant application of uncertainty visualisation techniques.
-
-##### People dont visualise uncertainty but it is considered important
-Despite these benefits, there is evidence that we don't visualise uncertainty as often as we should.
-A survey conducted by @Hullman2020a found that majority of visualisation authors agreed that expressing uncertainty is important and should be done more often than it currently is, some even agreed that failing to do so is tantamount to fraud.
-Despite this, only a quarter of respondents included uncertainty in 50% or more of their visualisations [@Hullman2020a].
-Meaning participants were convinced that visualising uncertainty is morally important but were able to provide self sufficient reasoning that allows them to avoid doing it.
-Some economists suggest that visualisation authors are responding to incentives that make it tempting to avoid visualising uncertainty, even if those incentives are based more in perception than reality [@Manski2020].
-The study by @Hullman2020a found that the most common reasons authors don't visualise uncertainty despite knowing it's moral importance are: not wanting to overwhelm the audience; an inability to calculate the uncertainty; a lack of access to the uncertainty information; and not wanting to make their data seem questionable [@Hullman2020a].
-
-##### Effect of uncertainty visualisation on decision making
-If decision markers are not presented with the uncertainty about an estimate the data analysts have, for all intents and purposes, made the decision for the decision maker.
-Upon further interviews @Hullman2020a found that authors believed uncertainty would overwhelm the audience and make their data seem questionable because decision makers are unable to understand uncertainty.
-This belief, while pervasive, is not true.
-There is some research that suggests laypeople cannot understand complicated concepts in statistical thinking (such as trick questions on hypothesis tests or the difference between Frequentist and Bayesian thinking) [@Hoekstra2014; @Bella2005] but there is a large amount of research suggesting that presenting uncertainty information improves decision making, both experimentally [@Joslyn2012; @Savelli2013; @Kay2016; @Fernandes2018] and in practice [@Al-Kassab2014].
-As a matter of fact, doing what many authors currently do (providing only a deterministic outcome with no uncertainty) causes decision makers to be *less* decisive and have completely unbounded expectations on an outcome [@Savelli2013].
-This reality cannot be avoided by providing secondary or non-specific information such as explaining calculations [@Joslyn2012], explaining the advantages of a recommendation [@Joslyn2012], or expressing uncertainty in vague terms [@Erev_1990; @Olson_1997], all of which are undesirable for decision makers and lead to measurably worse decisions [@Joslyn2012; @Erev_1990; @Olson_1997].
-One of the most popular depictions of uncertainty for decision making is a quantile dotplot, shown in @fig-quantdot.
-It is important to avoid whittling down the problem **too** much.
-Providing a categorical decision alone is somewhat useless [@Joslyn2012], and visualising a single estimate is akin to providing a decision or expressing no uncertainty at all.
-
-```{r}
-#| echo: false
-#| warning: false
-#| message: false
-#| label: fig-quantdot
-#| fig-cap: "This plot depicts an example of a a quantile dotplot that expresses the uncertainty associated with a daily maximum temperature.
-The probability associated with each temperature is expressed with discrete countable bins and the predicted temperature is expressed with a line.
-Discretised depictions such as this, make decision making in the face of uncertainty easier for the viewer."
-set.seed(1)
-dotplot_data <- tibble(temp = 10 + round(rnorm(40, mean=10, sd=2))) 
-mid <- round(mean(dotplot_data$temp), 2)
-dotplot_data %>%
-  ggplot(aes(x=temp)) +
-  geom_dotplot(binwidth = 0.75, fill="grey") +
-  theme_classic() +
-  geom_segment(x=round(mid), xend=round(mid), 
-               y=-1, yend=0.7, size=2) +
-  geom_label(x=round(mid), y=0.7, label=paste0(mid, "°C"),
-             fill = "black", fontface = "bold",
-             colour="white") +
-  scale_x_continuous(breaks = seq(10, 30, 5), 
-                     limits = c(10, 30)) +
-  theme(axis.line.y = element_blank(),
-        axis.ticks.y = element_blank(),
-        axis.title.y = element_blank(),
-        axis.text.y = element_blank()) +
-  labs(x = "Tomorrow's Daily Max Temperature")
-```
-
-##### Avoiding uncertainty creates a lack of trust
-Not only does communicating uncertainty improve decisions but the mistrust created by communicating certainty in uncertain situations can be exploited.
-A 6-month survey of anti-mask groups on Facebook during the COVID-19 pandemic showed that the anti-maskers thought carefully about their grammar of graphics and made persuasive visualisations using the same data as pro-mask groups.
-They did this by exploiting information ignored by the pro-maskers [@Lee2021].
-It is understood that deceptive plots can lead viewers to come to incorrect conclusions or significantly overstate effects or risks [@Pandey2015; @Padilla2022] but these incorrect takeaways cannot be mitigated with instructions in how to correctly understand the plot [@Boone2018].
-This evidence indicates we are more likely than not to hurt our message when we ignore uncertainty information and trying to raise the general public's plot literacy is an insufficient strategy to curb conspiracy theories and misguided scientific communication.
-In direct contrast to this, displaying numerical estimates of uncertainty information has shown to lead to greater trust in predictions [@Joslyn2012; @VanderBles2020].
-While @Han2009 found people have more worry when presented with uncertainty regarding health outcomes, this worry is not a bad thing if the concern is warranted given the ambiguous situation.
-
-##### Poeple might simply be chosing to not visualise uncertainty
-The disconnect between the research in support of visualising uncertainty and the consensus against it may not be entirely driven by a lack of understanding of the literature.
-For example, at least one interviewee from the study by @Hullman2020a claimed that expertise implies that the signal being conveyed is significant, but also said they would omit uncertainty if it obfuscated the message they were trying to convey.
-Other authors who were capable of calculating and and representing uncertainty well did not do it, and were unable to provide a self-satisfying reason why [@Hullman2020a].
-These conflicting motivations are acknowledged in the paper itself where @Hullman2020a says:
-
-> "It is worth noting that many authors seemed confident in stating rationales, as though they perceived them to be truths that do not require examples to demonstrate.
-It is possible that rationales for omission represent ingrained beliefs more than conclusions authors have drawn from concrete experiences attempting to convey uncertainty".
-
-#### Continuous vs discrete displays
-There are other distributional features, such as discreteness, that are important to consider in uncertainty visualisation.
-There is a reasonable amount of evidence that cumulative displays or discrete displays (such as a quantile dot-plots or histograms) are the best ways to express mass for decision making and probability estimates [@Fernandes2018; @Hofmann2012; Kay2016; @Hullman2018; @kale2019decision].
-
-#### Distributions shoud be in a single plot
-Elements from a single distribution should be displayed using a single plot, since displaying the features of one distribution across multiple plots makes the information hard to combine and results in some details (such as the estimate error) being completely ignored [@moritz2017trust; Correll2018].
-
-### Perceptual tasks wrt uncertainty
-Not only should we considered the hierarchy in the information we display, but we also need to consider the heuristics that connect some pieces of information to specific elementary tasks.
-For example @Hofmann2012 showed that polar co-ordinates are more effective than cartesian co-ordinates when considering data that depicts a 360 degree direction (a case where polar co-ordinates has a natural interpretation).
-Uncertainty also has a natural mapping that should be considered when we express it in a plot.
-Error is best mapped to fuzziness, location, and colour value; arrangement, size and transparency are an OK second choice; but saturation, hue, orientation and shape are unacceptable and have no intuitive connection to variance [@Maceachren2012].
-No only do the graphical elements we map our features to matter, but the direction matters too.
-Graphical elements that are more fuzzy (fuzziness), further from centre (location), lighter (colour value), poorly arranged (arrangement), smaller (size), more transparent (transparency) are perceived to be more uncertain [@Maceachren2012].
-This idea also extends to interval estimation, where questions about probability are best answered with gradients and questions about start and end times are easiest to answer with ambiguation [@Gschwandtnei2016].
-Heuristics can work against us just as much as they can work for us.
-The sine illusion can cause the confidence interval of a smoothed sine curve to seem wider at the peaks than the troughs, causing us to underestimate uncertainty associated with changing values [@Vanderplas2015].
-Therefore we should not only keep the hierarchy of information in mind when we map features of our distribution, but also take advantage of these intuitive mappings when we can.
 
@@
 	- Participant literacy
+
+# General thoughts/questions
+- Does the flip (hypothesis test vs uncertainty visualisation) have mathematical implications in how the plots should be assessed?
+This implies a visible signal in an uncertainty plot should indicate some divergence from the null.
+- This increased sensisitivity could be the result of another aspect of uncertainty that is ignored.
+Statstical tests are typically built upon a series of assumptions, and it is difficult to identify *which* assumption failure caused the data to fail the statistical tests.
+Lineup plots allow us to see *how* our data is different to the assumed distribution, and better understand *why* our data may have failled a test ^[Did Patricks paper compare statistical tests to lineup protocols in the event an earlier assumption of the tests, such as, independent observation?
+I think he did check some.
+I wonder if you can quantify the information difference you get from a rejected statistical test vs a rejected line plot].
+- If information that can be used to calculate a statistic, but not the statistic itself, is shown, people will need to use a heuristic to extract that information which will decrease the precision with which that information is communicated.
+- put uncertainty in the context of other meta-data of variables
+
+# Removed Sections

### 2024-04-01 writingdumpster.qmd
removed paper checklist

@@
 
-# Experiment paper checklist
-General
-- Specific Field of Research
-- Question asked
-- Risk or Uncertainty: 
-- Noise vs signal (risk vs uncertainty)
-- Major task goal
-    - e.g. identify a bias or misunderstanding, compare two visualisation methods
-- Minor task goal
-    - Communication(value retrieval)
-    - Analytical(complex consideration)
-    - Exploration(free for all)
-
-Uncertainty Considerations
-	- Source
-	- Level (statistical-ignorance)
-	- Nature (Epistemic/Aleatory)
-
-Visual Uncertainty taxonomy
-- Explicit/implicit
-    - (directly mapping or showing multiple outcomes)
-- Intrinsic/Extrinsic 
-    - (using existing symbols e.g. colour value, or new objects e.g. grids)
-- Visually integral/separable 
-    - (can or cannot be separated from the data) 
-- Coincidence/adjacent 
-    - (if data and uncertainty are in integrated or separate views
-- Static/dynamic 
-    - (animation/interaction)
-
-Other plot considerations
-	- Dimension of data
-	- Dimension of uncertainty
-	- Feature mappings
-
-Extra (not necessarily recorded)
-	- Metrics used (and recorded)
-	- Possible Heuristics
-	- What is the ground truth?
-	- Participant literacy
-
 # General thoughts/questions

### 2024-04-01 paper.qmd
removed old paragraphs after Di edit

@@
 - convert to American spelling
-- should it be data-driven or two words
 - May not need to reference the same paper again and again
@@
 
-<!-- ## Introduction-->
 From entertainment choices to news articles to insurance plans, the modern citizen is so over run with information in every aspect of their life it can be overwhelming.
@@
 Each of the pairwise plots in these data sets have the same summary statistics but strikingly different information, as revealed when visualised.
-<!-- Anscombe quartet is shown in @fig-anscombe, because describing the data is never the same as seeing it.
-Instead of having to repeatedly check endless hypothesis to find interesting numerical features: DON'T NEED MUCH EMPHASIS ON THIS DATA - VERY WELL-KNOWN -->
 Visualisations can **tell** us what is important about our data and **how** it might diverge from pre-conceived notions or what we expect.
-
-```{r}
-#| echo: false
-#| warning: false
-#| message: false
-#| eval: false
-#| label: fig-anscombe
-#| fig-cap: "The four scatter plots that make up Anscombe's quartet.
-The four scatter plots are visually distinct but have the same mean, standard deviation, and correlation.
-The visualisation highlights the importance of plotting your data to identify interesting features that are hidden by other summary statistics."
-library(tidyverse)
-tibble(x = c(anscombe$x1, anscombe$x2,
-             anscombe$x3, anscombe$x4),
-       y = c(anscombe$y1, anscombe$y2,
-             anscombe$y3, anscombe$y4),
-       Plot = c(rep("Plot 1",11), rep("Plot 2",11), 
-                rep("Plot 3",11), rep("Plot 4",11))) %>%
-  ggplot(aes(x,y)) +
-  geom_point(aes(fill=Plot), colour="black", 
-             size=3, pch=21, alpha=0.75) +
-  facet_wrap(~Plot) +
-  theme_classic() +
-  theme(aspect.ratio = 1,
-        legend.position = "none") +
-  scale_fill_brewer(type = "qual", palette = 4)
-```
 
@@
 For example, box plots or histograms can be considered to be displaying uncertainty in the sense of variability in observations sampled from a population distribution.
-<!--In the modern day, uncertainty visualisation has exploded into its own field that applies to a wide range of contexts.-->
-
-<!-- ## Similar work-->
-
-<!-- Any field that experiences an explosion of new research will also invite a series of literature reviews that seek to combine and summaries that information.
-Given that uncertainty visualisation has been around for a few decades now, there is already a wealth of uncertainty visualisation literature reviews.
--->
 
@@
 @Spiegelhalter2017 commented that different plots are good for different things, and disagreed with the goal of identifying a universal "best" plot for all people and circumstances.
-@Griethe2006 was unable to find common themes, but instead listed the findings and opinions of <!--each uncertainty visualisation--> a collection of papers.
+@Griethe2006 was unable to find common themes, but instead listed the findings and opinions of a collection of papers.
 
-There are several reasons provided to explain the difficulty in generalising <!--the findings in--> the uncertainty visualisation literature.
+There are several reasons provided to explain the difficulty in generalising the uncertainty visualisation literature.
 Some suggested or agree with the idea that visualisation typologies should move away from data types, uncertainty categories, and representation types and towards "task-centred typologies" [@Kinkeldey2014; @Hullman2016].
-Other papers indirectly hint at this <!--Some authors may not attribute the task to the *reason* the literature cannot be easily summarised, but do mention that--> by arguing that the choice of best visualisation is highly dependent on the specific goal [@Griethe2006; @Spiegelhalter2017].
+Other papers indirectly hint at this by arguing that the choice of best visualisation is highly dependent on the specific goal [@Griethe2006; @Spiegelhalter2017].
 Others contend that visualisations are highly dependent on the audience and there is no such thing as a "best" visualisation that will be accessible to all members [@Kinkeldey2014; @Spiegelhalter2017].
-These concerns appear <!--difficulties in noisy uncertainty visualisation findings arises regardless of the direction the analysis was approached, be it looking--> across all application areas,  <!--, in geospatial data [@Kinkeldey2014], or uncertainty visualisation as a whole [@Hullman2016; @Spiegelhalter2017; @Griethe2006].--> and an underlying thread is that the problem arises from a lack of a cohesive and encompassing **definition of uncertainty**.
-
-<!--This review believe all these sources of noise have a single root cause, a lack of a cohesive and encompassing definition of uncertainty.
-The absence of an encompassing definition of uncertainty is mentioned by every uncertainty visualisation directly [@Spiegelhalter2017; @Griethe2006] or indirectly by describing a myriad of ways it can be considered in the literature [@Kinkeldey2014; @Hullman2016], although it is never commented on as a source of the noise in the field.-->
+These concerns appear across all application areas, and an underlying thread is that the problem arises from a lack of a cohesive and encompassing **definition of uncertainty**.
 
@@
 The ideal role of this review to help synthesize this noisy field, by collecting together essential components, summarising different viewpoints the current literature, and provide a guide for potential research topics.
-<!-- In understanding this purpose, we will also consider whether or not "uncertainty visualisation" is a valid sub-field of visualisation itself, or if that distinction has only arisen as a by product of the confused definition of uncertainty.
-Ultimately, this paper should serve as a guide for future uncertainty visualisation authors to prevent their work from becoming noise in what is already an incredibly noisy field.-->
 
 # What is Uncertainty?
-<!--## Colloquial definitions - what's in this section is not colloquial definitions, it's more overview-->
+Uncertainty visualisation is directed towards communication - to the general public when disasters may hit, to decision makers in control centers, and a whole host of other public-facing tasks - needing an understanding of what it is when the term "uncertainty" is used.
 
-Uncertainty visualisation is directed towards communication - to the general public when disasters may hit, to decision makers in control centers, and a whole host of other public-facing tasks - needing an understanding of what it is when the term "uncertainty" is used.
-<!--is made particularly difficult by the term "uncertainty" lacking a commonly accepted definition in the literature.
-This mishmash of terminology leads to a large body of work, all claiming to finding the best visualisation or expression of of "uncertainty" but most don't even seem to agree on what uncertainty is.
--->
 Definitions in the literature appear seemingly view it to be encompassing and include everything the general public might consider to be uncertain.
-Some works (XXX eg) focus more narrowly on specific terms defined mathematically, such as probability, variance, error, or precision.
-Others (XXX eg)  <!--consider it to be an encompassing umbrella term of which for these mathematically defined objects are only examples, and-->  include broader loosely related elements, such as missing values (*Cite Gap 3: Organise paper definitions*).
-<!--This wide array of uncertainty definitions becomes a snake eating its own tail, as many authors opt to redefine uncertainty themselves[], ignore defining uncertainty entirely [], or pick a definition from a seemingly randomly previously published papers [], all of which make variance in uncertainty definitions even worse (*Cite Gap 3: Organise paper definitions*).
-
-While these vague definitions might be OK for polite conversation, they are *not* foundation on which to build an entire sub-field of visualisation.
-In order to visualise uncertainty, it needs to be quantifiable [@Griethe2006; Leland2005], in order to quantify uncertainty, it needs to be mathematically defined.
-The uncertainty visualisation literature is completely awash with papers that define uncertainty using a vague encompassing phrase, (e.g.
-@utypo defined it as any deviation from complete determinism) but go on to *quantify* uncertainty as a PDF, error, or some other easily quantified mathematical object when it comes time to do the visualisation (*Cite Gap 4: Organise uncertainty quantification*).
-This definition swap happens so subtly that it seems to go unnoticed by authors writing the papers, and the fact that "uncertainty visualisation" methods don't even notice they are exclusively focus on depictions of easily quantifiable uncertainty, such as error [], mass [], or variance [] (*Cite Gap 4: Organise uncertainty quantification*).
-This has left the expressions of uncertainty that are hard to quantify or visually differentiate untouched, despite many papers calling for their invention (*Cite Gap 6*).
-There is likely only confusion and obfuscation if the vague definitions of "uncertainty" continue to be used and authors should try to be more specific with the mathematical object they are visualising, *or* a strict definition of uncertainty needs to be defined.
-The current method of titling papers and defining uncertainty forces readers to comb through the method of a particular visualisation experiment to understand if the uncertainty visualised is amenable to their needs.
--->
+Some works (*XXX eg*) focus more narrowly on specific terms defined mathematically, such as probability, variance, error, or precision.
+Others (*XXX eg*) include broader loosely related elements, such as missing values.
 Several attempts to address this gap are made in the papers describing taxonomies.
-<!-- Some papers have recognised these colloquial definitions are lacking and have made attempts to formalise the space with taxonomies, however these taxonomies come with their own issues. -->
 
 ## Definitions in taxonomies
-<!--Much in the same way that there are almost as many definitions of uncertainty as there are papers on the topic, the field is also over run with taxonomies.--> Taxonomies split uncertainty based on an endless stream of ever changing boundaries, such as whether the uncertainty is due to true randomness or a lack of knowledge [@Spiegelhalter2017], if the uncertainty is in the  attribute, spatial elements, or temporal element of the data [], whether the uncertainty is scientific (e.g.
+Taxonomies split uncertainty based on an endless stream of ever changing boundaries, such as whether the uncertainty is due to true randomness or a lack of knowledge [@Spiegelhalter2017], if the uncertainty is in the  attribute, spatial elements, or temporal element of the data [], whether the uncertainty is scientific (e.g.
 error) or human (e.g.
@@
 This paper has identified gaps in the uncertainty visualisation literature that could be filled to progress the field.
-<!--and with those issues, we have several suggestions for future work that would likely be fruitful for the field.-->
 
-<!--Our first suggestion for future work is--> *Each new development should be accompanied by a mathematical definition of the uncertainty being addressed.* Utilizing bias and variance to define uncertainty is well-established, but it is absent concerns about imputed data, assumptions, sampling methods, and even the human element of analyst decisions.
-<!--There is currently a lot of work that quantifies the uncertainty through bias and variance at various stages of the data analysis pipeline separately.
-Elements such as  imputed data, assumptions, sampling methods, and analyst decisions all have their own independent work, quantification and discussions, but methods to brand between these concepts are few and far between.--> A conceptual framework that allows us to combine these components is needed.
+*Each new development should be accompanied by a mathematical definition of the uncertainty being addressed.* Utilizing bias and variance to define uncertainty is well-established, but it is absent concerns about imputed data, assumptions, sampling methods, and even the human element of analyst decisions.
+A conceptual framework that allows us to combine these components is needed.
 
@@
 Other fields of science employ marginal changes when designing experiments to ensure it is well understood *what* aspect of their experiment is contributing to their results, and a better sense of what "marginal" is in the case of uncertainty visualisation would greatly help the field.
-
-XXX Is data pipeline connected with the grammar of graphics? Should this be a recommendation?
+(*XXX Is data pipeline connected with the grammar of graphics?
+Should this be a recommendation?*)
 
@@
 Regardless, these are directions of research would be fruitful to the uncertainty visualisation community even if it appears on the surface to be research that is only beneficial to the "normal" visualisation community.
-XXX Not sure what this paragraph is recommending?
+(*XXX Not sure what this paragraph is recommending?*)
 

### 2024-04-01 writingdumpster.qmd
removed old paragraphs after Di edit

@@
 # Removed Sections
+## Defining Uncertainty
+The absence of an encompassing definition of uncertainty is mentioned by every uncertainty visualisation directly [@Spiegelhalter2017; @Griethe2006] or indirectly by describing a myriad of ways it can be considered in the literature [@Kinkeldey2014; @Hullman2016], although it is never commented on as a source of the noise in the field.
+
+In order to visualise uncertainty, it needs to be quantifiable [@Griethe2006; Leland2005], in order to quantify uncertainty, it needs to be mathematically defined.
+The uncertainty visualisation literature is completely awash with papers that define uncertainty using a vague encompassing phrase, (e.g.
+@utypo defined it as any deviation from complete determinism) but go on to quantify uncertainty as a PDF, error, or some other easily quantified mathematical object when it comes time to do the visualisation.
+This definition swap happens so subtly that it seems to go unnoticed by authors

### 2024-04-03 writingdumpster.qmd
added place for new notes
added notes from @vanderplas2020

@@
 
+# New Notes
+## @Vanderplas2020
+
+1. Introduction
+- The development of graphs and charts has been linked to the development of coordinate systems and abstract representations of data
+- the aim of most experimental research in statistical graphics focuses on the initial perception and graph comprehension, very little work has been done to understand the effect of charts and graphs on higher cognitive processes such as learning or analysis
+1.1 Design of Statistical Graphs
+- improved tech allow charts to be made easily for exploratory purposes
+- consider purposes of presentation, entertainmen, analysis. Also considr a continum of utalitarian to artistry as a purpose.
+- infographic tells a story (has chart junk) vs a statistical graphic is one of intent (and any visual enhancements should contribute to that aim)
+- Points out that the ocean colour and state boundaries in the hurriance map only exist for geographic purposes. The map depicted here has the warnings information along with the cone.
+1.2 Statistics Mapping using a Grammar of Graphics
+- Systems that attempt to categorize charts based on their geometric representations generally make no effort to include all types of graphics, and they have difficulty accommodating charts that may fall into two or more categories.
+The classifications of graphics based on the underlying components and their relationships, as in the grammar of graphics developed byWilkinson, are more robust; they also provide an elegant framework for comparing different types of graphical representations separate from the underlying data structure.
+The former is like treating plots like creatures in a zoo, with a unique name for each, while the latter is analogous to having a phylogeny based on genetic data showing how plots are related.
+- Has a visual explanation of the grammar of graphics.
+- The grammar of graphics creates visual statistics (and conversely, allows plots to be considered statistics).
+The data plot can then be treated like other statistics: by imagining what the plot might look like in the absence of any structure, we can use the plot of the actual observed data to test for the likelihood of any perceived structure being significant.
+- Using the grammar ofgraphics, it is easy for experimenters to compare different types of charts using the same data, as the underlying structure of the graph remains the same. 
+- this approach to transformations and scales allows experimenters to easily test judgments made utilizing different axis transformations and color scales to compare perceptual accuracy
+
+2. Testing Methods
+- explicit = partiicpants answer specific questions
+- implicitly structured=participants must infer the question of interest from the provided stimuli
+2.1 Explicitly structured graphical tests
+- Preattentive perceptual effects are those that do not require sustained cognitive attention; they are processed automati- cally within the first 500 ms of viewing a chart or graph.
+Components processed preattentively include color and shape, as well as some basic information about coarse relationships between individual components
+- After the preattentive stage, attention is necessary for subsequent processing.
+Most of the insights we gain from charts and graphs are due to the cognitive processes that occur after attention is focused on specific aspects of the graph.
+2.1.1 Preattentive graph perception
+- Preattentively perceived features show a near-constant reaction time over increasing display size, while features that are processed attentively show an increasing reaction time with increased dis- play size
+- A primary question in the discussion of preattentive graph perception is whether there are advantages in designing a graph to promote the preattentive perception of features, ideally reducing cognitive load.
+- We distinguish between tests that use graphical forms and more primitive tests that use basic geometric elements during the testing process.
+The results from more primitive experimental designs still apply to the design ofgraphs and charts, but the experimental design does not involve any display of actual data
+- Preattentively processed features include shape, angle, size, and texture; however, combinations of preattentive features which represent separate features in the data are processed attentively, with at least one major exception hue and brightness are integrated, that is, that even though they can be separately manipulated, they are still perceived preattentively as a single unit.
+- Experiment on this topic are: HealeyCG,BoothKS, Enns JT.
+1996.
+High-speed visual estimation using preattentive processing; and Callaghan TC.
+1984.
+Dimensional interaction of hue and brightness in preattentive field segregation.
+Percept.
+2.1.2 Attention medicated testing methods
+2.1.2.1 Direct Observation: numerical estimation, speed, error rates
+- simplest ways to test the utility ofa graph is to verify that information can be accurately read from it
+- the participants are provided with a chart and asked to estimate some quantity or answer a predefined question using the information provided in the chart
+- It is important to ensure that the specific charts and questions used are aligned; studies are commonly critiqued on the basis that the charts or the questions were not appropriate for the task
+- Not always perceptual ranking.
+The point vs line example in figure 8 shows there is some cases where perceptual tasks do not perfectly differentiate?
+Or is it that the PDF you highlight is of importance.
+- open-ended estimation tasks elicit certain well-known biases such as the tendency to round to multiples of 5 or 10
+- To approach situations with more complicated graphics, or charts that are known to induce perceptual biases in the participants, consider using eye tracking to measure attention and motivation, or the use of verbal descriptions to assess more complicated graphs (since they facilitate specific parts of the perceptual process)
+2.1.2.2 Psychophysics and signal detection theory
+- Psychophysical experimental design is focused on whether an effect is detectable and whether the magnitude of the effect can be accurately estimated
+- Common methods, such as the method of constant stimuli and the method of adjustment, involve repeatedly presenting a participant with charts and asking them to evaluate the chart on the basis of a particular question of interest
+- In the method of adjustment, this is done with the control of the participant, who adjusts the stimuli interactively until the effect is just barely noticeable; in the method of constant stimuli, the effect size changes randomly from trial to trial to reduce continuity effects.
+- Psychophysics methods also seem to be relatively common in studies of map perception, particularly when the goal is to estimate the amount of exaggeration or other corrective distortion necessary for realistic perception of the map (for
+2.1.2.3 Thinking aloud
+- ask participants to talk through their thoughts as they read and use a graph in a realistic setting
+- allows experimenters to examine the use ofcom- plex graphics in the wild, or at least in situations that are less artificial than the paradigms allowed by numerical estimation and psychophysics methods
+- allow researchers to attempt to measure insight, and reasoning in complex situations such as experimental design, decision-making, or the process of weather fore- casting
+- While the data that result from the think-aloud protocol are typically more qualitative and less quantitative than results produced using other methods, they provide significant additional insight into the underlying cognitive processes affecting visualization, which cannot be obtained through other means
+2.1.2.4 Eye tracking
+- eye tracking facilitates insight into the process of visual attention, providing data on the approximate spatial location of visual focus
+- Eye tracking allows researchers to determine that viewers spend relatively little time examining the axes in scatter plots, but significant amounts of time examining the axes in parallel coordinates plots, suggesting that the process of reading these two chart types is fundamentally different
+- Can also be used to identify features that provide useful information during the graph reading process for several different types of charts
+2.1.2.5 Combination Experiemnts
+2.2 Implicit Graphical Tests Using Visual Inference
+- In an implicit graphical test, the user must identify both the purpose and function of the plot and use that information to evaluate the plot as shown
+- Explicit tests are typically conducted on plots that have been created to showcase specific structure in the data in order to present results; in contrast, implicit tests are designed to inform exploratory data analysis and the iterative model diagnostic process.
+- e.g. lineup plot
+- The advantage to implicit testing is that lineups do not require a specification of a feature of interest in the testing framework
+- Much ofthe historical research of comparing different types of charts has been criticized because the specific question phrasing does not provide readily generalizable results; the lineup protocol removes this obstacle by charging the user with the task of identifying the most different looking plot and thereby selecting the feature with the visually most salient difference compared with the other plots.
+- The real power ofthe lineup protocol is that when combined with the grammar of graphics, we can hold the underlying data and summary statistics constant, isolating the effect ofdifferent plot types, coordinate transformations, and aesthetic mappings on our ability to detect effects in the data.
+- It can be difficult to specify the null data-generating model in a way that adequately mimics the data plots, which suggests that visually, we are able to identify many more features than those typically tested using standard quantitative hypothesis tests.
+This implicit testing of many different hypotheses does make null distribution specification challenging, but it also highlights the power ofvisual cognition to detect subtle differences in data.
+
+3 Current Best Graphical Practice
+- no best plot can lead to a rather fragmented approach when describing best practice, and so in order to avoid this, we examine graphical practice using the principle of “first, do no harm” from the Hippocratic oath.
+3.1 Cognitive Principles
+- Take advantage of human visual system and allows us to make charts that require less cognitive effort to read
+3.1.1 Proximity
+- objects or shapes that are close to one another appear to form groups. For plot design, proximity is used to place items to compare close together, and less important comparisons further apart.
+3.1.2 Similarity
+- The gestalt principle of similarity suggests that we group things that have similar appearances and exclude objects with different appearances. 
+. Vanderplas & Hofmann (2017) showed that the addition of color and shape to a scatter plot increases the likelihood that individuals will perceive clustered groups of points.
+- the coloring of bars allows us to easily see that the sim- ilarly colored rectangles represent the same group of people, even though the bars are separated by facets and other groups.
+3.1.3 Common Region
+- The gestalt principle of common region suggests that elements contained within a common region belong together.
+- confidence bands and bounding ellipses also activate this gestalt principle by grouping points within the boundaries together (Vanderplas & Hofmann 2017), highlighting the presence of outliers that do not belong to the main group.
+3.1.4 Common Fate
+- The gestalt principle of common fate describes the tendency to group objects that are moving together in the same direction and at the same speed together
+- Common fate is certainly active in animated plots that use fading or transitions over time, but even in static plots, continuity can be activated when multiple time series plots are shown together (common fate creates bias in identifying correlation)
+- Strong negative association is not easily detected from overlaid line plots, but it is easily seen in a scatter plot (makes me think of that HOPS plot experiment)
+3.1.5 Working Memory
+- Typically, working memory is limited to approximately seven (plus or minus two) items, or chunks.
+In practice, this means that categorical scales with more than seven categories decrease readability, increase comprehension time, and require signifi- cant attentional resources, because it is not possible to hold the legend mapping in working memory.
+3.1.6 Change Blindness
+- change blindness, affects both static and interactive plots.
+In static plots, it can be difficult to compare between dif- ferent small multiples or facets because the contents of the plots are not reliably represented in workingmemorywhen switching attention between them.
+In animated plots, it is important to use transition effects to connect successive frames of the animation: This reduces change blindness and also activates the gestalt principle of common fate, allowing us to quickly identify groups of objects that are transitioning in the same direction
+3.1.7 Ease of Comparison
+- Much of the psychophysics research on statistical charts exam- ines the accuracy of comparisons and quantitative evaluations made during the process of under- standing a plot.
+-This ranking ofcognitive tasks provides some consistent guidance for chart design: If the same data can be represented in a way that allows the user to make a comparison more accurately (based on the hierarchy), then that design is preferable.
+- While the hierarchy of graphical comparisons provides some guidance, there are other design choices that can be informed by experimental research in a less systematic way.
+3.2 Chart Design
+- The use of redundant aesthetics that activate the same gestalt principles (such as color and shape in a scatter plot, which both activate similarity) results in higher identification ofcorresponding data features.
+- In addition, dual encoding increases the accessibility ofa chart to individuals who have impaired color vision or perceptual processing (e.g., dyslexia, dysgraphia).
+3.2.1 Colour
+- Cognitive load can also be reduced by selecting colors with cultural associations that match the data display, such as the use of blue for men and red (or pink) for women, or the use ofblue for cold temperatures and red/orange for warm temperatures.
+- It is also important to consider the human perceptual system, which does not perceive hues uniformly: We can distinguish more shades of green than any other hue, and fewer shades of yellow, so green univariate color schemes will provide finer discriminability than other colors be- cause the human perceptual system evolved to work in the natural world, where shades of green are plentiful.
+- the distance between points in a given color space may not be the same as the distance between points in perceptual space
+- Color-blindness is common enough that it is reasonable to expect that any given chart used in a presentation or publication will be read by someone with a color perception deficiency.
+The use of dual encoding allows color-blind individuals to more readily read graphics that utilize color, and as hue and lightness can be varied separately, it is possible to use dual encoding without adding another aesthetic.
+
+4. Open Questions and Future Research
+- Another open question is the acceptance of results—much historical research on best practices exists, but how much of it is being put into practice? One
+
+## @Tierney2023
+
 # General thoughts/questions

### 2024-04-04 writingdumpster.qmd
added dumpster notes

@@
 
+## @Pu2020
+
+## @Leland2005
+
 # General thoughts/questions
+- grammar of graphics describes what your visualisation *is* given the data you have.
+The reason the grammar of graphics is not equipt to handle uncertainty is the same reason the tidyverse is not equipt to handle it.
+It is generated and carries through between these independent steps.
+  - unclear how uncertainty from the source would make it in.
+  - Grammar of graphics describes what *is*. Uncertainty is what *might be*.
+  - Every uncertainty visualisation is entirely definable within the grammar of graphics.
+The *difficulty* with implementing uncertainty does not come from a failing within the grammar of graphics but from the absent definition of uncertainty.
+Just like normal statistics, visual statistics defined by the grammar of graphics *must* be well defined, if they are not, they cannot be implemented in the grammar.
 - Does the flip (hypothesis test vs uncertainty visualisation) have mathematical implications in how the plots should be assessed?

### 2024-04-09 paper.qmd
Added citations for 2 citation gaps

@@
 ## Definitions in taxonomies
-Taxonomies split uncertainty based on an endless stream of ever changing boundaries, such as whether the uncertainty is due to true randomness or a lack of knowledge [@Spiegelhalter2017], if the uncertainty is in the  attribute, spatial elements, or temporal element of the data [], whether the uncertainty is scientific (e.g.
+Taxonomies split uncertainty based on an endless stream of ever changing boundaries, such as whether the uncertainty is due to true randomness or a lack of knowledge [@Spiegelhalter2017; @Hullman2016; @utypo], if the uncertainty is in the  attribute, spatial elements, or temporal element of the data [@Kinkeldey2014], whether the uncertainty is scientific (e.g.
 error) or human (e.g.
-disagreement among parties) [], if the uncertainty is random or systematic [], statistical or bounded [], accuracy or precision [], if the uncertainty is about the past or the future [], the stage of the data analysis pipeline it comes from [], the severity of the uncertainty [], how quantifiable the uncertainty is [@Spiegelhalter2017], etc., etc., etc.
-(*Cite Gap 7: Uncertainty Taxonomies*).
+disagreement among parties) [@Benjamin2018], if the uncertainty is random or systematic [@Sanyal2009], statistical or bounded [@Gschwandtnei2016; @Olston2002], recorded as accuracy or precision [@Griethe2006; @Benjamin2018], which stage of the data analysis pipeline the uncertainty comes from [@utypo], how quantifiable the uncertainty is [@Spiegelhalter2017; @utypo], etc.
 In their own way, each of these taxonomies show an aspect of uncertainty that an author felt was important to differentiate.
@@
 @Meng2014 commented what is kept as data and what is tossed away is determined by the motivation of an analysis and what was previously noise can be shown to become signal depending on the the question we seek to answer.
+@Kale2019 discussed how the choices we make in our analysis impact our outcomes and introduce uncertainty.
 @Carlin2023 mentions that each research question can be can be categorised as descriptive, predictive, or causal, each of which has its own appropriate statistical methods and motivation agnostic model selection leads to statistical analysis that is devoid of meaning.
@@
 This appears in the literature as a mountain of seemingly conflicting statements about what is, or is not, an "uncertainty visualisation".
-For example @Leland2005 mentions that popular graphics, such as pie charts and bar charts omit uncertainty, however at least one or both of these charts are used in most "uncertainty visualisation" experiments [@Ibrekk1987] (*Cite Gap 8: Bar/pie uncertainty graphics*).
+For example @Leland2005 mentions that popular graphics, such as pie charts and bar charts omit uncertainty, however at least one or both of these charts are used in a significant number of uncertainty visualisation experiments [@Ibrekk1987; @Olston2002; @Zhao2023; @Hofmann2012].
 @Wickham2011 suggests their product plot framework, which includes histograms, should have a way to measure uncertainty, but does not consider that a histogram is *already* a depiction of mass and would already be considered an uncertainty visualisation were our statistic of interest the population mean.

### 2024-04-10 paper.qmd
started working on the uncert vis taxonomy visualisation

@@
 This consideration will become more relevant later in this review.
+
+```{r}
+#| echo: false
+#| eval: false
+#| warning: false
+#| message: false
+#| label: fig-vistax
+#| fig-cap: "Planned visualisation: Add an R coded visualisation that shows the 5 considerations for showing uncertainty in a plot according to @Kinkeldey2014.
+One that has the consideration, and one that doesn't (try and keep as much the same between visuals as possible)"
+# Library
+library(tidyverse)
+
+# Generate Data
+data_short <- tibble(x = rnorm(100, sd = 1),
+            y = rnorm(100, sd = 2))
+data_long <- data_short |>
+  pivot_longer(everything())
+
+# implicit 
+ggplot(data_short, aes(x=x, y="X")) + geom_jitter() 
+# explicit
+ggplot(data_short, aes(x=x)) + geom_density()
+
+# intrinsic
+# avg data
+avg_data <- data_long |> 
+  group_by(name) |>
+  summarise(mean = mean(value),
+            sd = sd(value))
+# plot
+ggplot(avg_data, aes(x=mean, y=name, colour=sd)) +
+  geom_point()
+
+# extrinsic
+ggplot(data_long, aes(x=value, y=name)) +
+  geom_violin() + 
+  geom_point(avg_data, aes(x=mean))
+
+# coincidence
+
+# adjacent
+
+# static
+
+# dynamic
+
+```
 

### 2024-04-10 writingdumpster.qmd
started working on the uncert vis taxonomy visualisation

@@
 # General thoughts/questions
+- Integrable vs separable
+  - kinsley did not mean it in the way that the integrable seperable paper means it. They basically meant colour on the same axis
+  - should read [this paper](https://dl.acm.org/doi/fullHtml/10.1145/3290605.3300899#:~:text=Two%20visual%20channels%20are%20considered,from%20the%20other%20%5B48%5D.) and others along the same lines.
 - grammar of graphics describes what your visualisation *is* given the data you have.

### 2024-04-16 paper.qmd
updated some plots

@@
 
-# implicit 
+# implicit + visually integral
 ggplot(data_short, aes(x=x, y="X")) + geom_jitter() 
@@
 
-# intrinsic
+# intrinsic + separable
 # avg data
@@
 # coincidence
+## Basically all above plots
 
 # adjacent
+avg_data |>
+  pivot_longer()
+ggplot(avg_data, aes(x=mean, y=name, colour=sd)) +
+  geom_point()
 

### 2024-04-17 paper.qmd
Handful of cuts made when working out what figures to put in each section

@@
 In their own way, each of these taxonomies show an aspect of uncertainty that an author felt was important to differentiate.
-@utypo identified many common themes among this wide array of taxonomies and used them to design an encapsulating taxonomy, depicted in @fig-taxonomy.
-This definition organises uncertainty based on three axis, location (where the uncertainty is coming from), level (how quantifiable it is), and nature (if it comes from true randomness or incomplete knowledge).
-
-![Depicts an illustration of the taxonomy described in @utypo.
-From right to left the drawing shows the location, level and nature of uncertainty with examples of that category underneath.
-A specific source of uncertainty from the location can be mapped to a level of ignorance that can increase or decrease (i.e.
-moving up or down the green line) depending on the nature of the uncertainty.
-Identifying the location, level and nature of your uncertainty allows you to better understand it.
-](taxonomyvis.jpeg){#fig-taxonomy}
-
-When considering the use of taxonomies to define something inherently vague, such as the uncertainty in an uncertainty visualisation, it is helpful to consider this quote from *The Grammar of Graphics*:
-
-> "Taxonomies are useful to scientists when they lead to new theory or stimulate insights into a problem that previous theorizing might conceal.
-Classification for its own sake, however, is as unproductive in design as it is in science.
-[@Leland2005]
-
-Therefore, while taxonomies can be helpful to map out the space of "things that we consider to be related to uncertainty", a taxonomy is *not* a definition and do little to help statisticians untangle and estimate the uncertainty in their projects.
-This is obvious in practice, as not understanding how to calculate uncertainty is one of the leading reasons cited by visualisation authors when explaining why they don't include it in their visualisation [@Hullman2020a].
-That being said, these taxonomies can hint towards what is important when we think about uncertainty.
-The location axis of the @utypo taxonomy lines up neatly with the typical data analysis pipeline and multiple authors in the uncertainty literature have commented on the need to consider quantifying and expressing uncertainty at every stage of a project [@Kinkeldey2014; @Hullman2016; @Refsgaard2007] including stages as early as conceptualising the problem [@Otsuka2023] or collecting the data [@Meng2021].
-The level axis from the @utypo taxonomy also hints towards an important consideration in uncertainty, because establishing how quantifiable uncertainty is informs us how it can be communicated, something that @Spiegelhalter2017 identified in the form of "precision".
-It is commonly noted that each source of uncertainty in an analysis must be discussed in isolation, but combining the uncertainty from every stage is near impossible [@Spiegelhalter2017].
-It is clear that elements of these taxonomies are identified in many other comments on uncertainty and represent *real* considerations that need to be made when defining uncertainty, even if that definition is yet to be published.
+@utypo identified many common themes among this wide array of taxonomies and used them to design an encapsulating taxonomy that was boldly referred to as a "definition of uncertainty".
+While taxonomies can be helpful to map out the space of "things that we consider to be related to uncertainty", a taxonomy is *not* a precise definition that would allow statisticians to estimate the uncertainty in their projects.
+The ramifications of this is seen throughout the literature, as not understanding how to calculate uncertainty is one of the leading reasons cited by visualisation authors when explaining why they don't include it in their visualisation [@Hullman2020a].
 
@@
 This philosophical understanding of statistics highlights that descriptive statistics do not have uncertainty, however some readers may disagree with that statement.
-Variance and probability are typically considered stand ins for "uncertainty", it is often how we choose to measure it, and since probability and variance exist in descriptive statistics, descriptive statistics *must* have uncertainty.
-This is not necessarily true, and related distinction was made by @Spiegelhalter2017  commented on a differentiation between precise random events (such as the probability of a coin flip), and uncertainty (such as the estimated probability associated with a coin that might be biased).
-A sample variance is not unknown, and therefore it is not uncertain, rather it is a precise description of dispersion.
-If we were to discuss drawing a new observation, or estimating the true mean of a population *then* the the variance would become relevant in our discussions of uncertainty, but in isolation it is not uncertain.
+Variance and probability are typically considered stand ins for "uncertainty", it is often how we choose to measure it.
+Additionally because probability and variance exist in descriptive statistics, descriptive statistics *must* have uncertainty.
+This is not necessarily true.
+@Begg2014 highlight that uncertainty is related to not knowing a specific value, while variability refers to the range of values a quantity can take at different locations, times or instances.
+Similar distinctions have been made by other authors, including @Spiegelhalter2017 who did not directly discuss variance vs uncertainty, but did differentiate between precise random events (such as the probability associated with a coin flip), and uncertainty (such as the estimated probability associated with a coin that might be biased).
+The variance of a sample variance can be calculated and know, therefore it is not uncertain but rather it a precise description of dispersion.
+If we were to discuss drawing a new observation, or estimating the true mean of a population *then* the the variance would become relevant in our discussions of uncertainty.
 
-The idea that inference is related to uncertainty pops up in most discussions of uncertainty, however, due to the now *implicit* understanding we are performing inferential statistics, it is often brought up as a *task* or *goal* dependence.
-This is mentioned both by authors at a specific stage of an analysis, and by authors looking at an entire field or at all stages of the analysis pipeline.
+The idea that inference is related to uncertainty is not uncommon in the uncertainty literature, however it is often mentioned as a *task* or *goal* dependence.
+As the "goal" shapes every stage of an analysis, multiple authors in the uncertainty literature have commented on the need to consider quantifying and expressing uncertainty at every stage of a project [@Kinkeldey2014; @Hullman2016; @Refsgaard2007] while others highlight the importance of specific stages.
 @Otsuka2023 suggested that the process of observing data to perform statistics is largely dependent on our goals, because the process of boiling real world entities down into probabilistic objects (or "probabilistic kind" as he puts it) depends on the relationship we seek to identify with our data.
@@
 @Wallsten1997 argue that the best method for evaluating or combining subjective probabilities depends on the uncertainty the decision maker wants to represent and why it matters.
-@Munzner2009 created a nested model for visualisation that highlighted how the first mistake that can be made in a visualisation is in the problem characterisation, and failing to do it well can cause downstream effects and damage the effectiveness of a visualisation.
 @Fischhoff2014 looks at uncertainty visualisation for decision making decides that we should have different ways of communicating uncertainty based off what the user is supposed to do with it.
-These examples show that authors that discuss uncertainty believe it is there is an important relationship between uncertainty and what we decide to keep in our analysis and throw away, i.e.
-the task or goal.
-The task or goal is, in essence, the *statistic you are drawing inference on*.
-Were this not true, we would have no reason to differentiate between a prediction interval and a sampling distribution because they could both be considered "uncertainty about the mean", however they are different because their inferential goals are different.
-
-These examples highlight two key concepts that seem to be true about the relationship between uncertainty and inference:    
-1) Uncertainty is the by-product of inference, as we seek to draw conclusions about something unknowable.    
-2) Uncertainty must be defined with respect to a *specific* statistic we wish to draw inference on, there is no such thing as a *general* uncertainty that is not linked to a motivating question.
+It is clear that the distinction between noise and signal is important at every stage in an analysis, however combining the uncertainty from every stage is near impossible [@Spiegelhalter2017].
+This means that uncertainty is often discussed in isolation at multiple stages of a project, which further contributes to our fractured understanding of uncertainty.
 
 With this understanding it becomes clear to see why uncertainty is tied to an endless string of examples in the data analysis pipeline.
-Uncertainty examples include imputed data, model selection, inherent randomness, biased sampling, etc, not because these things *are* uncertainty, but because they *create* uncertainty *when we perform inference*.
-These things are not "uncertainty" in of themselves, but rather contribute to the distance between the final estimate and the statistic we want to perform inference on.
-The field of uncertainty communication is in desperate need of a unifying mathematical definition that firmly identifies exactly how each of these things contribute to uncertainty.
-This may be a considerable task, but as of right now, the list of things that are "uncertainty" continues to grow, and the best methods of understanding and quantifying them are ad-hoc at best.
-@Thomson2005 suggests a mathematical formula for *examples* of uncertainty, @Meng2014 mathematically defined the variance introduced to a model by the array of model choices, information theory tries to quantify uncertainty using the idea of entropy.
-None of these existing methods are thorough enough for an analyst to understand what causes uncertainty, and quantify it for communication.
-The need for a mathematical definition of this field is put simply by Freeman Dyson in his famous Birds and Frogs speech:
-
-> "Rigorous theorems are the best way to give a subject intellectual depth and precision. Until you can prove rigorous theorems, you do not fully understand the meaning of your concepts." @Dyson2010
-
-In order to visualise something, it needs to be quantifiable.
-In order to quantify something, it needs to be mathematically defined.
-The broad classes of what is considered "uncertainty" is not currently quantifiable in any way that is not ad-hoc.
-The only aspects of uncertainty that are currently quantifiable are confidence intervals, prediction intervals, and related terms.
-Even then, these quantifications often only capture the uncertainty in our model or its assumptions, it ignores bias often introduced in earlier stages of the analysis.
-Broader and more complicated concepts such as the effects of assumptions, imputed missing variables, and model choices remain difficult to meaningfully quantify beyond ad-hoc methods.
+Uncertainty examples include imputed data, model selection, inherent randomness, biased sampling, etc, not because these things *are* uncertainty, but because they *create* uncertainty when we perform inference.
+Existing mathematical definition of uncertainty (or related concepts) have been unable to unify this range of diverse concepts.
+For example, @Thomson2005 suggests a mathematical formula for *examples* of uncertainty, @Meng2014 mathematically defined the variance introduced to a model by the array of model choices, and information theory tries to quantify uncertainty using the idea of entropy.
+These existing methods are not thorough enough for an analyst to understand what causes uncertainty, and quantify it for communication.
 
@@
 It is already established practice in medicine to use a wide range of metrics to communicate different risks to best communicate the relevant information to a patient for a particular decision [@Spiegelhalter2017].
-The perceived importance of this work by @Hofman2020 hinges on a misunderstanding that the prediction and confidence interval *are* interchangeable despite performing inference on *very different* statistics.
+The perceived importance of this work by @Hofman2020 hinges on a misunderstanding that the prediction and confidence interval *are* interchangeable despite performing inference on *very different* latent variables.
 This misunderstanding is so widespread that this finding, that shows the confidence and prediction intervals are *not* interchangeable, is considered groundbreaking enough to be published in CHI.
@@
 
-## Establishing misunderstood concepts
+## Definitions of "uncertainty visualisation"
 Before discussing the issues facing uncertainty visualisation due to a poor concept *of* uncertainty visualisation, we want to establish two ideas that should be considered in conjunction with the lacking definition of uncertainty.
@@
 We will spend the next section explaining why that is not the case by discussing several taxonomies of uncertainty visualisation.
-The second is a conflation of the contextual information with the visual information.
-This problem is not, in isolation an issue, however it does seem to be a *key* contributor to the misunderstanding that uncertainty is an attribute of data, and not a by product of inference.
-Both of these issues need to be understood in order for us to fully understand the issues surrounding uncertainty visualisation.
 
-### Definitions of "uncertainty visualisation"
+In order to visualise something, it needs to be quantifiable.
+In order to quantify something, it needs to be mathematically defined.
+The broad classes of what is considered "uncertainty" rarely have mathematical defintions, and therefore, cannot be meaningfully visualised.
+For these reasons simple quantifiable uncertainties, such as confidence intervals  dominate the literature, while visualisations for more complicated sources of uncertainty, such as the effects of assumptions, imputed missing variables, and model choices remain elusive.
+
 A large drive of the field of uncertainty visualisation, has been the idea that "uncertainty visualisation" (defined as one word) is somehow different to "uncertainty" "visualisation" (separately defined as two words).
@@
 For example @Leland2005 mentions that popular graphics, such as pie charts and bar charts omit uncertainty, however at least one or both of these charts are used in a significant number of uncertainty visualisation experiments [@Ibrekk1987; @Olston2002; @Zhao2023; @Hofmann2012].
-@Wickham2011 suggests their product plot framework, which includes histograms, should have a way to measure uncertainty, but does not consider that a histogram is *already* a depiction of mass and would already be considered an uncertainty visualisation were our statistic of interest the population mean.
+@Wickham2011 suggests their product plot framework, which includes histograms, should have a way to measure uncertainty, but does not consider that a histogram is *already* a depiction of PMF and would already be considered an uncertainty visualisation were our statistic of interest a new oberservation from our data set.
 These conflicting ideas and lacking definition of what is means for a visualisation to be an "uncertainty visualisation" is a cornerstone of some clearly questionable results within this field.
@@
 They also tend to blur the line between the mathematical elements of uncertainty and the visual depiction of those mathematical objects.
-While we will not focus on these taxonomies, it is still important to mention them and highlight what we can learn from their success' and failures..
+While we will not focus on these taxonomies, it is still important to mention them and highlight what we can learn from their success' and failures.
 
@@
 @Grewal2021 created a taxonomy that mapped uncertainty visualisations to some point in a 2D space defined by the "domain expertise" and "continuum of discreteness" (that scaled from "point estimate" to "continuous distribution").
-This conceptualisation could be considered a visual extension of the law of large numbers, because as a point estimate is a single sample, as the sample size gets larger the distribution will appear to be a continuous.
+This conceptualization could be considered a visual extension of the law of large numbers, because as a point estimate is a single sample, as the sample size gets larger the distribution will appear to be a continuous.
 However, assumes every point estimate is a sample from the same distribution (a mean is not technically a sample from a prediction interval, but it is a sample from the sampling distribution) and it conflates discreteness from a too small a number of observations, discreteness from a precision problem, discreteness because that is the true nature of the variable, and discreteness as a visualisation choice.
@@
 
-### Contextual Information
-One of the key components of uncertainty and inference is the noise and signal dichotomy.
-When @Otsuka2023 discusses the information we boil away to create a "probabilistic kind" he uses the example of a coin flip and the sex of baby births as two things that are *very* different in reality, but it is only through the lens of a Bernoulli distribution that we can view these two things as the same.
-The fact is that this boiling down, of a real world event into data, is not something that happens independent of the statisticians, but is just as much as product of our interests (the particular relationship we hope to draw inference on) as it is a product of the event's natural properties.
-This is expressed simply in the text with an extension to the coin flipping example:
-
-> "Recall that we modeled a coin toss by a Bernoulli distribution.
-That was because we supposed the coin will land either heads or tails and ignored the possibility of its landing on its edge.
-If we are to take the latter possibility into account, we should use a multinational distribution with three categories.
-As this example shows, the question of which statistical model we should use is not just determined by the physical nature of the object under study, but rather depends on the modeler’s interests and intentions." @Otsuka2023
-
-With this in mind, let us turn to the concept of "contextual" information.
-For the purposes of this review, we will consider contextual information to be anything that makes it easy for the audience to connect the data from a probabilistic kind to its real world parallel, but is largely irrelevant in the abstract probability space.
-A key example of this is spatial information, such as borders, road locations, terrains, etc.
-Spatial information is *highly relevant* when we translate from the abstract domain back into real world for communication.
-It can help provide reasons for *why* there may be specific characteristics in our data, however it is not required information for the model.
-@fig-datapipe shows the process of boiling information down into data, visualising it, and choosing whether or not to including contextual information to improve our scientific reasoning.
-The concept that contextual information is not the same as statistical information is also highlighted in *The Grammar of Graphics* which states:
-
-> "Geography is anchored in real space-time and statistics in abstract dimensions.
-This is a distinction along a continuum rather than a sharp break...
-but this difference in focus clearly means that a system optimized to handle geography will not be graceful when dealing with statistical graphics." @Leland2005
-
-::: {#fig-datapipe layout-ncol=2}
-
-![Australia example](spatialinfo.jpeg)
-
-![USA example](spatialinfo2.jpeg)
-
-Data pipeline with visualisations to show why we are ignoring contextual information in the case of uncertainty visualisation.
-The visualisations that depict the data are similar to typical statistical graphics.
-The Australia example depicts a heat map for temperature in Australia.
-The statistical level of this graphic is no different than a typical density plot that could be made with  exist with any data.
-The geographic information is merely used to provide real world context to the statistical message we extract from graphic, however that addition is of great importance.
-The USA example depicts a map showing the voting proportions in a US election.
-This map illustrates another example of a spatial graphic, however the location of the data is only loosely related to the real world geographic location.
-The graphic maintains readability because the proportion of votes matters as the US state level, and the number of people in a state and its proportion of votes are far more important for drawing inference on who won an election than specific spatial information.
-Notice: Both Graphic needs to be fixed up.
-:::
-
-For the remaining of this paper, we are not going to differentiate uncertainty visualisation using contextual information, despite the fact that it is common in the field.
-Our reasoning for this is that organising the field according to contextual information, promotes the idea that uncertainty is a concept specific to data qualities and ignores the role of inferential statistics.
-
-While other discussions of uncertainty may consider this distinction more relevant to their work, here we will consider this contextual information.
-Contextual information is important for interpretation and understanding of graphics, but it does not in of itself generate uncertainty.
-Its relevance to the field of "uncertainty visualisation" is through the perceived lack of available channels.
-For example, a choropleth map has already used colour to represent an estimate, and position to represent spatial information, which is sometimes relevant to the probabilistic kind, and other times it is simply contextual information (such as when the modelling is done by using a state as a category rather than its geographic information).
-These two cases, when the spatial information is model relevant and when it is contextual, are rarely separated.
-It can lead authors to believe the spatial information is always of the highest importance, and thus it is always given the position axis, even when it is not, especially with respect to the uncertainty in a model.
-A confidence interval is a confidence interval whether it came from spatial data, temporal, or cross sectional data.
-
-Similar to the confusing definition of uncertainty, the constant conflation of statistical and contextual information also creates a wide range of problems.
-@Kinkeldey2014 discussed how uncertainty  can be represented by three components attribute (what) position (where) and temporal (when) and that studied typically deal with uncertainty around attribute but rarely position and time, however it is never specified what considerations should cause attribute uncertainty to be different to position or temporal uncertainty.
-@Kay2016 did an experiment that showed uncertainty around bus arrival times, however the visualisation used in the experiment, shown in @fig-bus, is indistinguishable from most work that would be considered "attribute" uncertainty.
-
-![The uncertainty visualisation used by @Kay2016 to show the uncertainty around a bus arrival time prediction.
-The data being a "time" estimate does not prohibit the authors from using a density plot, a visualisation method that is typically used for attributes.
-The time element is irrelevant because the viewer is concerned about their estimated wait time NOW, not how this estimate will change over time.
-The predictions relationship to any predictors is assumed to be irrelevant in this visualisation.](bustime.png){#fig-bus width=50%}
-
-This paper focuses on uncertainty, specifically uncertainty that is still in the the abstract *statistics* domain.
-Spatial and temporal uncertainty papers limit how they communicate their uncertainty because they are considering what can be visualised inside the context provided by maps and, not what they should *try* to visualise based on the uncertainty abstraction.
-When we view a spatial-temporal visualisation through the lens of trying to visualise the distribution of $P(X|longitude, latitude, time)$ there is no reason this is any different from any other high dimensional visualisation problem.
-
-In this sense we are not proposing that the current conceptualisation of data according to its physical properties is necessarily wrong, or that it leads to conclusions that are not necessarily specific to spatial or temporal data, and it may lead us to miss relevant information that is outside our domain.
-This is not only the case for uncertainty data.
-@Slingsby2023 discussed using a gridded glyph map for spatial modelling, however the statistical elements are identical to a fluctuation diagram depicting $f(infection|time, age, longitude, latitude)$ presented in @Wickham2011.
-A side by side depiction of the two graphics is displayed in @fig-prodplots.
-The plot by @Slingsby2023 is simply a special case of the graphic by @Wickham2011, where a separate visualisation is made for each time point, the continuous function $infection | age$ has replaced the categorical variable $happy$, a fade is added for the population size, and a map is shown in the background.
-@Slingsby2023 presents other maps in the paper, however they are also special cases of the product plots framework.
-This is not to say this extension is uninteresting, on the contrary we believe identifying cases where particular graphics are useful is a worthwhile endeavour, but these plots seem to be have been developed unaware of the existing statistical framework.
-When we separate the field of visualisation along lines that may not be of particular relevance, we increase the likelihood of "reinventing the wheel" in each sub-discipline.
-
-::: {#fig-prodplots}
-
-![Gridded glyphmap](prodplots/img1.png)
-![Fluctuation diagram](prodplots/img2.png)
-
-A side by side comparison of the gridded glyphmap and the fluctuation diagram.
-One is considered a spatial visualisation technique and the other is considered a normal statistical diagram.
-The product plot and the gridded glyph map are visualisations depicting the relationship $f(w|x,y,z)$ and the similarity in the two plots shows that this three (or 4) variable relationship is the key feature of statistical importance, not the spatial nature of the data.
-The limitations faced by spatial visualisation are often identical to the limitations faces by information visualisation and creating unnecessary divides within the field makes it more difficult to navigate and find relevant statistical visualisations.
-:::
-
-The focus on contextual information has also created a rift in exactly *why* uncertainty visualisation is a field of interest. The two most common justifications are:
+## Why does the field of "uncertainty visualisation" exist at all?
+It is unclear exactly *why* uncertainty visualisation is a field of interest. The two most common justifications are:
 1) Uncertainty is fundamentally differently to other variables due to psychological heuristics involved in the interpretation of uncertainty (e.g. risk aversion).
@@
 
-For these reasons, the implication that uncertainty is not inference related, the barriers that cause authors to ignore existing visualisation methods, and the confusion created in the motivation of uncertainty visualization as a whole, we do not believe it is fruitful to continue separating the field of uncertainty visualisation according to its contextual information.
-Therefore, the rest of this review will not separate uncertainty visualisations based on this information.
+Uncertainty is noise, but most evaluation experiments measure it as signal.
+Uncertainty acts as a form of statistical "hedging" for signals found in an analysis.
+Interviews with experts in statistic back up this primary motivation, as ignoring uncertainty information of often expressed as being similar to fraud or lying [@Hullman2020a; @Manski2020].
+Noise and signal are not an inherent property of variance, or meta-data or the aspects of an analysis that are usually secondary considerations, it is a property of the data that *promotes* a message we are trying to infer and the data that *suppresses* it, if that message is invalid.
+When we ask the viewer of a plot to look at data and extract a value, we are asking them to perform inference on that value.
+There will be noise associated with that answer and that is uncertainty.
+If we ask direct question about some uncertainty metric, we have turned the uncertainty into signal because that is what the participants are drawing inference on.
+
+This is likely the cause of the contradiction between uncertainty visualisation papers following identical results to normal information visualisation papers, and authors believing uncertainty is inherently *different* to normal visualisation.
+Each of the sections in this paper likely contribute in some way to this contradiction in conceptualisation of uncertainty visualisation and the results of the studies in the field.
+Uncertainty is poorly defined, therefore authors do not understand it is not an inherent property of the data and is actually related to a particular inference question, this incorrect view that uncertainty is always variance or probabilities leads to the class of "uncertainty visualisations" that depict different mathematical objects or visual features with little in the sense of consistent rules, this complicated class has caused visualisation authors to miss that they are simply repeating the established evaluation results from normal information visualisations.
+This situation is a bit of a mess.
+
+If uncertainty is inherently different, we would expect evaluation studies of uncertainty visualisations to produce different results to similar evaluation studies on "normal" variables.
+This does not seem to be the case.
 
@@
 
-# Uncertainty is noise, not signal
-Uncertainty is noise, but most evaluation experiments measure it as signal.
-Uncertainty acts as a form of statistical "hedging" for signals found in an analysis.
-Interviews with experts in statistic back up this primary motivation, as ignoring uncertainty information of often expressed as being similar to fraud or lying [@Hullman2020a; @Manski2020].
-Noise and signal are not an inherent property of variance, or meta-data or the aspects of an analysis that are usually secondary considerations, it is a property of the data that *promotes* a message we are trying to infer and the data that *suppresses* it, if that message is invalid.
-When we ask the viewer of a plot to look at data and extract a value, we are asking them to perform inference on that value.
-There will be noise associated with that answer and that is uncertainty.
-If we ask direct question about some uncertainty metric, we have turned the uncertainty into signal because that is what the participants are drawing inference on.
-
-This is likely the cause of the contradiction between uncertainty visualisation papers following identical results to normal information visualisation papers, and authors believing uncertainty is inherently *different* to normal visualisation.
-Each of the sections in this paper likely contribute in some way to this contradiction in conceptualisation of uncertainty visualisation and the results of the studies in the field.
-Uncertainty is poorly defined, therefore authors do not understand it is not an inherent property of the data and is actually related to a particular inference question, this incorrect view that uncertainty is always variance or probabilities leads to the class of "uncertainty visualisations" that depict different mathematical objects or visual features with little in the sense of consistent rules, this complicated class has caused visualisation authors to miss that they are simply repeating the established evaluation results from normal information visualisations.
-This situation is a bit of a mess.
-It also highlights a unique and fascinating problem faced by uncertainty visualisation.
+# Issues with measuring uncertainty as noise, not signal
+The previous section highlights a unique and fascinating problem faced by uncertainty visualisation.
 If asking direct questions about uncertainty causes us to treat it as a signal, how do we evaluate uncertainty as *noise*?
 
-Uncertainty, acting as it is intended, is fundamentally transparency to the viewer about a statistical analysis.
+Uncertainty, acting as it is intended, is transparency.
 There are many secondary benefits that come with this improved transparency, such as better decisions, more trust in the results and more confidence in the authors.
-These secondary benefits are, however, *not* uncertainty.
+These secondary benefits are, however, *not* the immediate goal of uncertainty.
 The following sections will discuss the issues and limitations in measuring uncertainty through these secondary metrics and provide suggestions as to how future studies should consider measuring uncertainty.
@@
 
-*Each new development should be accompanied by a mathematical definition of the uncertainty being addressed.* Utilizing bias and variance to define uncertainty is well-established, but it is absent concerns about imputed data, assumptions, sampling methods, and even the human element of analyst decisions.
-A conceptual framework that allows us to combine these components is needed.
+*Each new development should be accompanied by a mathematical definition of the uncertainty being addressed.* Ideally, a mathematically definition of uncertainty that allows us to combine these components would be developed, but in the absence of that, authors should be more specific about what aspect of "uncertainty" they are covering with their visualisation.
 

### 2024-04-17 writingdumpster.qmd
Handful of cuts made when working out what figures to put in each section

@@
 # Removed Sections
-## Defining Uncertainty
+@Munzner2009 created a nested model for visualisation that highlighted how the first mistake that can be made in a visualisation is in the problem characterisation, and failing to do it well can cause downstream effects and damage the effectiveness of a visualisation.
+
 The absence of an encompassing definition of uncertainty is mentioned by every uncertainty visualisation directly [@Spiegelhalter2017; @Griethe2006] or indirectly by describing a myriad of ways it can be considered in the literature [@Kinkeldey2014; @Hullman2016], although it is never commented on as a source of the noise in the field.
@@
 This definition swap happens so subtly that it seems to go unnoticed by authors
+
+Utilizing bias and variance to define uncertainty is well-established, but it is absent concerns about imputed data, assumptions, sampling methods, and even the human element of analyst decisions

### 2024-04-22 paper.qmd
edits to the grammar of graphics section

@@
 
-Unlike uncertainty, the concept of the "information" in a plot is not entirely avoided and works such as *The Grammar of Graphics* attempt to outline exactly what information is contained within a plot.
-When creating a graphic there are several tasks that must be completed in a specific order, regardless of whether or not we are working with a uncertainty visualisation or not, this pipeline is illustrated in @fig-pipeline, and contains all the steps of the grammar of graphics [@Leland2005].
-First the mathematical information in a plot is established in the "varset" steps, and then that mathematical information is visually represented, which is defined in the "graph" steps.
+Works such as *The Grammar of Graphics* attempt to outline exactly what information is contained within a plot, and in doing so allow plots to be considered a statistic [@Vanderplas2020].
+This statistic that *The Grammar of Graphics* defines fundamentally a descriptive statistic.
+@fig-pipeline, and contains the set of necessary and sufficient steps that are required to make a graphic, as defined in *The Grammar of Graphics* [@Leland2005].
+It can be seen that the grammar assumes we have our data, performs a set of functions on that data, and produces a statistic.
+The grammar of graphics fundamentally describes the graphic as it is, not as it *would* be were our sample size to increase to infinity.
+As a matter of fact, the graphics assumption that we start from a data set that has an unchanging sample size is at odds with inference that typically involves assuming an infinitely large sample size.
+@fig-infintegraph shows that as n approaches infinity, two graphics that are very similar according to the *The Grammar of Graphics* can diverge into different graphics, while two graphics that are very different, can converge into the same graphic.
+*The Grammar of Graphics* itself does not recognize this issue, as uncertainty is referred to as "semantics" and is ultimately hand waved away.
+This does not mean describing a graphic using inferential statistics is completely unheard of.
+Some graphical descriptors, such as Product Plots [@Wickham2011] describe graphics using the underlying distribution, however the framework is not a substitute for *The Grammar of Graphics*.
+Other authors have noticed difficulties with inferential statistics and and created an addition to the grammar that should make depicting a distribution easier [@Pu2020; @Kay2023].
+Unfortunately, this addition falls short as it does not recognize *why* latent distributions are difficult to describe using *The Grammar of Graphics* and carries over many of the same problems.
+
+This does not mean *The Grammar of Graphics* should be abandoned when we visualize uncertainty, rather, these difficulties are something we should keep in mind when
+
+If uncertainty is fundamentally a by-product of inference, then the visualisations as described by *The Grammar of Graphics*
+
+Most uncertainty visualisation experiments are written with a seeming indifference to the grammar of Graphics, despite it being necessary to untangle experimental results.
+The primary reason for this seems to be the absent definition of uncertainty.
+Uncertainty is fundamentally related to inference and descriptive statistics cannot have uncertainty.
+
 Since two plots cannot be compared on their visual elements if they contain different information (because you will never be able to identify if the difference is because of informational asymmetry or visual processing) we could understand two plots as being the same if their "varset" steps are identical.
 However because uncertainty is not defined mathematically, we cannot reliably add it in at the "statistic" stage.
-Additionally, the grammar of graphics was not designed to extend to "uncertainty visualisations" as @Leland2005 referred to uncertainty as "semantics", so concepts that are common to uncertainty mathematically, such as resampling do not quite fit within the framework.
+, such as resampling do not quite fit within the framework.
 For example, if we have one set of data and two graphics, where one graphic represents the data as a series of points, and the the other depicts a smoothed density using a line.
@@
 the "varset" stages are the mathematical components of the graphic, the mathematical components define the "information" in the graph, while the "graph" stages are the visual elements that determine the "computational" cost of the graph.](grammarofgraphicspipeline.png){#fig-pipeline}
+
+ ![(infintiegraph](infinitegraph.png){#fig-infinitegraph}
 

### 2024-04-23 paper.qmd
added small comment in grammar of graphics section
addded some hand drawn graphics

@@
 
+@Tierney2023 generates a visualisation for missing values, however because it fits nicely into the grammar of graphics framework, it is inherrently a descriptive statistic.
+
 If uncertainty is fundamentally a by-product of inference, then the visualisations as described by *The Grammar of Graphics*
@@
 
-![(Notice: Need to replace with a more relevant data pipeline that includes all the uncertainty stuff and how uncertainty arises at each stage.) The grammar of graphics data analysis pipeline from @Leland2005.
-the "varset" stages are the mathematical components of the graphic, the mathematical components define the "information" in the graph, while the "graph" stages are the visual elements that determine the "computational" cost of the graph.](grammarofgraphicspipeline.png){#fig-pipeline}
+![(The grammar of graphics data analysis pipeline from @Leland2005 written in black, with the tidy data equivalent written in blue.
+The product plot framework and the grammar of probabilistic graphics frameworks are in green.
+It can be seen that the grammar of graphics provides a set of instructions that create a graphic, however the inferential power of this graphic is not considered and relies on many additional assumptions that are not defined by the grammar.](grammarofgraphicspipeline.jpeg){#fig-pipeline}
 
- ![(infintiegraph](infinitegraph.png){#fig-infinitegraph}
+ ![This graphic shows three simple graphics made from the same data set that diverge according to the grammar of graphics.
+Despite being distinct graphics from a stage as low down as the statistic (or data manipulation in the tidy data framework), as our sample size increases to infinity, the grahics converge to the same distribution and become visually indistinguishable.
+](infinitegraph.jpeg){#fig-infinitegraph}
 
@@
 #### Decision making and risk aversion
+
+![Visualisation showing the difference ](extractvsdecision.jpeg){#fig-extract}
+
 Decision making tasks are often described by authors as a more realistic version how plots are used in practice.
@@
 The reject or do not reject concepts in hypothesis testing do not offer a complete image of uncertainty, and exploration of uncertainty visualisation largely stems from a desire to move away from this binary framework.
+
+![The visualisation showing the kind of signal that we would expect an uncertainty plot to supress](signalsupression.jpeg){#fig-sigsupress}
+
 Additionally, setting up an uncertainty visualisation where the participants are expected to notice the signal once the data behind the visualisation passes a hypothesis test implies the signal *is* noticeable to a human at that level.
@@
 
-![Notice: I will fix this image up, it is currently a quick scribble.
-Difference between the null plot vs the uncertainty plot.
-A line-up protocol displays the uncertainty about the null and identifies if the true data plot is identifiable (and therefore significantly difference) while an uncertainty plot displays the variance of the estimated value and assesses if the null of "no signal" is within this plot (e.g.
-if the error bars overlap with zero).](hypvs.jpeg){#fig-hypvs}
+![A line-up protocol displays the uncertainty about the null and identifies if the true data plot is identifiable (and therefore significantly difference).
+This can be considered the graphical equivalent of a standard hypothesis test.
+Uncertainty is frequently used to descibe the area around the estimate that we think might contain the true value, and we assesses if the null of "no signal" is within this plot (e.g.
+if the error bars overlap with zero).
+Uncertainty is technically the distance between the true value and the estimate.
+](hypvs.jpeg){#fig-hypvs}
 

### 2024-04-26 paper.qmd
Moved over more hand drawn images

@@
 
-```{r}
-#| echo: false
-#| eval: false
-#| warning: false
-#| message: false
-#| label: fig-vistax
-#| fig-cap: "Planned visualisation: Add an R coded visualisation that shows the 5 considerations for showing uncertainty in a plot according to @Kinkeldey2014.
-One that has the consideration, and one that doesn't (try and keep as much the same between visuals as possible)"
-# Library
-library(tidyverse)
-
-# Generate Data
-data_short <- tibble(x = rnorm(100, sd = 1),
-            y = rnorm(100, sd = 2))
-data_long <- data_short |>
-  pivot_longer(everything())
-
-# implicit + visually integral
-ggplot(data_short, aes(x=x, y="X")) + geom_jitter() 
-# explicit
-ggplot(data_short, aes(x=x)) + geom_density()
-
-# intrinsic + separable
-# avg data
-avg_data <- data_long |> 
-  group_by(name) |>
-  summarise(mean = mean(value),
-            sd = sd(value))
-# plot
-ggplot(avg_data, aes(x=mean, y=name, colour=sd)) +
-  geom_point()
-
-# extrinsic
-ggplot(data_long, aes(x=value, y=name)) +
-  geom_violin() + 
-  geom_point(avg_data, aes(x=mean))
-
-# coincidence
-## Basically all above plots
-
-# adjacent
-avg_data |>
-  pivot_longer()
-ggplot(avg_data, aes(x=mean, y=name, colour=sd)) +
-  geom_point()
-
-# static
-
-# dynamic
-
-```
-
-![Planned visualisation: Add an R coded visualisation that shows the 5 considerations for showing uncertainty in a plot according to @Kinkeldey2014.
-One that has the consideration, and one that doesn't (try and keep as much the same between visuals as possible)](notice.jpeg){#fig-location width=50%}
+![Visualisation of @Kinkeldey2014 uncertainty visualisation taxonomy.
+It can be seen that most of the categories can be considered a change to the typical error bar chart.
+Many of the distinctions conflate changes of the input data and statistics calculated and with visual changes, highlighting how defining uncertainty and visualisation together conflates several elements of a graphic.](uncertvistax.jpeg){#fig-location width=70%}
 
@@
 
-![Visualisation showing the difference ](extractvsdecision.jpeg){#fig-extract}
+![Visualisation of the process of transforming raw data to a visualisation insight.
+Summarising the data using common statistics removes information from our data.
+Once the data is in a format that represents the varibales of interest, organising that information in a visualisation makes it easier to extract insights from these statistics and convert them to insights.
+If information is removed at the statistics stage, it cannot be added back in at the visualisation stage, as it is meerly an efficient technique to organise and present information to enable a large range of insights.
+](visprocess.jpeg){#fig-location width=70%}
 
@@
 .
+
+```{r}
+
+```
 

### 2024-04-26 paper.qmd
started doing the r code for the maps

@@
 
-![This shows the user interface for the experiment performed by @Hofmann2012. The question of "furthest to the right" is open to interpretation.  ](furthestright.png){#fig-right width=50%}
+![This shows the user interface for the experiment performed by @Hofmann2012. The question of "furthest to the right" is open to interpretation.](furthestright.png){#fig-right width=50%}
 
@@
 ```{r}
+#| eval: false
+#| echo: false
 
+# seed for sampling
+set.seed(1997)
+
+# load libraries
+library(strayr)
+library(tidyverse)
+library(biscale)
+
+# Get map data
+sa4 <- read_absmap("sa42021")
+state <- read_absmap("state2021")
+
+# Add Spatial trend
+state <- state |>
+  dplyr::mutate(temp = 30 - 0.001*(cent_long-110)^(2) - 0.003*(cent_lat-12.52294)^(2), # trend
+         notrend = rnorm(1), # no trend
+         highvar = runif(10, min=4, max=5), # high variance
+         lowvar = runif(10, min=0, max=1), # low variance
+         )
+
+sa4 <- sa4 |>
+  dplyr::mutate(temp = 30 - 0.001*(cent_long-123.7799)^(2) - 0.003*(cent_lat-12.52294)^(2), # trend
+         notrend = rnorm(1), # no trend
+         highvar = runif(1, min=4, max=5), # high variance
+         lowvar = runif(1, min=0, max=1), # low variance
+         )
+
+# Choropleth Map
+state %>%
+  ggplot() +
+  geom_sf(aes(fill = temp, geometry = geometry)) + 
+  scale_fill_distiller(type = "seq", palette = "YlOrRd", direction = 1)
+
+# Bivariate Map
+# Method 1
+d<-expand.grid(x=1:100,y=1:100)
+ggplot(d, aes(x,y,fill=atan(y/x),alpha=x+y))+
+  geom_tile()+
+  scale_fill_distiller(type = "seq", palette = "YlOrRd", direction = 1) +
+  theme(legend.position="none",
+        panel.background=element_blank())+
+  labs(title="A bivariate color scheme (Viridis)")
+
+state <- state %>%
+  dplyr::mutate(x=100*(temp-min(temp))/max(temp),
+         y=100*highvar/max(highvar))
+
+ggplot() +
+  geom_map(data =cmap, map = cmap,
+           aes(x = long, y = lat, map_id = id),
+           color = "#2b2b2b", size = 0.05, fill = NA) +
+  geom_map(data = filter(counties@data,year==2015), map = cmap,
+           aes(fill =atan(y/x),alpha=x+y, map_id = fips),
+           color = "gray50") +
+  #add black state borders (just to see if things are working)
+  geom_map(data = smap, map = smap,
+           aes(x = long, y = lat, map_id = id),
+           color = "black", size = .5, fill = NA) +
+  theme_map(base_size = 12) +
+  theme(plot.title=element_text(size = 16, face="bold",margin=margin(b=10))) +
+  theme(plot.subtitle=element_text(size = 14, margin=margin(b=-20))) +
+  theme(plot.caption=element_text(size = 9, margin=margin(t=-15),hjust=0)) +
+ # scale_fill_gradient(low="red",high="blue")
+  scale_fill_viridis()+guides(alpha=F,fill=F)+
+  labs(caption="@lenkiefer Source: U.S. Census Bureau",
+       title="Housing units and Population growth 2010-2015",
+       subtitle="Bivariate choropleth")
+
+# Method 2
+
+custom_pal <- c(
+  "1-1" = "#d3d3d3", # low x, low y
+  "2-1" = "#9e3547", 
+  "3-1" = , 
+  "4-1" = , # high x, low y
+  "1-2" = "#4279b0", # low x, high y
+  "2-2" = "#311e3b" # high x, high y
+)
+
+library(RColorBrewer)
+rev(brewer.pal(9, name = "YlOrRd"))
+names <- paste(rep(1:9, 9), "-" , rep(1:9, each=9), sep="")
+
+# Pixel Map
+# Sample Map
 ```

### 2024-04-27 paper.qmd
added bivariate pallette map (kind of) and attempted VSUP map

--- 2024-04-26 paper.qmd
@@
 ```{r}
+#| eval: false
+#| echo: false
 
+# seed for sampling
+set.seed(1997)
+
+# load libraries
+library(strayr)
+library(tidyverse)
+library(biscale)
+library(RColorBrewer)
+
+# Get map data
+#sa4 <- read_absmap("sa42021")
+state <- read_absmap("state2021") |>
+  na.omit() |>
+  dplyr::mutate(temp = 31 - 0.001*(cent_long-110)^(2) - 0.003*(cent_lat-12.52294)^(2), # trend
+         notrend = rnorm(1), # no trend
+         highvar = runif(9, min=2, max=4), # high variance
+         lowvar = runif(9, min=0, max=2), # low variance
+         ) |>
+  pivot_longer(cols=highvar:lowvar, names_to = "variance_class", values_to = "variance") |>
+  # add bivariate classes to data
+  mutate(bitemp = cut(temp, breaks=20:28, labels=seq(8)),
+         bivar = cut(variance, breaks=0:4, labels=seq(4)),
+         biclass = paste(bitemp, bivar, sep="-")
+  )
+
+#sa4 <- sa4 |>
+#  dplyr::mutate(temp = 30 - 0.001*(cent_long-123.7799)^(2) - 0.003*(cent_lat-12.52294)^(2), # trend
+#         notrend = rnorm(1), # no trend
+#         highvar = runif(1, min=4, max=5), # high variance
+#         lowvar = runif(1, min=0, max=1), # low variance
+#         )
+
+# Choropleth Map
+state %>%
+  ggplot() +
+  geom_sf(aes(fill = temp, geometry = geometry)) + 
+  scale_fill_distiller(type = "seq", palette = "YlOrRd", direction = 1)+
+  facet_wrap(vars(variance_class))
+
+# Bivariate Map
+# Make bivariate palette
+# Function to devalue by a certain amount
+colsupress <- function(basecols, hue=1, sat=1, val=1) {
+    X <- diag(c(hue, sat, val)) %*% rgb2hsv(col2rgb(basecols))
+    hsv(pmin(X[1,], 1), pmin(X[2,], 1), pmin(X[3,], 1))
+}
+
+# set Base colours
+basecols <- brewer.pal(8, name = "YlOrRd")
+# recurvisely decrease value
+v_val = 0.5
+bivariatepal <- c(basecols,
+                  colsupress(basecols, sat=v_val),
+                  colsupress(colsupress(basecols, sat=v_val), sat=v_val),
+                  colsupress(colsupress(colsupress(basecols, sat=v_val), sat=v_val), sat=v_val))
+# establish levels of palette
+names(bivariatepal) <- paste(rep(1:8, 4), "-" , rep(1:4, each=8), sep="")
+# preview palette
+show_col(bivariatepal, ncol=8)
+
+# Bivariate maps
+state |>
+  ggplot() +
+  geom_sf(aes(fill = biclass, geometry = geometry)) + 
+  scale_fill_manual(values = bivariatepal) +
+  facet_wrap(vars(variance_class))
+
+# VSUP
+# Function to combine hue values
+evenhueblend <- function(basecols, p_length, nblend) {
+    X <- rgb2hsv(col2rgb(unique(basecols)))
+    v1 <- X[,seq(1,dim(X)[2], 2)]
+    v2 <- X[,seq(2,dim(X)[2], 2)]
+    v3 <- (v1 + v2)
+    if("matrix" %in% class(v3)){
+      v3["h",] <- ifelse(v3["h",]>=1, v3["h",]-1 ,v3["h",])/2 # hue is on a circle so 0 wraps around to 1, need to do differently
+      v3 <- v3/2
+      nsup <- nblend
+      hsv(rep(v3[1,], each=nsup), rep(v3[2,], each=nsup), rep(v3[3,], each=nsup))
+      } else {
+        rep(hsv(h=v3[1], s=v3[2], v=v3[3]), p_length)
+        }
+}
+VSUPfunc <- function(basecols, p_length, nblend){
+  evenhueblend(colsupress(basecols, sat=0.5), p_length, nblend)
+}
+# VSUP
+p = length(basecols)
+VSUP <- c(basecols,
+          VSUPfunc(basecols, p, 2),
+          VSUPfunc(VSUPfunc(basecols, p, 2), p, 4),
+          VSUPfunc(VSUPfunc(VSUPfunc(basecols, p, 2), p, 4), p, 8))
+names(VSUP) <- paste(rep(1:8, 4), "-" , rep(1:4, each=8), sep="")
+show_col(VSUP, ncol=8)
+# Pixel Map
+# Sample Map
 ```

### 2024-04-29 paper.qmd
finished making maps (kind of) and decided I hate making them

--- 2024-04-26 paper.qmd
@@
 This statistic can then be expressed in a one dimensional colour space, which is a method adopted by the Bayesian surprise metric map [@Ndlovu2023] and the exceedance probability map [Lucchesi2017].
-.
 
 ```{r}
+#| echo: false
+#| message: false
+#| warning: false
+#| label: fig-maps
+#| fig-cap: "A visualisation of a typical choropleth map, as well as three other maps that are used to display uncertainty on the choropleth map.
+There is a high variance and low variance example for each type of map to show how well the technique supresses signal.
+Each map was created using the same data with the same base palette.
+At first glance, the high uncertainty bivariate and VSUP maps just look like maps with a low saturation colour palette, rather than map with high uncertainty.
+This visualisation makes it clear that supression methods that plot uncertainty to a second axis, such as hue, make uncertainty appear as a second variable, rather than a signal supression on our estimate.
+"
+#| fig-subcap: 
+#|   - "Choropleth Map: Low Variance"
+#|   - "Choropleth Map: High Variance"
+#|   - "Bivariate Map: Low Variance"
+#|   - "Bivariate Map: High Variance"
+#|   - "VSUP Map: Low Variance"
+#|   - "VSUP Map: High Variance"
+#|   - "Pixel Map: Low Variance"
+#|   - "Pixel Map: High Variance"
+#| layout-ncol: 2
 
+# seed for sampling
+set.seed(1997)
+
+# load libraries
+library(strayr)
+library(tidyverse)
+library(biscale)
+library(RColorBrewer)
+library(scales)
+library(sf)
+
+# Get map data
+#sa4 <- read_absmap("sa42021")
+state <- read_absmap("state2021") |>
+  na.omit() |>
+  dplyr::mutate(temp = 31 - 0.001*(cent_long-110)^(2) - 0.003*(cent_lat-12.52294)^(2), # trend
+         notrend = rnorm(1), # no trend
+         highvar = runif(9, min=2, max=4), # high variance
+         lowvar = runif(9, min=0, max=2), # low variance
+         ) |>
+  pivot_longer(cols=highvar:lowvar, names_to = "variance_class", values_to = "variance") |>
+  # add bivariate classes to data
+  mutate(bitemp = cut(temp, breaks=20:28, labels=seq(8)),
+         bivar = cut(variance, breaks=0:4, labels=seq(4)),
+         biclass = paste(bitemp, bivar, sep="-")
+  )
+
+#sa4 <- sa4 |>
+#  dplyr::mutate(temp = 30 - 0.001*(cent_long-123.7799)^(2) - 0.003*(cent_lat-12.52294)^(2), # trend
+#         notrend = rnorm(1), # no trend
+#         highvar = runif(1, min=4, max=5), # high variance
+#         lowvar = runif(1, min=0, max=1), # low variance
+#         )
+
+# Choropleth Map
+p1a <- state |>
+  filter(variance_class=="lowvar") %>%
+  ggplot() +
+  geom_sf(aes(fill = temp, geometry = geometry)) + 
+  scale_fill_distiller(type = "seq", 
+                       palette = "YlOrRd", 
+                       direction = 1)+
+  theme_minimal() + 
+  theme(legend.position = "none")
+
+p1b <- p1a %+% filter(state, variance_class=="highvar")
+# Bivariate Map
+# Make bivariate palette
+# Function to devalue by a certain amount
+colsupress <- function(basecols, hue=1, sat=1, val=1) {
+    X <- diag(c(hue, sat, val)) %*% rgb2hsv(col2rgb(basecols))
+    hsv(pmin(X[1,], 1), pmin(X[2,], 1), pmin(X[3,], 1))
+}
+
+# set Base colours
+basecols <- brewer.pal(8, name = "YlOrRd")
+# recurvisely decrease value
+v_val = 0.5
+bivariatepal <- c(basecols,
+                  colsupress(basecols, sat=v_val),
+                  colsupress(colsupress(basecols, sat=v_val), sat=v_val),
+                  colsupress(colsupress(colsupress(basecols, sat=v_val), sat=v_val), sat=v_val))
+# establish levels of palette
+names(bivariatepal) <- paste(rep(1:8, 4), "-" , rep(1:4, each=8), sep="")
+# preview palette
+#show_col(bivariatepal, ncol=8)
+
+# Bivariate maps
+p2a <- state |>
+  filter(variance_class=="lowvar") %>%
+  ggplot() +
+  geom_sf(aes(fill = biclass, geometry = geometry)) + 
+  scale_fill_manual(values = bivariatepal) +
+  theme_minimal() + 
+  theme(legend.position = "none")
+
+p2b <- p2a %+% filter(state, variance_class=="highvar")
+
+# VSUP
+# Function to combine colours for VSUP
+colourblend <- function(basecols, p_length, nblend) {
+    X <- rgb2hsv(col2rgb(unique(basecols)))
+    v1 <- X[,seq(1,dim(X)[2], 2)]
+    v2 <- X[,seq(2,dim(X)[2], 2)]
+    if("matrix" %in% class(v1)){
+      # hue issue wrap around pt 1
+      v3 <- (v1+v2)
+      v3["h",] <- ifelse(abs(v1["h",]-v2["h",])>0.5, v3["h",]+1, v3["h",])
+      v3 <- v3/2
+      # hue issue wrap around pt 2
+      v3["h",] <- ifelse(v3["h",]>=1 , v3["h",]-1 ,v3["h",])
+      hsv(rep(v3[1,], each=nblend), rep(v3[2,], each=nblend), rep(v3[3,], each=nblend))
+      } else {
+        v3 <- (v1+v2)
+        v3["h"] <- ifelse(abs(v1["h"]-v2["h"])>0.5, v3["h"]+1, v3["h"])
+        v3 <- v3/2
+        v3["h"] <- ifelse(v3["h"]>=1 , v3["h"]-1 ,v3["h"])
+        rep(hsv(h=v3[1], s=v3[2], v=v3[3]), p_length)
+        }
+}
+
+VSUPfunc <- function(basecols, p_length, nblend){
+  colourblend(colsupress(basecols, sat=0.5), p_length, nblend)
+}
+
+# VSUP
+p = length(basecols)
+VSUP <- c(basecols,
+          VSUPfunc(basecols, p, 2),
+          VSUPfunc(VSUPfunc(basecols, p, 2), p, 4),
+          VSUPfunc(VSUPfunc(VSUPfunc(basecols, p, 2), p, 4), p, 8))
+
+names(VSUP) <- paste(rep(1:8, 4), "-" , rep(1:4, each=8), sep="")
+#show_col(VSUP, ncol=8)
+
+# VSUP maps
+p3a <- state |>
+  filter(variance_class=="lowvar") %>%
+  ggplot() +
+  geom_sf(aes(fill = biclass, geometry = geometry)) + 
+  scale_fill_manual(values = VSUP) +
+  theme_minimal() + 
+  theme(legend.position = "none")
+
+p3b <- p3a %+% filter(state, variance_class=="highvar")
+
+# Maps I haven't been able to get working yet
+
+# Pixel Map
+library(Vizumap)
+# Low variance map
+state_a <- state %>%
+  filter(variance_class == "lowvar") %>%
+  mutate(my_id = seq(9),
+         error = variance)
+
+pixel_1a <- state_a %>%
+  as.data.frame() %>%
+  select(my_id, temp, error) %>%
+  read.uv(estimate="temp", error="error")
+
+pixel_2a <- state_a %>% as("Spatial")
+
+pix_a <- pixelate(pixel_2a, pixelSize = 3, id = "my_id")
+
+pmap_a <- build_pmap(data = pixel_1a, distribution = "normal", pixelGeo = pix_a, id = "my_id", border = pixel_2a)
+
+mypixelmap <- function(obj){
+
+  if(class(obj)[1] != "pixelmap")
+    stop("Object is not of class 'pixelmap'.")
+
+  p <- ggplot()  +
+    geom_polygon(
+      data = obj$output_data,
+      aes_string(
+        x = 'long',
+        y = 'lat',
+        fill = 'values',
+        colour = 'values',
+        group = 'group'
+      )
+    )  +
+    scale_fill_distiller(
+      type = "seq", 
+      palette = "YlOrRd", 
+      direction = 1,
+      name = obj$key_label,
+      limits = obj$limits) +
+
+    scale_colour_distiller(
+      type = "seq", 
+      palette = "YlOrRd", 
+      direction = 1,
+      limits = obj$limits
+    ) +
+    guides(colour = "none")  +
+    geom_path(
+      data = obj$bord,
+      aes_string(x = 'long', y = 'lat', group = 'group'),
+      colour = "black"
+    )  +
+    coord_fixed(xlim = c(obj$bbox[1], obj$bbox[3]),
+                    ylim = c(obj$bbox[2], obj$bbox[4])) +
+    theme(
+      axis.line = element_blank(),
+      axis.text.x = element_blank(),
+      axis.text.y = element_blank(),
+      axis.ticks = element_blank(),
+      axis.title.x = element_blank(),
+      axis.title.y = element_blank(),
+      panel.background = element_blank()
+    )
+
+  p
+
+}
+
+p4a <- mypixelmap(pmap_a) + theme(legend.position="none")
+
+# High variance
+state_b <- state %>%
+  filter(variance_class == "highvar") %>%
+  mutate(my_id = seq(9),
+         error = variance)
+
+pixel_1b <- state_b %>%
+  as.data.frame() %>%
+  select(my_id, temp, error) %>%
+  read.uv(estimate="temp", error="error")
+
+pixel_2b <- state_b %>% 
+  as("Spatial")
+
+pix_b <- pixelate(pixel_2b, pixelSize = 3, id = "my_id")
+
+pmap_b <- build_pmap(data = pixel_1b, distribution = "normal", pixelGeo = pix_b, id = "my_id", border = pixel_2b)
+
+p4b <- mypixelmap(pmap_b) + theme(legend.position="none")
+
+p1a
+p1b
+p2a
+p2b
+p3a
+p3b
+p4a
+p4b
 ```
-
-::: {#fig-maps layout-ncol=2}
-
-![Choropleth map](mappals/img1.png)
-
-![Bivariate map](mappals/img2.png)
-
-![VSUP map](mappals/img4.png)
-
-![Bayesian surprise map](mappals/img3.png)
-
-![Exceedance probability map](mappals/img5.png)
-
-Five different ways to include error information in a map.
-A typical choropleth map is considered the map that ignores uncertainty and simply visualises the signal.
-The bivariate map visualises the uncertainty, but does so using two different visual channels, which gives the perception of two different signals (or one signal in the assumptions the colours contrast with each other).
-An adjustment to the bivariate map is the VSUP which is similar to the bivariate plot except it merges high uncertainty values in the palette to reduce our ability to discern values that have a high enough error to not be a valid signal.
-The Bayesian surprise metric is similar to the line-up protocol, where the colour value is the difference between the observers posterior and prior beliefs.
-The exceedance probability map is a direct visualisation of the probability of a specific hypothesis.
-Notice: This plot is the highest priority to for replacing with my own R coded maps, because the effects of these different mappings is really not visible using these methods.
-:::
 

### 2024-05-01 paper.qmd
adjusted pixelmap code

@@
 # preview palette
-#show_col(bivariatepal, ncol=8)
+show_col(bivariatepal, ncol=8)
 
@@
 names(VSUP) <- paste(rep(1:8, 4), "-" , rep(1:4, each=8), sep="")
-#show_col(VSUP, ncol=8)
+show_col(VSUP, ncol=8)
 
@@
 
-mypixelmap <- function(obj){
-
-  if(class(obj)[1] != "pixelmap")
-    stop("Object is not of class 'pixelmap'.")
-
-  p <- ggplot()  +
-    geom_polygon(
-      data = obj$output_data,
-      aes_string(
-        x = 'long',
-        y = 'lat',
-        fill = 'values',
-        colour = 'values',
-        group = 'group'
-      )
-    )  +
+p4b <- view(pmap_a) +
     scale_fill_distiller(
@@
       palette = "YlOrRd", 
-      direction = 1,
-      name = obj$key_label,
-      limits = obj$limits) +
-
+      direction = 1) +
   scale_colour_distiller(
@@
       palette = "YlOrRd", 
-      direction = 1,
-      limits = obj$limits
-    ) +
-    guides(colour = "none")  +
-    geom_path(
-      data = obj$bord,
-      aes_string(x = 'long', y = 'lat', group = 'group'),
-      colour = "black"
-    )  +
-    coord_fixed(xlim = c(obj$bbox[1], obj$bbox[3]),
-                    ylim = c(obj$bbox[2], obj$bbox[4])) +
-    theme(
-      axis.line = element_blank(),
-      axis.text.x = element_blank(),
-      axis.text.y = element_blank(),
-      axis.ticks = element_blank(),
-      axis.title.x = element_blank(),
-      axis.title.y = element_blank(),
-      panel.background = element_blank()
-    )
-
-  p
-
-}
-
-p4a <- mypixelmap(pmap_a) + theme(legend.position="none")
+      direction = 1) +
+  theme(legend.position="none")
 
@@
 
-p4b <- mypixelmap(pmap_b) + theme(legend.position="none")
+p4b <- view(pmap_b) +
+    scale_fill_distiller(
+      type = "seq", 
+      palette = "YlOrRd", 
+      direction = 1) +
+  scale_colour_distiller(
+      type = "seq", 
+      palette = "YlOrRd", 
+      direction = 1) +
+  theme(legend.position="none")
 

### 2024-05-01 paper.qmd
small edit + figures for background

@@
 
+```{r}
+#| echo: false
+#| message: false
+#| warning: false
+
+# load Libraries
+library(strayr)
+library(tidyverse)
+library(biscale)
+library(RColorBrewer)
+library(scales)
+library(sf)
+
+```
 # Background
@@
 From entertainment choices to news articles to insurance plans, the modern citizen is so over run with information in every aspect of their life it can be overwhelming.
-In this overflow of information, tools that can effectively summarise information down into simple and clear ideas become more valuable.
+In this overflow of information, tools that can effectively summarize information down into simple and clear ideas become more valuable.
 Information visualisations remain one of the most powerful tools for fast and reliable science communication.
@@
 There are many stages in our analysis that might benefit from the power of data visualisation, however this does not mean it is appropriately utilized or done with with effectiveness.
-(@Vanderplas2020 has helpful advice here.) Visualization is an important step in exploratory data analysis and it is often utilised to **learn** what is important about a data set.
-The importance of data-driven discovery has been highlighted through the data sets Anscombe's quartet [@anscombe] and the Datasaurus Dozen [@datasaurpkg].
-Each of the pairwise plots in these data sets have the same summary statistics but strikingly different information, as revealed when visualised.
-Visualisations can **tell** us what is important about our data and **how** it might diverge from pre-conceived notions or what we expect.
+Unlike an infographic, a statistical graphic is one of intent, and all enhancements made on a statistical graphic should be in pursuit of that goal (@Vanderplas2020).
+Visualization is an important step in exploratory data analysis and communication.
+At the exploration stage, it is utilized to **learn** what is important about a data set while the communication stage is utilized to **tell** others about what we have learned.
+Each of these visualisation goals utilize an important attribute of visualisation.
+In both these cases, visualisations are powerful tools that can help us understand **how** our data or results might diverge from pre-conceived notions or what we expect.
 
@@
 
+```{r}
+#| echo: false
+#| message: false
+# Generate data
+ib_data <- tibble(x=rnorm(1000, 8, 4))
+
+# Picture 1
+p1 <- ib_data %>%
+  summarise(avg = mean(x),
+          conf_95a = quantile(x, probs=c(0.025)),
+          conf_95b = quantile(x, probs=c(0.975))) %>%
+  ggplot(aes(y="NA")) +
+  geom_point(aes(x=avg)) +
+  geom_errorbar(aes(xmin = conf_95a, xmax = conf_95b), width = 0.1) +
+  scale_x_continuous(name = "INCHES OF SNOW",
+                     breaks=seq(0,19),
+                     labels= ggplot2:::interleave(as.character(c(seq(0,18, 2), 19)), rep("", 11))[c(0:19, 21)],
+                     limits=c(0,19)) +
+  theme_classic() +
+  theme(axis.line.y=element_blank(),
+        axis.title.y = element_blank(),
+        axis.text.y = element_blank(),
+        axis.ticks.y = element_blank(),
+        aspect.ratio=1/10)
+
+# Picture 2
+# Picture 3
+# Picture 4
+# Picture 5
+# Picture 6
+# Picture 7
+# Picture 8
+# Picture 9
+
+```
 However, uncertainty visualisation takes many forms today, It can be considered to expand the application of data visualisation towards making better decisions, and research is concerned about how the construction of plots and information contained might affect the resulting decisions.
@@
 
-# load libraries
-library(strayr)
-library(tidyverse)
-library(biscale)
-library(RColorBrewer)
-library(scales)
-library(sf)
-
 # Get map data

### 2024-05-01 writingdumpster.qmd
small edit + figures for background

@@
 - Systems that attempt to categorize charts based on their geometric representations generally make no effort to include all types of graphics, and they have difficulty accommodating charts that may fall into two or more categories.
-The classifications of graphics based on the underlying components and their relationships, as in the grammar of graphics developed byWilkinson, are more robust; they also provide an elegant framework for comparing different types of graphical representations separate from the underlying data structure.
+The classifications of graphics based on the underlying components and their relationships, as in the grammar of graphics developed by Wilkinson, are more robust; they also provide an elegant framework for comparing different types of graphical representations separate from the underlying data structure.
 The former is like treating plots like creatures in a zoo, with a unique name for each, while the latter is analogous to having a phylogeny based on genetic data showing how plots are related.

### 2024-05-02 paper.qmd
worked on figure 1

@@
 #| message: false
+#| label: fig-maps
+#| fig-cap: "A replication of the the uncertainty visualisations shown by @Ibrekk1987 in one of the earliest uncertainty visualisation experiments.
+This early experiment is a good example of many of the issues that are still common in uncertainty visualisation today.
+For example, the '95% confidence interval' is more accurately a '95% prediction interval'.
+Additionally graphics that depict different mathematical objects that are also different on their visual components are compared because of a percieved relation to uncertainty.
+So visualisations of the mean, PDF and CDF, are all discussed as though they all contain the relevant statistical information.
+The axis also have different scales, and visualisation methods that are now unpopular for displaying proportions, such as a pie chart, are used."
+#| fig-subcap: 
+#|   - "Mean with a 95% prediction interval"
+#|   - "Picture 2"
+#|   - "Picture 3"
+#|   - "Picture 4"
+#|   - "Picture 5"
+#|   - "Picture 6"
+#|   - "Picture 7"
+#|   - "Picture 8"
+#|   - "Picture 9"
+#| layout-ncol: 3
 # Generate data
-ib_data <- tibble(x=rnorm(1000, 8, 4))
-
+set.seed(6)
+ib_data <- tibble(x=c(rpois(500, 8)) 
+ib_data %>%
+  ggplot(aes(x=x)) +
+  geom_histogram(fill="black", col="white") +
+  theme_classic()
 # Picture 1
@@
 # Picture 2
+p2 <- ib_data |>
+  mutate(x = ifelse(x>18, 18, x),
+         binx = cut(x, breaks=seq(0,18,2))) |>
+  group_by(binx) |>
+  summarise(n = n()) |>
+  mutate(Probability = n / sum(n)) |>
+  ggplot(aes(x=binx, y=Probability)) +
+  geom_col(fill="black", colour="white") +
+  scale_x_discrete(name = "INCHES OF SNOW",
+                   labels= paste0(seq(0,16,2), sep = "-", seq(2,18,2))) +
+  scale_y_continuous(breaks = seq(0.00, 0.25, 0.05)) +
+  theme_classic() + 
+  theme(aspect.ratio=0.33)
+
 # Picture 3
+library(ggpattern)
+p3 <- ib_data |>
+  mutate(x = ifelse(x>18, 18, x),
+         binx = cut(x, 
+                     breaks=seq(0,18,2), 
+                     labels= paste0(seq(0,16,2), sep = "-", seq(2,18,2)))) |>
+  group_by(binx) |>
+  summarise(n = n()) |>
+  mutate(Probability = n / sum(n),
+         texture_groups = ) |>
+  ggplot(aes(x="", y=Probability)) +
+  geom_bar_pattern(aes(pattern=binx, pattern_angle = binx, pattern_spacing = binx),
+                   pattern = 'stripe',
+                   fill    = 'white',
+                   colour  = 'black',
+                   stat="identity", width=1) +
+  scale_pattern_discrete(choices = c("none", "stripe", "circle")) +
+  coord_polar("y", start=0) +
+  labs(fill = "INCHES OF SNOW") + 
+  theme_void()
+
 # Picture 4
@@
 # Picture 9
-
+# Display Plots
+p1
+p2
+p3
+p4
+p5
+p6
+p7
+p8
+p9
 ```

### 2024-05-06 paper.qmd
replicate ibrek figure up to picture 8

--- 2024-05-01 paper.qmd
@@
 library(sf)
+library(ggrepel)
 
@@
 #| message: false
+#| label: fig-maps
+#| fig-cap: "A replication of the the uncertainty visualisations shown by @Ibrekk1987 in one of the earliest uncertainty visualisation experiments.
+This early experiment is a good example of many of the issues that are still common in uncertainty visualisation today.
+For example, the '95% confidence interval' is more accurately a '95% prediction interval'.
+Additionally graphics that depict different mathematical objects that are also different on their visual components are compared because of a percieved relation to uncertainty.
+So visualisations of the mean, PDF and CDF, are all discussed as though they all contain the relevant statistical information.
+The axis also have different scales, and visualisation methods that are now unpopular for displaying proportions, such as a pie chart, are used."
+#| fig-subcap: 
+#|   - "Mean with a 95% prediction interval"
+#|   - "Picture 2"
+#|   - "Picture 3"
+#|   - "Picture 4"
+#|   - "Picture 5"
+#|   - "Picture 6"
+#|   - "Picture 7"
+#|   - "Picture 8"
+#|   - "Picture 9"
+#| layout-ncol: 3
+
 # Generate data
-ib_data <- tibble(x=rnorm(1000, 8, 4))
+set.seed(1)
+x=rnorm(1000, 8, 4)
+ib_data <- tibble(x=ifelse(x<0, -x, x))
+ib_data %>%
+  ggplot(aes(x=x)) + geom_histogram()
 
@@
 # Picture 2
+p2 <- ib_data |>
+  mutate(x = ifelse(x>18, 18, x),
+         binx = cut(x, breaks=seq(0,18,2))) |>
+  group_by(binx) |>
+  summarise(n = n()) |>
+  mutate(Probability = n / sum(n)) |>
+  ggplot(aes(x=binx, y=Probability)) +
+  geom_col(fill="black", colour="white") +
+  scale_x_discrete(name = "INCHES OF SNOW",
+                   labels= paste0(seq(0,16,2), sep = "-", seq(2,18,2))) +
+  scale_y_continuous(breaks = seq(0.00, 0.25, 0.05)) +
+  theme_classic() + 
+  theme(aspect.ratio=0.33)
+
 # Picture 3
+# library(ggpattern)
+p3 <- ib_data |>
+  mutate(x = ifelse(x>18, 18, x),
+         binx = cut(x, 
+                     breaks=seq(0,18,2), 
+                     labels= paste0(seq(0,16,2), sep = "-", seq(2,18,2)))) |>
+  group_by(binx) |>
+  summarise(n = n()) |>
+  mutate(Probability = n / sum(n),
+         csum = rev(cumsum(rev(Probability))), 
+         pos = Probability/2 + lead(csum, 1),
+         pos = if_else(is.na(pos), Probability/2, pos)) |>
+  ggplot(aes(x="", y=Probability, fill=binx)) +
+  geom_bar(stat="identity", width=1) +
+  geom_text_repel(aes(y = pos, label = paste0(round(Probability*100), sep="", "%")),
+                   size = 3, nudge_x = 0.6, show.legend = FALSE, segment.color = 'transparent') +
+  #geom_label(aes(label = paste0(round(Probability*100), sep="", "%")),
+  #          position = position_stack(vjust = 0.5)) + 
+  scale_fill_grey() +
+  coord_polar("y", start=0) +
+  labs(fill = "INCHES OF SNOW") + 
+  theme_void() + 
+  theme(aspect.ratio=1)
+
 # Picture 4
+p4 <- ib_data |>
+  ggplot(aes(x=x)) +
+  geom_density() + 
+  scale_x_continuous(name = "INCHES OF SNOW",
+                     breaks = seq(0,20,2),
+                     labels= paste0(seq(0,20,2))) +
+  scale_y_continuous(name = "Probability density",
+                     breaks = seq(0.00, 0.20, 0.02)) +
+  theme_classic() + 
+  theme(aspect.ratio=0.33)
+
 # Picture 5
+p5 <- ib_data |>
+  ggplot(aes(y="", x=x)) +
+  geom_violin() + 
+  scale_x_continuous(name = "INCHES OF SNOW",
+                     breaks = seq(0,20,2),
+                     labels= paste0(seq(0,20,2)),
+                     limits=c(0,20)) +
+  theme_classic() + 
+  theme(axis.line.y=element_blank(),
+        axis.title.y = element_blank(),
+        axis.text.y = element_blank(),
+        axis.ticks.y = element_blank(),
+        aspect.ratio=0.4)
+
 # Picture 6
+set.seed(1)
+x=rnorm(5000, 8, 4)
+ib_data2 <- tibble(x=ifelse(x<0, -x, x)) |>
+  mutate(x=ifelse(x>=18, 18-rexp(5000,rate=0), x))
+p6 <- ib_data2 |>
+  ggplot(aes(y="", x=x)) +
+  geom_jitter(size=0.05) + 
+  scale_x_continuous(name = "INCHES OF SNOW",
+                     breaks = seq(0,20,2),
+                     labels= paste0(seq(0,20,2)),
+                     limits=c(0,20)) +
+  theme_classic() + 
+  theme(axis.line.y=element_blank(),
+        axis.title.y = element_blank(),
+        axis.text.y = element_blank(),
+        axis.ticks.y = element_blank(),
+        aspect.ratio=0.1)
+
 # Picture 7
+p7 <- ib_data2 |>
+  arrange(x) |>
+  mutate(group = rep(1:50, each=100))|>
+  group_by(group) |>
+  summarise(x = max(x, na.rm=TRUE)) |>
+  add_row(group=c(0,51), x = c(0,20)) |>
+  ggplot(aes(x=x)) +
+  geom_linerange(ymin = 0.1, ymax = 1) + 
+  geom_linerange(y=1, xmin = -0.03, xmax = 20.03)+ 
+  geom_linerange(y=0.1, xmin = -0.03, xmax = 20.03)+ 
+  scale_x_continuous(name = "INCHES OF SNOW",
+                     breaks = seq(0,20,2),
+                     labels= paste0(seq(0,20,2)),
+                     limits=c(0,20)) +
+  scale_y_continuous(limits=c(0,1)) + 
+  theme_classic() + 
+  theme(axis.line.y=element_blank(),
+        axis.title.y = element_blank(),
+        axis.text.y = element_blank(),
+        axis.ticks.y = element_blank(),
+        aspect.ratio=0.1)
+
 # Picture 8
+p8 <- ib_data %>%
+  reframe(x = quantile(x, probs=c(0.25, 0.50, 0.75)))|>
+  add_row(x = c(0,20)) |>
+  arrange(x) |>
+  mutate(quantile = c("min", "q1", "med", "q3", "max")) |>
+  pivot_wider(names_from = quantile, values_from = x) |>
+  ggplot(aes(y="")) +
+  #geom_point(aes(x=med)) +
+  geom_errorbar(aes(y="", xmin = min, xmax = max), width = 0.2) +
+  geom_crossbar(aes(y="", x=med, xmin = q1, xmax = q3), width = 0.5) +
+  scale_x_continuous(name = "INCHES OF SNOW",
+                     breaks=seq(0,20),
+                     labels= ggplot2:::interleave(as.character(c(seq(0,20, 2))), rep("", 11))[1:21],
+                     limits=c(0,20)) +
+  theme_classic() +
+  theme(axis.line.y=element_blank(),
+        axis.title.y = element_blank(),
+        axis.text.y = element_blank(),
+        axis.ticks.y = element_blank(),
+        aspect.ratio=1/10)
+
 # Picture 9
-
+ggplot(df, aes(x)) +
+  stat_ecdf(geom = "step")
+
+# Display Plots
+p1
+p2
+p3
+p4
+p5
+p6
+p7
+p8
+p9
 ```

### 2024-05-06 paper.qmd
finish ibrekk images

@@
 #| message: false
-#| label: fig-maps
+#| warning: false
+#| label: fig-ibrekk
 #| fig-cap: "A replication of the the uncertainty visualisations shown by @Ibrekk1987 in one of the earliest uncertainty visualisation experiments.
@@
 #| fig-subcap: 
-#|   - "Mean with a 95% prediction interval"
+#|   - "Picture 1"
 #|   - "Picture 2"
@@
 ib_data <- tibble(x=ifelse(x<0, -x, x))
-ib_data %>%
-  ggplot(aes(x=x)) + geom_histogram()
 
@@
 # Picture 9
-ggplot(df, aes(x)) +
-  stat_ecdf(geom = "step")
+p9 <- ib_data |>
+  ggplot(aes(x)) +
+  stat_ecdf(geom = "step") +
+  scale_x_continuous(name = "INCHES OF SNOW",
+                     breaks=seq(0,20),
+                     labels= ggplot2:::interleave(as.character(c(seq(0,20, 2))), rep("", 11))[1:21],
+                     limits=c(0,20)) +
+  scale_y_continuous(name = "Cumulative probability",
+                     breaks=seq(0,1,0.1),
+                     labels= seq(0,1,0.1),
+                     limits=c(0,1)) +
+  theme_classic() +
+  theme(aspect.ratio=4/10)
 
@@
 
-p4b <- view(pmap_a) +
+p4a <- view(pmap_a) +
     scale_fill_distiller(

### 2024-05-06 paper.qmd
life is a nightmare and i hate pixel maps

@@
 # load Libraries
-library(strayr)
 library(tidyverse)
@@
 library(ggrepel)
+library(urbnmapr)
 
@@
 #|   - "Choropleth Map: High Variance"
+#|   - "Choropleth Palette"
 #|   - "Bivariate Map: Low Variance"
 #|   - "Bivariate Map: High Variance"
+#|   - "Bivariate Palette"
 #|   - "VSUP Map: Low Variance"
 #|   - "VSUP Map: High Variance"
+#|   - "VSU Palette"
 #|   - "Pixel Map: Low Variance"
 #|   - "Pixel Map: High Variance"
-#| layout-ncol: 2
+#|   - "Pixel Palette"
+#| layout-ncol: 3
 
@@
 # Get map data
-#sa4 <- read_absmap("sa42021")
-state <- read_absmap("state2021") |>
-  na.omit() |>
-  dplyr::mutate(temp = 31 - 0.001*(cent_long-110)^(2) - 0.003*(cent_lat-12.52294)^(2), # trend
-         notrend = rnorm(1), # no trend
-         highvar = runif(9, min=2, max=4), # high variance
-         lowvar = runif(9, min=0, max=2), # low variance
+my_map_data <- get_urbn_map("counties", sf = TRUE) |>
+  dplyr::filter(state_name=="Iowa")
+centroids <- as_tibble(rgeos::gCentroid(as(my_map_data$geometry, "Spatial"), byid = TRUE))
+my_map_data$cent_long <- centroids$x
+my_map_data$cent_lat <- centroids$y
+n <- dim(my_map_data)[1]
+
+# Make standard Palette
+# set Base colours
+basecols <- brewer.pal(8, name = "YlOrRd")
+#breaks <- seq(0,41, length.out = 9)
+breaks <- 21:29 #c(0,15,24,25,26,27,30,35,41)
+names(basecols) <- seq(8)
+limits <- c(10,41)
+
+my_map_data <- my_map_data |>
+  dplyr::mutate(temp = 28 - ((scale(cent_long))^(2) + (scale(cent_lat))^(2))[,1], # trend
+         notrend = rnorm(n), # no trend
+         highvar = runif(n, min=2, max=4), # high variance
+         lowvar = runif(n, min=0, max=2), # low variance
          ) |>
@@
   # add bivariate classes to data
-  mutate(bitemp = cut(temp, breaks=20:28, labels=seq(8)),
+  mutate(bitemp = cut(temp, breaks=breaks, labels=seq(8)),
          bivar = cut(variance, breaks=0:4, labels=seq(4)),
@@
 
-#sa4 <- sa4 |>
-#  dplyr::mutate(temp = 30 - 0.001*(cent_long-123.7799)^(2) - 0.003*(cent_lat-12.52294)^(2), # trend
-#         notrend = rnorm(1), # no trend
-#         highvar = runif(1, min=4, max=5), # high variance
-#         lowvar = runif(1, min=0, max=1), # low variance
-#         )
-
 # Choropleth Map
-p1a <- state |>
+p1a <- my_map_data |>
   filter(variance_class=="lowvar") %>%
   ggplot() +
-  geom_sf(aes(fill = temp, geometry = geometry)) + 
-  scale_fill_distiller(type = "seq", 
-                       palette = "YlOrRd", 
-                       direction = 1)+
-  theme_minimal() + 
+  geom_sf(aes(fill = bitemp, 
+              geometry = geometry)) + 
+  scale_fill_manual(values = basecols) +
+  #scale_fill_gradientn(colours = basecols, 
+  #                     values=breaks/limits[2],
+  #                     limits=limits) +
+  theme_void() + 
   theme(legend.position = "none")
 
-p1b <- p1a %+% filter(state, variance_class=="highvar")
+p1b <- p1a %+% filter(my_map_data, variance_class=="highvar")
 # Bivariate Map
@@
 
-# set Base colours
-basecols <- brewer.pal(8, name = "YlOrRd")
 # recurvisely decrease value
@@
 names(bivariatepal) <- paste(rep(1:8, 4), "-" , rep(1:4, each=8), sep="")
-# preview palette
-show_col(bivariatepal, ncol=8)
 
 # Bivariate maps
-p2a <- state |>
+p2a <- my_map_data |>
   filter(variance_class=="lowvar") %>%
@@
   scale_fill_manual(values = bivariatepal) +
-  theme_minimal() + 
+  theme_void() + 
   theme(legend.position = "none")
 
-p2b <- p2a %+% filter(state, variance_class=="highvar")
+p2b <- p2a %+% filter(my_map_data, variance_class=="highvar")
 
@@
 names(VSUP) <- paste(rep(1:8, 4), "-" , rep(1:4, each=8), sep="")
-show_col(VSUP, ncol=8)
 
 # VSUP maps
-p3a <- state |>
+p3a <- my_map_data |>
   filter(variance_class=="lowvar") %>%
@@
   scale_fill_manual(values = VSUP) +
-  theme_minimal() + 
+  theme_void() + 
   theme(legend.position = "none")
 
-p3b <- p3a %+% filter(state, variance_class=="highvar")
-
-# Maps I haven't been able to get working yet
+p3b <- p3a %+% filter(my_map_data, variance_class=="highvar")
 
@@
 # Low variance map
-state_a <- state %>%
+my_map_data_a <- my_map_data %>%
   filter(variance_class == "lowvar") %>%
-  mutate(my_id = seq(9),
+  mutate(my_id = seq(n),
          error = variance)
 
-pixel_1a <- state_a %>%
+# quantile
+q_a <- my_map_data_a |> 
+  as.tibble() |>
+  mutate(bitemp=as.numeric(bitemp)) |>
+  with(data.frame(p0.05 = qnorm(0.05, mean=bitemp, sd=sqrt(variance)),
+                  p0.25 = qnorm(0.25, mean=bitemp, sd=sqrt(variance)), 
+                  p0.5 = qnorm(0.5, mean=bitemp, sd=sqrt(variance)), 
+                  p0.75 = qnorm(0.75, mean=bitemp, sd=sqrt(variance)), 
+                  p0.95 = qnorm(0.95, mean=bitemp, sd=sqrt(variance))))|>
+  apply(2, function(x) ifelse(x>8, 8, x)) |>
+  apply(2, function(x) ifelse(x<0, 0, x))
+
+pixel_1a <- my_map_data_a %>%
   as.data.frame() %>%
-  select(my_id, temp, error) %>%
-  read.uv(estimate="temp", error="error")
-
-pixel_2a <- state_a %>% as("Spatial")
+  select(my_id, bitemp, error) %>%
+  read.uv(estimate="bitemp", error="error")
+
+pixel_2a <- my_map_data_a %>% as("Spatial")
 
@@
 
-pmap_a <- build_pmap(data = pixel_1a, distribution = "normal", pixelGeo = pix_a, id = "my_id", border = pixel_2a)
+pmap_a <- build_pmap(data = pixel_1a, distribution = "discrete", pixelGeo = pix_a, id = "my_id", border = pixel_2a, q=q_a)
 
 p4a <- view(pmap_a) +
-    scale_fill_distiller(
-      type = "seq", 
-      palette = "YlOrRd", 
-      direction = 1) +
-  scale_colour_distiller(
-      type = "seq", 
-      palette = "YlOrRd", 
-      direction = 1) +
+  scale_fill_gradientn(colours = basecols) +
+  scale_colour_gradientn(colours = basecols) +
   theme(legend.position="none")
@@
 # High variance
-state_b <- state %>%
+my_map_data_b <- my_map_data %>%
   filter(variance_class == "highvar") %>%
-  mutate(my_id = seq(9),
+  mutate(my_id = seq(n),
          error = variance)
 
-pixel_1b <- state_b %>%
+q_b <- my_map_data_b |> 
+  as.tibble() |>
+  mutate(bitemp=as.numeric(bitemp)) |>
+  with(data.frame(p0.05 = qnorm(0.05, mean=bitemp, sd=sqrt(variance)),
+                  p0.25 = qnorm(0.25, mean=bitemp, sd=sqrt(variance)), 
+                  p0.5 = qnorm(0.5, mean=bitemp, sd=sqrt(variance)), 
+                  p0.75 = qnorm(0.75, mean=bitemp, sd=sqrt(variance)), 
+                  p0.95 = qnorm(0.95, mean=bitemp, sd=sqrt(variance))))|>
+  apply(2, function(x) ifelse(x>8, 8, x)) |> #this will mess with the variance
+  apply(2, function(x) ifelse(x<0, 0, x))
+
+pixel_1b <- my_map_data_b %>%
   as.data.frame() %>%
-  select(my_id, temp, error) %>%
-  read.uv(estimate="temp", error="error")
-
-pixel_2b <- state_b %>% 
+  select(my_id, bitemp, error) %>%
+  read.uv(estimate="bitemp", error="error")
+
+pixel_2b <- my_map_data_b %>% 
   as("Spatial")
@@
 
-pmap_b <- build_pmap(data = pixel_1b, distribution = "normal", pixelGeo = pix_b, id = "my_id", border = pixel_2b)
+pmap_b <- build_pmap(data = pixel_1b, distribution = "discrete", pixelGeo = pix_b, id = "my_id", border = pixel_2b, q=q_b)
 
 p4b <- view(pmap_b) +
-    scale_fill_distiller(
-      type = "seq", 
-      palette = "YlOrRd", 
-      direction = 1) +
-  scale_colour_distiller(
-      type = "seq", 
-      palette = "YlOrRd", 
-      direction = 1) +
+  scale_fill_gradientn(colours = basecols) +
+  scale_colour_gradientn(colours = basecols) +
   theme(legend.position="none")
 
+# Display plots and palettes
 p1a
 p1b
+show_col(basecols, ncol=8, labels=FALSE)
 p2a
 p2b
+show_col(bivariatepal, ncol=8, labels=FALSE)
 p3a
 p3b
+show_col(VSUP, ncol=8, labels=FALSE)
 p4a
 p4b
+show_col(basecols, ncol=8, labels=FALSE)
 ```

### 2024-05-07 paper.qmd
edited background

@@
 Early papers that specifically reference "uncertainty visualisation" appear in the late 80s [@Ibrekk1987], with geospatial information visualisation literature in the early 90s declaring this to be essential aspect of information display [@MacEachren1992; @Carr1992].
+These early experiments typically involved showing participants a distribution, such as those depicted in @fig-ibrekk, and asking the viewers to extract a probability or average.
 Despite the new terminology visualisation of uncertainty has been present since the earliest times.
@@
 ```
+
 However, uncertainty visualisation takes many forms today, It can be considered to expand the application of data visualisation towards making better decisions, and research is concerned about how the construction of plots and information contained might affect the resulting decisions.
@@
 The review provided here attempts to address these issues.
-We begin with understanding the definition problem, to obtain and understanding of the *purpose* of visualising uncertainty, and understand if the current visualisation tools are sufficient to achieve that purpose and if not, where there is need for new methodology.
-We also posit that "uncertainty visualisation" is too diffuse to be considered a field in itself, but rather should be dissected into smaller elements.
-The ideal role of this review to help synthesize this noisy field, by collecting together essential components, summarising different viewpoints the current literature, and provide a guide for potential research topics.
+We highlight the three key issues facing the uncertainty visualisation literature, specifically the lack of a definition of uncertainty, the confusion as to what *is* an "uncertainty visualisation", and the difficulties with measuring "uncertainty".
+When disucssing each of these issues, we will discuss how this concept is understood in existing literature, why this understanding may be flawed, and the issues that arrise due to these flaws.
+By considering the *purpose* of visualizing uncertainty we avoid providing a simple list of the current methods, and instead consider if the current visualisation tools are sufficient to achieve that purpose and if not, where there is need for new methodology.
+The ideal role of this review to help synthesize this noisy field, by collecting together essential components, summarizing different viewpoints the current literature, and provide a guide for potential research topics.
 

### 2024-05-08 paper.qmd
cleaned up definition secion parts

@@
 # What is Uncertainty?
-Uncertainty visualisation is directed towards communication - to the general public when disasters may hit, to decision makers in control centers, and a whole host of other public-facing tasks - needing an understanding of what it is when the term "uncertainty" is used.
-
-Definitions in the literature appear seemingly view it to be encompassing and include everything the general public might consider to be uncertain.
-Some works (*XXX eg*) focus more narrowly on specific terms defined mathematically, such as probability, variance, error, or precision.
-Others (*XXX eg*) include broader loosely related elements, such as missing values.
+In order to visualise uncertainty, we need to know what we are trying to communicate through visualisation.
+Definitions in the literature take the view that the definition should be encompassing and include everything the general public might consider to be uncertain.
+Some works [@Hullman2018; @Maceachren2012; @Thomson2005] focus more narrowly on specific terms defined mathematically, such as probability, variance, error, or precision.
+Others [@Griethe2006; @Pang1997; @Pham2009; @Boukhelifa2017] include broader loosely related elements, such as missing values.
 Several attempts to address this gap are made in the papers describing taxonomies.
 
-## Definitions in taxonomies
+## Definitions of Uncertainty in the Literature
 Taxonomies split uncertainty based on an endless stream of ever changing boundaries, such as whether the uncertainty is due to true randomness or a lack of knowledge [@Spiegelhalter2017; @Hullman2016; @utypo], if the uncertainty is in the  attribute, spatial elements, or temporal element of the data [@Kinkeldey2014], whether the uncertainty is scientific (e.g.
@@
 @utypo identified many common themes among this wide array of taxonomies and used them to design an encapsulating taxonomy that was boldly referred to as a "definition of uncertainty".
+
 While taxonomies can be helpful to map out the space of "things that we consider to be related to uncertainty", a taxonomy is *not* a precise definition that would allow statisticians to estimate the uncertainty in their projects.
-The ramifications of this is seen throughout the literature, as not understanding how to calculate uncertainty is one of the leading reasons cited by visualisation authors when explaining why they don't include it in their visualisation [@Hullman2020a].
-
-The task of providing an encapsulating mathematical definition of uncertainty is far beyond the scope of this work, however we will discuss an important but commonly misunderstood feature of uncertainty; its relationship to inference.
+The ramifications of this is seen throughout the literature, as a survey of visualisation authors cited "not knowing how to calculate uncertainty" as one of the primary reasons they did not include it in visualisations [@Hullman2020a].
+The difficulties authors have with designing a clear definition of uncertainty likely comes from the fact that these definition fail to identify precicely what it is that *creates* uncertainty.
+These authors have failled to identify that uncertainty is simply a by product of inference.
 
 ## Defining uncertainty with respect to inference
-What exactly is uncertainty, then?
-If we were to consider making this overarching definition, what would it need in order to be "encapsulating"?
-Well, let us consider what might *not* be considered uncertain in order to understand this concept a little better.
-
-@Otsuka2023 spends the first chapter of his book discussing the place of descriptive statistics in the philosophy of the field and in doing so, highlights an interesting connection between inference and uncertainty.
-Descriptive statistics simply describe our sample as it is and summarises large data down into an easy to swallow format.
-Descriptive statistics are not seen as the primary goal of modern statistics, however this was not always the case.
+It is easy to see uncertainties relationship to inference when we consider what might *not* be considered uncertain rather than just trying to think about what *is*.
+
+Descriptive statistics simply describe our sample as it is and summarizes large data down into an easy to swallow format.
+Descriptive statistics are not seen as the primary goal of modern statistics, however, this was not always the case.
 Around the 19th century in England, *positivism* was the popular philosophical approach to science (positivists included famous statisticians such as Francis Galton and Karl Pearson) and practitioners of the approach believed statistics ended with descriptive statistics as science must be based on actual experience and observations, therefore anything that refers to the unobservable (such as new observations or population statistics) is not true science [@Otsuka2023].
-By its very nature, descriptive statistics *cannot* be used to make inferences about the data because it simply exists to summaries the data, to use it to make statements about *new* data is incorrect usage.
 In order to make statements about population statistics, future values, or new observations we need to perform inference, which requires the assumption of the "uniformity of nature" (i.e.
 that unobserved phenomena should be similar to observed phenomena) [@Otsuka2023].
-This subtle shift, from descriptive statistics to inferential statistics was historically shunned *due to the fact it introduced the unknowable*, or in other words, uncertainty.
-
-This philosophical understanding of statistics highlights that descriptive statistics do not have uncertainty, however some readers may disagree with that statement.
-Variance and probability are typically considered stand ins for "uncertainty", it is often how we choose to measure it.
-Additionally because probability and variance exist in descriptive statistics, descriptive statistics *must* have uncertainty.
-This is not necessarily true.
+This subtle shift, from descriptive statistics to inferential statistics was shunned during the positivism era *due to the fact it introduced the unknowable*, or in other words, uncertainty.
+
+This approach to uncertainty is embedded in the idea that descriptive statistics do not have uncertainty, which some readers may disagree with.
+Specifically, because it means uncertainty is *not* a latent attribute of data, but rather an attribute of a specific hypothesis or estimate.
+Deniers of this fact follow a consistent logical path and it is easy to identify the common mistake.
+We know that variance and probability are typically considered types of "uncertainty" *and* descriptive statistics can have variance and probabilities *therefore* descriptive statistics must have uncertainty.
+The flaw in this logic comes from the first step, assuming that the tools with which we measure uncertainty *are* uncertainty in of themselves.
+This confusion is common and there are many papers that spend a great deal of time clarifying the difference.
 @Begg2014 highlight that uncertainty is related to not knowing a specific value, while variability refers to the range of values a quantity can take at different locations, times or instances.
-Similar distinctions have been made by other authors, including @Spiegelhalter2017 who did not directly discuss variance vs uncertainty, but did differentiate between precise random events (such as the probability associated with a coin flip), and uncertainty (such as the estimated probability associated with a coin that might be biased).
+@Spiegelhalter2017 made sure to comment on the difference between precise random events (such as the probability associated with a coin flip), and uncertainty (such as the estimated probability associated with a coin that might be biased).
 The variance of a sample variance can be calculated and know, therefore it is not uncertain but rather it a precise description of dispersion.
-If we were to discuss drawing a new observation, or estimating the true mean of a population *then* the the variance would become relevant in our discussions of uncertainty.
-
-The idea that inference is related to uncertainty is not uncommon in the uncertainty literature, however it is often mentioned as a *task* or *goal* dependence.
-As the "goal" shapes every stage of an analysis, multiple authors in the uncertainty literature have commented on the need to consider quantifying and expressing uncertainty at every stage of a project [@Kinkeldey2014; @Hullman2016; @Refsgaard2007] while others highlight the importance of specific stages.
+If we were to discuss drawing a new observation, or estimating the true mean of a population *then* the variance would become relevant in our discussions of uncertainty.
+
+It is clear that uncertainties dependence on inference has gone unnoticed in the literature because authors often mention the by products of this fact as a *task* or *goal* dependence.
+Multiple authors have commented on the need to consider quantifying and expressing uncertainty at every stage of a project as the "goal" shapes every step of the analysis [@Kinkeldey2014; @Hullman2016; @Refsgaard2007].
 @Otsuka2023 suggested that the process of observing data to perform statistics is largely dependent on our goals, because the process of boiling real world entities down into probabilistic objects (or "probabilistic kind" as he puts it) depends on the relationship we seek to identify with our data.
@@
 @Fischhoff2014 looks at uncertainty visualisation for decision making decides that we should have different ways of communicating uncertainty based off what the user is supposed to do with it.
-It is clear that the distinction between noise and signal is important at every stage in an analysis, however combining the uncertainty from every stage is near impossible [@Spiegelhalter2017].
-This means that uncertainty is often discussed in isolation at multiple stages of a project, which further contributes to our fractured understanding of uncertainty.
-
+The importance of inference when discussing uncertainty is never directly acknowledge, but it is always present.
+It is clear that the relationship between uncertainty and inference is noticed at every stage in an analysis, however combining these uncertainties into a single "uncertainty" value is near impossible [@Spiegelhalter2017].
 With this understanding it becomes clear to see why uncertainty is tied to an endless string of examples in the data analysis pipeline.
 Uncertainty examples include imputed data, model selection, inherent randomness, biased sampling, etc, not because these things *are* uncertainty, but because they *create* uncertainty when we perform inference.
+
 Existing mathematical definition of uncertainty (or related concepts) have been unable to unify this range of diverse concepts.
 For example, @Thomson2005 suggests a mathematical formula for *examples* of uncertainty, @Meng2014 mathematically defined the variance introduced to a model by the array of model choices, and information theory tries to quantify uncertainty using the idea of entropy.
-These existing methods are not thorough enough for an analyst to understand what causes uncertainty, and quantify it for communication.
+These existing methods are useful for calculating uncertainty as a specific stage of the analysis, but not thorough enough for an analyst to understand the causes of uncertainty and quantify it.
+The task of providing an encapsulating mathematical definition of uncertainty is far beyond the scope of this work, however focusing on its relationship to inference is likely a fruitful step forward.
 
 ## Mistakes made when we misunderstand uncertainty
-The previous section established that uncertainty is a by product of the assumptions we make when we perform inference, additionally, our quantification of uncertainty is largely task dependent, therefore the uncertainty associated with *one* distribution (which we can also view as a task or goal dependence).
-When uncertainty visualisation authors directly stare at this concept, they almost unanimously agree with these concepts, however when these details are pushed the the periphery of a problem, we start to get research that confuses itself.
-In these cases, the authors are more confused about what they are doing than the participants.
-
-There are enough papers that fail to understand these basic principals of uncertainty that this entire paper could simply be a list with explanations. Instead, we have a handful of highlights.
-
-@Hofman2020 performed an experiment where some participants were shown a sampling distribution of the mean, and others were shown a prediction interval around the mean, for the effectiveness of a particular treatment.
-The authors reported that the group that was shown the sampling distribution were willing to pay more and ascribed a higher likelihood to the belief that the treatment was more effective than a control, than the group that was shown a prediction interval.
+The previous section established that uncertainty is a by product of the assumptions we make when we perform inference.
+This means that "uncertainty" is not defined by our data, but rather by the *questions we ask about it*.
+Unfortunately, uncertainty visualisation papers are not free from this issue.
+Usually papers that are confused on this concept will ask a question that is impossible to answer or does not make sense.
+Confusion about this core concept is incredibly common and rarely noticed, so much so that this entire paper could simply be a list of papers that have this issue.
+Instead, we have a handful of highlights.
+
+@Hofman2020 performed an experiment where some participants asked to judge the effectiveness of a particular treatment when shown a sampling distribution of the mean or a prediction interval around the mean.
 A prediction interval indicates the uncertainty associated with treatment *for a particular person* and the sampling interval shown the uncertainty associated with the *average effectiveness of the treatment*.
-It is already established practice in medicine to use a wide range of metrics to communicate different risks to best communicate the relevant information to a patient for a particular decision [@Spiegelhalter2017].
-The perceived importance of this work by @Hofman2020 hinges on a misunderstanding that the prediction and confidence interval *are* interchangeable despite performing inference on *very different* latent variables.
-This misunderstanding is so widespread that this finding, that shows the confidence and prediction intervals are *not* interchangeable, is considered groundbreaking enough to be published in CHI.
+The authors reported that the group that was shown the sampling distribution judged the treatment as more effective than those that were shown a prediction interval, very likely because the prediction interval is always wider than the confidence interval.
+The importance of displaying the *most relevant* statistic and associated uncertainty is already well established in medicine, where doctors are required to understand a wide range of metrics that each communicate different risks to a patient [@Spiegelhalter2017].
+Therefore, this work is not designed to commuicate to the field of medicine, but rather to the visualisation community who have historically struggled with the idea of uncertainty being related to inference.
+The perceived importance of this work by @Hofman2020 hinges on a widespread misunderstanding that the prediction and confidence interval *are* interchangeable despite performing inference on *very different* variables.
 
 @Boukhelifa2012 tried to quantify the strength of the intuitive connection between a line attribute called "sketchiness" and uncertainty.
-The authors of this study were aware that there was some kind of task dependence with respect to displaying uncertainty, so they tested the connection for multiple "tasks".
-The authors did not seem to understand the "task" dependence is likely a dependence on the inference for a particular statistic, so the "task" dependence was interpreted as *context* dependence.
-@fig-sketchy depicts the six scenarios that participants were shown and asked to interpret what they believe the squiggly line indicates.
-The authors misunderstanding led to "sketchiness" being used to depict uncertainty in the categories of bar charts, in graphic networks, and in train lines, with little consideration or explanation as to *how* these things would be uncertainty.
-It seems a tad ridiculous for the creators of a rail network map to be "uncertain" about the existence of a train line, however we are unable to guess at what the authors of the paper believed the uncertainty would be.
-The most likely scenario is that the authors did not consider the specific estimate that the uncertainty was representing, which is what led to visualisations that do not make conceptual sense.
+The authors of this study were aware that uncertainty experiments were suffering from some kind of task dependence issue (due to other papers on the topic discussing it), but interpreted this as *context* dependence rather than dependence an inferential statistic.
+Participants were shown the six scenarios depicted in @fig-sketchy and asked to interpret what they believed the squiggly line indicates.
+A failing to understand the relationship between uncertainty and inference led the authors to add "sketchiness" to aspects of the visualization that could not possibly have an inferential statistic associated with it.
+It seems the authors did not consider the specific estimate that the uncertainty should represent, which is what led to visualisations that do not make conceptual sense.
 Participants rightfully assumed the sketchiness therefore represented something else, such alternative options or simply ignored it.
@@
 This leads to the images being difficult to connect to uncertainty even if we assume tha that is what it is supposed to represent.
-For example, it is unclear how you can be uncertain about a train line (S5).](sketchiness.png){#fig-sketchy width=50%}
+For example, (S5) implies that participants should have read the plot to mean that the creators of a rail network map would be "uncertain" about the existence of a train line, something that is defies common logic.](sketchiness.png){#fig-sketchy width=50%}
 

### 2024-05-08 paper.qmd
updated definition mistakes made section

@@
 
-@Hofman2020 performed an experiment where some participants asked to judge the effectiveness of a particular treatment when shown a sampling distribution of the mean or a prediction interval around the mean.
-A prediction interval indicates the uncertainty associated with treatment *for a particular person* and the sampling interval shown the uncertainty associated with the *average effectiveness of the treatment*.
-The authors reported that the group that was shown the sampling distribution judged the treatment as more effective than those that were shown a prediction interval, very likely because the prediction interval is always wider than the confidence interval.
-The importance of displaying the *most relevant* statistic and associated uncertainty is already well established in medicine, where doctors are required to understand a wide range of metrics that each communicate different risks to a patient [@Spiegelhalter2017].
-Therefore, this work is not designed to commuicate to the field of medicine, but rather to the visualisation community who have historically struggled with the idea of uncertainty being related to inference.
-The perceived importance of this work by @Hofman2020 hinges on a widespread misunderstanding that the prediction and confidence interval *are* interchangeable despite performing inference on *very different* variables.
-
+The first common type of mistake is to present participants with insufficient information to answer a question due to the underlying belief that uncertainty is a latent product of the data.
 @Boukhelifa2012 tried to quantify the strength of the intuitive connection between a line attribute called "sketchiness" and uncertainty.
-The authors of this study were aware that uncertainty experiments were suffering from some kind of task dependence issue (due to other papers on the topic discussing it), but interpreted this as *context* dependence rather than dependence an inferential statistic.
 Participants were shown the six scenarios depicted in @fig-sketchy and asked to interpret what they believed the squiggly line indicates.
-A failing to understand the relationship between uncertainty and inference led the authors to add "sketchiness" to aspects of the visualization that could not possibly have an inferential statistic associated with it.
-It seems the authors did not consider the specific estimate that the uncertainty should represent, which is what led to visualisations that do not make conceptual sense.
-Participants rightfully assumed the sketchiness therefore represented something else, such alternative options or simply ignored it.
+The authors were aware that uncertainty seemed to have some "task" dependence but interpreted this as *context* dependence rather than dependence on a particular inferential statistic.
+Therefore, "sketchiness" was added to aspects of the visualization that had no obvious inferential statistic associated with it, so participants simply ignored the "sketchiness" or assumed it represented something else, such as an alternative option.
+@Hofman2020 performed an experiment where participants were asked to judge the effectiveness of a particular treatment.
+One group was shown the prediction interval around the mean, which indicates the uncertainty associated with treatment *for a particular person* and another group was shown the the sampling interval which depicted the uncertainty associated with the *average effectiveness of the treatment*.
+There was no critical assessment by the authors on *why* these two distributions should be compared despite performing inference on completely different distributions.
+The contribution of this experiment hinges on the belief that sharing a common descriptive statistic (in this case, the mean) is grounds to compare two unrelated inferential statistics, a belief which is clearly false when viewing uncertainty through the lens of inference.
+Both studies highlight a clear belief that uncertainty is related to descriptive statistics, not inferential statistics.
 

### 2024-05-13 paper.qmd
fixed up other bits

@@
 - May not need to reference the same paper again and again
+- add visually integrable to fig-tax once I work out what it is
+- look through for papers that can be used to cite motivation for uncertainty visualisation experiments
 -->
@@
 
-# What is Uncertainty?
+# What is uncertainty?
 In order to visualise uncertainty, we need to know what we are trying to communicate through visualisation.
@@
 
-Existing mathematical definition of uncertainty (or related concepts) have been unable to unify this range of diverse concepts.
-For example, @Thomson2005 suggests a mathematical formula for *examples* of uncertainty, @Meng2014 mathematically defined the variance introduced to a model by the array of model choices, and information theory tries to quantify uncertainty using the idea of entropy.
-These existing methods are useful for calculating uncertainty as a specific stage of the analysis, but not thorough enough for an analyst to understand the causes of uncertainty and quantify it.
-The task of providing an encapsulating mathematical definition of uncertainty is far beyond the scope of this work, however focusing on its relationship to inference is likely a fruitful step forward.
-
-## Mistakes made when we misunderstand uncertainty
-The previous section established that uncertainty is a by product of the assumptions we make when we perform inference.
-This means that "uncertainty" is not defined by our data, but rather by the *questions we ask about it*.
-Unfortunately, uncertainty visualisation papers are not free from this issue.
-Usually papers that are confused on this concept will ask a question that is impossible to answer or does not make sense.
-Confusion about this core concept is incredibly common and rarely noticed, so much so that this entire paper could simply be a list of papers that have this issue.
-Instead, we have a handful of highlights.
-
-The first common type of mistake is to present participants with insufficient information to answer a question due to the underlying belief that uncertainty is a latent product of the data.
-@Boukhelifa2012 tried to quantify the strength of the intuitive connection between a line attribute called "sketchiness" and uncertainty.
+The belief that uncertainty is a feature of the data, not the variable of interest, leads to several flabbergasting experiments in the uncertainty visualisation space.
+In these experiments, it is unclear how the participants were supposed to relate what they were shown to the uncertainty distribution the authors had in mind (or if the authors had any distribution in mind at all).
+For example, @Boukhelifa2012 tried to quantify the strength of the intuitive connection between a line attribute called "sketchiness" and uncertainty.
 Participants were shown the six scenarios depicted in @fig-sketchy and asked to interpret what they believed the squiggly line indicates.
@@
 Therefore, "sketchiness" was added to aspects of the visualization that had no obvious inferential statistic associated with it, so participants simply ignored the "sketchiness" or assumed it represented something else, such as an alternative option.
-@Hofman2020 performed an experiment where participants were asked to judge the effectiveness of a particular treatment.
+In another experiment, @Hofman2020 asked participants to judge the effectiveness of a particular treatment.
 One group was shown the prediction interval around the mean, which indicates the uncertainty associated with treatment *for a particular person* and another group was shown the the sampling interval which depicted the uncertainty associated with the *average effectiveness of the treatment*.
-There was no critical assessment by the authors on *why* these two distributions should be compared despite performing inference on completely different distributions.
+There was no critical assessment by the authors on *why* these two distributions should be compared.
 The contribution of this experiment hinges on the belief that sharing a common descriptive statistic (in this case, the mean) is grounds to compare two unrelated inferential statistics, a belief which is clearly false when viewing uncertainty through the lens of inference.
-Both studies highlight a clear belief that uncertainty is related to descriptive statistics, not inferential statistics.
 
@@
 
-Many authors do not understand bias *is* uncertainty (something that has been proven mathematically by @Meng2014) and whether we view uncertainty as bias or variance depends on the resolution of our problem.
-@Padilla2021 found that high uncertainty in the model estimates (calculated uncertainty) and low forecaster confidence (which is typically an expression of suspected bias in the model) both caused participants to have decreased confidence in their results and suggested modellers express both if they are relevant.
-@Kale2019 discussed the importance of communicating decisions made in the data analysis pipeline and being aware of the alternatives.
+Additionally, if we view uncertainty through the lens of inference then it is obvious that bias and variance both contribute to uncertainty regardless as to which stage of the analysis they appear.
+Many papers in uncertainty visualisation are not working from this null hypothesis.
+For example, @Padilla2021 found that high uncertainty in the model estimates (calculated uncertainty) and low forecaster confidence (which is typically an expression of suspected bias in the model) both caused participants to have decreased confidence in their results and suggested modelers express both if they are relevant.
+In another example, @Kale2019 discusses the importance of communicating decisions made in the data analysis pipeline and being aware of the alternatives.
 While there is nothing wrong with the results of these papers, the fact that they were published shows some level of surprise by these results.
-This work indicates that there was a significant number of authors in the field were not already aware that choices introduced early in the data analysis pipeline create bias and therefore uncertainty in our final values.
-The surprise that both bias and variance contribute to final uncertainty shows that authors do not understand both will contribute to the *distance* of your estimate to the value you are drawing inference on (AKA error).
-
-The misunderstanding that descriptive statistics also extends to visualisation.
-The exploration step of an analysis, which includes descriptive statistics, exploratory data visualisation, and unsupervised machine learning techniques, is performed without a prior hypothesis, however misunderstandings of this fact appears frequently in the literature.
-@Griethe2006 commented that "if visualization is used as a means to explore a data volume or to communicate its contents the uncertainty has to be included".
-@Potter2010 aimed to create a summary plot that "concisely presented data with uncertainty information" to create an exploratory visualisation tool that visualised uncertainty.
-Exploratory visualisations often differ from descriptive statistics because the explicit statistic we are drawing inference on is less explicit.
-A versatile visualisation such as a scatter plot allows for a viewer to consider several hypothesis at once, each of which may have its own associated uncertainty.
-@Hullman2021 argue that there is no such thing as a "model-free" visualisation, therefore visualisation require robust visualisations of uncertainty as we are always performing inference.
-While we agree with this sentiment, it is clear that uncertainty *cannot* be defined without a specific motivating question, and therefore trying to include uncertainty in an exploratory visualisation stage (which by definition does not have a hypothesis) is not possible.
-
-This confusion (or indifference) also creates many imprecisely named papers.
-Many papers will boast a title that claims to be about uncertainty visualisation, but simply depicts different visual representations of a PDF, however it takes reading the methodology to find this out (*Cite Gap 4: Organise uncertainty quantification*).
-While this is a minor by product of the definition issue discussed earlier, it does result in a literature that is needlessly difficult to navigate.
-
-The reality is that a lot of uncertainty visualisation authors do not seem to have an intuitive understanding of uncertainties connection to inference, when inference is being performed, and how to design experiments that capture this relationship.
-This misunderstanding is not due to laziness or the fault of the authors, but rather is likely caused by the absence of a strict definition of uncertainty.The imprecise and confusing set of existing definitions are creating a field in which the authors themselves do not know what they are testing.
-
-# What is Uncertainty Visualisation?
-If we simply built the idea of uncertainty visualisation up from "uncertainty" which we already know has a precarious definition we would understand the field of "uncertainty" visualisation must also be precarious.
-You cannot quantify something that is not defined, and you can't visualise something you cannot quantify.
-If we cannot quantify the full concept uncertainty in any meaningful way (other than with examples such as missing values or a PDF) then there is no such thing as an "uncertainty" visualisation.
-Unfortunately this problem is not quite that simple.
+This work indicates that there was a significant number of authors in the field were not already aware that choices introduced early in the data analysis pipeline create bias and therefore uncertainty in our final values that should be communicated.
+
+Existing mathematical definition of uncertainty (or related concepts) have been unable to unify this range of diverse concepts.
+For example, @Thomson2005 suggests a mathematical formula for *examples* of uncertainty, @Meng2014 mathematically defined the variance introduced to a model by the array of model choices, and information theory tries to quantify uncertainty using the idea of entropy.
+These existing methods are useful for calculating uncertainty as a specific stage of the analysis, but not thorough enough for an analyst to understand the causes of uncertainty and quantify it.
+The reality is that a lot of uncertainty visualisation authors do not seem to have an intuitive understanding of uncertainties connection to inference, when inference is being performed, and how to communicate this complicated by product of inference.
+This misunderstanding is not due to laziness or the fault of the authors, but rather is likely caused by the confusing set of existing definitions that leads authors to be unaware what they are supposed to be visualising.
+Therefore, it is clear the field would greatly benefit from a strict definition of uncertainty.
+The task of providing an encapsulating mathematical definition of uncertainty is far beyond the scope of this work, however focusing on its relationship to inference is likely a fruitful step forward.
+
+# What is an uncertainty visualisation?
+Given that there is no definition of "uncertainty" readers may wonder how we can define "uncertainty visualisations".
+This is a reasonable issue given that you cannot quantify something that is not defined, and you can't visualise something you cannot quantify.
+This section discusses how the problems caused by the absent definition of uncertainty bleed into our understandings of uncertainty visualisation and creates a field that cannot clearly state what is or is not within its domain.
 
 ## Definitions of "uncertainty visualisation"
-Before discussing the issues facing uncertainty visualisation due to a poor concept *of* uncertainty visualisation, we want to establish two ideas that should be considered in conjunction with the lacking definition of uncertainty.
-The first is the sidestepping of the "uncertainty" definition problem by suggesting "uncertainty visualisation" is inherently different.
-Some authors may consider "uncertainty" to be poorly defined, but "uncertainty visualisation" to be well established.
-We will spend the next section explaining why that is not the case by discussing several taxonomies of uncertainty visualisation.
-
-In order to visualise something, it needs to be quantifiable.
-In order to quantify something, it needs to be mathematically defined.
-The broad classes of what is considered "uncertainty" rarely have mathematical defintions, and therefore, cannot be meaningfully visualised.
-For these reasons simple quantifiable uncertainties, such as confidence intervals  dominate the literature, while visualisations for more complicated sources of uncertainty, such as the effects of assumptions, imputed missing variables, and model choices remain elusive.
-
-A large drive of the field of uncertainty visualisation, has been the idea that "uncertainty visualisation" (defined as one word) is somehow different to "uncertainty" "visualisation" (separately defined as two words).
-When "uncertainty visualisation" is defined as a single word we lose some nuance in our understanding of the driving force behind some issues, and problems with visualisation and problems with uncertainty become conflated.
+What exactly is or is not an uncertainty visualisation is just as poorly defined as the definition of uncertainty, so authors seem to decide for themselves what is an uncertainty visualisation.
 This appears in the literature as a mountain of seemingly conflicting statements about what is, or is not, an "uncertainty visualisation".
 For example @Leland2005 mentions that popular graphics, such as pie charts and bar charts omit uncertainty, however at least one or both of these charts are used in a significant number of uncertainty visualisation experiments [@Ibrekk1987; @Olston2002; @Zhao2023; @Hofmann2012].
-@Wickham2011 suggests their product plot framework, which includes histograms, should have a way to measure uncertainty, but does not consider that a histogram is *already* a depiction of PMF and would already be considered an uncertainty visualisation were our statistic of interest a new oberservation from our data set.
-These conflicting ideas and lacking definition of what is means for a visualisation to be an "uncertainty visualisation" is a cornerstone of some clearly questionable results within this field.
-
-Since the concept of "uncertainty visualisation" is harder to define than "uncertainty", "uncertainty visualisation" taxonomies are typically less thorough than taxonomies for "uncertainty".
-They also tend to blur the line between the mathematical elements of uncertainty and the visual depiction of those mathematical objects.
-While we will not focus on these taxonomies, it is still important to mention them and highlight what we can learn from their success' and failures.
-
-Some taxonomies of uncertainty highlight how confusing most authors find the concept, and how difficult it is to articulate the rules of an uncertainty visualisation when uncertainty itself if not defined.
-@Potter2012 organised several existing uncertainty visualisations into groups based on the dimensionality of the data (1D, 2D, 3D, and No Dimension) and the dimensionality of the uncertainty (Scalar, Vector, Tensor).
-However, because the term "PDF", a statistical object that describes a random variable that is typically a one dimensional function, is used to describe both the data and the uncertainty for all dimensions, it is hard to understand how this should work.
-@Grewal2021 created a taxonomy that mapped uncertainty visualisations to some point in a 2D space defined by the "domain expertise" and "continuum of discreteness" (that scaled from "point estimate" to "continuous distribution").
-This conceptualization could be considered a visual extension of the law of large numbers, because as a point estimate is a single sample, as the sample size gets larger the distribution will appear to be a continuous.
-However, assumes every point estimate is a sample from the same distribution (a mean is not technically a sample from a prediction interval, but it is a sample from the sampling distribution) and it conflates discreteness from a too small a number of observations, discreteness from a precision problem, discreteness because that is the true nature of the variable, and discreteness as a visualisation choice.
-@Griethe2006 organised uncertainty visualisations into two cases (1) a hypothesis test was preformed to confirm the validity of the visualisation and (2) the visualisation has uncertainty depicted.
-This conceptualization conflates hypothesis testing and confidence intervals, and while they are related concepts, a hypothesis test is not necessarily uncertainty.
-Of course the authors of these papers did not intend for the taxonomy to be a complete description of uncertainty visualisations, however these distinctions that are often ignored are subtly important.
-(*Cite Gap 9: Other uncertainty vis taxonomies*)
-
-There are a small handful of taxonomies that highlight some interesting features in the way uncertainty is visualised. @Kinkeldey2014 categorised uncertainty according to whether or not it was:
-1. Implicit (a sample) or explicit (a depiction of mass)
-2. Intrinsic (alter existing symbols symbols to represent uncertainty) or extrinsic (add new objects to represent uncertainty)
-3. visually integral or separable (the uncertainty can be separated from the data and read independently) 
-4. coincidence (uncertainty and data are represented in the same plot) or adjacent
-5. static or dynamic (animated, interactive etc)
-Of the groups they identified, they actually only used (4), (2), and (5), left out (1) because most visualisations are explicit, and (3) corresponds to (1) in most cases [@Kinkeldey2014].
-This taxonomy suggests interesting considerations for a visualisation of a distribution (even though it was not intended to be a visualisation of a distribution).
-A similar version of this taxonomy was presented by @uncertchap2022 who commented that visualisation can be organised into two categories, "graphical annotations of distributional properties" and "visual encodings of uncertainty" which functionally align with the intrinsic/extrinsic distinction.
-The first four categories boil down to the question "should uncertainty and signal be conveyed together as one variable or separately as two".
-This consideration will become more relevant later in this review.
-
-![Visualisation of @Kinkeldey2014 uncertainty visualisation taxonomy.
-It can be seen that most of the categories can be considered a change to the typical error bar chart.
-Many of the distinctions conflate changes of the input data and statistics calculated and with visual changes, highlighting how defining uncertainty and visualisation together conflates several elements of a graphic.](uncertvistax.jpeg){#fig-location width=70%}
-
-Most of these taxonomies are created by observing existing uncertainty visualisations, which means they suffer from many of the same conceptual issues that the field has with uncertainty in general.
-Additionally, there does not seem to be anything these taxonomies offer that would not be better established by separate taxonomies for uncertainty and visualisation.
-The difference between "uncertainty visualisations" and "data visualisations" is not technically in the visual element, it is mathematical.
-@Kinkeldey2014 almost acknowledges this in their own paper that discusses an uncertainty visualisation taxonomy when they claim "future typologies should take different categories of tasks into account (1) communication tasks (2) analytical tasks (3) exploratory task", a common typology for information visualisation in general.
-The process of understanding and estimating uncertainty requires knowledge of the data, the statistical methods used to make an estimate, and the assumptions of a model.
-Visualising the statistics that represent uncertainty should be no different than depicting the statistics that represent any other element of a graphic, this is something we will show in the following sections of this paper.
-
-## Why does the field of "uncertainty visualisation" exist at all?
-It is unclear exactly *why* uncertainty visualisation is a field of interest. The two most common justifications are:
-1) Uncertainty is fundamentally differently to other variables due to psychological heuristics involved in the interpretation of uncertainty (e.g. risk aversion).
-2) Uncertainty is not of interest in of itself and is typically layered on at the end of a the visualisation pipeline, so uncertainty visualisation is the field of adding error information into already established graphics and *thus* uncertainty information is a high dimensional visualisation problem.
-(*Cite Gap 10: Reason uncertainty visualisation is different*)
-Papers will often discuss uncertainty in relation to one of these motivating reasons, the evaluation experiments motivated by (1) often perform different visualisations of PDFs, while the papers motivated by (2) will focus on trying to impute uncertainty as error within one of the existing channels, such as colour (*Cite Gap 10*).
-These dual motivations that co-exist with uncertainty visualisation authors rarely mentioning the others, create confusion in the field as to its purpose.
-@Kinkeldey2014 also identified this issue when they highlighted that the literature makes it unclear if uncertainty is a variable in of itself, something that should be interpreted with the main variable, or if it is metadata.
-
-Uncertainty is noise, but most evaluation experiments measure it as signal.
-Uncertainty acts as a form of statistical "hedging" for signals found in an analysis.
-Interviews with experts in statistic back up this primary motivation, as ignoring uncertainty information of often expressed as being similar to fraud or lying [@Hullman2020a; @Manski2020].
-Noise and signal are not an inherent property of variance, or meta-data or the aspects of an analysis that are usually secondary considerations, it is a property of the data that *promotes* a message we are trying to infer and the data that *suppresses* it, if that message is invalid.
-When we ask the viewer of a plot to look at data and extract a value, we are asking them to perform inference on that value.
-There will be noise associated with that answer and that is uncertainty.
-If we ask direct question about some uncertainty metric, we have turned the uncertainty into signal because that is what the participants are drawing inference on.
-
-This is likely the cause of the contradiction between uncertainty visualisation papers following identical results to normal information visualisation papers, and authors believing uncertainty is inherently *different* to normal visualisation.
-Each of the sections in this paper likely contribute in some way to this contradiction in conceptualisation of uncertainty visualisation and the results of the studies in the field.
-Uncertainty is poorly defined, therefore authors do not understand it is not an inherent property of the data and is actually related to a particular inference question, this incorrect view that uncertainty is always variance or probabilities leads to the class of "uncertainty visualisations" that depict different mathematical objects or visual features with little in the sense of consistent rules, this complicated class has caused visualisation authors to miss that they are simply repeating the established evaluation results from normal information visualisations.
-This situation is a bit of a mess.
-
-If uncertainty is inherently different, we would expect evaluation studies of uncertainty visualisations to produce different results to similar evaluation studies on "normal" variables.
-This does not seem to be the case.
-
-## Mistakes made when we misunderstand uncertainty visualisation
-In the previous section, we highlighted several problematic experimental designs or pieces of research that were the result of misunderstandings around the definition of uncertainty.
-Here, we will discuss the issues that arise due to a misunderstanding of "uncertainty visualisation".
-The two primary problems are the regular comparisons of mathematically incomparable plots, and reproducing results that are already widely understood in the larger information visualisation literature.
-
-### Information asymmetry in evaluated plots
-Visualisation authors are almost unanimous in commenting that the "information" in two plots must be the same in order for the visual techniques to be compared [@Cleveland1984; @Kinkeldey2014] (*Cite gap 12: Equal information*).
-However, what makes two visualisations equal in "information" is not consistent and is regularly mentioned without it being clear exactly what this information is.
-This is easily illustrated by the wide array of comments made by visualisation authors when they explain why two representations are being compared in their evaluation study.
-@Ibrekk1987 displayed 6 PDFs in their experiment because "formally equivalent representations are often not psychologically equivalent", but do not explain why what makes a representation "formally equivalent" or why only one representation was presented for the 95% CI, box plot, and CDF.
-@Hofman2020 comments that "theoretically" the sampling distribution of the mean and the prediction interval of a new observation are equal "so long as one knows the sample size", but does not recognise the several assumptions and statistical knowledge that is required to compare the two.
-(*Cite Gap 13: examples of what is defined as "information"*).
+@Wickham2011 suggests their product plot framework, which includes histograms, should have a way to measure uncertainty, but does not consider that a histogram is *already* a depiction of PDF, something that is often considered an uncertainty visualisation by other authors.
+This completely untethered concept of an "uncertainty visualiation" leads to many user experiments where the visualisations used have a shocking information asymetry.
+
+Visualisation authors are almost unanimous in commenting that the "information" in two plots must be the same in order for the visual techniques to be compared [@Cleveland1984; @Kinkeldey2014].
 @Kinkeldey2014 adopts an existing definition also that suggests two graphics are informationally equivalent if all the information in one plot is inferable from the other and vice-versa, but adds that two plots are computationally equivalent if that information can be extracted from both plots with similar easy and speed.
-It is clear this lacking definition of "information" in a visual is reminiscent of our problems with the definition of "uncertainty".
-
-Works such as *The Grammar of Graphics* attempt to outline exactly what information is contained within a plot, and in doing so allow plots to be considered a statistic [@Vanderplas2020].
-This statistic that *The Grammar of Graphics* defines fundamentally a descriptive statistic.
-@fig-pipeline, and contains the set of necessary and sufficient steps that are required to make a graphic, as defined in *The Grammar of Graphics* [@Leland2005].
-It can be seen that the grammar assumes we have our data, performs a set of functions on that data, and produces a statistic.
-The grammar of graphics fundamentally describes the graphic as it is, not as it *would* be were our sample size to increase to infinity.
-As a matter of fact, the graphics assumption that we start from a data set that has an unchanging sample size is at odds with inference that typically involves assuming an infinitely large sample size.
-@fig-infintegraph shows that as n approaches infinity, two graphics that are very similar according to the *The Grammar of Graphics* can diverge into different graphics, while two graphics that are very different, can converge into the same graphic.
-*The Grammar of Graphics* itself does not recognize this issue, as uncertainty is referred to as "semantics" and is ultimately hand waved away.
-This does not mean describing a graphic using inferential statistics is completely unheard of.
-Some graphical descriptors, such as Product Plots [@Wickham2011] describe graphics using the underlying distribution, however the framework is not a substitute for *The Grammar of Graphics*.
-Other authors have noticed difficulties with inferential statistics and and created an addition to the grammar that should make depicting a distribution easier [@Pu2020; @Kay2023].
-Unfortunately, this addition falls short as it does not recognize *why* latent distributions are difficult to describe using *The Grammar of Graphics* and carries over many of the same problems.
-
-This does not mean *The Grammar of Graphics* should be abandoned when we visualize uncertainty, rather, these difficulties are something we should keep in mind when
-
-@Tierney2023 generates a visualisation for missing values, however because it fits nicely into the grammar of graphics framework, it is inherrently a descriptive statistic.
-
-If uncertainty is fundamentally a by-product of inference, then the visualisations as described by *The Grammar of Graphics*
-
-Most uncertainty visualisation experiments are written with a seeming indifference to the grammar of Graphics, despite it being necessary to untangle experimental results.
-The primary reason for this seems to be the absent definition of uncertainty.
-Uncertainty is fundamentally related to inference and descriptive statistics cannot have uncertainty.
-
-Since two plots cannot be compared on their visual elements if they contain different information (because you will never be able to identify if the difference is because of informational asymmetry or visual processing) we could understand two plots as being the same if their "varset" steps are identical.
-However because uncertainty is not defined mathematically, we cannot reliably add it in at the "statistic" stage.
-, such as resampling do not quite fit within the framework.
-For example, if we have one set of data and two graphics, where one graphic represents the data as a series of points, and the the other depicts a smoothed density using a line.
-These two plots differ in *both* the statistic and geometry stages of the grammar of graphics, however as n (the number of observations in our sample) gets larger, these two plots will become visually indistinguishable (depending on the collision modifier used).
-This result aligns nicely with large sample theory, because as n increases the data and mass should become mathematically indistinguishable, and many uncertainty visualisation experiments notice an interaction between sample size and graphic effectiveness so this result is likely also practically true [@Kale2018; @Newburger2022; @Hofmann2012] but despite the increasing similarity in what is actually depicted in the graphic, the sample and mass visualisations would remain different visualisations within the grammar of graphics framework.
-Additionally, authors rarely attribute this visual version of large sample theory to results where it is clearly applicable.
-@Kale2018 found that users were more sensitive to the underlying trend when shown the HOPs plot over the static outcomes plot, this effect went away when the speed of the animation increased, indicating that performance difference might be due to overplotting in the static case and a smaller sample size in the static condition might produce the same result.
-The issue of blurred elements of the grammar of graphics in the case of distribution visualisations does not end here.
-Visualisations that depict mass such as a confidence interval, a boxplot, a PDF, a letterbox plot, a violin plot, a histogram, a quantile dot plot, etc, all show some version of the mass of a variable at different resolutions, but whether these graphics differ in something as low down in the grammar pipeline as the variable stage, or later at the geometry stage is unclear.
+Many visualisation authors ignore this concept and simply compare two visualisations because they both belong to the class of "uncertainty visualisations".
+@Ibrekk1987 compared a 6 visualisations of a PDF,  a box plot, a CDF and a mean with a confidence interval on this basis.
+They found that people are better at extracting the mean from a plot when they are shown a plot that contains a mean with a confidence interval than when they are shown a PDF, a box plot, or a CDF (where the mean could not be read off the plot).
+@Hullman2015 compared the static error bars and violin plots of the marginal distributions of two variables ($A$ and $B$) to an animated plot that depicted outcomes of the joint distribution of $A$ and $B$ in each frame and found that the visualisation of the joint distibution was better at answering questions about the joint distribution than the visualisations of the marginal distribution.
+@Hofman2020 commented that "theoretically" the sampling distribution of the mean and the prediction interval of a new observation are equal "so long as one knows the sample size", but does not seem to provide participants with that sample size, or recognise the assumptions and background knowledge that would be required to compare the two.
+@Hofman2020 and @Zhang2022 compared prediction and sampling distributions because they are both "uncertainty" that is typically depicted around the mean.
+They found that people are better at answering questions about a prediction interval when shown a prediction interval instead of a sampling distribution.
+
+This collection of examples starts to paint a pretty clear picture.
+Visualisations with rather shocking information asymmetry are regularly compared because they are both "uncertainty visualisations".
+This results in a series of experiments where the visual aspects of the graphic are not even required to anticipate the experimental results.
+Ideally this problem would be avoided by defining the information in a graphic using *The Grammar of Graphics* since it is a sufficient summary of the information in a graphic and it allows a plot to be considered a statistic [@Vanderplas2020].
+The process for summarising the information in a plot using *The Grammar of Graphics* is depicted in @fig-pipeline [@Leland2005].
+Unfortunately, *The Grammar of Graphics* is not perfectly suited to summarise the information in an uncertainty visualisation since it assumes that we start from a finite data set.
+Very often uncertainty is expressed using a distribution or a resampling method that does not have an upper limit on the "sample size".
+Additionally many uncertainty visualisation experiments notice an interaction between sample size and graphic effectiveness [@Kale2018; @Newburger2022; @Hofmann2012], an effect that is likely caused by a difference in human perception of mass versus sample.
+Additionally, *The Grammar of Graphics* behaves strangely when the sample size is adjusted and it the distance between two graphics becomes less consistent.
+@fig-infintegraph shows how three graphics defined by *The Grammar of Graphics* can change as n approaches infinity.
+The two graphics that are very similar according to the *The Grammar of Graphics* diverge into different graphics, and two graphics that are very different, converge into the same graphic.
+This means the statistics visualised in the plot as well as the *The Grammar of Graphics* may not be an adequate to summarise the information in a plot that is being used for uncertainty visualisation.
+Some authors have noticed difficulties with visualising distributions using `ggplot2` and created extensions that make showing distributions more intuitive [@Pu2020; @Kay2023], however these extensions do not address the core discrepancies with uncertainty visualisation and *The Grammar of Graphics*.
+This does not mean *The Grammar of Graphics* should be abandoned when we attempt to summaise the information in an uncertainty visualisation, very often visations use the same sample to generate all their plots, so this issue is not always present, however these limitations should be kept in mind when analysing uncertainty visualisations using *The Grammar of Graphics*.
 
@@
 
-Alternative information consideration within well defined frameworks may also be worthwhile to consider.
-@wu2023rational tried to eliminate information asymmetry by essentially looking at a visualisation and identifying what information can be extracted and then using a "rational agent benchmark" to determine how a rational person would use that information for a decision making task.
-While this solution is interesting, it may have different results depending on who is applying it as they may use different methods or notice different elements when extracting information from a graphic.
-The mathematically defined concept of sufficient statistics is also scarcely mentioned likely because it is specific to a particular ground truth statistic, however uncertainty *is* specific to a particular statistic, so it could be a useful avenue for understanding when two plots differ in their ability to answer a particular question.
-
-Ultimately, this all points to confusion as to when two visualisation contain information asymmetry, a problem that becomes more pronounced in uncertainty visualisation.
-@Ibrekk1987 compared a 6 visualisations of a PDF,  a box plot, a CDF and a mean with a confidence interval on the basis that all of them were "uncertainty visualisations".
-They found that people are better at extracting the mean from a plot when they are shown a plot that contains a mean with a confidence interval than when they are shown a PDF, a box plot, or a CDF (where the mean could not be read off the plot).
-@Hofman2020 and @Zhang2022 compared prediction and sampling distributions because they are both "uncertainty" that is typically depicted around the mean.
-They found that people are better at answering questions about a prediction interval when shown a prediction interval instead of a sampling distribution.
-@Hullman2015 compared the static error bars and violin plots of the marginal distributions of two variables ($A$ and $B$) to an animated plot that depicted a single outcome of the joint distribution of $A$ and $B$ in each frame of an animation, a comparison that was justified because all plots were "uncertainty visualisations".
-They found that the plot that depicted outcomes from $P(A, B)$ where the distance between the two outcomes was equivalent to $a-b$ was better at answering the question "What is $P(B<A)$" than violin or error bar plots depicting $P(A)$ and $P(B)$ separately.
-(*Cite Gap 14: Examples of mathematically obvious results*).
-
-This collection of examples starts to paint a pretty clear picture.
-Visualisations with rather shocking information asymmetry are regularly compared because they are both "uncertainty visualisations".
-This results in a series of experiments where the visual aspects of the graphic are not even required to anticipate the experimental results.
-This issue likely results in the lacking mathematical definition preventing us from defining and quantifying "uncertainty" mathematically and then including this quantification in a well defined framework such as the grammar of graphics.
-
-### Repeating perceptual task experiments
-Given the previous section, one might consider problems in uncertainty visualisation to be fixed if we could restrict evaluation experiments to cases where the information in each graphic is almost identical.
-Unfortunately this still often results in uninteresting or obvious results.
-
-If two graphics are visually different but identical in the information they contain, they must differ in how that information is depicted. 
-Once we have a mathematical expression of uncertainty, the visualisation of that uncertainty is theoretically identical to the visualisation process of any other variable.
-For simple tasks such as value extraction, there is a hierarchy to perceptual tasks where extracting visual information in some forms is easier than others.
-The hierarchy was originally established 40 years ago by @Cleveland1984, below is an updated version summarised by @Vanderplas2020:
-
-1) Position along a common scale. 
-2) Position along a non-aligned scale. 
-3) Length, direction, angle, slope
-4) Area
-5) Volume, density, curvature
-6) Shading, colour saturation, colour hue
-7) Discriminable shape
-8) Indiscriminable shape
-
-This hierarchy is a good general rule, however it can change from person to person [@Davis2022] Additionally, there are other graphical rules to consider such as gestalt principles, broader methods of extraction, and attention principles [@Vanderplas2020].
-These established visualisation concepts allow us to anticipate the ease with which certain pieces of information will be extracted from a plot.
-We can use these concepts to understand the computational complexity of a graphic.
-A bizarre feature of the uncertainty visualisation literature is that it does not work to build upon these existing principles or identify the ways in which uncertainty visualisations may diverge from these rules.
-These building block concepts of visualisation are seldom mentioned.
-
-It is difficult to find examples of uncertainty visualisation experiments where the plots do contain the same information, however when they do, the results align with existing information visualisation research.
-Technically, a PDF and a mean with confidence intervals both have enough information to extract the mean of the distribution, however they both have a very different computational cost.
-To extract the mean using a PDF, a participant would need to identify the point along the x axis that splits the area under the curve in half.
-If a participant is provided with a mean with a confidence interval, extracting the average is a simple task of reading the position on an aligned scale.
-@Ibrekk1987 found that when asking participants for the "best estimate" (which they thought should be interpreted as the mean of the distribution) of a skewed distribution, participants provided the mean when given a mean with confidence intervals and the mode when given a PDF [@Ibrekk1987].
-Similar results to this occur over and over again in the uncertainty visualisation literature.
-@Gschwandtnei2016 found that visualisations where the start time of an interval could literally be read off the plot (error bars, centred error bars, and ambiguation) performed better than the plots (accumulated probability, gradient, and violin) where the start time involved some guesswork because the drop off was gradual.
-@Cheong2016 found that participants were better at answering questions when they were explicitly given the relevant probability in text rather than when they needed to read it off a map.
-(*Cite Gap 15: Examples of replicated perceptual task experiments*).
-
-These results show that uncertainty is not technically different to any other variable.
-When trying to anticipate the results of these studies, we can use the same principles of information equivalence and difficulty of relevant mental tasks to understand which plots will outperform others.
-This leads us to wonder...
-why is uncertainty visualisation a field at all?
-
-It is clear that authors intuitively *feel* that uncertainty is different even though this difference has not been captured in existing uncertainty visualisation experiments.
-Some authors cited uncertainty was special because of its psychological qualities, but we have yet to see any evidence of that here.
-@Kinkeldey2014 also noticed that it is not clear whether or not uncertainty is just another variable, as many value extraction papers treats it as such in their experimental design and uncertainty may need to be in a variable class of its own.
-@Hullman2016 commented that it is straightforward to show a value but it is much more complex to show uncertainty, but did not explain *why* uncertainty information faces more difficulty.
-After all, confidence intervals, minimums, maximums, variance, and any other statistic that can be considered related to "uncertainty" is just that, a statistic.
-If we want to visualise the relationship between a temperature estimate and our forecasted variance on that estimate, can we not just use a scatter plot, as we would any other pair of variables?
-There is nothing special about these statistics that should differentiate them from a mean or median and imply their visualisation should be "special" in a way that would warrant its own field.
-Especially when the core takeaways from uncertainty evaluation experiments do not seem to differ at all from the established results in information visualisation.
-So, what is it that makes authors believe uncertainty is different?
-
-# Issues with measuring uncertainty as noise, not signal
-The previous section highlights a unique and fascinating problem faced by uncertainty visualisation.
-If asking direct questions about uncertainty causes us to treat it as a signal, how do we evaluate uncertainty as *noise*?
-
-Uncertainty, acting as it is intended, is transparency.
-There are many secondary benefits that come with this improved transparency, such as better decisions, more trust in the results and more confidence in the authors.
-These secondary benefits are, however, *not* the immediate goal of uncertainty.
-The following sections will discuss the issues and limitations in measuring uncertainty through these secondary metrics and provide suggestions as to how future studies should consider measuring uncertainty.
-
-## Issues with the current methods of measuring uncertainty
-Uncertainty visualisation papers can be organised according to the *goal* of the experiment.
-Evaluation experiments are the standard rule for visualisations because the human brain is not as reliable as mathematical calculation.
-Therefore, user studies often aim to assess the limitations, biases, and heuristics of our mental calculator so that we can better understand the problems we may encounter when we plot our data.
-This is not to say any paper that suggests a visualisation without an evaluation experiment is completely lacking in justification, and there are many papers that suggest a novel visualisation without an evaluation study.
-Sometimes these papers are a preliminary step in finding a solution for common problems and intend to evaluate the visualisation in later work.
-The reasoning for this is obvious.
-There are often heuristics and biases that are not obvious to us when designing visualisation.
-Additionally, these heuristics and biases can change depending on the larger scope of the graphic and the population we are communicating with [@Spiegelhalter2017; @Kinkeldey2014].
-
-The overarching goal of most uncertainty visualisation evaluation papers belong to one of: performance (how effectively a participant can extract information from a plot), interpretation and semantics (the ease with which a particular visualisation is associated with uncertainty), and quality of user experience (if the participants liked the plot or not)[@Hullman2019].
-@Hullman2019 found that the majority of papers evaluate visualisations on performance (approximately 65% of the papers they surveyed) or interpretation and semantics (approximately 17%) and both of these evaluation goals will run into problems because of their conceptualisation of uncertainty.
-
-### Performance
-There are many metrics experimenters use to evaluate the performance of a visualisation.
-Each metric, such as can use accuracy, decision quality, confidence, trust and others each have their own issues which prevent them from accurately capturing the effects of uncertainty in a visualisation.
-
-By far the most common metric used is accuracy, which is used by approximately 36% of evaluation studies [@Hullman2019].
-In previous sections, we pointed out issues that arise when accuracy (of extracting probability or variance) is used as the primary metric in evaluating uncertainty visualisations, the work simply replicates known concepts in information visualisation.
-Subjective metrics, such as how much participants liked the aesthetic representation of a visualisation, or metrics with an unclear motivation or use, such as how memorable the information in a plot is or how much a user interacted with a plot, are ignored in this section.
-Below we will explain in detail why experiments that try to measure uncertainty using methods such as decision making, trust, or other types of questions, also fail to capture the effects of visualising uncertainty in a plot.
-
-#### Decision making and risk aversion
-
-![Visualisation of the process of transforming raw data to a visualisation insight.
-Summarising the data using common statistics removes information from our data.
-Once the data is in a format that represents the varibales of interest, organising that information in a visualisation makes it easier to extract insights from these statistics and convert them to insights.
-If information is removed at the statistics stage, it cannot be added back in at the visualisation stage, as it is meerly an efficient technique to organise and present information to enable a large range of insights.
-](visprocess.jpeg){#fig-location width=70%}
-
-Decision making tasks are often described by authors as a more realistic version how plots are used in practice.
-While we do not disagree with this statement, exactly *why* these tasks are different to value extraction tasks is never explained.
-While the task may be more similar to how plots are actually used, that does not necessarily mean it is a better *experimental* environment for evaluating plots.
-Let us consider this in more detail.
-A normal value extraction task involves:
-
-1) correctly interpret the question
-2) extract the specified value
-3) report the value
-
-On the other hand, a decision making task using an uncertainty visualisation involves:
-
-1) Correctly interpret the question
-2) Use some risk utility function to set a threshold for an "acceptable level of risk"
-3) Extract the value for from the plot that will be compared to the threshold
-4) Make a judgement based off that value
-
-The key aspect that separates a decision making task from a value extraction task, for uncertainty visualisations, is the inclusion of a utility function.
-While this might seem like a small change, it actually completely warps the experiment in unexpected.
-
-The first issue is that individuals will all have their own individual risk utility function, that is, how much they want to avoid or engage with risks or uncertainty.
-Therefore decision making experiments that have an additional layer of noise that value extraction experiments do not.
-We cannot be sure if participants answered differently from the ground truth because a visualisation was difficult to read *or* because the participants risk utility function did not align with the one set by the authors.
-Several authors have offered solutions to this issue, however the problem is deeper than any of them seem to realise.
-
-@Hullman2016 suggested providing a utility framework for each experiment, to instruct the participants in how to account for the uncertainty information.
-This is a method that seems to have been adopted by @Fernandes2018 who describe the following scenario to participants who are trying to maximise the coins in their experiment:
-> Subjects gain coins for every minute they are able to continue an activity that is valuable to them (e.g., watching TV at home) before going to the bus stop, and gain a bonus for arriving at their intended destination early.
-Subjects incur a coin penalty for time spent waiting at a stop for a bus to arrive.
-
-Any payment scheme that is incorporated into an experiment will also implicitly set up a utility framework, as it is assumed participants will try to maximise their payments.
-In tasks such as this, the ground truth of each question is typically selected to be the value a ration agent would select and visualisations are evaluated on the basis of how far the participants responses are from that ground truth.
-The problem with this, is that all the complexity of the question is in *working out* the best response, and it has little to do with the visualisation itself.
-The visualisation aspect of these studies becomes a value extraction experiment.
-
-@Cheong2016 tested multiple different visual representations of uncertainty for representing the likelihood of a house being burned down based on its location.
-Their payment scheme, which paid out $0.10 for a correct choice (i.e.
-staying when the house was not burned down or leaving when the house was burned down) and 0 for an incorrect choice (i.e.leaving when the house didn't burn down or staying when the house burned down), meant participants were incentivised to base their entire leave/stay decision on whether or not the likelihood of a fire at their house is above or below 50%.
-If the participants *did* correctly identify the optimum strategy and answer accordingly, the "decision making experiment" was actually *just a value extraction experiment*.
-The decision making aspect ceased to matter, all that mattered was identifying if the probability of a fire was greater than 50%.
-Interestingly, despite this very obvious and simple tactic to maximize payout from the experiment, it seems like many participants did not adopt it, instead acting much more warily as though they were considering whether or not they would *actually* evacuate in the event of a fire.
-This means it is likely that introducing a payment scheme or a utility function creates too much mental labour for the participants to answer "correctly", and they would rather stick with their own risk utility function that feels *natural* than go through the mental labour of adopting another.
-Setting up an experiment that requires participants to do laborious calculations or strenuous mental effort for a small marginal benefit is unlikely to result in successful participant responses because it is not in the "spirit" of visualisation.
-(*Cite Gap 16: Examples of value extraction decision making tasks*)
-
-It is clear that audiences primarily interact with visualisations through their "System 1" brain [@daniel2017thinking].
-This aligns with other authors comments that visualisation is primarily about "gists" [Spiegelhalter2017].
-In this sense, asking participants a question that requires them to shift to their "System 2" is entirely contrary to *why* we use visualisation.
-The exact reasons we describe visualisation to be so powerful, that they are able to communicate complicated information fast and efficiently *is a product of the system 1 brain* [@daniel2017thinking], however this means displaying information using visualisations have the same weakness' as system 1 thinking, and visualisation authors need to be aware of this.
-Asking participants to answer questions that require complicated calculations does not reflect how visualisations are used, and will likely be ignored by the participants.
-While we agree with @Hullman2016 who suggested that what determines an appropriate ground truth is largely a philosophical exercise, we differ in the sense that we do not think it should be done, lest you create a needlessly complicated value extraction task.
-
-Additionally, several papers mention "risk-aversion" in participant decisions as a negative by-product of a particular visualisation design, however that displays a failure to understand that risk aversion is *not* a mistake, but rather it *reveals the utility of decreased uncertainty*.
-It may be interesting to investigate why a particular visualisation elicits increased risk-aversion, for example is it due to poorer estimates or increased awareness of negative outcomes, however simply reporting this as "risk aversion" conflates these factors.
-The work that advocates for transparency in risk-communication gives you whiplash when it's subtext argues that transparency leads to "incorrect" conclusions.
-There have been other discussions on the appropriateness of discussing risk-aversion as a bias [@Vranas2000], but few have made the point that if uncertainty holds some *value* there is not technically a "right" decision at all.
-
-The final issue with decision making tasks is that calculating "optimum" choices is not a task we use uncertainty for, it is a task that utilises *risk*.
-Risk and uncertainty are slightly different, as risk is known probabilities and uncertainty is unknown probabilities, so uncertainty is what you get when you cannot accurately define risk [@Spiegelhalter2017].
-Communicating risk allows people to weigh up options and make an optimum decision, uncertainty hedges information to let people know there may not even be an optimum strategy.
-In this sense, many "uncertainty" decision making experiments are actually just risk decision making experiments.
-Not only this, but the inclusion of uncertainty information should *not* impact the choices of a rational agent.
-A well behaved rational agent *should* ignore uncertainty information, so long as the estimate provided is still the best case.
-This point seems to be lost on many uncertainty visualisation authors.
-@Zhao2023 displayed a model's prediction and its estimated uncertainty and asked participants if they wanted to submit the model estimate as their answer, or make their own prediction.
-At no point did the authors seem to realise a rational agent would accept the model estimate every time, no matter how uncertain the information was, and participants were actually disqualified from the study for taking this approach to answering the questions.
-(*Cite Gap 17: Examples of studies where the correct choice was to ignore the uncertainty information*)
-
-This leaves us with several issues in decision making experiments.
-They are noisy because of participants individual utility functions and when that is not the case they are thinly veiled and over complicated value extraction experiments.
-Additionally, authors don't even seem to agree how participants should incorporate uncertainty information, with some authors designing complicated utility functions, and others not realising the best response to their experiment is to ignore uncertainty information entirely.
-For these reasons, uncertainty visualisation experiments should either be designed with these caveats in mind and be designed to untangle why a particular visualisation will elicit a specific decision, or the field should steer away using decision quality and related metrics as a basis for evaluating visualisation performance.
-
-#### Trust and confidence
-Authors also often measure the impact of visualising uncertainty on participants trust in an estimate.
-Trust is a by product of displaying uncertainty rather than the goal of it, and viewing the relationship in the converse direction can lead to misguided research.
-
-If the purpose of displaying uncertainty information is to appropriately hedge a signal with noise, then it should be assumed that trust is only related to uncertainty communication through increased transparency and honesty.
-Considering trust, and not transparency, as the metric of importance in uncertainty communication can lead to a questionable subtext that argues against transparency, something that has been noticed by several other authors [@Spiegelhalter2017; ONeill2018].
-@Hullman2020a found that author simultaneously argued that failing to visualise uncertainty was akin to fraud, but also many avoided uncertainty visualisation because they didn't want their work to come across as "untrustworthy".
-These authors are optimising *trust* rather than *transparency*, which means they opt to leave out uncertainty information when it does exactly what it is supposed to, decrease certainty in conclusions.
-
-Science communication should be primarily concerned with accuracy.
-Setting trust and risk-aversion as the variables of interest implicitly encourages statisticians to set trust and risk-aversion as the primary goals of communication.
-The issue of trust being divorced from trustworthiness has been commented on by other authors [@ONeill2018], however the issue still persists in the uncertainty visualisation literature.
-@Zhao2023 displayed a several variations of a visualisation of a model prediction and its uncertainty and took participants using the model prediction as a sign of trust.
-They reported that visualising uncertainty information caused participants to trust the model in the low variance case, but the results in the high variance case were inconclusive.
-The discussion made it clear the authors thought the uncertainty information should make the visualisation more trustworthy, but conflating trust and the use of a prediction implied uncertainty information should somehow influence participants to use their own prediction, even though a prediction being uncertainty does not necessarily mean it is incorrect.
-Despite this, the authors seemed to assume that the uncertainty information *should* have an influence on that, showing they had not deeply considered *how* uncertainty information should influence the choices of the participants.
-(*Cite Gap 18: examples of studies where authors measure trust*)
-
-A similar measure to trust is using "confidence" in an extracted value or a decision.
-Interestingly, "confidence" is also used to try and capture the clarity of a message in a normal visualisation.
-Confidence cannot simultaneously be a measure of clarity of visualisation *and* a way to capture the uncertainty expressed in a visualisation.
-
-#### Other (questionable) attempts to capture uncertainty
-There is also a swath of studies that are aware a question that boils down to a value extraction experiment, or a question that should be answered by literally ignoring uncertainty information is not what we want when we consider uncertainty visualisations.
-These papers often try to ask a question that should utilise both the uncertainty and signal in the response, however this is rarely what actually occurs.
-This method typically results in in cryptic or confusing questions that create a large amount of noise on the interpretation side of the analysis [@Hullman2016].
-
-Some authors opt for asking slightly vague questions that imply a use of uncertainty, but compare it to a ground truth that is very specific.
-@Ibrekk1987 asked participants for the "best estimate" which was evaluated in accuracy by comparing it to the mean, however the "best estimate" depends on the loss function we are using, and a loss function of minimised error was not implied by the question.
-@Hofmann2012 showed two distributions in 20 different visualisations (a line-up protocol) using a jittered sample, a density plot, a histogram, and a box plot and asked participants.
-Participants were asked to report in which of the plots was "the blue group furthest to the right" The experiment set up is shown in @fig-right.
-The participants answers were then compared to a ground truth where the correct plot had a blue distribution with a right shifted mean.
-By comparing the results to a ground truth statistic and marking participants as "wrong" or "right", the error from the participants that had an alternative interpretation to the concept of "furthest right" was conflated with the error from a the visualisation choice.
-These papers make it unclear if the participants got the answers wrong because they misunderstood the question or because of something related to the plot.
-Therefore, this method leads to inconclusive results about the plot design, and is not advised.
-
-![This shows the user interface for the experiment performed by @Hofmann2012. The question of "furthest to the right" is open to interpretation.](furthestright.png){#fig-right width=50%}
-
-Another method used by authors is to ask a deterministic question about a random event.
-@Padilla2017 provided participants with a visualisation of the cone of uncertainty and asked then to "decide which oil rig will receive more damage based on the depicted forecast of the hurricane path".
-The cone of uncertainty provides a 60% confidence interval for the location of the eye of a hurricane, which allows us to know the area where the eye of the storm will go, it does not given any information about the intensity of a storm, the size of a storm, or even if a location will be hit.
-This inclusion of determinism seems to cause the authors to stumble themselves, as they are not consistent with their assumptions.
-In their first experiment @Padilla2017 indicated the correct answer was to assume that the storm was equally intense no matter how far from the centre of the distribution an oil rig was, however answering their third experiment correctly hinged on assuming the intensity of the storm at a particular point (which in this experiment they phrased as damage) *does* change in intensity as you move away from the centre of the distribution.
-Given these conflicting assumptions, it is unclear how the participants were supposed to adjust the probabilistic path information to answer a deterministic question about which oil rig would receive the most damage.
-Other authors have commented on the complexity of communicating hurricane risk because the path, storm surge and wind speed are all important and cannot be ignored [@Spiegelhalter2017].
-The flip side of this is asking participants for a deterministic answer to a probabilistic question.
-@Correll2014 asked participants "how likely is candidate B to win the election?" when the two distributions indicated voter preference.
-Participants were not able to answer the question about likelihood in term of probability, but were instead given seven options from 1=Outcome will be most in favour of A to 7=Outcome will be most in favour of B.
-The ground truth statistic for this question was a scalar multiple of Cohen’s d, indicating participants were supposed to incorporate uncertainty information using a very specific formula that was likely unknown to them but assumed to be used implicitly.
-
-The final method used by authors is to just explicitly ask about uncertainty and signal information separately.
-@Sanyal2009 mapped uncertainty to dots and signal to a 3D surface and asked participants to identify areas of high and low signal and high and low uncertainty.
-Participants were not asked to combine that information in any way, and the signal and the noise were treated as separate variables.
-@Correll2014 asked participants to separately extract the mean and variance from four uncertainty visualisations.
-These methods explicitly view the uncertainty and signal as two separate variables that should be extracted from a plot, and not two variables that should be interpreted together.
-Even viewing these questions as a routine check to make sure the signal information isn't impacted by the uncertainty is counter intuitive, because the whole point *of* the uncertainty is to impact the signal information.
-
-These examples are a bit complete mishmash of methods, however they point to a larger issue that goes beyond decision making, trust and confidence experiments.
-Authors *have no idea* how to evaluate the effects of uncertainty in an uncertainty visualisation.
+An alternative way authors consider describing an uncertainty visualisation is to use taxonomies.
+However, these taxonomies tend to blur the line between the mathematical elements of uncertainty and the visual depiction of those mathematical objects.
+Some taxonomies of uncertainty highlight how confusing most authors find the concept, and how difficult it is to articulate the rules of an uncertainty visualisation when uncertainty itself if not defined.
+@Kinkeldey2014 categorised uncertainty according to five criteria depicted in @fig-tax which considers if a visualisation is implicit or explicit, intrinsic or extrinsic, visually integrable or seperable, coincident or adjacent, and static or dynamic.
+A similar version of this taxonomy was presented by @uncertchap2022 who commented that visualisation can be organised into two categories, "graphical annotations of distributional properties" and "visual encodings of uncertainty" which seems to functionally align with the intrinsic/extrinsic distinction by @Kinkeldey2014.
+@Griethe2006 organised uncertainty visualisations into two cases (1) a hypothesis test was preformed to confirm the validity of the visualisation and (2) the visualisation has uncertainty depicted.
+@Potter2012 organised several existing uncertainty visualisations into groups based on the dimensionality of the data (1D, 2D, 3D, and No Dimension) and the dimensionality of the uncertainty (Scalar, Vector, Tensor).
+However, because the term "PDF", function that is used to describe a random variable, is used to describe both the data and the uncertainty for all dimensions.
+@Grewal2021 created a taxonomy that mapped uncertainty visualisations to some point in a 2D space defined by the "domain expertise" and "continuum of discreteness" (that scaled from "point estimate" to "continuous distribution").
+
+![Visualisation of @Kinkeldey2014 uncertainty visualisation taxonomy.
+It can be seen that most of the categories can be considered a change to the typical error bar chart.
+The taxonomy categorises the visualisations based on five criteria, implicit (a sample) or explicit (a depiction of mass); Intrinsic (alter existing symbols symbols to represent uncertainty) or extrinsic (add new objects to represent uncertainty); visually integral or separable (the uncertainty can be separated from the data and read independently) ; coincidence (uncertainty and data are represented in the same plot) or adjacent; static or dynamic (animated, interactive etc).
+Many of the distinctions conflate changes of the input data and statistics calculated and with visual changes, highlighting how defining uncertainty and visualisation together conflates several elements of a graphic.](uncertvistax.jpeg){#fig-tax}
+
+The authors of these taxonomies did not intend for them to be complete descriptions of uncertainty visualisations, however they do paint a picture of the what authors consider to be important.
+There are common themes in the taxonomies that ensure we differentiate between a sample and a mass, consider if uncertainty should be depicted with the estimate or as a second variable, identify the parallel of hypothesis testing, and consider the dimensionality, sample size and precision depicted by the visualisation.
+These taxonomies may highlight perceived important differences, but they do not help us understand what *is* or *is not* an uncertainty visualisation.
+For that we need to understand *why* uncertainty visualisation exists as a sub-field at all.
+
+## Uncertainty visualisations should perform signal supression
+It is rarely made clear exactly how uncertainty visualisation is different to normal information visualisation.
+Visualising a graphic is nothing more than computing a statistic, mapping that statistic to a visual feature and rendering that graphic.
+Nothing in this process suggests visualising an estimate of the variance should be any different to visualising an estimate of the mean, however this field *does* implicitly treat them different.
+Why?
+
+Uncertainty visualisation papers usually provide one of two primary justifications for their existence.
+The first justification is that uncertainty is fundamentally differently to other variables due to the psychological heuristics involved in interpreting uncertainty.
+Therefore uncertainty visualisation is different to normal visualisation as authors must consider the psychological effects of what they are visualising [@Spiegelhalter2017; @Hullman2019].
+The second justification assets that uncertainty is an additional variable and it is of vital importance to interpreting an estimate.
+Therefore uncertainty visualisation is a high dimensional visualisation problem, as we need to figure out how to seamlessly add uncertainty into already existing graphics [@moritz2017trust; @Griethe2006].
+Why some authors pick one motivation over another is rarely explained and both leave the role of uncertainty in the visualisation unclear.
+The literature never explains if uncertainty should be treated as a variable, as metadata, or as something else entirely [@Kinkeldey2014].
+@Hullman2016 commented that it is straightforward to show a value but it is much more complex to show uncertainty, but did not explain *why* Interestingly, these motivations do not remotely line up with the most common reason that people *outside* the uncertainty visualisation community thinks visualising uncertainty is important.
+The true reason that uncertainty visualisation is differerent to normal information visualisation, is that normal information visualisation is designed to *find* signals in our data, while uncertainty visualisation should be designed to *supress* it.
+
+Interviews with experts in visualisation and statistics often back up "signal suppression" as the primary motivation for uncertainty visualisation.
+Failling to communicate uncertainty is often considered to be akin to fraud or lying [@Hullman2020a; @Manski2020] indicating that uncertainty should exist to prevent undue certainty in visual inference.
+The relationship between signal supression and uncertainty visualisation is made more explicity by other authors [@Griethe2006].
+Therefore uncertainty acts as a form of statistical "hedging" for signals found in an analysis.
+This aligns nicely with our understanding of uncertainty as a by product of inference.
+The variable we are trying to perform inference on is the signal, while the while the visualisation of noise that prevents the signal from being read is the uncertainty.
+It is nice that the connection to inference easily extends from uncertainty to uncertainty visualisation, however many of the confusions extend as well.
+
+There are two primary reasons for making a visualisation, to perform exploratory data analysis (EDA) or communication.
+Communication involves identifying a signal we want to communicate and designing the visualisation that best conveys that, while EDA involves creating a versatile visualisation without an explicit purpose and using it to extract several signals.
+Since EDA is the visual parallel to descriptive statistics, it is performed without an explicit hypothesis which means there *is* no uncertainty in the visualisation.
+Similarly to our issues with descritive statistics, this is not well understood by the uncertainty visualisation community.
+Some authors simply confuse varianc for uncertainty.
+For example @Potter2010 aimed to create a summary plot that "concisely presented data with uncertainty information" to create an exploratory visualisation tool that visualised uncertainty.
+Other authors regonise inference will occur (in some shape or form) and believe uncertainty *should* be visualised but do not recognise *how* uncertainty would be visualised.
+@Hullman2021 argued that there is no such thing as a "model-free" visualisation, therefore visualisation require robust visualisations of uncertainty as we are always performing inference.
+@Griethe2006 commented that "if visualization is used as a means to explore a data volume or to communicate its contents the uncertainty has to be included".
+While we agree people cannot prevent themselves from performing inference, that does not mean it is possible for uncertainty to be included in an EDA visualisation.
+A versatile visualisation such as a scatter plot allows for a viewer to consider several hypothesis at once, each of which will be a different inferential statistic with a different distribution that depicts its uncertainty.
+It is likely impossible to suppress all possible signals at once.
+
+This leads us to wonder *how* signal suppression should be tested.
 
@@
 
+## Mistakes made when we misunderstand uncertainty visualisation
+In the previous section, we highlighted several problematic experimental designs or pieces of research that were the result of misunderstandings around the definition of uncertainty.
+Here, we will discuss the issues that arise due to a misunderstanding of "uncertainty visualisation".
+The two primary problems are the regular comparisons of mathematically incomparable plots, and reproducing results that are already widely understood in the larger information visualisation literature.
+
+### Repeating perceptual task experiments
+Given the previous section, one might consider problems in uncertainty visualisation to be fixed if we could restrict evaluation experiments to cases where the information in each graphic is almost identical.
+Unfortunately this still often results in uninteresting or obvious results.
+
+If two graphics are visually different but identical in the information they contain, they must differ in how that information is depicted. 
+Once we have a mathematical expression of uncertainty, the visualisation of that uncertainty is theoretically identical to the visualisation process of any other variable.
+For simple tasks such as value extraction, there is a hierarchy to perceptual tasks where extracting visual information in some forms is easier than others.
+The hierarchy was originally established 40 years ago by @Cleveland1984, below is an updated version summarised by @Vanderplas2020:
+
+1) Position along a common scale. 
+2) Position along a non-aligned scale. 
+3) Length, direction, angle, slope
+4) Area
+5) Volume, density, curvature
+6) Shading, colour saturation, colour hue
+7) Discriminable shape
+8) Indiscriminable shape
+
+This hierarchy is a good general rule, however it can change from person to person [@Davis2022] Additionally, there are other graphical rules to consider such as gestalt principles, broader methods of extraction, and attention principles [@Vanderplas2020].
+These established visualisation concepts allow us to anticipate the ease with which certain pieces of information will be extracted from a plot.
+We can use these concepts to understand the computational complexity of a graphic.
+A bizarre feature of the uncertainty visualisation literature is that it does not work to build upon these existing principles or identify the ways in which uncertainty visualisations may diverge from these rules.
+These building block concepts of visualisation are seldom mentioned.
+
+It is difficult to find examples of uncertainty visualisation experiments where the plots do contain the same information, however when they do, the results align with existing information visualisation research.
+Technically, a PDF and a mean with confidence intervals both have enough information to extract the mean of the distribution, however they both have a very different computational cost.
+To extract the mean using a PDF, a participant would need to identify the point along the x axis that splits the area under the curve in half.
+If a participant is provided with a mean with a confidence interval, extracting the average is a simple task of reading the position on an aligned scale.
+@Ibrekk1987 found that when asking participants for the "best estimate" (which they thought should be interpreted as the mean of the distribution) of a skewed distribution, participants provided the mean when given a mean with confidence intervals and the mode when given a PDF [@Ibrekk1987].
+Similar results to this occur over and over again in the uncertainty visualisation literature.
+@Gschwandtnei2016 found that visualisations where the start time of an interval could literally be read off the plot (error bars, centred error bars, and ambiguation) performed better than the plots (accumulated probability, gradient, and violin) where the start time involved some guesswork because the drop off was gradual.
+@Cheong2016 found that participants were better at answering questions when they were explicitly given the relevant probability in text rather than when they needed to read it off a map.
+(*Cite Gap 15: Examples of replicated perceptual task experiments*).
+
+These results show that uncertainty is not technically different to any other variable.
+When trying to anticipate the results of these studies, we can use the same principles of information equivalence and difficulty of relevant mental tasks to understand which plots will outperform others.
+This leads us to wonder...
+why is uncertainty visualisation a field at all?
+
+# How do we measure uncertainty?
+If uncertainty is inherently different, we would expect evaluation studies of uncertainty visualisations to produce different results to similar evaluation studies on "normal" variables.
+This does not seem to be the case.
+Uncertainty is noise, but most evaluation experiments measure it as signal.
+The previous section highlights a unique and fascinating problem faced by uncertainty visualisation.
+If asking direct questions about uncertainty causes us to treat it as a signal, how do we evaluate uncertainty as *noise*?
+ When we ask the viewer of a plot to look at data and extract a value, we are asking them to perform inference on that value.
+There will be noise associated with that answer and that is uncertainty.
+If we ask direct question about some uncertainty metric, we have turned the uncertainty into signal because that is what the participants are drawing inference on.
+
+Uncertainty, acting as it is intended, is transparency.
+There are many secondary benefits that come with this improved transparency, such as better decisions, more trust in the results and more confidence in the authors.
+These secondary benefits are, however, *not* the immediate goal of uncertainty.
+The following sections will discuss the issues and limitations in measuring uncertainty through these secondary metrics and provide suggestions as to how future studies should consider measuring uncertainty.
+
+## Measuring uncertainty as signal
+Uncertainty visualisation papers can be organised according to the *goal* of the experiment.
+Evaluation experiments are the standard rule for visualisations because the human brain is not as reliable as mathematical calculation.
+Therefore, user studies often aim to assess the limitations, biases, and heuristics of our mental calculator so that we can better understand the problems we may encounter when we plot our data.
+This is not to say any paper that suggests a visualisation without an evaluation experiment is completely lacking in justification, and there are many papers that suggest a novel visualisation without an evaluation study.
+Sometimes these papers are a preliminary step in finding a solution for common problems and intend to evaluate the visualisation in later work.
+The reasoning for this is obvious.
+There are often heuristics and biases that are not obvious to us when designing visualisation.
+Additionally, these heuristics and biases can change depending on the larger scope of the graphic and the population we are communicating with [@Spiegelhalter2017; @Kinkeldey2014].
+
+The overarching goal of most uncertainty visualisation evaluation papers belong to one of: performance (how effectively a participant can extract information from a plot), interpretation and semantics (the ease with which a particular visualisation is associated with uncertainty), and quality of user experience (if the participants liked the plot or not)[@Hullman2019].
+@Hullman2019 found that the majority of papers evaluate visualisations on performance (approximately 65% of the papers they surveyed) or interpretation and semantics (approximately 17%) and both of these evaluation goals will run into problems because of their conceptualisation of uncertainty.
+
+### Performance
+There are many metrics experimenters use to evaluate the performance of a visualisation.
+Each metric, such as can use accuracy, decision quality, confidence, trust and others each have their own issues which prevent them from accurately capturing the effects of uncertainty in a visualisation.
+
+By far the most common metric used is accuracy, which is used by approximately 36% of evaluation studies [@Hullman2019].
+In previous sections, we pointed out issues that arise when accuracy (of extracting probability or variance) is used as the primary metric in evaluating uncertainty visualisations, the work simply replicates known concepts in information visualisation.
+Subjective metrics, such as how much participants liked the aesthetic representation of a visualisation, or metrics with an unclear motivation or use, such as how memorable the information in a plot is or how much a user interacted with a plot, are ignored in this section.
+Below we will explain in detail why experiments that try to measure uncertainty using methods such as decision making, trust, or other types of questions, also fail to capture the effects of visualising uncertainty in a plot.
+
+#### Decision making and risk aversion
+
+![Visualisation of the process of transforming raw data to a visualisation insight.
+Summarising the data using common statistics removes information from our data.
+Once the data is in a format that represents the varibales of interest, organising that information in a visualisation makes it easier to extract insights from these statistics and convert them to insights.
+If information is removed at the statistics stage, it cannot be added back in at the visualisation stage, as it is meerly an efficient technique to organise and present information to enable a large range of insights.
+](visprocess.jpeg){#fig-location width=70%}
+
+Decision making tasks are often described by authors as a more realistic version how plots are used in practice.
+While we do not disagree with this statement, exactly *why* these tasks are different to value extraction tasks is never explained.
+While the task may be more similar to how plots are actually used, that does not necessarily mean it is a better *experimental* environment for evaluating plots.
+Let us consider this in more detail.
+A normal value extraction task involves:
+
+1) correctly interpret the question
+2) extract the specified value
+3) report the value
+
+On the other hand, a decision making task using an uncertainty visualisation involves:
+
+1) Correctly interpret the question
+2) Use some risk utility function to set a threshold for an "acceptable level of risk"
+3) Extract the value for from the plot that will be compared to the threshold
+4) Make a judgement based off that value
+
+The key aspect that separates a decision making task from a value extraction task, for uncertainty visualisations, is the inclusion of a utility function.
+While this might seem like a small change, it actually completely warps the experiment in unexpected.
+
+The first issue is that individuals will all have their own individual risk utility function, that is, how much they want to avoid or engage with risks or uncertainty.
+Therefore decision making experiments that have an additional layer of noise that value extraction experiments do not.
+We cannot be sure if participants answered differently from the ground truth because a visualisation was difficult to read *or* because the participants risk utility function did not align with the one set by the authors.
+Several authors have offered solutions to this issue, however the problem is deeper than any of them seem to realise.
+
+@Hullman2016 suggested providing a utility framework for each experiment, to instruct the participants in how to account for the uncertainty information.
+This is a method that seems to have been adopted by @Fernandes2018 who describe the following scenario to participants who are trying to maximise the coins in their experiment:
+> Subjects gain coins for every minute they are able to continue an activity that is valuable to them (e.g., watching TV at home) before going to the bus stop, and gain a bonus for arriving at their intended destination early.
+Subjects incur a coin penalty for time spent waiting at a stop for a bus to arrive.
+
+Any payment scheme that is incorporated into an experiment will also implicitly set up a utility framework, as it is assumed participants will try to maximise their payments.
+In tasks such as this, the ground truth of each question is typically selected to be the value a ration agent would select and visualisations are evaluated on the basis of how far the participants responses are from that ground truth.
+The problem with this, is that all the complexity of the question is in *working out* the best response, and it has little to do with the visualisation itself.
+The visualisation aspect of these studies becomes a value extraction experiment.
+
+@Cheong2016 tested multiple different visual representations of uncertainty for representing the likelihood of a house being burned down based on its location.
+Their payment scheme, which paid out $0.10 for a correct choice (i.e.
+staying when the house was not burned down or leaving when the house was burned down) and 0 for an incorrect choice (i.e.leaving when the house didn't burn down or staying when the house burned down), meant participants were incentivised to base their entire leave/stay decision on whether or not the likelihood of a fire at their house is above or below 50%.
+If the participants *did* correctly identify the optimum strategy and answer accordingly, the "decision making experiment" was actually *just a value extraction experiment*.
+The decision making aspect ceased to matter, all that mattered was identifying if the probability of a fire was greater than 50%.
+Interestingly, despite this very obvious and simple tactic to maximize payout from the experiment, it seems like many participants did not adopt it, instead acting much more warily as though they were considering whether or not they would *actually* evacuate in the event of a fire.
+This means it is likely that introducing a payment scheme or a utility function creates too much mental labour for the participants to answer "correctly", and they would rather stick with their own risk utility function that feels *natural* than go through the mental labour of adopting another.
+Setting up an experiment that requires participants to do laborious calculations or strenuous mental effort for a small marginal benefit is unlikely to result in successful participant responses because it is not in the "spirit" of visualisation.
+(*Cite Gap 16: Examples of value extraction decision making tasks*)
+
+It is clear that audiences primarily interact with visualisations through their "System 1" brain [@daniel2017thinking].
+This aligns with other authors comments that visualisation is primarily about "gists" [Spiegelhalter2017].
+In this sense, asking participants a question that requires them to shift to their "System 2" is entirely contrary to *why* we use visualisation.
+The exact reasons we describe visualisation to be so powerful, that they are able to communicate complicated information fast and efficiently *is a product of the system 1 brain* [@daniel2017thinking], however this means displaying information using visualisations have the same weakness' as system 1 thinking, and visualisation authors need to be aware of this.
+Asking participants to answer questions that require complicated calculations does not reflect how visualisations are used, and will likely be ignored by the participants.
+While we agree with @Hullman2016 who suggested that what determines an appropriate ground truth is largely a philosophical exercise, we differ in the sense that we do not think it should be done, lest you create a needlessly complicated value extraction task.
+
+Additionally, several papers mention "risk-aversion" in participant decisions as a negative by-product of a particular visualisation design, however that displays a failure to understand that risk aversion is *not* a mistake, but rather it *reveals the utility of decreased uncertainty*.
+It may be interesting to investigate why a particular visualisation elicits increased risk-aversion, for example is it due to poorer estimates or increased awareness of negative outcomes, however simply reporting this as "risk aversion" conflates these factors.
+The work that advocates for transparency in risk-communication gives you whiplash when it's subtext argues that transparency leads to "incorrect" conclusions.
+There have been other discussions on the appropriateness of discussing risk-aversion as a bias [@Vranas2000], but few have made the point that if uncertainty holds some *value* there is not technically a "right" decision at all.
+
+The final issue with decision making tasks is that calculating "optimum" choices is not a task we use uncertainty for, it is a task that utilises *risk*.
+Risk and uncertainty are slightly different, as risk is known probabilities and uncertainty is unknown probabilities, so uncertainty is what you get when you cannot accurately define risk [@Spiegelhalter2017].
+Communicating risk allows people to weigh up options and make an optimum decision, uncertainty hedges information to let people know there may not even be an optimum strategy.
+In this sense, many "uncertainty" decision making experiments are actually just risk decision making experiments.
+Not only this, but the inclusion of uncertainty information should *not* impact the choices of a rational agent.
+A well behaved rational agent *should* ignore uncertainty information, so long as the estimate provided is still the best case.
+This point seems to be lost on many uncertainty visualisation authors.
+@Zhao2023 displayed a model's prediction and its estimated uncertainty and asked participants if they wanted to submit the model estimate as their answer, or make their own prediction.
+At no point did the authors seem to realise a rational agent would accept the model estimate every time, no matter how uncertain the information was, and participants were actually disqualified from the study for taking this approach to answering the questions.
+(*Cite Gap 17: Examples of studies where the correct choice was to ignore the uncertainty information*)
+
+This leaves us with several issues in decision making experiments.
+They are noisy because of participants individual utility functions and when that is not the case they are thinly veiled and over complicated value extraction experiments.
+Additionally, authors don't even seem to agree how participants should incorporate uncertainty information, with some authors designing complicated utility functions, and others not realising the best response to their experiment is to ignore uncertainty information entirely.
+For these reasons, uncertainty visualisation experiments should either be designed with these caveats in mind and be designed to untangle why a particular visualisation will elicit a specific decision, or the field should steer away using decision quality and related metrics as a basis for evaluating visualisation performance.
+
+#### Trust and confidence
+Authors also often measure the impact of visualising uncertainty on participants trust in an estimate.
+Trust is a by product of displaying uncertainty rather than the goal of it, and viewing the relationship in the converse direction can lead to misguided research.
+
+If the purpose of displaying uncertainty information is to appropriately hedge a signal with noise, then it should be assumed that trust is only related to uncertainty communication through increased transparency and honesty.
+Considering trust, and not transparency, as the metric of importance in uncertainty communication can lead to a questionable subtext that argues against transparency, something that has been noticed by several other authors [@Spiegelhalter2017; ONeill2018].
+@Hullman2020a found that author simultaneously argued that failing to visualise uncertainty was akin to fraud, but also many avoided uncertainty visualisation because they didn't want their work to come across as "untrustworthy".
+These authors are optimising *trust* rather than *transparency*, which means they opt to leave out uncertainty information when it does exactly what it is supposed to, decrease certainty in conclusions.
+
+Science communication should be primarily concerned with accuracy.
+Setting trust and risk-aversion as the variables of interest implicitly encourages statisticians to set trust and risk-aversion as the primary goals of communication.
+The issue of trust being divorced from trustworthiness has been commented on by other authors [@ONeill2018], however the issue still persists in the uncertainty visualisation literature.
+@Zhao2023 displayed a several variations of a visualisation of a model prediction and its uncertainty and took participants using the model prediction as a sign of trust.
+They reported that visualising uncertainty information caused participants to trust the model in the low variance case, but the results in the high variance case were inconclusive.
+The discussion made it clear the authors thought the uncertainty information should make the visualisation more trustworthy, but conflating trust and the use of a prediction implied uncertainty information should somehow influence participants to use their own prediction, even though a prediction being uncertainty does not necessarily mean it is incorrect.
+Despite this, the authors seemed to assume that the uncertainty information *should* have an influence on that, showing they had not deeply considered *how* uncertainty information should influence the choices of the participants.
+(*Cite Gap 18: examples of studies where authors measure trust*)
+
+A similar measure to trust is using "confidence" in an extracted value or a decision.
+Interestingly, "confidence" is also used to try and capture the clarity of a message in a normal visualisation.
+Confidence cannot simultaneously be a measure of clarity of visualisation *and* a way to capture the uncertainty expressed in a visualisation.
+
+#### Other (questionable) attempts to capture uncertainty
+There is also a swath of studies that are aware a question that boils down to a value extraction experiment, or a question that should be answered by literally ignoring uncertainty information is not what we want when we consider uncertainty visualisations.
+These papers often try to ask a question that should utilise both the uncertainty and signal in the response, however this is rarely what actually occurs.
+This method typically results in in cryptic or confusing questions that create a large amount of noise on the interpretation side of the analysis [@Hullman2016].
+
+Some authors opt for asking slightly vague questions that imply a use of uncertainty, but compare it to a ground truth that is very specific.
+@Ibrekk1987 asked participants for the "best estimate" which was evaluated in accuracy by comparing it to the mean, however the "best estimate" depends on the loss function we are using, and a loss function of minimised error was not implied by the question.
+@Hofmann2012 showed two distributions in 20 different visualisations (a line-up protocol) using a jittered sample, a density plot, a histogram, and a box plot and asked participants.
+Participants were asked to report in which of the plots was "the blue group furthest to the right" The experiment set up is shown in @fig-right.
+The participants answers were then compared to a ground truth where the correct plot had a blue distribution with a right shifted mean.
+By comparing the results to a ground truth statistic and marking participants as "wrong" or "right", the error from the participants that had an alternative interpretation to the concept of "furthest right" was conflated with the error from a the visualisation choice.
+These papers make it unclear if the participants got the answers wrong because they misunderstood the question or because of something related to the plot.
+Therefore, this method leads to inconclusive results about the plot design, and is not advised.
+
+![This shows the user interface for the experiment performed by @Hofmann2012. The question of "furthest to the right" is open to interpretation.](furthestright.png){#fig-right width=50%}
+
+Another method used by authors is to ask a deterministic question about a random event.
+@Padilla2017 provided participants with a visualisation of the cone of uncertainty and asked then to "decide which oil rig will receive more damage based on the depicted forecast of the hurricane path".
+The cone of uncertainty provides a 60% confidence interval for the location of the eye of a hurricane, which allows us to know the area where the eye of the storm will go, it does not given any information about the intensity of a storm, the size of a storm, or even if a location will be hit.
+This inclusion of determinism seems to cause the authors to stumble themselves, as they are not consistent with their assumptions.
+In their first experiment @Padilla2017 indicated the correct answer was to assume that the storm was equally intense no matter how far from the centre of the distribution an oil rig was, however answering their third experiment correctly hinged on assuming the intensity of the storm at a particular point (which in this experiment they phrased as damage) *does* change in intensity as you move away from the centre of the distribution.
+Given these conflicting assumptions, it is unclear how the participants were supposed to adjust the probabilistic path information to answer a deterministic question about which oil rig would receive the most damage.
+Other authors have commented on the complexity of communicating hurricane risk because the path, storm surge and wind speed are all important and cannot be ignored [@Spiegelhalter2017].
+The flip side of this is asking participants for a deterministic answer to a probabilistic question.
+@Correll2014 asked participants "how likely is candidate B to win the election?" when the two distributions indicated voter preference.
+Participants were not able to answer the question about likelihood in term of probability, but were instead given seven options from 1=Outcome will be most in favour of A to 7=Outcome will be most in favour of B.
+The ground truth statistic for this question was a scalar multiple of Cohen’s d, indicating participants were supposed to incorporate uncertainty information using a very specific formula that was likely unknown to them but assumed to be used implicitly.
+
+The final method used by authors is to just explicitly ask about uncertainty and signal information separately.
+@Sanyal2009 mapped uncertainty to dots and signal to a 3D surface and asked participants to identify areas of high and low signal and high and low uncertainty.
+Participants were not asked to combine that information in any way, and the signal and the noise were treated as separate variables.
+@Correll2014 asked participants to separately extract the mean and variance from four uncertainty visualisations.
+These methods explicitly view the uncertainty and signal as two separate variables that should be extracted from a plot, and not two variables that should be interpreted together.
+Even viewing these questions as a routine check to make sure the signal information isn't impacted by the uncertainty is counter intuitive, because the whole point *of* the uncertainty is to impact the signal information.
+
+These examples are a bit complete mishmash of methods, however they point to a larger issue that goes beyond decision making, trust and confidence experiments.
+Authors *have no idea* how to evaluate the effects of uncertainty in an uncertainty visualisation.
+
 ## Suggestions to measure uncertainty

### 2024-05-13 writingdumpster.qmd
fixed up other bits

@@
 # Removed Sections
+The exploration step of an analysis, which includes descriptive statistics, exploratory data visualisation, and unsupervised machine learning techniques, is performed without a prior hypothesis, however misunderstandings of this fact appears frequently in the literature.
+
+@wu2023rational recognised and tried to eliminate information asymmetry by looking at a visualisation and identifying what information can be extracted and then using a "rational agent benchmark" to determine how a rational person would use that information for a decision making task.
+While this solution is interesting, it may have different results depending on who is applying it.
+
+@Tierney2023 generates a visualisation for missing values, however because it fits nicely into the grammar of graphics framework, it is inherrently a descriptive statistic.
+
+Signal can be considered the information that *promotes* a message we are trying to infer and while noise is the information that *suppresses* that message, if that message is invalid.
+
+This is why quantifiable uncertainties, such as confidence intervals, dominate the literature while visualisations for more complicated sources of uncertainty, such as the effects of assumptions, imputed missing variables, and model choices, remain elusive.
+
 @Munzner2009 created a nested model for visualisation that highlighted how the first mistake that can be made in a visualisation is in the problem characterisation, and failing to do it well can cause downstream effects and damage the effectiveness of a visualisation.

### 2024-05-13 paper.qmd
minor edits + knitted paper

@@
 
-This leads us to wonder *how* signal suppression should be tested.
+{{< pagebreak >}}
+
+**The following sections are in the middle of being reorganized and rewritten so it will not flow on from the previous sections**
+
+{{< pagebreak >}}
 
@@
 
-## Mistakes made when we misunderstand uncertainty visualisation
-In the previous section, we highlighted several problematic experimental designs or pieces of research that were the result of misunderstandings around the definition of uncertainty.
-Here, we will discuss the issues that arise due to a misunderstanding of "uncertainty visualisation".
-The two primary problems are the regular comparisons of mathematically incomparable plots, and reproducing results that are already widely understood in the larger information visualisation literature.
+# How do we measure uncertainty?
+If uncertainty is inherently different, we would expect evaluation studies of uncertainty visualisations to produce different results to similar evaluation studies on "normal" variables.
+This does not seem to be the case.
+Uncertainty is noise, but most evaluation experiments measure it as signal.
+The previous section highlights a unique and fascinating problem faced by uncertainty visualisation.
+If asking direct questions about uncertainty causes us to treat it as a signal, how do we evaluate uncertainty as *noise*?
+ When we ask the viewer of a plot to look at data and extract a value, we are asking them to perform inference on that value.
+There will be noise associated with that answer and that is uncertainty.
+If we ask direct question about some uncertainty metric, we have turned the uncertainty into signal because that is what the participants are drawing inference on.
+
+Uncertainty, acting as it is intended, is transparency.
+There are many secondary benefits that come with this improved transparency, such as better decisions, more trust in the results and more confidence in the authors.
+These secondary benefits are, however, *not* the immediate goal of uncertainty.
+The following sections will discuss the issues and limitations in measuring uncertainty through these secondary metrics and provide suggestions as to how future studies should consider measuring uncertainty.
+
+## Measuring uncertainty as signal
+Uncertainty visualisation papers can be organised according to the *goal* of the experiment.
+Evaluation experiments are the standard rule for visualisations because the human brain is not as reliable as mathematical calculation.
+Therefore, user studies often aim to assess the limitations, biases, and heuristics of our mental calculator so that we can better understand the problems we may encounter when we plot our data.
+This is not to say any paper that suggests a visualisation without an evaluation experiment is completely lacking in justification, and there are many papers that suggest a novel visualisation without an evaluation study.
+Sometimes these papers are a preliminary step in finding a solution for common problems and intend to evaluate the visualisation in later work.
+The reasoning for this is obvious.
+There are often heuristics and biases that are not obvious to us when designing visualisation.
+Additionally, these heuristics and biases can change depending on the larger scope of the graphic and the population we are communicating with [@Spiegelhalter2017; @Kinkeldey2014].
+
+The overarching goal of most uncertainty visualisation evaluation papers belong to one of: performance (how effectively a participant can extract information from a plot), interpretation and semantics (the ease with which a particular visualisation is associated with uncertainty), and quality of user experience (if the participants liked the plot or not)[@Hullman2019].
+@Hullman2019 found that the majority of papers evaluate visualisations on performance (approximately 65% of the papers they surveyed) or interpretation and semantics (approximately 17%) and both of these evaluation goals will run into problems because of their conceptualisation of uncertainty.
 
@@
 why is uncertainty visualisation a field at all?
-
-# How do we measure uncertainty?
-If uncertainty is inherently different, we would expect evaluation studies of uncertainty visualisations to produce different results to similar evaluation studies on "normal" variables.
-This does not seem to be the case.
-Uncertainty is noise, but most evaluation experiments measure it as signal.
-The previous section highlights a unique and fascinating problem faced by uncertainty visualisation.
-If asking direct questions about uncertainty causes us to treat it as a signal, how do we evaluate uncertainty as *noise*?
- When we ask the viewer of a plot to look at data and extract a value, we are asking them to perform inference on that value.
-There will be noise associated with that answer and that is uncertainty.
-If we ask direct question about some uncertainty metric, we have turned the uncertainty into signal because that is what the participants are drawing inference on.
-
-Uncertainty, acting as it is intended, is transparency.
-There are many secondary benefits that come with this improved transparency, such as better decisions, more trust in the results and more confidence in the authors.
-These secondary benefits are, however, *not* the immediate goal of uncertainty.
-The following sections will discuss the issues and limitations in measuring uncertainty through these secondary metrics and provide suggestions as to how future studies should consider measuring uncertainty.
-
-## Measuring uncertainty as signal
-Uncertainty visualisation papers can be organised according to the *goal* of the experiment.
-Evaluation experiments are the standard rule for visualisations because the human brain is not as reliable as mathematical calculation.
-Therefore, user studies often aim to assess the limitations, biases, and heuristics of our mental calculator so that we can better understand the problems we may encounter when we plot our data.
-This is not to say any paper that suggests a visualisation without an evaluation experiment is completely lacking in justification, and there are many papers that suggest a novel visualisation without an evaluation study.
-Sometimes these papers are a preliminary step in finding a solution for common problems and intend to evaluate the visualisation in later work.
-The reasoning for this is obvious.
-There are often heuristics and biases that are not obvious to us when designing visualisation.
-Additionally, these heuristics and biases can change depending on the larger scope of the graphic and the population we are communicating with [@Spiegelhalter2017; @Kinkeldey2014].
-
-The overarching goal of most uncertainty visualisation evaluation papers belong to one of: performance (how effectively a participant can extract information from a plot), interpretation and semantics (the ease with which a particular visualisation is associated with uncertainty), and quality of user experience (if the participants liked the plot or not)[@Hullman2019].
-@Hullman2019 found that the majority of papers evaluate visualisations on performance (approximately 65% of the papers they surveyed) or interpretation and semantics (approximately 17%) and both of these evaluation goals will run into problems because of their conceptualisation of uncertainty.
 

### 2024-05-14 paper.qmd
two small edits

@@
 @Ibrekk1987 compared a 6 visualisations of a PDF,  a box plot, a CDF and a mean with a confidence interval on this basis.
-They found that people are better at extracting the mean from a plot when they are shown a plot that contains a mean with a confidence interval than when they are shown a PDF, a box plot, or a CDF (where the mean could not be read off the plot).
+They found that people are better at extracting the mean from a plot when they are shown a plot that contains a mean with a confidence interval than when they are shown a box plot, or other visualisations that did not allow for the mean to just be read off the plot.
 @Hullman2015 compared the static error bars and violin plots of the marginal distributions of two variables ($A$ and $B$) to an animated plot that depicted outcomes of the joint distribution of $A$ and $B$ in each frame and found that the visualisation of the joint distibution was better at answering questions about the joint distribution than the visualisations of the marginal distribution.
@@
 
-**The following sections are in the middle of being reorganized and rewritten so it will not flow on from the previous sections**
-
-{{< pagebreak >}}
+**UP TO HERE FOR RESTRUCTURE/REWRITE**
 

### 2024-06-20 paper.qmd
started signal supression restructure

@@
 
-# What is uncertainty?
-In order to visualise uncertainty, we need to know what we are trying to communicate through visualisation.
-Definitions in the literature take the view that the definition should be encompassing and include everything the general public might consider to be uncertain.
-Some works [@Hullman2018; @Maceachren2012; @Thomson2005] focus more narrowly on specific terms defined mathematically, such as probability, variance, error, or precision.
+# The purpose of uncertainty visualisation
+## Other motivations 
+## Signal supression
+
+# When is signal supression appropriate?
+Signal suppression is not always possible in a visualisation.
+Whether or not uncertainty should be used for signal-suppression largely depends on the purpose of the visualisation and the definition of uncertainty.
+
+## What is uncertainty
+### Current definitions of uncertainty
+Most definitions of uncertainty treat it as a feature of the data, as something that exists in our data no matter it's purpose. This is not entirely accurate.
+
+This view of uncertainty leads to a wide range of definitions that are designed to be encompassing, but struggle to capture the full picture.
+Some works [@Hullman2018; @Maceachren2012; @Thomson2005] focus narrowly on specific terms defined mathematically, such as probability, variance, error, or precision.
 Others [@Griethe2006; @Pang1997; @Pham2009; @Boukhelifa2017] include broader loosely related elements, such as missing values.
-Several attempts to address this gap are made in the papers describing taxonomies.
-
-## Definitions of Uncertainty in the Literature
-Taxonomies split uncertainty based on an endless stream of ever changing boundaries, such as whether the uncertainty is due to true randomness or a lack of knowledge [@Spiegelhalter2017; @Hullman2016; @utypo], if the uncertainty is in the  attribute, spatial elements, or temporal element of the data [@Kinkeldey2014], whether the uncertainty is scientific (e.g.
+Some authors use taxonomies to split uncertainty based on an endless stream of ever changing boundaries, such as whether the uncertainty is due to true randomness or a lack of knowledge [@Spiegelhalter2017; @Hullman2016; @utypo], if the uncertainty is in the  attribute, spatial elements, or temporal element of the data [@Kinkeldey2014], whether the uncertainty is scientific (e.g.
 error) or human (e.g.
 disagreement among parties) [@Benjamin2018], if the uncertainty is random or systematic [@Sanyal2009], statistical or bounded [@Gschwandtnei2016; @Olston2002], recorded as accuracy or precision [@Griethe2006; @Benjamin2018], which stage of the data analysis pipeline the uncertainty comes from [@utypo], how quantifiable the uncertainty is [@Spiegelhalter2017; @utypo], etc.
-In their own way, each of these taxonomies show an aspect of uncertainty that an author felt was important to differentiate.
-@utypo identified many common themes among this wide array of taxonomies and used them to design an encapsulating taxonomy that was boldly referred to as a "definition of uncertainty".
-
-While taxonomies can be helpful to map out the space of "things that we consider to be related to uncertainty", a taxonomy is *not* a precise definition that would allow statisticians to estimate the uncertainty in their projects.
+In their own way, each of these approaches show an aspect of uncertainty that an author felt was important to differentiate.
+
+The constantly shifting definitions make it hard to pin down and quantify uncertainty.
 The ramifications of this is seen throughout the literature, as a survey of visualisation authors cited "not knowing how to calculate uncertainty" as one of the primary reasons they did not include it in visualisations [@Hullman2020a].
-The difficulties authors have with designing a clear definition of uncertainty likely comes from the fact that these definition fail to identify precicely what it is that *creates* uncertainty.
-These authors have failled to identify that uncertainty is simply a by product of inference.
-
-## Defining uncertainty with respect to inference
+
+The constanly shifting definitions might be a result of the view of uncertainty as a latent variable that can exist without an explicit purpose.
+This is not true, as uncertainty only exists as a by-product of inference.
+
+### The case for uncertainty as a by product of inference
 It is easy to see uncertainties relationship to inference when we consider what might *not* be considered uncertain rather than just trying to think about what *is*.
@@
 
-It is clear that uncertainties dependence on inference has gone unnoticed in the literature because authors often mention the by products of this fact as a *task* or *goal* dependence.
+Several authors have connected the issues around defining uncertainty to inference, however it is often discussed as a *task* or *goal* dependence.
 Multiple authors have commented on the need to consider quantifying and expressing uncertainty at every stage of a project as the "goal" shapes every step of the analysis [@Kinkeldey2014; @Hullman2016; @Refsgaard2007].
@@
 It is clear that the relationship between uncertainty and inference is noticed at every stage in an analysis, however combining these uncertainties into a single "uncertainty" value is near impossible [@Spiegelhalter2017].
+
 With this understanding it becomes clear to see why uncertainty is tied to an endless string of examples in the data analysis pipeline.
 Uncertainty examples include imputed data, model selection, inherent randomness, biased sampling, etc, not because these things *are* uncertainty, but because they *create* uncertainty when we perform inference.
-
-The belief that uncertainty is a feature of the data, not the variable of interest, leads to several flabbergasting experiments in the uncertainty visualisation space.
-In these experiments, it is unclear how the participants were supposed to relate what they were shown to the uncertainty distribution the authors had in mind (or if the authors had any distribution in mind at all).
+Whether or not these elements are relevant is highly dependent on what statistic you are trying to draw inference on, and by extension, the purpose of your visualisation.
+
+## EDA vs Communication
+## If signal supression is not the goal
+
+# How should be test uncertainty visualisations?
+## Uncertainty as signal
+## Uncertainty as noise
+## Other uncertainty experiments 
+-----------------------------------------------
+
+# UP TO HERE FOR SORTING
+
+# What is uncertainty?
+
+## Defining uncertainty with respect to inference
+
 For example, @Boukhelifa2012 tried to quantify the strength of the intuitive connection between a line attribute called "sketchiness" and uncertainty.
@@
 Therefore, "sketchiness" was added to aspects of the visualization that had no obvious inferential statistic associated with it, so participants simply ignored the "sketchiness" or assumed it represented something else, such as an alternative option.
-In another experiment, @Hofman2020 asked participants to judge the effectiveness of a particular treatment.
+In another experiment,
+
+@Hofman2020 asked participants to judge the effectiveness of a particular treatment.
 One group was shown the prediction interval around the mean, which indicates the uncertainty associated with treatment *for a particular person* and another group was shown the the sampling interval which depicted the uncertainty associated with the *average effectiveness of the treatment*.
@@
 
-**UP TO HERE FOR RESTRUCTURE/REWRITE**
+-----------------------------------------------
+
+# UP TO HERE FOR RESTRUCTURE/REWRITE
 

### 2024-06-20 paper.qmd
a large amount of restructuring

@@
 - May not need to reference the same paper again and again
-- add visually integrable to fig-tax once I work out what it is
-- look through for papers that can be used to cite motivation for uncertainty visualisation experiments
+- Update background to reflect new structure of paper
 -->
@@
 Others contend that visualisations are highly dependent on the audience and there is no such thing as a "best" visualisation that will be accessible to all members [@Kinkeldey2014; @Spiegelhalter2017].
-These concerns appear across all application areas, and an underlying thread is that the problem arises from a lack of a cohesive and encompassing **definition of uncertainty**.
+These concerns appear across all application areas, and an underlying thread is that the literature on uncertainty visualisations have become disconnected from their purpose.
 
 The review provided here attempts to address these issues.
-We highlight the three key issues facing the uncertainty visualisation literature, specifically the lack of a definition of uncertainty, the confusion as to what *is* an "uncertainty visualisation", and the difficulties with measuring "uncertainty".
+We start by discussing why uncertainty visualisation has become it's own sub-field, distinct from "normal" information visualisation.
+In this discussion we establish the purpose of uncertainty visualisation is to perform "signal-suppression", that is, hide statistically insignificant signals while promoting statistically rigorous signals.
+**EDIT FROM HERE** We then move highlight the three key issues facing the uncertainty visualisation literature, specifically the lack of a definition of uncertainty, the confusion as to what *is* an "uncertainty visualisation", and the difficulties with measuring "uncertainty".
 When disucssing each of these issues, we will discuss how this concept is understood in existing literature, why this understanding may be flawed, and the issues that arrise due to these flaws.
@@
 The ideal role of this review to help synthesize this noisy field, by collecting together essential components, summarizing different viewpoints the current literature, and provide a guide for potential research topics.
+**TO HERE**
 
 # The purpose of uncertainty visualisation
-## Other motivations 
-## Signal supression
-
-# When is signal supression appropriate?
-Signal suppression is not always possible in a visualisation.
-Whether or not uncertainty should be used for signal-suppression largely depends on the purpose of the visualisation and the definition of uncertainty.
-
-## What is uncertainty
-### Current definitions of uncertainty
-Most definitions of uncertainty treat it as a feature of the data, as something that exists in our data no matter it's purpose. This is not entirely accurate.
-
-This view of uncertainty leads to a wide range of definitions that are designed to be encompassing, but struggle to capture the full picture.
-Some works [@Hullman2018; @Maceachren2012; @Thomson2005] focus narrowly on specific terms defined mathematically, such as probability, variance, error, or precision.
-Others [@Griethe2006; @Pang1997; @Pham2009; @Boukhelifa2017] include broader loosely related elements, such as missing values.
-Some authors use taxonomies to split uncertainty based on an endless stream of ever changing boundaries, such as whether the uncertainty is due to true randomness or a lack of knowledge [@Spiegelhalter2017; @Hullman2016; @utypo], if the uncertainty is in the  attribute, spatial elements, or temporal element of the data [@Kinkeldey2014], whether the uncertainty is scientific (e.g.
+It is rarely made clear exactly how uncertainty visualisation is different to normal information visualisation.
+@Hullman2016 commented that it is straightforward to show a value but it is much more complex to show uncertainty, but did not explain *why*.
+A visualisation is nothing more than computing a statistic, mapping that statistic to a visual feature and rendering that graphic.
+Nothing in this process suggests visualising an estimate of the variance should be any different to visualising an estimate of the mean, however this field *does* implicitly treat them different.
+Why?
+
+Uncertainty visualisation papers usually provide one of two primary justifications for their existence.
+The first justification is that uncertainty is fundamentally differently to other variables due to the psychological heuristics involved in interpreting uncertainty.
+Therefore uncertainty visualisation is different to normal visualisation as authors must consider the psychological effects of what they are visualising [@Spiegelhalter2017; @Hullman2019].
+The second justification assets that uncertainty is an additional variable and it is of vital importance to interpreting an estimate.
+Therefore uncertainty visualisation is a high dimensional visualisation problem, as we need to figure out how to seamlessly add uncertainty into already existing graphics [@moritz2017trust; @Griethe2006].
+
+Why some authors pick one motivation over another is rarely explained and both leave the role of uncertainty in the visualisation unclear.
+The literature never explains if uncertainty should be treated as a variable, as metadata, or as something else entirely [@Kinkeldey2014].
+
+Uncertainty, acting as it is intended, is transparency.
+There are many secondary benefits that come with this improved transparency, such as better decisions, more trust in the results and more confidence in the authors.
+These secondary benefits are, however, *not* the immediate goal of uncertainty.
+The following sections will discuss the issues and limitations in measuring uncertainty through these secondary metrics and provide suggestions as to how future studies should consider measuring uncertainty.
+
+Trust is a by product of displaying uncertainty rather than the goal of it, and viewing the relationship in the converse direction can lead to misguided research.
+Considering trust, and not transparency, as the metric of importance in uncertainty communication can lead to a questionable subtext that argues against transparency, something that has been noticed by several other authors [@Spiegelhalter2017; ONeill2018].
+@Hullman2020a found that author simultaneously argued that failing to visualise uncertainty was akin to fraud, but also many avoided uncertainty visualisation because they didn't want their work to come across as "untrustworthy".
+These authors are optimising *trust* rather than *transparency*, which means they opt to leave out uncertainty information when it does exactly what it is supposed to, decrease certainty in conclusions.
+
+Science communication should be primarily concerned with accuracy.
+Setting trust and risk-aversion as the variables of interest implicitly encourages statisticians to set trust and risk-aversion as the primary goals of communication.
+The issue of trust being divorced from trustworthiness has been commented on by other authors [@ONeill2018], however the issue still persists in the uncertainty visualisation literature.
+@Zhao2023 displayed a several variations of a visualisation of a model prediction and its uncertainty and took participants using the model prediction as a sign of trust.
+They reported that visualising uncertainty information caused participants to trust the model in the low variance case, but the results in the high variance case were inconclusive.
+The discussion made it clear the authors thought the uncertainty information should make the visualisation more trustworthy, but conflating trust and the use of a prediction implied uncertainty information should somehow influence participants to use their own prediction, even though a prediction being uncertainty does not necessarily mean it is incorrect.
+Despite this, the authors seemed to assume that the uncertainty information *should* have an influence on that, showing they had not deeply considered *how* uncertainty information should influence the choices of the participants.
+(*Cite Gap 18: examples of studies where authors measure trust*)
+
+A similar measure to trust is using "confidence" in an extracted value or a decision.
+Interestingly, "confidence" is also used to try and capture the clarity of a message in a normal visualisation.
+Confidence cannot simultaneously be a measure of clarity of visualisation *and* a way to capture the uncertainty expressed in a visualisation.
+
+# Signal-Supression
+The true reason that uncertainty visualisation is differerent to normal information visualisation, is that normal information visualisation is designed to *find* signals in our data, while uncertainty visualisation should be designed to *supress* it.
+
+Interviews with experts in visualisation and statistics often back up "signal suppression" as the primary motivation for uncertainty visualisation.
+Failling to communicate uncertainty is often considered to be akin to fraud or lying [@Hullman2020a; @Manski2020] indicating that uncertainty should exist to prevent undue certainty in visual inference.
+The relationship between signal supression and uncertainty visualisation is made more explicity by other authors [@Griethe2006].
+Therefore uncertainty acts as a form of statistical "hedging" for signals found in an analysis.
+This aligns nicely with our understanding of uncertainty as a by product of inference.
+The variable we are trying to perform inference on is the signal, while the while the visualisation of noise that prevents the signal from being read is the uncertainty.
+It is nice that the connection to inference easily extends from uncertainty to uncertainty visualisation, however many of the confusions extend as well.
+
+## Uncertainty does not exist without signal
+It is obvious that we cannot design a visualisation for signal suppression without a clear signal.
+This may be considered a limitation by some readers, however it is less of a restriction when we recognize that *uncertainty* itself does not exist without a signal.
+
+#### The current definitions of uncertainty
+Uncertainty is a concept that is famously hard to pin down, with constantly shifting definitions that mean it is quantified differently from author to author.
+
+The ramifications of this is seen throughout the literature, as a survey of visualisation authors cited "not knowing how to calculate uncertainty" as one of the primary reasons they did not include it in visualisations [@Hullman2020a].
+Some works [@Hullman2018; @Maceachren2012; @Thomson2005] get around this definition issue by focusing narrowly on specific terms defined mathematically, such as probability, variance, error, or precision.
+Others [@Griethe2006; @Pang1997; @Pham2009; @Boukhelifa2017] include broader loosely related elements, such as missing values, but ignore these concepts when translating that uncertainty to a visualisation.
+Existing mathematical definition of uncertainty (or related concepts) have been unable to unify this range of diverse concepts.
+For example, @Thomson2005 suggests a mathematical formula for *examples* of uncertainty, and information theory tries to quantify uncertainty using the idea of entropy, but the disconnect between the broad concept of uncertainty and what we can reliably quantify remains a problem.
+
+Some authors prefer to define uncertainty in qualitative terms, often through the use of taxonomies.
+In these cases, uncertainty is split using an endless stream of ever changing boundaries, such as whether the uncertainty is due to true randomness or a lack of knowledge [@Spiegelhalter2017; @Hullman2016; @utypo], if the uncertainty is in the attribute, spatial elements, or temporal element of the data [@Kinkeldey2014], whether the uncertainty is scientific (e.g.
 error) or human (e.g.
@@
 
-The constantly shifting definitions make it hard to pin down and quantify uncertainty.
-The ramifications of this is seen throughout the literature, as a survey of visualisation authors cited "not knowing how to calculate uncertainty" as one of the primary reasons they did not include it in visualisations [@Hullman2020a].
-
-The constanly shifting definitions might be a result of the view of uncertainty as a latent variable that can exist without an explicit purpose.
-This is not true, as uncertainty only exists as a by-product of inference.
-
-### The case for uncertainty as a by product of inference
+What is interesting about all these views of uncertainty, is that they all imply that it is a latent feature of our data.
+That is, it is something that can be revealed through exploration, and not something that needs to be explicitly calculated.
+The constant stream of new definitions that are unable to capture the true essence of uncertainty might be a result of this false assumption that underpins ever definition.
+Uncertainty cannot be defined without a signal because uncertainty does not exist without a signal, is a by-product of inference.
+
+#### Uncertainty as a by-product of inference
 It is easy to see uncertainties relationship to inference when we consider what might *not* be considered uncertain rather than just trying to think about what *is*.
@@
 ## EDA vs Communication
-## If signal supression is not the goal
-
-# How should be test uncertainty visualisations?
-## Uncertainty as signal
-## Uncertainty as noise
-## Other uncertainty experiments 
------------------------------------------------
-
-# UP TO HERE FOR SORTING
-
-# What is uncertainty?
-
-## Defining uncertainty with respect to inference
-
-For example, @Boukhelifa2012 tried to quantify the strength of the intuitive connection between a line attribute called "sketchiness" and uncertainty.
-Participants were shown the six scenarios depicted in @fig-sketchy and asked to interpret what they believed the squiggly line indicates.
-The authors were aware that uncertainty seemed to have some "task" dependence but interpreted this as *context* dependence rather than dependence on a particular inferential statistic.
-Therefore, "sketchiness" was added to aspects of the visualization that had no obvious inferential statistic associated with it, so participants simply ignored the "sketchiness" or assumed it represented something else, such as an alternative option.
-In another experiment,
-
-@Hofman2020 asked participants to judge the effectiveness of a particular treatment.
-One group was shown the prediction interval around the mean, which indicates the uncertainty associated with treatment *for a particular person* and another group was shown the the sampling interval which depicted the uncertainty associated with the *average effectiveness of the treatment*.
-There was no critical assessment by the authors on *why* these two distributions should be compared.
-The contribution of this experiment hinges on the belief that sharing a common descriptive statistic (in this case, the mean) is grounds to compare two unrelated inferential statistics, a belief which is clearly false when viewing uncertainty through the lens of inference.
-
-![The graphics displayed by @Boukhelifa2012 to identify if there is an intuitive connection between sketchiness and uncertainty.
-These graphic were made without concern as to what (i.e.
-which statistic) in the image is supposed to be uncertain.
-This leads to the images being difficult to connect to uncertainty even if we assume tha that is what it is supposed to represent.
-For example, (S5) implies that participants should have read the plot to mean that the creators of a rail network map would be "uncertain" about the existence of a train line, something that is defies common logic.](sketchiness.png){#fig-sketchy width=50%}
-
-Additionally, if we view uncertainty through the lens of inference then it is obvious that bias and variance both contribute to uncertainty regardless as to which stage of the analysis they appear.
-Many papers in uncertainty visualisation are not working from this null hypothesis.
-For example, @Padilla2021 found that high uncertainty in the model estimates (calculated uncertainty) and low forecaster confidence (which is typically an expression of suspected bias in the model) both caused participants to have decreased confidence in their results and suggested modelers express both if they are relevant.
-In another example, @Kale2019 discusses the importance of communicating decisions made in the data analysis pipeline and being aware of the alternatives.
-While there is nothing wrong with the results of these papers, the fact that they were published shows some level of surprise by these results.
-This work indicates that there was a significant number of authors in the field were not already aware that choices introduced early in the data analysis pipeline create bias and therefore uncertainty in our final values that should be communicated.
-
-Existing mathematical definition of uncertainty (or related concepts) have been unable to unify this range of diverse concepts.
-For example, @Thomson2005 suggests a mathematical formula for *examples* of uncertainty, @Meng2014 mathematically defined the variance introduced to a model by the array of model choices, and information theory tries to quantify uncertainty using the idea of entropy.
-These existing methods are useful for calculating uncertainty as a specific stage of the analysis, but not thorough enough for an analyst to understand the causes of uncertainty and quantify it.
-The reality is that a lot of uncertainty visualisation authors do not seem to have an intuitive understanding of uncertainties connection to inference, when inference is being performed, and how to communicate this complicated by product of inference.
-This misunderstanding is not due to laziness or the fault of the authors, but rather is likely caused by the confusing set of existing definitions that leads authors to be unaware what they are supposed to be visualising.
-Therefore, it is clear the field would greatly benefit from a strict definition of uncertainty.
-The task of providing an encapsulating mathematical definition of uncertainty is far beyond the scope of this work, however focusing on its relationship to inference is likely a fruitful step forward.
-
-# What is an uncertainty visualisation?
-Given that there is no definition of "uncertainty" readers may wonder how we can define "uncertainty visualisations".
-This is a reasonable issue given that you cannot quantify something that is not defined, and you can't visualise something you cannot quantify.
-This section discusses how the problems caused by the absent definition of uncertainty bleed into our understandings of uncertainty visualisation and creates a field that cannot clearly state what is or is not within its domain.
-
-## Definitions of "uncertainty visualisation"
-What exactly is or is not an uncertainty visualisation is just as poorly defined as the definition of uncertainty, so authors seem to decide for themselves what is an uncertainty visualisation.
-This appears in the literature as a mountain of seemingly conflicting statements about what is, or is not, an "uncertainty visualisation".
-For example @Leland2005 mentions that popular graphics, such as pie charts and bar charts omit uncertainty, however at least one or both of these charts are used in a significant number of uncertainty visualisation experiments [@Ibrekk1987; @Olston2002; @Zhao2023; @Hofmann2012].
-@Wickham2011 suggests their product plot framework, which includes histograms, should have a way to measure uncertainty, but does not consider that a histogram is *already* a depiction of PDF, something that is often considered an uncertainty visualisation by other authors.
-This completely untethered concept of an "uncertainty visualiation" leads to many user experiments where the visualisations used have a shocking information asymetry.
-
-Visualisation authors are almost unanimous in commenting that the "information" in two plots must be the same in order for the visual techniques to be compared [@Cleveland1984; @Kinkeldey2014].
-@Kinkeldey2014 adopts an existing definition also that suggests two graphics are informationally equivalent if all the information in one plot is inferable from the other and vice-versa, but adds that two plots are computationally equivalent if that information can be extracted from both plots with similar easy and speed.
-Many visualisation authors ignore this concept and simply compare two visualisations because they both belong to the class of "uncertainty visualisations".
-@Ibrekk1987 compared a 6 visualisations of a PDF,  a box plot, a CDF and a mean with a confidence interval on this basis.
-They found that people are better at extracting the mean from a plot when they are shown a plot that contains a mean with a confidence interval than when they are shown a box plot, or other visualisations that did not allow for the mean to just be read off the plot.
-@Hullman2015 compared the static error bars and violin plots of the marginal distributions of two variables ($A$ and $B$) to an animated plot that depicted outcomes of the joint distribution of $A$ and $B$ in each frame and found that the visualisation of the joint distibution was better at answering questions about the joint distribution than the visualisations of the marginal distribution.
-@Hofman2020 commented that "theoretically" the sampling distribution of the mean and the prediction interval of a new observation are equal "so long as one knows the sample size", but does not seem to provide participants with that sample size, or recognise the assumptions and background knowledge that would be required to compare the two.
-@Hofman2020 and @Zhang2022 compared prediction and sampling distributions because they are both "uncertainty" that is typically depicted around the mean.
-They found that people are better at answering questions about a prediction interval when shown a prediction interval instead of a sampling distribution.
-
-This collection of examples starts to paint a pretty clear picture.
-Visualisations with rather shocking information asymmetry are regularly compared because they are both "uncertainty visualisations".
-This results in a series of experiments where the visual aspects of the graphic are not even required to anticipate the experimental results.
-Ideally this problem would be avoided by defining the information in a graphic using *The Grammar of Graphics* since it is a sufficient summary of the information in a graphic and it allows a plot to be considered a statistic [@Vanderplas2020].
-The process for summarising the information in a plot using *The Grammar of Graphics* is depicted in @fig-pipeline [@Leland2005].
-Unfortunately, *The Grammar of Graphics* is not perfectly suited to summarise the information in an uncertainty visualisation since it assumes that we start from a finite data set.
-Very often uncertainty is expressed using a distribution or a resampling method that does not have an upper limit on the "sample size".
-Additionally many uncertainty visualisation experiments notice an interaction between sample size and graphic effectiveness [@Kale2018; @Newburger2022; @Hofmann2012], an effect that is likely caused by a difference in human perception of mass versus sample.
-Additionally, *The Grammar of Graphics* behaves strangely when the sample size is adjusted and it the distance between two graphics becomes less consistent.
-@fig-infintegraph shows how three graphics defined by *The Grammar of Graphics* can change as n approaches infinity.
-The two graphics that are very similar according to the *The Grammar of Graphics* diverge into different graphics, and two graphics that are very different, converge into the same graphic.
-This means the statistics visualised in the plot as well as the *The Grammar of Graphics* may not be an adequate to summarise the information in a plot that is being used for uncertainty visualisation.
-Some authors have noticed difficulties with visualising distributions using `ggplot2` and created extensions that make showing distributions more intuitive [@Pu2020; @Kay2023], however these extensions do not address the core discrepancies with uncertainty visualisation and *The Grammar of Graphics*.
-This does not mean *The Grammar of Graphics* should be abandoned when we attempt to summaise the information in an uncertainty visualisation, very often visations use the same sample to generate all their plots, so this issue is not always present, however these limitations should be kept in mind when analysing uncertainty visualisations using *The Grammar of Graphics*.
-
-![(The grammar of graphics data analysis pipeline from @Leland2005 written in black, with the tidy data equivalent written in blue.
-The product plot framework and the grammar of probabilistic graphics frameworks are in green.
-It can be seen that the grammar of graphics provides a set of instructions that create a graphic, however the inferential power of this graphic is not considered and relies on many additional assumptions that are not defined by the grammar.](grammarofgraphicspipeline.jpeg){#fig-pipeline}
-
- ![This graphic shows three simple graphics made from the same data set that diverge according to the grammar of graphics.
-Despite being distinct graphics from a stage as low down as the statistic (or data manipulation in the tidy data framework), as our sample size increases to infinity, the grahics converge to the same distribution and become visually indistinguishable.
-](infinitegraph.jpeg){#fig-infinitegraph}
-
-An alternative way authors consider describing an uncertainty visualisation is to use taxonomies.
-However, these taxonomies tend to blur the line between the mathematical elements of uncertainty and the visual depiction of those mathematical objects.
-Some taxonomies of uncertainty highlight how confusing most authors find the concept, and how difficult it is to articulate the rules of an uncertainty visualisation when uncertainty itself if not defined.
-@Kinkeldey2014 categorised uncertainty according to five criteria depicted in @fig-tax which considers if a visualisation is implicit or explicit, intrinsic or extrinsic, visually integrable or seperable, coincident or adjacent, and static or dynamic.
-A similar version of this taxonomy was presented by @uncertchap2022 who commented that visualisation can be organised into two categories, "graphical annotations of distributional properties" and "visual encodings of uncertainty" which seems to functionally align with the intrinsic/extrinsic distinction by @Kinkeldey2014.
-@Griethe2006 organised uncertainty visualisations into two cases (1) a hypothesis test was preformed to confirm the validity of the visualisation and (2) the visualisation has uncertainty depicted.
-@Potter2012 organised several existing uncertainty visualisations into groups based on the dimensionality of the data (1D, 2D, 3D, and No Dimension) and the dimensionality of the uncertainty (Scalar, Vector, Tensor).
-However, because the term "PDF", function that is used to describe a random variable, is used to describe both the data and the uncertainty for all dimensions.
-@Grewal2021 created a taxonomy that mapped uncertainty visualisations to some point in a 2D space defined by the "domain expertise" and "continuum of discreteness" (that scaled from "point estimate" to "continuous distribution").
-
-![Visualisation of @Kinkeldey2014 uncertainty visualisation taxonomy.
-It can be seen that most of the categories can be considered a change to the typical error bar chart.
-The taxonomy categorises the visualisations based on five criteria, implicit (a sample) or explicit (a depiction of mass); Intrinsic (alter existing symbols symbols to represent uncertainty) or extrinsic (add new objects to represent uncertainty); visually integral or separable (the uncertainty can be separated from the data and read independently) ; coincidence (uncertainty and data are represented in the same plot) or adjacent; static or dynamic (animated, interactive etc).
-Many of the distinctions conflate changes of the input data and statistics calculated and with visual changes, highlighting how defining uncertainty and visualisation together conflates several elements of a graphic.](uncertvistax.jpeg){#fig-tax}
-
-The authors of these taxonomies did not intend for them to be complete descriptions of uncertainty visualisations, however they do paint a picture of the what authors consider to be important.
-There are common themes in the taxonomies that ensure we differentiate between a sample and a mass, consider if uncertainty should be depicted with the estimate or as a second variable, identify the parallel of hypothesis testing, and consider the dimensionality, sample size and precision depicted by the visualisation.
-These taxonomies may highlight perceived important differences, but they do not help us understand what *is* or *is not* an uncertainty visualisation.
-For that we need to understand *why* uncertainty visualisation exists as a sub-field at all.
-
-## Uncertainty visualisations should perform signal supression
-It is rarely made clear exactly how uncertainty visualisation is different to normal information visualisation.
-Visualising a graphic is nothing more than computing a statistic, mapping that statistic to a visual feature and rendering that graphic.
-Nothing in this process suggests visualising an estimate of the variance should be any different to visualising an estimate of the mean, however this field *does* implicitly treat them different.
-Why?
-
-Uncertainty visualisation papers usually provide one of two primary justifications for their existence.
-The first justification is that uncertainty is fundamentally differently to other variables due to the psychological heuristics involved in interpreting uncertainty.
-Therefore uncertainty visualisation is different to normal visualisation as authors must consider the psychological effects of what they are visualising [@Spiegelhalter2017; @Hullman2019].
-The second justification assets that uncertainty is an additional variable and it is of vital importance to interpreting an estimate.
-Therefore uncertainty visualisation is a high dimensional visualisation problem, as we need to figure out how to seamlessly add uncertainty into already existing graphics [@moritz2017trust; @Griethe2006].
-Why some authors pick one motivation over another is rarely explained and both leave the role of uncertainty in the visualisation unclear.
-The literature never explains if uncertainty should be treated as a variable, as metadata, or as something else entirely [@Kinkeldey2014].
-@Hullman2016 commented that it is straightforward to show a value but it is much more complex to show uncertainty, but did not explain *why* Interestingly, these motivations do not remotely line up with the most common reason that people *outside* the uncertainty visualisation community thinks visualising uncertainty is important.
-The true reason that uncertainty visualisation is differerent to normal information visualisation, is that normal information visualisation is designed to *find* signals in our data, while uncertainty visualisation should be designed to *supress* it.
-
-Interviews with experts in visualisation and statistics often back up "signal suppression" as the primary motivation for uncertainty visualisation.
-Failling to communicate uncertainty is often considered to be akin to fraud or lying [@Hullman2020a; @Manski2020] indicating that uncertainty should exist to prevent undue certainty in visual inference.
-The relationship between signal supression and uncertainty visualisation is made more explicity by other authors [@Griethe2006].
-Therefore uncertainty acts as a form of statistical "hedging" for signals found in an analysis.
-This aligns nicely with our understanding of uncertainty as a by product of inference.
-The variable we are trying to perform inference on is the signal, while the while the visualisation of noise that prevents the signal from being read is the uncertainty.
-It is nice that the connection to inference easily extends from uncertainty to uncertainty visualisation, however many of the confusions extend as well.
+The concept of uncertainty as a by-product of inference, not a latent feature of data, extends naturally to visualisation.
 
@@
 
-{{< pagebreak >}}
-
------------------------------------------------
-
-# UP TO HERE FOR RESTRUCTURE/REWRITE
-
-### Interpretation and semantics
+### No Such thing as an "uncertainty visualisation"
+If uncertainty is a by-product of inference, a visualisation being an "uncertainty visualisation" does not exist.
+
+What exactly is or is not an uncertainty visualisation is just as poorly defined as the definition of uncertainty, so authors seem to decide for themselves what is an uncertainty visualisation.
+For example @Leland2005 mentions that popular graphics, such as pie charts and bar charts omit uncertainty, however at least one or both of these charts are used in a significant number of uncertainty visualisation experiments [@Ibrekk1987; @Olston2002; @Zhao2023; @Hofmann2012].
+@Wickham2011 suggests their product plot framework, which includes histograms, should have a way to measure uncertainty, but does not consider that a histogram is *already* a depiction of PDF, something that is often considered an uncertainty visualisation by other authors.
+
+@Kinkeldey2014 categorised uncertainty according to five criteria depicted in @fig-tax which considers if a visualisation is implicit or explicit, intrinsic or extrinsic, visually integrable or seperable, coincident or adjacent, and static or dynamic.
+A similar version of this taxonomy was presented by @uncertchap2022 who commented that visualisation can be organised into two categories, "graphical annotations of distributional properties" and "visual encodings of uncertainty" which seems to functionally align with the intrinsic/extrinsic distinction by @Kinkeldey2014.
+@Griethe2006 organised uncertainty visualisations into two cases (1) a hypothesis test was preformed to confirm the validity of the visualisation and (2) the visualisation has uncertainty depicted.
+@Potter2012 organised several existing uncertainty visualisations into groups based on the dimensionality of the data (1D, 2D, 3D, and No Dimension) and the dimensionality of the uncertainty (Scalar, Vector, Tensor).
+However, because the term "PDF", function that is used to describe a random variable, is used to describe both the data and the uncertainty for all dimensions.
+@Grewal2021 created a taxonomy that mapped uncertainty visualisations to some point in a 2D space defined by the "domain expertise" and "continuum of discreteness" (that scaled from "point estimate" to "continuous distribution").
+
+![Visualisation of @Kinkeldey2014 uncertainty visualisation taxonomy.
+It can be seen that most of the categories can be considered a change to the typical error bar chart.
+The taxonomy categorises the visualisations based on five criteria, implicit (a sample) or explicit (a depiction of mass); Intrinsic (alter existing symbols symbols to represent uncertainty) or extrinsic (add new objects to represent uncertainty); visually integral or separable (the uncertainty can be separated from the data and read independently) ; coincidence (uncertainty and data are represented in the same plot) or adjacent; static or dynamic (animated, interactive etc).
+Many of the distinctions conflate changes of the input data and statistics calculated and with visual changes, highlighting how defining uncertainty and visualisation together conflates several elements of a graphic.](uncertvistax.jpeg){#fig-tax}
+
+The authors of these taxonomies did not intend for them to be complete descriptions of uncertainty visualisations, however they do paint a picture of the what authors consider to be important.
+There are common themes in the taxonomies that ensure we differentiate between a sample and a mass, consider if uncertainty should be depicted with the estimate or as a second variable, identify the parallel of hypothesis testing, and consider the dimensionality, sample size and precision depicted by the visualisation.
+These taxonomies may highlight perceived important differences, but they do not help us understand what *is* or *is not* an uncertainty visualisation.
+For that we need to understand *why* uncertainty visualisation exists as a sub-field at all.
+
+## Designing Visualisation for Signal supression
 Interpretation and semantics experiments are seeking to identify a dimension (or visual task) that uncertainty naturally maps to.
@@
 This problem has been noticed by others in the field, that typically use this method (specifically in the spatial uncertainty context) and a desire for representations that integrate uncertainty and signal is one of the reasons for the invention of the value-suppressing uncertainty pallet [@Correll2018].
-
 Value-suppressing uncertainty pallets (VSUP) were developed as a method that would allow the signal and the noise to be interpreted together such that insights gained by the viewers of a plot are appropriately *suppressed* by the uncertainty.
@@
 
-# How do we measure uncertainty?
+# Visualising uncertainty as signal
 If uncertainty is inherently different, we would expect evaluation studies of uncertainty visualisations to produce different results to similar evaluation studies on "normal" variables.
@@
 The previous section highlights a unique and fascinating problem faced by uncertainty visualisation.
-If asking direct questions about uncertainty causes us to treat it as a signal, how do we evaluate uncertainty as *noise*?
- When we ask the viewer of a plot to look at data and extract a value, we are asking them to perform inference on that value.
-There will be noise associated with that answer and that is uncertainty.
-If we ask direct question about some uncertainty metric, we have turned the uncertainty into signal because that is what the participants are drawing inference on.
-
-Uncertainty, acting as it is intended, is transparency.
-There are many secondary benefits that come with this improved transparency, such as better decisions, more trust in the results and more confidence in the authors.
-These secondary benefits are, however, *not* the immediate goal of uncertainty.
-The following sections will discuss the issues and limitations in measuring uncertainty through these secondary metrics and provide suggestions as to how future studies should consider measuring uncertainty.
-
-## Measuring uncertainty as signal
-Uncertainty visualisation papers can be organised according to the *goal* of the experiment.
-Evaluation experiments are the standard rule for visualisations because the human brain is not as reliable as mathematical calculation.
-Therefore, user studies often aim to assess the limitations, biases, and heuristics of our mental calculator so that we can better understand the problems we may encounter when we plot our data.
-This is not to say any paper that suggests a visualisation without an evaluation experiment is completely lacking in justification, and there are many papers that suggest a novel visualisation without an evaluation study.
-Sometimes these papers are a preliminary step in finding a solution for common problems and intend to evaluate the visualisation in later work.
-The reasoning for this is obvious.
-There are often heuristics and biases that are not obvious to us when designing visualisation.
-Additionally, these heuristics and biases can change depending on the larger scope of the graphic and the population we are communicating with [@Spiegelhalter2017; @Kinkeldey2014].
-
-The overarching goal of most uncertainty visualisation evaluation papers belong to one of: performance (how effectively a participant can extract information from a plot), interpretation and semantics (the ease with which a particular visualisation is associated with uncertainty), and quality of user experience (if the participants liked the plot or not)[@Hullman2019].
+
+Even though uncertainty does not exist without a signal, that does not mean every visualisation that includes a concept related to uncertainty (such as a visualisation or error or variance) must be designed for signal supression.
+However, if a visualisation is not designed for signal supression, it should be treated the same as any other variable.
+These visualisation are not "special" in the sense that signal supression visualisations are.
+
 @Hullman2019 found that the majority of papers evaluate visualisations on performance (approximately 65% of the papers they surveyed) or interpretation and semantics (approximately 17%) and both of these evaluation goals will run into problems because of their conceptualisation of uncertainty.
 
-### Repeating perceptual task experiments
-Given the previous section, one might consider problems in uncertainty visualisation to be fixed if we could restrict evaluation experiments to cases where the information in each graphic is almost identical.
-Unfortunately this still often results in uninteresting or obvious results.
+By far the most common metric used is accuracy, which is used by approximately 36% of evaluation studies [@Hullman2019].
+
+Visualisation authors are almost unanimous in commenting that the "information" in two plots must be the same in order for the visual techniques to be compared [@Cleveland1984; @Kinkeldey2014].
+@Kinkeldey2014 adopts an existing definition also that suggests two graphics are informationally equivalent if all the information in one plot is inferable from the other and vice-versa, but adds that two plots are computationally equivalent if that information can be extracted from both plots with similar easy and speed.
+Many visualisation authors ignore this concept and simply compare two visualisations because they both belong to the class of "uncertainty visualisations".
+@Ibrekk1987 compared a 6 visualisations of a PDF,  a box plot, a CDF and a mean with a confidence interval on this basis.
+They found that people are better at extracting the mean from a plot when they are shown a plot that contains a mean with a confidence interval than when they are shown a box plot, or other visualisations that did not allow for the mean to just be read off the plot.
+@Hullman2015 compared the static error bars and violin plots of the marginal distributions of two variables ($A$ and $B$) to an animated plot that depicted outcomes of the joint distribution of $A$ and $B$ in each frame and found that the visualisation of the joint distibution was better at answering questions about the joint distribution than the visualisations of the marginal distribution.
+@Hofman2020 commented that "theoretically" the sampling distribution of the mean and the prediction interval of a new observation are equal "so long as one knows the sample size", but does not seem to provide participants with that sample size, or recognise the assumptions and background knowledge that would be required to compare the two.
+@Hofman2020 and @Zhang2022 compared prediction and sampling distributions because they are both "uncertainty" that is typically depicted around the mean.
+They found that people are better at answering questions about a prediction interval when shown a prediction interval instead of a sampling distribution.
+
+This collection of examples starts to paint a pretty clear picture. Visualisations with rather shocking information asymmetry are regularly compared because they are both "uncertainty visualisations".
 
@@
 When trying to anticipate the results of these studies, we can use the same principles of information equivalence and difficulty of relevant mental tasks to understand which plots will outperform others.
-This leads us to wonder...
-why is uncertainty visualisation a field at all?
-
-### Performance
-There are many metrics experimenters use to evaluate the performance of a visualisation.
-Each metric, such as can use accuracy, decision quality, confidence, trust and others each have their own issues which prevent them from accurately capturing the effects of uncertainty in a visualisation.
-
-By far the most common metric used is accuracy, which is used by approximately 36% of evaluation studies [@Hullman2019].
-In previous sections, we pointed out issues that arise when accuracy (of extracting probability or variance) is used as the primary metric in evaluating uncertainty visualisations, the work simply replicates known concepts in information visualisation.
-Subjective metrics, such as how much participants liked the aesthetic representation of a visualisation, or metrics with an unclear motivation or use, such as how memorable the information in a plot is or how much a user interacted with a plot, are ignored in this section.
-Below we will explain in detail why experiments that try to measure uncertainty using methods such as decision making, trust, or other types of questions, also fail to capture the effects of visualising uncertainty in a plot.
-
-#### Decision making and risk aversion
-
-![Visualisation of the process of transforming raw data to a visualisation insight.
-Summarising the data using common statistics removes information from our data.
-Once the data is in a format that represents the varibales of interest, organising that information in a visualisation makes it easier to extract insights from these statistics and convert them to insights.
-If information is removed at the statistics stage, it cannot be added back in at the visualisation stage, as it is meerly an efficient technique to organise and present information to enable a large range of insights.
-](visprocess.jpeg){#fig-location width=70%}
-
-Decision making tasks are often described by authors as a more realistic version how plots are used in practice.
-While we do not disagree with this statement, exactly *why* these tasks are different to value extraction tasks is never explained.
-While the task may be more similar to how plots are actually used, that does not necessarily mean it is a better *experimental* environment for evaluating plots.
-Let us consider this in more detail.
-A normal value extraction task involves:
-
-1) correctly interpret the question
-2) extract the specified value
-3) report the value
-
-On the other hand, a decision making task using an uncertainty visualisation involves:
-
-1) Correctly interpret the question
-2) Use some risk utility function to set a threshold for an "acceptable level of risk"
-3) Extract the value for from the plot that will be compared to the threshold
-4) Make a judgement based off that value
-
-The key aspect that separates a decision making task from a value extraction task, for uncertainty visualisations, is the inclusion of a utility function.
-While this might seem like a small change, it actually completely warps the experiment in unexpected.
-
-The first issue is that individuals will all have their own individual risk utility function, that is, how much they want to avoid or engage with risks or uncertainty.
-Therefore decision making experiments that have an additional layer of noise that value extraction experiments do not.
-We cannot be sure if participants answered differently from the ground truth because a visualisation was difficult to read *or* because the participants risk utility function did not align with the one set by the authors.
-Several authors have offered solutions to this issue, however the problem is deeper than any of them seem to realise.
-
-@Hullman2016 suggested providing a utility framework for each experiment, to instruct the participants in how to account for the uncertainty information.
-This is a method that seems to have been adopted by @Fernandes2018 who describe the following scenario to participants who are trying to maximise the coins in their experiment:
-> Subjects gain coins for every minute they are able to continue an activity that is valuable to them (e.g., watching TV at home) before going to the bus stop, and gain a bonus for arriving at their intended destination early.
-Subjects incur a coin penalty for time spent waiting at a stop for a bus to arrive.
-
-Any payment scheme that is incorporated into an experiment will also implicitly set up a utility framework, as it is assumed participants will try to maximise their payments.
-In tasks such as this, the ground truth of each question is typically selected to be the value a ration agent would select and visualisations are evaluated on the basis of how far the participants responses are from that ground truth.
-The problem with this, is that all the complexity of the question is in *working out* the best response, and it has little to do with the visualisation itself.
-The visualisation aspect of these studies becomes a value extraction experiment.
-
-@Cheong2016 tested multiple different visual representations of uncertainty for representing the likelihood of a house being burned down based on its location.
-Their payment scheme, which paid out $0.10 for a correct choice (i.e.
-staying when the house was not burned down or leaving when the house was burned down) and 0 for an incorrect choice (i.e.leaving when the house didn't burn down or staying when the house burned down), meant participants were incentivised to base their entire leave/stay decision on whether or not the likelihood of a fire at their house is above or below 50%.
-If the participants *did* correctly identify the optimum strategy and answer accordingly, the "decision making experiment" was actually *just a value extraction experiment*.
-The decision making aspect ceased to matter, all that mattered was identifying if the probability of a fire was greater than 50%.
-Interestingly, despite this very obvious and simple tactic to maximize payout from the experiment, it seems like many participants did not adopt it, instead acting much more warily as though they were considering whether or not they would *actually* evacuate in the event of a fire.
-This means it is likely that introducing a payment scheme or a utility function creates too much mental labour for the participants to answer "correctly", and they would rather stick with their own risk utility function that feels *natural* than go through the mental labour of adopting another.
-Setting up an experiment that requires participants to do laborious calculations or strenuous mental effort for a small marginal benefit is unlikely to result in successful participant responses because it is not in the "spirit" of visualisation.
-(*Cite Gap 16: Examples of value extraction decision making tasks*)
-
-It is clear that audiences primarily interact with visualisations through their "System 1" brain [@daniel2017thinking].
-This aligns with other authors comments that visualisation is primarily about "gists" [Spiegelhalter2017].
-In this sense, asking participants a question that requires them to shift to their "System 2" is entirely contrary to *why* we use visualisation.
-The exact reasons we describe visualisation to be so powerful, that they are able to communicate complicated information fast and efficiently *is a product of the system 1 brain* [@daniel2017thinking], however this means displaying information using visualisations have the same weakness' as system 1 thinking, and visualisation authors need to be aware of this.
-Asking participants to answer questions that require complicated calculations does not reflect how visualisations are used, and will likely be ignored by the participants.
-While we agree with @Hullman2016 who suggested that what determines an appropriate ground truth is largely a philosophical exercise, we differ in the sense that we do not think it should be done, lest you create a needlessly complicated value extraction task.
-
-Additionally, several papers mention "risk-aversion" in participant decisions as a negative by-product of a particular visualisation design, however that displays a failure to understand that risk aversion is *not* a mistake, but rather it *reveals the utility of decreased uncertainty*.
-It may be interesting to investigate why a particular visualisation elicits increased risk-aversion, for example is it due to poorer estimates or increased awareness of negative outcomes, however simply reporting this as "risk aversion" conflates these factors.
-The work that advocates for transparency in risk-communication gives you whiplash when it's subtext argues that transparency leads to "incorrect" conclusions.
-There have been other discussions on the appropriateness of discussing risk-aversion as a bias [@Vranas2000], but few have made the point that if uncertainty holds some *value* there is not technically a "right" decision at all.
-
-The final issue with decision making tasks is that calculating "optimum" choices is not a task we use uncertainty for, it is a task that utilises *risk*.
-Risk and uncertainty are slightly different, as risk is known probabilities and uncertainty is unknown probabilities, so uncertainty is what you get when you cannot accurately define risk [@Spiegelhalter2017].
-Communicating risk allows people to weigh up options and make an optimum decision, uncertainty hedges information to let people know there may not even be an optimum strategy.
-In this sense, many "uncertainty" decision making experiments are actually just risk decision making experiments.
-Not only this, but the inclusion of uncertainty information should *not* impact the choices of a rational agent.
-A well behaved rational agent *should* ignore uncertainty information, so long as the estimate provided is still the best case.
-This point seems to be lost on many uncertainty visualisation authors.
-@Zhao2023 displayed a model's prediction and its estimated uncertainty and asked participants if they wanted to submit the model estimate as their answer, or make their own prediction.
-At no point did the authors seem to realise a rational agent would accept the model estimate every time, no matter how uncertain the information was, and participants were actually disqualified from the study for taking this approach to answering the questions.
-(*Cite Gap 17: Examples of studies where the correct choice was to ignore the uncertainty information*)
-
-This leaves us with several issues in decision making experiments.
-They are noisy because of participants individual utility functions and when that is not the case they are thinly veiled and over complicated value extraction experiments.
-Additionally, authors don't even seem to agree how participants should incorporate uncertainty information, with some authors designing complicated utility functions, and others not realising the best response to their experiment is to ignore uncertainty information entirely.
-For these reasons, uncertainty visualisation experiments should either be designed with these caveats in mind and be designed to untangle why a particular visualisation will elicit a specific decision, or the field should steer away using decision quality and related metrics as a basis for evaluating visualisation performance.
-
-#### Trust and confidence
-Authors also often measure the impact of visualising uncertainty on participants trust in an estimate.
-Trust is a by product of displaying uncertainty rather than the goal of it, and viewing the relationship in the converse direction can lead to misguided research.
-
-If the purpose of displaying uncertainty information is to appropriately hedge a signal with noise, then it should be assumed that trust is only related to uncertainty communication through increased transparency and honesty.
-Considering trust, and not transparency, as the metric of importance in uncertainty communication can lead to a questionable subtext that argues against transparency, something that has been noticed by several other authors [@Spiegelhalter2017; ONeill2018].
-@Hullman2020a found that author simultaneously argued that failing to visualise uncertainty was akin to fraud, but also many avoided uncertainty visualisation because they didn't want their work to come across as "untrustworthy".
-These authors are optimising *trust* rather than *transparency*, which means they opt to leave out uncertainty information when it does exactly what it is supposed to, decrease certainty in conclusions.
-
-Science communication should be primarily concerned with accuracy.
-Setting trust and risk-aversion as the variables of interest implicitly encourages statisticians to set trust and risk-aversion as the primary goals of communication.
-The issue of trust being divorced from trustworthiness has been commented on by other authors [@ONeill2018], however the issue still persists in the uncertainty visualisation literature.
-@Zhao2023 displayed a several variations of a visualisation of a model prediction and its uncertainty and took participants using the model prediction as a sign of trust.
-They reported that visualising uncertainty information caused participants to trust the model in the low variance case, but the results in the high variance case were inconclusive.
-The discussion made it clear the authors thought the uncertainty information should make the visualisation more trustworthy, but conflating trust and the use of a prediction implied uncertainty information should somehow influence participants to use their own prediction, even though a prediction being uncertainty does not necessarily mean it is incorrect.
-Despite this, the authors seemed to assume that the uncertainty information *should* have an influence on that, showing they had not deeply considered *how* uncertainty information should influence the choices of the participants.
-(*Cite Gap 18: examples of studies where authors measure trust*)
-
-A similar measure to trust is using "confidence" in an extracted value or a decision.
-Interestingly, "confidence" is also used to try and capture the clarity of a message in a normal visualisation.
-Confidence cannot simultaneously be a measure of clarity of visualisation *and* a way to capture the uncertainty expressed in a visualisation.
-
-#### Other (questionable) attempts to capture uncertainty
+This does not mean that visualising uncertainty as a signal is incorrect or bad, it just means that uncertainty as a signal is no different from any other information visualisation.
+
+- decision making experiments are usually displaying the wrong signals
+
+The final method used by authors is to just explicitly ask about uncertainty and signal information separately.
+@Sanyal2009 mapped uncertainty to dots and signal to a 3D surface and asked participants to identify areas of high and low signal and high and low uncertainty.
+Participants were not asked to combine that information in any way, and the signal and the noise were treated as separate variables.
+@Correll2014 asked participants to separately extract the mean and variance from four uncertainty visualisations.
+These methods explicitly view the uncertainty and signal as two separate variables that should be extracted from a plot, and not two variables that should be interpreted together.
+Even viewing these questions as a routine check to make sure the signal information isn't impacted by the uncertainty is counter intuitive, because the whole point *of* the uncertainty is to impact the signal information.
+
+# How should we test uncertainty visualisations?
+If asking direct questions about uncertainty causes us to treat it as a signal, how do we evaluate uncertainty as *noise*?
+ When we ask the viewer of a plot to look at data and extract a value, we are asking them to perform inference on that value.
+There will be noise associated with that answer and that is uncertainty.
+If we ask direct question about some uncertainty metric, we have turned the uncertainty into signal because that is what the participants are drawing inference on.
+
+## Current attempts to measure uncertainty
 There is also a swath of studies that are aware a question that boils down to a value extraction experiment, or a question that should be answered by literally ignoring uncertainty information is not what we want when we consider uncertainty visualisations.
@@
 
-The final method used by authors is to just explicitly ask about uncertainty and signal information separately.
-@Sanyal2009 mapped uncertainty to dots and signal to a 3D surface and asked participants to identify areas of high and low signal and high and low uncertainty.
-Participants were not asked to combine that information in any way, and the signal and the noise were treated as separate variables.
-@Correll2014 asked participants to separately extract the mean and variance from four uncertainty visualisations.
-These methods explicitly view the uncertainty and signal as two separate variables that should be extracted from a plot, and not two variables that should be interpreted together.
-Even viewing these questions as a routine check to make sure the signal information isn't impacted by the uncertainty is counter intuitive, because the whole point *of* the uncertainty is to impact the signal information.
-
 These examples are a bit complete mishmash of methods, however they point to a larger issue that goes beyond decision making, trust and confidence experiments.
@@
 
-## Suggestions to measure uncertainty
+## How to test uncertainty as noise
 So, the current methods of measuring or understanding the role of uncertainty in a visualisation is questionable at best, however this is not because visualisation authors are missing the mark, but rather uncertainty is *particularly* difficult to express in a visualisation.
@@
 
-### More specific hypothesis
+### Qualitative Studies
+Alternatively visualisation research could shift away from the accuracy concept all together ask questions that allow for open ended responses.
+This method can enlighten authors as to *how* the uncertainty information was used by the participants.
+@Hofmann2012 tried to capture this by asking participants why they considered a particular plot to be more "right shifted", however this qualitative assessment does not seem to have made it into the final paper.
+@Daradkeh2015 presented participants with ten investment alternatives and asked participants "from among available alternatives, which alternative do you prefer the most", and were asked to think aloud and consider the uncertainty in their decision making.
+The experimenters goal was to observe and organise the methods people use when making decisions in the face of uncertainty.
+This study was an excellent example in a useful experimental design.
+They highlighted the specific aspects of uncertainty that participants typically considered, such as the range of outcomes that are above/below a certain threshold, minimum and maximum values, the risk of a loss, etc, and mapped where in the decision making process participants made these considerations.
+Data visualisation is commonly utilised as a tool in data exploration, so it is not uncommon for a data analyst to make a plot with only a vague goal and pull out a large number of adjacent observations.
+This experimental framework could replicate this process.
+
+### Just noticeable signal
+It could be argued that a well done uncertainty visualisations should have an imperceptible signal unless the signal would be identified with a hypothesis test, almost like a reverse line up protocol, but this idea also has some issues that should be considered.
+The reject or do not reject concepts in hypothesis testing do not offer a complete image of uncertainty, and exploration of uncertainty visualisation largely stems from a desire to move away from this binary framework.
+
+![The visualisation showing the kind of signal that we would expect an uncertainty plot to supress](signalsupression.jpeg){#fig-sigsupress}
+
+Additionally, setting up an uncertainty visualisation where the participants are expected to notice the signal once the data behind the visualisation passes a hypothesis test implies the signal *is* noticeable to a human at that level.
+@Patrick2023 compared people ability to recognise patterns in a residual plot to typical statistical tests and found human viewers looking at a plot were less sensitive than the typical residual tests.
+These experiments utilised the line-up protocol which has been suggested as a method to check if perceived patterns are real or merely the result of chance [@Buja2009; @Wickham2010; @Chowdhury].
+This concept bears similarity to the goal of uncertainty visualisation, but it is not quite the same.
+@fig-hypvs shows the conceptual difference between the line-up protocol and uncertainty visualisation.
+
+![A line-up protocol displays the uncertainty about the null and identifies if the true data plot is identifiable (and therefore significantly difference).
+This can be considered the graphical equivalent of a standard hypothesis test.
+Uncertainty is frequently used to descibe the area around the estimate that we think might contain the true value, and we assesses if the null of "no signal" is within this plot (e.g.
+if the error bars overlap with zero).
+Uncertainty is technically the distance between the true value and the estimate.
+](hypvs.jpeg){#fig-hypvs}
+
+## Other elements of uncertainty visualisations
+- Things that are true for both normal visualisation and uncertainty, but of particular interest for uncertainty
+
+### Heuristics
+Additionally, these heuristics and biases can change depending on the larger scope of the graphic and the population we are communicating with [@Spiegelhalter2017; @Kinkeldey2014].
+
 Heuristic checks are useful because they look at unknown pitfalls that might exist in interpretation of current plots.
@@
 
+##
 In a similar vein, experiments that verify smaller aspects of plot design might be more useful to the field in the long run because it helps contribute to a larger working theory of "how do we see visualisations".
@@
 Small perceptual tasks that seek to answer small but highly relevant questions (for example, if colour hue and colour value can be perceived as a single signal suppressed variable) would be useful to the field.
-
-### Qualitative Studies
-Alternatively visualisation research could shift away from the accuracy concept all together ask questions that allow for open ended responses.
-This method can enlighten authors as to *how* the uncertainty information was used by the participants.
-@Hofmann2012 tried to capture this by asking participants why they considered a particular plot to be more "right shifted", however this qualitative assessment does not seem to have made it into the final paper.
-@Daradkeh2015 presented participants with ten investment alternatives and asked participants "from among available alternatives, which alternative do you prefer the most", and were asked to think aloud and consider the uncertainty in their decision making.
-The experimenters goal was to observe and organise the methods people use when making decisions in the face of uncertainty.
-This study was an excellent example in a useful experimental design.
-They highlighted the specific aspects of uncertainty that participants typically considered, such as the range of outcomes that are above/below a certain threshold, minimum and maximum values, the risk of a loss, etc, and mapped where in the decision making process participants made these considerations.
-Data visualisation is commonly utilised as a tool in data exploration, so it is not uncommon for a data analyst to make a plot with only a vague goal and pull out a large number of adjacent observations.
-This experimental framework could replicate this process.
-
-### Just noticeable signal
-It could be argued that a well done uncertainty visualisations should have an imperceptible signal unless the signal would be identified with a hypothesis test, almost like a reverse line up protocol, but this idea also has some issues that should be considered.
-The reject or do not reject concepts in hypothesis testing do not offer a complete image of uncertainty, and exploration of uncertainty visualisation largely stems from a desire to move away from this binary framework.
-
-![The visualisation showing the kind of signal that we would expect an uncertainty plot to supress](signalsupression.jpeg){#fig-sigsupress}
-
-Additionally, setting up an uncertainty visualisation where the participants are expected to notice the signal once the data behind the visualisation passes a hypothesis test implies the signal *is* noticeable to a human at that level.
-@Patrick2023 compared people ability to recognise patterns in a residual plot to typical statistical tests and found human viewers looking at a plot were less sensitive than the typical residual tests.
-These experiments utilised the line-up protocol which has been suggested as a method to check if perceived patterns are real or merely the result of chance [@Buja2009; @Wickham2010; @Chowdhury].
-This concept bears similarity to the goal of uncertainty visualisation, but it is not quite the same.
-@fig-hypvs shows the conceptual difference between the line-up protocol and uncertainty visualisation.
-
-![A line-up protocol displays the uncertainty about the null and identifies if the true data plot is identifiable (and therefore significantly difference).
-This can be considered the graphical equivalent of a standard hypothesis test.
-Uncertainty is frequently used to descibe the area around the estimate that we think might contain the true value, and we assesses if the null of "no signal" is within this plot (e.g.
-if the error bars overlap with zero).
-Uncertainty is technically the distance between the true value and the estimate.
-](hypvs.jpeg){#fig-hypvs}
-
-# Great Examples 
-Despite the common problems detailed in the previous sections, there is some interesting work in the uncertainty visualisation space.
-This can come in the form of an uncertainty visualisation that attempts to visualise a typically ignored aspect of uncertainty, or an experiment that avoids the pitfalls detailed in the previous sections.
-
-The literature focus on "quantifiable" uncertainty leaves the variance that occurs at early stages in our analysis that is difficult to quantify ignored and forgotten.
-Some authors have chosen to focus on these unquantifiable cases and look at expressing more complicated cases of uncertainty.
-@Tierney2023 builds upon the tidy data principles to allow users to handle missing values.
-This includes data plots with a missing value "shadow" that allows visualisation authors to identify if the variables used in a plot have any structure in their missing values, which would contribute to uncertainty.
-Another example of uncertainty that is often ignored is the uncertainty resulting from human choices.
-Climate scenario uncertainty, shown in @fig-climatescenario, attempts to display the range of climate change outcomes that can result from a range of best and worse case human choices.
-
-![Notice: I will make my own Eversion of a climate scenario plot, this one is just here from a random site as a place holder.
-A climate scenario plot shows the forecasts of different emission scenarios depending on human choices.
-The model visualisation wants to show the impact of human choice on the visualisation, rather than assume it away, so five different cases are shown in the model.
-Each of the case studies also depicts the statistical uncertainty of the model using a confidence interval.
-](climatescenario.png){#fig-climatescenario}
-
-Another visualisation method that has a lot of potential is visualisation of samples.
-Visualisations that opt to express a signal as a sample rather than an estimate have the potential to suppress signal since it is not explicitly visualised, however this has yet to be shown in evaluation experiments.
-This is not to say that visualisations of mass would not be able to perform signal suppression, but a sample can easily be expressed using aesthetics such as colour on a map and mass visualisation often struggling with issues such as over or under smoothing.
-These sampling methods can show more of the messiness of the data that sits behind a model.
-This may not have a detrimental effect on the viewers ability to extract global statistics, as it seems they can be extracted from a visualisation of a sample with ease [Franconeri2021].
-Sample visualisations have been used in maps with the pixel map, shown in @fig-samples [Lucchesi2017], but is more commonly used in animation with the HOPs plots [@Hullman2015] or similar concepts [@Blenkinsop2000].
-This method has also been adopted by visualisation authors outside of academia as can be seen in the  and the New York Times class mobility figure, shown in @fig-samples, an animated version of which can be found [here](https://www.nytimes.com/interactive/2018/03/19/upshot/race-class-white-and-black-men.html).
-
-::: {#fig-samples layout-ncol=2}
-
-![Pixel map](pixelmap.png){#fig-pixelmap}
-
-![Outcome plot](classmobility.png){#fig-nyt}
-
-Notice: Will replace both plots with my own r visualisations.
-The pixel map depicts a map of the different poverty rates in local government areas in the US state of Missouri.
-The NYT visualisation shows an screencap of an animation that depicts the economic status of white and black boys who were raised in wealthy families.
-By displaying a psudo-sample rather than an estimate and a variance, the visualisation does some signal suppression as the true poverty rate is masked and can only be extracted by taking the global statistic of a sample.
-You can also depict the signal using explicit values, as the the NYT visualisation did, but it is important that the main visual features depict the uncertainty in the estimate.
-:::
-
-A final method to perform signal suppression is simply to visualise the data if it is available and relevant to the uncertainty distribution.
-An example of this is shown in @fig-census for racial distributions spatially in America, an interactive version of the plot can be found [here](https://edition.cnn.com/interactive/2021/us/census-race-ethnicity-map/).
-This map shows the typical causes of uncertainty in a spatial model, (e.g.
-regions where data is sparse, ethnically diverse areas, uneven distribution of points within boundaries, etc) but it avoids the need to create a visualisation with a specific signal in mind.
-This is the technique typically employed by exploratory data analysis, which means it's lack of a specified signal means there is both *no* uncertainty (since we are technically not performing inference) but in the event we *do* implicitly perform inference, there is some hedging.
-In this sense, the best uncertainty visualisation you can get without specifying a signal you want to convey is visualising the data itself.
-The census dot map's addition of interactivity also allows users to zoom in and see the details that caused "uncertainty" in the form of inconsistent colours at lower resolutions when they were zoomed out.
-This does not mean that visualising raw data instead of implementing sampling techniques is always a valid uncertainty visualisation that will prevent insignificant signal from getting through.
-@Buja2009 illustrated how groups that appear linearly separable in a linear discriminant analysis (LDA) visualisation of the data can actually be the result of a LDA performed on too many variables, something that was not clear from the visualisation until the line-up protocol was implemented.
-However it is simple but effective option that seems to be largely overlooked by the uncertainty visualisation community.
-
-![Census dot map](censusdotmap.png){#fig-census}
 

### 2024-06-20 writingdumpster.qmd
a large amount of restructuring

@@
 
+For example, @Hofman2020 asked participants to judge the effectiveness of a particular treatment.
+One group was shown the prediction interval around the mean, which indicates the uncertainty associated with treatment *for a particular person* and another group was shown the the sampling interval which depicted the uncertainty associated with the *average effectiveness of the treatment*.
+The contribution of this experiment hinges on the belief that sharing a common descriptive statistic (in this case, the mean) is grounds to compare two unrelated inferential statistics, a belief which is clearly false when viewing uncertainty through the lens of inference.
+
+@Boukhelifa2012 tried to quantify the strength of the intuitive connection between a line attribute called "sketchiness" and uncertainty.
+Participants were shown the six scenarios depicted in @fig-sketchy and asked to interpret what they believed the squiggly line indicates.
+The authors were aware that uncertainty seemed to have some "task" dependence but interpreted this as *context* dependence rather than dependence on a particular inferential statistic.
+Therefore, "sketchiness" was added to aspects of the visualization that had no obvious inferential statistic associated with it, so participants simply ignored the "sketchiness" or assumed it represented something else, such as an alternative option.
+
+![The graphics displayed by @Boukhelifa2012 to identify if there is an intuitive connection between sketchiness and uncertainty.
+These graphic were made without concern as to what (i.e.
+which statistic) in the image is supposed to be uncertain.
+This leads to the images being difficult to connect to uncertainty even if we assume tha that is what it is supposed to represent.
+For example, (S5) implies that participants should have read the plot to mean that the creators of a rail network map would be "uncertain" about the existence of a train line, something that is defies common logic.](sketchiness.png){#fig-sketchy width=50%}
+
+@Padilla2021 found that high uncertainty in the model estimates and low forecaster confidence  both caused participants to have decreased confidence in their results and suggested modelers express both if they are relevant.
+
+@Kale2019 discusses the importance of communicating decisions made in the data analysis pipeline and being aware of the alternatives.
+
 @Tierney2023 generates a visualisation for missing values, however because it fits nicely into the grammar of graphics framework, it is inherrently a descriptive statistic.
@@
 Utilizing bias and variance to define uncertainty is well-established, but it is absent concerns about imputed data, assumptions, sampling methods, and even the human element of analyst decisions
+
+# Great Examples 
+Despite the common problems detailed in the previous sections, there is some interesting work in the uncertainty visualisation space.
+This can come in the form of an uncertainty visualisation that attempts to visualise a typically ignored aspect of uncertainty, or an experiment that avoids the pitfalls detailed in the previous sections.
+
+The literature focus on "quantifiable" uncertainty leaves the variance that occurs at early stages in our analysis that is difficult to quantify ignored and forgotten.
+Some authors have chosen to focus on these unquantifiable cases and look at expressing more complicated cases of uncertainty.
+@Tierney2023 builds upon the tidy data principles to allow users to handle missing values.
+This includes data plots with a missing value "shadow" that allows visualisation authors to identify if the variables used in a plot have any structure in their missing values, which would contribute to uncertainty.
+Another example of uncertainty that is often ignored is the uncertainty resulting from human choices.
+Climate scenario uncertainty, shown in @fig-climatescenario, attempts to display the range of climate change outcomes that can result from a range of best and worse case human choices.
+
+![Notice: I will make my own Eversion of a climate scenario plot, this one is just here from a random site as a place holder.
+A climate scenario plot shows the forecasts of different emission scenarios depending on human choices.
+The model visualisation wants to show the impact of human choice on the visualisation, rather than assume it away, so five different cases are shown in the model.
+Each of the case studies also depicts the statistical uncertainty of the model using a confidence interval.
+](climatescenario.png){#fig-climatescenario}
+
+Another visualisation method that has a lot of potential is visualisation of samples.
+Visualisations that opt to express a signal as a sample rather than an estimate have the potential to suppress signal since it is not explicitly visualised, however this has yet to be shown in evaluation experiments.
+This is not to say that visualisations of mass would not be able to perform signal suppression, but a sample can easily be expressed using aesthetics such as colour on a map and mass visualisation often struggling with issues such as over or under smoothing.
+These sampling methods can show more of the messiness of the data that sits behind a model.
+This may not have a detrimental effect on the viewers ability to extract global statistics, as it seems they can be extracted from a visualisation of a sample with ease [Franconeri2021].
+Sample visualisations have been used in maps with the pixel map, shown in @fig-samples [Lucchesi2017], but is more commonly used in animation with the HOPs plots [@Hullman2015] or similar concepts [@Blenkinsop2000].
+This method has also been adopted by visualisation authors outside of academia as can be seen in the  and the New York Times class mobility figure, shown in @fig-samples, an animated version of which can be found [here](https://www.nytimes.com/interactive/2018/03/19/upshot/race-class-white-and-black-men.html).
+
+::: {#fig-samples layout-ncol=2}
+
+![Pixel map](pixelmap.png){#fig-pixelmap}
+
+![Outcome plot](classmobility.png){#fig-nyt}
+
+Notice: Will replace both plots with my own r visualisations.
+The pixel map depicts a map of the different poverty rates in local government areas in the US state of Missouri.
+The NYT visualisation shows an screencap of an animation that depicts the economic status of white and black boys who were raised in wealthy families.
+By displaying a psudo-sample rather than an estimate and a variance, the visualisation does some signal suppression as the true poverty rate is masked and can only be extracted by taking the global statistic of a sample.
+You can also depict the signal using explicit values, as the the NYT visualisation did, but it is important that the main visual features depict the uncertainty in the estimate.
+:::
+
+A final method to perform signal suppression is simply to visualise the data if it is available and relevant to the uncertainty distribution.
+An example of this is shown in @fig-census for racial distributions spatially in America, an interactive version of the plot can be found [here](https://edition.cnn.com/interactive/2021/us/census-race-ethnicity-map/).
+This map shows the typical causes of uncertainty in a spatial model, (e.g.
+regions where data is sparse, ethnically diverse areas, uneven distribution of points within boundaries, etc) but it avoids the need to create a visualisation with a specific signal in mind.
+This is the technique typically employed by exploratory data analysis, which means it's lack of a specified signal means there is both *no* uncertainty (since we are technically not performing inference) but in the event we *do* implicitly perform inference, there is some hedging.
+In this sense, the best uncertainty visualisation you can get without specifying a signal you want to convey is visualising the data itself.
+The census dot map's addition of interactivity also allows users to zoom in and see the details that caused "uncertainty" in the form of inconsistent colours at lower resolutions when they were zoomed out.
+This does not mean that visualising raw data instead of implementing sampling techniques is always a valid uncertainty visualisation that will prevent insignificant signal from getting through.
+@Buja2009 illustrated how groups that appear linearly separable in a linear discriminant analysis (LDA) visualisation of the data can actually be the result of a LDA performed on too many variables, something that was not clear from the visualisation until the line-up protocol was implemented.
+However it is simple but effective option that seems to be largely overlooked by the uncertainty visualisation community.
+
+![Census dot map](censusdotmap.png){#fig-census}

### 2024-06-20 paper.qmd
finished general restructure for signal supression perspective

@@
 - May not need to reference the same paper again and again
-- Update background to reflect new structure of paper
+- Update background to reflect new motivation
 -->
@@
 Unlike an infographic, a statistical graphic is one of intent, and all enhancements made on a statistical graphic should be in pursuit of that goal (@Vanderplas2020).
-Visualization is an important step in exploratory data analysis and communication.
+Visualization is an important step in exploratory data analysis (EDA) and communication.
 At the exploration stage, it is utilized to **learn** what is important about a data set while the communication stage is utilized to **tell** others about what we have learned.
@@
 
-The review provided here attempts to address these issues.
-We start by discussing why uncertainty visualisation has become it's own sub-field, distinct from "normal" information visualisation.
-In this discussion we establish the purpose of uncertainty visualisation is to perform "signal-suppression", that is, hide statistically insignificant signals while promoting statistically rigorous signals.
-**EDIT FROM HERE** We then move highlight the three key issues facing the uncertainty visualisation literature, specifically the lack of a definition of uncertainty, the confusion as to what *is* an "uncertainty visualisation", and the difficulties with measuring "uncertainty".
-When disucssing each of these issues, we will discuss how this concept is understood in existing literature, why this understanding may be flawed, and the issues that arrise due to these flaws.
+The review provided here attempts to address these issues by offering a novel perspective on the uncertainty visualisation problem.
+We start by discussing the purpose of uncertainty visualisations and why it has become it's own sub-field, distinct from "normal" information visualisation.
+Following this, we discuss what is an uncertainty definition and what differentiates it from a "normal" visualisation.
+Finally, we discuss how uncertainty visualisations can be evaluated and why evaluation of uncertainty visualisations can be more difficult than evaluations of "normal" visualisations.
 By considering the *purpose* of visualizing uncertainty we avoid providing a simple list of the current methods, and instead consider if the current visualisation tools are sufficient to achieve that purpose and if not, where there is need for new methodology.
-The ideal role of this review to help synthesize this noisy field, by collecting together essential components, summarizing different viewpoints the current literature, and provide a guide for potential research topics.
-**TO HERE**
 
-# The purpose of uncertainty visualisation
+# Why is uncertainty special?
 It is rarely made clear exactly how uncertainty visualisation is different to normal information visualisation.
@@
 A visualisation is nothing more than computing a statistic, mapping that statistic to a visual feature and rendering that graphic.
-Nothing in this process suggests visualising an estimate of the variance should be any different to visualising an estimate of the mean, however this field *does* implicitly treat them different.
-Why?
+Nothing in this process suggests visualising an estimate of the variance should be any different to visualising an estimate of the mean, median, maximum, or of any other statistic, however this field *does* implicitly treat them different.
+There is no such thing as "maximum visualiation" or "median visualisation" so there *must* be something about uncertainty that causes authors to believe it is special.
+Here we seek to answer this question.
 
+### The purpose of uncertainty visualisations
 Uncertainty visualisation papers usually provide one of two primary justifications for their existence.
@@
 
-# Signal-Supression
+## What is signal-supression?
 The true reason that uncertainty visualisation is differerent to normal information visualisation, is that normal information visualisation is designed to *find* signals in our data, while uncertainty visualisation should be designed to *supress* it.
@@
 
-## Uncertainty does not exist without signal
-It is obvious that we cannot design a visualisation for signal suppression without a clear signal.
-This may be considered a limitation by some readers, however it is less of a restriction when we recognize that *uncertainty* itself does not exist without a signal.
-
-#### The current definitions of uncertainty
-Uncertainty is a concept that is famously hard to pin down, with constantly shifting definitions that mean it is quantified differently from author to author.
-
-The ramifications of this is seen throughout the literature, as a survey of visualisation authors cited "not knowing how to calculate uncertainty" as one of the primary reasons they did not include it in visualisations [@Hullman2020a].
-Some works [@Hullman2018; @Maceachren2012; @Thomson2005] get around this definition issue by focusing narrowly on specific terms defined mathematically, such as probability, variance, error, or precision.
-Others [@Griethe2006; @Pang1997; @Pham2009; @Boukhelifa2017] include broader loosely related elements, such as missing values, but ignore these concepts when translating that uncertainty to a visualisation.
-Existing mathematical definition of uncertainty (or related concepts) have been unable to unify this range of diverse concepts.
-For example, @Thomson2005 suggests a mathematical formula for *examples* of uncertainty, and information theory tries to quantify uncertainty using the idea of entropy, but the disconnect between the broad concept of uncertainty and what we can reliably quantify remains a problem.
-
-Some authors prefer to define uncertainty in qualitative terms, often through the use of taxonomies.
-In these cases, uncertainty is split using an endless stream of ever changing boundaries, such as whether the uncertainty is due to true randomness or a lack of knowledge [@Spiegelhalter2017; @Hullman2016; @utypo], if the uncertainty is in the attribute, spatial elements, or temporal element of the data [@Kinkeldey2014], whether the uncertainty is scientific (e.g.
-error) or human (e.g.
-disagreement among parties) [@Benjamin2018], if the uncertainty is random or systematic [@Sanyal2009], statistical or bounded [@Gschwandtnei2016; @Olston2002], recorded as accuracy or precision [@Griethe2006; @Benjamin2018], which stage of the data analysis pipeline the uncertainty comes from [@utypo], how quantifiable the uncertainty is [@Spiegelhalter2017; @utypo], etc.
-In their own way, each of these approaches show an aspect of uncertainty that an author felt was important to differentiate.
-
-What is interesting about all these views of uncertainty, is that they all imply that it is a latent feature of our data.
-That is, it is something that can be revealed through exploration, and not something that needs to be explicitly calculated.
-The constant stream of new definitions that are unable to capture the true essence of uncertainty might be a result of this false assumption that underpins ever definition.
-Uncertainty cannot be defined without a signal because uncertainty does not exist without a signal, is a by-product of inference.
-
-#### Uncertainty as a by-product of inference
-It is easy to see uncertainties relationship to inference when we consider what might *not* be considered uncertain rather than just trying to think about what *is*.
-
-Descriptive statistics simply describe our sample as it is and summarizes large data down into an easy to swallow format.
-Descriptive statistics are not seen as the primary goal of modern statistics, however, this was not always the case.
-Around the 19th century in England, *positivism* was the popular philosophical approach to science (positivists included famous statisticians such as Francis Galton and Karl Pearson) and practitioners of the approach believed statistics ended with descriptive statistics as science must be based on actual experience and observations, therefore anything that refers to the unobservable (such as new observations or population statistics) is not true science [@Otsuka2023].
-In order to make statements about population statistics, future values, or new observations we need to perform inference, which requires the assumption of the "uniformity of nature" (i.e.
-that unobserved phenomena should be similar to observed phenomena) [@Otsuka2023].
-This subtle shift, from descriptive statistics to inferential statistics was shunned during the positivism era *due to the fact it introduced the unknowable*, or in other words, uncertainty.
-
-This approach to uncertainty is embedded in the idea that descriptive statistics do not have uncertainty, which some readers may disagree with.
-Specifically, because it means uncertainty is *not* a latent attribute of data, but rather an attribute of a specific hypothesis or estimate.
-Deniers of this fact follow a consistent logical path and it is easy to identify the common mistake.
-We know that variance and probability are typically considered types of "uncertainty" *and* descriptive statistics can have variance and probabilities *therefore* descriptive statistics must have uncertainty.
-The flaw in this logic comes from the first step, assuming that the tools with which we measure uncertainty *are* uncertainty in of themselves.
-This confusion is common and there are many papers that spend a great deal of time clarifying the difference.
-@Begg2014 highlight that uncertainty is related to not knowing a specific value, while variability refers to the range of values a quantity can take at different locations, times or instances.
-@Spiegelhalter2017 made sure to comment on the difference between precise random events (such as the probability associated with a coin flip), and uncertainty (such as the estimated probability associated with a coin that might be biased).
-The variance of a sample variance can be calculated and know, therefore it is not uncertain but rather it a precise description of dispersion.
-If we were to discuss drawing a new observation, or estimating the true mean of a population *then* the variance would become relevant in our discussions of uncertainty.
-
-Several authors have connected the issues around defining uncertainty to inference, however it is often discussed as a *task* or *goal* dependence.
-Multiple authors have commented on the need to consider quantifying and expressing uncertainty at every stage of a project as the "goal" shapes every step of the analysis [@Kinkeldey2014; @Hullman2016; @Refsgaard2007].
-@Otsuka2023 suggested that the process of observing data to perform statistics is largely dependent on our goals, because the process of boiling real world entities down into probabilistic objects (or "probabilistic kind" as he puts it) depends on the relationship we seek to identify with our data.
-@Meng2014 commented what is kept as data and what is tossed away is determined by the motivation of an analysis and what was previously noise can be shown to become signal depending on the the question we seek to answer.
-@Kale2019 discussed how the choices we make in our analysis impact our outcomes and introduce uncertainty.
-@Carlin2023 mentions that each research question can be can be categorised as descriptive, predictive, or causal, each of which has its own appropriate statistical methods and motivation agnostic model selection leads to statistical analysis that is devoid of meaning.
-@Wallsten1997 argue that the best method for evaluating or combining subjective probabilities depends on the uncertainty the decision maker wants to represent and why it matters.
-@Fischhoff2014 looks at uncertainty visualisation for decision making decides that we should have different ways of communicating uncertainty based off what the user is supposed to do with it.
-The importance of inference when discussing uncertainty is never directly acknowledge, but it is always present.
-It is clear that the relationship between uncertainty and inference is noticed at every stage in an analysis, however combining these uncertainties into a single "uncertainty" value is near impossible [@Spiegelhalter2017].
-
-With this understanding it becomes clear to see why uncertainty is tied to an endless string of examples in the data analysis pipeline.
-Uncertainty examples include imputed data, model selection, inherent randomness, biased sampling, etc, not because these things *are* uncertainty, but because they *create* uncertainty when we perform inference.
-Whether or not these elements are relevant is highly dependent on what statistic you are trying to draw inference on, and by extension, the purpose of your visualisation.
-
-## EDA vs Communication
-The concept of uncertainty as a by-product of inference, not a latent feature of data, extends naturally to visualisation.
-
-There are two primary reasons for making a visualisation, to perform exploratory data analysis (EDA) or communication.
-Communication involves identifying a signal we want to communicate and designing the visualisation that best conveys that, while EDA involves creating a versatile visualisation without an explicit purpose and using it to extract several signals.
-Since EDA is the visual parallel to descriptive statistics, it is performed without an explicit hypothesis which means there *is* no uncertainty in the visualisation.
-Similarly to our issues with descritive statistics, this is not well understood by the uncertainty visualisation community.
-Some authors simply confuse varianc for uncertainty.
-For example @Potter2010 aimed to create a summary plot that "concisely presented data with uncertainty information" to create an exploratory visualisation tool that visualised uncertainty.
-Other authors regonise inference will occur (in some shape or form) and believe uncertainty *should* be visualised but do not recognise *how* uncertainty would be visualised.
-@Hullman2021 argued that there is no such thing as a "model-free" visualisation, therefore visualisation require robust visualisations of uncertainty as we are always performing inference.
-@Griethe2006 commented that "if visualization is used as a means to explore a data volume or to communicate its contents the uncertainty has to be included".
-While we agree people cannot prevent themselves from performing inference, that does not mean it is possible for uncertainty to be included in an EDA visualisation.
-A versatile visualisation such as a scatter plot allows for a viewer to consider several hypothesis at once, each of which will be a different inferential statistic with a different distribution that depicts its uncertainty.
-It is likely impossible to suppress all possible signals at once.
-
-### No Such thing as an "uncertainty visualisation"
-If uncertainty is a by-product of inference, a visualisation being an "uncertainty visualisation" does not exist.
-
-What exactly is or is not an uncertainty visualisation is just as poorly defined as the definition of uncertainty, so authors seem to decide for themselves what is an uncertainty visualisation.
-For example @Leland2005 mentions that popular graphics, such as pie charts and bar charts omit uncertainty, however at least one or both of these charts are used in a significant number of uncertainty visualisation experiments [@Ibrekk1987; @Olston2002; @Zhao2023; @Hofmann2012].
-@Wickham2011 suggests their product plot framework, which includes histograms, should have a way to measure uncertainty, but does not consider that a histogram is *already* a depiction of PDF, something that is often considered an uncertainty visualisation by other authors.
-
-@Kinkeldey2014 categorised uncertainty according to five criteria depicted in @fig-tax which considers if a visualisation is implicit or explicit, intrinsic or extrinsic, visually integrable or seperable, coincident or adjacent, and static or dynamic.
-A similar version of this taxonomy was presented by @uncertchap2022 who commented that visualisation can be organised into two categories, "graphical annotations of distributional properties" and "visual encodings of uncertainty" which seems to functionally align with the intrinsic/extrinsic distinction by @Kinkeldey2014.
-@Griethe2006 organised uncertainty visualisations into two cases (1) a hypothesis test was preformed to confirm the validity of the visualisation and (2) the visualisation has uncertainty depicted.
-@Potter2012 organised several existing uncertainty visualisations into groups based on the dimensionality of the data (1D, 2D, 3D, and No Dimension) and the dimensionality of the uncertainty (Scalar, Vector, Tensor).
-However, because the term "PDF", function that is used to describe a random variable, is used to describe both the data and the uncertainty for all dimensions.
-@Grewal2021 created a taxonomy that mapped uncertainty visualisations to some point in a 2D space defined by the "domain expertise" and "continuum of discreteness" (that scaled from "point estimate" to "continuous distribution").
-
-![Visualisation of @Kinkeldey2014 uncertainty visualisation taxonomy.
-It can be seen that most of the categories can be considered a change to the typical error bar chart.
-The taxonomy categorises the visualisations based on five criteria, implicit (a sample) or explicit (a depiction of mass); Intrinsic (alter existing symbols symbols to represent uncertainty) or extrinsic (add new objects to represent uncertainty); visually integral or separable (the uncertainty can be separated from the data and read independently) ; coincidence (uncertainty and data are represented in the same plot) or adjacent; static or dynamic (animated, interactive etc).
-Many of the distinctions conflate changes of the input data and statistics calculated and with visual changes, highlighting how defining uncertainty and visualisation together conflates several elements of a graphic.](uncertvistax.jpeg){#fig-tax}
-
-The authors of these taxonomies did not intend for them to be complete descriptions of uncertainty visualisations, however they do paint a picture of the what authors consider to be important.
-There are common themes in the taxonomies that ensure we differentiate between a sample and a mass, consider if uncertainty should be depicted with the estimate or as a second variable, identify the parallel of hypothesis testing, and consider the dimensionality, sample size and precision depicted by the visualisation.
-These taxonomies may highlight perceived important differences, but they do not help us understand what *is* or *is not* an uncertainty visualisation.
-For that we need to understand *why* uncertainty visualisation exists as a sub-field at all.
-
-## Designing Visualisation for Signal supression
 Interpretation and semantics experiments are seeking to identify a dimension (or visual task) that uncertainty naturally maps to.
@@
 
-# Visualising uncertainty as signal
+## Uncertainty visualisations vs information visualisation
 If uncertainty is inherently different, we would expect evaluation studies of uncertainty visualisations to produce different results to similar evaluation studies on "normal" variables.
@@
 
-# How should we test uncertainty visualisations?
+# What even is an uncertainty visualisation?
+## The literatures definition of uncertainty visualisations
+What exactly is an uncertainty visulisation is never explained and certainly does not have a consistent definition.
+For example @Leland2005 mentions that popular graphics, such as pie charts and bar charts omit uncertainty, however at least one or both of these charts are used in a significant number of uncertainty visualisation experiments [@Ibrekk1987; @Olston2002; @Zhao2023; @Hofmann2012].
+@Wickham2011 suggests their product plot framework, which includes histograms, should have a way to measure uncertainty, but does not consider that a histogram is *already* a depiction of PDF, something that is often considered an uncertainty visualisation by other authors.
+
+### What is an uncertainty visualisation?
+@Kinkeldey2014 categorised uncertainty according to five criteria depicted in @fig-tax which considers if a visualisation is implicit or explicit, intrinsic or extrinsic, visually integrable or seperable, coincident or adjacent, and static or dynamic.
+A similar version of this taxonomy was presented by @uncertchap2022 who commented that visualisation can be organised into two categories, "graphical annotations of distributional properties" and "visual encodings of uncertainty" which seems to functionally align with the intrinsic/extrinsic distinction by @Kinkeldey2014.
+@Griethe2006 organised uncertainty visualisations into two cases (1) a hypothesis test was preformed to confirm the validity of the visualisation and (2) the visualisation has uncertainty depicted.
+@Potter2012 organised several existing uncertainty visualisations into groups based on the dimensionality of the data (1D, 2D, 3D, and No Dimension) and the dimensionality of the uncertainty (Scalar, Vector, Tensor).
+However, because the term "PDF", function that is used to describe a random variable, is used to describe both the data and the uncertainty for all dimensions.
+@Grewal2021 created a taxonomy that mapped uncertainty visualisations to some point in a 2D space defined by the "domain expertise" and "continuum of discreteness" (that scaled from "point estimate" to "continuous distribution").
+
+![Visualisation of @Kinkeldey2014 uncertainty visualisation taxonomy.
+It can be seen that most of the categories can be considered a change to the typical error bar chart.
+The taxonomy categorises the visualisations based on five criteria, implicit (a sample) or explicit (a depiction of mass); Intrinsic (alter existing symbols symbols to represent uncertainty) or extrinsic (add new objects to represent uncertainty); visually integral or separable (the uncertainty can be separated from the data and read independently) ; coincidence (uncertainty and data are represented in the same plot) or adjacent; static or dynamic (animated, interactive etc).
+Many of the distinctions conflate changes of the input data and statistics calculated and with visual changes, highlighting how defining uncertainty and visualisation together conflates several elements of a graphic.](uncertvistax.jpeg){#fig-tax}
+
+The authors of these taxonomies did not intend for them to be complete descriptions of uncertainty visualisations, however they do paint a picture of the what authors consider to be important.
+There are common themes in the taxonomies that ensure we differentiate between a sample and a mass, consider if uncertainty should be depicted with the estimate or as a second variable, identify the parallel of hypothesis testing, and consider the dimensionality, sample size and precision depicted by the visualisation.
+These taxonomies may highlight perceived important differences, but they do not help us understand what *is* or *is not* an uncertainty visualisation.
+For that we need to understand *why* uncertainty visualisation exists as a sub-field at all.
+
+### What is uncertainty?
+
+Uncertainty is a concept that is famously hard to pin down, with constantly shifting definitions that mean it is quantified differently from author to author.
+
+The ramifications of this is seen throughout the literature, as a survey of visualisation authors cited "not knowing how to calculate uncertainty" as one of the primary reasons they did not include it in visualisations [@Hullman2020a].
+Some works [@Hullman2018; @Maceachren2012; @Thomson2005] get around this definition issue by focusing narrowly on specific terms defined mathematically, such as probability, variance, error, or precision.
+Others [@Griethe2006; @Pang1997; @Pham2009; @Boukhelifa2017] include broader loosely related elements, such as missing values, but ignore these concepts when translating that uncertainty to a visualisation.
+Existing mathematical definition of uncertainty (or related concepts) have been unable to unify this range of diverse concepts.
+For example, @Thomson2005 suggests a mathematical formula for *examples* of uncertainty, and information theory tries to quantify uncertainty using the idea of entropy, but the disconnect between the broad concept of uncertainty and what we can reliably quantify remains a problem.
+
+Some authors prefer to define uncertainty in qualitative terms, often through the use of taxonomies.
+In these cases, uncertainty is split using an endless stream of ever changing boundaries, such as whether the uncertainty is due to true randomness or a lack of knowledge [@Spiegelhalter2017; @Hullman2016; @utypo], if the uncertainty is in the attribute, spatial elements, or temporal element of the data [@Kinkeldey2014], whether the uncertainty is scientific (e.g.
+error) or human (e.g.
+disagreement among parties) [@Benjamin2018], if the uncertainty is random or systematic [@Sanyal2009], statistical or bounded [@Gschwandtnei2016; @Olston2002], recorded as accuracy or precision [@Griethe2006; @Benjamin2018], which stage of the data analysis pipeline the uncertainty comes from [@utypo], how quantifiable the uncertainty is [@Spiegelhalter2017; @utypo], etc.
+In their own way, each of these approaches show an aspect of uncertainty that an author felt was important to differentiate.
+
+What is interesting about all these views of uncertainty, is that they all imply that it is a latent feature of our data.
+That is, it is something that can be revealed through exploration, and not something that needs to be explicitly calculated.
+The constant stream of new definitions that are unable to capture the true essence of uncertainty might be a result of this false assumption that underpins ever definition.
+Uncertainty cannot be defined without a signal because uncertainty does not exist without a signal, is a by-product of inference.
+
+#### Uncertainty as a by-product of inference
+It is easy to see uncertainties relationship to inference when we consider what might *not* be considered uncertain rather than just trying to think about what *is*.
+
+Descriptive statistics simply describe our sample as it is and summarizes large data down into an easy to swallow format.
+Descriptive statistics are not seen as the primary goal of modern statistics, however, this was not always the case.
+Around the 19th century in England, *positivism* was the popular philosophical approach to science (positivists included famous statisticians such as Francis Galton and Karl Pearson) and practitioners of the approach believed statistics ended with descriptive statistics as science must be based on actual experience and observations, therefore anything that refers to the unobservable (such as new observations or population statistics) is not true science [@Otsuka2023].
+In order to make statements about population statistics, future values, or new observations we need to perform inference, which requires the assumption of the "uniformity of nature" (i.e.
+that unobserved phenomena should be similar to observed phenomena) [@Otsuka2023].
+This subtle shift, from descriptive statistics to inferential statistics was shunned during the positivism era *due to the fact it introduced the unknowable*, or in other words, uncertainty.
+
+This approach to uncertainty is embedded in the idea that descriptive statistics do not have uncertainty, which some readers may disagree with.
+Specifically, because it means uncertainty is *not* a latent attribute of data, but rather an attribute of a specific hypothesis or estimate.
+Deniers of this fact follow a consistent logical path and it is easy to identify the common mistake.
+We know that variance and probability are typically considered types of "uncertainty" *and* descriptive statistics can have variance and probabilities *therefore* descriptive statistics must have uncertainty.
+The flaw in this logic comes from the first step, assuming that the tools with which we measure uncertainty *are* uncertainty in of themselves.
+This confusion is common and there are many papers that spend a great deal of time clarifying the difference.
+@Begg2014 highlight that uncertainty is related to not knowing a specific value, while variability refers to the range of values a quantity can take at different locations, times or instances.
+@Spiegelhalter2017 made sure to comment on the difference between precise random events (such as the probability associated with a coin flip), and uncertainty (such as the estimated probability associated with a coin that might be biased).
+The variance of a sample variance can be calculated and know, therefore it is not uncertain but rather it a precise description of dispersion.
+If we were to discuss drawing a new observation, or estimating the true mean of a population *then* the variance would become relevant in our discussions of uncertainty.
+
+Several authors have connected the issues around defining uncertainty to inference, however it is often discussed as a *task* or *goal* dependence.
+Multiple authors have commented on the need to consider quantifying and expressing uncertainty at every stage of a project as the "goal" shapes every step of the analysis [@Kinkeldey2014; @Hullman2016; @Refsgaard2007].
+@Otsuka2023 suggested that the process of observing data to perform statistics is largely dependent on our goals, because the process of boiling real world entities down into probabilistic objects (or "probabilistic kind" as he puts it) depends on the relationship we seek to identify with our data.
+@Meng2014 commented what is kept as data and what is tossed away is determined by the motivation of an analysis and what was previously noise can be shown to become signal depending on the the question we seek to answer.
+@Kale2019 discussed how the choices we make in our analysis impact our outcomes and introduce uncertainty.
+@Carlin2023 mentions that each research question can be can be categorised as descriptive, predictive, or causal, each of which has its own appropriate statistical methods and motivation agnostic model selection leads to statistical analysis that is devoid of meaning.
+@Wallsten1997 argue that the best method for evaluating or combining subjective probabilities depends on the uncertainty the decision maker wants to represent and why it matters.
+@Fischhoff2014 looks at uncertainty visualisation for decision making decides that we should have different ways of communicating uncertainty based off what the user is supposed to do with it.
+The importance of inference when discussing uncertainty is never directly acknowledge, but it is always present.
+It is clear that the relationship between uncertainty and inference is noticed at every stage in an analysis, however combining these uncertainties into a single "uncertainty" value is near impossible [@Spiegelhalter2017].
+
+With this understanding it becomes clear to see why uncertainty is tied to an endless string of examples in the data analysis pipeline.
+Uncertainty examples include imputed data, model selection, inherent randomness, biased sampling, etc, not because these things *are* uncertainty, but because they *create* uncertainty when we perform inference.
+Whether or not these elements are relevant is highly dependent on what statistic you are trying to draw inference on, and by extension, the purpose of your visualisation.
+
+## What is an uncertainty visualisation?
+The concept of uncertainty as a by-product of inference, not a latent feature of data, extends naturally to visualisation. 
+There are two primary reasons for making a visualisation, to perform exploratory data analysis (EDA) or communication.
+Communication involves identifying a signal we want to communicate and designing the visualisation that best conveys that, while EDA involves creating a versatile visualisation without an explicit purpose and using it to extract several signals.
+
+ Since EDA is the visual parallel to descriptive statistics, it is performed without an explicit hypothesis which means there *is* no uncertainty in the visualisation.
+Similarly to our issues with descritive statistics, this is not well understood by the uncertainty visualisation community.
+Some authors simply confuse varianc for uncertainty.
+For example @Potter2010 aimed to create a summary plot that "concisely presented data with uncertainty information" to create an exploratory visualisation tool that visualised uncertainty.
+Other authors regonise inference will occur (in some shape or form) and believe uncertainty *should* be visualised but do not recognise *how* uncertainty would be visualised.
+@Hullman2021 argued that there is no such thing as a "model-free" visualisation, therefore visualisation require robust visualisations of uncertainty as we are always performing inference.
+@Griethe2006 commented that "if visualization is used as a means to explore a data volume or to communicate its contents the uncertainty has to be included".
+While we agree people cannot prevent themselves from performing inference, that does not mean it is possible for uncertainty to be included in an EDA visualisation.
+A versatile visualisation such as a scatter plot allows for a viewer to consider several hypothesis at once, each of which will be a different inferential statistic with a different distribution that depicts its uncertainty.
+It is likely impossible to suppress all possible signals at once.
+
+# How do we evaluate uncertainty visualisations?
 If asking direct questions about uncertainty causes us to treat it as a signal, how do we evaluate uncertainty as *noise*?

### 2024-07-17 paper.qmd
redid background

@@
 - May not need to reference the same paper again and again
-- Update background to reflect new motivation
+- Expand background "why visualize" section
 -->
@@
 
-There are many stages in our analysis that might benefit from the power of data visualisation, however this does not mean it is appropriately utilized or done with with effectiveness.
 Unlike an infographic, a statistical graphic is one of intent, and all enhancements made on a statistical graphic should be in pursuit of that goal (@Vanderplas2020).
-Visualization is an important step in exploratory data analysis (EDA) and communication.
-At the exploration stage, it is utilized to **learn** what is important about a data set while the communication stage is utilized to **tell** others about what we have learned.
-Each of these visualisation goals utilize an important attribute of visualisation.
-In both these cases, visualisations are powerful tools that can help us understand **how** our data or results might diverge from pre-conceived notions or what we expect.
 
@@
 For example, box plots or histograms can be considered to be displaying uncertainty in the sense of variability in observations sampled from a population distribution.
+Today, there is an abundance of publications on the topic which makes it is timely to construct a review of the field.
+In fact, there have already been several reviews published.
 
@@
 
-However, uncertainty visualisation takes many forms today, It can be considered to expand the application of data visualisation towards making better decisions, and research is concerned about how the construction of plots and information contained might affect the resulting decisions.
-With the abundance of publications it is timely to consider a review of the state-of-the-art.
-In fact, there have already been several reviews published.
-
-Interestingly, these reviews often do not offer overarching rules for tried and tested uncertainty visualisation, but rather comment on the *difficulties* faced when trying to summarise the papers from this field.
+Reviews on uncertainty visualisation rarely offer tried and tested rules for effective uncertainty visualisation, but rather comment on the *difficulties* faced when trying to summarize the field.
 @Kinkeldey2014 found most experiments on the methods for uncertainty visualisation evaluation to be ad hoc, with no commonly agreed upon methodology or formalisation and no greater goal of describing general principals.
-@Hullman2016 commented on the difficulty in taking overarching themes from uncertainty visualisation, as several conflated issues make it unclear if subjects did poorly in an experiment because they misunderstood a visualisation, because the question was misinterpreted, or because they used a specific heuristic.
+Additionally, they noticed that the literature leaves@Hullman2016 commented on the difficulty in taking overarching themes from uncertainty visualisation, as several conflated issues make it unclear if subjects did poorly in an experiment because they misunderstood a visualisation, because the question was misinterpreted, or because they used a specific heuristic.
 @Spiegelhalter2017 commented that different plots are good for different things, and disagreed with the goal of identifying a universal "best" plot for all people and circumstances.
 @Griethe2006 was unable to find common themes, but instead listed the findings and opinions of a collection of papers.
+@uncertchap2022 summarized several cognitive effects that have repeatedly arised in uncertainty literature, however these effects were each discussed in isolation as a list of considerations an author might make.
+While these reviews are thorough in scope, none discuss how the existing literature contribute to the commonly state goal of uncertainty visualisation, scientific transparency.
+The problem faced by the literature is easily summarized with a famous quote by Henri Poincaré.
 
-There are several reasons provided to explain the difficulty in generalising the uncertainty visualisation literature.
-Some suggested or agree with the idea that visualisation typologies should move away from data types, uncertainty categories, and representation types and towards "task-centred typologies" [@Kinkeldey2014; @Hullman2016].
-Other papers indirectly hint at this by arguing that the choice of best visualisation is highly dependent on the specific goal [@Griethe2006; @Spiegelhalter2017].
-Others contend that visualisations are highly dependent on the audience and there is no such thing as a "best" visualisation that will be accessible to all members [@Kinkeldey2014; @Spiegelhalter2017].
-These concerns appear across all application areas, and an underlying thread is that the literature on uncertainty visualisations have become disconnected from their purpose.
+> "Science is built up of facts, as a house is built of stones; but an accumulation of facts is no more a science than a heap of stones is a house." - Henri Poincaré (1905)
 
-The review provided here attempts to address these issues by offering a novel perspective on the uncertainty visualisation problem.
-We start by discussing the purpose of uncertainty visualisations and why it has become it's own sub-field, distinct from "normal" information visualisation.
-Following this, we discuss what is an uncertainty definition and what differentiates it from a "normal" visualisation.
-Finally, we discuss how uncertainty visualisations can be evaluated and why evaluation of uncertainty visualisations can be more difficult than evaluations of "normal" visualisations.
-By considering the *purpose* of visualizing uncertainty we avoid providing a simple list of the current methods, and instead consider if the current visualisation tools are sufficient to achieve that purpose and if not, where there is need for new methodology.
+That is to say, despite the wealth of reviews, the field of uncertainty visualisation remains a heap of stones.
+This review attempts to address this issue by offering a novel perspective on the uncertainty visualisation problem, and hopefully laying the foundations on which we can build a house.
 
+This review is broken into several parts that each reflect a different approach to uncertainty visualisation.
+First we look at graphics that ignore uncertainty entirely and discuss why uncertainty should be included at all.
+Second, we look at methods that consider uncertainty to be just another variable and discuss the characteristics of uncertainty that make it a unique visualisation problem.
+Third, we look at methods that explicitly combine our estimate and its uncertainty and discuss if the visualisations created by these transformations are still "uncertainty visualisations".
+Fourth, we will discuss methods that implicitly include uncertainty by depicting a sample or original data in place of an estimate.
+Finally, we discuss how uncertainty visualisations can be effectively evaluated.
+When discussing each of these methods, we consider the *purpose* of uncertainty visualisation and comment on how effective each visualisation is at fulfilling that purpose.
+
+Due to the fields origins and focus in geospatial information visualisation, there have been a large number of suggested variations on the choropleth map that allow authors to include uncertainty.
+We will use these maps to provide simple examples for each approach that can be easily compared to the "no uncertainty" choropleth map to better understand the costs and benefits of each approach.
+Despite the example of each method focusing on variations of the choropleth map, it is important to understand that the approaches we are discussing are universal and are not unique to maps.
+
+# Ignoring uncertainty
+
+## Choropleth Map
+
+## Why include uncertainty?
+
+> Do we treat uncertainty as ‘just another variable’ to be visually represented or does it need to be treated differently?
+For example, when we picture a map showing air pressure distribution in combination with temperature, these two variables are certainly dependent on each other (in physical terms).
+The same is true with air pressure and its uncertainty, but we see a stronger dependency: Uncertainty can be seen as metadata of air- pressure which we argue makes a difference.
+Thus, we support Edwards and Nelson (2001, p.
+35) who stated that ‘[p]erhaps data certainty information is unique and will require a new type of framework for designing symboliza- tion’.
+Most studies that assess the usability of uncertainty visualisations do not contribute to this aspect since they test the retrieval of data and uncertainty separately, but from the perspective that there is nothing special about uncertainty.
+Traditionally, studies in this field focus on the ability to read both the map content and its uncertainty at the same time.
+This may be the mandatory criterion for a successful use of uncertainty, but the question that remains is whether this is sufficient to ensure that a user does not only have two separate values in mind but an integrated uncertain data value.
+
+# Visualising Uncertainty as a Variable
+## Bivariate Maps
+## Signal Supression
+## Uncertainty without signal supression
+
+# Explicitly Combining Uncertainty and Signal
+## Using Statistics
+### Exceedance Probability Map
+### What is an uncertainty visualisation?
+## Using Aesthetics
+### Value Supressing Uncertainty Palettes
+### There is no such thing as a test for everything
+
+# Implicitly Combining Uncertainty and Signal
+## Pixel maps
+## Just visualise your data
+
+# Evaluating uncertainty
+# Current methods
+# Testing signal supression
+
+# ------------------- BELOW HERE IS UNSORTED ------------------------
 # Why is uncertainty special?

### 2024-07-18 paper.qmd
wrote the ignoring uncertainty section

@@
 library(urbnmapr)
-
 ```
@@
 # Ignoring uncertainty
+Before we look at different cases where uncertainty is included, it is beneficial for us to look at cases where it is not. After all, why include uncertainty at all?
 
-## Choropleth Map
-
-## Why include uncertainty?
-
-> Do we treat uncertainty as ‘just another variable’ to be visually represented or does it need to be treated differently?
-For example, when we picture a map showing air pressure distribution in combination with temperature, these two variables are certainly dependent on each other (in physical terms).
-The same is true with air pressure and its uncertainty, but we see a stronger dependency: Uncertainty can be seen as metadata of air- pressure which we argue makes a difference.
-Thus, we support Edwards and Nelson (2001, p.
-35) who stated that ‘[p]erhaps data certainty information is unique and will require a new type of framework for designing symboliza- tion’.
-Most studies that assess the usability of uncertainty visualisations do not contribute to this aspect since they test the retrieval of data and uncertainty separately, but from the perspective that there is nothing special about uncertainty.
-Traditionally, studies in this field focus on the ability to read both the map content and its uncertainty at the same time.
-This may be the mandatory criterion for a successful use of uncertainty, but the question that remains is whether this is sufficient to ensure that a user does not only have two separate values in mind but an integrated uncertain data value.
-
-# Visualising Uncertainty as a Variable
-## Bivariate Maps
-## Signal Supression
-## Uncertainty without signal supression
-
-# Explicitly Combining Uncertainty and Signal
-## Using Statistics
-### Exceedance Probability Map
-### What is an uncertainty visualisation?
-## Using Aesthetics
-### Value Supressing Uncertainty Palettes
-### There is no such thing as a test for everything
-
-# Implicitly Combining Uncertainty and Signal
-## Pixel maps
-## Just visualise your data
-
-# Evaluating uncertainty
-# Current methods
-# Testing signal supression
-
-# ------------------- BELOW HERE IS UNSORTED ------------------------
-# Why is uncertainty special?
-It is rarely made clear exactly how uncertainty visualisation is different to normal information visualisation.
-@Hullman2016 commented that it is straightforward to show a value but it is much more complex to show uncertainty, but did not explain *why*.
-A visualisation is nothing more than computing a statistic, mapping that statistic to a visual feature and rendering that graphic.
-Nothing in this process suggests visualising an estimate of the variance should be any different to visualising an estimate of the mean, median, maximum, or of any other statistic, however this field *does* implicitly treat them different.
-There is no such thing as "maximum visualiation" or "median visualisation" so there *must* be something about uncertainty that causes authors to believe it is special.
-Here we seek to answer this question.
-
-### The purpose of uncertainty visualisations
-Uncertainty visualisation papers usually provide one of two primary justifications for their existence.
-The first justification is that uncertainty is fundamentally differently to other variables due to the psychological heuristics involved in interpreting uncertainty.
-Therefore uncertainty visualisation is different to normal visualisation as authors must consider the psychological effects of what they are visualising [@Spiegelhalter2017; @Hullman2019].
-The second justification assets that uncertainty is an additional variable and it is of vital importance to interpreting an estimate.
-Therefore uncertainty visualisation is a high dimensional visualisation problem, as we need to figure out how to seamlessly add uncertainty into already existing graphics [@moritz2017trust; @Griethe2006].
-
-Why some authors pick one motivation over another is rarely explained and both leave the role of uncertainty in the visualisation unclear.
-The literature never explains if uncertainty should be treated as a variable, as metadata, or as something else entirely [@Kinkeldey2014].
-
-Uncertainty, acting as it is intended, is transparency.
-There are many secondary benefits that come with this improved transparency, such as better decisions, more trust in the results and more confidence in the authors.
-These secondary benefits are, however, *not* the immediate goal of uncertainty.
-The following sections will discuss the issues and limitations in measuring uncertainty through these secondary metrics and provide suggestions as to how future studies should consider measuring uncertainty.
-
-Trust is a by product of displaying uncertainty rather than the goal of it, and viewing the relationship in the converse direction can lead to misguided research.
-Considering trust, and not transparency, as the metric of importance in uncertainty communication can lead to a questionable subtext that argues against transparency, something that has been noticed by several other authors [@Spiegelhalter2017; ONeill2018].
-@Hullman2020a found that author simultaneously argued that failing to visualise uncertainty was akin to fraud, but also many avoided uncertainty visualisation because they didn't want their work to come across as "untrustworthy".
-These authors are optimising *trust* rather than *transparency*, which means they opt to leave out uncertainty information when it does exactly what it is supposed to, decrease certainty in conclusions.
-
-Science communication should be primarily concerned with accuracy.
-Setting trust and risk-aversion as the variables of interest implicitly encourages statisticians to set trust and risk-aversion as the primary goals of communication.
-The issue of trust being divorced from trustworthiness has been commented on by other authors [@ONeill2018], however the issue still persists in the uncertainty visualisation literature.
-@Zhao2023 displayed a several variations of a visualisation of a model prediction and its uncertainty and took participants using the model prediction as a sign of trust.
-They reported that visualising uncertainty information caused participants to trust the model in the low variance case, but the results in the high variance case were inconclusive.
-The discussion made it clear the authors thought the uncertainty information should make the visualisation more trustworthy, but conflating trust and the use of a prediction implied uncertainty information should somehow influence participants to use their own prediction, even though a prediction being uncertainty does not necessarily mean it is incorrect.
-Despite this, the authors seemed to assume that the uncertainty information *should* have an influence on that, showing they had not deeply considered *how* uncertainty information should influence the choices of the participants.
-(*Cite Gap 18: examples of studies where authors measure trust*)
-
-A similar measure to trust is using "confidence" in an extracted value or a decision.
-Interestingly, "confidence" is also used to try and capture the clarity of a message in a normal visualisation.
-Confidence cannot simultaneously be a measure of clarity of visualisation *and* a way to capture the uncertainty expressed in a visualisation.
-
-## What is signal-supression?
-The true reason that uncertainty visualisation is differerent to normal information visualisation, is that normal information visualisation is designed to *find* signals in our data, while uncertainty visualisation should be designed to *supress* it.
-
-Interviews with experts in visualisation and statistics often back up "signal suppression" as the primary motivation for uncertainty visualisation.
-Failling to communicate uncertainty is often considered to be akin to fraud or lying [@Hullman2020a; @Manski2020] indicating that uncertainty should exist to prevent undue certainty in visual inference.
-The relationship between signal supression and uncertainty visualisation is made more explicity by other authors [@Griethe2006].
-Therefore uncertainty acts as a form of statistical "hedging" for signals found in an analysis.
-This aligns nicely with our understanding of uncertainty as a by product of inference.
-The variable we are trying to perform inference on is the signal, while the while the visualisation of noise that prevents the signal from being read is the uncertainty.
-It is nice that the connection to inference easily extends from uncertainty to uncertainty visualisation, however many of the confusions extend as well.
-
-Interpretation and semantics experiments are seeking to identify a dimension (or visual task) that uncertainty naturally maps to.
-These experiments inherently view uncertainty as a variable that is separate to the variable on which we have mapped our signal.
-For example, lets say we have a map were maximum daily temperature is presented using points where the colour (red for hot and blue for cold) of the point is associated with the temperature, and the blurriness of that point is associated with the variance *of* that temperature.
-It is highly likely that our brains will not flatten that into a single variable depicting noise and signal, but rather *separately* extract the temperature (colour) and uncertainty (blur) information as two independent variables.
-If the variables are extracted separately, there is no guarantee that the uncertainty will act as an appropriate signal suppressor.
-This problem has been noticed by others in the field, that typically use this method (specifically in the spatial uncertainty context) and a desire for representations that integrate uncertainty and signal is one of the reasons for the invention of the value-suppressing uncertainty pallet [@Correll2018].
-Value-suppressing uncertainty pallets (VSUP) were developed as a method that would allow the signal and the noise to be interpreted together such that insights gained by the viewers of a plot are appropriately *suppressed* by the uncertainty.
-Hence the name of the pallet.
-@fig-maps depicted this map colouring approach and several other extensions on the typical choropleth map that differ in where in the visualisation process they combine the noise and signal information into a single concept of "valid signal".
-The first and most basic map is a simple choropleth map where each value (the colouring of each local government area) has no associated "uncertainty".
-The next map (which embodies the approach taken by the semantics experiments) is the bivariate map which maps the signal to colour value and the uncertainty to colour hue.
-If visualisation was performing signal suppression then that would mean the two dimensional space defined by colour value and colour hue can be mentally "flattened" into a single dimension of "valid value" that can captures the signal the our brain would need to be able to flatted this two dimensional space into a single space of "signal validity".
-The idea of two perceptual tasks flattening into one variable in the mind of the viewer may be wishful thinking, but it is not impossible given we are not certain on how the perceptual tasks are mapped within the human brain.
-@Sterzik2023 found that when a value was mapped to the textures of stippling, hatching, and triangles, and found that the difference between two points on this one dimensional texture was actually a 2D space (likely "texture business" and "light/darkness").
-That being said, if we look at the visual signals presented by the bivariate map, where the contrasting light and dark areas actually has no important meaning, it is unlikely this occurs for colour value and hue (or at least it doesn't occur in a way that is useful for uncertainty visualisation).
-Instead of hoping that uncertainty might collapse signal values into a single dimension, we can do some of that work ourselves, by using a VSUP which collapses the colour space such that high uncertainty values cannot be extracted.
-It is unclear how useful VSUPs are at actually combining signal and noise and therefore suppressing plot level insights, as they have only been tested on simple value extraction tasks that require evaluating a single point [@Correll2018; @Ndlovu2023] rather than looking for spatial relationships (which is arguably what maps are for).
-Following along with this trend, the next way we might consider visualising uncertainty is to combine uncertainty and signal at the earlier stage so the "suppressed signal" is represented by a single variable.
-This statistic can then be expressed in a one dimensional colour space, which is a method adopted by the Bayesian surprise metric map [@Ndlovu2023] and the exceedance probability map [Lucchesi2017].
+@fig-choropleth depicts a choropleth map of the counties of Iowa.
+Each of these counties are colored according to an estimate of average daily temperature that was simulated so that the values followed a clear spatial trend (hot in the middle of the map, and cold on the outside).
+The variance of these estimates were simulated such that a hypothesis test would indicate the existence of a spatial trend in the low variance map, while the trend in the high variance map should be indistinguishable from noise.
+Is this distinction in validity of the spatial trend clear in in the map?
+Is the validity of the trend communicated through the visualisation?
 
@@
 #| warning: false
-#| label: fig-maps
-#| fig-cap: "A visualisation of a typical choropleth map, as well as three other maps that are used to display uncertainty on the choropleth map.
-There is a high variance and low variance example for each type of map to show how well the technique supresses signal.
-Each map was created using the same data with the same base palette.
-At first glance, the high uncertainty bivariate and VSUP maps just look like maps with a low saturation colour palette, rather than map with high uncertainty.
-This visualisation makes it clear that supression methods that plot uncertainty to a second axis, such as hue, make uncertainty appear as a second variable, rather than a signal supression on our estimate.
-"
-#| fig-subcap: 
-#|   - "Choropleth Map: Low Variance"
-#|   - "Choropleth Map: High Variance"
-#|   - "Choropleth Palette"
-#|   - "Bivariate Map: Low Variance"
-#|   - "Bivariate Map: High Variance"
-#|   - "Bivariate Palette"
-#|   - "VSUP Map: Low Variance"
-#|   - "VSUP Map: High Variance"
-#|   - "VSU Palette"
-#|   - "Pixel Map: Low Variance"
-#|   - "Pixel Map: High Variance"
-#|   - "Pixel Palette"
-#| layout-ncol: 3
+#| label: fig-choropleth
+#| fig-cap: "Two choropleth maps that depicts the counties of Iowa where each country coloured acording to a simulated average temperature.
+Both maps depict a spatial trend, where counties closer to the center of the map are hotter than counties on the edge of the map.
+This trend is only technically there in the low variance condition, i.e.
+if we performed a hypothesis test we would conclude the trend exists.
+The high variance condition displays a spatial trend that could simply be the result of noise, which means the plot is displaying a false conclusion."
 
@@
 p1b <- p1a %+% filter(my_map_data, variance_class=="highvar")
+
+p1a
+p1b
+show_col(basecols, ncol=8, labels=FALSE)
+```
+
+The two choropleth maps appearing to be identical in @fig-choropleth highlights the need for uncertainty visualisations.
+Uncertainty visualisation is is required for transparency and this sentiment has been repeated.
+Some authors suggest uncertainty is important to include as it communicates the legitimacy (or illegitimacy) of the conclusion drawn from visual inference [@Correll2014; @Kale2018; @Griethe2006].
+Some authors have said that uncertainty should be included to degree of confidence or trust in the data [@Boukhelifa2012; @Zhao2023].
+Some authors directly connect uncertainty visualisation to hypothesis testing as it ensures the "validity" of a statement [@Hullman2020a; @Griethe2006], but allows for a proportional level of trust that is more detailed than the binary results of a hypothesis test [@Correll2014; @Correll2018].
+Some authors even go so far as to claim that failing to include uncertainty is akin to fraud or lying [@Hullman2020a; @Manski2020].
+
+This consensus leads us to understand that uncertainty visualisation is motivated by the need for a sort of "visual hypothesis test".
+A successful uncertainty visualisation would act as a "statistical hedge" for any inference me make using the graphic.
+This implies three key criteria for a good uncertainty visualisation:
+
+  1) Reinforce justified conclusions to encourage confidence in results
+  2) Prevent unjustified conclusions that are indistinguishable from noise
+  3) Perform tasks 1) and 2) in a way that is proportional to the level of confidence in those conclusions.
+
+As @fig-choropleth showed, visualisations that are unconcerned with uncertainty have no issue showing justified signals, but struggle with the display of unjustified signals.
+Therefore, we call this approach "signal-supression" as it primarily differentiates itself from from a typical visualisation approach through criteria (2), that is, an uncertainty visualisation should prevent us from drawing unjustified conclusions.
+We will investigate several approaches for uncertainty visualisation in the following sections and in doing so, discuss how successful each approach is in fulfilling the signal-suppression criteria.
+
+# Visualising Uncertainty as a Variable
+Upon hearing that uncertainty needs to be included for transparency, the solutions may seem obvious.
+You may think "well, I will just add a dimension to my plot that includes uncertainty".
+This is not a silly choice, but lets look at what happens when we do this.
+
+## Bivariate Map
+## Signal Supression
+
+The true reason that uncertainty visualisation is differerent to normal information visualisation, is that normal information visualisation is designed to *find* signals in our data, while uncertainty visualisation should be designed to *supress* it.
+
+It is rarely made clear exactly how uncertainty visualisation is different to normal information visualisation.
+@Hullman2016 commented that it is straightforward to show a value but it is much more complex to show uncertainty, but did not explain *why*.
+A visualisation is nothing more than computing a statistic, mapping that statistic to a visual feature and rendering that graphic.
+Nothing in this process suggests visualising an estimate of the variance should be any different to visualising an estimate of the mean, median, maximum, or of any other statistic, however this field *does* implicitly treat them different.
+There is no such thing as "maximum visualiation" or "median visualisation" so there *must* be something about uncertainty that causes authors to believe it is special.
+Here we seek to answer this question.
+
+Uncertainty visualisation papers usually provide one of two primary justifications for their existence.
+The first justification is that uncertainty is fundamentally differently to other variables due to the psychological heuristics involved in interpreting uncertainty.
+Therefore uncertainty visualisation is different to normal visualisation as authors must consider the psychological effects of what they are visualising [@Spiegelhalter2017; @Hullman2019].
+The second justification assets that uncertainty is an additional variable and it is of vital importance to interpreting an estimate.
+Therefore uncertainty visualisation is a high dimensional visualisation problem, as we need to figure out how to seamlessly add uncertainty into already existing graphics [@moritz2017trust; @Griethe2006].
+
+Why some authors pick one motivation over another is rarely explained and both leave the role of uncertainty in the visualisation unclear.
+The literature never explains if uncertainty should be treated as a variable, as metadata, or as something else entirely [@Kinkeldey2014].
+
+> Do we treat uncertainty as ‘just another variable’ to be visually represented or does it need to be treated differently?
+For example, when we picture a map showing air pressure distribution in combination with temperature, these two variables are certainly dependent on each other (in physical terms).
+The same is true with air pressure and its uncertainty, but we see a stronger dependency: Uncertainty can be seen as metadata of air- pressure which we argue makes a difference.
+Thus, we support Edwards and Nelson (2001, p.
+35) who stated that ‘[p]erhaps data certainty information is unique and will require a new type of framework for designing symboliza- tion’.
+Most studies that assess the usability of uncertainty visualisations do not contribute to this aspect since they test the retrieval of data and uncertainty separately, but from the perspective that there is nothing special about uncertainty.
+Traditionally, studies in this field focus on the ability to read both the map content and its uncertainty at the same time.
+This may be the mandatory criterion for a successful use of uncertainty, but the question that remains is whether this is sufficient to ensure that a user does not only have two separate values in mind but an integrated uncertain data value.
+
+## Can we have uncertainty without signal supression?
+Some authors do not describe uncertainty as a means of transparency, but rather as a variable of importance in of itself [@Blenkinsop2000].
+
+It is not entirely true that uncertainty is only important for transparency, however it is not clear that the "uncertainty" provided in other cases is uncertainty at all.
+Some authors claim the purpose of uncertainty is not transparency, but rather important information for effective decision making [@Ibrekk1987; @uncertchap2022; @Hullman2016; @Cheong2016; @Boone2018; @Padilla2017].
+While this can be true, the uncertainty provided for this purpose is *not* the same as the uncertainty provided for transparency.
+It arguably may not be uncertainty at all.
+This is easiest to understand with an example.
+
+Let's say my weather prediction app only provides me with the predicted daily rainfall when I am trying to decide if I should bring an umbrella.
+While my decision will be improved with the inclusion of uncertainty information, that is not because of a special quality of uncertainty, but rather because the uncertainty information can be used to calculate the true statistic of interest, the probability that it will rain at all.
+If I use uncertainty to improve decision making, it is not *true* uncertainty, rather it is supplementary material that is required to correct for an insufficient statistic.
+I was provided with was not the statistic required to make the decison.
+
+# Explicitly Combining Uncertainty and Signal
+## Using Statistics
+### Exceedance Probability Map
+### What is an uncertainty visualisation?
+## Using Aesthetics
+### Value Supressing Uncertainty Palettes
+### There is no such thing as a test for everything
+
+# Implicitly Combining Uncertainty and Signal
+## Pixel maps
+## Just visualise your data
+
+# Evaluating uncertainty
+# Current methods
+# Testing signal supression
+## Current methods
+There are many secondary benefits that come with this improved transparency, such as better decisions, more trust in the results and more confidence in the authors.
+These secondary benefits are, however, *not* the immediate goal of uncertainty.
+The following sections will discuss the issues and limitations in measuring uncertainty through these secondary metrics and provide suggestions as to how future studies should consider measuring uncertainty.
+
+Trust is a by product of displaying uncertainty rather than the goal of it, and viewing the relationship in the converse direction can lead to misguided research.
+Considering trust, and not transparency, as the metric of importance in uncertainty communication can lead to a questionable subtext that argues against transparency, something that has been noticed by several other authors [@Spiegelhalter2017; ONeill2018].
+@Hullman2020a found that author simultaneously argued that failing to visualise uncertainty was akin to fraud, but also many avoided uncertainty visualisation because they didn't want their work to come across as "untrustworthy".
+These authors are optimising *trust* rather than *transparency*, which means they opt to leave out uncertainty information when it does exactly what it is supposed to, decrease certainty in conclusions.
+
+Science communication should be primarily concerned with accuracy.
+Setting trust and risk-aversion as the variables of interest implicitly encourages statisticians to set trust and risk-aversion as the primary goals of communication.
+The issue of trust being divorced from trustworthiness has been commented on by other authors [@ONeill2018], however the issue still persists in the uncertainty visualisation literature.
+@Zhao2023 displayed a several variations of a visualisation of a model prediction and its uncertainty and took participants using the model prediction as a sign of trust.
+They reported that visualising uncertainty information caused participants to trust the model in the low variance case, but the results in the high variance case were inconclusive.
+The discussion made it clear the authors thought the uncertainty information should make the visualisation more trustworthy, but conflating trust and the use of a prediction implied uncertainty information should somehow influence participants to use their own prediction, even though a prediction being uncertainty does not necessarily mean it is incorrect.
+Despite this, the authors seemed to assume that the uncertainty information *should* have an influence on that, showing they had not deeply considered *how* uncertainty information should influence the choices of the participants.
+(*Cite Gap 18: examples of studies where authors measure trust*)
+
+A similar measure to trust is using "confidence" in an extracted value or a decision.
+Interestingly, "confidence" is also used to try and capture the clarity of a message in a normal visualisation.
+Confidence cannot simultaneously be a measure of clarity of visualisation *and* a way to capture the uncertainty expressed in a visualisation.
+
+# ------------------- BELOW HERE IS UNSORTED ------------------------
+# Why is uncertainty special?
+
+Interpretation and semantics experiments are seeking to identify a dimension (or visual task) that uncertainty naturally maps to.
+These experiments inherently view uncertainty as a variable that is separate to the variable on which we have mapped our signal.
+For example, lets say we have a map were maximum daily temperature is presented using points where the colour (red for hot and blue for cold) of the point is associated with the temperature, and the blurriness of that point is associated with the variance *of* that temperature.
+It is highly likely that our brains will not flatten that into a single variable depicting noise and signal, but rather *separately* extract the temperature (colour) and uncertainty (blur) information as two independent variables.
+If the variables are extracted separately, there is no guarantee that the uncertainty will act as an appropriate signal suppressor.
+This problem has been noticed by others in the field, that typically use this method (specifically in the spatial uncertainty context) and a desire for representations that integrate uncertainty and signal is one of the reasons for the invention of the value-suppressing uncertainty pallet [@Correll2018].
+
+Value-suppressing uncertainty pallets (VSUP) were developed as a method that would allow the signal and the noise to be interpreted together such that insights gained by the viewers of a plot are appropriately *suppressed* by the uncertainty.
+Hence the name of the pallet.
+@fig-maps depicted this map colouring approach and several other extensions on the typical choropleth map that differ in where in the visualisation process they combine the noise and signal information into a single concept of "valid signal".
+The first and most basic map is a simple choropleth map where each value (the colouring of each local government area) has no associated "uncertainty".
+The next map (which embodies the approach taken by the semantics experiments) is the bivariate map which maps the signal to colour value and the uncertainty to colour hue.
+If visualisation was performing signal suppression then that would mean the two dimensional space defined by colour value and colour hue can be mentally "flattened" into a single dimension of "valid value" that can captures the signal the our brain would need to be able to flatted this two dimensional space into a single space of "signal validity".
+The idea of two perceptual tasks flattening into one variable in the mind of the viewer may be wishful thinking, but it is not impossible given we are not certain on how the perceptual tasks are mapped within the human brain.
+@Sterzik2023 found that when a value was mapped to the textures of stippling, hatching, and triangles, and found that the difference between two points on this one dimensional texture was actually a 2D space (likely "texture business" and "light/darkness").
+That being said, if we look at the visual signals presented by the bivariate map, where the contrasting light and dark areas actually has no important meaning, it is unlikely this occurs for colour value and hue (or at least it doesn't occur in a way that is useful for uncertainty visualisation).
+Instead of hoping that uncertainty might collapse signal values into a single dimension, we can do some of that work ourselves, by using a VSUP which collapses the colour space such that high uncertainty values cannot be extracted.
+It is unclear how useful VSUPs are at actually combining signal and noise and therefore suppressing plot level insights, as they have only been tested on simple value extraction tasks that require evaluating a single point [@Correll2018; @Ndlovu2023] rather than looking for spatial relationships (which is arguably what maps are for).
+Following along with this trend, the next way we might consider visualising uncertainty is to combine uncertainty and signal at the earlier stage so the "suppressed signal" is represented by a single variable.
+This statistic can then be expressed in a one dimensional colour space, which is a method adopted by the Bayesian surprise metric map [@Ndlovu2023] and the exceedance probability map [Lucchesi2017].
+
+```{r}
+#| echo: false
+#| message: false
+#| warning: false
+#| label: fig-maps
+#| fig-cap: "A visualisation of a typical choropleth map, as well as three other maps that are used to display uncertainty on the choropleth map.
+There is a high variance and low variance example for each type of map to show how well the technique supresses signal.
+Each map was created using the same data with the same base palette.
+At first glance, the high uncertainty bivariate and VSUP maps just look like maps with a low saturation colour palette, rather than map with high uncertainty.
+This visualisation makes it clear that supression methods that plot uncertainty to a second axis, such as hue, make uncertainty appear as a second variable, rather than a signal supression on our estimate.
+"
+
 # Bivariate Map
@@
 # Display plots and palettes
-p1a
-p1b
-show_col(basecols, ncol=8, labels=FALSE)
 p2a

### 2024-07-19 paper.qmd
fixed choropleth legend

@@
 format: pdf
+fig-valign: bottom
 editor_options: 
@@
 ```
+
+{{< pagebreak >}}
+
 # Background
@@
 
-Unlike an infographic, a statistical graphic is one of intent, and all enhancements made on a statistical graphic should be in pursuit of that goal (@Vanderplas2020).
+Visualization is an important step in exploratory data analysis and it is often utilised to **learn** what is important about a data set.
+Datasets such as Anscombe's quartet [@anscombe] or the Datasaurus Dozen [@datasaurpkg] highlight this power in visualisation.
+Additionally, visualisations allow for efficient and memorable communication.
+Even something as simple as sketching a distribution before recalling statistics or making predictions can greatly increase the accuracy of those measures [@Hullman2018; @Goldstein2014].
 
@@
 #|   - "Picture 9"
+#| layout-valign: bottom
 #| layout-ncol: 3
@@
 @Kinkeldey2014 found most experiments on the methods for uncertainty visualisation evaluation to be ad hoc, with no commonly agreed upon methodology or formalisation and no greater goal of describing general principals.
-Additionally, they noticed that the literature leaves@Hullman2016 commented on the difficulty in taking overarching themes from uncertainty visualisation, as several conflated issues make it unclear if subjects did poorly in an experiment because they misunderstood a visualisation, because the question was misinterpreted, or because they used a specific heuristic.
+@Hullman2016 commented on the difficulty in taking overarching themes from uncertainty visualisation, as several conflated issues make it unclear if subjects did poorly in an experiment because they misunderstood a visualisation, because the question was misinterpreted, or because they used a specific heuristic.
 @Spiegelhalter2017 commented that different plots are good for different things, and disagreed with the goal of identifying a universal "best" plot for all people and circumstances.
@@
 
+{{< pagebreak >}}
+
 # Ignoring uncertainty
@@
 The high variance condition displays a spatial trend that could simply be the result of noise, which means the plot is displaying a false conclusion."
+#| fig-subcap: 
+#|   - "High Variance Data"
+#|   - "Low Variance Data"
+#|   - ""
+#| layout-ncol: 3
+#| layout-valign: "bottom"
 
@@
 
+show_pal1 <- function (colours, borders = NULL, cex_label = 1, 
+    ncol = NULL) {
+  # Set dimensions of palette
+  n <- length(colours)
+  ncol <- ncol %||% ceiling(sqrt(length(colours)))
+  nrow <- ceiling(n/ncol)
+  # make matrix with null values (if not full)
+  colours <- c(colours, rep(NA, nrow * ncol - length(colours)))
+  colours <- matrix(colours, ncol = ncol, byrow = TRUE)
+  # set graphical parameters (?)
+  old <- par(pty = "s", mar = c(0, 0, 0, 0))
+  on.exit(par(old))
+  size <- max(dim(colours))
+  plot(c(0, size), c(0, -size), type = "n", xlab = "", ylab = "", 
+      axes = FALSE)
+  rect(col(colours) - 1, -row(colours) + 1, col(colours), -row(colours), 
+       col = colours, border = borders)
+  text(c(0,col(colours)), -c(1,row(colours))-0.15, 21:29, 
+              cex = 1, col = "black")
+  text(4, -1.55, "Temperature",cex = 1.5, col = "black")
+}
+
 p1a
 p1b
-show_col(basecols, ncol=8, labels=FALSE)
+show_pal1(basecols, ncol=8, borders=NA)
+
 ```
@@
 We will investigate several approaches for uncertainty visualisation in the following sections and in doing so, discuss how successful each approach is in fulfilling the signal-suppression criteria.
+
+{{< pagebreak >}}
 

### 2024-07-19 paper.qmd
tried to get bottom align to work (didnt) and added some decision bit

@@
 #|   - "Picture 9"
-#| layout-valign: bottom
 #| layout-ncol: 3
+#| fig-cap-location: "margin"
 
@@
 This review attempts to address this issue by offering a novel perspective on the uncertainty visualisation problem, and hopefully laying the foundations on which we can build a house.
-
 This review is broken into several parts that each reflect a different approach to uncertainty visualisation.
@@
 # Ignoring uncertainty
-Before we look at different cases where uncertainty is included, it is beneficial for us to look at cases where it is not. After all, why include uncertainty at all?
+A good place to start might be at deceptively straight forward question, why should we include uncertainty at all?
 
@@
 #| label: fig-choropleth
-#| fig-cap: "Two choropleth maps that depicts the counties of Iowa where each country coloured acording to a simulated average temperature.
+#| fig-cap: "Two choropleth maps that depict the counties of Iowa where each country coloured acording to a simulated average temperature.
 Both maps depict a spatial trend, where counties closer to the center of the map are hotter than counties on the edge of the map.
@@
 #|   - "Low Variance Data"
-#|   - ""
+#|   - "Choropleth Palette"
 #| layout-ncol: 3
@@
 
+## Signal-supression
 The two choropleth maps appearing to be identical in @fig-choropleth highlights the need for uncertainty visualisations.
@@
 
+We can refer to the we draw from a graphic to be its "signal" and the variance that makes this signal harder to identify as the "noise".
 As @fig-choropleth showed, visualisations that are unconcerned with uncertainty have no issue showing justified signals, but struggle with the display of unjustified signals.
-Therefore, we call this approach "signal-supression" as it primarily differentiates itself from from a typical visualisation approach through criteria (2), that is, an uncertainty visualisation should prevent us from drawing unjustified conclusions.
+Therefore, we coin this approach to uncertainty visualisation as "signal-supression" since it primarily differentiates itself from from the "noiseless" visualisation approach through criteria (2).
+That is, the main difference between an uncertainty visualisation and a "normal" visualisation, is that an uncertainty visualisation should prevent us from drawing unjustified conclusions.
+
+## Decision making does not need uncertainty visualisation
+Of course, uncertainty visualisation is not only motivated by signal-suppression.
+Some authors claim the purpose of uncertainty is not transparency, but rather important information for effective decision making [@Ibrekk1987; @uncertchap2022; @Hullman2016; @Cheong2016; @Boone2018; @Padilla2017].
+While uncertainty can provide useful information in decision making, that information is actually signal that is required due to incomplete information.
+This is easiest to understand with an example.
+
+Imagine you are trying to decide if you want to bring an umbrella with you to work.
+An umbrella is annoying to bring and you only want to bring it if there is a 10% chance of rain (or higher).
+Unfortunately, your weather prediction app only provides you with the predicted daily rainfall.
+Therefore, your decision will be improved with the inclusion of uncertainty, *not* because uncertainty is important for your decision, but because it gives you the tools required to calculate the *actual* statistic you are basing your decision on.
+
+While my decision will be improved with the inclusion of uncertainty information, that is not because of a special quality of uncertainty, but rather because the uncertainty information can be used to calculate the true statistic of interest, the probability that it will rain at all.
+If I use uncertainty to improve decision making, it is not *true* uncertainty, rather it is supplementary material that is required because the original information was not sufficient.
+In this sense uncertainty is no different from any other variable.
+If someone is asked to calculate a persons BMI and they are only given their height, of *course* their calculation will be improved if you provide their weight too.
+
+We can actually prove this result is true mathematically.
+Let $A$ be a rational agent that needs to make a decision and let $U(p)$ be the utility function of agent $A$ that depends on the random variable, $p$ where $p$ is a probability.
+Additionally $p$ can be calculated using a density function $p=f(\mu, \sigma)$
+
+The ma
+
+The best "uncertainty visualisation" in decision making experiments is reliably the visualisation that makes the calculation of
+
+For these reasons decision making as a motivation for uncertainty visualisation will not be the focus of this review, as the uncertainty in this case is not "noise" and does not require any special visualisation.
+
+### Exceedance Probability Map
+For these reasons, if the statistic required to make the decision is visualised, there is no reason to include uncertainty on *that* graphic.
+
+the uncertainty provided for this purpose is *not* the same as the uncertainty provided for transparency.
+
 We will investigate several approaches for uncertainty visualisation in the following sections and in doing so, discuss how successful each approach is in fulfilling the signal-suppression criteria.
@@
 ## Bivariate Map
-## Signal Supression
+## Uncertainty is not just another variable
+
+Some authors do not describe uncertainty as a means of transparency, but rather as a variable of importance in of itself [@Blenkinsop2000].
 
@@
 
-## Can we have uncertainty without signal supression?
-Some authors do not describe uncertainty as a means of transparency, but rather as a variable of importance in of itself [@Blenkinsop2000].
-
-It is not entirely true that uncertainty is only important for transparency, however it is not clear that the "uncertainty" provided in other cases is uncertainty at all.
-Some authors claim the purpose of uncertainty is not transparency, but rather important information for effective decision making [@Ibrekk1987; @uncertchap2022; @Hullman2016; @Cheong2016; @Boone2018; @Padilla2017].
-While this can be true, the uncertainty provided for this purpose is *not* the same as the uncertainty provided for transparency.
-It arguably may not be uncertainty at all.
-This is easiest to understand with an example.
-
-Let's say my weather prediction app only provides me with the predicted daily rainfall when I am trying to decide if I should bring an umbrella.
-While my decision will be improved with the inclusion of uncertainty information, that is not because of a special quality of uncertainty, but rather because the uncertainty information can be used to calculate the true statistic of interest, the probability that it will rain at all.
-If I use uncertainty to improve decision making, it is not *true* uncertainty, rather it is supplementary material that is required to correct for an insufficient statistic.
-I was provided with was not the statistic required to make the decison.
-
 # Explicitly Combining Uncertainty and Signal
 ## Using Statistics
-### Exceedance Probability Map
 ### What is an uncertainty visualisation?

### 2024-07-20 paper.qmd
wrote most of decision making section

@@
 - May not need to reference the same paper again and again
-- Expand background "why visualize" section
+- I can do a more formal proof of the decision making thing using decision science. I left it out but it can be added if that section needs to be more thorough.
 -->
@@
 
-show_pal1 <- function (colours, borders = NULL, cex_label = 1, 
-    ncol = NULL) {
+show_pal <- function (colours, borders = NULL, cex_label = 1, ncol = NULL, myxlab, breaks) {
   # Set dimensions of palette
@@
        col = colours, border = borders)
-  text(c(0,col(colours)), -c(1,row(colours))-0.15, 21:29, 
+  text(c(0,col(colours)), -c(1,row(colours))-0.15, breaks, 
               cex = 1, col = "black")
-  text(4, -1.55, "Temperature",cex = 1.5, col = "black")
+  text(4, -1.55, myxlab ,cex = 1.5, col = "black")
 }
@@
 p1b
-show_pal1(basecols, ncol=8, borders=NA)
+
+show_pal(basecols, ncol=8, borders=NA, myxlab = "Temperature", breaks = 21:29)
 
@@
 
-## Decision making does not need uncertainty visualisation
-Of course, uncertainty visualisation is not only motivated by signal-suppression.
-Some authors claim the purpose of uncertainty is not transparency, but rather important information for effective decision making [@Ibrekk1987; @uncertchap2022; @Hullman2016; @Cheong2016; @Boone2018; @Padilla2017].
-While uncertainty can provide useful information in decision making, that information is actually signal that is required due to incomplete information.
+# Visualising Uncertainty that is not uncertainty
+Uncertainty visualisation is not only motivated by signal-suppression, and we would be remiss to ignore these alternative motivations.
+Some authors do not describe uncertainty as a means of transparency, but rather as a variable of importance in of itself [@Blenkinsop2000].
+Other authors claim the purpose of uncertainty is to improve decision making [@Ibrekk1987; @uncertchap2022; @Hullman2016; @Cheong2016; @Boone2018; @Padilla2017].
+While uncertainty can provide useful as its own variable of interest, or in decision making, it is important to recognize the "uncertainty" in these cases are not acting as "uncertainty" at all.
+The "uncertainty" in these cases is acting as signal.
 This is easiest to understand with an example.
@@
 Imagine you are trying to decide if you want to bring an umbrella with you to work.
-An umbrella is annoying to bring and you only want to bring it if there is a 10% chance of rain (or higher).
+An umbrella is annoying to bring with you, so you only want to pack it if the chance of rain is greater than 10%.
 Unfortunately, your weather prediction app only provides you with the predicted daily rainfall.
 Therefore, your decision will be improved with the inclusion of uncertainty, *not* because uncertainty is important for your decision, but because it gives you the tools required to calculate the *actual* statistic you are basing your decision on.
-
-While my decision will be improved with the inclusion of uncertainty information, that is not because of a special quality of uncertainty, but rather because the uncertainty information can be used to calculate the true statistic of interest, the probability that it will rain at all.
-If I use uncertainty to improve decision making, it is not *true* uncertainty, rather it is supplementary material that is required because the original information was not sufficient.
-In this sense uncertainty is no different from any other variable.
-If someone is asked to calculate a persons BMI and they are only given their height, of *course* their calculation will be improved if you provide their weight too.
-
-We can actually prove this result is true mathematically.
-Let $A$ be a rational agent that needs to make a decision and let $U(p)$ be the utility function of agent $A$ that depends on the random variable, $p$ where $p$ is a probability.
-Additionally $p$ can be calculated using a density function $p=f(\mu, \sigma)$
-
-The ma
-
-The best "uncertainty visualisation" in decision making experiments is reliably the visualisation that makes the calculation of
-
-For these reasons decision making as a motivation for uncertainty visualisation will not be the focus of this review, as the uncertainty in this case is not "noise" and does not require any special visualisation.
-
-### Exceedance Probability Map
-For these reasons, if the statistic required to make the decision is visualised, there is no reason to include uncertainty on *that* graphic.
-
-the uncertainty provided for this purpose is *not* the same as the uncertainty provided for transparency.
-
-We will investigate several approaches for uncertainty visualisation in the following sections and in doing so, discuss how successful each approach is in fulfilling the signal-suppression criteria.
+In this sense, uncertainty is no more "special" to decision making than weight is in a BMI calculation.
+
+If this stipulation is true, then we would expect the visualisation methods and design principles of "normal" information visualisation to apply to these "uncertainty" visualisations.
+This is exactly what the literature shows.
+
+Experiments that evaluate graphics on how useful they are for value extraction and decision making rarely consider how difficult this "true" statistic is to extract from the chart.
+Graphics are compared on the basis of being "uncertainty visualisations" [@Ibrekk1987; @Hullman2015; @Hofman2020]  even though the bounds of that class is rife with disagreement.
+For example @Leland2005 mentions that popular graphics, such as pie charts and bar charts omit uncertainty, however at least one or both of these charts are used in a significant number of uncertainty visualisation experiments [@Ibrekk1987; @Olston2002; @Zhao2023; @Hofmann2012].
+@Wickham2011 suggests their product plot framework, which includes histograms, should have a way to measure uncertainty even though histograms are frequently discussed as an uncertainty visualisation.
+
+When we view these evaluation studies through the lens of "How do you *actually* extract the *statistic required to answer the question*?" the results often become obvious, or easily explained using existing work in information visualisation and statistics.
+@Ibrekk1987 found that participants were more accurate at extracting a statistic when it could be directly read off the graphic, than when it required an area estimate (which is the case if using the PDF), or when there was no visual indicator for the statistic at all (which is the case when using the CDF of an asymmetric function).
+@Hullman2015 found that a visualisation that allows viewers to estimate a probability through counting outperformed one that required a complicated area calculation.@Hofman2020 and @Zhang2022 found that participants were better at answering questions about a prediction intervals when shown a prediction interval instead of a sampling distribution.
+@Gschwandtnei2016 found that graphics where the required statistic could be directly read off the plot outperformed those that involved guesswork due to a gradually decreasing line.
+@Cheong2016 found that participants made better decisions when they were explicitly given the relevant probability in text rather than when they needed to read it off a map.
+The entire uncertainty visualisation literature is awash with examples like this.
+*****
+
+@fig-exceed depicts an exceedance probability map that has was designed as an alternative to the choropleth map to improve decision making under uncertainty.
+A keen viewer may notice that the "exceedance probability map" is actually just a choropleth map.
+While the input into the graphic has changed, graphics, just like statistics, are not defined by their input data.
+
+```{r}
+#| echo: false
+#| message: false
+#| warning: false
+#| label: fig-exceed
+#| fig-cap: "An exceedance probability map that depict the counties of Iowa where each country coloured acording to the probability that the average temperature exceeds 27.
+This map is a choropleth map where the variable of interest is a probability."
+#| fig-subcap: 
+#|   - "High Variance Data"
+#|   - "Low Variance Data"
+#|   - "Exceedance Probability Palette"
+#| layout-ncol: 3
+#| layout-valign: "bottom"
+
+# quantile
+prob_breaks <- seq(-0.1,1.1, length.out=9)
+exeed_data <- my_map_data |> 
+  as_tibble() |>
+  mutate(xprob = 1- pnorm(27, mean=temp, sd=sqrt(variance))) |>
+  mutate(xprob = cut(xprob, breaks=prob_breaks, labels=seq(8)))
+
+# Exceed Prob Map
+p2a <- exeed_data |>
+  filter(variance_class=="lowvar") %>%
+  ggplot() +
+  geom_sf(aes(fill = xprob, 
+              geometry = geometry)) + 
+  scale_fill_manual(values = basecols) +
+  theme_void() + 
+  theme(legend.position = "none")
+
+p2b <- p2a %+% filter(exeed_data, variance_class=="highvar")
+
+p2a
+p2b
+show_pal(basecols, ncol=8, borders=NA, myxlab = "Probability Temp>27", breaks = c(0, prob_breaks[2:8], 1))
+```
+
+There seems to be a divide between those that describe *any* visualisation of error as an "uncertainty visualisation" and those that believe "uncertainty visualisation" is a function that can be *applied* to any graphic that depicts an estimate.
+The former view uncertainty as signal (i.e.
+as a variable of interest on its own, or as required information for decision making) and the later view uncertainty as noise (i.e.
+as a variable needed to suppress the signal in the plot).
+
+Of course, there is nothing wrong with visualizing variance, error, bias, and any other statistic used to depict uncertainty as a signal.
+Just like any other statistic, these metrics provide important and useful information for analysis and decisions.
+However, there is no interesting visualisation challenge associated with these graphics, and they do not require any special visualisation, they are acting an estimate and they should be treated as such.
+For these reasons decision making and visualisations that explicitly depict uncertainty measures will not be the focus of this review.
+The following sections will investigate several approaches to uncertainty visualisation and discuss how successful each approach is in fulfilling the signal-suppression criteria.
 
@@
 ## Bivariate Map
-## Uncertainty is not just another variable
-
-Some authors do not describe uncertainty as a means of transparency, but rather as a variable of importance in of itself [@Blenkinsop2000].
+## What makes uncertainty visualisation difficult?
 
@@
 
+More accurate to class this as "why is uncertainty visualisation difficult"
 Uncertainty visualisation papers usually provide one of two primary justifications for their existence.
@@
 
-# Evaluating uncertainty
+# Evaluating uncertainty visualisations
 # Current methods
@@
 ## Current methods
+### Secondary effects
+
+- trust, confidence, etc
+
+### Direct questions
+@Hullman2019 found that the majority of papers evaluate visualisations on performance (approximately 65% of the papers they surveyed).
+
+By far the most common metric used is accuracy, which is used by approximately 36% of evaluation studies [@Hullman2019].
+
+The final method used by authors is to just explicitly ask about uncertainty and signal information separately.
+@Sanyal2009 mapped uncertainty to dots and signal to a 3D surface and asked participants to identify areas of high and low signal and high and low uncertainty.
+Participants were not asked to combine that information in any way, and the signal and the noise were treated as separate variables.
+@Correll2014 asked participants to separately extract the mean and variance from four uncertainty visualisations.
+These methods explicitly view the uncertainty and signal as two separate variables that should be extracted from a plot, and not two variables that should be interpreted together.
+Even viewing these questions as a routine check to make sure the signal information isn't impacted by the uncertainty is counter intuitive, because the whole point *of* the uncertainty is to impact the signal information.
+
+- definition of uncertainty
+
 There are many secondary benefits that come with this improved transparency, such as better decisions, more trust in the results and more confidence in the authors.
@@
 
-## Uncertainty visualisations vs information visualisation
-If uncertainty is inherently different, we would expect evaluation studies of uncertainty visualisations to produce different results to similar evaluation studies on "normal" variables.
-This does not seem to be the case.
-Uncertainty is noise, but most evaluation experiments measure it as signal.
-The previous section highlights a unique and fascinating problem faced by uncertainty visualisation.
-
-Even though uncertainty does not exist without a signal, that does not mean every visualisation that includes a concept related to uncertainty (such as a visualisation or error or variance) must be designed for signal supression.
-However, if a visualisation is not designed for signal supression, it should be treated the same as any other variable.
-These visualisation are not "special" in the sense that signal supression visualisations are.
-
-@Hullman2019 found that the majority of papers evaluate visualisations on performance (approximately 65% of the papers they surveyed) or interpretation and semantics (approximately 17%) and both of these evaluation goals will run into problems because of their conceptualisation of uncertainty.
-
-By far the most common metric used is accuracy, which is used by approximately 36% of evaluation studies [@Hullman2019].
-
-Visualisation authors are almost unanimous in commenting that the "information" in two plots must be the same in order for the visual techniques to be compared [@Cleveland1984; @Kinkeldey2014].
-@Kinkeldey2014 adopts an existing definition also that suggests two graphics are informationally equivalent if all the information in one plot is inferable from the other and vice-versa, but adds that two plots are computationally equivalent if that information can be extracted from both plots with similar easy and speed.
-Many visualisation authors ignore this concept and simply compare two visualisations because they both belong to the class of "uncertainty visualisations".
-@Ibrekk1987 compared a 6 visualisations of a PDF,  a box plot, a CDF and a mean with a confidence interval on this basis.
-They found that people are better at extracting the mean from a plot when they are shown a plot that contains a mean with a confidence interval than when they are shown a box plot, or other visualisations that did not allow for the mean to just be read off the plot.
-@Hullman2015 compared the static error bars and violin plots of the marginal distributions of two variables ($A$ and $B$) to an animated plot that depicted outcomes of the joint distribution of $A$ and $B$ in each frame and found that the visualisation of the joint distibution was better at answering questions about the joint distribution than the visualisations of the marginal distribution.
-@Hofman2020 commented that "theoretically" the sampling distribution of the mean and the prediction interval of a new observation are equal "so long as one knows the sample size", but does not seem to provide participants with that sample size, or recognise the assumptions and background knowledge that would be required to compare the two.
-@Hofman2020 and @Zhang2022 compared prediction and sampling distributions because they are both "uncertainty" that is typically depicted around the mean.
-They found that people are better at answering questions about a prediction interval when shown a prediction interval instead of a sampling distribution.
-
-This collection of examples starts to paint a pretty clear picture. Visualisations with rather shocking information asymmetry are regularly compared because they are both "uncertainty visualisations".
-
-If two graphics are visually different but identical in the information they contain, they must differ in how that information is depicted. 
-Once we have a mathematical expression of uncertainty, the visualisation of that uncertainty is theoretically identical to the visualisation process of any other variable.
-For simple tasks such as value extraction, there is a hierarchy to perceptual tasks where extracting visual information in some forms is easier than others.
-The hierarchy was originally established 40 years ago by @Cleveland1984, below is an updated version summarised by @Vanderplas2020:
-
-1) Position along a common scale. 
-2) Position along a non-aligned scale. 
-3) Length, direction, angle, slope
-4) Area
-5) Volume, density, curvature
-6) Shading, colour saturation, colour hue
-7) Discriminable shape
-8) Indiscriminable shape
-
-This hierarchy is a good general rule, however it can change from person to person [@Davis2022] Additionally, there are other graphical rules to consider such as gestalt principles, broader methods of extraction, and attention principles [@Vanderplas2020].
-These established visualisation concepts allow us to anticipate the ease with which certain pieces of information will be extracted from a plot.
-We can use these concepts to understand the computational complexity of a graphic.
-A bizarre feature of the uncertainty visualisation literature is that it does not work to build upon these existing principles or identify the ways in which uncertainty visualisations may diverge from these rules.
-These building block concepts of visualisation are seldom mentioned.
-
-It is difficult to find examples of uncertainty visualisation experiments where the plots do contain the same information, however when they do, the results align with existing information visualisation research.
-Technically, a PDF and a mean with confidence intervals both have enough information to extract the mean of the distribution, however they both have a very different computational cost.
-To extract the mean using a PDF, a participant would need to identify the point along the x axis that splits the area under the curve in half.
-If a participant is provided with a mean with a confidence interval, extracting the average is a simple task of reading the position on an aligned scale.
-@Ibrekk1987 found that when asking participants for the "best estimate" (which they thought should be interpreted as the mean of the distribution) of a skewed distribution, participants provided the mean when given a mean with confidence intervals and the mode when given a PDF [@Ibrekk1987].
-Similar results to this occur over and over again in the uncertainty visualisation literature.
-@Gschwandtnei2016 found that visualisations where the start time of an interval could literally be read off the plot (error bars, centred error bars, and ambiguation) performed better than the plots (accumulated probability, gradient, and violin) where the start time involved some guesswork because the drop off was gradual.
-@Cheong2016 found that participants were better at answering questions when they were explicitly given the relevant probability in text rather than when they needed to read it off a map.
-(*Cite Gap 15: Examples of replicated perceptual task experiments*).
-
-These results show that uncertainty is not technically different to any other variable.
-When trying to anticipate the results of these studies, we can use the same principles of information equivalence and difficulty of relevant mental tasks to understand which plots will outperform others.
-This does not mean that visualising uncertainty as a signal is incorrect or bad, it just means that uncertainty as a signal is no different from any other information visualisation.
-
-- decision making experiments are usually displaying the wrong signals
-
-The final method used by authors is to just explicitly ask about uncertainty and signal information separately.
-@Sanyal2009 mapped uncertainty to dots and signal to a 3D surface and asked participants to identify areas of high and low signal and high and low uncertainty.
-Participants were not asked to combine that information in any way, and the signal and the noise were treated as separate variables.
-@Correll2014 asked participants to separately extract the mean and variance from four uncertainty visualisations.
-These methods explicitly view the uncertainty and signal as two separate variables that should be extracted from a plot, and not two variables that should be interpreted together.
-Even viewing these questions as a routine check to make sure the signal information isn't impacted by the uncertainty is counter intuitive, because the whole point *of* the uncertainty is to impact the signal information.
-
 # What even is an uncertainty visualisation?
-## The literatures definition of uncertainty visualisations
-What exactly is an uncertainty visulisation is never explained and certainly does not have a consistent definition.
-For example @Leland2005 mentions that popular graphics, such as pie charts and bar charts omit uncertainty, however at least one or both of these charts are used in a significant number of uncertainty visualisation experiments [@Ibrekk1987; @Olston2002; @Zhao2023; @Hofmann2012].
-@Wickham2011 suggests their product plot framework, which includes histograms, should have a way to measure uncertainty, but does not consider that a histogram is *already* a depiction of PDF, something that is often considered an uncertainty visualisation by other authors.
 

### 2024-07-20 writingdumpster.qmd
wrote most of decision making section

@@
 # Removed Sections
+Visualisation authors are almost unanimous in commenting that the "information" in two plots must be the same in order for the visual techniques to be compared [@Cleveland1984; @Kinkeldey2014].
+@Kinkeldey2014 adopts an existing definition also that suggests two graphics are informationally equivalent if all the information in one plot is inferable from the other and vice-versa, but adds that two plots are computationally equivalent if that information can be extracted from both plots with similar easy and speed.
+
+ @Hofman2020 commented that "theoretically" the sampling distribution of the mean and the prediction interval of a new observation are equal "so long as one knows the sample size", but does not seem to provide participants with that sample size, or recognise the assumptions and background knowledge that would be required to compare the two.
+
+If two graphics are visually different but identical in the information they contain, they must differ in how that information is depicted. 
+Once we have a mathematical expression of uncertainty, the visualisation of that uncertainty is theoretically identical to the visualisation process of any other variable.
+For simple tasks such as value extraction, there is a hierarchy to perceptual tasks where extracting visual information in some forms is easier than others.
+The hierarchy was originally established 40 years ago by @Cleveland1984, below is an updated version summarised by @Vanderplas2020:
+
+1) Position along a common scale. 
+2) Position along a non-aligned scale. 
+3) Length, direction, angle, slope
+4) Area
+5) Volume, density, curvature
+6) Shading, colour saturation, colour hue
+7) Discriminable shape
+8) Indiscriminable shape
+
+This hierarchy is a good general rule, however it can change from person to person [@Davis2022] Additionally, there are other graphical rules to consider such as gestalt principles, broader methods of extraction, and attention principles [@Vanderplas2020].
+These established visualisation concepts allow us to anticipate the ease with which certain pieces of information will be extracted from a plot.
+We can use these concepts to understand the computational complexity of a graphic.
+A bizarre feature of the uncertainty visualisation literature is that it does not work to build upon these existing principles or identify the ways in which uncertainty visualisations may diverge from these rules.
+These building block concepts of visualisation are seldom mentioned.
+
+It is difficult to find examples of uncertainty visualisation experiments where the plots do contain the same information, however when they do, the results align with existing information visualisation research.
+Technically, a PDF and a mean with confidence intervals both have enough information to extract the mean of the distribution, however they both have a very different computational cost.
+To extract the mean using a PDF, a participant would need to identify the point along the x axis that splits the area under the curve in half.
+If a participant is provided with a mean with a confidence interval, extracting the average is a simple task of reading the position on an aligned scale.
+
+Similar results to this occur over and over again in the uncertainty visualisation literature.   (*Cite Gap 15: Examples of replicated perceptual task experiments*).
+
+These results show that uncertainty is not technically different to any other variable.
+When trying to anticipate the results of these studies, we can use the same principles of information equivalence and difficulty of relevant mental tasks to understand which plots will outperform others.
+This does not mean that visualising uncertainty as a signal is incorrect or bad, it just means that uncertainty as a signal is no different from any other information visualisation.
+
+- decision making experiments are usually displaying the wrong signals
+
 The exploration step of an analysis, which includes descriptive statistics, exploratory data visualisation, and unsupervised machine learning techniques, is performed without a prior hypothesis, however misunderstandings of this fact appears frequently in the literature.

### 2024-07-22 paper.qmd
finished the Visualising uncertainty that is not uncertainty section

@@
 
-# Visualising Uncertainty that is not uncertainty
+## Visualising uncertainty that is not uncertainty
 Uncertainty visualisation is not only motivated by signal-suppression, and we would be remiss to ignore these alternative motivations.
-Some authors do not describe uncertainty as a means of transparency, but rather as a variable of importance in of itself [@Blenkinsop2000].
-Other authors claim the purpose of uncertainty is to improve decision making [@Ibrekk1987; @uncertchap2022; @Hullman2016; @Cheong2016; @Boone2018; @Padilla2017].
-While uncertainty can provide useful as its own variable of interest, or in decision making, it is important to recognize the "uncertainty" in these cases are not acting as "uncertainty" at all.
-The "uncertainty" in these cases is acting as signal.
+Some authors claim the purpose of uncertainty is to improve decision making [@Ibrekk1987; @uncertchap2022; @Hullman2016; @Cheong2016; @Boone2018; @Padilla2017].
+While uncertainty can provide useful information in decision making, it is important to recognize the "uncertainty" in these cases is not acting as "uncertainty" at all.
+It is acting as signal.
 This is easiest to understand with an example.
@@
 
-If this stipulation is true, then we would expect the visualisation methods and design principles of "normal" information visualisation to apply to these "uncertainty" visualisations.
+Other authors do not describe uncertainty as important for decision making, but rather explicitly state it as variable of importance in of itself [@Blenkinsop2000].
+This group are *also* viewing uncertainty as a signal, albeit more explicitly than the decision making group.
+If the purpose of uncertainty in these cases is to convey information rather than obfuscate it, then there is no reason to assume these "uncertainty visualisations" would behave any differently to normal information visualisations.
 This is exactly what the literature shows.
 
-Experiments that evaluate graphics on how useful they are for value extraction and decision making rarely consider how difficult this "true" statistic is to extract from the chart.
-Graphics are compared on the basis of being "uncertainty visualisations" [@Ibrekk1987; @Hullman2015; @Hofman2020]  even though the bounds of that class is rife with disagreement.
-For example @Leland2005 mentions that popular graphics, such as pie charts and bar charts omit uncertainty, however at least one or both of these charts are used in a significant number of uncertainty visualisation experiments [@Ibrekk1987; @Olston2002; @Zhao2023; @Hofmann2012].
-@Wickham2011 suggests their product plot framework, which includes histograms, should have a way to measure uncertainty even though histograms are frequently discussed as an uncertainty visualisation.
-
-When we view these evaluation studies through the lens of "How do you *actually* extract the *statistic required to answer the question*?" the results often become obvious, or easily explained using existing work in information visualisation and statistics.
+Uncertainty visualisation experiments often compare graphics on the basis of being "uncertainty visualisations" [@Ibrekk1987; @Hullman2015; @Hofman2020] with little consideration for how difficult the information is to extract from the chat.
+Experiments where users are asked to make a decision based on a probability, and experiments where users need to extract a variance or some feature of the distribution both have the implicit belief that uncertainty visualisation is a "class" of visualisation, and a "good" uncertainty visualisation allows us to extract all information with ease.
+This is the logic that results in predictable results.
 @Ibrekk1987 found that participants were more accurate at extracting a statistic when it could be directly read off the graphic, than when it required an area estimate (which is the case if using the PDF), or when there was no visual indicator for the statistic at all (which is the case when using the CDF of an asymmetric function).
-@Hullman2015 found that a visualisation that allows viewers to estimate a probability through counting outperformed one that required a complicated area calculation.@Hofman2020 and @Zhang2022 found that participants were better at answering questions about a prediction intervals when shown a prediction interval instead of a sampling distribution.
+@Hullman2015 found that a visualisation that allows viewers to count outcomes to estimate a probability outperformed one that required a complicated area calculation.
+@Hofman2020 and @Zhang2022 found that participants were better at answering questions about a prediction intervals when shown a prediction interval instead of a sampling distribution.
 @Gschwandtnei2016 found that graphics where the required statistic could be directly read off the plot outperformed those that involved guesswork due to a gradually decreasing line.
 @Cheong2016 found that participants made better decisions when they were explicitly given the relevant probability in text rather than when they needed to read it off a map.
-The entire uncertainty visualisation literature is awash with examples like this.
-*****
-
-@fig-exceed depicts an exceedance probability map that has was designed as an alternative to the choropleth map to improve decision making under uncertainty.
+These experiments all have the same fundamental finding, if we create a problem by obfuscating information, then the representation that makes the obfuscated information easier to extract outperforms those that don't.
+This is why text representations or graphics that allow users to simply count a value to extract the required statistic outperform those that require any perceptual tasks.
+We have known perceptual tasks have more noise than text representations for a long time [@Cleveland1984] and it has little to do with uncertainty.
+
+Additionally, if uncertainty visualisation for decision making is a "wrong information" problem, then we would expect uncertainty visualisations made for this purpose to simply adjust the statistic they display of rather than a change the visual elements of the graphic.
+This is precicely what we observe.
+@fig-exceed depicts an exceedance probability map that has was designed as an alternative to the choropleth map to improve decision making under uncertainty [@Kuhnert2018; @Lucchesi2021].
 A keen viewer may notice that the "exceedance probability map" is actually just a choropleth map.
 While the input into the graphic has changed, graphics, just like statistics, are not defined by their input data.
+
+This seems to be an accepted fact for all visualisations *except* for uncertainty visualisation, and this creates disagreement in what is even considered to be an uncertainty visualisation.
+For example @Leland2005 mentions that popular graphics, such as pie charts and bar charts omit uncertainty, and @Wickham2011 suggests their product plot framework, which includes histograms and bar charts, should be extended to include uncertainty however at least one or both of these charts are used in a significant number of uncertainty visualisation experiments [@Ibrekk1987; @Olston2002; @Zhao2023; @Hofmann2012].
+This disagreement comes from two different conceptualisations of what an uncertainty visualisation *is*.
+There are those that describe *any* visualisation of error as an "uncertainty visualisation" and those that believe "uncertainty visualisation" is a function that can be *applied* to any graphic that depicts an estimate.
+The former group view the purpose of uncertainty visualisation to provide signal about a distribution, while the later believe it should act as noise to obfuscate a signal.
 
@@
 
-There seems to be a divide between those that describe *any* visualisation of error as an "uncertainty visualisation" and those that believe "uncertainty visualisation" is a function that can be *applied* to any graphic that depicts an estimate.
-The former view uncertainty as signal (i.e.
-as a variable of interest on its own, or as required information for decision making) and the later view uncertainty as noise (i.e.
-as a variable needed to suppress the signal in the plot).
-
-Of course, there is nothing wrong with visualizing variance, error, bias, and any other statistic used to depict uncertainty as a signal.
+The easily explained results in evaluation experiments, and the exceedance probabiltiy map show the problem of viewing uncertainty as a signal instead of noise.
+If we view the purpose of viewing uncertainty visualisation to simply extract a probability or variance (reguardless of whether or not it is for value extraction or decision making) then the best performing visualsiation will be one that makes that statistic easy to extract.
+Even better is to avoid displaying a visualisation all together and simply provide the probability or variance with text.
+
+Of course, we do not beleive there is anything wrong with explicitly visualizing variance, error, bias, or any other statistic used to depict uncertainty as a signal.
 Just like any other statistic, these metrics provide important and useful information for analysis and decisions.
 However, there is no interesting visualisation challenge associated with these graphics, and they do not require any special visualisation, they are acting an estimate and they should be treated as such.
+
 For these reasons decision making and visualisations that explicitly depict uncertainty measures will not be the focus of this review.
+We will instead focus on the perspective that uncertainty visualisation serves to obfuscate signal, and should be considered a function that can be *applied* to any graphic to suppress false signals.
 The following sections will investigate several approaches to uncertainty visualisation and discuss how successful each approach is in fulfilling the signal-suppression criteria.

### 2024-07-22 paper.qmd
changed section title

@@
 
-## Visualising uncertainty that is not uncertainty
+# Visualising uncertainty as signal
 Uncertainty visualisation is not only motivated by signal-suppression, and we would be remiss to ignore these alternative motivations.

### 2024-07-22 paper.qmd
rewrote a lot of "uncertainty as a variable stuff"

@@
 fig-valign: bottom
+cap-location: bottom
 editor_options: 
@@
 #| layout-ncol: 3
-#| fig-cap-location: "margin"
 
@@
 #| layout-valign: "bottom"
+#| cap-location: "bottom"
 
@@
 
+{{< pagebreak >}}
+
 # Visualising uncertainty as signal
@@
 For example @Leland2005 mentions that popular graphics, such as pie charts and bar charts omit uncertainty, and @Wickham2011 suggests their product plot framework, which includes histograms and bar charts, should be extended to include uncertainty however at least one or both of these charts are used in a significant number of uncertainty visualisation experiments [@Ibrekk1987; @Olston2002; @Zhao2023; @Hofmann2012].
-This disagreement comes from two different conceptualisations of what an uncertainty visualisation *is*.
+We belive this disagreement comes from two different conceptualisations of what an uncertainty  visualisation *is*.
+This issue causing confusion in the literature was also commented on by @Kinkeldey2014, who explained it was unclear if uncertainty should be treated as a variable, as metadata, or as something else entirely [@Kinkeldey2014].
 There are those that describe *any* visualisation of error as an "uncertainty visualisation" and those that believe "uncertainty visualisation" is a function that can be *applied* to any graphic that depicts an estimate.
@@
 
-# Visualising Uncertainty as a Variable
+# Including uncertainty as an indepdendent variable
 Upon hearing that uncertainty needs to be included for transparency, the solutions may seem obvious.
 You may think "well, I will just add a dimension to my plot that includes uncertainty".
-This is not a silly choice, but lets look at what happens when we do this.
-
-## Bivariate Map
+This makes sense, as this is the simplest way to add uncertainty to an existing graphic is to simply map uncertainty to an unused visual channel.
+
+@fig-bivariate a variation of the choropleth map, where we have a two dimensional colour palette.
+Not only is temperature mapped to hue, but variance is also mapped to saturation.
+While these two maps *do* look visually different (which was not the case in the choropleth map) the spatial trend is still clearly visible in both graphics.
+This means the uncertainty *is* being communicated, however the primary take away in the graphic is the spatial trend (that does not exist).
+The graphic no longer has a transparency issue as the uncertainty *is* technically included, but is this a uncertainty visualisation?
+And if not, why?
+
+```{r}
+#| echo: false
+#| message: false
+#| warning: false
+#| label: fig-bivariate
+#| fig-cap: "A bivariate map that depict the counties of Iowa where each county is coloured acording to it's average daily temperature and the variance in temperature.
+This map is a choropleth map with a two dimensional colour palette where temperature is represented by colour hue, and variance is represented by colour saturation.
+Even though uncertainty has been added to the graphic the spatial trend is still clearly visible in the case where the spatial trend could be attributed to noise."
+#| fig-subcap: 
+#|   - "High Variance Data"
+#|   - "Low Variance Data"
+#|   - "Bivariate Palette"
+#| layout-ncol: 3
+#| layout-valign: "bottom"
+
+# Bivariate Map
+# Make bivariate palette
+# Function to devalue by a certain amount
+colsupress <- function(basecols, hue=1, sat=1, val=1) {
+    X <- diag(c(hue, sat, val)) %*% rgb2hsv(col2rgb(basecols))
+    hsv(pmin(X[1,], 1), pmin(X[2,], 1), pmin(X[3,], 1))
+}
+
+# recurvisely decrease value
+v_val = 0.5
+bivariatepal <- c(basecols,
+                  colsupress(basecols, sat=v_val),
+                  colsupress(colsupress(basecols, sat=v_val), sat=v_val),
+                  colsupress(colsupress(colsupress(basecols, sat=v_val), sat=v_val), sat=v_val))
+# establish levels of palette
+names(bivariatepal) <- paste(rep(1:8, 4), "-" , rep(1:4, each=8), sep="")
+
+# Bivariate maps
+p2a <- my_map_data |>
+  filter(variance_class=="lowvar") %>%
+  ggplot() +
+  geom_sf(aes(fill = biclass, geometry = geometry)) + 
+  scale_fill_manual(values = bivariatepal) +
+  theme_void() + 
+  theme(legend.position = "none")
+
+p2b <- p2a %+% filter(my_map_data, variance_class=="highvar")
+
+show_pal2 <- function (colours, borders = NULL, cex_label = 1, ncol = NULL, myxlab, myylab, breaks, breaks2) {
+  # Set dimensions of palette
+  n <- length(colours)
+  ncol <- ncol %||% ceiling(sqrt(length(colours)))
+  nrow <- ceiling(n/ncol)
+  # make matrix with null values (if not full)
+  colours <- c(colours, rep(NA, nrow * ncol - length(colours)))
+  colours <- matrix(colours, ncol = ncol, byrow = TRUE)
+  # set graphical parameters (?)
+  old <- par(pty = "s", mar = c(0, 0, 0, 0))
+  on.exit(par(old))
+  size <- max(dim(colours))
+  plot(c(-1.5, size), c(0, -size), type = "n", xlab = "", ylab = "", 
+      axes = FALSE)
+  rect(col(colours) - 1, -row(colours) + 1, col(colours), -row(colours), 
+       col = colours, border = borders)
+  text(c(0,col(colours)[nrow,]) + c(0.2, 0.1, 0,0,0,0,0,-0.1,-0.2) , -4.5, 
+       breaks, cex = 1, col = "black")
+  text(-0.25, -c(0,row(colours)[,ncol]) + c(-0.2, -0.1, 0, 0.1, 0.2),
+       breaks2, cex = 1, col = "black")
+  text(4, -5.5, myxlab ,cex = 1.5, col = "black")
+  text(x=-1.25,y=-2, myylab, srt=270, cex = 1.5, col = "black")
+}
+
+p2a
+p2b
+show_pal2(colours = bivariatepal, ncol=8, borders=NA, myxlab = "Temperature", myylab = "Variance", breaks = 21:29, breaks2 = 0:4)
+```
+
 ## What makes uncertainty visualisation difficult?
-
-The true reason that uncertainty visualisation is differerent to normal information visualisation, is that normal information visualisation is designed to *find* signals in our data, while uncertainty visualisation should be designed to *supress* it.
-
-It is rarely made clear exactly how uncertainty visualisation is different to normal information visualisation.
-@Hullman2016 commented that it is straightforward to show a value but it is much more complex to show uncertainty, but did not explain *why*.
-A visualisation is nothing more than computing a statistic, mapping that statistic to a visual feature and rendering that graphic.
-Nothing in this process suggests visualising an estimate of the variance should be any different to visualising an estimate of the mean, median, maximum, or of any other statistic, however this field *does* implicitly treat them different.
-There is no such thing as "maximum visualiation" or "median visualisation" so there *must* be something about uncertainty that causes authors to believe it is special.
-Here we seek to answer this question.
-
-More accurate to class this as "why is uncertainty visualisation difficult"
-Uncertainty visualisation papers usually provide one of two primary justifications for their existence.
-The first justification is that uncertainty is fundamentally differently to other variables due to the psychological heuristics involved in interpreting uncertainty.
-Therefore uncertainty visualisation is different to normal visualisation as authors must consider the psychological effects of what they are visualising [@Spiegelhalter2017; @Hullman2019].
-The second justification assets that uncertainty is an additional variable and it is of vital importance to interpreting an estimate.
-Therefore uncertainty visualisation is a high dimensional visualisation problem, as we need to figure out how to seamlessly add uncertainty into already existing graphics [@moritz2017trust; @Griethe2006].
-
-Why some authors pick one motivation over another is rarely explained and both leave the role of uncertainty in the visualisation unclear.
-The literature never explains if uncertainty should be treated as a variable, as metadata, or as something else entirely [@Kinkeldey2014].
-
-> Do we treat uncertainty as ‘just another variable’ to be visually represented or does it need to be treated differently?
-For example, when we picture a map showing air pressure distribution in combination with temperature, these two variables are certainly dependent on each other (in physical terms).
-The same is true with air pressure and its uncertainty, but we see a stronger dependency: Uncertainty can be seen as metadata of air- pressure which we argue makes a difference.
-Thus, we support Edwards and Nelson (2001, p.
-35) who stated that ‘[p]erhaps data certainty information is unique and will require a new type of framework for designing symboliza- tion’.
-Most studies that assess the usability of uncertainty visualisations do not contribute to this aspect since they test the retrieval of data and uncertainty separately, but from the perspective that there is nothing special about uncertainty.
-Traditionally, studies in this field focus on the ability to read both the map content and its uncertainty at the same time.
-This may be the mandatory criterion for a successful use of uncertainty, but the question that remains is whether this is sufficient to ensure that a user does not only have two separate values in mind but an integrated uncertain data value.
-
-# Explicitly Combining Uncertainty and Signal
-## Using Statistics
-### What is an uncertainty visualisation?
-## Using Aesthetics
+Uncertainty visualisation is a uniquely difficult problem in visualisation.
+@Hullman2016 commented that it is straightforward to show a value but it is much more complex to show uncertainty.
+There are two primary theories as to *why* uncertainty visualisation is unique.
+
+The first group believes uncertainty visualisation are unique due to the psychological heuristics involved in interpreting uncertainty [@Spiegelhalter2017; @Hullman2019].
+While it is true that heuristics lead people to avoid uncertainty [@Spiegelhalter2017] there is no evidence that this psychological effect translates to a problem in visualisation.
+Were this the case, visualisations of death, poverty, and any other variable that elicits poor feelings would require its own field of visualisation.
+
+The second group sees uncertainty as a high dimensional visualisation problem as the difficulty is in working out how to add uncertainty into already existing graphics [@moritz2017trust; @Griethe2006].
+If we view uncertainty visualisation as a high dimension problem we include uncertainty as we would any other variable, and allow for the signal to be interpreted independent of the variance.
+This prevents uncertainty and signal being read together as a "single integrated uncertain value" [@Kinkeldey2014] and allows for uncertainty to be ignored [@uncertchap2022].
+Additioanlly, this integration needs to be communicated through the "gist" of the graphic, as that is primary purpose of a visualisation [@Spiegelhalter2017].
+The @fig-bivariate plot's "gist" is "this data has a spatial trend and the estimates have a large variance". It is clear the uncertainty and signal are being communicated separately.
+
+Uncertainty visualisation is difficult because we need the uncertainty axis to interfere with the signal axis.
+When we typically visualise two independent variables we *do not* want them to interfere with each other, however the opposite is true for uncertainty visualisation.
+For uncertainty to perform signal suppression, it must interfere with the visual channel that signal was mapped to.
+It needs to interfere enough that viewers do not see signals that are simply the product of noise.
+
+Lets say we have a map were maximum daily temperature is presented using points where the colour (red for hot and blue for cold) of the point is associated with the temperature, and the blurriness of that point is associated with the variance *of* that temperature.
+It is highly likely that our brains will not flatten that into a single variable depicting noise and signal, but rather *separately* extract the temperature (colour) and uncertainty (blur) information as two independent variables.
+If the variables are extracted separately, there is no guarantee that the uncertainty will act as an appropriate signal suppressor.
+
+ *** up to here ***
+
+The next map (which embodies the approach taken by the semantics experiments) is the bivariate map which maps the signal to colour value and the uncertainty to colour hue.
+If visualisation was performing signal suppression then that would mean the two dimensional space defined by colour value and colour hue can be mentally "flattened" into a single dimension of "valid value" that can captures the signal the our brain would need to be able to flatted this two dimensional space into a single space of "signal validity".
+
+The idea of two perceptual tasks flattening into one variable in the mind of the viewer may be wishful thinking, but it is not impossible given we are not certain on how the perceptual tasks are mapped within the human brain.
+@Sterzik2023 found that when a value was mapped to the textures of stippling, hatching, and triangles, and found that the difference between two points on this one dimensional texture was actually a 2D space (likely "texture business" and "light/darkness").
+That being said, if we look at the visual signals presented by the bivariate map, where the contrasting light and dark areas actually has no important meaning, it is unlikely this occurs for colour value and hue (or at least it doesn't occur in a way that is useful for uncertainty visualisation).
+
+- seperability experiments (susan, animation, scatter size)
+- suprise map (I think they had something about comparing light and dark)
+- Something else on accessibility or colour space (check susan's work)
+
+Instead of hoping that uncertainty might collapse signal values into a single dimension, we can do some of that work ourselves, by using a VSUP which collapses the colour space such that high uncertainty values cannot be extracted.
+It is unclear how useful VSUPs are at actually combining signal and noise and therefore suppressing plot level insights, as they have only been tested on simple value extraction tasks that require evaluating a single point [@Correll2018; @Ndlovu2023] rather than looking for spatial relationships (which is arguably what maps are for).
+Following along with this trend, the next way we might consider visualising uncertainty is to combine uncertainty and signal at the earlier stage so the "suppressed signal" is represented by a single variable.
+This statistic can then be expressed in a one dimensional colour space, which is a method adopted by the Bayesian surprise metric map [@Ndlovu2023] and the exceedance probability map [Lucchesi2017].
+
+These maps make the importance of combining uncertainty and signal in a single visual channel clear.
+A choropleth map will show signal that is not valid inference because of high uncertainty.
+At the other end of the spectrum, the bivariate map will show signal that is not always interesting because it forces us to interpret uncertainty and signal separately.
+As we move through these methods, it seems that the validity of any overarching insight becomes more visible at the cost of our ability to extract particular values of signal or noise.
+Therefore, given that the primary goal of visualisation *is* insights [@North2006], visualisation authors should err on the side of representing suppressed signal as a single variable, rather than visualising uncertainty separately using two different channels.
+
+{{< pagebreak >}}
+
+# Visualising uncertainty and signal in a new space
+This problem has been noticed by others in the field, that typically use this method (specifically in the spatial uncertainty context) and a desire for representations that integrate uncertainty and signal is one of the reasons for the invention of the value-suppressing uncertainty pallet [@Correll2018].
+
+Value-suppressing uncertainty pallets (VSUP) were developed as a method that would allow the signal and the noise to be interpreted together such that insights gained by the viewers of a plot are appropriately *suppressed* by the uncertainty.
+Hence the name of the pallet.
+depicted this map colouring approach and several other extensions on the typical choropleth map that differ in where in the visualisation process they combine the noise and signal information into a single concept of "valid signal".
+
 ### Value Supressing Uncertainty Palettes
@@
 
+{{< pagebreak >}}
+
 # Implicitly Combining Uncertainty and Signal
 ## Pixel maps
-## Just visualise your data
+## Show me the data
+
+{{< pagebreak >}}
 
@@
 ### Direct questions
+
+> Most studies that assess the usability of uncertainty visualisations do not contribute to this aspect since they test the retrieval of data and uncertainty separately, but from the perspective that there is nothing special about uncertainty.
+Traditionally, studies in this field focus on the ability to read both the map content and its uncertainty at the same time.
+This may be the mandatory criterion for a successful use of uncertainty, but the question that remains is whether this is sufficient to ensure that a user does not only have two separate values in mind but an integrated uncertain data value.
+[@Kinkeldey2014]
+
 @Hullman2019 found that the majority of papers evaluate visualisations on performance (approximately 65% of the papers they surveyed).
@@
 
+{{< pagebreak >}}
+
+# Bibliography
+
+```{r, include=FALSE, eval=FALSE}
+library(spelling)
+qmd <- "paper.qmd"
+ignore <- readLines("WORDLIST")
+check_spelling <- spell_check_files(
+  qmd,
+  ignore = ignore,
+  lang = "en_GB"
+)
+if (nrow(check_spelling) > 0) {
+  print(check_spelling)
+  stop("Check spelling in Qmd files!")
+}
+```
+
 # ------------------- BELOW HERE IS UNSORTED ------------------------
-# Why is uncertainty special?
-
-Interpretation and semantics experiments are seeking to identify a dimension (or visual task) that uncertainty naturally maps to.
-These experiments inherently view uncertainty as a variable that is separate to the variable on which we have mapped our signal.
-For example, lets say we have a map were maximum daily temperature is presented using points where the colour (red for hot and blue for cold) of the point is associated with the temperature, and the blurriness of that point is associated with the variance *of* that temperature.
-It is highly likely that our brains will not flatten that into a single variable depicting noise and signal, but rather *separately* extract the temperature (colour) and uncertainty (blur) information as two independent variables.
-If the variables are extracted separately, there is no guarantee that the uncertainty will act as an appropriate signal suppressor.
-This problem has been noticed by others in the field, that typically use this method (specifically in the spatial uncertainty context) and a desire for representations that integrate uncertainty and signal is one of the reasons for the invention of the value-suppressing uncertainty pallet [@Correll2018].
-
-Value-suppressing uncertainty pallets (VSUP) were developed as a method that would allow the signal and the noise to be interpreted together such that insights gained by the viewers of a plot are appropriately *suppressed* by the uncertainty.
-Hence the name of the pallet.
-@fig-maps depicted this map colouring approach and several other extensions on the typical choropleth map that differ in where in the visualisation process they combine the noise and signal information into a single concept of "valid signal".
-The first and most basic map is a simple choropleth map where each value (the colouring of each local government area) has no associated "uncertainty".
-The next map (which embodies the approach taken by the semantics experiments) is the bivariate map which maps the signal to colour value and the uncertainty to colour hue.
-If visualisation was performing signal suppression then that would mean the two dimensional space defined by colour value and colour hue can be mentally "flattened" into a single dimension of "valid value" that can captures the signal the our brain would need to be able to flatted this two dimensional space into a single space of "signal validity".
-The idea of two perceptual tasks flattening into one variable in the mind of the viewer may be wishful thinking, but it is not impossible given we are not certain on how the perceptual tasks are mapped within the human brain.
-@Sterzik2023 found that when a value was mapped to the textures of stippling, hatching, and triangles, and found that the difference between two points on this one dimensional texture was actually a 2D space (likely "texture business" and "light/darkness").
-That being said, if we look at the visual signals presented by the bivariate map, where the contrasting light and dark areas actually has no important meaning, it is unlikely this occurs for colour value and hue (or at least it doesn't occur in a way that is useful for uncertainty visualisation).
-Instead of hoping that uncertainty might collapse signal values into a single dimension, we can do some of that work ourselves, by using a VSUP which collapses the colour space such that high uncertainty values cannot be extracted.
-It is unclear how useful VSUPs are at actually combining signal and noise and therefore suppressing plot level insights, as they have only been tested on simple value extraction tasks that require evaluating a single point [@Correll2018; @Ndlovu2023] rather than looking for spatial relationships (which is arguably what maps are for).
-Following along with this trend, the next way we might consider visualising uncertainty is to combine uncertainty and signal at the earlier stage so the "suppressed signal" is represented by a single variable.
-This statistic can then be expressed in a one dimensional colour space, which is a method adopted by the Bayesian surprise metric map [@Ndlovu2023] and the exceedance probability map [Lucchesi2017].
+
+# MAPS
 
@@
 #| warning: false
-#| label: fig-maps
-#| fig-cap: "A visualisation of a typical choropleth map, as well as three other maps that are used to display uncertainty on the choropleth map.
-There is a high variance and low variance example for each type of map to show how well the technique supresses signal.
-Each map was created using the same data with the same base palette.
-At first glance, the high uncertainty bivariate and VSUP maps just look like maps with a low saturation colour palette, rather than map with high uncertainty.
-This visualisation makes it clear that supression methods that plot uncertainty to a second axis, such as hue, make uncertainty appear as a second variable, rather than a signal supression on our estimate.
-"
-
-# Bivariate Map
-# Make bivariate palette
-# Function to devalue by a certain amount
-colsupress <- function(basecols, hue=1, sat=1, val=1) {
-    X <- diag(c(hue, sat, val)) %*% rgb2hsv(col2rgb(basecols))
-    hsv(pmin(X[1,], 1), pmin(X[2,], 1), pmin(X[3,], 1))
-}
-
-# recurvisely decrease value
-v_val = 0.5
-bivariatepal <- c(basecols,
-                  colsupress(basecols, sat=v_val),
-                  colsupress(colsupress(basecols, sat=v_val), sat=v_val),
-                  colsupress(colsupress(colsupress(basecols, sat=v_val), sat=v_val), sat=v_val))
-# establish levels of palette
-names(bivariatepal) <- paste(rep(1:8, 4), "-" , rep(1:4, each=8), sep="")
-
-# Bivariate maps
-p2a <- my_map_data |>
-  filter(variance_class=="lowvar") %>%
-  ggplot() +
-  geom_sf(aes(fill = biclass, geometry = geometry)) + 
-  scale_fill_manual(values = bivariatepal) +
-  theme_void() + 
-  theme(legend.position = "none")
-
-p2b <- p2a %+% filter(my_map_data, variance_class=="highvar")
+#| eval: false
 
@@
 # Display plots and palettes
-p2a
-p2b
-show_col(bivariatepal, ncol=8, labels=FALSE)
 p3a
@@
 
-These maps make the importance of combining uncertainty and signal in a single visual channel clear.
-A choropleth map will show signal that is not valid inference because of high uncertainty.
-At the other end of the spectrum, the bivariate map will show signal that is not always interesting because it forces us to interpret uncertainty and signal separately.
-As we move through these methods, it seems that the validity of any overarching insight becomes more visible at the cost of our ability to extract particular values of signal or noise.
-Therefore, given that the primary goal of visualisation *is* insights [@North2006], visualisation authors should err on the side of representing suppressed signal as a single variable, rather than visualising uncertainty separately using two different channels.
-
-# What even is an uncertainty visualisation?
-
-### What is an uncertainty visualisation?
-@Kinkeldey2014 categorised uncertainty according to five criteria depicted in @fig-tax which considers if a visualisation is implicit or explicit, intrinsic or extrinsic, visually integrable or seperable, coincident or adjacent, and static or dynamic.
+# What is an uncertainty visualisation?
+@Kinkeldey2014 categorised uncertainty according to five criteria depicted in fig-tax which considers if a visualisation is implicit or explicit, intrinsic or extrinsic, visually integrable or seperable, coincident or adjacent, and static or dynamic.
 A similar version of this taxonomy was presented by @uncertchap2022 who commented that visualisation can be organised into two categories, "graphical annotations of distributional properties" and "visual encodings of uncertainty" which seems to functionally align with the intrinsic/extrinsic distinction by @Kinkeldey2014.
@@
 
-![Visualisation of @Kinkeldey2014 uncertainty visualisation taxonomy.
-It can be seen that most of the categories can be considered a change to the typical error bar chart.
-The taxonomy categorises the visualisations based on five criteria, implicit (a sample) or explicit (a depiction of mass); Intrinsic (alter existing symbols symbols to represent uncertainty) or extrinsic (add new objects to represent uncertainty); visually integral or separable (the uncertainty can be separated from the data and read independently) ; coincidence (uncertainty and data are represented in the same plot) or adjacent; static or dynamic (animated, interactive etc).
-Many of the distinctions conflate changes of the input data and statistics calculated and with visual changes, highlighting how defining uncertainty and visualisation together conflates several elements of a graphic.](uncertvistax.jpeg){#fig-tax}
-
 The authors of these taxonomies did not intend for them to be complete descriptions of uncertainty visualisations, however they do paint a picture of the what authors consider to be important.
@@
 ### What is uncertainty?
-
-Uncertainty is a concept that is famously hard to pin down, with constantly shifting definitions that mean it is quantified differently from author to author.
 
@@
 @Hofmann2012 showed two distributions in 20 different visualisations (a line-up protocol) using a jittered sample, a density plot, a histogram, and a box plot and asked participants.
-Participants were asked to report in which of the plots was "the blue group furthest to the right" The experiment set up is shown in @fig-right.
+Participants were asked to report in which of the plots was "the blue group furthest to the right" The experiment set up is shown in fig-right.
 The participants answers were then compared to a ground truth where the correct plot had a blue distribution with a right shifted mean.
@@
 Therefore, this method leads to inconclusive results about the plot design, and is not advised.
-
-![This shows the user interface for the experiment performed by @Hofmann2012. The question of "furthest to the right" is open to interpretation.](furthestright.png){#fig-right width=50%}
 
@@
 
-![The visualisation showing the kind of signal that we would expect an uncertainty plot to supress](signalsupression.jpeg){#fig-sigsupress}
-
 Additionally, setting up an uncertainty visualisation where the participants are expected to notice the signal once the data behind the visualisation passes a hypothesis test implies the signal *is* noticeable to a human at that level.
@@
 This concept bears similarity to the goal of uncertainty visualisation, but it is not quite the same.
-@fig-hypvs shows the conceptual difference between the line-up protocol and uncertainty visualisation.
-
-![A line-up protocol displays the uncertainty about the null and identifies if the true data plot is identifiable (and therefore significantly difference).
-This can be considered the graphical equivalent of a standard hypothesis test.
-Uncertainty is frequently used to descibe the area around the estimate that we think might contain the true value, and we assesses if the null of "no signal" is within this plot (e.g.
-if the error bars overlap with zero).
-Uncertainty is technically the distance between the true value and the estimate.
-](hypvs.jpeg){#fig-hypvs}
+fig-hypvs shows the conceptual difference between the line-up protocol and uncertainty visualisation.
 
@@
 
-# Bibliography
-
-```{r, include=FALSE, eval=FALSE}
-library(spelling)
-qmd <- "paper.qmd"
-ignore <- readLines("WORDLIST")
-check_spelling <- spell_check_files(
-  qmd,
-  ignore = ignore,
-  lang = "en_GB"
-)
-if (nrow(check_spelling) > 0) {
-  print(check_spelling)
-  stop("Check spelling in Qmd files!")
-}
+## Transparneyc map
+```{r}
+#| echo: false
+#| message: false
+#| warning: false
+#| eval: false
+#| label: fig-transparency
+#| fig-cap: ""
+#| fig-subcap: 
+#|   - "High Variance Data"
+#|   - "Low Variance Data"
+#|   - "Choropleth Palette"
+#| layout-ncol: 3
+#| layout-valign: "bottom"
+
+trans_map <- my_map_data |>
+  mutate(scaled_var = (variance - min(variance))/(max(variance)-min(variance)))
+
+# Transparency map
+pta <- trans_map |>
+  filter(variance_class=="lowvar") |>
+  ggplot() +
+  geom_sf(aes(fill = bitemp, 
+              geometry = geometry,
+              alpha=scaled_var)) + 
+  scale_fill_manual(values = basecols) +
+  theme_void() + 
+  theme(legend.position = "none")
+
+ptb <- pta %+% filter(trans_map, variance_class=="highvar")
+
 ```

### 2024-07-23 paper.qmd
added colour value map and finished integrability section

@@
 - I can do a more formal proof of the decision making thing using decision science. I left it out but it can be added if that section needs to be more thorough.
+- fix the colour value map
 -->
@@
 This prevents uncertainty and signal being read together as a "single integrated uncertain value" [@Kinkeldey2014] and allows for uncertainty to be ignored [@uncertchap2022].
-Additioanlly, this integration needs to be communicated through the "gist" of the graphic, as that is primary purpose of a visualisation [@Spiegelhalter2017].
-The @fig-bivariate plot's "gist" is "this data has a spatial trend and the estimates have a large variance". It is clear the uncertainty and signal are being communicated separately.
-
-Uncertainty visualisation is difficult because we need the uncertainty axis to interfere with the signal axis.
-When we typically visualise two independent variables we *do not* want them to interfere with each other, however the opposite is true for uncertainty visualisation.
-For uncertainty to perform signal suppression, it must interfere with the visual channel that signal was mapped to.
-It needs to interfere enough that viewers do not see signals that are simply the product of noise.
-
-Lets say we have a map were maximum daily temperature is presented using points where the colour (red for hot and blue for cold) of the point is associated with the temperature, and the blurriness of that point is associated with the variance *of* that temperature.
-It is highly likely that our brains will not flatten that into a single variable depicting noise and signal, but rather *separately* extract the temperature (colour) and uncertainty (blur) information as two independent variables.
-If the variables are extracted separately, there is no guarantee that the uncertainty will act as an appropriate signal suppressor.
-
- *** up to here ***
-
-The next map (which embodies the approach taken by the semantics experiments) is the bivariate map which maps the signal to colour value and the uncertainty to colour hue.
-If visualisation was performing signal suppression then that would mean the two dimensional space defined by colour value and colour hue can be mentally "flattened" into a single dimension of "valid value" that can captures the signal the our brain would need to be able to flatted this two dimensional space into a single space of "signal validity".
-
-The idea of two perceptual tasks flattening into one variable in the mind of the viewer may be wishful thinking, but it is not impossible given we are not certain on how the perceptual tasks are mapped within the human brain.
-@Sterzik2023 found that when a value was mapped to the textures of stippling, hatching, and triangles, and found that the difference between two points on this one dimensional texture was actually a 2D space (likely "texture business" and "light/darkness").
-That being said, if we look at the visual signals presented by the bivariate map, where the contrasting light and dark areas actually has no important meaning, it is unlikely this occurs for colour value and hue (or at least it doesn't occur in a way that is useful for uncertainty visualisation).
-
-- seperability experiments (susan, animation, scatter size)
-- suprise map (I think they had something about comparing light and dark)
-- Something else on accessibility or colour space (check susan's work)
-
-Instead of hoping that uncertainty might collapse signal values into a single dimension, we can do some of that work ourselves, by using a VSUP which collapses the colour space such that high uncertainty values cannot be extracted.
+Additionally, even when reading a map that has a more complicated colour palette than a choropleth map (such as a bivariate map) participants still read the map according to the same principles of the choropleth map [@Ndlovu2023].
+This means that even when uncertainty is technically included in the graphic, the take away can still be invalid if the map is read incorrectly.
+The signal-suppression needs to be communicated through the "gist" of the graphic, as that is primary purpose of a visualisation [@Spiegelhalter2017].
+The @fig-bivariate plot's "gist" is "this data has a spatial trend and the estimates have a large variance".
+It is clear the uncertainty and signal are being communicated separately rather that as a single uncertain value.
+
+This highlights the key problem of viewing uncertainty visualisation as a simple high dimensional problem.
+While we are trying to include a new variable, uncertainty cannot be treated the same as any other variable.
+Typically we want visual channels to be separable, that is, the data represented in one visual channel does not interfere with the others, and most research in this area is for this goal [@Smart2019].
+Uncertainty visualisation, however, should endeavor for integrability, that is, the channels interfere with one another to allow the uncertainty to suppress the signal.
+While most aesthetics *are* read separately, colour hue and brightness seem to be integrable [@Vanderplas2020] so even though they can be individually manipulated, they are processed by the brain as a single visual channel.
+Additionally color value has a more intuitive connection to uncertainty than hue [@Maceachren2012], but it should be mapped such that lighter colours are more uncertain than darker colours.
+If we could utilise integrability effectively, them the two dimensional space defined by colour value and colour hue could be mentally "flattened" into a single dimension of "statistically valid signal".
+
+@fig-bivariate2 shows a bivariate colour map where uncertainty is mapped to colour value, and temperature is mapped to colour hue.
+It is clear the spatial trend is still visible in the high uncertainty case.
+This does not necessarily mean integrability will never be a useful tool for signal suppression, but without a strong understanding of how these visual channels collapse down into a single channel, it may not be a reliable option.
+Additionally, adding so many variations to colour in a map might make the graphics harder to read and suppress statistically significant signals, while also hurting the plots accessibility [@Vanderplas2015].
+A more viable option might be to warp the space uncertainty and signal are mapped to ourselves.
+
+```{r}
+#| echo: false
+#| message: false
+#| warning: false
+#| label: fig-bivariate2
+#| fig-cap: "A bivariate map that depict the counties of Iowa where each county is coloured acording to it's average daily temperature and the variance in temperature.
+This map is a variation on the previous bivariate map where instead of variance being mapped to colour saturation, it is mapped to colour value.
+Even though colour value is intergrable with colour hue, and colour value is naturally accociated with uncertainty, the spatial trend is still visible in the map."
+#| fig-subcap: 
+#|   - "High Variance Data"
+#|   - "Low Variance Data"
+#|   - "Bivariate Palette"
+#| layout-ncol: 3
+#| layout-valign: "bottom"
+
+# Bivariate Map
+# Make bivariate palette
+# Function to devalue by a certain amount
+colsupress2 <- function(basecols, hue=1, sat=1, val=1) {
+    X <- diag(c(hue, sat, val)) %*% rgb2hsv(col2rgb(basecols))
+    hsv(pmin(X[1,], 1), pmin(X[2,], 1), pmin(X[3,], 1))
+}
+
+# recurvisely decrease value
+v_val = 1.1
+bivariatepal2 <- c(basecols,
+                   colsupress2(basecols, val=v_val), 
+                   colsupress2(colsupress2(basecols, val=v_val), val=v_val),
+                   colsupress2(colsupress2(colsupress2(basecols, val=v_val), val=v_val), val=v_val)
+                   )
+
+# establish levels of palette
+names(bivariatepal2) <- paste(rep(1:8, 4), "-" , rep(1:4, each=8), sep="")
+
+# Bivariate maps 2
+p2a1 <- my_map_data |>
+  filter(variance_class=="lowvar") %>%
+  ggplot() +
+  geom_sf(aes(fill = biclass, geometry = geometry)) + 
+  scale_fill_manual(values = bivariatepal2) +
+  theme_void() + 
+  theme(legend.position = "none")
+
+p2b2 <- p2a1 %+% filter(my_map_data, variance_class=="highvar")
+
+p2a1
+p2b2
+show_pal2(colours = bivariatepal2, ncol=8, borders=NA, myxlab = "Temperature", myylab = "Variance", breaks = 21:29, breaks2 = 0:4)
+
+```
+
+```{r}
+#| echo: false
+#| message: false
+#| warning: false
+#| eval: false
+#| label: fig-transparency
+#| layout-ncol: 3
+#| layout-valign: "bottom"
+
+# ALTERNATIVE TRANSPARENCY INSTEAD OF COLOUR VALUE
+
+trans_map <- my_map_data |>
+  mutate(scaled_var = (variance - min(variance))/(max(variance)-min(variance)))
+
+# Transparency map
+pta <- trans_map |>
+  filter(variance_class=="lowvar") |>
+  ggplot() +
+  geom_sf(aes(fill = bitemp, 
+              geometry = geometry,
+              alpha=scaled_var)) + 
+  scale_fill_manual(values = basecols) +
+  theme_void() + 
+  theme(legend.position = "none")
+
+ptb <- pta %+% filter(trans_map, variance_class=="highvar")
+```
+
+{{< pagebreak >}}
+
+# Visualising uncertainty and signal in a new space
+Instead of hoping that uncertainty might collapse signal values into a single dimension, we can do some of that work ourselves.
+
+by using a VSUP which collapses the colour space such that high uncertainty values cannot be extracted.
 It is unclear how useful VSUPs are at actually combining signal and noise and therefore suppressing plot level insights, as they have only been tested on simple value extraction tasks that require evaluating a single point [@Correll2018; @Ndlovu2023] rather than looking for spatial relationships (which is arguably what maps are for).
@@
 
-These maps make the importance of combining uncertainty and signal in a single visual channel clear.
-A choropleth map will show signal that is not valid inference because of high uncertainty.
-At the other end of the spectrum, the bivariate map will show signal that is not always interesting because it forces us to interpret uncertainty and signal separately.
-As we move through these methods, it seems that the validity of any overarching insight becomes more visible at the cost of our ability to extract particular values of signal or noise.
-Therefore, given that the primary goal of visualisation *is* insights [@North2006], visualisation authors should err on the side of representing suppressed signal as a single variable, rather than visualising uncertainty separately using two different channels.
-
-{{< pagebreak >}}
-
-# Visualising uncertainty and signal in a new space
 This problem has been noticed by others in the field, that typically use this method (specifically in the spatial uncertainty context) and a desire for representations that integrate uncertainty and signal is one of the reasons for the invention of the value-suppressing uncertainty pallet [@Correll2018].
@@
 ### There is no such thing as a test for everything
+As we move through these methods, it seems that the validity of any overarching insight becomes more visible at the cost of our ability to extract particular values of signal or noise.
+Therefore, given that the primary goal of visualisation *is* insights [@North2006], visualisation authors should err on the side of representing suppressed signal as a single variable, rather than visualising uncertainty separately using two different channels.
 
@@
 (*XXX Not sure what this paragraph is recommending?*)
-
-## Transparneyc map
-```{r}
-#| echo: false
-#| message: false
-#| warning: false
-#| eval: false
-#| label: fig-transparency
-#| fig-cap: ""
-#| fig-subcap: 
-#|   - "High Variance Data"
-#|   - "Low Variance Data"
-#|   - "Choropleth Palette"
-#| layout-ncol: 3
-#| layout-valign: "bottom"
-
-trans_map <- my_map_data |>
-  mutate(scaled_var = (variance - min(variance))/(max(variance)-min(variance)))
-
-# Transparency map
-pta <- trans_map |>
-  filter(variance_class=="lowvar") |>
-  ggplot() +
-  geom_sf(aes(fill = bitemp, 
-              geometry = geometry,
-              alpha=scaled_var)) + 
-  scale_fill_manual(values = basecols) +
-  theme_void() + 
-  theme(legend.position = "none")
-
-ptb <- pta %+% filter(trans_map, variance_class=="highvar")
-
-```

### 2024-07-24 paper.qmd
edited "why is uncertainty visualisation difficult" section

@@
 A successful uncertainty visualisation would act as a "statistical hedge" for any inference me make using the graphic.
-This implies three key criteria for a good uncertainty visualisation:
-
-  1) Reinforce justified conclusions to encourage confidence in results
-  2) Prevent unjustified conclusions that are indistinguishable from noise
+Additionally, since the purpose of a visualisation is to give a quick "gist" of the information [@Spiegelhalter2017], this hedging needs to be communicated visually without any need for computation from the viewer.
+Additionally, @Ndlovu2023 found that participants applied the same methods they used for simple choropleth maps to complicated uncertainty maps even if that take away was invalid.
+Therefore, this hedging effect needs to be communicated simply through the visualisation.
+If we refer to the conclusion we draw from a graphic to be its "signal" and the variance that makes this signal harder to identify as the "noise", we can summaries this information into three key requirements.
+A good uncertainty visualisation needs to:
+
+  1) Reinforce justified signals to encourage confidence in results
+  2) Hide signals that are just noise to prevent unjustified conclusions
   3) Perform tasks 1) and 2) in a way that is proportional to the level of confidence in those conclusions.
 
-We can refer to the we draw from a graphic to be its "signal" and the variance that makes this signal harder to identify as the "noise".
 As @fig-choropleth showed, visualisations that are unconcerned with uncertainty have no issue showing justified signals, but struggle with the display of unjustified signals.
-Therefore, we coin this approach to uncertainty visualisation as "signal-supression" since it primarily differentiates itself from from the "noiseless" visualisation approach through criteria (2).
-That is, the main difference between an uncertainty visualisation and a "normal" visualisation, is that an uncertainty visualisation should prevent us from drawing unjustified conclusions.
+Therefore, we coin this approach to uncertainty visualisation as "signal-suppression" since it primarily differentiates itself from from the "noiseless" visualisation approach through criteria (2).
+That is, the main difference between an uncertainty visualisation and a "normal" visualisation is that an uncertainty visualisation should prevent us from drawing unjustified conclusions.
 
@@
 This means the uncertainty *is* being communicated, however the primary take away in the graphic is the spatial trend (that does not exist).
-The graphic no longer has a transparency issue as the uncertainty *is* technically included, but is this a uncertainty visualisation?
+The graphic no longer has a transparency issue as the uncertainty *is* technically included, but is this a good uncertainty visualisation?
 And if not, why?
@@
 ## What makes uncertainty visualisation difficult?
-Uncertainty visualisation is a uniquely difficult problem in visualisation.
+Uncertainty visualisation is a uniquely difficult problem.
 @Hullman2016 commented that it is straightforward to show a value but it is much more complex to show uncertainty.
-There are two primary theories as to *why* uncertainty visualisation is unique.
+There does not seem to be a consensus on *why* this is the case, but there are two main groups of thought that emerge when trying to justify the difficulty in uncertainty visualisation.
 
@@
 While it is true that heuristics lead people to avoid uncertainty [@Spiegelhalter2017] there is no evidence that this psychological effect translates to a problem in visualisation.
-Were this the case, visualisations of death, poverty, and any other variable that elicits poor feelings would require its own field of visualisation.
+Were this the case, visualisations of death, poverty, and any other variable that elicits poor feelings would require their own field of visualisation.
 
 The second group sees uncertainty as a high dimensional visualisation problem as the difficulty is in working out how to add uncertainty into already existing graphics [@moritz2017trust; @Griethe2006].
-If we view uncertainty visualisation as a high dimension problem we include uncertainty as we would any other variable, and allow for the signal to be interpreted independent of the variance.
-This prevents uncertainty and signal being read together as a "single integrated uncertain value" [@Kinkeldey2014] and allows for uncertainty to be ignored [@uncertchap2022].
-Additionally, even when reading a map that has a more complicated colour palette than a choropleth map (such as a bivariate map) participants still read the map according to the same principles of the choropleth map [@Ndlovu2023].
-This means that even when uncertainty is technically included in the graphic, the take away can still be invalid if the map is read incorrectly.
-The signal-suppression needs to be communicated through the "gist" of the graphic, as that is primary purpose of a visualisation [@Spiegelhalter2017].
-The @fig-bivariate plot's "gist" is "this data has a spatial trend and the estimates have a large variance".
-It is clear the uncertainty and signal are being communicated separately rather that as a single uncertain value.
-
-This highlights the key problem of viewing uncertainty visualisation as a simple high dimensional problem.
-While we are trying to include a new variable, uncertainty cannot be treated the same as any other variable.
-Typically we want visual channels to be separable, that is, the data represented in one visual channel does not interfere with the others, and most research in this area is for this goal [@Smart2019].
-Uncertainty visualisation, however, should endeavor for integrability, that is, the channels interfere with one another to allow the uncertainty to suppress the signal.
-While most aesthetics *are* read separately, colour hue and brightness seem to be integrable [@Vanderplas2020] so even though they can be individually manipulated, they are processed by the brain as a single visual channel.
-Additionally color value has a more intuitive connection to uncertainty than hue [@Maceachren2012], but it should be mapped such that lighter colours are more uncertain than darker colours.
-If we could utilise integrability effectively, them the two dimensional space defined by colour value and colour hue could be mentally "flattened" into a single dimension of "statistically valid signal".
-
-@fig-bivariate2 shows a bivariate colour map where uncertainty is mapped to colour value, and temperature is mapped to colour hue.
-It is clear the spatial trend is still visible in the high uncertainty case.
-This does not necessarily mean integrability will never be a useful tool for signal suppression, but without a strong understanding of how these visual channels collapse down into a single channel, it may not be a reliable option.
-Additionally, adding so many variations to colour in a map might make the graphics harder to read and suppress statistically significant signals, while also hurting the plots accessibility [@Vanderplas2015].
-A more viable option might be to warp the space uncertainty and signal are mapped to ourselves.
+If we view uncertainty visualisation as a high dimension problem we include uncertainty as we would any other variable and use high-dimension visualisation techniques to incorporate it.
+This is the approach taken in @fig-bivariate.
+
+The problem with the high dimension approach is that it treats uncertainty the same as we would any other variable.
+Typically we want visual channels to be separable, that is, the data represented in one visual channel does not interfere with the others [@Smart2019].
+Unfortunately, this means that applying normal visualisation methods to "uncertainty visualisation" will prevent uncertainty and signal being read together as a "single integrated uncertain value" [@Kinkeldey2014], which is the goal of signal-suppression.
+Separating the uncertainty and the signal also allows for the uncertainty to be ignored, which is a common problem faced by most uncertainty visualisations [@uncertchap2022].
+We can see this problem in @fig-bivariate, as it sends the message "this data has a spatial trend and the estimates have a large variance" rather than hiding the statistically invalid signal.
+
+If thses high-dimensional approaches were to work, the uncertainty and the estimates would need to be mapped to integrable channels.
+That is, the channel that uncertainty and the estimates are mapped to would need to be separately manipulable, but read as a single channel by the human brain.
+While most aesthetics *are* seperable, there are some variables that have been shown to be integrable, such as colour hue and brightness [@Vanderplas2020].
+We can more effectively approach this method by also leveraging visual semiotics and mapping more uncertain values to lighter colors [@Maceachren2012].
 
@@
 ```
+
+@fig-bivariate2 shows a bivariate color map where uncertainty is mapped to color value, and temperature is mapped to color hue.
+It is clear the spatial trend is still visible in the high uncertainty case.
+This does not necessarily mean integrability will never be a useful tool for signal suppression, but without a strong understanding of how these visual channels collapse down into a single channel, it may not be a reliable option.
+Relying on integrability could create unintended consequences of uncertainty visualization, such as displaying phantom signals or hiding justified signals.
+Additionally, multi-dimensional colour palettes can make the graphics harder to read and hurt the accessibility of the plots [@Vanderplas2015].
+A more viable option might be to warp the space uncertainty and signal are mapped to ourselves.
 

### 2024-07-24 paper.qmd
started moving stuff around

@@
 ### Value Supressing Uncertainty Palettes
-### There is no such thing as a test for everything
-As we move through these methods, it seems that the validity of any overarching insight becomes more visible at the cost of our ability to extract particular values of signal or noise.
-Therefore, given that the primary goal of visualisation *is* insights [@North2006], visualisation authors should err on the side of representing suppressed signal as a single variable, rather than visualising uncertainty separately using two different channels.
-
-{{< pagebreak >}}
-
-# Implicitly Combining Uncertainty and Signal
-## Pixel maps
-## Show me the data
-
-{{< pagebreak >}}
-
-# Evaluating uncertainty visualisations
-# Current methods
-# Testing signal supression
-## Current methods
-### Secondary effects
-
-- trust, confidence, etc
-
-### Direct questions
-
-> Most studies that assess the usability of uncertainty visualisations do not contribute to this aspect since they test the retrieval of data and uncertainty separately, but from the perspective that there is nothing special about uncertainty.
-Traditionally, studies in this field focus on the ability to read both the map content and its uncertainty at the same time.
-This may be the mandatory criterion for a successful use of uncertainty, but the question that remains is whether this is sufficient to ensure that a user does not only have two separate values in mind but an integrated uncertain data value.
-[@Kinkeldey2014]
-
-@Hullman2019 found that the majority of papers evaluate visualisations on performance (approximately 65% of the papers they surveyed).
-
-By far the most common metric used is accuracy, which is used by approximately 36% of evaluation studies [@Hullman2019].
-
-The final method used by authors is to just explicitly ask about uncertainty and signal information separately.
-@Sanyal2009 mapped uncertainty to dots and signal to a 3D surface and asked participants to identify areas of high and low signal and high and low uncertainty.
-Participants were not asked to combine that information in any way, and the signal and the noise were treated as separate variables.
-@Correll2014 asked participants to separately extract the mean and variance from four uncertainty visualisations.
-These methods explicitly view the uncertainty and signal as two separate variables that should be extracted from a plot, and not two variables that should be interpreted together.
-Even viewing these questions as a routine check to make sure the signal information isn't impacted by the uncertainty is counter intuitive, because the whole point *of* the uncertainty is to impact the signal information.
-
-- definition of uncertainty
-
-There are many secondary benefits that come with this improved transparency, such as better decisions, more trust in the results and more confidence in the authors.
-These secondary benefits are, however, *not* the immediate goal of uncertainty.
-The following sections will discuss the issues and limitations in measuring uncertainty through these secondary metrics and provide suggestions as to how future studies should consider measuring uncertainty.
-
-Trust is a by product of displaying uncertainty rather than the goal of it, and viewing the relationship in the converse direction can lead to misguided research.
-Considering trust, and not transparency, as the metric of importance in uncertainty communication can lead to a questionable subtext that argues against transparency, something that has been noticed by several other authors [@Spiegelhalter2017; ONeill2018].
-@Hullman2020a found that author simultaneously argued that failing to visualise uncertainty was akin to fraud, but also many avoided uncertainty visualisation because they didn't want their work to come across as "untrustworthy".
-These authors are optimising *trust* rather than *transparency*, which means they opt to leave out uncertainty information when it does exactly what it is supposed to, decrease certainty in conclusions.
-
-Science communication should be primarily concerned with accuracy.
-Setting trust and risk-aversion as the variables of interest implicitly encourages statisticians to set trust and risk-aversion as the primary goals of communication.
-The issue of trust being divorced from trustworthiness has been commented on by other authors [@ONeill2018], however the issue still persists in the uncertainty visualisation literature.
-@Zhao2023 displayed a several variations of a visualisation of a model prediction and its uncertainty and took participants using the model prediction as a sign of trust.
-They reported that visualising uncertainty information caused participants to trust the model in the low variance case, but the results in the high variance case were inconclusive.
-The discussion made it clear the authors thought the uncertainty information should make the visualisation more trustworthy, but conflating trust and the use of a prediction implied uncertainty information should somehow influence participants to use their own prediction, even though a prediction being uncertainty does not necessarily mean it is incorrect.
-Despite this, the authors seemed to assume that the uncertainty information *should* have an influence on that, showing they had not deeply considered *how* uncertainty information should influence the choices of the participants.
-(*Cite Gap 18: examples of studies where authors measure trust*)
-
-A similar measure to trust is using "confidence" in an extracted value or a decision.
-Interestingly, "confidence" is also used to try and capture the clarity of a message in a normal visualisation.
-Confidence cannot simultaneously be a measure of clarity of visualisation *and* a way to capture the uncertainty expressed in a visualisation.
-
-{{< pagebreak >}}
-
-# Bibliography
-
-```{r, include=FALSE, eval=FALSE}
-library(spelling)
-qmd <- "paper.qmd"
-ignore <- readLines("WORDLIST")
-check_spelling <- spell_check_files(
-  qmd,
-  ignore = ignore,
-  lang = "en_GB"
-)
-if (nrow(check_spelling) > 0) {
-  print(check_spelling)
-  stop("Check spelling in Qmd files!")
-}
-```
-
-# ------------------- BELOW HERE IS UNSORTED ------------------------
-
-# MAPS
 
@@
 #| warning: false
-#| eval: false
+#| label: fig-vsup
+#| fig-cap: "A map made with a VSUP.
+The counties of Iowa are coloured acording to it's average daily temperature and the variance in temperature, although the space they have been mapped to is slightly warped."
+#| fig-subcap: 
+#|   - "High Variance Data"
+#|   - "Low Variance Data"
+#|   - "VSUP Palette"
+#| layout-ncol: 3
+#| layout-valign: "bottom"
 
@@
 
+p3a
+p3b
+show_col(VSUP, ncol=8, labels=FALSE)
+
+```
+
+### There is no such thing as a test for everything
+As we move through these methods, it seems that the validity of any overarching insight becomes more visible at the cost of our ability to extract particular values of signal or noise.
+Therefore, given that the primary goal of visualisation *is* insights [@North2006], visualisation authors should err on the side of representing suppressed signal as a single variable, rather than visualising uncertainty separately using two different channels.
+
+{{< pagebreak >}}
+
+# Implicitly Combining Uncertainty and Signal
+## Pixel maps
+```{r}
+#| echo: false
+#| message: false
+#| warning: false
+#| label: fig-pixel
+#| fig-cap: "A pixel map of the counties of Iowa.
+In this map, each county is broken up into several smaller regions and coloured according to a potential daily temperature, given the its average daily temperature and its sampling distribution.
+This results in each county being represented by a sample rather than a single value."
+#| fig-subcap: 
+#|   - "High Variance Data"
+#|   - "Low Variance Data"
+#|   - "Pixel Map Palette"
+#| layout-ncol: 3
+#| layout-valign: "bottom"
+
 # Pixel Map
@@
 # Display plots and palettes
-p3a
-p3b
-show_col(VSUP, ncol=8, labels=FALSE)
 p4a
@@
 show_col(basecols, ncol=8, labels=FALSE)
+
 ```
+## Show me the data
+
+{{< pagebreak >}}
+
+# Evaluating uncertainty visualisations
+# Current methods
+# Testing signal supression
+## Current methods
+### Secondary effects
+
+- trust, confidence, etc
+
+### Direct questions
+
+> Most studies that assess the usability of uncertainty visualisations do not contribute to this aspect since they test the retrieval of data and uncertainty separately, but from the perspective that there is nothing special about uncertainty.
+Traditionally, studies in this field focus on the ability to read both the map content and its uncertainty at the same time.
+This may be the mandatory criterion for a successful use of uncertainty, but the question that remains is whether this is sufficient to ensure that a user does not only have two separate values in mind but an integrated uncertain data value.
+[@Kinkeldey2014]
+
+@Hullman2019 found that the majority of papers evaluate visualisations on performance (approximately 65% of the papers they surveyed).
+
+By far the most common metric used is accuracy, which is used by approximately 36% of evaluation studies [@Hullman2019].
+
+The final method used by authors is to just explicitly ask about uncertainty and signal information separately.
+@Sanyal2009 mapped uncertainty to dots and signal to a 3D surface and asked participants to identify areas of high and low signal and high and low uncertainty.
+Participants were not asked to combine that information in any way, and the signal and the noise were treated as separate variables.
+@Correll2014 asked participants to separately extract the mean and variance from four uncertainty visualisations.
+These methods explicitly view the uncertainty and signal as two separate variables that should be extracted from a plot, and not two variables that should be interpreted together.
+Even viewing these questions as a routine check to make sure the signal information isn't impacted by the uncertainty is counter intuitive, because the whole point *of* the uncertainty is to impact the signal information.
+
+- definition of uncertainty
+
+There are many secondary benefits that come with this improved transparency, such as better decisions, more trust in the results and more confidence in the authors.
+These secondary benefits are, however, *not* the immediate goal of uncertainty.
+The following sections will discuss the issues and limitations in measuring uncertainty through these secondary metrics and provide suggestions as to how future studies should consider measuring uncertainty.
+
+Trust is a by product of displaying uncertainty rather than the goal of it, and viewing the relationship in the converse direction can lead to misguided research.
+Considering trust, and not transparency, as the metric of importance in uncertainty communication can lead to a questionable subtext that argues against transparency, something that has been noticed by several other authors [@Spiegelhalter2017; ONeill2018].
+@Hullman2020a found that author simultaneously argued that failing to visualise uncertainty was akin to fraud, but also many avoided uncertainty visualisation because they didn't want their work to come across as "untrustworthy".
+These authors are optimising *trust* rather than *transparency*, which means they opt to leave out uncertainty information when it does exactly what it is supposed to, decrease certainty in conclusions.
+
+Science communication should be primarily concerned with accuracy.
+Setting trust and risk-aversion as the variables of interest implicitly encourages statisticians to set trust and risk-aversion as the primary goals of communication.
+The issue of trust being divorced from trustworthiness has been commented on by other authors [@ONeill2018], however the issue still persists in the uncertainty visualisation literature.
+@Zhao2023 displayed a several variations of a visualisation of a model prediction and its uncertainty and took participants using the model prediction as a sign of trust.
+They reported that visualising uncertainty information caused participants to trust the model in the low variance case, but the results in the high variance case were inconclusive.
+The discussion made it clear the authors thought the uncertainty information should make the visualisation more trustworthy, but conflating trust and the use of a prediction implied uncertainty information should somehow influence participants to use their own prediction, even though a prediction being uncertainty does not necessarily mean it is incorrect.
+Despite this, the authors seemed to assume that the uncertainty information *should* have an influence on that, showing they had not deeply considered *how* uncertainty information should influence the choices of the participants.
+(*Cite Gap 18: examples of studies where authors measure trust*)
+
+A similar measure to trust is using "confidence" in an extracted value or a decision.
+Interestingly, "confidence" is also used to try and capture the clarity of a message in a normal visualisation.
+Confidence cannot simultaneously be a measure of clarity of visualisation *and* a way to capture the uncertainty expressed in a visualisation.
+
+{{< pagebreak >}}
+
+# Bibliography
+
+```{r, include=FALSE, eval=FALSE}
+library(spelling)
+qmd <- "paper.qmd"
+ignore <- readLines("WORDLIST")
+check_spelling <- spell_check_files(
+  qmd,
+  ignore = ignore,
+  lang = "en_GB"
+)
+if (nrow(check_spelling) > 0) {
+  print(check_spelling)
+  stop("Check spelling in Qmd files!")
+}
+```
+
+# ------------------- BELOW HERE IS UNSORTED ------------------------
 

### 2024-07-24 paper.qmd
reorganised "visualising uncertainy as signal" section

@@
 Uncertainty visualisation is not only motivated by signal-suppression, and we would be remiss to ignore these alternative motivations.
-Some authors claim the purpose of uncertainty is to improve decision making [@Ibrekk1987; @uncertchap2022; @Hullman2016; @Cheong2016; @Boone2018; @Padilla2017].
+In this section we discuss alternative motivations for uncertainty visualization, and explain why they are not the focus of this review.
+
+## Uncertainty as a variable of interest
+Other authors do not describe uncertainty as important for decision making, but rather explicitly state it as variable of importance in of itself [@Blenkinsop2000].
 While uncertainty can provide useful information in decision making, it is important to recognize the "uncertainty" in these cases is not acting as "uncertainty" at all.
@@
 This is easiest to understand with an example.
+
+Extracting information using a graphic requires a perceptual task, so it will always be less accurate than explicitly reading the value provided in text form [@Cleveland1984].
+
+This is why text representations or graphics that allow users to simply count a value to extract the required statistic outperform those that require any perceptual tasks.
+
+## Uncertainty for decision making
+Some authors claim the purpose of uncertainty is to improve decision making [@Ibrekk1987; @uncertchap2022; @Hullman2016; @Cheong2016; @Boone2018; @Padilla2017].
+This group are *also* viewing uncertainty as a signal, although it is less obvious in this case.
 
@@
 
-Other authors do not describe uncertainty as important for decision making, but rather explicitly state it as variable of importance in of itself [@Blenkinsop2000].
-This group are *also* viewing uncertainty as a signal, albeit more explicitly than the decision making group.
-If the purpose of uncertainty in these cases is to convey information rather than obfuscate it, then there is no reason to assume these "uncertainty visualisations" would behave any differently to normal information visualisations.
-This is exactly what the literature shows.
-
-Uncertainty visualisation experiments often compare graphics on the basis of being "uncertainty visualisations" [@Ibrekk1987; @Hullman2015; @Hofman2020] with little consideration for how difficult the information is to extract from the chat.
-Experiments where users are asked to make a decision based on a probability, and experiments where users need to extract a variance or some feature of the distribution both have the implicit belief that uncertainty visualisation is a "class" of visualisation, and a "good" uncertainty visualisation allows us to extract all information with ease.
-This is the logic that results in predictable results.
-@Ibrekk1987 found that participants were more accurate at extracting a statistic when it could be directly read off the graphic, than when it required an area estimate (which is the case if using the PDF), or when there was no visual indicator for the statistic at all (which is the case when using the CDF of an asymmetric function).
-@Hullman2015 found that a visualisation that allows viewers to count outcomes to estimate a probability outperformed one that required a complicated area calculation.
-@Hofman2020 and @Zhang2022 found that participants were better at answering questions about a prediction intervals when shown a prediction interval instead of a sampling distribution.
-@Gschwandtnei2016 found that graphics where the required statistic could be directly read off the plot outperformed those that involved guesswork due to a gradually decreasing line.
-@Cheong2016 found that participants made better decisions when they were explicitly given the relevant probability in text rather than when they needed to read it off a map.
-These experiments all have the same fundamental finding, if we create a problem by obfuscating information, then the representation that makes the obfuscated information easier to extract outperforms those that don't.
-This is why text representations or graphics that allow users to simply count a value to extract the required statistic outperform those that require any perceptual tasks.
-We have known perceptual tasks have more noise than text representations for a long time [@Cleveland1984] and it has little to do with uncertainty.
+This obfuscation problem translates to unanswerable questions in uncertainty evaluation experiments.
+
+If we create a problem by obfuscating information, then the representation that makes the obfuscated information easiest to extract outperforms those that don't.
 
@@
 While the input into the graphic has changed, graphics, just like statistics, are not defined by their input data.
-
-This seems to be an accepted fact for all visualisations *except* for uncertainty visualisation, and this creates disagreement in what is even considered to be an uncertainty visualisation.
-For example @Leland2005 mentions that popular graphics, such as pie charts and bar charts omit uncertainty, and @Wickham2011 suggests their product plot framework, which includes histograms and bar charts, should be extended to include uncertainty however at least one or both of these charts are used in a significant number of uncertainty visualisation experiments [@Ibrekk1987; @Olston2002; @Zhao2023; @Hofmann2012].
-We belive this disagreement comes from two different conceptualisations of what an uncertainty  visualisation *is*.
-This issue causing confusion in the literature was also commented on by @Kinkeldey2014, who explained it was unclear if uncertainty should be treated as a variable, as metadata, or as something else entirely [@Kinkeldey2014].
-There are those that describe *any* visualisation of error as an "uncertainty visualisation" and those that believe "uncertainty visualisation" is a function that can be *applied* to any graphic that depicts an estimate.
-The former group view the purpose of uncertainty visualisation to provide signal about a distribution, while the later believe it should act as noise to obfuscate a signal.
 
@@
 
+## What is an uncertainty visualisation?
+
+This seems to be an accepted fact for all visualisations *except* for uncertainty visualisation, and this creates disagreement in what is even considered to be an uncertainty visualisation.
+For example @Leland2005 mentions that popular graphics, such as pie charts and bar charts omit uncertainty, and @Wickham2011 suggests their product plot framework, which includes histograms and bar charts, should be extended to include uncertainty however at least one or both of these charts are used in a significant number of uncertainty visualisation experiments [@Ibrekk1987; @Olston2002; @Zhao2023; @Hofmann2012].
+
+We belive this disagreement comes from two different conceptualizations of what an uncertainty  visualisation *is*.
+This issue causing confusion in the literature was also commented on by @Kinkeldey2014, who explained it was unclear if uncertainty should be treated as a variable, as metadata, or as something else entirely [@Kinkeldey2014].
+There are those that describe *any* visualisation of error as an "uncertainty visualisation" and those that believe "uncertainty visualisation" is a function that can be *applied* to any graphic that depicts an estimate.
+The former group view the purpose of uncertainty visualisation to provide signal about a distribution, while the later believe it should act as noise to obfuscate a signal.
+
 The easily explained results in evaluation experiments, and the exceedance probabiltiy map show the problem of viewing uncertainty as a signal instead of noise.
@@
 
-# Including uncertainty as an indepdendent variable
+# Visualising uncertainty as a variable
 Upon hearing that uncertainty needs to be included for transparency, the solutions may seem obvious.
@@
 
-### There is no such thing as a test for everything
+### Uncertainty and inference
+The ramifications of this is seen throughout the literature, as a survey of visualisation authors cited "not knowing how to calculate uncertainty" as one of the primary reasons they did not include it in visualisations [@Hullman2020a].
+Some works [@Hullman2018; @Maceachren2012; @Thomson2005] get around this definition issue by focusing narrowly on specific terms defined mathematically, such as probability, variance, error, or precision.
+Others [@Griethe2006; @Pang1997; @Pham2009; @Boukhelifa2017] include broader loosely related elements, such as missing values, but ignore these concepts when translating that uncertainty to a visualisation.
+Existing mathematical definition of uncertainty (or related concepts) have been unable to unify this range of diverse concepts.
+For example, @Thomson2005 suggests a mathematical formula for *examples* of uncertainty, and information theory tries to quantify uncertainty using the idea of entropy, but the disconnect between the broad concept of uncertainty and what we can reliably quantify remains a problem.
+
+Some authors prefer to define uncertainty in qualitative terms, often through the use of taxonomies.
+In these cases, uncertainty is split using an endless stream of ever changing boundaries, such as whether the uncertainty is due to true randomness or a lack of knowledge [@Spiegelhalter2017; @Hullman2016; @utypo], if the uncertainty is in the attribute, spatial elements, or temporal element of the data [@Kinkeldey2014], whether the uncertainty is scientific (e.g.
+error) or human (e.g.
+disagreement among parties) [@Benjamin2018], if the uncertainty is random or systematic [@Sanyal2009], statistical or bounded [@Gschwandtnei2016; @Olston2002], recorded as accuracy or precision [@Griethe2006; @Benjamin2018], which stage of the data analysis pipeline the uncertainty comes from [@utypo], how quantifiable the uncertainty is [@Spiegelhalter2017; @utypo], etc.
+In their own way, each of these approaches show an aspect of uncertainty that an author felt was important to differentiate.
+
+What is interesting about all these views of uncertainty, is that they all imply that it is a latent feature of our data.
+That is, it is something that can be revealed through exploration, and not something that needs to be explicitly calculated.
+The constant stream of new definitions that are unable to capture the true essence of uncertainty might be a result of this false assumption that underpins ever definition.
+Uncertainty cannot be defined without a signal because uncertainty does not exist without a signal, is a by-product of inference.
+
+It is easy to see uncertainties relationship to inference when we consider what might *not* be considered uncertain rather than just trying to think about what *is*.
+
+Descriptive statistics simply describe our sample as it is and summarizes large data down into an easy to swallow format.
+Descriptive statistics are not seen as the primary goal of modern statistics, however, this was not always the case.
+Around the 19th century in England, *positivism* was the popular philosophical approach to science (positivists included famous statisticians such as Francis Galton and Karl Pearson) and practitioners of the approach believed statistics ended with descriptive statistics as science must be based on actual experience and observations, therefore anything that refers to the unobservable (such as new observations or population statistics) is not true science [@Otsuka2023].
+In order to make statements about population statistics, future values, or new observations we need to perform inference, which requires the assumption of the "uniformity of nature" (i.e.
+that unobserved phenomena should be similar to observed phenomena) [@Otsuka2023].
+This subtle shift, from descriptive statistics to inferential statistics was shunned during the positivism era *due to the fact it introduced the unknowable*, or in other words, uncertainty.
+
+This approach to uncertainty is embedded in the idea that descriptive statistics do not have uncertainty, which some readers may disagree with.
+Specifically, because it means uncertainty is *not* a latent attribute of data, but rather an attribute of a specific hypothesis or estimate.
+Deniers of this fact follow a consistent logical path and it is easy to identify the common mistake.
+We know that variance and probability are typically considered types of "uncertainty" *and* descriptive statistics can have variance and probabilities *therefore* descriptive statistics must have uncertainty.
+The flaw in this logic comes from the first step, assuming that the tools with which we measure uncertainty *are* uncertainty in of themselves.
+This confusion is common and there are many papers that spend a great deal of time clarifying the difference.
+@Begg2014 highlight that uncertainty is related to not knowing a specific value, while variability refers to the range of values a quantity can take at different locations, times or instances.
+@Spiegelhalter2017 made sure to comment on the difference between precise random events (such as the probability associated with a coin flip), and uncertainty (such as the estimated probability associated with a coin that might be biased).
+The variance of a sample variance can be calculated and know, therefore it is not uncertain but rather it a precise description of dispersion.
+If we were to discuss drawing a new observation, or estimating the true mean of a population *then* the variance would become relevant in our discussions of uncertainty.
+
+With this understanding it becomes clear to see why uncertainty is tied to an endless string of examples in the data analysis pipeline.
+Uncertainty examples include imputed data, model selection, inherent randomness, biased sampling, etc, not because these things *are* uncertainty, but because they *create* uncertainty when we perform inference.
+Whether or not these elements are relevant is highly dependent on what statistic you are trying to draw inference on, and by extension, the purpose of your visualisation.
+
 As we move through these methods, it seems that the validity of any overarching insight becomes more visible at the cost of our ability to extract particular values of signal or noise.
 Therefore, given that the primary goal of visualisation *is* insights [@North2006], visualisation authors should err on the side of representing suppressed signal as a single variable, rather than visualising uncertainty separately using two different channels.
+
+The concept of uncertainty as a by-product of inference, not a latent feature of data, extends naturally to visualisation. 
+There are two primary reasons for making a visualisation, to perform exploratory data analysis (EDA) or communication.
+Communication involves identifying a signal we want to communicate and designing the visualisation that best conveys that, while EDA involves creating a versatile visualisation without an explicit purpose and using it to extract several signals.
+
+ Since EDA is the visual parallel to descriptive statistics, it is performed without an explicit hypothesis which means there *is* no uncertainty in the visualisation.
+Similarly to our issues with descritive statistics, this is not well understood by the uncertainty visualisation community.
+Some authors simply confuse varianc for uncertainty.
+For example @Potter2010 aimed to create a summary plot that "concisely presented data with uncertainty information" to create an exploratory visualisation tool that visualised uncertainty.
+Other authors regonise inference will occur (in some shape or form) and believe uncertainty *should* be visualised but do not recognise *how* uncertainty would be visualised.
+@Hullman2021 argued that there is no such thing as a "model-free" visualisation, therefore visualisation require robust visualisations of uncertainty as we are always performing inference.
+@Griethe2006 commented that "if visualization is used as a means to explore a data volume or to communicate its contents the uncertainty has to be included".
+While we agree people cannot prevent themselves from performing inference, that does not mean it is possible for uncertainty to be included in an EDA visualisation.
+A versatile visualisation such as a scatter plot allows for a viewer to consider several hypothesis at once, each of which will be a different inferential statistic with a different distribution that depicts its uncertainty.
+It is likely impossible to suppress all possible signals at once.
 
@@
 # Evaluating uncertainty visualisations
+
+## Uncertainty disappears if you stare directly at it
+
 # Current methods
+
 # Testing signal supression
+If asking direct questions about uncertainty causes us to treat it as a signal, how do we evaluate uncertainty as *noise*?
+ When we ask the viewer of a plot to look at data and extract a value, we are asking them to perform inference on that value.
+There will be noise associated with that answer and that is uncertainty.
+If we ask direct question about some uncertainty metric, we have turned the uncertainty into signal because that is what the participants are drawing inference on.
+
 ## Current methods
@@
 ### Direct questions
+Several authors have connected the issues around defining uncertainty to inference, however it is often discussed as a *task* or *goal* dependence.
+Multiple authors have commented on the need to consider quantifying and expressing uncertainty at every stage of a project as the "goal" shapes every step of the analysis [@Kinkeldey2014; @Hullman2016; @Refsgaard2007].
+@Otsuka2023 suggested that the process of observing data to perform statistics is largely dependent on our goals, because the process of boiling real world entities down into probabilistic objects (or "probabilistic kind" as he puts it) depends on the relationship we seek to identify with our data.
+@Meng2014 commented what is kept as data and what is tossed away is determined by the motivation of an analysis and what was previously noise can be shown to become signal depending on the the question we seek to answer.
+@Kale2019 discussed how the choices we make in our analysis impact our outcomes and introduce uncertainty.
+@Carlin2023 mentions that each research question can be can be categorised as descriptive, predictive, or causal, each of which has its own appropriate statistical methods and motivation agnostic model selection leads to statistical analysis that is devoid of meaning.
+@Wallsten1997 argue that the best method for evaluating or combining subjective probabilities depends on the uncertainty the decision maker wants to represent and why it matters.
+@Fischhoff2014 looks at uncertainty visualisation for decision making decides that we should have different ways of communicating uncertainty based off what the user is supposed to do with it.
+The importance of inference when discussing uncertainty is never directly acknowledge, but it is always present.
+It is clear that the relationship between uncertainty and inference is noticed at every stage in an analysis, however combining these uncertainties into a single "uncertainty" value is near impossible [@Spiegelhalter2017].
+
+Experiments where users are asked to make a decision based on a probability, and experiments where users need to extract a variance or some feature of the distribution both have the implicit belief that uncertainty visualisation is a "class" of visualisation, and a "good" uncertainty visualisation allows us to extract all information with ease.
 
@@
 [@Kinkeldey2014]
+
+Uncertainty visualisation experiments often compare graphics on the basis of being "uncertainty visualisations" [@Ibrekk1987; @Hullman2015; @Hofman2020] with little consideration for how difficult the information is to extract from the chat.
+For example, @Ibrekk1987 found that participants were more accurate at extracting a statistic when it could be directly read off the graphic, than when it required an area estimate (which is the case if using the PDF), or when there was no visual indicator for the statistic at all (which is the case when using the CDF of an asymmetric function).
+@Hullman2015 found that a visualisation that allows viewers to count outcomes to estimate a probability outperformed one that required a complicated area calculation.
+@Hofman2020 and @Zhang2022 found that participants were better at answering questions about a prediction intervals when shown a prediction interval instead of a sampling distribution.
+@Gschwandtnei2016 found that graphics where the required statistic could be directly read off the plot outperformed those that involved guesswork due to a gradually decreasing line.
+@Cheong2016 found that participants made better decisions when they were explicitly given the relevant probability in text rather than when they needed to read it off a map.
 
@@
 
-# What is an uncertainty visualisation?
-@Kinkeldey2014 categorised uncertainty according to five criteria depicted in fig-tax which considers if a visualisation is implicit or explicit, intrinsic or extrinsic, visually integrable or seperable, coincident or adjacent, and static or dynamic.
-A similar version of this taxonomy was presented by @uncertchap2022 who commented that visualisation can be organised into two categories, "graphical annotations of distributional properties" and "visual encodings of uncertainty" which seems to functionally align with the intrinsic/extrinsic distinction by @Kinkeldey2014.
-@Griethe2006 organised uncertainty visualisations into two cases (1) a hypothesis test was preformed to confirm the validity of the visualisation and (2) the visualisation has uncertainty depicted.
-@Potter2012 organised several existing uncertainty visualisations into groups based on the dimensionality of the data (1D, 2D, 3D, and No Dimension) and the dimensionality of the uncertainty (Scalar, Vector, Tensor).
-However, because the term "PDF", function that is used to describe a random variable, is used to describe both the data and the uncertainty for all dimensions.
-@Grewal2021 created a taxonomy that mapped uncertainty visualisations to some point in a 2D space defined by the "domain expertise" and "continuum of discreteness" (that scaled from "point estimate" to "continuous distribution").
-
-The authors of these taxonomies did not intend for them to be complete descriptions of uncertainty visualisations, however they do paint a picture of the what authors consider to be important.
-There are common themes in the taxonomies that ensure we differentiate between a sample and a mass, consider if uncertainty should be depicted with the estimate or as a second variable, identify the parallel of hypothesis testing, and consider the dimensionality, sample size and precision depicted by the visualisation.
-These taxonomies may highlight perceived important differences, but they do not help us understand what *is* or *is not* an uncertainty visualisation.
-For that we need to understand *why* uncertainty visualisation exists as a sub-field at all.
-
-### What is uncertainty?
-
-The ramifications of this is seen throughout the literature, as a survey of visualisation authors cited "not knowing how to calculate uncertainty" as one of the primary reasons they did not include it in visualisations [@Hullman2020a].
-Some works [@Hullman2018; @Maceachren2012; @Thomson2005] get around this definition issue by focusing narrowly on specific terms defined mathematically, such as probability, variance, error, or precision.
-Others [@Griethe2006; @Pang1997; @Pham2009; @Boukhelifa2017] include broader loosely related elements, such as missing values, but ignore these concepts when translating that uncertainty to a visualisation.
-Existing mathematical definition of uncertainty (or related concepts) have been unable to unify this range of diverse concepts.
-For example, @Thomson2005 suggests a mathematical formula for *examples* of uncertainty, and information theory tries to quantify uncertainty using the idea of entropy, but the disconnect between the broad concept of uncertainty and what we can reliably quantify remains a problem.
-
-Some authors prefer to define uncertainty in qualitative terms, often through the use of taxonomies.
-In these cases, uncertainty is split using an endless stream of ever changing boundaries, such as whether the uncertainty is due to true randomness or a lack of knowledge [@Spiegelhalter2017; @Hullman2016; @utypo], if the uncertainty is in the attribute, spatial elements, or temporal element of the data [@Kinkeldey2014], whether the uncertainty is scientific (e.g.
-error) or human (e.g.
-disagreement among parties) [@Benjamin2018], if the uncertainty is random or systematic [@Sanyal2009], statistical or bounded [@Gschwandtnei2016; @Olston2002], recorded as accuracy or precision [@Griethe2006; @Benjamin2018], which stage of the data analysis pipeline the uncertainty comes from [@utypo], how quantifiable the uncertainty is [@Spiegelhalter2017; @utypo], etc.
-In their own way, each of these approaches show an aspect of uncertainty that an author felt was important to differentiate.
-
-What is interesting about all these views of uncertainty, is that they all imply that it is a latent feature of our data.
-That is, it is something that can be revealed through exploration, and not something that needs to be explicitly calculated.
-The constant stream of new definitions that are unable to capture the true essence of uncertainty might be a result of this false assumption that underpins ever definition.
-Uncertainty cannot be defined without a signal because uncertainty does not exist without a signal, is a by-product of inference.
-
-#### Uncertainty as a by-product of inference
-It is easy to see uncertainties relationship to inference when we consider what might *not* be considered uncertain rather than just trying to think about what *is*.
-
-Descriptive statistics simply describe our sample as it is and summarizes large data down into an easy to swallow format.
-Descriptive statistics are not seen as the primary goal of modern statistics, however, this was not always the case.
-Around the 19th century in England, *positivism* was the popular philosophical approach to science (positivists included famous statisticians such as Francis Galton and Karl Pearson) and practitioners of the approach believed statistics ended with descriptive statistics as science must be based on actual experience and observations, therefore anything that refers to the unobservable (such as new observations or population statistics) is not true science [@Otsuka2023].
-In order to make statements about population statistics, future values, or new observations we need to perform inference, which requires the assumption of the "uniformity of nature" (i.e.
-that unobserved phenomena should be similar to observed phenomena) [@Otsuka2023].
-This subtle shift, from descriptive statistics to inferential statistics was shunned during the positivism era *due to the fact it introduced the unknowable*, or in other words, uncertainty.
-
-This approach to uncertainty is embedded in the idea that descriptive statistics do not have uncertainty, which some readers may disagree with.
-Specifically, because it means uncertainty is *not* a latent attribute of data, but rather an attribute of a specific hypothesis or estimate.
-Deniers of this fact follow a consistent logical path and it is easy to identify the common mistake.
-We know that variance and probability are typically considered types of "uncertainty" *and* descriptive statistics can have variance and probabilities *therefore* descriptive statistics must have uncertainty.
-The flaw in this logic comes from the first step, assuming that the tools with which we measure uncertainty *are* uncertainty in of themselves.
-This confusion is common and there are many papers that spend a great deal of time clarifying the difference.
-@Begg2014 highlight that uncertainty is related to not knowing a specific value, while variability refers to the range of values a quantity can take at different locations, times or instances.
-@Spiegelhalter2017 made sure to comment on the difference between precise random events (such as the probability associated with a coin flip), and uncertainty (such as the estimated probability associated with a coin that might be biased).
-The variance of a sample variance can be calculated and know, therefore it is not uncertain but rather it a precise description of dispersion.
-If we were to discuss drawing a new observation, or estimating the true mean of a population *then* the variance would become relevant in our discussions of uncertainty.
-
-Several authors have connected the issues around defining uncertainty to inference, however it is often discussed as a *task* or *goal* dependence.
-Multiple authors have commented on the need to consider quantifying and expressing uncertainty at every stage of a project as the "goal" shapes every step of the analysis [@Kinkeldey2014; @Hullman2016; @Refsgaard2007].
-@Otsuka2023 suggested that the process of observing data to perform statistics is largely dependent on our goals, because the process of boiling real world entities down into probabilistic objects (or "probabilistic kind" as he puts it) depends on the relationship we seek to identify with our data.
-@Meng2014 commented what is kept as data and what is tossed away is determined by the motivation of an analysis and what was previously noise can be shown to become signal depending on the the question we seek to answer.
-@Kale2019 discussed how the choices we make in our analysis impact our outcomes and introduce uncertainty.
-@Carlin2023 mentions that each research question can be can be categorised as descriptive, predictive, or causal, each of which has its own appropriate statistical methods and motivation agnostic model selection leads to statistical analysis that is devoid of meaning.
-@Wallsten1997 argue that the best method for evaluating or combining subjective probabilities depends on the uncertainty the decision maker wants to represent and why it matters.
-@Fischhoff2014 looks at uncertainty visualisation for decision making decides that we should have different ways of communicating uncertainty based off what the user is supposed to do with it.
-The importance of inference when discussing uncertainty is never directly acknowledge, but it is always present.
-It is clear that the relationship between uncertainty and inference is noticed at every stage in an analysis, however combining these uncertainties into a single "uncertainty" value is near impossible [@Spiegelhalter2017].
-
-With this understanding it becomes clear to see why uncertainty is tied to an endless string of examples in the data analysis pipeline.
-Uncertainty examples include imputed data, model selection, inherent randomness, biased sampling, etc, not because these things *are* uncertainty, but because they *create* uncertainty when we perform inference.
-Whether or not these elements are relevant is highly dependent on what statistic you are trying to draw inference on, and by extension, the purpose of your visualisation.
-
-## What is an uncertainty visualisation?
-The concept of uncertainty as a by-product of inference, not a latent feature of data, extends naturally to visualisation. 
-There are two primary reasons for making a visualisation, to perform exploratory data analysis (EDA) or communication.
-Communication involves identifying a signal we want to communicate and designing the visualisation that best conveys that, while EDA involves creating a versatile visualisation without an explicit purpose and using it to extract several signals.
-
- Since EDA is the visual parallel to descriptive statistics, it is performed without an explicit hypothesis which means there *is* no uncertainty in the visualisation.
-Similarly to our issues with descritive statistics, this is not well understood by the uncertainty visualisation community.
-Some authors simply confuse varianc for uncertainty.
-For example @Potter2010 aimed to create a summary plot that "concisely presented data with uncertainty information" to create an exploratory visualisation tool that visualised uncertainty.
-Other authors regonise inference will occur (in some shape or form) and believe uncertainty *should* be visualised but do not recognise *how* uncertainty would be visualised.
-@Hullman2021 argued that there is no such thing as a "model-free" visualisation, therefore visualisation require robust visualisations of uncertainty as we are always performing inference.
-@Griethe2006 commented that "if visualization is used as a means to explore a data volume or to communicate its contents the uncertainty has to be included".
-While we agree people cannot prevent themselves from performing inference, that does not mean it is possible for uncertainty to be included in an EDA visualisation.
-A versatile visualisation such as a scatter plot allows for a viewer to consider several hypothesis at once, each of which will be a different inferential statistic with a different distribution that depicts its uncertainty.
-It is likely impossible to suppress all possible signals at once.
-
 # How do we evaluate uncertainty visualisations?
-If asking direct questions about uncertainty causes us to treat it as a signal, how do we evaluate uncertainty as *noise*?
- When we ask the viewer of a plot to look at data and extract a value, we are asking them to perform inference on that value.
-There will be noise associated with that answer and that is uncertainty.
-If we ask direct question about some uncertainty metric, we have turned the uncertainty into signal because that is what the participants are drawing inference on.
 

### 2024-07-24 writingdumpster.qmd
reorganised "visualising uncertainy as signal" section

@@
 # Removed Sections
+@Kinkeldey2014 categorised uncertainty according to five criteria depicted in fig-tax which considers if a visualisation is implicit or explicit, intrinsic or extrinsic, visually integrable or seperable, coincident or adjacent, and static or dynamic.
+A similar version of this taxonomy was presented by @uncertchap2022 who commented that visualisation can be organised into two categories, "graphical annotations of distributional properties" and "visual encodings of uncertainty" which seems to functionally align with the intrinsic/extrinsic distinction by @Kinkeldey2014.
+@Griethe2006 organised uncertainty visualisations into two cases (1) a hypothesis test was preformed to confirm the validity of the visualisation and (2) the visualisation has uncertainty depicted.
+@Potter2012 organised several existing uncertainty visualisations into groups based on the dimensionality of the data (1D, 2D, 3D, and No Dimension) and the dimensionality of the uncertainty (Scalar, Vector, Tensor).
+However, because the term "PDF", function that is used to describe a random variable, is used to describe both the data and the uncertainty for all dimensions.
+@Grewal2021 created a taxonomy that mapped uncertainty visualisations to some point in a 2D space defined by the "domain expertise" and "continuum of discreteness" (that scaled from "point estimate" to "continuous distribution").
+
 Visualisation authors are almost unanimous in commenting that the "information" in two plots must be the same in order for the visual techniques to be compared [@Cleveland1984; @Kinkeldey2014].

### 2024-07-25 paper.qmd
edited the visualising uncertainty as signal section again

@@
 
-# Visualising uncertainty as signal
+## Visualising uncertainty as signal
 Uncertainty visualisation is not only motivated by signal-suppression, and we would be remiss to ignore these alternative motivations.
-In this section we discuss alternative motivations for uncertainty visualization, and explain why they are not the focus of this review.
-
-## Uncertainty as a variable of interest
+Some authors claim the purpose of uncertainty is to improve decision making [@Ibrekk1987; @uncertchap2022; @Hullman2016; @Cheong2016; @Boone2018; @Padilla2017].
 Other authors do not describe uncertainty as important for decision making, but rather explicitly state it as variable of importance in of itself [@Blenkinsop2000].
@@
 It is acting as signal.
+
+This is obvious for the cases where we are explicitly interested in the variance or error, as we are literally trying to draw conclusions about an "uncertainty" statistic.
+The same is true for "decision making" experiments, but it is less overt.
 This is easiest to understand with an example.
-
-Extracting information using a graphic requires a perceptual task, so it will always be less accurate than explicitly reading the value provided in text form [@Cleveland1984].
-
-This is why text representations or graphics that allow users to simply count a value to extract the required statistic outperform those that require any perceptual tasks.
-
-## Uncertainty for decision making
-Some authors claim the purpose of uncertainty is to improve decision making [@Ibrekk1987; @uncertchap2022; @Hullman2016; @Cheong2016; @Boone2018; @Padilla2017].
-This group are *also* viewing uncertainty as a signal, although it is less obvious in this case.
-
 Imagine you are trying to decide if you want to bring an umbrella with you to work.
@@
 
-This obfuscation problem translates to unanswerable questions in uncertainty evaluation experiments.
-
-If we create a problem by obfuscating information, then the representation that makes the obfuscated information easiest to extract outperforms those that don't.
-
-Additionally, if uncertainty visualisation for decision making is a "wrong information" problem, then we would expect uncertainty visualisations made for this purpose to simply adjust the statistic they display of rather than a change the visual elements of the graphic.
-This is precicely what we observe.
+Uncertainty visualisation's made for these purposes should simply display the uncertainty statistic we are interested in, such as the variance, or probability of an event.
+This is precisely what we observe.
 @fig-exceed depicts an exceedance probability map that has was designed as an alternative to the choropleth map to improve decision making under uncertainty [@Kuhnert2018; @Lucchesi2021].
-A keen viewer may notice that the "exceedance probability map" is actually just a choropleth map.
-While the input into the graphic has changed, graphics, just like statistics, are not defined by their input data.
+A keen viewer may notice that the "exceedance probability map" is actually just a choropleth map, only the statistic being displayed has changed.
+We do not believe this graphic be considered an uncertainty visualisation.
 
@@
 
-## What is an uncertainty visualisation?
-
-This seems to be an accepted fact for all visualisations *except* for uncertainty visualisation, and this creates disagreement in what is even considered to be an uncertainty visualisation.
+There seem to be two different definitions of "uncertainty visualisation" floating around in the lierature.
+The first considers *any* visualisation of error, variance, or probability to be an uncertainty visualisation.
+The second believes an "uncertainty visualisation" is the output of a function that takes a normal visualisation as an input, and transforms it to include the uncertainty information.
+The former group believe the purpose of uncertainty visualisation to provide signal about a distribution, while the later believe it should act as noise to obfuscate a signal.
+The lack of explicit distinction between these two motivations leaves the literature muddled and reviewers struggle to understand if uncertainty should be treated as a variable, as metadata, or as something else entirely [@Kinkeldey2014].
+This disagreement creates constant contradictions in what the literature considers to be an "uncertainty visualisation".
 For example @Leland2005 mentions that popular graphics, such as pie charts and bar charts omit uncertainty, and @Wickham2011 suggests their product plot framework, which includes histograms and bar charts, should be extended to include uncertainty however at least one or both of these charts are used in a significant number of uncertainty visualisation experiments [@Ibrekk1987; @Olston2002; @Zhao2023; @Hofmann2012].
-
-We belive this disagreement comes from two different conceptualizations of what an uncertainty  visualisation *is*.
-This issue causing confusion in the literature was also commented on by @Kinkeldey2014, who explained it was unclear if uncertainty should be treated as a variable, as metadata, or as something else entirely [@Kinkeldey2014].
-There are those that describe *any* visualisation of error as an "uncertainty visualisation" and those that believe "uncertainty visualisation" is a function that can be *applied* to any graphic that depicts an estimate.
-The former group view the purpose of uncertainty visualisation to provide signal about a distribution, while the later believe it should act as noise to obfuscate a signal.
-
-The easily explained results in evaluation experiments, and the exceedance probabiltiy map show the problem of viewing uncertainty as a signal instead of noise.
-If we view the purpose of viewing uncertainty visualisation to simply extract a probability or variance (reguardless of whether or not it is for value extraction or decision making) then the best performing visualsiation will be one that makes that statistic easy to extract.
-Even better is to avoid displaying a visualisation all together and simply provide the probability or variance with text.
-
-Of course, we do not beleive there is anything wrong with explicitly visualizing variance, error, bias, or any other statistic used to depict uncertainty as a signal.
+If you view uncertainty as a function applied to an existing graphic, then you would believe a pie chart and bar chart are not uncertainty visualisations, as they are yet to have the "uncertainty visualisation function" applied to them.
+If you view uncertainty as any graphic that depicts an "uncertainty statistic" then there are no limitations on which graphics can or cannot be uncertainty visualisations.
+
+When we use "uncertainty visualisaion" to refer to graphics that simply communicate a variance or probability, we are classifying  visualisations by the data they display, not their visual features.
+Graphics, just like statistics, are not defined by their input data.
+A scatter plot that compares mean and a scatter plot that compares variances are both scatter plots.
+Given that there is no special class of visualisation for *other* statistics (such the median or maximum) there is no reason to assume visualisations that simply depict a variance, error, or probability to be special.
+Some authors implicitly suggest that that visualisations of variance or probability are differentiated due to the psychological heuristics involved in interpreting uncertainty [@Hullman2019].
+While it is true that heuristics lead people to avoid uncertainty [@Spiegelhalter2017] there is no evidence that this psychological effect translates to issues with the visual representation of uncertainty.
+Again, given that we do not make these same visual considerations for other variables that elicit distaste or irrational behavior, there is no reason to assume this is what makes uncertainty visualisation so special.
+
+This leads us to the conclusion that the visualisations made for the purpose of displaying information about uncertainty statistics are not uncertainty visualisations.
+These graphics are just normal information visualisations, and authors can follow existing principles of graphical design.
+We will focus on the perspective that uncertainty visualisation serves to obfuscate signal, and an uncertainty visualisation is a variation on an existing graphic that gives it the ability to suppress false signals.
+
+Of course, we do not believe there is anything wrong with explicitly visualizing variance, error, bias, or any other statistic used to depict uncertainty as a signal.
 Just like any other statistic, these metrics provide important and useful information for analysis and decisions.
-However, there is no interesting visualisation challenge associated with these graphics, and they do not require any special visualisation, they are acting an estimate and they should be treated as such.
-
-For these reasons decision making and visualisations that explicitly depict uncertainty measures will not be the focus of this review.
-We will instead focus on the perspective that uncertainty visualisation serves to obfuscate signal, and should be considered a function that can be *applied* to any graphic to suppress false signals.
-The following sections will investigate several approaches to uncertainty visualisation and discuss how successful each approach is in fulfilling the signal-suppression criteria.
+However, there is no interesting visualisation challenge associated with these graphics, and they do not require any special visualisation.
+The uncertainty in these graphics are acting as a signal variable, and they should be treated as such.
 
@@
 @Hullman2016 commented that it is straightforward to show a value but it is much more complex to show uncertainty.
-There does not seem to be a consensus on *why* this is the case, but there are two main groups of thought that emerge when trying to justify the difficulty in uncertainty visualisation.
-
-The first group believes uncertainty visualisation are unique due to the psychological heuristics involved in interpreting uncertainty [@Spiegelhalter2017; @Hullman2019].
-While it is true that heuristics lead people to avoid uncertainty [@Spiegelhalter2017] there is no evidence that this psychological effect translates to a problem in visualisation.
-Were this the case, visualisations of death, poverty, and any other variable that elicits poor feelings would require their own field of visualisation.
+There does not seem to be a consensus on *why* this is the case.
 
@@
 
-## Uncertainty disappears if you stare directly at it
-
-# Current methods
-
-# Testing signal supression
+##  Current methods
+@Kinkeldey2014 expressed that uncertainty visualization experiments are idiosyncratic, so it is difficult to generalise the results.
+
+Extracting information using a graphic requires a perceptual task, so it will always be less accurate than explicitly reading the value provided in text form [@Cleveland1984].
+
+## The Hermann grid illusion of visualisation experiments
+A Herman grid is an optical illusion where a black and white grid appears to have grey dots, but they disappear when you look directly at them.
+This optical illusion is a fitting description of the issues in evaluating uncertainty visualisations.
+
+##  Testing signal supression
 If asking direct questions about uncertainty causes us to treat it as a signal, how do we evaluate uncertainty as *noise*?

### 2024-07-25 writingdumpster.qmd
edited the visualising uncertainty as signal section again

@@
 ![Census dot map](censusdotmap.png){#fig-census}
+
+# Statistic notes
+X
+E[X] 
+n 
+Var(X) 
+Max(Var(X))
+Min(Var(X))
+
+# Experiment notes
+
+@Sanyal2009 colored glyphs outperformed error bars in a search task study, which boils down to asking users to find the most extreme color, largest size, or longest length.
+They found people were better at finding the value with the most extreme color than the longest length.
+
+@Kale2018 found users were better at identifying if there was a general trend in a series of line plots when shown hops rather than error bars or ensembles.
+
+“growth” or “no growth” judgement about one bar chart and provided + a rating of their confidence on a scale of 50
+
+How likely is the red outcome?
+
+How likely is the outcome where can- didate A gets 55% of the vote?
+
+How likely is candi- date B to win the election?
+
+# Vis inference stuff
+objects or shapes that are close to one another appear to form groups.
+  proximity is used to place items to compare close together, and less important comparisons further apart
+
+we group things that have similar appearances and exclude objects with different appearances.
+
+elements con- tained within a common region belong together
+  confidence bands and bounding ellipses also activate this gestalt principle by grouping points within the boundaries together
+
+tendency to group objects that are moving together in the same direction and at the same speed together
+
+working memory is limited to approximately seven (plus or minus two) items, or chunks.
+
+If comparisons and quantitative evaluations made during the process of under- standing a plot. This research can be distilled into a hierarchy of comparisons 
+1. Position (common scale) 
+2. Position (nonaligned scale)
+3. Length, direction, angle, slope 
+4. Area 
+5. Volume, density, curvature 
+6. Shading, color saturation, color hue 
+7. Discriminable shape 
+8. Indiscriminable shape
+This ranking ofcognitive tasks provides some consistent guidance for chart design: If the same
+data can be represented in a way that allows the user to make a comparison more accurately (based on the hierarchy), then that design is preferable.

### 2024-07-25 paper.qmd
FINISHED visualisation as a variable section

@@
 
-## Visualising uncertainty as signal
+## Uncertainty as a signal
 Uncertainty visualisation is not only motivated by signal-suppression, and we would be remiss to ignore these alternative motivations.
@@
 
-There seem to be two different definitions of "uncertainty visualisation" floating around in the lierature.
+There seem to be two different definitions of "uncertainty visualisation" floating around in the literature.
 The first considers *any* visualisation of error, variance, or probability to be an uncertainty visualisation.
@@
 
-When we use "uncertainty visualisaion" to refer to graphics that simply communicate a variance or probability, we are classifying  visualisations by the data they display, not their visual features.
+When we use "uncertainty visualization" to refer to graphics that simply communicate a variance or probability, we are classifying  visualisations by the data they display, not their visual features.
 Graphics, just like statistics, are not defined by their input data.
@@
 
-@fig-bivariate a variation of the choropleth map, where we have a two dimensional colour palette.
+@fig-bivariate a variation of the choropleth map, where we have a two dimensional color palette.
 Not only is temperature mapped to hue, but variance is also mapped to saturation.
@@
 @Hullman2016 commented that it is straightforward to show a value but it is much more complex to show uncertainty.
-There does not seem to be a consensus on *why* this is the case.
-
-The second group sees uncertainty as a high dimensional visualisation problem as the difficulty is in working out how to add uncertainty into already existing graphics [@moritz2017trust; @Griethe2006].
-If we view uncertainty visualisation as a high dimension problem we include uncertainty as we would any other variable and use high-dimension visualisation techniques to incorporate it.
-This is the approach taken in @fig-bivariate.
-
-The problem with the high dimension approach is that it treats uncertainty the same as we would any other variable.
-Typically we want visual channels to be separable, that is, the data represented in one visual channel does not interfere with the others [@Smart2019].
-Unfortunately, this means that applying normal visualisation methods to "uncertainty visualisation" will prevent uncertainty and signal being read together as a "single integrated uncertain value" [@Kinkeldey2014], which is the goal of signal-suppression.
-Separating the uncertainty and the signal also allows for the uncertainty to be ignored, which is a common problem faced by most uncertainty visualisations [@uncertchap2022].
-We can see this problem in @fig-bivariate, as it sends the message "this data has a spatial trend and the estimates have a large variance" rather than hiding the statistically invalid signal.
-
-If thses high-dimensional approaches were to work, the uncertainty and the estimates would need to be mapped to integrable channels.
-That is, the channel that uncertainty and the estimates are mapped to would need to be separately manipulable, but read as a single channel by the human brain.
-While most aesthetics *are* seperable, there are some variables that have been shown to be integrable, such as colour hue and brightness [@Vanderplas2020].
-We can more effectively approach this method by also leveraging visual semiotics and mapping more uncertain values to lighter colors [@Maceachren2012].
+This difficulty is often described as a high-dimensional visualisation problem, as the difficulty in uncertainty visualisation is in working out how to add uncertainty into already existing graphics [@moritz2017trust; @Griethe2006].
+The problem with this approach to uncertainty visualisation is that it treats uncertainty the same as we would any other variable, but we are not interested in visualising uncertainty as though it were any other variable.
+
+Typically, when making visualisations, we want the visual channels to be separable, that is, we don't want the data represented through one visual channel to interfere with the others [@Smart2019].
+This is not the goal of uncertainty visualisation.
+The goal of uncertainty visualisations is to display signal and uncertainty together as a "single integrated uncertain value" [@Kinkeldey2014].
+Separability allows for the signal and the uncertainty to be read separately which allows for the uncertainty information to simply be ignored when reading the signal, which is a pervasive issue in current uncertainty visualisation methods [@uncertchap2022].
+We can see this problem in @fig-bivariate, as it sends the message "this data has a spatial trend and the estimates have a large variance" as we read the signal and the uncertainty separately.
+
+The need for interference from the uncertainty channel does not necessarily mean high-dimensional visualisation techniques would be useless for uncertainty visualisation.
+High-dimensional visualisation methods could work if the uncertainty and the estimates were mapped to integrable channels.
+That is, the visual channels would need to be separately manipulable, but read as a single channel by the human brain.
+While most visual aesthetics *are* separable, there are some variables that have been shown to be integrable, such as color hue and brightness [@Vanderplas2020].
+When visualising uncertainty using its own visual channel, we can also consider visual semiotics and make sure to map uncertainty to intuitive visual channels, such as mapping more uncertain values to lighter colors [@Maceachren2012].
 
@@
 #| warning: false
+#| eval: false
 #| label: fig-bivariate2
@@
 #| warning: false
-#| eval: false
 #| label: fig-transparency
+#| fig-cap: "A bivariate map that depict the counties of Iowa where each county is coloured acording to it's average daily temperature and variance.
+This map is a variation on the previous bivariate map where instead of variance being mapped to colour saturation, it is mapped to transparency."
+#| fig-subcap: 
+#|   - "High Variance Data"
+#|   - "Low Variance Data"
+#|   - "Transparency Palette"
 #| layout-ncol: 3
@@
 # ALTERNATIVE TRANSPARENCY INSTEAD OF COLOUR VALUE
-
-trans_map <- my_map_data |>
-  mutate(scaled_var = (variance - min(variance))/(max(variance)-min(variance)))
-
-# Transparency map
-pta <- trans_map |>
-  filter(variance_class=="lowvar") |>
+# get colour values from transparency transformation
+transpal <- lapply(basecols, scales::alpha, alpha=c(0.95, 0.65, 0.35, 0.05))
+
+transpal <- as.vector(matrix(unlist(x), ncol = 4, byrow = TRUE))
+
+names(transpal) <- paste(rep(1:8, 4), "-" , rep(1:4, each=8), sep="")
+
+# Transparency maps
+pta <- my_map_data |>
+  filter(variance_class=="lowvar") %>%
   ggplot() +
-  geom_sf(aes(fill = bitemp, 
-              geometry = geometry,
-              alpha=scaled_var)) + 
-  scale_fill_manual(values = basecols) +
+  geom_sf(aes(fill = biclass, geometry = geometry)) + 
+  scale_fill_manual(values = transpal) +
   theme_void() + 
@@
 
-ptb <- pta %+% filter(trans_map, variance_class=="highvar")
+ptb <- pta %+% filter(my_map_data, variance_class=="highvar")
+
+pta
+ptb
+show_pal2(colours = transpal, ncol=8, borders=NA, myxlab = "Temperature", myylab = "Variance", breaks = 21:29, breaks2 = 0:4)
+
 ```
 
-@fig-bivariate2 shows a bivariate color map where uncertainty is mapped to color value, and temperature is mapped to color hue.
-It is clear the spatial trend is still visible in the high uncertainty case.
-This does not necessarily mean integrability will never be a useful tool for signal suppression, but without a strong understanding of how these visual channels collapse down into a single channel, it may not be a reliable option.
-Relying on integrability could create unintended consequences of uncertainty visualization, such as displaying phantom signals or hiding justified signals.
+@fig-transparency is an example of a variations of @fig-bivariate where uncertainty is mapped to transparency, and temperature is mapped to color hue to leverage these visualisation concepts.
+This method achieves signal suppression quite well.
+The spatial trend is clearly visible in the low variance case and that trend it becomes much harder to identify in the high variance case.
+While this was an effective approach for this graphic, relying on integrability may not give us the amount of control we want over our signal-supresison.
+Without a strong understanding of how these visual channels collapse down into a single channel, relying on integrability could create unintended consequences such as displaying phantom signals or hiding justified signals.
 Additionally, multi-dimensional colour palettes can make the graphics harder to read and hurt the accessibility of the plots [@Vanderplas2015].
-A more viable option might be to warp the space uncertainty and signal are mapped to ourselves.
+
+There is another reason @fig-transparency is better at signal-suppression than @fig-bivariate, and it may not be due to integrability.
+Colour value has a second desirable quality for signal-supression, which is that the colours become harder to distinguish as the value decreases.
+This means high uncertainty values are harder to differentiate than low uncertainty values.
+This implicit feature of colour value can generalised to other aesthetics by transforming the visual feature space ourselves.
 
@@
 p3b
-show_col(VSUP, ncol=8, labels=FALSE)
+show_pal2(colours = VSUP, ncol=8, borders=NA, myxlab = "Temperature", myylab = "Variance", breaks = 21:29, breaks2 = 0:4)
 

### 2024-07-25 paper.qmd
Made some final changes to ""visualising uncertainty as signal" and started "transforming space" section

@@
 - May not need to reference the same paper again and again
-- I can do a more formal proof of the decision making thing using decision science. I left it out but it can be added if that section needs to be more thorough.
-- fix the colour value map
+- examples that aren't maps? Sure are a lot of maps.
 -->
@@
 
+## The choropleth map
 @fig-choropleth depicts a choropleth map of the counties of Iowa.
@@
 
+## The bivariate map
 @fig-bivariate a variation of the choropleth map, where we have a two dimensional color palette.
@@
 
-p2a1
-p2b2
-show_pal2(colours = bivariatepal2, ncol=8, borders=NA, myxlab = "Temperature", myylab = "Variance", breaks = 21:29, breaks2 = 0:4)
+#p2a1
+#p2b2
+#show_pal2(colours = bivariatepal2, ncol=8, borders=NA, myxlab = "Temperature", myylab = "Variance", breaks = 21:29, breaks2 = 0:4)
 
@@
 
-transpal <- as.vector(matrix(unlist(x), ncol = 4, byrow = TRUE))
+transpal <- as.vector(matrix(unlist(transpal), ncol = 4, byrow = TRUE))
 
@@
 # Visualising uncertainty and signal in a new space
-Instead of hoping that uncertainty might collapse signal values into a single dimension, we can do some of that work ourselves.
-
-by using a VSUP which collapses the colour space such that high uncertainty values cannot be extracted.
-It is unclear how useful VSUPs are at actually combining signal and noise and therefore suppressing plot level insights, as they have only been tested on simple value extraction tasks that require evaluating a single point [@Correll2018; @Ndlovu2023] rather than looking for spatial relationships (which is arguably what maps are for).
-Following along with this trend, the next way we might consider visualising uncertainty is to combine uncertainty and signal at the earlier stage so the "suppressed signal" is represented by a single variable.
-This statistic can then be expressed in a one dimensional colour space, which is a method adopted by the Bayesian surprise metric map [@Ndlovu2023] and the exceedance probability map [Lucchesi2017].
-
-This problem has been noticed by others in the field, that typically use this method (specifically in the spatial uncertainty context) and a desire for representations that integrate uncertainty and signal is one of the reasons for the invention of the value-suppressing uncertainty pallet [@Correll2018].
-
-Value-suppressing uncertainty pallets (VSUP) were developed as a method that would allow the signal and the noise to be interpreted together such that insights gained by the viewers of a plot are appropriately *suppressed* by the uncertainty.
-Hence the name of the pallet.
-depicted this map colouring approach and several other extensions on the typical choropleth map that differ in where in the visualisation process they combine the noise and signal information into a single concept of "valid signal".
+Instead of hoping that uncertainty might collapse signal values into a single dimension, we can do some of that work ourselves, and uncertainty visualisation authors already have.
 
 ### Value Supressing Uncertainty Palettes
+The Value Suppressing Uncertainty Palette (VSUP) [@Correll2018], was designed with the intention of preventing high uncertainty values from being extracted from a map.
+Since the palette was designed with the extraction of individual values in mind and it has only been tested on simple value extraction tasks [@Correll2018] or search tasks [@Ndlovu2023], it is unclear how effective the method is at suppressing broader insights such as spatial trends.
+
+@fig-vsup is a visualisation of the Iowa temperature data using a VSUP to color the counties.
+The low uncertainty case still has a visible spatial trend, while the spatial trend in the high uncertainty map has functionally disappeared.
+This means the VSUP has successfully performed signal suppression for our data.
 
@@
 #| fig-cap: "A map made with a VSUP.
-The counties of Iowa are coloured acording to it's average daily temperature and the variance in temperature, although the space they have been mapped to is slightly warped."
+The counties of Iowa are coloured acording to it's average daily temperature and the variance in temperature, although the space they have been mapped to is slightly warped.
+Similar to the bivariate map, temperature is mapped to hue while variance is mapped to saturation.
+This map successfully reduces the visibility of the spatial trend in the high uncertainty case while maintaining the visibility of the spatial trend in the low uncertainty case."
 #| fig-subcap: 
@@
 
-### Uncertainty and inference
-The ramifications of this is seen throughout the literature, as a survey of visualisation authors cited "not knowing how to calculate uncertainty" as one of the primary reasons they did not include it in visualisations [@Hullman2020a].
-Some works [@Hullman2018; @Maceachren2012; @Thomson2005] get around this definition issue by focusing narrowly on specific terms defined mathematically, such as probability, variance, error, or precision.
+Now that we have successfully suppressed our invalid signal, we must return to the original criteria we had for signal supression and ask ourselves if we meet all of them.
+We want to make sure that in our efforts to suppress invalid signals, we have not also suppressed statistically *valid* signals in the process.
+
+## Can we have uncertinty visualisations for EDA?
+A slight problem with uncertainty visualisation is that the uncertainty and the purpose of visualisation are somewhat at odds with one another.
+This problem becomes much more explicit through the use of the VSUP.
+
+### What is uncertainty?
+There does not seem to be a clear agreed upon definition of uncertainty which would obviously create some issues when we want to visualize it.
+You cant measure something you cant define, you cant quantify something you cant measure, and you cant visualise something you cant quantify.
+
+The ramifications of this problem is seen repeatedly in the uncertaint visualisation literature.
+A survey of visualisation authors cited "not knowing how to calculate uncertainty" as one of the primary reasons they did not include it in visualisations [@Hullman2020a].
+Some works [@Hullman2018; @Maceachren2012; @Thomson2005] ignore the problem get around this definition issue by focusing narrowly on specific terms defined mathematically, such as probability, variance, error, or precision.
 Others [@Griethe2006; @Pang1997; @Pham2009; @Boukhelifa2017] include broader loosely related elements, such as missing values, but ignore these concepts when translating that uncertainty to a visualisation.

### 2024-07-25 writingdumpster.qmd
Made some final changes to ""visualising uncertainty as signal" and started "transforming space" section

@@
 # Removed Sections
+After all, if we consider a choropleth map, which ignores uncertainty entirely to be at one end of the spectrum (it is a map that only cares about showing signals reguardless of whether or not they are valid) then a completely colourless map is at the other end of the spectrum (it is a map that only cares about suppressing signals regardless of whether or not they are valid).
+
 @Kinkeldey2014 categorised uncertainty according to five criteria depicted in fig-tax which considers if a visualisation is implicit or explicit, intrinsic or extrinsic, visually integrable or seperable, coincident or adjacent, and static or dynamic.

### 2024-07-26 paper.qmd
meeting update

@@
 #| label: fig-bivariate2
-#| fig-cap: "A bivariate map that depict the counties of Iowa where each county is coloured acording to it's average daily temperature and the variance in temperature.
+#| layout-ncol: 3
+#| layout-valign: "bottom"
+
+#fig-cap: "A bivariate map that depict the counties of Iowa where each county is coloured acording to it's average daily temperature and the variance in temperature.
 This map is a variation on the previous bivariate map where instead of variance being mapped to colour saturation, it is mapped to colour value.
 Even though colour value is intergrable with colour hue, and colour value is naturally accociated with uncertainty, the spatial trend is still visible in the map."
-#| fig-subcap: 
-#|   - "High Variance Data"
-#|   - "Low Variance Data"
-#|   - "Bivariate Palette"
-#| layout-ncol: 3
-#| layout-valign: "bottom"
 
@@
   ggplot() +
-  geom_sf(aes(fill = biclass, geometry = geometry)) + 
+  geom_sf(aes(fill = biclass, geometry = geometry), colour=NA) + 
   scale_fill_manual(values = bivariatepal2) +
@@
 
-#p2a1
-#p2b2
+p2a1
+p2b2
 #show_pal2(colours = bivariatepal2, ncol=8, borders=NA, myxlab = "Temperature", myylab = "Variance", breaks = 21:29, breaks2 = 0:4)

### 2024-07-29 paper.qmd
made small map changes from meeting

@@
   geom_sf(aes(fill = bitemp, 
-              geometry = geometry)) + 
+              geometry = geometry), colour=NA) + 
   scale_fill_manual(values = basecols) +
@@
        col = colours, border = borders)
-  text(c(0,col(colours)), -c(1,row(colours))-0.15, breaks, 
+  text(c(0,col(colours)) + c(0.2, 0.1, 0,0,0,0,0,-0.1,-0.2), -c(1,row(colours))-0.25, breaks, 
               cex = 1, col = "black")
-  text(4, -1.55, myxlab ,cex = 1.5, col = "black")
+  text(4, -1.75, myxlab ,cex = 1.5, col = "black")
 }
@@
   geom_sf(aes(fill = xprob, 
-              geometry = geometry)) + 
+              geometry = geometry), colour=NA) + 
   scale_fill_manual(values = basecols) +
@@
   ggplot() +
-  geom_sf(aes(fill = biclass, geometry = geometry)) + 
+  geom_sf(aes(fill = biclass, geometry = geometry), colour=NA) + 
   scale_fill_manual(values = bivariatepal) +
@@
   ggplot() +
-  geom_sf(aes(fill = biclass, geometry = geometry)) + 
+  geom_sf(aes(fill = biclass, geometry = geometry), colour=NA) + 
   scale_fill_manual(values = transpal) +
@@
   ggplot() +
-  geom_sf(aes(fill = biclass, geometry = geometry)) + 
+  geom_sf(aes(fill = biclass, geometry = geometry), colour=NA) + 
   scale_fill_manual(values = VSUP) +

### 2024-07-29 paper.qmd
EDA section

@@
 ## Can we have uncertinty visualisations for EDA?
-A slight problem with uncertainty visualisation is that the uncertainty and the purpose of visualisation are somewhat at odds with one another.
-This problem becomes much more explicit through the use of the VSUP.
+A slight problem with uncertainty visualisation is that the uncertainty and the purpose of visualisation are somewhat at odds with one another. The VSUP highlights this problem quite well.
+
+There are two primary motivations behind visualisation, communication and exploratory data analysis (EDA).
+Communication involves identifying a signal we want to communicate and designing a visualisation that best conveys that, while EDA involves creating a versatile visualisation using it to extract several signals.
+It is not uncommon for authors to express a desire for uncertainty visualisations that perform signal-suppression in visualisations made for EDA [@Sarma2024; @Griethe2006], however it is not clear that these uncertainty visualisations *exist at all*.
+
+The impossible task of making this graphic is easy to see with a simple example.
+Let's say we have a graphic that depicts a set of coefficients from a linear regression and the value of the coefficient is shown using a single colour.
+We do a series of t-tests on these estimates, and all of them fail to reject the null hypothesis that they are different from 0.
+This would indicate that each value needs to be visualised using a single color that is indistinguishable from the colour assigned to 0.
+However, we also perform an f-test on our model, and this test *is* rejected.
+This test indicates that the set of estimates should not be visually indistinguishable from 0.
+This is clearly a contradiction.
+
+This t-test, f-test example highlights a fundamental problem with the VSUP and even the bivariate map.
+When we blend colors to make them visually indistinguishable, we need to decide at what level of *uncertainty* to blend these colors together.
+There is no intrinsic mapping of "uncertainty", so we need to make a decision as to how much "suppressing" the visualisation does.
+We did this implicitly in the bivarate map by deciding how much "variance"
+
+- we could consider this a power thing?
+
+Using the methods discussed thus far, where each cell is colours by a single value, it would be impossible to succeed in this method. However, this may not be true if we change our approach entirely.
 
@@
 
-The ramifications of this problem is seen repeatedly in the uncertaint visualisation literature.
+The ramifications of this problem is seen repeatedly in the uncertainty visualisation literature.
 A survey of visualisation authors cited "not knowing how to calculate uncertainty" as one of the primary reasons they did not include it in visualisations [@Hullman2020a].

### 2024-07-30 paper.qmd
added comment to eda
FInished eda section

@@
 When we blend colors to make them visually indistinguishable, we need to decide at what level of *uncertainty* to blend these colors together.
-There is no intrinsic mapping of "uncertainty", so we need to make a decision as to how much "suppressing" the visualisation does.
-We did this implicitly in the bivarate map by deciding how much "variance"
-
-- we could consider this a power thing?
-
-Using the methods discussed thus far, where each cell is colours by a single value, it would be impossible to succeed in this method. However, this may not be true if we change our approach entirely.
-
-### What is uncertainty?
-There does not seem to be a clear agreed upon definition of uncertainty which would obviously create some issues when we want to visualize it.
-You cant measure something you cant define, you cant quantify something you cant measure, and you cant visualise something you cant quantify.
-
-The ramifications of this problem is seen repeatedly in the uncertainty visualisation literature.
-A survey of visualisation authors cited "not knowing how to calculate uncertainty" as one of the primary reasons they did not include it in visualisations [@Hullman2020a].
-Some works [@Hullman2018; @Maceachren2012; @Thomson2005] ignore the problem get around this definition issue by focusing narrowly on specific terms defined mathematically, such as probability, variance, error, or precision.
-Others [@Griethe2006; @Pang1997; @Pham2009; @Boukhelifa2017] include broader loosely related elements, such as missing values, but ignore these concepts when translating that uncertainty to a visualisation.
-Existing mathematical definition of uncertainty (or related concepts) have been unable to unify this range of diverse concepts.
-For example, @Thomson2005 suggests a mathematical formula for *examples* of uncertainty, and information theory tries to quantify uncertainty using the idea of entropy, but the disconnect between the broad concept of uncertainty and what we can reliably quantify remains a problem.
-
-Some authors prefer to define uncertainty in qualitative terms, often through the use of taxonomies.
-In these cases, uncertainty is split using an endless stream of ever changing boundaries, such as whether the uncertainty is due to true randomness or a lack of knowledge [@Spiegelhalter2017; @Hullman2016; @utypo], if the uncertainty is in the attribute, spatial elements, or temporal element of the data [@Kinkeldey2014], whether the uncertainty is scientific (e.g.
-error) or human (e.g.
-disagreement among parties) [@Benjamin2018], if the uncertainty is random or systematic [@Sanyal2009], statistical or bounded [@Gschwandtnei2016; @Olston2002], recorded as accuracy or precision [@Griethe2006; @Benjamin2018], which stage of the data analysis pipeline the uncertainty comes from [@utypo], how quantifiable the uncertainty is [@Spiegelhalter2017; @utypo], etc.
-In their own way, each of these approaches show an aspect of uncertainty that an author felt was important to differentiate.
-
-What is interesting about all these views of uncertainty, is that they all imply that it is a latent feature of our data.
-That is, it is something that can be revealed through exploration, and not something that needs to be explicitly calculated.
-The constant stream of new definitions that are unable to capture the true essence of uncertainty might be a result of this false assumption that underpins ever definition.
-Uncertainty cannot be defined without a signal because uncertainty does not exist without a signal, is a by-product of inference.
-
-It is easy to see uncertainties relationship to inference when we consider what might *not* be considered uncertain rather than just trying to think about what *is*.
-
-Descriptive statistics simply describe our sample as it is and summarizes large data down into an easy to swallow format.
-Descriptive statistics are not seen as the primary goal of modern statistics, however, this was not always the case.
-Around the 19th century in England, *positivism* was the popular philosophical approach to science (positivists included famous statisticians such as Francis Galton and Karl Pearson) and practitioners of the approach believed statistics ended with descriptive statistics as science must be based on actual experience and observations, therefore anything that refers to the unobservable (such as new observations or population statistics) is not true science [@Otsuka2023].
-In order to make statements about population statistics, future values, or new observations we need to perform inference, which requires the assumption of the "uniformity of nature" (i.e.
-that unobserved phenomena should be similar to observed phenomena) [@Otsuka2023].
-This subtle shift, from descriptive statistics to inferential statistics was shunned during the positivism era *due to the fact it introduced the unknowable*, or in other words, uncertainty.
-
-This approach to uncertainty is embedded in the idea that descriptive statistics do not have uncertainty, which some readers may disagree with.
-Specifically, because it means uncertainty is *not* a latent attribute of data, but rather an attribute of a specific hypothesis or estimate.
-Deniers of this fact follow a consistent logical path and it is easy to identify the common mistake.
-We know that variance and probability are typically considered types of "uncertainty" *and* descriptive statistics can have variance and probabilities *therefore* descriptive statistics must have uncertainty.
-The flaw in this logic comes from the first step, assuming that the tools with which we measure uncertainty *are* uncertainty in of themselves.
-This confusion is common and there are many papers that spend a great deal of time clarifying the difference.
-@Begg2014 highlight that uncertainty is related to not knowing a specific value, while variability refers to the range of values a quantity can take at different locations, times or instances.
-@Spiegelhalter2017 made sure to comment on the difference between precise random events (such as the probability associated with a coin flip), and uncertainty (such as the estimated probability associated with a coin that might be biased).
-The variance of a sample variance can be calculated and know, therefore it is not uncertain but rather it a precise description of dispersion.
-If we were to discuss drawing a new observation, or estimating the true mean of a population *then* the variance would become relevant in our discussions of uncertainty.
-
-With this understanding it becomes clear to see why uncertainty is tied to an endless string of examples in the data analysis pipeline.
-Uncertainty examples include imputed data, model selection, inherent randomness, biased sampling, etc, not because these things *are* uncertainty, but because they *create* uncertainty when we perform inference.
-Whether or not these elements are relevant is highly dependent on what statistic you are trying to draw inference on, and by extension, the purpose of your visualisation.
-
-As we move through these methods, it seems that the validity of any overarching insight becomes more visible at the cost of our ability to extract particular values of signal or noise.
-Therefore, given that the primary goal of visualisation *is* insights [@North2006], visualisation authors should err on the side of representing suppressed signal as a single variable, rather than visualising uncertainty separately using two different channels.
-
-The concept of uncertainty as a by-product of inference, not a latent feature of data, extends naturally to visualisation. 
-There are two primary reasons for making a visualisation, to perform exploratory data analysis (EDA) or communication.
-Communication involves identifying a signal we want to communicate and designing the visualisation that best conveys that, while EDA involves creating a versatile visualisation without an explicit purpose and using it to extract several signals.
-
- Since EDA is the visual parallel to descriptive statistics, it is performed without an explicit hypothesis which means there *is* no uncertainty in the visualisation.
-Similarly to our issues with descritive statistics, this is not well understood by the uncertainty visualisation community.
-Some authors simply confuse varianc for uncertainty.
-For example @Potter2010 aimed to create a summary plot that "concisely presented data with uncertainty information" to create an exploratory visualisation tool that visualised uncertainty.
-Other authors regonise inference will occur (in some shape or form) and believe uncertainty *should* be visualised but do not recognise *how* uncertainty would be visualised.
-@Hullman2021 argued that there is no such thing as a "model-free" visualisation, therefore visualisation require robust visualisations of uncertainty as we are always performing inference.
-@Griethe2006 commented that "if visualization is used as a means to explore a data volume or to communicate its contents the uncertainty has to be included".
-While we agree people cannot prevent themselves from performing inference, that does not mean it is possible for uncertainty to be included in an EDA visualisation.
-A versatile visualisation such as a scatter plot allows for a viewer to consider several hypothesis at once, each of which will be a different inferential statistic with a different distribution that depicts its uncertainty.
-It is likely impossible to suppress all possible signals at once.
+There is no intrinsic mapping of uncertainty, so we need to make a decision as to how much signal-suppression the visualisation should perform.
+There is no existing rule or accepted practice in how to go about this, but it is present in every uncertainty graphic.
+Even though the bivariate map was not explicitly combining temperature values at certain variance levels, the mapping of variance to color saturation does this implicitly.
+That is, at certain saturation values the colors in a bivariate map are imperceptibly different and appear as though they are mapped to the same value, regardless of whether or not that is actually the case.
+
+Regardless as to whether the map colors are implicitly or explicitly indistinguishable, the point at which the colors blend together can only be validated by a single hypothesis.
+Therefore, using the methods discussed thus far, it would be impossible to design signal-suppression for EDA.
+If we are designing a graphic for communication then we can suppress the signal we were seeking to communicate, but this would rely on the distinction between EDA and communication to be somewhat clear cut.
+Visualization authors may design a graphic with a particular purpose, but what the visualization *is actually* used for is determined by the viewer and some authors have discussed the validity of there being any distinction between visualisation motivation's at all [@Hullman2021].
+However, this problem only exists when we color each cell using a single color.
+If we can color a cell using multiple colors, this limitation may disappear entirely.
 
@@
 # Implicitly Combining Uncertainty and Signal
+
 ## Pixel maps
+
 ```{r}
@@
 ```
+
+## What is uncertainty?
+
 ## Show me the data
+@Kim2019 looked at how different uncertainty visualisation techniques influenced the percieved sample size
 
@@
 
+### What is uncertainty?
+There does not seem to be a clear agreed upon definition of uncertainty which would obviously create some issues when we want to visualize it.
+You cant measure something you cant define, you cant quantify something you cant measure, and you cant visualise something you cant quantify.
+
+The ramifications of this problem is seen repeatedly in the uncertainty visualisation literature.
+A survey of visualisation authors cited "not knowing how to calculate uncertainty" as one of the primary reasons they did not include it in visualisations [@Hullman2020a].
+Some works [@Hullman2018; @Maceachren2012; @Thomson2005] ignore the problem get around this definition issue by focusing narrowly on specific terms defined mathematically, such as probability, variance, error, or precision.
+Others [@Griethe2006; @Pang1997; @Pham2009; @Boukhelifa2017] include broader loosely related elements, such as missing values, but ignore these concepts when translating that uncertainty to a visualisation.
+Existing mathematical definition of uncertainty (or related concepts) have been unable to unify this range of diverse concepts.
+For example, @Thomson2005 suggests a mathematical formula for *examples* of uncertainty, and information theory tries to quantify uncertainty using the idea of entropy, but the disconnect between the broad concept of uncertainty and what we can reliably quantify remains a problem.
+
+Some authors prefer to define uncertainty in qualitative terms, often through the use of taxonomies.
+In these cases, uncertainty is split using an endless stream of ever changing boundaries, such as whether the uncertainty is due to true randomness or a lack of knowledge [@Spiegelhalter2017; @Hullman2016; @utypo], if the uncertainty is in the attribute, spatial elements, or temporal element of the data [@Kinkeldey2014], whether the uncertainty is scientific (e.g.
+error) or human (e.g.
+disagreement among parties) [@Benjamin2018], if the uncertainty is random or systematic [@Sanyal2009], statistical or bounded [@Gschwandtnei2016; @Olston2002], recorded as accuracy or precision [@Griethe2006; @Benjamin2018], which stage of the data analysis pipeline the uncertainty comes from [@utypo], how quantifiable the uncertainty is [@Spiegelhalter2017; @utypo], etc.
+In their own way, each of these approaches show an aspect of uncertainty that an author felt was important to differentiate.
+
+What is interesting about all these views of uncertainty, is that they all imply that it is a latent feature of our data.
+That is, it is something that can be revealed through exploration, and not something that needs to be explicitly calculated.
+The constant stream of new definitions that are unable to capture the true essence of uncertainty might be a result of this false assumption that underpins ever definition.
+Uncertainty cannot be defined without a signal because uncertainty does not exist without a signal, is a by-product of inference.
+
+It is easy to see uncertainties relationship to inference when we consider what might *not* be considered uncertain rather than just trying to think about what *is*.
+
+Descriptive statistics simply describe our sample as it is and summarizes large data down into an easy to swallow format.
+Descriptive statistics are not seen as the primary goal of modern statistics, however, this was not always the case.
+Around the 19th century in England, *positivism* was the popular philosophical approach to science (positivists included famous statisticians such as Francis Galton and Karl Pearson) and practitioners of the approach believed statistics ended with descriptive statistics as science must be based on actual experience and observations, therefore anything that refers to the unobservable (such as new observations or population statistics) is not true science [@Otsuka2023].
+In order to make statements about population statistics, future values, or new observations we need to perform inference, which requires the assumption of the "uniformity of nature" (i.e.
+that unobserved phenomena should be similar to observed phenomena) [@Otsuka2023].
+This subtle shift, from descriptive statistics to inferential statistics was shunned during the positivism era *due to the fact it introduced the unknowable*, or in other words, uncertainty.
+
+This approach to uncertainty is embedded in the idea that descriptive statistics do not have uncertainty, which some readers may disagree with.
+Specifically, because it means uncertainty is *not* a latent attribute of data, but rather an attribute of a specific hypothesis or estimate.
+Deniers of this fact follow a consistent logical path and it is easy to identify the common mistake.
+We know that variance and probability are typically considered types of "uncertainty" *and* descriptive statistics can have variance and probabilities *therefore* descriptive statistics must have uncertainty.
+The flaw in this logic comes from the first step, assuming that the tools with which we measure uncertainty *are* uncertainty in of themselves.
+This confusion is common and there are many papers that spend a great deal of time clarifying the difference.
+@Begg2014 highlight that uncertainty is related to not knowing a specific value, while variability refers to the range of values a quantity can take at different locations, times or instances.
+@Spiegelhalter2017 made sure to comment on the difference between precise random events (such as the probability associated with a coin flip), and uncertainty (such as the estimated probability associated with a coin that might be biased).
+The variance of a sample variance can be calculated and know, therefore it is not uncertain but rather it a precise description of dispersion.
+If we were to discuss drawing a new observation, or estimating the true mean of a population *then* the variance would become relevant in our discussions of uncertainty.
+
+With this understanding it becomes clear to see why uncertainty is tied to an endless string of examples in the data analysis pipeline.
+Uncertainty examples include imputed data, model selection, inherent randomness, biased sampling, etc, not because these things *are* uncertainty, but because they *create* uncertainty when we perform inference.
+Whether or not these elements are relevant is highly dependent on what statistic you are trying to draw inference on, and by extension, the purpose of your visualisation.
+
+As we move through these methods, it seems that the validity of any overarching insight becomes more visible at the cost of our ability to extract particular values of signal or noise.
+Therefore, given that the primary goal of visualisation *is* insights [@North2006], visualisation authors should err on the side of representing suppressed signal as a single variable, rather than visualising uncertainty separately using two different channels.
+
+The concept of uncertainty as a by-product of inference, not a latent feature of data, extends naturally to visualisation. 
+There are two primary reasons for making a visualisation, to perform exploratory data analysis (EDA) or communication.
+Communication involves identifying a signal we want to communicate and designing the visualisation that best conveys that, while EDA involves creating a versatile visualisation without an explicit purpose and using it to extract several signals.
+
+ Since EDA is the visual parallel to descriptive statistics, it is performed without an explicit hypothesis which means there *is* no uncertainty in the visualisation.
+Similarly to our issues with descritive statistics, this is not well understood by the uncertainty visualisation community.
+Some authors simply confuse varianc for uncertainty.
+For example @Potter2010 aimed to create a summary plot that "concisely presented data with uncertainty information" to create an exploratory visualisation tool that visualised uncertainty.
+Other authors regonise inference will occur (in some shape or form) and believe uncertainty *should* be visualised but do not recognise *how* uncertainty would be visualised.
+@Hullman2021 argued that there is no such thing as a "model-free" visualisation, therefore visualisation require robust visualisations of uncertainty as we are always performing inference.
+@Griethe2006 commented that "if visualization is used as a means to explore a data volume or to communicate its contents the uncertainty has to be included".
+While we agree people cannot prevent themselves from performing inference, that does not mean it is possible for uncertainty to be included in an EDA visualisation.
+A versatile visualisation such as a scatter plot allows for a viewer to consider several hypothesis at once, each of which will be a different inferential statistic with a different distribution that depicts its uncertainty.
+It is likely impossible to suppress all possible signals at once.
+
 # How do we evaluate uncertainty visualisations?
@@
 
-Additionally, setting up an uncertainty visualisation where the participants are expected to notice the signal once the data behind the visualisation passes a hypothesis test implies the signal *is* noticeable to a human at that level.
-@Patrick2023 compared people ability to recognise patterns in a residual plot to typical statistical tests and found human viewers looking at a plot were less sensitive than the typical residual tests.
-These experiments utilised the line-up protocol which has been suggested as a method to check if perceived patterns are real or merely the result of chance [@Buja2009; @Wickham2010; @Chowdhury].
-This concept bears similarity to the goal of uncertainty visualisation, but it is not quite the same.
-fig-hypvs shows the conceptual difference between the line-up protocol and uncertainty visualisation.
+@Patrick2023 found that human viewers were less sensitive to deviations from the null hypothesis than the typical statistical tests.
+
+What this all means, is that there will always be some trade off when we are designing a visualisation for signal-suppression.
+There will be a trade off in *which* signal we suppress, as well as *how much* we suppress it.
+This is true for any graphic that is comparable in some way to a statistical test or process.
+A good example is the line-up protocol, which is a visualisation tool that used to check if perceived patterns are real or merely the result of chance [@Buja2009; @Wickham2010].
+This motivation is similar to our signal-suppression goal.
+@Patrick2023 compared standard statistical tests to the lineup-protocol, and evaluated the visualisations using the power curves that are typical for hypothesis testing.
+In a similar vein, @Kim2019 investigated how different uncertainty visualisation methods influenced user's prior beliefs, and evaluated the graphics by comparing their results to those from Bayesian inference.
+While these methods apply different statistical philosophies, they both investigated the sensitivity of visualisation methods relative to existing evaluation methods.
+This approach may prove useful when considering
 

### 2024-07-31 paper.qmd
finished pixel map intro + swapped high/low variance labels on maps + added white borders to pixel map

@@
 #| fig-subcap: 
+#|   - "Low Variance Data"
 #|   - "High Variance Data"
-#|   - "Low Variance Data"
 #|   - "Choropleth Palette"
@@
 #| fig-subcap: 
+#|   - "Low Variance Data"
 #|   - "High Variance Data"
-#|   - "Low Variance Data"
 #|   - "Exceedance Probability Palette"
@@
 #| fig-subcap: 
+#|   - "Low Variance Data"
 #|   - "High Variance Data"
-#|   - "Low Variance Data"
 #|   - "Bivariate Palette"
@@
 #| fig-subcap: 
+#|   - "Low Variance Data"
 #|   - "High Variance Data"
-#|   - "Low Variance Data"
 #|   - "Transparency Palette"
@@
 
-### Value Supressing Uncertainty Palettes
+## Value Supressing Uncertainty Palettes
 The Value Suppressing Uncertainty Palette (VSUP) [@Correll2018], was designed with the intention of preventing high uncertainty values from being extracted from a map.
@@
 #| fig-subcap: 
+#|   - "Low Variance Data"
 #|   - "High Variance Data"
-#|   - "Low Variance Data"
 #|   - "VSUP Palette"
@@
 # Implicitly Combining Uncertainty and Signal
-
-## Pixel maps
+There is technically a stage of our analysis where the estimate and variance are not separate, when we only have a sample.
+Rather than trying to figure out how to combine signal and uncertainty into a single color, we can just display a sample instead and allow the viewer to extract *both* the estimate and the variance.
+
+## Pixel map
+@fig-pixel displays a pixel map [@Lucchesi2021], which is a variation of the choropleth map where each area is divided up into several smaller areas, each colored using outcomes from the larger area's temperature sampling distribution.
+The spatial trend is clearly visible in the low variance case, but functionally disappears in the low variance case.
+While the spatial trend is just barely visible in the high uncertainty case, it is much harder to see.
+This means the graphic also achieves the third criteria for signal-suppression, i.e.
+our difficulty in seeing the distribution is proportional to the level of uncertainty in the graphic.
 
@@
 In this map, each county is broken up into several smaller regions and coloured according to a potential daily temperature, given the its average daily temperature and its sampling distribution.
-This results in each county being represented by a sample rather than a single value."
+This results in each county being represented by a sample rather than a single value.
+In this graphic, we can clearly see the spatial trend in the low variance case, while the spatial trend is much harder to identify in the high variance case.
+"
 #| fig-subcap: 
+#|   - "Low Variance Data"
 #|   - "High Variance Data"
-#|   - "Low Variance Data"
 #|   - "Pixel Map Palette"
@@
 library(Vizumap)
+
 # Low variance map
@@
 p4a <- view(pmap_a) +
+  geom_path(
+      data = pmap_a$bord,
+      aes_string(x = 'long', y = 'lat', group = 'group'),
+      colour = "white"
+    )  +
   scale_fill_gradientn(colours = basecols) +
@@
 p4b <- view(pmap_b) +
+  geom_path(
+      data = pmap_b$bord,
+      aes_string(x = 'long', y = 'lat', group = 'group'),
+      colour = "white"
+    )  +
   scale_fill_gradientn(colours = basecols) +
@@
 p4b
-show_col(basecols, ncol=8, labels=FALSE)
-
+show_pal(basecols, ncol=8, borders="white", myxlab = "Temperature", breaks = 21:29)
 ```
 
-## What is uncertainty?
+It is clear that the pixel-map is not only "suppressing" the false information, but it is doing so by providing *more* information.
+Not only does this method help readers understand the plot level "gist", it is also unlikely to damage the viewers ability to extract individual estimates.
+Extracting global statistics, such as the mean or variance, from a sample can be done with relative each, especially when those values are mapped to color [@Franconeri2021].
+Therefore, the pixel map performs signal suppression, without sacrificing the viewers ability to extract general statistics, unless those statistics *should* be harder to extract due to the uncertainty in the value.
+So is this the best uncertainty visualisation?
+And if so, why?
+
+## What even is "uncertainty" anyway?
+The pixel map is not the best uncertainty visualisation, but it is trying to *replicate* the best uncertainty visualisation.
+The best uncertainty visualisation is probably the the raw data, to understand why, we need to understand what "uncertainty" actually is.
+
+Uncertainty is not just variance, and
+
+## The distribution problem
 
 ## Show me the data
+If we want to show uncertainty for the sake of
+
 @Kim2019 looked at how different uncertainty visualisation techniques influenced the percieved sample size

### 2024-07-31 paper.qmd
moved the last section around a bit

@@
 
+## What is uncertainty
+
 ## The choropleth map
@@
 While these two maps *do* look visually different (which was not the case in the choropleth map) the spatial trend is still clearly visible in both graphics.
-This means the uncertainty *is* being communicated, however the primary take away in the graphic is the spatial trend (that does not exist).
-The graphic no longer has a transparency issue as the uncertainty *is* technically included, but is this a good uncertainty visualisation?
-And if not, why?
+This means the uncertainty *is technically* being communicated, however the primary take away in the graphic is the spatial trend (that does not exist).
+The graphic did not hide the invalid signal, so it is not performing signal-suppression as we would like.
+At this point, it might be reasonable to ask, why?
+Why is including the uncertainty as a variable insufficient to achieve signal-suppression, and what changes should we make to ensure signal-suppression occurs?
 
@@
 
-## What makes uncertainty visualisation difficult?
+## Why this approach may (or may not) work
+Unfortunately it is not enough to only understand the purpose of uncertainty visualisation.
+To do signal-suppression successfully we need to understand what uncertainty actually *is*, as this will inform us on how it should be integrated into the visualisaiton.
+
+### It's a variable... it's a metadata... it's uncertainty?
+Describing what uncertainty *is* is actually suprisingly hard.
+Most authors simply avoid the problem and describe the characteristics of uncertainty, of which there are plenty.
+Often, uncertainty is split using an endless stream of ever changing boundaries, such as whether the uncertainty is due to true randomness or a lack of knowledge [@Spiegelhalter2017; @Hullman2016; @utypo], if the uncertainty is in the attribute, spatial elements, or temporal element of the data [@Kinkeldey2014], whether the uncertainty is scientific (e.g.
+error) or human (e.g.
+disagreement among parties) [@Benjamin2018], if the uncertainty is random or systematic [@Sanyal2009], statistical or bounded [@Gschwandtnei2016; @Olston2002], recorded as accuracy or precision [@Griethe2006; @Benjamin2018], which stage of the data analysis pipeline the uncertainty comes from [@utypo], how quantifiable the uncertainty is [@Spiegelhalter2017; @utypo], etc.
+
+There are enough qualitative descriptors of uncertainty to fill a paper, but, none of this is particularly helpful in understanding how to integrate it into a visualisation.
+Rather than trying to define uncertainty by what it *is* it may be easier to try and describe what uncertainty *is not*.
+Thankfully, the literature on what uncertainty is *not* is much more specific.
+
+Descriptive statistics describe our sample as it is and summarizes large data down into an easy to swallow format.
+Descriptive statistics are not seen as the primary goal of modern statistics, however, this was not always the case.
+In 19th century England, *positivism* was the popular philosophical approach to science (positivists included famous statisticians such as Francis Galton and Karl Pearson).
+Practitioners of the approach believed statistics ended with descriptive statistics as science must be based on actual experience and observations [@Otsuka2023].
+In order to make statements about population statistics, future values, or new observations we need to perform inference, which requires the assumption of the "uniformity of nature", i.e.
+we need to assume that unobserved phenomena should be similar to observed phenomena [@Otsuka2023].
+Positivists abore the assumption of the "uniformity of nature" as they believed that reference to the unobservable is not true science.
+In other words, these scientists embraces descriptive statistics and shunned inferential statistics due to the fact inferential statistics introduces uncertainty.
+
+This history lesson illustrates what uncertainty actually is.
+A by-product of inference.
+This means that anything that could lead us to violate our uniformity of nature assumption, is a source of uncertainty.
+The qualitative characteristics of uncertainty provided by our taxonomies all highlight aspect
+
+feared 
+This subtle shift, from descriptive statistics to inferential statistics was shunned during the positivism era *due to the fact it introduced the unknowable*, or in other words, uncertainty. 
+While we may agree on the *purpose* of uncertainty, it is an entirely different can of worms to agree on what uncertainty actually *is*.
+@fig-bivariate uses variance as a proxy for uncertainty, however those terms are not perfect synonyms.
+
+### Visualising the "single integrated uncertain value"
 Uncertainty visualisation is a uniquely difficult problem.
@@
 The pixel map is not the best uncertainty visualisation, but it is trying to *replicate* the best uncertainty visualisation.
-The best uncertainty visualisation is probably the the raw data, to understand why, we need to understand what "uncertainty" actually is.
-
-Uncertainty is not just variance, and
-
-## The distribution problem
-
-## Show me the data
-If we want to show uncertainty for the sake of
-
-@Kim2019 looked at how different uncertainty visualisation techniques influenced the percieved sample size
-
-{{< pagebreak >}}
-
-# Evaluating uncertainty visualisations
-
-##  Current methods
-@Kinkeldey2014 expressed that uncertainty visualization experiments are idiosyncratic, so it is difficult to generalise the results.
-
-Extracting information using a graphic requires a perceptual task, so it will always be less accurate than explicitly reading the value provided in text form [@Cleveland1984].
-
-## The Hermann grid illusion of visualisation experiments
-A Herman grid is an optical illusion where a black and white grid appears to have grey dots, but they disappear when you look directly at them.
-This optical illusion is a fitting description of the issues in evaluating uncertainty visualisations.
-
-##  Testing signal supression
-If asking direct questions about uncertainty causes us to treat it as a signal, how do we evaluate uncertainty as *noise*?
- When we ask the viewer of a plot to look at data and extract a value, we are asking them to perform inference on that value.
-There will be noise associated with that answer and that is uncertainty.
-If we ask direct question about some uncertainty metric, we have turned the uncertainty into signal because that is what the participants are drawing inference on.
-
-## Current methods
-### Secondary effects
-
-- trust, confidence, etc
-
-### Direct questions
+The best uncertainty visualisation is the visualisation that best captures the limitations of of our raw data.
+
+## variance is not uncertainty
+This approach to uncertainty is embedded in the idea that descriptive statistics do not have uncertainty, which some readers may disagree with.
+Specifically, because it means uncertainty is *not* a latent attribute of data, but rather an attribute of a specific hypothesis or estimate.
+Deniers of this fact follow a consistent logical path and it is easy to identify the common mistake.
+We know that variance and probability are typically considered types of "uncertainty" *and* descriptive statistics can have variance and probabilities *therefore* descriptive statistics must have uncertainty.
+The flaw in this logic comes from the first step, assuming that the tools with which we measure uncertainty *are* uncertainty in of themselves.
+This confusion is common and there are many papers that spend a great deal of time clarifying the difference.
+@Begg2014 highlight that uncertainty is related to not knowing a specific value, while variability refers to the range of values a quantity can take at different locations, times or instances.
+@Spiegelhalter2017 made sure to comment on the difference between precise random events (such as the probability associated with a coin flip), and uncertainty (such as the estimated probability associated with a coin that might be biased).
+The variance of a sample variance can be calculated and know, therefore it is not uncertain but rather it a precise description of dispersion.
+If we were to discuss drawing a new observation, or estimating the true mean of a population *then* the variance would become relevant in our discussions of uncertainty.
+
+You cant measure something you cant define, you cant quantify something you cant measure, and you cant visualise something you cant quantify.
+
+### Goal/task dependence in each stage
 Several authors have connected the issues around defining uncertainty to inference, however it is often discussed as a *task* or *goal* dependence.
@@
 
+With this understanding it becomes clear to see why uncertainty is tied to an endless string of examples in the data analysis pipeline.
+Uncertainty examples include imputed data, model selection, inherent randomness, biased sampling, etc, not because these things *are* uncertainty, but because they *create* uncertainty when we perform inference.
+Whether or not these elements are relevant is highly dependent on what statistic you are trying to draw inference on, and by extension, the purpose of your visualisation.
+
+### Problems with uncertainty having no definition
+The ramifications of this problem is seen repeatedly in the uncertainty visualisation literature.
+A survey of visualisation authors cited "not knowing how to calculate uncertainty" as one of the primary reasons they did not include it in visualisations [@Hullman2020a].
+Some works [@Hullman2018; @Maceachren2012; @Thomson2005] ignore the problem get around this definition issue by focusing narrowly on specific terms defined mathematically, such as probability, variance, error, or precision.
+Others [@Griethe2006; @Pang1997; @Pham2009; @Boukhelifa2017] include broader loosely related elements, such as missing values, but ignore these concepts when translating that uncertainty to a visualisation.
+Existing mathematical definition of uncertainty (or related concepts) have been unable to unify this range of diverse concepts.
+For example, @Thomson2005 suggests a mathematical formula for *examples* of uncertainty, and information theory tries to quantify uncertainty using the idea of entropy, but the disconnect between the broad concept of uncertainty and what we can reliably quantify remains a problem.
+
+As we move through these methods, it seems that the validity of any overarching insight becomes more visible at the cost of our ability to extract particular values of signal or noise.
+Therefore, given that the primary goal of visualisation *is* insights [@North2006], visualisation authors should err on the side of representing suppressed signal as a single variable, rather than visualising uncertainty separately using two different channels.
+
+The concept of uncertainty as a by-product of inference, not a latent feature of data, extends naturally to visualisation. 
+There are two primary reasons for making a visualisation, to perform exploratory data analysis (EDA) or communication.
+Communication involves identifying a signal we want to communicate and designing the visualisation that best conveys that, while EDA involves creating a versatile visualisation without an explicit purpose and using it to extract several signals.
+
+### EDA is descriptive statistics
+ Since EDA is the visual parallel to descriptive statistics, it is performed without an explicit hypothesis which means there *is* no uncertainty in the visualisation.
+Similarly to our issues with descritive statistics, this is not well understood by the uncertainty visualisation community.
+Some authors simply confuse varianc for uncertainty.
+For example @Potter2010 aimed to create a summary plot that "concisely presented data with uncertainty information" to create an exploratory visualisation tool that visualised uncertainty.
+Other authors regonise inference will occur (in some shape or form) and believe uncertainty *should* be visualised but do not recognise *how* uncertainty would be visualised.
+@Hullman2021 argued that there is no such thing as a "model-free" visualisation, therefore visualisation require robust visualisations of uncertainty as we are always performing inference.
+@Griethe2006 commented that "if visualization is used as a means to explore a data volume or to communicate its contents the uncertainty has to be included".
+While we agree people cannot prevent themselves from performing inference, that does not mean it is possible for uncertainty to be included in an EDA visualisation.
+A versatile visualisation such as a scatter plot allows for a viewer to consider several hypothesis at once, each of which will be a different inferential statistic with a different distribution that depicts its uncertainty.
+It is likely impossible to suppress all possible signals at once.
+
+## The distribution problem
+
+## Show me the data
+If we want to show uncertainty for the sake of
+
+If you have uncertainty information, the best approach to create an "uncertainty visualisation" might be to incorperate that information into the visualisation to attempt to replicate the original sample.
+
+@Kim2019 looked at how different uncertainty visualisation techniques influenced the percieved sample size
+
+{{< pagebreak >}}
+
+# Evaluating uncertainty visualisations
+
+##  Current methods
+@Kinkeldey2014 expressed that uncertainty visualization experiments are idiosyncratic, so it is difficult to generalise the results.
+
+Extracting information using a graphic requires a perceptual task, so it will always be less accurate than explicitly reading the value provided in text form [@Cleveland1984].
+
 Experiments where users are asked to make a decision based on a probability, and experiments where users need to extract a variance or some feature of the distribution both have the implicit belief that uncertainty visualisation is a "class" of visualisation, and a "good" uncertainty visualisation allows us to extract all information with ease.
@@
 
+## The Hermann grid illusion of visualisation experiments
+A Herman grid is an optical illusion where a black and white grid appears to have grey dots, but they disappear when you look directly at them.
+This optical illusion is a fitting description of the issues in evaluating uncertainty visualisations.
+
+There is also a swath of studies that are aware a question that boils down to a value extraction experiment, or a question that should be answered by literally ignoring uncertainty information is not what we want when we consider uncertainty visualisations.
+These papers often try to ask a question that should utilise both the uncertainty and signal in the response, however this is rarely what actually occurs.
+This method typically results in in cryptic or confusing questions that create a large amount of noise on the interpretation side of the analysis [@Hullman2016].
+
+Some authors opt for asking slightly vague questions that imply a use of uncertainty, but compare it to a ground truth that is very specific.
+@Ibrekk1987 asked participants for the "best estimate" which was evaluated in accuracy by comparing it to the mean, however the "best estimate" depends on the loss function we are using, and a loss function of minimised error was not implied by the question.
+@Hofmann2012 showed two distributions in 20 different visualisations (a line-up protocol) using a jittered sample, a density plot, a histogram, and a box plot and asked participants.
+Participants were asked to report in which of the plots was "the blue group furthest to the right" The experiment set up is shown in fig-right.
+The participants answers were then compared to a ground truth where the correct plot had a blue distribution with a right shifted mean.
+By comparing the results to a ground truth statistic and marking participants as "wrong" or "right", the error from the participants that had an alternative interpretation to the concept of "furthest right" was conflated with the error from a the visualisation choice.
+These papers make it unclear if the participants got the answers wrong because they misunderstood the question or because of something related to the plot.
+Therefore, this method leads to inconclusive results about the plot design, and is not advised.
+
+Another method used by authors is to ask a deterministic question about a random event.
+@Padilla2017 provided participants with a visualisation of the cone of uncertainty and asked then to "decide which oil rig will receive more damage based on the depicted forecast of the hurricane path".
+The cone of uncertainty provides a 60% confidence interval for the location of the eye of a hurricane, which allows us to know the area where the eye of the storm will go, it does not given any information about the intensity of a storm, the size of a storm, or even if a location will be hit.
+This inclusion of determinism seems to cause the authors to stumble themselves, as they are not consistent with their assumptions.
+In their first experiment @Padilla2017 indicated the correct answer was to assume that the storm was equally intense no matter how far from the centre of the distribution an oil rig was, however answering their third experiment correctly hinged on assuming the intensity of the storm at a particular point (which in this experiment they phrased as damage) *does* change in intensity as you move away from the centre of the distribution.
+Given these conflicting assumptions, it is unclear how the participants were supposed to adjust the probabilistic path information to answer a deterministic question about which oil rig would receive the most damage.
+Other authors have commented on the complexity of communicating hurricane risk because the path, storm surge and wind speed are all important and cannot be ignored [@Spiegelhalter2017].
+The flip side of this is asking participants for a deterministic answer to a probabilistic question.
+@Correll2014 asked participants "how likely is candidate B to win the election?" when the two distributions indicated voter preference.
+Participants were not able to answer the question about likelihood in term of probability, but were instead given seven options from 1=Outcome will be most in favour of A to 7=Outcome will be most in favour of B.
+The ground truth statistic for this question was a scalar multiple of Cohen’s d, indicating participants were supposed to incorporate uncertainty information using a very specific formula that was likely unknown to them but assumed to be used implicitly.
+
+These examples are a bit complete mishmash of methods, however they point to a larger issue that goes beyond decision making, trust and confidence experiments.
+Authors *have no idea* how to evaluate the effects of uncertainty in an uncertainty visualisation.
+
+### Heuristics
+- Things that are true for both normal visualisation and uncertainty, but of particular interest for uncertainty
+
+Additionally, these heuristics and biases can change depending on the larger scope of the graphic and the population we are communicating with [@Spiegelhalter2017; @Kinkeldey2014].
+
+Heuristic checks are useful because they look at unknown pitfalls that might exist in interpretation of current plots.
+Since the hypothesis for these experiments are usually quite specific, e.g.
+"do people perceive a an outcome that is within the bar as more likely than one outside it, even if both outcomes are the same distance from the mean?".
+This means they are less likely to fall into the trap of trying to answer questions that are *far* too broad to be answered with a single experiment (e.g.
+"is a scatter plot better at showing uncertainty than a box plot").
+This work also provides useful insights for experiments by highlights pitfalls participants might fall into when they review the results of evaluation experiments [@Hullman2016].
+@Newman2012 found that participants were more likely to view points within the bar as more likely than points outside of the bar in bar charts with error bars.
+Similar effects have been identified in other types of uncertainty displayed.
+@Padilla2017 found that points that were on an outcome of an ensemble display were perceived as more likely than points not on an outcome, even when the point that was not on a specific outcome of the ensemble was closer to the mean of the uncertainty distribution.
+The sine illusion can cause the confidence interval of a smoothed sine curve to seem wider at the peaks than the troughs, causing us to underestimate uncertainty associated with changing values [@Vanderplas2015].
+
+In a similar vein, experiments that verify smaller aspects of plot design might be more useful to the field in the long run because it helps contribute to a larger working theory of "how do we see visualisations".
+Many visualisation experiments try to compare two plots with several differences, but do not seem to be interested in the mechanisms by which we extract information from visualisations.
+Small perceptual tasks that seek to answer small but highly relevant questions (for example, if colour hue and colour value can be perceived as a single signal suppressed variable) would be useful to the field.
+
+##  Testing signal supression
+If asking direct questions about uncertainty causes us to treat it as a signal, how do we evaluate uncertainty as *noise*?
+ When we ask the viewer of a plot to look at data and extract a value, we are asking them to perform inference on that value.
+There will be noise associated with that answer and that is uncertainty.
+If we ask direct question about some uncertainty metric, we have turned the uncertainty into signal because that is what the participants are drawing inference on.
+
+So, the current methods of measuring or understanding the role of uncertainty in a visualisation is questionable at best, however this is not because visualisation authors are missing the mark, but rather uncertainty is *particularly* difficult to express in a visualisation.
+In simple estimates or verbal communication, the signal is often easy to identify because it is what we are explicitly saying.
+Unlike statistical models, visualisations are used in both data exploration and communication.
+This means what exactly is a *signal* in any particular visualisation is hard to identify, since we often let the visualisation *tell us* what the signal is.
+Additionally, you cannot add noise to *every single possible* signal one might take from a visualisation.
+Two people looking at the same visualisation might, just by chance, develop two entirely different insights and draw inference on two completely different statistics.
+These unique and fascinating challenges that are faced by uncertainty visualisation have been completely untouched by the literature.
+This section will cover some interesting research in uncertainty visualisation and suggestions for better ways to measure uncertainty.
+
+### Qualitative Studies
+Alternatively visualisation research could shift away from the accuracy concept all together ask questions that allow for open ended responses.
+This method can enlighten authors as to *how* the uncertainty information was used by the participants.
+@Hofmann2012 tried to capture this by asking participants why they considered a particular plot to be more "right shifted", however this qualitative assessment does not seem to have made it into the final paper.
+@Daradkeh2015 presented participants with ten investment alternatives and asked participants "from among available alternatives, which alternative do you prefer the most", and were asked to think aloud and consider the uncertainty in their decision making.
+The experimenters goal was to observe and organise the methods people use when making decisions in the face of uncertainty.
+This study was an excellent example in a useful experimental design.
+They highlighted the specific aspects of uncertainty that participants typically considered, such as the range of outcomes that are above/below a certain threshold, minimum and maximum values, the risk of a loss, etc, and mapped where in the decision making process participants made these considerations.
+Data visualisation is commonly utilised as a tool in data exploration, so it is not uncommon for a data analyst to make a plot with only a vague goal and pull out a large number of adjacent observations.
+This experimental framework could replicate this process.
+
+### Just noticeable signal
+It could be argued that a well done uncertainty visualisations should have an imperceptible signal unless the signal would be identified with a hypothesis test, almost like a reverse line up protocol, but this idea also has some issues that should be considered.
+The reject or do not reject concepts in hypothesis testing do not offer a complete image of uncertainty, and exploration of uncertainty visualisation largely stems from a desire to move away from this binary framework.
+
+@Patrick2023 found that human viewers were less sensitive to deviations from the null hypothesis than the typical statistical tests.
+
+What this all means, is that there will always be some trade off when we are designing a visualisation for signal-suppression.
+There will be a trade off in *which* signal we suppress, as well as *how much* we suppress it.
+This is true for any graphic that is comparable in some way to a statistical test or process.
+A good example is the line-up protocol, which is a visualisation tool that used to check if perceived patterns are real or merely the result of chance [@Buja2009; @Wickham2010].
+This motivation is similar to our signal-suppression goal.
+@Patrick2023 compared standard statistical tests to the lineup-protocol, and evaluated the visualisations using the power curves that are typical for hypothesis testing.
+In a similar vein, @Kim2019 investigated how different uncertainty visualisation methods influenced user's prior beliefs, and evaluated the graphics by comparing their results to those from Bayesian inference.
+While these methods apply different statistical philosophies, they both investigated the sensitivity of visualisation methods relative to existing evaluation methods.
+This approach may prove useful when considering
+
 {{< pagebreak >}}
+
+# Future work
+This paper has identified gaps in the uncertainty visualisation literature that could be filled to progress the field.
+
+*Each new development should be accompanied by a mathematical definition of the uncertainty being addressed.* Ideally, a mathematically definition of uncertainty that allows us to combine these components would be developed, but in the absence of that, authors should be more specific about what aspect of "uncertainty" they are covering with their visualisation.
+
+*The concept of uncertainty should be formalised within the grammar of graphics.* This formalisation would allow uncertainty visualisation authors to have a clear understanding of what is or is not an uncertainty visualisation.
+Additionally placing uncertainty visualisation in the framework that is used to understand existing information visualisation research would help authors understand when existing methods can be used to explain their results.
+incorporating uncertainty into the grammar of graphics will also give a more precise concept of the information contained within a plot.
+Other fields of science employ marginal changes when designing experiments to ensure it is well understood *what* aspect of their experiment is contributing to their results, and a better sense of what "marginal" is in the case of uncertainty visualisation would greatly help the field.
+(*XXX Is data pipeline connected with the grammar of graphics?
+Should this be a recommendation?*)
+
+*Experimental practices on uncertainty visualisation need to be standardised.* If we are going to consider uncertainty as noise, not signal, there needs to be a way to identify this signal suppression in an experimental design.
+As the literature currently exists, there is no way to combine papers to get a meaningful sense of how uncertainty information is understood by a viewer.
+There is also the possibility that uncertainty visualisation evaluations will need to swap to a qualitative methodology where participants are allowed to freely comment on what they notice in graphics until we establish how the existence of noise can be observed.
+
+If an uncertainty visualisation researcher would prefer to perform experiments rather than formalise methods, there are options there too.
+It would be interesting to know if any perceptual tasks that can be mapped to two different visual tasks condense into a single dimension when looking for overarching signal in a plot.
+Alternatively, the task dependency many authors in uncertainty visualisation mention would be a useful direction to consider.
+It is clear that the the number of potential tasks that can be performed on a visualisation increases with with the number of observations.
+A single observation is limited to value extraction, two observations can be compared, multiple observations allow for shapes or global statistics to be extracted.
+The interaction between sample size and task is of particular interest to the uncertainty visualisation community, as uncertainty can be expressed through multiple observations using a sample, or through a single value using an error.
+Of course, this is limited by the fact that there also isn't a definition for what is a "task" and given the mess created by the lack of formalisation in uncertainty visualisation, it may be wise to formalise that concept before performing these experiments.
+@Amar2005 suggested a taxonomy for information visualisation based on the types of tasks we use visualisations for and suggest 10 "analytical primitives" that we can then map to visualisations, which could be a good starting point.
+Regardless, these are directions of research would be fruitful to the uncertainty visualisation community even if it appears on the surface to be research that is only beneficial to the "normal" visualisation community.
+(*XXX Not sure what this paragraph is recommending?*)
 
@@
 ```
-
-# ------------------- BELOW HERE IS UNSORTED ------------------------
-
-### What is uncertainty?
-There does not seem to be a clear agreed upon definition of uncertainty which would obviously create some issues when we want to visualize it.
-You cant measure something you cant define, you cant quantify something you cant measure, and you cant visualise something you cant quantify.
-
-The ramifications of this problem is seen repeatedly in the uncertainty visualisation literature.
-A survey of visualisation authors cited "not knowing how to calculate uncertainty" as one of the primary reasons they did not include it in visualisations [@Hullman2020a].
-Some works [@Hullman2018; @Maceachren2012; @Thomson2005] ignore the problem get around this definition issue by focusing narrowly on specific terms defined mathematically, such as probability, variance, error, or precision.
-Others [@Griethe2006; @Pang1997; @Pham2009; @Boukhelifa2017] include broader loosely related elements, such as missing values, but ignore these concepts when translating that uncertainty to a visualisation.
-Existing mathematical definition of uncertainty (or related concepts) have been unable to unify this range of diverse concepts.
-For example, @Thomson2005 suggests a mathematical formula for *examples* of uncertainty, and information theory tries to quantify uncertainty using the idea of entropy, but the disconnect between the broad concept of uncertainty and what we can reliably quantify remains a problem.
-
-Some authors prefer to define uncertainty in qualitative terms, often through the use of taxonomies.
-In these cases, uncertainty is split using an endless stream of ever changing boundaries, such as whether the uncertainty is due to true randomness or a lack of knowledge [@Spiegelhalter2017; @Hullman2016; @utypo], if the uncertainty is in the attribute, spatial elements, or temporal element of the data [@Kinkeldey2014], whether the uncertainty is scientific (e.g.
-error) or human (e.g.
-disagreement among parties) [@Benjamin2018], if the uncertainty is random or systematic [@Sanyal2009], statistical or bounded [@Gschwandtnei2016; @Olston2002], recorded as accuracy or precision [@Griethe2006; @Benjamin2018], which stage of the data analysis pipeline the uncertainty comes from [@utypo], how quantifiable the uncertainty is [@Spiegelhalter2017; @utypo], etc.
-In their own way, each of these approaches show an aspect of uncertainty that an author felt was important to differentiate.
-
-What is interesting about all these views of uncertainty, is that they all imply that it is a latent feature of our data.
-That is, it is something that can be revealed through exploration, and not something that needs to be explicitly calculated.
-The constant stream of new definitions that are unable to capture the true essence of uncertainty might be a result of this false assumption that underpins ever definition.
-Uncertainty cannot be defined without a signal because uncertainty does not exist without a signal, is a by-product of inference.
-
-It is easy to see uncertainties relationship to inference when we consider what might *not* be considered uncertain rather than just trying to think about what *is*.
-
-Descriptive statistics simply describe our sample as it is and summarizes large data down into an easy to swallow format.
-Descriptive statistics are not seen as the primary goal of modern statistics, however, this was not always the case.
-Around the 19th century in England, *positivism* was the popular philosophical approach to science (positivists included famous statisticians such as Francis Galton and Karl Pearson) and practitioners of the approach believed statistics ended with descriptive statistics as science must be based on actual experience and observations, therefore anything that refers to the unobservable (such as new observations or population statistics) is not true science [@Otsuka2023].
-In order to make statements about population statistics, future values, or new observations we need to perform inference, which requires the assumption of the "uniformity of nature" (i.e.
-that unobserved phenomena should be similar to observed phenomena) [@Otsuka2023].
-This subtle shift, from descriptive statistics to inferential statistics was shunned during the positivism era *due to the fact it introduced the unknowable*, or in other words, uncertainty.
-
-This approach to uncertainty is embedded in the idea that descriptive statistics do not have uncertainty, which some readers may disagree with.
-Specifically, because it means uncertainty is *not* a latent attribute of data, but rather an attribute of a specific hypothesis or estimate.
-Deniers of this fact follow a consistent logical path and it is easy to identify the common mistake.
-We know that variance and probability are typically considered types of "uncertainty" *and* descriptive statistics can have variance and probabilities *therefore* descriptive statistics must have uncertainty.
-The flaw in this logic comes from the first step, assuming that the tools with which we measure uncertainty *are* uncertainty in of themselves.
-This confusion is common and there are many papers that spend a great deal of time clarifying the difference.
-@Begg2014 highlight that uncertainty is related to not knowing a specific value, while variability refers to the range of values a quantity can take at different locations, times or instances.
-@Spiegelhalter2017 made sure to comment on the difference between precise random events (such as the probability associated with a coin flip), and uncertainty (such as the estimated probability associated with a coin that might be biased).
-The variance of a sample variance can be calculated and know, therefore it is not uncertain but rather it a precise description of dispersion.
-If we were to discuss drawing a new observation, or estimating the true mean of a population *then* the variance would become relevant in our discussions of uncertainty.
-
-With this understanding it becomes clear to see why uncertainty is tied to an endless string of examples in the data analysis pipeline.
-Uncertainty examples include imputed data, model selection, inherent randomness, biased sampling, etc, not because these things *are* uncertainty, but because they *create* uncertainty when we perform inference.
-Whether or not these elements are relevant is highly dependent on what statistic you are trying to draw inference on, and by extension, the purpose of your visualisation.
-
-As we move through these methods, it seems that the validity of any overarching insight becomes more visible at the cost of our ability to extract particular values of signal or noise.
-Therefore, given that the primary goal of visualisation *is* insights [@North2006], visualisation authors should err on the side of representing suppressed signal as a single variable, rather than visualising uncertainty separately using two different channels.
-
-The concept of uncertainty as a by-product of inference, not a latent feature of data, extends naturally to visualisation. 
-There are two primary reasons for making a visualisation, to perform exploratory data analysis (EDA) or communication.
-Communication involves identifying a signal we want to communicate and designing the visualisation that best conveys that, while EDA involves creating a versatile visualisation without an explicit purpose and using it to extract several signals.
-
- Since EDA is the visual parallel to descriptive statistics, it is performed without an explicit hypothesis which means there *is* no uncertainty in the visualisation.
-Similarly to our issues with descritive statistics, this is not well understood by the uncertainty visualisation community.
-Some authors simply confuse varianc for uncertainty.
-For example @Potter2010 aimed to create a summary plot that "concisely presented data with uncertainty information" to create an exploratory visualisation tool that visualised uncertainty.
-Other authors regonise inference will occur (in some shape or form) and believe uncertainty *should* be visualised but do not recognise *how* uncertainty would be visualised.
-@Hullman2021 argued that there is no such thing as a "model-free" visualisation, therefore visualisation require robust visualisations of uncertainty as we are always performing inference.
-@Griethe2006 commented that "if visualization is used as a means to explore a data volume or to communicate its contents the uncertainty has to be included".
-While we agree people cannot prevent themselves from performing inference, that does not mean it is possible for uncertainty to be included in an EDA visualisation.
-A versatile visualisation such as a scatter plot allows for a viewer to consider several hypothesis at once, each of which will be a different inferential statistic with a different distribution that depicts its uncertainty.
-It is likely impossible to suppress all possible signals at once.
-
-# How do we evaluate uncertainty visualisations?
-
-## Current attempts to measure uncertainty
-There is also a swath of studies that are aware a question that boils down to a value extraction experiment, or a question that should be answered by literally ignoring uncertainty information is not what we want when we consider uncertainty visualisations.
-These papers often try to ask a question that should utilise both the uncertainty and signal in the response, however this is rarely what actually occurs.
-This method typically results in in cryptic or confusing questions that create a large amount of noise on the interpretation side of the analysis [@Hullman2016].
-
-Some authors opt for asking slightly vague questions that imply a use of uncertainty, but compare it to a ground truth that is very specific.
-@Ibrekk1987 asked participants for the "best estimate" which was evaluated in accuracy by comparing it to the mean, however the "best estimate" depends on the loss function we are using, and a loss function of minimised error was not implied by the question.
-@Hofmann2012 showed two distributions in 20 different visualisations (a line-up protocol) using a jittered sample, a density plot, a histogram, and a box plot and asked participants.
-Participants were asked to report in which of the plots was "the blue group furthest to the right" The experiment set up is shown in fig-right.
-The participants answers were then compared to a ground truth where the correct plot had a blue distribution with a right shifted mean.
-By comparing the results to a ground truth statistic and marking participants as "wrong" or "right", the error from the participants that had an alternative interpretation to the concept of "furthest right" was conflated with the error from a the visualisation choice.
-These papers make it unclear if the participants got the answers wrong because they misunderstood the question or because of something related to the plot.
-Therefore, this method leads to inconclusive results about the plot design, and is not advised.
-
-Another method used by authors is to ask a deterministic question about a random event.
-@Padilla2017 provided participants with a visualisation of the cone of uncertainty and asked then to "decide which oil rig will receive more damage based on the depicted forecast of the hurricane path".
-The cone of uncertainty provides a 60% confidence interval for the location of the eye of a hurricane, which allows us to know the area where the eye of the storm will go, it does not given any information about the intensity of a storm, the size of a storm, or even if a location will be hit.
-This inclusion of determinism seems to cause the authors to stumble themselves, as they are not consistent with their assumptions.
-In their first experiment @Padilla2017 indicated the correct answer was to assume that the storm was equally intense no matter how far from the centre of the distribution an oil rig was, however answering their third experiment correctly hinged on assuming the intensity of the storm at a particular point (which in this experiment they phrased as damage) *does* change in intensity as you move away from the centre of the distribution.
-Given these conflicting assumptions, it is unclear how the participants were supposed to adjust the probabilistic path information to answer a deterministic question about which oil rig would receive the most damage.
-Other authors have commented on the complexity of communicating hurricane risk because the path, storm surge and wind speed are all important and cannot be ignored [@Spiegelhalter2017].
-The flip side of this is asking participants for a deterministic answer to a probabilistic question.
-@Correll2014 asked participants "how likely is candidate B to win the election?" when the two distributions indicated voter preference.
-Participants were not able to answer the question about likelihood in term of probability, but were instead given seven options from 1=Outcome will be most in favour of A to 7=Outcome will be most in favour of B.
-The ground truth statistic for this question was a scalar multiple of Cohen’s d, indicating participants were supposed to incorporate uncertainty information using a very specific formula that was likely unknown to them but assumed to be used implicitly.
-
-These examples are a bit complete mishmash of methods, however they point to a larger issue that goes beyond decision making, trust and confidence experiments.
-Authors *have no idea* how to evaluate the effects of uncertainty in an uncertainty visualisation.
-
-## How to test uncertainty as noise
-So, the current methods of measuring or understanding the role of uncertainty in a visualisation is questionable at best, however this is not because visualisation authors are missing the mark, but rather uncertainty is *particularly* difficult to express in a visualisation.
-In simple estimates or verbal communication, the signal is often easy to identify because it is what we are explicitly saying.
-Unlike statistical models, visualisations are used in both data exploration and communication.
-This means what exactly is a *signal* in any particular visualisation is hard to identify, since we often let the visualisation *tell us* what the signal is.
-Additionally, you cannot add noise to *every single possible* signal one might take from a visualisation.
-Two people looking at the same visualisation might, just by chance, develop two entirely different insights and draw inference on two completely different statistics.
-These unique and fascinating challenges that are faced by uncertainty visualisation have been completely untouched by the literature.
-This section will cover some interesting research in uncertainty visualisation and suggestions for better ways to measure uncertainty.
-
-### Qualitative Studies
-Alternatively visualisation research could shift away from the accuracy concept all together ask questions that allow for open ended responses.
-This method can enlighten authors as to *how* the uncertainty information was used by the participants.
-@Hofmann2012 tried to capture this by asking participants why they considered a particular plot to be more "right shifted", however this qualitative assessment does not seem to have made it into the final paper.
-@Daradkeh2015 presented participants with ten investment alternatives and asked participants "from among available alternatives, which alternative do you prefer the most", and were asked to think aloud and consider the uncertainty in their decision making.
-The experimenters goal was to observe and organise the methods people use when making decisions in the face of uncertainty.
-This study was an excellent example in a useful experimental design.
-They highlighted the specific aspects of uncertainty that participants typically considered, such as the range of outcomes that are above/below a certain threshold, minimum and maximum values, the risk of a loss, etc, and mapped where in the decision making process participants made these considerations.
-Data visualisation is commonly utilised as a tool in data exploration, so it is not uncommon for a data analyst to make a plot with only a vague goal and pull out a large number of adjacent observations.
-This experimental framework could replicate this process.
-
-### Just noticeable signal
-It could be argued that a well done uncertainty visualisations should have an imperceptible signal unless the signal would be identified with a hypothesis test, almost like a reverse line up protocol, but this idea also has some issues that should be considered.
-The reject or do not reject concepts in hypothesis testing do not offer a complete image of uncertainty, and exploration of uncertainty visualisation largely stems from a desire to move away from this binary framework.
-
-@Patrick2023 found that human viewers were less sensitive to deviations from the null hypothesis than the typical statistical tests.
-
-What this all means, is that there will always be some trade off when we are designing a visualisation for signal-suppression.
-There will be a trade off in *which* signal we suppress, as well as *how much* we suppress it.
-This is true for any graphic that is comparable in some way to a statistical test or process.
-A good example is the line-up protocol, which is a visualisation tool that used to check if perceived patterns are real or merely the result of chance [@Buja2009; @Wickham2010].
-This motivation is similar to our signal-suppression goal.
-@Patrick2023 compared standard statistical tests to the lineup-protocol, and evaluated the visualisations using the power curves that are typical for hypothesis testing.
-In a similar vein, @Kim2019 investigated how different uncertainty visualisation methods influenced user's prior beliefs, and evaluated the graphics by comparing their results to those from Bayesian inference.
-While these methods apply different statistical philosophies, they both investigated the sensitivity of visualisation methods relative to existing evaluation methods.
-This approach may prove useful when considering
-
-## Other elements of uncertainty visualisations
-- Things that are true for both normal visualisation and uncertainty, but of particular interest for uncertainty
-
-### Heuristics
-Additionally, these heuristics and biases can change depending on the larger scope of the graphic and the population we are communicating with [@Spiegelhalter2017; @Kinkeldey2014].
-
-Heuristic checks are useful because they look at unknown pitfalls that might exist in interpretation of current plots.
-Since the hypothesis for these experiments are usually quite specific, e.g.
-"do people perceive a an outcome that is within the bar as more likely than one outside it, even if both outcomes are the same distance from the mean?".
-This means they are less likely to fall into the trap of trying to answer questions that are *far* too broad to be answered with a single experiment (e.g.
-"is a scatter plot better at showing uncertainty than a box plot").
-This work also provides useful insights for experiments by highlights pitfalls participants might fall into when they review the results of evaluation experiments [@Hullman2016].
-@Newman2012 found that participants were more likely to view points within the bar as more likely than points outside of the bar in bar charts with error bars.
-Similar effects have been identified in other types of uncertainty displayed.
-@Padilla2017 found that points that were on an outcome of an ensemble display were perceived as more likely than points not on an outcome, even when the point that was not on a specific outcome of the ensemble was closer to the mean of the uncertainty distribution.
-The sine illusion can cause the confidence interval of a smoothed sine curve to seem wider at the peaks than the troughs, causing us to underestimate uncertainty associated with changing values [@Vanderplas2015].
-
-##
-In a similar vein, experiments that verify smaller aspects of plot design might be more useful to the field in the long run because it helps contribute to a larger working theory of "how do we see visualisations".
-Many visualisation experiments try to compare two plots with several differences, but do not seem to be interested in the mechanisms by which we extract information from visualisations.
-Small perceptual tasks that seek to answer small but highly relevant questions (for example, if colour hue and colour value can be perceived as a single signal suppressed variable) would be useful to the field.
-
-# Future work
-This paper has identified gaps in the uncertainty visualisation literature that could be filled to progress the field.
-
-*Each new development should be accompanied by a mathematical definition of the uncertainty being addressed.* Ideally, a mathematically definition of uncertainty that allows us to combine these components would be developed, but in the absence of that, authors should be more specific about what aspect of "uncertainty" they are covering with their visualisation.
-
-*The concept of uncertainty should be formalised within the grammar of graphics.* This formalisation would allow uncertainty visualisation authors to have a clear understanding of what is or is not an uncertainty visualisation.
-Additionally placing uncertainty visualisation in the framework that is used to understand existing information visualisation research would help authors understand when existing methods can be used to explain their results.
-incorporating uncertainty into the grammar of graphics will also give a more precise concept of the information contained within a plot.
-Other fields of science employ marginal changes when designing experiments to ensure it is well understood *what* aspect of their experiment is contributing to their results, and a better sense of what "marginal" is in the case of uncertainty visualisation would greatly help the field.
-(*XXX Is data pipeline connected with the grammar of graphics?
-Should this be a recommendation?*)
-
-*Experimental practices on uncertainty visualisation need to be standardised.* If we are going to consider uncertainty as noise, not signal, there needs to be a way to identify this signal suppression in an experimental design.
-As the literature currently exists, there is no way to combine papers to get a meaningful sense of how uncertainty information is understood by a viewer.
-There is also the possibility that uncertainty visualisation evaluations will need to swap to a qualitative methodology where participants are allowed to freely comment on what they notice in graphics until we establish how the existence of noise can be observed.
-
-If an uncertainty visualisation researcher would prefer to perform experiments rather than formalise methods, there are options there too.
-It would be interesting to know if any perceptual tasks that can be mapped to two different visual tasks condense into a single dimension when looking for overarching signal in a plot.
-Alternatively, the task dependency many authors in uncertainty visualisation mention would be a useful direction to consider.
-It is clear that the the number of potential tasks that can be performed on a visualisation increases with with the number of observations.
-A single observation is limited to value extraction, two observations can be compared, multiple observations allow for shapes or global statistics to be extracted.
-The interaction between sample size and task is of particular interest to the uncertainty visualisation community, as uncertainty can be expressed through multiple observations using a sample, or through a single value using an error.
-Of course, this is limited by the fact that there also isn't a definition for what is a "task" and given the mess created by the lack of formalisation in uncertainty visualisation, it may be wise to formalise that concept before performing these experiments.
-@Amar2005 suggested a taxonomy for information visualisation based on the types of tasks we use visualisations for and suggest 10 "analytical primitives" that we can then map to visualisations, which could be a good starting point.
-Regardless, these are directions of research would be fruitful to the uncertainty visualisation community even if it appears on the surface to be research that is only beneficial to the "normal" visualisation community.
-(*XXX Not sure what this paragraph is recommending?*)

### 2024-08-02 paper.qmd
BIG edit (I forgot to commit). Finished visualising uncertainty as a variable *and* the EDA issue with VSUP

@@
 - May not need to reference the same paper again and again
+  - I am not sure when I need a citation and when I dont. Sometimes I could literally cite hundreds of papers to make a point, so I'm not sure if i should cite any at all.
 - examples that aren't maps? Sure are a lot of maps.
@@
 ## Why this approach may (or may not) work
-Unfortunately it is not enough to only understand the purpose of uncertainty visualisation.
-To do signal-suppression successfully we need to understand what uncertainty actually *is*, as this will inform us on how it should be integrated into the visualisaiton.
+The difficulty in incorporating uncertainty into a visualisation is frequently mentioned but seldom explained.
+For example @Hullman2016 commented that it is straightforward to show a value but it is much more complex to show uncertainty but did not explain why.
+Many authors seem to believe uncertainty visualisation is a simple high-dimensional visualisation problem, as the difficulty comes from working out how to add uncertainty into already existing graphics [@moritz2017trust; @Griethe2006].
+The problem with this approach to uncertainty visualisation is that it treats uncertainty the same as we would any other variable.
+However, @fig-bivariate makes it clear that simply including uncertainty as a variable is insufficient to perform signal-suppresion.
+If we cannot treat uncertainty as any other variable, what should we treat it as?
+We need to understand what uncertainty actually *is*, in order to understand how to integrate it into a visualisation.
 
 ### It's a variable... it's a metadata... it's uncertainty?
-Describing what uncertainty *is* is actually suprisingly hard.
+Describing what uncertainty actually is, is surprisingly hard.
 Most authors simply avoid the problem and describe the characteristics of uncertainty, of which there are plenty.
@@
 disagreement among parties) [@Benjamin2018], if the uncertainty is random or systematic [@Sanyal2009], statistical or bounded [@Gschwandtnei2016; @Olston2002], recorded as accuracy or precision [@Griethe2006; @Benjamin2018], which stage of the data analysis pipeline the uncertainty comes from [@utypo], how quantifiable the uncertainty is [@Spiegelhalter2017; @utypo], etc.
-
 There are enough qualitative descriptors of uncertainty to fill a paper, but, none of this is particularly helpful in understanding how to integrate it into a visualisation.
+
 Rather than trying to define uncertainty by what it *is* it may be easier to try and describe what uncertainty *is not*.
-Thankfully, the literature on what uncertainty is *not* is much more specific.
-
 Descriptive statistics describe our sample as it is and summarizes large data down into an easy to swallow format.
@@
 we need to assume that unobserved phenomena should be similar to observed phenomena [@Otsuka2023].
-Positivists abore the assumption of the "uniformity of nature" as they believed that reference to the unobservable is not true science.
-In other words, these scientists embraces descriptive statistics and shunned inferential statistics due to the fact inferential statistics introduces uncertainty.
+Positivists abhore the assumption of the "uniformity of nature" as they believed referencing the unobservable is bad science.
+In other words, these scientists embraces descriptive statistics and shunned inferential statistics due to the inherent uncertainty that came with them.
+Uncertainty is a by-product of inference.
 
 This history lesson illustrates what uncertainty actually is.
-A by-product of inference.
-This means that anything that could lead us to violate our uniformity of nature assumption, is a source of uncertainty.
-The qualitative characteristics of uncertainty provided by our taxonomies all highlight aspect
-
-feared 
-This subtle shift, from descriptive statistics to inferential statistics was shunned during the positivism era *due to the fact it introduced the unknowable*, or in other words, uncertainty. 
-While we may agree on the *purpose* of uncertainty, it is an entirely different can of worms to agree on what uncertainty actually *is*.
-@fig-bivariate uses variance as a proxy for uncertainty, however those terms are not perfect synonyms.
+At several stages in a statistical analysis, we will violate the "uniformity of nature" assumption.
+Each of these violations will impact the statistic we have calculated and push it further from the population parameter we wish to draw inference on.
+Uncertainty is the amalgamation of these impacts.
+If we do not violate the uniformity of nature assumption at any point in our analysis, we do not have any uncertainty.
+
+This interpretation of uncertainty indicates that the uncertainty on a statistic is not of value in of itself.
+Uncertainty is metadata about our statistic that is required for valid inference.
+This means uncertainty should not be visualised by itself and we should seek to display signal and uncertainty together as a "single integrated uncertain value" [@Kinkeldey2014].
+This aspect of uncertainty visualization makes it a uniquely difficult problem.
+This is something frequently mentioned
 
 ### Visualising the "single integrated uncertain value"
-Uncertainty visualisation is a uniquely difficult problem.
-@Hullman2016 commented that it is straightforward to show a value but it is much more complex to show uncertainty.
-This difficulty is often described as a high-dimensional visualisation problem, as the difficulty in uncertainty visualisation is in working out how to add uncertainty into already existing graphics [@moritz2017trust; @Griethe2006].
-The problem with this approach to uncertainty visualisation is that it treats uncertainty the same as we would any other variable, but we are not interested in visualising uncertainty as though it were any other variable.
-
 Typically, when making visualisations, we want the visual channels to be separable, that is, we don't want the data represented through one visual channel to interfere with the others [@Smart2019].
-This is not the goal of uncertainty visualisation.
-The goal of uncertainty visualisations is to display signal and uncertainty together as a "single integrated uncertain value" [@Kinkeldey2014].
-Separability allows for the signal and the uncertainty to be read separately which allows for the uncertainty information to simply be ignored when reading the signal, which is a pervasive issue in current uncertainty visualisation methods [@uncertchap2022].
+Mapping uncertainty and signal to separable channels allows them to be read separately, which does not align with the goal of communicating them as a "single integrated channel".
+Additionally, visualizing uncertainty and signal separately allows the uncertainty information to simply be ignored, which is a pervasive issue in current uncertainty visualisation methods [@uncertchap2022].
 We can see this problem in @fig-bivariate, as it sends the message "this data has a spatial trend and the estimates have a large variance" as we read the signal and the uncertainty separately.
 
-The need for interference from the uncertainty channel does not necessarily mean high-dimensional visualisation techniques would be useless for uncertainty visualisation.
-High-dimensional visualisation methods could work if the uncertainty and the estimates were mapped to integrable channels.
-That is, the visual channels would need to be separately manipulable, but read as a single channel by the human brain.
+This means effective uncertainty visualisation should be leveraging integrability.
+That is, the visual channels of the uncertainty and the signal would need to be separately manipulable, but read as a single channel by the human brain.
 While most visual aesthetics *are* separable, there are some variables that have been shown to be integrable, such as color hue and brightness [@Vanderplas2020].
-When visualising uncertainty using its own visual channel, we can also consider visual semiotics and make sure to map uncertainty to intuitive visual channels, such as mapping more uncertain values to lighter colors [@Maceachren2012].
+When visualizing uncertainty using its own visual channel, we can also consider visual semiotics and make sure to map uncertainty to intuitive visual channels, such as mapping more uncertain values to lighter colors [@Maceachren2012].
 
@@
 The low uncertainty case still has a visible spatial trend, while the spatial trend in the high uncertainty map has functionally disappeared.
-This means the VSUP has successfully performed signal suppression for our data.
+This means the VSUP has successfully suppressed the spatial trend in the data.
+However the spatial trend may not be the only signal of concern in our graphic.
+Now we must return to the original signal-suppression criteria and ask ourselves if they have all been met.
+Are all the justified signals reinforced, while all the unjustified signals are suppressed?
+Is a graphic that performs perfect signal-suppression even possible?
 
@@
 
-Now that we have successfully suppressed our invalid signal, we must return to the original criteria we had for signal supression and ask ourselves if we meet all of them.
-We want to make sure that in our efforts to suppress invalid signals, we have not also suppressed statistically *valid* signals in the process.
-
-## Can we have uncertinty visualisations for EDA?
-A slight problem with uncertainty visualisation is that the uncertainty and the purpose of visualisation are somewhat at odds with one another. The VSUP highlights this problem quite well.
-
+## What can and cannot be suppressed?
+The methods used by the VSUP bring to light a slight problem with uncertainty visualisation.
+Specifically that uncertainty and the purpose of visualisation are somewhat at odds with one another.
 There are two primary motivations behind visualisation, communication and exploratory data analysis (EDA).
 Communication involves identifying a signal we want to communicate and designing a visualisation that best conveys that, while EDA involves creating a versatile visualisation using it to extract several signals.
-It is not uncommon for authors to express a desire for uncertainty visualisations that perform signal-suppression in visualisations made for EDA [@Sarma2024; @Griethe2006], however it is not clear that these uncertainty visualisations *exist at all*.
-
-The impossible task of making this graphic is easy to see with a simple example.
-Let's say we have a graphic that depicts a set of coefficients from a linear regression and the value of the coefficient is shown using a single colour.
-We do a series of t-tests on these estimates, and all of them fail to reject the null hypothesis that they are different from 0.
-This would indicate that each value needs to be visualised using a single color that is indistinguishable from the colour assigned to 0.
-However, we also perform an f-test on our model, and this test *is* rejected.
-This test indicates that the set of estimates should not be visually indistinguishable from 0.
-This is clearly a contradiction.
-
-This t-test, f-test example highlights a fundamental problem with the VSUP and even the bivariate map.
-When we blend colors to make them visually indistinguishable, we need to decide at what level of *uncertainty* to blend these colors together.
-There is no intrinsic mapping of uncertainty, so we need to make a decision as to how much signal-suppression the visualisation should perform.
-There is no existing rule or accepted practice in how to go about this, but it is present in every uncertainty graphic.
-Even though the bivariate map was not explicitly combining temperature values at certain variance levels, the mapping of variance to color saturation does this implicitly.
-That is, at certain saturation values the colors in a bivariate map are imperceptibly different and appear as though they are mapped to the same value, regardless of whether or not that is actually the case.
-
-Regardless as to whether the map colors are implicitly or explicitly indistinguishable, the point at which the colors blend together can only be validated by a single hypothesis.
-Therefore, using the methods discussed thus far, it would be impossible to design signal-suppression for EDA.
-If we are designing a graphic for communication then we can suppress the signal we were seeking to communicate, but this would rely on the distinction between EDA and communication to be somewhat clear cut.
-Visualization authors may design a graphic with a particular purpose, but what the visualization *is actually* used for is determined by the viewer and some authors have discussed the validity of there being any distinction between visualisation motivation's at all [@Hullman2021].
-However, this problem only exists when we color each cell using a single color.
-If we can color a cell using multiple colors, this limitation may disappear entirely.
+If we are designing an uncertainty visualisation for communication then we can just suppress the specific signal we are seeking to communicate.
+In the map example, we would consider @fig-vsup to be a success as the only signal we are concerned with is the spatial trend.
+However, it is not uncommon for authors to express a desire for uncertainty visualisations that perform signal-suppression in visualisations made for EDA [@Sarma2024; @Griethe2006].
+Unfortunately it is not clear whether or not uncertainty visualisations for EDA are possible.
+
+### There is no uncertainty in EDA
+Earlier we established that uncertainty is a by-product of inference, which means without inference, there is no uncertainty.
+Often EDA is used to give us an understanding of our data and identify which signals are worth pursuing.
+In this sense, EDA is the visual parallel to descriptive statistics, as it is performed without an explicit hypothesis which means there is no inference, and by extension, there is no uncertainty.
+
+Some authors recognize inference will occur (in some shape or form) and believe uncertainty *should* be visualised but do not recognize *how* uncertainty would be visualised.
+@Hullman2021 argued that there is no such thing as a "model-free" visualisation, therefore all visualisations require uncertainty as we are always performing inference.
+While we agree that people cannot prevent themselves from performing inference, this does not mean it is possible to include uncertainty in a visualisation designed for EDA.
+However, this does mean we should endeavor for a versatile uncertainty visualisation method that is able to perform signal-suppression on all the signals displayed in the visualisation.
+It is unlikely that the methods discussed thus far are able to achieve this.
+
+### The limitations of explicitly visualising uncertainty and signal
+The lack of versatility of the VSUP is easy to see with a simple example.
+Let's say we have a graphic that depicts a set of coefficients from a linear regression and the value of the coefficient is shown using a single color.
+We want to know "Which of these coefficients are different from 0?" as well as "Which of these coefficients are different from each other?".
+To answer this question we do a series of t-tests on these estimates.
+
+All of the individual t-tests of fail to reject the null hypothesis that the coefficients are different from 0.
+We then make a visualisation that suppresses this signal and ensures that all of the estimates are visually indistinguishable from 0.
+We then do a comparison of two means t-test and find that several of the values need to be visually distinguishable from each other.
+The VSUP method must pick a single color for each estimate, and these colours must be *either* visually distinguishable or indistinguishable from each other.
+We cannot perform signal-suppression on both these signals simultaneously.
+
+This example highlights a fundamental problem with the VSUP that extends to the bivariate map as well.
+When we blend these colors, we need to decide at what level of *uncertainty* to blend these colors together.
+Even though the bivariate map does not explicitly combine color values at certain variance levels, the mapping of variance to color saturation does this implicitly.
+That is, at certain saturation values the colors in a bivariate map are imperceptibly different to the human brain and appear as though they are mapped to the same value.
+At this point, it is irrelevant whether or not the colours are technically different, they are the same color in the human brain.
+Which hypothesis are suppressed and which are not largely depends on the method used to combining colors in the palette [@Kay2019].
+The VSUP here used a tree based method as that is what was used by @Correll2018, but there are alternatives that are more appropriate for different hypothesis.
+
+If we only use a single value to express each signal-suppressed statistic, we will always need to decide which signals we suppress and which we do not.
+However, If we could express the statistic of a cell using multiple colors, this limitation may disappear entirely.
 
@@
 
-## What even is "uncertainty" anyway?
+## Where is all this uncertainty coming from?
 The pixel map is not the best uncertainty visualisation, but it is trying to *replicate* the best uncertainty visualisation.
@@
 
-## variance is not uncertainty
+### variance is not uncertainty
 This approach to uncertainty is embedded in the idea that descriptive statistics do not have uncertainty, which some readers may disagree with.
@@
 You cant measure something you cant define, you cant quantify something you cant measure, and you cant visualise something you cant quantify.
+
+Some authors simply confuse variance for uncertainty.
+For example @Potter2010 aimed to create a summary plot that "concisely presented data with uncertainty information" to create an exploratory visualisation tool that visualised uncertainty.
 
@@
 A survey of visualisation authors cited "not knowing how to calculate uncertainty" as one of the primary reasons they did not include it in visualisations [@Hullman2020a].
-Some works [@Hullman2018; @Maceachren2012; @Thomson2005] ignore the problem get around this definition issue by focusing narrowly on specific terms defined mathematically, such as probability, variance, error, or precision.
-Others [@Griethe2006; @Pang1997; @Pham2009; @Boukhelifa2017] include broader loosely related elements, such as missing values, but ignore these concepts when translating that uncertainty to a visualisation.
+
+Unfortunately "the amalgamations of the impacts of violations to the assumption of the uniformity of nature" is a rather broad concept that does not easily roll off the tongue.
+Regardless, this definition of uncertainty does seem to align with what the literature considers to be included in the uncertainty umbrella.
+Some works [@Hullman2018; @Maceachren2012; @Thomson2005] focus narrowly on specific terms with mathematical definitions, such as probability, confidence intervals, variance, error, or precision.
+These works are only concerned with quantifying the final impact of uncertainty on our statistics.
+That is, how large should the bound around our statistic be, such that our "true" statistic can be inferred.
+Others [@Griethe2006; @Leland2005; @Pang1997; @Pham2009; @Boukhelifa2017] include broader loosely related elements, such as missing values, reliability, model validity, or source integrity.
+These broader and harder to quantify concepts are concerned about potential sources of uncertainty, that is, they describe violations to the assumption of the uniformity of nature.
+
 Existing mathematical definition of uncertainty (or related concepts) have been unable to unify this range of diverse concepts.
@@
 Communication involves identifying a signal we want to communicate and designing the visualisation that best conveys that, while EDA involves creating a versatile visualisation without an explicit purpose and using it to extract several signals.
-
-### EDA is descriptive statistics
- Since EDA is the visual parallel to descriptive statistics, it is performed without an explicit hypothesis which means there *is* no uncertainty in the visualisation.
-Similarly to our issues with descritive statistics, this is not well understood by the uncertainty visualisation community.
-Some authors simply confuse varianc for uncertainty.
-For example @Potter2010 aimed to create a summary plot that "concisely presented data with uncertainty information" to create an exploratory visualisation tool that visualised uncertainty.
-Other authors regonise inference will occur (in some shape or form) and believe uncertainty *should* be visualised but do not recognise *how* uncertainty would be visualised.
-@Hullman2021 argued that there is no such thing as a "model-free" visualisation, therefore visualisation require robust visualisations of uncertainty as we are always performing inference.
-@Griethe2006 commented that "if visualization is used as a means to explore a data volume or to communicate its contents the uncertainty has to be included".
-While we agree people cannot prevent themselves from performing inference, that does not mean it is possible for uncertainty to be included in an EDA visualisation.
-A versatile visualisation such as a scatter plot allows for a viewer to consider several hypothesis at once, each of which will be a different inferential statistic with a different distribution that depicts its uncertainty.
-It is likely impossible to suppress all possible signals at once.
 
@@
 ### Just noticeable signal
+
+there are some papers that do identify signal suppress effects, but they discuss it as a failure rather than a success.
+
 It could be argued that a well done uncertainty visualisations should have an imperceptible signal unless the signal would be identified with a hypothesis test, almost like a reverse line up protocol, but this idea also has some issues that should be considered.

### 2024-08-02 writingdumpster.qmd
BIG edit (I forgot to commit). Finished visualising uncertainty as a variable *and* the EDA issue with VSUP

@@
 # Removed Sections
+@Griethe2006 commented that "if visualization is used as a means to explore a data volume or to communicate its contents the uncertainty has to be included".
+
 After all, if we consider a choropleth map, which ignores uncertainty entirely to be at one end of the spectrum (it is a map that only cares about showing signals reguardless of whether or not they are valid) then a completely colourless map is at the other end of the spectrum (it is a map that only cares about suppressing signals regardless of whether or not they are valid).
@@
 data can be represented in a way that allows the user to make a comparison more accurately (based on the hierarchy), then that design is preferable.
+
+# Data description
+Consider our Iowa temperature map, let $X$ be the 12pm temperature taken within each county on for each day in May from 2000 to 2024.
+Unfortunately, we are given a data set that does not include $X$, instead we have the average of $X$, $\hat{X_k}$ and the variance of $X$, $\hat{VarX}$ for each county, $\forall k=1,...,n$ where $n$ is the total number of counties.
+Theoretically, the estimate and its variance should provide enough information for us to
+
+# EDA
+Regardless as to whether the map colors are implicitly or explicitly indistinguishable, the point at which the colors blend together can only be validated by a single hypothesis.
+Therefore, using the methods discussed thus far, it would be impossible to design signal-suppression for EDA.
+If we are designing a graphic for communication then we can suppress the signal we were seeking to communicate, but this would rely on the distinction between EDA and communication to be somewhat clear cut.
+Visualization authors may design a graphic with a particular purpose, but what the visualization *is actually* used for is determined by the viewer and some authors have discussed the validity of there being any distinction between visualisation motivation's at all [@Hullman2021].
+
+# Bivariate map saturation mapping
+Since changes in the color value and hue can also cause the colors to blend together, the mapping of uncertainty to saturation would also affect the signal-suppression in these maps.
+To correctly map uncertainty, we would need to be able to answer how many units of saturation a 1 unit change in variance worth?
+There are no existing guidelines, but similar to the suppression method, the choice in colour mapping would depend on the signals we hope to suppress.

### 2024-08-03 paper.qmd
finished writing "show me the data" section

@@
 @Hullman2021 argued that there is no such thing as a "model-free" visualisation, therefore all visualisations require uncertainty as we are always performing inference.
+However, even something as simple as calculating the uncertainty that is used to suppress our visualisation is not model free, as we need to identify if the sampling variance or the sample variance is more appropriate [@Hofman2020].
 While we agree that people cannot prevent themselves from performing inference, this does not mean it is possible to include uncertainty in a visualisation designed for EDA.
@@
 
-It is clear that the pixel-map is not only "suppressing" the false information, but it is doing so by providing *more* information.
+It is clear that the pixel-map is not only "suppressing" the false information, but it is doing so by simulating *more* information.
+The efficacy of this method means that visualisations of simulated samples pop up repeatedly in the literature, with examples including samples that are animated over time [@Hullman2015; @Blenkinsop2000], pixel-maps, and spaghetti time series plots.
 Not only does this method help readers understand the plot level "gist", it is also unlikely to damage the viewers ability to extract individual estimates.
@@
 
-## Where is all this uncertainty coming from?
-The pixel map is not the best uncertainty visualisation, but it is trying to *replicate* the best uncertainty visualisation.
+## Show me the data
+The pixel map is not the best uncertainty visualisation, but it is trying to *imitate* the best uncertainty visualisation.
 The best uncertainty visualisation is the visualisation that best captures the limitations of of our raw data.
-
-### variance is not uncertainty
-This approach to uncertainty is embedded in the idea that descriptive statistics do not have uncertainty, which some readers may disagree with.
-Specifically, because it means uncertainty is *not* a latent attribute of data, but rather an attribute of a specific hypothesis or estimate.
-Deniers of this fact follow a consistent logical path and it is easy to identify the common mistake.
-We know that variance and probability are typically considered types of "uncertainty" *and* descriptive statistics can have variance and probabilities *therefore* descriptive statistics must have uncertainty.
-The flaw in this logic comes from the first step, assuming that the tools with which we measure uncertainty *are* uncertainty in of themselves.
-This confusion is common and there are many papers that spend a great deal of time clarifying the difference.
-@Begg2014 highlight that uncertainty is related to not knowing a specific value, while variability refers to the range of values a quantity can take at different locations, times or instances.
-@Spiegelhalter2017 made sure to comment on the difference between precise random events (such as the probability associated with a coin flip), and uncertainty (such as the estimated probability associated with a coin that might be biased).
-The variance of a sample variance can be calculated and know, therefore it is not uncertain but rather it a precise description of dispersion.
-If we were to discuss drawing a new observation, or estimating the true mean of a population *then* the variance would become relevant in our discussions of uncertainty.
-
-You cant measure something you cant define, you cant quantify something you cant measure, and you cant visualise something you cant quantify.
-
-Some authors simply confuse variance for uncertainty.
-For example @Potter2010 aimed to create a summary plot that "concisely presented data with uncertainty information" to create an exploratory visualisation tool that visualised uncertainty.
+As we discussed in previous sections, we can consider uncertainty to be "the amalgamations of the impacts of violations to the assumption of the uniformity of nature".
+It's not a definition that rolls off the tongue, but we can work with it.
+
+Thankfully, this definition of uncertainty aligns nicely with all the concepts that are included in the uncertainty umbrella.
+Some works [@Hullman2018; @Maceachren2012; @Thomson2005] focus narrowly on specific terms with mathematical definitions, such as probability, confidence intervals, variance, error, or precision.
+These works are only concerned with quantifying the final impact of uncertainty on our statistics.
+That is, how large should the bound around our statistic be, such that our "true" statistic can be inferred.
+Others [@Griethe2006; @Leland2005; @Pang1997; @Pham2009; @Boukhelifa2017] include broader loosely related elements, such as missing values, reliability, model validity, or source integrity.
+These broader and harder to quantify concepts are concerned about potential sources of uncertainty, that is, they describe violations to the assumption of the uniformity of nature.
+
+What this means is that we have two types of uncertainty, but one is more "processed" than the other.
+Quantifiable uncertainties are just assumption violations expressed as an effect on our final statistic.
+The disconnect between these two expressions of uncertainty creates a huge problem for authors trying to visualise it.
+A survey of visualisation authors cited "not knowing how to calculate uncertainty" as one of the primary reasons they did not include it in visualisations [@Hullman2020a].
+
+There are two reasons we might leave our uncertainty as an assumption violation rather quantifying the effect.
+The first reason is that we may be unable to translate the assumption violation to a quantifiable uncertainty.
+There is no blanket rule that allows us to reliably quantify all uncertainty for every statistic, although some researchers have tangled with the idea.
+For example, @Thomson2005 suggests a mathematical formula for *examples* of uncertainty, and information theory tries to quantify uncertainty using the idea of entropy, but they ignore the disconnect between the broad concept of uncertainty and what we can reliably quantify.
+Some authors don't believe that it is even possible to quantify all the assumption violations.
+@Spiegelhalter2017 mentioned that combining the uncertainties that appear at each stage of an analysis into a single "uncertainty" value is near impossible.
+
+The second reason to leave uncertainty as a potential violation of our assumptions, is that we might not know the final statistic we are seeking to calculate.
+This is the case for visualisations made for EDA, and a large number of developments in EDA visualisation have been in displaying these difficult to quantify violations.
+For example, @Tierney2023 builds upon the tidy data principles to allow users to handle missing values.
+This includes data plots with a missing value "shadow" that allows visualisation authors to identify if the variables used in a plot have any structure in their missing values, which would contribute to uncertainty.
+
+With this understanding it becomes clear to see why uncertainty is tied to an endless string of examples in the data analysis pipeline.
+Uncertainty examples include imputed data, model selection, inherent randomness, biased sampling, etc, not because these things *are* uncertainty, but because they *create* uncertainty when we perform inference.
+Whether or not these elements are relevant is highly dependent on what statistic you are trying to draw inference on, and by extension, the purpose of your visualisation.
+Therefore the best uncertainty visualisation is not am "uncertainty visualisation" at all, but simply the most accurate depiction of our raw data, that gives us a good idea of the limitations of our model.
+
+This does not mean that visualizing raw data instead of implementing sampling techniques will always prevent insignificant signal from getting through.
+@Buja2009 illustrated how groups that appear linearly separable in a linear discriminant analysis (LDA) visualisation of the data can actually be the result of a LDA performed on too many variables, something that was not clear from the visualisation until the line-up protocol was implemented.
+However just showing the data it is simple but effective option for uncertainty visualisation that that seems to be largely overlooked.
+While it is not always possible, it should always be considered as an effective uncertainty visualisation when the raw data is available.
+
+{{< pagebreak >}}
+
+# Evaluating uncertainty visualisations
 
@@
 @Meng2014 commented what is kept as data and what is tossed away is determined by the motivation of an analysis and what was previously noise can be shown to become signal depending on the the question we seek to answer.
-@Kale2019 discussed how the choices we make in our analysis impact our outcomes and introduce uncertainty.
-@Carlin2023 mentions that each research question can be can be categorised as descriptive, predictive, or causal, each of which has its own appropriate statistical methods and motivation agnostic model selection leads to statistical analysis that is devoid of meaning.
 @Wallsten1997 argue that the best method for evaluating or combining subjective probabilities depends on the uncertainty the decision maker wants to represent and why it matters.
@@
 The importance of inference when discussing uncertainty is never directly acknowledge, but it is always present.
-It is clear that the relationship between uncertainty and inference is noticed at every stage in an analysis, however combining these uncertainties into a single "uncertainty" value is near impossible [@Spiegelhalter2017].
-
-With this understanding it becomes clear to see why uncertainty is tied to an endless string of examples in the data analysis pipeline.
-Uncertainty examples include imputed data, model selection, inherent randomness, biased sampling, etc, not because these things *are* uncertainty, but because they *create* uncertainty when we perform inference.
-Whether or not these elements are relevant is highly dependent on what statistic you are trying to draw inference on, and by extension, the purpose of your visualisation.
-
-### Problems with uncertainty having no definition
-The ramifications of this problem is seen repeatedly in the uncertainty visualisation literature.
-A survey of visualisation authors cited "not knowing how to calculate uncertainty" as one of the primary reasons they did not include it in visualisations [@Hullman2020a].
-
-Unfortunately "the amalgamations of the impacts of violations to the assumption of the uniformity of nature" is a rather broad concept that does not easily roll off the tongue.
-Regardless, this definition of uncertainty does seem to align with what the literature considers to be included in the uncertainty umbrella.
-Some works [@Hullman2018; @Maceachren2012; @Thomson2005] focus narrowly on specific terms with mathematical definitions, such as probability, confidence intervals, variance, error, or precision.
-These works are only concerned with quantifying the final impact of uncertainty on our statistics.
-That is, how large should the bound around our statistic be, such that our "true" statistic can be inferred.
-Others [@Griethe2006; @Leland2005; @Pang1997; @Pham2009; @Boukhelifa2017] include broader loosely related elements, such as missing values, reliability, model validity, or source integrity.
-These broader and harder to quantify concepts are concerned about potential sources of uncertainty, that is, they describe violations to the assumption of the uniformity of nature.
-
-Existing mathematical definition of uncertainty (or related concepts) have been unable to unify this range of diverse concepts.
-For example, @Thomson2005 suggests a mathematical formula for *examples* of uncertainty, and information theory tries to quantify uncertainty using the idea of entropy, but the disconnect between the broad concept of uncertainty and what we can reliably quantify remains a problem.
 
@@
 
-The concept of uncertainty as a by-product of inference, not a latent feature of data, extends naturally to visualisation. 
-There are two primary reasons for making a visualisation, to perform exploratory data analysis (EDA) or communication.
-Communication involves identifying a signal we want to communicate and designing the visualisation that best conveys that, while EDA involves creating a versatile visualisation without an explicit purpose and using it to extract several signals.
-
-## The distribution problem
-
-## Show me the data
-If we want to show uncertainty for the sake of
-
-If you have uncertainty information, the best approach to create an "uncertainty visualisation" might be to incorperate that information into the visualisation to attempt to replicate the original sample.
-
-@Kim2019 looked at how different uncertainty visualisation techniques influenced the percieved sample size
-
-{{< pagebreak >}}
-
-# Evaluating uncertainty visualisations
-
 ##  Current methods
@@
 Extracting information using a graphic requires a perceptual task, so it will always be less accurate than explicitly reading the value provided in text form [@Cleveland1984].
+
+@moritz2017trust did some trust EDA stuff? check the paper
 

### 2024-08-03 writingdumpster.qmd
finished writing "show me the data" section

@@
 The census dot map's addition of interactivity also allows users to zoom in and see the details that caused "uncertainty" in the form of inconsistent colours at lower resolutions when they were zoomed out.
-This does not mean that visualising raw data instead of implementing sampling techniques is always a valid uncertainty visualisation that will prevent insignificant signal from getting through.
-@Buja2009 illustrated how groups that appear linearly separable in a linear discriminant analysis (LDA) visualisation of the data can actually be the result of a LDA performed on too many variables, something that was not clear from the visualisation until the line-up protocol was implemented.
-However it is simple but effective option that seems to be largely overlooked by the uncertainty visualisation community.
 
@@
 There are no existing guidelines, but similar to the suppression method, the choice in colour mapping would depend on the signals we hope to suppress.
+
+# Variance is not uncertainty
+Some authors simply confuse variance for uncertainty.
+For example @Potter2010 aimed to create a summary plot that "concisely presented data with uncertainty information" to create an exploratory visualisation tool that visualised uncertainty.
+
+This approach to uncertainty is embedded in the idea that descriptive statistics do not have uncertainty, which some readers may disagree with.
+Specifically, because it means uncertainty is *not* a latent attribute of data, but rather an attribute of a specific hypothesis or estimate.
+Deniers of this fact follow a consistent logical path and it is easy to identify the common mistake.
+We know that variance and probability are typically considered types of "uncertainty" *and* descriptive statistics can have variance and probabilities *therefore* descriptive statistics must have uncertainty.
+The flaw in this logic comes from the first step, assuming that the tools with which we measure uncertainty *are* uncertainty in of themselves.
+This confusion is common and there are many papers that spend a great deal of time clarifying the difference.
+@Begg2014 highlight that uncertainty is related to not knowing a specific value, while variability refers to the range of values a quantity can take at different locations, times or instances.
+@Spiegelhalter2017 made sure to comment on the difference between precise random events (such as the probability associated with a coin flip), and uncertainty (such as the estimated probability associated with a coin that might be biased).
+The variance of a sample variance can be calculated and know, therefore it is not uncertain but rather it a precise description of dispersion.
+If we were to discuss drawing a new observation, or estimating the true mean of a population *then* the variance would become relevant in our discussions of uncertainty.

### 2024-08-04 paper.qmd
sorted evaluation section and wrote intro

@@
 This review attempts to address this issue by offering a novel perspective on the uncertainty visualisation problem, and hopefully laying the foundations on which we can build a house.
+
 This review is broken into several parts that each reflect a different approach to uncertainty visualisation.
@@
 # Evaluating uncertainty visualisations
-
-### Goal/task dependence in each stage
-Several authors have connected the issues around defining uncertainty to inference, however it is often discussed as a *task* or *goal* dependence.
+If we want to make conclusions about how effective any uncertainty visualisation method is we need to look at the results of evaluation experiments.
+Unfortunately the illustrative methods we have used thus far, i.e.
+showing a graphic and saying "wow look at this", are lacking if we want any generalizable results.
+However, despite the abundance of uncertainty visualisation evaluation experiments, existing literature reviews have struggled to synthesise them into any common rules [@Kinkeldey2014; @Hullman2016].
+Additionally, several authors have linked this problem to a  "task" or "goal" dependence in the evaluations.
+
+Here we discuss the cause of this goal dependence, the current methods for evaluating uncertainty visualisation, and we ask if these methods are sufficient to evaluate graphics on their ability to perform singal-suppression.
+
+## The Hermann grid illusion of visualisation experiments
+When you look at a graphic, what statistics are you inferring from the visulisation?
+Depending on the graphic, you might extract several statistics at once.
+However, this is assuming I have just presented a graphic and said nothing.
+What is I asked you to extract the mean of $X$?
+Then it might be safe to assume you are doing inference on $\bar{X}$.
+
+By far the most common metric used is accuracy, which is used by approximately 36% of evaluation studies [@Hullman2019].
+
+Experiments where users are asked to make a decision based on a probability, and experiments where users need to extract a variance or some feature of the distribution both have the implicit belief that uncertainty visualisation is a "class" of visualisation, and a "good" uncertainty visualisation allows us to extract all information with ease.
+
+A Herman grid is an optical illusion where a black and white grid appears to have grey dots, but they disappear when you look directly at them.
+This optical illusion is a fitting description of the issues in evaluating uncertainty visualisations.
+
 Multiple authors have commented on the need to consider quantifying and expressing uncertainty at every stage of a project as the "goal" shapes every step of the analysis [@Kinkeldey2014; @Hullman2016; @Refsgaard2007].
@@
 ## Current methods
-@Kinkeldey2014 expressed that uncertainty visualization experiments are idiosyncratic, so it is difficult to generalise the results.
+
+@Kinkeldey2014 noted that most studies assessed users ability to extract the estimate and the uncertainty as separate signals, and expressed doubt that this was an effective evaluation tool for conveying their "single integrated value".
+
+There are many secondary benefits that come with this improved transparency, such as better decisions, more trust in the results and more confidence in the authors.
+These secondary benefits are, however, *not* the immediate goal of uncertainty.
+The following sections will discuss the issues and limitations in measuring uncertainty through these secondary metrics and provide suggestions as to how future studies should consider measuring uncertainty.
+
+### Value extraction
+
+It is unclear if this sufficient to assess a visualisation on its ability to communicate a
+
+There is also a swath of studies that are aware a question that boils down to a value extraction experiment, or a question that should be answered by literally ignoring uncertainty information is not what we want when we consider uncertainty visualisations.
+These papers often try to ask a question that should utilise both the uncertainty and signal in the response, however this is rarely what actually occurs.
+This method typically results in in cryptic or confusing questions that create a large amount of noise on the interpretation side of the analysis [@Hullman2016].
+
+@Hullman2019 found that the majority of papers evaluate visualisations on performance (approximately 65% of the papers they surveyed).
 
 Extracting information using a graphic requires a perceptual task, so it will always be less accurate than explicitly reading the value provided in text form [@Cleveland1984].
-
-@moritz2017trust did some trust EDA stuff? check the paper
-
-Experiments where users are asked to make a decision based on a probability, and experiments where users need to extract a variance or some feature of the distribution both have the implicit belief that uncertainty visualisation is a "class" of visualisation, and a "good" uncertainty visualisation allows us to extract all information with ease.
-
-> Most studies that assess the usability of uncertainty visualisations do not contribute to this aspect since they test the retrieval of data and uncertainty separately, but from the perspective that there is nothing special about uncertainty.
-Traditionally, studies in this field focus on the ability to read both the map content and its uncertainty at the same time.
-This may be the mandatory criterion for a successful use of uncertainty, but the question that remains is whether this is sufficient to ensure that a user does not only have two separate values in mind but an integrated uncertain data value.
-[@Kinkeldey2014]
 
@@
 
-@Hullman2019 found that the majority of papers evaluate visualisations on performance (approximately 65% of the papers they surveyed).
-
-By far the most common metric used is accuracy, which is used by approximately 36% of evaluation studies [@Hullman2019].
-
 The final method used by authors is to just explicitly ask about uncertainty and signal information separately.
@@
 
-- definition of uncertainty
-
-There are many secondary benefits that come with this improved transparency, such as better decisions, more trust in the results and more confidence in the authors.
-These secondary benefits are, however, *not* the immediate goal of uncertainty.
-The following sections will discuss the issues and limitations in measuring uncertainty through these secondary metrics and provide suggestions as to how future studies should consider measuring uncertainty.
+### Trust and confidence
+
+@moritz2017trust did some trust EDA stuff? check the paper
+
+A similar measure to trust is using "confidence" in an extracted value or a decision.
+Interestingly, "confidence" is also used to try and capture the clarity of a message in a normal visualisation.
+Confidence cannot simultaneously be a measure of clarity of visualisation *and* a way to capture the uncertainty expressed in a visualisation.
 
@@
 
-A similar measure to trust is using "confidence" in an extracted value or a decision.
-Interestingly, "confidence" is also used to try and capture the clarity of a message in a normal visualisation.
-Confidence cannot simultaneously be a measure of clarity of visualisation *and* a way to capture the uncertainty expressed in a visualisation.
-
-## The Hermann grid illusion of visualisation experiments
-A Herman grid is an optical illusion where a black and white grid appears to have grey dots, but they disappear when you look directly at them.
-This optical illusion is a fitting description of the issues in evaluating uncertainty visualisations.
-
-There is also a swath of studies that are aware a question that boils down to a value extraction experiment, or a question that should be answered by literally ignoring uncertainty information is not what we want when we consider uncertainty visualisations.
-These papers often try to ask a question that should utilise both the uncertainty and signal in the response, however this is rarely what actually occurs.
-This method typically results in in cryptic or confusing questions that create a large amount of noise on the interpretation side of the analysis [@Hullman2016].
+### Vague questions
 
@@
 
+### Deterministic questions
 Another method used by authors is to ask a deterministic question about a random event.
@@
 
-These examples are a bit complete mishmash of methods, however they point to a larger issue that goes beyond decision making, trust and confidence experiments.
-Authors *have no idea* how to evaluate the effects of uncertainty in an uncertainty visualisation.
-
-### Heuristics
+### Identifying heuristics
 - Things that are true for both normal visualisation and uncertainty, but of particular interest for uncertainty

### 2024-08-05 paper.qmd
wrote value extraction bit

@@
 Whether or not these elements are relevant is highly dependent on what statistic you are trying to draw inference on, and by extension, the purpose of your visualisation.
-Therefore the best uncertainty visualisation is not am "uncertainty visualisation" at all, but simply the most accurate depiction of our raw data, that gives us a good idea of the limitations of our model.
+
+This relationship between uncertainty and the "purpose" of our analysis is littered throughout the literature.
+Multiple authors have commented on the need to consider quantifying and expressing uncertainty at every stage of a project as the "goal" shapes every step of the analysis [@Kinkeldey2014; @Hullman2016; @Refsgaard2007].
+@Otsuka2023 suggested that the process of observing data to perform statistics is largely dependent on our goals, because the process of boiling real world entities down into probabilistic objects (or "probabilistic kind" as he puts it) depends on the relationship we seek to identify with our data.
+@Meng2014 commented what is kept as data and what is tossed away is determined by the motivation of an analysis and what was previously noise can be shown to become signal depending on the the question we seek to answer.
+@Wallsten1997 argue that the best method for evaluating or combining subjective probabilities depends on the uncertainty the decision maker wants to represent and why it matters.
+@Fischhoff2014 looks at uncertainty visualisation for decision making decides that we should have different ways of communicating uncertainty based off what the user is supposed to do with it.
+
+This makes it very difficult to move quantified uncertainty through the the layers of our analysis, especially when designing a visualisation for EDA.
+If we don't know what the final statistic is, we cannot quantify the effects of our assumptions.
+Therefore, often the best uncertainty visualisation is not an "uncertainty visualisation" at all, but simply the most accurate depiction of our raw data as it gives us a good idea of it's limitations.
 
@@
 However, despite the abundance of uncertainty visualisation evaluation experiments, existing literature reviews have struggled to synthesise them into any common rules [@Kinkeldey2014; @Hullman2016].
-Additionally, several authors have linked this problem to a  "task" or "goal" dependence in the evaluations.
-
-Here we discuss the cause of this goal dependence, the current methods for evaluating uncertainty visualisation, and we ask if these methods are sufficient to evaluate graphics on their ability to perform singal-suppression.
-
-## The Hermann grid illusion of visualisation experiments
-When you look at a graphic, what statistics are you inferring from the visulisation?
-Depending on the graphic, you might extract several statistics at once.
-However, this is assuming I have just presented a graphic and said nothing.
-What is I asked you to extract the mean of $X$?
-Then it might be safe to assume you are doing inference on $\bar{X}$.
-
-By far the most common metric used is accuracy, which is used by approximately 36% of evaluation studies [@Hullman2019].
-
-Experiments where users are asked to make a decision based on a probability, and experiments where users need to extract a variance or some feature of the distribution both have the implicit belief that uncertainty visualisation is a "class" of visualisation, and a "good" uncertainty visualisation allows us to extract all information with ease.
-
-A Herman grid is an optical illusion where a black and white grid appears to have grey dots, but they disappear when you look directly at them.
-This optical illusion is a fitting description of the issues in evaluating uncertainty visualisations.
-
-Multiple authors have commented on the need to consider quantifying and expressing uncertainty at every stage of a project as the "goal" shapes every step of the analysis [@Kinkeldey2014; @Hullman2016; @Refsgaard2007].
-@Otsuka2023 suggested that the process of observing data to perform statistics is largely dependent on our goals, because the process of boiling real world entities down into probabilistic objects (or "probabilistic kind" as he puts it) depends on the relationship we seek to identify with our data.
-@Meng2014 commented what is kept as data and what is tossed away is determined by the motivation of an analysis and what was previously noise can be shown to become signal depending on the the question we seek to answer.
-@Wallsten1997 argue that the best method for evaluating or combining subjective probabilities depends on the uncertainty the decision maker wants to represent and why it matters.
-@Fischhoff2014 looks at uncertainty visualisation for decision making decides that we should have different ways of communicating uncertainty based off what the user is supposed to do with it.
-The importance of inference when discussing uncertainty is never directly acknowledge, but it is always present.
-
-As we move through these methods, it seems that the validity of any overarching insight becomes more visible at the cost of our ability to extract particular values of signal or noise.
-Therefore, given that the primary goal of visualisation *is* insights [@North2006], visualisation authors should err on the side of representing suppressed signal as a single variable, rather than visualising uncertainty separately using two different channels.
+
+Here we discuss common evaluation methods, why these methods might struggle to create a cohesive set of recommendations for uncertainty visualisations, and consier how we to best evaluate visualisations on their ability to perform signal-suppression.
 
 ## Current methods
-
-@Kinkeldey2014 noted that most studies assessed users ability to extract the estimate and the uncertainty as separate signals, and expressed doubt that this was an effective evaluation tool for conveying their "single integrated value".
-
-There are many secondary benefits that come with this improved transparency, such as better decisions, more trust in the results and more confidence in the authors.
-These secondary benefits are, however, *not* the immediate goal of uncertainty.
-The following sections will discuss the issues and limitations in measuring uncertainty through these secondary metrics and provide suggestions as to how future studies should consider measuring uncertainty.
-
-### Value extraction
-
-It is unclear if this sufficient to assess a visualisation on its ability to communicate a
-
-There is also a swath of studies that are aware a question that boils down to a value extraction experiment, or a question that should be answered by literally ignoring uncertainty information is not what we want when we consider uncertainty visualisations.
-These papers often try to ask a question that should utilise both the uncertainty and signal in the response, however this is rarely what actually occurs.
-This method typically results in in cryptic or confusing questions that create a large amount of noise on the interpretation side of the analysis [@Hullman2016].
-
-@Hullman2019 found that the majority of papers evaluate visualisations on performance (approximately 65% of the papers they surveyed).
-
-Extracting information using a graphic requires a perceptual task, so it will always be less accurate than explicitly reading the value provided in text form [@Cleveland1984].
-
-Uncertainty visualisation experiments often compare graphics on the basis of being "uncertainty visualisations" [@Ibrekk1987; @Hullman2015; @Hofman2020] with little consideration for how difficult the information is to extract from the chat.
+Including uncertainty in a visualisation comes with many secondary benefits.
+Examples of these benefits include better decisions, more trust in the results and the ability to extract additional statistics, such as the variance.
+Ultimately, these secondary benefits are not the primary goal of uncertainty visualisation, and evaluating uncertainty visualisations on these criteria often has unintended consequences.
+
+### Value extraction of uncertainty statistics
+Uncertainty visualisations are most commonly evaluated on how accurately readers can extract information from them [@Hullman2019].
+This most commonly comes in the form of separate questions about estimates variance [@Kinkeldey2014].
+This means a significant chunk of evaluation studies boil down to showing a participant a visualisation and asking questions such as "what is the variance of $X$?", or "what is the mean of $X$?".
+Additionally, uncertainty visualisation experiments often compare graphics on the basis of being "uncertainty visualisations" [@Ibrekk1987; @Hullman2015; @Hofman2020] regardless of what information can actually be extracted by the chart.
+This seems like a relatively straight forward approach, and it is similar to how non-uncertainty visualisations are evaluated, but is this appropriate for uncertainty visualisations?
+Uncertainty is not something we are trying to infer, uncertainty is a by-product of inference.
+By shifting the focus of our inference to $Var(X)$ or $P(X)$ we are just evaluating visualisations on their ability to convey uncertainty statistics.
+These graphics may not always set out to communicate uncertainty as a signal, but they are certainty evaluating uncertainty as a signal.
+
+The problem with evaluating uncertainty as a signal are identical to the problems associated with displaying uncertainty as a signal.
+There is no reason to assume uncertainty would behave any differently to any other variable when we evaluate them in this way.
 For example, @Ibrekk1987 found that participants were more accurate at extracting a statistic when it could be directly read off the graphic, than when it required an area estimate (which is the case if using the PDF), or when there was no visual indicator for the statistic at all (which is the case when using the CDF of an asymmetric function).
@@
 @Cheong2016 found that participants made better decisions when they were explicitly given the relevant probability in text rather than when they needed to read it off a map.
-
-The final method used by authors is to just explicitly ask about uncertainty and signal information separately.
-@Sanyal2009 mapped uncertainty to dots and signal to a 3D surface and asked participants to identify areas of high and low signal and high and low uncertainty.
-Participants were not asked to combine that information in any way, and the signal and the noise were treated as separate variables.
-@Correll2014 asked participants to separately extract the mean and variance from four uncertainty visualisations.
-These methods explicitly view the uncertainty and signal as two separate variables that should be extracted from a plot, and not two variables that should be interpreted together.
-Even viewing these questions as a routine check to make sure the signal information isn't impacted by the uncertainty is counter intuitive, because the whole point *of* the uncertainty is to impact the signal information.
+It is well established that extracting information from a graphic using a perceptual task will always be less accurate than explicitly reading the value provided in text form [@Cleveland1984].
+
+Directly asking about uncertainty evaluates uncertainty as a signal, not as noise, so it is not a reliable evaluation method when we are designing graphics for signal suppression.
+Many uncertainty visualisation researchers are already aware of this, so direct questions are not the only methods used to evaluate uncertainty visualisations.
+We will discuss these alternatives in the following section.
 
@@
 ### Vague questions
+There is also a swath of studies that are aware a question that boils down to a value extraction experiment, or a question that should be answered by literally ignoring uncertainty information is not what we want when we consider uncertainty visualisations.
+These papers often try to ask a question that should utilise both the uncertainty and signal in the response, however this is rarely what actually occurs.
+This method typically results in in cryptic or confusing questions that create a large amount of noise on the interpretation side of the analysis [@Hullman2016].
 
@@
 ## Testing signal supression
+
+If a graphic was performing signal suppression we would expect the values to be harder to read at higher levels of uncertainty, something that is often seen as a bad thing in current evaluation studies.
+- trust experiment
+- weird animation of grass
+
+Since the primary goal of visualisation *is* developing insights [@North2006], you will probably extract several signals and use them to make a broader conclusion.
+A simple example is the spatial trend in
+
 If asking direct questions about uncertainty causes us to treat it as a signal, how do we evaluate uncertainty as *noise*?

### 2024-08-05 writingdumpster.qmd
wrote value extraction bit

@@
 # Removed Sections
+The beauty of visualisation is that it *tells* us what we should be concerned about in our analysis. This aspect disappears the second we ask a direct question about our visualisation.
+
 @Griethe2006 commented that "if visualization is used as a means to explore a data volume or to communicate its contents the uncertainty has to be included".
@@
 If we were to discuss drawing a new observation, or estimating the true mean of a population *then* the variance would become relevant in our discussions of uncertainty.
+
+# Removed comments on experiments
+The final method used by authors is to just explicitly ask about uncertainty and signal information separately.
+@Sanyal2009 mapped uncertainty to dots and signal to a 3D surface and asked participants to identify areas of high and low signal and high and low uncertainty.
+Participants were not asked to combine that information in any way, and the signal and the noise were treated as separate variables.
+@Correll2014 asked participants to separately extract the mean and variance from four uncertainty visualisations.
+These methods explicitly view the uncertainty and signal as two separate variables that should be extracted from a plot, and not two variables that should be interpreted together.
+Even viewing these questions as a routine check to make sure the signal information isn't impacted by the uncertainty is counter intuitive, because the whole point *of* the uncertainty is to impact the signal information.
+
+# Alternative purpose section
+This purpose or task dependency appears repeatedly in the literature, although its connection to inference is seldom mentioned.
+Multiple authors have commented on the need to consider quantifying and expressing uncertainty at every stage of a project as the "goal" shapes every step of the analysis [@Kinkeldey2014; @Hullman2016; @Refsgaard2007].
+This is because "signal" or "noise" are not inherent qualities of our data, they are simply the terms we give to the information we wants to keep, and the information we want to throw away.
+Whether or not information is signal or noise depends entirely on the questions we are are seeking to answer [@Meng2014].
+This distinction between noise and signal occurs as early as data collection, where the entire process of deciding how to boil real world entities down into data depends on the relationships we seek to identify [@Otsuka2023].
+This is true for our entire analysis, right up to deciding how to display information, as the way we display information should be based on that the user is supposed to do with it [@Fischhoff2014].
+Even the process we use to combine uncertainty from different stages of an analysis depends on why the uncertainty matters [@Wallsten1997].
+It is clear that as we move through the stages of our analysis noise is the status given to the information we throw away that inhibits our ability to identify signal.
+We want to reintroduce some noise back into our visualisation such that we can perform signal-suppression.
+
+# Example for perceptual task problem
+Consider the typical uncertainty visualisations shown in @fig-mean, which includes a density plot, a sample, and an mean with confidence intervals.
+These graphics each show the distribution of the standard normal random variable X using a density plot, a sample, and a mean with confidence bars.
+Each of these technically show a piece of information with zero uncertainty.
+The density plot provides a deterministic interpretation of the PDF, the sample shows exactly the values we observed, and the mean with error bars shows three explicitly calculated statistics ($X_{0.05}, \bar{x}, X_{0.95}$).
+If we were to use these visualisations to extract the mean, the role of signal in the chart shifts, and instead of directly reading the values, we need to do mental calculations.
+The density plot requires us to find the midpoint along the x axis that could spit the area under the density curve perfect in half.
+The sample requires us to find the point that minimizes the sum of squares errors, i.e.
+the total lengths from each point to the mean.
+The mean with error bars allows us to simply read the mean off the graphic using a basic perceptual task.
+If we were to use the visualisations to extract the variance instead, we could describe a similar process.
+
+```{r}
+#| echo: false
+#| message: false
+#| warning: false
+#| label: fig-mean
+#| fig-cap: "This visualisation "
+#| fig-subcap: 
+#|   - "Probability density function"
+#|   - "Sample"
+#|   - "Mean with error bars"
+#| layout-ncol: 3
+#| layout-valign: "bottom"
+set.seed(1)
+xdata <- tibble(x=rnorm(100))
+
+# Density function
+x1 <- xdata |>
+  ggplot(aes(x=x)) +
+  geom_density(colour="#1e81b0", fill="#abdbe3", alpha=0.5) + 
+  scale_x_continuous(name = "x") +
+  scale_y_continuous(name = "P(X)") +
+  theme_minimal() + 
+  theme(aspect.ratio=1)
+
+# Sample
+
+# Mean with error bars
+x3 <- xdata %>%
+  summarise(avg = mean(x),
+          conf_95a = quantile(x, probs=c(0.025)),
+          conf_95b = quantile(x, probs=c(0.975))) %>%
+  ggplot(aes(y="NA")) +
+  geom_errorbar(aes(xmin = conf_95a, xmax = conf_95b), width = 0.1, colour="#abdbe3") +
+  geom_point(aes(x=avg), colour="#1e81b0") +
+  scale_x_continuous(name = "X",
+                     breaks = seq(-2,2, 0.5)) +
+  theme_minimal() +
+  theme(axis.line.y=element_blank(),
+        axis.title.y = element_blank(),
+        axis.text.y = element_blank(),
+        axis.ticks.y = element_blank(),
+        aspect.ratio=1)
+
+```

### 2024-08-05 paper.qmd
finished trust and confidence section

@@
 For example @Hullman2016 commented that it is straightforward to show a value but it is much more complex to show uncertainty but did not explain why.
-Many authors seem to believe uncertainty visualisation is a simple high-dimensional visualisation problem, as the difficulty comes from working out how to add uncertainty into already existing graphics [@moritz2017trust; @Griethe2006].
+Many authors seem to believe uncertainty visualisation is a simple high-dimensional visualisation problem, as the difficulty comes from working out how to add uncertainty into already existing graphics [@Griethe2006].
 The problem with this approach to uncertainty visualisation is that it treats uncertainty the same as we would any other variable.
@@
 ### Trust and confidence
-
-@moritz2017trust did some trust EDA stuff? check the paper
-
-A similar measure to trust is using "confidence" in an extracted value or a decision.
-Interestingly, "confidence" is also used to try and capture the clarity of a message in a normal visualisation.
-Confidence cannot simultaneously be a measure of clarity of visualisation *and* a way to capture the uncertainty expressed in a visualisation.
-
-Trust is a by product of displaying uncertainty rather than the goal of it, and viewing the relationship in the converse direction can lead to misguided research.
+Trust is a by-product of displaying uncertainty and it commonly measured in uncertainty evaluation studies [@Hullman2019].
 Considering trust, and not transparency, as the metric of importance in uncertainty communication can lead to a questionable subtext that argues against transparency, something that has been noticed by several other authors [@Spiegelhalter2017; ONeill2018].
+Science communication should be primarily concerned with accuracy.
+
+Setting trust as the variable of interest implicitly encourages statisticians to set trust and as the primary goal of communication.
 @Hullman2020a found that author simultaneously argued that failing to visualise uncertainty was akin to fraud, but also many avoided uncertainty visualisation because they didn't want their work to come across as "untrustworthy".
-These authors are optimising *trust* rather than *transparency*, which means they opt to leave out uncertainty information when it does exactly what it is supposed to, decrease certainty in conclusions.
-
-Science communication should be primarily concerned with accuracy.
-Setting trust and risk-aversion as the variables of interest implicitly encourages statisticians to set trust and risk-aversion as the primary goals of communication.
-The issue of trust being divorced from trustworthiness has been commented on by other authors [@ONeill2018], however the issue still persists in the uncertainty visualisation literature.
+This is a classic example of the negative impacts of placing direct importance on *trust* rather than *transparency*.
+In cases of high uncertainty, authors will opt to leave out uncertainty information because it decreases confidence in the authors conclusions.
+
+The disconnect between trust and transparency extends to evaluation experiments as well.
 @Zhao2023 displayed a several variations of a visualisation of a model prediction and its uncertainty and took participants using the model prediction as a sign of trust.
 They reported that visualising uncertainty information caused participants to trust the model in the low variance case, but the results in the high variance case were inconclusive.
-The discussion made it clear the authors thought the uncertainty information should make the visualisation more trustworthy, but conflating trust and the use of a prediction implied uncertainty information should somehow influence participants to use their own prediction, even though a prediction being uncertainty does not necessarily mean it is incorrect.
-Despite this, the authors seemed to assume that the uncertainty information *should* have an influence on that, showing they had not deeply considered *how* uncertainty information should influence the choices of the participants.
-(*Cite Gap 18: examples of studies where authors measure trust*)
-
-### Vague questions
+The discussion made it clear the authors thought the uncertainty information should make the visualisation more trustworthy.
+However, this perspective conflates transparency and trustworthiness.
+If a model has very high uncertainty because of a very low sample size, the heuristics a human might use to make the same decision could be more accurate.
+
+Another metric that is similar to trust is the participants confidence in their decision or extracted value.
+Confidence has many of the same issues as trust, but it has an additional confounding factor.
+In non-uncertainty visualisation evaluation experiments, "confidence" is used as a proxy for the clarity of the visualisation.
+Confidence cannot simultaneously be a measure of clarity of visualisation *and* a way to capture the uncertainty expressed in a visualisation.
+Uncertainty visualisations conflate these two measures when they ask about confidence.
+
+### Questions that attempts to capture signal-suppression
 There is also a swath of studies that are aware a question that boils down to a value extraction experiment, or a question that should be answered by literally ignoring uncertainty information is not what we want when we consider uncertainty visualisations.
@@
 
-### Deterministic questions
 Another method used by authors is to ask a deterministic question about a random event.

### 2024-08-07 paper.qmd
finished vague question and added risk-aversion paragraph

@@
 
-### Trust and confidence
+### Trust, confidence, and risk aversion
 Trust is a by-product of displaying uncertainty and it commonly measured in uncertainty evaluation studies [@Hullman2019].
@@
 
+Unfortunately, experiments where the best response comes from participants who outright ignore the uncertainty information are not uncommon.
+A simple example of this is experiments where participants are expectd to extract the mean of the distribution.
+In these cases, we would expect the best performing visualisation to be one that does not include uncertainty at all, as it would simply show the mean.
+Similarly, many uncertainty visualisations are evaluated on how much risk aversion the participants display [@Hullman2019].
+Additionally, comparing participant responses to that of a ration agent has been suggested as a benchmark for uncertainty visualization experiments [@wu2023rational].
+This standard seems to go against the purpose of uncertainty visualisation as a rational agent, *by definition* should *ignore* uncertainty information.
+Risk aversion is considered to be irrational because it means the economic agent *is* considering the uncertainty information.
+The only cases where a rational agent should not ignore the uncertainty information is when it is required to optimise their utility function, which would mean the uncertainty is not acting as noise, it is acting as signal.
+
 Another metric that is similar to trust is the participants confidence in their decision or extracted value.
@@
 
+For these reasons we do not believe trust, confidence, or risk-aversion are useful measures to evaluate uncertainty visualisations.
+While they are designed to capture the secondary effects of uncertainty, using them as primary measures of visualisation performance incentives the use of visualisations that *do not* include uncertainty.
+
 ### Questions that attempts to capture signal-suppression
-There is also a swath of studies that are aware a question that boils down to a value extraction experiment, or a question that should be answered by literally ignoring uncertainty information is not what we want when we consider uncertainty visualisations.
-These papers often try to ask a question that should utilise both the uncertainty and signal in the response, however this is rarely what actually occurs.
-This method typically results in in cryptic or confusing questions that create a large amount of noise on the interpretation side of the analysis [@Hullman2016].
-
-Some authors opt for asking slightly vague questions that imply a use of uncertainty, but compare it to a ground truth that is very specific.
-@Ibrekk1987 asked participants for the "best estimate" which was evaluated in accuracy by comparing it to the mean, however the "best estimate" depends on the loss function we are using, and a loss function of minimised error was not implied by the question.
-@Hofmann2012 showed two distributions in 20 different visualisations (a line-up protocol) using a jittered sample, a density plot, a histogram, and a box plot and asked participants.
-Participants were asked to report in which of the plots was "the blue group furthest to the right" The experiment set up is shown in fig-right.
-The participants answers were then compared to a ground truth where the correct plot had a blue distribution with a right shifted mean.
-By comparing the results to a ground truth statistic and marking participants as "wrong" or "right", the error from the participants that had an alternative interpretation to the concept of "furthest right" was conflated with the error from a the visualisation choice.
-These papers make it unclear if the participants got the answers wrong because they misunderstood the question or because of something related to the plot.
-Therefore, this method leads to inconclusive results about the plot design, and is not advised.
-
-Another method used by authors is to ask a deterministic question about a random event.
-@Padilla2017 provided participants with a visualisation of the cone of uncertainty and asked then to "decide which oil rig will receive more damage based on the depicted forecast of the hurricane path".
+There is a collection of studies that seem to be aware of the issues behind using trust or value extraction to evaluate uncertainty visualisations.
+These studies try to measure the effect of some kind of "single integrated value" but the methodology is often ad-hoc with varying levels of success.
+
+The first method is what we call the "vague question" approach.
+The authors of these studies will ask the participants a question that implies that they should use  a use of uncertainty, but compare the readers answer to a very specific ground truth.
+This means the participants are being *evaluated* as though they are performing a value extraction task, but they are not being asked the *questions* that are asked in the value extraction task.
+Ultimately this approach results in strangely cryptic questions that create a large amount of noise due to the varied interpretation of the questions [@Hullman2016].
+For example @Hofmann2012 showed study participants 20 plots, where each plot displayed two distributiosn (as a pair of jittered samples, density plots, histograms, or box plots) and asked them to identify the plot where "the blue group furthest to the right".
+The mean of the two groups was used to decide which group was actually "furthest to the right" and the participants were evaluated against that ground truth.
+In another example @Ibrekk1987 asked participants for the "best estimate", but the participant's responses were evaluated against the mean of the distribution.
+Ultimately, the term "best" is up to the users interpretation, and the estimate that minimized the sum of squared errors was not implied by the question.
+This vague question approach leads to inconclusive results, as we are left unclear if it was the phrasing of the question or the plot design that caused the participants to answer incorrectly.
+
+A variation of the "vague question" problem, is a series of studies that ask questions that are impossible to answer using the information given to the participants.
+These studies frequently expect participants to give deterministic answers for probabilistic questions.
+For example @Correll2014 showed participants the distribution of voter preferences for two candidates in an election, and aksed them "how likely is candidate B to win the election?".
+Participants were not able to answer the question about likelihood in term of probability, but were instead given seven options from 1 = "Outcome will be most in favour of A" to 7 = "Outcome will be most in favour of B".
+The ground truth statistic for this question was a scalar multiple of Cohen’s d, indicating participants were supposed to incorporate uncertainty information using a very specific formula that was likely unknown to them but assumed to be used implicitly.
+In another example, @Padilla2017 provided participants with a visualisation of the cone of uncertainty and asked then to "decide which oil rig will receive more damage based on the depicted forecast of the hurricane path".
 The cone of uncertainty provides a 60% confidence interval for the location of the eye of a hurricane, which allows us to know the area where the eye of the storm will go, it does not given any information about the intensity of a storm, the size of a storm, or even if a location will be hit.
-This inclusion of determinism seems to cause the authors to stumble themselves, as they are not consistent with their assumptions.
-In their first experiment @Padilla2017 indicated the correct answer was to assume that the storm was equally intense no matter how far from the centre of the distribution an oil rig was, however answering their third experiment correctly hinged on assuming the intensity of the storm at a particular point (which in this experiment they phrased as damage) *does* change in intensity as you move away from the centre of the distribution.
-Given these conflicting assumptions, it is unclear how the participants were supposed to adjust the probabilistic path information to answer a deterministic question about which oil rig would receive the most damage.
 Other authors have commented on the complexity of communicating hurricane risk because the path, storm surge and wind speed are all important and cannot be ignored [@Spiegelhalter2017].
-The flip side of this is asking participants for a deterministic answer to a probabilistic question.
-@Correll2014 asked participants "how likely is candidate B to win the election?" when the two distributions indicated voter preference.
-Participants were not able to answer the question about likelihood in term of probability, but were instead given seven options from 1=Outcome will be most in favour of A to 7=Outcome will be most in favour of B.
-The ground truth statistic for this question was a scalar multiple of Cohen’s d, indicating participants were supposed to incorporate uncertainty information using a very specific formula that was likely unknown to them but assumed to be used implicitly.
-
-### Identifying heuristics
+Interestingly it is unclear how the participants were supposed to answer this question, as @Padilla2017 did three experiments and the first and third experiment had conflicting assumptions in their ground truth.
+In the first study, participants were required assume to the storm was equally intense reguardless of the probability of the probability of the oil rig being hit (i.e.
+more likely to be hit *does not* mean more damage) but in the second study required participants to assume the opposite (i.e.
+more likely to be hit *does* mean more damage).
+Similar to the vague question problem, these studies seem to be aware that we should be evaluating an uncertainty visualization based on a deterministic observation (such as our example of "does this map have a spatial trend") but are unsure how to incorporate or evaluate it.
+
+While these approaches are certainly a step in the right direction, the experiments end up having far too much noise in their results.
+Additionally, by having a ground truth, they end up implicitly asking users to either treat uncertainty as a signal, or to ignore it entirely, neither of which is advisable for evaluating graphics on signal-suppression.
+So, if we cannot ask direct questions about uncertainty, *and* we can't use uncertainty statistics as ground truths in our experiments, how *should* we evaluate uncertainty?
+
+## Testing signal supression
+
+If a graphic was performing signal suppression we would expect the values to be harder to read at higher levels of uncertainty, something that is often seen as a bad thing in current evaluation studies.
+- trust experiment
+- weird animation of grass
+
+Since the primary goal of visualisation *is* developing insights [@North2006], you will probably extract several signals and use them to make a broader conclusion.
+A simple example is the spatial trend in
+
+If asking direct questions about uncertainty causes us to treat it as a signal, how do we evaluate uncertainty as *noise*?
+ When we ask the viewer of a plot to look at data and extract a value, we are asking them to perform inference on that value.
+There will be noise associated with that answer and that is uncertainty.
+If we ask direct question about some uncertainty metric, we have turned the uncertainty into signal because that is what the participants are drawing inference on.
+
+So, the current methods of measuring or understanding the role of uncertainty in a visualisation is questionable at best, however this is not because visualisation authors are missing the mark, but rather uncertainty is *particularly* difficult to express in a visualisation.
+In simple estimates or verbal communication, the signal is often easy to identify because it is what we are explicitly saying.
+Unlike statistical models, visualisations are used in both data exploration and communication.
+This means what exactly is a *signal* in any particular visualisation is hard to identify, since we often let the visualisation *tell us* what the signal is.
+Additionally, you cannot add noise to *every single possible* signal one might take from a visualisation.
+Two people looking at the same visualisation might, just by chance, develop two entirely different insights and draw inference on two completely different statistics.
+These unique and fascinating challenges that are faced by uncertainty visualisation have been completely untouched by the literature.
+This section will cover some interesting research in uncertainty visualisation and suggestions for better ways to measure uncertainty.
+
+### Qualitative Studies
+Alternatively visualisation research could shift away from the accuracy concept all together ask questions that allow for open ended responses.
+This method can enlighten authors as to *how* the uncertainty information was used by the participants.
+@Hofmann2012 tried to capture this by asking participants why they considered a particular plot to be more "right shifted", however this qualitative assessment does not seem to have made it into the final paper.
+@Daradkeh2015 presented participants with ten investment alternatives and asked participants "from among available alternatives, which alternative do you prefer the most", and were asked to think aloud and consider the uncertainty in their decision making.
+The experimenters goal was to observe and organise the methods people use when making decisions in the face of uncertainty.
+This study was an excellent example in a useful experimental design.
+They highlighted the specific aspects of uncertainty that participants typically considered, such as the range of outcomes that are above/below a certain threshold, minimum and maximum values, the risk of a loss, etc, and mapped where in the decision making process participants made these considerations.
+Data visualisation is commonly utilised as a tool in data exploration, so it is not uncommon for a data analyst to make a plot with only a vague goal and pull out a large number of adjacent observations.
+This experimental framework could replicate this process.
+
+### Just noticeable signal
+
+there are some papers that do identify signal suppress effects, but they discuss it as a failure rather than a success.
+
+It could be argued that a well done uncertainty visualisations should have an imperceptible signal unless the signal would be identified with a hypothesis test, almost like a reverse line up protocol, but this idea also has some issues that should be considered.
+The reject or do not reject concepts in hypothesis testing do not offer a complete image of uncertainty, and exploration of uncertainty visualisation largely stems from a desire to move away from this binary framework.
+
+@Patrick2023 found that human viewers were less sensitive to deviations from the null hypothesis than the typical statistical tests.
+
+What this all means, is that there will always be some trade off when we are designing a visualisation for signal-suppression.
+There will be a trade off in *which* signal we suppress, as well as *how much* we suppress it.
+This is true for any graphic that is comparable in some way to a statistical test or process.
+A good example is the line-up protocol, which is a visualisation tool that used to check if perceived patterns are real or merely the result of chance [@Buja2009; @Wickham2010].
+This motivation is similar to our signal-suppression goal.
+@Patrick2023 compared standard statistical tests to the lineup-protocol, and evaluated the visualisations using the power curves that are typical for hypothesis testing.
+In a similar vein, @Kim2019 investigated how different uncertainty visualisation methods influenced user's prior beliefs, and evaluated the graphics by comparing their results to those from Bayesian inference.
+While these methods apply different statistical philosophies, they both investigated the sensitivity of visualisation methods relative to existing evaluation methods.
+This approach may prove useful when considering
+
+## Identifying heuristics
 - Things that are true for both normal visualisation and uncertainty, but of particular interest for uncertainty
@@
 Small perceptual tasks that seek to answer small but highly relevant questions (for example, if colour hue and colour value can be perceived as a single signal suppressed variable) would be useful to the field.
-
-## Testing signal supression
-
-If a graphic was performing signal suppression we would expect the values to be harder to read at higher levels of uncertainty, something that is often seen as a bad thing in current evaluation studies.
-- trust experiment
-- weird animation of grass
-
-Since the primary goal of visualisation *is* developing insights [@North2006], you will probably extract several signals and use them to make a broader conclusion.
-A simple example is the spatial trend in
-
-If asking direct questions about uncertainty causes us to treat it as a signal, how do we evaluate uncertainty as *noise*?
- When we ask the viewer of a plot to look at data and extract a value, we are asking them to perform inference on that value.
-There will be noise associated with that answer and that is uncertainty.
-If we ask direct question about some uncertainty metric, we have turned the uncertainty into signal because that is what the participants are drawing inference on.
-
-So, the current methods of measuring or understanding the role of uncertainty in a visualisation is questionable at best, however this is not because visualisation authors are missing the mark, but rather uncertainty is *particularly* difficult to express in a visualisation.
-In simple estimates or verbal communication, the signal is often easy to identify because it is what we are explicitly saying.
-Unlike statistical models, visualisations are used in both data exploration and communication.
-This means what exactly is a *signal* in any particular visualisation is hard to identify, since we often let the visualisation *tell us* what the signal is.
-Additionally, you cannot add noise to *every single possible* signal one might take from a visualisation.
-Two people looking at the same visualisation might, just by chance, develop two entirely different insights and draw inference on two completely different statistics.
-These unique and fascinating challenges that are faced by uncertainty visualisation have been completely untouched by the literature.
-This section will cover some interesting research in uncertainty visualisation and suggestions for better ways to measure uncertainty.
-
-### Qualitative Studies
-Alternatively visualisation research could shift away from the accuracy concept all together ask questions that allow for open ended responses.
-This method can enlighten authors as to *how* the uncertainty information was used by the participants.
-@Hofmann2012 tried to capture this by asking participants why they considered a particular plot to be more "right shifted", however this qualitative assessment does not seem to have made it into the final paper.
-@Daradkeh2015 presented participants with ten investment alternatives and asked participants "from among available alternatives, which alternative do you prefer the most", and were asked to think aloud and consider the uncertainty in their decision making.
-The experimenters goal was to observe and organise the methods people use when making decisions in the face of uncertainty.
-This study was an excellent example in a useful experimental design.
-They highlighted the specific aspects of uncertainty that participants typically considered, such as the range of outcomes that are above/below a certain threshold, minimum and maximum values, the risk of a loss, etc, and mapped where in the decision making process participants made these considerations.
-Data visualisation is commonly utilised as a tool in data exploration, so it is not uncommon for a data analyst to make a plot with only a vague goal and pull out a large number of adjacent observations.
-This experimental framework could replicate this process.
-
-### Just noticeable signal
-
-there are some papers that do identify signal suppress effects, but they discuss it as a failure rather than a success.
-
-It could be argued that a well done uncertainty visualisations should have an imperceptible signal unless the signal would be identified with a hypothesis test, almost like a reverse line up protocol, but this idea also has some issues that should be considered.
-The reject or do not reject concepts in hypothesis testing do not offer a complete image of uncertainty, and exploration of uncertainty visualisation largely stems from a desire to move away from this binary framework.
-
-@Patrick2023 found that human viewers were less sensitive to deviations from the null hypothesis than the typical statistical tests.
-
-What this all means, is that there will always be some trade off when we are designing a visualisation for signal-suppression.
-There will be a trade off in *which* signal we suppress, as well as *how much* we suppress it.
-This is true for any graphic that is comparable in some way to a statistical test or process.
-A good example is the line-up protocol, which is a visualisation tool that used to check if perceived patterns are real or merely the result of chance [@Buja2009; @Wickham2010].
-This motivation is similar to our signal-suppression goal.
-@Patrick2023 compared standard statistical tests to the lineup-protocol, and evaluated the visualisations using the power curves that are typical for hypothesis testing.
-In a similar vein, @Kim2019 investigated how different uncertainty visualisation methods influenced user's prior beliefs, and evaluated the graphics by comparing their results to those from Bayesian inference.
-While these methods apply different statistical philosophies, they both investigated the sensitivity of visualisation methods relative to existing evaluation methods.
-This approach may prove useful when considering
 

### 2024-08-07 paper.qmd
finished the testing signal suppression section

@@
 However, it is not uncommon for authors to express a desire for uncertainty visualisations that perform signal-suppression in visualisations made for EDA [@Sarma2024; @Griethe2006].
+For uncertainty visualisation for EDA to work, we would need to assume that suppressing individual estimates using their variance should naturally extend to broader suppression of plot level insights.
 Unfortunately it is not clear whether or not uncertainty visualisations for EDA are possible.
@@
 However, this does mean we should endeavor for a versatile uncertainty visualisation method that is able to perform signal-suppression on all the signals displayed in the visualisation.
-It is unlikely that the methods discussed thus far are able to achieve this.
+
+Uncertainty visualisation for EDA would be possible if we designed a plot in such a way that suppressing individual estimates using their variance would naturally extend to broader suppression of plot level insights.
+This assumption is commonly made by visualisation researchers in normal visualisation experiments [@North2006], however achieving it would largely depend on the methods use to perform signal suppression.
+By manually combining the values in @fig-vsup, we violated this requirement, so the VSUP is not versatile enough to act as an uncertainty visualisation for EDA.
 
@@
 We then do a comparison of two means t-test and find that several of the values need to be visually distinguishable from each other.
-The VSUP method must pick a single color for each estimate, and these colours must be *either* visually distinguishable or indistinguishable from each other.
+The VSUP method must pick a single color for each estimate, and these colors must be *either* visually distinguishable or indistinguishable from each other.
 We cannot perform signal-suppression on both these signals simultaneously.
@@
 ### Value extraction of uncertainty statistics
-Uncertainty visualisations are most commonly evaluated on how accurately readers can extract information from them [@Hullman2019].
-This most commonly comes in the form of separate questions about estimates variance [@Kinkeldey2014].
+Uncertainty visualisations are often evaluated based on how accurately [@Hullman2019] viewers can separately extract the estimate and the variance [@Kinkeldey2014].
 This means a significant chunk of evaluation studies boil down to showing a participant a visualisation and asking questions such as "what is the variance of $X$?", or "what is the mean of $X$?".
-Additionally, uncertainty visualisation experiments often compare graphics on the basis of being "uncertainty visualisations" [@Ibrekk1987; @Hullman2015; @Hofman2020] regardless of what information can actually be extracted by the chart.
 This seems like a relatively straight forward approach, and it is similar to how non-uncertainty visualisations are evaluated, but is this appropriate for uncertainty visualisations?
-Uncertainty is not something we are trying to infer, uncertainty is a by-product of inference.
-By shifting the focus of our inference to $Var(X)$ or $P(X)$ we are just evaluating visualisations on their ability to convey uncertainty statistics.
-These graphics may not always set out to communicate uncertainty as a signal, but they are certainty evaluating uncertainty as a signal.
+The role of uncertainty is rarely evaluated in these studies as the graphics are often compared on the basis of being "uncertainty visualisations" [@Ibrekk1987; @Hullman2015; @Hofman2020], a class that has no established definition.
+By shifting the focus of our inference from $\bar{X}$ to $Var(X)$ or $P(X)$ we end up evaluating visualisations on their ability to convey uncertainty statistics, rather than on uncertainties ability to suppress statistics.
+This leads to a series of experiments where the uncertainty is evaluated as a signal even if that was not the goal of the experiment.
 
@@
 
-Directly asking about uncertainty evaluates uncertainty as a signal, not as noise, so it is not a reliable evaluation method when we are designing graphics for signal suppression.
-Many uncertainty visualisation researchers are already aware of this, so direct questions are not the only methods used to evaluate uncertainty visualisations.
-We will discuss these alternatives in the following section.
+The biggest failing of this evaluation method is not the predictable outcomes, but that it encourages us to see successful examples of signal-suppression as a failing.
+@Blenkinsop2000 commented that visually integrable depictions of uncertainty should be avoided, as they decrease the viewers confidence in their extracted data values.
+This conclusion is antithetical to the goals of signal-suppression and occurs because these methods evaluate uncertainty as a signal, not as noise.
 
@@
 Setting trust as the variable of interest implicitly encourages statisticians to set trust and as the primary goal of communication.
+Evaluating visualisations on trust conflates trust and transparency and ultimately discourages signal-suppression.
+We can see this effect pop up in the results of evaluation studies.
+For example, @Zhao2023 found that participants were more trusting of model estimates with low uncertainty, but this effect did not carry over to estimates with high uncertainty.
+Despite decreased trust being a desired outcome of signal suppression, the authors discussion implied this result was not desired.
+This perspective ends up extending to visualisation authors as well.
 @Hullman2020a found that author simultaneously argued that failing to visualise uncertainty was akin to fraud, but also many avoided uncertainty visualisation because they didn't want their work to come across as "untrustworthy".
@@
 In cases of high uncertainty, authors will opt to leave out uncertainty information because it decreases confidence in the authors conclusions.
-
-The disconnect between trust and transparency extends to evaluation experiments as well.
-@Zhao2023 displayed a several variations of a visualisation of a model prediction and its uncertainty and took participants using the model prediction as a sign of trust.
-They reported that visualising uncertainty information caused participants to trust the model in the low variance case, but the results in the high variance case were inconclusive.
-The discussion made it clear the authors thought the uncertainty information should make the visualisation more trustworthy.
-However, this perspective conflates transparency and trustworthiness.
-If a model has very high uncertainty because of a very low sample size, the heuristics a human might use to make the same decision could be more accurate.
-
-Unfortunately, experiments where the best response comes from participants who outright ignore the uncertainty information are not uncommon.
-A simple example of this is experiments where participants are expectd to extract the mean of the distribution.
-In these cases, we would expect the best performing visualisation to be one that does not include uncertainty at all, as it would simply show the mean.
-Similarly, many uncertainty visualisations are evaluated on how much risk aversion the participants display [@Hullman2019].
-Additionally, comparing participant responses to that of a ration agent has been suggested as a benchmark for uncertainty visualization experiments [@wu2023rational].
-This standard seems to go against the purpose of uncertainty visualisation as a rational agent, *by definition* should *ignore* uncertainty information.
-Risk aversion is considered to be irrational because it means the economic agent *is* considering the uncertainty information.
-The only cases where a rational agent should not ignore the uncertainty information is when it is required to optimise their utility function, which would mean the uncertainty is not acting as noise, it is acting as signal.
+This is the end result of designing uncertainty visualisations for increased trust, rather than signal-suppression.
 
@@
 
+Risk-aversion is another secondary effect of uncertainty visualization that is used to evaluate uncertainty visualisations [@Hullman2019].
+Risk aversion is an economics term used to describe an agent who would chose a random variable with a lower expected payout because it also has a lower variance.
+Risk aversion is considered irrational behaviour because it is a deviation from the behavior of a rational agent.
+Comparing participant responses to that of a ration agent has even been suggested as a benchmark for uncertainty visualization experiments [@wu2023rational].
+However, just like trust, evaluating graphics on risk-aversion discourages signal-suppression.
+Rational agents *by definition* should *ignore* uncertainty information.
+Risk aversion is considered to be irrational because it means the economic agent *is* considering the uncertainty information.
+The only cases where a rational agent should not ignore the uncertainty information is when the uncertainty is signal, not noise.
+Designing graphics that encourage choices that align with that of a rational agent, is to encourage graphics that do not include uncertainty at all.
+
 For these reasons we do not believe trust, confidence, or risk-aversion are useful measures to evaluate uncertainty visualisations.
-While they are designed to capture the secondary effects of uncertainty, using them as primary measures of visualisation performance incentives the use of visualisations that *do not* include uncertainty.
+While they are designed to capture the secondary effects of uncertainty, using them as primary measures of visualisation performance is in direct conflict with designing visualisations for signal-suppression.
 
@@
 Additionally, by having a ground truth, they end up implicitly asking users to either treat uncertainty as a signal, or to ignore it entirely, neither of which is advisable for evaluating graphics on signal-suppression.
-So, if we cannot ask direct questions about uncertainty, *and* we can't use uncertainty statistics as ground truths in our experiments, how *should* we evaluate uncertainty?
 
 ## Testing signal supression
-
-If a graphic was performing signal suppression we would expect the values to be harder to read at higher levels of uncertainty, something that is often seen as a bad thing in current evaluation studies.
-- trust experiment
-- weird animation of grass
-
-Since the primary goal of visualisation *is* developing insights [@North2006], you will probably extract several signals and use them to make a broader conclusion.
-A simple example is the spatial trend in
-
-If asking direct questions about uncertainty causes us to treat it as a signal, how do we evaluate uncertainty as *noise*?
- When we ask the viewer of a plot to look at data and extract a value, we are asking them to perform inference on that value.
-There will be noise associated with that answer and that is uncertainty.
-If we ask direct question about some uncertainty metric, we have turned the uncertainty into signal because that is what the participants are drawing inference on.
-
-So, the current methods of measuring or understanding the role of uncertainty in a visualisation is questionable at best, however this is not because visualisation authors are missing the mark, but rather uncertainty is *particularly* difficult to express in a visualisation.
-In simple estimates or verbal communication, the signal is often easy to identify because it is what we are explicitly saying.
-Unlike statistical models, visualisations are used in both data exploration and communication.
-This means what exactly is a *signal* in any particular visualisation is hard to identify, since we often let the visualisation *tell us* what the signal is.
-Additionally, you cannot add noise to *every single possible* signal one might take from a visualisation.
-Two people looking at the same visualisation might, just by chance, develop two entirely different insights and draw inference on two completely different statistics.
-These unique and fascinating challenges that are faced by uncertainty visualisation have been completely untouched by the literature.
-This section will cover some interesting research in uncertainty visualisation and suggestions for better ways to measure uncertainty.
+If we cannot ask direct questions about uncertainty, we cant measure the secondary effects of uncertainty, and we can't ask indirect questions about the uncertainty, how are we supposed to evaluate uncertainty visualisations?
+How do you measure something that disappears the second you look directly at it?
+To evaluate uncertainty visualisations, we need an experimental design that evaluates uncertainty as noise, not as signal.
+This means we need to measure *uncertainty's impact on the signal*, not the uncertainty itself.
+
+### Comparing to hypothesis tests
+The most obvious way to evaluate uncertainty visualizations is to compare the visualisations to statistical tests.
+If a graphic was performing signal-suppression we would expect the signal to be harder to see at higher levels of uncertainty.
+The ideal outcome is a an uncertainty visualisation where the signal is only perceivable when it would be identified by a hypothesis test.
+Evaluating visualisations as though they are akin to hypothesis tests is a well established concept in visualisation.
+The lineup protocol is a good example of this approach, which is a confirmatory visualisation tool that can be used to check if perceived patterns are real or merely the result of chance [@Buja2009; @Wickham2010].
+The motivation behind the lineup protocol is similar to the motivations behind signal-suppression, although the lineup protocol is more explicitly tied to a specific hypothesis.
+@Patrick2023 compared standard statistical tests to the lineup-protocol, and evaluated the visualisations using the power curves that are typical for hypothesis testing.
+In a similar vein, @Kim2019 investigated how different uncertainty visualisation methods influenced user's prior beliefs, and evaluated the graphics by comparing their results to those from Bayesian inference.
+This approach is similar to the lineup protocol as it also evaluates a visualisation by comparing it to an analogous statistical calculation, however it does so using a different statistical philosophy.
+
+We can use a similar evaluation method to evaluate uncertainty visualisations, however, special care would need to be taken for their versatility requirement.
+While lineup protocols often have a specific hypothesis that is used to create the null distribution, uncertainty visualisations are intended to be more flexible and can be created without visualizing a specific null distribution.
+Human viewers using the lineup protocol are less sensitive to deviations from the null hypothesis than the typical statistical tests [@Patrick2023].
+This means swapping to a visual format and giving viewers the ability to check several hypothesis at once can reduce the effectiveness of the hypothesis test.
+Therefore it is highly likely that uncertainty visualisations would not produce identical results to standard hypothesis tests or even the lineup-protocol.
+Additionally, uncertainty visualisations would need to be simultaneously checked against several hypothesis that come from a variety of null distributions.
 
@@
 This method can enlighten authors as to *how* the uncertainty information was used by the participants.
-@Hofmann2012 tried to capture this by asking participants why they considered a particular plot to be more "right shifted", however this qualitative assessment does not seem to have made it into the final paper.
+@Hofmann2012 tried to capture this by asking participants why they considered a particular plot to be more "right shifted", even though this qualitative analysis did not make it into the final paper.
 @Daradkeh2015 presented participants with ten investment alternatives and asked participants "from among available alternatives, which alternative do you prefer the most", and were asked to think aloud and consider the uncertainty in their decision making.
 The experimenters goal was to observe and organise the methods people use when making decisions in the face of uncertainty.
-This study was an excellent example in a useful experimental design.
-They highlighted the specific aspects of uncertainty that participants typically considered, such as the range of outcomes that are above/below a certain threshold, minimum and maximum values, the risk of a loss, etc, and mapped where in the decision making process participants made these considerations.
-Data visualisation is commonly utilised as a tool in data exploration, so it is not uncommon for a data analyst to make a plot with only a vague goal and pull out a large number of adjacent observations.
-This experimental framework could replicate this process.
-
-### Just noticeable signal
-
-there are some papers that do identify signal suppress effects, but they discuss it as a failure rather than a success.
-
-It could be argued that a well done uncertainty visualisations should have an imperceptible signal unless the signal would be identified with a hypothesis test, almost like a reverse line up protocol, but this idea also has some issues that should be considered.
-The reject or do not reject concepts in hypothesis testing do not offer a complete image of uncertainty, and exploration of uncertainty visualisation largely stems from a desire to move away from this binary framework.
-
-@Patrick2023 found that human viewers were less sensitive to deviations from the null hypothesis than the typical statistical tests.
-
-What this all means, is that there will always be some trade off when we are designing a visualisation for signal-suppression.
-There will be a trade off in *which* signal we suppress, as well as *how much* we suppress it.
-This is true for any graphic that is comparable in some way to a statistical test or process.
-A good example is the line-up protocol, which is a visualisation tool that used to check if perceived patterns are real or merely the result of chance [@Buja2009; @Wickham2010].
-This motivation is similar to our signal-suppression goal.
-@Patrick2023 compared standard statistical tests to the lineup-protocol, and evaluated the visualisations using the power curves that are typical for hypothesis testing.
-In a similar vein, @Kim2019 investigated how different uncertainty visualisation methods influenced user's prior beliefs, and evaluated the graphics by comparing their results to those from Bayesian inference.
-While these methods apply different statistical philosophies, they both investigated the sensitivity of visualisation methods relative to existing evaluation methods.
-This approach may prove useful when considering
-
-## Identifying heuristics
-- Things that are true for both normal visualisation and uncertainty, but of particular interest for uncertainty
-
-Additionally, these heuristics and biases can change depending on the larger scope of the graphic and the population we are communicating with [@Spiegelhalter2017; @Kinkeldey2014].
-
-Heuristic checks are useful because they look at unknown pitfalls that might exist in interpretation of current plots.
-Since the hypothesis for these experiments are usually quite specific, e.g.
-"do people perceive a an outcome that is within the bar as more likely than one outside it, even if both outcomes are the same distance from the mean?".
-This means they are less likely to fall into the trap of trying to answer questions that are *far* too broad to be answered with a single experiment (e.g.
-"is a scatter plot better at showing uncertainty than a box plot").
-This work also provides useful insights for experiments by highlights pitfalls participants might fall into when they review the results of evaluation experiments [@Hullman2016].
-@Newman2012 found that participants were more likely to view points within the bar as more likely than points outside of the bar in bar charts with error bars.
-Similar effects have been identified in other types of uncertainty displayed.
-@Padilla2017 found that points that were on an outcome of an ensemble display were perceived as more likely than points not on an outcome, even when the point that was not on a specific outcome of the ensemble was closer to the mean of the uncertainty distribution.
+They highlighted the specific aspects of uncertainty that participants typically considered, such as the range of outcomes that are above/below a certain threshold, minimum and maximum values, the risk of a loss, etc, and identified where in the decision making process participants made these considerations.
+
+### Heuristics
+While experiments that explicitly identify heuristics in current methods are not technically measuring signal-suppression, they are still a useful consideration to keep in mind when designing experiments for signal suppression.
+Heuristic checks are look at unknown pitfalls that might exist in interpretation of current plots [@Hullman2016] and can change depending on the larger scope of the graphic and the population we are communicating with [@Spiegelhalter2017; @Kinkeldey2014].
+
+Several heuristics are of particular importance for uncertainty visualisations, and are likely to impact how well different methods performing signal suppression.
 The sine illusion can cause the confidence interval of a smoothed sine curve to seem wider at the peaks than the troughs, causing us to underestimate uncertainty associated with changing values [@Vanderplas2015].
-
-In a similar vein, experiments that verify smaller aspects of plot design might be more useful to the field in the long run because it helps contribute to a larger working theory of "how do we see visualisations".
-Many visualisation experiments try to compare two plots with several differences, but do not seem to be interested in the mechanisms by which we extract information from visualisations.
-Small perceptual tasks that seek to answer small but highly relevant questions (for example, if colour hue and colour value can be perceived as a single signal suppressed variable) would be useful to the field.
+Points that were on an outcome of an ensemble display were perceived as more likely than points not on an outcome, even when the point that was not on an outcome was closer to the center of the distribution (and therefore more likely) [@Padilla2017].
+This can be considered an extension of the within bar bias, where participants looking at a bar chat with error bars view outcomes within the bar as more likely than those outside it [@Newman2012].
+Several studies have found that viewers use a heuristic where they compare the distance between two estimates to estimate if they are different, and in doing so, ignore the uncertainty information [@uncertchap2022].
+
+These heuristics have the potential to create noise in signal suppression evaluations, and should be kept in mind when designing an evaluation experiment.
 

### 2024-08-08 paper.qmd
started temperature description

@@
 
+## Example Data
+Throughout this review, we will be looking at
+
+```{r}
+#| echo: false
+#| message: false
+#| warning: false
+
+# seed for sampling
+set.seed(1997)
+
+# Get map data
+my_map_data <- get_urbn_map("counties", sf = TRUE) |>
+  dplyr::filter(state_name=="Iowa")
+centroids <- as_tibble(rgeos::gCentroid(as(my_map_data$geometry, "Spatial"), byid = TRUE))
+my_map_data$cent_long <- centroids$x
+my_map_data$cent_lat <- centroids$y
+n <- dim(my_map_data)[1]
+
+# Make standard Palette
+# set Base colours
+basecols <- brewer.pal(8, name = "YlOrRd")
+#breaks <- seq(0,41, length.out = 9)
+breaks <- 21:29 #c(0,15,24,25,26,27,30,35,41)
+names(basecols) <- seq(8)
+limits <- c(10,41)
+
+my_map_data <- my_map_data |>
+  dplyr::mutate(temp = 28 - ((scale(cent_long))^(2) + (scale(cent_lat))^(2))[,1], # trend
+         notrend = rnorm(n), # no trend
+         highvar = runif(n, min=2, max=4), # high variance
+         lowvar = runif(n, min=0, max=2), # low variance
+         ) |>
+  pivot_longer(cols=highvar:lowvar, names_to = "variance_class", values_to = "variance") |>
+  # add bivariate classes to data
+  mutate(bitemp = cut(temp, breaks=breaks, labels=seq(8)),
+         bivar = cut(variance, breaks=0:4, labels=seq(4)),
+         biclass = paste(bitemp, bivar, sep="-")
+  )
+
+```
 Due to the fields origins and focus in geospatial information visualisation, there have been a large number of suggested variations on the choropleth map that allow authors to include uncertainty.
@@
 #| cap-location: "bottom"
-
-# seed for sampling
-set.seed(1997)
-
-# Get map data
-my_map_data <- get_urbn_map("counties", sf = TRUE) |>
-  dplyr::filter(state_name=="Iowa")
-centroids <- as_tibble(rgeos::gCentroid(as(my_map_data$geometry, "Spatial"), byid = TRUE))
-my_map_data$cent_long <- centroids$x
-my_map_data$cent_lat <- centroids$y
-n <- dim(my_map_data)[1]
-
-# Make standard Palette
-# set Base colours
-basecols <- brewer.pal(8, name = "YlOrRd")
-#breaks <- seq(0,41, length.out = 9)
-breaks <- 21:29 #c(0,15,24,25,26,27,30,35,41)
-names(basecols) <- seq(8)
-limits <- c(10,41)
-
-my_map_data <- my_map_data |>
-  dplyr::mutate(temp = 28 - ((scale(cent_long))^(2) + (scale(cent_lat))^(2))[,1], # trend
-         notrend = rnorm(n), # no trend
-         highvar = runif(n, min=2, max=4), # high variance
-         lowvar = runif(n, min=0, max=2), # low variance
-         ) |>
-  pivot_longer(cols=highvar:lowvar, names_to = "variance_class", values_to = "variance") |>
-  # add bivariate classes to data
-  mutate(bitemp = cut(temp, breaks=breaks, labels=seq(8)),
-         bivar = cut(variance, breaks=0:4, labels=seq(4)),
-         biclass = paste(bitemp, bivar, sep="-")
-  )
 

### 2024-08-08 paper.qmd
added notes
wrote future work

--- 2024-08-07 paper.qmd
@@
 
+## Example Data
+Throughout this review, we will be looking at
+
+```{r}
+#| echo: false
+#| message: false
+#| warning: false
+
+# seed for sampling
+set.seed(1997)
+
+# Get map data
+my_map_data <- get_urbn_map("counties", sf = TRUE) |>
+  dplyr::filter(state_name=="Iowa")
+centroids <- as_tibble(rgeos::gCentroid(as(my_map_data$geometry, "Spatial"), byid = TRUE))
+my_map_data$cent_long <- centroids$x
+my_map_data$cent_lat <- centroids$y
+n <- dim(my_map_data)[1]
+
+# Make standard Palette
+# set Base colours
+basecols <- brewer.pal(8, name = "YlOrRd")
+#breaks <- seq(0,41, length.out = 9)
+breaks <- 21:29 #c(0,15,24,25,26,27,30,35,41)
+names(basecols) <- seq(8)
+limits <- c(10,41)
+
+my_map_data <- my_map_data |>
+  dplyr::mutate(temp = 28 - ((scale(cent_long))^(2) + (scale(cent_lat))^(2))[,1], # trend
+         notrend = rnorm(n), # no trend
+         highvar = runif(n, min=2, max=4), # high variance
+         lowvar = runif(n, min=0, max=2), # low variance
+         ) |>
+  pivot_longer(cols=highvar:lowvar, names_to = "variance_class", values_to = "variance") |>
+  # add bivariate classes to data
+  mutate(bitemp = cut(temp, breaks=breaks, labels=seq(8)),
+         bivar = cut(variance, breaks=0:4, labels=seq(4)),
+         biclass = paste(bitemp, bivar, sep="-")
+  )
+
+```
+
+- add map that is just outline
+- include function used to design temperature
+
 Due to the fields origins and focus in geospatial information visualisation, there have been a large number of suggested variations on the choropleth map that allow authors to include uncertainty.
@@
 This paper has identified gaps in the uncertainty visualisation literature that could be filled to progress the field.
-
-*Each new development should be accompanied by a mathematical definition of the uncertainty being addressed.* Ideally, a mathematically definition of uncertainty that allows us to combine these components would be developed, but in the absence of that, authors should be more specific about what aspect of "uncertainty" they are covering with their visualisation.
-
-*The concept of uncertainty should be formalised within the grammar of graphics.* This formalisation would allow uncertainty visualisation authors to have a clear understanding of what is or is not an uncertainty visualisation.
-Additionally placing uncertainty visualisation in the framework that is used to understand existing information visualisation research would help authors understand when existing methods can be used to explain their results.
-incorporating uncertainty into the grammar of graphics will also give a more precise concept of the information contained within a plot.
-Other fields of science employ marginal changes when designing experiments to ensure it is well understood *what* aspect of their experiment is contributing to their results, and a better sense of what "marginal" is in the case of uncertainty visualisation would greatly help the field.
-(*XXX Is data pipeline connected with the grammar of graphics?
-Should this be a recommendation?*)
-
-*Experimental practices on uncertainty visualisation need to be standardised.* If we are going to consider uncertainty as noise, not signal, there needs to be a way to identify this signal suppression in an experimental design.
+We formalized the uncertainty visualisation problem, and in doing so highlighted an untouched area of uncertainty visualisation research.
+
+*Experimental practices on uncertainty visualisation need to be standardised.* Uncertainty visualisations for decision making treat uncertainty as signal, while visualisations for signal-suppression treat uncertainty as noise.
 As the literature currently exists, there is no way to combine papers to get a meaningful sense of how uncertainty information is understood by a viewer.
-There is also the possibility that uncertainty visualisation evaluations will need to swap to a qualitative methodology where participants are allowed to freely comment on what they notice in graphics until we establish how the existence of noise can be observed.
-
-If an uncertainty visualisation researcher would prefer to perform experiments rather than formalise methods, there are options there too.
-It would be interesting to know if any perceptual tasks that can be mapped to two different visual tasks condense into a single dimension when looking for overarching signal in a plot.
-Alternatively, the task dependency many authors in uncertainty visualisation mention would be a useful direction to consider.
-It is clear that the the number of potential tasks that can be performed on a visualisation increases with with the number of observations.
-A single observation is limited to value extraction, two observations can be compared, multiple observations allow for shapes or global statistics to be extracted.
-The interaction between sample size and task is of particular interest to the uncertainty visualisation community, as uncertainty can be expressed through multiple observations using a sample, or through a single value using an error.
-Of course, this is limited by the fact that there also isn't a definition for what is a "task" and given the mess created by the lack of formalisation in uncertainty visualisation, it may be wise to formalise that concept before performing these experiments.
-@Amar2005 suggested a taxonomy for information visualisation based on the types of tasks we use visualisations for and suggest 10 "analytical primitives" that we can then map to visualisations, which could be a good starting point.
-Regardless, these are directions of research would be fruitful to the uncertainty visualisation community even if it appears on the surface to be research that is only beneficial to the "normal" visualisation community.
-(*XXX Not sure what this paragraph is recommending?*)
+Researchers need to ensure that when they identify the motivation behind their visualisation technique, and that their evaluation methods match the motivations of the paper.
+Additionally *evaluation methods that evaluate uncertainty as noise need to be developed.*
+
+*Experimenters should consider evaluating visual aesthetics on how well they suppress information in a graphic.* Research into separability and integrability is of particular interest to uncertainty visualisation, as it allows interference from the uncertainty variable.
+When designing experiments, authors often choose aesthetics that are visually distinguishable, uncertainty visualisation authors should consider doing the opposite.
+
+*Software that allows users to easily perform signal suppression should be developed.* Existing uncertainty visualisation methods view a distribution as its own object and there are no software options for the "an uncertainty visualisation is a function of a visualisation" philosophy.
 

### 2024-08-08 writingdumpster.qmd
wrote future work

@@
 # Removed Sections
+@Amar2005 suggested a taxonomy for information visualisation based on the types of tasks we use visualisations for and suggest 10 "analytical primitives" that we can then map to visualisations, which could be a good starting point.
+
 The beauty of visualisation is that it *tells* us what we should be concerned about in our analysis. This aspect disappears the second we ask a direct question about our visualisation.

### 2024-08-19 paper.qmd
spent ages making a GT table and it doesnt even work in the PDF file. I hate this.

@@
 library(urbnmapr)
+library(gt)
 ```
@@
 
-## Example Data
-Throughout this review, we will be looking at
+## Spatial examples
+There is an endless number of uncertainty visualisations and it would be impossible to show all of them in this review.
+While the theoretical approach we outline is useful to all uncertainty visualisations, regardless of their application, we feel that showing the variations on a single visualisation will help isolate the ideas we are trying to convey.
+Due to the fields origins and focus in geospatial information visualisation, there have been a large number of suggested variations on the choropleth map that allow authors to include uncertainty.
+We will use these maps to highlight the costs and benefits of each approach.
+
+@fig-data shows the first six rows of our example data, along with the geographical boundaries of our data set.
+The temperature variable was generated using a quadratic spatial trend, that is $Temperature = 28-SLongitude^2 + SLatitude^2$ where the $SLongitide$ and $SLatitude$ are the scaled (to standard normal) longitude and latitude of the county's centroid.
+The high and low cases for the sampling variance are randomly sampled from a uniform distribution.
+The low standard errors are drawn from a $U_{[0,1]}$ distribution, while the high standard error cases are drawn from a a $U_{[1,2]}$ distribution.
+This is the data we will be using in our spatial uncertaity examples for the rest of the paper.
 
@@
 #| warning: false
+#| label: fig-data
+#| layout-nrow: 1
+#| fig-cap: "The"
+#| cap-location: "bottom"
 
@@
          notrend = rnorm(n), # no trend
-         highvar = runif(n, min=2, max=4), # high variance
-         lowvar = runif(n, min=0, max=2), # low variance
+         highvar = runif(n, min=1, max=2), # high variance
+         lowvar = runif(n, min=0, max=1), # low variance
          ) |>
+  mutate(highvar = highvar^2,
+         lowvar = lowvar^2,
+         count_id = row_number()) |>
   pivot_longer(cols=highvar:lowvar, names_to = "variance_class", values_to = "variance") |>
@@
          bivar = cut(variance, breaks=0:4, labels=seq(4)),
-         biclass = paste(bitemp, bivar, sep="-")
-  )
+         biclass = paste(bitemp, bivar, sep="-"))
+
+# Make nice table
+example_table <- my_map_data |>
+  mutate(variance = sqrt(variance)) |>
+  select(c(count_id, county_name, temp, variance_class, variance)) |>
+  as_tibble()|>
+  pivot_wider(id_cols=c(count_id, county_name, temp,), 
+              names_from = variance_class, 
+              values_from = variance) |>
+  head()|>
+  gt() |>
+   tab_header(
+    title = "Average Daily High Temperatures of Iowa Counties",
+    subtitle = "Psudo-data with high standard error and low standard error cases") |>
+  tab_spanner(
+    label = "Standard Error",
+    columns = c(highvar, lowvar)
+  ) |>
+  cols_label( 
+    count_id = "ID", 
+    county_name  = "County",
+    temp = "Temperature (°C)",
+    highvar = "High",
+    lowvar = "Low"
+    ) |>
+  fmt_number(columns = c(temp, highvar, lowvar)) |>
+  data_color(
+    columns = temp,
+    method = "numeric",
+    palette = "Oranges",
+    domain = c(21, 29)
+  ) |>
+  data_color(
+    columns = c(highvar, lowvar),
+    method = "numeric",
+    palette = "Greens",
+    domain = c(0, 3)
+  ) |>
+  as_latex()
+
+# make blank map 
+example_map <- my_map_data |>
+  filter(variance_class=="lowvar") |>
+  ggplot() +
+  geom_sf(aes(geometry = geometry), fill="white") + 
+  #geom_text(aes(x=cent_long, y=cent_lat, label=count_id), size=2) +
+  theme_void()
+
+example_table
+
+example_map
 
 ```
-
-- add map that is just outline
-- include function used to design temperature
-
-Due to the fields origins and focus in geospatial information visualisation, there have been a large number of suggested variations on the choropleth map that allow authors to include uncertainty.
-We will use these maps to provide simple examples for each approach that can be easily compared to the "no uncertainty" choropleth map to better understand the costs and benefits of each approach.
-Despite the example of each method focusing on variations of the choropleth map, it is important to understand that the approaches we are discussing are universal and are not unique to maps.
 

### 2024-08-19 writingdumpster.qmd
added gt table to writing dumpster

@@
 ```
+
+# GT table code
+```{r}
+#| echo: false
+#| message: false
+#| warning: false
+# Make nice table
+example_table <- my_map_data |>
+  mutate(variance = sqrt(variance)) |>
+  select(c(count_id, county_name, temp, variance_class, variance)) |>
+  as_tibble()|>
+  pivot_wider(id_cols=c(count_id, county_name, temp,), 
+              names_from = variance_class, 
+              values_from = variance) |>
+  head()|>
+  gt() |>
+   tab_header(
+    title = "Average Daily High Temperatures of Iowa Counties",
+    subtitle = "Psudo-data with high standard error and low standard error cases") |>
+  tab_spanner(
+    label = "Standard Error",
+    columns = c(highvar, lowvar)
+  ) |>
+  cols_label( 
+    count_id = "ID", 
+    county_name  = "County",
+    temp = "Temperature (°C)",
+    highvar = "High",
+    lowvar = "Low"
+    ) |>
+  fmt_number(columns = c(temp, highvar, lowvar)) |>
+  data_color(
+    columns = temp,
+    method = "numeric",
+    palette = "Oranges",
+    domain = c(21, 29)
+  ) |>
+  data_color(
+    columns = c(highvar, lowvar),
+    method = "numeric",
+    palette = "Greens",
+    domain = c(0, 3)
+  ) |>
+  as_latex()
+```

### 2024-08-19 paper.qmd
replaced DT with flextable. Still have formatting issues. I'll deal with them later

@@
 library(urbnmapr)
-library(gt)
+library(flextable)
 ```
@@
 The low standard errors are drawn from a $U_{[0,1]}$ distribution, while the high standard error cases are drawn from a a $U_{[1,2]}$ distribution.
-This is the data we will be using in our spatial uncertaity examples for the rest of the paper.
+As we are dealing with an average, the sampling distribution would be normally distributed, so each county temperature estimate is assumed to come from a $N(Temp_i, SE_{case,i})$.
+This is the data we will be using in our spatial uncertainty examples for the rest of the paper.
 
@@
 #| label: fig-data
-#| layout-nrow: 1
 #| fig-cap: "The"
@@
 
-# Make nice table
+# Make nice example table
 example_table <- my_map_data |>
@@
   head()|>
-  gt() |>
-   tab_header(
-    title = "Average Daily High Temperatures of Iowa Counties",
-    subtitle = "Psudo-data with high standard error and low standard error cases") |>
-  tab_spanner(
-    label = "Standard Error",
-    columns = c(highvar, lowvar)
-  ) |>
-  cols_label( 
-    count_id = "ID", 
+  flextable() |>
+  set_caption(caption = "Average Daily High Temperatures of Iowa Counties") |>
+  add_header_row(colwidths = c(3, 2),
+                 values = c("", "Standard Error")) |>
+  colformat_double(digits = 2) |>
+  set_header_labels(count_id = "ID", 
                     county_name  = "County",
@@
                     highvar = "High",
-    lowvar = "Low"
+                    lowvar = "Low") |>
+  theme_vanilla() |>
+  vline(i=c(1,2), j=3, part="header")|>
+  align(align = "left", part = "all") |>
+  bg(j = "temp",
+    bg = col_numeric(palette = brewer.pal(8, name = "Oranges"),
+                     domain = c(21, 30)),
+    part = "body"
   ) |>
-  fmt_number(columns = c(temp, highvar, lowvar)) |>
-  data_color(
-    columns = temp,
-    method = "numeric",
-    palette = "Oranges",
-    domain = c(21, 29)
-  ) |>
-  data_color(
-    columns = c(highvar, lowvar),
-    method = "numeric",
-    palette = "Greens",
-    domain = c(0, 3)
-  ) |>
-  as_latex()
+  bg(j = c("highvar", "lowvar"),
+    bg = col_numeric(palette = brewer.pal(8, name = "Greens"),
+                     domain = c(0, 3)),
+    part = "body"
+  )
 

### 2024-08-19 paper.qmd
fixed table issue (it is what it is)

@@
 #| label: fig-data
-#| fig-cap: "The"
+#| fig-cap: "The the first 5 observations of the data used for the spatial uncertainty examples along with the boundaries of each county.
+The map boundaries are the Iowa county boundaries, however the 'temperature' data is not representative of the average temperature in Iowa.
+The temperature and standard error represent the average of the daily high temperature and the standard error of that average respectively."
+#| fig-subcap: 
+#|   - "Data Table"
+#|   - "Map Boundaries"
+#| layout-nrow: 1
 #| cap-location: "bottom"
@@
               values_from = variance) |>
-  head()|>
+  head(5)|>
   flextable() |>
@@
                     county_name  = "County",
-                    temp = "Temperature (°C)",
+                    temp = "Average Temperature (°C)",
                     highvar = "High",
@@
 
-example_table
+plot(example_table)
 

### 2024-08-22 paper.qmd
read over + edited background

@@
 From entertainment choices to news articles to insurance plans, the modern citizen is so over run with information in every aspect of their life it can be overwhelming.
-In this overflow of information, tools that can effectively summarize information down into simple and clear ideas become more valuable.
-Information visualisations remain one of the most powerful tools for fast and reliable science communication.
-
-Visualization is an important step in exploratory data analysis and it is often utilised to **learn** what is important about a data set.
-Datasets such as Anscombe's quartet [@anscombe] or the Datasaurus Dozen [@datasaurpkg] highlight this power in visualisation.
+In the face of this overflow of information, tools that effectively summarize information down into simple and clear ideas become more valuable.
+
+Among these summary tools, information visualisations are one of the most powerful.
+Visualisations allow us to **learn** what is important about our data set and datasets such as Anscombe's quartet [@anscombe] or the Datasaurus Dozen [@datasaurpkg] highlight this power.
 Additionally, visualisations allow for efficient and memorable communication.
@@
 Early papers that specifically reference "uncertainty visualisation" appear in the late 80s [@Ibrekk1987], with geospatial information visualisation literature in the early 90s declaring this to be essential aspect of information display [@MacEachren1992; @Carr1992].
-These early experiments typically involved showing participants a distribution, such as those depicted in @fig-ibrekk, and asking the viewers to extract a probability or average.
+@fig-ibrekk depicts an example of the uncertainty visualisations discussed in these early papers.
 Despite the new terminology visualisation of uncertainty has been present since the earliest times.
@@
 #| fig-cap: "A replication of the the uncertainty visualisations shown by @Ibrekk1987 in one of the earliest uncertainty visualisation experiments.
-This early experiment is a good example of many of the issues that are still common in uncertainty visualisation today.
-For example, the '95% confidence interval' is more accurately a '95% prediction interval'.
-Additionally graphics that depict different mathematical objects that are also different on their visual components are compared because of a percieved relation to uncertainty.
-So visualisations of the mean, PDF and CDF, are all discussed as though they all contain the relevant statistical information.
-The axis also have different scales, and visualisation methods that are now unpopular for displaying proportions, such as a pie chart, are used."
+Several visualisation methods that are now unpopular (such as the pie chart) are used throughout this paper."
 #| fig-subcap: 
@@
 
-Reviews on uncertainty visualisation rarely offer tried and tested rules for effective uncertainty visualisation, but rather comment on the *difficulties* faced when trying to summarize the field.
-@Kinkeldey2014 found most experiments on the methods for uncertainty visualisation evaluation to be ad hoc, with no commonly agreed upon methodology or formalisation and no greater goal of describing general principals.
-@Hullman2016 commented on the difficulty in taking overarching themes from uncertainty visualisation, as several conflated issues make it unclear if subjects did poorly in an experiment because they misunderstood a visualisation, because the question was misinterpreted, or because they used a specific heuristic.
+Reviews on uncertainty visualisation rarely offer tried and tested rules for effective uncertainty visualisation, but rather, they comment on the *difficulties* faced when trying to summarize the field.
+@Kinkeldey2014 found most experimental methods to be ad hoc, with no commonly agreed upon methodology, formalisations, or greater goal of describing general principals.
+@Hullman2016 noticed there is a serious noise issue in the data coming from uncertainty visualisation experiments.
+She commented on the prevalence of confounding variables that make it unclear as to what exactly caused a subjects poor performance on a particular questions.
+Mistakes due to misunderstood a visualisations, misinterpreted questions, and incorrectly applied heuristics are all combined into a single error value.
 @Spiegelhalter2017 commented that different plots are good for different things, and disagreed with the goal of identifying a universal "best" plot for all people and circumstances.
 @Griethe2006 was unable to find common themes, but instead listed the findings and opinions of a collection of papers.
-@uncertchap2022 summarized several cognitive effects that have repeatedly arised in uncertainty literature, however these effects were each discussed in isolation as a list of considerations an author might make.
-While these reviews are thorough in scope, none discuss how the existing literature contribute to the commonly state goal of uncertainty visualisation, scientific transparency.
+@uncertchap2022 summarized several cognitive effects that repeatedly arise in uncertainty visualisation experiments, however these effects were each discussed in isolation as a list of considerations an author might make rather than an overarching theory of rules for effective uncertainty visualisation.
+While these reviews are thorough in scope, none discuss how the existing literature contribute to the commonly stated goal of uncertainty visualisation, scientific transparency.
 The problem faced by the literature is easily summarized with a famous quote by Henri Poincaré.
@@
 That is to say, despite the wealth of reviews, the field of uncertainty visualisation remains a heap of stones.
+Lovely stones, but stones nonetheless.
 This review attempts to address this issue by offering a novel perspective on the uncertainty visualisation problem, and hopefully laying the foundations on which we can build a house.
@@
 Second, we look at methods that consider uncertainty to be just another variable and discuss the characteristics of uncertainty that make it a unique visualisation problem.
-Third, we look at methods that explicitly combine our estimate and its uncertainty and discuss if the visualisations created by these transformations are still "uncertainty visualisations".
-Fourth, we will discuss methods that implicitly include uncertainty by depicting a sample or original data in place of an estimate.
+Third, we look at methods that explicitly combine our estimate and its uncertainty and discuss the limitations of these approaches.
+Fourth, we will discuss methods that implicitly include uncertainty by depicting a sample in place of an estimate.
 Finally, we discuss how uncertainty visualisations can be effectively evaluated.
-When discussing each of these methods, we consider the *purpose* of uncertainty visualisation and comment on how effective each visualisation is at fulfilling that purpose.
-
-## Spatial examples
-There is an endless number of uncertainty visualisations and it would be impossible to show all of them in this review.
-While the theoretical approach we outline is useful to all uncertainty visualisations, regardless of their application, we feel that showing the variations on a single visualisation will help isolate the ideas we are trying to convey.
+When discussing each of these methods, we will repeatedly return to the *purpose* of uncertainty visualisation and the effectiveness of each approach in fulfilling tha purpose.
+
+## Spatial example
+It would be impossible to display every single uncertainty visualisation in this review, there are far too many so for the sake of clarity, we will focus on the changes made to a single plot.
+The choropleth map.
 Due to the fields origins and focus in geospatial information visualisation, there have been a large number of suggested variations on the choropleth map that allow authors to include uncertainty.
-We will use these maps to highlight the costs and benefits of each approach.
-
-@fig-data shows the first six rows of our example data, along with the geographical boundaries of our data set.
-The temperature variable was generated using a quadratic spatial trend, that is $Temperature = 28-SLongitude^2 + SLatitude^2$ where the $SLongitide$ and $SLatitude$ are the scaled (to standard normal) longitude and latitude of the county's centroid.
-The high and low cases for the sampling variance are randomly sampled from a uniform distribution.
-The low standard errors are drawn from a $U_{[0,1]}$ distribution, while the high standard error cases are drawn from a a $U_{[1,2]}$ distribution.
-As we are dealing with an average, the sampling distribution would be normally distributed, so each county temperature estimate is assumed to come from a $N(Temp_i, SE_{case,i})$.
+While the theoretical approach we outline is useful to all uncertainty visualisations, regardless of their application, we feel that focusing on a single example will help isolate the ideas we are trying to convey.
+We will use these maps to show the effects we are talking about, and highlight the costs and benefits of each approach.
+These examples will need
+
+@fig-data shows the first six rows and the geographical boundaries of our data set.
+The temperature variable was generated using a quadratic spatial trend, that is $Temperature_i = 28-Longitude_i^2 + Latitude_i^2$ where the $Longitide$ and $Latitude$ are the longitude and latitude of the county's centroid scaled to a standard normal distribution.
+Each county's variance is independently, randomly sampled from a uniform distribution.
+The low standard errors are drawn from a $U_{[0,1]}$ distribution, while the high standard errors are drawn from a a $U_{[1,2]}$ distribution.
+As we are dealing with an average, the sampling distribution would be approximately normal by the central limit theorem, so each county temperature estimate is assumed to come from a $N(Temp_i, SE_{case,i})$.
 This is the data we will be using in our spatial uncertainty examples for the rest of the paper.
@@
 
-{{< pagebreak >}}
-
 # Ignoring uncertainty
 A good place to start might be at deceptively straight forward question, why should we include uncertainty at all?
-
-## What is uncertainty
 
@@
 
-# Visualising uncertainty and signal in a new space
+# Combining uncertainty and signal in a transformed space
 Instead of hoping that uncertainty might collapse signal values into a single dimension, we can do some of that work ourselves, and uncertainty visualisation authors already have.

### 2024-08-22 paper.qmd
read over + edited visualising uncertainty as a variable

@@
 @fig-choropleth depicts a choropleth map of the counties of Iowa.
-Each of these counties are colored according to an estimate of average daily temperature that was simulated so that the values followed a clear spatial trend (hot in the middle of the map, and cold on the outside).
-The variance of these estimates were simulated such that a hypothesis test would indicate the existence of a spatial trend in the low variance map, while the trend in the high variance map should be indistinguishable from noise.
-Is this distinction in validity of the spatial trend clear in in the map?
-Is the validity of the trend communicated through the visualisation?
+Each of these counties are colored according to an estimate of average daily temperature that was generated so that the values followed a clear spatial trend (hot in the middle of the map, and cold on the outside).
+The variance of these estimates were simulated such that the trend accounts for most of the variance in the plot in the low variance case (so we should expect the trend to be visible) while the high variance case has significantly more noise, so much so that it overwhelms the spatial trend.
+Is this aspect of the data and the spatial data it communicates clear in in the map?
+Is the strength of the trend communicated through the visualisation?
 
@@
 Both maps depict a spatial trend, where counties closer to the center of the map are hotter than counties on the edge of the map.
-This trend is only technically there in the low variance condition, i.e.
-if we performed a hypothesis test we would conclude the trend exists.
-The high variance condition displays a spatial trend that could simply be the result of noise, which means the plot is displaying a false conclusion."
+In the low variance condition, the trend accounts for most of the variation in the data, in the high variance case, the variance on the temperature estimate accounts for most of the variance.
+The high variance condition displays a spatial trend that could simply be spurious, which means the plot is displaying a false conclusion."
 #| fig-subcap: 
@@
 ## Signal-supression
-The two choropleth maps appearing to be identical in @fig-choropleth highlights the need for uncertainty visualisations.
-Uncertainty visualisation is is required for transparency and this sentiment has been repeated.
+Uncertainty visualisation is required for transparency.
+The two choropleth maps that appear to be identical in @fig-choropleth highlights the issues with simply electing to ignore uncertainty.
+This sentiment appears frequently in the uncertainty visualisation literature.
 Some authors suggest uncertainty is important to include as it communicates the legitimacy (or illegitimacy) of the conclusion drawn from visual inference [@Correll2014; @Kale2018; @Griethe2006].
@@
 A successful uncertainty visualisation would act as a "statistical hedge" for any inference me make using the graphic.
-Additionally, since the purpose of a visualisation is to give a quick "gist" of the information [@Spiegelhalter2017], this hedging needs to be communicated visually without any need for computation from the viewer.
-Additionally, @Ndlovu2023 found that participants applied the same methods they used for simple choropleth maps to complicated uncertainty maps even if that take away was invalid.
-Therefore, this hedging effect needs to be communicated simply through the visualisation.
-If we refer to the conclusion we draw from a graphic to be its "signal" and the variance that makes this signal harder to identify as the "noise", we can summaries this information into three key requirements.
+Additionally, since the purpose of a visualisation is to give a quick "gist" of the information [@Spiegelhalter2017], this hedging needs to be communicated visually without the need for complicated calculations.
+If we refer to the conclusion we draw from a graphic to be its "signal" and the variance that makes this signal harder to identify as the "noise", we can summaries the above information into three key requirements.
 A good uncertainty visualisation needs to:
@@
   1) Reinforce justified signals to encourage confidence in results
-  2) Hide signals that are just noise to prevent unjustified conclusions
+  2) Hide signals that are overwhelmed by noise to prevent unjustified conclusions
   3) Perform tasks 1) and 2) in a way that is proportional to the level of confidence in those conclusions.
@@
 As @fig-choropleth showed, visualisations that are unconcerned with uncertainty have no issue showing justified signals, but struggle with the display of unjustified signals.
-Therefore, we coin this approach to uncertainty visualisation as "signal-suppression" since it primarily differentiates itself from from the "noiseless" visualisation approach through criteria (2).
+Therefore, we call this approach to uncertainty visualisation as "signal-suppression" since it primarily differentiates itself from from the "noiseless" visualisation approach through criteria (2).
 That is, the main difference between an uncertainty visualisation and a "normal" visualisation is that an uncertainty visualisation should prevent us from drawing unjustified conclusions.
 
-{{< pagebreak >}}
-
 ## Uncertainty as a signal
-Uncertainty visualisation is not only motivated by signal-suppression, and we would be remiss to ignore these alternative motivations.
+Uncertainty visualisation is not only motivated by signal-suppression, and we would be remiss not to mention these alternative approaches.
 Some authors claim the purpose of uncertainty is to improve decision making [@Ibrekk1987; @uncertchap2022; @Hullman2016; @Cheong2016; @Boone2018; @Padilla2017].
 Other authors do not describe uncertainty as important for decision making, but rather explicitly state it as variable of importance in of itself [@Blenkinsop2000].
-While uncertainty can provide useful information in decision making, it is important to recognize the "uncertainty" in these cases is not acting as "uncertainty" at all.
+While uncertainty can provide useful information in decision making, it is important to recognize that the "uncertainty" in these cases is not acting as "uncertainty" at all.
 It is acting as signal.
@@
 This is easiest to understand with an example.
+
 Imagine you are trying to decide if you want to bring an umbrella with you to work.
@@
 Unfortunately, your weather prediction app only provides you with the predicted daily rainfall.
-Therefore, your decision will be improved with the inclusion of uncertainty, *not* because uncertainty is important for your decision, but because it gives you the tools required to calculate the *actual* statistic you are basing your decision on.
-In this sense, uncertainty is no more "special" to decision making than weight is in a BMI calculation.
-
-Uncertainty visualisation's made for these purposes should simply display the uncertainty statistic we are interested in, such as the variance, or probability of an event.
-This is precisely what we observe.
+Therefore, your decision will be improved with the inclusion of uncertainty.
+This is *not* because uncertainty in general is important for decision making, but because it gives you the tools required to calculate the *actual* statistic you are basing your decision on.
+In this sense, uncertainty is no more "special" to decision making than weight special to a BMI calculation.
+
+This means the uncertainty visualisations that would perform the best in decision making would simply display the uncertainty statistic we are interested in, such as the variance, or probability of an event, using existing visualisation principles.
+This is precisely what we observe in the literature.
 @fig-exceed depicts an exceedance probability map that has was designed as an alternative to the choropleth map to improve decision making under uncertainty [@Kuhnert2018; @Lucchesi2021].
 A keen viewer may notice that the "exceedance probability map" is actually just a choropleth map, only the statistic being displayed has changed.
-We do not believe this graphic be considered an uncertainty visualisation.
+We are not sure it is productive to categorise this visualisation as an uncertainty visualisation.
 
@@
 The first considers *any* visualisation of error, variance, or probability to be an uncertainty visualisation.
-The second believes an "uncertainty visualisation" is the output of a function that takes a normal visualisation as an input, and transforms it to include the uncertainty information.
+The second believes an "uncertainty visualisation" is the output of a function that takes a normal visualisation as an input, and transforms it to include uncertainty information.
 The former group believe the purpose of uncertainty visualisation to provide signal about a distribution, while the later believe it should act as noise to obfuscate a signal.
@@
 This disagreement creates constant contradictions in what the literature considers to be an "uncertainty visualisation".
-For example @Leland2005 mentions that popular graphics, such as pie charts and bar charts omit uncertainty, and @Wickham2011 suggests their product plot framework, which includes histograms and bar charts, should be extended to include uncertainty however at least one or both of these charts are used in a significant number of uncertainty visualisation experiments [@Ibrekk1987; @Olston2002; @Zhao2023; @Hofmann2012].
+For example @Leland2005 mentions that popular graphics, such as pie charts and bar charts omit uncertainty, and @Wickham2011 suggests their product plot framework, which includes histograms and bar charts, should be extended to include uncertainty.
+Additionally, at least one of these charts have been used in a significant number of uncertainty visualisation experiments [@Ibrekk1987; @Olston2002; @Zhao2023; @Hofmann2012].
 If you view uncertainty as a function applied to an existing graphic, then you would believe a pie chart and bar chart are not uncertainty visualisations, as they are yet to have the "uncertainty visualisation function" applied to them.
@@
 The uncertainty in these graphics are acting as a signal variable, and they should be treated as such.
-
-{{< pagebreak >}}
 
@@
 Extracting global statistics, such as the mean or variance, from a sample can be done with relative each, especially when those values are mapped to color [@Franconeri2021].
+Additionally, @Ndlovu2023 found that participants applied the same methods they used for simple choropleth maps to complicated uncertainty maps even if that take away was invalid.
+The pixel map can be read the same way as a choropleth map, which allows us to leverage that automatic response.
 Therefore, the pixel map performs signal suppression, without sacrificing the viewers ability to extract general statistics, unless those statistics *should* be harder to extract due to the uncertainty in the value.

### 2024-08-22 paper.qmd
edit evaluation bit

@@
 You may think "well, I will just add a dimension to my plot that includes uncertainty".
-This makes sense, as this is the simplest way to add uncertainty to an existing graphic is to simply map uncertainty to an unused visual channel.
+This is a reasonable approach.
+This is the simplest way to add uncertainty to an existing graphic is to simply map uncertainty to an unused visual channel.
+However, it is unclear if this approach is will for our purposes.
 
 ## The bivariate map
-@fig-bivariate a variation of the choropleth map, where we have a two dimensional color palette.
-Not only is temperature mapped to hue, but variance is also mapped to saturation.
+@fig-bivariate depicts a variation of the choropleth map, where we have a two dimensional color palette.
+In this graphic, temperature is still mapped to hue, but the variance is included by utilising colour saturation.
 While these two maps *do* look visually different (which was not the case in the choropleth map) the spatial trend is still clearly visible in both graphics.
-This means the uncertainty *is technically* being communicated, however the primary take away in the graphic is the spatial trend (that does not exist).
-The graphic did not hide the invalid signal, so it is not performing signal-suppression as we would like.
+This means the uncertainty *is technically* being communicated, however the main message of the graphic is still the spatial trend (that may not exist).
+The graphic did not suppress the invalid signal, so it is not performing signal-suppression as we would like.
 At this point, it might be reasonable to ask, why?
@@
 This map is a choropleth map with a two dimensional colour palette where temperature is represented by colour hue, and variance is represented by colour saturation.
-Even though uncertainty has been added to the graphic the spatial trend is still clearly visible in the case where the spatial trend could be attributed to noise."
+Even though uncertainty has been added to the graphic the spatial trend is still clearly visible in the high variance case."
 #| fig-subcap: 
@@
 Many authors seem to believe uncertainty visualisation is a simple high-dimensional visualisation problem, as the difficulty comes from working out how to add uncertainty into already existing graphics [@Griethe2006].
-The problem with this approach to uncertainty visualisation is that it treats uncertainty the same as we would any other variable.
-However, @fig-bivariate makes it clear that simply including uncertainty as a variable is insufficient to perform signal-suppresion.
-If we cannot treat uncertainty as any other variable, what should we treat it as?
+While this is part of the problem in uncertainty visualisation, it is not the complete picture.
+@fig-bivariate makes it clear that simply including uncertainty as a variable is insufficient to perform signal-suppresion.
+If we cannot treat uncertainty the same as we would any other variable, how should we treat it as?
 We need to understand what uncertainty actually *is*, in order to understand how to integrate it into a visualisation.
@@
 There are enough qualitative descriptors of uncertainty to fill a paper, but, none of this is particularly helpful in understanding how to integrate it into a visualisation.
-
 Rather than trying to define uncertainty by what it *is* it may be easier to try and describe what uncertainty *is not*.
+
 Descriptive statistics describe our sample as it is and summarizes large data down into an easy to swallow format.
@@
 Practitioners of the approach believed statistics ended with descriptive statistics as science must be based on actual experience and observations [@Otsuka2023].
-In order to make statements about population statistics, future values, or new observations we need to perform inference, which requires the assumption of the "uniformity of nature", i.e.
-we need to assume that unobserved phenomena should be similar to observed phenomena [@Otsuka2023].
+In order to make statements about population statistics, future values, or new observations we need to perform inference, which requires the assumption of the "uniformity of nature", that is, we need to assume that unobserved phenomena should be similar to observed phenomena [@Otsuka2023].
 Positivists abhore the assumption of the "uniformity of nature" as they believed referencing the unobservable is bad science.
 In other words, these scientists embraces descriptive statistics and shunned inferential statistics due to the inherent uncertainty that came with them.
-Uncertainty is a by-product of inference.
+Since uncertainty is non-existent in descriptive statistics, it is clear that uncertainty is a by-product of inference.
 
@@
 
-This interpretation of uncertainty indicates that the uncertainty on a statistic is not of value in of itself.
+This interpretation of uncertainty indicates that the uncertainty is not a variable of importance in of itself.
 Uncertainty is metadata about our statistic that is required for valid inference.
@@
 This aspect of uncertainty visualization makes it a uniquely difficult problem.
-This is something frequently mentioned
 
@@
 When visualizing uncertainty using its own visual channel, we can also consider visual semiotics and make sure to map uncertainty to intuitive visual channels, such as mapping more uncertain values to lighter colors [@Maceachren2012].
-
-```{r}
-#| echo: false
-#| message: false
-#| warning: false
-#| eval: false
-#| label: fig-bivariate2
-#| layout-ncol: 3
-#| layout-valign: "bottom"
-
-#fig-cap: "A bivariate map that depict the counties of Iowa where each county is coloured acording to it's average daily temperature and the variance in temperature.
-This map is a variation on the previous bivariate map where instead of variance being mapped to colour saturation, it is mapped to colour value.
-Even though colour value is intergrable with colour hue, and colour value is naturally accociated with uncertainty, the spatial trend is still visible in the map."
-
-# Bivariate Map
-# Make bivariate palette
-# Function to devalue by a certain amount
-colsupress2 <- function(basecols, hue=1, sat=1, val=1) {
-    X <- diag(c(hue, sat, val)) %*% rgb2hsv(col2rgb(basecols))
-    hsv(pmin(X[1,], 1), pmin(X[2,], 1), pmin(X[3,], 1))
-}
-
-# recurvisely decrease value
-v_val = 1.1
-bivariatepal2 <- c(basecols,
-                   colsupress2(basecols, val=v_val), 
-                   colsupress2(colsupress2(basecols, val=v_val), val=v_val),
-                   colsupress2(colsupress2(colsupress2(basecols, val=v_val), val=v_val), val=v_val)
-                   )
-
-# establish levels of palette
-names(bivariatepal2) <- paste(rep(1:8, 4), "-" , rep(1:4, each=8), sep="")
-
-# Bivariate maps 2
-p2a1 <- my_map_data |>
-  filter(variance_class=="lowvar") %>%
-  ggplot() +
-  geom_sf(aes(fill = biclass, geometry = geometry), colour=NA) + 
-  scale_fill_manual(values = bivariatepal2) +
-  theme_void() + 
-  theme(legend.position = "none")
-
-p2b2 <- p2a1 %+% filter(my_map_data, variance_class=="highvar")
-
-p2a1
-p2b2
-#show_pal2(colours = bivariatepal2, ncol=8, borders=NA, myxlab = "Temperature", myylab = "Variance", breaks = 21:29, breaks2 = 0:4)
-
-```
-
-```{r}
-#| echo: false
-#| message: false
-#| warning: false
-#| label: fig-transparency
-#| fig-cap: "A bivariate map that depict the counties of Iowa where each county is coloured acording to it's average daily temperature and variance.
-This map is a variation on the previous bivariate map where instead of variance being mapped to colour saturation, it is mapped to transparency."
-#| fig-subcap: 
-#|   - "Low Variance Data"
-#|   - "High Variance Data"
-#|   - "Transparency Palette"
-#| layout-ncol: 3
-#| layout-valign: "bottom"
-
-# ALTERNATIVE TRANSPARENCY INSTEAD OF COLOUR VALUE
-# get colour values from transparency transformation
-transpal <- lapply(basecols, scales::alpha, alpha=c(0.95, 0.65, 0.35, 0.05))
-
-transpal <- as.vector(matrix(unlist(transpal), ncol = 4, byrow = TRUE))
-
-names(transpal) <- paste(rep(1:8, 4), "-" , rep(1:4, each=8), sep="")
-
-# Transparency maps
-pta <- my_map_data |>
-  filter(variance_class=="lowvar") %>%
-  ggplot() +
-  geom_sf(aes(fill = biclass, geometry = geometry), colour=NA) + 
-  scale_fill_manual(values = transpal) +
-  theme_void() + 
-  theme(legend.position = "none")
-
-ptb <- pta %+% filter(my_map_data, variance_class=="highvar")
-
-pta
-ptb
-show_pal2(colours = transpal, ncol=8, borders=NA, myxlab = "Temperature", myylab = "Variance", breaks = 21:29, breaks2 = 0:4)
-
-```
-
-@fig-transparency is an example of a variations of @fig-bivariate where uncertainty is mapped to transparency, and temperature is mapped to color hue to leverage these visualisation concepts.
-This method achieves signal suppression quite well.
-The spatial trend is clearly visible in the low variance case and that trend it becomes much harder to identify in the high variance case.
-While this was an effective approach for this graphic, relying on integrability may not give us the amount of control we want over our signal-supresison.
+Unfortunately relying on integrability may not give us the amount of control we want over our signal-supresison.
 Without a strong understanding of how these visual channels collapse down into a single channel, relying on integrability could create unintended consequences such as displaying phantom signals or hiding justified signals.
@@
 
-There is another reason @fig-transparency is better at signal-suppression than @fig-bivariate, and it may not be due to integrability.
-Colour value has a second desirable quality for signal-supression, which is that the colours become harder to distinguish as the value decreases.
-This means high uncertainty values are harder to differentiate than low uncertainty values.
-This implicit feature of colour value can generalised to other aesthetics by transforming the visual feature space ourselves.
-
-{{< pagebreak >}}
+There is another benefit to mapping uncertainty to stauration that is not directly related to integrability.
+Colour value has a second desirable quality as the colours become harder to distinguish as the saturation decreases.
+This means high uncertainty saturations are harder to differentiate than low uncertainty values.
+We can leverage this implicit feature of colour value by transforming the visual feature space ourselves.
 
 # Combining uncertainty and signal in a transformed space
-Instead of hoping that uncertainty might collapse signal values into a single dimension, we can do some of that work ourselves, and uncertainty visualisation authors already have.
+Instead of hoping that uncertainty might collapse signal values into a single dimension, we can do some of that work ourselves.
+As a matter of fact, some uncertainty visualisation authors already have.
 
@@
 This means the VSUP has successfully suppressed the spatial trend in the data.
-However the spatial trend may not be the only signal of concern in our graphic.
+However, the spatial trend may not be the only signal of concern in our graphic.
 Now we must return to the original signal-suppression criteria and ask ourselves if they have all been met.
@@
 There are two primary motivations behind visualisation, communication and exploratory data analysis (EDA).
-Communication involves identifying a signal we want to communicate and designing a visualisation that best conveys that, while EDA involves creating a versatile visualisation using it to extract several signals.
+Communication involves identifying a signal we want to communicate and designing a visualisation that best conveys that, while EDA involves creating a versatile visualisation and using it to extract several signals.
 If we are designing an uncertainty visualisation for communication then we can just suppress the specific signal we are seeking to communicate.
@@
 For uncertainty visualisation for EDA to work, we would need to assume that suppressing individual estimates using their variance should naturally extend to broader suppression of plot level insights.
-Unfortunately it is not clear whether or not uncertainty visualisations for EDA are possible.
+Unfortunately it is unclear how reliably this would work.
 
@@
 
-Some authors recognize inference will occur (in some shape or form) and believe uncertainty *should* be visualised but do not recognize *how* uncertainty would be visualised.
+Some authors recognize inference will always occur (in some shape or form) and believe uncertainty *should* be visualised but do not recognize *how* uncertainty would be visualised.
 @Hullman2021 argued that there is no such thing as a "model-free" visualisation, therefore all visualisations require uncertainty as we are always performing inference.
 However, even something as simple as calculating the uncertainty that is used to suppress our visualisation is not model free, as we need to identify if the sampling variance or the sample variance is more appropriate [@Hofman2020].
-While we agree that people cannot prevent themselves from performing inference, this does not mean it is possible to include uncertainty in a visualisation designed for EDA.
-However, this does mean we should endeavor for a versatile uncertainty visualisation method that is able to perform signal-suppression on all the signals displayed in the visualisation.
-
-Uncertainty visualisation for EDA would be possible if we designed a plot in such a way that suppressing individual estimates using their variance would naturally extend to broader suppression of plot level insights.
-This assumption is commonly made by visualisation researchers in normal visualisation experiments [@North2006], however achieving it would largely depend on the methods use to perform signal suppression.
-By manually combining the values in @fig-vsup, we violated this requirement, so the VSUP is not versatile enough to act as an uncertainty visualisation for EDA.
+This means an ideal uncertainty visualisations should not only meet the signal suppression requirements, but should also endeavor to be versatile enough to meet those requirements for all the signals displayed in the graphic.
 
@@
 The lack of versatility of the VSUP is easy to see with a simple example.
+
 Let's say we have a graphic that depicts a set of coefficients from a linear regression and the value of the coefficient is shown using a single color.
@@
 To answer this question we do a series of t-tests on these estimates.
-
 All of the individual t-tests of fail to reject the null hypothesis that the coefficients are different from 0.
@@
 Even though the bivariate map does not explicitly combine color values at certain variance levels, the mapping of variance to color saturation does this implicitly.
-That is, at certain saturation values the colors in a bivariate map are imperceptibly different to the human brain and appear as though they are mapped to the same value.
+That is, at certain saturation values the colors in a bivariate map are imperceptibly different and appear as though they are mapped to the same value.
 At this point, it is irrelevant whether or not the colours are technically different, they are the same color in the human brain.
+
+If we only use a single value to express each signal-suppressed statistic, we will always need to decide which signals we suppress and which we do not.
+This issue has already been raised in the literature.
 Which hypothesis are suppressed and which are not largely depends on the method used to combining colors in the palette [@Kay2019].
-The VSUP here used a tree based method as that is what was used by @Correll2018, but there are alternatives that are more appropriate for different hypothesis.
-
-If we only use a single value to express each signal-suppressed statistic, we will always need to decide which signals we suppress and which we do not.
-However, If we could express the statistic of a cell using multiple colors, this limitation may disappear entirely.
-
-{{< pagebreak >}}
+The VSUP in @fig-vsup used the tree based method that was used by @Correll2018, but there are alternatives that are more appropriate for different hypothesis.
+
+Uncertainty visualisation for EDA would be possible if we designed a plot in such a way that suppressing individual estimates using their variance would naturally extend to broader suppression of plot level insights.
+This assumption is commonly made by visualisation researchers in normal visualisation experiments [@North2006].
+If we could express the statistic of a cell using multiple colors, this limitation may disappear entirely.
 
 # Implicitly Combining Uncertainty and Signal
-There is technically a stage of our analysis where the estimate and variance are not separate, when we only have a sample.
 Rather than trying to figure out how to combine signal and uncertainty into a single color, we can just display a sample instead and allow the viewer to extract *both* the estimate and the variance.
@@
 ## Pixel map
-@fig-pixel displays a pixel map [@Lucchesi2021], which is a variation of the choropleth map where each area is divided up into several smaller areas, each colored using outcomes from the larger area's temperature sampling distribution.
+@fig-pixel displays a pixel map [@Lucchesi2021], which is a variation of the choropleth map where each area is divided up into several smaller areas, each colored using outcomes from the larger area's (i.e.
+the county's) average temperature sampling distribution.
 The spatial trend is clearly visible in the low variance case, but functionally disappears in the low variance case.
@@
 ## Show me the data
-The pixel map is not the best uncertainty visualisation, but it is trying to *imitate* the best uncertainty visualisation.
+The pixel map is not the best uncertainty visualisation, but it does share a lot in common with the best uncertainty visualisation.
 The best uncertainty visualisation is the visualisation that best captures the limitations of of our raw data.
@@
 What this means is that we have two types of uncertainty, but one is more "processed" than the other.
-Quantifiable uncertainties are just assumption violations expressed as an effect on our final statistic.
-The disconnect between these two expressions of uncertainty creates a huge problem for authors trying to visualise it.
+Quantifiable uncertainties are just assumption violations expressed as an effect on our final statistic (or true variation of our data).
+The disconnect between quantifiable and unqualitifable uncertainty creates a huge problem for authors trying to visualize it.
 A survey of visualisation authors cited "not knowing how to calculate uncertainty" as one of the primary reasons they did not include it in visualisations [@Hullman2020a].
@@
 For example, @Thomson2005 suggests a mathematical formula for *examples* of uncertainty, and information theory tries to quantify uncertainty using the idea of entropy, but they ignore the disconnect between the broad concept of uncertainty and what we can reliably quantify.
-Some authors don't believe that it is even possible to quantify all the assumption violations.
-@Spiegelhalter2017 mentioned that combining the uncertainties that appear at each stage of an analysis into a single "uncertainty" value is near impossible.
+Some authors don't believe that it is even possible to quantify all the assumption violations [@Spiegelhalter2017].
 
@@
 
-This does not mean that visualizing raw data instead of implementing sampling techniques will always prevent insignificant signal from getting through.
-@Buja2009 illustrated how groups that appear linearly separable in a linear discriminant analysis (LDA) visualisation of the data can actually be the result of a LDA performed on too many variables, something that was not clear from the visualisation until the line-up protocol was implemented.
+This does not mean that visualizing raw data will always prevent insignificant signal from getting through.
+@Chowdhury illustrated how groups that appear linearly separable in a linear discriminant analysis (LDA) visualisation of the data can actually be the result of a LDA performed on too many variables, something that was not clear from the visualisation until the line-up protocol was implemented.
 However just showing the data it is simple but effective option for uncertainty visualisation that that seems to be largely overlooked.
-While it is not always possible, it should always be considered as an effective uncertainty visualisation when the raw data is available.
-
-{{< pagebreak >}}
+
+While the pixel map does not literally show the data, it does imitate showing the data and give us details on where the variability is coming from.
+In order to uncerstand *why* pixel maps seem to be successful at signal suppression at first glace would likely require a more thorough discussion than what what we can offer here.
 
@@
 Unfortunately the illustrative methods we have used thus far, i.e.
-showing a graphic and saying "wow look at this", are lacking if we want any generalizable results.
+showing a graphic and saying "wow look at this", is lacking if we want any serious results.
 However, despite the abundance of uncertainty visualisation evaluation experiments, existing literature reviews have struggled to synthesise them into any common rules [@Kinkeldey2014; @Hullman2016].
 
-Here we discuss common evaluation methods, why these methods might struggle to create a cohesive set of recommendations for uncertainty visualisations, and consier how we to best evaluate visualisations on their ability to perform signal-suppression.
+Here we discuss common evaluation methods, why these methods might struggle to create a cohesive set of recommendations for uncertainty visualisations, and consider how to evaluate visualisations on their ability to perform signal-suppression.
 
@@
 ### Trust, confidence, and risk aversion
-Trust is a by-product of displaying uncertainty and it commonly measured in uncertainty evaluation studies [@Hullman2019].
-Considering trust, and not transparency, as the metric of importance in uncertainty communication can lead to a questionable subtext that argues against transparency, something that has been noticed by several other authors [@Spiegelhalter2017; ONeill2018].
+Trust is a by-product of displaying uncertainty so it is commonly measured in uncertainty evaluation studies [@Hullman2019].
+Considering trust, and not transparency, as the metric of importance in uncertainty communication can lead to a questionable subtext that argues against transparency, something that has been noticed by several other authors [@Spiegelhalter2017; @ONeill2018].
 Science communication should be primarily concerned with accuracy.
@@
 Confidence cannot simultaneously be a measure of clarity of visualisation *and* a way to capture the uncertainty expressed in a visualisation.
-Uncertainty visualisations conflate these two measures when they ask about confidence.
+Uncertainty visualisations conflate these two measures.
 
@@
 The first method is what we call the "vague question" approach.
-The authors of these studies will ask the participants a question that implies that they should use  a use of uncertainty, but compare the readers answer to a very specific ground truth.
+The authors of these studies will ask the participants a question that implies that they should use the uncertainty information, but it is unclear how.
+The authors then go on to compare the readers answer to a very specific ground truth.
 This means the participants are being *evaluated* as though they are performing a value extraction task, but they are not being asked the *questions* that are asked in the value extraction task.
@@
 Other authors have commented on the complexity of communicating hurricane risk because the path, storm surge and wind speed are all important and cannot be ignored [@Spiegelhalter2017].
-Interestingly it is unclear how the participants were supposed to answer this question, as @Padilla2017 did three experiments and the first and third experiment had conflicting assumptions in their ground truth.
+Interestingly it is unclear how the participants were supposed to answer this question.
+@Padilla2017 did three experiments and the first and third experiment had conflicting assumptions in their ground truth.
 In the first study, participants were required assume to the storm was equally intense reguardless of the probability of the probability of the oil rig being hit (i.e.
@@
 While these approaches are certainly a step in the right direction, the experiments end up having far too much noise in their results.
-Additionally, by having a ground truth, they end up implicitly asking users to either treat uncertainty as a signal, or to ignore it entirely, neither of which is advisable for evaluating graphics on signal-suppression.
+Additionally, by having a ground truth, they end up implicitly asking users to treat uncertainty as a signal which goes against the purposes of signal-suppression.
 

### 2024-08-22 writingdumpster.qmd
edit evaluation bit

@@
 ```
+
+# Alternative maps
+
+```{r}
+#| echo: false
+#| message: false
+#| warning: false
+#| eval: false
+#| label: fig-bivariate2
+#| layout-ncol: 3
+#| layout-valign: "bottom"
+
+#fig-cap: "A bivariate map that depict the counties of Iowa where each county is coloured acording to it's average daily temperature and the variance in temperature.
+This map is a variation on the previous bivariate map where instead of variance being mapped to colour saturation, it is mapped to colour value.
+Even though colour value is intergrable with colour hue, and colour value is naturally accociated with uncertainty, the spatial trend is still visible in the map."
+
+# Bivariate Map
+# Make bivariate palette
+# Function to devalue by a certain amount
+colsupress2 <- function(basecols, hue=1, sat=1, val=1) {
+    X <- diag(c(hue, sat, val)) %*% rgb2hsv(col2rgb(basecols))
+    hsv(pmin(X[1,], 1), pmin(X[2,], 1), pmin(X[3,], 1))
+}
+
+# recurvisely decrease value
+v_val = 1.1
+bivariatepal2 <- c(basecols,
+                   colsupress2(basecols, val=v_val), 
+                   colsupress2(colsupress2(basecols, val=v_val), val=v_val),
+                   colsupress2(colsupress2(colsupress2(basecols, val=v_val), val=v_val), val=v_val)
+                   )
+
+# establish levels of palette
+names(bivariatepal2) <- paste(rep(1:8, 4), "-" , rep(1:4, each=8), sep="")
+
+# Bivariate maps 2
+p2a1 <- my_map_data |>
+  filter(variance_class=="lowvar") %>%
+  ggplot() +
+  geom_sf(aes(fill = biclass, geometry = geometry), colour=NA) + 
+  scale_fill_manual(values = bivariatepal2) +
+  theme_void() + 
+  theme(legend.position = "none")
+
+p2b2 <- p2a1 %+% filter(my_map_data, variance_class=="highvar")
+
+p2a1
+p2b2
+#show_pal2(colours = bivariatepal2, ncol=8, borders=NA, myxlab = "Temperature", myylab = "Variance", breaks = 21:29, breaks2 = 0:4)
+
+```
+
+```{r}
+#| echo: false
+#| message: false
+#| warning: false
+#| label: fig-transparency
+#| fig-cap: "A bivariate map that depict the counties of Iowa where each county is coloured acording to it's average daily temperature and variance.
+This map is a variation on the previous bivariate map where instead of variance being mapped to colour saturation, it is mapped to transparency."
+#| fig-subcap: 
+#|   - "Low Variance Data"
+#|   - "High Variance Data"
+#|   - "Transparency Palette"
+#| layout-ncol: 3
+#| layout-valign: "bottom"
+
+# ALTERNATIVE TRANSPARENCY INSTEAD OF COLOUR VALUE
+# get colour values from transparency transformation
+transpal <- lapply(basecols, scales::alpha, alpha=c(0.95, 0.65, 0.35, 0.05))
+
+transpal <- as.vector(matrix(unlist(transpal), ncol = 4, byrow = TRUE))
+
+names(transpal) <- paste(rep(1:8, 4), "-" , rep(1:4, each=8), sep="")
+
+# Transparency maps
+pta <- my_map_data |>
+  filter(variance_class=="lowvar") %>%
+  ggplot() +
+  geom_sf(aes(fill = biclass, geometry = geometry), colour=NA) + 
+  scale_fill_manual(values = transpal) +
+  theme_void() + 
+  theme(legend.position = "none")
+
+ptb <- pta %+% filter(my_map_data, variance_class=="highvar")
+
+pta
+ptb
+show_pal2(colours = transpal, ncol=8, borders=NA, myxlab = "Temperature", myylab = "Variance", breaks = 21:29, breaks2 = 0:4)
+
+```

### 2024-09-14 paper.qmd
Major changes: adding line breaks at each sentence - makes it easier for git to track changes without merge issues. Added a citation to Vanderplas 2017, in part because it's important to say that lineups can be used for hypothesis testing, but they can also be used to compare visual power of different charts for actually seeing differences in data.
Minor changes: Fixed a few typos and grammar issues.

@@
 - convert to American spelling 
+  (Susan) Is this necessary? I'm not sure it matters as long as you're consistent...
 - May not need to reference the same paper again and again
@@
 
-From entertainment choices to news articles to insurance plans, the modern citizen is so over run with information in every aspect of their life it can be overwhelming.
-In the face of this overflow of information, tools that effectively summarize information down into simple and clear ideas become more valuable.
+From entertainment choices to news articles to insurance plans, the modern citizen is so over-run with information in every aspect of their life it can be overwhelming. 
+In the face of this overflow of information, tools that effectively reduce piles of information to simple and clear ideas become more valuable.
 
@@
 For example, box plots or histograms can be considered to be displaying uncertainty in the sense of variability in observations sampled from a population distribution. 
-Today, there is an abundance of publications on the topic which makes it is timely to construct a review of the field.
+Today, there are an abundance of publications on the topic which makes it timely to construct a review of the field. 
 In fact, there have already been several reviews published.
+
+<!-- Susan: I think there's a missed opportunity here to reference the premise you started with, that when there are heaps of information, it's valuable to distill it down into simple facts -->
 
@@
 This review is broken into several parts that each reflect a different approach to uncertainty visualisation. 
-First we look at graphics that ignore uncertainty entirely and discuss why uncertainty should be included at all.
+First, we look at graphics that ignore uncertainty entirely and discuss why uncertainty should be included at all. 
 Second, we look at methods that consider uncertainty to be just another variable and discuss the characteristics of uncertainty that make it a unique visualisation problem. 
 Third, we look at methods that explicitly combine our estimate and its uncertainty and discuss the limitations of these approaches. 
-Fourth, we will discuss methods that implicitly include uncertainty by depicting a sample in place of an estimate.
+Fourth, we discuss methods that implicitly include uncertainty by depicting a sample in place of an estimate. 
 Finally, we discuss how uncertainty visualisations can be effectively evaluated.
-When discussing each of these methods, we will repeatedly return to the *purpose* of uncertainty visualisation and the effectiveness of each approach in fulfilling tha purpose.
+When discussing each of these methods, we will repeatedly return to the *purpose* of uncertainty visualisation and the effectiveness of each approach in fulfilling that purpose.
 
 ## Spatial example
-It would be impossible to display every single uncertainty visualisation in this review, there are far too many so for the sake of clarity, we will focus on the changes made to a single plot.
-The choropleth map.
-Due to the fields origins and focus in geospatial information visualisation, there have been a large number of suggested variations on the choropleth map that allow authors to include uncertainty.
+
+It would be impossible to display every single uncertainty visualisation in this review, there are far too many so for the sake of clarity, we will focus on the changes made to a single plot, the choropleth map.
+Due to the field's origins and focus in geospatial information visualisation, there have been a large number of suggested variations on the choropleth map that allow authors to include uncertainty. 
 While the theoretical approach we outline is useful to all uncertainty visualisations, regardless of their application, we feel that focusing on a single example will help isolate the ideas we are trying to convey.
 We will use these maps to show the effects we are talking about, and highlight the costs and benefits of each approach. 
-These examples will need
+These examples will need XXX incomplete sentence!! XXX
 
@@
 Uncertainty visualisation is required for transparency. 
-The two choropleth maps that appear to be identical in @fig-choropleth highlights the issues with simply electing to ignore uncertainty.
+The two choropleth maps that appear to be identical in @fig-choropleth highlight the issues with simply electing to ignore uncertainty. 
 This sentiment appears frequently in the uncertainty visualisation literature.
@@
 This is *not* because uncertainty in general is important for decision making, but because it gives you the tools required to calculate the *actual* statistic you are basing your decision on. 
-In this sense, uncertainty is no more "special" to decision making than weight special to a BMI calculation.
+In this sense, uncertainty is no more "special" to decision making than weight is special to a BMI calculation.
 
@@
 Some authors implicitly suggest that that visualisations of variance or probability are differentiated due to the psychological heuristics involved in interpreting uncertainty [@Hullman2019].
-While it is true that heuristics lead people to avoid uncertainty [@Spiegelhalter2017] there is no evidence that this psychological effect translates to issues with the visual representation of uncertainty.
+While it is true that heuristics lead people to avoid uncertainty [@Spiegelhalter2017], there is no evidence that this psychological effect translates to issues with the visual representation of uncertainty.
 Again, given that we do not make these same visual considerations for other variables that elicit distaste or irrational behavior, there is no reason to assume this is what makes uncertainty visualisation so special.
@@
 This is a reasonable approach. 
-This is the simplest way to add uncertainty to an existing graphic is to simply map uncertainty to an unused visual channel.
-However, it is unclear if this approach is will for our purposes.
+The simplest way to add uncertainty to an existing graphic is to simply map uncertainty to an unused visual channel. 
+However, it is unclear if this approach is 
+sufficient XXX Word choice??? XXX
+<!-- I changed will to sufficient, but I'm not sure that's what you were going for. -->
+for our purposes.
 
@@
 In order to make statements about population statistics, future values, or new observations we need to perform inference, which requires the assumption of the "uniformity of nature", that is, we need to assume that unobserved phenomena should be similar to observed phenomena [@Otsuka2023].
-Positivists abhore the assumption of the "uniformity of nature" as they believed referencing the unobservable is bad science.
+Positivists abhor the assumption of the "uniformity of nature" as they believed referencing the unobservable is bad science. 
 In other words, these scientists embraces descriptive statistics and shunned inferential statistics due to the inherent uncertainty that came with them. 
 Since uncertainty is non-existent in descriptive statistics, it is clear that uncertainty is a by-product of inference.
+<!-- Inferential statistics didn't yet exist when positivists were dominant - they weren't making a choice between description and inference, as the choice didn't exist.
+This is slightly misleading as written, even if I generally agree with the concept you're trying to communicate.
+-->
 
@@
 
+<!-- You've promised to describe what uncertainty is not...
+did you achieve that goal?
+If not, you probably should revisit how you're introducing the philosophical bit about descriptive statistics and positivism.
+-->
+
 ### Visualising the "single integrated uncertain value"
-Typically, when making visualisations, we want the visual channels to be separable, that is, we don't want the data represented through one visual channel to interfere with the others [@Smart2019].
+Typically, when making visualisations, we want the visual channels to be separable. 
+That is, we don't want the data represented through one visual channel to interfere with the others [@Smart2019]. 
 Mapping uncertainty and signal to separable channels allows them to be read separately, which does not align with the goal of communicating them as a "single integrated channel". 
@@
 When visualizing uncertainty using its own visual channel, we can also consider visual semiotics and make sure to map uncertainty to intuitive visual channels, such as mapping more uncertain values to lighter colors [@Maceachren2012].
-Unfortunately relying on integrability may not give us the amount of control we want over our signal-supresison.
+Unfortunately relying on integrability may not give us the amount of control we want over our signal-supression. 
 Without a strong understanding of how these visual channels collapse down into a single channel, relying on integrability could create unintended consequences such as displaying phantom signals or hiding justified signals.
@@
 The methods used by the VSUP bring to light a slight problem with uncertainty visualisation. 
-Specifically that uncertainty and the purpose of visualisation are somewhat at odds with one another.
+Specifically, uncertainty and the purpose of visualisation are somewhat at odds with one another. 
 There are two primary motivations behind visualisation, communication and exploratory data analysis (EDA). 
@@
 For uncertainty visualisation for EDA to work, we would need to assume that suppressing individual estimates using their variance should naturally extend to broader suppression of plot level insights.
-Unfortunately it is unclear how reliably this would work.
+Unfortunately, it is unclear how reliably this would work.
 
@@
 Often EDA is used to give us an understanding of our data and identify which signals are worth pursuing. 
-In this sense, EDA is the visual parallel to descriptive statistics, as it is performed without an explicit hypothesis which means there is no inference, and by extension, there is no uncertainty.
+In this sense, EDA is the visual parallel to descriptive statistics, as it is performed without an explicit hypothesis, which means there is no inference, and by extension, there is no uncertainty.
 
@@
 However, even something as simple as calculating the uncertainty that is used to suppress our visualisation is not model free, as we need to identify if the sampling variance or the sample variance is more appropriate [@Hofman2020].
+<!-- Ok, so this is an additional challenge on top of Hullman's claim, but it's not refuting Hullman's claim.
+I think the easier way to do that is to say that while you can certainly think of visualizations as containing implicit inferential properties, there are many potential inferences in any single visualization; until the viewer chooses one and thinks about it (which isn't easily observable), the particular variety of uncertainty which would need to be displayed can't be calculated - it's a quantum physics type issue where until something is observed, the singularity doesn't collapse and everything exists only in probabilistic terms.
+The cat isn't dead or alive until the viewer starts thinking inferentially in one way or another.
+-->
 This means an ideal uncertainty visualisations should not only meet the signal suppression requirements, but should also endeavor to be versatile enough to meet those requirements for all the signals displayed in the graphic.
@@
 We then make a visualisation that suppresses this signal and ensures that all of the estimates are visually indistinguishable from 0. 
-We then do a comparison of two means t-test and find that several of the values need to be visually distinguishable from each other.
+Next, we conduct two sample t-test and find that several of the values need to be visually distinguishable from each other. 
 The VSUP method must pick a single color for each estimate, and these colors must be *either* visually distinguishable or indistinguishable from each other. 
@@
 At this point, it is irrelevant whether or not the colours are technically different, they are the same color in the human brain. 
+<!-- This is of course complicated by the fact that human colour perception varies at an individual level: some individuals have been documented to have 4 different cones and can thus perceive a different range of colors; others have only two or one functioning cone and have colour deficiencies.
+For VSUP to function for all individuals, we must calibrate each plot to an individual's ability to perceive colour.
+-->
 
@@
 While the spatial trend is just barely visible in the high uncertainty case, it is much harder to see. 
-This means the graphic also achieves the third criteria for signal-suppression, i.e.
-our difficulty in seeing the distribution is proportional to the level of uncertainty in the graphic.
+This means the graphic also achieves the third criteria for signal-suppression, i.e. our difficulty in seeing the distribution is proportional to the level of uncertainty in the graphic.
 
@@
 @Chowdhury illustrated how groups that appear linearly separable in a linear discriminant analysis (LDA) visualisation of the data can actually be the result of a LDA performed on too many variables, something that was not clear from the visualisation until the line-up protocol was implemented.
-However just showing the data it is simple but effective option for uncertainty visualisation that that seems to be largely overlooked.
+However, just showing the data it is simple but effective option for uncertainty visualisation that that seems to be largely overlooked. 
+<!-- Does showing the data actually show uncertainty, or does it just show variance?
+I'm thinking of inferential statistics here - you can show a sample, but it doesn't necessarily show you anything about the uncertainty of the population-based inferential statistics, visually speaking.
+Some of the dotmap style visualizations you're thinking of are actually census visualizations - there isn't any inferential uncertainty, but there is variance.
+I'm not sure you're adequately separating these concepts here, even if I generally agree with the broad points you're making.
+-->
 
@@
 If we want to make conclusions about how effective any uncertainty visualisation method is we need to look at the results of evaluation experiments. 
-Unfortunately the illustrative methods we have used thus far, i.e.
-showing a graphic and saying "wow look at this", is lacking if we want any serious results.
+Unfortunately the illustrative methods we have used thus far, i.e. showing a graphic and saying "wow look at this", is lacking if we want any serious results.
 However, despite the abundance of uncertainty visualisation evaluation experiments, existing literature reviews have struggled to synthesise them into any common rules [@Kinkeldey2014; @Hullman2016].
@@
 The role of uncertainty is rarely evaluated in these studies as the graphics are often compared on the basis of being "uncertainty visualisations" [@Ibrekk1987; @Hullman2015; @Hofman2020], a class that has no established definition.
-By shifting the focus of our inference from $\bar{X}$ to $Var(X)$ or $P(X)$ we end up evaluating visualisations on their ability to convey uncertainty statistics, rather than on uncertainties ability to suppress statistics.
+By shifting the focus of our inference from $\bar{X}$ to $Var(X)$ or $P(X)$ we end up evaluating visualisations on their ability to convey uncertainty statistics, rather than on uncertainty's ability to suppress statistics.
 This leads to a series of experiments where the uncertainty is evaluated as a signal even if that was not the goal of the experiment.
@@
 Science communication should be primarily concerned with accuracy. 
+<!-- Ok, but you can be completely accurate in how you render a plot and still convey information that isn't accurate (e.g.
+sine illusion/line width issues).
+Do we care about accuracy of the visual form, or accuracy of the perception?
+Because accuracy of the perception is potentially influenced by trust/transparency and potentially influences that metric as well.
+Rachel Rogers and I have studied how much weight people give testimony in a courtroom - it strikes me that trust/transparency are a separate dimension that is important and related to accuracy, in that both accuracy and trust/transparency will influence how much mental weight I ascribe to new information gained from a visualization.
+-->
 
@@
 Risk aversion is an economics term used to describe an agent who would chose a random variable with a lower expected payout because it also has a lower variance. 
-Risk aversion is considered irrational behaviour because it is a deviation from the behavior of a rational agent.
+Risk aversion is considered irrational behaviour because it is a deviation from the behaviour of a rational agent. 
 Comparing participant responses to that of a ration agent has even been suggested as a benchmark for uncertainty visualization experiments [@wu2023rational].
@@
 Ultimately this approach results in strangely cryptic questions that create a large amount of noise due to the varied interpretation of the questions [@Hullman2016]. 
-For example @Hofmann2012 showed study participants 20 plots, where each plot displayed two distributiosn (as a pair of jittered samples, density plots, histograms, or box plots) and asked them to identify the plot where "the blue group furthest to the right".
+For example @Hofmann2012 showed study participants 20 plots, where each plot displayed two distributions (as a pair of jittered samples, density plots, histograms, or box plots) and asked them to identify the plot where "the blue group is furthest to the right".
 The mean of the two groups was used to decide which group was actually "furthest to the right" and the participants were evaluated against that ground truth. 
+<!-- Participants weren't evaluated against that ground truth, though - the plots were what was actually being evaluated in that study, not participant accuracy.
+Lineup prompts are typically vague (sometimes, I agree, too vauge) because the participant has to figure out what metric to use to evaluate the data first, and then complete the evaluation task.
+We're as interested in the metric they're using as we are in their successful identification of the target plot.
+-->
 In another example @Ibrekk1987 asked participants for the "best estimate", but the participant's responses were evaluated against the mean of the distribution.
@@
 These studies frequently expect participants to give deterministic answers for probabilistic questions. 
-For example @Correll2014 showed participants the distribution of voter preferences for two candidates in an election, and aksed them "how likely is candidate B to win the election?".
+For example @Correll2014 showed participants the distribution of voter preferences for two candidates in an election, and asked them "how likely is candidate B to win the election?". 
 Participants were not able to answer the question about likelihood in term of probability, but were instead given seven options from 1 = "Outcome will be most in favour of A" to 7 = "Outcome will be most in favour of B".
 The ground truth statistic for this question was a scalar multiple of Cohen’s d, indicating participants were supposed to incorporate uncertainty information using a very specific formula that was likely unknown to them but assumed to be used implicitly.
+<!-- Might be worth referencing the fact that visual statistics and the computational equivalents often differ, and this is even before you start talking measurement error when using a Likert scale.
+-->
 In another example, @Padilla2017 provided participants with a visualisation of the cone of uncertainty and asked then to "decide which oil rig will receive more damage based on the depicted forecast of the hurricane path".
-The cone of uncertainty provides a 60% confidence interval for the location of the eye of a hurricane, which allows us to know the area where the eye of the storm will go, it does not given any information about the intensity of a storm, the size of a storm, or even if a location will be hit.
+The cone of uncertainty provides a 60% confidence interval for the location of the eye of a hurricane, which allows us to know the area where the eye of the storm will go, but it does not given any information about the intensity of a storm, the size of a storm, or even if a location will be hit.
+<!-- It's not unreasonable to use forms of display that are used in gov't communications, even if they're admittedly not optimal -- it can be reasonable to *show* that they're not optimal, and Padilla is usually advocating for different risk displays as a result, including spaghetti maps for the track and probabilistic wind speed/flood risk maps for the impacts.
+The problem is that gulf coast residents are used to the bad version and not used to interpreting the other versions, so there's a ton of resistance...
+and I say this as someone who grew up plotting hurricane paths on maps we got from the grocery store in May at the beginning of hurricane season :).
+These displays are also complicated by the asymmetry of wind/flood risk around the eye of the storm - the northwest side gets hit harder in the northern hemisphere, which leads to natives referring to the "dirty side" and the "clean side" of the storm.
+Land topology also shapes flood risk, which is yet another complexity.
+All of this is to say that the explicitly calculated risk maps do a lot of external cognition, but when you're used to doing that calculus mentally, it can be a bit hard to trust the predigested information.
+-->
 Other authors have commented on the complexity of communicating hurricane risk because the path, storm surge and wind speed are all important and cannot be ignored [@Spiegelhalter2017]. 
-Interestingly it is unclear how the participants were supposed to answer this question.
+Interestingly, it is unclear how the participants were supposed to answer this question. 
 @Padilla2017 did three experiments and the first and third experiment had conflicting assumptions in their ground truth. 
-In the first study, participants were required assume to the storm was equally intense reguardless of the probability of the probability of the oil rig being hit (i.e.
-more likely to be hit *does not* mean more damage) but in the second study required participants to assume the opposite (i.e.
+In the first study, participants were required assume to the storm was equally intense regardless of the  probability of the oil rig being hit (i.e.
+more likely to be hit *does not* mean more damage), but in the second study required participants to assume the opposite (i.e.
 more likely to be hit *does* mean more damage).
@@
 Evaluating visualisations as though they are akin to hypothesis tests is a well established concept in visualisation.
-The lineup protocol is a good example of this approach, which is a confirmatory visualisation tool that can be used to check if perceived patterns are real or merely the result of chance [@Buja2009; @Wickham2010].
+The lineup protocol is a good example of this approach.
+Lineups are a confirmatory visualisation tool that can be used to check if perceived patterns are real or merely the result of chance [@Buja2009; @Wickham2010]. 
+Lineups can also be used to evaluate the effectiveness of different types of plots [@hofmann2012] and design decisions [@vanderplas2017].
 The motivation behind the lineup protocol is similar to the motivations behind signal-suppression, although the lineup protocol is more explicitly tied to a specific hypothesis. 
@@
 We can use a similar evaluation method to evaluate uncertainty visualisations, however, special care would need to be taken for their versatility requirement.
-While lineup protocols often have a specific hypothesis that is used to create the null distribution, uncertainty visualisations are intended to be more flexible and can be created without visualizing a specific null distribution.
-Human viewers using the lineup protocol are less sensitive to deviations from the null hypothesis than the typical statistical tests [@Patrick2023].
+While lineup protocols often have a specific hypothesis that is used to create the null distribution, uncertainty visualisations are intended to be more flexible and can be created without identifying a specific null distribution.
+Human viewers using the lineup protocol are less sensitive to deviations from the null hypothesis than statistical tests [@Patrick2023]. 
 This means swapping to a visual format and giving viewers the ability to check several hypothesis at once can reduce the effectiveness of the hypothesis test.
-Therefore it is highly likely that uncertainty visualisations would not produce identical results to standard hypothesis tests or even the lineup-protocol.
+<!-- I'm not convinced this reduces the effectiveness - it increases the power of the test to detect "something is not right" even if that something is unspecified.
+See Vanderplas 2017 (Clusters beats trend...) for an example of where viewers found something that wasn't right that was absolutely not what we intended, but was still very visually relevant.
+-->
+Therefore, it is highly likely that uncertainty visualisations would not produce identical results to standard hypothesis tests or even the lineup protocol.
 Additionally, uncertainty visualisations would need to be simultaneously checked against several hypothesis that come from a variety of null distributions.
@@
 # Future work
-This paper has identified gaps in the uncertainty visualisation literature that could be filled to progress the field.
+This paper has identified gaps in the uncertainty visualisation literature that must be filled for the field to progress. 
 We formalized the uncertainty visualisation problem, and in doing so highlighted an untouched area of uncertainty visualisation research.
@@
 
-*Experimenters should consider evaluating visual aesthetics on how well they suppress information in a graphic.* Research into separability and integrability is of particular interest to uncertainty visualisation, as it allows interference from the uncertainty variable.
-When designing experiments, authors often choose aesthetics that are visually distinguishable, uncertainty visualisation authors should consider doing the opposite.
-
-*Software that allows users to easily perform signal suppression should be developed.* Existing uncertainty visualisation methods view a distribution as its own object and there are no software options for the "an uncertainty visualisation is a function of a visualisation" philosophy.
+*Experimenters should consider evaluating visual aesthetics on how well they suppress information in a graphic.* 
+Research into separability and integrability is of particular interest to uncertainty visualisation, as it allows interference from the uncertainty variable. 
+When designing experiments, authors often choose aesthetics that are visually distinguishable; uncertainty visualisation authors should consider doing the opposite.
+
+*Software that allows users to easily perform signal suppression should be developed.* 
+Existing uncertainty visualisation methods view a distribution as its own object and there are no software options for the "an uncertainty visualisation is a function of a visualisation" philosophy.
 

### 2024-09-17 paper.qmd
made some edits based on Susans comments

@@
 Today, there are an abundance of publications on the topic which makes it timely to construct a review of the field.
+That is, now that there is an overwhelming amount of information, it is valuable to distill it into simple facts. 
 In fact, there have already been several reviews published.
-
-<!-- Susan: I think there's a missed opportunity here to reference the premise you started with, that when there are heaps of information, it's valuable to distill it down into simple facts -->
 
@@
 The simplest way to add uncertainty to an existing graphic is to simply map uncertainty to an unused visual channel. 
-However, it is unclear if this approach is 
-sufficient XXX Word choice??? XXX
-<!-- I changed will to sufficient, but I'm not sure that's what you were going for. -->
-for our purposes.
+However, it is unclear if this approach is sufficient for our purposes.
 
@@
 In order to make statements about population statistics, future values, or new observations we need to perform inference, which requires the assumption of the "uniformity of nature", that is, we need to assume that unobserved phenomena should be similar to observed phenomena [@Otsuka2023].
-Positivists abhor the assumption of the "uniformity of nature" as they believed referencing the unobservable is bad science. 
-In other words, these scientists embraces descriptive statistics and shunned inferential statistics due to the inherent uncertainty that came with them. 
+Positivists believed referencing the unobservable was bad science. 
+In other words, these scientists embraced descriptive statistics due to the inherent certainty that came with them.
 Since uncertainty is non-existent in descriptive statistics, it is clear that uncertainty is a by-product of inference.
-<!-- Inferential statistics didn't yet exist when positivists were dominant - they weren't making a choice between description and inference, as the choice didn't exist.
-This is slightly misleading as written, even if I generally agree with the concept you're trying to communicate.
--->
 

### 2024-09-18 paper.qmd
fixed/replied to a few more of Susan's comments

@@
 There are enough qualitative descriptors of uncertainty to fill a paper, but, none of this is particularly helpful in understanding how to integrate it into a visualisation. 
-Rather than trying to define uncertainty by what it *is* it may be easier to try and describe what uncertainty *is not*.
+Rather than trying to define uncertainty by looking at the mirriad of ways in which it *does* appear in an analysis, we may find it easier to look at where it *does not*.
 
@@
 This aspect of uncertainty visualization makes it a uniquely difficult problem.
-
-<!-- You've promised to describe what uncertainty is not...
-did you achieve that goal?
-If not, you probably should revisit how you're introducing the philosophical bit about descriptive statistics and positivism.
--->
 
@@
 @Hullman2021 argued that there is no such thing as a "model-free" visualisation, therefore all visualisations require uncertainty as we are always performing inference. 
-However, even something as simple as calculating the uncertainty that is used to suppress our visualisation is not model free, as we need to identify if the sampling variance or the sample variance is more appropriate [@Hofman2020].
-<!-- Ok, so this is an additional challenge on top of Hullman's claim, but it's not refuting Hullman's claim.
-I think the easier way to do that is to say that while you can certainly think of visualizations as containing implicit inferential properties, there are many potential inferences in any single visualization; until the viewer chooses one and thinks about it (which isn't easily observable), the particular variety of uncertainty which would need to be displayed can't be calculated - it's a quantum physics type issue where until something is observed, the singularity doesn't collapse and everything exists only in probabilistic terms.
-The cat isn't dead or alive until the viewer starts thinking inferentially in one way or another.
--->
-This means an ideal uncertainty visualisations should not only meet the signal suppression requirements, but should also endeavor to be versatile enough to meet those requirements for all the signals displayed in the graphic.
+While it is true that we can think of visualisations as containing implicit inferential properties, there are many potential inferences in any single visualization. 
+This makes it a little difficult ensure uncertainty is always included.
+For example, if we have a visualisation that shows an average, we would need to identify if the signal suppression should be performed using the sampling variance or the sample variance [@Hofman2020].
+The distribution we use depends on the inferential statistic but until the viewer chooses one and thinks about it (which isn't easily observable), the particular variety of uncertainty which would need to be displayed can't be calculated.
+This means the ideal uncertainty visualisation should not only meet the signal suppression requirements, but should also endeavor to be versatile enough to meet those requirements for all the signals displayed in the graphic.
 
@@
 At this point, it is irrelevant whether or not the colours are technically different, they are the same color in the human brain. 
-<!-- This is of course complicated by the fact that human colour perception varies at an individual level: some individuals have been documented to have 4 different cones and can thus perceive a different range of colors; others have only two or one functioning cone and have colour deficiencies.
+This is of course complicated by the fact that human colour perception varies at an individual level: some individuals have been documented to have 4 different cones and can thus perceive a different range of colors; others have only two or one functioning cone and have colour deficiencies.
 For VSUP to function for all individuals, we must calibrate each plot to an individual's ability to perceive colour.
--->
-
-If we only use a single value to express each signal-suppressed statistic, we will always need to decide which signals we suppress and which we do not. 
+
+If we only use a single colour to express each signal-suppressed statistic, we will always need to decide which signals we suppress and which we do not. 
 This issue has already been raised in the literature. 
@@
 -->
-
+<!--    
+- I think the point I’m trying to make here is not that the raw data shows uncertainty by showing variation, but rather that it shows uncertainty by allowing people to see violations to the assumptions they might have made when mentally calculating an inferrential statistic
+  - e.g. if i show the raw data and the areas are unevenly spaced or there are a lot of missing observations in some areas.
+- I agree that the variability in the raw data is not related to the variability in the inferential statistic even though this section certainly implies that. I need to fix that.
+- The connection with the pixel map was supposed to be that it would theoretically give a sense of the shape of the distribution.
+When I said “where the variability is coming from” I was trying to refer to that shape, but that doesn’t really translate from what I’ve written and the connecton between the pixel map and showing the raw data isn’t really there.
+  - I think the logic line I was going for was
+    Pixel map shows collection of hypothetical outcomes of sampling distribution 
+    => These outcomes let us see the shape of the distribution 
+    => When seeing the shape of the distribution, we may become aware of errors in our assumptions, (e.g. such as normality) 
+    => If we want to be aware of where our uncertainty is coming from (e.g. underlying variability or assumption violations) then the raw data is what best shows this
+- I did consider including a visualisation that shows the data I used for the pixel map as a final map visualisation to better separate these concepts, but I didn’t technically use any real data (since I just made the normality assumption which is excatly what this map should highlight)
+  - It might be better to change the statistic and give the sampling distribution in each area a different shape to make the pixel map benefit clearer.
+  - I just want to make sure this is an accurate description of the issue you were highlighting before I fix it up this section (also I think it will take me a bit to figure out how to fix it) 
+-->
 While the pixel map does not literally show the data, it does imitate showing the data and give us details on where the variability is coming from. 
@@
 -->
-
 Setting trust as the variable of interest implicitly encourages statisticians to set trust and as the primary goal of communication. 

### 2024-09-19 paper.qmd
added reply/question to susans comment on trust section

@@
 -->
+<!--
+I see there are three “directions of influence” we can use to think about the relationship between trust and accuracy:
+  1) Accuracy impact trustworthiness:
+     - This is basically the main consideration of this point of this section.
+The point is that historical accuracy impacts “trustworthiness” which then impacts trust.
+Since trust is impacted by the trustworthiness of the source, so you should endeavor to be as accurate as possible to establish trustworthiness if you want your information to be trustworthy.
+  2) Trustworthiness impacts accuracy:
+    - I agree that the accuracy can be influenced by trust.
+We had a climate scientist discussing this exact thing at last week’s numbat seminar and had a good citation for this.
+I could probably go find but it used political beliefs as a proxy rather than “trust” so it might not be exactly right but it is thinking along these lines.
+  3) Trustworthiness and accuracy combined impact trust (or belief or something):
+    - The final statistic of “trust” is a combination of sourth trustworthiness and accuracy.
+Since trust can be impacted by accuracy or trustworthiness, pursuing trust rather than trustworthiness can encourage inaccuracy.
+I think that commenting on (2) is a good addition to this section, but it doesn’t really change my opinion on (1) and (3) (I’m not 100% sure if this comment was an “in addition” or “in dispute” comment and I want to check before I change it)
+
+I agree with your point about the accuracy in display vs perception but I don’t see its connection to the greater point about trust and I’m not 100% sure it fits in this section.
+Accuracy in the visual form (e.g.
+correcting for sine illusion) is still a pursuit of accuracy, just accuracy in the visual statistic.
+Like, if we use Frequentist or Bayesian statistics, they also have different inputs, processes and outputs and they both give us different estimates.
+It is also common to adjust those statistics to correct for a bias in the calculation (e.g.
+we might weight the observations).
+I don’t really see the difference between these methods and editing a graphic to correct for the sine illusion.
+I don’t think any of these things conflict with the pursuit of scientific accuracy.
+I’m not against adding this as an additional comment (i.e.
+that accuracy in a visualisation may not be what the reader thinks it is) but I’m not certain it has any serious implications for the trust
+-->
 Setting trust as the variable of interest implicitly encourages statisticians to set trust and as the primary goal of communication. 
@@
 This perspective ends up extending to visualisation authors as well.
-@Hullman2020a found that author simultaneously argued that failing to visualise uncertainty was akin to fraud, but also many avoided uncertainty visualisation because they didn't want their work to come across as "untrustworthy".
+@Hullman2020a found that authors simultaneously argued that failing to visualise uncertainty was akin to fraud, but also many avoided uncertainty visualisation because they didn't want their work to come across as "untrustworthy".
 This is a classic example of the negative impacts of placing direct importance on *trust* rather than *transparency*. 

### 2024-09-20 paper.qmd
Updated comments/responses

@@
 - I agree that the variability in the raw data is not related to the variability in the inferential statistic even though this section certainly implies that. I need to fix that.
+-->
+<!-- Ok, but variability in the raw data is also related to variability in the inferential statistic, it's just that as n -> infinity, the variability in the inferential statistic goes to 0.  -->
+<!--
 - The connection with the pixel map was supposed to be that it would theoretically give a sense of the shape of the distribution.
@@
 -->
+<!-- Yes, this looks like you're thinking along the same lines as I was -->
 While the pixel map does not literally show the data, it does imitate showing the data and give us details on where the variability is coming from. 
@@
 -->
+<!-- That's totally fair :) Just quibbling - not something you have to change unless someone else calls it out! -->
 Setting trust as the variable of interest implicitly encourages statisticians to set trust and as the primary goal of communication. 

### 2024-09-20 paper.qmd
spelling fix

@@
 There are enough qualitative descriptors of uncertainty to fill a paper, but, none of this is particularly helpful in understanding how to integrate it into a visualisation. 
-Rather than trying to define uncertainty by looking at the mirriad of ways in which it *does* appear in an analysis, we may find it easier to look at where it *does not*.
+Rather than trying to define uncertainty by looking at the myriad ways in which it *does* appear in an analysis, we may find it easier to look at where it *does not*.
 

### 2024-09-23 paper.qmd
replied to comment 9

--- 2024-09-19 paper.qmd
@@
 For example @Hofmann2012 showed study participants 20 plots, where each plot displayed two distributions (as a pair of jittered samples, density plots, histograms, or box plots) and asked them to identify the plot where "the blue group is furthest to the right".
-The mean of the two groups was used to decide which group was actually "furthest to the right" and the participants were evaluated against that ground truth. 
+The mean of the two groups was used to decide which group was actually "furthest to the right" and the different graphical approaches were evaluated against that ground truth. 
 <!-- Participants weren't evaluated against that ground truth, though - the plots were what was actually being evaluated in that study, not participant accuracy.
@@
 -->
+<!-- Good point about the plots being evaluated, not the participants. I changed that.
+I agree with the point about lineups being useful in identifying how participants complete evaluation tasks, and I did comment on that with respect to this particular paper later in the “Qualitative studies” section with this comment:
+   “@Hofmann2012 tried to capture this by asking participants why they considered a particular plot to be more "right shifted", even though this qualitative analysis did not make it into the final paper.
+“
+Do you think I should move that comment to this section to keep all comments on that paper together, leave it where it is, or do you think there should be a different edit? -->
 In another example @Ibrekk1987 asked participants for the "best estimate", but the participant's responses were evaluated against the mean of the distribution.

### 2024-09-23 paper.qmd
editted for susan comment 10

@@
 Ultimately this approach results in strangely cryptic questions that create a large amount of noise due to the varied interpretation of the questions [@Hullman2016].
+
 For example @Hofmann2012 showed study participants 20 plots, where each plot displayed two distributions (as a pair of jittered samples, density plots, histograms, or box plots) and asked them to identify the plot where "the blue group is furthest to the right".
@@
 The ground truth statistic for this question was a scalar multiple of Cohen’s d, indicating participants were supposed to incorporate uncertainty information using a very specific formula that was likely unknown to them but assumed to be used implicitly.
-<!-- Might be worth referencing the fact that visual statistics and the computational equivalents often differ, and this is even before you start talking measurement error when using a Likert scale.
--->
 In another example, @Padilla2017 provided participants with a visualisation of the cone of uncertainty and asked then to "decide which oil rig will receive more damage based on the depicted forecast of the hurricane path".
@@
 While these approaches are certainly a step in the right direction, the experiments end up having far too much noise in their results. 
+This is further complicated by the fact that visual statistics and their computational equivalents often differ, even before we consider these measurement errors.
 Additionally, by having a ground truth, they end up implicitly asking users to treat uncertainty as a signal which goes against the purposes of signal-suppression.

### 2024-09-23 paper.qmd
edited for susan comment 11

@@
 The cone of uncertainty provides a 60% confidence interval for the location of the eye of a hurricane, which allows us to know the area where the eye of the storm will go, but it does not given any information about the intensity of a storm, the size of a storm, or even if a location will be hit.
-<!-- It's not unreasonable to use forms of display that are used in gov't communications, even if they're admittedly not optimal -- it can be reasonable to *show* that they're not optimal, and Padilla is usually advocating for different risk displays as a result, including spaghetti maps for the track and probabilistic wind speed/flood risk maps for the impacts.
-The problem is that gulf coast residents are used to the bad version and not used to interpreting the other versions, so there's a ton of resistance...
-and I say this as someone who grew up plotting hurricane paths on maps we got from the grocery store in May at the beginning of hurricane season :).
-These displays are also complicated by the asymmetry of wind/flood risk around the eye of the storm - the northwest side gets hit harder in the northern hemisphere, which leads to natives referring to the "dirty side" and the "clean side" of the storm.
-Land topology also shapes flood risk, which is yet another complexity.
-All of this is to say that the explicitly calculated risk maps do a lot of external cognition, but when you're used to doing that calculus mentally, it can be a bit hard to trust the predigested information.
--->
 Other authors have commented on the complexity of communicating hurricane risk because the path, storm surge and wind speed are all important and cannot be ignored [@Spiegelhalter2017]. 
-Interestingly, it is unclear how the participants were supposed to answer this question. 
-@Padilla2017 did three experiments and the first and third experiment had conflicting assumptions in their ground truth. 
-In the first study, participants were required assume to the storm was equally intense regardless of the  probability of the oil rig being hit (i.e.
-more likely to be hit *does not* mean more damage), but in the second study required participants to assume the opposite (i.e.
-more likely to be hit *does* mean more damage).
+@Padilla2017 seem to be aware of the issues in the visualisation and are using the user study to show that current government communications are not the best choice when communicating hurricane risk, as it is unreasonable to assume that "more likely to be hit" should automatically translate to "more damage".
+While this is useful to highlight, the authors end up caught by their own trap, as answering the questions in their second study correctly required participants to assume an area being more likely to be hit *does* translate to receiving more damage.
+The assumptions required to answer these deterministic questions are often invisible, even to the authors of the paper, which means they often create more problems than they solve when they are used as a tool for graphical evaluation.
 Similar to the vague question problem, these studies seem to be aware that we should be evaluating an uncertainty visualization based on a deterministic observation (such as our example of "does this map have a spatial trend") but are unsure how to incorporate or evaluate it.

### 2024-09-23 paper.qmd
edited according to susan comment 12

@@
 This approach is similar to the lineup protocol as it also evaluates a visualisation by comparing it to an analogous statistical calculation, however it does so using a different statistical philosophy.
-
-We can use a similar evaluation method to evaluate uncertainty visualisations, however, special care would need to be taken for their versatility requirement.
-While lineup protocols often have a specific hypothesis that is used to create the null distribution, uncertainty visualisations are intended to be more flexible and can be created without identifying a specific null distribution.
-Human viewers using the lineup protocol are less sensitive to deviations from the null hypothesis than statistical tests [@Patrick2023]. 
-This means swapping to a visual format and giving viewers the ability to check several hypothesis at once can reduce the effectiveness of the hypothesis test.
-<!-- I'm not convinced this reduces the effectiveness - it increases the power of the test to detect "something is not right" even if that something is unspecified.
-See Vanderplas 2017 (Clusters beats trend...) for an example of where viewers found something that wasn't right that was absolutely not what we intended, but was still very visually relevant.
--->
-Therefore, it is highly likely that uncertainty visualisations would not produce identical results to standard hypothesis tests or even the lineup protocol.
-Additionally, uncertainty visualisations would need to be simultaneously checked against several hypothesis that come from a variety of null distributions.
+The connection with visualisation and statistical tests even extends to EDA, where @Hullman2021 commented that a visualisation can be considered a model check where the null hypothesis is "the expected" and we are looking for "the unexpected".
+This is all to say that comparing visualisations to hypothesis tests is not uncharted territory.
+
+We could use a similar evaluation method to evaluate uncertainty visualisations, however, it is not entirely clear what the computational equivalent of an uncertainty visualisation is.
+Even the lineup protocol, which is much closer to a standard statistical test, runs into some difficulty when trying to directly compare it to its computational equivalent.
+@Patrick2023 found that human viewers using the lineup protocol are less sensitive to deviations from the null hypothesis than standard statistical tests.
+However, this finding does not directly translate to the lineup protocol being worse than its computational equivalent.
+The lineup protocol allows viewers to detect "something is not right" even if that something is unspecified, and even if that "something" is not what see the authors intended (for an example see @vanderplas2017).
+While standard hypothesis tests may outperform the lineup protocol on power, computational tests have the benefit of knowing exactly what they are looking for, while visualisations are often explicitly made when we *don't* know what we are looking for.
+Additionally, while lineup protocols often have a specific hypothesis that is used to create the null distribution, uncertainty visualisations are intended to be more flexible and can be created without identifying a specific null distribution.
+Therefore, while we could evaluate uncertainty visualisations by comparing them to a selection of hypothesis tests, we should not expect the same level of sensitivity that we get in standard hypothesis tests as the assumed knowledge about our data is quite different.
+This approach could give us a good idea of the use cases of different uncertainty visualisation designs, but we should be mindful of it's limitations in fully describing the effectivness of these graphics.
 

### 2024-09-24 paper.qmd
added comment about changing broom version to get pdf to knit

@@
 library(flextable)
+
+# Broom returns an error with Vizumap, run this if can't compile
+# devtools::install_version(package = "broom", version = "1.0.4")
 ```
@@
 Lineups are a confirmatory visualisation tool that can be used to check if perceived patterns are real or merely the result of chance [@Buja2009; @Wickham2010]. 
-Lineups can also be used to evaluate the effectiveness of different types of plots [@hofmann2012] and design decisions [@vanderplas2017].
+Lineups can also be used to evaluate the effectiveness of different types of plots [@Hofmann2012] and design decisions [@vanderplas2017].
 The motivation behind the lineup protocol is similar to the motivations behind signal-suppression, although the lineup protocol is more explicitly tied to a specific hypothesis. 

### 2024-09-24 paper.qmd
fixed pixel map palette so it isnt truncated anymore

@@
 
-# Make standard Palette
-# set Base colours
-basecols <- brewer.pal(8, name = "YlOrRd")
-#breaks <- seq(0,41, length.out = 9)
-breaks <- 21:29 #c(0,15,24,25,26,27,30,35,41)
+# Make palettes
+longpal <- rev(colorspace::sequential_hcl(13, palette = "YlOrRd"))
+basecols <- longpal[3:10]
+breaks <- 21:29 
+breakslong <- 18:32 
 names(basecols) <- seq(8)
-limits <- c(10,41)
+names(longpal) <- -1:11
 
@@
 
-show_pal <- function (colours, borders = NULL, cex_label = 1, ncol = NULL, myxlab, breaks) {
+show_pal <- function (colours, borders = NULL, cex_label = 1, ncol = NULL, myxlab, breaks, textnudge, xlabx, xlaby) {
   # Set dimensions of palette
@@
        col = colours, border = borders)
-  text(c(0,col(colours)) + c(0.2, 0.1, 0,0,0,0,0,-0.1,-0.2), -c(1,row(colours))-0.25, breaks, 
+  text(c(0,col(colours)) + textnudge, -c(1,row(colours))-0.25, breaks, 
               cex = 1, col = "black")
-  text(4, -1.75, myxlab ,cex = 1.5, col = "black")
+  text(xlabx, xlaby, myxlab ,cex = 1.5, col = "black")
 }
@@
 
-show_pal(basecols, ncol=8, borders=NA, myxlab = "Temperature", breaks = 21:29)
+show_pal(basecols, ncol=8, borders=NA, myxlab = "Temperature", breaks = 21:29, textnudge = c(0.2, 0.1, 0,0,0,0,0,-0.1,-0.2), xlabx= 4, xlaby=-1.75)
 
@@
 p2b
-show_pal(basecols, ncol=8, borders=NA, myxlab = "Probability Temp>27", breaks = c(0, prob_breaks[2:8], 1))
+show_pal(basecols, ncol=8, borders=NA, myxlab = "Probability Temp>27", breaks = c(0, prob_breaks[2:8], 1), textnudge = c(0.2, 0.1, 0,0,0,0,0,-0.1,-0.2), xlabx= 4, xlaby=-1.75)
 ```
@@
                   p0.75 = qnorm(0.75, mean=bitemp, sd=sqrt(variance)), 
-                  p0.95 = qnorm(0.95, mean=bitemp, sd=sqrt(variance))))|>
-  apply(2, function(x) ifelse(x>8, 8, x)) |>
-  apply(2, function(x) ifelse(x<0, 0, x))
+                  p0.95 = qnorm(0.95, mean=bitemp, sd=sqrt(variance))))
 
@@
     )  +
-  scale_fill_gradientn(colours = basecols) +
-  scale_colour_gradientn(colours = basecols) +
+  scale_fill_gradientn(colours = longpal) +
+  scale_colour_gradientn(colours = longpal) +
   theme(legend.position="none")
@@
                   p0.75 = qnorm(0.75, mean=bitemp, sd=sqrt(variance)), 
-                  p0.95 = qnorm(0.95, mean=bitemp, sd=sqrt(variance))))|>
-  apply(2, function(x) ifelse(x>8, 8, x)) |> #this will mess with the variance
-  apply(2, function(x) ifelse(x<0, 0, x))
+                  p0.95 = qnorm(0.95, mean=bitemp, sd=sqrt(variance))))
 
@@
 p4b
-show_pal(basecols, ncol=8, borders="white", myxlab = "Temperature", breaks = 21:29)
+show_pal(longpal, ncol=13, borders="white", myxlab = "Temperature", breaks = 19:32, textnudge = c(0.2, 0.1, 0,0,0,0,0,0,0,0,0,0,-0.1,-0.2), xlabx= 6.5, xlaby=-2.2)
 ```

### 2024-09-24 paper.qmd
changed underlying spatial trend

@@
 my_map_data <- my_map_data |>
-  dplyr::mutate(temp = 28 - ((scale(cent_long))^(2) + (scale(cent_lat))^(2))[,1], # trend
+  dplyr::mutate(temp = 29 - 3*abs(scale(cent_lat) - sin(2*(scale(cent_long)))[,1]), # trend
          notrend = rnorm(n), # no trend
@@
   geom_sf(aes(geometry = geometry), fill="white") + 
-  #geom_text(aes(x=cent_long, y=cent_lat, label=count_id), size=2) +
+  geom_text(aes(x=cent_long, y=cent_lat, label=count_id), size=2) +
   theme_void()
@@
 p1b
-
 show_pal(basecols, ncol=8, borders=NA, myxlab = "Temperature", breaks = 21:29, textnudge = c(0.2, 0.1, 0,0,0,0,0,-0.1,-0.2), xlabx= 4, xlaby=-1.75)

### 2024-09-24 paper.qmd
got rid of half written sentance (I dont remember what it was supposed to say)
changed weird variance calculation

@@
 We will use these maps to show the effects we are talking about, and highlight the costs and benefits of each approach.
-These examples will need XXX incomplete sentence!! XXX
 
@@
   dplyr::mutate(temp = 29 - 3*abs(scale(cent_lat) - sin(2*(scale(cent_long)))[,1]), # trend
-         notrend = rnorm(n), # no trend
-         highvar = runif(n, min=1, max=2), # high variance
-         lowvar = runif(n, min=0, max=1), # low variance
-         ) |>
-  mutate(highvar = highvar^2,
-         lowvar = lowvar^2,
+         highvar = runif(n, min=2, max=4), # high variance
+         lowvar = runif(n, min=0, max=2), # low variance
          count_id = row_number()) |>

### 2024-09-25 paper.qmd
Updated with pages 1-14 of Emi's comments

@@
 
-# Background
+# Introduction
 
@@
 Among these summary tools, information visualisations are one of the most powerful. 
-Visualisations allow us to **learn** what is important about our data set and datasets such as Anscombe's quartet [@anscombe] or the Datasaurus Dozen [@datasaurpkg] highlight this power. 
+Visualisations allow us to *learn* what is important about our data set and datasets such as Anscombe's quartet [@anscombe] or the Datasaurus Dozen [@datasaurpkg] highlight this power. 
 Additionally, visualisations allow for efficient and memorable communication. 
@@
 Despite the new terminology visualisation of uncertainty has been present since the earliest times. 
-For example, box plots or histograms can be considered to be displaying uncertainty in the sense of variability in observations sampled from a population distribution. 
+For example, box plots and histograms display variation which becomes synonymous with uncertainty when we are using them to depict the variation of an estimate. 
 Today, there are an abundance of publications on the topic which makes it timely to construct a review of the field.
 That is, now that there is an overwhelming amount of information, it is valuable to distill it into simple facts. 
-In fact, there have already been several reviews published.
+In fact, there have already been several reviews published but a central piece of discussion is missing.
 
@@
 @Hullman2016 noticed there is a serious noise issue in the data coming from uncertainty visualisation experiments. 
-She commented on the prevalence of confounding variables that make it unclear as to what exactly caused a subjects poor performance on a particular questions.
-Mistakes due to misunderstood a visualisations, misinterpreted questions, and incorrectly applied heuristics are all combined into a single error value.
+She commented on the prevalence of confounding variables that make it unclear as to what exactly caused a subjects poor performance on a set of particular questions.
+Mistakes due to misunderstanding a visualisations, misinterpreted questions, and incorrectly applied heuristics are all combined into a single error value.
 @Spiegelhalter2017 commented that different plots are good for different things, and disagreed with the goal of identifying a universal "best" plot for all people and circumstances.
@@
 @uncertchap2022 summarized several cognitive effects that repeatedly arise in uncertainty visualisation experiments, however these effects were each discussed in isolation as a list of considerations an author might make rather than an overarching theory of rules for effective uncertainty visualisation.
-While these reviews are thorough in scope, none discuss how the existing literature contribute to the commonly stated goal of uncertainty visualisation, scientific transparency. 
+While these reviews are thorough in scope, none discuss how the existing literature contribute to the borader goal of uncertainty visualisation. 
 The problem faced by the literature is easily summarized with a famous quote by Henri Poincaré.
@@
 
-It would be impossible to display every single uncertainty visualisation in this review, there are far too many so for the sake of clarity, we will focus on the changes made to a single plot, the choropleth map.
+There are far too many uncertainty visualisations to exhaustively discuss them all. Instead we focus on the changes made to a single plot, the choropleth map. 
 Due to the field's origins and focus in geospatial information visualisation, there have been a large number of suggested variations on the choropleth map that allow authors to include uncertainty. 
-While the theoretical approach we outline is useful to all uncertainty visualisations, regardless of their application, we feel that focusing on a single example will help isolate the ideas we are trying to convey.
+While the theoretical approach we outline is useful to all uncertainty visualisations regardless of their application, focusing on a single example will help isolate the ideas we are trying to convey.
 We will use these maps to show the effects we are talking about, and highlight the costs and benefits of each approach.
@@
 @fig-data shows the first six rows and the geographical boundaries of our data set. 
-The temperature variable was generated using a quadratic spatial trend, that is $Temperature_i = 28-Longitude_i^2 + Latitude_i^2$ where the $Longitide$ and $Latitude$ are the longitude and latitude of the county's centroid scaled to a standard normal distribution.
+The temperature variable was generated using a sine wave, that is $Temperature_i = 29-3\cdot{|Latitude_i - \sin{2 \cdot Longitude_i}|}$ where the $Longitide$ and $Latitude$ are the longitude and latitude of the county's centroid scaled to a standard normal distribution.
 Each county's variance is independently, randomly sampled from a uniform distribution. 
@@
 # Ignoring uncertainty
-A good place to start might be at deceptively straight forward question, why should we include uncertainty at all?
+A good place to start might be at a deceptively straight forward question, why should we include uncertainty at all?
 
@@
 This is *not* because uncertainty in general is important for decision making, but because it gives you the tools required to calculate the *actual* statistic you are basing your decision on. 
-In this sense, uncertainty is no more "special" to decision making than weight is special to a BMI calculation.
+In this sense, uncertainty is no more "special" to decision making than weight is special to a body mass index calculation.
 
@@
 These graphics are just normal information visualisations, and authors can follow existing principles of graphical design. 
-We will focus on the perspective that uncertainty visualisation serves to obfuscate signal, and an uncertainty visualisation is a variation on an existing graphic that gives it the ability to suppress false signals.
-
-Of course, we do not believe there is anything wrong with explicitly visualizing variance, error, bias, or any other statistic used to depict uncertainty as a signal. 
+We focus on the perspective that uncertainty visualisation serves to obfuscate signal, and an uncertainty visualisation is a variation on an existing graphic that gives it the ability to suppress false signals.
+
+Of course, there is nothing wrong with explicitly visualizing variance, error, bias, or any other statistic used to depict uncertainty as a signal. 
 Just like any other statistic, these metrics provide important and useful information for analysis and decisions. 
@@
 
-There is another benefit to mapping uncertainty to stauration that is not directly related to integrability. 
+There is another benefit to mapping uncertainty to saturation that is not directly related to integrability. 
 Colour value has a second desirable quality as the colours become harder to distinguish as the saturation decreases. 
-This means high uncertainty saturations are harder to differentiate than low uncertainty values. 
+This means high uncertainty values are harder to differentiate than low uncertainty values. 
 We can leverage this implicit feature of colour value by transforming the visual feature space ourselves.
@@
 #| fig-cap: "A map made with a VSUP.
-The counties of Iowa are coloured acording to it's average daily temperature and the variance in temperature, although the space they have been mapped to is slightly warped.
+The counties of Iowa are coloured acording to its average daily temperature and the variance in temperature, although the space they have been mapped to is slightly warped.
 Similar to the bivariate map, temperature is mapped to hue while variance is mapped to saturation.

### 2024-09-28 paper.qmd
finished small emi edit suggestions

@@
 While it is true that we can think of visualisations as containing implicit inferential properties, there are many potential inferences in any single visualization. 
-This makes it a little difficult ensure uncertainty is always included.
+This makes it a little difficult to ensure uncertainty is always included.
 For example, if we have a visualisation that shows an average, we would need to identify if the signal suppression should be performed using the sampling variance or the sample variance [@Hofman2020].
@@
 We want to know "Which of these coefficients are different from 0?" as well as "Which of these coefficients are different from each other?". 
-To answer this question we do a series of t-tests on these estimates. 
-All of the individual t-tests of fail to reject the null hypothesis that the coefficients are different from 0. 
+To answer this question we do a series of $t$-tests on these estimates. 
+All of the individual $t$-tests of fail to reject the null hypothesis that the coefficients are different from 0. 
 We then make a visualisation that suppresses this signal and ensures that all of the estimates are visually indistinguishable from 0. 
-Next, we conduct two sample t-test and find that several of the values need to be visually distinguishable from each other. 
+Next, we conduct two sample $t$-test and find that several of the values need to be visually distinguishable from each other. 
 The VSUP method must pick a single color for each estimate, and these colors must be *either* visually distinguishable or indistinguishable from each other. 
@@
 At this point, it is irrelevant whether or not the colours are technically different, they are the same color in the human brain. 
-This is of course complicated by the fact that human colour perception varies at an individual level: some individuals have been documented to have 4 different cones and can thus perceive a different range of colors; others have only two or one functioning cone and have colour deficiencies.
+This is of course complicated by the fact that human colour perception varies at an individual level. 
+Some women are believed to have four different cone cells which allows them to perceive a greater range of colors, while others only have one or two types of cone cells and have colour deficiencies [@simunovic2010]
 For VSUP to function for all individuals, we must calibrate each plot to an individual's ability to perceive colour.
@@
 Not only does this method help readers understand the plot level "gist", it is also unlikely to damage the viewers ability to extract individual estimates.
-Extracting global statistics, such as the mean or variance, from a sample can be done with relative each, especially when those values are mapped to color [@Franconeri2021]. 
+Extracting global statistics, such as the mean or variance, from a sample can be done with relative ease, especially when those values are mapped to color [@Franconeri2021]. 
 Additionally, @Ndlovu2023 found that participants applied the same methods they used for simple choropleth maps to complicated uncertainty maps even if that take away was invalid. 
@@
 @Chowdhury illustrated how groups that appear linearly separable in a linear discriminant analysis (LDA) visualisation of the data can actually be the result of a LDA performed on too many variables, something that was not clear from the visualisation until the line-up protocol was implemented.
-However, just showing the data it is simple but effective option for uncertainty visualisation that that seems to be largely overlooked. 
+However, just showing the data is a simple but effective option for uncertainty visualisation that that seems to be largely overlooked. 
 <!-- Does showing the data actually show uncertainty, or does it just show variance?
@@
 While the pixel map does not literally show the data, it does imitate showing the data and give us details on where the variability is coming from. 
-In order to uncerstand *why* pixel maps seem to be successful at signal suppression at first glace would likely require a more thorough discussion than what what we can offer here.
+In order to understand *why* pixel maps seem to be successful at signal suppression at first glance would likely require a more thorough discussion than what what we can offer here.
 
@@
 ### Value extraction of uncertainty statistics
-Uncertainty visualisations are often evaluated based on how accurately [@Hullman2019] viewers can separately extract the estimate and the variance [@Kinkeldey2014]. 
+Uncertainty visualisations are often evaluated based on how accurately [@Hullman2019] viewers can separately extract the estimate and its variance [@Kinkeldey2014]. 
 This means a significant chunk of evaluation studies boil down to showing a participant a visualisation and asking questions such as "what is the variance of $X$?", or "what is the mean of $X$?". 
@@
 There is no reason to assume uncertainty would behave any differently to any other variable when we evaluate them in this way. 
-For example, @Ibrekk1987 found that participants were more accurate at extracting a statistic when it could be directly read off the graphic, than when it required an area estimate (which is the case if using the PDF), or when there was no visual indicator for the statistic at all (which is the case when using the CDF of an asymmetric function).
+For example, @Ibrekk1987 found that participants were more accurate at extracting a statistic when it could be directly read off the graphic, than when it required an area estimate (which is the case if using the probability density function), or when there was no visual indicator for the statistic at all (which is the case when using the cumulative distribution function of a skewed random variable).
 @Hullman2015 found that a visualisation that allows viewers to count outcomes to estimate a probability outperformed one that required a complicated area calculation. 
-@Hofman2020 and @Zhang2022 found that participants were better at answering questions about a prediction intervals when shown a prediction interval instead of a sampling distribution. 
+@Hofman2020 and @Zhang2022 found that participants were better at answering questions about prediction intervals when shown a prediction interval instead of a sampling distribution. 
 @Gschwandtnei2016 found that graphics where the required statistic could be directly read off the plot outperformed those that involved guesswork due to a gradually decreasing line. 
@@
 <!-- That's totally fair :) Just quibbling - not something you have to change unless someone else calls it out! -->
-Setting trust as the variable of interest implicitly encourages statisticians to set trust and as the primary goal of communication. 
+Setting trust as the variable of interest implicitly encourages statisticians to set trust as the primary goal of communication. 
 Evaluating visualisations on trust conflates trust and transparency and ultimately discourages signal-suppression. 
@@
 For example, @Zhao2023 found that participants were more trusting of model estimates with low uncertainty, but this effect did not carry over to estimates with high uncertainty. 
-Despite decreased trust being a desired outcome of signal suppression, the authors discussion implied this result was not desired. 
+Despite decreased trust being a desired outcome of signal suppression, the author's discussion implied this result was not desired. 
 This perspective ends up extending to visualisation authors as well.
@@
 Risk aversion is considered irrational behaviour because it is a deviation from the behaviour of a rational agent. 
-Comparing participant responses to that of a ration agent has even been suggested as a benchmark for uncertainty visualization experiments [@wu2023rational].
+Comparing participant responses to that of a rational agent has even been suggested as a benchmark for uncertainty visualization experiments [@wu2023rational].
 However, just like trust, evaluating graphics on risk-aversion discourages signal-suppression. 
@@
 Risk aversion is considered to be irrational because it means the economic agent *is* considering the uncertainty information.
-The only cases where a rational agent should not ignore the uncertainty information is when the uncertainty is signal, not noise. 
+The only case where a rational agent should not ignore the uncertainty information is when the uncertainty is signal, not noise. 
 Designing graphics that encourage choices that align with that of a rational agent, is to encourage graphics that do not include uncertainty at all.
@@
 ## Testing signal supression
-If we cannot ask direct questions about uncertainty, we cant measure the secondary effects of uncertainty, and we can't ask indirect questions about the uncertainty, how are we supposed to evaluate uncertainty visualisations?
+If we cannot ask direct questions about uncertainty, we cannot measure the secondary effects of uncertainty, and we cannot ask indirect questions about the uncertainty, how are we supposed to evaluate uncertainty visualisations?
 How do you measure something that disappears the second you look directly at it?
@@
 Therefore, while we could evaluate uncertainty visualisations by comparing them to a selection of hypothesis tests, we should not expect the same level of sensitivity that we get in standard hypothesis tests as the assumed knowledge about our data is quite different.
-This approach could give us a good idea of the use cases of different uncertainty visualisation designs, but we should be mindful of it's limitations in fully describing the effectivness of these graphics.
+This approach could give us a good idea of the use cases of different uncertainty visualisation designs, but we should be mindful of it's limitations in fully describing the effectiveness of these graphics.
 
 ### Qualitative Studies
-Alternatively visualisation research could shift away from the accuracy concept all together ask questions that allow for open ended responses.
+Alternatively visualisation research could shift away from the accuracy concept all together and ask questions that allow for open ended responses.
 This method can enlighten authors as to *how* the uncertainty information was used by the participants.
@@
 While experiments that explicitly identify heuristics in current methods are not technically measuring signal-suppression, they are still a useful consideration to keep in mind when designing experiments for signal suppression.
-Heuristic checks are look at unknown pitfalls that might exist in interpretation of current plots [@Hullman2016] and can change depending on the larger scope of the graphic and the population we are communicating with [@Spiegelhalter2017; @Kinkeldey2014].
+Heuristic checks look at unknown pitfalls that might exist in interpretation of current plots [@Hullman2016] and can change depending on the larger scope of the graphic and the population we are communicating with [@Spiegelhalter2017; @Kinkeldey2014].
 

### 2024-09-28 paper.qmd
made first half of sarahs small edits

@@
 
-From entertainment choices to news articles to insurance plans, the modern citizen is so over-run with information in every aspect of their life it can be overwhelming. 
+From entertainment choices to news articles to insurance plans, the modern citizen is so inundated with information in every aspect of their life it can be overwhelming. 
 In the face of this overflow of information, tools that effectively reduce piles of information to simple and clear ideas become more valuable.
@@
 Among these summary tools, information visualisations are one of the most powerful. 
-Visualisations allow us to *learn* what is important about our data set and datasets such as Anscombe's quartet [@anscombe] or the Datasaurus Dozen [@datasaurpkg] highlight this power. 
+Visualisations allow us to *learn* what is important about our data and datasets such as Anscombe's quartet [@anscombe] or the Datasaurus Dozen [@datasaurpkg] highlight this power. 
 Additionally, visualisations allow for efficient and memorable communication. 
@@
 
-Reviews on uncertainty visualisation rarely offer tried and tested rules for effective uncertainty visualisation, but rather, they comment on the *difficulties* faced when trying to summarize the field.
+Reviews on uncertainty visualisation rarely offer tried and tested rules for effective uncertainty visualisation, but rather, they comment on the *difficulties* faced when trying to summarise the field.
 @Kinkeldey2014 found most experimental methods to be ad hoc, with no commonly agreed upon methodology, formalisations, or greater goal of describing general principals. 
@@
 She commented on the prevalence of confounding variables that make it unclear as to what exactly caused a subjects poor performance on a set of particular questions.
-Mistakes due to misunderstanding a visualisations, misinterpreted questions, and incorrectly applied heuristics are all combined into a single error value.
+Mistakes due to misunderstanding visualisations, misinterpreting questions, and incorrectly applying heuristics are all combined into a single error value.
 @Spiegelhalter2017 commented that different plots are good for different things, and disagreed with the goal of identifying a universal "best" plot for all people and circumstances.
-@Griethe2006 was unable to find common themes, but instead listed the findings and opinions of a collection of papers. 
-@uncertchap2022 summarized several cognitive effects that repeatedly arise in uncertainty visualisation experiments, however these effects were each discussed in isolation as a list of considerations an author might make rather than an overarching theory of rules for effective uncertainty visualisation.
+@Griethe2006 did not identify common themes, but instead listed the findings and opinions of a collection of papers. 
+@uncertchap2022 summarised several cognitive effects that repeatedly arise in uncertainty visualisation experiments, however these effects were each discussed in isolation as a list of considerations an author might make rather than an overarching theory of rules for effective uncertainty visualisation.
 While these reviews are thorough in scope, none discuss how the existing literature contribute to the borader goal of uncertainty visualisation. 
-The problem faced by the literature is easily summarized with a famous quote by Henri Poincaré.
+The problem faced by the literature is easily summarised with a famous quote by Henri Poincaré.
 
@@
 @fig-data shows the first six rows and the geographical boundaries of our data set. 
-The temperature variable was generated using a sine wave, that is $Temperature_i = 29-3\cdot{|Latitude_i - \sin{2 \cdot Longitude_i}|}$ where the $Longitide$ and $Latitude$ are the longitude and latitude of the county's centroid scaled to a standard normal distribution.
+The temperature variable was generated using a sine wave, that is $Temperature_i = 29-3\cdot{|Latitude_i - \sin{2 \cdot Longitude_i}|}$ where the $Longitude$ and $Latitude$ are the longitude and latitude of the county's centroid scaled to a standard normal distribution.
 Each county's variance is independently, randomly sampled from a uniform distribution. 
-The low standard errors are drawn from a $U_{[0,1]}$ distribution, while the high standard errors are drawn from a a $U_{[1,2]}$ distribution. 
+The low standard errors are drawn from a $U_{[0,1]}$ distribution, while the high standard errors are drawn from a $U_{[1,2]}$ distribution. 
 As we are dealing with an average, the sampling distribution would be approximately normal by the central limit theorem, so each county temperature estimate is assumed to come from a $N(Temp_i, SE_{case,i})$.
@@
 #| label: fig-data
-#| fig-cap: "The the first 5 observations of the data used for the spatial uncertainty examples along with the boundaries of each county.
+#| fig-cap: "The first 5 observations of the data used for the spatial uncertainty examples along with the boundaries of each county.
 The map boundaries are the Iowa county boundaries, however the 'temperature' data is not representative of the average temperature in Iowa.
@@
 @fig-choropleth depicts a choropleth map of the counties of Iowa. 
-Each of these counties are colored according to an estimate of average daily temperature that was generated so that the values followed a clear spatial trend (hot in the middle of the map, and cold on the outside).
+Each of these counties are coloured according to an estimate of average daily temperature that was generated so that the values followed a clear spatial trend (hot in the middle of the map, and cold on the outside).
 The variance of these estimates were simulated such that the trend accounts for most of the variance in the plot in the low variance case (so we should expect the trend to be visible) while the high variance case has significantly more noise, so much so that it overwhelms the spatial trend.
@@
 This consensus leads us to understand that uncertainty visualisation is motivated by the need for a sort of "visual hypothesis test". 
-A successful uncertainty visualisation would act as a "statistical hedge" for any inference me make using the graphic. 
+A successful uncertainty visualisation would act as a "statistical hedge" for any inference we make using the graphic. 
 Additionally, since the purpose of a visualisation is to give a quick "gist" of the information [@Spiegelhalter2017], this hedging needs to be communicated visually without the need for complicated calculations.
-If we refer to the conclusion we draw from a graphic to be its "signal" and the variance that makes this signal harder to identify as the "noise", we can summaries the above information into three key requirements.
+If we refer to the conclusion we draw from a graphic to be its "signal" and the variance that makes this signal harder to identify as the "noise", we can summarise the above information into three key requirements.
 A good uncertainty visualisation needs to:
@@
 ## Uncertainty as a signal
-Uncertainty visualisation is not only motivated by signal-suppression, and we would be remiss not to mention these alternative approaches. 
+Uncertainty visualisation is not only motivated by signal-suppression, and we would be remiss if we did not mention these alternative approaches. 
 Some authors claim the purpose of uncertainty is to improve decision making [@Ibrekk1987; @uncertchap2022; @Hullman2016; @Cheong2016; @Boone2018; @Padilla2017]. 
-Other authors do not describe uncertainty as important for decision making, but rather explicitly state it as variable of importance in of itself [@Blenkinsop2000]. 
-While uncertainty can provide useful information in decision making, it is important to recognize that the "uncertainty" in these cases is not acting as "uncertainty" at all. 
+Other authors do not describe uncertainty as important for decision making, but rather explicitly state it as a variable of importance in of itself [@Blenkinsop2000]. 
+While uncertainty can provide useful information in decision making, it is important to recognise that the "uncertainty" in these cases is not acting as "uncertainty" at all. 
 It is acting as signal.
@@
 This is precisely what we observe in the literature. 
-@fig-exceed depicts an exceedance probability map that has was designed as an alternative to the choropleth map to improve decision making under uncertainty [@Kuhnert2018; @Lucchesi2021]. 
+@fig-exceed depicts an exceedance probability map that was designed as an alternative to the choropleth map to improve decision making under uncertainty [@Kuhnert2018; @Lucchesi2021]. 
 A keen viewer may notice that the "exceedance probability map" is actually just a choropleth map, only the statistic being displayed has changed. 
@@
 
-Of course, there is nothing wrong with explicitly visualizing variance, error, bias, or any other statistic used to depict uncertainty as a signal. 
+Of course, there is nothing wrong with explicitly visualising variance, error, bias, or any other statistic used to depict uncertainty as a signal. 
 Just like any other statistic, these metrics provide important and useful information for analysis and decisions. 
-However, there is no interesting visualisation challenge associated with these graphics, and they do not require any special visualisation. 
+However, there is no interesting visualisation challenge associated with these graphics, and they do not require any special visualisation techniques. 
 The uncertainty in these graphics are acting as a signal variable, and they should be treated as such.
@@
 ## The bivariate map
-@fig-bivariate depicts a variation of the choropleth map, where we have a two dimensional color palette. 
+@fig-bivariate depicts a variation of the choropleth map, where we have a two dimensional colour palette. 
 In this graphic, temperature is still mapped to hue, but the variance is included by utilising colour saturation. 
@@
 @fig-bivariate makes it clear that simply including uncertainty as a variable is insufficient to perform signal-suppresion. 
-If we cannot treat uncertainty the same as we would any other variable, how should we treat it as? 
+If we cannot treat uncertainty the same as we would any other variable, how should we treat it? 
 We need to understand what uncertainty actually *is*, in order to understand how to integrate it into a visualisation.
 
-### It's a variable... it's a metadata... it's uncertainty?
+### It's a variable... it's metadata... it's uncertainty?
 Describing what uncertainty actually is, is surprisingly hard. 
@@
 
-Descriptive statistics describe our sample as it is and summarizes large data down into an easy to swallow format. 
+Descriptive statistics describe our sample as it is and summarises large data down into an easy to swallow format. 
 Descriptive statistics are not seen as the primary goal of modern statistics, however, this was not always the case. 
@@
 Mapping uncertainty and signal to separable channels allows them to be read separately, which does not align with the goal of communicating them as a "single integrated channel". 
-Additionally, visualizing uncertainty and signal separately allows the uncertainty information to simply be ignored, which is a pervasive issue in current uncertainty visualisation methods [@uncertchap2022].
+Additionally, visualising uncertainty and signal separately allows the uncertainty information to simply be ignored, which is a pervasive issue in current uncertainty visualisation methods [@uncertchap2022].
 We can see this problem in @fig-bivariate, as it sends the message "this data has a spatial trend and the estimates have a large variance" as we read the signal and the uncertainty separately.
@@
 That is, the visual channels of the uncertainty and the signal would need to be separately manipulable, but read as a single channel by the human brain. 
-While most visual aesthetics *are* separable, there are some variables that have been shown to be integrable, such as color hue and brightness [@Vanderplas2020].
-When visualizing uncertainty using its own visual channel, we can also consider visual semiotics and make sure to map uncertainty to intuitive visual channels, such as mapping more uncertain values to lighter colors [@Maceachren2012].
+While most visual aesthetics *are* separable, there are some variables that have been shown to be integrable, such as colour hue and brightness [@Vanderplas2020].
+When visualising uncertainty using its own visual channel, we can also consider visual semiotics and make sure to map uncertainty to intuitive visual channels, such as mapping more uncertain values to lighter colours [@Maceachren2012].
 Unfortunately relying on integrability may not give us the amount of control we want over our signal-supression. 
@@
 There is another benefit to mapping uncertainty to saturation that is not directly related to integrability. 
-Colour value has a second desirable quality as the colours become harder to distinguish as the saturation decreases. 
+As the saturation decreases colours become harder to distinguish. 
 This means high uncertainty values are harder to differentiate than low uncertainty values. 
@@
 
-@fig-vsup is a visualisation of the Iowa temperature data using a VSUP to color the counties. 
+@fig-vsup is a visualisation of the Iowa temperature data using a VSUP to colour the counties. 
 The low uncertainty case still has a visible spatial trend, while the spatial trend in the high uncertainty map has functionally disappeared. 
@@
 
-Some authors recognize inference will always occur (in some shape or form) and believe uncertainty *should* be visualised but do not recognize *how* uncertainty would be visualised. 
+Some authors recognise inference will always occur (in some shape or form) and believe uncertainty *should* be visualised but do not recognise *how* uncertainty would be visualised. 
 @Hullman2021 argued that there is no such thing as a "model-free" visualisation, therefore all visualisations require uncertainty as we are always performing inference. 
@@
 
-Let's say we have a graphic that depicts a set of coefficients from a linear regression and the value of the coefficient is shown using a single color. 
+Let's say we have a graphic that depicts a set of coefficients from a linear regression and the value of the coefficient is shown using a single colour. 
 We want to know "Which of these coefficients are different from 0?" as well as "Which of these coefficients are different from each other?". 
@@
 Next, we conduct two sample $t$-test and find that several of the values need to be visually distinguishable from each other. 
-The VSUP method must pick a single color for each estimate, and these colors must be *either* visually distinguishable or indistinguishable from each other. 
+The VSUP method must pick a single colour for each estimate, and these colours must be *either* visually distinguishable or indistinguishable from each other. 
 We cannot perform signal-suppression on both these signals simultaneously.
@@
 This example highlights a fundamental problem with the VSUP that extends to the bivariate map as well. 
-When we blend these colors, we need to decide at what level of *uncertainty* to blend these colors together. 
-Even though the bivariate map does not explicitly combine color values at certain variance levels, the mapping of variance to color saturation does this implicitly. 
-That is, at certain saturation values the colors in a bivariate map are imperceptibly different and appear as though they are mapped to the same value.
-At this point, it is irrelevant whether or not the colours are technically different, they are the same color in the human brain. 
+When we blend these colours, we need to decide at what level of *uncertainty* to blend these colours together. 
+Even though the bivariate map does not explicitly combine colour values at certain variance levels, the mapping of variance to colour saturation does this implicitly. 
+That is, at certain saturation values the colours in a bivariate map are imperceptibly different and appear as though they are mapped to the same value.
+At this point, it is irrelevant whether or not the colours are technically different, they are the same colour in the human brain. 
 This is of course complicated by the fact that human colour perception varies at an individual level. 
-Some women are believed to have four different cone cells which allows them to perceive a greater range of colors, while others only have one or two types of cone cells and have colour deficiencies [@simunovic2010]
+Some women are believed to have four different cone cells which allows them to perceive a greater range of colours, while others only have one or two types of cone cells and have colour deficiencies [@simunovic2010]
 For VSUP to function for all individuals, we must calibrate each plot to an individual's ability to perceive colour.
@@
 This issue has already been raised in the literature. 
-Which hypothesis are suppressed and which are not largely depends on the method used to combining colors in the palette [@Kay2019]. 
+Which hypothesis are suppressed and which are not largely depends on the method used to combining colours in the palette [@Kay2019]. 
 The VSUP in @fig-vsup used the tree based method that was used by @Correll2018, but there are alternatives that are more appropriate for different hypothesis.
@@
 This assumption is commonly made by visualisation researchers in normal visualisation experiments [@North2006]. 
-If we could express the statistic of a cell using multiple colors, this limitation may disappear entirely.
+If we could express the statistic of a cell using multiple colours, this limitation may disappear entirely.
 
 # Implicitly Combining Uncertainty and Signal
-Rather than trying to figure out how to combine signal and uncertainty into a single color, we can just display a sample instead and allow the viewer to extract *both* the estimate and the variance.
+Rather than trying to figure out how to combine signal and uncertainty into a single colour, we can just display a sample instead and allow the viewer to extract *both* the estimate and the variance.
 
 ## Pixel map
-@fig-pixel displays a pixel map [@Lucchesi2021], which is a variation of the choropleth map where each area is divided up into several smaller areas, each colored using outcomes from the larger area's (i.e.
+@fig-pixel displays a pixel map [@Lucchesi2021], which is a variation of the choropleth map where each area is divided up into several smaller areas, each coloured using outcomes from the larger area's (i.e.
 the county's) average temperature sampling distribution.
@@
 Not only does this method help readers understand the plot level "gist", it is also unlikely to damage the viewers ability to extract individual estimates.
-Extracting global statistics, such as the mean or variance, from a sample can be done with relative ease, especially when those values are mapped to color [@Franconeri2021]. 
+Extracting global statistics, such as the mean or variance, from a sample can be done with relative ease, especially when those values are mapped to colour [@Franconeri2021]. 
 Additionally, @Ndlovu2023 found that participants applied the same methods they used for simple choropleth maps to complicated uncertainty maps even if that take away was invalid. 
@@
 Quantifiable uncertainties are just assumption violations expressed as an effect on our final statistic (or true variation of our data). 
-The disconnect between quantifiable and unqualitifable uncertainty creates a huge problem for authors trying to visualize it. 
+The disconnect between quantifiable and unqualitifable uncertainty creates a huge problem for authors trying to visualise it. 
 A survey of visualisation authors cited "not knowing how to calculate uncertainty" as one of the primary reasons they did not include it in visualisations [@Hullman2020a].
@@
 
-This does not mean that visualizing raw data will always prevent insignificant signal from getting through. 
+This does not mean that visualising raw data will always prevent insignificant signal from getting through. 
 @Chowdhury illustrated how groups that appear linearly separable in a linear discriminant analysis (LDA) visualisation of the data can actually be the result of a LDA performed on too many variables, something that was not clear from the visualisation until the line-up protocol was implemented.
@@
 In another example @Ibrekk1987 asked participants for the "best estimate", but the participant's responses were evaluated against the mean of the distribution.
-Ultimately, the term "best" is up to the users interpretation, and the estimate that minimized the sum of squared errors was not implied by the question. 
+Ultimately, the term "best" is up to the users interpretation, and the estimate that minimised the sum of squared errors was not implied by the question. 
 This vague question approach leads to inconclusive results, as we are left unclear if it was the phrasing of the question or the plot design that caused the participants to answer incorrectly.
@@
 
-# Future work
+# Conclusions and Future work
 This paper has identified gaps in the uncertainty visualisation literature that must be filled for the field to progress. 
-We formalized the uncertainty visualisation problem, and in doing so highlighted an untouched area of uncertainty visualisation research.
+We formalised the uncertainty visualisation problem, and in doing so highlighted an untouched area of uncertainty visualisation research.
 

### 2024-10-01 paper.qmd
fixed more -ize

@@
 
-When we use "uncertainty visualization" to refer to graphics that simply communicate a variance or probability, we are classifying visualisations by the data they display, not their visual features. 
+When we use "uncertainty visualation" to refer to graphics that simply communicate a variance or probability, we are classifying visualisations by the data they display, not their visual features. 
 Graphics, just like statistics, are not defined by their input data. 
@@
 This means uncertainty should not be visualised by itself and we should seek to display signal and uncertainty together as a "single integrated uncertain value" [@Kinkeldey2014]. 
-This aspect of uncertainty visualization makes it a uniquely difficult problem.
+This aspect of uncertainty visualation makes it a uniquely difficult problem.
 
@@
 @Hullman2021 argued that there is no such thing as a "model-free" visualisation, therefore all visualisations require uncertainty as we are always performing inference. 
-While it is true that we can think of visualisations as containing implicit inferential properties, there are many potential inferences in any single visualization. 
+While it is true that we can think of visualisations as containing implicit inferential properties, there are many potential inferences in any single visualisation. 
 This makes it a little difficult to ensure uncertainty is always included.
@@
 
-Risk-aversion is another secondary effect of uncertainty visualization that is used to evaluate uncertainty visualisations [@Hullman2019]. 
+Risk-aversion is another secondary effect of uncertainty visualisation that is used to evaluate uncertainty visualisations [@Hullman2019]. 
 Risk aversion is an economics term used to describe an agent who would chose a random variable with a lower expected payout because it also has a lower variance. 
 Risk aversion is considered irrational behaviour because it is a deviation from the behaviour of a rational agent. 
-Comparing participant responses to that of a rational agent has even been suggested as a benchmark for uncertainty visualization experiments [@wu2023rational].
+Comparing participant responses to that of a rational agent has even been suggested as a benchmark for uncertainty visualisation experiments [@wu2023rational].
 However, just like trust, evaluating graphics on risk-aversion discourages signal-suppression. 
@@
 The assumptions required to answer these deterministic questions are often invisible, even to the authors of the paper, which means they often create more problems than they solve when they are used as a tool for graphical evaluation.
-Similar to the vague question problem, these studies seem to be aware that we should be evaluating an uncertainty visualization based on a deterministic observation (such as our example of "does this map have a spatial trend") but are unsure how to incorporate or evaluate it.
+Similar to the vague question problem, these studies seem to be aware that we should be evaluating an uncertainty visualation based on a deterministic observation (such as our example of "does this map have a spatial trend") but are unsure how to incorporate or evaluate it.
 
@@
 ### Comparing to hypothesis tests
-The most obvious way to evaluate uncertainty visualizations is to compare the visualisations to statistical tests. 
+The most obvious way to evaluate uncertainty visualations is to compare the visualisations to statistical tests. 
 If a graphic was performing signal-suppression we would expect the signal to be harder to see at higher levels of uncertainty. 

### 2024-10-01 paper.qmd
implemented sarah's small edits

@@
 #| label: fig-ibrekk
-#| fig-cap: "A replication of the the uncertainty visualisations shown by @Ibrekk1987 in one of the earliest uncertainty visualisation experiments.
+#| fig-cap: "A replication of the uncertainty visualisations shown by @Ibrekk1987 in one of the earliest uncertainty visualisation experiments.
 Several visualisation methods that are now unpopular (such as the pie chart) are used throughout this paper."
@@
 Given that there is no special class of visualisation for *other* statistics (such the median or maximum) there is no reason to assume visualisations that simply depict a variance, error, or probability to be special.
-Some authors implicitly suggest that that visualisations of variance or probability are differentiated due to the psychological heuristics involved in interpreting uncertainty [@Hullman2019].
+Some authors implicitly suggest that visualisations of variance or probability are differentiated due to the psychological heuristics involved in interpreting uncertainty [@Hullman2019].
 While it is true that heuristics lead people to avoid uncertainty [@Spiegelhalter2017], there is no evidence that this psychological effect translates to issues with the visual representation of uncertainty.
@@
 #| fig-cap: "A pixel map of the counties of Iowa.
-In this map, each county is broken up into several smaller regions and coloured according to a potential daily temperature, given the its average daily temperature and its sampling distribution.
+In this map, each county is broken up into several small areas and coloured according to a potential daily temperature, given the its average daily temperature and its sampling distribution.
 This results in each county being represented by a sample rather than a single value.
@@
 The pixel map is not the best uncertainty visualisation, but it does share a lot in common with the best uncertainty visualisation. 
-The best uncertainty visualisation is the visualisation that best captures the limitations of of our raw data.  
+The best uncertainty visualisation is the visualisation that best captures the limitations of our raw data.  
 As we discussed in previous sections, we can consider uncertainty to be "the amalgamations of the impacts of violations to the assumption of the uniformity of nature". 
@@
 @Otsuka2023 suggested that the process of observing data to perform statistics is largely dependent on our goals, because the process of boiling real world entities down into probabilistic objects (or "probabilistic kind" as he puts it) depends on the relationship we seek to identify with our data.
-@Meng2014 commented what is kept as data and what is tossed away is determined by the motivation of an analysis and what was previously noise can be shown to become signal depending on the the question we seek to answer.
+@Meng2014 commented what is kept as data and what is tossed away is determined by the motivation of an analysis and what was previously noise can be shown to become signal depending on the question we seek to answer.
 @Wallsten1997 argue that the best method for evaluating or combining subjective probabilities depends on the uncertainty the decision maker wants to represent and why it matters. 
@@
 
-This makes it very difficult to move quantified uncertainty through the the layers of our analysis, especially when designing a visualisation for EDA.
+This makes it very difficult to move quantified uncertainty through the layers of our analysis, especially when designing a visualisation for EDA.
 If we don't know what the final statistic is, we cannot quantify the effects of our assumptions.
@@
 @Chowdhury illustrated how groups that appear linearly separable in a linear discriminant analysis (LDA) visualisation of the data can actually be the result of a LDA performed on too many variables, something that was not clear from the visualisation until the line-up protocol was implemented.
-However, just showing the data is a simple but effective option for uncertainty visualisation that that seems to be largely overlooked. 
+However, just showing the data is a simple but effective option for uncertainty visualisation that seems to be largely overlooked. 
 <!-- Does showing the data actually show uncertainty, or does it just show variance?
@@
 <!-- Yes, this looks like you're thinking along the same lines as I was -->
-While the pixel map does not literally show the data, it does imitate showing the data and give us details on where the variability is coming from. 
-In order to understand *why* pixel maps seem to be successful at signal suppression at first glance would likely require a more thorough discussion than what what we can offer here.
+While the pixel map does not literally show the data, it does imitate showing the data and gives us details on where the variability is coming from. 
+In order to understand *why* pixel maps seem to be successful at signal suppression at first glance would likely require a more thorough discussion than what we can offer here.
 
@@
 
-### Questions that attempts to capture signal-suppression
+### Questions that attempt to capture signal-suppression
 There is a collection of studies that seem to be aware of the issues behind using trust or value extraction to evaluate uncertainty visualisations. 
@@
 ## Testing signal supression
-If we cannot ask direct questions about uncertainty, we cannot measure the secondary effects of uncertainty, and we cannot ask indirect questions about the uncertainty, how are we supposed to evaluate uncertainty visualisations?
+If we cannot ask direct questions about uncertainty, or measure the secondary effects of uncertainty, or ask indirect questions about the uncertainty; then how are we supposed to evaluate uncertainty visualisations?
 How do you measure something that disappears the second you look directly at it?

### 2024-10-01 paper.qmd
went back and edited for susan comment 7

@@
 Therefore, the pixel map performs signal suppression, without sacrificing the viewers ability to extract general statistics, unless those statistics *should* be harder to extract due to the uncertainty in the value.
-So is this the best uncertainty visualisation? 
-And if so, why?
+So, what is it about @fig-pixel that allows it to be such a successful uncertainty visualisation?
+Is this a question we can easily answer?
 
 ## Show me the data
-The pixel map is not the best uncertainty visualisation, but it does share a lot in common with the best uncertainty visualisation. 
-The best uncertainty visualisation is the visualisation that best captures the limitations of our raw data.  
+To thoroughly understand why pixel maps seem to successfully perform signal suppression, we would need a more thorough discussion than what can be offered here. 
+However, this does not prohibit us from discussing potential reasons for the graph's effectiveness. 
+There is a second type of uncertainty that the pixel map quietly communicates but we have not touched on yet, and this uncertainty is also conveyed by visualisations of our raw data. 
+That is, these graphics can show assumption violations.
+
 As we discussed in previous sections, we can consider uncertainty to be "the amalgamations of the impacts of violations to the assumption of the uniformity of nature". 
 It's not a definition that rolls off the tongue, but we can work with it.
-
 Thankfully, this definition of uncertainty aligns nicely with all the concepts that are included in the uncertainty umbrella. 
@@
 If we don't know what the final statistic is, we cannot quantify the effects of our assumptions. 
-Therefore, often the best uncertainty visualisation is not an "uncertainty visualisation" at all, but simply the most accurate depiction of our raw data as it gives us a good idea of it's limitations.
-
+Therefore, it is often the case that the best uncertainty visualisation is not an "uncertainty visualisation" at all, but simply the most accurate depiction of our raw data.
 This does not mean that visualising raw data will always prevent insignificant signal from getting through. 
@@
 However, just showing the data is a simple but effective option for uncertainty visualisation that seems to be largely overlooked.
-<!-- Does showing the data actually show uncertainty, or does it just show variance?
-I'm thinking of inferential statistics here - you can show a sample, but it doesn't necessarily show you anything about the uncertainty of the population-based inferential statistics, visually speaking.
-Some of the dotmap style visualizations you're thinking of are actually census visualizations - there isn't any inferential uncertainty, but there is variance.
-I'm not sure you're adequately separating these concepts here, even if I generally agree with the broad points you're making.
--->
-<!--    
-- I think the point I’m trying to make here is not that the raw data shows uncertainty by showing variation, but rather that it shows uncertainty by allowing people to see violations to the assumptions they might have made when mentally calculating an inferrential statistic
-  - e.g. if i show the raw data and the areas are unevenly spaced or there are a lot of missing observations in some areas.
-- I agree that the variability in the raw data is not related to the variability in the inferential statistic even though this section certainly implies that. I need to fix that.
--->
-<!-- Ok, but variability in the raw data is also related to variability in the inferential statistic, it's just that as n -> infinity, the variability in the inferential statistic goes to 0.  -->
-<!--
-- The connection with the pixel map was supposed to be that it would theoretically give a sense of the shape of the distribution.
-When I said “where the variability is coming from” I was trying to refer to that shape, but that doesn’t really translate from what I’ve written and the connecton between the pixel map and showing the raw data isn’t really there.
-  - I think the logic line I was going for was
-    Pixel map shows collection of hypothetical outcomes of sampling distribution 
-    => These outcomes let us see the shape of the distribution 
-    => When seeing the shape of the distribution, we may become aware of errors in our assumptions, (e.g. such as normality) 
-    => If we want to be aware of where our uncertainty is coming from (e.g. underlying variability or assumption violations) then the raw data is what best shows this
-- I did consider including a visualisation that shows the data I used for the pixel map as a final map visualisation to better separate these concepts, but I didn’t technically use any real data (since I just made the normality assumption which is excatly what this map should highlight)
-  - It might be better to change the statistic and give the sampling distribution in each area a different shape to make the pixel map benefit clearer.
-  - I just want to make sure this is an accurate description of the issue you were highlighting before I fix it up this section (also I think it will take me a bit to figure out how to fix it) 
--->
-<!-- Yes, this looks like you're thinking along the same lines as I was -->
-While the pixel map does not literally show the data, it does imitate showing the data and gives us details on where the variability is coming from. 
-In order to understand *why* pixel maps seem to be successful at signal suppression at first glance would likely require a more thorough discussion than what we can offer here.
+
+Of course, visualising the raw data is not always possible nor suitable.
+For example, if we have a specific statistic in mind, it may be more appropriate to show the sampling distribution of that statistic rather than the variability of all observations.
+If we visualise that sampling distribution in a way that allows us to see it's shape, and therefore recognise any assumption violations (such as a non-normal distribution), we maintain some of the benefits that came with visualising the raw data.
+This is exactly what the pixel map does, and likely contributes to it's success as a signal suppression method.
 

### 2024-10-01 paper.qmd
Just left section from comment 9 as it was

@@
 The mean of the two groups was used to decide which group was actually "furthest to the right" and the different graphical approaches were evaluated against that ground truth. 
-<!-- Participants weren't evaluated against that ground truth, though - the plots were what was actually being evaluated in that study, not participant accuracy.
-Lineup prompts are typically vague (sometimes, I agree, too vauge) because the participant has to figure out what metric to use to evaluate the data first, and then complete the evaluation task.
-We're as interested in the metric they're using as we are in their successful identification of the target plot.
--->
-<!-- Good point about the plots being evaluated, not the participants. I changed that.
-I agree with the point about lineups being useful in identifying how participants complete evaluation tasks, and I did comment on that with respect to this particular paper later in the “Qualitative studies” section with this comment:
-   “@Hofmann2012 tried to capture this by asking participants why they considered a particular plot to be more "right shifted", even though this qualitative analysis did not make it into the final paper.
-“
-Do you think I should move that comment to this section to keep all comments on that paper together, leave it where it is, or do you think there should be a different edit? -->
 In another example @Ibrekk1987 asked participants for the "best estimate", but the participant's responses were evaluated against the mean of the distribution.
@@
 This method can enlighten authors as to *how* the uncertainty information was used by the participants.
-@Hofmann2012 tried to capture this by asking participants why they considered a particular plot to be more "right shifted", even though this qualitative analysis did not make it into the final paper. 
+@Hofmann2012 captured this by asking participants why they considered a particular plot to be more "right shifted", even though this qualitative analysis did not make it into the final paper. 
 @Daradkeh2015 presented participants with ten investment alternatives and asked participants "from among available alternatives, which alternative do you prefer the most", and were asked to think aloud and consider the uncertainty in their decision making.

### 2024-10-01 paper.qmd
made edits for susan comment 8

@@
 Science communication should be primarily concerned with accuracy. 
-<!-- Ok, but you can be completely accurate in how you render a plot and still convey information that isn't accurate (e.g.
-sine illusion/line width issues).
-Do we care about accuracy of the visual form, or accuracy of the perception?
-Because accuracy of the perception is potentially influenced by trust/transparency and potentially influences that metric as well.
-Rachel Rogers and I have studied how much weight people give testimony in a courtroom - it strikes me that trust/transparency are a separate dimension that is important and related to accuracy, in that both accuracy and trust/transparency will influence how much mental weight I ascribe to new information gained from a visualization.
--->
-<!--
-I see there are three “directions of influence” we can use to think about the relationship between trust and accuracy:
-  1) Accuracy impact trustworthiness:
-     - This is basically the main consideration of this point of this section.
-The point is that historical accuracy impacts “trustworthiness” which then impacts trust.
-Since trust is impacted by the trustworthiness of the source, so you should endeavor to be as accurate as possible to establish trustworthiness if you want your information to be trustworthy.
-  2) Trustworthiness impacts accuracy:
-    - I agree that the accuracy can be influenced by trust.
-We had a climate scientist discussing this exact thing at last week’s numbat seminar and had a good citation for this.
-I could probably go find but it used political beliefs as a proxy rather than “trust” so it might not be exactly right but it is thinking along these lines.
-  3) Trustworthiness and accuracy combined impact trust (or belief or something):
-    - The final statistic of “trust” is a combination of sourth trustworthiness and accuracy.
-Since trust can be impacted by accuracy or trustworthiness, pursuing trust rather than trustworthiness can encourage inaccuracy.
-I think that commenting on (2) is a good addition to this section, but it doesn’t really change my opinion on (1) and (3) (I’m not 100% sure if this comment was an “in addition” or “in dispute” comment and I want to check before I change it)
-
-I agree with your point about the accuracy in display vs perception but I don’t see its connection to the greater point about trust and I’m not 100% sure it fits in this section.
-Accuracy in the visual form (e.g.
-correcting for sine illusion) is still a pursuit of accuracy, just accuracy in the visual statistic.
-Like, if we use Frequentist or Bayesian statistics, they also have different inputs, processes and outputs and they both give us different estimates.
-It is also common to adjust those statistics to correct for a bias in the calculation (e.g.
-we might weight the observations).
-I don’t really see the difference between these methods and editing a graphic to correct for the sine illusion.
-I don’t think any of these things conflict with the pursuit of scientific accuracy.
-I’m not against adding this as an additional comment (i.e.
-that accuracy in a visualisation may not be what the reader thinks it is) but I’m not certain it has any serious implications for the trust
--->
-<!-- That's totally fair :) Just quibbling - not something you have to change unless someone else calls it out! -->
+This accuracy may not always involve showing the data exactly as it is, for example we may want to adjust our graphic to correct for visual heuristics, but these changes should still be in pursuit of accuracy.
 Setting trust as the variable of interest implicitly encourages statisticians to set trust as the primary goal of communication. 
@@
 This is the end result of designing uncertainty visualisations for increased trust, rather than signal-suppression.
+Of course, this does not mean measuring trust or perceived trustworthiness has no place in visualisation experiments.
+After all, if participants political motivations can influence their ability to accurately read a plot [@nurse2020], then their level of trust in the information could have a similar effect.
+However, it does mean trust should not be the primary goal of uncertainty visualisation evaluation experiments.
 

### 2024-10-01 paper.qmd
fix for susan comment 12

@@
 However, this finding does not directly translate to the lineup protocol being worse than its computational equivalent.
-The lineup protocol allows viewers to detect "something is not right" even if that something is unspecified, and even if that "something" is not what see the authors intended (for an example see @vanderplas2017).
+The lineup protocol allows viewers to detect "something is not right" even if that something is unspecified, and even if that "something" is not what see the authors intended.
+For example, @vanderplas2017 did an experiment where participants were drawn to plots with unequal cluster sizes, even though the size of the cluster was not one of the features the authors were testing.
 While standard hypothesis tests may outperform the lineup protocol on power, computational tests have the benefit of knowing exactly what they are looking for, while visualisations are often explicitly made when we *don't* know what we are looking for.

### 2024-10-04 paper.qmd
updated based on sarah comments

@@
          bivar = cut(variance, breaks=0:4, labels=seq(4)),
-         biclass = paste(bitemp, bivar, sep="-"))
+         biclass = paste(bitemp, bivar, sep="-"))|>
+  mutate(highlight = ifelse(count_id <= 5, TRUE, FALSE))
 
@@
                     lowvar = "Low") |>
+  add_footer_row(values = rep("..."), colwidths = 5) |>
   theme_vanilla() |>
@@
   ggplot() +
-  geom_sf(aes(geometry = geometry), fill="white") + 
-  geom_text(aes(x=cent_long, y=cent_lat, label=count_id), size=2) +
-  theme_void()
+  geom_sf(aes(geometry = geometry, fill=highlight)) + 
+  scale_fill_manual(values=c("white", "#fbfba2")) +
+  geom_text(data=filter(my_map_data, highlight==TRUE), aes(x=cent_long, y=cent_lat, label=count_id), size=3) +
+  theme_void() +
+  theme(legend.position = 'none')
 
@@
 In the low variance condition, the trend accounts for most of the variation in the data, in the high variance case, the variance on the temperature estimate accounts for most of the variance.
+This distinction is not clear in the map as they both appear identical.
 The high variance condition displays a spatial trend that could simply be spurious, which means the plot is displaying a false conclusion."
@@
 
-show_pal <- function (colours, borders = NULL, cex_label = 1, ncol = NULL, myxlab, breaks, textnudge, xlabx, xlaby) {
+show_pal <- function (colours, borders = NULL, cex_label = 1, ncol = NULL, myxlab, breaks, textnudge, xlabx, xlaby, tsize=1.2) {
   # Set dimensions of palette
@@
               cex = 1, col = "black")
-  text(xlabx, xlaby, myxlab ,cex = 1.5, col = "black")
+  text(xlabx, xlaby, myxlab ,cex = tsize, col = "black")
 }
@@
 p1b
-show_pal(basecols, ncol=8, borders=NA, myxlab = "Temperature", breaks = 21:29, textnudge = c(0.2, 0.1, 0,0,0,0,0,-0.1,-0.2), xlabx= 4, xlaby=-1.75)
+show_pal(basecols, ncol=8, borders=NA, myxlab = "Temperature", breaks = 21:29, textnudge = c(0.2, 0.1, 0,0,0,0,0,-0.1,-0.2), xlabx= 4, xlaby=-1.75, tsize=1.5)
 
@@
 #|   - "High Variance Data"
-#|   - "Exceedance Probability Palette"
+#|   - "Exceedance Probability Map Palette"
 #| layout-ncol: 3
-#| layout-valign: "bottom"
 
@@
 p2b
-show_pal(basecols, ncol=8, borders=NA, myxlab = "Probability Temp>27", breaks = c(0, prob_breaks[2:8], 1), textnudge = c(0.2, 0.1, 0,0,0,0,0,-0.1,-0.2), xlabx= 4, xlaby=-1.75)
+show_pal(basecols, ncol=8, borders=NA, myxlab = "P(Temperature>27)", breaks = c(0, prob_breaks[2:8], 1), textnudge = c(0.2, 0.1, 0,0,0,0,0,-0.1,-0.2), xlabx= 4, xlaby=-1.75, tsize=1.2)
 ```
@@
 
-show_pal2 <- function (colours, borders = NULL, cex_label = 1, ncol = NULL, myxlab, myylab, breaks, breaks2) {
+show_pal2 <- function (colours, borders = NULL, cex_label = 1, ncol = NULL, myxlab, myylab, breaks, breaks2, tsize1=1.2, tsize2=1.2) {
   # Set dimensions of palette
@@
        breaks2, cex = 1, col = "black")
-  text(4, -5.5, myxlab ,cex = 1.5, col = "black")
-  text(x=-1.25,y=-2, myylab, srt=270, cex = 1.5, col = "black")
+  text(4, -5.5, myxlab ,cex = tsize1, col = "black")
+  text(x=-1.25,y=-2, myylab, srt=270, cex = tsize2, col = "black")
 }
@@
 #| fig-cap: "A map made with a VSUP.
-The counties of Iowa are coloured acording to its average daily temperature and the variance in temperature, although the space they have been mapped to is slightly warped.
+The counties of Iowa are coloured acording to its average daily temperature and the variance in temperature.
 Similar to the bivariate map, temperature is mapped to hue while variance is mapped to saturation.
+Unlike the bivariance map, the colour space we are mapping our variables to has been transformed so that high variance estimates are harder to discern from each other.
 This map successfully reduces the visibility of the spatial trend in the high uncertainty case while maintaining the visibility of the spatial trend in the low uncertainty case."
@@
 
+Signal suppression is an untouched area of visualisation research and developing methods for the practice may require us to challenge our entire notion of what makes a good visualisation.
+We are interested to see how this challenge plays out.
+
 # Bibliography

### 2024-10-07 paper.qmd
Fixed vizumap issue

@@
 library(flextable)
-
-# Broom returns an error with Vizumap, run this if can't compile
-# devtools::install_version(package = "broom", version = "1.0.4")
 ```
@@
 
-pix_a <- pixelate(pixel_2a, pixelSize = 3, id = "my_id")
+pix_a <- pixelate(pixel_2a, pixelSize = 70, id = "my_id")
 
@@
 
-pix_b <- pixelate(pixel_2b, pixelSize = 3, id = "my_id")
+pix_b <- pixelate(pixel_2b, pixelSize = 70, id = "my_id")
 

### 2024-10-07 paper.qmd
checked libraries were all doing something

@@
 library(tidyverse)
-library(biscale)
+library(Vizumap)
+library(ggplot2)
 library(RColorBrewer)
@@
 library(flextable)
+library(colorspace)
+library(rgeos)
 ```
@@
 # Picture 1
-p1 <- ib_data %>%
+p1 <- ib_data |>
   summarise(avg = mean(x),
           conf_95a = quantile(x, probs=c(0.025)),
-          conf_95b = quantile(x, probs=c(0.975))) %>%
+          conf_95b = quantile(x, probs=c(0.975))) |>
   ggplot(aes(y="NA")) +
@@
 # Picture 3
-# library(ggpattern)
 p3 <- ib_data |>
@@
 # Picture 8
-p8 <- ib_data %>%
+p8 <- ib_data |>
   reframe(x = quantile(x, probs=c(0.25, 0.50, 0.75)))|>
@@
 my_map_data <- get_urbn_map("counties", sf = TRUE) |>
-  dplyr::filter(state_name=="Iowa")
-centroids <- as_tibble(rgeos::gCentroid(as(my_map_data$geometry, "Spatial"), byid = TRUE))
+  filter(state_name=="Iowa")
+centroids <- as_tibble(gCentroid(as(my_map_data$geometry, "Spatial"), byid = TRUE))
 my_map_data$cent_long <- centroids$x
@@
 # Make palettes
-longpal <- rev(colorspace::sequential_hcl(13, palette = "YlOrRd"))
+longpal <- rev(sequential_hcl(13, palette = "YlOrRd"))
 basecols <- longpal[3:10]
@@
 my_map_data <- my_map_data |>
-  dplyr::mutate(temp = 29 - 3*abs(scale(cent_lat) - sin(2*(scale(cent_long)))[,1]), # trend
+  mutate(temp = 29 - 3*abs(scale(cent_lat) - sin(2*(scale(cent_long)))[,1]), # trend
          highvar = runif(n, min=2, max=4), # high variance
@@
 p1a <- my_map_data |>
-  filter(variance_class=="lowvar") %>%
+  filter(variance_class=="lowvar") |>
   ggplot() +
@@
 p2a <- exeed_data |>
-  filter(variance_class=="lowvar") %>%
+  filter(variance_class=="lowvar") |>
   ggplot() +
@@
 p2a <- my_map_data |>
-  filter(variance_class=="lowvar") %>%
+  filter(variance_class=="lowvar") |>
   ggplot() +
@@
 p3a <- my_map_data |>
-  filter(variance_class=="lowvar") %>%
+  filter(variance_class=="lowvar") |>
   ggplot() +
@@
 # Pixel Map
-library(Vizumap)
 
 # Low variance map
-my_map_data_a <- my_map_data %>%
-  filter(variance_class == "lowvar") %>%
+my_map_data_a <- my_map_data |>
+  filter(variance_class == "lowvar") |>
   mutate(my_id = seq(n),
@@
 
-pixel_1a <- my_map_data_a %>%
-  as.data.frame() %>%
-  select(my_id, bitemp, error) %>%
+pixel_1a <- my_map_data_a |>
+  as.data.frame() |>
+  select(my_id, bitemp, error) |>
   read.uv(estimate="bitemp", error="error")
 
-pixel_2a <- my_map_data_a %>% as("Spatial")
+pixel_2a <- my_map_data_a |> as("Spatial")
 
@@
 # High variance
-my_map_data_b <- my_map_data %>%
-  filter(variance_class == "highvar") %>%
+my_map_data_b <- my_map_data |>
+  filter(variance_class == "highvar") |>
   mutate(my_id = seq(n),
@@
 
-pixel_1b <- my_map_data_b %>%
-  as.data.frame() %>%
-  select(my_id, bitemp, error) %>%
+pixel_1b <- my_map_data_b |>
+  as.data.frame() |>
+  select(my_id, bitemp, error) |>
   read.uv(estimate="bitemp", error="error")
 
-pixel_2b <- my_map_data_b %>% 
+pixel_2b <- my_map_data_b |> 
   as("Spatial")

### 2024-10-08 paper.qmd
added acknowledgement section

@@
 library(Vizumap)
-library(ggplot2)
 library(RColorBrewer)
@@
 
+# Acknowledgements
+This work is supported in part by a scholarship from the the Australian Energy Market Operator. 
+The R packages were used for this work were: `tidyverse` [@tidyverse], `Vizumap` [@Vizumap], `RColorBrewer` [@RColorBrewer], `scales` [@scales], `sf` [@sf], `ggrepel` [@ggrepel], `urbnmapr` [@urbnmapr], `flextable` [@flextable], `colorspace` [@colorspace], and `rgeos` [@rgeos].
+The GitHub repository for this paper can be found at https://github.com/harriet-mason/ARSA-UncertaintyLitReview which contains the files required to reproduce this article in full.
+
 # Bibliography

### 2024-10-08 paper.qmd
finished emi edits

@@
 In the face of this overflow of information, tools that effectively reduce piles of information to simple and clear ideas become more valuable. 
+Even more so if they can this information with a healthy amount of skepticism.
 
@@
 That is to say, despite the wealth of reviews, the field of uncertainty visualisation remains a heap of stones. 
-Lovely stones, but stones nonetheless. 
-This review attempts to address this issue by offering a novel perspective on the uncertainty visualisation problem, and hopefully laying the foundations on which we can build a house.
+There is a mountain of work that identifies common heuristics found in uncertainty visualisations, evaluate competing plot designs, or start a theoretical discussion on a niche aspect of the field.
+This is important work that needs to be done, but each of these papers offers up their own bespoke motivation and methodology, with little reference to the uncertainty visualisation papers outside their periphery.
+This becomes even more difficult to manage when these studies are in conflict. 
+The field is in desperate need of a unifying theory that can tie this swath of research together.
+This review attempts to address this issue by offering a novel perspective on the uncertainty visualisation problem.
+That is, we are going to use the wealth of established stone to construct the foundations on which we can build a house.
 

### 2024-10-10 paper.qmd
decreased trend variance so within vs between group variance statement is true (also added authors)

@@
 title: "The Noisy Work of Uncertainty Visualisation Research: A Review"
-author: Harriet Mason
+author: Harriet Mason, Dianne Cook, Sarah Goodwin, Emi Tanaka, and Susan Vanderplus
 bibliography: references.bib
@@
 my_map_data <- my_map_data |>
-  mutate(temp = 29 - 3*abs(scale(cent_lat) - sin(2*(scale(cent_long)))[,1]), # trend
+  mutate(temp = 29 - 2*abs(scale(cent_lat) - sin(2*(scale(cent_long)))[,1]), # trend
          highvar = runif(n, min=2, max=4), # high variance
@@
 @fig-choropleth depicts a choropleth map of the counties of Iowa. 
-Each of these counties are coloured according to an estimate of average daily temperature that was generated so that the values followed a clear spatial trend (hot in the middle of the map, and cold on the outside).
-The variance of these estimates were simulated such that the trend accounts for most of the variance in the plot in the low variance case (so we should expect the trend to be visible) while the high variance case has significantly more noise, so much so that it overwhelms the spatial trend.
+Each of these counties are coloured according to an estimate of average daily temperature that was generated so that the values followed a clear spatial trend. 
+The variance of these estimates were simulated such that the trend accounts for most of the variance in the plot in the low variance case (so we should expect the trend to be visible) while in the high variance, there is more variance within each county than between all the counties, so we should expect it to (at least in some capacity) overwhelm the spatial trend.
 Is this aspect of the data and the spatial data it communicates clear in in the map? 

### 2024-10-10 paper.qmd
updated general comments

@@
 <!-- TODO 
-- convert to American spelling 
+- (Di) convert to American spelling 
   (Susan) Is this necessary? I'm not sure it matters as long as you're consistent...
-- May not need to reference the same paper again and again
-  - I am not sure when I need a citation and when I dont. Sometimes I could literally cite hundreds of papers to make a point, so I'm not sure if i should cite any at all.
-- examples that aren't maps? Sure are a lot of maps.
+- (Susan) examples that aren't maps? Sure are a lot of maps.
 -->

### 2024-10-10 paper.qmd
fixed variance formula

@@
 @fig-data shows the first six rows and the geographical boundaries of our data set. 
-The temperature variable was generated using a sine wave, that is $Temperature_i = 29-3\cdot{|Latitude_i - \sin{2 \cdot Longitude_i}|}$ where the $Longitude$ and $Latitude$ are the longitude and latitude of the county's centroid scaled to a standard normal distribution.
+The temperature variable was generated using a sine wave, that is $Temperature_i = 29-2\cdot{|Latitude_i - \sin{2 \cdot Longitude_i}|}$ where the $Longitude$ and $Latitude$ are the longitude and latitude of the county's centroid scaled to a standard normal distribution.
 Each county's variance is independently, randomly sampled from a uniform distribution. 

### 2024-10-10 paper.qmd
added abstract to paper

@@
 bibliography: references.bib
+abstract: Uncertainty visualisation is quickly becomming a hot topic in information visualisation.
+Exisiting reviews in the field take the definition and purpose of an uncertainty visualisation to be self evident which results in a large amout of conflicting information.
+This conflict largely stems from a conflation between uncertainty visualisations designed for decision making and those designed to prevent false conclusions.
+We coin the term "signal suppression" to describe a visualisation that is designed for preventing false conclusions, as the approach demands that the signal (i.e.
+the collective take away of the estimates) is suppressed by the noise (i.e.
+the variance on those estimates).
+We argue that the current standards in visualisation suggest that uncertainty visualisations designed for decision making should not be considered uncertainty visualisations at all.
+Therefore, future work should focus on signal suppression.
+Effective signal suppression requires us to communicate the signal and the noise as a single "validity of signal" variable, and doing so proves to be difficult with current methods.
+We illustrate current approaches to uncertainty visualisation by showing how they would change the visual apprearance of a choropleth map.
+These maps allow us to see why some methods succeed at signal suppression, while others fall short.
+Evaluating visualisations on how well they perform signal suppression also proves to be difficult, as it involves measuring the effect of noise, a variable we typically try to ignore.
+We suggest authors use qualitative studies or compare uncertainty visualisations to the relevant hypothesis tests.
 date: last-modified
@@
 While it is true that heuristics lead people to avoid uncertainty [@Spiegelhalter2017], there is no evidence that this psychological effect translates to issues with the visual representation of uncertainty.
-Again, given that we do not make these same visual considerations for other variables that elicit distaste or irrational behavior, there is no reason to assume this is what makes uncertainty visualisation so special.
+Again, given that we do not make these same visual considerations for other variables that elicit distaste or irrational behaviour, there is no reason to assume this is what makes uncertainty visualisation so special.
 

### 2024-10-10 paper.qmd
spell check paper

@@
 In the face of this overflow of information, tools that effectively reduce piles of information to simple and clear ideas become more valuable. 
-Even more so if they can this information with a healthy amount of skepticism.
+Even more so if they can this information with a healthy amount of scepticism.
 
@@
 Today, there are an abundance of publications on the topic which makes it timely to construct a review of the field.
-That is, now that there is an overwhelming amount of information, it is valuable to distill it into simple facts. 
+That is, now that there is an overwhelming amount of information, it is valuable to distil it into simple facts. 
 In fact, there have already been several reviews published but a central piece of discussion is missing.
@@
 @uncertchap2022 summarised several cognitive effects that repeatedly arise in uncertainty visualisation experiments, however these effects were each discussed in isolation as a list of considerations an author might make rather than an overarching theory of rules for effective uncertainty visualisation.
-While these reviews are thorough in scope, none discuss how the existing literature contribute to the borader goal of uncertainty visualisation. 
+While these reviews are thorough in scope, none discuss how the existing literature contribute to the broader goal of uncertainty visualisation. 
 The problem faced by the literature is easily summarised with a famous quote by Henri Poincaré.
@@
 #| fig-cap: "Two choropleth maps that depict the counties of Iowa where each country coloured acording to a simulated average temperature.
-Both maps depict a spatial trend, where counties closer to the center of the map are hotter than counties on the edge of the map.
+Both maps depict a spatial trend, where counties closer to the centre of the map are hotter than counties on the edge of the map.
 In the low variance condition, the trend accounts for most of the variation in the data, in the high variance case, the variance on the temperature estimate accounts for most of the variance.
@@
 
-## Signal-supression
+## Signal-suppression
 Uncertainty visualisation is required for transparency. 
@@
 
-When we use "uncertainty visualation" to refer to graphics that simply communicate a variance or probability, we are classifying visualisations by the data they display, not their visual features. 
+When we use "uncertainty visualisation" to refer to graphics that simply communicate a variance or probability, we are classifying visualisations by the data they display, not their visual features. 
 Graphics, just like statistics, are not defined by their input data. 
@@
 While this is part of the problem in uncertainty visualisation, it is not the complete picture. 
-@fig-bivariate makes it clear that simply including uncertainty as a variable is insufficient to perform signal-suppresion. 
+@fig-bivariate makes it clear that simply including uncertainty as a variable is insufficient to perform signal-suppression. 
 If we cannot treat uncertainty the same as we would any other variable, how should we treat it? 
@@
 This means uncertainty should not be visualised by itself and we should seek to display signal and uncertainty together as a "single integrated uncertain value" [@Kinkeldey2014]. 
-This aspect of uncertainty visualation makes it a uniquely difficult problem.
+This aspect of uncertainty visualisation makes it a uniquely difficult problem.
 
@@
 When visualising uncertainty using its own visual channel, we can also consider visual semiotics and make sure to map uncertainty to intuitive visual channels, such as mapping more uncertain values to lighter colours [@Maceachren2012].
-Unfortunately relying on integrability may not give us the amount of control we want over our signal-supression. 
+Unfortunately relying on integrability may not give us the amount of control we want over our signal-suppression. 
 Without a strong understanding of how these visual channels collapse down into a single channel, relying on integrability could create unintended consequences such as displaying phantom signals or hiding justified signals.
@@
 
-## Value Supressing Uncertainty Palettes
+## Value Suppressing Uncertainty Palettes
 The Value Suppressing Uncertainty Palette (VSUP) [@Correll2018], was designed with the intention of preventing high uncertainty values from being extracted from a map.
@@
 Quantifiable uncertainties are just assumption violations expressed as an effect on our final statistic (or true variation of our data). 
-The disconnect between quantifiable and unqualitifable uncertainty creates a huge problem for authors trying to visualise it.
+The disconnect between quantifiable and unquantifiable uncertainty creates a huge problem for authors trying to visualise it.
 A survey of visualisation authors cited "not knowing how to calculate uncertainty" as one of the primary reasons they did not include it in visualisations [@Hullman2020a].
@@
 With this understanding it becomes clear to see why uncertainty is tied to an endless string of examples in the data analysis pipeline. 
-Uncertainty examples include imputed data, model selection, inherent randomness, biased sampling, etc, not because these things *are* uncertainty, but because they *create* uncertainty when we perform inference.
+Uncertainty examples include imputed data, model selection, inherent randomness, biased sampling, etc., not because these things *are* uncertainty, but because they *create* uncertainty when we perform inference.
 Whether or not these elements are relevant is highly dependent on what statistic you are trying to draw inference on, and by extension, the purpose of your visualisation.
@@
 The assumptions required to answer these deterministic questions are often invisible, even to the authors of the paper, which means they often create more problems than they solve when they are used as a tool for graphical evaluation.
-Similar to the vague question problem, these studies seem to be aware that we should be evaluating an uncertainty visualation based on a deterministic observation (such as our example of "does this map have a spatial trend") but are unsure how to incorporate or evaluate it.
+Similar to the vague question problem, these studies seem to be aware that we should be evaluating an uncertainty visualisation based on a deterministic observation (such as our example of "does this map have a spatial trend") but are unsure how to incorporate or evaluate it.
 
@@
 
-## Testing signal supression
+## Testing signal suppression
 If we cannot ask direct questions about uncertainty, or measure the secondary effects of uncertainty, or ask indirect questions about the uncertainty; then how are we supposed to evaluate uncertainty visualisations?
@@
 ### Comparing to hypothesis tests
-The most obvious way to evaluate uncertainty visualations is to compare the visualisations to statistical tests. 
+The most obvious way to evaluate uncertainty visualisations is to compare the visualisations to statistical tests. 
 If a graphic was performing signal-suppression we would expect the signal to be harder to see at higher levels of uncertainty. 
@@
 The experimenters goal was to observe and organise the methods people use when making decisions in the face of uncertainty. 
-They highlighted the specific aspects of uncertainty that participants typically considered, such as the range of outcomes that are above/below a certain threshold, minimum and maximum values, the risk of a loss, etc, and identified where in the decision making process participants made these considerations.
+They highlighted the specific aspects of uncertainty that participants typically considered, such as the range of outcomes that are above/below a certain threshold, minimum and maximum values, the risk of a loss, etc., and identified where in the decision making process participants made these considerations.
 
@@
 The sine illusion can cause the confidence interval of a smoothed sine curve to seem wider at the peaks than the troughs, causing us to underestimate uncertainty associated with changing values [@Vanderplas2015].
-Points that were on an outcome of an ensemble display were perceived as more likely than points not on an outcome, even when the point that was not on an outcome was closer to the center of the distribution (and therefore more likely) [@Padilla2017].
+Points that were on an outcome of an ensemble display were perceived as more likely than points not on an outcome, even when the point that was not on an outcome was closer to the centre of the distribution (and therefore more likely) [@Padilla2017].
 This can be considered an extension of the within bar bias, where participants looking at a bar chat with error bars view outcomes within the bar as more likely than those outside it [@Newman2012]. 
@@
 These heuristics have the potential to create noise in signal suppression evaluations, and should be kept in mind when designing an evaluation experiment.
-
-{{< pagebreak >}}
 

### 2024-10-16 paper.qmd
fixed up author list + author acknowledgement edit

@@
 title: "The Noisy Work of Uncertainty Visualisation Research: A Review"
-author: Harriet Mason, Dianne Cook, Sarah Goodwin, Emi Tanaka, and Susan Vanderplus
+author: 
+  - name: Harriet Mason
+    url: https://harrietmason.netlify.app/
+    orcid: 0009-0007-4568-8215
+    email:  harriet.m.mason@gmail.com
+    affiliation:
+    - name: Monash University
+      department: Department of Econometrics and Business Statistics
+      city: Melbourne
+      country: Australia
+  - name: Dianne Cook
+    url: https://dicook.org
+    orcid: 0000-0002-3813-7155
+    email: dicook@monash.edu
+    affiliation:
+    - name: Monash University
+      department: Department of Econometrics and Business Statistics
+      city: Melbourne
+      country: Australia
+  - name: Sarah Goodwin
+    url: www.linkedin.com/in/smgoodwin/
+    orcid: 0000-0001-8894-8282
+    email: Sarah.Goodwin@monash.edu
+    affiliation:
+    - name: Monash University
+      department: Department of Human Centred Computing
+      city: Melbourne
+      country: Australia
+  - name: Emi Tanaka
+    url: emitanaka.org
+    orcid: 0000-0002-1455-259X
+    email: Emi.Tanaka@anu.edu.au
+    affiliation: 
+    - name: The Australian National University
+      department: Biological Data Science Institute
+      city: Canberra
+      country: Australia
+  - name: Susan Vanderplus
+    url: srvanderplas.github.io
+    orcid: 0000-0002-3803-0972
+    email: susan.vanderplas@unl.edu
+    affiliation:
+    - name: University of Nebraska–Lincoln
+      department: Statistics Department
+      city:  Lincoln
+      country: United States
 bibliography: references.bib
@@
 ---
-
-<!-- TODO 
-- (Di) convert to American spelling 
-  (Susan) Is this necessary? I'm not sure it matters as long as you're consistent...
-- (Susan) examples that aren't maps? Sure are a lot of maps.
--->
 
@@
 # Acknowledgements
-This work is supported in part by a scholarship from the the Australian Energy Market Operator. 
+The first author of this paper is supported in part by a scholarship from the the Australian Energy Market Operator. 
 The R packages were used for this work were: `tidyverse` [@tidyverse], `Vizumap` [@Vizumap], `RColorBrewer` [@RColorBrewer], `scales` [@scales], `sf` [@sf], `ggrepel` [@ggrepel], `urbnmapr` [@urbnmapr], `flextable` [@flextable], `colorspace` [@colorspace], and `rgeos` [@rgeos].

### 2024-10-16 paper.qmd
edits sections with unecessary "additionally"

@@
 
-Among these summary tools, information visualisations are one of the most powerful. 
-Visualisations allow us to *learn* what is important about our data and datasets such as Anscombe's quartet [@anscombe] or the Datasaurus Dozen [@datasaurpkg] highlight this power. 
-Additionally, visualisations allow for efficient and memorable communication. 
-Even something as simple as sketching a distribution before recalling statistics or making predictions can greatly increase the accuracy of those measures [@Hullman2018; @Goldstein2014].
-
-Uncertainty visualisation is a relatively new field in research. 
-Early papers that specifically reference "uncertainty visualisation" appear in the late 80s [@Ibrekk1987], with geospatial information visualisation literature in the early 90s declaring this to be essential aspect of information display [@MacEachren1992; @Carr1992].
+Among these summary tools, information visualisations are one of the most powerful as they allow for quick and memorable communication that allow us to identify quirks in our data that we didn't know to look for.
+Datasets such as Anscombe's quartet [@anscombe] or the Datasaurus Dozen [@datasaurpkg] show a case where visual statistics highlight elements of the data that are invisible to the typical summary statistics.
+Something as simple as sketching a distribution before recalling statistics or making predictions can greatly increase the accuracy of those measures [@Hullman2018; @Goldstein2014].
+
+"Uncertainty visualisation" is a relatively new field in research. 
+Early papers that specifically reference uncertainty visualisation appear in the late 80s [@Ibrekk1987], with geospatial information visualisation literature in the early 90s declaring this to be essential aspect of information display [@MacEachren1992; @Carr1992].
 @fig-ibrekk depicts an example of the uncertainty visualisations discussed in these early papers. 
@@
 Mistakes due to misunderstanding visualisations, misinterpreting questions, and incorrectly applying heuristics are all combined into a single error value.
-@Spiegelhalter2017 commented that different plots are good for different things, and disagreed with the goal of identifying a universal "best" plot for all people and circumstances.
+@Spiegelhalter2017 commented that different plots are good for different things, and disagreed with the goal of identifying a universal best plot for all people and circumstances.
 @Griethe2006 did not identify common themes, but instead listed the findings and opinions of a collection of papers. 
@@
 Some authors have said that uncertainty should be included to degree of confidence or trust in the data [@Boukhelifa2012; @Zhao2023]. 
-Some authors directly connect uncertainty visualisation to hypothesis testing as it ensures the "validity" of a statement [@Hullman2020a; @Griethe2006], but allows for a proportional level of trust that is more detailed than the binary results of a hypothesis test [@Correll2014; @Correll2018].
+Some authors directly connect uncertainty visualisation to hypothesis testing as it ensures the validity of a statement [@Hullman2020a; @Griethe2006], but allows for a proportional level of trust that is more detailed than the binary results of a hypothesis test [@Correll2014; @Correll2018].
 Some authors even go so far as to claim that failing to include uncertainty is akin to fraud or lying [@Hullman2020a; @Manski2020].
@@
 A successful uncertainty visualisation would act as a "statistical hedge" for any inference we make using the graphic. 
-Additionally, since the purpose of a visualisation is to give a quick "gist" of the information [@Spiegelhalter2017], this hedging needs to be communicated visually without the need for complicated calculations.
-If we refer to the conclusion we draw from a graphic to be its "signal" and the variance that makes this signal harder to identify as the "noise", we can summarise the above information into three key requirements.
+Since the purpose of a visualisation is to give a quick gist of the information [@Spiegelhalter2017], this hedging needs to be communicated visually without the need for complicated calculations. 
+If we refer to the conclusion we draw from a graphic to be its signal and the variance that makes this signal harder to identify as the "noise", we can summarise the above information into three key requirements.
 A good uncertainty visualisation needs to:
@@
 As @fig-choropleth showed, visualisations that are unconcerned with uncertainty have no issue showing justified signals, but struggle with the display of unjustified signals. 
-Therefore, we call this approach to uncertainty visualisation as "signal-suppression" since it primarily differentiates itself from from the "noiseless" visualisation approach through criteria (2). 
-That is, the main difference between an uncertainty visualisation and a "normal" visualisation is that an uncertainty visualisation should prevent us from drawing unjustified conclusions.
+Therefore, we call this approach to uncertainty visualisation as "signal-suppression" since it primarily differentiates itself from from the normal "noiseless" visualisation approach through criteria (2).
+That is, the main difference between an uncertainty visualisation and a normal visualisation is that an uncertainty visualisation should prevent us from drawing unjustified conclusions.
 
@@
 Other authors do not describe uncertainty as important for decision making, but rather explicitly state it as a variable of importance in of itself [@Blenkinsop2000]. 
-While uncertainty can provide useful information in decision making, it is important to recognise that the "uncertainty" in these cases is not acting as "uncertainty" at all. 
+While uncertainty can provide useful information in decision making, it is important to recognise that the "uncertainty" in these cases is not acting as uncertainty at all. 
 It is acting as signal.
 
-This is obvious for the cases where we are explicitly interested in the variance or error, as we are literally trying to draw conclusions about an "uncertainty" statistic. 
-The same is true for "decision making" experiments, but it is less overt. 
+This is obvious for the cases where we are explicitly interested in the variance or error, as we are literally trying to draw conclusions about an statistic that is used to describe uncertainty. 
+The same is true for visualisations made for decision making, but it is less overt. 
 This is easiest to understand with an example.
@@
 This is *not* because uncertainty in general is important for decision making, but because it gives you the tools required to calculate the *actual* statistic you are basing your decision on. 
-In this sense, uncertainty is no more "special" to decision making than weight is special to a body mass index calculation.
+In this sense, uncertainty is no more special to decision making than weight is special to a body mass index calculation.
 
@@
 @fig-exceed depicts an exceedance probability map that was designed as an alternative to the choropleth map to improve decision making under uncertainty [@Kuhnert2018; @Lucchesi2021]. 
-A keen viewer may notice that the "exceedance probability map" is actually just a choropleth map, only the statistic being displayed has changed. 
+A keen viewer may notice that the exceedance probability map is actually just a choropleth map, only the statistic being displayed has changed. 
 We are not sure it is productive to categorise this visualisation as an uncertainty visualisation.
@@
 
-There seem to be two different definitions of "uncertainty visualisation" floating around in the literature. 
+There seem to be two different definitions of uncertainty visualisation floating around in the literature. 
 The first considers *any* visualisation of error, variance, or probability to be an uncertainty visualisation. 
-The second believes an "uncertainty visualisation" is the output of a function that takes a normal visualisation as an input, and transforms it to include uncertainty information. 
+The second believes an uncertainty visualisation is the output of a function that takes a normal visualisation as an input, and transforms it to include uncertainty information. 
 The former group believe the purpose of uncertainty visualisation to provide signal about a distribution, while the later believe it should act as noise to obfuscate a signal. 
 The lack of explicit distinction between these two motivations leaves the literature muddled and reviewers struggle to understand if uncertainty should be treated as a variable, as metadata, or as something else entirely [@Kinkeldey2014].
-This disagreement creates constant contradictions in what the literature considers to be an "uncertainty visualisation". 
+This disagreement creates constant contradictions in what the literature considers to be an uncertainty visualisation. 
 For example @Leland2005 mentions that popular graphics, such as pie charts and bar charts omit uncertainty, and @Wickham2011 suggests their product plot framework, which includes histograms and bar charts, should be extended to include uncertainty.
-Additionally, at least one of these charts have been used in a significant number of uncertainty visualisation experiments [@Ibrekk1987; @Olston2002; @Zhao2023; @Hofmann2012]. 
-If you view uncertainty as a function applied to an existing graphic, then you would believe a pie chart and bar chart are not uncertainty visualisations, as they are yet to have the "uncertainty visualisation function" applied to them.
-If you view uncertainty as any graphic that depicts an "uncertainty statistic" then there are no limitations on which graphics can or cannot be uncertainty visualisations.
-
-When we use "uncertainty visualisation" to refer to graphics that simply communicate a variance or probability, we are classifying visualisations by the data they display, not their visual features. 
+However, pie charts, bar charts and histograms have all been used in a significant number of uncertainty visualisation experiments [@Ibrekk1987; @Olston2002; @Zhao2023; @Hofmann2012]. 
+If you view an uncertainty visualisation as a function applied to an existing graphic, then you would not see a pie chart or bar chart as uncertainty visualisations.
+These charts are they are yet to have the uncertainty visualisation function applied to them. 
+If you view an uncertainty visualisation as any graphic that depicts a statistic then there are no limitations on which graphics can or cannot be uncertainty visualisations.
+
+When we use the term uncertainty visualisation to refer to graphics that simply communicate a variance or probability, we are classifying visualisations by the data they display, not their visual features.
 Graphics, just like statistics, are not defined by their input data. 
@@
 This history lesson illustrates what uncertainty actually is. 
-At several stages in a statistical analysis, we will violate the "uniformity of nature" assumption. 
+At several stages in a statistical analysis, we will violate the uniformity of nature assumption. 
 Each of these violations will impact the statistic we have calculated and push it further from the population parameter we wish to draw inference on. 
@@
 That is, we don't want the data represented through one visual channel to interfere with the others [@Smart2019]. 
-Mapping uncertainty and signal to separable channels allows them to be read separately, which does not align with the goal of communicating them as a "single integrated channel". 
-Additionally, visualising uncertainty and signal separately allows the uncertainty information to simply be ignored, which is a pervasive issue in current uncertainty visualisation methods [@uncertchap2022].
+Mapping uncertainty and signal to separable channels allows them to be read separately, which does not align with the goal of communicating them as a single integrated channel. 
+Visualising uncertainty and signal separately allows the uncertainty information to simply be ignored, which is a pervasive issue in current uncertainty visualisation methods [@uncertchap2022]. 
 We can see this problem in @fig-bivariate, as it sends the message "this data has a spatial trend and the estimates have a large variance" as we read the signal and the uncertainty separately.
@@
 
-It is clear that the pixel-map is not only "suppressing" the false information, but it is doing so by simulating *more* information. 
+It is clear that the pixel-map is not only suppressing the false information, but it is doing so by simulating *more* information. 
 The efficacy of this method means that visualisations of simulated samples pop up repeatedly in the literature, with examples including samples that are animated over time [@Hullman2015; @Blenkinsop2000], pixel-maps, and spaghetti time series plots.
-Not only does this method help readers understand the plot level "gist", it is also unlikely to damage the viewers ability to extract individual estimates.
+Not only does this method help readers understand the plot level gist, it is also unlikely to damage the viewers ability to extract individual estimates.
 Extracting global statistics, such as the mean or variance, from a sample can be done with relative ease, especially when those values are mapped to colour [@Franconeri2021]. 
@@
 
-This relationship between uncertainty and the "purpose" of our analysis is littered throughout the literature. 
-Multiple authors have commented on the need to consider quantifying and expressing uncertainty at every stage of a project as the "goal" shapes every step of the analysis [@Kinkeldey2014; @Hullman2016; @Refsgaard2007].
+This relationship between uncertainty and the purpose of our analysis is littered throughout the literature. 
+Multiple authors have commented on the need to consider quantifying and expressing uncertainty at every stage of a project as the goal shapes every step of the analysis [@Kinkeldey2014; @Hullman2016; @Refsgaard2007].
 @Otsuka2023 suggested that the process of observing data to perform statistics is largely dependent on our goals, because the process of boiling real world entities down into probabilistic objects (or "probabilistic kind" as he puts it) depends on the relationship we seek to identify with our data.
@@
 If we don't know what the final statistic is, we cannot quantify the effects of our assumptions. 
-Therefore, it is often the case that the best uncertainty visualisation is not an "uncertainty visualisation" at all, but simply the most accurate depiction of our raw data.
+Therefore, it is often the case that the best uncertainty visualisation is not an uncertainty visualisation at all, but simply the most accurate depiction of our raw data.
 This does not mean that visualising raw data will always prevent insignificant signal from getting through. 
@@
 This seems like a relatively straight forward approach, and it is similar to how non-uncertainty visualisations are evaluated, but is this appropriate for uncertainty visualisations? 
-The role of uncertainty is rarely evaluated in these studies as the graphics are often compared on the basis of being "uncertainty visualisations" [@Ibrekk1987; @Hullman2015; @Hofman2020], a class that has no established definition.
+The role of uncertainty is rarely evaluated in these studies as the graphics are often compared on the basis of being uncertainty visualisations [@Ibrekk1987; @Hullman2015; @Hofman2020], a class that has no established definition.
 By shifting the focus of our inference from $\bar{X}$ to $Var(X)$ or $P(X)$ we end up evaluating visualisations on their ability to convey uncertainty statistics, rather than on uncertainty's ability to suppress statistics.
@@
 This perspective ends up extending to visualisation authors as well.
-@Hullman2020a found that authors simultaneously argued that failing to visualise uncertainty was akin to fraud, but also many avoided uncertainty visualisation because they didn't want their work to come across as "untrustworthy".
+@Hullman2020a found that authors simultaneously argued that failing to visualise uncertainty was akin to fraud, but also many avoided uncertainty visualisation because they didn't want their work to come across as untrustworthy.
 This is a classic example of the negative impacts of placing direct importance on *trust* rather than *transparency*. 
@@
 Confidence has many of the same issues as trust, but it has an additional confounding factor. 
-In non-uncertainty visualisation evaluation experiments, "confidence" is used as a proxy for the clarity of the visualisation. 
+In non-uncertainty visualisation evaluation experiments, confidence is used as a proxy for the clarity of the visualisation. 
 Confidence cannot simultaneously be a measure of clarity of visualisation *and* a way to capture the uncertainty expressed in a visualisation. 
@@
 There is a collection of studies that seem to be aware of the issues behind using trust or value extraction to evaluate uncertainty visualisations. 
-These studies try to measure the effect of some kind of "single integrated value" but the methodology is often ad-hoc with varying levels of success.
+These studies try to measure the effect of some kind of single integrated value, but the methodology is often ad-hoc with varying levels of success.
 
@@
 
-A variation of the "vague question" problem, is a series of studies that ask questions that are impossible to answer using the information given to the participants. 
+A variation of the vague question problem, is a series of studies that ask questions that are impossible to answer using the information given to the participants. 
 These studies frequently expect participants to give deterministic answers for probabilistic questions. 
@@
 This is further complicated by the fact that visual statistics and their computational equivalents often differ, even before we consider these measurement errors.
-Additionally, by having a ground truth, they end up implicitly asking users to treat uncertainty as a signal which goes against the purposes of signal-suppression.
+By having a ground truth, these experiments implicitly ask users to treat uncertainty as a signal which goes against the purposes of signal-suppression.
 
@@
 This approach is similar to the lineup protocol as it also evaluates a visualisation by comparing it to an analogous statistical calculation, however it does so using a different statistical philosophy.
-The connection with visualisation and statistical tests even extends to EDA, where @Hullman2021 commented that a visualisation can be considered a model check where the null hypothesis is "the expected" and we are looking for "the unexpected".
+The connection with visualisation and statistical tests even extends to EDA, where @Hullman2021 commented that a visualisation can be considered a model check where the null hypothesis is the expected and we are looking for the unexpected.
 This is all to say that comparing visualisations to hypothesis tests is not uncharted territory.
@@
 However, this finding does not directly translate to the lineup protocol being worse than its computational equivalent.
-The lineup protocol allows viewers to detect "something is not right" even if that something is unspecified, and even if that "something" is not what see the authors intended.
+The lineup protocol allows viewers to detect if "something" is not right even if that something is unspecified, or what the authors intended.
 For example, @vanderplas2017 did an experiment where participants were drawn to plots with unequal cluster sizes, even though the size of the cluster was not one of the features the authors were testing.
-While standard hypothesis tests may outperform the lineup protocol on power, computational tests have the benefit of knowing exactly what they are looking for, while visualisations are often explicitly made when we *don't* know what we are looking for.
-Additionally, while lineup protocols often have a specific hypothesis that is used to create the null distribution, uncertainty visualisations are intended to be more flexible and can be created without identifying a specific null distribution.
+While standard hypothesis tests may outperform the lineup protocol on power, computational tests have the benefit of knowing exactly what they are looking for. 
+Visualisations are often explicitly made when we *don't* know what we are looking for.
+Additionally, lineup protocols often have a specific hypothesis that is used to create the null distribution, while uncertainty visualisations are intended to be more flexible and can be created without identifying a specific null distribution.
 Therefore, while we could evaluate uncertainty visualisations by comparing them to a selection of hypothesis tests, we should not expect the same level of sensitivity that we get in standard hypothesis tests as the assumed knowledge about our data is quite different.
@@
 This method can enlighten authors as to *how* the uncertainty information was used by the participants.
-@Hofmann2012 captured this by asking participants why they considered a particular plot to be more "right shifted", even though this qualitative analysis did not make it into the final paper. 
+@Hofmann2012 captured this by asking participants why they considered a particular plot to be more right-shifted, even though this qualitative analysis did not make it into the final paper. 
 @Daradkeh2015 presented participants with ten investment alternatives and asked participants "from among available alternatives, which alternative do you prefer the most", and were asked to think aloud and consider the uncertainty in their decision making.
@@
 *Software that allows users to easily perform signal suppression should be developed.* 
-Existing uncertainty visualisation methods view a distribution as its own object and there are no software options for the "an uncertainty visualisation is a function of a visualisation" philosophy.
+Existing uncertainty visualisation methods view a distribution as its own object and there are no software options for the "an uncertainty visualisation is a function of an existing visualisation" philosophy.
 

### 2024-10-16 paper.qmd
fix define uncertainty thing

@@
 In the face of this overflow of information, tools that effectively reduce piles of information to simple and clear ideas become more valuable. 
-Even more so if they can this information with a healthy amount of scepticism.
+That is, we need tools that can sort the signal from the noise.
 
@@
 "Uncertainty visualisation" is a relatively new field in research. 
-Early papers that specifically reference uncertainty visualisation appear in the late 80s [@Ibrekk1987], with geospatial information visualisation literature in the early 90s declaring this to be essential aspect of information display [@MacEachren1992; @Carr1992].
+Early mentions of uncertainty visualisation start to appear in the late 1980s [@Ibrekk1987], with geospatial information visualisation literature from the early 1990s declaring this to be essential aspect of any information display [@MacEachren1992; @Carr1992].
 @fig-ibrekk depicts an example of the uncertainty visualisations discussed in these early papers. 
-Despite the new terminology visualisation of uncertainty has been present since the earliest times. 
+Despite kicking off the field, these papers did not define uncertainty visualisation.
+This has led to a lack of consensus on what it means for a graphic to visualise uncertainty, an issue we will return to later.
+Therefore, while the field is considered to be quite new, many of the graphics used for uncertainty visualisation have been around for much longer. 
 For example, box plots and histograms display variation which becomes synonymous with uncertainty when we are using them to depict the variation of an estimate. 
@@
 
-*Experimental practices on uncertainty visualisation need to be standardised.* Uncertainty visualisations for decision making treat uncertainty as signal, while visualisations for signal-suppression treat uncertainty as noise.
+*Experimental practices on uncertainty visualisation need to be standardised.* 
+Uncertainty visualisations for decision making treat uncertainty as signal, while visualisations for signal-suppression treat uncertainty as noise. 
 As the literature currently exists, there is no way to combine papers to get a meaningful sense of how uncertainty information is understood by a viewer.

### 2024-10-16 paper.qmd
rewrote example to highlight scope of work, + edited author links

@@
   - name: Sarah Goodwin
-    url: www.linkedin.com/in/smgoodwin/
+    url: https://www.linkedin.com/in/smgoodwin/
     orcid: 0000-0001-8894-8282
@@
   - name: Emi Tanaka
-    url: emitanaka.org
+    url: https://emitanaka.org/
     orcid: 0000-0002-1455-259X
@@
   - name: Susan Vanderplus
-    url: srvanderplas.github.io
+    url: https://srvanderplas.github.io
     orcid: 0000-0002-3803-0972
@@
 ## Spatial example
-
-There are far too many uncertainty visualisations to exhaustively discuss them all. Instead we focus on the changes made to a single plot, the choropleth map. 
+There are far too many uncertainty visualisations to exhaustively discuss them all. 
+Instead we focus on the changes made to a single plot, the choropleth map. 
 Due to the field's origins and focus in geospatial information visualisation, there have been a large number of suggested variations on the choropleth map that allow authors to include uncertainty. 
-While the theoretical approach we outline is useful to all uncertainty visualisations regardless of their application, focusing on a single example will help isolate the ideas we are trying to convey.
-We will use these maps to show the effects we are talking about, and highlight the costs and benefits of each approach.
+Utilising a single example will help isolate the ideas we are trying to convey. 
+Therefore, it is important to remember that even though we focus our discussion on the choropleth map, the theoretical approach we outline in this review is useful to all uncertainty visualisations regardless of their application.
+Additionally, our examples focus on incorporating uncertainty through colour manipulation, as that is the key visual channel used in a choropleth map.
+However, the methods we discuss go beyond variations in a colour palette.
+Even though they are not explicitly shown, visualisations that depict uncertainty using layers such as position or shape, and more complicated graphics that incorporate animation or interactivity are also within the scope of this review.
+We will use the choropleth maps as a tool to clearly highlight the costs and benefits of each approach.
 

### 2024-10-18 paper.qmd
formatting

@@
 date: last-modified
-toc: true
+toc: false
 number-sections: true
-format: pdf
+format:
+  jasa-pdf:
+    keep-tex: true  
+    journal:
+      blinded: false
+  jasa-html: default
 fig-valign: bottom
@@
 ```{r}
+#| eval: false
+#| echo: false
+# Get map data: do this once and save
+my_map_data <- get_urbn_map("counties", sf = TRUE) |>
+  filter(state_name=="Iowa")
+save(my_map_data, file="data/my_map_data.rda")
+```
+
+```{r}
 #| echo: false
@@
 #| cap-location: "bottom"
+load("data/my_map_data.rda")
 

### 2024-10-18 paper.qmd
added comment for package install

@@
 library(tidyverse)
+# devtools::install_github("lydialucchesi/Vizumap")
 library(Vizumap)
@@
 library(ggrepel)
+# devtools::install_github("UrbanInstitute/urbnmapr")
 library(urbnmapr)

### 2024-10-18 paper.qmd
separate and save questionable code chunks (api or depreciated)

@@
   filter(state_name=="Iowa")
+save(my_map_data, file="data/iowa_map.rda")
+```
+
+```{r}
+#| eval: false
+#| echo: false
+# Get centroid because rgeos is depreciated
+load("data/iowa_map.rda")
+centroids <- as_tibble(gCentroid(as(my_map_data$geometry, "Spatial"), byid = TRUE))
+my_map_data$cent_long <- centroids$x
+my_map_data$cent_lat <- centroids$y
 save(my_map_data, file="data/my_map_data.rda")
@@
 #| cap-location: "bottom"
+
+# get data
 load("data/my_map_data.rda")
@@
 
-# Get map data
-my_map_data <- get_urbn_map("counties", sf = TRUE) |>
-  filter(state_name=="Iowa")
-centroids <- as_tibble(gCentroid(as(my_map_data$geometry, "Spatial"), byid = TRUE))
-my_map_data$cent_long <- centroids$x
-my_map_data$cent_lat <- centroids$y
+# data dimension for sampling
 n <- dim(my_map_data)[1]
@@
 ```{r}
+#| eval: false
+#| echo: false
+# Make + save pixel map (in case of depreciation)
+# Low variance map
+my_map_data_a <- my_map_data |>
+  filter(variance_class == "lowvar") |>
+  mutate(my_id = seq(n),
+         error = variance)
+
+# quantile
+q_a <- my_map_data_a |> 
+  as.tibble() |>
+  mutate(bitemp=as.numeric(bitemp)) |>
+  with(data.frame(p0.05 = qnorm(0.05, mean=bitemp, sd=sqrt(variance)),
+                  p0.25 = qnorm(0.25, mean=bitemp, sd=sqrt(variance)), 
+                  p0.5 = qnorm(0.5, mean=bitemp, sd=sqrt(variance)), 
+                  p0.75 = qnorm(0.75, mean=bitemp, sd=sqrt(variance)), 
+                  p0.95 = qnorm(0.95, mean=bitemp, sd=sqrt(variance))))
+
+pixel_1a <- my_map_data_a |>
+  as.data.frame() |>
+  select(my_id, bitemp, error) |>
+  read.uv(estimate="bitemp", error="error")
+
+pixel_2a <- my_map_data_a |> as("Spatial")
+
+pix_a <- pixelate(pixel_2a, pixelSize = 70, id = "my_id")
+
+pmap_a <- build_pmap(data = pixel_1a, distribution = "discrete", pixelGeo = pix_a, id = "my_id", border = pixel_2a, q=q_a)
+
+p4a <- view(pmap_a) +
+  geom_path(
+      data = pmap_a$bord,
+      aes_string(x = 'long', y = 'lat', group = 'group'),
+      colour = "white"
+    )  +
+  scale_fill_gradientn(colours = longpal) +
+  scale_colour_gradientn(colours = longpal) +
+  theme(legend.position="none")
+
+# High variance
+my_map_data_b <- my_map_data |>
+  filter(variance_class == "highvar") |>
+  mutate(my_id = seq(n),
+         error = variance)
+
+q_b <- my_map_data_b |> 
+  as.tibble() |>
+  mutate(bitemp=as.numeric(bitemp)) |>
+  with(data.frame(p0.05 = qnorm(0.05, mean=bitemp, sd=sqrt(variance)),
+                  p0.25 = qnorm(0.25, mean=bitemp, sd=sqrt(variance)), 
+                  p0.5 = qnorm(0.5, mean=bitemp, sd=sqrt(variance)), 
+                  p0.75 = qnorm(0.75, mean=bitemp, sd=sqrt(variance)), 
+                  p0.95 = qnorm(0.95, mean=bitemp, sd=sqrt(variance))))
+
+pixel_1b <- my_map_data_b |>
+  as.data.frame() |>
+  select(my_id, bitemp, error) |>
+  read.uv(estimate="bitemp", error="error")
+
+pixel_2b <- my_map_data_b |> 
+  as("Spatial")
+
+pix_b <- pixelate(pixel_2b, pixelSize = 70, id = "my_id")
+
+pmap_b <- build_pmap(data = pixel_1b, distribution = "discrete", pixelGeo = pix_b, id = "my_id", border = pixel_2b, q=q_b)
+
+p4b <- view(pmap_b) +
+  geom_path(
+      data = pmap_b$bord,
+      aes_string(x = 'long', y = 'lat', group = 'group'),
+      colour = "white"
+    )  +
+  scale_fill_gradientn(colours = longpal) +
+  scale_colour_gradientn(colours = longpal) +
+  theme(legend.position="none")
+
+save(p4a, file="data/p4a.rda")
+save(p4b, file="data/p4b.rda")
+```
+
+```{r}
 #| echo: false
@@
 
-# Pixel Map
-
-# Low variance map
-my_map_data_a <- my_map_data |>
-  filter(variance_class == "lowvar") |>
-  mutate(my_id = seq(n),
-         error = variance)
-
-# quantile
-q_a <- my_map_data_a |> 
-  as.tibble() |>
-  mutate(bitemp=as.numeric(bitemp)) |>
-  with(data.frame(p0.05 = qnorm(0.05, mean=bitemp, sd=sqrt(variance)),
-                  p0.25 = qnorm(0.25, mean=bitemp, sd=sqrt(variance)), 
-                  p0.5 = qnorm(0.5, mean=bitemp, sd=sqrt(variance)), 
-                  p0.75 = qnorm(0.75, mean=bitemp, sd=sqrt(variance)), 
-                  p0.95 = qnorm(0.95, mean=bitemp, sd=sqrt(variance))))
-
-pixel_1a <- my_map_data_a |>
-  as.data.frame() |>
-  select(my_id, bitemp, error) |>
-  read.uv(estimate="bitemp", error="error")
-
-pixel_2a <- my_map_data_a |> as("Spatial")
-
-pix_a <- pixelate(pixel_2a, pixelSize = 70, id = "my_id")
-
-pmap_a <- build_pmap(data = pixel_1a, distribution = "discrete", pixelGeo = pix_a, id = "my_id", border = pixel_2a, q=q_a)
-
-p4a <- view(pmap_a) +
-  geom_path(
-      data = pmap_a$bord,
-      aes_string(x = 'long', y = 'lat', group = 'group'),
-      colour = "white"
-    )  +
-  scale_fill_gradientn(colours = longpal) +
-  scale_colour_gradientn(colours = longpal) +
-  theme(legend.position="none")
-
-# High variance
-my_map_data_b <- my_map_data |>
-  filter(variance_class == "highvar") |>
-  mutate(my_id = seq(n),
-         error = variance)
-
-q_b <- my_map_data_b |> 
-  as.tibble() |>
-  mutate(bitemp=as.numeric(bitemp)) |>
-  with(data.frame(p0.05 = qnorm(0.05, mean=bitemp, sd=sqrt(variance)),
-                  p0.25 = qnorm(0.25, mean=bitemp, sd=sqrt(variance)), 
-                  p0.5 = qnorm(0.5, mean=bitemp, sd=sqrt(variance)), 
-                  p0.75 = qnorm(0.75, mean=bitemp, sd=sqrt(variance)), 
-                  p0.95 = qnorm(0.95, mean=bitemp, sd=sqrt(variance))))
-
-pixel_1b <- my_map_data_b |>
-  as.data.frame() |>
-  select(my_id, bitemp, error) |>
-  read.uv(estimate="bitemp", error="error")
-
-pixel_2b <- my_map_data_b |> 
-  as("Spatial")
-
-pix_b <- pixelate(pixel_2b, pixelSize = 70, id = "my_id")
-
-pmap_b <- build_pmap(data = pixel_1b, distribution = "discrete", pixelGeo = pix_b, id = "my_id", border = pixel_2b, q=q_b)
-
-p4b <- view(pmap_b) +
-  geom_path(
-      data = pmap_b$bord,
-      aes_string(x = 'long', y = 'lat', group = 'group'),
-      colour = "white"
-    )  +
-  scale_fill_gradientn(colours = longpal) +
-  scale_colour_gradientn(colours = longpal) +
-  theme(legend.position="none")
-
+# load pixel maps
+load("data/p4a.rda")
+load("data/p4b.rda")
 # Display plots and palettes

### 2024-10-18 paper.qmd
changed format to jasa and edited reference so that it would work

@@
 
-{{< pagebreak >}}
-
-# Introduction
+## Introduction
 
@@
 
-## Spatial example
+### Spatial example
 There are far too many uncertainty visualisations to exhaustively discuss them all. 
@@
 
-# Ignoring uncertainty
+## Ignoring uncertainty
 A good place to start might be at a deceptively straight forward question, why should we include uncertainty at all?
 
-## The choropleth map
+### The choropleth map
 @fig-choropleth depicts a choropleth map of the counties of Iowa. 
@@
 
-## Signal-suppression
+### Signal-suppression
 Uncertainty visualisation is required for transparency. 
@@
 
-## Uncertainty as a signal
+### Uncertainty as a signal
 Uncertainty visualisation is not only motivated by signal-suppression, and we would be remiss if we did not mention these alternative approaches. 
@@
 
-# Visualising uncertainty as a variable
+## Visualising uncertainty as a variable
 Upon hearing that uncertainty needs to be included for transparency, the solutions may seem obvious. 
@@
 
-## The bivariate map
+### The bivariate map
 @fig-bivariate depicts a variation of the choropleth map, where we have a two dimensional colour palette. 
@@
 
-## Why this approach may (or may not) work
+### Why this approach may (or may not) work
 The difficulty in incorporating uncertainty into a visualisation is frequently mentioned but seldom explained. 
@@
 
-### It's a variable... it's metadata... it's uncertainty?
+#### It's a variable... it's metadata... it's uncertainty?
 Describing what uncertainty actually is, is surprisingly hard. 
@@
 
-### Visualising the "single integrated uncertain value"
+#### Visualising the "single integrated uncertain value"
 Typically, when making visualisations, we want the visual channels to be separable. 
@@
 
-# Combining uncertainty and signal in a transformed space
+## Combining uncertainty and signal in a transformed space
 Instead of hoping that uncertainty might collapse signal values into a single dimension, we can do some of that work ourselves. 
@@
 
-## Value Suppressing Uncertainty Palettes
+### Value Suppressing Uncertainty Palettes
 The Value Suppressing Uncertainty Palette (VSUP) [@Correll2018], was designed with the intention of preventing high uncertainty values from being extracted from a map.
@@
 
-## What can and cannot be suppressed?
+### What can and cannot be suppressed?
 The methods used by the VSUP bring to light a slight problem with uncertainty visualisation. 
@@
 
-### There is no uncertainty in EDA
+#### There is no uncertainty in EDA
 Earlier we established that uncertainty is a by-product of inference, which means without inference, there is no uncertainty. 
@@
 
-### The limitations of explicitly visualising uncertainty and signal
+#### The limitations of explicitly visualising uncertainty and signal
 The lack of versatility of the VSUP is easy to see with a simple example.
@@
 
-# Implicitly Combining Uncertainty and Signal
+## Implicitly Combining Uncertainty and Signal
 Rather than trying to figure out how to combine signal and uncertainty into a single colour, we can just display a sample instead and allow the viewer to extract *both* the estimate and the variance.
 
-## Pixel map
+### Pixel map
 @fig-pixel displays a pixel map [@Lucchesi2021], which is a variation of the choropleth map where each area is divided up into several smaller areas, each coloured using outcomes from the larger area's (i.e.
@@
 
-## Show me the data
+### Show me the data
 To thoroughly understand why pixel maps seem to successfully perform signal suppression, we would need a more thorough discussion than what can be offered here. 
@@
 
-# Evaluating uncertainty visualisations
+## Evaluating uncertainty visualisations
 If we want to make conclusions about how effective any uncertainty visualisation method is we need to look at the results of evaluation experiments. 
@@
 
-## Current methods
+### Current methods
 Including uncertainty in a visualisation comes with many secondary benefits.
@@
 
-### Value extraction of uncertainty statistics
+#### Value extraction of uncertainty statistics
 Uncertainty visualisations are often evaluated based on how accurately [@Hullman2019] viewers can separately extract the estimate and its variance [@Kinkeldey2014]. 
@@
 
-### Trust, confidence, and risk aversion
+#### Trust, confidence, and risk aversion
 Trust is a by-product of displaying uncertainty so it is commonly measured in uncertainty evaluation studies [@Hullman2019]. 
@@
 
-### Questions that attempt to capture signal-suppression
+#### Questions that attempt to capture signal-suppression
 There is a collection of studies that seem to be aware of the issues behind using trust or value extraction to evaluate uncertainty visualisations. 
@@
 
-## Testing signal suppression
+### Testing signal suppression
 If we cannot ask direct questions about uncertainty, or measure the secondary effects of uncertainty, or ask indirect questions about the uncertainty; then how are we supposed to evaluate uncertainty visualisations?
@@
 
-### Comparing to hypothesis tests
+#### Comparing to hypothesis tests
 The most obvious way to evaluate uncertainty visualisations is to compare the visualisations to statistical tests. 
@@
 
-### Qualitative Studies
+#### Qualitative Studies
 Alternatively visualisation research could shift away from the accuracy concept all together and ask questions that allow for open ended responses.
@@
 
-### Heuristics
+#### Heuristics
 While experiments that explicitly identify heuristics in current methods are not technically measuring signal-suppression, they are still a useful consideration to keep in mind when designing experiments for signal suppression.
@@
 
-# Conclusions and Future work
+## Conclusions and Future work
 This paper has identified gaps in the uncertainty visualisation literature that must be filled for the field to progress. 
@@
 
-# Acknowledgements
+## Acknowledgements
+
 The first author of this paper is supported in part by a scholarship from the the Australian Energy Market Operator. 
@@
 
-# Bibliography
+## Bibliography
 

### 2024-10-18 paper.tex
changed format to jasa and edited reference so that it would work

--- 2024-10-18 paper.qmd
@@
----
-title: "The Noisy Work of Uncertainty Visualisation Research: A Review"
-author: 
-  - name: Harriet Mason
-    url: https://harrietmason.netlify.app/
-    orcid: 0009-0007-4568-8215
-    email:  harriet.m.mason@gmail.com
-    affiliation:
-    - name: Monash University
-      department: Department of Econometrics and Business Statistics
-      city: Melbourne
-      country: Australia
-  - name: Dianne Cook
-    url: https://dicook.org
-    orcid: 0000-0002-3813-7155
-    email: dicook@monash.edu
-    affiliation:
-    - name: Monash University
-      department: Department of Econometrics and Business Statistics
-      city: Melbourne
-      country: Australia
-  - name: Sarah Goodwin
-    url: https://www.linkedin.com/in/smgoodwin/
-    orcid: 0000-0001-8894-8282
-    email: Sarah.Goodwin@monash.edu
-    affiliation:
-    - name: Monash University
-      department: Department of Human Centred Computing
-      city: Melbourne
-      country: Australia
-  - name: Emi Tanaka
-    url: https://emitanaka.org/
-    orcid: 0000-0002-1455-259X
-    email: Emi.Tanaka@anu.edu.au
-    affiliation: 
-    - name: The Australian National University
-      department: Biological Data Science Institute
-      city: Canberra
-      country: Australia
-  - name: Susan Vanderplus
-    url: https://srvanderplas.github.io
-    orcid: 0000-0002-3803-0972
-    email: susan.vanderplas@unl.edu
-    affiliation:
-    - name: University of Nebraska–Lincoln
-      department: Statistics Department
-      city:  Lincoln
-      country: United States
-bibliography: references.bib
-abstract: Uncertainty visualisation is quickly becomming a hot topic in information visualisation.
-Exisiting reviews in the field take the definition and purpose of an uncertainty visualisation to be self evident which results in a large amout of conflicting information.
-This conflict largely stems from a conflation between uncertainty visualisations designed for decision making and those designed to prevent false conclusions.
-We coin the term "signal suppression" to describe a visualisation that is designed for preventing false conclusions, as the approach demands that the signal (i.e.
-the collective take away of the estimates) is suppressed by the noise (i.e.
-the variance on those estimates).
-We argue that the current standards in visualisation suggest that uncertainty visualisations designed for decision making should not be considered uncertainty visualisations at all.
-Therefore, future work should focus on signal suppression.
-Effective signal suppression requires us to communicate the signal and the noise as a single "validity of signal" variable, and doing so proves to be difficult with current methods.
-We illustrate current approaches to uncertainty visualisation by showing how they would change the visual apprearance of a choropleth map.
-These maps allow us to see why some methods succeed at signal suppression, while others fall short.
-Evaluating visualisations on how well they perform signal suppression also proves to be difficult, as it involves measuring the effect of noise, a variable we typically try to ignore.
-We suggest authors use qualitative studies or compare uncertainty visualisations to the relevant hypothesis tests.
-date: last-modified
-toc: false
-number-sections: true
-format:
-  jasa-pdf:
-    keep-tex: true  
-    journal:
-      blinded: false
-  jasa-html: default
-fig-valign: bottom
-cap-location: bottom
-editor_options: 
-  chunk_output_type: console
----
-
-```{r}
-#| echo: false
-#| message: false
-#| warning: false
-
-# load Libraries
-library(tidyverse)
-# devtools::install_github("lydialucchesi/Vizumap")
-library(Vizumap)
-library(RColorBrewer)
-library(scales)
-library(sf)
-library(ggrepel)
-# devtools::install_github("UrbanInstitute/urbnmapr")
-library(urbnmapr)
-library(flextable)
-library(colorspace)
-library(rgeos)
-```
-
-{{< pagebreak >}}
-
-# Introduction
-
-From entertainment choices to news articles to insurance plans, the modern citizen is so inundated with information in every aspect of their life it can be overwhelming. 
-In the face of this overflow of information, tools that effectively reduce piles of information to simple and clear ideas become more valuable. 
-That is, we need tools that can sort the signal from the noise.
-
-Among these summary tools, information visualisations are one of the most powerful as they allow for quick and memorable communication that allow us to identify quirks in our data that we didn't know to look for.
-Datasets such as Anscombe's quartet [@anscombe] or the Datasaurus Dozen [@datasaurpkg] show a case where visual statistics highlight elements of the data that are invisible to the typical summary statistics.
-Something as simple as sketching a distribution before recalling statistics or making predictions can greatly increase the accuracy of those measures [@Hullman2018; @Goldstein2014].
-
-"Uncertainty visualisation" is a relatively new field in research. 
-Early mentions of uncertainty visualisation start to appear in the late 1980s [@Ibrekk1987], with geospatial information visualisation literature from the early 1990s declaring this to be essential aspect of any information display [@MacEachren1992; @Carr1992].
-@fig-ibrekk depicts an example of the uncertainty visualisations discussed in these early papers. 
-Despite kicking off the field, these papers did not define uncertainty visualisation.
-This has led to a lack of consensus on what it means for a graphic to visualise uncertainty, an issue we will return to later.
-Therefore, while the field is considered to be quite new, many of the graphics used for uncertainty visualisation have been around for much longer. 
-For example, box plots and histograms display variation which becomes synonymous with uncertainty when we are using them to depict the variation of an estimate. 
-Today, there are an abundance of publications on the topic which makes it timely to construct a review of the field.
-That is, now that there is an overwhelming amount of information, it is valuable to distil it into simple facts. 
-In fact, there have already been several reviews published but a central piece of discussion is missing.
-
-```{r}
-#| echo: false
-#| message: false
-#| warning: false
-#| label: fig-ibrekk
-#| fig-cap: "A replication of the uncertainty visualisations shown by @Ibrekk1987 in one of the earliest uncertainty visualisation experiments.
-Several visualisation methods that are now unpopular (such as the pie chart) are used throughout this paper."
-#| fig-subcap: 
-#|   - "Picture 1"
-#|   - "Picture 2"
-#|   - "Picture 3"
-#|   - "Picture 4"
-#|   - "Picture 5"
-#|   - "Picture 6"
-#|   - "Picture 7"
-#|   - "Picture 8"
-#|   - "Picture 9"
-#| layout-ncol: 3
-
-# Generate data
-set.seed(1)
-x=rnorm(1000, 8, 4)
-ib_data <- tibble(x=ifelse(x<0, -x, x))
-
-# Picture 1
-p1 <- ib_data |>
-  summarise(avg = mean(x),
-          conf_95a = quantile(x, probs=c(0.025)),
-          conf_95b = quantile(x, probs=c(0.975))) |>
-  ggplot(aes(y="NA")) +
-  geom_point(aes(x=avg)) +
-  geom_errorbar(aes(xmin = conf_95a, xmax = conf_95b), width = 0.1) +
-  scale_x_continuous(name = "INCHES OF SNOW",
-                     breaks=seq(0,19),
-                     labels= ggplot2:::interleave(as.character(c(seq(0,18, 2), 19)), rep("", 11))[c(0:19, 21)],
-                     limits=c(0,19)) +
-  theme_classic() +
-  theme(axis.line.y=element_blank(),
-        axis.title.y = element_blank(),
-        axis.text.y = element_blank(),
-        axis.ticks.y = element_blank(),
-        aspect.ratio=1/10)
-
-# Picture 2
-p2 <- ib_data |>
-  mutate(x = ifelse(x>18, 18, x),
-         binx = cut(x, breaks=seq(0,18,2))) |>
-  group_by(binx) |>
-  summarise(n = n()) |>
-  mutate(Probability = n / sum(n)) |>
-  ggplot(aes(x=binx, y=Probability)) +
-  geom_col(fill="black", colour="white") +
-  scale_x_discrete(name = "INCHES OF SNOW",
-                   labels= paste0(seq(0,16,2), sep = "-", seq(2,18,2))) +
-  scale_y_continuous(breaks = seq(0.00, 0.25, 0.05)) +
-  theme_classic() + 
-  theme(aspect.ratio=0.33)
-
-# Picture 3
-p3 <- ib_data |>
-  mutate(x = ifelse(x>18, 18, x),
-         binx = cut(x, 
-                     breaks=seq(0,18,2), 
-                     labels= paste0(seq(0,16,2), sep = "-", seq(2,18,2)))) |>
-  group_by(binx) |>
-  summarise(n = n()) |>
-  mutate(Probability = n / sum(n),
-         csum = rev(cumsum(rev(Probability))), 
-         pos = Probability/2 + lead(csum, 1),
-         pos = if_else(is.na(pos), Probability/2, pos)) |>
-  ggplot(aes(x="", y=Probability, fill=binx)) +
-  geom_bar(stat="identity", width=1) +
-  geom_text_repel(aes(y = pos, label = paste0(round(Probability*100), sep="", "%")),
-                   size = 3, nudge_x = 0.6, show.legend = FALSE, segment.color = 'transparent') +
-  #geom_label(aes(label = paste0(round(Probability*100), sep="", "%")),
-  #          position = position_stack(vjust = 0.5)) + 
-  scale_fill_grey() +
-  coord_polar("y", start=0) +
-  labs(fill = "INCHES OF SNOW") + 
-  theme_void() + 
-  theme(aspect.ratio=1)
-
-# Picture 4
-p4 <- ib_data |>
-  ggplot(aes(x=x)) +
-  geom_density() + 
-  scale_x_continuous(name = "INCHES OF SNOW",
-                     breaks = seq(0,20,2),
-                     labels= paste0(seq(0,20,2))) +
-  scale_y_continuous(name = "Probability density",
-                     breaks = seq(0.00, 0.20, 0.02)) +
-  theme_classic() + 
-  theme(aspect.ratio=0.33)
-
-# Picture 5
-p5 <- ib_data |>
-  ggplot(aes(y="", x=x)) +
-  geom_violin() + 
-  scale_x_continuous(name = "INCHES OF SNOW",
-                     breaks = seq(0,20,2),
-                     labels= paste0(seq(0,20,2)),
-                     limits=c(0,20)) +
-  theme_classic() + 
-  theme(axis.line.y=element_blank(),
-        axis.title.y = element_blank(),
-        axis.text.y = element_blank(),
-        axis.ticks.y = element_blank(),
-        aspect.ratio=0.4)
-
-# Picture 6
-set.seed(1)
-x=rnorm(5000, 8, 4)
-ib_data2 <- tibble(x=ifelse(x<0, -x, x)) |>
-  mutate(x=ifelse(x>=18, 18-rexp(5000,rate=0), x))
-p6 <- ib_data2 |>
-  ggplot(aes(y="", x=x)) +
-  geom_jitter(size=0.05) + 
-  scale_x_continuous(name = "INCHES OF SNOW",
-                     breaks = seq(0,20,2),
-                     labels= paste0(seq(0,20,2)),
-                     limits=c(0,20)) +
-  theme_classic() + 
-  theme(axis.line.y=element_blank(),
-        axis.title.y = element_blank(),
-        axis.text.y = element_blank(),
-        axis.ticks.y = element_blank(),
-        aspect.ratio=0.1)
-
-# Picture 7
-p7 <- ib_data2 |>
-  arrange(x) |>
-  mutate(group = rep(1:50, each=100))|>
-  group_by(group) |>
-  summarise(x = max(x, na.rm=TRUE)) |>
-  add_row(group=c(0,51), x = c(0,20)) |>
-  ggplot(aes(x=x)) +
-  geom_linerange(ymin = 0.1, ymax = 1) + 
-  geom_linerange(y=1, xmin = -0.03, xmax = 20.03)+ 
-  geom_linerange(y=0.1, xmin = -0.03, xmax = 20.03)+ 
-  scale_x_continuous(name = "INCHES OF SNOW",
-                     breaks = seq(0,20,2),
-                     labels= paste0(seq(0,20,2)),
-                     limits=c(0,20)) +
-  scale_y_continuous(limits=c(0,1)) + 
-  theme_classic() + 
-  theme(axis.line.y=element_blank(),
-        axis.title.y = element_blank(),
-        axis.text.y = element_blank(),
-        axis.ticks.y = element_blank(),
-        aspect.ratio=0.1)
-
-# Picture 8
-p8 <- ib_data |>
-  reframe(x = quantile(x, probs=c(0.25, 0.50, 0.75)))|>
-  add_row(x = c(0,20)) |>
-  arrange(x) |>
-  mutate(quantile = c("min", "q1", "med", "q3", "max")) |>
-  pivot_wider(names_from = quantile, values_from = x) |>
-  ggplot(aes(y="")) +
-  #geom_point(aes(x=med)) +
-  geom_errorbar(aes(y="", xmin = min, xmax = max), width = 0.2) +
-  geom_crossbar(aes(y="", x=med, xmin = q1, xmax = q3), width = 0.5) +
-  scale_x_continuous(name = "INCHES OF SNOW",
-                     breaks=seq(0,20),
-                     labels= ggplot2:::interleave(as.character(c(seq(0,20, 2))), rep("", 11))[1:21],
-                     limits=c(0,20)) +
-  theme_classic() +
-  theme(axis.line.y=element_blank(),
-        axis.title.y = element_blank(),
-        axis.text.y = element_blank(),
-        axis.ticks.y = element_blank(),
-        aspect.ratio=1/10)
-
-# Picture 9
-p9 <- ib_data |>
-  ggplot(aes(x)) +
-  stat_ecdf(geom = "step") +
-  scale_x_continuous(name = "INCHES OF SNOW",
-                     breaks=seq(0,20),
-                     labels= ggplot2:::interleave(as.character(c(seq(0,20, 2))), rep("", 11))[1:21],
-                     limits=c(0,20)) +
-  scale_y_continuous(name = "Cumulative probability",
-                     breaks=seq(0,1,0.1),
-                     labels= seq(0,1,0.1),
-                     limits=c(0,1)) +
-  theme_classic() +
-  theme(aspect.ratio=4/10)
-
-# Display Plots
-p1
-p2
-p3
-p4
-p5
-p6
-p7
-p8
-p9
-```
-
-Reviews on uncertainty visualisation rarely offer tried and tested rules for effective uncertainty visualisation, but rather, they comment on the *difficulties* faced when trying to summarise the field.
-@Kinkeldey2014 found most experimental methods to be ad hoc, with no commonly agreed upon methodology, formalisations, or greater goal of describing general principals. 
-@Hullman2016 noticed there is a serious noise issue in the data coming from uncertainty visualisation experiments. 
-She commented on the prevalence of confounding variables that make it unclear as to what exactly caused a subjects poor performance on a set of particular questions.
-Mistakes due to misunderstanding visualisations, misinterpreting questions, and incorrectly applying heuristics are all combined into a single error value.
-@Spiegelhalter2017 commented that different plots are good for different things, and disagreed with the goal of identifying a universal best plot for all people and circumstances.
-@Griethe2006 did not identify common themes, but instead listed the findings and opinions of a collection of papers. 
-@uncertchap2022 summarised several cognitive effects that repeatedly arise in uncertainty visualisation experiments, however these effects were each discussed in isolation as a list of considerations an author might make rather than an overarching theory of rules for effective uncertainty visualisation.
-While these reviews are thorough in scope, none discuss how the existing literature contribute to the broader goal of uncertainty visualisation. 
-The problem faced by the literature is easily summarised with a famous quote by Henri Poincaré.
-
-> "Science is built up of facts, as a house is built of stones; but an accumulation of facts is no more a science than a heap of stones is a house." - Henri Poincaré (1905)
-
-That is to say, despite the wealth of reviews, the field of uncertainty visualisation remains a heap of stones. 
-There is a mountain of work that identifies common heuristics found in uncertainty visualisations, evaluate competing plot designs, or start a theoretical discussion on a niche aspect of the field.
-This is important work that needs to be done, but each of these papers offers up their own bespoke motivation and methodology, with little reference to the uncertainty visualisation papers outside their periphery.
-This becomes even more difficult to manage when these studies are in conflict. 
-The field is in desperate need of a unifying theory that can tie this swath of research together.
-This review attempts to address this issue by offering a novel perspective on the uncertainty visualisation problem.
-That is, we are going to use the wealth of established stone to construct the foundations on which we can build a house.
-
-This review is broken into several parts that each reflect a different approach to uncertainty visualisation. 
-First, we look at graphics that ignore uncertainty entirely and discuss why uncertainty should be included at all. 
-Second, we look at methods that consider uncertainty to be just another variable and discuss the characteristics of uncertainty that make it a unique visualisation problem. 
-Third, we look at methods that explicitly combine our estimate and its uncertainty and discuss the limitations of these approaches. 
-Fourth, we discuss methods that implicitly include uncertainty by depicting a sample in place of an estimate. 
-Finally, we discuss how uncertainty visualisations can be effectively evaluated.
-When discussing each of these methods, we will repeatedly return to the *purpose* of uncertainty visualisation and the effectiveness of each approach in fulfilling that purpose.
-
-## Spatial example
-There are far too many uncertainty visualisations to exhaustively discuss them all. 
-Instead we focus on the changes made to a single plot, the choropleth map. 
-Due to the field's origins and focus in geospatial information visualisation, there have been a large number of suggested variations on the choropleth map that allow authors to include uncertainty. 
-Utilising a single example will help isolate the ideas we are trying to convey. 
-Therefore, it is important to remember that even though we focus our discussion on the choropleth map, the theoretical approach we outline in this review is useful to all uncertainty visualisations regardless of their application.
-Additionally, our examples focus on incorporating uncertainty through colour manipulation, as that is the key visual channel used in a choropleth map.
-However, the methods we discuss go beyond variations in a colour palette.
-Even though they are not explicitly shown, visualisations that depict uncertainty using layers such as position or shape, and more complicated graphics that incorporate animation or interactivity are also within the scope of this review.
-We will use the choropleth maps as a tool to clearly highlight the costs and benefits of each approach.
-
-@fig-data shows the first six rows and the geographical boundaries of our data set. 
-The temperature variable was generated using a sine wave, that is $Temperature_i = 29-2\cdot{|Latitude_i - \sin{2 \cdot Longitude_i}|}$ where the $Longitude$ and $Latitude$ are the longitude and latitude of the county's centroid scaled to a standard normal distribution.
-Each county's variance is independently, randomly sampled from a uniform distribution. 
-The low standard errors are drawn from a $U_{[0,1]}$ distribution, while the high standard errors are drawn from a $U_{[1,2]}$ distribution. 
-As we are dealing with an average, the sampling distribution would be approximately normal by the central limit theorem, so each county temperature estimate is assumed to come from a $N(Temp_i, SE_{case,i})$.
-This is the data we will be using in our spatial uncertainty examples for the rest of the paper.
-
-```{r}
-#| eval: false
-#| echo: false
-# Get map data: do this once and save
-my_map_data <- get_urbn_map("counties", sf = TRUE) |>
-  filter(state_name=="Iowa")
-save(my_map_data, file="data/my_map_data.rda")
-```
-
-```{r}
-#| echo: false
-#| message: false
-#| warning: false
-#| label: fig-data
-#| fig-cap: "The first 5 observations of the data used for the spatial uncertainty examples along with the boundaries of each county.
-The map boundaries are the Iowa county boundaries, however the 'temperature' data is not representative of the average temperature in Iowa.
-The temperature and standard error represent the average of the daily high temperature and the standard error of that average respectively."
-#| fig-subcap: 
-#|   - "Data Table"
-#|   - "Map Boundaries"
-#| layout-nrow: 1
-#| cap-location: "bottom"
-load("data/my_map_data.rda")
-
-# seed for sampling
-set.seed(1997)
-
-# Get map data
-my_map_data <- get_urbn_map("counties", sf = TRUE) |>
-  filter(state_name=="Iowa")
-centroids <- as_tibble(gCentroid(as(my_map_data$geometry, "Spatial"), byid = TRUE))
-my_map_data$cent_long <- centroids$x
-my_map_data$cent_lat <- centroids$y
-n <- dim(my_map_data)[1]
-
-# Make palettes
-longpal <- rev(sequential_hcl(13, palette = "YlOrRd"))
-basecols <- longpal[3:10]
-breaks <- 21:29 
-breakslong <- 18:32 
-names(basecols) <- seq(8)
-names(longpal) <- -1:11
-
-my_map_data <- my_map_data |>
-  mutate(temp = 29 - 2*abs(scale(cent_lat) - sin(2*(scale(cent_long)))[,1]), # trend
-         highvar = runif(n, min=2, max=4), # high variance
-         lowvar = runif(n, min=0, max=2), # low variance
-         count_id = row_number()) |>
-  pivot_longer(cols=highvar:lowvar, names_to = "variance_class", values_to = "variance") |>
-  # add bivariate classes to data
-  mutate(bitemp = cut(temp, breaks=breaks, labels=seq(8)),
-         bivar = cut(variance, breaks=0:4, labels=seq(4)),
-         biclass = paste(bitemp, bivar, sep="-"))|>
-  mutate(highlight = ifelse(count_id <= 5, TRUE, FALSE))
-
-# Make nice example table
-example_table <- my_map_data |>
-  mutate(variance = sqrt(variance)) |>
-  select(c(count_id, county_name, temp, variance_class, variance)) |>
-  as_tibble()|>
-  pivot_wider(id_cols=c(count_id, county_name, temp,), 
-              names_from = variance_class, 
-              values_from = variance) |>
-  head(5)|>
-  flextable() |>
-  set_caption(caption = "Average Daily High Temperatures of Iowa Counties") |>
-  add_header_row(colwidths = c(3, 2),
-                 values = c("", "Standard Error")) |>
-  colformat_double(digits = 2) |>
-  set_header_labels(count_id = "ID", 
-                    county_name  = "County",
-                    temp = "Average Temperature (°C)",
-                    highvar = "High",
-                    lowvar = "Low") |>
-  add_footer_row(values = rep("..."), colwidths = 5) |>
-  theme_vanilla() |>
-  vline(i=c(1,2), j=3, part="header")|>
-  align(align = "left", part = "all") |>
-  bg(j = "temp",
-    bg = col_numeric(palette = brewer.pal(8, name = "Oranges"),
-                     domain = c(21, 30)),
-    part = "body"
-  ) |>
-  bg(j = c("highvar", "lowvar"),
-    bg = col_numeric(palette = brewer.pal(8, name = "Greens"),
-                     domain = c(0, 3)),
-    part = "body"
-  )
-
-# make blank map 
-example_map <- my_map_data |>
-  filter(variance_class=="lowvar") |>
-  ggplot() +
-  geom_sf(aes(geometry = geometry, fill=highlight)) + 
-  scale_fill_manual(values=c("white", "#fbfba2")) +
-  geom_text(data=filter(my_map_data, highlight==TRUE), aes(x=cent_long, y=cent_lat, label=count_id), size=3) +
-  theme_void() +
-  theme(legend.position = 'none')
-
-plot(example_table)
-
-example_map
-
-```
-
-# Ignoring uncertainty
-A good place to start might be at a deceptively straight forward question, why should we include uncertainty at all?
-
-## The choropleth map
-@fig-choropleth depicts a choropleth map of the counties of Iowa. 
-Each of these counties are coloured according to an estimate of average daily temperature that was generated so that the values followed a clear spatial trend. 
-The variance of these estimates were simulated such that the trend accounts for most of the variance in the plot in the low variance case (so we should expect the trend to be visible) while in the high variance, there is more variance within each county than between all the counties, so we should expect it to (at least in some capacity) overwhelm the spatial trend.
-Is this aspect of the data and the spatial data it communicates clear in in the map? 
-Is the strength of the trend communicated through the visualisation?
-
-```{r}
-#| echo: false
-#| message: false
-#| warning: false
-#| label: fig-choropleth
-#| fig-cap: "Two choropleth maps that depict the counties of Iowa where each country coloured acording to a simulated average temperature.
-Both maps depict a spatial trend, where counties closer to the centre of the map are hotter than counties on the edge of the map.
-In the low variance condition, the trend accounts for most of the variation in the data, in the high variance case, the variance on the temperature estimate accounts for most of the variance.
-This distinction is not clear in the map as they both appear identical.
-The high variance condition displays a spatial trend that could simply be spurious, which means the plot is displaying a false conclusion."
-#| fig-subcap: 
-#|   - "Low Variance Data"
-#|   - "High Variance Data"
-#|   - "Choropleth Palette"
-#| layout-ncol: 3
-#| layout-valign: "bottom"
-#| cap-location: "bottom"
-
-# Choropleth Map
-p1a <- my_map_data |>
-  filter(variance_class=="lowvar") |>
-  ggplot() +
-  geom_sf(aes(fill = bitemp, 
-              geometry = geometry), colour=NA) + 
-  scale_fill_manual(values = basecols) +
-  #scale_fill_gradientn(colours = basecols, 
-  #                     values=breaks/limits[2],
-  #                     limits=limits) +
-  theme_void() + 
-  theme(legend.position = "none")
-
-p1b <- p1a %+% filter(my_map_data, variance_class=="highvar")
-
-show_pal <- function (colours, borders = NULL, cex_label = 1, ncol = NULL, myxlab, breaks, textnudge, xlabx, xlaby, tsize=1.2) {
-  # Set dimensions of palette
-  n <- length(colours)
-  ncol <- ncol %||% ceiling(sqrt(length(colours)))
-  nrow <- ceiling(n/ncol)
-  # make matrix with null values (if not full)
-  colours <- c(colours, rep(NA, nrow * ncol - length(colours)))
-  colours <- matrix(colours, ncol = ncol, byrow = TRUE)
-  # set graphical parameters (?)
-  old <- par(pty = "s", mar = c(0, 0, 0, 0))
-  on.exit(par(old))
-  size <- max(dim(colours))
-  plot(c(0, size), c(0, -size), type = "n", xlab = "", ylab = "", 
-      axes = FALSE)
-  rect(col(colours) - 1, -row(colours) + 1, col(colours), -row(colours), 
-       col = colours, border = borders)
-  text(c(0,col(colours)) + textnudge, -c(1,row(colours))-0.25, breaks, 
-              cex = 1, col = "black")
-  text(xlabx, xlaby, myxlab ,cex = tsize, col = "black")
-}
-
-p1a
-p1b
-show_pal(basecols, ncol=8, borders=NA, myxlab = "Temperature", breaks = 21:29, textnudge = c(0.2, 0.1, 0,0,0,0,0,-0.1,-0.2), xlabx= 4, xlaby=-1.75, tsize=1.5)
-
-```
-
-## Signal-suppression
-Uncertainty visualisation is required for transparency. 
-The two choropleth maps that appear to be identical in @fig-choropleth highlight the issues with simply electing to ignore uncertainty. 
-This sentiment appears frequently in the uncertainty visualisation literature.
-Some authors suggest uncertainty is important to include as it communicates the legitimacy (or illegitimacy) of the conclusion drawn from visual inference [@Correll2014; @Kale2018; @Griethe2006]. 
-Some authors have said that uncertainty should be included to degree of confidence or trust in the data [@Boukhelifa2012; @Zhao2023]. 
-Some authors directly connect uncertainty visualisation to hypothesis testing as it ensures the validity of a statement [@Hullman2020a; @Griethe2006], but allows for a proportional level of trust that is more detailed than the binary results of a hypothesis test [@Correll2014; @Correll2018].
-Some authors even go so far as to claim that failing to include uncertainty is akin to fraud or lying [@Hullman2020a; @Manski2020].
-
-This consensus leads us to understand that uncertainty visualisation is motivated by the need for a sort of "visual hypothesis test". 
-A successful uncertainty visualisation would act as a "statistical hedge" for any inference we make using the graphic. 
-Since the purpose of a visualisation is to give a quick gist of the information [@Spiegelhalter2017], this hedging needs to be communicated visually without the need for complicated calculations. 
-If we refer to the conclusion we draw from a graphic to be its signal and the variance that makes this signal harder to identify as the "noise", we can summarise the above information into three key requirements.
-A good uncertainty visualisation needs to:
-
-  1) Reinforce justified signals to encourage confidence in results
-  2) Hide signals that are overwhelmed by noise to prevent unjustified conclusions
-  3) Perform tasks 1) and 2) in a way that is proportional to the level of confidence in those conclusions.
-
-As @fig-choropleth showed, visualisations that are unconcerned with uncertainty have no issue showing justified signals, but struggle with the display of unjustified signals. 
-Therefore, we call this approach to uncertainty visualisation as "signal-suppression" since it primarily differentiates itself from from the normal "noiseless" visualisation approach through criteria (2).
-That is, the main difference between an uncertainty visualisation and a normal visualisation is that an uncertainty visualisation should prevent us from drawing unjustified conclusions.
-
-## Uncertainty as a signal
-Uncertainty visualisation is not only motivated by signal-suppression, and we would be remiss if we did not mention these alternative approaches. 
-Some authors claim the purpose of uncertainty is to improve decision making [@Ibrekk1987; @uncertchap2022; @Hullman2016; @Cheong2016; @Boone2018; @Padilla2017]. 
-Other authors do not describe uncertainty as important for decision making, but rather explicitly state it as a variable of importance in of itself [@Blenkinsop2000]. 
-While uncertainty can provide useful information in decision making, it is important to recognise that the "uncertainty" in these cases is not acting as uncertainty at all. 
-It is acting as signal.
-
-This is obvious for the cases where we are explicitly interested in the variance or error, as we are literally trying to draw conclusions about an statistic that is used to describe uncertainty. 
-The same is true for visualisations made for decision making, but it is less overt. 
-This is easiest to understand with an example.
-
-Imagine you are trying to decide if you want to bring an umbrella with you to work. 
-An umbrella is annoying to bring with you, so you only want to pack it if the chance of rain is greater than 10%. 
-Unfortunately, your weather prediction app only provides you with the predicted daily rainfall. 
-Therefore, your decision will be improved with the inclusion of uncertainty. 
-This is *not* because uncertainty in general is important for decision making, but because it gives you the tools required to calculate the *actual* statistic you are basing your decision on. 
-In this sense, uncertainty is no more special to decision making than weight is special to a body mass index calculation.
-
-This means the uncertainty visualisations that would perform the best in decision making would simply display the uncertainty statistic we are interested in, such as the variance, or probability of an event, using existing visualisation principles.
-This is precisely what we observe in the literature. 
-@fig-exceed depicts an exceedance probability map that was designed as an alternative to the choropleth map to improve decision making under uncertainty [@Kuhnert2018; @Lucchesi2021]. 
-A keen viewer may notice that the exceedance probability map is actually just a choropleth map, only the statistic being displayed has changed. 
-We are not sure it is productive to categorise this visualisation as an uncertainty visualisation.
-
-```{r}
-#| echo: false
-#| message: false
-#| warning: false
-#| label: fig-exceed
-#| fig-cap: "An exceedance probability map that depict the counties of Iowa where each country coloured acording to the probability that the average temperature exceeds 27.
-This map is a choropleth map where the variable of interest is a probability."
-#| fig-subcap: 
-#|   - "Low Variance Data"
-#|   - "High Variance Data"
-#|   - "Exceedance Probability Map Palette"
-#| layout-ncol: 3
-
-# quantile
-prob_breaks <- seq(-0.1,1.1, length.out=9)
-exeed_data <- my_map_data |> 
-  as_tibble() |>
-  mutate(xprob = 1- pnorm(27, mean=temp, sd=sqrt(variance))) |>
-  mutate(xprob = cut(xprob, breaks=prob_breaks, labels=seq(8)))
-
-# Exceed Prob Map
-p2a <- exeed_data |>
-  filter(variance_class=="lowvar") |>
-  ggplot() +
-  geom_sf(aes(fill = xprob, 
-              geometry = geometry), colour=NA) + 
-  scale_fill_manual(values = basecols) +
-  theme_void() + 
-  theme(legend.position = "none")
-
-p2b <- p2a %+% filter(exeed_data, variance_class=="highvar")
-
-p2a
-p2b
-show_pal(basecols, ncol=8, borders=NA, myxlab = "P(Temperature>27)", breaks = c(0, prob_breaks[2:8], 1), textnudge = c(0.2, 0.1, 0,0,0,0,0,-0.1,-0.2), xlabx= 4, xlaby=-1.75, tsize=1.2)
-```
-
-There seem to be two different definitions of uncertainty visualisation floating around in the literature. 
-The first considers *any* visualisation of error, variance, or probability to be an uncertainty visualisation. 
-The second believes an uncertainty visualisation is the output of a function that takes a normal visualisation as an input, and transforms it to include uncertainty information. 
-The former group believe the purpose of uncertainty visualisation to provide signal about a distribution, while the later believe it should act as noise to obfuscate a signal. 
-The lack of explicit distinction between these two motivations leaves the literature muddled and reviewers struggle to understand if uncertainty should be treated as a variable, as metadata, or as something else entirely [@Kinkeldey2014].
-This disagreement creates constant contradictions in what the literature considers to be an uncertainty visualisation. 
-For example @Leland2005 mentions that popular graphics, such as pie charts and bar charts omit uncertainty, and @Wickham2011 suggests their product plot framework, which includes histograms and bar charts, should be extended to include uncertainty.
-However, pie charts, bar charts and histograms have all been used in a significant number of uncertainty visualisation experiments [@Ibrekk1987; @Olston2002; @Zhao2023; @Hofmann2012]. 
-If you view an uncertainty visualisation as a function applied to an existing graphic, then you would not see a pie chart or bar chart as uncertainty visualisations.
-These charts are they are yet to have the uncertainty visualisation function applied to them. 
-If you view an uncertainty visualisation as any graphic that depicts a statistic then there are no limitations on which graphics can or cannot be uncertainty visualisations.
-
-When we use the term uncertainty visualisation to refer to graphics that simply communicate a variance or probability, we are classifying visualisations by the data they display, not their visual features.
-Graphics, just like statistics, are not defined by their input data. 
-A scatter plot that compares mean and a scatter plot that compares variances are both scatter plots. 
-Given that there is no special class of visualisation for *other* statistics (such the median or maximum) there is no reason to assume visualisations that simply depict a variance, error, or probability to be special.
-Some authors implicitly suggest that visualisations of variance or probability are differentiated due to the psychological heuristics involved in interpreting uncertainty [@Hullman2019].
-While it is true that heuristics lead people to avoid uncertainty [@Spiegelhalter2017], there is no evidence that this psychological effect translates to issues with the visual representation of uncertainty.
-Again, given that we do not make these same visual considerations for other variables that elicit distaste or irrational behaviour, there is no reason to assume this is what makes uncertainty visualisation so special.
-
-This leads us to the conclusion that the visualisations made for the purpose of displaying information about uncertainty statistics are not uncertainty visualisations. 
-These graphics are just normal information visualisations, and authors can follow existing principles of graphical design. 
-We focus on the perspective that uncertainty visualisation serves to obfuscate signal, and an uncertainty visualisation is a variation on an existing graphic that gives it the ability to suppress false signals.
-
-Of course, there is nothing wrong with explicitly visualising variance, error, bias, or any other statistic used to depict uncertainty as a signal. 
-Just like any other statistic, these metrics provide important and useful information for analysis and decisions. 
-However, there is no interesting visualisation challenge associated with these graphics, and they do not require any special visualisation techniques. 
-The uncertainty in these graphics are acting as a signal variable, and they should be treated as such.
-
-# Visualising uncertainty as a variable
-Upon hearing that uncertainty needs to be included for transparency, the solutions may seem obvious. 
-You may think "well, I will just add a dimension to my plot that includes uncertainty". 
-This is a reasonable approach. 
-The simplest way to add uncertainty to an existing graphic is to simply map uncertainty to an unused visual channel. 
-However, it is unclear if this approach is sufficient for our purposes.
-
-## The bivariate map
-@fig-bivariate depicts a variation of the choropleth map, where we have a two dimensional colour palette. 
-In this graphic, temperature is still mapped to hue, but the variance is included by utilising colour saturation. 
-While these two maps *do* look visually different (which was not the case in the choropleth map) the spatial trend is still clearly visible in both graphics. 
-This means the uncertainty *is technically* being communicated, however the main message of the graphic is still the spatial trend (that may not exist). 
-The graphic did not suppress the invalid signal, so it is not performing signal-suppression as we would like. 
-At this point, it might be reasonable to ask, why? 
-Why is including the uncertainty as a variable insufficient to achieve signal-suppression, and what changes should we make to ensure signal-suppression occurs?
-
-```{r}
-#| echo: false
-#| message: false
-#| warning: false
-#| label: fig-bivariate
-#| fig-cap: "A bivariate map that depict the counties of Iowa where each county is coloured acording to it's average daily temperature and the variance in temperature.
-This map is a choropleth map with a two dimensional colour palette where temperature is represented by colour hue, and variance is represented by colour saturation.
-Even though uncertainty has been added to the graphic the spatial trend is still clearly visible in the high variance case."
-#| fig-subcap: 
-#|   - "Low Variance Data"
-#|   - "High Variance Data"
-#|   - "Bivariate Palette"
-#| layout-ncol: 3
-#| layout-valign: "bottom"
-
-# Bivariate Map
-# Make bivariate palette
-# Function to devalue by a certain amount
-colsupress <- function(basecols, hue=1, sat=1, val=1) {
-    X <- diag(c(hue, sat, val)) %*% rgb2hsv(col2rgb(basecols))
-    hsv(pmin(X[1,], 1), pmin(X[2,], 1), pmin(X[3,], 1))
-}
-
-# recurvisely decrease value
-v_val = 0.5
-bivariatepal <- c(basecols,
-                  colsupress(basecols, sat=v_val),
-                  colsupress(colsupress(basecols, sat=v_val), sat=v_val),
-                  colsupress(colsupress(colsupress(basecols, sat=v_val), sat=v_val), sat=v_val))
-# establish levels of palette
-names(bivariatepal) <- paste(rep(1:8, 4), "-" , rep(1:4, each=8), sep="")
-
-# Bivariate maps
-p2a <- my_map_data |>
-  filter(variance_class=="lowvar") |>
-  ggplot() +
-  geom_sf(aes(fill = biclass, geometry = geometry), colour=NA) + 
-  scale_fill_manual(values = bivariatepal) +
-  theme_void() + 
-  theme(legend.position = "none")
-
-p2b <- p2a %+% filter(my_map_data, variance_class=="highvar")
-
-show_pal2 <- function (colours, borders = NULL, cex_label = 1, ncol = NULL, myxlab, myylab, breaks, breaks2, tsize1=1.2, tsize2=1.2) {
-  # Set dimensions of palette
-  n <- length(colours)
-  ncol <- ncol %||% ceiling(sqrt(length(colours)))
-  nrow <- ceiling(n/ncol)
-  # make matrix with null values (if not full)
-  colours <- c(colours, rep(NA, nrow * ncol - length(colours)))
-  colours <- matrix(colours, ncol = ncol, byrow = TRUE)
-  # set graphical parameters (?)
-  old <- par(pty = "s", mar = c(0, 0, 0, 0))
-  on.exit(par(old))
-  size <- max(dim(colours))
-  plot(c(-1.5, size), c(0, -size), type = "n", xlab = "", ylab = "", 
-      axes = FALSE)
-  rect(col(colours) - 1, -row(colours) + 1, col(colours), -row(colours), 
-       col = colours, border = borders)
-  text(c(0,col(colours)[nrow,]) + c(0.2, 0.1, 0,0,0,0,0,-0.1,-0.2) , -4.5, 
-       breaks, cex = 1, col = "black")
-  text(-0.25, -c(0,row(colours)[,ncol]) + c(-0.2, -0.1, 0, 0.1, 0.2),
-       breaks2, cex = 1, col = "black")
-  text(4, -5.5, myxlab ,cex = tsize1, col = "black")
-  text(x=-1.25,y=-2, myylab, srt=270, cex = tsize2, col = "black")
-}
-
-p2a
-p2b
-show_pal2(colours = bivariatepal, ncol=8, borders=NA, myxlab = "Temperature", myylab = "Variance", breaks = 21:29, breaks2 = 0:4)
-```
-
-## Why this approach may (or may not) work
-The difficulty in incorporating uncertainty into a visualisation is frequently mentioned but seldom explained. 
-For example @Hullman2016 commented that it is straightforward to show a value but it is much more complex to show uncertainty but did not explain why. 
-Many authors seem to believe uncertainty visualisation is a simple high-dimensional visualisation problem, as the difficulty comes from working out how to add uncertainty into already existing graphics [@Griethe2006].
-While this is part of the problem in uncertainty visualisation, it is not the complete picture. 
-@fig-bivariate makes it clear that simply including uncertainty as a variable is insufficient to perform signal-suppression. 
-If we cannot treat uncertainty the same as we would any other variable, how should we treat it? 
-We need to understand what uncertainty actually *is*, in order to understand how to integrate it into a visualisation.
-
-### It's a variable... it's metadata... it's uncertainty?
-Describing what uncertainty actually is, is surprisingly hard. 
-Most authors simply avoid the problem and describe the characteristics of uncertainty, of which there are plenty. 
-Often, uncertainty is split using an endless stream of ever changing boundaries, such as whether the uncertainty is due to true randomness or a lack of knowledge [@Spiegelhalter2017; @Hullman2016; @utypo], if the uncertainty is in the attribute, spatial elements, or temporal element of the data [@Kinkeldey2014], whether the uncertainty is scientific (e.g.
-error) or human (e.g.
-disagreement among parties) [@Benjamin2018], if the uncertainty is random or systematic [@Sanyal2009], statistical or bounded [@Gschwandtnei2016; @Olston2002], recorded as accuracy or precision [@Griethe2006; @Benjamin2018], which stage of the data analysis pipeline the uncertainty comes from [@utypo], how quantifiable the uncertainty is [@Spiegelhalter2017; @utypo], etc.
-There are enough qualitative descriptors of uncertainty to fill a paper, but, none of this is particularly helpful in understanding how to integrate it into a visualisation. 
-Rather than trying to define uncertainty by looking at the myriad ways in which it *does* appear in an analysis, we may find it easier to look at where it *does not*.
-
-Descriptive statistics describe our sample as it is and summarises large data down into an easy to swallow format. 
-Descriptive statistics are not seen as the primary goal of modern statistics, however, this was not always the case. 
-In 19th century England, *positivism* was the popular philosophical approach to science (positivists included famous statisticians such as Francis Galton and Karl Pearson). 
-Practitioners of the approach believed statistics ended with descriptive statistics as science must be based on actual experience and observations [@Otsuka2023]. 
-In order to make statements about population statistics, future values, or new observations we need to perform inference, which requires the assumption of the "uniformity of nature", that is, we need to assume that unobserved phenomena should be similar to observed phenomena [@Otsuka2023].
-Positivists believed referencing the unobservable was bad science. 
-In other words, these scientists embraced descriptive statistics due to the inherent certainty that came with them.
-Since uncertainty is non-existent in descriptive statistics, it is clear that uncertainty is a by-product of inference.
-
-This history lesson illustrates what uncertainty actually is. 
-At several stages in a statistical analysis, we will violate the uniformity of nature assumption. 
-Each of these violations will impact the statistic we have calculated and push it further from the population parameter we wish to draw inference on. 
-Uncertainty is the amalgamation of these impacts. 
-If we do not violate the uniformity of nature assumption at any point in our analysis, we do not have any uncertainty.
-
-This interpretation of uncertainty indicates that the uncertainty is not a variable of importance in of itself. 
-Uncertainty is metadata about our statistic that is required for valid inference.
-This means uncertainty should not be visualised by itself and we should seek to display signal and uncertainty together as a "single integrated uncertain value" [@Kinkeldey2014]. 
-This aspect of uncertainty visualisation makes it a uniquely difficult problem.
-
-### Visualising the "single integrated uncertain value"
-Typically, when making visualisations, we want the visual channels to be separable. 
-That is, we don't want the data represented through one visual channel to interfere with the others [@Smart2019]. 
-Mapping uncertainty and signal to separable channels allows them to be read separately, which does not align with the goal of communicating them as a single integrated channel. 
-Visualising uncertainty and signal separately allows the uncertainty information to simply be ignored, which is a pervasive issue in current uncertainty visualisation methods [@uncertchap2022]. 
-We can see this problem in @fig-bivariate, as it sends the message "this data has a spatial trend and the estimates have a large variance" as we read the signal and the uncertainty separately.
-
-This means effective uncertainty visualisation should be leveraging integrability. 
-That is, the visual channels of the uncertainty and the signal would need to be separately manipulable, but read as a single channel by the human brain. 
-While most visual aesthetics *are* separable, there are some variables that have been shown to be integrable, such as colour hue and brightness [@Vanderplas2020].
-When visualising uncertainty using its own visual channel, we can also consider visual semiotics and make sure to map uncertainty to intuitive visual channels, such as mapping more uncertain values to lighter colours [@Maceachren2012].
-Unfortunately relying on integrability may not give us the amount of control we want over our signal-suppression. 
-Without a strong understanding of how these visual channels collapse down into a single channel, relying on integrability could create unintended consequences such as displaying phantom signals or hiding justified signals.
-Additionally, multi-dimensional colour palettes can make the graphics harder to read and hurt the accessibility of the plots [@Vanderplas2015].
-
-There is another benefit to mapping uncertainty to saturation that is not directly related to integrability. 
-As the saturation decreases colours become harder to distinguish. 
-This means high uncertainty values are harder to differentiate than low uncertainty values. 
-We can leverage this implicit feature of colour value by transforming the visual feature space ourselves.
-
-# Combining uncertainty and signal in a transformed space
-Instead of hoping that uncertainty might collapse signal values into a single dimension, we can do some of that work ourselves. 
-As a matter of fact, some uncertainty visualisation authors already have.
-
-## Value Suppressing Uncertainty Palettes
-The Value Suppressing Uncertainty Palette (VSUP) [@Correll2018], was designed with the intention of preventing high uncertainty values from being extracted from a map.
-Since the palette was designed with the extraction of individual values in mind and it has only been tested on simple value extraction tasks [@Correll2018] or search tasks [@Ndlovu2023], it is unclear how effective the method is at suppressing broader insights such as spatial trends.
-
-@fig-vsup is a visualisation of the Iowa temperature data using a VSUP to colour the counties. 
-The low uncertainty case still has a visible spatial trend, while the spatial trend in the high uncertainty map has functionally disappeared. 
-This means the VSUP has successfully suppressed the spatial trend in the data.
-However, the spatial trend may not be the only signal of concern in our graphic.
-Now we must return to the original signal-suppression criteria and ask ourselves if they have all been met. 
-Are all the justified signals reinforced, while all the unjustified signals are suppressed? 
-Is a graphic that performs perfect signal-suppression even possible?
-
-```{r}
-#| echo: false
-#| message: false
-#| warning: false
-#| label: fig-vsup
-#| fig-cap: "A map made with a VSUP.
-The counties of Iowa are coloured acording to its average daily temperature and the variance in temperature.
-Similar to the bivariate map, temperature is mapped to hue while variance is mapped to saturation.
-Unlike the bivariance map, the colour space we are mapping our variables to has been transformed so that high variance estimates are harder to discern from each other.
-This map successfully reduces the visibility of the spatial trend in the high uncertainty case while maintaining the visibility of the spatial trend in the low uncertainty case."
-#| fig-subcap: 
-#|   - "Low Variance Data"
-#|   - "High Variance Data"
-#|   - "VSUP Palette"
-#| layout-ncol: 3
-#| layout-valign: "bottom"
-
-# VSUP
-# Function to combine colours for VSUP
-colourblend <- function(basecols, p_length, nblend) {
-    X <- rgb2hsv(col2rgb(unique(basecols)))
-    v1 <- X[,seq(1,dim(X)[2], 2)]
-    v2 <- X[,seq(2,dim(X)[2], 2)]
-    if("matrix" %in% class(v1)){
-      # hue issue wrap around pt 1
-      v3 <- (v1+v2)
-      v3["h",] <- ifelse(abs(v1["h",]-v2["h",])>0.5, v3["h",]+1, v3["h",])
-      v3 <- v3/2
-      # hue issue wrap around pt 2
-      v3["h",] <- ifelse(v3["h",]>=1 , v3["h",]-1 ,v3["h",])
-      hsv(rep(v3[1,], each=nblend), rep(v3[2,], each=nblend), rep(v3[3,], each=nblend))
-      } else {
-        v3 <- (v1+v2)
-        v3["h"] <- ifelse(abs(v1["h"]-v2["h"])>0.5, v3["h"]+1, v3["h"])
-        v3 <- v3/2
-        v3["h"] <- ifelse(v3["h"]>=1 , v3["h"]-1 ,v3["h"])
-        rep(hsv(h=v3[1], s=v3[2], v=v3[3]), p_length)
-        }
-}
-
-VSUPfunc <- function(basecols, p_length, nblend){
-  colourblend(colsupress(basecols, sat=0.5), p_length, nblend)
-}
-
-# VSUP
-p = length(basecols)
-VSUP <- c(basecols,
-          VSUPfunc(basecols, p, 2),
-          VSUPfunc(VSUPfunc(basecols, p, 2), p, 4),
-          VSUPfunc(VSUPfunc(VSUPfunc(basecols, p, 2), p, 4), p, 8))
-
-names(VSUP) <- paste(rep(1:8, 4), "-" , rep(1:4, each=8), sep="")
-
-# VSUP maps
-p3a <- my_map_data |>
-  filter(variance_class=="lowvar") |>
-  ggplot() +
-  geom_sf(aes(fill = biclass, geometry = geometry), colour=NA) + 
-  scale_fill_manual(values = VSUP) +
-  theme_void() + 
-  theme(legend.position = "none")
-
-p3b <- p3a %+% filter(my_map_data, variance_class=="highvar")
-
-p3a
-p3b
-show_pal2(colours = VSUP, ncol=8, borders=NA, myxlab = "Temperature", myylab = "Variance", breaks = 21:29, breaks2 = 0:4)
-
-```
-
-## What can and cannot be suppressed?
-The methods used by the VSUP bring to light a slight problem with uncertainty visualisation. 
-Specifically, uncertainty and the purpose of visualisation are somewhat at odds with one another. 
-There are two primary motivations behind visualisation, communication and exploratory data analysis (EDA). 
-Communication involves identifying a signal we want to communicate and designing a visualisation that best conveys that, while EDA involves creating a versatile visualisation and using it to extract several signals.
-If we are designing an uncertainty visualisation for communication then we can just suppress the specific signal we are seeking to communicate. 
-In the map example, we would consider @fig-vsup to be a success as the only signal we are concerned with is the spatial trend. 
-However, it is not uncommon for authors to express a desire for uncertainty visualisations that perform signal-suppression in visualisations made for EDA [@Sarma2024; @Griethe2006]. 
-For uncertainty visualisation for EDA to work, we would need to assume that suppressing individual estimates using their variance should naturally extend to broader suppression of plot level insights.
-Unfortunately, it is unclear how reliably this would work.
-
-### There is no uncertainty in EDA
-Earlier we established that uncertainty is a by-product of inference, which means without inference, there is no uncertainty. 
-Often EDA is used to give us an understanding of our data and identify which signals are worth pursuing. 
-In this sense, EDA is the visual parallel to descriptive statistics, as it is performed without an explicit hypothesis, which means there is no inference, and by extension, there is no uncertainty.
-
-Some authors recognise inference will always occur (in some shape or form) and believe uncertainty *should* be visualised but do not recognise *how* uncertainty would be visualised. 
-@Hullman2021 argued that there is no such thing as a "model-free" visualisation, therefore all visualisations require uncertainty as we are always performing inference. 
-While it is true that we can think of visualisations as containing implicit inferential properties, there are many potential inferences in any single visualisation. 
-This makes it a little difficult to ensure uncertainty is always included.
-For example, if we have a visualisation that shows an average, we would need to identify if the signal suppression should be performed using the sampling variance or the sample variance [@Hofman2020].
-The distribution we use depends on the inferential statistic but until the viewer chooses one and thinks about it (which isn't easily observable), the particular variety of uncertainty which would need to be displayed can't be calculated.
-This means the ideal uncertainty visualisation should not only meet the signal suppression requirements, but should also endeavor to be versatile enough to meet those requirements for all the signals displayed in the graphic.
-
-### The limitations of explicitly visualising uncertainty and signal
-The lack of versatility of the VSUP is easy to see with a simple example.
-
-Let's say we have a graphic that depicts a set of coefficients from a linear regression and the value of the coefficient is shown using a single colour. 
-We want to know "Which of these coefficients are different from 0?" as well as "Which of these coefficients are different from each other?". 
-To answer this question we do a series of $t$-tests on these estimates. 
-All of the individual $t$-tests of fail to reject the null hypothesis that the coefficients are different from 0. 
-We then make a visualisation that suppresses this signal and ensures that all of the estimates are visually indistinguishable from 0. 
-Next, we conduct two sample $t$-test and find that several of the values need to be visually distinguishable from each other. 
-The VSUP method must pick a single colour for each estimate, and these colours must be *either* visually distinguishable or indistinguishable from each other. 
-We cannot perform signal-suppression on both these signals simultaneously.
-
-This example highlights a fundamental problem with the VSUP that extends to the bivariate map as well. 
-When we blend these colours, we need to decide at what level of *uncertainty* to blend these colours together. 
-Even though the bivariate map does not explicitly combine colour values at certain variance levels, the mapping of variance to colour saturation does this implicitly. 
-That is, at certain saturation values the colours in a bivariate map are imperceptibly different and appear as though they are mapped to the same value.
-At this point, it is irrelevant whether or not the colours are technically different, they are the same colour in the human brain. 
-This is of course complicated by the fact that human colour perception varies at an individual level. 
-Some women are believed to have four different cone cells which allows them to perceive a greater range of colours, while others only have one or two types of cone cells and have colour deficiencies [@simunovic2010]
-For VSUP to function for all individuals, we must calibrate each plot to an individual's ability to perceive colour.
-
-If we only use a single colour to express each signal-suppressed statistic, we will always need to decide which signals we suppress and which we do not. 
-This issue has already been raised in the literature. 
-Which hypothesis are suppressed and which are not largely depends on the method used to combining colours in the palette [@Kay2019]. 
-The VSUP in @fig-vsup used the tree based method that was used by @Correll2018, but there are alternatives that are more appropriate for different hypothesis.
-
-Uncertainty visualisation for EDA would be possible if we designed a plot in such a way that suppressing individual estimates using their variance would naturally extend to broader suppression of plot level insights.
-This assumption is commonly made by visualisation researchers in normal visualisation experiments [@North2006]. 
-If we could express the statistic of a cell using multiple colours, this limitation may disappear entirely.
-
-# Implicitly Combining Uncertainty and Signal
-Rather than trying to figure out how to combine signal and uncertainty into a single colour, we can just display a sample instead and allow the viewer to extract *both* the estimate and the variance.
-
-## Pixel map
-@fig-pixel displays a pixel map [@Lucchesi2021], which is a variation of the choropleth map where each area is divided up into several smaller areas, each coloured using outcomes from the larger area's (i.e.
-the county's) average temperature sampling distribution.
-The spatial trend is clearly visible in the low variance case, but functionally disappears in the low variance case. 
-While the spatial trend is just barely visible in the high uncertainty case, it is much harder to see. 
-This means the graphic also achieves the third criteria for signal-suppression, i.e. our difficulty in seeing the distribution is proportional to the level of uncertainty in the graphic.
-
-```{r}
-#| echo: false
-#| message: false
-#| warning: false
-#| label: fig-pixel
-#| fig-cap: "A pixel map of the counties of Iowa.
-In this map, each county is broken up into several small areas and coloured according to a potential daily temperature, given the its average daily temperature and its sampling distribution.
-This results in each county being represented by a sample rather than a single value.
-In this graphic, we can clearly see the spatial trend in the low variance case, while the spatial trend is much harder to identify in the high variance case.
-"
-#| fig-subcap: 
-#|   - "Low Variance Data"
-#|   - "High Variance Data"
-#|   - "Pixel Map Palette"
-#| layout-ncol: 3
-#| layout-valign: "bottom"
-
-# Pixel Map
-
-# Low variance map
-my_map_data_a <- my_map_data |>
-  filter(variance_class == "lowvar") |>
-  mutate(my_id = seq(n),
-         error = variance)
-
-# quantile
-q_a <- my_map_data_a |> 
-  as.tibble() |>
-  mutate(bitemp=as.numeric(bitemp)) |>
-  with(data.frame(p0.05 = qnorm(0.05, mean=bitemp, sd=sqrt(variance)),
-                  p0.25 = qnorm(0.25, mean=bitemp, sd=sqrt(variance)), 
-                  p0.5 = qnorm(0.5, mean=bitemp, sd=sqrt(variance)), 
-                  p0.75 = qnorm(0.75, mean=bitemp, sd=sqrt(variance)), 
-                  p0.95 = qnorm(0.95, mean=bitemp, sd=sqrt(variance))))
-
-pixel_1a <- my_map_data_a |>
-  as.data.frame() |>
-  select(my_id, bitemp, error) |>
-  read.uv(estimate="bitemp", error="error")
-
-pixel_2a <- my_map_data_a |> as("Spatial")
-
-pix_a <- pixelate(pixel_2a, pixelSize = 70, id = "my_id")
-
-pmap_a <- build_pmap(data = pixel_1a, distribution = "discrete", pixelGeo = pix_a, id = "my_id", border = pixel_2a, q=q_a)
-
-p4a <- view(pmap_a) +
-  geom_path(
-      data = pmap_a$bord,
-      aes_string(x = 'long', y = 'lat', group = 'group'),
-      colour = "white"
-    )  +
-  scale_fill_gradientn(colours = longpal) +
-  scale_colour_gradientn(colours = longpal) +
-  theme(legend.position="none")
-
-# High variance
-my_map_data_b <- my_map_data |>
-  filter(variance_class == "highvar") |>
-  mutate(my_id = seq(n),
-         error = variance)
-
-q_b <- my_map_data_b |> 
-  as.tibble() |>
-  mutate(bitemp=as.numeric(bitemp)) |>
-  with(data.frame(p0.05 = qnorm(0.05, mean=bitemp, sd=sqrt(variance)),
-                  p0.25 = qnorm(0.25, mean=bitemp, sd=sqrt(variance)), 
-                  p0.5 = qnorm(0.5, mean=bitemp, sd=sqrt(variance)), 
-                  p0.75 = qnorm(0.75, mean=bitemp, sd=sqrt(variance)), 
-                  p0.95 = qnorm(0.95, mean=bitemp, sd=sqrt(variance))))
-
-pixel_1b <- my_map_data_b |>
-  as.data.frame() |>
-  select(my_id, bitemp, error) |>
-  read.uv(estimate="bitemp", error="error")
-
-pixel_2b <- my_map_data_b |> 
-  as("Spatial")
-
-pix_b <- pixelate(pixel_2b, pixelSize = 70, id = "my_id")
-
-pmap_b <- build_pmap(data = pixel_1b, distribution = "discrete", pixelGeo = pix_b, id = "my_id", border = pixel_2b, q=q_b)
-
-p4b <- view(pmap_b) +
-  geom_path(
-      data = pmap_b$bord,
-      aes_string(x = 'long', y = 'lat', group = 'group'),
-      colour = "white"
-    )  +
-  scale_fill_gradientn(colours = longpal) +
-  scale_colour_gradientn(colours = longpal) +
-  theme(legend.position="none")
-
-# Display plots and palettes
-p4a
-p4b
-show_pal(longpal, ncol=13, borders="white", myxlab = "Temperature", breaks = 19:32, textnudge = c(0.2, 0.1, 0,0,0,0,0,0,0,0,0,0,-0.1,-0.2), xlabx= 6.5, xlaby=-2.2)
-```
-
-It is clear that the pixel-map is not only suppressing the false information, but it is doing so by simulating *more* information. 
-The efficacy of this method means that visualisations of simulated samples pop up repeatedly in the literature, with examples including samples that are animated over time [@Hullman2015; @Blenkinsop2000], pixel-maps, and spaghetti time series plots.
-Not only does this method help readers understand the plot level gist, it is also unlikely to damage the viewers ability to extract individual estimates.
-Extracting global statistics, such as the mean or variance, from a sample can be done with relative ease, especially when those values are mapped to colour [@Franconeri2021]. 
-Additionally, @Ndlovu2023 found that participants applied the same methods they used for simple choropleth maps to complicated uncertainty maps even if that take away was invalid. 
-The pixel map can be read the same way as a choropleth map, which allows us to leverage that automatic response. 
-Therefore, the pixel map performs signal suppression, without sacrificing the viewers ability to extract general statistics, unless those statistics *should* be harder to extract due to the uncertainty in the value.
-So, what is it about @fig-pixel that allows it to be such a successful uncertainty visualisation?
-Is this a question we can easily answer?
-
-## Show me the data
-To thoroughly understand why pixel maps seem to successfully perform signal suppression, we would need a more thorough discussion than what can be offered here. 
-However, this does not prohibit us from discussing potential reasons for the graph's effectiveness. 
-There is a second type of uncertainty that the pixel map quietly communicates but we have not touched on yet, and this uncertainty is also conveyed by visualisations of our raw data. 
-That is, these graphics can show assumption violations.
-
-As we discussed in previous sections, we can consider uncertainty to be "the amalgamations of the impacts of violations to the assumption of the uniformity of nature". 
-It's not a definition that rolls off the tongue, but we can work with it.
-Thankfully, this definition of uncertainty aligns nicely with all the concepts that are included in the uncertainty umbrella. 
-Some works [@Hullman2018; @Maceachren2012; @Thomson2005] focus narrowly on specific terms with mathematical definitions, such as probability, confidence intervals, variance, error, or precision. 
-These works are only concerned with quantifying the final impact of uncertainty on our statistics. 
-That is, how large should the bound around our statistic be, such that our "true" statistic can be inferred. 
-Others [@Griethe2006; @Leland2005; @Pang1997; @Pham2009; @Boukhelifa2017] include broader loosely related elements, such as missing values, reliability, model validity, or source integrity. 
-These broader and harder to quantify concepts are concerned about potential sources of uncertainty, that is, they describe violations to the assumption of the uniformity of nature.
-
-What this means is that we have two types of uncertainty, but one is more "processed" than the other. 
-Quantifiable uncertainties are just assumption violations expressed as an effect on our final statistic (or true variation of our data). 
-The disconnect between quantifiable and unquantifiable uncertainty creates a huge problem for authors trying to visualise it.
-A survey of visualisation authors cited "not knowing how to calculate uncertainty" as one of the primary reasons they did not include it in visualisations [@Hullman2020a].
-
-There are two reasons we might leave our uncertainty as an assumption violation rather quantifying the effect. 
-The first reason is that we may be unable to translate the assumption violation to a quantifiable uncertainty. 
-There is no blanket rule that allows us to reliably quantify all uncertainty for every statistic, although some researchers have tangled with the idea. 
-For example, @Thomson2005 suggests a mathematical formula for *examples* of uncertainty, and information theory tries to quantify uncertainty using the idea of entropy, but they ignore the disconnect between the broad concept of uncertainty and what we can reliably quantify.
-Some authors don't believe that it is even possible to quantify all the assumption violations [@Spiegelhalter2017].
-
-The second reason to leave uncertainty as a potential violation of our assumptions, is that we might not know the final statistic we are seeking to calculate. 
-This is the case for visualisations made for EDA, and a large number of developments in EDA visualisation have been in displaying these difficult to quantify violations. 
-For example, @Tierney2023 builds upon the tidy data principles to allow users to handle missing values. 
-This includes data plots with a missing value "shadow" that allows visualisation authors to identify if the variables used in a plot have any structure in their missing values, which would contribute to uncertainty.
-
-With this understanding it becomes clear to see why uncertainty is tied to an endless string of examples in the data analysis pipeline. 
-Uncertainty examples include imputed data, model selection, inherent randomness, biased sampling, etc., not because these things *are* uncertainty, but because they *create* uncertainty when we perform inference.
-Whether or not these elements are relevant is highly dependent on what statistic you are trying to draw inference on, and by extension, the purpose of your visualisation.
-
-This relationship between uncertainty and the purpose of our analysis is littered throughout the literature. 
-Multiple authors have commented on the need to consider quantifying and expressing uncertainty at every stage of a project as the goal shapes every step of the analysis [@Kinkeldey2014; @Hullman2016; @Refsgaard2007].
-@Otsuka2023 suggested that the process of observing data to perform statistics is largely dependent on our goals, because the process of boiling real world entities down into probabilistic objects (or "probabilistic kind" as he puts it) depends on the relationship we seek to identify with our data.
-@Meng2014 commented what is kept as data and what is tossed away is determined by the motivation of an analysis and what was previously noise can be shown to become signal depending on the question we seek to answer.
-@Wallsten1997 argue that the best method for evaluating or combining subjective probabilities depends on the uncertainty the decision maker wants to represent and why it matters. 
-@Fischhoff2014 looks at uncertainty visualisation for decision making decides that we should have different ways of communicating uncertainty based off what the user is supposed to do with it.
-
-This makes it very difficult to move quantified uncertainty through the layers of our analysis, especially when designing a visualisation for EDA. 
-If we don't know what the final statistic is, we cannot quantify the effects of our assumptions. 
-Therefore, it is often the case that the best uncertainty visualisation is not an uncertainty visualisation at all, but simply the most accurate depiction of our raw data.
-This does not mean that visualising raw data will always prevent insignificant signal from getting through. 
-@Chowdhury illustrated how groups that appear linearly separable in a linear discriminant analysis (LDA) visualisation of the data can actually be the result of a LDA performed on too many variables, something that was not clear from the visualisation until the line-up protocol was implemented.
-However, just showing the data is a simple but effective option for uncertainty visualisation that seems to be largely overlooked.
+% Options for packages loaded elsewhere
+\PassOptionsToPackage{unicode}{hyperref}
+\PassOptionsToPackage{hyphens}{url}
+\PassOptionsToPackage{dvipsnames,svgnames,x11names}{xcolor}
+%
+\documentclass[
+  12pt]{article}
+
+\usepackage{amsmath,amssymb}
+\usepackage{iftex}
+\ifPDFTeX
+  \usepackage[T1]{fontenc}
+  \usepackage[utf8]{inputenc}
+  \usepackage{textcomp} % provide euro and other symbols
+\else % if luatex or xetex
+  \usepackage{unicode-math}
+  \defaultfontfeatures{Scale=MatchLowercase}
+  \defaultfontfeatures[\rmfamily]{Ligatures=TeX,Scale=1}
+\fi
+\usepackage{lmodern}
+\ifPDFTeX\else  
+    % xetex/luatex font selection
+\fi
+% Use upquote if available, for straight quotes in verbatim environments
+\IfFileExists{upquote.sty}{\usepackage{upquote}}{}
+\IfFileExists{microtype.sty}{% use microtype if available
+  \usepackage[]{microtype}
+  \UseMicrotypeSet[protrusion]{basicmath} % disable protrusion for tt fonts
+}{}
+\makeatletter
+\@ifundefined{KOMAClassName}{% if non-KOMA class
+  \IfFileExists{parskip.sty}{%
+    \usepackage{parskip}
+  }{% else
+    \setlength{\parindent}{0pt}
+    \setlength{\parskip}{6pt plus 2pt minus 1pt}}
+}{% if KOMA class
+  \KOMAoptions{parskip=half}}
+\makeatother
+\usepackage{xcolor}
+\setlength{\emergencystretch}{3em} % prevent overfull lines
+\setcounter{secnumdepth}{5}
+% Make \paragraph and \subparagraph free-standing
+\makeatletter
+\ifx\paragraph\undefined\else
+  \let\oldparagraph\paragraph
+  \renewcommand{\paragraph}{
+    \@ifstar
+      \xxxParagraphStar
+      \xxxParagraphNoStar
+  }
+  \newcommand{\xxxParagraphStar}[1]{\oldparagraph*{#1}\mbox{}}
+  \newcommand{\xxxParagraphNoStar}[1]{\oldparagraph{#1}\mbox{}}
+\fi
+\ifx\subparagraph\undefined\else
+  \let\oldsubparagraph\subparagraph
+  \renewcommand{\subparagraph}{
+    \@ifstar
+      \xxxSubParagraphStar
+      \xxxSubParagraphNoStar
+  }
+  \newcommand{\xxxSubParagraphStar}[1]{\oldsubparagraph*{#1}\mbox{}}
+  \newcommand{\xxxSubParagraphNoStar}[1]{\oldsubparagraph{#1}\mbox{}}
+\fi
+\makeatother
+
+\providecommand{\tightlist}{%
+  \setlength{\itemsep}{0pt}\setlength{\parskip}{0pt}}\usepackage{longtable,booktabs,array}
+\usepackage{calc} % for calculating minipage widths
+% Correct order of tables after \paragraph or \subparagraph
+\usepackage{etoolbox}
+\makeatletter
+\patchcmd\longtable{\par}{\if@noskipsec\mbox{}\fi\par}{}{}
+\makeatother
+% Allow footnotes in longtable head/foot
+\IfFileExists{footnotehyper.sty}{\usepackage{footnotehyper}}{\usepackage{footnote}}
+\makesavenoteenv{longtable}
+\usepackage{graphicx}
+\makeatletter
+\def\maxwidth{\ifdim\Gin@nat@width>\linewidth\linewidth\else\Gin@nat@width\fi}
+\def\maxheight{\ifdim\Gin@nat@height>\textheight\textheight\else\Gin@nat@height\fi}
+\makeatother
+% Scale images if necessary, so that they will not overflow the page
+% margins by default, and it is still possible to overwrite the defaults
+% using explicit options in \includegraphics[width, height, ...]{}
+\setkeys{Gin}{width=\maxwidth,height=\maxheight,keepaspectratio}
+% Set default figure placement to htbp
+\makeatletter
+\def\fps@figure{htbp}
+\makeatother
+
+\addtolength{\oddsidemargin}{-.5in}%
+\addtolength{\evensidemargin}{-1in}%
+\addtolength{\textwidth}{1in}%
+\addtolength{\textheight}{1.7in}%
+\addtolength{\topmargin}{-1in}%
+\makeatletter
+\@ifpackageloaded{caption}{}{\usepackage{caption}}
+\AtBeginDocument{%
+\ifdefined\contentsname
+  \renewcommand*\contentsname{Table of contents}
+\else
+  \newcommand\contentsname{Table of contents}
+\fi
+\ifdefined\listfigurename
+  \renewcommand*\listfigurename{List of Figures}
+\else
+  \newcommand\listfigurename{List of Figures}
+\fi
+\ifdefined\listtablename
+  \renewcommand*\listtablename{List of Tables}
+\else
+  \newcommand\listtablename{List of Tables}
+\fi
+\ifdefined\figurename
+  \renewcommand*\figurename{Figure}
+\else
+  \newcommand\figurename{Figure}
+\fi
+\ifdefined\tablename
+  \renewcommand*\tablename{Table}
+\else
+  \newcommand\tablename{Table}
+\fi
+}
+\@ifpackageloaded{float}{}{\usepackage{float}}
+\floatstyle{ruled}
+\@ifundefined{c@chapter}{\newfloat{codelisting}{h}{lop}}{\newfloat{codelisting}{h}{lop}[chapter]}
+\floatname{codelisting}{Listing}
+\newcommand*\listoflistings{\listof{codelisting}{List of Listings}}
+\makeatother
+\makeatletter
+\makeatother
+\makeatletter
+\@ifpackageloaded{caption}{}{\usepackage{caption}}
+\@ifpackageloaded{subcaption}{}{\usepackage{subcaption}}
+\makeatother
+
+\ifLuaTeX
+  \usepackage{selnolig}  % disable illegal ligatures
+\fi
+\usepackage[]{natbib}
+\bibliographystyle{agsm}
+\usepackage{bookmark}
+
+\IfFileExists{xurl.sty}{\usepackage{xurl}}{} % add URL line breaks if available
+\urlstyle{same} % disable monospaced font for URLs
+\hypersetup{
+  pdftitle={The Noisy Work of Uncertainty Visualisation Research: A Review},
+  pdfauthor={Harriet Mason; Dianne Cook; Sarah Goodwin; Emi Tanaka; Susan Vanderplus},
+  colorlinks=true,
+  linkcolor={blue},
+  filecolor={Maroon},
+  citecolor={Blue},
+  urlcolor={Blue},
+  pdfcreator={LaTeX via pandoc}}
+
+\begin{document}
+
+\def\spacingset#1{\renewcommand{\baselinestretch}%
+{#1}\small\normalsize} \spacingset{1}
+
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+
+\date{October 18, 2024}
+\title{\bf The Noisy Work of Uncertainty Visualisation Research: A
+Review}
+\author{
+Harriet Mason\\
+Department of Econometrics and Business Statistics, Monash University\\
+and\\Dianne Cook\\
+Department of Econometrics and Business Statistics, Monash University\\
+and\\Sarah Goodwin\\
+Department of Human Centred Computing, Monash University\\
+and\\Emi Tanaka\\
+Biological Data Science Institute, The Australian National University\\
+and\\Susan Vanderplus\\
+Statistics Department, University of Nebraska--Lincoln\\
+}
+\maketitle
+
+\bigskip
+\bigskip
+\begin{abstract}
+Uncertainty visualisation is quickly becomming a hot topic in
+information visualisation. Exisiting reviews in the field take the
+definition and purpose of an uncertainty visualisation to be self
+evident which results in a large amout of conflicting information. This
+conflict largely stems from a conflation between uncertainty
+visualisations designed for decision making and those designed to
+prevent false conclusions. We coin the term ``signal suppression'' to
+describe a visualisation that is designed for preventing false
+conclusions, as the approach demands that the signal (i.e.~the
+collective take away of the estimates) is suppressed by the noise
+(i.e.~the variance on those estimates). We argue that the current
+standards in visualisation suggest that uncertainty visualisations
+designed for decision making should not be considered uncertainty
+visualisations at all. Therefore, future work should focus on signal
+suppression. Effective signal suppression requires us to communicate the
+signal and the noise as a single ``validity of signal'' variable, and
+doing so proves to be difficult with current methods. We illustrate
+current approaches to uncertainty visualisation by showing how they
+would change the visual apprearance of a choropleth map. These maps
+allow us to see why some methods succeed at signal suppression, while
+others fall short. Evaluating visualisations on how well they perform
+signal suppression also proves to be difficult, as it involves measuring
+the effect of noise, a variable we typically try to ignore. We suggest
+authors use qualitative studies or compare uncertainty visualisations to
+the relevant hypothesis tests.
+\end{abstract}
+
+\newpage
+\spacingset{1.9} % DON'T change the spacing!
+
+\section{Introduction}\label{introduction}
+
+From entertainment choices to news articles to insurance plans, the
+modern citizen is so inundated with information in every aspect of their
+life it can be overwhelming. In the face of this overflow of
+information, tools that effectively reduce piles of information to
+simple and clear ideas become more valuable. That is, we need tools that
+can sort the signal from the noise.
+
+Among these summary tools, information visualisations are one of the
+most powerful as they allow for quick and memorable communication that
+allow us to identify quirks in our data that we didn't know to look for.
+Datasets such as Anscombe's quartet \citep{anscombe} or the Datasaurus
+Dozen \citep{datasaurpkg} show a case where visual statistics highlight
+elements of the data that are invisible to the typical summary
+statistics. Something as simple as sketching a distribution before
+recalling statistics or making predictions can greatly increase the
+accuracy of those measures \citep{Hullman2018, Goldstein2014}.
+
+``Uncertainty visualisation'' is a relatively new field in research.
+Early mentions of uncertainty visualisation start to appear in the late
+1980s \citep{Ibrekk1987}, with geospatial information visualisation
+literature from the early 1990s declaring this to be essential aspect of
+any information display \citep{MacEachren1992, Carr1992}.
+Figure~\ref{fig-ibrekk} depicts an example of the uncertainty
+visualisations discussed in these early papers. Despite kicking off the
+field, these papers did not define uncertainty visualisation. This has
+led to a lack of consensus on what it means for a graphic to visualise
+uncertainty, an issue we will return to later. Therefore, while the
+field is considered to be quite new, many of the graphics used for
+uncertainty visualisation have been around for much longer. For example,
+box plots and histograms display variation which becomes synonymous with
+uncertainty when we are using them to depict the variation of an
+estimate. Today, there are an abundance of publications on the topic
+which makes it timely to construct a review of the field. That is, now
+that there is an overwhelming amount of information, it is valuable to
+distil it into simple facts. In fact, there have already been several
+reviews published but a central piece of discussion is missing.
+
+\begin{figure}
+
+\begin{minipage}{0.33\linewidth}
+
+\centering{
+
+\includegraphics{paper_files/figure-pdf/fig-ibrekk-1.pdf}
+
+}
+
+\subcaption{\label{fig-ibrekk-1}Picture 1}
+
+\end{minipage}%
+%
+\begin{minipage}{0.33\linewidth}
+
+\centering{
+
+\includegraphics{paper_files/figure-pdf/fig-ibrekk-2.pdf}
+
+}
+
+\subcaption{\label{fig-ibrekk-2}Picture 2}
+
+\end{minipage}%
+%
+\begin{minipage}{0.33\linewidth}
+
+\centering{
+
+\includegraphics{paper_files/figure-pdf/fig-ibrekk-3.pdf}
+
+}
+
+\subcaption{\label{fig-ibrekk-3}Picture 3}
+
+\end{minipage}%
+\newline
+\begin{minipage}{0.33\linewidth}
+
+\centering{
+
+\includegraphics{paper_files/figure-pdf/fig-ibrekk-4.pdf}
+
+}
+
+\subcaption{\label{fig-ibrekk-4}Picture 4}
+
+\end{minipage}%
+%
+\begin{minipage}{0.33\linewidth}
+
+\centering{
+
+\includegraphics{paper_files/figure-pdf/fig-ibrekk-5.pdf}
+
+}
+
+\subcaption{\label{fig-ibrekk-5}Picture 5}
+
+\end{minipage}%
+%
+\begin{minipage}{0.33\linewidth}
+
+\centering{
+
+\includegraphics{paper_files/figure-pdf/fig-ibrekk-6.pdf}
+
+}
+
+\subcaption{\label{fig-ibrekk-6}Picture 6}
+
+\end{minipage}%
+\newline
+\begin{minipage}{0.33\linewidth}
+
+\centering{
+
+\includegraphics{paper_files/figure-pdf/fig-ibrekk-7.pdf}
+
+}
+
+\subcaption{\label{fig-ibrekk-7}Picture 7}
+
+\end{minipage}%
+%
+\begin{minipage}{0.33\linewidth}
+
+\centering{
+
+\includegraphics{paper_files/figure-pdf/fig-ibrekk-8.pdf}
+
+}
+
+\subcaption{\label{fig-ibrekk-8}Picture 8}
+
+\end{minipage}%
+%
+\begin{minipage}{0.33\linewidth}
+
+\centering{
+
+\includegraphics{paper_files/figure-pdf/fig-ibrekk-9.pdf}
+
+}
+
+\subcaption{\label{fig-ibrekk-9}Picture 9}
+
+\end{minipage}%
+
+\caption{\label{fig-ibrekk}A replication of the uncertainty
+visualisations shown by \citet{Ibrekk1987} in one of the earliest
+uncertainty visualisation experiments. Several visualisation methods
+that are now unpopular (such as the pie chart) are used throughout this
+paper.}
+
+\end{figure}%
+
+Reviews on uncertainty visualisation rarely offer tried and tested rules
+for effective uncertainty visualisation, but rather, they comment on the
+\emph{difficulties} faced when trying to summarise the field.
+\citet{Kinkeldey2014} found most experimental methods to be ad hoc, with
+no commonly agreed upon methodology, formalisations, or greater goal of
+describing general principals. \citet{Hullman2016} noticed there is a
+serious noise issue in the data coming from uncertainty visualisation
+experiments. She commented on the prevalence of confounding variables
+that make it unclear as to what exactly caused a subjects poor
+performance on a set of particular questions. Mistakes due to
+misunderstanding visualisations, misinterpreting questions, and
+incorrectly applying heuristics are all combined into a single error
+value. \citet{Spiegelhalter2017} commented that different plots are good
+for different things, and disagreed with the goal of identifying a
+universal best plot for all people and circumstances.
+\citet{Griethe2006} did not identify common themes, but instead listed
+the findings and opinions of a collection of papers.
+\citet{uncertchap2022} summarised several cognitive effects that
+repeatedly arise in uncertainty visualisation experiments, however these
+effects were each discussed in isolation as a list of considerations an
+author might make rather than an overarching theory of rules for
+effective uncertainty visualisation. While these reviews are thorough in
+scope, none discuss how the existing literature contribute to the
+broader goal of uncertainty visualisation. The problem faced by the
+literature is easily summarised with a famous quote by Henri Poincaré.
+
+\begin{quote}
+``Science is built up of facts, as a house is built of stones; but an
+accumulation of facts is no more a science than a heap of stones is a
+house.'' - Henri Poincaré (1905)
+\end{quote}
+
+That is to say, despite the wealth of reviews, the field of uncertainty
+visualisation remains a heap of stones. There is a mountain of work that
+identifies common heuristics found in uncertainty visualisations,
+evaluate competing plot designs, or start a theoretical discussion on a
+niche aspect of the field. This is important work that needs to be done,
+but each of these papers offers up their own bespoke motivation and
+methodology, with little reference to the uncertainty visualisation
+papers outside their periphery. This becomes even more difficult to
+manage when these studies are in conflict. The field is in desperate
+need of a unifying theory that can tie this swath of research together.
+This review attempts to address this issue by offering a novel
+perspective on the uncertainty visualisation problem. That is, we are
+going to use the wealth of established stone to construct the
+foundations on which we can build a house.
+
+This review is broken into several parts that each reflect a different
+approach to uncertainty visualisation. First, we look at graphics that
+ignore uncertainty entirely and discuss why uncertainty should be
+included at all. Second, we look at methods that consider uncertainty to
+be just another variable and discuss the characteristics of uncertainty
+that make it a unique visualisation problem. Third, we look at methods
+that explicitly combine our estimate and its uncertainty and discuss the
+limitations of these approaches. Fourth, we discuss methods that
+implicitly include uncertainty by depicting a sample in place of an
+estimate. Finally, we discuss how uncertainty visualisations can be
+effectively evaluated. When discussing each of these methods, we will
+repeatedly return to the \emph{purpose} of uncertainty visualisation and
+the effectiveness of each approach in fulfilling that purpose.
+
+\subsection{Spatial example}\label{spatial-example}
+
+There are far too many uncertainty visualisations to exhaustively
+discuss them all. Instead we focus on the changes made to a single plot,
+the choropleth map. Due to the field's origins and focus in geospatial
+information visualisation, there have been a large number of suggested
+variations on the choropleth map that allow authors to include
+uncertainty. Utilising a single example will help isolate the ideas we
+are trying to convey. Therefore, it is important to remember that even
+though we focus our discussion on the choropleth map, the theoretical
+approach we outline in this review is useful to all uncertainty
+visualisations regardless of their application. Additionally, our
+examples focus on incorporating uncertainty through colour manipulation,
+as that is the key visual channel used in a choropleth map. However, the
+methods we discuss go beyond variations in a colour palette. Even though
+they are not explicitly shown, visualisations that depict uncertainty
+using layers such as position or shape, and more complicated graphics
+that incorporate animation or interactivity are also within the scope of
+this review. We will use the choropleth maps as a tool to clearly
+highlight the costs and benefits of each approach.
+
+Figure~\ref{fig-data} shows the first six rows and the geographical
+boundaries of our data set. The temperature variable was generated using
+a sine wave, that is
+\(Temperature_i = 29-2\cdot{|Latitude_i - \sin{2 \cdot Longitude_i}|}\)
+where the \(Longitude\) and \(Latitude\) are the longitude and latitude
+of the county's centroid scaled to a standard normal distribution. Each
+county's variance is independently, randomly sampled from a uniform
+distribution. The low standard errors are drawn from a \(U_{[0,1]}\)
+distribution, while the high standard errors are drawn from a
+\(U_{[1,2]}\) distribution. As we are dealing with an average, the
+sampling distribution would be approximately normal by the central limit
+theorem, so each county temperature estimate is assumed to come from a
+\(N(Temp_i, SE_{case,i})\). This is the data we will be using in our
+spatial uncertainty examples for the rest of the paper.
+
+\begin{figure}
+
+\begin{minipage}{0.50\linewidth}
+
+\centering{
+
+\includegraphics{paper_files/figure-pdf/fig-data-1.pdf}
+
+}
+
+\subcaption{\label{fig-data-1}Data Table}
+
+\end{minipage}%
+%
+\begin{minipage}{0.50\linewidth}
+
+\centering{
+
+\includegraphics{paper_files/figure-pdf/fig-data-2.pdf}
+
+}
+
+\subcaption{\label{fig-data-2}Map Boundaries}
+
+\end{minipage}%
+
+\caption{\label{fig-data}The first 5 observations of the data used for
+the spatial uncertainty examples along with the boundaries of each
+county. The map boundaries are the Iowa county boundaries, however the
+`temperature' data is not representative of the average temperature in
+Iowa. The temperature and standard error represent the average of the
+daily high temperature and the standard error of that average
+respectively.}
+
+\end{figure}%
+
+\section{Ignoring uncertainty}\label{ignoring-uncertainty}
+
+A good place to start might be at a deceptively straight forward
+question, why should we include uncertainty at all?
+
+\subsection{The choropleth map}\label{the-choropleth-map}
+
+Figure~\ref{fig-choropleth} depicts a choropleth map of the counties of
+Iowa. Each of these counties are coloured according to an estimate of
+average daily temperature that was generated so that the values followed
+a clear spatial trend. The variance of these estimates were simulated
+such that the trend accounts for most of the variance in the plot in the
+low variance case (so we should expect the trend to be visible) while in
+the high variance, there is more variance within each county than
+between all the counties, so we should expect it to (at least in some
+capacity) overwhelm the spatial trend. Is this aspect of the data and
+the spatial data it communicates clear in in the map? Is the strength of
+the trend communicated through the visualisation?
+
+\begin{figure}
+
+\begin{minipage}{0.33\linewidth}
+
+\centering{
+
+\includegraphics{paper_files/figure-pdf/fig-choropleth-1.pdf}
+
+}
+
+\subcaption{\label{fig-choropleth-1}Low Variance Data}
+
+\end{minipage}%
+%
+\begin{minipage}{0.33\linewidth}
+
+\centering{
+
+\includegraphics{paper_files/figure-pdf/fig-choropleth-2.pdf}
+
+}
+
+\subcaption{\label{fig-choropleth-2}High Variance Data}
+
+\end{minipage}%
+%
+\begin{minipage}{0.33\linewidth}
+
+\centering{
+
+\includegraphics{paper_files/figure-pdf/fig-choropleth-3.pdf}
+
+}
+
+\subcaption{\label{fig-choropleth-3}Choropleth Palette}
+
+\end{minipage}%
+
+\caption{\label{fig-choropleth}Two choropleth maps that depict the
+counties of Iowa where each country coloured acording to a simulated
+average temperature. Both maps depict a spatial trend, where counties
+closer to the centre of the map are hotter than counties on the edge of
+the map. In the low variance condition, the trend accounts for most of
+the variation in the data, in the high variance case, the variance on
+the temperature estimate accounts for most of the variance. This
+distinction is not clear in the map as they both appear identical. The
+high variance condition displays a spatial trend that could simply be
+spurious, which means the plot is displaying a false conclusion.}
+
+\end{figure}%
+
+\subsection{Signal-suppression}\label{signal-suppression}
+
+Uncertainty visualisation is required for transparency. The two
+choropleth maps that appear to be identical in
+Figure~\ref{fig-choropleth} highlight the issues with simply electing to
+ignore uncertainty. This sentiment appears frequently in the uncertainty
+visualisation literature. Some authors suggest uncertainty is important
+to include as it communicates the legitimacy (or illegitimacy) of the
+conclusion drawn from visual inference
+\citep{Correll2014, Kale2018, Griethe2006}. Some authors have said that
+uncertainty should be included to degree of confidence or trust in the
+data \citep{Boukhelifa2012, Zhao2023}. Some authors directly connect
+uncertainty visualisation to hypothesis testing as it ensures the
+validity of a statement \citep{Hullman2020a, Griethe2006}, but allows
+for a proportional level of trust that is more detailed than the binary
+results of a hypothesis test \citep{Correll2014, Correll2018}. Some
+authors even go so far as to claim that failing to include uncertainty
+is akin to fraud or lying \citep{Hullman2020a, Manski2020}.
+
+This consensus leads us to understand that uncertainty visualisation is
+motivated by the need for a sort of ``visual hypothesis test''. A
+successful uncertainty visualisation would act as a ``statistical
+hedge'' for any inference we make using the graphic. Since the purpose
+of a visualisation is to give a quick gist of the information
+\citep{Spiegelhalter2017}, this hedging needs to be communicated
+visually without the need for complicated calculations. If we refer to
+the conclusion we draw from a graphic to be its signal and the variance
+that makes this signal harder to identify as the ``noise'', we can
+summarise the above information into three key requirements. A good
+uncertainty visualisation needs to:
+
+\begin{enumerate}
+\def\labelenumi{\arabic{enumi})}
+\tightlist
+\item
+  Reinforce justified signals to encourage confidence in results
+\item
+  Hide signals that are overwhelmed by noise to prevent unjustified
+  conclusions
+\item
+  Perform tasks 1) and 2) in a way that is proportional to the level of
+  confidence in those conclusions.
+\end{enumerate}
+
+As Figure~\ref{fig-choropleth} showed, visualisations that are
+unconcerned with uncertainty have no issue showing justified signals,
+but struggle with the display of unjustified signals. Therefore, we call
+this approach to uncertainty visualisation as ``signal-suppression''
+since it primarily differentiates itself from from the normal
+``noiseless'' visualisation approach through criteria (2). That is, the
+main difference between an uncertainty visualisation and a normal
+visualisation is that an uncertainty visualisation should prevent us
+from drawing unjustified conclusions.
+
+\subsection{Uncertainty as a signal}\label{uncertainty-as-a-signal}
+
+Uncertainty visualisation is not only motivated by signal-suppression,
+and we would be remiss if we did not mention these alternative
+approaches. Some authors claim the purpose of uncertainty is to improve
+decision making
+\citep{Ibrekk1987, uncertchap2022, Hullman2016, Cheong2016, Boone2018, Padilla2017}.
+Other authors do not describe uncertainty as important for decision
+making, but rather explicitly state it as a variable of importance in of
+itself \citep{Blenkinsop2000}. While uncertainty can provide useful
+information in decision making, it is important to recognise that the
+``uncertainty'' in these cases is not acting as uncertainty at all. It
+is acting as signal.
+
+This is obvious for the cases where we are explicitly interested in the
+variance or error, as we are literally trying to draw conclusions about
+an statistic that is used to describe uncertainty. The same is true for
+visualisations made for decision making, but it is less overt. This is
+easiest to understand with an example.
+
+Imagine you are trying to decide if you want to bring an umbrella with
+you to work. An umbrella is annoying to bring with you, so you only want
+to pack it if the chance of rain is greater than 10\%. Unfortunately,
+your weather prediction app only provides you with the predicted daily
+rainfall. Therefore, your decision will be improved with the inclusion
+of uncertainty. This is \emph{not} because uncertainty in general is
+important for decision making, but because it gives you the tools
+required to calculate the \emph{actual} statistic you are basing your
+decision on. In this sense, uncertainty is no more special to decision
+making than weight is special to a body mass index calculation.
+
+This means the uncertainty visualisations that would perform the best in
+decision making would simply display the uncertainty statistic we are
+interested in, such as the variance, or probability of an event, using
+existing visualisation principles. This is precisely what we observe in
+the literature. Figure~\ref{fig-exceed} depicts an exceedance
+probability map that was designed as an alternative to the choropleth
+map to improve decision making under uncertainty
+\citep{Kuhnert2018, Lucchesi2021}. A keen viewer may notice that the
+exceedance probability map is actually just a choropleth map, only the
+statistic being displayed has changed. We are not sure it is productive
+to categorise this visualisation as an uncertainty visualisation.
+
+\begin{figure}
+
+\begin{minipage}{0.33\linewidth}
+
+\centering{
+
+\includegraphics{paper_files/figure-pdf/fig-exceed-1.pdf}
+
+}
+
+\subcaption{\label{fig-exceed-1}Low Variance Data}
+
+\end{minipage}%
+%
+\begin{minipage}{0.33\linewidth}
+
+\centering{
+
+\includegraphics{paper_files/figure-pdf/fig-exceed-2.pdf}
+
+}
+
+\subcaption{\label{fig-exceed-2}High Variance Data}
+
+\end{minipage}%
+%
+\begin{minipage}{0.33\linewidth}
+
+\centering{
+
+\includegraphics{paper_files/figure-pdf/fig-exceed-3.pdf}
+
+}
+
+\subcaption{\label{fig-exceed-3}Exceedance Probability Map Palette}
+
+\end{minipage}%
+
+\caption{\label{fig-exceed}An exceedance probability map that depict the
+counties of Iowa where each country coloured acording to the probability
+that the average temperature exceeds 27. This map is a choropleth map
+where the variable of interest is a probability.}
+
+\end{figure}%
+
+There seem to be two different definitions of uncertainty visualisation
+floating around in the literature. The first considers \emph{any}
+visualisation of error, variance, or probability to be an uncertainty
+visualisation. The second believes an uncertainty visualisation is the
+output of a function that takes a normal visualisation as an input, and
+transforms it to include uncertainty information. The former group
+believe the purpose of uncertainty visualisation to provide signal about
+a distribution, while the later believe it should act as noise to
+obfuscate a signal. The lack of explicit distinction between these two
+motivations leaves the literature muddled and reviewers struggle to
+understand if uncertainty should be treated as a variable, as metadata,
+or as something else entirely \citep{Kinkeldey2014}. This disagreement
+creates constant contradictions in what the literature considers to be
+an uncertainty visualisation. For example \citet{Leland2005} mentions
+that popular graphics, such as pie charts and bar charts omit
+uncertainty, and \citet{Wickham2011} suggests their product plot
+framework, which includes histograms and bar charts, should be extended
+to include uncertainty. However, pie charts, bar charts and histograms
+have all been used in a significant number of uncertainty visualisation
+experiments \citep{Ibrekk1987, Olston2002, Zhao2023, Hofmann2012}. If
+you view an uncertainty visualisation as a function applied to an
+existing graphic, then you would not see a pie chart or bar chart as
+uncertainty visualisations. These charts are they are yet to have the
+uncertainty visualisation function applied to them. If you view an
+uncertainty visualisation as any graphic that depicts a statistic then
+there are no limitations on which graphics can or cannot be uncertainty
+visualisations.
+
+When we use the term uncertainty visualisation to refer to graphics that
+simply communicate a variance or probability, we are classifying
+visualisations by the data they display, not their visual features.
+Graphics, just like statistics, are not defined by their input data. A
+scatter plot that compares mean and a scatter plot that compares
+variances are both scatter plots. Given that there is no special class
+of visualisation for \emph{other} statistics (such the median or
+maximum) there is no reason to assume visualisations that simply depict
+a variance, error, or probability to be special. Some authors implicitly
+suggest that visualisations of variance or probability are
+differentiated due to the psychological heuristics involved in
+interpreting uncertainty \citep{Hullman2019}. While it is true that
+heuristics lead people to avoid uncertainty \citep{Spiegelhalter2017},
+there is no evidence that this psychological effect translates to issues
+with the visual representation of uncertainty. Again, given that we do
+not make these same visual considerations for other variables that
+elicit distaste or irrational behaviour, there is no reason to assume
+this is what makes uncertainty visualisation so special.
+
+This leads us to the conclusion that the visualisations made for the
+purpose of displaying information about uncertainty statistics are not
+uncertainty visualisations. These graphics are just normal information
+visualisations, and authors can follow existing principles of graphical
+design. We focus on the perspective that uncertainty visualisation
+serves to obfuscate signal, and an uncertainty visualisation is a
+variation on an existing graphic that gives it the ability to suppress
+false signals.
+
+Of course, there is nothing wrong with explicitly visualising variance,
+error, bias, or any other statistic used to depict uncertainty as a
+signal. Just like any other statistic, these metrics provide important
+and useful information for analysis and decisions. However, there is no
+interesting visualisation challenge associated with these graphics, and
+they do not require any special visualisation techniques. The
+uncertainty in these graphics are acting as a signal variable, and they
+should be treated as such.
+
+\section{Visualising uncertainty as a
+variable}\label{visualising-uncertainty-as-a-variable}
+
+Upon hearing that uncertainty needs to be included for transparency, the
+solutions may seem obvious. You may think ``well, I will just add a
+dimension to my plot that includes uncertainty''. This is a reasonable
+approach. The simplest way to add uncertainty to an existing graphic is
+to simply map uncertainty to an unused visual channel. However, it is
+unclear if this approach is sufficient for our purposes.
+
+\subsection{The bivariate map}\label{the-bivariate-map}
+
+Figure~\ref{fig-bivariate} depicts a variation of the choropleth map,
+where we have a two dimensional colour palette. In this graphic,
+temperature is still mapped to hue, but the variance is included by
+utilising colour saturation. While these two maps \emph{do} look
+visually different (which was not the case in the choropleth map) the
+spatial trend is still clearly visible in both graphics. This means the
+uncertainty \emph{is technically} being communicated, however the main
+message of the graphic is still the spatial trend (that may not exist).
+The graphic did not suppress the invalid signal, so it is not performing
+signal-suppression as we would like. At this point, it might be
+reasonable to ask, why? Why is including the uncertainty as a variable
+insufficient to achieve signal-suppression, and what changes should we
+make to ensure signal-suppression occurs?
+
+\begin{figure}
+
+\begin{minipage}{0.33\linewidth}
+
+\centering{
+
+\includegraphics{paper_files/figure-pdf/fig-bivariate-1.pdf}
+
+}
+
+\subcaption{\label{fig-bivariate-1}Low Variance Data}
+
+\end{minipage}%
+%
+\begin{minipage}{0.33\linewidth}
+
+\centering{
+
+\includegraphics{paper_files/figure-pdf/fig-bivariate-2.pdf}
+
+}
+
+\subcaption{\label{fig-bivariate-2}High Variance Data}
+
+\end{minipage}%
+%
+\begin{minipage}{0.33\linewidth}
+
+\centering{
+
+\includegraphics{paper_files/figure-pdf/fig-bivariate-3.pdf}
+
+}
+
+\subcaption{\label{fig-bivariate-3}Bivariate Palette}
+
+\end{minipage}%
+
+\caption{\label{fig-bivariate}A bivariate map that depict the counties
+of Iowa where each county is coloured acording to it's average daily
+temperature and the variance in temperature. This map is a choropleth
+map with a two dimensional colour palette where temperature is
+represented by colour hue, and variance is represented by colour
+saturation. Even though uncertainty has been added to the graphic the
+spatial trend is still clearly visible in the high variance case.}
+
+\end{figure}%
+
+\subsection{Why this approach may (or may not)
+work}\label{why-this-approach-may-or-may-not-work}
+
+The difficulty in incorporating uncertainty into a visualisation is
+frequently mentioned but seldom explained. For example
+\citet{Hullman2016} commented that it is straightforward to show a value
+but it is much more complex to show uncertainty but did not explain why.
+Many authors seem to believe uncertainty visualisation is a simple
+high-dimensional visualisation problem, as the difficulty comes from
+working out how to add uncertainty into already existing graphics
+\citep{Griethe2006}. While this is part of the problem in uncertainty
+visualisation, it is not the complete picture.
+Figure~\ref{fig-bivariate} makes it clear that simply including
+uncertainty as a variable is insufficient to perform signal-suppression.
+If we cannot treat uncertainty the same as we would any other variable,
+how should we treat it? We need to understand what uncertainty actually
+\emph{is}, in order to understand how to integrate it into a
+visualisation.
+
+\subsubsection{It's a variable\ldots{} it's metadata\ldots{} it's
+uncertainty?}\label{its-a-variable-its-metadata-its-uncertainty}
+
+Describing what uncertainty actually is, is surprisingly hard. Most
+authors simply avoid the problem and describe the characteristics of
+uncertainty, of which there are plenty. Often, uncertainty is split
+using an endless stream of ever changing boundaries, such as whether the
+uncertainty is due to true randomness or a lack of knowledge
+\citep{Spiegelhalter2017, Hullman2016, utypo}, if the uncertainty is in
+the attribute, spatial elements, or temporal element of the data
+\citep{Kinkeldey2014}, whether the uncertainty is scientific
+(e.g.~error) or human (e.g.~disagreement among parties)
+\citep{Benjamin2018}, if the uncertainty is random or systematic
+\citep{Sanyal2009}, statistical or bounded
+\citep{Gschwandtnei2016, Olston2002}, recorded as accuracy or precision
+\citep{Griethe2006, Benjamin2018}, which stage of the data analysis
+pipeline the uncertainty comes from \citep{utypo}, how quantifiable the
+uncertainty is \citep{Spiegelhalter2017, utypo}, etc. There are enough
+qualitative descriptors of uncertainty to fill a paper, but, none of
+this is particularly helpful in understanding how to integrate it into a
+visualisation. Rather than trying to define uncertainty by looking at
+the myriad ways in which it \emph{does} appear in an analysis, we may
+find it easier to look at where it \emph{does not}.
+
+Descriptive statistics describe our sample as it is and summarises large
+data down into an easy to swallow format. Descriptive statistics are not
+seen as the primary goal of modern statistics, however, this was not
+always the case. In 19th century England, \emph{positivism} was the
+popular philosophical approach to science (positivists included famous
+statisticians such as Francis Galton and Karl Pearson). Practitioners of
+the approach believed statistics ended with descriptive statistics as
+science must be based on actual experience and observations
+\citep{Otsuka2023}. In order to make statements about population
+statistics, future values, or new observations we need to perform
+inference, which requires the assumption of the ``uniformity of
+nature'', that is, we need to assume that unobserved phenomena should be
+similar to observed phenomena \citep{Otsuka2023}. Positivists believed
+referencing the unobservable was bad science. In other words, these
+scientists embraced descriptive statistics due to the inherent certainty
+that came with them. Since uncertainty is non-existent in descriptive
+statistics, it is clear that uncertainty is a by-product of inference.
+
+This history lesson illustrates what uncertainty actually is. At several
+stages in a statistical analysis, we will violate the uniformity of
+nature assumption. Each of these violations will impact the statistic we
+have calculated and push it further from the population parameter we
+wish to draw inference on. Uncertainty is the amalgamation of these
+impacts. If we do not violate the uniformity of nature assumption at any
+point in our analysis, we do not have any uncertainty.
+
+This interpretation of uncertainty indicates that the uncertainty is not
+a variable of importance in of itself. Uncertainty is metadata about our
+statistic that is required for valid inference. This means uncertainty
+should not be visualised by itself and we should seek to display signal
+and uncertainty together as a ``single integrated uncertain value''
+\citep{Kinkeldey2014}. This aspect of uncertainty visualisation makes it
+a uniquely difficult problem.
+
+\subsubsection{Visualising the ``single integrated uncertain
+value''}\label{visualising-the-single-integrated-uncertain-value}
+
+Typically, when making visualisations, we want the visual channels to be
+separable. That is, we don't want the data represented through one
+visual channel to interfere with the others \citep{Smart2019}. Mapping
+uncertainty and signal to separable channels allows them to be read
+separately, which does not align with the goal of communicating them as
+a single integrated channel. Visualising uncertainty and signal
+separately allows the uncertainty information to simply be ignored,
+which is a pervasive issue in current uncertainty visualisation methods
+\citep{uncertchap2022}. We can see this problem in
+Figure~\ref{fig-bivariate}, as it sends the message ``this data has a
+spatial trend and the estimates have a large variance'' as we read the
+signal and the uncertainty separately.
+
+This means effective uncertainty visualisation should be leveraging
+integrability. That is, the visual channels of the uncertainty and the
+signal would need to be separately manipulable, but read as a single
+channel by the human brain. While most visual aesthetics \emph{are}
+separable, there are some variables that have been shown to be
+integrable, such as colour hue and brightness \citep{Vanderplas2020}.
+When visualising uncertainty using its own visual channel, we can also
+consider visual semiotics and make sure to map uncertainty to intuitive
+visual channels, such as mapping more uncertain values to lighter
+colours \citep{Maceachren2012}. Unfortunately relying on integrability
+may not give us the amount of control we want over our
+signal-suppression. Without a strong understanding of how these visual
+channels collapse down into a single channel, relying on integrability
+could create unintended consequences such as displaying phantom signals
+or hiding justified signals. Additionally, multi-dimensional colour
+palettes can make the graphics harder to read and hurt the accessibility
+of the plots \citep{Vanderplas2015}.
+
+There is another benefit to mapping uncertainty to saturation that is
+not directly related to integrability. As the saturation decreases
+colours become harder to distinguish. This means high uncertainty values
+are harder to differentiate than low uncertainty values. We can leverage
+this implicit feature of colour value by transforming the visual feature
+space ourselves.
+
+\section{Combining uncertainty and signal in a transformed
+space}\label{combining-uncertainty-and-signal-in-a-transformed-space}
+
+Instead of hoping that uncertainty might collapse signal values into a
+single dimension, we can do some of that work ourselves. As a matter of
+fact, some uncertainty visualisation authors already have.
+
+\subsection{Value Suppressing Uncertainty
+Palettes}\label{value-suppressing-uncertainty-palettes}
+
+The Value Suppressing Uncertainty Palette (VSUP) \citep{Correll2018},
+was designed with the intention of preventing high uncertainty values
+from being extracted from a map. Since the palette was designed with the
+extraction of individual values in mind and it has only been tested on
+simple value extraction tasks \citep{Correll2018} or search tasks
+\citep{Ndlovu2023}, it is unclear how effective the method is at
+suppressing broader insights such as spatial trends.
+
+Figure~\ref{fig-vsup} is a visualisation of the Iowa temperature data
+using a VSUP to colour the counties. The low uncertainty case still has
+a visible spatial trend, while the spatial trend in the high uncertainty
+map has functionally disappeared. This means the VSUP has successfully
+suppressed the spatial trend in the data. However, the spatial trend may
+not be the only signal of concern in our graphic. Now we must return to
+the original signal-suppression criteria and ask ourselves if they have
+all been met. Are all the justified signals reinforced, while all the
+unjustified signals are suppressed? Is a graphic that performs perfect
+signal-suppression even possible?
+
+\begin{figure}
+
+\begin{minipage}{0.33\linewidth}
+
+\centering{
+
+\includegraphics{paper_files/figure-pdf/fig-vsup-1.pdf}
+
+}
+
+\subcaption{\label{fig-vsup-1}Low Variance Data}
+
+\end{minipage}%
+%
+\begin{minipage}{0.33\linewidth}
+
+\centering{
+
+\includegraphics{paper_files/figure-pdf/fig-vsup-2.pdf}
+
+}
+
+\subcaption{\label{fig-vsup-2}High Variance Data}
+
+\end{minipage}%
+%
+\begin{minipage}{0.33\linewidth}
+
+\centering{
+
+\includegraphics{paper_files/figure-pdf/fig-vsup-3.pdf}
+
+}
+
+\subcaption{\label{fig-vsup-3}VSUP Palette}
+
+\end{minipage}%
+
+\caption{\label{fig-vsup}A map made with a VSUP. The counties of Iowa
+are coloured acording to its average daily temperature and the variance
+in temperature. Similar to the bivariate map, temperature is mapped to
+hue while variance is mapped to saturation. Unlike the bivariance map,
+the colour space we are mapping our variables to has been transformed so
+that high variance estimates are harder to discern from each other. This
+map successfully reduces the visibility of the spatial trend in the high
+uncertainty case while maintaining the visibility of the spatial trend
+in the low uncertainty case.}
+
+\end{figure}%
+
+\subsection{What can and cannot be
+suppressed?}\label{what-can-and-cannot-be-suppressed}
+
+The methods used by the VSUP bring to light a slight problem with
+uncertainty visualisation. Specifically, uncertainty and the purpose of
+visualisation are somewhat at odds with one another. There are two
+primary motivations behind visualisation, communication and exploratory
+data analysis (EDA). Communication involves identifying a signal we want
+to communicate and designing a visualisation that best conveys that,
+while EDA involves creating a versatile visualisation and using it to
+extract several signals. If we are designing an uncertainty
+visualisation for communication then we can just suppress the specific
+signal we are seeking to communicate. In the map example, we would
+consider Figure~\ref{fig-vsup} to be a success as the only signal we are
+concerned with is the spatial trend. However, it is not uncommon for
+authors to express a desire for uncertainty visualisations that perform
+signal-suppression in visualisations made for EDA
+\citep{Sarma2024, Griethe2006}. For uncertainty visualisation for EDA to
+work, we would need to assume that suppressing individual estimates
+using their variance should naturally extend to broader suppression of
+plot level insights. Unfortunately, it is unclear how reliably this
+would work.
+
+\subsubsection{There is no uncertainty in
+EDA}\label{there-is-no-uncertainty-in-eda}
+
+Earlier we established that uncertainty is a by-product of inference,
+which means without inference, there is no uncertainty. Often EDA is
+used to give us an understanding of our data and identify which signals
+are worth pursuing. In this sense, EDA is the visual parallel to
+descriptive statistics, as it is performed without an explicit
+hypothesis, which means there is no inference, and by extension, there
+is no uncertainty.
+
+Some authors recognise inference will always occur (in some shape or
+form) and believe uncertainty \emph{should} be visualised but do not
+recognise \emph{how} uncertainty would be visualised.
+\citet{Hullman2021} argued that there is no such thing as a
+``model-free'' visualisation, therefore all visualisations require
+uncertainty as we are always performing inference. While it is true that
+we can think of visualisations as containing implicit inferential
+properties, there are many potential inferences in any single
+visualisation. This makes it a little difficult to ensure uncertainty is
+always included. For example, if we have a visualisation that shows an
+average, we would need to identify if the signal suppression should be
+performed using the sampling variance or the sample variance
+\citep{Hofman2020}. The distribution we use depends on the inferential
+statistic but until the viewer chooses one and thinks about it (which
+isn't easily observable), the particular variety of uncertainty which
+would need to be displayed can't be calculated. This means the ideal
+uncertainty visualisation should not only meet the signal suppression
+requirements, but should also endeavor to be versatile enough to meet
+those requirements for all the signals displayed in the graphic.
+
+\subsubsection{The limitations of explicitly visualising uncertainty and
+signal}\label{the-limitations-of-explicitly-visualising-uncertainty-and-signal}
+
+The lack of versatility of the VSUP is easy to see with a simple
+example.
+
+Let's say we have a graphic that depicts a set of coefficients from a
+linear regression and the value of the coefficient is shown using a
+single colour. We want to know ``Which of these coefficients are
+different from 0?'' as well as ``Which of these coefficients are
+different from each other?''. To answer this question we do a series of
+\(t\)-tests on these estimates. All of the individual \(t\)-tests of
+fail to reject the null hypothesis that the coefficients are different
+from 0. We then make a visualisation that suppresses this signal and
+ensures that all of the estimates are visually indistinguishable from 0.
+Next, we conduct two sample \(t\)-test and find that several of the
+values need to be visually distinguishable from each other. The VSUP
+method must pick a single colour for each estimate, and these colours
+must be \emph{either} visually distinguishable or indistinguishable from
+each other. We cannot perform signal-suppression on both these signals
+simultaneously.
+
+This example highlights a fundamental problem with the VSUP that extends
+to the bivariate map as well. When we blend these colours, we need to
+decide at what level of \emph{uncertainty} to blend these colours
+together. Even though the bivariate map does not explicitly combine
+colour values at certain variance levels, the mapping of variance to
+colour saturation does this implicitly. That is, at certain saturation
+values the colours in a bivariate map are imperceptibly different and
+appear as though they are mapped to the same value. At this point, it is
+irrelevant whether or not the colours are technically different, they
+are the same colour in the human brain. This is of course complicated by
+the fact that human colour perception varies at an individual level.
+Some women are believed to have four different cone cells which allows
+them to perceive a greater range of colours, while others only have one
+or two types of cone cells and have colour deficiencies
+\citep{simunovic2010} For VSUP to function for all individuals, we must
+calibrate each plot to an individual's ability to perceive colour.
+
+If we only use a single colour to express each signal-suppressed
+statistic, we will always need to decide which signals we suppress and
+which we do not. This issue has already been raised in the literature.
+Which hypothesis are suppressed and which are not largely depends on the
+method used to combining colours in the palette \citep{Kay2019}. The
+VSUP in Figure~\ref{fig-vsup} used the tree based method that was used
+by \citet{Correll2018}, but there are alternatives that are more
+appropriate for different hypothesis.
+
+Uncertainty visualisation for EDA would be possible if we designed a
+plot in such a way that suppressing individual estimates using their
+variance would naturally extend to broader suppression of plot level
+insights. This assumption is commonly made by visualisation researchers
+in normal visualisation experiments \citep{North2006}. If we could
+express the statistic of a cell using multiple colours, this limitation
+may disappear entirely.
+
+\section{Implicitly Combining Uncertainty and
+Signal}\label{implicitly-combining-uncertainty-and-signal}
+
+Rather than trying to figure out how to combine signal and uncertainty
+into a single colour, we can just display a sample instead and allow the
+viewer to extract \emph{both} the estimate and the variance.
+
+\subsection{Pixel map}\label{pixel-map}
+
+Figure~\ref{fig-pixel} displays a pixel map \citep{Lucchesi2021}, which
+is a variation of the choropleth map where each area is divided up into
+several smaller areas, each coloured using outcomes from the larger
+area's (i.e.~the county's) average temperature sampling distribution.
+The spatial trend is clearly visible in the low variance case, but
+functionally disappears in the low variance case. While the spatial
+trend is just barely visible in the high uncertainty case, it is much
+harder to see. This means the graphic also achieves the third criteria
+for signal-suppression, i.e.~our difficulty in seeing the distribution
+is proportional to the level of uncertainty in the graphic.
+
+\begin{figure}
+
+\begin{minipage}{0.33\linewidth}
+
+\centering{
+
+\includegraphics{paper_files/figure-pdf/fig-pixel-1.pdf}
+
+}
+
+\subcaption{\label{fig-pixel-1}Low Variance Data}
+
+\end{minipage}%
+%
+\begin{minipage}{0.33\linewidth}
+
+\centering{
+
+\includegraphics{paper_files/figure-pdf/fig-pixel-2.pdf}
+
+}
+
+\subcaption{\label{fig-pixel-2}High Variance Data}
+
+\end{minipage}%
+%
+\begin{minipage}{0.33\linewidth}
+
+\centering{
+
+\includegraphics{paper_files/figure-pdf/fig-pixel-3.pdf}
+
+}
+
+\subcaption{\label{fig-pixel-3}Pixel Map Palette}
+
+\end{minipage}%
+
+\caption{\label{fig-pixel}A pixel map of the counties of Iowa. In this
+map, each county is broken up into several small areas and coloured
+according to a potential daily temperature, given the its average daily
+temperature and its sampling distribution. This results in each county
+being represented by a sample rather than a single value. In this
+graphic, we can clearly see the spatial trend in the low variance case,
+while the spatial trend is much harder to identify in the high variance
+case.}
+
+\end{figure}%
+
+It is clear that the pixel-map is not only suppressing the false
+information, but it is doing so by simulating \emph{more} information.
+The efficacy of this method means that visualisations of simulated
+samples pop up repeatedly in the literature, with examples including
+samples that are animated over time \citep{Hullman2015, Blenkinsop2000},
+pixel-maps, and spaghetti time series plots. Not only does this method
+help readers understand the plot level gist, it is also unlikely to
+damage the viewers ability to extract individual estimates. Extracting
+global statistics, such as the mean or variance, from a sample can be
+done with relative ease, especially when those values are mapped to
+colour \citep{Franconeri2021}. Additionally, \citet{Ndlovu2023} found
+that participants applied the same methods they used for simple
+choropleth maps to complicated uncertainty maps even if that take away
+was invalid. The pixel map can be read the same way as a choropleth map,
+which allows us to leverage that automatic response. Therefore, the
+pixel map performs signal suppression, without sacrificing the viewers
+ability to extract general statistics, unless those statistics
+\emph{should} be harder to extract due to the uncertainty in the value.
+So, what is it about Figure~\ref{fig-pixel} that allows it to be such a
+successful uncertainty visualisation? Is this a question we can easily
+answer?
+
+\subsection{Show me the data}\label{show-me-the-data}
+
+To thoroughly understand why pixel maps seem to successfully perform
+signal suppression, we would need a more thorough discussion than what
+can be offered here. However, this does not prohibit us from discussing
+potential reasons for the graph's effectiveness. There is a second type
+of uncertainty that the pixel map quietly communicates but we have not
+touched on yet, and this uncertainty is also conveyed by visualisations
+of our raw data. That is, these graphics can show assumption violations.
+
+As we discussed in previous sections, we can consider uncertainty to be
+``the amalgamations of the impacts of violations to the assumption of
+the uniformity of nature''. It's not a definition that rolls off the
+tongue, but we can work with it. Thankfully, this definition of
+uncertainty aligns nicely with all the concepts that are included in the
+uncertainty umbrella. Some works
+\citep{Hullman2018, Maceachren2012, Thomson2005} focus narrowly on
+specific terms with mathematical definitions, such as probability,
+confidence intervals, variance, error, or precision. These works are
+only concerned with quantifying the final impact of uncertainty on our
+statistics. That is, how large should the bound around our statistic be,
+such that our ``true'' statistic can be inferred. Others
+\citep{Griethe2006, Leland2005, Pang1997, Pham2009, Boukhelifa2017}
+include broader loosely related elements, such as missing values,
+reliability, model validity, or source integrity. These broader and
+harder to quantify concepts are concerned about potential sources of
+uncertainty, that is, they describe violations to the assumption of the
+uniformity of nature.
+
+What this means is that we have two types of uncertainty, but one is
+more ``processed'' than the other. Quantifiable uncertainties are just
+assumption violations expressed as an effect on our final statistic (or
+true variation of our data). The disconnect between quantifiable and
+unquantifiable uncertainty creates a huge problem for authors trying to
+visualise it. A survey of visualisation authors cited ``not knowing how
+to calculate uncertainty'' as one of the primary reasons they did not
+include it in visualisations \citep{Hullman2020a}.
+
+There are two reasons we might leave our uncertainty as an assumption
+violation rather quantifying the effect. The first reason is that we may
+be unable to translate the assumption violation to a quantifiable
+uncertainty. There is no blanket rule that allows us to reliably
+quantify all uncertainty for every statistic, although some researchers
+have tangled with the idea. For example, \citet{Thomson2005} suggests a
+mathematical formula for \emph{examples} of uncertainty, and information
+theory tries to quantify uncertainty using the idea of entropy, but they
+ignore the disconnect between the broad concept of uncertainty and what
+we can reliably quantify. Some authors don't believe that it is even
+possible to quantify all the assumption violations
+\citep{Spiegelhalter2017}.
+
+The second reason to leave uncertainty as a potential violation of our
+assumptions, is that we might not know the final statistic we are
+seeking to calculate. This is the case for visualisations made for EDA,
+and a large number of developments in EDA visualisation have been in
+displaying these difficult to quantify violations. For example,
+\citet{Tierney2023} builds upon the tidy data principles to allow users
+to handle missing values. This includes data plots with a missing value
+``shadow'' that allows visualisation authors to identify if the
+variables used in a plot have any structure in their missing values,
+which would contribute to uncertainty.
+
+With this understanding it becomes clear to see why uncertainty is tied
+to an endless string of examples in the data analysis pipeline.
+Uncertainty examples include imputed data, model selection, inherent
+randomness, biased sampling, etc., not because these things \emph{are}
+uncertainty, but because they \emph{create} uncertainty when we perform
+inference. Whether or not these elements are relevant is highly
+dependent on what statistic you are trying to draw inference on, and by
+extension, the purpose of your visualisation.
+
+This relationship between uncertainty and the purpose of our analysis is
+littered throughout the literature. Multiple authors have commented on
+the need to consider quantifying and expressing uncertainty at every
+stage of a project as the goal shapes every step of the analysis
+\citep{Kinkeldey2014, Hullman2016, Refsgaard2007}. \citet{Otsuka2023}
+suggested that the process of observing data to perform statistics is
+largely dependent on our goals, because the process of boiling real
+world entities down into probabilistic objects (or ``probabilistic
+kind'' as he puts it) depends on the relationship we seek to identify
+with our data. \citet{Meng2014} commented what is kept as data and what
+is tossed away is determined by the motivation of an analysis and what
+was previously noise can be shown to become signal depending on the
+question we seek to answer.\\
+\citet{Wallsten1997} argue that the best method for evaluating or
+combining subjective probabilities depends on the uncertainty the
+decision maker wants to represent and why it matters.
+\citet{Fischhoff2014} looks at uncertainty visualisation for decision
+making decides that we should have different ways of communicating
+uncertainty based off what the user is supposed to do with it.
+
+This makes it very difficult to move quantified uncertainty through the
+layers of our analysis, especially when designing a visualisation for
+EDA. If we don't know what the final statistic is, we cannot quantify
+the effects of our assumptions. Therefore, it is often the case that the
+best uncertainty visualisation is not an uncertainty visualisation at
+all, but simply the most accurate depiction of our raw data. This does
+not mean that visualising raw data will always prevent insignificant
+signal from getting through. \citet{Chowdhury} illustrated how groups
+that appear linearly separable in a linear discriminant analysis (LDA)
+visualisation of the data can actually be the result of a LDA performed
+on too many variables, something that was not clear from the
+visualisation until the line-up protocol was implemented. However, just
+showing the data is a simple but effective option for uncertainty
+visualisation that seems to be largely overlooked.
 
 Of course, visualising the raw data is not always possible nor suitable.
-For example, if we have a specific statistic in mind, it may be more appropriate to show the sampling distribution of that statistic rather than the variability of all observations.
-If we visualise that sampling distribution in a way that allows us to see it's shape, and therefore recognise any assumption violations (such as a non-normal distribution), we maintain some of the benefits that came with visualising the raw data.
-This is exactly what the pixel map does, and likely contributes to it's success as a signal suppression method.
-
-# Evaluating uncertainty visualisations
-If we want to make conclusions about how effective any uncertainty visualisation method is we need to look at the results of evaluation experiments. 
-Unfortunately the illustrative methods we have used thus far, i.e. showing a graphic and saying "wow look at this", is lacking if we want any serious results.
-However, despite the abundance of uncertainty visualisation evaluation experiments, existing literature reviews have struggled to synthesise them into any common rules [@Kinkeldey2014; @Hullman2016].
-
-Here we discuss common evaluation methods, why these methods might struggle to create a cohesive set of recommendations for uncertainty visualisations, and consider how to evaluate visualisations on their ability to perform signal-suppression.
-
-## Current methods
-Including uncertainty in a visualisation comes with many secondary benefits.
-Examples of these benefits include better decisions, more trust in the results and the ability to extract additional statistics, such as the variance.
-Ultimately, these secondary benefits are not the primary goal of uncertainty visualisation, and evaluating uncertainty visualisations on these criteria often has unintended consequences.
-
-### Value extraction of uncertainty statistics
-Uncertainty visualisations are often evaluated based on how accurately [@Hullman2019] viewers can separately extract the estimate and its variance [@Kinkeldey2014]. 
-This means a significant chunk of evaluation studies boil down to showing a participant a visualisation and asking questions such as "what is the variance of $X$?", or "what is the mean of $X$?". 
-This seems like a relatively straight forward approach, and it is similar to how non-uncertainty visualisations are evaluated, but is this appropriate for uncertainty visualisations? 
-The role of uncertainty is rarely evaluated in these studies as the graphics are often compared on the basis of being uncertainty visualisations [@Ibrekk1987; @Hullman2015; @Hofman2020], a class that has no established definition.
-By shifting the focus of our inference from $\bar{X}$ to $Var(X)$ or $P(X)$ we end up evaluating visualisations on their ability to convey uncertainty statistics, rather than on uncertainty's ability to suppress statistics.
-This leads to a series of experiments where the uncertainty is evaluated as a signal even if that was not the goal of the experiment.
-
-The problem with evaluating uncertainty as a signal are identical to the problems associated with displaying uncertainty as a signal. 
-There is no reason to assume uncertainty would behave any differently to any other variable when we evaluate them in this way. 
-For example, @Ibrekk1987 found that participants were more accurate at extracting a statistic when it could be directly read off the graphic, than when it required an area estimate (which is the case if using the probability density function), or when there was no visual indicator for the statistic at all (which is the case when using the cumulative distribution function of a skewed random variable).
-@Hullman2015 found that a visualisation that allows viewers to count outcomes to estimate a probability outperformed one that required a complicated area calculation. 
-@Hofman2020 and @Zhang2022 found that participants were better at answering questions about prediction intervals when shown a prediction interval instead of a sampling distribution. 
-@Gschwandtnei2016 found that graphics where the required statistic could be directly read off the plot outperformed those that involved guesswork due to a gradually decreasing line. 
-@Cheong2016 found that participants made better decisions when they were explicitly given the relevant probability in text rather than when they needed to read it off a map. 
-It is well established that extracting information from a graphic using a perceptual task will always be less accurate than explicitly reading the value provided in text form [@Cleveland1984].
-
-The biggest failing of this evaluation method is not the predictable outcomes, but that it encourages us to see successful examples of signal-suppression as a failing. 
-@Blenkinsop2000 commented that visually integrable depictions of uncertainty should be avoided, as they decrease the viewers confidence in their extracted data values. 
-This conclusion is antithetical to the goals of signal-suppression and occurs because these methods evaluate uncertainty as a signal, not as noise.
-
-### Trust, confidence, and risk aversion
-Trust is a by-product of displaying uncertainty so it is commonly measured in uncertainty evaluation studies [@Hullman2019]. 
-Considering trust, and not transparency, as the metric of importance in uncertainty communication can lead to a questionable subtext that argues against transparency, something that has been noticed by several other authors [@Spiegelhalter2017; @ONeill2018].
-Science communication should be primarily concerned with accuracy. 
-This accuracy may not always involve showing the data exactly as it is, for example we may want to adjust our graphic to correct for visual heuristics, but these changes should still be in pursuit of accuracy.
-Setting trust as the variable of interest implicitly encourages statisticians to set trust as the primary goal of communication. 
-Evaluating visualisations on trust conflates trust and transparency and ultimately discourages signal-suppression. 
-We can see this effect pop up in the results of evaluation studies. 
-For example, @Zhao2023 found that participants were more trusting of model estimates with low uncertainty, but this effect did not carry over to estimates with high uncertainty. 
-Despite decreased trust being a desired outcome of signal suppression, the author's discussion implied this result was not desired. 
-This perspective ends up extending to visualisation authors as well.
-@Hullman2020a found that authors simultaneously argued that failing to visualise uncertainty was akin to fraud, but also many avoided uncertainty visualisation because they didn't want their work to come across as untrustworthy.
-This is a classic example of the negative impacts of placing direct importance on *trust* rather than *transparency*. 
-In cases of high uncertainty, authors will opt to leave out uncertainty information because it decreases confidence in the authors conclusions. 
-This is the end result of designing uncertainty visualisations for increased trust, rather than signal-suppression.
-Of course, this does not mean measuring trust or perceived trustworthiness has no place in visualisation experiments.
-After all, if participants political motivations can influence their ability to accurately read a plot [@nurse2020], then their level of trust in the information could have a similar effect.
-However, it does mean trust should not be the primary goal of uncertainty visualisation evaluation experiments.
-
-Another metric that is similar to trust is the participants confidence in their decision or extracted value. 
-Confidence has many of the same issues as trust, but it has an additional confounding factor. 
-In non-uncertainty visualisation evaluation experiments, confidence is used as a proxy for the clarity of the visualisation. 
-Confidence cannot simultaneously be a measure of clarity of visualisation *and* a way to capture the uncertainty expressed in a visualisation. 
-Uncertainty visualisations conflate these two measures.
-
-Risk-aversion is another secondary effect of uncertainty visualisation that is used to evaluate uncertainty visualisations [@Hullman2019]. 
-Risk aversion is an economics term used to describe an agent who would chose a random variable with a lower expected payout because it also has a lower variance. 
-Risk aversion is considered irrational behaviour because it is a deviation from the behaviour of a rational agent. 
-Comparing participant responses to that of a rational agent has even been suggested as a benchmark for uncertainty visualisation experiments [@wu2023rational].
-However, just like trust, evaluating graphics on risk-aversion discourages signal-suppression. 
-Rational agents *by definition* should *ignore* uncertainty information. 
-Risk aversion is considered to be irrational because it means the economic agent *is* considering the uncertainty information.
-The only case where a rational agent should not ignore the uncertainty information is when the uncertainty is signal, not noise. 
-Designing graphics that encourage choices that align with that of a rational agent, is to encourage graphics that do not include uncertainty at all.
-
-For these reasons we do not believe trust, confidence, or risk-aversion are useful measures to evaluate uncertainty visualisations. 
-While they are designed to capture the secondary effects of uncertainty, using them as primary measures of visualisation performance is in direct conflict with designing visualisations for signal-suppression.
-
-### Questions that attempt to capture signal-suppression
-There is a collection of studies that seem to be aware of the issues behind using trust or value extraction to evaluate uncertainty visualisations. 
-These studies try to measure the effect of some kind of single integrated value, but the methodology is often ad-hoc with varying levels of success.
-
-The first method is what we call the "vague question" approach. 
-The authors of these studies will ask the participants a question that implies that they should use the uncertainty information, but it is unclear how. 
-The authors then go on to compare the readers answer to a very specific ground truth. 
-This means the participants are being *evaluated* as though they are performing a value extraction task, but they are not being asked the *questions* that are asked in the value extraction task. 
-Ultimately this approach results in strangely cryptic questions that create a large amount of noise due to the varied interpretation of the questions [@Hullman2016].
-
-For example @Hofmann2012 showed study participants 20 plots, where each plot displayed two distributions (as a pair of jittered samples, density plots, histograms, or box plots) and asked them to identify the plot where "the blue group is furthest to the right".
-The mean of the two groups was used to decide which group was actually "furthest to the right" and the different graphical approaches were evaluated against that ground truth. 
-In another example @Ibrekk1987 asked participants for the "best estimate", but the participant's responses were evaluated against the mean of the distribution.
-Ultimately, the term "best" is up to the users interpretation, and the estimate that minimised the sum of squared errors was not implied by the question. 
-This vague question approach leads to inconclusive results, as we are left unclear if it was the phrasing of the question or the plot design that caused the participants to answer incorrectly.
-
-A variation of the vague question problem, is a series of studies that ask questions that are impossible to answer using the information given to the participants. 
-These studies frequently expect participants to give deterministic answers for probabilistic questions. 
-For example @Correll2014 showed participants the distribution of voter preferences for two candidates in an election, and asked them "how likely is candidate B to win the election?". 
-Participants were not able to answer the question about likelihood in term of probability, but were instead given seven options from 1 = "Outcome will be most in favour of A" to 7 = "Outcome will be most in favour of B".
-The ground truth statistic for this question was a scalar multiple of Cohen’s d, indicating participants were supposed to incorporate uncertainty information using a very specific formula that was likely unknown to them but assumed to be used implicitly.
-In another example, @Padilla2017 provided participants with a visualisation of the cone of uncertainty and asked then to "decide which oil rig will receive more damage based on the depicted forecast of the hurricane path".
-The cone of uncertainty provides a 60% confidence interval for the location of the eye of a hurricane, which allows us to know the area where the eye of the storm will go, but it does not given any information about the intensity of a storm, the size of a storm, or even if a location will be hit.
-Other authors have commented on the complexity of communicating hurricane risk because the path, storm surge and wind speed are all important and cannot be ignored [@Spiegelhalter2017]. 
-@Padilla2017 seem to be aware of the issues in the visualisation and are using the user study to show that current government communications are not the best choice when communicating hurricane risk, as it is unreasonable to assume that "more likely to be hit" should automatically translate to "more damage".
-While this is useful to highlight, the authors end up caught by their own trap, as answering the questions in their second study correctly required participants to assume an area being more likely to be hit *does* translate to receiving more damage.
-The assumptions required to answer these deterministic questions are often invisible, even to the authors of the paper, which means they often create more problems than they solve when they are used as a tool for graphical evaluation.
-Similar to the vague question problem, these studies seem to be aware that we should be evaluating an uncertainty visualisation based on a deterministic observation (such as our example of "does this map have a spatial trend") but are unsure how to incorporate or evaluate it.
-
-While these approaches are certainly a step in the right direction, the experiments end up having far too much noise in their results. 
-This is further complicated by the fact that visual statistics and their computational equivalents often differ, even before we consider these measurement errors.
-By having a ground truth, these experiments implicitly ask users to treat uncertainty as a signal which goes against the purposes of signal-suppression.
-
-## Testing signal suppression
-If we cannot ask direct questions about uncertainty, or measure the secondary effects of uncertainty, or ask indirect questions about the uncertainty; then how are we supposed to evaluate uncertainty visualisations?
-How do you measure something that disappears the second you look directly at it?
-To evaluate uncertainty visualisations, we need an experimental design that evaluates uncertainty as noise, not as signal. 
-This means we need to measure *uncertainty's impact on the signal*, not the uncertainty itself.
-
-### Comparing to hypothesis tests
-The most obvious way to evaluate uncertainty visualisations is to compare the visualisations to statistical tests. 
-If a graphic was performing signal-suppression we would expect the signal to be harder to see at higher levels of uncertainty. 
-The ideal outcome is a an uncertainty visualisation where the signal is only perceivable when it would be identified by a hypothesis test.
-Evaluating visualisations as though they are akin to hypothesis tests is a well established concept in visualisation.
-The lineup protocol is a good example of this approach.
-Lineups are a confirmatory visualisation tool that can be used to check if perceived patterns are real or merely the result of chance [@Buja2009; @Wickham2010]. 
-Lineups can also be used to evaluate the effectiveness of different types of plots [@Hofmann2012] and design decisions [@vanderplas2017].
-The motivation behind the lineup protocol is similar to the motivations behind signal-suppression, although the lineup protocol is more explicitly tied to a specific hypothesis. 
-@Patrick2023 compared standard statistical tests to the lineup-protocol, and evaluated the visualisations using the power curves that are typical for hypothesis testing. 
-In a similar vein, @Kim2019 investigated how different uncertainty visualisation methods influenced user's prior beliefs, and evaluated the graphics by comparing their results to those from Bayesian inference.
-This approach is similar to the lineup protocol as it also evaluates a visualisation by comparing it to an analogous statistical calculation, however it does so using a different statistical philosophy.
-The connection with visualisation and statistical tests even extends to EDA, where @Hullman2021 commented that a visualisation can be considered a model check where the null hypothesis is the expected and we are looking for the unexpected.
-This is all to say that comparing visualisations to hypothesis tests is not uncharted territory.
-
-We could use a similar evaluation method to evaluate uncertainty visualisations, however, it is not entirely clear what the computational equivalent of an uncertainty visualisation is.
-Even the lineup protocol, which is much closer to a standard statistical test, runs into some difficulty when trying to directly compare it to its computational equivalent.
-@Patrick2023 found that human viewers using the lineup protocol are less sensitive to deviations from the null hypothesis than standard statistical tests.
-However, this finding does not directly translate to the lineup protocol being worse than its computational equivalent.
-The lineup protocol allows viewers to detect if "something" is not right even if that something is unspecified, or what the authors intended.
-For example, @vanderplas2017 did an experiment where participants were drawn to plots with unequal cluster sizes, even though the size of the cluster was not one of the features the authors were testing.
-While standard hypothesis tests may outperform the lineup protocol on power, computational tests have the benefit of knowing exactly what they are looking for. 
-Visualisations are often explicitly made when we *don't* know what we are looking for.
-Additionally, lineup protocols often have a specific hypothesis that is used to create the null distribution, while uncertainty visualisations are intended to be more flexible and can be created without identifying a specific null distribution.
-Therefore, while we could evaluate uncertainty visualisations by comparing them to a selection of hypothesis tests, we should not expect the same level of sensitivity that we get in standard hypothesis tests as the assumed knowledge about our data is quite different.
-This approach could give us a good idea of the use cases of different uncertainty visualisation designs, but we should be mindful of it's limitations in fully describing the effectiveness of these graphics.
-
-### Qualitative Studies
-Alternatively visualisation research could shift away from the accuracy concept all together and ask questions that allow for open ended responses.
-This method can enlighten authors as to *how* the uncertainty information was used by the participants.
-@Hofmann2012 captured this by asking participants why they considered a particular plot to be more right-shifted, even though this qualitative analysis did not make it into the final paper. 
-@Daradkeh2015 presented participants with ten investment alternatives and asked participants "from among available alternatives, which alternative do you prefer the most", and were asked to think aloud and consider the uncertainty in their decision making.
-The experimenters goal was to observe and organise the methods people use when making decisions in the face of uncertainty. 
-They highlighted the specific aspects of uncertainty that participants typically considered, such as the range of outcomes that are above/below a certain threshold, minimum and maximum values, the risk of a loss, etc., and identified where in the decision making process participants made these considerations.
-
-### Heuristics
-While experiments that explicitly identify heuristics in current methods are not technically measuring signal-suppression, they are still a useful consideration to keep in mind when designing experiments for signal suppression.
-Heuristic checks look at unknown pitfalls that might exist in interpretation of current plots [@Hullman2016] and can change depending on the larger scope of the graphic and the population we are communicating with [@Spiegelhalter2017; @Kinkeldey2014].
-
-Several heuristics are of particular importance for uncertainty visualisations, and are likely to impact how well different methods performing signal suppression. 
-The sine illusion can cause the confidence interval of a smoothed sine curve to seem wider at the peaks than the troughs, causing us to underestimate uncertainty associated with changing values [@Vanderplas2015].
-Points that were on an outcome of an ensemble display were perceived as more likely than points not on an outcome, even when the point that was not on an outcome was closer to the centre of the distribution (and therefore more likely) [@Padilla2017].
-This can be considered an extension of the within bar bias, where participants looking at a bar chat with error bars view outcomes within the bar as more likely than those outside it [@Newman2012]. 
-Several studies have found that viewers use a heuristic where they compare the distance between two estimates to estimate if they are different, and in doing so, ignore the uncertainty information [@uncertchap2022].
-
-These heuristics have the potential to create noise in signal suppression evaluations, and should be kept in mind when designing an evaluation experiment.
-
-# Conclusions and Future work
-This paper has identified gaps in the uncertainty visualisation literature that must be filled for the field to progress. 
-We formalised the uncertainty visualisation problem, and in doing so highlighted an untouched area of uncertainty visualisation research.
-
-*Experimental practices on uncertainty visualisation need to be standardised.* 
-Uncertainty visualisations for decision making treat uncertainty as signal, while visualisations for signal-suppression treat uncertainty as noise. 
-As the literature currently exists, there is no way to combine papers to get a meaningful sense of how uncertainty information is understood by a viewer.
-Researchers need to ensure that when they identify the motivation behind their visualisation technique, and that their evaluation methods match the motivations of the paper. 
-Additionally *evaluation methods that evaluate uncertainty as noise need to be developed.*
-
-*Experimenters should consider evaluating visual aesthetics on how well they suppress information in a graphic.* 
-Research into separability and integrability is of particular interest to uncertainty visualisation, as it allows interference from the uncertainty variable. 
-When designing experiments, authors often choose aesthetics that are visually distinguishable; uncertainty visualisation authors should consider doing the opposite.
-
-*Software that allows users to easily perform signal suppression should be developed.* 
-Existing uncertainty visualisation methods view a distribution as its own object and there are no software options for the "an uncertainty visualisation is a function of an existing visualisation" philosophy.
-
-Signal suppression is an untouched area of visualisation research and developing methods for the practice may require us to challenge our entire notion of what makes a good visualisation.
-We are interested to see how this challenge plays out.
-
-# Acknowledgements
-The first author of this paper is supported in part by a scholarship from the the Australian Energy Market Operator. 
-The R packages were used for this work were: `tidyverse` [@tidyverse], `Vizumap` [@Vizumap], `RColorBrewer` [@RColorBrewer], `scales` [@scales], `sf` [@sf], `ggrepel` [@ggrepel], `urbnmapr` [@urbnmapr], `flextable` [@flextable], `colorspace` [@colorspace], and `rgeos` [@rgeos].
-The GitHub repository for this paper can be found at https://github.com/harriet-mason/ARSA-UncertaintyLitReview which contains the files required to reproduce this article in full.
-
-# Bibliography
-
-```{r, include=FALSE, eval=FALSE}
-library(spelling)
-qmd <- "paper.qmd"
-ignore <- readLines("WORDLIST")
-check_spelling <- spell_check_files(
-  qmd,
-  ignore = ignore,
-  lang = "en_GB"
-)
-if (nrow(check_spelling) > 0) {
-  print(check_spelling)
-  stop("Check spelling in Qmd files!")
-}
-```
+For example, if we have a specific statistic in mind, it may be more
+appropriate to show the sampling distribution of that statistic rather
+than the variability of all observations. If we visualise that sampling
+distribution in a way that allows us to see it's shape, and therefore
+recognise any assumption violations (such as a non-normal distribution),
+we maintain some of the benefits that came with visualising the raw
+data. This is exactly what the pixel map does, and likely contributes to
+it's success as a signal suppression method.
+
+\section{Evaluating uncertainty
+visualisations}\label{evaluating-uncertainty-visualisations}
+
+If we want to make conclusions about how effective any uncertainty
+visualisation method is we need to look at the results of evaluation
+experiments. Unfortunately the illustrative methods we have used thus
+far, i.e.~showing a graphic and saying ``wow look at this'', is lacking
+if we want any serious results. However, despite the abundance of
+uncertainty visualisation evaluation experiments, existing literature
+reviews have struggled to synthesise them into any common rules
+\citep{Kinkeldey2014, Hullman2016}.
+
+Here we discuss common evaluation methods, why these methods might
+struggle to create a cohesive set of recommendations for uncertainty
+visualisations, and consider how to evaluate visualisations on their
+ability to perform signal-suppression.
+
+\subsection{Current methods}\label{current-methods}
+
+Including uncertainty in a visualisation comes with many secondary
+benefits. Examples of these benefits include better decisions, more
+trust in the results and the ability to extract additional statistics,
+such as the variance. Ultimately, these secondary benefits are not the
+primary goal of uncertainty visualisation, and evaluating uncertainty
+visualisations on these criteria often has unintended consequences.
+
+\subsubsection{Value extraction of uncertainty
+statistics}\label{value-extraction-of-uncertainty-statistics}
+
+Uncertainty visualisations are often evaluated based on how accurately
+\citep{Hullman2019} viewers can separately extract the estimate and its
+variance \citep{Kinkeldey2014}. This means a significant chunk of
+evaluation studies boil down to showing a participant a visualisation
+and asking questions such as ``what is the variance of \(X\)?'', or
+``what is the mean of \(X\)?''. This seems like a relatively straight
+forward approach, and it is similar to how non-uncertainty
+visualisations are evaluated, but is this appropriate for uncertainty
+visualisations? The role of uncertainty is rarely evaluated in these
+studies as the graphics are often compared on the basis of being
+uncertainty visualisations \citep{Ibrekk1987, Hullman2015, Hofman2020},
+a class that has no established definition. By shifting the focus of our
+inference from \(\bar{X}\) to \(Var(X)\) or \(P(X)\) we end up
+evaluating visualisations on their ability to convey uncertainty
+statistics, rather than on uncertainty's ability to suppress statistics.
+This leads to a series of experiments where the uncertainty is evaluated
+as a signal even if that was not the goal of the experiment.
+
+The problem with evaluating uncertainty as a signal are identical to the
+problems associated with displaying uncertainty as a signal. There is no
+reason to assume uncertainty would behave any differently to any other
+variable when we evaluate them in this way. For example,
+\citet{Ibrekk1987} found that participants were more accurate at
+extracting a statistic when it could be directly read off the graphic,
+than when it required an area estimate (which is the case if using the
+probability density function), or when there was no visual indicator for
+the statistic at all (which is the case when using the cumulative
+distribution function of a skewed random variable). \citet{Hullman2015}
+found that a visualisation that allows viewers to count outcomes to
+estimate a probability outperformed one that required a complicated area
+calculation. \citet{Hofman2020} and \citet{Zhang2022} found that
+participants were better at answering questions about prediction
+intervals when shown a prediction interval instead of a sampling
+distribution. \citet{Gschwandtnei2016} found that graphics where the
+required statistic could be directly read off the plot outperformed
+those that involved guesswork due to a gradually decreasing line.
+\citet{Cheong2016} found that participants made better decisions when
+they were explicitly given the relevant probability in text rather than
+when they needed to read it off a map. It is well established that
+extracting information from a graphic using a perceptual task will
+always be less accurate than explicitly reading the value provided in
+text form \citep{Cleveland1984}.
+
+The biggest failing of this evaluation method is not the predictable
+outcomes, but that it encourages us to see successful examples of
+signal-suppression as a failing. \citet{Blenkinsop2000} commented that
+visually integrable depictions of uncertainty should be avoided, as they
+decrease the viewers confidence in their extracted data values. This
+conclusion is antithetical to the goals of signal-suppression and occurs
+because these methods evaluate uncertainty as a signal, not as noise.
+
+\subsubsection{Trust, confidence, and risk
+aversion}\label{trust-confidence-and-risk-aversion}
+
+Trust is a by-product of displaying uncertainty so it is commonly
+measured in uncertainty evaluation studies \citep{Hullman2019}.
+Considering trust, and not transparency, as the metric of importance in
+uncertainty communication can lead to a questionable subtext that argues
+against transparency, something that has been noticed by several other
+authors \citep{Spiegelhalter2017, ONeill2018}. Science communication
+should be primarily concerned with accuracy. This accuracy may not
+always involve showing the data exactly as it is, for example we may
+want to adjust our graphic to correct for visual heuristics, but these
+changes should still be in pursuit of accuracy. Setting trust as the
+variable of interest implicitly encourages statisticians to set trust as
+the primary goal of communication. Evaluating visualisations on trust
+conflates trust and transparency and ultimately discourages
+signal-suppression. We can see this effect pop up in the results of
+evaluation studies. For example, \citet{Zhao2023} found that
+participants were more trusting of model estimates with low uncertainty,
+but this effect did not carry over to estimates with high uncertainty.
+Despite decreased trust being a desired outcome of signal suppression,
+the author's discussion implied this result was not desired. This
+perspective ends up extending to visualisation authors as well.
+\citet{Hullman2020a} found that authors simultaneously argued that
+failing to visualise uncertainty was akin to fraud, but also many
+avoided uncertainty visualisation because they didn't want their work to
+come across as untrustworthy. This is a classic example of the negative
+impacts of placing direct importance on \emph{trust} rather than
+\emph{transparency}. In cases of high uncertainty, authors will opt to
+leave out uncertainty information because it decreases confidence in the
+authors conclusions. This is the end result of designing uncertainty
+visualisations for increased trust, rather than signal-suppression. Of
+course, this does not mean measuring trust or perceived trustworthiness
+has no place in visualisation experiments. After all, if participants
+political motivations can influence their ability to accurately read a
+plot \citep{nurse2020}, then their level of trust in the information
+could have a similar effect. However, it does mean trust should not be
+the primary goal of uncertainty visualisation evaluation experiments.
+
+Another metric that is similar to trust is the participants confidence
+in their decision or extracted value. Confidence has many of the same
+issues as trust, but it has an additional confounding factor. In
+non-uncertainty visualisation evaluation experiments, confidence is used
+as a proxy for the clarity of the visualisation. Confidence cannot
+simultaneously be a measure of clarity of visualisation \emph{and} a way
+to capture the uncertainty expressed in a visualisation. Uncertainty
+visualisations conflate these two measures.
+
+Risk-aversion is another secondary effect of uncertainty visualisation
+that is used to evaluate uncertainty visualisations \citep{Hullman2019}.
+Risk aversion is an economics term used to describe an agent who would
+chose a random variable with a lower expected payout because it also has
+a lower variance. Risk aversion is considered irrational behaviour
+because it is a deviation from the behaviour of a rational agent.
+Comparing participant responses to that of a rational agent has even
+been suggested as a benchmark for uncertainty visualisation experiments
+\citep{wu2023rational}. However, just like trust, evaluating graphics on
+risk-aversion discourages signal-suppression. Rational agents \emph{by
+definition} should \emph{ignore} uncertainty information. Risk aversion
+is considered to be irrational because it means the economic agent
+\emph{is} considering the uncertainty information. The only case where a
+rational agent should not ignore the uncertainty information is when the
+uncertainty is signal, not noise. Designing graphics that encourage
+choices that align with that of a rational agent, is to encourage
+graphics that do not include uncertainty at all.
+
+For these reasons we do not believe trust, confidence, or risk-aversion
+are useful measures to evaluate uncertainty visualisations. While they
+are designed to capture the secondary effects of uncertainty, using them
+as primary measures of visualisation performance is in direct conflict
+with designing visualisations for signal-suppression.
+
+\subsubsection{Questions that attempt to capture
+signal-suppression}\label{questions-that-attempt-to-capture-signal-suppression}
+
+There is a collection of studies that seem to be aware of the issues
+behind using trust or value extraction to evaluate uncertainty
+visualisations. These studies try to measure the effect of some kind of
+single integrated value, but the methodology is often ad-hoc with
+varying levels of success.
+
+The first method is what we call the ``vague question'' approach. The
+authors of these studies will ask the participants a question that
+implies that they should use the uncertainty information, but it is
+unclear how. The authors then go on to compare the readers answer to a
+very specific ground truth. This means the participants are being
+\emph{evaluated} as though they are performing a value extraction task,
+but they are not being asked the \emph{questions} that are asked in the
+value extraction task. Ultimately this approach results in strangely
+cryptic questions that create a large amount of noise due to the varied
+interpretation of the questions \citep{Hullman2016}.
+
+For example \citet{Hofmann2012} showed study participants 20 plots,
+where each plot displayed two distributions (as a pair of jittered
+samples, density plots, histograms, or box plots) and asked them to
+identify the plot where ``the blue group is furthest to the right''. The
+mean of the two groups was used to decide which group was actually
+``furthest to the right'' and the different graphical approaches were
+evaluated against that ground truth. In another example
+\citet{Ibrekk1987} asked participants for the ``best estimate'', but the
+participant's responses were evaluated against the mean of the
+distribution. Ultimately, the term ``best'' is up to the users
+interpretation, and the estimate that minimised the sum of squared
+errors was not implied by the question. This vague question approach
+leads to inconclusive results, as we are left unclear if it was the
+phrasing of the question or the plot design that caused the participants
+to answer incorrectly.
+
+A variation of the vague question problem, is a series of studies that
+ask questions that are impossible to answer using the information given
+to the participants. These studies frequently expect participants to
+give deterministic answers for probabilistic questions. For example
+\citet{Correll2014} showed participants the distribution of voter
+preferences for two candidates in an election, and asked them ``how
+likely is candidate B to win the election?''. Participants were not able
+to answer the question about likelihood in term of probability, but were
+instead given seven options from 1 = ``Outcome will be most in favour of
+A'' to 7 = ``Outcome will be most in favour of B''. The ground truth
+statistic for this question was a scalar multiple of Cohen's d,
+indicating participants were supposed to incorporate uncertainty
+information using a very specific formula that was likely unknown to
+them but assumed to be used implicitly.\\
+In another example, \citet{Padilla2017} provided participants with a
+visualisation of the cone of uncertainty and asked then to ``decide
+which oil rig will receive more damage based on the depicted forecast of
+the hurricane path''. The cone of uncertainty provides a 60\% confidence
+interval for the location of the eye of a hurricane, which allows us to
+know the area where the eye of the storm will go, but it does not given
+any information about the intensity of a storm, the size of a storm, or
+even if a location will be hit. Other authors have commented on the
+complexity of communicating hurricane risk because the path, storm surge
+and wind speed are all important and cannot be ignored
+\citep{Spiegelhalter2017}. \citet{Padilla2017} seem to be aware of the
+issues in the visualisation and are using the user study to show that
+current government communications are not the best choice when
+communicating hurricane risk, as it is unreasonable to assume that
+``more likely to be hit'' should automatically translate to ``more
+damage''. While this is useful to highlight, the authors end up caught
+by their own trap, as answering the questions in their second study
+correctly required participants to assume an area being more likely to
+be hit \emph{does} translate to receiving more damage. The assumptions
+required to answer these deterministic questions are often invisible,
+even to the authors of the paper, which means they often create more
+problems than they solve when they are used as a tool for graphical
+evaluation. Similar to the vague question problem, these studies seem to
+be aware that we should be evaluating an uncertainty visualisation based
+on a deterministic observation (such as our example of ``does this map
+have a spatial trend'') but are unsure how to incorporate or evaluate
+it.
+
+While these approaches are certainly a step in the right direction, the
+experiments end up having far too much noise in their results. This is
+further complicated by the fact that visual statistics and their
+computational equivalents often differ, even before we consider these
+measurement errors. By having a ground truth, these experiments
+implicitly ask users to treat uncertainty as a signal which goes against
+the purposes of signal-suppression.
+
+\subsection{Testing signal
+suppression}\label{testing-signal-suppression}
+
+If we cannot ask direct questions about uncertainty, or measure the
+secondary effects of uncertainty, or ask indirect questions about the
+uncertainty; then how are we supposed to evaluate uncertainty
+visualisations? How do you measure something that disappears the second
+you look directly at it? To evaluate uncertainty visualisations, we need
+an experimental design that evaluates uncertainty as noise, not as
+signal. This means we need to measure \emph{uncertainty's impact on the
+signal}, not the uncertainty itself.
+
+\subsubsection{Comparing to hypothesis
+tests}\label{comparing-to-hypothesis-tests}
+
+The most obvious way to evaluate uncertainty visualisations is to
+compare the visualisations to statistical tests. If a graphic was
+performing signal-suppression we would expect the signal to be harder to
+see at higher levels of uncertainty. The ideal outcome is a an
+uncertainty visualisation where the signal is only perceivable when it
+would be identified by a hypothesis test. Evaluating visualisations as
+though they are akin to hypothesis tests is a well established concept
+in visualisation. The lineup protocol is a good example of this
+approach. Lineups are a confirmatory visualisation tool that can be used
+to check if perceived patterns are real or merely the result of chance
+\citep{Buja2009, Wickham2010}. Lineups can also be used to evaluate the
+effectiveness of different types of plots \citep{Hofmann2012} and design
+decisions \citep{vanderplas2017}. The motivation behind the lineup
+protocol is similar to the motivations behind signal-suppression,
+although the lineup protocol is more explicitly tied to a specific
+hypothesis. \citet{Patrick2023} compared standard statistical tests to
+the lineup-protocol, and evaluated the visualisations using the power
+curves that are typical for hypothesis testing. In a similar vein,
+\citet{Kim2019} investigated how different uncertainty visualisation
+methods influenced user's prior beliefs, and evaluated the graphics by
+comparing their results to those from Bayesian inference. This approach
+is similar to the lineup protocol as it also evaluates a visualisation
+by comparing it to an analogous statistical calculation, however it does
+so using a different statistical philosophy. The connection with
+visualisation and statistical tests even extends to EDA, where
+\citet{Hullman2021} commented that a visualisation can be considered a
+model check where the null hypothesis is the expected and we are looking
+for the unexpected. This is all to say that comparing visualisations to
+hypothesis tests is not uncharted territory.
+
+We could use a similar evaluation method to evaluate uncertainty
+visualisations, however, it is not entirely clear what the computational
+equivalent of an uncertainty visualisation is. Even the lineup protocol,
+which is much closer to a standard statistical test, runs into some
+difficulty when trying to directly compare it to its computational
+equivalent. \citet{Patrick2023} found that human viewers using the
+lineup protocol are less sensitive to deviations from the null
+hypothesis than standard statistical tests. However, this finding does
+not directly translate to the lineup protocol being worse than its
+computational equivalent. The lineup protocol allows viewers to detect
+if ``something'' is not right even if that something is unspecified, or
+what the authors intended. For example, \citet{vanderplas2017} did an
+experiment where participants were drawn to plots with unequal cluster
+sizes, even though the size of the cluster was not one of the features
+the authors were testing. While standard hypothesis tests may outperform
+the lineup protocol on power, computational tests have the benefit of
+knowing exactly what they are looking for. Visualisations are often
+explicitly made when we \emph{don't} know what we are looking for.
+Additionally, lineup protocols often have a specific hypothesis that is
+used to create the null distribution, while uncertainty visualisations
+are intended to be more flexible and can be created without identifying
+a specific null distribution. Therefore, while we could evaluate
+uncertainty visualisations by comparing them to a selection of
+hypothesis tests, we should not expect the same level of sensitivity
+that we get in standard hypothesis tests as the assumed knowledge about
+our data is quite different. This approach could give us a good idea of
+the use cases of different uncertainty visualisation designs, but we
+should be mindful of it's limitations in fully describing the
+effectiveness of these graphics.
+
+\subsubsection{Qualitative Studies}\label{qualitative-studies}
+
+Alternatively visualisation research could shift away from the accuracy
+concept all together and ask questions that allow for open ended
+responses. This method can enlighten authors as to \emph{how} the
+uncertainty information was used by the participants.
+\citet{Hofmann2012} captured this by asking participants why they
+considered a particular plot to be more right-shifted, even though this
+qualitative analysis did not make it into the final paper.
+\citet{Daradkeh2015} presented participants with ten investment
+alternatives and asked participants ``from among available alternatives,
+which alternative do you prefer the most'', and were asked to think
+aloud and consider the uncertainty in their decision making. The
+experimenters goal was to observe and organise the methods people use
+when making decisions in the face of uncertainty. They highlighted the
+specific aspects of uncertainty that participants typically considered,
+such as the range of outcomes that are above/below a certain threshold,
+minimum and maximum values, the risk of a loss, etc., and identified
+where in the decision making process participants made these
+considerations.
+
+\subsubsection{Heuristics}\label{heuristics}
+
+While experiments that explicitly identify heuristics in current methods
+are not technically measuring signal-suppression, they are still a
+useful consideration to keep in mind when designing experiments for
+signal suppression. Heuristic checks look at unknown pitfalls that might
+exist in interpretation of current plots \citep{Hullman2016} and can
+change depending on the larger scope of the graphic and the population
+we are communicating with \citep{Spiegelhalter2017, Kinkeldey2014}.
+
+Several heuristics are of particular importance for uncertainty
+visualisations, and are likely to impact how well different methods
+performing signal suppression. The sine illusion can cause the
+confidence interval of a smoothed sine curve to seem wider at the peaks
+than the troughs, causing us to underestimate uncertainty associated
+with changing values \citep{Vanderplas2015}. Points that were on an
+outcome of an ensemble display were perceived as more likely than points
+not on an outcome, even when the point that was not on an outcome was
+closer to the centre of the distribution (and therefore more likely)
+\citep{Padilla2017}. This can be considered an extension of the within
+bar bias, where participants looking at a bar chat with error bars view
+outcomes within the bar as more likely than those outside it
+\citep{Newman2012}. Several studies have found that viewers use a
+heuristic where they compare the distance between two estimates to
+estimate if they are different, and in doing so, ignore the uncertainty
+information \citep{uncertchap2022}.
+
+These heuristics have the potential to create noise in signal
+suppression evaluations, and should be kept in mind when designing an
+evaluation experiment.
+
+\section{Conclusions and Future work}\label{conclusions-and-future-work}
+
+This paper has identified gaps in the uncertainty visualisation
+literature that must be filled for the field to progress. We formalised
+the uncertainty visualisation problem, and in doing so highlighted an
+untouched area of uncertainty visualisation research.
+
+\emph{Experimental practices on uncertainty visualisation need to be
+standardised.} Uncertainty visualisations for decision making treat
+uncertainty as signal, while visualisations for signal-suppression treat
+uncertainty as noise. As the literature currently exists, there is no
+way to combine papers to get a meaningful sense of how uncertainty
+information is understood by a viewer. Researchers need to ensure that
+when they identify the motivation behind their visualisation technique,
+and that their evaluation methods match the motivations of the paper.
+Additionally \emph{evaluation methods that evaluate uncertainty as noise
+need to be developed.}
+
+\emph{Experimenters should consider evaluating visual aesthetics on how
+well they suppress information in a graphic.} Research into separability
+and integrability is of particular interest to uncertainty
+visualisation, as it allows interference from the uncertainty variable.
+When designing experiments, authors often choose aesthetics that are
+visually distinguishable; uncertainty visualisation authors should
+consider doing the opposite.
+
+\emph{Software that allows users to easily perform signal suppression
+should be developed.} Existing uncertainty visualisation methods view a
+distribution as its own object and there are no software options for the
+``an uncertainty visualisation is a function of an existing
+visualisation'' philosophy.
+
+Signal suppression is an untouched area of visualisation research and
+developing methods for the practice may require us to challenge our
+entire notion of what makes a good visualisation. We are interested to
+see how this challenge plays out.
+
+\section{Acknowledgements}\label{acknowledgements}
+
+The first author of this paper is supported in part by a scholarship
+from the the Australian Energy Market Operator. The R packages were used
+for this work were: \texttt{tidyverse} \citep{tidyverse},
+\texttt{Vizumap} \citep{Vizumap}, \texttt{RColorBrewer}
+\citep{RColorBrewer}, \texttt{scales} \citep{scales}, \texttt{sf}
+\citep{sf}, \texttt{ggrepel} \citep{ggrepel}, \texttt{urbnmapr}
+\citep{urbnmapr}, \texttt{flextable} \citep{flextable},
+\texttt{colorspace} \citep{colorspace}, and \texttt{rgeos}
+\citep{rgeos}. The GitHub repository for this paper can be found at
+https://github.com/harriet-mason/ARSA-UncertaintyLitReview which
+contains the files required to reproduce this article in full.
+
+\renewcommand\refname{Bibliography}
+  \bibliography{references.bib}
+
+\end{document}

### 2024-11-13 paper.qmd
repeatedly reknit paper to submit to arxive

@@
       country: United States
-bibliography: references.bib
+bibliography: paper.bib
 abstract: Uncertainty visualisation is quickly becomming a hot topic in information visualisation.

### 2024-11-20 paper.qmd
omg I cant believe I spelt susans name wrong

@@
       country: Australia
-  - name: Susan Vanderplus
+  - name: Susan VanderPlas
     url: https://srvanderplas.github.io

### 2024-11-20 paper.tex
repeatedly reknit paper to submit to arxive
omg I cant believe I spelt susans name wrong

@@
   pdftitle={The Noisy Work of Uncertainty Visualisation Research: A Review},
-  pdfauthor={Harriet Mason; Dianne Cook; Sarah Goodwin; Emi Tanaka; Susan Vanderplus},
+  pdfauthor={Harriet Mason; Dianne Cook; Sarah Goodwin; Emi Tanaka; Susan VanderPlas},
   colorlinks=true,
@@
 
-\date{October 18, 2024}
+\date{November 20, 2024}
 \title{\bf The Noisy Work of Uncertainty Visualisation Research: A
@@
 Biological Data Science Institute, The Australian National University\\
-and\\Susan Vanderplus\\
+and\\Susan VanderPlas\\
 Statistics Department, University of Nebraska--Lincoln\\
@@
 \renewcommand\refname{Bibliography}
-  \bibliography{references.bib}
+  \bibliography{paper.bib}
 

### 2024-11-21 paper.qmd
updated paper with correct spelling and sent to arxiv

@@
 number-sections: true
+latex-clean: true
 format:

### 2025-09-06 paper.qmd
corrected a reference and spelling

@@
 bibliography: paper.bib
-abstract: Uncertainty visualisation is quickly becomming a hot topic in information visualisation.
-Exisiting reviews in the field take the definition and purpose of an uncertainty visualisation to be self evident which results in a large amout of conflicting information.
+abstract: Uncertainty visualisation is quickly becoming a hot topic in information visualisation.
+Existing reviews in the field take the definition and purpose of an uncertainty visualisation to be self evident which results in a large amout of conflicting information.
 This conflict largely stems from a conflation between uncertainty visualisations designed for decision making and those designed to prevent false conclusions.
@@
 library(colorspace)
-library(rgeos)
+#library(rgeos)
 ```
@@
 Among these summary tools, information visualisations are one of the most powerful as they allow for quick and memorable communication that allow us to identify quirks in our data that we didn't know to look for.
-Datasets such as Anscombe's quartet [@anscombe] or the Datasaurus Dozen [@datasaurpkg] show a case where visual statistics highlight elements of the data that are invisible to the typical summary statistics.
+Datasets such as Anscombe's quartet [@anscombe] or the Datasaurus Dozen [@datasaurus; @datasaurpkg] show a case where visual statistics highlight elements of the data that are invisible to the typical summary statistics.
 Something as simple as sketching a distribution before recalling statistics or making predictions can greatly increase the accuracy of those measures [@Hullman2018; @Goldstein2014].
@@
 Today, there are an abundance of publications on the topic which makes it timely to construct a review of the field.
-That is, now that there is an overwhelming amount of information, it is valuable to distil it into simple facts. 
+That is, now that there is an overwhelming amount of information, it is valuable to distill it into simple facts. 
 In fact, there have already been several reviews published but a central piece of discussion is missing.
@@
 
-# recurvisely decrease value
+# recursively decrease value
 v_val = 0.5

### 2025-09-06 paper.tex
updated paper with correct spelling and sent to arxiv
corrected a reference and spelling

@@
+% Options for packages loaded elsewhere
 % Options for packages loaded elsewhere
@@
   12pt]{article}
-
+\usepackage{xcolor}
 \usepackage{amsmath,amssymb}
+\setcounter{secnumdepth}{5}
 \usepackage{iftex}
@@
 \else % if luatex or xetex
-  \usepackage{unicode-math}
+  \usepackage{unicode-math} % this also loads fontspec
   \defaultfontfeatures{Scale=MatchLowercase}
@@
 \makeatother
-\usepackage{xcolor}
-\setlength{\emergencystretch}{3em} % prevent overfull lines
-\setcounter{secnumdepth}{5}
 % Make \paragraph and \subparagraph free-standing
@@
 
-\providecommand{\tightlist}{%
-  \setlength{\itemsep}{0pt}\setlength{\parskip}{0pt}}\usepackage{longtable,booktabs,array}
+\usepackage{longtable,booktabs,array}
 \usepackage{calc} % for calculating minipage widths
@@
 \makeatletter
-\def\maxwidth{\ifdim\Gin@nat@width>\linewidth\linewidth\else\Gin@nat@width\fi}
-\def\maxheight{\ifdim\Gin@nat@height>\textheight\textheight\else\Gin@nat@height\fi}
-\makeatother
-% Scale images if necessary, so that they will not overflow the page
-% margins by default, and it is still possible to overwrite the defaults
-% using explicit options in \includegraphics[width, height, ...]{}
-\setkeys{Gin}{width=\maxwidth,height=\maxheight,keepaspectratio}
+\newsavebox\pandoc@box
+\newcommand*\pandocbounded[1]{% scales image to fit in text height/width
+  \sbox\pandoc@box{#1}%
+  \Gscale@div\@tempa{\textheight}{\dimexpr\ht\pandoc@box+\dp\pandoc@box\relax}%
+  \Gscale@div\@tempb{\linewidth}{\wd\pandoc@box}%
+  \ifdim\@tempb\p@<\@tempa\p@\let\@tempa\@tempb\fi% select the smaller of both
+  \ifdim\@tempa\p@<\p@\scalebox{\@tempa}{\usebox\pandoc@box}%
+  \else\usebox{\pandoc@box}%
+  \fi%
+}
 % Set default figure placement to htbp
-\makeatletter
 \def\fps@figure{htbp}
 \makeatother
+
+\setlength{\emergencystretch}{3em} % prevent overfull lines
+
+\providecommand{\tightlist}{%
+  \setlength{\itemsep}{0pt}\setlength{\parskip}{0pt}}
+
+\usepackage[]{natbib}
+\bibliographystyle{agsm}
 
@@
 \makeatother
-
-\ifLuaTeX
-  \usepackage{selnolig}  % disable illegal ligatures
-\fi
-\usepackage[]{natbib}
-\bibliographystyle{agsm}
 \usepackage{bookmark}
-
 \IfFileExists{xurl.sty}{\usepackage{xurl}}{} % add URL line breaks if available
-\urlstyle{same} % disable monospaced font for URLs
+\urlstyle{same}
 \hypersetup{
@@
 
-\date{November 20, 2024}
+\date{September 7, 2025}
 \title{\bf The Noisy Work of Uncertainty Visualisation Research: A
@@
 \begin{abstract}
-Uncertainty visualisation is quickly becomming a hot topic in
-information visualisation. Exisiting reviews in the field take the
-definition and purpose of an uncertainty visualisation to be self
-evident which results in a large amout of conflicting information. This
-conflict largely stems from a conflation between uncertainty
-visualisations designed for decision making and those designed to
-prevent false conclusions. We coin the term ``signal suppression'' to
-describe a visualisation that is designed for preventing false
-conclusions, as the approach demands that the signal (i.e.~the
-collective take away of the estimates) is suppressed by the noise
-(i.e.~the variance on those estimates). We argue that the current
-standards in visualisation suggest that uncertainty visualisations
-designed for decision making should not be considered uncertainty
-visualisations at all. Therefore, future work should focus on signal
-suppression. Effective signal suppression requires us to communicate the
-signal and the noise as a single ``validity of signal'' variable, and
-doing so proves to be difficult with current methods. We illustrate
-current approaches to uncertainty visualisation by showing how they
-would change the visual apprearance of a choropleth map. These maps
-allow us to see why some methods succeed at signal suppression, while
-others fall short. Evaluating visualisations on how well they perform
-signal suppression also proves to be difficult, as it involves measuring
-the effect of noise, a variable we typically try to ignore. We suggest
-authors use qualitative studies or compare uncertainty visualisations to
-the relevant hypothesis tests.
+Uncertainty visualisation is quickly becoming a hot topic in information
+visualisation. Existing reviews in the field take the definition and
+purpose of an uncertainty visualisation to be self evident which results
+in a large amout of conflicting information. This conflict largely stems
+from a conflation between uncertainty visualisations designed for
+decision making and those designed to prevent false conclusions. We coin
+the term ``signal suppression'' to describe a visualisation that is
+designed for preventing false conclusions, as the approach demands that
+the signal (i.e.~the collective take away of the estimates) is
+suppressed by the noise (i.e.~the variance on those estimates). We argue
+that the current standards in visualisation suggest that uncertainty
+visualisations designed for decision making should not be considered
+uncertainty visualisations at all. Therefore, future work should focus
+on signal suppression. Effective signal suppression requires us to
+communicate the signal and the noise as a single ``validity of signal''
+variable, and doing so proves to be difficult with current methods. We
+illustrate current approaches to uncertainty visualisation by showing
+how they would change the visual apprearance of a choropleth map. These
+maps allow us to see why some methods succeed at signal suppression,
+while others fall short. Evaluating visualisations on how well they
+perform signal suppression also proves to be difficult, as it involves
+measuring the effect of noise, a variable we typically try to ignore. We
+suggest authors use qualitative studies or compare uncertainty
+visualisations to the relevant hypothesis tests.
 \end{abstract}
@@
 Datasets such as Anscombe's quartet \citep{anscombe} or the Datasaurus
-Dozen \citep{datasaurpkg} show a case where visual statistics highlight
-elements of the data that are invisible to the typical summary
-statistics. Something as simple as sketching a distribution before
-recalling statistics or making predictions can greatly increase the
-accuracy of those measures \citep{Hullman2018, Goldstein2014}.
+Dozen \citep{datasaurus, datasaurpkg} show a case where visual
+statistics highlight elements of the data that are invisible to the
+typical summary statistics. Something as simple as sketching a
+distribution before recalling statistics or making predictions can
+greatly increase the accuracy of those measures
+\citep{Hullman2018, Goldstein2014}.
 
@@
 that there is an overwhelming amount of information, it is valuable to
-distil it into simple facts. In fact, there have already been several
+distill it into simple facts. In fact, there have already been several
 reviews published but a central piece of discussion is missing.
@@
 
-\includegraphics{paper_files/figure-pdf/fig-ibrekk-1.pdf}
+\pandocbounded{\includegraphics[keepaspectratio]{paper_files/figure-pdf/fig-ibrekk-1.pdf}}
 
@@
 
-\includegraphics{paper_files/figure-pdf/fig-ibrekk-2.pdf}
+\pandocbounded{\includegraphics[keepaspectratio]{paper_files/figure-pdf/fig-ibrekk-2.pdf}}
 
@@
 
-\includegraphics{paper_files/figure-pdf/fig-ibrekk-3.pdf}
+\pandocbounded{\includegraphics[keepaspectratio]{paper_files/figure-pdf/fig-ibrekk-3.pdf}}
 
@@
 
-\includegraphics{paper_files/figure-pdf/fig-ibrekk-4.pdf}
+\pandocbounded{\includegraphics[keepaspectratio]{paper_files/figure-pdf/fig-ibrekk-4.pdf}}
 
@@
 
-\includegraphics{paper_files/figure-pdf/fig-ibrekk-5.pdf}
+\pandocbounded{\includegraphics[keepaspectratio]{paper_files/figure-pdf/fig-ibrekk-5.pdf}}
 
@@
 
-\includegraphics{paper_files/figure-pdf/fig-ibrekk-6.pdf}
+\pandocbounded{\includegraphics[keepaspectratio]{paper_files/figure-pdf/fig-ibrekk-6.pdf}}
 
@@
 
-\includegraphics{paper_files/figure-pdf/fig-ibrekk-7.pdf}
+\pandocbounded{\includegraphics[keepaspectratio]{paper_files/figure-pdf/fig-ibrekk-7.pdf}}
 
@@
 
-\includegraphics{paper_files/figure-pdf/fig-ibrekk-8.pdf}
+\pandocbounded{\includegraphics[keepaspectratio]{paper_files/figure-pdf/fig-ibrekk-8.pdf}}
 
@@
 
-\includegraphics{paper_files/figure-pdf/fig-ibrekk-9.pdf}
+\pandocbounded{\includegraphics[keepaspectratio]{paper_files/figure-pdf/fig-ibrekk-9.pdf}}
 
@@
 
-\includegraphics{paper_files/figure-pdf/fig-data-1.pdf}
+\pandocbounded{\includegraphics[keepaspectratio]{paper_files/figure-pdf/fig-data-1.pdf}}
 
@@
 
-\includegraphics{paper_files/figure-pdf/fig-data-2.pdf}
+\pandocbounded{\includegraphics[keepaspectratio]{paper_files/figure-pdf/fig-data-2.pdf}}
 
@@
 
-\includegraphics{paper_files/figure-pdf/fig-choropleth-1.pdf}
+\pandocbounded{\includegraphics[keepaspectratio]{paper_files/figure-pdf/fig-choropleth-1.pdf}}
 
@@
 
-\includegraphics{paper_files/figure-pdf/fig-choropleth-2.pdf}
+\pandocbounded{\includegraphics[keepaspectratio]{paper_files/figure-pdf/fig-choropleth-2.pdf}}
 
@@
 
-\includegraphics{paper_files/figure-pdf/fig-choropleth-3.pdf}
+\pandocbounded{\includegraphics[keepaspectratio]{paper_files/figure-pdf/fig-choropleth-3.pdf}}
 
@@
 
-\includegraphics{paper_files/figure-pdf/fig-exceed-1.pdf}
+\pandocbounded{\includegraphics[keepaspectratio]{paper_files/figure-pdf/fig-exceed-1.pdf}}
 
@@
 
-\includegraphics{paper_files/figure-pdf/fig-exceed-2.pdf}
+\pandocbounded{\includegraphics[keepaspectratio]{paper_files/figure-pdf/fig-exceed-2.pdf}}
 
@@
 
-\includegraphics{paper_files/figure-pdf/fig-exceed-3.pdf}
+\pandocbounded{\includegraphics[keepaspectratio]{paper_files/figure-pdf/fig-exceed-3.pdf}}
 
@@
 
-\includegraphics{paper_files/figure-pdf/fig-bivariate-1.pdf}
+\pandocbounded{\includegraphics[keepaspectratio]{paper_files/figure-pdf/fig-bivariate-1.pdf}}
 
@@
 
-\includegraphics{paper_files/figure-pdf/fig-bivariate-2.pdf}
+\pandocbounded{\includegraphics[keepaspectratio]{paper_files/figure-pdf/fig-bivariate-2.pdf}}
 
@@
 
-\includegraphics{paper_files/figure-pdf/fig-bivariate-3.pdf}
+\pandocbounded{\includegraphics[keepaspectratio]{paper_files/figure-pdf/fig-bivariate-3.pdf}}
 
@@
 
-\includegraphics{paper_files/figure-pdf/fig-vsup-1.pdf}
+\pandocbounded{\includegraphics[keepaspectratio]{paper_files/figure-pdf/fig-vsup-1.pdf}}
 
@@
 
-\includegraphics{paper_files/figure-pdf/fig-vsup-2.pdf}
+\pandocbounded{\includegraphics[keepaspectratio]{paper_files/figure-pdf/fig-vsup-2.pdf}}
 
@@
 
-\includegraphics{paper_files/figure-pdf/fig-vsup-3.pdf}
+\pandocbounded{\includegraphics[keepaspectratio]{paper_files/figure-pdf/fig-vsup-3.pdf}}
 
@@
 
-\includegraphics{paper_files/figure-pdf/fig-pixel-1.pdf}
+\pandocbounded{\includegraphics[keepaspectratio]{paper_files/figure-pdf/fig-pixel-1.pdf}}
 
@@
 
-\includegraphics{paper_files/figure-pdf/fig-pixel-2.pdf}
+\pandocbounded{\includegraphics[keepaspectratio]{paper_files/figure-pdf/fig-pixel-2.pdf}}
 
@@
 
-\includegraphics{paper_files/figure-pdf/fig-pixel-3.pdf}
+\pandocbounded{\includegraphics[keepaspectratio]{paper_files/figure-pdf/fig-pixel-3.pdf}}
 

### 2025-11-30 twopagesummary/summary.qmd
set up summary document

--- 2024-11-21 paper.qmd
@@
 ---
-title: "The Noisy Work of Uncertainty Visualisation Research: A Review"
+title: "Paper Summary: Uncertainty Visualisation"
 author: 
@@
       country: United States
-bibliography: paper.bib
-abstract: Uncertainty visualisation is quickly becomming a hot topic in information visualisation.
-Exisiting reviews in the field take the definition and purpose of an uncertainty visualisation to be self evident which results in a large amout of conflicting information.
-This conflict largely stems from a conflation between uncertainty visualisations designed for decision making and those designed to prevent false conclusions.
+bibliography: summary.bib
+nocite: |
+  @*
+abstract: Uncertainty visualisation reviews often take the definition and purpose of an uncertainty visualisation to be self evident which results in a large amout of conflicting information.
+This conflict largely stems from a conflation between uncertainty visualisations designed for to visualise uncertainty as signal and those used to visualise uncertainty as noise.
 We coin the term "signal suppression" to describe a visualisation that is designed for preventing false conclusions, as the approach demands that the signal (i.e.
@@
 the variance on those estimates).
-We argue that the current standards in visualisation suggest that uncertainty visualisations designed for decision making should not be considered uncertainty visualisations at all.
+We argue that the current standards in visualisation suggest that uncertainty visualisations designed to extract uncertain information should not be considered uncertainty visualisations at all.
 Therefore, future work should focus on signal suppression.
@@
 ---
-
-```{r}
-#| echo: false
-#| message: false
-#| warning: false
-
-# load Libraries
-library(tidyverse)
-# devtools::install_github("lydialucchesi/Vizumap")
-library(Vizumap)
-library(RColorBrewer)
-library(scales)
-library(sf)
-library(ggrepel)
-# devtools::install_github("UrbanInstitute/urbnmapr")
-library(urbnmapr)
-library(flextable)
-library(colorspace)
-library(rgeos)
-```
-
-## Introduction
-
-From entertainment choices to news articles to insurance plans, the modern citizen is so inundated with information in every aspect of their life it can be overwhelming. 
-In the face of this overflow of information, tools that effectively reduce piles of information to simple and clear ideas become more valuable. 
-That is, we need tools that can sort the signal from the noise.
-
-Among these summary tools, information visualisations are one of the most powerful as they allow for quick and memorable communication that allow us to identify quirks in our data that we didn't know to look for.
-Datasets such as Anscombe's quartet [@anscombe] or the Datasaurus Dozen [@datasaurpkg] show a case where visual statistics highlight elements of the data that are invisible to the typical summary statistics.
-Something as simple as sketching a distribution before recalling statistics or making predictions can greatly increase the accuracy of those measures [@Hullman2018; @Goldstein2014].
-
-"Uncertainty visualisation" is a relatively new field in research. 
-Early mentions of uncertainty visualisation start to appear in the late 1980s [@Ibrekk1987], with geospatial information visualisation literature from the early 1990s declaring this to be essential aspect of any information display [@MacEachren1992; @Carr1992].
-@fig-ibrekk depicts an example of the uncertainty visualisations discussed in these early papers. 
-Despite kicking off the field, these papers did not define uncertainty visualisation.
-This has led to a lack of consensus on what it means for a graphic to visualise uncertainty, an issue we will return to later.
-Therefore, while the field is considered to be quite new, many of the graphics used for uncertainty visualisation have been around for much longer. 
-For example, box plots and histograms display variation which becomes synonymous with uncertainty when we are using them to depict the variation of an estimate. 
-Today, there are an abundance of publications on the topic which makes it timely to construct a review of the field.
-That is, now that there is an overwhelming amount of information, it is valuable to distil it into simple facts. 
-In fact, there have already been several reviews published but a central piece of discussion is missing.
-
-```{r}
-#| echo: false
-#| message: false
-#| warning: false
-#| label: fig-ibrekk
-#| fig-cap: "A replication of the uncertainty visualisations shown by @Ibrekk1987 in one of the earliest uncertainty visualisation experiments.
-Several visualisation methods that are now unpopular (such as the pie chart) are used throughout this paper."
-#| fig-subcap: 
-#|   - "Picture 1"
-#|   - "Picture 2"
-#|   - "Picture 3"
-#|   - "Picture 4"
-#|   - "Picture 5"
-#|   - "Picture 6"
-#|   - "Picture 7"
-#|   - "Picture 8"
-#|   - "Picture 9"
-#| layout-ncol: 3
-
-# Generate data
-set.seed(1)
-x=rnorm(1000, 8, 4)
-ib_data <- tibble(x=ifelse(x<0, -x, x))
-
-# Picture 1
-p1 <- ib_data |>
-  summarise(avg = mean(x),
-          conf_95a = quantile(x, probs=c(0.025)),
-          conf_95b = quantile(x, probs=c(0.975))) |>
-  ggplot(aes(y="NA")) +
-  geom_point(aes(x=avg)) +
-  geom_errorbar(aes(xmin = conf_95a, xmax = conf_95b), width = 0.1) +
-  scale_x_continuous(name = "INCHES OF SNOW",
-                     breaks=seq(0,19),
-                     labels= ggplot2:::interleave(as.character(c(seq(0,18, 2), 19)), rep("", 11))[c(0:19, 21)],
-                     limits=c(0,19)) +
-  theme_classic() +
-  theme(axis.line.y=element_blank(),
-        axis.title.y = element_blank(),
-        axis.text.y = element_blank(),
-        axis.ticks.y = element_blank(),
-        aspect.ratio=1/10)
-
-# Picture 2
-p2 <- ib_data |>
-  mutate(x = ifelse(x>18, 18, x),
-         binx = cut(x, breaks=seq(0,18,2))) |>
-  group_by(binx) |>
-  summarise(n = n()) |>
-  mutate(Probability = n / sum(n)) |>
-  ggplot(aes(x=binx, y=Probability)) +
-  geom_col(fill="black", colour="white") +
-  scale_x_discrete(name = "INCHES OF SNOW",
-                   labels= paste0(seq(0,16,2), sep = "-", seq(2,18,2))) +
-  scale_y_continuous(breaks = seq(0.00, 0.25, 0.05)) +
-  theme_classic() + 
-  theme(aspect.ratio=0.33)
-
-# Picture 3
-p3 <- ib_data |>
-  mutate(x = ifelse(x>18, 18, x),
-         binx = cut(x, 
-                     breaks=seq(0,18,2), 
-                     labels= paste0(seq(0,16,2), sep = "-", seq(2,18,2)))) |>
-  group_by(binx) |>
-  summarise(n = n()) |>
-  mutate(Probability = n / sum(n),
-         csum = rev(cumsum(rev(Probability))), 
-         pos = Probability/2 + lead(csum, 1),
-         pos = if_else(is.na(pos), Probability/2, pos)) |>
-  ggplot(aes(x="", y=Probability, fill=binx)) +
-  geom_bar(stat="identity", width=1) +
-  geom_text_repel(aes(y = pos, label = paste0(round(Probability*100), sep="", "%")),
-                   size = 3, nudge_x = 0.6, show.legend = FALSE, segment.color = 'transparent') +
-  #geom_label(aes(label = paste0(round(Probability*100), sep="", "%")),
-  #          position = position_stack(vjust = 0.5)) + 
-  scale_fill_grey() +
-  coord_polar("y", start=0) +
-  labs(fill = "INCHES OF SNOW") + 
-  theme_void() + 
-  theme(aspect.ratio=1)
-
-# Picture 4
-p4 <- ib_data |>
-  ggplot(aes(x=x)) +
-  geom_density() + 
-  scale_x_continuous(name = "INCHES OF SNOW",
-                     breaks = seq(0,20,2),
-                     labels= paste0(seq(0,20,2))) +
-  scale_y_continuous(name = "Probability density",
-                     breaks = seq(0.00, 0.20, 0.02)) +
-  theme_classic() + 
-  theme(aspect.ratio=0.33)
-
-# Picture 5
-p5 <- ib_data |>
-  ggplot(aes(y="", x=x)) +
-  geom_violin() + 
-  scale_x_continuous(name = "INCHES OF SNOW",
-                     breaks = seq(0,20,2),
-                     labels= paste0(seq(0,20,2)),
-                     limits=c(0,20)) +
-  theme_classic() + 
-  theme(axis.line.y=element_blank(),
-        axis.title.y = element_blank(),
-        axis.text.y = element_blank(),
-        axis.ticks.y = element_blank(),
-        aspect.ratio=0.4)
-
-# Picture 6
-set.seed(1)
-x=rnorm(5000, 8, 4)
-ib_data2 <- tibble(x=ifelse(x<0, -x, x)) |>
-  mutate(x=ifelse(x>=18, 18-rexp(5000,rate=0), x))
-p6 <- ib_data2 |>
-  ggplot(aes(y="", x=x)) +
-  geom_jitter(size=0.05) + 
-  scale_x_continuous(name = "INCHES OF SNOW",
-                     breaks = seq(0,20,2),
-                     labels= paste0(seq(0,20,2)),
-                     limits=c(0,20)) +
-  theme_classic() + 
-  theme(axis.line.y=element_blank(),
-        axis.title.y = element_blank(),
-        axis.text.y = element_blank(),
-        axis.ticks.y = element_blank(),
-        aspect.ratio=0.1)
-
-# Picture 7
-p7 <- ib_data2 |>
-  arrange(x) |>
-  mutate(group = rep(1:50, each=100))|>
-  group_by(group) |>
-  summarise(x = max(x, na.rm=TRUE)) |>
-  add_row(group=c(0,51), x = c(0,20)) |>
-  ggplot(aes(x=x)) +
-  geom_linerange(ymin = 0.1, ymax = 1) + 
-  geom_linerange(y=1, xmin = -0.03, xmax = 20.03)+ 
-  geom_linerange(y=0.1, xmin = -0.03, xmax = 20.03)+ 
-  scale_x_continuous(name = "INCHES OF SNOW",
-                     breaks = seq(0,20,2),
-                     labels= paste0(seq(0,20,2)),
-                     limits=c(0,20)) +
-  scale_y_continuous(limits=c(0,1)) + 
-  theme_classic() + 
-  theme(axis.line.y=element_blank(),
-        axis.title.y = element_blank(),
-        axis.text.y = element_blank(),
-        axis.ticks.y = element_blank(),
-        aspect.ratio=0.1)
-
-# Picture 8
-p8 <- ib_data |>
-  reframe(x = quantile(x, probs=c(0.25, 0.50, 0.75)))|>
-  add_row(x = c(0,20)) |>
-  arrange(x) |>
-  mutate(quantile = c("min", "q1", "med", "q3", "max")) |>
-  pivot_wider(names_from = quantile, values_from = x) |>
-  ggplot(aes(y="")) +
-  #geom_point(aes(x=med)) +
-  geom_errorbar(aes(y="", xmin = min, xmax = max), width = 0.2) +
-  geom_crossbar(aes(y="", x=med, xmin = q1, xmax = q3), width = 0.5) +
-  scale_x_continuous(name = "INCHES OF SNOW",
-                     breaks=seq(0,20),
-                     labels= ggplot2:::interleave(as.character(c(seq(0,20, 2))), rep("", 11))[1:21],
-                     limits=c(0,20)) +
-  theme_classic() +
-  theme(axis.line.y=element_blank(),
-        axis.title.y = element_blank(),
-        axis.text.y = element_blank(),
-        axis.ticks.y = element_blank(),
-        aspect.ratio=1/10)
-
-# Picture 9
-p9 <- ib_data |>
-  ggplot(aes(x)) +
-  stat_ecdf(geom = "step") +
-  scale_x_continuous(name = "INCHES OF SNOW",
-                     breaks=seq(0,20),
-                     labels= ggplot2:::interleave(as.character(c(seq(0,20, 2))), rep("", 11))[1:21],
-                     limits=c(0,20)) +
-  scale_y_continuous(name = "Cumulative probability",
-                     breaks=seq(0,1,0.1),
-                     labels= seq(0,1,0.1),
-                     limits=c(0,1)) +
-  theme_classic() +
-  theme(aspect.ratio=4/10)
-
-# Display Plots
-p1
-p2
-p3
-p4
-p5
-p6
-p7
-p8
-p9
-```
-
-Reviews on uncertainty visualisation rarely offer tried and tested rules for effective uncertainty visualisation, but rather, they comment on the *difficulties* faced when trying to summarise the field.
-@Kinkeldey2014 found most experimental methods to be ad hoc, with no commonly agreed upon methodology, formalisations, or greater goal of describing general principals. 
-@Hullman2016 noticed there is a serious noise issue in the data coming from uncertainty visualisation experiments. 
-She commented on the prevalence of confounding variables that make it unclear as to what exactly caused a subjects poor performance on a set of particular questions.
-Mistakes due to misunderstanding visualisations, misinterpreting questions, and incorrectly applying heuristics are all combined into a single error value.
-@Spiegelhalter2017 commented that different plots are good for different things, and disagreed with the goal of identifying a universal best plot for all people and circumstances.
-@Griethe2006 did not identify common themes, but instead listed the findings and opinions of a collection of papers. 
-@uncertchap2022 summarised several cognitive effects that repeatedly arise in uncertainty visualisation experiments, however these effects were each discussed in isolation as a list of considerations an author might make rather than an overarching theory of rules for effective uncertainty visualisation.
-While these reviews are thorough in scope, none discuss how the existing literature contribute to the broader goal of uncertainty visualisation. 
-The problem faced by the literature is easily summarised with a famous quote by Henri Poincaré.
-
-> "Science is built up of facts, as a house is built of stones; but an accumulation of facts is no more a science than a heap of stones is a house." - Henri Poincaré (1905)
-
-That is to say, despite the wealth of reviews, the field of uncertainty visualisation remains a heap of stones. 
-There is a mountain of work that identifies common heuristics found in uncertainty visualisations, evaluate competing plot designs, or start a theoretical discussion on a niche aspect of the field.
-This is important work that needs to be done, but each of these papers offers up their own bespoke motivation and methodology, with little reference to the uncertainty visualisation papers outside their periphery.
-This becomes even more difficult to manage when these studies are in conflict. 
-The field is in desperate need of a unifying theory that can tie this swath of research together.
-This review attempts to address this issue by offering a novel perspective on the uncertainty visualisation problem.
-That is, we are going to use the wealth of established stone to construct the foundations on which we can build a house.
-
-This review is broken into several parts that each reflect a different approach to uncertainty visualisation. 
-First, we look at graphics that ignore uncertainty entirely and discuss why uncertainty should be included at all. 
-Second, we look at methods that consider uncertainty to be just another variable and discuss the characteristics of uncertainty that make it a unique visualisation problem. 
-Third, we look at methods that explicitly combine our estimate and its uncertainty and discuss the limitations of these approaches. 
-Fourth, we discuss methods that implicitly include uncertainty by depicting a sample in place of an estimate. 
-Finally, we discuss how uncertainty visualisations can be effectively evaluated.
-When discussing each of these methods, we will repeatedly return to the *purpose* of uncertainty visualisation and the effectiveness of each approach in fulfilling that purpose.
-
-### Spatial example
-There are far too many uncertainty visualisations to exhaustively discuss them all. 
-Instead we focus on the changes made to a single plot, the choropleth map. 
-Due to the field's origins and focus in geospatial information visualisation, there have been a large number of suggested variations on the choropleth map that allow authors to include uncertainty. 
-Utilising a single example will help isolate the ideas we are trying to convey. 
-Therefore, it is important to remember that even though we focus our discussion on the choropleth map, the theoretical approach we outline in this review is useful to all uncertainty visualisations regardless of their application.
-Additionally, our examples focus on incorporating uncertainty through colour manipulation, as that is the key visual channel used in a choropleth map.
-However, the methods we discuss go beyond variations in a colour palette.
-Even though they are not explicitly shown, visualisations that depict uncertainty using layers such as position or shape, and more complicated graphics that incorporate animation or interactivity are also within the scope of this review.
-We will use the choropleth maps as a tool to clearly highlight the costs and benefits of each approach.
-
-@fig-data shows the first six rows and the geographical boundaries of our data set. 
-The temperature variable was generated using a sine wave, that is $Temperature_i = 29-2\cdot{|Latitude_i - \sin{2 \cdot Longitude_i}|}$ where the $Longitude$ and $Latitude$ are the longitude and latitude of the county's centroid scaled to a standard normal distribution.
-Each county's variance is independently, randomly sampled from a uniform distribution. 
-The low standard errors are drawn from a $U_{[0,1]}$ distribution, while the high standard errors are drawn from a $U_{[1,2]}$ distribution. 
-As we are dealing with an average, the sampling distribution would be approximately normal by the central limit theorem, so each county temperature estimate is assumed to come from a $N(Temp_i, SE_{case,i})$.
-This is the data we will be using in our spatial uncertainty examples for the rest of the paper.
-
-```{r}
-#| eval: false
-#| echo: false
-# Get map data: do this once and save
-my_map_data <- get_urbn_map("counties", sf = TRUE) |>
-  filter(state_name=="Iowa")
-save(my_map_data, file="data/iowa_map.rda")
-```
-
-```{r}
-#| eval: false
-#| echo: false
-# Get centroid because rgeos is depreciated
-load("data/iowa_map.rda")
-centroids <- as_tibble(gCentroid(as(my_map_data$geometry, "Spatial"), byid = TRUE))
-my_map_data$cent_long <- centroids$x
-my_map_data$cent_lat <- centroids$y
-save(my_map_data, file="data/my_map_data.rda")
-```
-
-```{r}
-#| echo: false
-#| message: false
-#| warning: false
-#| label: fig-data
-#| fig-cap: "The first 5 observations of the data used for the spatial uncertainty examples along with the boundaries of each county.
-The map boundaries are the Iowa county boundaries, however the 'temperature' data is not representative of the average temperature in Iowa.
-The temperature and standard error represent the average of the daily high temperature and the standard error of that average respectively."
-#| fig-subcap: 
-#|   - "Data Table"
-#|   - "Map Boundaries"
-#| layout-nrow: 1
-#| cap-location: "bottom"
-
-# get data
-load("data/my_map_data.rda")
-
-# seed for sampling
-set.seed(1997)
-
-# data dimension for sampling
-n <- dim(my_map_data)[1]
-
-# Make palettes
-longpal <- rev(sequential_hcl(13, palette = "YlOrRd"))
-basecols <- longpal[3:10]
-breaks <- 21:29 
-breakslong <- 18:32 
-names(basecols) <- seq(8)
-names(longpal) <- -1:11
-
-my_map_data <- my_map_data |>
-  mutate(temp = 29 - 2*abs(scale(cent_lat) - sin(2*(scale(cent_long)))[,1]), # trend
-         highvar = runif(n, min=2, max=4), # high variance
-         lowvar = runif(n, min=0, max=2), # low variance
-         count_id = row_number()) |>
-  pivot_longer(cols=highvar:lowvar, names_to = "variance_class", values_to = "variance") |>
-  # add bivariate classes to data
-  mutate(bitemp = cut(temp, breaks=breaks, labels=seq(8)),
-         bivar = cut(variance, breaks=0:4, labels=seq(4)),
-         biclass = paste(bitemp, bivar, sep="-"))|>
-  mutate(highlight = ifelse(count_id <= 5, TRUE, FALSE))
-
-# Make nice example table
-example_table <- my_map_data |>
-  mutate(variance = sqrt(variance)) |>
-  select(c(count_id, county_name, temp, variance_class, variance)) |>
-  as_tibble()|>
-  pivot_wider(id_cols=c(count_id, county_name, temp,), 
-              names_from = variance_class, 
-              values_from = variance) |>
-  head(5)|>
-  flextable() |>
-  set_caption(caption = "Average Daily High Temperatures of Iowa Counties") |>
-  add_header_row(colwidths = c(3, 2),
-                 values = c("", "Standard Error")) |>
-  colformat_double(digits = 2) |>
-  set_header_labels(count_id = "ID", 
-                    county_name  = "County",
-                    temp = "Average Temperature (°C)",
-                    highvar = "High",
-                    lowvar = "Low") |>
-  add_footer_row(values = rep("..."), colwidths = 5) |>
-  theme_vanilla() |>
-  vline(i=c(1,2), j=3, part="header")|>
-  align(align = "left", part = "all") |>
-  bg(j = "temp",
-    bg = col_numeric(palette = brewer.pal(8, name = "Oranges"),
-                     domain = c(21, 30)),
-    part = "body"
-  ) |>
-  bg(j = c("highvar", "lowvar"),
-    bg = col_numeric(palette = brewer.pal(8, name = "Greens"),
-                     domain = c(0, 3)),
-    part = "body"
-  )
-
-# make blank map 
-example_map <- my_map_data |>
-  filter(variance_class=="lowvar") |>
-  ggplot() +
-  geom_sf(aes(geometry = geometry, fill=highlight)) + 
-  scale_fill_manual(values=c("white", "#fbfba2")) +
-  geom_text(data=filter(my_map_data, highlight==TRUE), aes(x=cent_long, y=cent_lat, label=count_id), size=3) +
-  theme_void() +
-  theme(legend.position = 'none')
-
-plot(example_table)
-
-example_map
-
-```
-
-## Ignoring uncertainty
-A good place to start might be at a deceptively straight forward question, why should we include uncertainty at all?
-
-### The choropleth map
-@fig-choropleth depicts a choropleth map of the counties of Iowa. 
-Each of these counties are coloured according to an estimate of average daily temperature that was generated so that the values followed a clear spatial trend. 
-The variance of these estimates were simulated such that the trend accounts for most of the variance in the plot in the low variance case (so we should expect the trend to be visible) while in the high variance, there is more variance within each county than between all the counties, so we should expect it to (at least in some capacity) overwhelm the spatial trend.
-Is this aspect of the data and the spatial data it communicates clear in in the map? 
-Is the strength of the trend communicated through the visualisation?
-
-```{r}
-#| echo: false
-#| message: false
-#| warning: false
-#| label: fig-choropleth
-#| fig-cap: "Two choropleth maps that depict the counties of Iowa where each country coloured acording to a simulated average temperature.
-Both maps depict a spatial trend, where counties closer to the centre of the map are hotter than counties on the edge of the map.
-In the low variance condition, the trend accounts for most of the variation in the data, in the high variance case, the variance on the temperature estimate accounts for most of the variance.
-This distinction is not clear in the map as they both appear identical.
-The high variance condition displays a spatial trend that could simply be spurious, which means the plot is displaying a false conclusion."
-#| fig-subcap: 
-#|   - "Low Variance Data"
-#|   - "High Variance Data"
-#|   - "Choropleth Palette"
-#| layout-ncol: 3
-#| layout-valign: "bottom"
-#| cap-location: "bottom"
-
-# Choropleth Map
-p1a <- my_map_data |>
-  filter(variance_class=="lowvar") |>
-  ggplot() +
-  geom_sf(aes(fill = bitemp, 
-              geometry = geometry), colour=NA) + 
-  scale_fill_manual(values = basecols) +
-  #scale_fill_gradientn(colours = basecols, 
-  #                     values=breaks/limits[2],
-  #                     limits=limits) +
-  theme_void() + 
-  theme(legend.position = "none")
-
-p1b <- p1a %+% filter(my_map_data, variance_class=="highvar")
-
-show_pal <- function (colours, borders = NULL, cex_label = 1, ncol = NULL, myxlab, breaks, textnudge, xlabx, xlaby, tsize=1.2) {
-  # Set dimensions of palette
-  n <- length(colours)
-  ncol <- ncol %||% ceiling(sqrt(length(colours)))
-  nrow <- ceiling(n/ncol)
-  # make matrix with null values (if not full)
-  colours <- c(colours, rep(NA, nrow * ncol - length(colours)))
-  colours <- matrix(colours, ncol = ncol, byrow = TRUE)
-  # set graphical parameters (?)
-  old <- par(pty = "s", mar = c(0, 0, 0, 0))
-  on.exit(par(old))
-  size <- max(dim(colours))
-  plot(c(0, size), c(0, -size), type = "n", xlab = "", ylab = "", 
-      axes = FALSE)
-  rect(col(colours) - 1, -row(colours) + 1, col(colours), -row(colours), 
-       col = colours, border = borders)
-  text(c(0,col(colours)) + textnudge, -c(1,row(colours))-0.25, breaks, 
-              cex = 1, col = "black")
-  text(xlabx, xlaby, myxlab ,cex = tsize, col = "black")
-}
-
-p1a
-p1b
-show_pal(basecols, ncol=8, borders=NA, myxlab = "Temperature", breaks = 21:29, textnudge = c(0.2, 0.1, 0,0,0,0,0,-0.1,-0.2), xlabx= 4, xlaby=-1.75, tsize=1.5)
-
-```
-
-### Signal-suppression
-Uncertainty visualisation is required for transparency. 
-The two choropleth maps that appear to be identical in @fig-choropleth highlight the issues with simply electing to ignore uncertainty. 
-This sentiment appears frequently in the uncertainty visualisation literature.
-Some authors suggest uncertainty is important to include as it communicates the legitimacy (or illegitimacy) of the conclusion drawn from visual inference [@Correll2014; @Kale2018; @Griethe2006]. 
-Some authors have said that uncertainty should be included to degree of confidence or trust in the data [@Boukhelifa2012; @Zhao2023]. 
-Some authors directly connect uncertainty visualisation to hypothesis testing as it ensures the validity of a statement [@Hullman2020a; @Griethe2006], but allows for a proportional level of trust that is more detailed than the binary results of a hypothesis test [@Correll2014; @Correll2018].
-Some authors even go so far as to claim that failing to include uncertainty is akin to fraud or lying [@Hullman2020a; @Manski2020].
-
-This consensus leads us to understand that uncertainty visualisation is motivated by the need for a sort of "visual hypothesis test". 
-A successful uncertainty visualisation would act as a "statistical hedge" for any inference we make using the graphic. 
-Since the purpose of a visualisation is to give a quick gist of the information [@Spiegelhalter2017], this hedging needs to be communicated visually without the need for complicated calculations. 
-If we refer to the conclusion we draw from a graphic to be its signal and the variance that makes this signal harder to identify as the "noise", we can summarise the above information into three key requirements.
-A good uncertainty visualisation needs to:
-
-  1) Reinforce justified signals to encourage confidence in results
-  2) Hide signals that are overwhelmed by noise to prevent unjustified conclusions
-  3) Perform tasks 1) and 2) in a way that is proportional to the level of confidence in those conclusions.
-
-As @fig-choropleth showed, visualisations that are unconcerned with uncertainty have no issue showing justified signals, but struggle with the display of unjustified signals. 
-Therefore, we call this approach to uncertainty visualisation as "signal-suppression" since it primarily differentiates itself from from the normal "noiseless" visualisation approach through criteria (2).
-That is, the main difference between an uncertainty visualisation and a normal visualisation is that an uncertainty visualisation should prevent us from drawing unjustified conclusions.
-
-### Uncertainty as a signal
-Uncertainty visualisation is not only motivated by signal-suppression, and we would be remiss if we did not mention these alternative approaches. 
-Some authors claim the purpose of uncertainty is to improve decision making [@Ibrekk1987; @uncertchap2022; @Hullman2016; @Cheong2016; @Boone2018; @Padilla2017]. 
-Other authors do not describe uncertainty as important for decision making, but rather explicitly state it as a variable of importance in of itself [@Blenkinsop2000]. 
-While uncertainty can provide useful information in decision making, it is important to recognise that the "uncertainty" in these cases is not acting as uncertainty at all. 
-It is acting as signal.
-
-This is obvious for the cases where we are explicitly interested in the variance or error, as we are literally trying to draw conclusions about an statistic that is used to describe uncertainty. 
-The same is true for visualisations made for decision making, but it is less overt. 
-This is easiest to understand with an example.
-
-Imagine you are trying to decide if you want to bring an umbrella with you to work. 
-An umbrella is annoying to bring with you, so you only want to pack it if the chance of rain is greater than 10%. 
-Unfortunately, your weather prediction app only provides you with the predicted daily rainfall. 
-Therefore, your decision will be improved with the inclusion of uncertainty. 
-This is *not* because uncertainty in general is important for decision making, but because it gives you the tools required to calculate the *actual* statistic you are basing your decision on. 
-In this sense, uncertainty is no more special to decision making than weight is special to a body mass index calculation.
-
-This means the uncertainty visualisations that would perform the best in decision making would simply display the uncertainty statistic we are interested in, such as the variance, or probability of an event, using existing visualisation principles.
-This is precisely what we observe in the literature. 
-@fig-exceed depicts an exceedance probability map that was designed as an alternative to the choropleth map to improve decision making under uncertainty [@Kuhnert2018; @Lucchesi2021]. 
-A keen viewer may notice that the exceedance probability map is actually just a choropleth map, only the statistic being displayed has changed. 
-We are not sure it is productive to categorise this visualisation as an uncertainty visualisation.
-
-```{r}
-#| echo: false
-#| message: false
-#| warning: false
-#| label: fig-exceed
-#| fig-cap: "An exceedance probability map that depict the counties of Iowa where each country coloured acording to the probability that the average temperature exceeds 27.
-This map is a choropleth map where the variable of interest is a probability."
-#| fig-subcap: 
-#|   - "Low Variance Data"
-#|   - "High Variance Data"
-#|   - "Exceedance Probability Map Palette"
-#| layout-ncol: 3
-
-# quantile
-prob_breaks <- seq(-0.1,1.1, length.out=9)
-exeed_data <- my_map_data |> 
-  as_tibble() |>
-  mutate(xprob = 1- pnorm(27, mean=temp, sd=sqrt(variance))) |>
-  mutate(xprob = cut(xprob, breaks=prob_breaks, labels=seq(8)))
-
-# Exceed Prob Map
-p2a <- exeed_data |>
-  filter(variance_class=="lowvar") |>
-  ggplot() +
-  geom_sf(aes(fill = xprob, 
-              geometry = geometry), colour=NA) + 
-  scale_fill_manual(values = basecols) +
-  theme_void() + 
-  theme(legend.position = "none")
-
-p2b <- p2a %+% filter(exeed_data, variance_class=="highvar")
-
-p2a
-p2b
-show_pal(basecols, ncol=8, borders=NA, myxlab = "P(Temperature>27)", breaks = c(0, prob_breaks[2:8], 1), textnudge = c(0.2, 0.1, 0,0,0,0,0,-0.1,-0.2), xlabx= 4, xlaby=-1.75, tsize=1.2)
-```
-
-There seem to be two different definitions of uncertainty visualisation floating around in the literature. 
-The first considers *any* visualisation of error, variance, or probability to be an uncertainty visualisation. 
-The second believes an uncertainty visualisation is the output of a function that takes a normal visualisation as an input, and transforms it to include uncertainty information. 
-The former group believe the purpose of uncertainty visualisation to provide signal about a distribution, while the later believe it should act as noise to obfuscate a signal. 
-The lack of explicit distinction between these two motivations leaves the literature muddled and reviewers struggle to understand if uncertainty should be treated as a variable, as metadata, or as something else entirely [@Kinkeldey2014].
-This disagreement creates constant contradictions in what the literature considers to be an uncertainty visualisation. 
-For example @Leland2005 mentions that popular graphics, such as pie charts and bar charts omit uncertainty, and @Wickham2011 suggests their product plot framework, which includes histograms and bar charts, should be extended to include uncertainty.
-However, pie charts, bar charts and histograms have all been used in a significant number of uncertainty visualisation experiments [@Ibrekk1987; @Olston2002; @Zhao2023; @Hofmann2012]. 
-If you view an uncertainty visualisation as a function applied to an existing graphic, then you would not see a pie chart or bar chart as uncertainty visualisations.
-These charts are they are yet to have the uncertainty visualisation function applied to them. 
-If you view an uncertainty visualisation as any graphic that depicts a statistic then there are no limitations on which graphics can or cannot be uncertainty visualisations.
-
-When we use the term uncertainty visualisation to refer to graphics that simply communicate a variance or probability, we are classifying visualisations by the data they display, not their visual features.
-Graphics, just like statistics, are not defined by their input data. 
-A scatter plot that compares mean and a scatter plot that compares variances are both scatter plots. 
-Given that there is no special class of visualisation for *other* statistics (such the median or maximum) there is no reason to assume visualisations that simply depict a variance, error, or probability to be special.
-Some authors implicitly suggest that visualisations of variance or probability are differentiated due to the psychological heuristics involved in interpreting uncertainty [@Hullman2019].
-While it is true that heuristics lead people to avoid uncertainty [@Spiegelhalter2017], there is no evidence that this psychological effect translates to issues with the visual representation of uncertainty.
-Again, given that we do not make these same visual considerations for other variables that elicit distaste or irrational behaviour, there is no reason to assume this is what makes uncertainty visualisation so special.
-
-This leads us to the conclusion that the visualisations made for the purpose of displaying information about uncertainty statistics are not uncertainty visualisations. 
-These graphics are just normal information visualisations, and authors can follow existing principles of graphical design. 
-We focus on the perspective that uncertainty visualisation serves to obfuscate signal, and an uncertainty visualisation is a variation on an existing graphic that gives it the ability to suppress false signals.
-
-Of course, there is nothing wrong with explicitly visualising variance, error, bias, or any other statistic used to depict uncertainty as a signal. 
-Just like any other statistic, these metrics provide important and useful information for analysis and decisions. 
-However, there is no interesting visualisation challenge associated with these graphics, and they do not require any special visualisation techniques. 
-The uncertainty in these graphics are acting as a signal variable, and they should be treated as such.
-
-## Visualising uncertainty as a variable
-Upon hearing that uncertainty needs to be included for transparency, the solutions may seem obvious. 
-You may think "well, I will just add a dimension to my plot that includes uncertainty". 
-This is a reasonable approach. 
-The simplest way to add uncertainty to an existing graphic is to simply map uncertainty to an unused visual channel. 
-However, it is unclear if this approach is sufficient for our purposes.
-
-### The bivariate map
-@fig-bivariate depicts a variation of the choropleth map, where we have a two dimensional colour palette. 
-In this graphic, temperature is still mapped to hue, but the variance is included by utilising colour saturation. 
-While these two maps *do* look visually different (which was not the case in the choropleth map) the spatial trend is still clearly visible in both graphics. 
-This means the uncertainty *is technically* being communicated, however the main message of the graphic is still the spatial trend (that may not exist). 
-The graphic did not suppress the invalid signal, so it is not performing signal-suppression as we would like. 
-At this point, it might be reasonable to ask, why? 
-Why is including the uncertainty as a variable insufficient to achieve signal-suppression, and what changes should we make to ensure signal-suppression occurs?
-
-```{r}
-#| echo: false
-#| message: false
-#| warning: false
-#| label: fig-bivariate
-#| fig-cap: "A bivariate map that depict the counties of Iowa where each county is coloured acording to it's average daily temperature and the variance in temperature.
-This map is a choropleth map with a two dimensional colour palette where temperature is represented by colour hue, and variance is represented by colour saturation.
-Even though uncertainty has been added to the graphic the spatial trend is still clearly visible in the high variance case."
-#| fig-subcap: 
-#|   - "Low Variance Data"
-#|   - "High Variance Data"
-#|   - "Bivariate Palette"
-#| layout-ncol: 3
-#| layout-valign: "bottom"
-
-# Bivariate Map
-# Make bivariate palette
-# Function to devalue by a certain amount
-colsupress <- function(basecols, hue=1, sat=1, val=1) {
-    X <- diag(c(hue, sat, val)) %*% rgb2hsv(col2rgb(basecols))
-    hsv(pmin(X[1,], 1), pmin(X[2,], 1), pmin(X[3,], 1))
-}
-
-# recurvisely decrease value
-v_val = 0.5
-bivariatepal <- c(basecols,
-                  colsupress(basecols, sat=v_val),
-                  colsupress(colsupress(basecols, sat=v_val), sat=v_val),
-                  colsupress(colsupress(colsupress(basecols, sat=v_val), sat=v_val), sat=v_val))
-# establish levels of palette
-names(bivariatepal) <- paste(rep(1:8, 4), "-" , rep(1:4, each=8), sep="")
-
-# Bivariate maps
-p2a <- my_map_data |>
-  filter(variance_class=="lowvar") |>
-  ggplot() +
-  geom_sf(aes(fill = biclass, geometry = geometry), colour=NA) + 
-  scale_fill_manual(values = bivariatepal) +
-  theme_void() + 
-  theme(legend.position = "none")
-
-p2b <- p2a %+% filter(my_map_data, variance_class=="highvar")
-
-show_pal2 <- function (colours, borders = NULL, cex_label = 1, ncol = NULL, myxlab, myylab, breaks, breaks2, tsize1=1.2, tsize2=1.2) {
-  # Set dimensions of palette
-  n <- length(colours)
-  ncol <- ncol %||% ceiling(sqrt(length(colours)))
-  nrow <- ceiling(n/ncol)
-  # make matrix with null values (if not full)
-  colours <- c(colours, rep(NA, nrow * ncol - length(colours)))
-  colours <- matrix(colours, ncol = ncol, byrow = TRUE)
-  # set graphical parameters (?)
-  old <- par(pty = "s", mar = c(0, 0, 0, 0))
-  on.exit(par(old))
-  size <- max(dim(colours))
-  plot(c(-1.5, size), c(0, -size), type = "n", xlab = "", ylab = "", 
-      axes = FALSE)
-  rect(col(colours) - 1, -row(colours) + 1, col(colours), -row(colours), 
-       col = colours, border = borders)
-  text(c(0,col(colours)[nrow,]) + c(0.2, 0.1, 0,0,0,0,0,-0.1,-0.2) , -4.5, 
-       breaks, cex = 1, col = "black")
-  text(-0.25, -c(0,row(colours)[,ncol]) + c(-0.2, -0.1, 0, 0.1, 0.2),
-       breaks2, cex = 1, col = "black")
-  text(4, -5.5, myxlab ,cex = tsize1, col = "black")
-  text(x=-1.25,y=-2, myylab, srt=270, cex = tsize2, col = "black")
-}
-
-p2a
-p2b
-show_pal2(colours = bivariatepal, ncol=8, borders=NA, myxlab = "Temperature", myylab = "Variance", breaks = 21:29, breaks2 = 0:4)
-```
-
-### Why this approach may (or may not) work
-The difficulty in incorporating uncertainty into a visualisation is frequently mentioned but seldom explained. 
-For example @Hullman2016 commented that it is straightforward to show a value but it is much more complex to show uncertainty but did not explain why. 
-Many authors seem to believe uncertainty visualisation is a simple high-dimensional visualisation problem, as the difficulty comes from working out how to add uncertainty into already existing graphics [@Griethe2006].
-While this is part of the problem in uncertainty visualisation, it is not the complete picture. 
-@fig-bivariate makes it clear that simply including uncertainty as a variable is insufficient to perform signal-suppression. 
-If we cannot treat uncertainty the same as we would any other variable, how should we treat it? 
-We need to understand what uncertainty actually *is*, in order to understand how to integrate it into a visualisation.
-
-#### It's a variable... it's metadata... it's uncertainty?
-Describing what uncertainty actually is, is surprisingly hard. 
-Most authors simply avoid the problem and describe the characteristics of uncertainty, of which there are plenty. 
-Often, uncertainty is split using an endless stream of ever changing boundaries, such as whether the uncertainty is due to true randomness or a lack of knowledge [@Spiegelhalter2017; @Hullman2016; @utypo], if the uncertainty is in the attribute, spatial elements, or temporal element of the data [@Kinkeldey2014], whether the uncertainty is scientific (e.g.
-error) or human (e.g.
-disagreement among parties) [@Benjamin2018], if the uncertainty is random or systematic [@Sanyal2009], statistical or bounded [@Gschwandtnei2016; @Olston2002], recorded as accuracy or precision [@Griethe2006; @Benjamin2018], which stage of the data analysis pipeline the uncertainty comes from [@utypo], how quantifiable the uncertainty is [@Spiegelhalter2017; @utypo], etc.
-There are enough qualitative descriptors of uncertainty to fill a paper, but, none of this is particularly helpful in understanding how to integrate it into a visualisation. 
-Rather than trying to define uncertainty by looking at the myriad ways in which it *does* appear in an analysis, we may find it easier to look at where it *does not*.
-
-Descriptive statistics describe our sample as it is and summarises large data down into an easy to swallow format. 
-Descriptive statistics are not seen as the primary goal of modern statistics, however, this was not always the case. 
-In 19th century England, *positivism* was the popular philosophical approach to science (positivists included famous statisticians such as Francis Galton and Karl Pearson). 
-Practitioners of the approach believed statistics ended with descriptive statistics as science must be based on actual experience and observations [@Otsuka2023]. 
-In order to make statements about population statistics, future values, or new observations we need to perform inference, which requires the assumption of the "uniformity of nature", that is, we need to assume that unobserved phenomena should be similar to observed phenomena [@Otsuka2023].
-Positivists believed referencing the unobservable was bad science. 
-In other words, these scientists embraced descriptive statistics due to the inherent certainty that came with them.
-Since uncertainty is non-existent in descriptive statistics, it is clear that uncertainty is a by-product of inference.
-
-This history lesson illustrates what uncertainty actually is. 
-At several stages in a statistical analysis, we will violate the uniformity of nature assumption. 
-Each of these violations will impact the statistic we have calculated and push it further from the population parameter we wish to draw inference on. 
-Uncertainty is the amalgamation of these impacts. 
-If we do not violate the uniformity of nature assumption at any point in our analysis, we do not have any uncertainty.
-
-This interpretation of uncertainty indicates that the uncertainty is not a variable of importance in of itself. 
-Uncertainty is metadata about our statistic that is required for valid inference.
-This means uncertainty should not be visualised by itself and we should seek to display signal and uncertainty together as a "single integrated uncertain value" [@Kinkeldey2014]. 
-This aspect of uncertainty visualisation makes it a uniquely difficult problem.
-
-#### Visualising the "single integrated uncertain value"
-Typically, when making visualisations, we want the visual channels to be separable. 
-That is, we don't want the data represented through one visual channel to interfere with the others [@Smart2019]. 
-Mapping uncertainty and signal to separable channels allows them to be read separately, which does not align with the goal of communicating them as a single integrated channel. 
-Visualising uncertainty and signal separately allows the uncertainty information to simply be ignored, which is a pervasive issue in current uncertainty visualisation methods [@uncertchap2022]. 
-We can see this problem in @fig-bivariate, as it sends the message "this data has a spatial trend and the estimates have a large variance" as we read the signal and the uncertainty separately.
-
-This means effective uncertainty visualisation should be leveraging integrability. 
-That is, the visual channels of the uncertainty and the signal would need to be separately manipulable, but read as a single channel by the human brain. 
-While most visual aesthetics *are* separable, there are some variables that have been shown to be integrable, such as colour hue and brightness [@Vanderplas2020].
-When visualising uncertainty using its own visual channel, we can also consider visual semiotics and make sure to map uncertainty to intuitive visual channels, such as mapping more uncertain values to lighter colours [@Maceachren2012].
-Unfortunately relying on integrability may not give us the amount of control we want over our signal-suppression. 
-Without a strong understanding of how these visual channels collapse down into a single channel, relying on integrability could create unintended consequences such as displaying phantom signals or hiding justified signals.
-Additionally, multi-dimensional colour palettes can make the graphics harder to read and hurt the accessibility of the plots [@Vanderplas2015].
-
-There is another benefit to mapping uncertainty to saturation that is not directly related to integrability. 
-As the saturation decreases colours become harder to distinguish. 
-This means high uncertainty values are harder to differentiate than low uncertainty values. 
-We can leverage this implicit feature of colour value by transforming the visual feature space ourselves.
-
-## Combining uncertainty and signal in a transformed space
-Instead of hoping that uncertainty might collapse signal values into a single dimension, we can do some of that work ourselves. 
-As a matter of fact, some uncertainty visualisation authors already have.
-
-### Value Suppressing Uncertainty Palettes
-The Value Suppressing Uncertainty Palette (VSUP) [@Correll2018], was designed with the intention of preventing high uncertainty values from being extracted from a map.
-Since the palette was designed with the extraction of individual values in mind and it has only been tested on simple value extraction tasks [@Correll2018] or search tasks [@Ndlovu2023], it is unclear how effective the method is at suppressing broader insights such as spatial trends.
-
-@fig-vsup is a visualisation of the Iowa temperature data using a VSUP to colour the counties. 
-The low uncertainty case still has a visible spatial trend, while the spatial trend in the high uncertainty map has functionally disappeared. 
-This means the VSUP has successfully suppressed the spatial trend in the data.
-However, the spatial trend may not be the only signal of concern in our graphic.
-Now we must return to the original signal-suppression criteria and ask ourselves if they have all been met. 
-Are all the justified signals reinforced, while all the unjustified signals are suppressed? 
-Is a graphic that performs perfect signal-suppression even possible?
-
-```{r}
-#| echo: false
-#| message: false
-#| warning: false
-#| label: fig-vsup
-#| fig-cap: "A map made with a VSUP.
-The counties of Iowa are coloured acording to its average daily temperature and the variance in temperature.
-Similar to the bivariate map, temperature is mapped to hue while variance is mapped to saturation.
-Unlike the bivariance map, the colour space we are mapping our variables to has been transformed so that high variance estimates are harder to discern from each other.
-This map successfully reduces the visibility of the spatial trend in the high uncertainty case while maintaining the visibility of the spatial trend in the low uncertainty case."
-#| fig-subcap: 
-#|   - "Low Variance Data"
-#|   - "High Variance Data"
-#|   - "VSUP Palette"
-#| layout-ncol: 3
-#| layout-valign: "bottom"
-
-# VSUP
-# Function to combine colours for VSUP
-colourblend <- function(basecols, p_length, nblend) {
-    X <- rgb2hsv(col2rgb(unique(basecols)))
-    v1 <- X[,seq(1,dim(X)[2], 2)]
-    v2 <- X[,seq(2,dim(X)[2], 2)]
-    if("matrix" %in% class(v1)){
-      # hue issue wrap around pt 1
-      v3 <- (v1+v2)
-      v3["h",] <- ifelse(abs(v1["h",]-v2["h",])>0.5, v3["h",]+1, v3["h",])
-      v3 <- v3/2
-      # hue issue wrap around pt 2
-      v3["h",] <- ifelse(v3["h",]>=1 , v3["h",]-1 ,v3["h",])
-      hsv(rep(v3[1,], each=nblend), rep(v3[2,], each=nblend), rep(v3[3,], each=nblend))
-      } else {
-        v3 <- (v1+v2)
-        v3["h"] <- ifelse(abs(v1["h"]-v2["h"])>0.5, v3["h"]+1, v3["h"])
-        v3 <- v3/2
-        v3["h"] <- ifelse(v3["h"]>=1 , v3["h"]-1 ,v3["h"])
-        rep(hsv(h=v3[1], s=v3[2], v=v3[3]), p_length)
-        }
-}
-
-VSUPfunc <- function(basecols, p_length, nblend){
-  colourblend(colsupress(basecols, sat=0.5), p_length, nblend)
-}
-
-# VSUP
-p = length(basecols)
-VSUP <- c(basecols,
-          VSUPfunc(basecols, p, 2),
-          VSUPfunc(VSUPfunc(basecols, p, 2), p, 4),
-          VSUPfunc(VSUPfunc(VSUPfunc(basecols, p, 2), p, 4), p, 8))
-
-names(VSUP) <- paste(rep(1:8, 4), "-" , rep(1:4, each=8), sep="")
-
-# VSUP maps
-p3a <- my_map_data |>
-  filter(variance_class=="lowvar") |>
-  ggplot() +
-  geom_sf(aes(fill = biclass, geometry = geometry), colour=NA) + 
-  scale_fill_manual(values = VSUP) +
-  theme_void() + 
-  theme(legend.position = "none")
-
-p3b <- p3a %+% filter(my_map_data, variance_class=="highvar")
-
-p3a
-p3b
-show_pal2(colours = VSUP, ncol=8, borders=NA, myxlab = "Temperature", myylab = "Variance", breaks = 21:29, breaks2 = 0:4)
-
-```
-
-### What can and cannot be suppressed?
-The methods used by the VSUP bring to light a slight problem with uncertainty visualisation. 
-Specifically, uncertainty and the purpose of visualisation are somewhat at odds with one another. 
-There are two primary motivations behind visualisation, communication and exploratory data analysis (EDA). 
-Communication involves identifying a signal we want to communicate and designing a visualisation that best conveys that, while EDA involves creating a versatile visualisation and using it to extract several signals.
-If we are designing an uncertainty visualisation for communication then we can just suppress the specific signal we are seeking to communicate. 
-In the map example, we would consider @fig-vsup to be a success as the only signal we are concerned with is the spatial trend. 
-However, it is not uncommon for authors to express a desire for uncertainty visualisations that perform signal-suppression in visualisations made for EDA [@Sarma2024; @Griethe2006]. 
-For uncertainty visualisation for EDA to work, we would need to assume that suppressing individual estimates using their variance should naturally extend to broader suppression of plot level insights.
-Unfortunately, it is unclear how reliably this would work.
-
-#### There is no uncertainty in EDA
-Earlier we established that uncertainty is a by-product of inference, which means without inference, there is no uncertainty. 
-Often EDA is used to give us an understanding of our data and identify which signals are worth pursuing. 
-In this sense, EDA is the visual parallel to descriptive statistics, as it is performed without an explicit hypothesis, which means there is no inference, and by extension, there is no uncertainty.
-
-Some authors recognise inference will always occur (in some shape or form) and believe uncertainty *should* be visualised but do not recognise *how* uncertainty would be visualised. 
-@Hullman2021 argued that there is no such thing as a "model-free" visualisation, therefore all visualisations require uncertainty as we are always performing inference. 
-While it is true that we can think of visualisations as containing implicit inferential properties, there are many potential inferences in any single visualisation. 
-This makes it a little difficult to ensure uncertainty is always included.
-For example, if we have a visualisation that shows an average, we would need to identify if the signal suppression should be performed using the sampling variance or the sample variance [@Hofman2020].
-The distribution we use depends on the inferential statistic but until the viewer chooses one and thinks about it (which isn't easily observable), the particular variety of uncertainty which would need to be displayed can't be calculated.
-This means the ideal uncertainty visualisation should not only meet the signal suppression requirements, but should also endeavor to be versatile enough to meet those requirements for all the signals displayed in the graphic.
-
-#### The limitations of explicitly visualising uncertainty and signal
-The lack of versatility of the VSUP is easy to see with a simple example.
-
-Let's say we have a graphic that depicts a set of coefficients from a linear regression and the value of the coefficient is shown using a single colour. 
-We want to know "Which of these coefficients are different from 0?" as well as "Which of these coefficients are different from each other?". 
-To answer this question we do a series of $t$-tests on these estimates. 
-All of the individual $t$-tests of fail to reject the null hypothesis that the coefficients are different from 0. 
-We then make a visualisation that suppresses this signal and ensures that all of the estimates are visually indistinguishable from 0. 
-Next, we conduct two sample $t$-test and find that several of the values need to be visually distinguishable from each other. 
-The VSUP method must pick a single colour for each estimate, and these colours must be *either* visually distinguishable or indistinguishable from each other. 
-We cannot perform signal-suppression on both these signals simultaneously.
-
-This example highlights a fundamental problem with the VSUP that extends to the bivariate map as well. 
-When we blend these colours, we need to decide at what level of *uncertainty* to blend these colours together. 
-Even though the bivariate map does not explicitly combine colour values at certain variance levels, the mapping of variance to colour saturation does this implicitly. 
-That is, at certain saturation values the colours in a bivariate map are imperceptibly different and appear as though they are mapped to the same value.
-At this point, it is irrelevant whether or not the colours are technically different, they are the same colour in the human brain. 
-This is of course complicated by the fact that human colour perception varies at an individual level. 
-Some women are believed to have four different cone cells which allows them to perceive a greater range of colours, while others only have one or two types of cone cells and have colour deficiencies [@simunovic2010]
-For VSUP to function for all individuals, we must calibrate each plot to an individual's ability to perceive colour.
-
-If we only use a single colour to express each signal-suppressed statistic, we will always need to decide which signals we suppress and which we do not. 
-This issue has already been raised in the literature. 
-Which hypothesis are suppressed and which are not largely depends on the method used to combining colours in the palette [@Kay2019]. 
-The VSUP in @fig-vsup used the tree based method that was used by @Correll2018, but there are alternatives that are more appropriate for different hypothesis.
-
-Uncertainty visualisation for EDA would be possible if we designed a plot in such a way that suppressing individual estimates using their variance would naturally extend to broader suppression of plot level insights.
-This assumption is commonly made by visualisation researchers in normal visualisation experiments [@North2006]. 
-If we could express the statistic of a cell using multiple colours, this limitation may disappear entirely.
-
-## Implicitly Combining Uncertainty and Signal
-Rather than trying to figure out how to combine signal and uncertainty into a single colour, we can just display a sample instead and allow the viewer to extract *both* the estimate and the variance.
-
-### Pixel map
-@fig-pixel displays a pixel map [@Lucchesi2021], which is a variation of the choropleth map where each area is divided up into several smaller areas, each coloured using outcomes from the larger area's (i.e.
-the county's) average temperature sampling distribution.
-The spatial trend is clearly visible in the low variance case, but functionally disappears in the low variance case. 
-While the spatial trend is just barely visible in the high uncertainty case, it is much harder to see. 
-This means the graphic also achieves the third criteria for signal-suppression, i.e. our difficulty in seeing the distribution is proportional to the level of uncertainty in the graphic.
-
-```{r}
-#| eval: false
-#| echo: false
-# Make + save pixel map (in case of depreciation)
-# Low variance map
-my_map_data_a <- my_map_data |>
-  filter(variance_class == "lowvar") |>
-  mutate(my_id = seq(n),
-         error = variance)
-
-# quantile
-q_a <- my_map_data_a |> 
-  as.tibble() |>
-  mutate(bitemp=as.numeric(bitemp)) |>
-  with(data.frame(p0.05 = qnorm(0.05, mean=bitemp, sd=sqrt(variance)),
-                  p0.25 = qnorm(0.25, mean=bitemp, sd=sqrt(variance)), 
-                  p0.5 = qnorm(0.5, mean=bitemp, sd=sqrt(variance)), 
-                  p0.75 = qnorm(0.75, mean=bitemp, sd=sqrt(variance)), 
-                  p0.95 = qnorm(0.95, mean=bitemp, sd=sqrt(variance))))
-
-pixel_1a <- my_map_data_a |>
-  as.data.frame() |>
-  select(my_id, bitemp, error) |>
-  read.uv(estimate="bitemp", error="error")
-
-pixel_2a <- my_map_data_a |> as("Spatial")
-
-pix_a <- pixelate(pixel_2a, pixelSize = 70, id = "my_id")
-
-pmap_a <- build_pmap(data = pixel_1a, distribution = "discrete", pixelGeo = pix_a, id = "my_id", border = pixel_2a, q=q_a)
-
-p4a <- view(pmap_a) +
-  geom_path(
-      data = pmap_a$bord,
-      aes_string(x = 'long', y = 'lat', group = 'group'),
-      colour = "white"
-    )  +
-  scale_fill_gradientn(colours = longpal) +
-  scale_colour_gradientn(colours = longpal) +
-  theme(legend.position="none")
-
-# High variance
-my_map_data_b <- my_map_data |>
-  filter(variance_class == "highvar") |>
-  mutate(my_id = seq(n),
-         error = variance)
-
-q_b <- my_map_data_b |> 
-  as.tibble() |>
-  mutate(bitemp=as.numeric(bitemp)) |>
-  with(data.frame(p0.05 = qnorm(0.05, mean=bitemp, sd=sqrt(variance)),
-                  p0.25 = qnorm(0.25, mean=bitemp, sd=sqrt(variance)), 
-                  p0.5 = qnorm(0.5, mean=bitemp, sd=sqrt(variance)), 
-                  p0.75 = qnorm(0.75, mean=bitemp, sd=sqrt(variance)), 
-                  p0.95 = qnorm(0.95, mean=bitemp, sd=sqrt(variance))))
-
-pixel_1b <- my_map_data_b |>
-  as.data.frame() |>
-  select(my_id, bitemp, error) |>
-  read.uv(estimate="bitemp", error="error")
-
-pixel_2b <- my_map_data_b |> 
-  as("Spatial")
-
-pix_b <- pixelate(pixel_2b, pixelSize = 70, id = "my_id")
-
-pmap_b <- build_pmap(data = pixel_1b, distribution = "discrete", pixelGeo = pix_b, id = "my_id", border = pixel_2b, q=q_b)
-
-p4b <- view(pmap_b) +
-  geom_path(
-      data = pmap_b$bord,
-      aes_string(x = 'long', y = 'lat', group = 'group'),
-      colour = "white"
-    )  +
-  scale_fill_gradientn(colours = longpal) +
-  scale_colour_gradientn(colours = longpal) +
-  theme(legend.position="none")
-
-save(p4a, file="data/p4a.rda")
-save(p4b, file="data/p4b.rda")
-```
-
-```{r}
-#| echo: false
-#| message: false
-#| warning: false
-#| label: fig-pixel
-#| fig-cap: "A pixel map of the counties of Iowa.
-In this map, each county is broken up into several small areas and coloured according to a potential daily temperature, given the its average daily temperature and its sampling distribution.
-This results in each county being represented by a sample rather than a single value.
-In this graphic, we can clearly see the spatial trend in the low variance case, while the spatial trend is much harder to identify in the high variance case.
-"
-#| fig-subcap: 
-#|   - "Low Variance Data"
-#|   - "High Variance Data"
-#|   - "Pixel Map Palette"
-#| layout-ncol: 3
-#| layout-valign: "bottom"
-
-# load pixel maps
-load("data/p4a.rda")
-load("data/p4b.rda")
-# Display plots and palettes
-p4a
-p4b
-show_pal(longpal, ncol=13, borders="white", myxlab = "Temperature", breaks = 19:32, textnudge = c(0.2, 0.1, 0,0,0,0,0,0,0,0,0,0,-0.1,-0.2), xlabx= 6.5, xlaby=-2.2)
-```
-
-It is clear that the pixel-map is not only suppressing the false information, but it is doing so by simulating *more* information. 
-The efficacy of this method means that visualisations of simulated samples pop up repeatedly in the literature, with examples including samples that are animated over time [@Hullman2015; @Blenkinsop2000], pixel-maps, and spaghetti time series plots.
-Not only does this method help readers understand the plot level gist, it is also unlikely to damage the viewers ability to extract individual estimates.
-Extracting global statistics, such as the mean or variance, from a sample can be done with relative ease, especially when those values are mapped to colour [@Franconeri2021]. 
-Additionally, @Ndlovu2023 found that participants applied the same methods they used for simple choropleth maps to complicated uncertainty maps even if that take away was invalid. 
-The pixel map can be read the same way as a choropleth map, which allows us to leverage that automatic response. 
-Therefore, the pixel map performs signal suppression, without sacrificing the viewers ability to extract general statistics, unless those statistics *should* be harder to extract due to the uncertainty in the value.
-So, what is it about @fig-pixel that allows it to be such a successful uncertainty visualisation?
-Is this a question we can easily answer?
-
-### Show me the data
-To thoroughly understand why pixel maps seem to successfully perform signal suppression, we would need a more thorough discussion than what can be offered here. 
-However, this does not prohibit us from discussing potential reasons for the graph's effectiveness. 
-There is a second type of uncertainty that the pixel map quietly communicates but we have not touched on yet, and this uncertainty is also conveyed by visualisations of our raw data. 
-That is, these graphics can show assumption violations.
-
-As we discussed in previous sections, we can consider uncertainty to be "the amalgamations of the impacts of violations to the assumption of the uniformity of nature". 
-It's not a definition that rolls off the tongue, but we can work with it.
-Thankfully, this definition of uncertainty aligns nicely with all the concepts that are included in the uncertainty umbrella. 
-Some works [@Hullman2018; @Maceachren2012; @Thomson2005] focus narrowly on specific terms with mathematical definitions, such as probability, confidence intervals, variance, error, or precision. 
-These works are only concerned with quantifying the final impact of uncertainty on our statistics. 
-That is, how large should the bound around our statistic be, such that our "true" statistic can be inferred. 
-Others [@Griethe2006; @Leland2005; @Pang1997; @Pham2009; @Boukhelifa2017] include broader loosely related elements, such as missing values, reliability, model validity, or source integrity. 
-These broader and harder to quantify concepts are concerned about potential sources of uncertainty, that is, they describe violations to the assumption of the uniformity of nature.
-
-What this means is that we have two types of uncertainty, but one is more "processed" than the other. 
-Quantifiable uncertainties are just assumption violations expressed as an effect on our final statistic (or true variation of our data). 
-The disconnect between quantifiable and unquantifiable uncertainty creates a huge problem for authors trying to visualise it.
-A survey of visualisation authors cited "not knowing how to calculate uncertainty" as one of the primary reasons they did not include it in visualisations [@Hullman2020a].
-
-There are two reasons we might leave our uncertainty as an assumption violation rather quantifying the effect. 
-The first reason is that we may be unable to translate the assumption violation to a quantifiable uncertainty. 
-There is no blanket rule that allows us to reliably quantify all uncertainty for every statistic, although some researchers have tangled with the idea. 
-For example, @Thomson2005 suggests a mathematical formula for *examples* of uncertainty, and information theory tries to quantify uncertainty using the idea of entropy, but they ignore the disconnect between the broad concept of uncertainty and what we can reliably quantify.
-Some authors don't believe that it is even possible to quantify all the assumption violations [@Spiegelhalter2017].
-
-The second reason to leave uncertainty as a potential violation of our assumptions, is that we might not know the final statistic we are seeking to calculate. 
-This is the case for visualisations made for EDA, and a large number of developments in EDA visualisation have been in displaying these difficult to quantify violations. 
-For example, @Tierney2023 builds upon the tidy data principles to allow users to handle missing values. 
-This includes data plots with a missing value "shadow" that allows visualisation authors to identify if the variables used in a plot have any structure in their missing values, which would contribute to uncertainty.
-
-With this understanding it becomes clear to see why uncertainty is tied to an endless string of examples in the data analysis pipeline. 
-Uncertainty examples include imputed data, model selection, inherent randomness, biased sampling, etc., not because these things *are* uncertainty, but because they *create* uncertainty when we perform inference.
-Whether or not these elements are relevant is highly dependent on what statistic you are trying to draw inference on, and by extension, the purpose of your visualisation.
-
-This relationship between uncertainty and the purpose of our analysis is littered throughout the literature. 
-Multiple authors have commented on the need to consider quantifying and expressing uncertainty at every stage of a project as the goal shapes every step of the analysis [@Kinkeldey2014; @Hullman2016; @Refsgaard2007].
-@Otsuka2023 suggested that the process of observing data to perform statistics is largely dependent on our goals, because the process of boiling real world entities down into probabilistic objects (or "probabilistic kind" as he puts it) depends on the relationship we seek to identify with our data.
-@Meng2014 commented what is kept as data and what is tossed away is determined by the motivation of an analysis and what was previously noise can be shown to become signal depending on the question we seek to answer.
-@Wallsten1997 argue that the best method for evaluating or combining subjective probabilities depends on the uncertainty the decision maker wants to represent and why it matters. 
-@Fischhoff2014 looks at uncertainty visualisation for decision making decides that we should have different ways of communicating uncertainty based off what the user is supposed to do with it.
-
-This makes it very difficult to move quantified uncertainty through the layers of our analysis, especially when designing a visualisation for EDA. 
-If we don't know what the final statistic is, we cannot quantify the effects of our assumptions. 
-Therefore, it is often the case that the best uncertainty visualisation is not an uncertainty visualisation at all, but simply the most accurate depiction of our raw data.
-This does not mean that visualising raw data will always prevent insignificant signal from getting through. 
-@Chowdhury illustrated how groups that appear linearly separable in a linear discriminant analysis (LDA) visualisation of the data can actually be the result of a LDA performed on too many variables, something that was not clear from the visualisation until the line-up protocol was implemented.
-However, just showing the data is a simple but effective option for uncertainty visualisation that seems to be largely overlooked.
-
-Of course, visualising the raw data is not always possible nor suitable.
-For example, if we have a specific statistic in mind, it may be more appropriate to show the sampling distribution of that statistic rather than the variability of all observations.
-If we visualise that sampling distribution in a way that allows us to see it's shape, and therefore recognise any assumption violations (such as a non-normal distribution), we maintain some of the benefits that came with visualising the raw data.
-This is exactly what the pixel map does, and likely contributes to it's success as a signal suppression method.
-
-## Evaluating uncertainty visualisations
-If we want to make conclusions about how effective any uncertainty visualisation method is we need to look at the results of evaluation experiments. 
-Unfortunately the illustrative methods we have used thus far, i.e. showing a graphic and saying "wow look at this", is lacking if we want any serious results.
-However, despite the abundance of uncertainty visualisation evaluation experiments, existing literature reviews have struggled to synthesise them into any common rules [@Kinkeldey2014; @Hullman2016].
-
-Here we discuss common evaluation methods, why these methods might struggle to create a cohesive set of recommendations for uncertainty visualisations, and consider how to evaluate visualisations on their ability to perform signal-suppression.
-
-### Current methods
-Including uncertainty in a visualisation comes with many secondary benefits.
-Examples of these benefits include better decisions, more trust in the results and the ability to extract additional statistics, such as the variance.
-Ultimately, these secondary benefits are not the primary goal of uncertainty visualisation, and evaluating uncertainty visualisations on these criteria often has unintended consequences.
-
-#### Value extraction of uncertainty statistics
-Uncertainty visualisations are often evaluated based on how accurately [@Hullman2019] viewers can separately extract the estimate and its variance [@Kinkeldey2014]. 
-This means a significant chunk of evaluation studies boil down to showing a participant a visualisation and asking questions such as "what is the variance of $X$?", or "what is the mean of $X$?". 
-This seems like a relatively straight forward approach, and it is similar to how non-uncertainty visualisations are evaluated, but is this appropriate for uncertainty visualisations? 
-The role of uncertainty is rarely evaluated in these studies as the graphics are often compared on the basis of being uncertainty visualisations [@Ibrekk1987; @Hullman2015; @Hofman2020], a class that has no established definition.
-By shifting the focus of our inference from $\bar{X}$ to $Var(X)$ or $P(X)$ we end up evaluating visualisations on their ability to convey uncertainty statistics, rather than on uncertainty's ability to suppress statistics.
-This leads to a series of experiments where the uncertainty is evaluated as a signal even if that was not the goal of the experiment.
-
-The problem with evaluating uncertainty as a signal are identical to the problems associated with displaying uncertainty as a signal. 
-There is no reason to assume uncertainty would behave any differently to any other variable when we evaluate them in this way. 
-For example, @Ibrekk1987 found that participants were more accurate at extracting a statistic when it could be directly read off the graphic, than when it required an area estimate (which is the case if using the probability density function), or when there was no visual indicator for the statistic at all (which is the case when using the cumulative distribution function of a skewed random variable).
-@Hullman2015 found that a visualisation that allows viewers to count outcomes to estimate a probability outperformed one that required a complicated area calculation. 
-@Hofman2020 and @Zhang2022 found that participants were better at answering questions about prediction intervals when shown a prediction interval instead of a sampling distribution. 
-@Gschwandtnei2016 found that graphics where the required statistic could be directly read off the plot outperformed those that involved guesswork due to a gradually decreasing line. 
-@Cheong2016 found that participants made better decisions when they were explicitly given the relevant probability in text rather than when they needed to read it off a map. 
-It is well established that extracting information from a graphic using a perceptual task will always be less accurate than explicitly reading the value provided in text form [@Cleveland1984].
-
-The biggest failing of this evaluation method is not the predictable outcomes, but that it encourages us to see successful examples of signal-suppression as a failing. 
-@Blenkinsop2000 commented that visually integrable depictions of uncertainty should be avoided, as they decrease the viewers confidence in their extracted data values. 
-This conclusion is antithetical to the goals of signal-suppression and occurs because these methods evaluate uncertainty as a signal, not as noise.
-
-#### Trust, confidence, and risk aversion
-Trust is a by-product of displaying uncertainty so it is commonly measured in uncertainty evaluation studies [@Hullman2019]. 
-Considering trust, and not transparency, as the metric of importance in uncertainty communication can lead to a questionable subtext that argues against transparency, something that has been noticed by several other authors [@Spiegelhalter2017; @ONeill2018].
-Science communication should be primarily concerned with accuracy. 
-This accuracy may not always involve showing the data exactly as it is, for example we may want to adjust our graphic to correct for visual heuristics, but these changes should still be in pursuit of accuracy.
-Setting trust as the variable of interest implicitly encourages statisticians to set trust as the primary goal of communication. 
-Evaluating visualisations on trust conflates trust and transparency and ultimately discourages signal-suppression. 
-We can see this effect pop up in the results of evaluation studies. 
-For example, @Zhao2023 found that participants were more trusting of model estimates with low uncertainty, but this effect did not carry over to estimates with high uncertainty. 
-Despite decreased trust being a desired outcome of signal suppression, the author's discussion implied this result was not desired. 
-This perspective ends up extending to visualisation authors as well.
-@Hullman2020a found that authors simultaneously argued that failing to visualise uncertainty was akin to fraud, but also many avoided uncertainty visualisation because they didn't want their work to come across as untrustworthy.
-This is a classic example of the negative impacts of placing direct importance on *trust* rather than *transparency*. 
-In cases of high uncertainty, authors will opt to leave out uncertainty information because it decreases confidence in the authors conclusions. 
-This is the end result of designing uncertainty visualisations for increased trust, rather than signal-suppression.
-Of course, this does not mean measuring trust or perceived trustworthiness has no place in visualisation experiments.
-After all, if participants political motivations can influence their ability to accurately read a plot [@nurse2020], then their level of trust in the information could have a similar effect.
-However, it does mean trust should not be the primary goal of uncertainty visualisation evaluation experiments.
-
-Another metric that is similar to trust is the participants confidence in their decision or extracted value. 
-Confidence has many of the same issues as trust, but it has an additional confounding factor. 
-In non-uncertainty visualisation evaluation experiments, confidence is used as a proxy for the clarity of the visualisation. 
-Confidence cannot simultaneously be a measure of clarity of visualisation *and* a way to capture the uncertainty expressed in a visualisation. 
-Uncertainty visualisations conflate these two measures.
-
-Risk-aversion is another secondary effect of uncertainty visualisation that is used to evaluate uncertainty visualisations [@Hullman2019]. 
-Risk aversion is an economics term used to describe an agent who would chose a random variable with a lower expected payout because it also has a lower variance. 
-Risk aversion is considered irrational behaviour because it is a deviation from the behaviour of a rational agent. 
-Comparing participant responses to that of a rational agent has even been suggested as a benchmark for uncertainty visualisation experiments [@wu2023rational].
-However, just like trust, evaluating graphics on risk-aversion discourages signal-suppression. 
-Rational agents *by definition* should *ignore* uncertainty information. 
-Risk aversion is considered to be irrational because it means the economic agent *is* considering the uncertainty information.
-The only case where a rational agent should not ignore the uncertainty information is when the uncertainty is signal, not noise. 
-Designing graphics that encourage choices that align with that of a rational agent, is to encourage graphics that do not include uncertainty at all.
-
-For these reasons we do not believe trust, confidence, or risk-aversion are useful measures to evaluate uncertainty visualisations. 
-While they are designed to capture the secondary effects of uncertainty, using them as primary measures of visualisation performance is in direct conflict with designing visualisations for signal-suppression.
-
-#### Questions that attempt to capture signal-suppression
-There is a collection of studies that seem to be aware of the issues behind using trust or value extraction to evaluate uncertainty visualisations. 
-These studies try to measure the effect of some kind of single integrated value, but the methodology is often ad-hoc with varying levels of success.
-
-The first method is what we call the "vague question" approach. 
-The authors of these studies will ask the participants a question that implies that they should use the uncertainty information, but it is unclear how. 
-The authors then go on to compare the readers answer to a very specific ground truth. 
-This means the participants are being *evaluated* as though they are performing a value extraction task, but they are not being asked the *questions* that are asked in the value extraction task. 
-Ultimately this approach results in strangely cryptic questions that create a large amount of noise due to the varied interpretation of the questions [@Hullman2016].
-
-For example @Hofmann2012 showed study participants 20 plots, where each plot displayed two distributions (as a pair of jittered samples, density plots, histograms, or box plots) and asked them to identify the plot where "the blue group is furthest to the right".
-The mean of the two groups was used to decide which group was actually "furthest to the right" and the different graphical approaches were evaluated against that ground truth. 
-In another example @Ibrekk1987 asked participants for the "best estimate", but the participant's responses were evaluated against the mean of the distribution.
-Ultimately, the term "best" is up to the users interpretation, and the estimate that minimised the sum of squared errors was not implied by the question. 
-This vague question approach leads to inconclusive results, as we are left unclear if it was the phrasing of the question or the plot design that caused the participants to answer incorrectly.
-
-A variation of the vague question problem, is a series of studies that ask questions that are impossible to answer using the information given to the participants. 
-These studies frequently expect participants to give deterministic answers for probabilistic questions. 
-For example @Correll2014 showed participants the distribution of voter preferences for two candidates in an election, and asked them "how likely is candidate B to win the election?". 
-Participants were not able to answer the question about likelihood in term of probability, but were instead given seven options from 1 = "Outcome will be most in favour of A" to 7 = "Outcome will be most in favour of B".
-The ground truth statistic for this question was a scalar multiple of Cohen’s d, indicating participants were supposed to incorporate uncertainty information using a very specific formula that was likely unknown to them but assumed to be used implicitly.
-In another example, @Padilla2017 provided participants with a visualisation of the cone of uncertainty and asked then to "decide which oil rig will receive more damage based on the depicted forecast of the hurricane path".
-The cone of uncertainty provides a 60% confidence interval for the location of the eye of a hurricane, which allows us to know the area where the eye of the storm will go, but it does not given any information about the intensity of a storm, the size of a storm, or even if a location will be hit.
-Other authors have commented on the complexity of communicating hurricane risk because the path, storm surge and wind speed are all important and cannot be ignored [@Spiegelhalter2017]. 
-@Padilla2017 seem to be aware of the issues in the visualisation and are using the user study to show that current government communications are not the best choice when communicating hurricane risk, as it is unreasonable to assume that "more likely to be hit" should automatically translate to "more damage".
-While this is useful to highlight, the authors end up caught by their own trap, as answering the questions in their second study correctly required participants to assume an area being more likely to be hit *does* translate to receiving more damage.
-The assumptions required to answer these deterministic questions are often invisible, even to the authors of the paper, which means they often create more problems than they solve when they are used as a tool for graphical evaluation.
-Similar to the vague question problem, these studies seem to be aware that we should be evaluating an uncertainty visualisation based on a deterministic observation (such as our example of "does this map have a spatial trend") but are unsure how to incorporate or evaluate it.
-
-While these approaches are certainly a step in the right direction, the experiments end up having far too much noise in their results. 
-This is further complicated by the fact that visual statistics and their computational equivalents often differ, even before we consider these measurement errors.
-By having a ground truth, these experiments implicitly ask users to treat uncertainty as a signal which goes against the purposes of signal-suppression.
-
-### Testing signal suppression
-If we cannot ask direct questions about uncertainty, or measure the secondary effects of uncertainty, or ask indirect questions about the uncertainty; then how are we supposed to evaluate uncertainty visualisations?
-How do you measure something that disappears the second you look directly at it?
-To evaluate uncertainty visualisations, we need an experimental design that evaluates uncertainty as noise, not as signal. 
-This means we need to measure *uncertainty's impact on the signal*, not the uncertainty itself.
-
-#### Comparing to hypothesis tests
-The most obvious way to evaluate uncertainty visualisations is to compare the visualisations to statistical tests. 
-If a graphic was performing signal-suppression we would expect the signal to be harder to see at higher levels of uncertainty. 
-The ideal outcome is a an uncertainty visualisation where the signal is only perceivable when it would be identified by a hypothesis test.
-Evaluating visualisations as though they are akin to hypothesis tests is a well established concept in visualisation.
-The lineup protocol is a good example of this approach.
-Lineups are a confirmatory visualisation tool that can be used to check if perceived patterns are real or merely the result of chance [@Buja2009; @Wickham2010]. 
-Lineups can also be used to evaluate the effectiveness of different types of plots [@Hofmann2012] and design decisions [@vanderplas2017].
-The motivation behind the lineup protocol is similar to the motivations behind signal-suppression, although the lineup protocol is more explicitly tied to a specific hypothesis. 
-@Patrick2023 compared standard statistical tests to the lineup-protocol, and evaluated the visualisations using the power curves that are typical for hypothesis testing. 
-In a similar vein, @Kim2019 investigated how different uncertainty visualisation methods influenced user's prior beliefs, and evaluated the graphics by comparing their results to those from Bayesian inference.
-This approach is similar to the lineup protocol as it also evaluates a visualisation by comparing it to an analogous statistical calculation, however it does so using a different statistical philosophy.
-The connection with visualisation and statistical tests even extends to EDA, where @Hullman2021 commented that a visualisation can be considered a model check where the null hypothesis is the expected and we are looking for the unexpected.
-This is all to say that comparing visualisations to hypothesis tests is not uncharted territory.
-
-We could use a similar evaluation method to evaluate uncertainty visualisations, however, it is not entirely clear what the computational equivalent of an uncertainty visualisation is.
-Even the lineup protocol, which is much closer to a standard statistical test, runs into some difficulty when trying to directly compare it to its computational equivalent.
-@Patrick2023 found that human viewers using the lineup protocol are less sensitive to deviations from the null hypothesis than standard statistical tests.
-However, this finding does not directly translate to the lineup protocol being worse than its computational equivalent.
-The lineup protocol allows viewers to detect if "something" is not right even if that something is unspecified, or what the authors intended.
-For example, @vanderplas2017 did an experiment where participants were drawn to plots with unequal cluster sizes, even though the size of the cluster was not one of the features the authors were testing.
-While standard hypothesis tests may outperform the lineup protocol on power, computational tests have the benefit of knowing exactly what they are looking for. 
-Visualisations are often explicitly made when we *don't* know what we are looking for.
-Additionally, lineup protocols often have a specific hypothesis that is used to create the null distribution, while uncertainty visualisations are intended to be more flexible and can be created without identifying a specific null distribution.
-Therefore, while we could evaluate uncertainty visualisations by comparing them to a selection of hypothesis tests, we should not expect the same level of sensitivity that we get in standard hypothesis tests as the assumed knowledge about our data is quite different.
-This approach could give us a good idea of the use cases of different uncertainty visualisation designs, but we should be mindful of it's limitations in fully describing the effectiveness of these graphics.
-
-#### Qualitative Studies
-Alternatively visualisation research could shift away from the accuracy concept all together and ask questions that allow for open ended responses.
-This method can enlighten authors as to *how* the uncertainty information was used by the participants.
-@Hofmann2012 captured this by asking participants why they considered a particular plot to be more right-shifted, even though this qualitative analysis did not make it into the final paper. 
-@Daradkeh2015 presented participants with ten investment alternatives and asked participants "from among available alternatives, which alternative do you prefer the most", and were asked to think aloud and consider the uncertainty in their decision making.
-The experimenters goal was to observe and organise the methods people use when making decisions in the face of uncertainty. 
-They highlighted the specific aspects of uncertainty that participants typically considered, such as the range of outcomes that are above/below a certain threshold, minimum and maximum values, the risk of a loss, etc., and identified where in the decision making process participants made these considerations.
-
-#### Heuristics
-While experiments that explicitly identify heuristics in current methods are not technically measuring signal-suppression, they are still a useful consideration to keep in mind when designing experiments for signal suppression.
-Heuristic checks look at unknown pitfalls that might exist in interpretation of current plots [@Hullman2016] and can change depending on the larger scope of the graphic and the population we are communicating with [@Spiegelhalter2017; @Kinkeldey2014].
-
-Several heuristics are of particular importance for uncertainty visualisations, and are likely to impact how well different methods performing signal suppression. 
-The sine illusion can cause the confidence interval of a smoothed sine curve to seem wider at the peaks than the troughs, causing us to underestimate uncertainty associated with changing values [@Vanderplas2015].
-Points that were on an outcome of an ensemble display were perceived as more likely than points not on an outcome, even when the point that was not on an outcome was closer to the centre of the distribution (and therefore more likely) [@Padilla2017].
-This can be considered an extension of the within bar bias, where participants looking at a bar chat with error bars view outcomes within the bar as more likely than those outside it [@Newman2012]. 
-Several studies have found that viewers use a heuristic where they compare the distance between two estimates to estimate if they are different, and in doing so, ignore the uncertainty information [@uncertchap2022].
-
-These heuristics have the potential to create noise in signal suppression evaluations, and should be kept in mind when designing an evaluation experiment.
-
-## Conclusions and Future work
-This paper has identified gaps in the uncertainty visualisation literature that must be filled for the field to progress. 
-We formalised the uncertainty visualisation problem, and in doing so highlighted an untouched area of uncertainty visualisation research.
-
-*Experimental practices on uncertainty visualisation need to be standardised.* 
-Uncertainty visualisations for decision making treat uncertainty as signal, while visualisations for signal-suppression treat uncertainty as noise. 
-As the literature currently exists, there is no way to combine papers to get a meaningful sense of how uncertainty information is understood by a viewer.
-Researchers need to ensure that when they identify the motivation behind their visualisation technique, and that their evaluation methods match the motivations of the paper. 
-Additionally *evaluation methods that evaluate uncertainty as noise need to be developed.*
-
-*Experimenters should consider evaluating visual aesthetics on how well they suppress information in a graphic.* 
-Research into separability and integrability is of particular interest to uncertainty visualisation, as it allows interference from the uncertainty variable. 
-When designing experiments, authors often choose aesthetics that are visually distinguishable; uncertainty visualisation authors should consider doing the opposite.
-
-*Software that allows users to easily perform signal suppression should be developed.* 
-Existing uncertainty visualisation methods view a distribution as its own object and there are no software options for the "an uncertainty visualisation is a function of an existing visualisation" philosophy.
-
-Signal suppression is an untouched area of visualisation research and developing methods for the practice may require us to challenge our entire notion of what makes a good visualisation.
-We are interested to see how this challenge plays out.
-
-## Acknowledgements
-
-The first author of this paper is supported in part by a scholarship from the the Australian Energy Market Operator. 
-The R packages were used for this work were: `tidyverse` [@tidyverse], `Vizumap` [@Vizumap], `RColorBrewer` [@RColorBrewer], `scales` [@scales], `sf` [@sf], `ggrepel` [@ggrepel], `urbnmapr` [@urbnmapr], `flextable` [@flextable], `colorspace` [@colorspace], and `rgeos` [@rgeos].
-The GitHub repository for this paper can be found at https://github.com/harriet-mason/ARSA-UncertaintyLitReview which contains the files required to reproduce this article in full.
-
-## Bibliography
-
-```{r, include=FALSE, eval=FALSE}
-library(spelling)
-qmd <- "paper.qmd"
-ignore <- readLines("WORDLIST")
-check_spelling <- spell_check_files(
-  qmd,
-  ignore = ignore,
-  lang = "en_GB"
-)
-if (nrow(check_spelling) > 0) {
-  print(check_spelling)
-  stop("Check spelling in Qmd files!")
-}
-```

### 2025-12-01 twopagesummary/summary.qmd
changed format and added outline to summary

@@
 ---
-title: "Paper Summary: Uncertainty Visualisation"
+title: "Paper Outline"
 author: 
@@
   @*
-abstract: Uncertainty visualisation reviews often take the definition and purpose of an uncertainty visualisation to be self evident which results in a large amout of conflicting information.
+toc: false
+number-sections: true
+latex-clean: true
+format: pdf
+fig-valign: bottom
+cap-location: bottom
+editor_options: 
+  chunk_output_type: console
+---
+
+# Proposed title: The Noisy Work of Uncertainty Visualisation Research
+
+# Overview:
+
+XXX needs re-writing: The literature on uncertainty visualisation treats the definition and purpose to be self evident which results in a large amount of conflicting information.
 This conflict largely stems from a conflation between uncertainty visualisations designed for to visualise uncertainty as signal and those used to visualise uncertainty as noise.
@@
 We suggest authors use qualitative studies or compare uncertainty visualisations to the relevant hypothesis tests.
-date: last-modified
-toc: false
-number-sections: true
-latex-clean: true
-format:
-  jasa-pdf:
-    keep-tex: true  
-    journal:
-      blinded: false
-  jasa-html: default
-fig-valign: bottom
-cap-location: bottom
-editor_options: 
-  chunk_output_type: console
----
+
+# Topics planned
+
+1. Common uses of the terms "uncertainty", and "uncertainty visualisation"
+2. Dangers of ignoring uncertainty representation in a visualisation
+3. Review of current approaches to representing uncertainty in plots, 
+4. Review of literature describing experiments testing effectiveness of different representations
+5. Examples: spatial and ??

### 2025-12-01 twopagesummary/summary.qmd
revised summary

--- 2025-11-30 twopagesummary/summary.qmd
@@
 ---
-title: "Paper Summary: Uncertainty Visualisation"
+title: "Proposed title: The Noisy Work of Uncertainty Visualisation Research"
 author: 
   - name: Harriet Mason
-    url: https://harrietmason.netlify.app/
-    orcid: 0009-0007-4568-8215
-    email:  harriet.m.mason@gmail.com
-    affiliation:
-    - name: Monash University
-      department: Department of Econometrics and Business Statistics
-      city: Melbourne
-      country: Australia
   - name: Dianne Cook
-    url: https://dicook.org
-    orcid: 0000-0002-3813-7155
-    email: dicook@monash.edu
-    affiliation:
-    - name: Monash University
-      department: Department of Econometrics and Business Statistics
-      city: Melbourne
-      country: Australia
   - name: Sarah Goodwin
-    url: https://www.linkedin.com/in/smgoodwin/
-    orcid: 0000-0001-8894-8282
-    email: Sarah.Goodwin@monash.edu
-    affiliation:
-    - name: Monash University
-      department: Department of Human Centred Computing
-      city: Melbourne
-      country: Australia
   - name: Emi Tanaka
-    url: https://emitanaka.org/
-    orcid: 0000-0002-1455-259X
-    email: Emi.Tanaka@anu.edu.au
-    affiliation: 
-    - name: The Australian National University
-      department: Biological Data Science Institute
-      city: Canberra
-      country: Australia
   - name: Susan VanderPlas
-    url: https://srvanderplas.github.io
-    orcid: 0000-0002-3803-0972
-    email: susan.vanderplas@unl.edu
-    affiliation:
-    - name: University of Nebraska–Lincoln
-      department: Statistics Department
-      city:  Lincoln
-      country: United States
 bibliography: summary.bib
@@
   @*
-abstract: Uncertainty visualisation reviews often take the definition and purpose of an uncertainty visualisation to be self evident which results in a large amout of conflicting information.
+toc: false
+number-sections: true
+latex-clean: true
+format: pdf
+fig-valign: bottom
+cap-location: bottom
+editor_options: 
+  chunk_output_type: console
+---
+
+# Rationale
+
+Better ways of including a representation of the uncertainty in a data visualisation is a focus of recent research activity.
+A problem with the current literature is that there is a lack of clarity in the definition of uncertainty, and what it means to represent it in a plot.
+This confusion can also be seen in literature describing experiments to test effectiveness of choices of different uncertainty representation.
+The purpose of this review article is to summarise the current literature, provide workable definitions, and illustrate with examples.
+It is hope that it will be useful to guide new graphics methodology and  experimental work.
+
+<!-- XXX needs re-writing: The literature on uncertainty visualisation treats the definition and purpose to be self evident which results in a large amount of conflicting information.
 This conflict largely stems from a conflation between uncertainty visualisations designed for to visualise uncertainty as signal and those used to visualise uncertainty as noise.
@@
 We suggest authors use qualitative studies or compare uncertainty visualisations to the relevant hypothesis tests.
-date: last-modified
-toc: false
-number-sections: true
-latex-clean: true
-format:
-  jasa-pdf:
-    keep-tex: true  
-    journal:
-      blinded: false
-  jasa-html: default
-fig-valign: bottom
-cap-location: bottom
-editor_options: 
-  chunk_output_type: console
----
+-->
+
+# Outline
+
+1. Common uses of the terms "uncertainty visualisation": Here we will provide some background to the use of the term "uncertainty visualisation" and tie it to statistical definitions of uncertainty.
+2. Dangers of ignoring uncertainty representation in a visualisation. This part motivates the reason for the importance of this area of research. 
+3. Review of current approaches to representing uncertainty in plots, as gathered from discipline journals in information visualisation, statistics, environmental science and psychology. 
+4.
+Review of literature describing experiments testing effectiveness of different representations.
+The information visualisation literature is especially prolific in describing experiments assessing the relative merits of different uncertainty visualisations, and this section will critically summarise these efforts, and propose refining of procedures.
+5. Examples:
+
+- spatial data poses substantial challenges because of the difficulties of using the page for the map coordinates, and hence little room for representing uncertatinty.
+It also has been a place with useful recent work, providing some good solutions.
+- comparing groups (boxplots, letter value plots, dotplots) OR
+- confidence intervals, vs rainclouds, vs dotplots, vs transparency
+
+6. Recommendations for research directions. Here we will gather the discussion into some suggestions for framing new work, developing methodology and designing experiments.
+
+# References

### 2025-12-01 twopagesummary/summary.qmd
outline refined

@@
 
-1. Common uses of the terms "uncertainty visualisation": Here we will provide some background to the use of the term "uncertainty visualisation" and tie it to statistical definitions of uncertainty.
+1. Common uses of the term "uncertainty visualisation": Here we will provide some background to the use of the term "uncertainty visualisation" and tie it to statistical definitions of uncertainty.
 2. Dangers of ignoring uncertainty representation in a visualisation. This part motivates the reason for the importance of this area of research. 
@@
 
-- spatial data poses substantial challenges because of the difficulties of using the page for the map coordinates, and hence little room for representing uncertatinty.
+- Comparing distributions with the purpose to illustrate definitions of uncertainty, and distinction between representations of signal vs noise.
+It is also a place where software exists providing some good choices but are not yet sufficient.
+- Spatial data poses substantial challenges because of the difficulties of using the page for the map coordinates, and hence little room for representing uncertainty.
 It also has been a place with useful recent work, providing some good solutions.
-- comparing groups (boxplots, letter value plots, dotplots) OR
-- confidence intervals, vs rainclouds, vs dotplots, vs transparency
 

### 2025-12-04 twopagesummary/summary.qmd
maybe i made some more changes and didn't push

@@
 
-Better ways of including a representation of the uncertainty in a data visualisation is a focus of recent research activity.
-A problem with the current literature is that there is a lack of clarity in the definition of uncertainty, and what it means to represent it in a plot.
-This confusion can also be seen in literature describing experiments to test effectiveness of choices of different uncertainty representation.
-The purpose of this review article is to summarise the current literature, provide workable definitions, and illustrate with examples.
-It is hope that it will be useful to guide new graphics methodology and  experimental work.
+Better ways of including representation of the uncertainty in a data visualisation is a focus of recent research activity.
+A problem with the current literature is that there is a lack of clarity about the definition of uncertainty, and what it means to represent it in a plot.
+This confusion can also be seen in the literature describing experiments on the effectiveness of different uncertainty representation.
+The purpose of this review article is to summarise the current literature, provide workable definitions, and illustrate these with examples.
+It is hoped that it will be useful for guiding new graphics methodology and experimental research.
 
